### 2024-08-12 misc/historical-background-gravity-model.Rmd
draft introduction

---
title: "Gravity model-historical background"
output: html_notebook
---

## Historical Background

Human interaction, be it migration, urban daily travel, marriage, information flows, and so on, has fascinated scholars for ages, and has been of interest for planning purposes to decision makers for almost as long. The conceptual foundations of the study of human interaction date back to at least the 19th century, when HC Carey wrote (as quoted in McKean, 1870):

>> Man, the molecule of society, is the subject of Social Science...The great law of _Molecular Gravitation_ (is) the indispensable condition of the existence of the being known as man...The greater the number collected in a given space, the greater is the attractive force that is there exerted...Gravitation is here, as everywhere, in the direct ratio of the mass and the _inverse_ one of distance

The fundamental idea is an analogy to Newton's Law of Universal Gravitation, originally published in Philosophiæ Naturalis Principia Mathematica in 1687. Carrothers insight that similar principles as in physics could be at work in human interaction did not receive widespread attention, and it was only partially used by Ravenstein (1889) for the study of migration.

It was not until the 20th century that these ideas would be rediscovered. Here, again, there are parallel attributions, which vary by discipline. In this way, EC Young investigated in the 1920s the movement of farm people under the hypothesis that the relative volume of migration is proportional to the attractiveness of a destination and inversely proportional to the square of the distance to the destination. It is notable that Young introduced an empirical proportionallity factor in this formulation of interaction.

In the decade of the 1930s, economist WJ Reilly proposed a law of retail gravitation that explained the point of indifference for two retail sites in terms of the attractiveness of the sites (i.e., bigger sites are more attractive than smaller sites), and the distance to each, with consumers generally having a preference to travel shorter distances, other things being equal. Accordingly:

$$
\frac{P_1}{d_1^2} = \frac{P_2}{d_2^2}
$$
where $P_i$ is the attractiveness (i.e., size) of retail site $i$ and $d_i$ the distance from an arbitrary point to the retail site.

It is easy to see that if we disregard size, the relationship is:
$$
\frac{d_1}{d_2} = 1
$$

This partial representation of Reilly's hypothesis suggests a connection to Voronoi polygons, since the point of indifference will be on the boundary of the polygon, equidistant from the two sites. The full formulation, by extension, can be seen as an alternative way of expressing weighted Voronoi polygons.

Closer approximations to the expression of Carey where obtain shortly after these works.

Astrophysicist JQ Stewart in a series of papers published in the decades of 1940 and 1950 posited a series of principles for so-called social physics, from a rule of inverse distance variation for social influence (1941) to a principle of demographic gravitation (1948). It is this latter work that most explicitly argues for a model of human interaction based on the physical law and put the principle in the notation that we currently use.

Around the same time, sociologist GK Zipf published a series of papers testing regularities in the size of settlements, including what he termed the $\frac{P_1P_2}{D}$ hypothesis. The hypothesis was that movement between two communities with populations $P_1$ and $P_2$ separated by a shortest path of length $D$ is proportional to the ratio noted above. Zipf empirically showed that this regularity held for the case of intercity personal movement (1946), for goods movement by railways (1946), and information (1946).

In Stewart's formulation, the gravity model for human interaction became:
$$
F = \frac{(\mu_1 N_1)(\mu_2N_2)}{d_{12}^2}
$$
where $F$ is the force of attraction between two population, $N_i$ is the size of the population of settlement $i$, $\mu_i$ is the _molecular weight_ of individuals in settlement $i$, and $d_{ij}$ is the distance between settlements $i$ and $j$. In the writting of Stewart (1948, p. 34), the molecular weight "...of the "average American" is taken as unity. Presumably the molecular weight of an Australian aborigine, for example, is on this scale much less than one", which sounds now shockingly racist, but that can reflect the interest/ability of individuals to interact with others. For instance, the molecular weight could be purchasing parity, or reflect individual preferences for interaction.

Many elements of the social physics movement were later rejected by the social science, including the comparison to molecules and the value-ladden assumptions about the molecular weight of individuals. Nonetheless, the gravity model continues to be widely used today in many applications, and the reason for that is that even if we reject the existence of "Laws" for human behavior, as a heuristic it still has remarkable power to capture regularities in behavior.

The work of Alan Wilson (1969) was invaluable in providing a theoretical footing for spatial interaction models that did not depend on physical analogies, but instead was based on information theory. Wilson demonstrated how spatial interaction models could be derived based on the principle of entropy maximization, and how the gravity model is a particular case of entropy maximization.

In this seminar we will introduce the gravity model and we will introduce different modelling methodologies for its application.

### 2024-08-12 paper/manuscript.qmd
draft introduction

---
title: "A family of accessibility measures"
format: pdf
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
library(geomtextpath) # Curved Text in 'ggplot2'
library(ggforce) # Accelerating 'ggplot2'
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
library(here)

# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(patchwork) # The Composer of Plots
# library(RColorBrewer) # ColorBrewer Palettes
# library(sf) # Simple Features for R
# library(scales) # Scale Functions for Visualization
# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# library(tmap) # Thematic Maps
```

# Introduction

<!--The inquiry into spatial interaction traces to Newts Universal Gravitation-->
The inquiry into people's interaction in space has been a popular subject of investigation for over a century. From as far back as @careyPrinciplesSocialScience1858's sentiments on "man [being] the molecule of society [and their interaction being subject to] the direct ratio of the mass and the inverse one of distance" [pg. 41,43], the investigation has often been coloured by the relationship between a pair of masses $M_i$ and $M_j$ (quantity of people, opportunities, or other _scales_ of mass), their separation distances $D_{ij}$, and some proportionality constant $G$ to yield the force of attraction $F$, the features of Newton's Law of Universal Gravitation (1687) @eq-1:

$$
F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
$${#eq-1}

<!--Using Newts Universal Gravitation: Ravenstein 1889 to Stewart 1948-->
Over a century ago, @ravensteinLawsMigration1889's proposed "Laws to Migration" from the empirical analysis of migration flows in early British censuses. This empirical investigation ignited many others looking to formulate models of human interaction based on the physical gravitation law (@eq-1). Notably, @reilly1929methods formulated the law of retail gravitation  <!--I can find the original text of this online, only discussed by others as important-->, Zipf's $\frac{P_1P_2}{D}$ hypothesis holding in the case of intercity personal movement @zipfHypothesisIntercityMovement1946, goods movement by railways @zipfHypothesisCaseRailway1946, and information @zipfDeterminantsCirculationInformation1946, and @stewartDemographicGravitationEvidence1948 introduced principles of demographic gravitation.

<!--Hansen leans on Stewart's 1948 Population Potential measure. Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient. Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent of then 1 for d_ij as it fits empirically better: the issue of units are never addressed. -->
Following this literature, accessibility as the  _"potential of opportunities for interaction"_ defined in @hansen1959 emerged. Today, @hansen1959 is frequently cited as the father of the term's scientific use [@reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure. @hansen1959 borrowed the "population potential" formulation used in @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to measure potential of opportunities (for interaction) (@eq-2). ^[The integral of $F_{ij}$ (@eq-1) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$. The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields 'potential population'. @hansen1959 takes this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.] $S_ij$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size. ^[Notably, @Hansen1959 finds that when the model is empirically calibrated, ${\beta}$ takes a value of between 0.5 to 3, not the value of 1 as consistent in the Newtonian potential energy relationship.]

$$
S_{ij} = O_j * d_{ij}^\beta \\
$${#eq-2}

Of important note, however is that @hansen1959's formulation omits a proportionality constant akin to the function of $G$ in @eq-1. Interestingly, as @hansen1959 draws from @stewartDemographicGravitationEvidence1948, @stewartDemographicGravitationEvidence1948 sets the proportionality constant to 1, stating _"the constant G is left for future determination: a suitable choice of other units can reduce it to unity"_. Without a proportionality constant, however, the units of $S_ij$ are unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units. @eq-3 is the form of gravity-based accessibility measure commonly presented today. It is re-written with $S_i$ being the relative measure of the accessibility at $i$ to all activities $j$, and instead of $d_{ij}^\beta$, $f(c_{ij})$ is any monotonically decreasing function of travel cost $c_{ij}$.

$$
S_{i} = \sum_j O_j * f(c_{ij})
$${#eq-3}

<!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to the authors knowledge, have not been concerned with explicitly determining a proportionality constant. The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure ()) and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations (). However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction, plagues it to this day and contributes to resistance to its application in practice(). The quantity and variations on @eq-3 compounds this issue of interpretability across studies. A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (gravity), utility-, location-, and person- based measures with different components, and @wuUnifyingAccess2020 demonstrates that the concept can be re-unified into its main features: the relationship between travel-cost and corresponding reachable opportunities and demonstrates how previous classifications are special cases of travel-cost or opportunities constraints. From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing - the proportionality constant $G$. The aim of this paper's focus is on the proportionality constant to re-unify accessibility research with its Newtonian gravitational roots.

<!--To address our paper's aim: we referr to the gravity model of today (Wilson) as it is based in the same Newtonian gravitational analogy as accessibility as well as Paez et al. 2019 and Soukhov et al. 2023,2024 -->
The aim of this work is to demonstrate the importance of the proportionality constant consideration within @eq-3 to balancing the units of accessibility into an interpretable value thereby clarifying its meaning. To address this answer this question, we need not reinvent the wheel, instead we refer to the general gravity model of today (@eq-4) as introduced in @wilson1971. The gravity model traces back to these early 20th century empirical investigations [@carrothersHistoricalReviewGravity1956; @griggRavensteinLawsMigration1977; @haynesGravitySpatialInteraction1985; @oshanSpatialStructureDebate2021]; it also retains the proportionality constant and the resulting output is a tangible unit.

$$
T_{ij} = k P_i^{\lambda} P_j^{\alpha} d_{ij}^{\beta}
$${#eq-4}

<!--Explain the features of the gravity model of today (Wilson) -->
Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between $n$ origins $i$ and $m$ destinations $j$, $P$ are a vector of mass attributes at origin $i$ and $j$ and cost to overcome physical separation, and $d_{ij}$ is a $n$ x $m$ matrix. $d_{ij}$ is also frequently replaced by $f(c_{ij})$ in the literature, some function of $c_{ij}$ which decreases as $c_{ij}$ increases. Furthermore, the exponents are introduced to modulate the effect size of the variables, where $\lambda$ and $\alpha$ modulate $P$ and $\beta$ modulates $d$; when variables $T$, $P$ and $d$ are known, the exponents can be estimated to summarize the variable's effect size contribution to the known flows in the system. Finally and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon. In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.

<!-- reiterate the paper's aim-->
This work aims to re-unify accessibility literature with its Newtonian roots by mapping it onto the family of gravity models. Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history. We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today. In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained. Then, we will provide empirical examples of the types of measures. Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.

# A family of accessibility measures

Derive the various accessibility measures. In my mind:
•	Non-constrained: it does not include a single proportionality constant.
•	Unconstrained/Total constrained: uses a single proportionality constant
•	Singly constrained: spatial availability
•	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)
•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)

# Examples

Use the TTS2016R package to illustrate every measure.

# Discussion

When to use each of these measures?

# Conclusions

### 2024-08-14 paper/manuscript.qmd
Additions regarding "The Law of Retail Gravitation"

@@
 Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient.
-Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent of then 1 for d_ij as it fits empirically better: the issue of units are never addressed.
+Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent other than 1 for d_ij as it fits empirically better: the issue of units are never addressed.
 -->
@@
 Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to the authors knowledge, have not been concerned with explicitly determining a proportionality constant.
-The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure ()) and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations ().
+The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
 However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction, plagues it to this day and contributes to resistance to its application in practice().
@@
 
+It's all about balancing factors! Lessons from the balancing factors of the gravity model (cool paper that explores this @kirbyNormalizingFactorsGravity1970)
+
 Derive the various accessibility measures. In my mind:
 •	Non-constrained: it does not include a single proportionality constant.
+
 •	Unconstrained/Total constrained: uses a single proportionality constant
+
 •	Singly constrained: spatial availability
+
+Origins in retail models, decision model for shops to optimally spatially locate to max.
+their potential consumer interaction.
+Starting with reilly's law of gravitational retail (1931) (can't find online), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
+
+Independently formed in people spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998;for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024
+
 •	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)
+
 •	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)
+ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
 

### 2024-09-09 TRB/workshop-abstract-and-letter.qmd
trb worshop proposal abstract and nomination letter draft

---
title: "A family of accessibility measures: a proposal for the TRB 2025 Doctoral Research in Transportation Modeling and Travel Behavior Analysis Workshop"
format: pdf
bibliography: bibliography.bib
---

Applicant: Anastasia Soukhov, PhD Candidate (Transport Geography) in the School of Earth, Environment and Society at McMaster University

Doctoral supervisor: Antonio Paez, full professor in the School of Earth, Environment and Society at McMaster University

# Research Abstract (500 words)

The inquiry into people's interaction in space has been a popular subject of investigation for over a century. From as far back as @careyPrinciplesSocialScience1858's sentiments on _"man [being] the molecule of society [and their interaction being subject to] the direct ratio of the mass and the inverse one of distance"_ [pg. 41,43], the investigation has been coloured by these relationships between a pair of masses (e.g., quantity of people, opportunities, or other scales of mass), their separation distances, and some **proportionality constant** to yield some force of attraction, the features of Newton's Law of Universal Gravitation (1687).

Following this tradition with a focus on _potential_ attraction, the concept of accessibility was defined as the  _"potential of opportunities for interaction"_ in @hansen1959. The work of @hansen1959 proved to be highly-influential, with literature demonstrating a link between Hansen-type measures (i.e., a mass term x distance function, also known as gravity-based accessibility measures) and positive population-based-, sustainability- and economic- outcomes (though often at odds). Moreover, accessibility planning counters mobility-based planning by centering the potential for mobility, arguably the ulterior 'good' provided by transportation systems, not mobility itself. For this theoretical reason, Hansen-type measures are becoming an increasingly popular transportation planning instrument signaling that accessibility-based planning over mobility-based perspectives may be on the horizon. However, an important critique still stands in the way: the difficultly in interpreting the accessibility values themselves.

Notably, the Hansen-type model of today is missing a proportionality constant. The doctoral research hypothesizes that this omission is the source of interpretation difficulty, and can be remedied by drawing from related 'gravity-based' transportation literature. Related literature includes spatial interaction models (the gravity model) [@wilson1971] as it also traces back to the same early 20th century investigations as the accessibility concept [@carrothersHistoricalReviewGravity1956; @griggRavensteinLawsMigration1977]. Unlike the gravity-based accessibility measure, the gravity model still carries all the features of the Newtonian analogy (i.e., balancing factors (i.e, a proportionality constant), mass term 1, mass term 2, and distance function [@wilson1971]): a proportionality constant is crucial for retaining the output as a tangible unit. For the gravity model, which measures _interaction_, the output units are often the number of trips generated between an origin and destination. For accessibility, the doctoral research reckons the units are _potential_ interaction, and the output could be the number of opportunities that can be potentially interacted with from the origin to all destinations.

If selected, the applicant will introduce a family of accessibility measures in a similar vein as the "family of spatial interaction models" [@wilson1971] by specifying types of balancing factors, demonstrating simple synthetic examples, and visualizations using data using origin-destination employment flows for Toronto, Canada. Though the research is ongoing, the applicant intends to defend this academic year, the empirical data for use in this work is already published as an open R data package paper [@soukhovTTS2016RDataSet2023], and the singly-constrained and composite singly-constrained examples are fully developed with associated peer-reviewed publications [@paez2019; @soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].

# Letter of nomination

<!-- this first para is largely copy-paste from the letter you submitted to SSHRC-->
Over the past few decades, a tension has been brewing between the importance of transportation for society and the economy, and its impacts on environment and people. The focus of transportation planning has traditionally been on mobility (facilitating the ease of movement over increasingly large regions), which has led to congestion in metropolitan regions that are home to large populations. As a result, for all its benefits, transportation is increasingly a tool for inefficient, energy-intensive movement. Researchers in the field of transportation has been sounding the alarm, and urging a change of focus from mobility to accessibility, that is, the ease of reaching destinations or _the potential of opportunities for interactions_. Transportation, after all, is seldom an end in and of itself, but rather a means to reach meaningful destinations.

This is why the development of accessibility planning practice is important; however, accessibility measures must be put in order. As mentioned in the research abstract, a significant barrier to the adoption of accessibility into planning has to do with the issue of unit interpretation. The described doctoral research hypothesises a solution to this interaction issue by bridging the spatial interaction modeling's balancing factor typology onto accessibility measures, to create a family of accessibility measures. In this way, the research can potentially be a highly significant and timely, especially since a deeper understanding of the outputs of _potential_ interaction is required to respond to the rapidly changing transportation and land-use in urban areas and beyond.

Their anticipated graduate date is August 2025.

### 2024-09-10 TRB/workshop-abstract-and-letter.qmd
Letter of nomination for TRB Doctoral Workshop

@@
 Though the research is ongoing, the applicant intends to defend this academic year, the empirical data for use in this work is already published as an open R data package paper [@soukhovTTS2016RDataSet2023], and the singly-constrained and composite singly-constrained examples are fully developed with associated peer-reviewed publications [@paez2019; @soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
-
-# Letter of nomination
-
-<!-- this first para is largely copy-paste from the letter you submitted to SSHRC-->
-Over the past few decades, a tension has been brewing between the importance of transportation for society and the economy, and its impacts on environment and people.
-The focus of transportation planning has traditionally been on mobility (facilitating the ease of movement over increasingly large regions), which has led to congestion in metropolitan regions that are home to large populations.
-As a result, for all its benefits, transportation is increasingly a tool for inefficient, energy-intensive movement.
-Researchers in the field of transportation has been sounding the alarm, and urging a change of focus from mobility to accessibility, that is, the ease of reaching destinations or _the potential of opportunities for interactions_.
-Transportation, after all, is seldom an end in and of itself, but rather a means to reach meaningful destinations.
-
-This is why the development of accessibility planning practice is important; however, accessibility measures must be put in order.
-As mentioned in the research abstract, a significant barrier to the adoption of accessibility into planning has to do with the issue of unit interpretation.
-The described doctoral research hypothesises a solution to this interaction issue by bridging the spatial interaction modeling's balancing factor typology onto accessibility measures, to create a family of accessibility measures.
-In this way, the research can potentially be a highly significant and timely, especially since a deeper understanding of the outputs of _potential_ interaction is required to respond to the rapidly changing transportation and land-use in urban areas and beyond.
-
-Their anticipated graduate date is August 2025.

### 2024-09-23 paper/manuscript.qmd
trb worshop proposal abstract and nomination letter draft
Edits to manuscript

@@
 format: pdf
+bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
+nocite: |
+  @careyPrinciplesSocialScience1858, @ravensteinLawsMigration1889, @ravensteinLawsMigration1885
 ---
@@
 <!--The inquiry into spatial interaction traces to Newts Universal Gravitation-->
-The inquiry into people's interaction in space has been a popular subject of investigation for over a century.
-From as far back as @careyPrinciplesSocialScience1858's sentiments on "man [being] the molecule of society [and their interaction being subject to] the direct ratio of the mass and the inverse one of distance" [pg.
-41,43], the investigation has often been coloured by the relationship between a pair of masses $M_i$ and $M_j$ (quantity of people, opportunities, or other _scales_ of mass), their separation distances $D_{ij}$, and some proportionality constant $G$ to yield the force of attraction $F$, the features of Newton's Law of Universal Gravitation (1687) @eq-1:
+The patterns of people's movement in space have been a subject of scientific inquiry for centuries.
+From as far back as 1858 with Carey's Principles of Social Science where he stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [see McKean's -@mckeanManual1883 Manual of Social Science, pp.
+37-38], investigations of how people interact in space has often been  explicitly coloured by the features of Newton's Law of Universal Gravitation, as posited in his _Principia Mathematica_ of 1687 (see @eq-grav-prop):
+
+$$
+F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
+$${#eq-grav-prop}
+
+@eq-grav-prop is one of the most famous equations in all of science, and it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$.
+Direct proportionality only means that as the product of the masses increases, so does the force.
+Likewise, inversely proportional only means that as the distance increases, the force decreases.
+To actually quantify the force, an empirical factor is required to convert the expression for the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the attraction between masses.
+Ultimately, the equation for the force is as seen in @eq-grav:
 
@@
 F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
-$${#eq-1}
+$${#eq-grav}
+
+\noindent where $G$ is a proportionality constant obtained empirically, and called _Cavendish gravitational constant_, after Henry Cavendish's experimental results of 1798--more than a century after Newton's _Principia_ was published--that brought it to within 1\% accuracy level.
 
 <!--Using Newts Universal Gravitation: Ravenstein 1889 to Stewart 1948-->
-Over a century ago, @ravensteinLawsMigration1889's proposed "Laws to Migration" from the empirical analysis of migration flows in early British censuses.
-This empirical investigation ignited many others looking to formulate models of human interaction based on the physical gravitation law (@eq-1).
-Notably, @reilly1929methods formulated the law of retail gravitation  <!--I can find the original text of this online, only discussed by others as important-->, Zipf's $\frac{P_1P_2}{D}$ hypothesis holding in the case of intercity personal movement @zipfHypothesisIntercityMovement1946, goods movement by railways @zipfHypothesisCaseRailway1946, and information @zipfDeterminantsCirculationInformation1946, and @stewartDemographicGravitationEvidence1948 introduced principles of demographic gravitation.
+Following Carey, research into human spatial interaction continued apace in different contexts.
+In 1885 and 1889 Ravenstein (who appears to have been unaware of Carey's treatise) proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries.
+In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination, a clear similitude to gravitational laws.
+Other researchers kept discovering, seemingly independently, principles of human spatial interaction that resembled Newton's universal law of gravitation.
+For example, @reilly1929methods (who seems to have been unaware of Carey and Ravenstein) formulated a law of retail gravitation  <!--Rilley does not cite Newton, nor Carey, nor Ravenstein-->, expressed in terms of equal attraction to competing retail destinations.
+Later, Zipf (seemingly unaware of Carey, Ravenstein, and Rilley) proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946.
+This hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
+
+A common theme among these early researchers of human spatial interaction is that a proportionality constant is missing.
+Of the researchers cited here, only Reilly and Zipf expressed their hypotheses in mathematical terms.
+In this way, Reilly expressed his hypothesis as follows:
+$$
+B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
+$${#eq-reilly}
+
+\noindent where $B_a$ is the businness drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
+Quantity $N$ was chosen by Reilly in a somewhat ad hoc fasion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
+
+Zipf, on the other hand, expressed his hypothesis as:
+$$
+C^2 = \frac{P_1\cdot P_2}{D}
+$${#eq-zipf}
+
+\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.
+
+After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law (Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists), and also recognition of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction.
+Stewart appears to be the first author to express his theorized relationships for human spatial interaction with a proportionality constant, as follows:
+$$
+F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
+$${#eq-stewart-force}
+
+\noindent where:
+
+- $F$ is the _demographic force_
+- $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
+- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
+- $d_{12}^2$ is the distance between $1$ and $2$
+- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]
+
+In addition to demographic force, Stewart defined a measure of "population potential" of $2$ with respect to $1$ as follows:
+$$
+V_1 = G\frac{M_2}{d}
+$${#eq-stewart-population-potential}
+
+For a system with more than 2 population bodies, Stewart formulated the population potential at $i$ as:
+$$
+V_i = \int\frac{D}{r} ds
+$${#eq-stewart-population-potential-integral}
+
+\noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. In more contemporary terms, this could be rewritten as:
+$$
+V_i = \sum_j \frac{M_j}{d_{ij}}
+$${#eq-stewart-population-potential-sum}
+
+\noindent which astute readers will identify as our modern definition of accessibility.
+
+Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic and had issues with inconsistent mathematical notation [e.g., in Stewart's 1948 paper, $G$ is used as a proportionality constant on p.
+34 and again for the demographic energy on p.
+54], but perhaps more seriously, it was permeated by a view of humans as particles and afflicted by unmitigated racism (Stewart assumed that the "molecular weight" $\mu$ was one for the average American, but "presumably...much less than one" for an Australian Aboriginal; p.
+34).
+Eventually, Stewart's ideas about "social physics" fell out of favor among social scientists, but not before influencing Walter G.
+Hansen, then a student of city and regional planning at Massachusetts Institute of Technology.
 
@@
 -->
-Following this literature, accessibility as the  _"potential of opportunities for interaction"_ defined in @hansen1959 emerged.
-Today, @hansen1959 is frequently cited as the father of the term's scientific use [@reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
-@hansen1959 borrowed the "population potential" formulation used in @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to measure potential of opportunities (for interaction) (@eq-2).
-^[The integral of $F_{ij}$ (@eq-1) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$.
-The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields 'potential population'.
-@hansen1959 takes this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.] $S_ij$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
+Hansen, was an engineer with the the Bureau of Roads in the United States, and preoccupied with transportation's power to shape land uses.
+As such, he was neither scientist nor social scientist, but rather an applied researcher.
+Hansen's [-@hansen1959] paper in the Journal of the American Planning Association was instrumental in shedding some of the less convincing, or frankly objectionable, aspects of "social physics", to recast accessibility not as a model of physics, but as a model useful to capture regularities in human spatial behavior.
+The result of this was accessibility as the  _"potential of opportunities for interaction"_.
+Today, @hansen1959 is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
+
+In his work, @hansen1959 borrowed the idea of a "population potential" of @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to the measurement of the potential of opportunities.
+As noted above, the integral of $F_{ij}$ (@eq-grav-prop) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$.
+The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields "potential population".
+@hansen1959 took this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.]
+
+(@eq-2):
+$$
+S_{ij} = O_j \cdot d_{ij}^\beta \\
+$${#eq-accessibility}
+
+$S_{ij}$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
 ^[Notably, @Hansen1959 finds that when the model is empirically calibrated, ${\beta}$ takes a value of between 0.5 to 3, not the value of 1 as consistent in the Newtonian potential energy relationship.]
 
-$$
-S_{ij} = O_j * d_{ij}^\beta \\
-$${#eq-2}
-
-Of important note, however is that @hansen1959's formulation omits a proportionality constant akin to the function of $G$ in @eq-1.
+Of note, however is that @hansen1959's formulation omits a proportionality constant akin to the function of $G$ in @eq-grav.
 Interestingly, as @hansen1959 draws from @stewartDemographicGravitationEvidence1948, @stewartDemographicGravitationEvidence1948 sets the proportionality constant to 1, stating _"the constant G is left for future determination: a suitable choice of other units can reduce it to unity"_.

### 2024-10-08 paper/manuscript.qmd
Antonio's edits before sending paper back to Anastasia

@@
 
+ANAASTASIA: A brief intro describing the aims of the paper and anticipating the conclusions needs to go here.
+
+<!-- the paper's aim-->
+This work aims to re-unify accessibility literature with its Newtonian roots by mapping it onto the family of gravity models.
+Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history.
+We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today.
+In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained.
+Then, we will provide empirical examples of the types of measures.
+Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.
+
+# Background
+
 <!--The inquiry into spatial interaction traces to Newts Universal Gravitation-->
@@
 Likewise, inversely proportional only means that as the distance increases, the force decreases.
-To actually quantify the force, an empirical factor is required to convert the expression for the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the attraction between masses.
+To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses.
 Ultimately, the equation for the force is as seen in @eq-grav:
@@
 
-\noindent where $G$ is a proportionality constant obtained empirically, and called _Cavendish gravitational constant_, after Henry Cavendish's experimental results of 1798--more than a century after Newton's _Principia_ was published--that brought it to within 1\% accuracy level.
+\noindent where $G$ is a proportionality constant obtained empirically and called _Cavendish gravitational constant_, after Henry Cavendish's experimental results of 1798--more than a century after Newton's _Principia_ was published--that brought it to within 1\% accuracy level.
 
@@
 In 1885 and 1889 Ravenstein (who appears to have been unaware of Carey's treatise) proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries.
-In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination, a clear similitude to gravitational laws.
-Other researchers kept discovering, seemingly independently, principles of human spatial interaction that resembled Newton's universal law of gravitation.
+In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination, in clear similitude to Newton's gravitational laws.
+Over time other researchers kept discovering, seemingly independently, some principles of human spatial interaction that resembled Newton's universal law of gravitation.
 For example, @reilly1929methods (who seems to have been unaware of Carey and Ravenstein) formulated a law of retail gravitation  <!--Rilley does not cite Newton, nor Carey, nor Ravenstein-->, expressed in terms of equal attraction to competing retail destinations.
@@
 
-A common theme among these early researchers of human spatial interaction is that a proportionality constant is missing.
+A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing.
 Of the researchers cited here, only Reilly and Zipf expressed their hypotheses in mathematical terms.
-In this way, Reilly expressed his hypothesis as follows:
+Reilly's hypothesis was expressed as follows:
 $$
@@
 
-\noindent where $B_a$ is the businness drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
-Quantity $N$ was chosen by Reilly in a somewhat ad hoc fasion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
+\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
+Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fasion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
 
-Zipf, on the other hand, expressed his hypothesis as:
+Zipf, on the other hand, wrote his hypothesis in mathematical form as:
 $$
@@
 
-After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law (Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists), and also recognition of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction.
-Stewart appears to be the first author to express his theorized relationships for human spatial interaction with a proportionality constant, as follows:
+After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law (Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists), besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction.
+Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
 $$
@@
 - $d_{12}^2$ is the distance between $1$ and $2$
-- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]
+- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34] but it is unclear whether he revisited
 
@@
 
-\noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. In more contemporary terms, this could be rewritten as:
+\noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. In more contemporary terms, this can be rewritten as:
 $$
@@
 34).
-Eventually, Stewart's ideas about "social physics" fell out of favor among social scientists, but not before influencing Walter G.
+It did not take long for Stewart's ideas about "social physics" to fall out of favor among social scientists, but not before influencing Walter G.
 Hansen, then a student of city and regional planning at Massachusetts Institute of Technology.
@@
 -->
-Hansen, was an engineer with the the Bureau of Roads in the United States, and preoccupied with transportation's power to shape land uses.
+Hansen's day job was as an engineer with the the Bureau of Roads in the United States, and preoccupied with transportation's power to shape land uses.
 As such, he was neither scientist nor social scientist, but rather an applied researcher.
-Hansen's [-@hansen1959] paper in the Journal of the American Planning Association was instrumental in shedding some of the less convincing, or frankly objectionable, aspects of "social physics", to recast accessibility not as a model of physics, but as a model useful to capture regularities in human spatial behavior.
-The result of this was accessibility as the  _"potential of opportunities for interaction"_.
+Hansen's [-@hansen1959] paper in the Journal of the American Planning Association was instrumental in shedding some of the less convincing, or frankly objectionable, aspects of "social physics", to recast accessibility not as a model of physics, but as a model of human behavior useful to capture regularities in human mobility patterns.
+In his work, @hansen1959 borrowed the idea of a "population potential" of @stewartDemographicGravitationEvidence1948, which closely followed the Newtonian definition of energy potential, and applied it to the measurement of the potential of opportunities.
+The potential energy of one mass, say $M_i$, when applied to population instead of celestial bodies yields "potential population":
+$$
+S_{ij} = \sum_j O_j \cdot d_{ij}^\beta \\
+$${#eq-accessibility}
+
+In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_.
+A more general form of this expression is as follows, to account for different impedance functions to reflect variations in travel behavior:
+$$
+S_{ij} = \sum_j O_j \cdot f(d_{ij})
+$${#eq-accessibility-general}
+
+$S_{ij}$ in @eq-accessibility is is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
 Today, @hansen1959 is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
 
-In his work, @hansen1959 borrowed the idea of a "population potential" of @stewartDemographicGravitationEvidence1948, which closely follows the Newtonian definition of energy potential, and applied it to the measurement of the potential of opportunities.
-As noted above, the integral of $F_{ij}$ (@eq-grav-prop) over distance $d_{ij}$: $U_{ij}=-G M_{i}M_{j} /d_{ij}$.
-The potential energy of one mass, say $M_i$, is hence $U_i = G M_j / d_{ij}$, when applied to population instead of mass this equation yields "potential population".
-@hansen1959 took this equation and replaces $M_j$ with _opportunities_ to derive _opportunity potential_ or, as Hansen defines, _potential of opportunities for interaction_.]
+Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.
 
-(@eq-2):
-$$
-S_{ij} = O_j \cdot d_{ij}^\beta \\
-$${#eq-accessibility}
-
-$S_{ij}$ is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
-^[Notably, @Hansen1959 finds that when the model is empirically calibrated, ${\beta}$ takes a value of between 0.5 to 3, not the value of 1 as consistent in the Newtonian potential energy relationship.]
-
-Of note, however is that @hansen1959's formulation omits a proportionality constant akin to the function of $G$ in @eq-grav.
-Interestingly, as @hansen1959 draws from @stewartDemographicGravitationEvidence1948, @stewartDemographicGravitationEvidence1948 sets the proportionality constant to 1, stating _"the constant G is left for future determination: a suitable choice of other units can reduce it to unity"_.
+There is some evidence that @hansen1959 was aware of the import of this constant, since in his description of the process of interest he wrote that "the formulation states that the accessibility at point 1 to a particular type of activity at area 2 (say employment) is _directly proportional_ to the size of the activity at area 2 (number of jobs) and _inversely proportional_ to some function of the distance separating point 1 from area 2" (p.
+73; our emphasis).
+The reason for this treatment of the proportionallity constant could simply be the empirical and theoretical limitations of the state of knowledge at the time when @hansen1959 drew from @stewartDemographicGravitationEvidence1948.
+After all, in the latter the proportionality constant was set to 1 and that _"the constant $G$ [was] left for future determination: a suitable choice of other units can reduce it to unity"_.
+In practice, this is still done today, and the proportionality constant continues to be implicitly set to $1$ even when the relationship is clearly of proportionality and not equality [see for instance the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
 Without a proportionality constant, however, the units of $S_ij$ are unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.
-@eq-3 is the form of gravity-based accessibility measure commonly presented today.
-It is re-written with $S_i$ being the relative measure of the accessibility at $i$ to all activities $j$, and instead of $d_{ij}^\beta$, $f(c_{ij})$ is any monotonically decreasing function of travel cost $c_{ij}$.
-
-$$
-S_{i} = \sum_j O_j * f(c_{ij})
-$${#eq-3}
 
 <!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
-Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to the authors knowledge, have not been concerned with explicitly determining a proportionality constant.
-The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$ (), and additions to the opportunity $O_j$ attribute such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
-However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction, plagues it to this day and contributes to resistance to its application in practice().
-The quantity and variations on @eq-3 compounds this issue of interpretability across studies.
-A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (gravity), utility-, location-, and person- based measures with different components, and @wuUnifyingAccess2020 demonstrates that the concept can be re-unified into its main features: the relationship between travel-cost and corresponding reachable opportunities and demonstrates how previous classifications are special cases of travel-cost or opportunities constraints.
-From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing - the proportionality constant $G$.
-The aim of this paper's focus is on the proportionality constant to re-unify accessibility research with its Newtonian gravitational roots.
+Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant.
+The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$, and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
+However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility), plagues this type of analysis to the date, and contributes to resistance to its application in practice [REFERENCE MISSING??].
+The quantity and variations on @eq-accessibility-general compounds the issue of interpretability across studies.
+A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components.
+@wuUnifyingAccess2020 demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities.
+Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints.
+From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing—the proportionality constant $G$.
 
-<!--To address our paper's aim: we referr to the gravity model of today (Wilson) as it is based in the same Newtonian gravitational analogy as accessibility as well as Paez et al.
+<!--
+Anastasia: Maybe the idea in the text below is better left for the discussion to note the exceptions to the rule that the proportionality constant is implicitly set to one:
+
+"To address our paper's aim: we refer to the gravity model of today (Wilson) as it is based in the same Newtonian gravitational analogy as accessibility as well as Paez et al.
 2019 and Soukhov et al.
-2023,2024 -->
-The aim of this work is to demonstrate the importance of the proportionality constant consideration within @eq-3 to balancing the units of accessibility into an interpretable value thereby clarifying its meaning.
-To address this answer this question, we need not reinvent the wheel, instead we refer to the general gravity model of today (@eq-4) as introduced in @wilson1971.
+2023,2024"
+-->
+
+The aim of this paper's focus is on the proportionality constant.
+We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings.
+To address this matter we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971.
+In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts.
+Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.
+
+# Enter Wilson and entropy maximization
+
 The gravity model traces back to these early 20th century empirical investigations [@carrothersHistoricalReviewGravity1956; @griggRavensteinLawsMigration1977; @haynesGravitySpatialInteraction1985; @oshanSpatialStructureDebate2021]; it also retains the proportionality constant and the resulting output is a tangible unit.
@@
 
-<!-- reiterate the paper's aim-->
-This work aims to re-unify accessibility literature with its Newtonian roots by mapping it onto the family of gravity models.
-Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history.
-We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today.
-In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained.
-Then, we will provide empirical examples of the types of measures.
-Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.
-
 # A family of accessibility measures
@@
 Derive the various accessibility measures. In my mind:
+
 •	Non-constrained: it does not include a single proportionality constant.
 
-•	Unconstrained/Total constrained: uses a single proportionality constant
+•	Total-constrained: uses a single proportionality constant
 
-•	Singly constrained: spatial availability
+•	Origin- or Destination-constrained: spatial availability
+
+•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)
+ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
 
@@
 •	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)
-
-•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)
-ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
 

### 2024-11-07 paper/manuscript.qmd
Draft of introduction and lit review, draft attempt at numerical formulation

@@
 
-ANAASTASIA: A brief intro describing the aims of the paper and anticipating the conclusions needs to go here.
+<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy.
+This has had problematic consequences.
+Conceptually, accessibility is a more complete and comprehensive evolution of mobility.
+As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->
+Accessibility is conceptually related to human mobility, it is often defined as the "potential of opportunities for interaction" [@hansen1959], opportunities such as places, events and social contacts that are important to people [@ferreiraReenactingMobilityAccessibility2020].
+Accessibility adds a more comprehensive and complete understanding onto mobility, which simplify emphasizes the ability to move.
+However, the motivation of transportation planning has historically centered mobility and treated accessibility as its proxy, leading to problematic consequences [@handy2020].
+In the rise of automobility-dominance, recent decades of mobility-based practice (e.g., highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost) often not having an impact on accessibility levels [@handyACCESSIBILITYVSMOBILITYENHANCING2002].
+In response, transportation planning practice has begin to adopt accessibility planning approaches, be it through the use of accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling].
+However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and tangible meaning of the outputs [@ferreiraReenactingMobilityAccessibility2020].
 
-<!-- the paper's aim-->
-This work aims to re-unify accessibility literature with its Newtonian roots by mapping it onto the family of gravity models.
+<!-- Considering population-level methods in transportation and city planning research, accessibility measures have been used to tackle a wide variety of research questions.
+They have taken on numerous definitions and analytical expressions; from simpler cumulative opportunities measures, to measures considering more complex travel behaviour and trip weighting characteristics, measures of _potential of opportunities for interaction_ are varied.
+Summarizing these definitions across studies have been challenging; a testament to these unsettled classifications are present in influential accessibility review literature [@handyMeasuringAccessibilityExploration1997; @geursAccessibilityEvaluationLanduse2004; @wuUnifyingAccess2020; @millerAccessibilityMeasurementApplication2018].
+-->
+
+<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs.
+We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.
+-->
+In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its gravity-based Newtonian roots.
+It proposes to do so by mapping accessibility onto the family of gravity models.
 Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history.
@@
 
-<!--The inquiry into spatial interaction traces to Newts Universal Gravitation-->
+## The inquiry into spatial interaction traces to Newton's Universal Gravitation
 The patterns of people's movement in space have been a subject of scientific inquiry for centuries.
-From as far back as 1858 with Carey's Principles of Social Science where he stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [see McKean's -@mckeanManual1883 Manual of Social Science, pp.
-37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation, as posited in his _Principia Mathematica_ of 1687 (see @eq-grav-prop):
+From as far back as 1858 in Henry C.
+Carey's _Principles of Social Science_ where "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
+37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-grav-prop].
 
@@
 To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses.
-Ultimately, the equation for the force is as seen in @eq-grav:
+Ultimately, the equation for the force is as seen in @eq-grav, where $G$ is a proportionality constant obtained empirically:
 
@@
 
-\noindent where $G$ is a proportionality constant obtained empirically and called _Cavendish gravitational constant_, after Henry Cavendish's experimental results of 1798--more than a century after Newton's _Principia_ was published--that brought it to within 1\% accuracy level.
-
-<!--Using Newts Universal Gravitation: Ravenstein 1889 to Stewart 1948-->
-Following Carey, research into human spatial interaction continued apace in different contexts.
-In 1885 and 1889 Ravenstein (who appears to have been unaware of Carey's treatise) proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries.
-In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination, in clear similitude to Newton's gravitational laws.
-Over time other researchers kept discovering, seemingly independently, some principles of human spatial interaction that resembled Newton's universal law of gravitation.
-For example, @reilly1929methods (who seems to have been unaware of Carey and Ravenstein) formulated a law of retail gravitation  <!--Rilley does not cite Newton, nor Carey, nor Ravenstein-->, expressed in terms of equal attraction to competing retail destinations.
-Later, Zipf (seemingly unaware of Carey, Ravenstein, and Rilley) proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946.
-This hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
+## Early human spatial interaction research: from Ravenstein 1889 to Stewart 1948
+<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
+This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy.
+-->
+Following Carey, research into human spatial interaction continued in different contexts.
+In 1885 and 1889, Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885;@ravensteinLawsMigration1889].
+In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination.
+These propositions are similar to Newton's gravitational laws.
+Over time, other researchers discovered similar relationships.
+<!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations.
+Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946.
+The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
 
 A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing.
-Of the researchers cited here, only Reilly and Zipf expressed their hypotheses in mathematical terms.
+Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms.
 Reilly's hypothesis was expressed as follows:
@@
 \noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
-Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fasion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
+Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
 
@@
 
-After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law (Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists), besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction.
-Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
+After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law.
+We suspect that may relate to academic background; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists.
+Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
+
 $$
@@
 - $d_{12}^2$ is the distance between $1$ and $2$
-- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34] but it is unclear whether he revisited
+- And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]
 
@@
 
-Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic and had issues with inconsistent mathematical notation [e.g., in Stewart's 1948 paper, $G$ is used as a proportionality constant on p.
-34 and again for the demographic energy on p.
-54], but perhaps more seriously, it was permeated by a view of humans as particles and afflicted by unmitigated racism (Stewart assumed that the "molecular weight" $\mu$ was one for the average American, but "presumably...much less than one" for an Australian Aboriginal; p.
-34).
-It did not take long for Stewart's ideas about "social physics" to fall out of favor among social scientists, but not before influencing Walter G.
-Hansen, then a student of city and regional planning at Massachusetts Institute of Technology.
+Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic.
+It had issues with inconsistent mathematical notation.
+For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p.
+34 <!--Antonio, is this what you meant?-->.
+And perhaps more seriously, it was permeated by a view of humans as particles, following physical laws, laws of which were afflicted by unmitigated racism.
+For instance, molecular weight $\mu$ was taken as one for the average American, but "presumably...much less than one....for an Australian aborigine" [p.
+35].
+Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the field of accessibility today.
 
-<!--Hansen leans on Stewart's 1948 Population Potential measure.
-Two crucial things: 1) Stewart set the proportional constant to 1 *for future determination* and 2) Stewart suggests that keeping the exponent value of d_ij as 1 is sufficient.
-Hansen doesn't address the proportional constant of 1 explicitly in his application and picks an exponent other than 1 for d_ij as it fits empirically better: the issue of units are never addressed.
+In summary: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$.
+Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.
+
+## Hansen's gravity-based accessibility
+<!-- SUMMARY: in Hansen's use of @stewartDemographicGravitationEvidence1948's Population Potential measure, we highlight two crucial omissions.
+1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it.
+2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient.
+In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed.
 -->
-Hansen's day job was as an engineer with the the Bureau of Roads in the United States, and preoccupied with transportation's power to shape land uses.
-As such, he was neither scientist nor social scientist, but rather an applied researcher.
-Hansen's [-@hansen1959] paper in the Journal of the American Planning Association was instrumental in shedding some of the less convincing, or frankly objectionable, aspects of "social physics", to recast accessibility not as a model of physics, but as a model of human behavior useful to capture regularities in human mobility patterns.
-In his work, @hansen1959 borrowed the idea of a "population potential" of @stewartDemographicGravitationEvidence1948, which closely followed the Newtonian definition of energy potential, and applied it to the measurement of the potential of opportunities.
-The potential energy of one mass, say $M_i$, when applied to population instead of celestial bodies yields "potential population":
+
+From @stewardDemographicGravitationEvidence1948, we arrive to Walter G.
+Hansen, the author of @hansen1959 which has been exceptionally influential in the accessibility literature.
+In this seminal paper @hansen1959, "accessibility is defined as the potential of opportunities for interaction" [p.73], it is "a generalization of the population-over-distance relationship or "population potential" concept developed by Stewart [@stewardDemographicGravitationEvidence1948]" [p.73], explicitly defined "as the potential of opportunities for interaction" [p.73].
+As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads likely preoccupied with applied research that can be used to qualify transportation's power to shape land uses.
+@hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in (@eq-stewart-population-potential-sum)), leaving out Stewart's other formulaic contributions and objectionable aspects of "social physics".
+@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful in capturing regularities in mobility patterns .
+In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_ as follows:
 $$
-S_{ij} = \sum_j O_j \cdot d_{ij}^\beta \\
+S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
 $${#eq-accessibility}
 
-In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_.
-A more general form of this expression is as follows, to account for different impedance functions to reflect variations in travel behavior:
+Alternatively, a more general form of this expression to account for different impedance functions:
 $$
-S_{ij} = \sum_j O_j \cdot f(d_{ij})
+S_{i} = \sum_j O_j \cdot f(d_{ij})
 $${#eq-accessibility-general}
 
-$S_{ij}$ in @eq-accessibility is is a relative measure of the accessibility at $i$ to an activity located at $j$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
-Today, @hansen1959 is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011] and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
+$S_{i}$ in @eq-accessibility is is a relative measure of the accessibility at $i$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
+Today, Hansen is is frequently cited as the father of modern accessibility analysis due to his work in @hansen1959 [e.g., @reggianiGuestEditorialNew2011], and the Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
 
 Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.
+There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance.
+Furthermore, those reading @hansen1959 could consult @stewartDemographicGravitationEvidence1948 in which the proportionality constant $G$  was set to 1 and that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
 
-There is some evidence that @hansen1959 was aware of the import of this constant, since in his description of the process of interest he wrote that "the formulation states that the accessibility at point 1 to a particular type of activity at area 2 (say employment) is _directly proportional_ to the size of the activity at area 2 (number of jobs) and _inversely proportional_ to some function of the distance separating point 1 from area 2" (p.
-73; our emphasis).
-The reason for this treatment of the proportionallity constant could simply be the empirical and theoretical limitations of the state of knowledge at the time when @hansen1959 drew from @stewartDemographicGravitationEvidence1948.
-After all, in the latter the proportionality constant was set to 1 and that _"the constant $G$ [was] left for future determination: a suitable choice of other units can reduce it to unity"_.
-In practice, this is still done today, and the proportionality constant continues to be implicitly set to $1$ even when the relationship is clearly of proportionality and not equality [see for instance the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
-Without a proportionality constant, however, the units of $S_ij$ are unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.
+## From Hansen-type accessibility to accessibility today
+<!-- SUMMARY: accessibility from Hansen to today is very diverse and has a lack of concern for the interpretability of output units; likewise, a lack of concern for balancing units such as a proportionality constant.
+Our answer: let's focus on those omitted (or implicit set to 1) proportionality constants.-->
 
-<!--Accessibility from Hansen to today: very diverse, hard to interpret, and its lack of concern for the proportionality constant.  Our answer: the proportionality constant.-->
-Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant.
-The majority of studies develop on the breath of opportunity $O_j$ studied (i.e., work, shopping, schools, leisure () and their geographic case study selection, the accuracy of parameters that capture distance-decay $f(c_{ij})$, and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
-However, the difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility), plagues this type of analysis to the date, and contributes to resistance to its application in practice [REFERENCE MISSING??].
-The quantity and variations on @eq-accessibility-general compounds the issue of interpretability across studies.
+Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant $G$.
+$G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.
+
+To overview recent accessibility literature development, the majority of studies develop on; the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
+Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.
+
+However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [].
+This interpretation plagues this type of analysis, as implications beyond high-or-low access hotspots are often desired.
+What does the accessibility value mean?
+Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020].
+As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.
+
 A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components.
@@
 Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints.
-From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing—the proportionality constant $G$.
-
-<!--
-Anastasia: Maybe the idea in the text below is better left for the discussion to note the exceptions to the rule that the proportionality constant is implicitly set to one:
-
-"To address our paper's aim: we refer to the gravity model of today (Wilson) as it is based in the same Newtonian gravitational analogy as accessibility as well as Paez et al.
-2019 and Soukhov et al.
-2023,2024"
--->
+From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing, the proportionality constant $G$.
 
@@
 We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings.
-To address this matter we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971.
+To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971.
 In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts.
@@
 
-# Enter Wilson and entropy maximization
+# Gravity-based accessibility's sibling: the gravity model
 
-The gravity model traces back to these early 20th century empirical investigations [@carrothersHistoricalReviewGravity1956; @griggRavensteinLawsMigration1977; @haynesGravitySpatialInteraction1985; @oshanSpatialStructureDebate2021]; it also retains the proportionality constant and the resulting output is a tangible unit.
+Today's gravity-based accessibility measure (@eq-accessibility-general) share the same early 20th century Newtonian-based  origins as the gravity model, the most widely used model in the study of spatial interaction [@ortuzar_2011_modelling;@wilson1971].
+However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science- adjacent researchers and planners.
+<!-- Antonio, this is my sense at this point..?-->
+
+There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the classic gravity model, which takes the following form:
 
 $$
-T_{ij} = k P_i^{\lambda} P_j^{\alpha} d_{ij}^{\beta}
-$${#eq-4}
+T_{ij} = k O_i D_j f(c_{ij})
+$${#eq-gravity-model}
 
 <!--Explain the features of the gravity model of today (Wilson) -->
-Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between $n$ origins $i$ and $m$ destinations $j$, $P$ are a vector of mass attributes at origin $i$ and $j$ and cost to overcome physical separation, and $d_{ij}$ is a $n$ x $m$ matrix.
-$d_{ij}$ is also frequently replaced by $f(c_{ij})$ in the literature, some function of $c_{ij}$ which decreases as $c_{ij}$ increases.
-Furthermore, the exponents are introduced to modulate the effect size of the variables, where $\lambda$ and $\alpha$ modulate $P$ and $\beta$ modulates $d$; when variables $T$, $P$ and $d$ are known, the exponents can be estimated to summarize the variable's effect size contribution to the known flows in the system.
-Finally and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
-In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
+Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
+In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
+It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
+Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
+
+Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
+As such, @eq-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$.
+As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$.
+These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:
+
+$$
+\sum_j T_{ij} = O_i
+$${#eq-constraint1-gravitymodel}
+
+$$
+\sum_i T_{ij} = D_j
+$${#eq-constraint2-gravitymodel}
+
+Using these constraints, the gravity model @eq-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+
+- neither @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold -- the unconstrained case;
+- @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
+- @eq-constraint2-gravitymodel holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
+- @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel both hold simultaneously -- the production-attraction constrained case (doubly-constrained).
+
+In each of these four cases, the single proportionality constant $k$ is replaced by two sets of balancing factors $A_i$ and $B_j$ to ensure the case-specific constraints are satisfied.
+For the doubly-constrained case:
+$$
+T_{ij} = A_i O_i B_j D_j f(c_{ij})
+$${#eq-doublyconstrained-gravitymodel}
+
+And the balancing factors equal:
+$$
+A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}
+$${#eq-bf-A-gravitymodel}
+
+$$
+B_j = \frac{1}{\sum_j A_i O_j f(c_{ij})}
+$${#eq-bf-B-gravitymodel}
+
+For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel.
+In a way, the singly-constrained and doubly-constrained cases are variants of the unconstrained case.
+
+## Entropy maximising: the gravity-model as statistical averages
+Instead of the Newtonian gravity analogy, entropy maximisation is a popular model building approach taken in the spatial interaction studies to generate the gravity model.
+Namely, it is mathetmatically regorouse and offers a sensible interpretation of the resulting mode as a statistical averaging procedure.
+This allows the measure to be effective in describing behaviour in large populations but less effective in smaller decision making units in the region of interest.
+
+The basis of the approach is to accept that there are macro, meso, and micro states -- where all micro states that are consistent with our information about macro states are equally likely to occur.
+For the _interaction_ variable of trips $T_{ij}$, as shown in @wilson1971, the number of micro states $W\{{T_{ij}}\}$ within a meso state $T_{ij}$ can be expressed as:
+$$ 
+W\{{T_{ij}}\} = \frac{T!}{\prod_{ij}T_{ij}!}
+$${#eq-micro-meso-states}
+<!--I LEFT OUT THE DERIVATION... NOT SURE IF ITS NECESSARY? -->
+Using the maximizing entropy approach, we must find a set of $T_{ij}$ that maximizes @eq-micro-meso-states subject to our sensible macro-state constraints @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and the following on total cost $C$ constraint:
+$$
+\sum_{ij}T_{ij}c_{ij} = C
+$${#eq-constraint3-gravitymodel}
+
+The natural logarithm of @eq-micro-meso-states is maximized under @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and @eq-constraint3-gravitymodel for mathematical convenience, as both $ln W$ and $T_{ij}$ share the same maximum [@wilson1971; @ortuzar_2011_modelling].
+The result is:
+
+$$
+T_{ij} = A_i O_i B_j D_j exp(-\beta c_{ij})
+$${#eq-maxentropy-gravitymodel}
+
+Where,
+$$
+A_i = \frac{1}{\sum_j B_j D_j  \exp(- \beta c_{ij})}
+$${#eq-bf-maxentropy-A-gravitymodel}
+
+and
+$$
+B_j = \frac{1}{\sum_j A_i O_j  \exp(- \beta c_{ij})}
+$${#eq-bf-maxentropy-B-gravitymodel}
+
+Results of maximisation are identical to @eq-doublyconstrained-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-B-gravitymodel except the general $f(c_{ij})$ is $exp(-\beta c_{ij})$.
+From this approach, the gravity model can be interpretted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
+There is then a most probable distribution of trips between zones (meso-states) that stasify all the macro-state constraints.
 
@@
 
-It's all about balancing factors! Lessons from the balancing factors of the gravity model (cool paper that explores this @kirbyNormalizingFactorsGravity1970)
+Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities.
+$T_{ij}$ in @eq-gravity-model can be replaced with $V_{ij}$ and presented as follows:
 
-Derive the various accessibility measures. In my mind:
+$$
+V_{ij} = k O_i D_j f(c_{ij})
+$${#eq-access}
 
-•	Non-constrained: it does not include a single proportionality constant.
+Where $V_{ij}$ represents a $n$ x $m$ matrix of potential reachable opportunities between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
+In other words, $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
+Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
 
-•	Total-constrained: uses a single proportionality constant
+Like with @eq-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel).
+To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used.
+These pieces of information can constrain $V_{ij}$ as follows, ensuring it reflects this macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
 
-•	Origin- or Destination-constrained: spatial availability
+$$
+\sum_j V_{ij} = O_i
+$${#eq-constraint1-access}
+\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the opportunities originating $O_i$ at each $i$ in the region.
 
-•	Doubly constrained (only when the units of the marginals are the same, therefore of more limited application)
+$$
+\sum_i V_{ij} = D_j
+$${#eq-constraint2-access}
+\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the opportunities ending $D_i$ at each $j$ in the region.
+
+$$
+\sum_{i}\sum_{j} V_{ij} c_{ij} = C
+$${#eq-constraint3-access}
+\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.
+
+For all FIVE cases, @eq-constraint3-access holds <!--in the event C is known, it could be used to identify parameters in f(c_ij)...
+though model calibration is probably a better idea.
+--> but are differentiated like:
+- neither @eq-constraint1-access or @eq-constraint2-access hold -- the unconstrained case;
+- @eq-constraint1-access holds -- the production constrained case (origin-constrained, a singly-constrained case); 
+- @eq-constraint2-access holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
+- @eq-constraint1-access and @eq-constraint2-access both hold simultaneously -- the production-attraction constrained case (doubly-constrained).
+- @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->
+
+Using the maximizing entropy approach... <!--DESCRIBING WHAT EACH MICRO/MESO/MACRO STATE IS HERE IN RELATION TO ACCESSIBILITY--> For each case...
+
+## Unconstrained case
+
+Neither @eq-constraint1-access or @eq-constraint2-access hold, so $V_ij'$ has no opportunity constraints!
+
+$$
+V_{ij}' = k O_i D_j f(c_{ij})
+$${#eq-access-unconstrained}
+
+Where $k$ is two sets of balancing factors, $A_i'$ and $B_j'$, and both are equal to 1.
+
+## Singly-constrained case (origin-constrained)
+
+Spatial availability; origin-constrained.
+
+Origins in human spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998; for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024
+
+$$
+V_{ij}'' = k O_i D_j f(c_{ij})
+$${#eq-access-singlyconstrained}
+
+<!--HOW DO WE FIT OUR EXISTING CONSTRAINTS INTO THESE BALANCING FACTORS..? $k O_i f(c_{ij}) = the balancing factors..?-->
+$$
+V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
+$$
+Population balancing factor:
+$$
+F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}
+$$
+Travel cost balancing factor:
+$$
+F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}
+$$
+## Singly-constrained case (destination-constrained)
+
+Origins are in retail models, decision model for shops to optimally spatially locate to max.
+their potential consumer interaction.
+Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
+Spatial availability as another example.
+
+## Doubly constrained case (origin-destination constrained)
+
+(only when the units of the marginals are the same, therefore of more limited application)
 ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
 
-Origins in retail models, decision model for shops to optimally spatially locate to max.
-their potential consumer interaction.
-Starting with reilly's law of gravitational retail (1931) (can't find online), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
+## Hybrid singly-constrained case: 
+<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit??-->
+Balanced floating catchment areas (not truly doubly constrained)
 
-Independently formed in people spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998;for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024
-
-•	Composite singly constrained: Balanced floating catchment areas (not truly doubly constrained)
-
-# Examples
+# Empirical examples
 
@@
 # Conclusions
+
+# References
+
+<!--the benefits of this formulation, and the inclusion of the propoartionality constant has been imense for spatial interaction modelling.
+For instance, the identification of the conovlution of spatial autocorrelation and and distance-decay @griffithSpatialStructureSpatial2007 -->
+
+<!-- Wilson in @wilsonSTATISTICALTHEORYSPATIAL1967 was the first to connect accessibility measures to balancing factors-->
+
+<!--
+@wilson1971 (pg. 28) summarises the following checklist for spatial interaction model builders:
+1. Nature of the spatial interaction variable; 
+2. Nature of the mass terms
+3. Measurement of mass terms
+4. Measurement of interaction cost
+5. Choice of cost function
+6. Other terms
+7. Design of spatial system
+8. Calibration
+9. Relation to general model framework
+-->
+
+<!-- (cool paper that explores this @kirbyNormalizingFactorsGravity1970 - he "shows that the Ai and Bj balancing factors are the mean values of the inverse of the function e^-N*F(c_ij) calculated by summing the function over i values for Ai factors and over j values for Bj factors." -->

### 2024-11-07 paper/manuscript.qmd
Document after discussion with Anastasia

@@
 bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-nocite: |
-  @careyPrinciplesSocialScience1858, @ravensteinLawsMigration1889, @ravensteinLawsMigration1885
 ---
@@
 As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->
-Accessibility is conceptually related to human mobility, it is often defined as the "potential of opportunities for interaction" [@hansen1959], opportunities such as places, events and social contacts that are important to people [@ferreiraReenactingMobilityAccessibility2020].
-Accessibility adds a more comprehensive and complete understanding onto mobility, which simplify emphasizes the ability to move.
-However, the motivation of transportation planning has historically centered mobility and treated accessibility as its proxy, leading to problematic consequences [@handy2020].
-In the rise of automobility-dominance, recent decades of mobility-based practice (e.g., highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost) often not having an impact on accessibility levels [@handyACCESSIBILITYVSMOBILITYENHANCING2002].
-In response, transportation planning practice has begin to adopt accessibility planning approaches, be it through the use of accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling].
-However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and tangible meaning of the outputs [@ferreiraReenactingMobilityAccessibility2020].
+Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape.
+In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959].
+The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more.
+In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
+Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].
+However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020].
+With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling].
+However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geurs2004; @ferreiraReenactingMobilityAccessibility2020].
 
@@
 -->
-In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its gravity-based Newtonian roots.
-It proposes to do so by mapping accessibility onto the family of gravity models.
-Not only to clarify accessibility's interpretation as a measure, but also to encourage future developments that draw from its physics roots, gravity modelling siblings, and spatial interaction history.
-We began by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to it's split into accessibility research from @hansen1959 to today.
-In the following section, we will derive different types of accessibility measures, following the family of gravity models namely: non-constrained, constrained, unconstrained/Total constrained, singly constrained, composite singly constrained, doubly constrained.
+In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction-based roots.
+It proposes to do so by mapping accessibility onto Wilson's family of spatial interactions models [@wilson1971].
+In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its gravity modelling siblings, and spatial interaction history.
+To this end, we begin by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to its split into accessibility research from @hansen1959 to today.
+After this, we derive different types of accessibility measures, following the family of gravity models namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained.
 Then, we will provide empirical examples of the types of measures.
-Finally, we'll discuss the use of each measure and broadly the implications of ignoring the issue of proportionality in accessibility analysis.
+Finally, we will discuss the use of each measure and broadly the implications of ignoring the role of constrains in accessibility analysis.
 
@@
 ## The inquiry into spatial interaction traces to Newton's Universal Gravitation
+
 The patterns of people's movement in space have been a subject of scientific inquiry for centuries.
-From as far back as 1858 in Henry C.
-Carey's _Principles of Social Science_ where "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
-37-38], investigations of how people interact in space has often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-grav-prop].
+From as far back as Henry C.
+Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space.
+It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
+37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].
 
@@
 F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
-$${#eq-grav-prop}
+$${#eq-phys-grav-prop}
 
-@eq-grav-prop is one of the most famous equations in all of science, and it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$.
+To be certain, the expression above, a proportionality, is one of the most famous in all of science.
+In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$.
 Direct proportionality only means that as the product of the masses increases, so does the force.
-Likewise, inversely proportional only means that as the distance increases, the force decreases.
-To quantify the magnitude of the force, an empirical factor is required to convert the proportionality into an equation, which ensures that the force $F$ matches empirical observations of the force of attraction between masses.
-Ultimately, the equation for the force is as seen in @eq-grav, where $G$ is a proportionality constant obtained empirically:
+Likewise, inverse proportionality only means that as the distance increases, the force decreases.
+To quantify the magnitude of the force, an empirical factor, say $G$, is required to convert the proportionality into an equality, which ensures that the force $F$ matches empirical observations of the force of attraction between masses.
+Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is a proportionality constant obtained empirically:
 
@@
 F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
-$${#eq-grav}
+$${#eq-phys-grav}
 
-## Early human spatial interaction research: from Ravenstein 1889 to Stewart 1948
+Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that Hutton's experiments of 1778 validated [@hutton_xxxiii_1778, p.
+783].
+Still, it took over a century from the publication of _Principia_ to refine the estimate of this constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
+
+## Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
+
 <!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
@@
 -->
-Following Carey, research into human spatial interaction continued in different contexts.
-In 1885 and 1889, Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885;@ravensteinLawsMigration1889].
-In these works, Ravenstein posited a directly proportional relationship with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination.
+Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts.
+In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
+In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination.
 These propositions are similar to Newton's gravitational laws.
@@
 
-A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-grav is missing.
+A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav is missing.
 Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms.
@@
 
-After Carey, it is in Stewart's [-@stewartDemographicGravitationEvidence1948] work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law.
-We suspect that may relate to academic background; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists.
+After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
+We suspect that this may relate to academic backgrounds; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists.
 Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
-
 $$
@@
 34 <!--Antonio, is this what you meant?-->.
-And perhaps more seriously, it was permeated by a view of humans as particles, following physical laws, laws of which were afflicted by unmitigated racism.
-For instance, molecular weight $\mu$ was taken as one for the average American, but "presumably...much less than one....for an Australian aborigine" [p.
+Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism.
+For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p.
 35].
-Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the field of accessibility today.
+Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.
 
-In summary: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$.
+In summary: from the 1850s to the 1940s, researchers attempted to theoretically and empirically characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$.
 Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.
@@
 
-From @stewardDemographicGravitationEvidence1948, we arrive to Walter G.
-Hansen, the author of @hansen1959 which has been exceptionally influential in the accessibility literature.
-In this seminal paper @hansen1959, "accessibility is defined as the potential of opportunities for interaction" [p.73], it is "a generalization of the population-over-distance relationship or "population potential" concept developed by Stewart [@stewardDemographicGravitationEvidence1948]" [p.73], explicitly defined "as the potential of opportunities for interaction" [p.73].
-As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads likely preoccupied with applied research that can be used to qualify transportation's power to shape land uses.
-@hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in (@eq-stewart-population-potential-sum)), leaving out Stewart's other formulaic contributions and objectionable aspects of "social physics".
-@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful in capturing regularities in mobility patterns .
-In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive _opportunity potential_, or more accurately, _potential of opportunities for interaction_ as follows:
+From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G.
+Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
+In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction" [p.73], as "a generalization of the population-over-distance relationship or 'population potential' concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73].
+As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses.
+@hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
+@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns .
+In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
 $$
@@
 
-$S_{i}$ in @eq-accessibility is is a relative measure of the accessibility at $i$, $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
-Today, Hansen is is frequently cited as the father of modern accessibility analysis due to his work in @hansen1959 [e.g., @reggianiGuestEditorialNew2011], and the Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
+$S_{i}$ in @eq-accessibility is a measure of the accessibility at $i$, whereas $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
+Today, Hansen is is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
 
-Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-grav vanished.
+Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
 There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance.
-Furthermore, those reading @hansen1959 could consult @stewartDemographicGravitationEvidence1948 in which the proportionality constant $G$  was set to 1 and that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
+Furthermore, those reading @hansen1959 should remember that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$  to 1, with a note that"$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
 
@@
 
-Since @hansen1959, accessibility analysis has been widely employed across planning disciplines, but to our knowledge, developments have not been concerned with explicitly determining a proportionality constant $G$.
-$G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [CITE MAJOR REVIEWS OVER LAST 30 YEARS].
+However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it.
+In this way, $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
 Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.
@@
 However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [].
-This interpretation plagues this type of analysis, as implications beyond high-or-low access hotspots are often desired.
+Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired.
 What does the accessibility value mean?
@@
 
-# Gravity-based accessibility's sibling: the gravity model
+# Gravity-based accessibility's sibling: Wilson's spatial interaction model
 
-Today's gravity-based accessibility measure (@eq-accessibility-general) share the same early 20th century Newtonian-based  origins as the gravity model, the most widely used model in the study of spatial interaction [@ortuzar_2011_modelling;@wilson1971].
-However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science- adjacent researchers and planners.
+Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used [@ortuzar_2011_modelling;@wilson1971].
+However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners.
 <!-- Antonio, this is my sense at this point..?-->
 
-There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the classic gravity model, which takes the following form:
+There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:
 
@@
 T_{ij} = k O_i D_j f(c_{ij})
-$${#eq-gravity-model}
+$${#eq-phys-gravity-model}
 
@@
 In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
-It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
+It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-phys-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
 Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
@@
 Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
-As such, @eq-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$.
+As such, @eq-phys-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$.
 As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$.
 These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:
+
+Total constrained:
+$$
+\sum_i\sum_j T_{ij} = T
+$${#eq-constraint0-gravitymodel}
 
@@
 
-Using these constraints, the gravity model @eq-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+```{r}
+T_1 <- matrix(c(95, 23, 42, 27, 378, 45, 38, 36, 106), ncol = 3, byrow = TRUE)
 
-- neither @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold -- the unconstrained case;
+T_1
+```
+
+```{r}
+C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
+
+C
+```
+
+```{r}
+O_i <- rowSums(T_1)
+D_j <- colSums(T_1)
+
+O_i
+D_j
+```
+
+```{r}
+T <- sum(sum(T_1))
+
+T
+```
+
+Unconstrained (note that there is no proportionality constant!):
+$$
+\hat{T_{ij}} = \frac{O_i\cdot D_j}{C_ij}
+$$
+
+```{r}
+T_unc <- O_i %*% t(D_j) / C
+
+T_unc
+```
+
+If we do the sums over rows, columns and all:
+```{r}
+rowSums(T_unc)
+colSums(T_unc)
+sum(sum(T_unc))
+```
+
+Total flow constrained (needs a single/global proportionality constant):
+$$
+\hat{T_{ij}} = k\frac{O_i\cdot D_j}{C_{ij}}
+$$
+
+```{r}
+k <- T/sum(sum(T_unc))
+k
+```
+
+```{r}
+T_tot <- k * T_unc
+
+T_tot
+```
+
+Let's check those marginals and total:
+```{r}
+rowSums(T_tot)
+colSums(T_tot)
+sum(sum(T_tot))
+```
+
+Destination constrained:
+$$
+\hat{T_{ij}} = \frac{B_j\cdot O_i\cdot D_j}{C_{ij}}
+$$
+```{r}
+B_j <- colSums(T_1)/colSums(T_unc)
+
+B_j <- matrix(rep(B_j, 3), ncol = 3, byrow = TRUE)
+
+B_j
+```
+
+```{r}
+T_dest <- B_j * T_unc
+
+T_dest
+```
+
+Let's check those marginals and total:
+```{r}
+rowSums(T_dest)
+colSums(T_dest)
+sum(sum(T_dest))
+```
+
+Origin constrained:
+$$
+\hat{T_{ij}} = \frac{A_i\cdot O_i\cdot D_j}{C_{ij}}
+$$
+
+```{r}
+A_i <- rowSums(T_1)/rowSums(T_unc)
+
+A_i <- matrix(rep(A_i, 3), ncol = 3, byrow = FALSE)
+
+A_i
+```
+
+```{r}
+T_orig <- A_i * T_unc
+
+T_orig
+```
+
+Let's check those marginals and total:
+```{r}
+rowSums(T_orig)
+colSums(T_orig)
+sum(sum(T_orig))
+```
+
+Now how does this translate to accessibility?
+
+Opportunities (say, grocery stores):
+```{r}
+D_j <- c(7, 3, 2)
+```
+
+Unconstrained accessibility (no proportionality constant):
+$$
+S_i = \sum_j\frac{D_j}{C_{ij}}
+$$
+
+```{r}
+S_i <- (1/C) %*% D_j
+```
+
+These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
+```{r}
+sum(S_i)
+```
+
+Total flow constrained accessibility, needs a balancing factor:
+```{r}
+k <- sum(D_j)/sum(S_i)
+
+k
+```
+
+Total flow constrained accessibility:
+Unconstrained accessibility (no proportionality constant):
+$$
+S_i = k\sum_j\frac{D_j}{C_{ij}}
+$$
+
+```{r}
+S_tot <- k * S_i
+
+S_tot
+```
+
+Now check the sum:
+```{r}
+sum(S_tot)
+```
+
+Using these constraints, the gravity model @eq-phys-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+
+- None of @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel hold -- the unconstrained case;
+- Only @eq-constraint0-gravitymodel holds -- total flow constrained
 - @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
@@
 Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities.
-$T_{ij}$ in @eq-gravity-model can be replaced with $V_{ij}$ and presented as follows:
+$T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:
 
@@
 
-Like with @eq-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel).
+Like with @eq-phys-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel).
 To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used.

### 2024-11-08 paper/manuscript.qmd
cleaned up some text before the toy trip example.

@@
 
-# Gravity-based accessibility's sibling: Wilson's spatial interaction model
+# Gravity-based accessibility's cousin: Wilson's spatial interaction model
 
@@
 However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners.
-<!-- Antonio, this is my sense at this point..?-->
 
@@
 $$
-T_{ij} = k O_i D_j f(c_{ij})
+T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
 $${#eq-phys-gravity-model}
@@
 <!--Explain the features of the gravity model of today (Wilson) -->
-Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
+Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $W_i^{(1)}$ and $W_j^{(2)}$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
 In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
@@
 
+## Entropy maximising: the gravity-model as statistical averages
+
 Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
-As such, @eq-phys-gravity-model can take account of any additional information that can be built into the model to constrain the _interaction_ variable $T_{ij}$.
-As mentioned, $T_ij$ is in the unit of trips, so additional knowledge can be the total trip flows $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$.
-These pieces of information can constrain $T_{ij}$ as follows, ensuring it reflects this knowledge:
+But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling].
+Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].
 
-Total constrained:
+As such, @eq-phys-gravity-model can be derived to account for additional information to include constraints that reflect known information about the system and is mathematically rigorous.
+The interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$.
+These pieces of information can constrain $T_{ij}$ as follows.
+
+The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$.
+Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:
 $$
@@
 $$
-\sum_j T_{ij} = O_i
+\sum_i\sum_j T_{ij} c_{ij} = C
+$${#eq-constraint0_5-gravitymodel}
+
+The production constraint: the sum of trips ending at each $j$ is equal to the total flow originating from each $i$:
+$$
+\sum_j T_{ij} = W_i^{(1)}
 $${#eq-constraint1-gravitymodel}
 
+The attraction constraint: the sum of trips starting at each $i$ is equal to the total flow ending at each $j$:
 $$
-\sum_i T_{ij} = D_j
+\sum_i T_{ij} = W_j^{(2)}
 $${#eq-constraint2-gravitymodel}
 
+The entropy maximation derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967).
+The basis of the approach is to accept that there are macro, meso, and micro states.
+In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of indiivudal trip makers, and the macro state is the entire study region.
+Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
+From this approach, the gravity model can be interpreted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
+There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.
+
+Using these constraints, the gravity model @eq-phys-gravity-model is can now be distinguished into four general cases.
+These cases are presented in increase order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+
+- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
+- The total constrained case: only @eq-constraint0-gravitymodel hold
+- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
+- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
+
+In each of these four cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
+$$
+A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}
+$${#eq-bf-A-gravitymodel}
+
+$$
+B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})}
+$${#eq-bf-B-gravitymodel}
+<!--PRETTY SURE YOU DON'T GET K FROM ENTROPY MAXIMAISATION...-->
+$$
+K = \frac{T}{\sum_i \sum_j W_i^{(1)} W_j^{(2)} f(c_{ij})}
+$${#eq-bf-K-gravitymodel}
+
+For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$.
+For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel.
+For both singly-constrained cases, $K$ also holds.
+For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
+And for the unconstrained case, in a way, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.
+
+The following are simple numerical examples of each case assuming the following $T_{ij}$ matrix and a $f(c_{ij}) = \frac{1}{c_{ij}}$.
+
+The variables:
 ```{r}
@@
 
-Unconstrained (note that there is no proportionality constant!):
+## Unconstrained interaction case
+
 $$
@@
 
-Total flow constrained (needs a single/global proportionality constant):
+## Total constrained interaction case
+
+Needs a single/global proportionality constant.
 $$
@@
 
-Using these constraints, the gravity model @eq-phys-gravity-model is typically distinguished in four cases, cases that are mathematically and conceptually consistent, so they have continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+# A family of accessibility measures
 
-- None of @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel hold -- the unconstrained case;
-- Only @eq-constraint0-gravitymodel holds -- total flow constrained
-- @eq-constraint1-gravitymodel holds -- the production constrained case (origin-constrained, a singly-constrained case); 
-- @eq-constraint2-gravitymodel holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
-- @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel both hold simultaneously -- the production-attraction constrained case (doubly-constrained).
-
-In each of these four cases, the single proportionality constant $k$ is replaced by two sets of balancing factors $A_i$ and $B_j$ to ensure the case-specific constraints are satisfied.
-For the doubly-constrained case:
-$$
-T_{ij} = A_i O_i B_j D_j f(c_{ij})
-$${#eq-doublyconstrained-gravitymodel}
-
-And the balancing factors equal:
-$$
-A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}
-$${#eq-bf-A-gravitymodel}
-
-$$
-B_j = \frac{1}{\sum_j A_i O_j f(c_{ij})}
-$${#eq-bf-B-gravitymodel}
-
-For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel.
-In a way, the singly-constrained and doubly-constrained cases are variants of the unconstrained case.
-
-## Entropy maximising: the gravity-model as statistical averages
-Instead of the Newtonian gravity analogy, entropy maximisation is a popular model building approach taken in the spatial interaction studies to generate the gravity model.
-Namely, it is mathetmatically regorouse and offers a sensible interpretation of the resulting mode as a statistical averaging procedure.
-This allows the measure to be effective in describing behaviour in large populations but less effective in smaller decision making units in the region of interest.
-
-The basis of the approach is to accept that there are macro, meso, and micro states -- where all micro states that are consistent with our information about macro states are equally likely to occur.
-For the _interaction_ variable of trips $T_{ij}$, as shown in @wilson1971, the number of micro states $W\{{T_{ij}}\}$ within a meso state $T_{ij}$ can be expressed as:
-$$ 
-W\{{T_{ij}}\} = \frac{T!}{\prod_{ij}T_{ij}!}
-$${#eq-micro-meso-states}
-<!--I LEFT OUT THE DERIVATION... NOT SURE IF ITS NECESSARY? -->
-Using the maximizing entropy approach, we must find a set of $T_{ij}$ that maximizes @eq-micro-meso-states subject to our sensible macro-state constraints @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and the following on total cost $C$ constraint:
-$$
-\sum_{ij}T_{ij}c_{ij} = C
-$${#eq-constraint3-gravitymodel}
-
-The natural logarithm of @eq-micro-meso-states is maximized under @eq-constraint1-gravitymodel, @eq-constraint1-gravitymodel and @eq-constraint3-gravitymodel for mathematical convenience, as both $ln W$ and $T_{ij}$ share the same maximum [@wilson1971; @ortuzar_2011_modelling].
-The result is:
-
-$$
-T_{ij} = A_i O_i B_j D_j exp(-\beta c_{ij})
-$${#eq-maxentropy-gravitymodel}
-
-Where,
-$$
-A_i = \frac{1}{\sum_j B_j D_j  \exp(- \beta c_{ij})}
-$${#eq-bf-maxentropy-A-gravitymodel}
-
-and
-$$
-B_j = \frac{1}{\sum_j A_i O_j  \exp(- \beta c_{ij})}
-$${#eq-bf-maxentropy-B-gravitymodel}
-
-Results of maximisation are identical to @eq-doublyconstrained-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-B-gravitymodel except the general $f(c_{ij})$ is $exp(-\beta c_{ij})$.
-From this approach, the gravity model can be interpretted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
-There is then a most probable distribution of trips between zones (meso-states) that stasify all the macro-state constraints.
-
-# A family of accessibility measures
+Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities. We don't have to change much!
 

### 2024-11-11 data-raw/00-bibliometrics-prepare-data.qmd
Added materials for bibliographical analysis

---
title: "Bibliometrics"
format: html
---

## Motivation

In the paper we make the argument that research on accessibility diverged from research on spatial interaction modelling, especially after Wilson's landmark paper "A family of spatial interaction models", and that this is one of the reasons was the notion of proportionality constants that vanished between Stewart and Hansen was never fully recovered, despite being key to Wilson's approach to spatial interaction.

Out of curiosity, it would be interesting to see how often Wilson is cited in the literature on accessibility, and how much overlap there is with citations to Hansen.

To this end, we completed a search using the Web of Science Core Collection. We used the "CITED REFERENCES" function, and searched by DOI, as follows:

- Hansen WG (1959) How Accessibility Shapes Land Use. doi:10.1080/01944365908978307 
- Wilson AG (1971) A Family of Spatial Interaction Models, and Associated Developments. doi:10.1068/a030001

![](00-WoS-search.png)

The search was conducted on November 9, 2024.

This search retrieved 2,134 documents that cite Hansen's paper and 368 documents that cite Wilson's paper.

The documents that cite these two papers were downloaded (in the case of Hanson in batches, because of a limit of 500 documents per export). The records were exported EndNote Desktop format, and the Record Contents were "Full Record and Cited References".

The resulting files were saved to folder `data-raw/`:

- hansen_01.ciw (records 1-500)
- hansen_01.ciw (records 501-1000)
- hansen_01.ciw (records 1001-1500)
- hansen_01.ciw (records 1501-2000)
- hansen_01.ciw (records 2001-2134)
- wilson.ciw (records 1-368)

## Preliminaries

For this piece of analysis, we will use package [{bibliometrix}](https://www.bibliometrix.org/home/)

```{r}
#| label: load packages

library(bibliometrix) # Comprehensive Science Mapping Analysis
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
library(stringr) # Simple, Consistent Wrappers for Common String Operations
```

## Read the bibliometric data

Read the bibliography that cites Hanson and collect in a single data frame:
```{r}
hansen_01 <- convert2df(file = glue::glue(here::here(),
                                          "/data-raw/hansen_01.ciw"),
                        dbsource = "wos")

hansen_02 <- convert2df(file = glue::glue(here::here(),
                                          "/data-raw/hansen_02.ciw"),
                        dbsource = "wos")

hansen_03 <- convert2df(file = glue::glue(here::here(),
                                          "/data-raw/hansen_03.ciw"),
                        dbsource = "wos")

hansen_04 <- convert2df(file = glue::glue(here::here(),
                                          "/data-raw/hansen_04.ciw"),
                        dbsource = "wos")

hansen_05 <- convert2df(file = glue::glue(here::here(),
                                          "/data-raw/hansen_05.ciw"),
                        dbsource = "wos")

hansen <- bind_rows(hansen_01,
                    hansen_02,
                    hansen_03,
                    hansen_04,
                    hansen_05)
```

Read the bibliography that cites Wilson:
```{r}
wilson <- convert2df(file = glue::glue(here::here(),
                                       "/data-raw/wilson.ciw"),
                     dbsource = "wos")
```

Bibliometric analysis of the two sets of references:
```{r}
hansen_results <- biblioAnalysis(hansen)
wilson_results <- biblioAnalysis(wilson)
```

Retrieve the documents cited in the two sets of references:
```{r}
hansen_citations <- citations(hansen, 
                              field = "article", 
                              sep = ";")

wilson_citations <- citations(wilson, 
                              field = "article", 
                              sep = ";")
```

As expected, the most cited paper in the `hansen` corpus is Hansen, since the corpus was selected by this criterion. At most, the number of citations to another article could have been a tie, but it is not close:
```{r}
cbind(hansen_citations$Cited[1:10])
```

The next most cited paper after Hansen is Geurs (2004) with 680 citations.

Again, as expected, the most cited paper in the `wilson` corpus is Wilson, since the corpus was selected by this criterion. At most, the number of citations to another article could have been a tie, but it is not close:
```{r}
cbind(wilson_citations$Cited[1:10])
```

The next most cited paper is Hansen, with 76 citations.

Add a column to the data frames to uniquely identify the documents in each corpus:
```{r}
hansen <- hansen |>
  mutate(HUID = 1:n(),
         HUID = paste0("HUID", HUID))

wilson <- wilson |>
  mutate(WUID = 1:n(),
         WUID = paste0("WUID", WUID))
```

Identify the overlap between the two corpora; these are all papers that cite both Hansen and Wilson:
```{r}
both <- hansen |>
  transmute(AU,
            DI, 
            HUID) |>
  inner_join(wilson |>
               transmute(AU,
                         DI, 
                         WUID),
             by = c("AU", "DI"))
```

Obtain the full corpora of documents that cite both Hansen and Wilson:
```{r}
both_temp <- hansen |>
  filter(HUID %in% both$HUID) |>
  mutate(corpus = "Both")
```

Remove the papers that cite both documents from the Hansen and Wilson corpora:
```{r}
hansen <- hansen |>
  filter(!(HUID %in% both$HUID))

wilson <- wilson |>
  filter(!(WUID %in% both$WUID))
```

Find duplicates:
```{r}
hansen_items <- data.frame(rn = rownames(hansen))
wilson_items <- data.frame(rn = rownames(wilson))
both_items <- data.frame(rn = rownames(both))

inner_join(hansen_items, wilson_items, by = "rn")
inner_join(hansen_items, both_items, by = "rn")
inner_join(wilson_items, both_items, by = "rn")
```

Merlin 2020 is duplicated in the `wilson` and `hansen` tables. Remove it from `hansen`:
```{r}
hansen <- hansen |>
  filter(DI != "10.5198/jtlu.2020.1626")
```

Add a column to identify the corpus and bind the data frames (and drop the unique identifiers now that they accomplished their purpose):
```{r}
corpus <- bind_rows(hansen |>
                      mutate(corpus = "Hansen") |>
                      select(-HUID) |>
                      select(corpus, everything()),
                    wilson |>
                      mutate(corpus = "Wilson") |>
                      select(-WUID) |>
                      select(corpus, everything()),
                    both_temp |>
                      mutate(corpus = "Both") |>
                      select(-c(HUID)) |>
                      select(corpus, everything()))
```

Save corpus; `coupling_net` is pretty heavy and it can be recreated from the corpus:
```{r}
save(corpus,
     file = glue::glue(here::here(),
                       "/data/corpus.rda"),
     compress = "xz")
```

### 2024-11-11 data-raw/01-bibliometrics-analysis.qmd
Added materials for bibliographical analysis

---
title: "Bibliometrics"
format: html
---

## Preliminaries

The bibliographic data were pre-processed in notebook `data-raw/00-bibliometrics-prepare-data.qmd`. The output is a corpus with papers that cite Hansen, Wilson, or both.

Also, a bibliometric coupling network was created that contains information about the

For this piece of analysis, we will use package [{bibliometrix}](https://www.bibliometrix.org/home/)

```{r}
#| label: load packages

library(bibliometrix) # Comprehensive Science Mapping Analysis
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
library(glue) # Interpreted String Literals
library(here) # A Simpler Way to Find Your Files
```

## Load data

Load data:
```{r}
load(glue::glue(here::here(),
                "/data/corpus.rda"))
```

Bibliographical coupling based on cited sources (see https://www.bibliometrix.org/vignettes/Introduction_to_bibliometrix.html). The bibliographical coupling is stronger when documents cite more of the same sources.

To compute the bibliographical coupling, we need an "adjacency" matrix of the documents and their cited references. This can be obtained as the bipartite network of paper and citations (`CR` is the field with cited references in the corpus):
```{r}
A <- cocMatrix(corpus, 
               Field = "CR", 
               sep = ";")
```

The output of `cocMatrix` is a sparse adjacency matrix with sources cited by each document in the corpus. The product of $A$ and its transpose gives the bibliographical coupling by cited references:
$$
B = A\cdot A^T
$$

$B$ is a square symmetric matrix where element $b_{ij}$ is the total number of cited references in common between document $i$ and document $j$. A higher value of $b_{ij}$ indicates a stronger relationship between documents $i$ and $j$ (more references in common). A value of zero indicates no references in common. Calculate $B$:
```{r}
B <- tcrossprod(A)
```

Remove the diagonal (papers cited within a single source):
```{r}
B_nd <- B - Matrix::band(B, 0, 0)
```

Examine this matrix without the diagonal:
```{r}
B_nd[1:10, 1:10]
```
As desired, the diagonal is zeros.

These matrices are sparse symmetric matrices of "adjacencies" between documents in the corpus. Convert `B` to an `igraph` network without the diagonal:
```{r}
coupling_net <- igraph::graph_from_adjacency_matrix(B, 
                                            mode = "directed", 
                                            weighted = TRUE,
                                            diag = FALSE)
```

Remove objects from memory:
```{r}
rm(A)
```

For convenience, convert the [{igraph}](https://r.igraph.org) object to a [{tidygraph}](https://www.data-imaginist.com/posts/2017-07-07-introducing-tidygraph/) object:
```{r}
coupling_net <- tidygraph::as_tbl_graph(coupling_net)
```

Check the result:
```{r}
coupling_net
```

Add the `corpus` variable to `coupling_net`. Remember to activate the nodes.
```{r}
coupling_net <- coupling_net |>
  tidygraph::activate("nodes") |>
  left_join(tibble::rownames_to_column(corpus, 
                                       var = "rowname") |> 
              transmute(rowname, 
                        corpus = corpus),
            by = c("name" = "rowname"))
```

Check the object:
```{r}
coupling_net |>
  tidygraph::activate("nodes") |>
  as_tibble() |>
  mutate(corpus = factor(corpus)) |>
  summary()
```

Detect communities based on the bibliographic coupling. Set the random seed because depending on the seed, the algorithm fails to detect the communities:
```{r}
set.seed(979834789)
#set.seed(97389)
#set.seed(973899887)

coupling_net <- coupling_net |>
  mutate(community = as.factor(tidygraph::group_infomap(weights = weight)))
```

Check the results:
```{r}
coupling_net |> 
  tidygraph::activate("nodes") |> 
  as_tibble() |> 
  summary()
```

Create a layout for plotting the results:
```{r}
layout <- create_layout(coupling_net,
                        layout = "stress",
                        weights = weight)
```

Plot:
```{r}
ggplot(data = layout) +
  geom_point(aes(x = x,
                 y = y,
                 shape = corpus,
                 color = corpus),
             size = 3,
             alpha = 0.7) +
  geom_point(data = layout |>
               filter(corpus == "Both"),
             aes(x = x,
                 y = y),
             shape = 1,
             size = 3) +
  geom_point(data = layout |>
               filter(corpus == "Wilson"),
             aes(x = x,
                 y = y),
             shape = 0,
             size = 3,
             alpha = 0.5) +

  theme_void()
```

Center of mass of Hansen:
```{r}
layout |>
  filter(corpus == "Hansen") |>
  summarize(x = mean(x),
            y = mean(y))
```

Center of mass of Wilson:
```{r}
layout |>
  filter(corpus == "Wilson") |>
  summarize(x = mean(x),
            y = mean(y))
```

Matrix `B` has the coupling.

Extract the row names by corpus:
```{r}
hansen_names <- rownames(corpus |>
                           filter(corpus == "Hansen"))

wilson_names <- rownames(corpus |>
                           filter(corpus == "Wilson"))

both_names <- rownames(corpus |>
                           filter(corpus == "Both"))
```

Subset `B` to look at the patterns of coupling. These give the coupling within the group of documents that cite only Hanson or only Wilson, and the coupling between the papers that cite only Hanson and only Wilson:
```{r}
hansen_to_wilson <- B_nd[hansen_names, wilson_names]

hansen_to_hansen <- B_nd[hansen_names, hansen_names]

wilson_to_wilson <- B_nd[wilson_names, wilson_names]

both_to_both <- B_nd[both_names, both_names]

both_to_hansen <- B_nd[both_names, hansen_names]

both_to_wilson <- B_nd[both_names, wilson_names]
```

The number of common citations per pair of documents can be calculated as the sum of all common citations in the matrix, divided by the number of pair-paper combinations:
```{r}
sum(sum(B_nd))/(dim(B_nd)[1] * (dim(B_nd)[2]))
sum(sum(both_to_both))/(dim(both_to_both)[1] * (dim(both_to_both)[2]))
sum(sum(hansen_to_hansen))/(dim(hansen_to_hansen)[1] * (dim(hansen_to_hansen)[2]))
sum(sum(wilson_to_wilson))/(dim(wilson_to_wilson)[1] * (dim(wilson_to_wilson)[2]))
sum(sum(both_to_hansen))/(dim(both_to_hansen)[1] * dim(both_to_hansen)[2])
sum(sum(both_to_wilson))/(dim(both_to_wilson)[1] * dim(both_to_wilson)[2])
sum(sum(hansen_to_wilson))/(dim(hansen_to_wilson)[1] * dim(hansen_to_wilson)[2])
```

From this we see that the coupling within the whole corpus is on average 1.18 (the number of sources that are typically cited in common by pair of documents). The coupling tends to be stronger for documents in the corpus that cite both Hansen and Wilson. The coupling is weakest between documents that cite only Hansen and only Wilson.

### 2024-11-11 data-raw/01-bibliometrics-analysis.qmd
Pushing latest changes to bibliometric analysis

@@
 library(here) # A Simpler Way to Find Your Files
+library(tidytext)
 ```
@@
 The coupling is weakest between documents that cite only Hansen and only Wilson.
+
+Who are the authors in that niche that cites both Wilson and Hansen, and what is the temporal pattern of publication?
+```{r}
+corpus |> 
+  filter(corpus == "Both") |> 
+  select(AU, PY) |> arrange(PY)
+```
+
+<!--
+```{r}
+CS <- conceptualStructure(corpus,
+                          field="DE",
+                          method="MCA",
+                          minDegree=4, 
+                          clust=5, 
+                          stemming = TRUE,
+                          labelsize = 10,
+                          documents = 10)
+```
+
+-->
+
+```{r}
+corpus |> tibble::rownames_to_column(var = "rownames") |> select(rownames, AB) |> unnest_tokens(word, input = "AB")
+```

### 2024-11-12 paper/manuscript.qmd
finished Tij calculations using simple example

@@
 library(dplyr) # A Grammar of Data Manipulation
-library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
-library(geomtextpath) # Curved Text in 'ggplot2'
-library(ggforce) # Accelerating 'ggplot2'
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
-library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
-library(ggspatial) # Spatial Data Framework for ggplot2
-library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
-library(here)
-
+library(flextable)
 # library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
@@
 # library(tmap) # Thematic Maps
+# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
+# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
+# library(ggspatial) # Spatial Data Framework for ggplot2
+# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
+# library(here)
 ```
@@
 
-The following are simple numerical examples of each case assuming the following $T_{ij}$ matrix and a $f(c_{ij}) = \frac{1}{c_{ij}}$.
+The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.
+For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below:
+```{r}
+known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
+#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
+W_i <- known_trip_matrix |> rowSums()
+W_j <- known_trip_matrix |> colSums()
+T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
+knowns_longer <- cbind(known_trip_matrix,W_i)
+knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))
 
-The variables:
-```{r}
-T_1 <- matrix(c(95, 23, 42, 27, 378, 45, 38, 36, 106), ncol = 3, byrow = TRUE)
+rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
+colnames(knowns_longer) <- c("1","2","3", "TOTAL")
 
-T_1
+knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
 ```
 
+And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
 ```{r}
 C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
-
-C
+C_longer <-C
+rownames(C_longer) <- c("1","2","3") 
+colnames(C_longer) <- c("1","2","3") 
+C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
 ```
 
+### Unconstrained interaction case
+
+$k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$
+
+$$
+T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
 ```{r}
-O_i <- rowSums(T_1)
-D_j <- colSums(T_1)
+T_unc <- W_i %*% t(W_j) / C
 
-O_i
-D_j
+W_i_unc <- T_unc |> rowSums()
+W_j_unc <- T_unc |> colSums()
+T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA)
+T_unc_longer <- cbind(T_unc,W_i_unc)
+T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc))
+
+rownames(T_unc_longer) <- c("1","2","3", "TOTAL") 
+colnames(T_unc_longer) <- c("1","2","3", "TOTAL")
+
+T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
 ```
 
-```{r}
-T <- sum(sum(T_1))
+As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends.
+Instead, they take on the assumptions of all balancing factors $k$ being equal to 1.
+When would this happen?
+It's hard to say...
+as the balancing factors themselves do not have practical meaning; what has meaning are the constraints.
+By satisfying the constraints, the model gains meaning.
+In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$.
+In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.
 
-T
-```
+### Total constrained interaction case
 
-## Unconstrained interaction case
-
+In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows:
 $$
-\hat{T_{ij}} = \frac{O_i\cdot D_j}{C_ij}
+T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+\where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made:
+$$
+K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\
+K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +... \frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}}
+}\\
+K = \frac{790}{166443} \\
+K = 0.00474636
 $$
@@
 ```{r}
-T_unc <- O_i %*% t(D_j) / C
+K <- T/sum(sum(T_unc))
+T_tot <- K * T_unc
 
-T_unc
+W_i_tot <- T_tot |> rowSums()
+W_j_tot <- T_tot |> colSums()
+T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA)
+T_tot_longer <- cbind(T_tot,W_i_tot)
+T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot))
+
+rownames(T_tot_longer) <- c("1","2","3", "TOTAL") 
+colnames(T_tot_longer) <- c("1","2","3", "TOTAL")
+
+T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
 ```
+While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model.
 
-If we do the sums over rows, columns and all:
-```{r}
-rowSums(T_unc)
-colSums(T_unc)
-sum(sum(T_unc))
-```
+### Singly-constrained
 
-## Total constrained interaction case
+Two versions of this case, either origin-constrained or destination-constrained.
 
-Needs a single/global proportionality constant.
+For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
 $$
-\hat{T_{ij}} = k\frac{O_i\cdot D_j}{C_{ij}}
+T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+\where In our example, $A_i$ is:
+$$
+A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\
+A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\
+...\\
+A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\
+A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\
+A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201 
 $$
@@
 ```{r}
-k <- T/sum(sum(T_unc))
-k
+A_i <- 1/t((matrix(W_j))) %*% (1/C)
+#A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns
+A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3)
+
+T_orig <-  A_i%*%T_unc
+
+W_i_orig <- T_orig |> rowSums()
+W_j_orig <- T_orig |> colSums()
+T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA)
+T_orig_longer <- cbind(T_orig,W_i_orig)
+T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig))
+
+rownames(T_orig_longer) <- c("1","2","3", "TOTAL") 
+colnames(T_orig_longer) <- c("1","2","3", "TOTAL")
+
+T_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
 ```
 
+We can see that the $W_i^{(1)}$ values are equal to the trip ends, but not $W_j^{(2)}$ values.
+
+Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel):
+$$
+T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+\where $B_j$ is, and equal to:
+$$
+B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
+$$
 ```{r}
-T_tot <- k * T_unc
+B_j <- 1/(t(matrix(W_i))) %*% (1/C)
 
-T_tot
+B_j <- matrix(B_j) |> t()
+B_j_longer <- B_j
+colnames(B_j_longer) <- c("1","2","3")
+B_j_longer  #|> as.data.frame()  |> gt()
+#B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns
 ```
 
-Let's check those marginals and total:
+And trips are:
 ```{r}
-rowSums(T_tot)
-colSums(T_tot)
-sum(sum(T_tot))
+B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3)
+
+T_dest <-  (B_j)%*%t(T_unc) |> t()
+
+W_i_dest <- T_dest |> rowSums()
+W_j_dest <- T_dest |> colSums()
+T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA)
+T_dest_longer <- cbind(T_dest,W_i_dest)
+T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest))
+
+rownames(T_dest_longer) <- c("1","2","3", "TOTAL") 
+colnames(T_dest_longer) <- c("1","2","3", "TOTAL")
+
+T_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
 ```
 
-Destination constrained:
+And as we can observe, the $W_j^{(2)}$ values are equal to the known trip ends, while $W_i^{(1)}$ is not.
+
+### Doubly-constrained case
+
+The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
 $$
-\hat{T_{ij}} = \frac{B_j\cdot O_i\cdot D_j}{C_{ij}}
+T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+\where In our example, $A_i$ and $B_j$ is:
+$$
+A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\
+B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})}
 $$
 ```{r}
-B_j <- colSums(T_1)/colSums(T_unc)
+doubly_constrained_gravity_model <- function(
+    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
+  O,  # Origin production weights vector
+  D,  # Destination attraction weights vector
+  impedance_matrix,  # the resulting travel cost impedance matrix
+  error_threshold = 0.001,  # Error threshold for stopping condition
+  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
+) {
 
-B_j <- matrix(rep(B_j, 3), ncol = 3, byrow = TRUE)
+  # Normalize O and D
+  sum_O <- sum(O)
+  sum_D <- sum(D)
+  cat("Checking production, attraction balancing:\n")
+  cat("Production: ", sum_O, "\n")
+  cat("Attraction: ", sum_D, "\n")
 
-B_j
+  if (sum_O != sum_D) {
+    cat("Productions and attractions do not balance, scaling to match.\n")
+    if (sum_O < sum_D) {
+      O <- O * (sum_D / sum_O)
+    } else {
+      D <- D * (sum_O / sum_D)
+    }
+  } else {
+    cat("Production, attraction balancing OK.\n")
+  }
+
+  n <- length(O)  # Number of i
+  T <- sum(O)  # Total trips
+  Ai <- rep(1, n)  # Initialize Ai
+  Bj <- rep(1, n)  # Initialize Bj
+
+  previous_error <- Inf
+  iteration_count <- 0
+  stop_reason <- ""
+
+  while (TRUE) {
+    iteration_count <- iteration_count + 1
+
+    # Update Ai
+    for (i in 1:n) {
+      Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9)
+    }
+
+    # Update Bj
+    Bj_new <- rep(1, n)
+    for (j in 1:n) {
+      Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9)
+    }
+
+    # Calculate Tij
+    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
+
+    # Print Ai and Bj for this iteration
+    cat("\nIteration:", iteration_count, "\n")
+    cat("Ai:", round(Ai, 3), "\n")
+    cat("Bj:", round(Bj_new, 3), "\n")
+
+    # Calculate error
+    error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T
+
+    # Calculate change in error
+    error_change <- abs(previous_error - error)
+
+    # Check stopping conditions
+    if (error < error_threshold) {
+      stop_reason <- "Error threshold met"
+      break
+    } else if (error_change < improvement_threshold) {
+      stop_reason <- "Slow improvement"
+      break
+    }
+
+    previous_error <- error
+    Bj <- Bj_new
+  }
+
+  # report stopping condition and final error threshold
+  cat("\nStopping Condition: ", stop_reason, "\n")
+  cat(sprintf("Final Error: %.3f%%\n", error * 100))
+
+  return(list(Ai = Ai, Bj = Bj, Tij = Tij))
+}
+
+dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C)
 ```
 
+The $A_i$ and $B_j$ matrices:
 ```{r}
-T_dest <- B_j * T_unc
+A_i_doubly <- dc_results$Ai
+B_j_doubly <- dc_results$Bj
 
-T_dest
+A_i_doubly |> matrix()
+B_j_doubly |> matrix() |> t()
 ```
 
-Let's check those marginals and total:
+The trip matrix (notice, all trip ends are equivalent to the known trip ends)
 ```{r}
-rowSums(T_dest)
-colSums(T_dest)
-sum(sum(T_dest))
+T_doubly <- dc_results$Tij
+
+W_i_doubly <- T_doubly |> rowSums()
+W_j_doubly <- T_doubly |> colSums()
+T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA)
+T_doubly_longer <- cbind(T_doubly,W_i_dest)
+T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest))
+
+rownames(T_doubly_longer) <- c("1","2","3", "TOTAL") 
+colnames(T_doubly_longer) <- c("1","2","3", "TOTAL")
+
+T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
 ```
 
-Origin constrained:
-$$
-\hat{T_{ij}} = \frac{A_i\cdot O_i\cdot D_j}{C_{ij}}
-$$
+### Mean absolute percentage error (MAPE) between all cases
 
 ```{r}
-A_i <- rowSums(T_1)/rowSums(T_unc)
+#unconstrained
+((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
 
-A_i <- matrix(rep(A_i, 3), ncol = 3, byrow = FALSE)
+#total constrained
+((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
 
-A_i
-```
+#singly-constrained (orig)
+((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
 
-```{r}
-T_orig <- A_i * T_unc
+#singly-constrained (dest)
+((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
 
-T_orig
-```
-
-Let's check those marginals and total:
-```{r}
-rowSums(T_orig)
-colSums(T_orig)
-sum(sum(T_orig))
-```
-
-Now how does this translate to accessibility?
-
-Opportunities (say, grocery stores):
-```{r}
-D_j <- c(7, 3, 2)
-```
-
-Unconstrained accessibility (no proportionality constant):
-$$
-S_i = \sum_j\frac{D_j}{C_{ij}}
-$$
-
-```{r}
-S_i <- (1/C) %*% D_j
-```
-
-These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
-```{r}
-sum(S_i)
-```
-
-Total flow constrained accessibility, needs a balancing factor:
-```{r}
-k <- sum(D_j)/sum(S_i)
-
-k
-```
-
-Total flow constrained accessibility:
-Unconstrained accessibility (no proportionality constant):
-$$
-S_i = k\sum_j\frac{D_j}{C_{ij}}
-$$
-
-```{r}
-S_tot <- k * S_i
-
-S_tot
-```
-
-Now check the sum:
-```{r}
-sum(S_tot)
+#doubly-constrained (dest)
+((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
 ```
@@
 
-Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities. We don't have to change much!
-
+Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities.
+We don't have to change much!
 Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities.
@@
 
-Using the maximizing entropy approach... <!--DESCRIBING WHAT EACH MICRO/MESO/MACRO STATE IS HERE IN RELATION TO ACCESSIBILITY--> For each case...
+Now how does this translate to accessibility?
 
-## Unconstrained case
+Opportunities (say, grocery stores):
+```{r}
+D_j <- c(7, 3, 2)
+```
+
+## Unconstrained accessibility 
+(no proportionality constant):
+$$
+S_i = \sum_j\frac{D_j}{C_{ij}}
+$$
+
+```{r}
+S_i <- (1/C) %*% D_j
+```
+
+These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
+```{r}
+sum(S_i)
+```
+
+##Total flow constrained accessibility, needs a balancing factor:
+```{r}
+k <- sum(D_j)/sum(S_i)
+
+k
+```
+
+## Unconstrained accessibility (no proportionality constant):
+$$
+S_i = k\sum_j\frac{D_j}{C_{ij}}
+$$
+
+```{r}
+S_tot <- k * S_i
+
+S_tot
+```
+
+Now check the sum:
+```{r}
+sum(S_tot)
+```
 

### 2024-11-12 paper/manuscript.qmd
draft of accessibility formulation

@@
 
-## Entropy maximising: the gravity-model as statistical averages
-
+**Entropy maximising: the gravity-model as statistical averages:**
 Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
@@
 
-### Unconstrained interaction case
+## Unconstrained interaction case
 
@@
 
-### Total constrained interaction case
+## Total constrained interaction case
 
@@
 
-### Singly-constrained
+## Singly-constrained
 
@@
 
-### Doubly-constrained case
+## Doubly-constrained case
 
@@
 
-### Mean absolute percentage error (MAPE) between all cases
+## Mean absolute percentage error (MAPE) between all cases
 
@@
 
+#unconstrained -- rescaled!
+W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums()
+W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums()
+
+W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums()
+W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums()
+
+((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01)
+```
+
+```{r}
 #total constrained
@@
 
+- doubly c is lowest error
+- singly constrained error is mid
+- total constrained has high error
+- notably, unconstrained unscaled is obviously much higher in error - there are no constraints!
+But if you scale the matrices, the error is both low (interesting!!
+why is this?).
+However as mentioned, the unconstrained case is hard to interpret literally -- in what situations would k=1..?
+
 # A family of accessibility measures
@@
 $$
-V_{ij} = k O_i D_j f(c_{ij})
+V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
 $${#eq-access}
 
-Where $V_{ij}$ represents a $n$ x $m$ matrix of potential reachable opportunities between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $O_i$ and $D_j$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
-In other words, $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
-Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
+Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
 
-Like with @eq-phys-gravity-model, we distinguish @eq-access into FIVE cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel).
-To inform each case of @eq-access with constraints that reflect knowledge of the macro-state, the total trip flow $O_i$ originating at each $i$ and the total trip flows $D_j$ terminating at $j$ are used.
-These pieces of information can constrain $V_{ij}$ as follows, ensuring it reflects this macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
+Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
 
 $$
-\sum_j V_{ij} = O_i
-$${#eq-constraint1-access}
-\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the opportunities originating $O_i$ at each $i$ in the region.
+\sum_i\sum_j V_{ij} = V
+$${#eq-constraint0-access}
 
 $$
-\sum_i V_{ij} = D_j
-$${#eq-constraint2-access}
-\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the opportunities ending $D_i$ at each $j$ in the region.
+\sum_i\sum_j V_{ij} c_{ij} = C
+$${#eq-constraint0_5-access}
+\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.
 
 $$
-\sum_{i}\sum_{j} V_{ij} c_{ij} = C
-$${#eq-constraint3-access}
-\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.
+\sum_j V_{ij} = W_i^{(1)}
+$${#eq-constraint1-access}
+\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.
 
-For all FIVE cases, @eq-constraint3-access holds <!--in the event C is known, it could be used to identify parameters in f(c_ij)...
-though model calibration is probably a better idea.
---> but are differentiated like:
-- neither @eq-constraint1-access or @eq-constraint2-access hold -- the unconstrained case;
-- @eq-constraint1-access holds -- the production constrained case (origin-constrained, a singly-constrained case); 
-- @eq-constraint2-access holds -- the attraction constrained case (destination-constrained, a singly-constrained case); 
-- @eq-constraint1-access and @eq-constraint2-access both hold simultaneously -- the production-attraction constrained case (doubly-constrained).
-- @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case... and doubly constrained case)-->
+$$
+\sum_i V_{ij} = W_j^{(2)}
+$${#eq-constraint2-access}
+\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.
 
-Now how does this translate to accessibility?
+All four <!--five?--> cases are differentiated like:
+- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
+- The total constrained case: only @eq-constraint0-gravitymodel hold
+- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
+- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
+- HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case...
+and doubly constrained case)-->
 
-Opportunities (say, grocery stores):
+We can use the same trip matrix, but reformulate its meaning.
+Let's say $W_i^{(1)}$ is population at $i$ who seek opportunities, lets recast this variable as $P_i$.
+Then let's say $W_j^{(2)}$ are the opportunities being sought, recast to $O_j$.
+For this simple example, we can say that the opportunities are the population's jobs.
+In our region, each person has a job -- but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost ($C$ and associated travel propensity $1/C$).
+(let's pretend our 'known_trip_matrix' is unknown).
+
 ```{r}
-D_j <- c(7, 3, 2)
+known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
+#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
+W_i <- known_trip_matrix |> rowSums()
+W_j <- known_trip_matrix |> colSums()
+T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
+knowns_longer <- cbind(known_trip_matrix,W_i)
+knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))
+
+rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
+colnames(knowns_longer) <- c("1","2","3", "TOTAL")
+
+knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
+```
+
+And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
+```{r}
+C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
+C_longer <-C
+rownames(C_longer) <- c("1","2","3") 
+colnames(C_longer) <- c("1","2","3") 
+C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
 ```
@@
 ## Unconstrained accessibility 
-(no proportionality constant):
+(no proportionality constant, k=1, no constraints hold - hard to interpret):
 $$
-S_i = \sum_j\frac{D_j}{C_{ij}}
+S_{ij} = P_i O_j f(c_{ij})
+$$
+In practice, $S_i$ is calculated for each $i$ so:
+$$
+S_{i} = P_i \sum_j O_j f(c_{ij})
+$$
+These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).
+
+Looks like Hansen, but multiplied by P_i.
+
+## Total flow constrained accessibility
+
+This needs a balancing factor, $k$ is replaced with $K$:
+$$
+S_{ij}' = K P_i O_j f(c_{ij})\\
+S_{i}' = K P_i \sum_j O_j f(c_{ij})
 $$
 
-```{r}
-S_i <- (1/C) %*% D_j
-```
+This approach is not seen in accessibility literature...
 
-These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number of stores. Check the sum:
-```{r}
-sum(S_i)
-```
+## Singly-constrained case
 
-##Total flow constrained accessibility, needs a balancing factor:
-```{r}
-k <- sum(D_j)/sum(S_i)
+Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
+For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
+$$
+V_{ij}''^{(1)} = A_i P_i O_j f(c_{ij})\\
+V_{ij}''^{(2)} = B_j P_i O_j f(c_{ij})\\
 
-k
-```
+V_{i}''^{(1)} = A_i P_i \sum_j O_j f(c_{ij})\\
+V_{i}''^{(2)} = P_i \sum_j B_j O_j f(c_{ij})\\
 
-## Unconstrained accessibility (no proportionality constant):
+$${#eq-access-singlyconstrained-origin}
+
+\noindent Where $A_i$ and $B_j$ are:
 $$
-S_i = k\sum_j\frac{D_j}{C_{ij}}
+A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
+B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
 $$
+For the origin-constrained, $V_{i}''^{(1)}$ are the number of opportunities (like jobs) that can be reached at $i$, and $V_{i}''^{(2)}$ are the number of opportunities (like people) that can be reached at $i$
 
-```{r}
-S_tot <- k * S_i
+In terms of origin-constrained -- spatial availability is explicitly so [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024].
+Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general [].
 
-S_tot
-```
+Regarding destination-constrained, has it's origins are in retail models.
+e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction.
+Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
 
-Now check the sum:
-```{r}
-sum(S_tot)
-```
-
-Neither @eq-constraint1-access or @eq-constraint2-access hold, so $V_ij'$ has no opportunity constraints!
-
-$$
-V_{ij}' = k O_i D_j f(c_{ij})
-$${#eq-access-unconstrained}
-
-Where $k$ is two sets of balancing factors, $A_i'$ and $B_j'$, and both are equal to 1.
-
-## Singly-constrained case (origin-constrained)
-
-Spatial availability; origin-constrained.
-
-Origins in human spatial interaction: "intervening opportunities and competing migrants" @stoufferInterveningOpportunitiesTheory1940 and @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960, "axiomatically"  @weibull_axiomatic_1976, "for multiple-modes, as a rate" @shen1998; for multiple-modes, as the number of opportunities" @soukhovMultimodalSpatialAvailability2024
-
-$$
-V_{ij}'' = k O_i D_j f(c_{ij})
-$${#eq-access-singlyconstrained}
-
-<!--HOW DO WE FIT OUR EXISTING CONSTRAINTS INTO THESE BALANCING FACTORS..? $k O_i f(c_{ij}) = the balancing factors..?-->
+For instance, the most recent example spatial availability is formulated like so:
 $$
@@
 $$
-Population balancing factor:
+Where the $F^p_{ij}$ Population balancing factor is:
 $$
@@
 $$
-Travel cost balancing factor:
+And the $F^c_{ij}$ travel cost balancing factor:
 $$
@@
 $$
-## Singly-constrained case (destination-constrained)
+It can be reformulated to fit into the singly-constrained (origin case) gravity model as the denominators in both those balancing factors cancel out:
+$$
+V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}\\
 
-Origins are in retail models, decision model for shops to optimally spatially locate to max.
-their potential consumer interaction.
-Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
-Spatial availability as another example.
+V_{ij}''^{(1)} =O_j \frac{1}{\sum_j O_j f(c_{ij})} P_i  f(c_{ij})\\
+$$
 
@@
 ## Hybrid singly-constrained case: 
-<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit??-->
+<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here-->
 Balanced floating catchment areas (not truly doubly constrained)
@@
 <!-- (cool paper that explores this @kirbyNormalizingFactorsGravity1970 - he "shows that the Ai and Bj balancing factors are the mean values of the inverse of the function e^-N*F(c_ij) calculated by summing the function over i values for Ai factors and over j values for Bj factors." -->
+
+<!--TO DISCUSS:
+-- should we use entropy maximisation to actually derive the models? I'm hazy on that.
+  -- relatedly, in the unconstrained case...
+is k=1 or does k just not exist..?
+I assume k=1 and say that the interpretation in reality is confusing.
+a consistent derivation of all cases can help address this.
+  -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->

### 2024-12-13 TRB/TRB-pres.qmd
5-slide trb pres draft

---
title: "Family of accessibility measures"
format: revealjs
---

## BIGGER QUESTIONS:
- why and/or when did accessibility diverge from spatial interaction modelling?
- examples of accessibility measures in the literature that fit into our 4 cases?

## PRESENTATION'S AGENDA:
Accessibility’s relationship can be made into one of equality -- with balancing factors! Accessibility shares spatial interaction modeling’s roots.
3.	A slide with 0) no constrained (Hansen, what we do now), 1) total constrained, 2) singly-constrained, 3) doubly constrained formulations
4.	Calculated simple example of the 3 types of accessibility. Three visualisations: 1st are the Vij matrices, 2nd is a small grid aggregation, and 3rd is bigger grid aggregation.
5.	Conclusions: 
a.	With balancing factors values are interpretable: each spatial unit is ‘potential opportunities that can reached’ (of all the opportunities)
b.	No treatment is needed to address MUAP across grid aggregations (?)

## A vanishing constant

- Gravity-based accessibility is derived from Newtonian Gravitation, it had a constant:
$F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}$

- Ravenstein, ... no one. But Stewart brought it back but assumed it was 1! With Hansen, he also assumed it was one (implicitly) and others have continued using Hansen without acknowledging.

- Gravity-based accessibility is a partial sum, no constant though. It has implications:
$S_{i} = \sum_j O_j \cdot f(d_{ij})$

## A family of gravity models

- Wilson provided a framework, showed proportionality constant is important (1. balances the units (from proportionality to equality); 2. meaning of the model is different depending on the constant(s) -- totally constrained, singly-, doubly- constrained). Both these points are important for _interpretation_, some accessibility suffers from.

$ T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$

- Wilson contributed this, but it never made it back to accessibility. Accessibility has continued to evolve independently.

## Skip over
(plot of how accessibility research interacts with gravity modelling)

## A family of accessibility measures

Borrowing the constants from gravity modelling:

Constraint 0:
$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}
\noindent Where the sum of all potentially reached opportunities in the region is equal to a total opportunities in the region $V$.

Constraint 1:
$$
\sum_j V_{ij} = W_i^{(1)} = P_i
$${#eq-constraint1-access}
\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

Constraint 2:
$$
\sum_i V_{ij} = W_j^{(2)} = O_j
$${#eq-constraint2-access}
\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

All four cases are differentiated like:
- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel are enforced (it may happen by chance);
- The total constrained case: only @eq-constraint0-gravitymodel are enforced 
  - (regional population competing for opportunities at the regional level)
- The singly-constrained case: either @eq-constraint1-gravitymodel are enforced (origin-constrained) or @eq-constraint2-gravitymodel are enforced (destination-constrained);
  - (origin-constrained: zonal population competing for opportunities at the regional level)
  - (destination-constrained: regional population competing for opportunities at the zonal level)
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel are enforced simultaneously
  -  (zonal population competition for opportunities at the zonal level)

## Why do proportality constants in accessibility matter?

All four cases are differentiated like:

- The unconstrained case (current practice - we're doing something but hard to compare proportions)
  - e.g.,: any opportunity

- The total constrained case (regional population competing for opportunities at the regional level)
  - e.g.,: any opportunity in the region for people in the region

- The singly-constrained case:
  - (origin-constrained: zonal population competing for opportunities at the regional level)
    - (e.g., any opportunity of 'doctor-spots' in the region -- for people in a fixed zone)
  - (destination-constrained: regional population competing for opportunities at the zonal level)
    - (e.g., only fixed number of 'doctor-spots' from each zone -- for anyone in the population)

- The doubly-constrained case:
  -  (zonal population competition for opportunities at the zonal level)
  - when population = opportunity (e.g., 1 person to 1 'doctor-spot'). quite rare. only fixed number of 'doctor-spots' from each zone -- for people in each zone.

- Accessibility suffers from interpretation: we think Wilson's framework solves this interpretation issue in mathematical terms:
  (1. balances the units (from proportionality to equality); 
   2. meaning of the model is different depending on the constant(s) -- totally constrained, singly-, doubly- constrained).

## Examples to include as extra slides

<!-- ## The roots of spatial interaction modelling: Newton's 1687 Law of Universal Gravitation -->
<!-- $$ -->
<!-- F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}} -->
<!-- $${#eq-phys-grav-prop}  -->
<!-- \noindent Where the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$. -->

<!-- After a century - $G$,  a proportionality constant, was obtained empirically in [@careyPrinciplesSocialScience1858]: -->
<!-- $$ -->
<!-- F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}} -->
<!-- $${#eq-phys-grav} -->

<!-- ## The roots of spatial interaction modelling: before Stewart 1948 -->
<!-- - In Ravenstein's "Laws of Migration", a directly proportional relationship of migration flows with the size of the destination is posited [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889]. Other researchers posited such a relationship. -->

<!-- Then mathematically expressed in:  -->
<!-- - @reilly1929methods as the law of retail gravitation,  the magnitude of business flows is proportional to the product of the populations of settlements, and inversely proportional to the distance between them.  -->
<!-- - Zipf also proposed a similar law of retail gravitation @reilly1929methods, then used this hypothesis ($\frac{P_1P_2}{D}$) for the case of goods movement by railways, intercity personal movement, and information flows [@zipfHypothesisCaseRailway1946; @zipfHypothesisIntercityMovement1946; @zipfDeterminantsCirculationInformation1946]. -->
<!-- - Neither Reily or Zipf included a proportionality constant, something like $G$. -->

<!-- ## The roots of spatial interaction modelling: Stewart 1948 -->

<!-- Stewart [@stewartDemographicGravitationEvidence1948] appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows: -->
<!-- $$ -->
<!-- F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2} -->
<!-- $${#eq-stewart-force} -->

<!-- \noindent where: -->

<!-- - $F$ is the _demographic force_ -->
<!-- - $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2 -->
<!-- - $\mu_1$ and $\mu_2$ are so-called _molecular weights_ -->
<!-- - $d_{12}^2$ is the distance between $1$ and $2$ -->
<!-- - And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34] -->

<!-- In addition to demographic force, Stewart defined a measure of "population potential" for a system with populations $i$ : -->
<!-- $$ -->
<!-- V_i = \int\frac{D}{r} ds \approx \sum_j \frac{M_j}{d_{ij}} -->
<!-- $${#eq-stewart-population-potential-integral-sum} -->

<!-- \noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. It can be rewritten as our modern definition of accessibility. -->

<!-- ## From same roots, we get accessibility à la Hansen -->

<!-- Using Stewart's population potential -- @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ rewritten as follows to acccept any impedance function: -->

<!-- $$ -->
<!-- S_{i} = \sum_j O_j \cdot f(d_{ij}) -->
<!-- $${#eq-accessibility-general} -->

<!-- - what do we see...? No proportionality constant! It was left for future determination in @stewartDemographicGravitationEvidence1948 (so set to 1). In @hansen1959, not mentioned but implicitly assumed to be set to 1.  -->

<!-- - This is very important: if set to 1, then we have equality: LHS = RHS. If not set then we only have proportionality.  -->

<!-- ## For the same roots, we also get the gravity model:  -->

<!-- As defined in @wilson1971: -->
<!-- $$ -->
<!-- T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
<!-- $${#eq-phys-gravity-model} -->

<!-- Where: -->
<!-- - $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$,  -->
<!-- - $W_i^{(1)}$ and $W_j^{(2)}$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively,  -->
<!-- - $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.  -->
<!-- - $k$ (and most important to this work) is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon. -->

<!-- Wilson went on to define a "family of measures": namely: singly-constrained and doubly-constrained. Replacing $k$ with balancing factors to keep equality (and hence units) as $T_{ij}$. -->

## A family of accessibility measures

Now imagine: what if we define the interaction variable not as the units of trips but as the unit of potential interaction with opportunities:

$$
V_{ij} = k * O_j f(c_{ij})
$${#eq-access}

## A family of accessibility measures: Unconstrained case

$k$ is replaced with 1:
$$
V_{ij} = 1 * O_j f(c_{ij})
$$

At the spatial unit:
$$
V_{i} = 1* \sum_j O_j f(c_{ij})
$$

And no constraints.

## A family of accessibility measures: Total constrained case

$k$ is replaced with $K$:
$$
V_{ij}' = K O_j f(c_{ij})\\
$$

At the spatial unit:
$$
V_{i}' = K * \sum_j O_j f(c_{ij})
$$
$$
V' =  K *\sum_i \sum_j O_j f(c_{ij})\\
V' = V\\
K = \frac{V}{\sum_i \sum_j O_j f(c_{ij})}
$$

And only one macro constraint:
$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}

## A family of accessibility measures: Singly-constrained case

$V_{ij}''^{(1)}$ = origin-constrained (the number of people are constrained)

$V_{ij}''^{(2)}$ = destination-constrained (the number of destinations (e.g., jobs) are constrained -- spatial availability)

$k$ is replaced with $A_i$ and $B_j$ respectively:
$$
V_{ij}''^{(1)} = A_i * O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_j * O_j f(c_{ij})\\
$${#eq-access-singlyconstrained-origin}

\noindent Where $A_i$ and $B_j$ are:
$$
A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
$$
At the zonal value:
$$
V_{i}''^{(1)} =  \sum_j A_i * O_j f(c_{ij})\\
V_{j}''^{(2)} =  \sum_i B_j * O_j f(c_{ij})\\
$$

And two macro constraints:

First:
$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}

And second, either:
$$
\sum_j V_{ij} = P_i
$${#eq-constraint1-access}
\noindent For origin-constrained ($V_{ij}''^{(1)}$), where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

or:
$$
\sum_i V_{ij} = O_j
$${#eq-constraint2-access}
\noindent For destination-constrained ($V_{ij}''^{(2)}$), where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

# A family of accessibility measures: Doubly-constrained case
(gravity model as a partial sum)

When 1 person = 1 opportunity.

The parameter $k$ is replaced with both $A_i$ and $B_j$:
$$
V_{ij}^{'''} = A_i B_j * O_j f(c_{ij}) \\
V_{i}^{'''} =  \sum_j A_i B_j * {O_j f(c_{ij})} \\
$$

\noindent Where $A_i$ and $B_j$ are:
$$
A_{i} = \frac{1}{\sum_j B_{i} O_j f(c_{ij})}\\
B_{j} = \frac{1}{\sum_i A_{i} P_i f(c_{ij})}\\
$$

And two macro constraints:

First:
$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}

And second, either:

$$
\sum_j V_{ij} = P_i
$${#eq-constraint1-access}
\noindent For origin-constrained ($V_{ij}''^{(1)}$), where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

or:
$$
\sum_i V_{ij} = O_j
$${#eq-constraint2-access}
\noindent For destination-constrained ($V_{ij}''^{(2)}$), where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

## Synthetic example

Let's say $W_i^{(1)}$ is population at $i$ who seek opportunities, lets recast this variable as $P_i$. Then let's say $W_j^{(2)}$ are the opportunities being sought, recast to $O_j$. For this simple example, we can say that the opportunities are the population's jobs. In our region, each person has a job -- but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost ($C$ and associated travel propensity $1/C$). (let's pretend our 'known_access_matrix' is unknown).

```{r}
known_access_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
W_i <- known_access_matrix |> rowSums()
W_j <- known_access_matrix |> colSums()
V <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
knowns_longer <- cbind(known_access_matrix,W_i)
knowns_longer <- rbind(knowns_longer,cbind(t(W_j),V))

rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
colnames(knowns_longer) <- c("1","2","3", "TOTAL")

knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
```

And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
```{r}
C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
C_longer <-C
rownames(C_longer) <- c("1","2","3") 
colnames(C_longer) <- c("1","2","3") 
C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
```

## Case 1: Unconstrained

AKA Hansen-type accessibility

$$
V_{ij} = k O_j f(c_{ij})
$$
\noindent Where no constraints hold, e.g., proportionality constant, $k$ is set to 1. This makes the interpretation of the units of $V_{ij}$ meaningless. The value's magnitude can only be interpreted.

```{r}
V_unc <- t(matrix(rep(W_j, 3), 3, 3)) / C

W_i_unc <- V_unc |> rowSums()
W_j_unc <- V_unc |> colSums()
V_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(V_unc), NA)
V_unc_longer <- cbind(V_unc,W_i_unc)
V_unc_longer <- rbind(V_unc_longer, cbind(t(W_j_unc),V_TOTAL_unc))

rownames(V_unc_longer) <- c("1","2","3", "TOTAL") 
colnames(V_unc_longer) <- c("1","2","3", "TOTAL")

V_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

We'd report i=1 has 148, i=2 has 248, and i=3 has 172 accessibility. (implicitly, this is saying j=1 123,...)

The sum of values at each origin or destination or overall, do not add up to anything we know. What we do know is that for the units to retain their meaning of 'number of potentially reachable opportunities', all balancing factors $k$ equal 1. When would this happen? It's hard to say... as the balancing factors themselves do not have practical meaning; what has meaning are the constraints. By satisfying the constraints, the model gains meaning. In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of potential interaction with $O_j$ assuming $f(c_{ij})$.

OF NOTE: in the context of spatial interaction modelling, the unconstrained model is not entirely useful - as the units of 'trips' -- but trips that satisfy the macro constraint of $k$ equaling 1, which is practically meaningless.

## Case 2: Total constrained

In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-access satisfying only constraint @eq-constraint0-access) so the model looks as follows:
$$
V_{ij}' = K O_j f(c_{ij})\\
$$
\where Solved for our example, $K$ is equal to the total number of 'access" divided by the sum of all the unconstrained access:
$$
K = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
K = \frac{V}{\frac{O_1}{c_{1,1}}+\frac{O_1}{c_{2,1}}+\frac{O_1}{c_{3,1}}+\frac{O_2}{c_{1,2}} +... \frac{O_3}{c_{3,3}}
}\\
K = \frac{790}{568.4} \\
K = 1.389866
$$

```{r}
K <- V/sum(V_unc)
V_tot <- K * V_unc

W_i_tot <- V_tot |> rowSums()
W_j_tot <- V_tot |> colSums()
V_TOTAL_tot <- ifelse(sum(W_i_tot)==sum(W_j_tot), sum(V_tot), NA)
V_tot_longer <- cbind(V_tot,W_i_tot)
V_tot_longer <- rbind(V_tot_longer, cbind(t(W_j_tot),V_TOTAL_tot))

rownames(V_tot_longer) <- c("1","2","3", "TOTAL") 
colnames(V_tot_longer) <- c("1","2","3", "TOTAL")

V_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```
While we can see that the 'number of potentially reachable opportunities' `r V` results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model).

## Case 3: Singly-constrained - destination-constrained access

Two versions of this case, either origin-constrained or destination-constrained.

For the destination-constrained model, where $k$ is replaced with $B_{ij}$ (@eq-bf-B-access satisfying only constraint @eq-constraint2-access  (e.g., $\sum_i V_{ij} = O_j$) and implicitly @eq-constraint0-access), the model looks as follows:
$$
V_{ij}''^{(1)} = B_{ij} O_j f(c_{ij})\\
$${#eq-access-singlyconstrained-origin}

\where In our example, $B_{ij}$ is:
$$
B_{ij} = \frac{P_{ij}}{\sum_i P_i f(c_{ij})}\\
B_{1,1} = \frac{P_{ij}}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
...\\
B_{1,1} = 160 *\frac{1}{160/2+450/15+180/5} = 1.09589 \\ 
B_{2,1} = 450 *\frac{1}{160/2+450/15+180/5} =3.082192 \\
B_{3,1} = 180 * \frac{1}{160/2+450/15+180/5} =1.232877 \\
...\\
B_{1,2} = 160 *\frac{1}{160/15+450/2+180/10} =0.630749 \\
B_{2,2} = 450 *\frac{1}{160/15+450/2+180/10} =1.773982 \\
B_{3,2} = 180 * \frac{1}{160/15+450/2+180/10} =0.7095926 \\
...\\
B_{1,3} = 160 *\frac{1}{160/5+450/10+180/2} = 0.9580838\\
B_{2,3} = 450 *\frac{1}{160/5+450/10+180/2} = 2.694611  \\
B_{3,3} = 180 *\frac{1}{160/5+450/10+180/2} = 1.077844 \\
$$
```{r}
B_j <-matrix(W_i) %*% (1/(t((matrix(W_i))) %*% (1/C)))  
B_j 
```

```{r}
V_dest <- (B_j) * V_unc # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))

W_i_dest <- V_dest |> rowSums()
W_j_dest <- V_dest |> colSums()

V_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(V_dest), NA)
V_dest_longer <- cbind(V_dest,W_i_dest)
V_dest_longer <- rbind(V_dest_longer, cbind(t(W_j_dest),V_TOTAL_dest))

rownames(V_dest_longer) <- c("1","2","3", "TOTAL") 
colnames(V_dest_longer) <- c("1","2","3", "TOTAL")

V_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

And as we can observe, the $O_j$ values are equal to the known opportunities, while $P_i$ is not.

NOTE: This is also spatial availability!

## Case 3: Singly-constrained access case - origin-constrained case

For the origin-constrained model,  $k$ is replaced with $A_{ij}$ (@eq-bf-A-access satisfying only constraint @eq-constraint1-access  (e.g., $\sum_j V_{ij} = P_i$ and) and implicitly @eq-constraint0-access), the model looks as follows:
$$
V_{ij}''^{(2)} = A_{ij} O_j f(c_{ij})\\
$${#eq-access-singlyconstrained-origin}

\where In our example, $A_{ij}$ is:
$$
A_{ij} = \frac{O_{ij}}{\sum_j O_j f(c_{ij})}\\
...\\
$$
```{r}
A_i <-matrix(W_j) %*% (1/(t((matrix(W_j))) %*% (1/C)))
#diag(A_i|> as.numeric(), nrow=3, ncol=3)
A_i 
```

```{r}
V_orig <- (A_i) * (1/C)* (matrix(rep(W_i, 3), 3, 3))  # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))

W_i_orig <- V_orig |> rowSums()
W_j_orig <- V_orig |> colSums()

V_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(V_orig), NA)
V_orig_longer <- cbind(V_orig, W_i_orig)
V_orig_longer <- rbind(V_orig_longer, cbind(t(W_j_orig),V_TOTAL_orig))

rownames(V_orig_longer) <- c("1","2","3", "TOTAL") 
colnames(V_orig_longer) <- c("1","2","3", "TOTAL")

V_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

And as we can observe, the $P_i$ values are equal to the known populations, while $O_j$ is not.

NOTE: This is also spatial availability -- except I switched $O_j$ for $P_i$. This is also related to retail modeling

## Case 4: Doubly-constrained case
NOTE: This one needs work... right now, it's equal to doubly-constrained gravity model.

The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-access and @eq-bf-B-access satisfying all three constraints @eq-constraint1-access,  @eq-constraint2-access and implicitly @eq-constraint0-access), the model looks as follows:
$$
V_{ij}^{'''} = A_i B_j O_j f(c_{ij})
$$
\where In our example, $A_i$ and $B_j$ is:
$$
A_{ij} = \frac{O_{ij} P_i}{\sum_j B_{ij} O_j f(c_{ij})}\\
B_{ij} = \frac{P_{ij}}{\sum_i A_{ij} P_i f(c_{ij})}\\
$$

```{r}
doubly_constrained_gravity_model <- function(
    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
  P,  # Origin production weights vector (POPULATION)
  O,  # Destination attraction weights vector (OPPORTUNITIES)
  impedance_matrix,  # the resulting travel cost impedance matrix
  error_threshold = 0.001,  # Error threshold for stopping condition
  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
) {

  # Normalize O and D
  sum_P <- sum(P)
  sum_O <- sum(O)
  cat("Checking population, opportunity balancing:\n")
  cat("Population: ", sum_P, "\n")
  cat("Opportunities: ", sum_O, "\n")

  if (sum_P != sum_O) {
    cat("Population and opportunities do not balance, scaling to match.\n")
    if (sum_P < sum_O) {
      P <- P * (sum_O / sum_P)
    } else {
      O <- O * (sum_P / sum_O)
    }
  } else {
    cat("Population, opportunities balancing OK.\n")
  }

  n <- length(P)  # Number of i
  V <- sum(P)  # Total access
  Ai <- rep(1, n)  # Initialize Ai
  Bj <- rep(1, n)  # Initialize Bj

  previous_error <- Inf
  iteration_count <- 0
  stop_reason <- ""

  while (TRUE) {
    iteration_count <- iteration_count + 1

    # Update Ai
    for (i in 1:n) {
      Ai[i] <- 1 / (sum(Bj * O * impedance_matrix[i, ]) + 1e-9)
    }

    # Update Bj
    Bj_new <- rep(1, n)
    for (j in 1:n) {
      Bj_new[j] <- 1 / (sum(Ai * P * impedance_matrix[, j]) + 1e-9)
    }

    # Calculate Vij
    #Vij <- outer(Ai * P, Bj_new * O) * impedance_matrix
    Vij <- Bj_new * Ai * O * impedance_matrix

    # Print Ai and Bj for this iteration
    cat("\nIteration:", iteration_count, "\n")
    cat("Ai:", round(Ai, 3), "\n")
    cat("Bj:", round(Bj_new, 3), "\n")

    # Calculate error
    error <- (sum(abs(P - rowSums(Vij))) + sum(abs(O - colSums(Vij)))) / V

    # Calculate change in error
    error_change <- abs(previous_error - error)

    # Check stopping conditions
    if (error < error_threshold) {
      stop_reason <- "Error threshold met"
      break
    } else if (error_change < improvement_threshold) {
      stop_reason <- "Slow improvement"
      break
    }

    previous_error <- error
    Bj <- Bj_new
  }

  # report stopping condition and final error threshold
  cat("\nStopping Condition: ", stop_reason, "\n")
  cat(sprintf("Final Error: %.3f%%\n", error * 100))

  return(list(Ai = Ai, Bj = Bj, Vij = Vij))
}

dc_results <- doubly_constrained_gravity_model(P=W_i, O=W_j, impedance_matrix = 1/C)
dc_results$Vij
```

The $A_i$ and $B_j$ matrices:
```{r}
A_i_doubly <- dc_results$Ai
B_j_doubly <- dc_results$Bj

A_i_doubly |> matrix()
B_j_doubly |> matrix() |> t()
```

The trip matrix (notice, all summed up opportunities and populations are equivalent to our knowns)
```{r}
V_doubly <- dc_results$Vij

W_i_doubly <- V_doubly |> rowSums()
W_j_doubly <- V_doubly |> colSums()
V_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(V_doubly), NA)
V_doubly_longer <- cbind(V_doubly,W_i_dest)
V_doubly_longer <- rbind(V_doubly_longer, cbind(t(W_j_doubly),V_TOTAL_dest))

rownames(V_doubly_longer) <- c("1","2","3", "TOTAL") 
colnames(V_doubly_longer) <- c("1","2","3", "TOTAL")

V_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
```

## Case 5: Hybrid / partial constraints

## next steps
- grocery stores for example; maybe 2 grids?
- bibliometric analysis 
- - draft of this end of jan 2025

- (future) intervening opportunities ? radiation model - deep gravity
- (future) accessibility pyramids 
- (future) MAUP addressed
- (future) comparing cities

### 2024-12-13 paper/manuscript.qmd
5-slide trb pres draft

@@
 
-The aim of this paper's focus is on the proportionality constant.
+This paper's focus is on the proportionality constant.
 We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings.
@@
 
-$$
-\sum_i\sum_j T_{ij} c_{ij} = C
-$${#eq-constraint0_5-gravitymodel}
+<!-- $$ -->
+<!-- \sum_i\sum_j T_{ij} c_{ij} = C -->
+<!-- $${#eq-constraint0_5-gravitymodel} -->
 
@@
 The basis of the approach is to accept that there are macro, meso, and micro states.
-In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of indiivudal trip makers, and the macro state is the entire study region.
+In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
 Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
@@
 
-- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
+- The unconstrained case: neither @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
 - The total constrained case: only @eq-constraint0-gravitymodel hold
@@
 For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
-And for the unconstrained case, in a way, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.
+And for the unconstrained case, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.
 
@@
 
-Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities.
-We don't have to change much!
-Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities.
-$T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:
+Instead of the interaction variable being in the unit of trips, it is in the unit of potential interaction with opportunities.
+This focus is the preoccupation of accessibility measures.
+The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of gravity models, however with different interpretation.
+In this section, we explicitly define these cases.
+
+In this context, accessibility is our _interaction_ variable of choice, so $T_{ij}$ in @eq-phys-gravity-model is replaced with $V_{ij}$.
+As the interaction variable represents _potential_, i.e., the number of potential reachable opportunities, as opposed to trips:
+
+<!-- W_i^{(1)} enters through k.. it is a quality of the destination -->
 
 $$
-V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
+V_{ij} = k W_j^{(2)} f(c_{ij})
 $${#eq-access}
@@
 Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
+Accessibility is also most often presented as a locational measure, the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone.
 
@@
 $${#eq-constraint0-access}
+\noindent Where the sum of all potentially reached opportunities in the region is equal to a total opportunities in the region $V$.
 
 $$
-\sum_i\sum_j V_{ij} c_{ij} = C
-$${#eq-constraint0_5-access}
-\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.
-
-$$
-\sum_j V_{ij} = W_i^{(1)}
+\sum_j V_{ij} = W_i^{(1)} = P_i
 $${#eq-constraint1-access}
@@
 $$
-\sum_i V_{ij} = W_j^{(2)}
+\sum_i V_{ij} = W_j^{(2)} = O_j
 $${#eq-constraint2-access}
@@
 ## Unconstrained accessibility
-(no proportionality constant, k=1, no constraints hold - hard to interpret):
+
+- No proportionality constant, $k$ is set to 1, so no constraints hold, making the interpretation of the units meaningless so the value's magnitude can only be interpreted:
 $$
-S_{ij} = P_i O_j f(c_{ij})
+V_{ij} = k O_j f(c_{ij})
 $$
-In practice, $S_i$ is calculated for each $i$ so:
+
+- In practice, $V_i$ is also calculated for each $i$. When thinking about the interpretation of the value, how many opportunities can be visited for each $i$, the $P_i$ is also set to 1:
 $$
-S_{i} = P_i \sum_j O_j f(c_{ij})
+V_{i} = \sum_j O_j f(c_{ij})
 $$
@@
 $$
-S_{ij}' = K P_i O_j f(c_{ij})\\
-S_{i}' = K P_i \sum_j O_j f(c_{ij})
+V_{ij}' = K P_i O_j f(c_{ij})\\
+V_{i}' = K P_i \sum_j O_j f(c_{ij})
 $$
@@
 $$
-V_{ij}''^{(1)} = A_i P_i O_j f(c_{ij})\\
-V_{ij}''^{(2)} = B_j P_i O_j f(c_{ij})\\
-
-V_{i}''^{(1)} = A_i P_i \sum_j O_j f(c_{ij})\\
-V_{i}''^{(2)} = P_i \sum_j B_j O_j f(c_{ij})\\
+V_{ij}''^{(1)} = A_i O_j f(c_{ij})\\
+V_{ij}''^{(2)} = B_j O_j f(c_{ij})\\
 
@@
 $$
-For the origin-constrained, $V_{i}''^{(1)}$ are the number of opportunities (like jobs) that can be reached at $i$, and $V_{i}''^{(2)}$ are the number of opportunities (like people) that can be reached at $i$
 
-In terms of origin-constrained -- spatial availability is explicitly so [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024].
-Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general [].
+In terms of origin-constrained -- ..? .
 
@@
 Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
+Spatial availability is explicitly also destination-constrained [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024].
+Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general []
 

### 2024-12-30 data-raw/01-bibliometrics-analysis.qmd
some updates to the manuscript

@@
              alpha = 0.5) +
-
   theme_void()
+ggsave("Han_Wilson_bib_coupling_plot.png")
 ```
@@
 ```{r}
+both_Han_Wilson <- 
   corpus |> 
   filter(corpus == "Both") |> 
-  select(AU, PY) |> arrange(PY)
+  select(AU, PY, TI) |> arrange(PY)
+
+write.csv(both_Han_Wilson, file = "both_Han_Wilson.csv")
 ```

### 2024-12-30 paper/manuscript.qmd
some updates to the manuscript

@@
 In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling].
-However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geurs2004; @ferreiraReenactingMobilityAccessibility2020].
+However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geursAccessibilityEvaluationLanduse2004; @ferreiraReenactingMobilityAccessibility2020].
 
@@
 ## The inquiry into spatial interaction traces to Newton's Universal Gravitation
-
+<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
 The patterns of people's movement in space have been a subject of scientific inquiry for centuries.
@@
 
-<!-- SUMMARY: from the 1880s to the 1940s, researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
+<!-- SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s.
+These researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
 This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy.
@@
 
-\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.
+\noindent Where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.
 
@@
 For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p.
-34 <!--Antonio, is this what you meant?-->.
+34.
 Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism.
@@
 
-## Hansen's gravity-based accessibility
-<!-- SUMMARY: in Hansen's use of @stewartDemographicGravitationEvidence1948's Population Potential measure, we highlight two crucial omissions.
-1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it.
-2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient.
+## Hansen's gravity-based accessibility to today
+<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up.
+Is now cited as the father of accessibility.
+However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day.
+The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it.
+The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient.
 In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed.
@@
 
-## From Hansen-type accessibility to accessibility today
-<!-- SUMMARY: accessibility from Hansen to today is very diverse and has a lack of concern for the interpretability of output units; likewise, a lack of concern for balancing units such as a proportionality constant.
-Our answer: let's focus on those omitted (or implicit set to 1) proportionality constants.-->
-
-Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [CITE MAJOR REVIEWS OVER LAST 30 YEARS].
+Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [@koenigIndicatorsUrbanAccessibility1980; @geursAccessibilityEvaluationLanduse2004; @millerAccessibilityMeasurementApplication2018; @wuUnifyingAccess2020].
 However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it.
@@
 
-To overview recent accessibility literature development, the majority of studies develop on; the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
-Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.
-
-However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [].
+However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [@millerAccessibilityMeasurementApplication2018].
 Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired.
@@
 Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020].
-As reviewed, quantity and variations on @eq-accessibility-general also compounds the issue of interpretability across studies.
+Furthermore, The majority of accessibility studies within the past few decades develop on the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
+Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.
+These variations also compounds the issue of interpretability across studies.
 
 A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components.
-@wuUnifyingAccess2020 demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities.
+@curtis2010planning further classifies accessibility measures into spatial separation, contour-, gravity-, competition-, time-space, utility-, and network- based measures [@curtis2010planning], though methodologies may clearly overlap.
+Broader approaches have also been taken: @paez2012measuring categories indicators on how they have been normatively and positively applied, and @levineCenturyEvolutionAccessibility2020 reviews accessibility's evolution along definition-based and application-based dimensions throughout the years.
+
+Among these reviews, we find @wuUnifyingAccess2020's contribution compelling for unifying accessibility measures: they demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities.
 Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints.
@@
 
-This paper's focus is on the proportionality constant.
+As such, this paper's focus is on the proportionality constant.
 We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings.
@@
 
-# Gravity-based accessibility's cousin: Wilson's spatial interaction model
+# Introducing a family of accessibility measures
+
+## Gravity-based accessibility's cousin: Wilson's spatial interaction model
 
@@
 However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners.
-
-There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:
+From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:
 
@@
 
-**Entropy maximising: the gravity-model as statistical averages:**
 Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
@@
 Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].
-
 As such, @eq-phys-gravity-model can be derived to account for additional information to include constraints that reflect known information about the system and is mathematically rigorous.
@@
 
-The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$.
-Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:
+The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!-- Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:-->
 $$
@@
 
-The entropy maximation derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967).
+The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967).
 The basis of the approach is to accept that there are macro, meso, and micro states.
 In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
-Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
-From this approach, the gravity model can be interpreted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
+Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
+From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
 There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.
 
-Using these constraints, the gravity model @eq-phys-gravity-model is can now be distinguished into four general cases.
-These cases are presented in increase order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+Using these constraints, the gravity model @eq-phys-gravity-model is distinguished into four general cases.
+These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
 
@@
 
-The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.
-For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below:
-```{r}
-known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
-#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
-W_i <- known_trip_matrix |> rowSums()
-W_j <- known_trip_matrix |> colSums()
-T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
-knowns_longer <- cbind(known_trip_matrix,W_i)
-knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))
+## Accessibilities' development seperate from the family of gravity models
 
-rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
-colnames(knowns_longer) <- c("1","2","3", "TOTAL")
+Upon detailing the gravity model, we can now discuss how accessibility has been developed within that field and how it has not.
+Gravity modellers (interaction variable being trips, a model of spatial interaction - mobility) continued to use accessibility within transportation demand modelling.
+As succinctly summarized in @morrisAccessibilityIndicatorsTransport1979, "Mobility and accessibility together influence an individual’s capacity to travel in daily life".
+However, the gravity model literature that focuses on Hansen's accessibility is relatively scarce.
+To capture this sentiment, see @fig-Fig1 which demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971.
+@fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].
 
-knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
+```{r plot-figure1, out.width=600, out.height=600}
+#| label: fig-Fig1
+#| fig-cap: "The bibliometric coupling of papers that cite Hansen (1959), Wilson (1971) and both Hansen (1959) and Wilson (1971)."
+include_graphics(glue(here(), "/data-raw/Han_Wilson_bib_coupling_plot.png"))
 ```
 
-And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
-```{r}
-C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
-C_longer <-C
-rownames(C_longer) <- c("1","2","3") 
-colnames(C_longer) <- c("1","2","3") 
-C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
-```
+Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the gravity model (mobility).
 
-## Unconstrained interaction case
+For instance, @battyMethodResiduesUrban1976 states that the relationship between the gravity model and Hansen's original accessibility-potential model may be worth exploring...
+There is a consistent acknowledgement that these two models are related.
+Specifically, that the Hansen-type accessibility model is embedded within the balancing factors [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
+This is a "common sense" approach at integrating accessibility [@morrisAccessibilityIndicatorsTransport1979].
 
-$k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$
+Though there are other approaches to consider accessibility with spatial interaction modelling beyond Hansen founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979].
+For instance, others explore entropy maximising solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].
 
-$$
-T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-```{r}
-T_unc <- W_i %*% t(W_j) / C
+Some work on location-allocation models also used both Hansen and Wilson.
+For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained gravity model, and then defines one using Hansen-type accessibility as the objective function subject.
 
-W_i_unc <- T_unc |> rowSums()
-W_j_unc <- T_unc |> colSums()
-T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA)
-T_unc_longer <- cbind(T_unc,W_i_unc)
-T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc))
+Then, @fotheringhamSpatialCompetitionAgglomeration1985 demonstrated that destination's accessibility should be explicitly specified within Wilson's framework, citing Hansen measure as convenient formulation.
+Without the consideration for accessibility explicitly, unexplained spatial pattern remains.
+This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in gravity models.
+[@oshanSpatialStructureDebate2021].
+Implicitly, we gather that at this point: it was understood balancing factors in the gravity model could not be understood as capturing accessibility.
+We also gather that in this moment of growing complexity of spatial interaction modelling, accessibility branched off and sprouted in the medium of budding GIS technology and the emergence of space-time dimensions [@millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
+With the possibility for showcasing complex GIS methodologies for varied case studies, accessibility became mature enough to branch off on its own.
 
-rownames(T_unc_longer) <- c("1","2","3", "TOTAL") 
-colnames(T_unc_longer) <- c("1","2","3", "TOTAL")
+In papers after @fotheringhamSpatialCompetitionAgglomeration1985, Hansen is typically cited as providing a definition of accessibility but how Wilson is cited changes through the years.
+Wilson is attributed for using different travel cost functions in depending on the context [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 
-T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
+Othertimes, Wilson and Hansen measures were progressively incorrectly conflated throughout the years [ @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring] - indeed they both measure spatial interaction so are somewhat related, but they are not the same.
+They are also co-cited to refer to the concept of 'gravity' in models [@liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]
 
-As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends.
-Instead, they take on the assumptions of all balancing factors $k$ being equal to 1.
-When would this happen?
-It's hard to say...
-as the balancing factors themselves do not have practical meaning; what has meaning are the constraints.
-By satisfying the constraints, the model gains meaning.
-In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$.
-In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.
+And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974a; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.
 
-## Total constrained interaction case
+THE SPATIAL STRUCTURE DEBATE: @oshanSpatialStructureDebate2021
 
-In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows:
-$$
-T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-\where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made:
-$$
-K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\
-K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +... \frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}}
-}\\
-K = \frac{790}{166443} \\
-K = 0.00474636
-$$
+Some papers also use gravity models to estimate trip or some other interaction flows alongside accessibility indicators [@clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the gravity model, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
 
-```{r}
-K <- T/sum(sum(T_unc))
-T_tot <- K * T_unc
+Hansen is a relatively straightforward simple application, while Wilon's contribution is methodological and dense - must be carefully read.
+Many takeaways can be drawn, so with accessibilities changing applications throughout the years -- different things about Wilon's contribution becomes relevant.
+The most contemprary connection between the two is in the relam of competitive accessibility -- in the interpretation of Wilson's balancing factors as a way to consider competition.
+We fit this contribution in this literature.
+Competition fits on a spectrum - by introducing balancing factors trips are _constrained_ to equal regional or zonal totals.
+This same application can be applied to accessibility.
 
-W_i_tot <- T_tot |> rowSums()
-W_j_tot <- T_tot |> colSums()
-T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA)
-T_tot_longer <- cbind(T_tot,W_i_tot)
-T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot))
+Some 
+Additionally, trending research topics of the time such as space-time and more  []
 
-rownames(T_tot_longer) <- c("1","2","3", "TOTAL") 
-colnames(T_tot_longer) <- c("1","2","3", "TOTAL")
+previously mentioned in accessibility development: the breadth of opportunity $O_j$ that can be studied [] as well as the accuracy of parameters that capture distance-decay $f(c_{ij})$ [].
+Outside of these factors, accessibility literature that delves into competition, does not cite Wilson, with some exception...
+Soukhov for the balancing factor.
 
-T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model.
+Overall, work that cities Wilson is closer related to transportation demand modelling than accessibility. However there are a few works of the 77 documents that are in the realm of accessibility. ...
 
-## Singly-constrained
+However there are a few papers that are accessibility focused that focus on
 
-Two versions of this case, either origin-constrained or destination-constrained.
+<!-- REVIEW CSV OF PAPERS-->
 
-For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
-$$
-T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-\where In our example, $A_i$ is:
-$$
-A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\
-A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\
-...\\
-A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\
-A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\
-A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201 
-$$
+There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models.
 
-```{r}
-A_i <- 1/t((matrix(W_j))) %*% (1/C)
-#A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns
-A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3)
-
-T_orig <-  A_i%*%T_unc
-
-W_i_orig <- T_orig |> rowSums()
-W_j_orig <- T_orig |> colSums()
-T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA)
-T_orig_longer <- cbind(T_orig,W_i_orig)
-T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig))
-
-rownames(T_orig_longer) <- c("1","2","3", "TOTAL") 
-colnames(T_orig_longer) <- c("1","2","3", "TOTAL")
-
-T_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-We can see that the $W_i^{(1)}$ values are equal to the trip ends, but not $W_j^{(2)}$ values.
-
-Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel):
-$$
-T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-\where $B_j$ is, and equal to:
-$$
-B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
-$$
-```{r}
-B_j <- 1/(t(matrix(W_i))) %*% (1/C)
-
-B_j <- matrix(B_j) |> t()
-B_j_longer <- B_j
-colnames(B_j_longer) <- c("1","2","3")
-B_j_longer  #|> as.data.frame()  |> gt()
-#B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns
-```
-
-And trips are:
-```{r}
-B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3)
-
-T_dest <-  (B_j)%*%t(T_unc) |> t()
-
-W_i_dest <- T_dest |> rowSums()
-W_j_dest <- T_dest |> colSums()
-T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA)
-T_dest_longer <- cbind(T_dest,W_i_dest)
-T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest))
-
-rownames(T_dest_longer) <- c("1","2","3", "TOTAL") 
-colnames(T_dest_longer) <- c("1","2","3", "TOTAL")
-
-T_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-And as we can observe, the $W_j^{(2)}$ values are equal to the known trip ends, while $W_i^{(1)}$ is not.
-
-## Doubly-constrained case
-
-The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows:
-$$
-T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-\where In our example, $A_i$ and $B_j$ is:
-$$
-A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\
-B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})}
-$$
-```{r}
-doubly_constrained_gravity_model <- function(
-    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
-  O,  # Origin production weights vector
-  D,  # Destination attraction weights vector
-  impedance_matrix,  # the resulting travel cost impedance matrix
-  error_threshold = 0.001,  # Error threshold for stopping condition
-  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
-) {
-
-  # Normalize O and D
-  sum_O <- sum(O)
-  sum_D <- sum(D)
-  cat("Checking production, attraction balancing:\n")
-  cat("Production: ", sum_O, "\n")
-  cat("Attraction: ", sum_D, "\n")
-
-  if (sum_O != sum_D) {
-    cat("Productions and attractions do not balance, scaling to match.\n")
-    if (sum_O < sum_D) {
-      O <- O * (sum_D / sum_O)
-    } else {
-      D <- D * (sum_O / sum_D)
-    }
-  } else {
-    cat("Production, attraction balancing OK.\n")
-  }
-
-  n <- length(O)  # Number of i
-  T <- sum(O)  # Total trips
-  Ai <- rep(1, n)  # Initialize Ai
-  Bj <- rep(1, n)  # Initialize Bj
-
-  previous_error <- Inf
-  iteration_count <- 0
-  stop_reason <- ""
-
-  while (TRUE) {
-    iteration_count <- iteration_count + 1
-
-    # Update Ai
-    for (i in 1:n) {
-      Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9)
-    }
-
-    # Update Bj
-    Bj_new <- rep(1, n)
-    for (j in 1:n) {
-      Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9)
-    }
-
-    # Calculate Tij
-    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
-
-    # Print Ai and Bj for this iteration
-    cat("\nIteration:", iteration_count, "\n")
-    cat("Ai:", round(Ai, 3), "\n")
-    cat("Bj:", round(Bj_new, 3), "\n")
-
-    # Calculate error
-    error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T
-
-    # Calculate change in error
-    error_change <- abs(previous_error - error)
-
-    # Check stopping conditions
-    if (error < error_threshold) {
-      stop_reason <- "Error threshold met"
-      break
-    } else if (error_change < improvement_threshold) {
-      stop_reason <- "Slow improvement"
-      break
-    }
-
-    previous_error <- error
-    Bj <- Bj_new
-  }
-
-  # report stopping condition and final error threshold
-  cat("\nStopping Condition: ", stop_reason, "\n")
-  cat(sprintf("Final Error: %.3f%%\n", error * 100))
-
-  return(list(Ai = Ai, Bj = Bj, Tij = Tij))
-}
-
-dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C)
-```
-
-The $A_i$ and $B_j$ matrices:
-```{r}
-A_i_doubly <- dc_results$Ai
-B_j_doubly <- dc_results$Bj
-
-A_i_doubly |> matrix()
-B_j_doubly |> matrix() |> t()
-```
-
-The trip matrix (notice, all trip ends are equivalent to the known trip ends)
-```{r}
-T_doubly <- dc_results$Tij
-
-W_i_doubly <- T_doubly |> rowSums()
-W_j_doubly <- T_doubly |> colSums()
-T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA)
-T_doubly_longer <- cbind(T_doubly,W_i_dest)
-T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest))
-
-rownames(T_doubly_longer) <- c("1","2","3", "TOTAL") 
-colnames(T_doubly_longer) <- c("1","2","3", "TOTAL")
-
-T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-## Mean absolute percentage error (MAPE) between all cases
-
-```{r}
-#unconstrained
-((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
-
-#unconstrained -- rescaled!
-W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums()
-W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums()
-
-W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums()
-W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums()
-
-((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01)
-```
-
-```{r}
-#total constrained
-((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
-
-#singly-constrained (orig)
-((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
-
-#singly-constrained (dest)
-((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
-
-#doubly-constrained (dest)
-((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01)
-```
-
-- doubly c is lowest error
-- singly constrained error is mid
-- total constrained has high error
-- notably, unconstrained unscaled is obviously much higher in error - there are no constraints!
-But if you scale the matrices, the error is both low (interesting!!
-why is this?).
-However as mentioned, the unconstrained case is hard to interpret literally -- in what situations would k=1..?
-
-# A family of accessibility measures
+## A family of accessibility measures
 
@@
 
-## Unconstrained accessibility
+### Unconstrained accessibility
 
@@
 
-## Total flow constrained accessibility
+### Total flow constrained accessibility
 
@@
 
-## Singly-constrained case
+### Singly-constrained case
 
@@
 
-## Doubly constrained case (origin-destination constrained)
+### Doubly constrained case (origin-destination constrained)
 
@@
 
-## Hybrid singly-constrained case: 
+### Hybrid singly-constrained case: 
 <!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here-->
@@
   -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->
+
+<!--THE FOLLOWING ARE SOLVED EXAMPLES OF ALL FOUR CASES OF THE GRAVITY MODEL -->
+<!-- The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.
+For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below: -->
+<!-- ```{r} -->
+<!-- known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE) -->
+<!-- #unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE) -->
+<!-- W_i <- known_trip_matrix |> rowSums() -->
+<!-- W_j <- known_trip_matrix |> colSums() -->
+<!-- T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA) -->
+<!-- knowns_longer <- cbind(known_trip_matrix,W_i) -->
+<!-- knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T)) -->
+
+<!-- rownames(knowns_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(knowns_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows: -->
+<!-- ```{r} -->
+<!-- C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE) -->
+<!-- C_longer <-C -->
+<!-- rownames(C_longer) <- c("1","2","3")  -->
+<!-- colnames(C_longer) <- c("1","2","3")  -->
+<!-- C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- For examples of how each type of case is solved, see many of the cited literutre, like @Wilson and @ortuza. -->
+
+<!-- ## Unconstrained interaction case -->
+
+<!-- $k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$ -->
+
+<!-- $$ -->
+<!-- T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+<!-- $$ -->
+<!-- ```{r} -->
+<!-- T_unc <- W_i %*% t(W_j) / C -->
+
+<!-- W_i_unc <- T_unc |> rowSums() -->
+<!-- W_j_unc <- T_unc |> colSums() -->
+<!-- T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA) -->
+<!-- T_unc_longer <- cbind(T_unc,W_i_unc) -->
+<!-- T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc)) -->
+
+<!-- rownames(T_unc_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(T_unc_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends.
+Instead, they take on the assumptions of all balancing factors $k$ being equal to 1.
+When would this happen?
+It's hard to say...
+as the balancing factors themselves do not have practical meaning; what has meaning are the constraints.
+By satisfying the constraints, the model gains meaning.
+In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$.
+In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.
+-->
+
+<!-- ## Total constrained interaction case -->
+
+<!-- In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
+<!-- $$ -->
+<!-- T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+<!-- $$ -->
+<!-- \where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made: -->
+<!-- $$ -->
+<!-- K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\ -->
+<!-- K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +...
+\frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}} -->
+<!-- }\\ -->
+<!-- K = \frac{790}{166443} \\ -->
+<!-- K = 0.00474636 -->
+<!-- $$ -->
+
+<!-- ```{r} -->
+<!-- K <- T/sum(sum(T_unc)) -->
+<!-- T_tot <- K * T_unc -->
+
+<!-- W_i_tot <- T_tot |> rowSums() -->
+<!-- W_j_tot <- T_tot |> colSums() -->
+<!-- T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA) -->
+<!-- T_tot_longer <- cbind(T_tot,W_i_tot) -->
+<!-- T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot)) -->
+
+<!-- rownames(T_tot_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(T_tot_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+<!-- While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model. -->
+
+<!-- ## Singly-constrained -->
+
+<!-- Two versions of this case, either origin-constrained or destination-constrained.  -->
+
+<!-- For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
+<!-- $$ -->
+<!-- T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+<!-- $$ -->
+<!-- \where In our example, $A_i$ is: -->
+<!-- $$ -->
+<!-- A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\ -->
+<!-- A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\ -->
+<!-- ...\\ -->
+<!-- A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\ -->
+<!-- A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\ -->
+<!-- A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201  -->
+<!-- $$ -->
+
+<!-- ```{r} -->
+<!-- A_i <- 1/t((matrix(W_j))) %*% (1/C) -->
+<!-- #A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns -->
+<!-- A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3) -->
+
+<!-- T_orig <-  A_i%*%T_unc -->
+
+<!-- W_i_orig <- T_orig |> rowSums() -->
+<!-- W_j_orig <- T_orig |> colSums() -->
+<!-- T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA) -->
+<!-- T_orig_longer <- cbind(T_orig,W_i_orig) -->
+<!-- T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig)) -->
+
+<!-- rownames(T_orig_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(T_orig_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- T_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- We can see that the $W_i^{(1)}$ values are equal to the trip ends, but not $W_j^{(2)}$ values.  -->
+
+<!-- Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel): -->
+<!-- $$ -->
+<!-- T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+<!-- $$ -->
+<!-- \where $B_j$ is, and equal to: -->
+<!-- $$ -->
+<!-- B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\ -->
+<!-- $$ -->
+<!-- ```{r} -->
+<!-- B_j <- 1/(t(matrix(W_i))) %*% (1/C)  -->
+
+<!-- B_j <- matrix(B_j) |> t() -->
+<!-- B_j_longer <- B_j -->
+<!-- colnames(B_j_longer) <- c("1","2","3") -->
+<!-- B_j_longer  #|> as.data.frame()  |> gt() -->
+<!-- #B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns -->
+<!-- ``` -->
+
+<!-- And trips are: -->
+<!-- ```{r} -->
+<!-- B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3) -->
+
+<!-- T_dest <-  (B_j)%*%t(T_unc) |> t() -->
+
+<!-- W_i_dest <- T_dest |> rowSums() -->
+<!-- W_j_dest <- T_dest |> colSums() -->
+<!-- T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA) -->
+<!-- T_dest_longer <- cbind(T_dest,W_i_dest) -->
+<!-- T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest)) -->
+
+<!-- rownames(T_dest_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(T_dest_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- T_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- And as we can observe, the $W_j^{(2)}$ values are equal to the known trip ends, while $W_i^{(1)}$ is not. -->
+
+<!-- ## Doubly-constrained case -->
+
+<!-- The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
+<!-- $$ -->
+<!-- T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+<!-- $$ -->
+<!-- \where In our example, $A_i$ and $B_j$ is: -->
+<!-- $$ -->
+<!-- A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\ -->
+<!-- B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})} -->
+<!-- $$ -->
+<!-- ```{r} -->
+<!-- doubly_constrained_gravity_model <- function( -->
+<!--     #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R -->
+<!--   O,  # Origin production weights vector -->
+<!--   D,  # Destination attraction weights vector -->
+<!--   impedance_matrix,  # the resulting travel cost impedance matrix -->
+<!--   error_threshold = 0.001,  # Error threshold for stopping condition -->
+<!--   improvement_threshold = 1e-6  # Improvement threshold for stopping condition -->
+<!-- ) { -->
+
+<!--   # Normalize O and D -->
+<!--   sum_O <- sum(O) -->
+<!--   sum_D <- sum(D) -->
+<!--   cat("Checking production, attraction balancing:\n") -->
+<!--   cat("Production: ", sum_O, "\n") -->
+<!--   cat("Attraction: ", sum_D, "\n") -->
+
+<!--   if (sum_O != sum_D) { -->
+<!--     cat("Productions and attractions do not balance, scaling to match.\n") -->
+<!--     if (sum_O < sum_D) { -->
+<!--       O <- O * (sum_D / sum_O) -->
+<!--     } else { -->
+<!--       D <- D * (sum_O / sum_D) -->
+<!--     } -->
+<!--   } else { -->
+<!--     cat("Production, attraction balancing OK.\n") -->
+<!--   } -->
+
+<!--   n <- length(O)  # Number of i -->
+<!--   T <- sum(O)  # Total trips -->
+<!--   Ai <- rep(1, n)  # Initialize Ai -->
+<!--   Bj <- rep(1, n)  # Initialize Bj -->
+
+<!--   previous_error <- Inf -->
+<!--   iteration_count <- 0 -->
+<!--   stop_reason <- "" -->
+
+<!--   while (TRUE) { -->
+<!--     iteration_count <- iteration_count + 1 -->
+
+<!--     # Update Ai -->
+<!--     for (i in 1:n) { -->
+<!--       Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9) -->
+<!--     } -->
+
+<!--     # Update Bj -->
+<!--     Bj_new <- rep(1, n) -->
+<!--     for (j in 1:n) { -->
+<!--       Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9) -->
+<!--     } -->
+
+<!--     # Calculate Tij -->
+<!--     Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix -->
+
+<!--     # Print Ai and Bj for this iteration -->
+<!--     cat("\nIteration:", iteration_count, "\n") -->
+<!--     cat("Ai:", round(Ai, 3), "\n") -->
+<!--     cat("Bj:", round(Bj_new, 3), "\n") -->
+
+<!--     # Calculate error -->
+<!--     error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T -->
+
+<!--     # Calculate change in error -->
+<!--     error_change <- abs(previous_error - error) -->
+
+<!--     # Check stopping conditions -->
+<!--     if (error < error_threshold) { -->
+<!--       stop_reason <- "Error threshold met" -->
+<!--       break -->
+<!--     } else if (error_change < improvement_threshold) { -->
+<!--       stop_reason <- "Slow improvement" -->
+<!--       break -->
+<!--     } -->
+
+<!--     previous_error <- error -->
+<!--     Bj <- Bj_new -->
+<!--   } -->
+
+<!--   # report stopping condition and final error threshold -->
+<!--   cat("\nStopping Condition: ", stop_reason, "\n") -->
+<!--   cat(sprintf("Final Error: %.3f%%\n", error * 100)) -->
+
+<!--   return(list(Ai = Ai, Bj = Bj, Tij = Tij)) -->
+<!-- } -->
+
+<!-- dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C) -->
+<!-- ``` -->
+
+<!-- The $A_i$ and $B_j$ matrices: -->
+<!-- ```{r} -->
+<!-- A_i_doubly <- dc_results$Ai -->
+<!-- B_j_doubly <- dc_results$Bj -->
+
+<!-- A_i_doubly |> matrix() -->
+<!-- B_j_doubly |> matrix() |> t() -->
+<!-- ``` -->
+
+<!-- The trip matrix (notice, all trip ends are equivalent to the known trip ends) -->
+<!-- ```{r} -->
+<!-- T_doubly <- dc_results$Tij -->
+
+<!-- W_i_doubly <- T_doubly |> rowSums() -->
+<!-- W_j_doubly <- T_doubly |> colSums() -->
+<!-- T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA) -->
+<!-- T_doubly_longer <- cbind(T_doubly,W_i_dest) -->
+<!-- T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest)) -->
+
+<!-- rownames(T_doubly_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(T_doubly_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!-- ## Mean absolute percentage error (MAPE) between all cases -->
+
+<!-- ```{r} -->
+<!-- #unconstrained -->
+<!-- ((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+
+<!-- #unconstrained -- rescaled! -->
+<!-- W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums() -->
+<!-- W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums() -->
+
+<!-- W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums() -->
+<!-- W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums() -->
+
+<!-- ((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01) -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- #total constrained -->
+<!-- ((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+
+<!-- #singly-constrained (orig) -->
+<!-- ((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+
+<!-- #singly-constrained (dest) -->
+<!-- ((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+
+<!-- #doubly-constrained (dest) -->
+<!-- ((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+<!-- ``` -->
+
+<!-- - doubly c is lowest error -->
+<!-- - singly constrained error is mid -->
+<!-- - total constrained has high error -->
+<!-- - notably, unconstrained unscaled is obviously much higher in error - there are no constraints!
+But if you scale the matrices, the error is both low (interesting!!
+why is this?).
+However as mentioned, the unconstrained case is hard to interpret literally -- in what situations would k=1..?
+-->
+
+<!-- So...
+our dear Mac alumni Fotheringham, in the early 1980s as a fresh PhD graduate demonstrates that the gravity model is misspecified.
+Responding to other authors at the time saying there is a 'map pattern' left over in estimated gravity models -- he demonstrated that this residual spatial variation can be reduced by including an additional independent variable into the gravity model, accessibility at the destination!
+This is the 'competing destination' model.
+In transport modelling, the CD model is one of a few ways they address this spatial In the wake of the debate, four generalizable approaches were proposed to account for the spatial structure effect (others include: the Box–Cox transform, spatial econometric models, and eigenvector spatial filtering (ESF)).
+Failing to account for destination accessibility when spatial structure effects are present generally biases origin-specific distance-decay parameter estimates upwards for accessible origins and downward for inaccessible origin.
+The strength of this bias is shown by Fotheringham (1984) to depend on the strength of two relationships: that between the volume of flows and destination accessibility and that between destination accessibility and distance from the origin to each destination.
+It is this bias that causes the unintuitive spatial patterns observed in origin-specific distance-decay parameter estimates, such as positive estimates for the most accessible origins (Fotheringham, 1981) -->
+
+<!-- So Aij is a relevant explanatory variable that can increase the accuracy of a Tij model!
+There's a relationship there.
+It makes me wonder if those spatial effects would be there if you used a constrained accessibility measure.
+Maybe using Aij (Hansen-type accessibility) doesn't completely solve this issue.
+-->

### 2024-12-30 TRB/TRB-pres.qmd
V1 of the TRB presentation

@@
 ---
+```{r}
+library(here)
+library(knitr)
+library(glue)
+```
 
@@
 
-## Skip over
-(plot of how accessibility research interacts with gravity modelling)
+Plot of how accessibility research (cites Hansen) interacts with gravity modelling (cites Wilson):
+```{r}
+include_graphics(glue(here(), "/data-raw/Han_Wilson_bib_coupling_plot.png"))
+```
 

### 2025-01-03 TRB/Musings-a-draft-pres.qmd
Cleaned up repo - latest version of the manuscript draft.

--- 2024-12-30 TRB/TRB-pres.qmd
@@
 ---
-title: "Family of accessibility measures"
+title: "Some musings Family of accessibility measures"
 format: revealjs
@@
 ```
-
-## BIGGER QUESTIONS:
-- why and/or when did accessibility diverge from spatial interaction modelling?
-- examples of accessibility measures in the literature that fit into our 4 cases?
-
-## PRESENTATION'S AGENDA:
-Accessibility’s relationship can be made into one of equality -- with balancing factors! Accessibility shares spatial interaction modeling’s roots.
-3.	A slide with 0) no constrained (Hansen, what we do now), 1) total constrained, 2) singly-constrained, 3) doubly constrained formulations
-4.	Calculated simple example of the 3 types of accessibility. Three visualisations: 1st are the Vij matrices, 2nd is a small grid aggregation, and 3rd is bigger grid aggregation.
-5.	Conclusions: 
-a.	With balancing factors values are interpretable: each spatial unit is ‘potential opportunities that can reached’ (of all the opportunities)
-b.	No treatment is needed to address MUAP across grid aggregations (?)
 
@@
 
-\noindent Where $A_i$ and $B_j$ are:
+\noindent Where $A_i$ (A_i may not make sense..?) and $B_j$ are:
 $$
-A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
-B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
+A_i = \frac{O_j}{\sum_j O_j f(c_{ij})}\\
+B_j = \frac{P_i}{\sum_i P_i f(c_{ij})}\\
 $$
@@
 
-\noindent Where $A_i$ and $B_j$ are:
+\noindent Where $A_i$ and $B_j$ are (this may not make sense..)
 $$
@@
 
-We'd report i=1 has 148, i=2 has 248, and i=3 has 172 accessibility. (implicitly, this is saying j=1 123,...)
-
+We'd report i=1 has 148, i=2 has 248, and i=3 has 172 accessibility.
 The sum of values at each origin or destination or overall, do not add up to anything we know.
@@
 ```{r}
-V_orig <- (A_i) * (1/C)* (matrix(rep(W_i, 3), 3, 3))  # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))
+V_orig <- (A_i) * (V_unc)  # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))
 
@@
 
-And as we can observe, the $P_i$ values are equal to the known populations, while $O_j$ is not.
-
-NOTE: This is also spatial availability -- except I switched $O_j$ for $P_i$. This is also related to retail modeling
+And as we can observe, the $P_i$ values are equal to the known populations, while $O_j$ is not. (well.. this doesn't really happen, this A_i may not make sense)
 
@@
 ```
-
-## Case 5: Hybrid / partial constraints
-
-## next steps
-- grocery stores for example; maybe 2 grids?
-- bibliometric analysis 
-- - draft of this end of jan 2025
-
-- (future) intervening opportunities ? radiation model - deep gravity
-- (future) accessibility pyramids 
-- (future) MAUP addressed
-- (future) comparing cities

### 2025-01-03 paper/manuscript.qmd
Cleaned up repo - latest version of the manuscript draft.

@@
 ---
-title: "A family of accessibility measures"
+title: "A family of accessibility measures (bringing balance back?)"
 format: pdf
@@
 library(flextable)
+library(knitr)
+library(glue)
+library(here)
 # library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
@@
 
-\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
+\noindent Where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
 Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
@@
 In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
-Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
+Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
 From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
@@
 
-And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974a; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
 This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.
@@
 $$
-V_{ij} = k W_j^{(2)} f(c_{ij})
+V_{ij} = k W_j^{(2)} f(c_{ij}) = k O_j f(c_{ij})
 $${#eq-access}
@@
 Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
-Accessibility is also most often presented as a locational measure, the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone.
-
-Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel, and @eq-constraint3-gravitymodel) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
+Notably, the mass vectors $W_i^{(1)}$ and $W_j^{(2)}$ are replaced by $P_i$ and $O_j$, representing the weight of the origin (population) and destination (opportunities), the weights common in accessibility literature.
+As accessibility is the sum of opportunities for a $j$ from $i$, when the mass of the origin ($P_i$) is relevant, it enters @eq-access through the $k$ constant.
+Accessibility is also most often presented as a locational measure, so the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone:
+
+$$
+V_{i} = k \sum_j O_j f(c_{ij})
+$${#eq-access-sum}
+
+Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint0-access,  @eq-constraint1-access, and @eq-constraint2-access) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
 
@@
 
-All four <!--five?--> cases are differentiated like:
-- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
-- The total constrained case: only @eq-constraint0-gravitymodel hold
-- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
-- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
-- HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case...
+All four cases are differentiated like:
+- The unconstrained case: neither @eq-constraint0-access, @eq-constraint1-access, or @eq-constraint2-access hold;
+- The total constrained case: only @eq-constraint0-access hold
+- The singly-constrained case: either @eq-constraint1-access holds (origin-constrained) <!-- production--> or @eq-constraint2-access holds (destination-constrained)<!-- attraction-->; 
+- The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously <!--production-attraction-->
+
+<!-- - HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case...
 and doubly constrained case)-->
 
-We can use the same trip matrix, but reformulate its meaning.
-Let's say $W_i^{(1)}$ is population at $i$ who seek opportunities, lets recast this variable as $P_i$.
-Then let's say $W_j^{(2)}$ are the opportunities being sought, recast to $O_j$.
+Like gravity model problem, we can solve these cases using spatial weights matrix and travel cost matrix, but with alternative meanings.
+$P_i$ (our $W_i^{(1)}$) is population at $i$ who seek opportunities.
+$O_j$ (our $W_j^{(2)}$) are the opportunities being sought.
 For this simple example, we can say that the opportunities are the population's jobs.
-In our region, each person has a job -- but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost ($C$ and associated travel propensity $1/C$).
-(let's pretend our 'known_trip_matrix' is unknown).
-
-```{r}
-known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
+In our region, each person has a job - but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost $c_{ij}$.
+We assume the the travel impedance ($f(c_{ij})$) is the associated travel propensity of $f(c_{ij}) = 1/C$.
+
+The following is the known accessibility matrix.
+The $V_{ij}$, the number of jobs spatially accessible at $i$ from destination $j$.
+The TOTALs are the $V_i$ and $V_j$, the total number of jobs spatially available at $i$ and the number of jobs spatially available from $j$.
+```{r}
+known_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
 #unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
-W_i <- known_trip_matrix |> rowSums()
-W_j <- known_trip_matrix |> colSums()
-T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA)
-knowns_longer <- cbind(known_trip_matrix,W_i)
-knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T))
+P_i <- known_matrix |> rowSums()
+O_j <- known_matrix |> colSums()
+V <- ifelse(sum(P_i)==sum(O_j), sum(O_j),NA)
+knowns_longer <- cbind(known_matrix,P_i)
+knowns_longer <- rbind(knowns_longer,cbind(t(O_j),T))
 
@@
 
-- No proportionality constant, $k$ is set to 1, so no constraints hold, making the interpretation of the units meaningless so the value's magnitude can only be interpreted:
-$$
-V_{ij} = k O_j f(c_{ij})
-$$
-
-- In practice, $V_i$ is also calculated for each $i$. When thinking about the interpretation of the value, how many opportunities can be visited for each $i$, the $P_i$ is also set to 1:
-$$
+This accessibility is standard practice.
+There is no explicit consideration of a proportionality constant ($k$).
+We can presume it is still set to 1 as @stewartPrinciples1947.
+No constraints hold, making the interpretation of the units of $V_{ij}$ simply some magnitude of $O_j$ weighted by some $f(c_ij)$, and $V_{i}$ the sum of these values for each $j$.
+$V_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O$ and/or population $P$.
+$$
+V_{ij} = O_j f(c_{ij})\\ 
 V_{i} = \sum_j O_j f(c_{ij})
 $$
-These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).
-
-Looks like Hansen, but multiplied by P_i.
+
+Each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin: 
+```{r}
+O_P_unc <- t(matrix(rep(O_j, 3), 3, 3))
+O_P_unc_longer <- O_P_unc
+rownames(O_P_unc_longer) <- c("1","2","3") 
+colnames(O_P_unc_longer) <- c("1","2","3")
+O_P_unc_longer
+```
+
+The resulting matrix of $V_{ij}$ along with the $V_i$ and $V_j$ totals for this example:
+```{r}
+V_unc <- O_P_unc / C
+
+P_i_unc <- V_unc |> rowSums()
+O_j_unc <- V_unc |> colSums()
+V_TOTAL_unc <- ifelse(sum(P_i_unc|>round())==sum(O_j_unc|>round()), sum(V_unc), NA)
+V_unc_longer <- cbind(V_unc,P_i_unc)
+V_unc_longer <- rbind(V_unc_longer, cbind(t(O_j_unc),V_TOTAL_unc))
+
+rownames(V_unc_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_unc_longer) <- c("1","2","3", "TOTAL")
+
+V_unc_longer |> round(digits = 2) #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
+We'd report origin 1 as having 148 accessibility (i.e., $V_{i=1} = 148$), origin 2 as 248 accessibility, and origin 3 as 172 accessibility.
+The total accessibility in the region is 568 ($\sum_i V_{i} = 568$), not equal to any meaningful value.
+This value is not often reported in the accessibility literature for this reason.
+One could also report the accessibility values that destinations offers ($V_{j}$), namely destination 1 offers 123, destination 2 offer 291, and destination 3 offers 154.
+However, these values are also not reported in the literature currently...
+as they aren't constrained to match up to the number of opportunities at each j (i.e., 160, 437 and 193).
+Assuming $k=1$, $V_i$ and $V_j$ represent magnitudes of accessibility concepts, but in an unconstrained way.
+The population matrix is no incorporated in the $V_{ij}$ calculation, so $V_i$ cannot be directly compared to population.
+From a different perspective, $V_j$ values also cannot be compared to known $O_j$ values as the equation unconstrained.
+
+OF NOTE: in the context of spatial interaction modelling, the unconstrained model is not often useful.
+While the resulting $T_{ij}$ values are in units of 'trips' - the assumption of $k$ equaling 1 is practically meaningless.
+
+Accessibility, or the _potential_ for interaction is different than interaction ($V_{ij}$ vs.
+$T_{ij}$), namely since the number of opportunities that are accessible at an origin from a destination cannot be known in the same way as a trip from an origin to a destination.
+However, we would argue that getting closer to knowning this number and planning for it - is what accessibility research should strive to achieve.
+<!-- yeah?
+-->
 
@@
 
-This needs a balancing factor, $k$ is replaced with $K$:
-$$
-V_{ij}' = K P_i O_j f(c_{ij})\\
-V_{i}' = K P_i \sum_j O_j f(c_{ij})
-$$
-
-This approach is not seen in accessibility literature...
+In the total constrained case, $k$ is replaced with $K$ (@eq-access satisfying only constraint @eq-constraint0-access) as follows for $V_{ij}$ and $V_{i}$:
+$$
+V_{ij}' = K O_j f(c_{ij})\\
+V_{i}' = K\sum_j O_j f(c_{ij})
+$$
+
+Solved for our example, $K$ is equal to the total accessibility $V$, divided by the sum of the unconstrained accessibility:
+$$
+K = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
+K = \frac{V}{\frac{O_1}{c_{1,1}}+\frac{O_1}{c_{2,1}}+\frac{O_1}{c_{3,1}}+\frac{O_2}{c_{1,2}} +... \frac{O_3}{c_{3,3}}
+}\\
+K = \frac{790}{568.4} \\
+K = 1.389866
+$$
+
+```{r}
+K <- V/sum(V_unc)
+V_tot <- K * V_unc
+
+P_i_tot <- V_tot |> rowSums()
+D_j_tot <- V_tot |> colSums()
+V_TOTAL_tot <- ifelse(sum(P_i_tot)==sum(D_j_tot), sum(V_tot), NA)
+V_tot_longer <- cbind(V_tot,P_i_tot)
+V_tot_longer <- rbind(V_tot_longer, cbind(t(D_j_tot),V_TOTAL_tot))
+
+rownames(V_tot_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_tot_longer) <- c("1","2","3", "TOTAL")
+
+V_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
+While we can see that the 'number of potentially reachable opportunities' `r V` results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model).
+This approach is not yet seen in accessibility literature, but could be an intelligible way to compare accessibility values ($V_i$ or $V_j$) without considering the population matrix.
 
@@
 Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
-For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
-$$
-V_{ij}''^{(1)} = A_i O_j f(c_{ij})\\
-V_{ij}''^{(2)} = B_j O_j f(c_{ij})\\
-
+For the origin- and destination- constrained versions, $k$ is replaced with $A_{ij}$ and $B_{ij}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):
+
+$$
+V_{ij}''^{(1)} = A_{ij} O_j f(c_{ij})\\
+V_{ij}''^{(2)} = B_{ij} O_j f(c_{ij})\\
 $${#eq-access-singlyconstrained-origin}
 
-\noindent Where $A_i$ and $B_j$ are:
-$$
-A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
-B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
-$$
-
-In terms of origin-constrained -- ..? .
-
-Regarding destination-constrained, has it's origins are in retail models.
-e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction.
-Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
-Spatial availability is explicitly also destination-constrained [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024].
-Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general []
-
-For instance, the most recent example spatial availability is formulated like so:
-$$
-V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
-$$
-Where the $F^p_{ij}$ Population balancing factor is:
-$$
-F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}
-$$
-And the $F^c_{ij}$ travel cost balancing factor:
-$$
-F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}
-$$
-It can be reformulated to fit into the singly-constrained (origin case) gravity model as the denominators in both those balancing factors cancel out:
-$$
-V_{ij}'' = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}\\
-
-V_{ij}''^{(1)} =O_j \frac{1}{\sum_j O_j f(c_{ij})} P_i  f(c_{ij})\\
-$$
+\noindent Where $A_{ij}$ and $B_{ij}$ are similar to the balancing factors of the gravity model for the singly-constrained cases, though with an addition to the numerator to maintain the associated enforcing constraints:
+$$
+A_{ij} = \frac{O_j}{\sum_j B_{ij} O_j f(c_{ij})}\\
+B_{ij} = \frac{P_i}{\sum_i A_{ij} P_i f(c_{ij})}\\
+$$
+In solving the origin-constrained $V_{ij}^{''(1)}$ case (e.g., @eq-constraint1-access of $\sum_j V_{ij} = P_i$ is enforced), $B_{ij} = 1$.
+In solving the destination-constrained $V_{ij}^{''(2)}$ case (e.g., @eq-constraint2-access of $\sum_i V_{ij} = O_j$ is enforced), $A_{ij} = 1$.
+
+Solving for the destination-constrained case $V_{ij}^{''(2)}$ may be helpful in an example where the number of people seeking opportunities is known ($P_i$), along with the number of opportunities $O_j$ offered by a destination as in unconstrained and total accessibility.
+An example could be the number of people at each residence seeking hospital-beds (origin) and hospital-beds (opportunities) at each healthcare location (destination) to calculate the opportunity access at each origin.
+Another case could be calculating the opportunity access at each origin using the quantity of goods available to purchase (opportunities) at each commercial location (destination).
+
+From our review, related versions of this defined destination-constrained accessibility are formulaically similar to retail accessibility models e.g., decision model for shops to optimally spatially locate to maximize their potential consumer interaction.
+Starting with reilly's law of gravitational retail [@reilly1929methods], formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
+Other literature also demonstrates  similar _competition_ considerations through its analytical formulation; for instance, in the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, and competition consideration expressed as a per population rate particularly in healthcare and job opportunity accessibility analysis [@shen1998], and 2FCA methods in general [@soukhovIntroducingSpatialAvailability2023].
+Most of these works do not explicitly focus on balancing accessibility units; rather, they center on _competition_ considerations.
+We suspect this may be why they have been used in case studies examining access to opportunities that are inherently competitive or where the concept of supply and demand resonates.
+It was only until the recent work of [@soukhovIntroducingSpatialAvailability2023], where the units of accessibility given competition, e.g., the spatial availability, was introduced with the focus on both competition and balanced units.
+NOTE: See Appendix where I demonstrate how this destination-constrained case is equivalent to spatial availability
+
+Returning to our simple example, the destination-constrained singly-constrained case would yield the following $B_{ij}$ values:
+$$
+B_{ij} = \frac{P_{i}}{\sum_i P_i f(c_{ij})}\\
+B_{1,1} = \frac{P_{i}}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
+...\\
+B_{1,1} = 160 *\frac{1}{160/2+450/15+180/5} = 1.09589 \\ 
+B_{2,1} = 450 *\frac{1}{160/2+450/15+180/5} =3.082192 \\
+B_{3,1} = 180 * \frac{1}{160/2+450/15+180/5} =1.232877 \\
+...\\
+B_{1,2} = 160 *\frac{1}{160/15+450/2+180/10} =0.630749 \\
+B_{2,2} = 450 *\frac{1}{160/15+450/2+180/10} =1.773982 \\
+B_{3,2} = 180 * \frac{1}{160/15+450/2+180/10} =0.7095926 \\
+...\\
+B_{1,3} = 160 *\frac{1}{160/5+450/10+180/2} = 0.9580838\\
+B_{2,3} = 450 *\frac{1}{160/5+450/10+180/2} = 2.694611  \\
+B_{3,3} = 180 *\frac{1}{160/5+450/10+180/2} = 1.077844 \\
+$$
+```{r}
+B_ij <- matrix(P_i) %*% (1/(t((matrix(P_i))) %*% (1/C))) 
+B_ij_longer <- B_ij
+rownames(B_ij_longer) <- c("1","2","3") 
+colnames(B_ij_longer) <- c("1","2","3") 
+B_ij_longer 
+```
+
+$V_{ij}$ values:
+```{r}
+V_dest <- (B_ij) *(V_unc) # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))
+
+P_i_dest <- V_dest |> rowSums()
+O_j_dest <- V_dest |> colSums()
+
+V_TOTAL_dest <- ifelse(sum(P_i_dest)==sum(O_j_dest), sum(V_dest), NA)
+V_dest_longer <- cbind(V_dest,P_i_dest)
+V_dest_longer <- rbind(V_dest_longer, cbind(t(O_j_dest),V_TOTAL_dest))
+
+rownames(V_dest_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_dest_longer) <- c("1","2","3", "TOTAL")
+
+V_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
+The $O_j$ values are equal to the known opportunities, whereas the $P_i$ values are not.
+The total accessibility in the region is maintained at 790.
+Each $V_{ij}$ can be interpreted as the number of opportunities accessible from each destination by anyone in the region all weighted by travel cost of course.
 
@@
 ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
+<!-- this done'st work.. for now!-->
+```{r}
+doubly_constrained_gravity_model <- function(
+    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
+  P,  # Origin production weights vector (POPULATION)
+  O,  # Destination attraction weights vector (OPPORTUNITIES)
+  impedance_matrix,  # the resulting travel cost impedance matrix
+  error_threshold = 0.001,  # Error threshold for stopping condition
+  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
+) {
+
+  # Normalize O and D
+  sum_P <- sum(P)
+  sum_O <- sum(O)
+  cat("Checking population, opportunity balancing:\n")
+  cat("Population: ", sum_P, "\n")
+  cat("Opportunities: ", sum_O, "\n")
+
+  if (sum_P != sum_O) {
+    cat("Population and opportunities do not balance, scaling to match.\n")
+    if (sum_P < sum_O) {
+      P <- P * (sum_O / sum_P)
+    } else {
+      O <- O * (sum_P / sum_O)
+    }
+  } else {
+    cat("Population, opportunities balancing OK.\n")
+  }
+
+  n <- length(P)  # Number of i
+  V <- sum(P)  # Total access
+  Ai <- rep(1, n)  # Initialize Ai
+  Bj <- rep(1, n)  # Initialize Bj
+
+  previous_error <- Inf
+  iteration_count <- 0
+  stop_reason <- ""
+
+  while (TRUE) {
+    iteration_count <- iteration_count + 1
+
+    # Update Ai
+    for (i in 1:n) {
+      Ai[i] <- 1 / (sum(Bj * O * impedance_matrix[i, ]) + 1e-9)
+    }
+
+    # Update Bj
+    Bj_new <- rep(1, n)
+    for (j in 1:n) {
+      Bj_new[j] <- 1 / (sum(Ai * P * impedance_matrix[, j]) + 1e-9)
+    }
+
+    # Calculate Vij
+    #Vij <- outer(Ai * P, Bj_new * O) * impedance_matrix
+    Vij <- Bj_new * Ai * O * impedance_matrix
+
+    # Print Ai and Bj for this iteration
+    cat("\nIteration:", iteration_count, "\n")
+    cat("Ai:", round(Ai, 3), "\n")
+    cat("Bj:", round(Bj_new, 3), "\n")
+
+    # Calculate error
+    error <- (sum(abs(P - rowSums(Vij))) + sum(abs(O - colSums(Vij)))) / V
+
+    # Calculate change in error
+    error_change <- abs(previous_error - error)
+
+    # Check stopping conditions
+    if (error < error_threshold) {
+      stop_reason <- "Error threshold met"
+      break
+    } else if (error_change < improvement_threshold) {
+      stop_reason <- "Slow improvement"
+      break
+    }
+
+    previous_error <- error
+    Bj <- Bj_new
+  }
+
+  # report stopping condition and final error threshold
+  cat("\nStopping Condition: ", stop_reason, "\n")
+  cat(sprintf("Final Error: %.3f%%\n", error * 100))
+
+  return(list(Ai = Ai, Bj = Bj, Vij = Vij))
+}
+
+dc_results <- doubly_constrained_gravity_model(P=P_i, O=O_j, impedance_matrix = 1/C)
+dc_results$Vij
+```
+
+The $A_i$ and $B_j$ matrices:
+```{r}
+A_i_doubly <- dc_results$Ai
+B_j_doubly <- dc_results$Bj
+
+A_i_doubly |> matrix()
+B_j_doubly |> matrix() |> t()
+```
+
+The trip matrix (notice, all summed up opportunities and populations are equivalent to our knowns)
+```{r}
+V_doubly <- dc_results$Vij
+
+P_i_doubly <- V_doubly |> rowSums()
+O_j_doubly <- V_doubly |> colSums()
+V_TOTAL_dest <- ifelse(sum(P_i_doubly)==sum(O_j_doubly), sum(V_doubly), NA)
+V_doubly_longer <- cbind(V_doubly,P_i_dest)
+V_doubly_longer <- rbind(V_doubly_longer, cbind(t(O_j_doubly),V_TOTAL_dest))
+
+rownames(V_doubly_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_doubly_longer) <- c("1","2","3", "TOTAL")
+
+V_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
 
@@
 
+# APPENDIX (some musings...)
+
+## Singly-constrained case is spatial availability
+
+The destination-constrained case is also spatial availability, see below!
+As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{ij}$ is removed).
+
+Spatial availability is formulated like so:
+$$
+V_{ij}''^{(2)}  = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
+$$
+It turns out that the denominators of the Population balancing factor $F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}$ and Travel cost balancing factor $F^c_{ij} = F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}$ cancel out, so spatial availability can be expressed as:
+$$
+V_{ij}''^{(2)} =O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
+$$
+
+Reformatted in the general family of accessibility measure form, it looks like:
+$$
+V_{ij}''^{(2)} =k O_j f(c_{ij})\\
+V_{ij}''^{(2)} =\frac{ P_i}{\sum_i P_i f(c_{ij})} O_j f(c_{ij})\\
+V_{ij}''^{(2)} =  O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} = O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
+$$
+## Checking singly-constrained case against spatial availability
+```{r}
+sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){
+
+  o_id <- rlang::enquo(o_id)
+  d_id <- rlang::enquo(d_id)
+  pop <- rlang::enquo(pop)
+  opp <- rlang::enquo(opp)
+  r <- rlang::enquo(r)
+  f <- rlang::enquo(f)
+
+  sum_pop <- x %>%
+    dplyr::distinct(!!o_id,
+                    .keep_all = TRUE) %>%
+    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
+    dplyr::pull(sum_pop) %>%
+    sum()
+
+  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop
+
+  sum_impedance <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_impedance = sum(!!f))
+
+  x <- x %>%
+    dplyr::left_join(sum_impedance,
+                     by = rlang::as_name(d_id))
+
+  f_c <- dplyr::pull(x, !!f) / x$sum_impedance
+
+  x$f_c <- f_c
+  x$f_p <- f_p
+
+  sum_pa <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_pa= sum(f_p * f_c))
+
+  x <- x %>%
+    dplyr::left_join(sum_pa,
+                     by = rlang::as_name(d_id))
+  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)
+
+  dplyr::pull(x, !!opp) * f_t
+}
+```
+
+```{r}
+
+df <- data.frame(i = c(1,1,1,2,2,2,3,3,3), 
+                 j= c(1,2,3,1,2,3,1,2,3),
+                 f = c(1/C[1,], 1/C[2,], 1/C[3,]),
+                 pop = c(P_i[1], P_i[1], P_i[1], P_i[2], P_i[2], P_i[2], P_i[3], P_i[3], P_i[3]),
+                 opp = c(O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3]),
+                 catch = c(1,1,1,1,1,1,1,1,1))
+
+calced <- sp_avail(df,
+                         o_id = i,
+                         d_id = j,
+                         pop = pop,
+                         opp = opp,
+                         r = catch,
+                         f = f)
+df$V_ij <- calced
+
+df |> group_by(j) |> summarise(V_j = sum(V_ij))
+```
+
+## Destination- (or ATTRACTION) constrained is the only singly-constrained case... population- (or PRODUCTION) constrained doesn't make sense!
+
+Then the origin-constrained (singly-constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
+Using this method, V_i values equal P_i values...?
+while O_j values are unconstrained.
+This case may be helpful in an example where the number of opportunities available to an origin is known and fixed.
+This could be a normative minimum threshold, such as 100 opportunities at each origin.
+
+Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{ij}$ values:
+$$
+A_{ij} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\
+A_{1,1} = \frac{O_{j}}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\
+...\\
+A_{1,1} = 160 *\frac{1}{160/2+437/15+193/5} = 1.083032 \\ 
+A_{2,1} = 160 *\frac{1}{160/15+437/2+193/10} =0.6439496 \\
+A_{3,1} = 160 * \frac{1}{160/5+437/10+193/2} =0.9291521 \\
+...\\
+A_{1,2} = 437 *\frac{1}{160/2+437/15+193/5} =2.958032 \\
+A_{2,2} = 437 *\frac{1}{160/15+437/2+193/10} =1.7587872 \\
+A_{3,2} = 437 * \frac{1}{160/5+437/10+193/2} =2.5377468 \\
+...\\
+A_{1,3} = 193 *\frac{1}{160/2+437/15+193/5} = 1.306408 \\
+A_{2,3} = 193 *\frac{1}{160/15+437/2+193/10} = 0.7767642 \\
+A_{3,3} = 193 *\frac{1}{160/5+437/10+193/2} = 1.1207898 \\
+$$
+
+```{r}
+A_ij <- matrix(O_j) %*% (1/(t((matrix(O_j))) %*% (1/C))) |> t()
+A_ij_longer <- A_ij
+rownames(A_ij_longer) <- c("1","2","3") 
+colnames(A_ij_longer) <- c("1","2","3") 
+A_ij_longer 
+```
+Starting off with this opportunity matrix, namely - at i=1, only a fixed number of opportunities are accessible to this origin. same for i=2, i=3.
+```{r}
+(matrix(rep(O_j, 3), 3, 3))
+```
+
+$V_{ij}$ values:
+```{r}
+V_orig <- (A_ij) * t(V_unc) # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.
+
+P_i_orig <- V_orig |> rowSums()
+O_j_orig<- V_orig |> colSums()
+V_TOTAL_orig <- ifelse(sum(P_i_orig)==sum(O_j_orig), sum(V_orig), NA)
+V_orig_longer <- cbind(V_orig,P_i_orig)
+V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig))
+
+rownames(V_orig_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_orig_longer) <- c("1","2","3", "TOTAL")
+
+V_orig_longer #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
 # References
+
+<!-- 
+## next steps
+- grocery stores for example; maybe 2 grids?
+- bibliometric analysis 
+- - draft of this end of jan 2025
+
+- (future) intervening opportunities ? radiation model - deep gravity
+- (future) accessibility pyramids 
+- (future) MAUP addressed
+- (future) comparing cities 
+-->
 

### 2025-01-03 paper/manuscript.qmd
Added the TRB presentation

@@
 <!-- 
+## QUESTIONS:
+- what does k=1 mean? What does assuming k=1, so the units can be kept mean for the model? Is it meaningless?
+- not sure if the singly-constrained (PRODUCTION case) makes sense.. It sort of does..? I believe this needs to be figured out before the doubly constrained case. Pi needs to be integrated somehow. 
+- doubly constrained still needs to be solved.
+
 ## next steps

### 2025-01-25 paper/manuscript.qmd
Latest state of the draft - need to update from the last discussion with AP (Jan 24)

@@
 ---
-title: "A family of accessibility measures (bringing balance back?)"
+title: "Accessibility, constraints and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
 format: pdf
@@
 -->
-In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction-based roots.
+In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction roots.
 It proposes to do so by mapping accessibility onto Wilson's family of spatial interactions models [@wilson1971].
-In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its gravity modelling siblings, and spatial interaction history.
+In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its spatial interaction history.
 To this end, we begin by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to its split into accessibility research from @hansen1959 to today.
-After this, we derive different types of accessibility measures, following the family of gravity models namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained.
+After this, we derive different types of accessibility measures, following the family of "spatial interaction models" (SIM) namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained.
 Then, we will provide empirical examples of the types of measures.
@@
 ## Gravity-based accessibility's cousin: Wilson's spatial interaction model
-
-Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used [@ortuzar_2011_modelling;@wilson1971].
-However, in some ways the gravity model and spatial interaction continued to evolve independently from @eq-accessibility-general, the former being developed into formalized transportation demand modelling practice to 'predict-and-provide', and the later taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners.
-From the perspective of gravity-based accessibility measure, this paper's focus is on the proportionality constant of the spatial interaction model, which takes the following form after Wilson [-@wilson1971]:
-
+<!--SUMMARY: State how accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models.
+Introduce SIM's formulation.
+Emphasis the interpretation that comes from $k$.
+Emphasis how while SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average.
+-->
+Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971].
+However, in many ways these spatial interaction models (AKA gravity models) continued to evolve independently from accessibility (@eq-accessibility-general) as will be discussed next section.
+Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM.
+This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs.
+This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.
+To begin, the SIM as defined in Wilson [-@wilson1971], takes the following form: 
 $$
@@
 In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
-It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the gravity model (@eq-phys-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
+It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the SIM (@eq-phys-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
 Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
 
-Traditionally, the development of the gravity model has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
-But instead of relying on the heuristic Newtonian gravity analogy, entropy maximisation has emerged as a popular model building approach taken in the spatial interaction studies to generate the gravity model [@ortuzar_2011_modelling].
-Namely, it is mathematically rigorousness and offers a sensible interpretation of the resulting model as a statistical averaging procedure [@wilson1971;@seniorGravityModellingEntropy1979].
-As such, @eq-phys-gravity-model can be derived to account for additional information to include constraints that reflect known information about the system and is mathematically rigorous.
-The interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$.
-These pieces of information can constrain $T_{ij}$ as follows.
-
-The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!-- Along a similar region-wide perspective, the total flows multiplied by their travel cost in the region is constrained to equal $C$:-->
+Traditionally, the development of the SIM has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
+But instead of relying on the heuristic Newtonian gravity analogy (e.g., some interaction between a mass at $i$ and a mass at $j$ seperated by some distance), entropy maximisation has emerged and continues to be a popular model building approach to generate the SIM [@ortuzar_2011_modelling].
+
+Namely, the entropy maximisation approach achieves stable results as a statistical average that represents the population.
+The approach works with individual interaction, assesses their probability of making a particular journey given given zonal and regional constraints, and the resulting interaction from @eq-phys-gravity-model is a statistical average [@wilson1971;@seniorGravityModellingEntropy1979].
+For instance, the interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$ can be integrated into zonal constraints as follows.
+
+The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel:-->
 $$
@@
 The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967).
-The basis of the approach is to accept that there are macro, meso, and micro states.
-In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
+The approach is based on partition functions from statistical mechanics (the physics of gases), not Newtonian gravitation.
+In short, the approach accepts that there are macro, meso, and micro states.
+In the case of 'trips' being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
 Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
-From this approach, the gravity model can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
-There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.
-
-Using these constraints, the gravity model @eq-phys-gravity-model is distinguished into four general cases.
-These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the gravity model in trip distribution modelling to this day [@ortuzar_2011_modelling]:
+From this approach, the SIM can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
+There is then a most probable distribution of trips between zones (meso-states) that satisfies all the macro-state constraints.
+
+Using these constraints, the SIM @eq-phys-gravity-model is distinguished into four general cases.
+These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling]:
 
 - The unconstrained case: neither @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
-- The total constrained case: only @eq-constraint0-gravitymodel hold
+- The total constrained case: only @eq-constraint0-gravitymodel hold <!--NOTE: not sure if this is consistent with entropy-maximisation?-->
 - The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
@@
 $${#eq-bf-B-gravitymodel}
-<!--PRETTY SURE YOU DON'T GET K FROM ENTROPY MAXIMAISATION...-->
+<!--NOTE: pretty sure K is not consistent with entropy-maximisation?-->
 $$
@@
 
-## Accessibilities' development seperate from the family of gravity models
-
-Upon detailing the gravity model, we can now discuss how accessibility has been developed within that field and how it has not.
-Gravity modellers (interaction variable being trips, a model of spatial interaction - mobility) continued to use accessibility within transportation demand modelling.
-As succinctly summarized in @morrisAccessibilityIndicatorsTransport1979, "Mobility and accessibility together influence an individual’s capacity to travel in daily life".
-However, the gravity model literature that focuses on Hansen's accessibility is relatively scarce.
-To capture this sentiment, see @fig-Fig1 which demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971.
+## Background on accessibilities' distinct development away from spatial interaction models
+<!--COME-BACK HERE-->
+<!--SUMMARY: Accessibility developed separate from SIM.
+
+SIM focuses on trips and was institutionalized to develop the US Interstate highway systems.
+Many guidelines and practical training was done - a lot of funding.
+SIM development has integrated accessibility in a few ways: 1) within the SIM itself as a balancing factor, 2) as an additional independent variable in the SIM, 3) SIM and HAM in parallel in the context of location allocation problems and within the steps of transportation demand modelling (trip distribution or trip generation).
+
+Accessibility research has separated SIM almost entirely.
+When Wilson is cited: 1) typically when using a different impedance function, 2) when using the SIM to predict trip flows for accessibility input or vice versa, 3) a few exceptional papers that interpret the doubly-constraints as an inverse HAM in the context of competitive accessibility, and 4) it was only until recently, that a singly-constrain was introduced in Soukhov et al.
+2023.
+-->
+
+The institutionalization of research on spatial interaction modelling was fueled by advancements in computer technologies and the substantive and unprecedented Federal-Aid Highway Act of 1956, which authorized $24.8 billion in spending at a 90% federal share from between 1957 to 1969 to begin and complete the US Interstate Highway System [@weinerUrbanTransportationPlanning2016].
+This ambitious project went on to shape transportation planning practice globally [].
+Within the same time period, accessibility measures were not applied practically at such a large scale.
+Instead, accessibility measures were taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners; remaining more in theory than in practice.
+However, due to growing critiques about traditional transportation modelling practice focus on  _mobility_, calls for a shift to accessibility are intensifying [@handy2020] and accessibility planning is beginning to be present in practice [] <!-- citation on where they are applied in practice?
+US guidelines.
+Netherlands.
+-->
+
+Despite the theoretical and observed reality that both mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature.
+To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971.
 @fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].
@@
 
-Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the gravity model (mobility).
-
-For instance, @battyMethodResiduesUrban1976 states that the relationship between the gravity model and Hansen's original accessibility-potential model may be worth exploring...
-There is a consistent acknowledgement that these two models are related.
-Specifically, that the Hansen-type accessibility model is embedded within the balancing factors [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
-This is a "common sense" approach at integrating accessibility [@morrisAccessibilityIndicatorsTransport1979].
-
-Though there are other approaches to consider accessibility with spatial interaction modelling beyond Hansen founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979].
-For instance, others explore entropy maximising solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].
+Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the SIM or interpreting HAM within the SIM.
+For instance, @battyMethodResiduesUrban1976 states that the relationship between the SIM and HAM may be worth exploring.
+There is also a consistent acknowledgement in the literature that HAM and SIM are related.
+Specifically, that HAM are within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
+Recognizing the balancing factors as HAM is a "common sense" approach at integrating accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979].
+
+There are other approaches to consider accessibility with SIM beyond HAM which can be founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979].
+For instance, others explore entropy maximizing solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].
 
 Some work on location-allocation models also used both Hansen and Wilson.
-For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained gravity model, and then defines one using Hansen-type accessibility as the objective function subject.
+For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained SIM, and then defines one using Hansen-type accessibility as the objective function subject.
 
@@
 Without the consideration for accessibility explicitly, unexplained spatial pattern remains.
-This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in gravity models.
+This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in SIMs.
 [@oshanSpatialStructureDebate2021].
-Implicitly, we gather that at this point: it was understood balancing factors in the gravity model could not be understood as capturing accessibility.
+Implicitly, we gather that at this point: it was understood balancing factors in the SIM could not be understood as capturing accessibility.
 We also gather that in this moment of growing complexity of spatial interaction modelling, accessibility branched off and sprouted in the medium of budding GIS technology and the emergence of space-time dimensions [@millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
@@
 
-Some papers also use gravity models to estimate trip or some other interaction flows alongside accessibility indicators [@clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the gravity model, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
+Some papers also use SIMs to estimate trip or some other interaction flows alongside accessibility indicators [@clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
 
@@
 
-<!-- REVIEW CSV OF PAPERS-->
+<!-- NOTE: you can validate SIM e.g., how well does the model predict a known Tij matrix? but what is a known Vij matrix, potential accessible opportunities at i for j? Harder to say. -->
+
+<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. -->
 
@@
 
-## A family of accessibility measures
-
-Instead of the interaction variable being in the unit of trips, it is in the unit of potential interaction with opportunities.
-This focus is the preoccupation of accessibility measures.
-The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of gravity models, however with different interpretation.
+## Defining accessibility constraints (a potential spatial interaction model?)
+
+Instead of an interaction variable as in the SIM (e.g., trips, spending, sales), we focus on _potential_ interaction (e.g., accessible opportunities).
+Potential interaction is the preoccupation of accessibility measures.
+The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of SIMs, however with different interpretation.
 In this section, we explicitly define these cases.
@@
 As the interaction variable represents _potential_, i.e., the number of potential reachable opportunities, as opposed to trips:
-
-<!-- W_i^{(1)} enters through k.. it is a quality of the destination -->
 
@@
 
-Like gravity model problem, we can solve these cases using spatial weights matrix and travel cost matrix, but with alternative meanings.
+Like the SIM, we can solve these cases using spatial weights matrix and travel cost matrix, but with alternative meanings.
 $P_i$ (our $W_i^{(1)}$) is population at $i$ who seek opportunities.
@@
 
-Solved for our example, $K$ is equal to the total accessibility $V$, divided by the sum of the unconstrained accessibility:
+Solved for our example, $K$ is equal to the total accessibility $V$ (ALL THE OPPORTUNITIES, OR ALL THE DESTINATIONS), divided by the sum of the unconstrained accessibility:
 $$
@@
 
-Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
+The singly-constrained case can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
 For the origin- and destination- constrained versions, $k$ is replaced with $A_{ij}$ and $B_{ij}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):
@@
 
-\noindent Where $A_{ij}$ and $B_{ij}$ are similar to the balancing factors of the gravity model for the singly-constrained cases, though with an addition to the numerator to maintain the associated enforcing constraints:
+\noindent Where $A_{ij}$ and $B_{ij}$ are similar to the balancing factors of the SIM for the singly-constrained cases, though with an addition to the numerator to maintain the associated enforcing constraints:
 $$
@@
 
-Returning to our simple example, the destination-constrained singly-constrained case would yield the following $B_{ij}$ values:
+Returning to our simple example, the opportunity-constrained singly-constrained case would yield the following $B_{ij}$ values:
 $$
@@
 
-$V_{ij}$ values:
+$V_{ij}''^{(2)}$ values:
 ```{r}
@@
 
-### Doubly constrained case (origin-destination constrained)
+Example of solving the population-constrained $V_i''^{(1)}$:
+<!--COMEBACKHERE-->
+
+### Doubly constrained case (population-opportunity constrained)
 
@@
 ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
-<!-- this done'st work.. for now!-->
 ```{r}
@@
     #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
-  P,  # Origin production weights vector (POPULATION)
-  O,  # Destination attraction weights vector (OPPORTUNITIES)
+  O,  # Origin production weights vector
+  D,  # Destination attraction weights vector
   impedance_matrix,  # the resulting travel cost impedance matrix
@@
   # Normalize O and D
-  sum_P <- sum(P)
   sum_O <- sum(O)
-  cat("Checking population, opportunity balancing:\n")
-  cat("Population: ", sum_P, "\n")
-  cat("Opportunities: ", sum_O, "\n")
-
-  if (sum_P != sum_O) {
-    cat("Population and opportunities do not balance, scaling to match.\n")
-    if (sum_P < sum_O) {
-      P <- P * (sum_O / sum_P)
+  sum_D <- sum(D)
+  cat("Checking production, attraction balancing:\n")
+  cat("Production: ", sum_O, "\n")
+  cat("Attraction: ", sum_D, "\n")
+
+  if (sum_O != sum_D) {
+    cat("Productions and attractions do not balance, scaling to match.\n")
+    if (sum_O < sum_D) {
+      O <- O * (sum_D / sum_O)
     } else {
-      O <- O * (sum_P / sum_O)
+      D <- D * (sum_O / sum_D)
     }
   } else {
-    cat("Population, opportunities balancing OK.\n")
+    cat("Production, attraction balancing OK.\n")
   }
 
-  n <- length(P)  # Number of i
-  V <- sum(P)  # Total access
+  n <- length(O)  # Number of i
+  T <- sum(O)  # Total trips
   Ai <- rep(1, n)  # Initialize Ai
@@
     for (i in 1:n) {
-      Ai[i] <- 1 / (sum(Bj * O * impedance_matrix[i, ]) + 1e-9)
+      Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9)
     }
@@
     for (j in 1:n) {
-      Bj_new[j] <- 1 / (sum(Ai * P * impedance_matrix[, j]) + 1e-9)
+      Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9)
     }
 
-    # Calculate Vij
-    #Vij <- outer(Ai * P, Bj_new * O) * impedance_matrix
-    Vij <- Bj_new * Ai * O * impedance_matrix
+    # Calculate Tij
+    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix #Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix -->
 
@@
     # Calculate error
-    error <- (sum(abs(P - rowSums(Vij))) + sum(abs(O - colSums(Vij)))) / V
+    error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T
 
@@
 
-  return(list(Ai = Ai, Bj = Bj, Vij = Vij))
+  return(list(Ai = Ai, Bj = Bj, Tij = Tij))
 }
 
-dc_results <- doubly_constrained_gravity_model(P=P_i, O=O_j, impedance_matrix = 1/C)
-dc_results$Vij
+dc_results <- doubly_constrained_gravity_model(O=P_i, D=O_j, impedance_matrix = 1/C)
+dc_results$Tij
 ```
@@
 ```{r}
-V_doubly <- dc_results$Vij
+V_doubly <- dc_results$Tij
 
@@
 V_TOTAL_dest <- ifelse(sum(P_i_doubly)==sum(O_j_doubly), sum(V_doubly), NA)
-V_doubly_longer <- cbind(V_doubly,P_i_dest)
+V_doubly_longer <- cbind(V_doubly,P_i_doubly)
 V_doubly_longer <- rbind(V_doubly_longer, cbind(t(O_j_doubly),V_TOTAL_dest))
@@
 
-<!--THE FOLLOWING ARE SOLVED EXAMPLES OF ALL FOUR CASES OF THE GRAVITY MODEL -->
+<!--THE FOLLOWING ARE SOLVED EXAMPLES OF ALL FOUR CASES OF THE SIM -->
 <!-- The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.

### 2025-02-04 paper/manuscript.qmd
Finished good-enough draft up until (and include) sub-section "##Re-interpreting SIM constraints for accessibility"

@@
 ---
-title: "Accessibility, constraints and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
+title: "Constraining accessibility and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
 format: pdf
@@
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
-In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction" [p.73], as "a generalization of the population-over-distance relationship or 'population potential' concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73].
+In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction...
+a generalization of the population-over-distance relationship or _population potential_ concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73].
 As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses.
@@
 ## Gravity-based accessibility's cousin: Wilson's spatial interaction model
-<!--SUMMARY: State how accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models.
+<!--SUMMARY: Accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models.
 Introduce SIM's formulation.
-Emphasis the interpretation that comes from $k$.
-Emphasis how while SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average.
+Emphasize the interpretation that comes from $k$.
+While SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average.
 -->
 Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971].
-However, in many ways these spatial interaction models (AKA gravity models) continued to evolve independently from accessibility (@eq-accessibility-general) as will be discussed next section.
+However, in many ways these spatial interaction models, commonly refered to as gravity models, continued to evolve independently from accessibility (@eq-accessibility-general).
 Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM.
@@
 This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.
+
 To begin, the SIM as defined in Wilson [-@wilson1971], takes the following form: 
@@
 
-The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel:-->
+The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel a TOTAL TRAVEL COST constraint:-->
 $$
@@
 
-Using these constraints, the SIM @eq-phys-gravity-model is distinguished into four general cases.
+Using these constraints, the SIM @eq-phys-gravity-model is distinguished into three general cases.
 These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling]:
 
-- The unconstrained case: neither @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
-- The total constrained case: only @eq-constraint0-gravitymodel hold <!--NOTE: not sure if this is consistent with entropy-maximisation?-->
+- The unconstrained case: neither @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold, only @eq-constraint0-gravitymodel - the total number of trips, is used;
 - The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
@@
 
-In each of these four cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
+In each of these cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
 $$
@@
 For both singly-constrained cases, $K$ also holds.
-For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
-And for the unconstrained case, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.
-
-## Background on accessibilities' distinct development away from spatial interaction models
-<!--COME-BACK HERE-->
-<!--SUMMARY: Accessibility developed separate from SIM.
+For the unconstrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
+
+These constraints (@eq-bf-A-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-K-gravitymodel) and the SIM @eq-phys-gravity-model can be expressed not just for the variable of _interaction_ (e.g., the number of trips) but also as a variable of _potential interaction_.
+Before detailing SIM's formulation for accessibility, however, background on the development of accessibility and SIM independent from eachother is explained.
+
+## Background on accessibilities' distinct development from spatial interaction models
+<!--
+NOTE: This section is a tough one... please suggest ideas...
+
+SUMMARY: Accessibility developed separate from SIM.
 
@@
 
-The institutionalization of research on spatial interaction modelling was fueled by advancements in computer technologies and the substantive and unprecedented Federal-Aid Highway Act of 1956, which authorized $24.8 billion in spending at a 90% federal share from between 1957 to 1969 to begin and complete the US Interstate Highway System [@weinerUrbanTransportationPlanning2016].
-This ambitious project went on to shape transportation planning practice globally [].
-Within the same time period, accessibility measures were not applied practically at such a large scale.
-Instead, accessibility measures were taken up to describe land-use and transportation interactions by geography and social science-adjacent researchers and planners; remaining more in theory than in practice.
-However, due to growing critiques about traditional transportation modelling practice focus on  _mobility_, calls for a shift to accessibility are intensifying [@handy2020] and accessibility planning is beginning to be present in practice [] <!-- citation on where they are applied in practice?
-US guidelines.
-Netherlands.
--->
-
-Despite the theoretical and observed reality that both mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature.
-To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959, @wilson1971, and both @hansen1959 and @wilson1971.
+Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+However, within the same time period, accessibility remained largely within the academic realm.
+For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., <!--studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.-->].
+We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
+Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
+
+However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
+A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
+Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
+While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
+
+To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959 (one of accessibility's "seminal" works), @wilson1971 (one of SIM's "seminal" works), and both @hansen1959 and @wilson1971.
 @fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].
@@
 
-Many of the papers that cite both @hansen1959 and @wilson1971 (77 journal articles) are transportation modelling papers working on integrating accessibility within the SIM or interpreting HAM within the SIM.
-For instance, @battyMethodResiduesUrban1976 states that the relationship between the SIM and HAM may be worth exploring.
-There is also a consistent acknowledgement in the literature that HAM and SIM are related.
-Specifically, that HAM are within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
-Recognizing the balancing factors as HAM is a "common sense" approach at integrating accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979].
-
-There are other approaches to consider accessibility with SIM beyond HAM which can be founded on micro-economic consumer behaviour [@morrisAccessibilityIndicatorsTransport1979].
-For instance, others explore entropy maximizing solutions (Wilson's framework and hence the balancing factors defined as Hansen's accessibility) when considering demand behavior described by random utility choice models [@leonardiRandomUtilityDemand1984].
-
-Some work on location-allocation models also used both Hansen and Wilson.
-For instance, @leonardiOptimumFacilityLocation1978 and @beaumontLocationallocationProblemsPlane1981 define a location-allocation problem using the singly-constrained SIM, and then defines one using Hansen-type accessibility as the objective function subject.
-
-Then, @fotheringhamSpatialCompetitionAgglomeration1985 demonstrated that destination's accessibility should be explicitly specified within Wilson's framework, citing Hansen measure as convenient formulation.
-Without the consideration for accessibility explicitly, unexplained spatial pattern remains.
-This later came to be the 'competition destination' model, and is one of a few techniques to address spatial variation in SIMs.
-[@oshanSpatialStructureDebate2021].
-Implicitly, we gather that at this point: it was understood balancing factors in the SIM could not be understood as capturing accessibility.
-We also gather that in this moment of growing complexity of spatial interaction modelling, accessibility branched off and sprouted in the medium of budding GIS technology and the emergence of space-time dimensions [@millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
-With the possibility for showcasing complex GIS methodologies for varied case studies, accessibility became mature enough to branch off on its own.
-
-In papers after @fotheringhamSpatialCompetitionAgglomeration1985, Hansen is typically cited as providing a definition of accessibility but how Wilson is cited changes through the years.
-Wilson is attributed for using different travel cost functions in depending on the context [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-
-Othertimes, Wilson and Hansen measures were progressively incorrectly conflated throughout the years [ @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring] - indeed they both measure spatial interaction so are somewhat related, but they are not the same.
-They are also co-cited to refer to the concept of 'gravity' in models [@liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending]
-
-And finally, Wilson's balancing factors are interpreted as inverse Hansen-type accessibility, especially in the context of competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-This interpretation is fraut, upon considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration, which many of these works do not contend with.
-
-THE SPATIAL STRUCTURE DEBATE: @oshanSpatialStructureDebate2021
-
-Some papers also use SIMs to estimate trip or some other interaction flows alongside accessibility indicators [@clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
-
-Hansen is a relatively straightforward simple application, while Wilon's contribution is methodological and dense - must be carefully read.
-Many takeaways can be drawn, so with accessibilities changing applications throughout the years -- different things about Wilon's contribution becomes relevant.
-The most contemprary connection between the two is in the relam of competitive accessibility -- in the interpretation of Wilson's balancing factors as a way to consider competition.
-We fit this contribution in this literature.
-Competition fits on a spectrum - by introducing balancing factors trips are _constrained_ to equal regional or zonal totals.
-This same application can be applied to accessibility.
-
-Some 
-Additionally, trending research topics of the time such as space-time and more  []
-
-previously mentioned in accessibility development: the breadth of opportunity $O_j$ that can be studied [] as well as the accuracy of parameters that capture distance-decay $f(c_{ij})$ [].
-Outside of these factors, accessibility literature that delves into competition, does not cite Wilson, with some exception...
-Soukhov for the balancing factor.
-
-Overall, work that cities Wilson is closer related to transportation demand modelling than accessibility. However there are a few works of the 77 documents that are in the realm of accessibility. ...
-
-However there are a few papers that are accessibility focused that focus on
-
-<!-- NOTE: you can validate SIM e.g., how well does the model predict a known Tij matrix? but what is a known Vij matrix, potential accessible opportunities at i for j? Harder to say. -->
-
-<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. -->
-
-There are plenty of insights to be shared across these sibling branches; they are both spatial interaction models.
-
-## Defining accessibility constraints (a potential spatial interaction model?)
+In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
+Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
+
+Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
+Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
+Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
+
+However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
+This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
+Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
+Some works went beyond HAM and considered other accessibility expressions of individual choice.
+For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
+
+Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility <!--eek?
+I don't think I can say this-->.
+We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
+With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
+
+In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
+@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
+
+Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+
+Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
+Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+
+And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
+However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
+Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
+
+Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
+Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
+However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
+A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interation_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
+How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
+
+## Re-interpreting SIM constraints for accessibility
 
@@
 Example of solving the population-constrained $V_i''^{(1)}$:
-<!--COMEBACKHERE-->
+<!-- COMEBACKHERE -->
 
@@
 
+<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. -->
+
+<!-- Despite the theoretical and observed findings that _both_ mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature.
+-->
+
 <!--the benefits of this formulation, and the inclusion of the propoartionality constant has been imense for spatial interaction modelling.

### 2025-02-06 paper/manuscript.qmd
Draft for AP

@@
 ---
-title: "Constraining accessibility and the family of spatial interaction models" #"A family of accessibility measures (bringing balance back?)"
+title: "Accessibility, a type of spatial interaction model" #"A family of accessibility measures (bringing balance back?)"
 format: pdf
@@
 
-- The unconstrained case: neither @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold, only @eq-constraint0-gravitymodel - the total number of trips, is used;
+- The total constrained case: neither @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold, only @eq-constraint0-gravitymodel - the total number of trips, is used;
 - The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
@@
 For both singly-constrained cases, $K$ also holds.
-For the unconstrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
+For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
 
@@
 However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
-A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interation_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
+A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
 How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
 
-## Re-interpreting SIM constraints for accessibility
+## Interpreting SIM constraints for accessibility
 
@@
 $$
-V_{ij} = k W_j^{(2)} f(c_{ij}) = k O_j f(c_{ij})
+V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
 $${#eq-access}
@@
 Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
-Notably, the mass vectors $W_i^{(1)}$ and $W_j^{(2)}$ are replaced by $P_i$ and $O_j$, representing the weight of the origin (population) and destination (opportunities), the weights common in accessibility literature.
+Notably, the mass vectors $W_i^{(1)}$ and $W_j^{(2)}$ can be replaced by $P_i$ and $O_j$, representing the weight of the origin (population) and destination (opportunities) as known, the weights common in accessibility literature.
 As accessibility is the sum of opportunities for a $j$ from $i$, when the mass of the origin ($P_i$) is relevant, it enters @eq-access through the $k$ constant.
@@
 $$
-V_{i} = k \sum_j O_j f(c_{ij})
+V_{i} = k W_i^{(1)} \sum_j W_j^{(2)} f(c_{ij})
 $${#eq-access-sum}
 
-Like with @eq-phys-gravity-model, we distinguish @eq-access into four cases based on versions of the gravity-model building constraints (@eq-constraint0-access,  @eq-constraint1-access, and @eq-constraint2-access) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
+Like with @eq-phys-gravity-model, we distinguish @eq-access into three cases based on versions of the gravity-model building constraints (@eq-constraint0-access,  @eq-constraint1-access, and @eq-constraint2-access) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
 
@@
 $$
-\sum_j V_{ij} = W_i^{(1)} = P_i
+\sum_j V_{ij} = P_i
 $${#eq-constraint1-access}
@@
 $$
-\sum_i V_{ij} = W_j^{(2)} = O_j
+\sum_i V_{ij} =  O_j
 $${#eq-constraint2-access}
@@
 
-All four cases are differentiated like:
-- The unconstrained case: neither @eq-constraint0-access, @eq-constraint1-access, or @eq-constraint2-access hold;
-- The total constrained case: only @eq-constraint0-access hold
-- The singly-constrained case: either @eq-constraint1-access holds (origin-constrained) <!-- production--> or @eq-constraint2-access holds (destination-constrained)<!-- attraction-->; 
+All three cases are differentiated as follows:
+- The total constrained case: only @eq-constraint0-access hold. In this case, $V$ is the number of potential for interaction 
+- The singly-constrained case: either @eq-constraint1-access holds (origin-constrained)  or @eq-constraint2-access holds (destination-constrained) -- can only be in any combination of "potential for interaction" units, e.g., units of "opportunities" or "population" or "population for opportunities" or "opportunities for population;
 - The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously <!--production-attraction-->
 
-<!-- - HYBRID CASE: @eq-constraint1-access or @eq-constraint2-access partly hold -- hybrid production constrained case <!-- (could be repeated for attraction constrained case...
-and doubly constrained case)-->
-
-Like the SIM, we can solve these cases using spatial weights matrix and travel cost matrix, but with alternative meanings.
-$P_i$ (our $W_i^{(1)}$) is population at $i$ who seek opportunities.
-$O_j$ (our $W_j^{(2)}$) are the opportunities being sought.
+Notably, $V_{ij}$ can be in any combination of "potential for interaction" units that reflect the meaning of the spatial weights matrix e.g., units of "opportunities" from $i$ to $j$, or "population" from $i$ to $j$, or "population for opportunities" from $i$ to $j$, or "opportunities for population" from $i$ to $j$.
+The meaning of the units depends on the constraints: e.g., if singly-constrained - the spatial weights matrix represents units of opportunity or population while in doubly-constrained population and opportunities are two sides of this same matrix so take on the meaning of "population for opportunities" or "opportunities for population" from $i$ to $j$.
+
+Also it is relevant to describe previous accessibility literature relation to this framework:
+- Nonconstrained cases: current practice, the HAM, where @eq-constraint0-access, @eq-constraint1-access, nor  @eq-constraint2-access hold.
+- HYBRID CASE: examples of where @eq-constraint1-access or @eq-constraint2-access partly hold <!-- BFCA? Others?-->
+
+In these cases, $V_{ij}$ is never explicitly defined but we work backwards to develop it and place it within the SIM framework.
+
+Like the conventional SIM, $V_{ij}$ can be solved using a travel cost matrix and a spatial weights matrix but with alternative meanings.
+$P_i$ (our $W_i^{(1)}$ if known) is population at $i$ who seek opportunities.
+$O_j$ (our $W_j^{(2)}$ if known) are the opportunities being sought.
 For this simple example, we can say that the opportunities are the population's jobs.
@@
 
-### Unconstrained accessibility
-
-This accessibility is standard practice.
-There is no explicit consideration of a proportionality constant ($k$).
-We can presume it is still set to 1 as @stewartPrinciples1947.
-No constraints hold, making the interpretation of the units of $V_{ij}$ simply some magnitude of $O_j$ weighted by some $f(c_ij)$, and $V_{i}$ the sum of these values for each $j$.
-$V_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O$ and/or population $P$.
-$$
-V_{ij} = O_j f(c_{ij})\\ 
-V_{i} = \sum_j O_j f(c_{ij})
-$$
-
-Each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin: 
-```{r}
-O_P_unc <- t(matrix(rep(O_j, 3), 3, 3))
-O_P_unc_longer <- O_P_unc
-rownames(O_P_unc_longer) <- c("1","2","3") 
-colnames(O_P_unc_longer) <- c("1","2","3")
-O_P_unc_longer
-```
-
-The resulting matrix of $V_{ij}$ along with the $V_i$ and $V_j$ totals for this example:
-```{r}
-V_unc <- O_P_unc / C
-
-P_i_unc <- V_unc |> rowSums()
-O_j_unc <- V_unc |> colSums()
-V_TOTAL_unc <- ifelse(sum(P_i_unc|>round())==sum(O_j_unc|>round()), sum(V_unc), NA)
-V_unc_longer <- cbind(V_unc,P_i_unc)
-V_unc_longer <- rbind(V_unc_longer, cbind(t(O_j_unc),V_TOTAL_unc))
-
-rownames(V_unc_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_unc_longer) <- c("1","2","3", "TOTAL")
-
-V_unc_longer |> round(digits = 2) #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-We'd report origin 1 as having 148 accessibility (i.e., $V_{i=1} = 148$), origin 2 as 248 accessibility, and origin 3 as 172 accessibility.
-The total accessibility in the region is 568 ($\sum_i V_{i} = 568$), not equal to any meaningful value.
+### Nonconstrained accessibility
+
+The following case is standard practice, the HAM.
+It is expressed as a value of the origin, a summary of the destination weight multiplied by the travel impedance.
+There is no explicit consideration of a proportionality constant ($k$), we can presume it is still set to 1 as in @stewartPrinciples1947.
+Neither are there origin weights $W_i^{(1)}$, we assume is also wrapped up in this proportionality and implicitly set to 1.
+Working backward from this expression and assumptions, $S_{ij}$ can be expressed more generally as well.
+
+$$
+S_{i} = 1*\sum_j O_j f(c_{ij}) \\
+S_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij}) = 1*1*O_j f(c_{ij}) 
+$$
+
+$S_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O_j$ and/or population $P_i$ (only in some case when $k$ and $W_i^{(1)}$ equal 1).
+
+In solving the example, each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin: 
+```{r}
+O_P_non <- t(matrix(rep(O_j, 3), 3, 3))
+O_P_non_longer <- O_P_non
+rownames(O_P_non_longer) <- c("1","2","3") 
+colnames(O_P_non_longer) <- c("1","2","3")
+O_P_non_longer
+```
+
+The resulting matrix of $S_{ij}$ along with the $S_i$ and $S_j$ totals for this example:
+```{r}
+S_non <- O_P_non / C
+
+P_i_non <- S_non |> rowSums()
+O_j_non <- S_non |> colSums()
+S_TOTAL_non <- ifelse(sum(P_i_non|>round())==sum(O_j_non|>round()), sum(S_non), NA)
+S_non_longer <- cbind(S_non,P_i_non)
+S_non_longer <- rbind(S_non_longer, cbind(t(O_j_non),S_TOTAL_non))
+
+rownames(S_non_longer) <- c("1","2","3", "TOTAL") 
+colnames(S_non_longer) <- c("1","2","3", "TOTAL")
+
+S_non_longer |> round(digits = 2) #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
+We'd report origin 1 as having 148 accessibility (i.e., $S_{i=1} = 148$), origin 2 as 248 accessibility, and origin 3 as 172 accessibility.
+The total accessibility in the region is 568 ($\sum_i S_{i} = 568$), not equal to any meaningful value.
 This value is not often reported in the accessibility literature for this reason.
-One could also report the accessibility values that destinations offers ($V_{j}$), namely destination 1 offers 123, destination 2 offer 291, and destination 3 offers 154.
+One could also report the accessibility values that destinations offers ($S_{j}$), namely destination 1 offers 123, destination 2 offer 291, and destination 3 offers 154.
 However, these values are also not reported in the literature currently...
-as they aren't constrained to match up to the number of opportunities at each j (i.e., 160, 437 and 193).
-Assuming $k=1$, $V_i$ and $V_j$ represent magnitudes of accessibility concepts, but in an unconstrained way.
-The population matrix is no incorporated in the $V_{ij}$ calculation, so $V_i$ cannot be directly compared to population.
-From a different perspective, $V_j$ values also cannot be compared to known $O_j$ values as the equation unconstrained.
-
-OF NOTE: in the context of spatial interaction modelling, the unconstrained model is not often useful.
-While the resulting $T_{ij}$ values are in units of 'trips' - the assumption of $k$ equaling 1 is practically meaningless.
+as they aren't constrained to match up to anything known, including the number of opportunities at each j (i.e., 160, 437 and 193).
+Assuming $k=1$, $S_i$ and $S_j$ represent magnitudes of accessibility concepts, but in an nonconstrained way.
+The population matrix is not incorporated in the $S_{ij}$ calculation, so $S_i$ cannot be directly compared to population.
+From a different perspective, $S_j$ values also cannot be compared to known $O_j$ values as the equation unconstrained.
+
+OF NOTE: in the context of spatial interaction modelling, this nonconstrained version of model is seldom useful.
+While the resulting $T_{ij}$ values are technically units of 'trips' - there are no constraints, so the assumption of $k$ and $W_i^(1)$ equaling 1 is practically meaningless.
 
@@
 $T_{ij}$), namely since the number of opportunities that are accessible at an origin from a destination cannot be known in the same way as a trip from an origin to a destination.
-However, we would argue that getting closer to knowning this number and planning for it - is what accessibility research should strive to achieve.
-<!-- yeah?
--->
-
-### Total flow constrained accessibility
+However, we would argue that getting closer to knowing this number and planning for it - is what accessibility research should strive to capture.
+An origin-destination based accessibility value $V_{ij}$.
+We demonstrate the following three cases sub-sequently.
+
+### Total constrained accessibility
 
@@
 $$
-V_{ij}' = K O_j f(c_{ij})\\
-V_{i}' = K\sum_j O_j f(c_{ij})
-$$
-
-Solved for our example, $K$ is equal to the total accessibility $V$ (ALL THE OPPORTUNITIES, OR ALL THE DESTINATIONS), divided by the sum of the unconstrained accessibility:
-$$
-K = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
+V_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})\\
+V_{i}' = K\sum_j W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+
+Solved for our example, $K$ is equal to the total accessibility $V$, the only thing we know: either all the opportunities $O$ or all the population $P$, divided by remaining part of the equation.
+In the following, we assume our spatial weights matrix only reflects $W_j^(2)$ as $O_j$ and $W_i^(1)$ is uniform (1):
+$$
+K = \frac{V}{\sum_{i}\sum_{j} 1* O_j f(c_{ij})}\\
 K = \frac{V}{\frac{O_1}{c_{1,1}}+\frac{O_1}{c_{2,1}}+\frac{O_1}{c_{3,1}}+\frac{O_2}{c_{1,2}} +... \frac{O_3}{c_{3,3}}
@@
 ```{r}
-K <- V/sum(V_unc)
-V_tot <- K * V_unc
+K <- V/sum(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3))
+V_tot <- K * S_non
 
@@
 
+This case can be reformulated if the spatial weights matrix reflects only weights at the origin $W_i^(1)$ such as population $P_i$ and $W_j^(2)$ is unknown/irrelevant.
+
 ### Singly-constrained case
@@
 The singly-constrained case can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
-For the origin- and destination- constrained versions, $k$ is replaced with $A_{ij}$ and $B_{ij}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):
-
-$$
-V_{ij}''^{(1)} = A_{ij} O_j f(c_{ij})\\
-V_{ij}''^{(2)} = B_{ij} O_j f(c_{ij})\\
+For the origin- and destination- constrained versions, $k$ is replaced with $A_{i}$ and $B_{j}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):
+
+$$
+V_{ij}''^{(1)} = A_{i} W_j^{(2)} P_i f(c_{ij})\\
+V_{ij}''^{(2)} = B_{j} O_j W_i^{(1)} f(c_{ij})\\
 $${#eq-access-singlyconstrained-origin}
 
-\noindent Where $A_{ij}$ and $B_{ij}$ are similar to the balancing factors of the SIM for the singly-constrained cases, though with an addition to the numerator to maintain the associated enforcing constraints:
-$$
-A_{ij} = \frac{O_j}{\sum_j B_{ij} O_j f(c_{ij})}\\
-B_{ij} = \frac{P_i}{\sum_i A_{ij} P_i f(c_{ij})}\\
-$$
-In solving the origin-constrained $V_{ij}^{''(1)}$ case (e.g., @eq-constraint1-access of $\sum_j V_{ij} = P_i$ is enforced), $B_{ij} = 1$.
-In solving the destination-constrained $V_{ij}^{''(2)}$ case (e.g., @eq-constraint2-access of $\sum_i V_{ij} = O_j$ is enforced), $A_{ij} = 1$.
-
-Solving for the destination-constrained case $V_{ij}^{''(2)}$ may be helpful in an example where the number of people seeking opportunities is known ($P_i$), along with the number of opportunities $O_j$ offered by a destination as in unconstrained and total accessibility.
+\noindent Where $A_{i}$ and $B_{j}$ are the same as the balancing factors of the SIM for the singly-constrained cases:
+$$
+A_{i} = \frac{1}{\sum_j B_{j} O_j f(c_{ij})}\\
+B_{j} = \frac{1}{\sum_i A_{i} P_i f(c_{ij})}\\
+$$
+
+\noindent NOTE: if the unknown spatial weight ($W_j^{(2)}$ for $A_{i}$ and $W_i^{(1)}$ for $B_{j}$) is moved from @eq-access-singlyconstrained-origin's numerator to $A_{i}$ or $B_{j}$ it is equivalent to a simplified version of Spatial Availability's formulation which is also demonstrated to be related to @shen1998 and the 2SFCA (see appendix @soukhovIntroducingSpatialAvailability2023).
+
+In solving the origin-constrained $V_{ij}^{''(1)}$ case (e.g., @eq-constraint1-access of $\sum_j V_{ij} = P_i$ is enforced), $B_{j} = 1$.
+In solving the destination-constrained $V_{ij}^{''(2)}$ case (e.g., @eq-constraint2-access of $\sum_i V_{ij} = O_j$ is enforced), $A_{i} = 1$.
+
+Solving for the destination-constrained case $V_{ij}^{''(2)}$ may be helpful in an example where the number of people seeking opportunities is known ($P_i$), along with the number of opportunities $O_j$ offered by a destination as in total accessibility.
 An example could be the number of people at each residence seeking hospital-beds (origin) and hospital-beds (opportunities) at each healthcare location (destination) to calculate the opportunity access at each origin.
@@
 It was only until the recent work of [@soukhovIntroducingSpatialAvailability2023], where the units of accessibility given competition, e.g., the spatial availability, was introduced with the focus on both competition and balanced units.
-NOTE: See Appendix where I demonstrate how this destination-constrained case is equivalent to spatial availability
-
-Returning to our simple example, the opportunity-constrained singly-constrained case would yield the following $B_{ij}$ values:
-$$
-B_{ij} = \frac{P_{i}}{\sum_i P_i f(c_{ij})}\\
-B_{1,1} = \frac{P_{i}}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
-...\\
-B_{1,1} = 160 *\frac{1}{160/2+450/15+180/5} = 1.09589 \\ 
-B_{2,1} = 450 *\frac{1}{160/2+450/15+180/5} =3.082192 \\
-B_{3,1} = 180 * \frac{1}{160/2+450/15+180/5} =1.232877 \\
-...\\
-B_{1,2} = 160 *\frac{1}{160/15+450/2+180/10} =0.630749 \\
-B_{2,2} = 450 *\frac{1}{160/15+450/2+180/10} =1.773982 \\
-B_{3,2} = 180 * \frac{1}{160/15+450/2+180/10} =0.7095926 \\
-...\\
-B_{1,3} = 160 *\frac{1}{160/5+450/10+180/2} = 0.9580838\\
-B_{2,3} = 450 *\frac{1}{160/5+450/10+180/2} = 2.694611  \\
-B_{3,3} = 180 *\frac{1}{160/5+450/10+180/2} = 1.077844 \\
-$$
-```{r}
-B_ij <- matrix(P_i) %*% (1/(t((matrix(P_i))) %*% (1/C))) 
-B_ij_longer <- B_ij
-rownames(B_ij_longer) <- c("1","2","3") 
-colnames(B_ij_longer) <- c("1","2","3") 
-B_ij_longer 
-```
+
+Returning to our simple example, the opportunity-constrained singly-constrained case would yield the following $B_{j}$ values assuming $A_{i}$ is 1:
+$$
+B_{j} = \frac{1}{\sum_i 1* P_i f(c_{ij})}\\
+B_{1} = \frac{1}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
+.., j=2, j=3\\
+B_{1} =  \frac{1}{160/2+450/15+180/5} = 0.006849315 \\ 
+B_{2} =  \frac{1}{160/15+450/2+180/10} = 0.003942181 \\
+B_{3} =  \frac{1}{160/5+450/10+180/2} = 0.005988024 \\
+$$
+```{r}
+B_j <- 1/(P_i %*% (1/C))
+B_j_longer <- B_j
+rownames(B_j_longer) <- c("i..") 
+colnames(B_j_longer) <- c("1","2","3") 
+B_j_longer 
+```
+(NOTE: notably, when $B_j$ is multiplied by $P_i*f(c_{ij})$, it is a proportion (like spatial availability's total population&travel-cost balancing factor) representing how much of $O_j$ is allocated to each $i$)
 
@@
 ```{r}
-V_dest <- (B_ij) *(V_unc) # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3))
+V_dest <- outer(P_i, as.numeric(B_j) * O_j) * (1/C) #we're not doing standard matrix multiplication here.
+We use 'outer' for element-wise multiplication, then element-wise multiplication of the resulting 3*3 matrix with the 3*3 impedance table
 
@@
 
-V_TOTAL_dest <- ifelse(sum(P_i_dest)==sum(O_j_dest), sum(V_dest), NA)
+V_TOTAL_dest <- ifelse(all.equal(sum(P_i_dest),sum(O_j_dest)), sum(V_dest), NA)
 V_dest_longer <- cbind(V_dest,P_i_dest)
@@
 
-The $O_j$ values are equal to the known opportunities, whereas the $P_i$ values are not.
+The $O_j$ values (bottom row) are equal to the known opportunities at each $j$, whereas the $P_i$ values (right-most column) are not.
 The total accessibility in the region is maintained at 790.
-Each $V_{ij}$ can be interpreted as the number of opportunities accessible from each destination by anyone in the region all weighted by travel cost of course.
-
-Example of solving the population-constrained $V_i''^{(1)}$:
-<!-- COMEBACKHERE -->
-
-### Doubly constrained case (population-opportunity constrained)
-
-(only when the units of the marginals are the same, therefore of more limited application)
+Each $V_{ij}$ can be interpreted as $O_j$ weighted by the proportional balancing factor e.g., "the number of opportunities accessible from $j$ by the population at $i$" all weighted by travel cost, of course.
+Use this formulation of the singly-constrained case when the number of opportunities, at each $j$, that are being competed for by the population is known but the precise number of population that is competiting for opportunities is not.
+For example, public schools as a destination in a region where all children attend them.
+All school-seats at $j$ are open to all children at all $i$s.
+However, not all population $i$ are equally interested in school-seats at all $j$s (e.g., some prefer special programs, teachers, etc.); and we do not know this precise population $i$ to $j$ number.
+Hence, an opportunity-constrained version of the singly-constrained case is appropriate.
+
+However, the population-constrained version of the singly-constrained case can be appropriate in the converse situation.
+Consider the case of public urgent care clinics in a region where all people use them equally.
+Unlike schools, all population at $i$ are equally interested in urgent-care-seats at all clinics $j$.
+If there's an emergency, or not, they will visit.
+However, from an operational perspective all clinics at $j$ are _not_ interested in serving all people (e.g., they'd rather serve people who have access to primary care medicine as they are more likely to seek urgent care in actual medical emergencies).
+From this perspective, using a _population-constrained_ version of the singly-constrained case $V_i''^{(1)}$, is appropriate.
+Each $V_{ij}$ can be interpreted as $P_i$ weighted by the proportional balancing factor e.g., "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
+
+In $V_i''^{(1)}$, the following is $A_i$ assuming $B_{j}$ is 1:
+$$
+A_{i} = \frac{1}{\sum_j 1* O_j f(c_{ij})}\\
+A_{i=1} = \frac{1}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\
+.., i=2, i=3\\
+A_{1} =  \frac{1}{160/2+437/15+193/5} = 0.006768953 \\ 
+A_{2} =  \frac{1}{160/15+437/2+193/10} = 0.004024685 \\
+A_{3} =  \frac{1}{160/5+437/10+193/2} = 0.005807201 \\
+$$
+```{r}
+A_i <- 1/((1/C) %*% O_j)
+A_i_longer <- A_i
+rownames(A_i_longer) <- c("1","2","3") 
+colnames(A_i_longer) <- c("j...") 
+A_i_longer 
+```
+(AGAIN NOTE: when $A_i$ is multiplied by $O_j*f(c_{ij})$, it is a proportion (like spatial availability's total balancing factor) representing how much of $P_i$ is allocated from each $i$, I think.
+e.g.
+each $V_{ij}$ is the proportional opportunity&travel-cost balancing factor of $A_{i} W_j^{(2)}f(c_{ij})$ multiplied by $P_i$)
+
+The resulting $V_{ij}''^{(1)}$ values:
+```{r}
+V_orig <- outer(as.numeric(A_i) * P_i, O_j) * (1/C) #we're not doing standard matrix multiplication here.
+We use 'outer' for element-wise multiplication, then element-wise multiplication of the resulting 3*3 matrix with the 3*3 impedance table
+
+P_i_orig <- V_orig |> rowSums()
+O_j_orig <- V_orig |> colSums()
+
+V_TOTAL_orig <- ifelse(all.equal(sum(P_i_orig),sum(O_j_orig)), sum(V_orig), NA)
+V_orig_longer <- cbind(V_orig,P_i_orig)
+V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig))
+
+rownames(V_orig_longer) <- c("1","2","3", "TOTAL") 
+colnames(V_orig_longer) <- c("1","2","3", "TOTAL")
+
+V_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
+```
+
+In this version of the singly-constrained case, the population is constrained so the $P_i$ values (right-most column) are equal to the known population, whereas the $O_j$ values (bottom row) is not.
+The total accessibility in the region is maintained at 790.
+To reiterate, each  $V_{ij}$ can be interpreted as  "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
+
+### Doubly-constrained case (population-opportunity constrained)
+
+In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
+In this case, the units of the population and opportunities must be able to be understood as the same.
+E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are saught by population at each $i$ must both be known to match.
+
+In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
+In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
+If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
+
+Literature has explored 'doubly-constrained' but -- not the same: ,etc. 
 ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
+
+In this case, $V_{ij}$ would be:
 ```{r}
@@
 
+Nice.
+
 ### Hybrid singly-constrained case: 
-<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here-->
+<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA? -->
 Balanced floating catchment areas (not truly doubly constrained)
@@
 
-# APPENDIX (some musings...)
+Accessibility development has a rich history.
+There have been developments in the cases/breath of travel impedance functions, competition considerations, and in incorporating individual choice as part of activity-based accessibility.
+However, no work yet has examined accessibilities explicit connecting to SIM.
+We examine this connection, and provide a origin-destination ($i$ to $j$) formulation of accessibility as part of the SIM framework.
+Novel stuff.
+
+# APPENDIX (some musings... destined for the commented-out graveyard)
 
@@
 The destination-constrained case is also spatial availability, see below!
-As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{ij}$ is removed).
+As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{j}$ is removed).
 
@@
 $$
+
 ## Checking singly-constrained case against spatial availability
@@
 
-Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{ij}$ values:
-$$
-A_{ij} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\
+Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{i}$ values:
+$$
+A_{i} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\
 A_{1,1} = \frac{O_{j}}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\
@@
 ```{r}
-V_orig <- (A_ij) * t(V_unc) # V_unc = (1/C)* t(matrix(rep(W_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.
+V_orig <- (A_ij) * t(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.
 

### 2025-02-09 paper/manuscript.qmd
Editing while proof-reading

@@
 ---
-title: "Accessibility, a type of spatial interaction model" #"A family of accessibility measures (bringing balance back?)"
+title: "A family of accessibility measures derived from spatial interaction principles"
 format: pdf
@@
 ```{r load-packages, include=FALSE, cache=FALSE}
+library(bibliometrix) # Comprehensive Science Mapping Analysis
 library(dplyr) # A Grammar of Data Manipulation
+library(flextable) # Functions for Tabular Reporting
 library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-library(flextable)
-library(knitr)
-library(glue)
-library(here)
+library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
+library(glue) # Interpreted String Literals
+library(gt) # Easily Create Presentation-Ready Display Tables
+library(here) # A Simpler Way to Find Your Files
+library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
+library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
 # library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
@@
 # library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
-# library(here)
+# library(here) # A Simpler Way to Find Your Files
 ```
@@
 As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->
+
+Historically, the focus of transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement.
+This has had problematic consequences: with the car seen as the ultimate mobility tool, this approach has led to the emergence and continued dominance of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013].
+Decades of mobility-based planning (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often have had limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+In response to this situation, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020] using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020].
 Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape.
 In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959].
-The range of opportunities considered in accessibility analysis has grown to include <!-- --> employment [REFS], health care [REFS], green spaces [REFS], schools [REFS], social contacts [REFS], emergency services [e.g., shelters, REFS], and many more.
-In other words, accessibility analysis is today used to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
-Compared to earlier measures of performance used in transportation (e.g., VKT, PKT, etc.) that benchmark the amount of movement, accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].
-However, the focus of transportation planning has historically prioritized mobility and treated accessibility as a by-product of movement, leading to problematic consequences [@handy2020].
-With the rise of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013], recent decades of mobility-based practice (e.g., road and highway expansion) have been marked by increased travel cost (time, effort, energy, financial cost), often with limited impact on accessibility levels [e.g., @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
-In response, transportation planning practice has begin to adopt accessibility planning approaches, via accessibility planning tools [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013] or through the traditional four-step transportation demand model [@ortuzar_2011_modelling].
-However, institutional and conceptual barriers remain, many of which relate to the diversity of accessibility definitions [@handy2020] and interpretability and communicability of the outputs [@geursAccessibilityEvaluationLanduse2004; @ferreiraReenactingMobilityAccessibility2020].
-
-<!-- Considering population-level methods in transportation and city planning research, accessibility measures have been used to tackle a wide variety of research questions.
-They have taken on numerous definitions and analytical expressions; from simpler cumulative opportunities measures, to measures considering more complex travel behaviour and trip weighting characteristics, measures of _potential of opportunities for interaction_ are varied.
-Summarizing these definitions across studies have been challenging; a testament to these unsettled classifications are present in influential accessibility review literature [@handyMeasuringAccessibilityExploration1997; @geursAccessibilityEvaluationLanduse2004; @wuUnifyingAccess2020; @millerAccessibilityMeasurementApplication2018].
+Compared to other measures of performance used in transportation that benchmark movement (e.g., VKT, PKT, etc.), accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].
+
+As a result, scholarly research on accessibility has experienced a remarkable boom, and has grown to include employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
+In other words, accessibility analysis is used today to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
+However, despite its growth in popularity in scholarly works, challenges remain to more widespread adoption of accessibility in planning practice.
+Several barriers to bridging the scholarly-practice accessibility gap have been identified.
+For example, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
+Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
+
+Adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
+The menu of accessibility measures has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
+How is a practitioner to choose among this myriad options?
+What differences in accessibility scores should matter, and how should they be communicated?
+[see @vanweeAccessible2016; p.
+14].
+
+In this respect, we welcome @wuUnifyingAccess2020's contribution for the way it provides a unifying framework to think about accessibility measures.
+By considering the key features of accessibility, namely travel-cost and the distribution of opportunities, these authors demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula.
+One needs only to judiciously change the way travel-cost and opportunities are formulated, to derive almost any known accessibility measure.
+
+With their research, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
+Still, we feel that there is room to further clarify another relevant aspect.
+@wuUnifyingAccess2020's treatment of competition deserves further scrutiny, since the unifying framework does not appear to accommodate many (or any) of the hugely popular floating catchment area methods.
+In recent research, however, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn an accessibility measure of the following type (the most general in Wu and Levinson's framework) into a competitive measure of accessibility _à la_ 2-Step Floating Catchment Area approach:
+
+$$
+S_i \propto \sum_j g(O_i)f(c_{ij})
+$$
+
+In this paper we contend that accessibility research needs to reconnect with its roots in spatial interaction modelling.
+It is by looking to the past that we believe accessibility analysis can take a new step to the future.
+In particular, we argue that an important aspect of spatial interaction modelling, namely constraining the results to match empirical observations, was never properly reincorporated into accessibility analysis after these two streams of research began to diverge after the work of @hansen1959 and @wilson1971.
+As Wilson demonstrated, applying empirical constraints led to various spatial interaction models, and in this work we aim to show that the same can be done for accessibility measures.
+On the contrary, the lack of constraints in accessibility analysis seems to have contributed to some of the interpretability issues that plague it, particularly the fuzziness of insights beyond making statements of proportion such as higher-than/lower-than.
+
+With this background in mind, the focus of the present paper is on the empirical constraints and the attendant proportionality constant missing from accessibility measures.
+We propose to map accessibility onto Wilson's family of spatial interactions models.
+To achieve this, we begin by tracing the development of accessibility from its origins in spatial interaction, beginning with @ravensteinLawsMigration1889 to the accessibility research of @hansen1959.
+We also provide evidence of a marked divergence of accessibility and spatial interaction modelling research after the work of @wilson1971.
+After this, we hark back to Wilson's spatial interaction model, and use it to derive a family of accessibility measures based on different types of constraints.
+We illustrate various members of this family with a simple numerical example and a real world data set.
+We then conclude by discussing the uses of these measure and their interpretation.
+
+<!--
+We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values, as well as converting accessibility indicators from ordinal variables into interval/ration ones.
+To address this matter, we need not reinvent the wheel.
+Stewart left the determination of $G$ for the future, possibly expecting it to be some sort of universal constant.
+However, the solution to this problem was found by @wilson1971, who discovered that there was not a single $G$, but rather various balancing factors that depend on 1) empirical observations of population and/or opportunities; and 2) whether (and how) they were used to constrain the analysis.
+This development is outlined next.
 -->
 
-<!-- PAPER'S AIM: improve interpretation of accessibility measure outputs.
-We do this by, 1) assuming accessibility is more complete understanding of mobiltiy, and connected them both mobility and accessibility are both based in the Netownian gravitational anology, lets introduce balancing factors like those  Gravity Model.
--->
-In adopting the perspective that accessibility is an evolution of mobility conceptually, this paper aims to re-unify accessibility measures with its spatial interaction roots.
-It proposes to do so by mapping accessibility onto Wilson's family of spatial interactions models [@wilson1971].
-In this, we aim not only to clarify accessibility's interpretation as a measure of performance, but also to encourage future developments that draw from its spatial interaction history.
-To this end, we begin by tracing the development of spatial interaction from @ravensteinLawsMigration1889 to its split into accessibility research from @hansen1959 to today.
-After this, we derive different types of accessibility measures, following the family of "spatial interaction models" (SIM) namely: unconstrained, total constrained, origin- and destination- constrained, composite destination-constrained, and doubly constrained.
-Then, we will provide empirical examples of the types of measures.
-Finally, we will discuss the use of each measure and broadly the implications of ignoring the role of constrains in accessibility analysis.
-
-# Background
-
-## The inquiry into spatial interaction traces to Newton's Universal Gravitation
+# Newtonian's roots of human spatial interaction research
+
 <!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
-The patterns of people's movement in space have been a subject of scientific inquiry for centuries.
+
+The patterns of people's movement in space have been a subject of scientific inquiry for almost a century and a half.
 From as far back as Henry C.
-Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] there was already a concern with the scientific study of human interaction over space.
-It is in this work where Carey states that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
-37-38], which shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation [posited in _Principia Mathematica_ of 1687 and expressed in @eq-phys-grav-prop].
+Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] a concern can be observed with the scientific study of human spatial interaction.
+It was in this work where Carey stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
+37-38].
+This statement shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation, first posited in 1687's _Principia Mathematica_ and expressed as in @eq-phys-grav-prop.
 
@@
 In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$.
-Direct proportionality only means that as the product of the masses increases, so does the force.
-Likewise, inverse proportionality only means that as the distance increases, the force decreases.
-To quantify the magnitude of the force, an empirical factor, say $G$, is required to convert the proportionality into an equality, which ensures that the force $F$ matches empirical observations of the force of attraction between masses.
-Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is a proportionality constant obtained empirically:
+Direct proportionality means that as the product of the masses increases, so does the force.
+Likewise, inverse proportionality means that as the distance increases, the force decreases.
+@eq-phys-grav-prop, however, does not quantify the magnitude of the force, for to do so an empirical constant is required to convert the proportionality into an equality, thus ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
+In other words, @eq-phys-grav-prop needs to be _constrained_ using empirical data.
+Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically callibrated proportionality constant:
 
@@
 
-Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that Hutton's experiments of 1778 validated [@hutton_xxxiii_1778, p.
+Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p.
 783].
-Still, it took over a century from the publication of _Principia_ to refine the estimate of this constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
-
-## Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
-
-<!-- SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s.
-These researchers have theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
-This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy.
+Still, it took over a century from the publication of _Principia_ to refine the estimate of the proportionality constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
+
+# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
+
+<!-- 
+SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s.
+A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
+This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
 -->
+
 Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts.
-In the late 1880s Ravenstein proposed his "Laws of Migration", based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
-In these works, Ravenstein posited a directly proportional relationship of migration flows with the size of the destination (i.e., centres of commerce and industry), and an inversely proportional relationship between the size of flows and separation between origin and destination.
-These propositions are similar to Newton's gravitational laws.
+In the late 1880s Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
+In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
+As with Carey, these propositions are similar to Newton's gravitational laws.
 Over time, other researchers discovered similar relationships.
 <!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations.
-Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways @zipfHypothesisCaseRailway1946, intercity personal movement @zipfHypothesisIntercityMovement1946,and information @zipfDeterminantsCirculationInformation1946.
+Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways [@zipfHypothesisCaseRailway1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and information [@zipfDeterminantsCirculationInformation1946].
 The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
 
-A common theme among these early researchers of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav is missing.
-Of the researchers cited, only Reilly and Zipf expressed their hypotheses in mathematical terms.
-Reilly's hypothesis was expressed as follows:
+A common feature of these early investigations of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav was never considered.
+Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms.
+Reilly's hypothesis was presented in the following form:
+
 $$
@@
 
-\noindent Where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
+\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
 Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
@@
 Zipf, on the other hand, wrote his hypothesis in mathematical form as:
-$$
-C^2 = \frac{P_1\cdot P_2}{D}
+
+$$
+C^2 = \frac{P_1\cdot P_2}{D_{12}}
 $${#eq-zipf}
 
-\noindent Where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.
+\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and  $D_{12}$ is the distance between them.
 
 After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
-We suspect that this may relate to academic backgrounds; Stewart was a physicist, whereas Ravenstein, Reilly, and Zipf were social scientists.
-Besides awareness of preceding research, as he cites Reilly and Zipf as predecessors in the analysis of human spatial interaction, Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
-$$
-F = G\frac{(N_1\mu_1)(N_2\mu_2)}{d_{12}^2}
+We suspect that this relates to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist.
+Besides awareness of preceding research (Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
+
+$$
+F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
 $${#eq-stewart-force}
@@
 - $\mu_1$ and $\mu_2$ are so-called _molecular weights_
+- $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
 - $d_{12}^2$ is the distance between $1$ and $2$
@@
 
-In addition to demographic force, Stewart defined a measure of "population potential" of $2$ with respect to $1$ as follows:
-$$
-V_1 = G\frac{M_2}{d}
+In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
+$$
+V_1 = G\frac{M_2}{d_{12}}
 $${#eq-stewart-population-potential}
 
-For a system with more than 2 population bodies, Stewart formulated the population potential at $i$ as:
+For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
 $$
@@
 
-\noindent where $D$ is population density over an infinitesimal area $ds$, and $r$ is the distance to $i$. In more contemporary terms, this can be rewritten as:
+\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
+In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
 $$
@@
 
-\noindent which astute readers will identify as our modern definition of accessibility.
+Alerted readers will notice that @eq-stewart-population-potential-sum is formally equivalent to our modern definition of accessibility.
 
@@
 It had issues with inconsistent mathematical notation.
-For example, e.g., in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant for both _demographic force_ and  _demographic energy_ on p.
-34.
-Perhaps more seriously, Stewart's work was permeated by a view of humans as particles following physical laws, but with ideas that were afflicted by unscientific racism.
-For instance, the molecular weight $\mu$ was presumed to be one for the average American, but "presumably...much less than one....for an Australian aborigine" [p.
+<!--For example, in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant p.
+34 and then later as _demographic energy_ on p.
+53.--> More seriously, though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism.
+For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" [p.
 35].
-Stewart's ideas about "social physics" fell out of favour among social scientists, but not before influencing the emerging field of accessibility research.
-
-In summary: from the 1850s to the 1940s, researchers attempted to theoretically and empirically characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M$ of a pair of locations ($i$ and $j$) and inversely proportional by some sort of separation distance $D_{ij}$.
-Those that captured the relationship empirically, approached the concept with different expressions, some including a proportionality constant and some not, but all tying back to the Newtonian gravity analogy.
-
-## Hansen's gravity-based accessibility to today
+Stewart's ideas about "social physics" soon fell out of favour among social scientists, but not before influencing the nascent field of accessibility research, as detailed next.
+
+# Hansen's gravity-based accessibility to today
+
 <!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up.
@@
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
-In his seminal paper, Hansen defined accessibility, i.e., "the potential of opportunities for interaction...
-a generalization of the population-over-distance relationship or _population potential_ concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" [p.73].
-As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the the Bureau of Roads, and likely preoccupied with applied research into the power of transportation to shape land uses.
+In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
+a generalization of the population-over-distance relationship or _population potential_ concept developed by @stewartDemographicGravitationEvidence1948" (p.
+73).
+As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
 @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
@@
 In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
-$$
-S_{i} = \sum_j O_j \cdot d_{ij}^\beta \\
+
+$$
+S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
 $${#eq-accessibility}
 
-Alternatively, a more general form of this expression to account for different impedance functions:
+A modern re-writing of the above accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
+
 $$
@@
 
-$S_{i}$ in @eq-accessibility is a measure of the accessibility at $i$, whereas $O_j$ is the size of the activity at $j$ and $d_{ij}$ is the distance or travel time between $i$ and $j$ with $\beta$ describing the effect size.
-Today, Hansen is is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to today as the gravity-based accessibility measure.
+$S_{i}$ in @eq-accessibility is a measure of the accessibility of site $i$.
+This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g.,  distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost).
+Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.
 
 Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _direct_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance.
-Furthermore, those reading @hansen1959 should remember that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$ to 1, with a note that"$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
-
-Since @hansen1959, accessibility analysis has been widely employed across planning disciplines [@koenigIndicatorsUrbanAccessibility1980; @geursAccessibilityEvaluationLanduse2004; @millerAccessibilityMeasurementApplication2018; @wuUnifyingAccess2020].
-However, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determining it.
-In this way, $G$ continues to be implicitly set to $1$ even when the fundamental relationship in accessibility is proportionality (e.g., $ S_{i}  \propto G \sum_j O_j \cdot f(d_{ij})$ and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
-Without a proportionality constant, however, the units of $S_i$ remain unclear: the unit of 'potential of opportunity for interaction' is left free to change as $\beta$ is calibrated for the population center with no discussion of a constant to balance the units.
-
-However, it has been raised that there is a difficulty in interpreting gravity-based accessibility's meaning as anything other than a relative index of potential interaction (higher accessibility/lower accessibility) [@millerAccessibilityMeasurementApplication2018].
-Interpretability issues plague this type of analysis, as implications beyond high-or-low access hotspots are often desired.
-What does the accessibility value mean?
-Difficultly in interpreting the output's tangible meaning contributes to resistance to its application in practice [@ferreiraReenactingMobilityAccessibility2020].
-Furthermore, The majority of accessibility studies within the past few decades develop on the breath and geographic case study selection of the opportunity $O_j$ studied and potentially impacted population []; the accuracy of parameters that capture distance-decay $f(c_{ij})$ []; and additions to the opportunity $O_j$ attributes such as competitive demand considerations [@stoufferINTERVENINGOPPORTUNITIESCOMPETING1960; @weibull_axiomatic_1976; @shen1998; @allenMeasureCompetitiveAccess2020; @soukhovMultimodalSpatialAvailability2024].
-Overall, due to the diverse range of questions accessibility measures have been applied to help clarify, a plethora of variations in @eq-accessibility-general are present in the literature.
-These variations also compounds the issue of interpretability across studies.
-
-A testament to this inconsistent interpretation are the unsettled classifications generated by influential review works of the accessibility literature; for instance, @handyMeasuringAccessibilityExploration1997 classifies studies into gravity- and utility-based measures, @geursAccessibilityEvaluationLanduse2004 introduces further classification by expanding categories into potential- (or gravity), utility-, location-, and person- based measures with different components.
-@curtis2010planning further classifies accessibility measures into spatial separation, contour-, gravity-, competition-, time-space, utility-, and network- based measures [@curtis2010planning], though methodologies may clearly overlap.
-Broader approaches have also been taken: @paez2012measuring categories indicators on how they have been normatively and positively applied, and @levineCenturyEvolutionAccessibility2020 reviews accessibility's evolution along definition-based and application-based dimensions throughout the years.
-
-Among these reviews, we find @wuUnifyingAccess2020's contribution compelling for unifying accessibility measures: they demonstrate that most or all of these concepts can be unified into its main features: the relationship between travel-cost and corresponding reachable opportunities.
-Indeed, these authors demonstrate how previous classifications are special cases of travel-cost or opportunities constraints.
-From the initial Newtonian analogy, we interpret @wuUnifyingAccess2020's work as offering a positive contribution by circling back to re-unify mass $M$ and separation distance $D$; however, an important component is still missing, the proportionality constant $G$.
-
-As such, this paper's focus is on the proportionality constant.
-We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings.
-To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971.
-In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts.
-Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.
-
-# Introducing a family of accessibility measures
-
-## Gravity-based accessibility's cousin: Wilson's spatial interaction model
-<!--SUMMARY: Accessibility has the same Newtonian roots as Wilsons' family of spatial interaction models.
-Introduce SIM's formulation.
-Emphasize the interpretation that comes from $k$.
-While SIM has Newtonian roots, its proportionality constant(s) are actually balancing factors derived via entropy-maximisation, obtaining the interaction variable as a statistical average.
--->
-Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971].
-However, in many ways these spatial interaction models, commonly refered to as gravity models, continued to evolve independently from accessibility (@eq-accessibility-general).
-Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM.
-This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs.
-This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.
-
-To begin, the SIM as defined in Wilson [-@wilson1971], takes the following form: 
+There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance.
+At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
+
+After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determine it.
+In this way, $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+Alas, without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
+Hansen-style accessibility, therefore, is technically an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].
+
+# Wilson's family of spatial interaction models
+
+In his groundbreaking work, Wilson [-@wilson1971] defined a general spatial interaction model as follows:
+
 $$
@@
 
-<!--Explain the features of the gravity model of today (Wilson) -->
-Where $T_{ij}$ represents a $n$ x $m$ matrix of flows between an $n$ number of origins $i$ and an $m$ number of destinations $j$, $W_i^{(1)}$ and $W_j^{(2)}$ is a vector of mass attributes representing origin $i$ and destination $j$ respectively, and $f(c_{ij})$ is a $n$ x $m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
-In other words, $T_{ij}$ explicitly measures _interaction_ as a unit of trips and the single proportionality constant $k$ ensures the total sum of $T_{ij}$ represents the total flows in the data.
-It should be noted that the early analogy with the gravitational law proved too simplistic, and improvements to the SIM (@eq-phys-gravity-model) are the use of total trip ends (number of originating trips from zone $i$ are $O_i$ and ending trips at zone $j$ are $D_j$) instead of total populations and separation distance calibrated by a power $n$ (such as $P_i$, $P_j$, and $n$ in @eq-zipf, @eq-reilly and @eq-stewart-force).
-Finally, and of most important note in this paper, $k$ is a scale parameter that makes the overall equation proportional to the rate characteristic of the modeled phenomenon.
-
-Traditionally, the development of the SIM has put an emphasis on the interpretability of the resulting units [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
-But instead of relying on the heuristic Newtonian gravity analogy (e.g., some interaction between a mass at $i$ and a mass at $j$ seperated by some distance), entropy maximisation has emerged and continues to be a popular model building approach to generate the SIM [@ortuzar_2011_modelling].
-
-Namely, the entropy maximisation approach achieves stable results as a statistical average that represents the population.
-The approach works with individual interaction, assesses their probability of making a particular journey given given zonal and regional constraints, and the resulting interaction from @eq-phys-gravity-model is a statistical average [@wilson1971;@seniorGravityModellingEntropy1979].
-For instance, the interaction variable $T_ij$ is often in the unit of trips, so additional knowledge can be the total trip flows originating at each zone $W_i^{(1)}$ and the total trip flows terminating at zone $W_j^{(2)}$ can be integrated into zonal constraints as follows.
-
-The total flow constraint: @eq-constraint0-gravitymodel, where the total number of flows in the region equals $T$: <!--NOTE: in Wilson's entropy-max derivation, this is actually eq-constraint0_5-gravitymodel a TOTAL TRAVEL COST constraint:-->
+The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations.
+Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
+In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of the single proportionality constant $k$ is to ensure the total sum of $T_{ij}$ represents the total flows in the data.
+In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
+
+Traditionally, development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
+But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
+Entropy maximisation in this cases achieves stable results as a statistical average that represents the population.
+The approach works by assuming undifferentiated individual interactions, and assessing their probabilities of making a particular journey.
+The result of @eq-phys-gravity-model then is a statistical average [@wilson1971; @seniorGravityModellingEntropy1979].
+
+To ensure that $T_{ij}$ in @eq-phys-gravity-model is in the unit of trips, additional knowledge about the system is required.
+At the very least, this framework assumes that the total number of trips in the system $T$ is known, and therefore:
+
 $$
@@
 
-<!-- $$ -->
-<!-- \sum_i\sum_j T_{ij} c_{ij} = C -->
-<!-- $${#eq-constraint0_5-gravitymodel} -->
-
-The production constraint: the sum of trips ending at each $j$ is equal to the total flow originating from each $i$:
-$$
-\sum_j T_{ij} = W_i^{(1)}
+Additional information can be introduced. For example, when information is available about the total number of trips produced by each origin ($O_i$), the following constraint can be used:
+
+$$
+\sum_j T_{ij} = O_i
 $${#eq-constraint1-gravitymodel}
 
-The attraction constraint: the sum of trips starting at each $i$ is equal to the total flow ending at each $j$:
-$$
-\sum_i T_{ij} = W_j^{(2)}
+Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used:
+$$
+\sum_i T_{ij} = D_j
 $${#eq-constraint2-gravitymodel}
 
-The entropy maximization derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967).
+It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints could be imposed on the model.
+
+Parting from this, it is possible to derive a family of spatial interaction models.
+In Wilson's terms, there is the "unconstrained" (but actually total flow constrained) model in equation @eq-phys-gravity-model.
+The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
+
+$$
+k=\frac{T}{\sum_i\sum_j T_{ij}}
+$${#eq-total-flow-balancing-factor}
+
+When only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel holds, the resulting models are, in Wilson's terms, singly-constrained.
+Entropy maximisation leads to the following production-constrained model:
+
+$$
+T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
+$${#eq-production-constrained-gravitymodel}
+
+Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced by the actual mass, as measured by the trips produced.
+Also, there is no longer a single system-wide proportionality constant, but rather a balancing factor specific to each origin, namely $A_i$, which Wilson found to be as follows:
+
+$$
+A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
+$${#eq-production-constrained-balancing-factor}
+
+The attraction-constrained model, in turn, takes the following form:
+
+$$
+T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
+$${#eq-attraction-constrained-gravitymodel}
+
+Notice how now the proxy for the mass at the destination $W_j^{(2)}$ is replaced by the actual mass, as measured by the trips attracted.
+As before, destination-specific balancing factors $B_j$ were derived by Wilson as:
+
+$$
+B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-attraction-constrained-balancing-factor}
+
+When both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold, the resulting models is, in Wilson's terms, doubly-constrained, and takes the following form:
+
+$$
+T_{ij} = A_i B_j O_i D_j f(c_{ij})
+$${#eq-doubly-constrained-gravitymodel}
+
+In this model both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination. And, now the balancing factors are:
+
+$$
+\begin{array}{l}
+A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
+B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
+\end{array}
+$${#eq-doubly-constrained-balancing-factors}
+
+Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
+It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.
+Why could this be?
+In the next section we aim to address this question.
+
+# Accessibility and spatial interaction modelling: two divergent research streams
+
+<!--
+SUMMARY: Accessibility developed separate from SIM.
+
+SIM focuses on trips and was institutionalized to develop the US Interstate highway systems.
+Many guidelines and practical training was done - a lot of funding.
+SIM development has integrated accessibility in a few ways: 1) within the SIM itself as a balancing factor, 2) as an additional independent variable in the SIM, 3) SIM and HAM in parallel in the context of location allocation problems and within the steps of transportation demand modelling (trip distribution or trip generation).
+
+Accessibility research has separated SIM almost entirely.
+When Wilson is cited: 1) typically when using a different impedance function, 2) when using the SIM to predict trip flows for accessibility input or vice versa, 3) a few exceptional papers that interpret the doubly-constraints as an inverse HAM in the context of competitive accessibility, and 4) it was only until recently, that a singly-constrain was introduced in Soukhov et al.
+2023.
+-->
+
+```{r}
+#| label: load-data
+
+load(glue::glue(here::here(),
+                "/data/corpus.rda"))
+```
+
+```{r}
+#| label: calculate-cocMatrix
+#| cache: true
+
+# $A$ is a sparse adjacency matrix with sources cited by each document in the corpus
+A <- cocMatrix(corpus, 
+               Field = "CR", 
+               sep = ";")
+```
+
+```{r}
+#| label: coupling-matrix
+
+# The product of $A$ and its transpose gives the bibliographical coupling by cited references $B = A\cdot A^T$
+
+# $B$ is a square symmetric matrix where element $b_{ij}$ is the total number of cited references in common between document $i$ and document $j$.
+A higher value of $b_{ij}$ indicates a stronger relationship between documents $i$ and $j$ (more references in common).
+A value of zero indicates no references in common.
+B <- tcrossprod(A)
+```
+
+```{r}
+#| label: clean-coupling-matrix
+
+# Remove the diagonal (papers cited within a single source)
+B_nd <- B - Matrix::band(B, 0, 0)
+```
+
+```{r}
+#| label: create-coupling-network
+
+#These matrices are sparse symmetric matrices of "adjacencies" between documents in the corpus. Convert `B` to an `igraph` network without the diagonal
+coupling_net <- igraph::graph_from_adjacency_matrix(B, 
+                                            mode = "directed", 
+                                            weighted = TRUE,
+                                            diag = FALSE)
+```
+
+```{r}
+#| label: remove-A-memory
+
+rm(A)
+```
+
+```{r}
+#| label: convert-to-tidygraph
+
+# For convenience, convert the [{igraph}](https://r.igraph.org) object to a [{tidygraph}](https://www.data-imaginist.com/posts/2017-07-07-introducing-tidygraph/) object:
+coupling_net <- tidygraph::as_tbl_graph(coupling_net)
+```
+
+```{r}
+#| label: add-corpus-label-to-coupling-network
+
+# Add the `corpus` variable to `coupling_net`. Remember to activate the nodes.
+coupling_net <- coupling_net |>
+  tidygraph::activate("nodes") |>
+  left_join(tibble::rownames_to_column(corpus, 
+                                       var = "rowname") |> 
+              transmute(rowname, 
+                        corpus = corpus),
+            by = c("name" = "rowname"))
+```
+
+```{r}
+#| label: corpus-summary
+
+summary_coupling_network <- coupling_net |>
+  tidygraph::activate("nodes") |>
+  as_tibble() |>
+  mutate(corpus = factor(corpus)) |>
+  group_by(corpus) |>
+  summarize(n = n())
+
+# Number of documents in each corpus
+# Hansen
+n_h <- summary_coupling_network |> 
+  filter(corpus == "Hansen") |> 
+  pull(n)
+# Wilson
+n_w <- summary_coupling_network |> 
+  filter(corpus == "Wilson") |> 
+  pull(n)
+# Both
+n_b <- summary_coupling_network |> 
+  filter(corpus == "Both") |> 
+  pull(n)
+```
+
+The work of @hansen1959 and @wilson1971 responded to important developments in the Anglo-American world at the time, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
+Said dictates and needs were far from trivial.
+In the US alone, the Federal-Aid Highway Act of 1956 authorized the creation of the US Interstate Highway System, with a budget of more than one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+Accessibility at the time did not quite have that power: it did not quantify trips, but rather something fuzzier: the less tangible "potential for interaction".
+In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
+
+```{r}
+#| label: plot-docs-per-year
+#| include: false
+
+docs_per_year_plot <- ggplot(data = corpus |> 
+         group_by(corpus, PY) |> 
+         summarize(n = n()),
+         .groups = "drop") + 
+  geom_col(aes(x = PY, 
+               y = n,
+               fill = corpus),
+           color = "black") + 
+  xlab("Publication year") + 
+  ylab("Number of documents") +
+  theme_minimal()
+
+ggsave(filename = glue::glue(here::here(), 
+                             "/figures/docs_per_year_plot.png"))
+```
+
+```{r}
+#| label: fig-docs-per-year
+#| out-width: 70%
+#| fig-cap: "Historical pattern of publication: documents per year."
+
+docs_per_year_plot
+```
+
+To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959 and/or @wilson1971.
+We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
+As a result of this search we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, `r n_w  |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
+The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
+The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
+In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971 or both.
+
+<!--
+Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+However, within the same time period, accessibility remained largely within the academic realm.
+For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.].
+We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
+Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
+
+However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
+A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
+Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
+While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
+
+```{r}
+#| label: community-detection
+#| cache: true
+
+# Detect communities based on the bibliographic coupling. For repeatability set the random seed
+set.seed(979834789)
+coupling_net <- coupling_net |>
+  mutate(community = as.factor(tidygraph::group_infomap(weights = weight)))
+```
+
+-->
+
+After compiling this corpus of documents, we used the {bibliometrix} package [@bibliometrix2017] to create a bibliographical coupling matrix.
+In this matrix two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common.
+@fig-Fig1 presents the results.
+In this figure each symbol represents a document, and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents, meaning those that share more items in their lists of references, are plotted closer to each other.
+
+```{r}
+#| label: layout-for-coupling-net
+#| cache: true
+
+layout <- create_layout(coupling_net,
+                        layout = "stress",
+                        weights = weight)
+```
+
+```{r}
+#| label: plot-coupling-network
+#| include: false
+
+coupling_net_plot <- ggplot(data = layout) +
+  geom_point(aes(x = x,
+                 y = y,
+                 shape = corpus,
+                 color = corpus),
+             size = 3,
+             alpha = 0.7) +
+  geom_point(data = layout |>
+               filter(corpus == "Both"),
+             aes(x = x,
+                 y = y),
+             shape = 1,
+             size = 3) +
+  geom_point(data = layout |>
+               filter(corpus == "Wilson"),
+             aes(x = x,
+                 y = y),
+             shape = 0,
+             size = 3,
+             alpha = 0.5) +
+  theme_void()
+
+ggsave(filename = glue::glue(here::here(), 
+                             "/figures/Han_Wilson_bib_coupling_plot.png"))
+```
+
+```{r}
+#| label: fig-Fig1
+#| out-width: 70%
+#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959) and/or Wilson (1971)."
+
+coupling_net_plot
+```
+
+```{r}
+#| label: center-of-mass-of-Hansen
+
+# This is the center of mass of the documents that cite Hansen
+hansen_center_of_mass <- layout |>
+  filter(corpus == "Hansen") |>
+  summarize(x = mean(x),
+            y = mean(y))
+```
+
+```{r}
+#| label: center-of-mass-of-Wilson
+
+# This is the center of mass of the documents that cite Wilson
+wilson_center_of_mass <- layout |>
+  filter(corpus == "Wilson") |>
+  summarize(x = mean(x),
+            y = mean(y))
+```
+
+```{r}
+#| label: row-names-by-corpus
+
+# Extract the row names by corpus to be able to partition the coupling matrix
+hansen_names <- rownames(corpus |>
+                           filter(corpus == "Hansen"))
+
+wilson_names <- rownames(corpus |>
+                           filter(corpus == "Wilson"))
+
+both_names <- rownames(corpus |>
+                           filter(corpus == "Both"))
+```
+
+```{r}
+#| label: partition-of-coupling-matrix
+
+# With the row names we can extract submatrices to examine the strength of coupling within and between the three corpora, i.e., documents that cite only Hansen, only Wilson, or both
+hansen_to_wilson <- B_nd[hansen_names, wilson_names]
+
+hansen_to_hansen <- B_nd[hansen_names, hansen_names]
+
+wilson_to_wilson <- B_nd[wilson_names, wilson_names]
+
+both_to_both <- B_nd[both_names, both_names]
+
+both_to_hansen <- B_nd[both_names, hansen_names]
+
+both_to_wilson <- B_nd[both_names, wilson_names]
+```
+
+```{r}
+#| label: summary-of-coupling-within-and-between-corpora
+
+# The number of common citations per pair of documents can be calculated as the sum of all common citations in the matrix, divided by the number of pair-paper combinations:
+# Average coupling for the total corpora
+c_t <- sum(sum(B_nd))/(dim(B_nd)[1] * (dim(B_nd)[2]))
+# Average coupling for papers that cite both Hansen and Wilson
+c_b <- sum(sum(both_to_both))/(dim(both_to_both)[1] * (dim(both_to_both)[2]))
+# Average coupling for papers that cite only Hansen
+c_h <- sum(sum(hansen_to_hansen))/(dim(hansen_to_hansen)[1] * (dim(hansen_to_hansen)[2]))
+# Average coupling for papers that cite only Wilson
+c_w <- sum(sum(wilson_to_wilson))/(dim(wilson_to_wilson)[1] * (dim(wilson_to_wilson)[2]))
+# Average coupling between papers that cite both Hansen and Wilson, and those that cite only Hansen
+c_h_to_b <- sum(sum(both_to_hansen))/(dim(both_to_hansen)[1] * dim(both_to_hansen)[2])
+# Average coupling between papers that cite both Hansen and Wilson, and those that cite only Wilson
+c_w_to_b <- sum(sum(both_to_wilson))/(dim(both_to_wilson)[1] * dim(both_to_wilson)[2])
+# Average coupling between papers that cite only Hansen, and those that cite only Wilson
+c_h_to_w <- sum(sum(hansen_to_wilson))/(dim(hansen_to_wilson)[1] * dim(hansen_to_wilson)[2])
+```
+
+Further examination of the bibliographical coupling matrix allows us to check the coupling strength within and between subgroups of documents (i.e., those that cite only Hansen, only Wilson, or both papers).
+First, we note that when we look at the three subgroups of documents pooled together, the average number of references shared by a pair of documents is `r round(c_t, 2)`.
+@tbl-coupling-results presents the average number of items cited in common within each subgroup, as well as between subgroups.
+As seen in the table, the coupling within subgroup tends to be higher than when we consider the full corpus with all three groups together.
+The tightest coupling is seen for the group of documents that cite both Hansen and Wilson, with an average number of references in common of `r prettyNum(c_b, digits = 3)`.
+The weakest coupling is between the Hansen and Wilson subgroups, with only `r prettyNum(c_h_to_w, digits = 1)` references in common on average, which indicates a large degree of decoupling (independence) between these bodies of work.
+
+```{r}
+#| label: tbl-coupling-results
+#| tbl-cap: "Average coupling strength within and between groups of documents. The average coupling strength is the number of references shared on average by a pair of documents."
+
+data.frame(corpus = c(glue::glue("Hansen (n = {n_h |> prettyNum(big.mark = ",")})"), 
+                      glue::glue("Wilson (n = {n_w |> prettyNum(big.mark = ",")})"),
+                      glue::glue("Both (n = {n_b  |> prettyNum(big.mark = ",")})")),
+           Hansen = c(prettyNum(c_h, digits = 3),
+                      prettyNum(c_h_to_w, digits = 1),
+                      prettyNum(c_h_to_b, digits = 3)),
+           Wilson = c("-",
+                      prettyNum(c_w, digits = 3),
+                      prettyNum(c_w_to_b, digits = 3)),
+           Both = c("-",
+                    "-",
+                    prettyNum(c_b, digits = 3))
+           ) |>
+  gt() |>
+  tab_footnote(footnote = glue::glue("Average coupling for the pooled set of documents (n = {(n_h + n_w + n_b) |> prettyNum(big.mark = ",")}) is {round(c_t, 2)}")) |>
+  tab_options(footnotes.font.size = "small") |>
+  as_latex()
+```
+
+<!-- 
+In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
+Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
+
+Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
+Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
+Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
+
+However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
+This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
+Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
+Some works went beyond HAM and considered other accessibility expressions of individual choice.
+For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
+
+Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility -eek?
+I don't think I can say this--.
+We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
+With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
+
+In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
+@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
+
+Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+
+Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
+Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+
+And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
+However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
+Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
+
+Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
+Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
+However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
+A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
+How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
+-->
+
+<!--
 The approach is based on partition functions from statistical mechanics (the physics of gases), not Newtonian gravitation.
@@
 Using these constraints, the SIM @eq-phys-gravity-model is distinguished into three general cases.
-These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling]:
-
-- The total constrained case: neither @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold, only @eq-constraint0-gravitymodel - the total number of trips, is used;
-- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
-- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->
-
-In each of these cases, the single proportionality constant $k$ is replaced by balancing factors $A_i$, $B_j$, and $K$ as follows to ensure the case-specific constraints are satisfied:
-$$
-A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}
-$${#eq-bf-A-gravitymodel}
-
-$$
-B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})}
-$${#eq-bf-B-gravitymodel}
-<!--NOTE: pretty sure K is not consistent with entropy-maximisation?-->
-$$
-K = \frac{T}{\sum_i \sum_j W_i^{(1)} W_j^{(2)} f(c_{ij})}
-$${#eq-bf-K-gravitymodel}
-
-For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$.
-For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel.
-For both singly-constrained cases, $K$ also holds.
-For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint.
-
-These constraints (@eq-bf-A-gravitymodel, @eq-bf-A-gravitymodel, and @eq-bf-K-gravitymodel) and the SIM @eq-phys-gravity-model can be expressed not just for the variable of _interaction_ (e.g., the number of trips) but also as a variable of _potential interaction_.
-Before detailing SIM's formulation for accessibility, however, background on the development of accessibility and SIM independent from eachother is explained.
-
-## Background on accessibilities' distinct development from spatial interaction models
+These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling].
+-->
+
+# A family of accessibility measures
+
+As we argued above, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact over the years.
+This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis.
+This is intriguing, considering that Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the similarity between the denominator in @eq-production-constrained-balancing-factor and Hansen's accessibility indicator.
+In fact, Wilson noted that Hansen's accessibility was the inverse of $A_i$, which offered a potential interpretation for the balancing factor (p.
+10):
+
+$$
+S_i = \frac{1}{A_i} = \sum_j W_j^{(2)} f(c_{ij})
+$${#eq-Ai-as-accessibility}
+
+This, however, only helps to illustrate the ease with which confusion may arise in this context.
+The relationship in @eq-Ai-as-accessibility appears to be misguided when we note that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force:
+
+$$
+F = G\frac{M_1\cdot M_2}{d_{12}^2}
+$$
+
+\noindent to the pairwise population potential:
+$$
+V_1 = G\frac{M_2}{d_{12}}
+$$
+
+\noindent and from there to the system-wide population potential:
+$$
+V_i = \sum_j \frac{M_j}{d_{ij}}
+$$
+
+It is useful to think of the accessibility as a partial sum of the spatial interaction, before we return to Wilson's general model:
+
+$$
+T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
+$$
+
+Then, we define the _potential for interaction_ (i.e., accessibility) as follows:
+
+$$
+V_{ij} = k W_j^{(2)} f(c_{ij})
+$${#eq-access}
+
+\noindent where $V_{ij}$ is the potential for interaction.
+Similar to @eq-phys-gravity-model, $W_j^{(2)}$ is the mass at the destination, and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+
 <!--
-NOTE: This section is a tough one... please suggest ideas...
-
-SUMMARY: Accessibility developed separate from SIM.
-
-SIM focuses on trips and was institutionalized to develop the US Interstate highway systems.
-Many guidelines and practical training was done - a lot of funding.
-SIM development has integrated accessibility in a few ways: 1) within the SIM itself as a balancing factor, 2) as an additional independent variable in the SIM, 3) SIM and HAM in parallel in the context of location allocation problems and within the steps of transportation demand modelling (trip distribution or trip generation).
-
-Accessibility research has separated SIM almost entirely.
-When Wilson is cited: 1) typically when using a different impedance function, 2) when using the SIM to predict trip flows for accessibility input or vice versa, 3) a few exceptional papers that interpret the doubly-constraints as an inverse HAM in the context of competitive accessibility, and 4) it was only until recently, that a singly-constrain was introduced in Soukhov et al.
-2023.
+.
+Potential interaction is the preoccupation of accessibility measures.
+The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of SIMs, however with different interpretation.
+In this section, we explicitly define these cases.
 -->
 
-Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-However, within the same time period, accessibility remained largely within the academic realm.
-For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., <!--studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.-->].
-We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
-Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
-
-However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
-A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
-Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
-While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
-
-To capture this sentiment in the literature, @fig-Fig1 demonstrates the bibliometric coupling of papers that cites @hansen1959 (one of accessibility's "seminal" works), @wilson1971 (one of SIM's "seminal" works), and both @hansen1959 and @wilson1971.
-@fig-Fig1 is based on a SCOPUS search of papers that cite @hansen1959's DOI (2,134 documents) and another SCOPUS search of papers that cite @wilson1971's DOI (368 documents) conducted on November 9, 2024 and a bibliometric coupling matrix created using the {bibliometrix} R package [@bibliometrix2017].
-
-```{r plot-figure1, out.width=600, out.height=600}
-#| label: fig-Fig1
-#| fig-cap: "The bibliometric coupling of papers that cite Hansen (1959), Wilson (1971) and both Hansen (1959) and Wilson (1971)."
-include_graphics(glue(here(), "/data-raw/Han_Wilson_bib_coupling_plot.png"))
-```
-
-In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
-Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
-
-Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
-Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
-Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
-
-However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
-This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
-Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
-Some works went beyond HAM and considered other accessibility expressions of individual choice.
-For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
-
-Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility <!--eek?
-I don't think I can say this-->.
-We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
-With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
-
-In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
-@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
-
-Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-
-Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
-Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-
-And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
-However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
-Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
-
-Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
-Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
-However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
-A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
-How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
-
+From here, we can define various accessibility measures depending on how we choose to constrain them.
+
+## Unconstrained accessibility
+
+Unconstrained accessibility results when we set the proportionality constant/balancing factor to one in @eq-access:
+
+$$
+V^0_{ij} = W_j^{(2)} f(c_{ij})
+$${#eq-unconstrained-access}
+
+Then, the partial sum becomes:
+
+$$
+V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
+$${#eq-unconstrained-accessibility}
+
+We can see that in this case we simply have Hansen's accessibility. In general, the following does not hold true:
+
+$$
+\sum_i V^0_{i} = O
+$${#eq-no-constraint0-accessibility}
+
+\noindent where $O$ is the total number of opportunities in the system.
+In other words, the sum of all accessibility scores will generally be different from the total of the opportunities available in the region.
+To complicate matters, the difference will depend on the choice of impedance function $f(c_{ij})$ and any associated parameters, as well as the number of locations $n$ for which an accessibility score is calculated.
+For this reason, unconstrained accessibility can only be appropriately used as an ordinal variable to make comparisons of size (greater than, less than, equal to), but not to calculate intervals (the magnitude of differences) or ratios.
+
+## Total opportunity constrained accessibility
+
+We can choose to retain the proportionality constant in @eq-access to obtain the following:
+
+$$
+V^T_{ij} = kW_j^{(2)} f(c_{ij})
+$${#eq-total-constrained-access}
+
+Our accessibility measure now becomes:
+
+$$
+V^T_i = \sum_j V^T_{ij} = k\sum_j W^{(2)}_jf(c_{ij}) = kV^0_i
+$${#eq-total-constrained-accessibility}
+
+The constraint that we impose in this case is the total number of opportunities in the region $O$ (compare to @eq-constraint0-gravitymodel):
+
+$$
+\sum_i\sum_j V^T_{ij} = O
+$${#eq-total-constraint-accessibility}
+
+Substituting @eq-total-constrained-accessibility in @eq-total-constraint-accessibility, and solving for $k$, we have that (compare to the constant in the total flow spatial interaction model, Equation 2.11 in @cliff_evaluating_1974):
+
+$$
+k=\frac{O}{\sum_i\sum_j V^0_{ij}}
+$${#eq-total-opportunity-balancing-factor}
+
+Let us write our total opportunity constrained accessibility as follows:
+
+$$
+V^T_i = k\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{Of(c_{ij})}{\sum_i\sum_j V^0_{ij}} = \sum_j W^{(2)}_j\frac{Of(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
+$$
+
+Further, we can see that, since $O$ and $W^{(2)}_j$ are opportunities, the following term is dimensionless:
+
+$$
+\kappa = \frac{Of(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination:
+
+$$
+V^T_i = \kappa\sum_j W^{(2)}_j 
+$$
+
+Notice that the role of $\kappa$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+Constant $\kappa$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+For this reason we call it a proportional allocation factor.
+
+## Singly-constrained accessibility: attraction-constrained
+
+If we decide to introduce additional information into our analysis, say, the number of opportunities _by destination_ (i.e., $O_j$), we can impose the following constraint (compare to @eq-constraint2-gravitymodel):
+
+$$
+\sum_i V^D_{ij} =  D_j
+$${#eq-attraction-constraint}
+
+The underlying spatial interaction model is now the attraction-constrained model in @eq-attraction-constrained-gravitymodel, and our accessibility measure becomes:
+
+$$
+V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
+$${#eq-attraction-constrained-accessibility}
+
+\noindent where $W_i^{(1)}$ is a measure of the mass at origin $i$ (i.e., the opportunity-seeking population). The corresponding balancing factor, as per Wilson, is:
+
+$$
+B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
+$$
+
+If we introduce this balancing factor in @eq-attraction-constrained-accessibility we obtain:
+
+$$
+V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-attraction-constrained-accessibility-with-balancing-factor}
+
+Further, we define the following proportional allocation factor:
+
+$$
+\kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-attraction-constrained-proportional-allocation-factor}
+
+After this, it is possible to rewrite @eq-attraction-constrained-accessibility-with-balancing-factor as:
+
+$$
+V^D_{i} = \sum_j  \kappa^D_{ij} D_j
+$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
+
+@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^D_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
+As before, $\kappa^D_{ij}$ is dimensionless and $V_i^D$ is in the units of opportunities $D_j$.
+The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that there is no competition for them.
+These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
+
+$$
+v^D_{i} = \frac{V^D_{i}}{W^{(1)}_i}
+$$
+
+As an additional point, the constraint in @eq-attraction-constraint ensures that the following also holds:
+
+$$
+\sum_i V^D_{i} = O
+$$
+
+For this reason, $\frac{V_i^D}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system
+
+## Singly-constrained accessibility: production-constrained
+
+Not the most common application of accessibility analysis, but if the origins and destinations are transposed, we can define a measure of market potential that preserves the population.
+The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $V^P_i$ becomes:
+
+$$
+V^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
+$${#eq-attraction-constrained-accessibility}
+
+In this case the measure is constrained by the population _by origin_ (i.e., $P_i$). Compare to @eq-constraint2-gravitymodel:
+
+$$
+\sum_j V^P_{ij} =  P_i
+$${#eq-attraction-constraint}
+
+The corresponding balancing factor, as per Wilson, is:
+
+$$
+A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
+$$
+
+Following the same logic as in the preceding section, one arrives at the following expression:
+
+$$
+V^P_{j} = \sum_i  \kappa^P_{ij} P_i
+$${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
+
+\noindent with:
+
+$$
+\kappa^P _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+$${#eq-attraction-constrained-proportional-allocation-factor}
+
+In addition to market potential, $V_j^P$ can be interpreted as a constrained measure of "facility crowdedness" as in @wang_inverted_2018.
+This "crowdedness" now satisfies, in addition to @eq-attraction-constraint, the following:
+
+$$
+\sum_j V^P_{j} = P
+$$
+
+\noindent where $P$ is the total population of the region. In this way, $\frac{V_j^P}{P}$ can be interpreted as the proportion of the total population serviced by location $j$.
+
+<!--
+
+## Gravity-based accessibility's cousin: Wilson's spatial interaction model
+Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971].
+However, in many ways these spatial interaction models, commonly refered to as gravity models, continued to evolve independently from accessibility (@eq-accessibility-general).
+Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM.
+This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs.
+This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.
+
+-->
+
+<!--
 ## Interpreting SIM constraints for accessibility
+
+SUMMARY: Introduce the SIM constraints generally, and explain them in terms of accessibility (e.g., P_i as weight at origin and O_j as weight at destination)
+
+Then, detail a synthetic example for 'non-constrained' (e.g., HAM), Total constrained, Singly-constrained (both versions), Doubly-constrained and finally Hybrid constrained (?maybe?).
 
@@
 $${#eq-constraint2-access}
+
 \noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.
@@
 - The singly-constrained case: either @eq-constraint1-access holds (origin-constrained)  or @eq-constraint2-access holds (destination-constrained) -- can only be in any combination of "potential for interaction" units, e.g., units of "opportunities" or "population" or "population for opportunities" or "opportunities for population;
-- The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously <!--production-attraction-->
+- The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously production-attraction
 
@@
 - Nonconstrained cases: current practice, the HAM, where @eq-constraint0-access, @eq-constraint1-access, nor  @eq-constraint2-access hold.
-- HYBRID CASE: examples of where @eq-constraint1-access or @eq-constraint2-access partly hold <!-- BFCA? Others?-->
+- HYBRID CASE: examples of where @eq-constraint1-access or @eq-constraint2-access partly hold BFCA? Others?
 
@@
 We assume the the travel impedance ($f(c_{ij})$) is the associated travel propensity of $f(c_{ij}) = 1/C$.
+-->
+
+## Doubly-constrained
+
+In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
+In this case, the units of the population and opportunities must be able to be understood as the same.
+E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are saught by population at each $i$ must both be known to match.
+
+In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
+In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
+If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
+
+Literature has explored 'doubly-constrained' but -- not the same: , etc. Ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
+
+# Simple numerical example
 
@@
 
-### Nonconstrained accessibility
+## Unconstrained accessibility
 
@@
 Working backward from this expression and assumptions, $S_{ij}$ can be expressed more generally as well.
-
-$$
-S_{i} = 1*\sum_j O_j f(c_{ij}) \\
-S_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij}) = 1*1*O_j f(c_{ij}) 
-$$
 
@@
 
-### Total constrained accessibility
-
-In the total constrained case, $k$ is replaced with $K$ (@eq-access satisfying only constraint @eq-constraint0-access) as follows for $V_{ij}$ and $V_{i}$:
-$$
-V_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij})\\
-V_{i}' = K\sum_j W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
+## Total constrained accessibility
 
 Solved for our example, $K$ is equal to the total accessibility $V$, the only thing we know: either all the opportunities $O$ or all the population $P$, divided by remaining part of the equation.
-In the following, we assume our spatial weights matrix only reflects $W_j^(2)$ as $O_j$ and $W_i^(1)$ is uniform (1):
-$$
-K = \frac{V}{\sum_{i}\sum_{j} 1* O_j f(c_{ij})}\\
-K = \frac{V}{\frac{O_1}{c_{1,1}}+\frac{O_1}{c_{2,1}}+\frac{O_1}{c_{3,1}}+\frac{O_2}{c_{1,2}} +... \frac{O_3}{c_{3,3}}
+In the following, we assume our spatial weights matrix only reflects $W_j^(2)$ as $O_j$ and $W_i^{(1)}$ is uniform (1):
+$$
+\begin{array}{l}
+k = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
+k = \frac{V}{\frac{O_1}{c_{11}}+\frac{O_1}{c_{21}}+\frac{O_1}{c_{31}}+\frac{O_2}{c_{12}} +... \frac{O_3}{c_{33}}
 }\\
-K = \frac{790}{568.4} \\
-K = 1.389866
+k = \frac{790}{568.4} \\
+k = 1.389866
+\end{array}
 $$
@@
 
-### Singly-constrained case
-
+## Singly-constrained accessibility: attraction-constrained
+
+<!--
 The singly-constrained case can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
@@
 It was only until the recent work of [@soukhovIntroducingSpatialAvailability2023], where the units of accessibility given competition, e.g., the spatial availability, was introduced with the focus on both competition and balanced units.
+-->
 
 Returning to our simple example, the opportunity-constrained singly-constrained case would yield the following $B_{j}$ values assuming $A_{i}$ is 1:
-$$
-B_{j} = \frac{1}{\sum_i 1* P_i f(c_{ij})}\\
-B_{1} = \frac{1}{P_1 f(c_{1,1}) + P_2 f(c_{1,2}) + P_3 f(c_{1,3})}\\
-.., j=2, j=3\\
+
+$$
+B_{j} = \frac{1}{\sum_i P_i f(c_{ij})}\\
+$$
+
+$$
+\begin{array}{l}
 B_{1} =  \frac{1}{160/2+450/15+180/5} = 0.006849315 \\ 
 B_{2} =  \frac{1}{160/15+450/2+180/10} = 0.003942181 \\
-B_{3} =  \frac{1}{160/5+450/10+180/2} = 0.005988024 \\
-$$
+B_{3} =  \frac{1}{160/5+450/10+180/2} = 0.005988024
+\end{array}
+$$
+
 ```{r}
@@
 ```
+
 (NOTE: notably, when $B_j$ is multiplied by $P_i*f(c_{ij})$, it is a proportion (like spatial availability's total population&travel-cost balancing factor) representing how much of $O_j$ is allocated to each $i$)
@@
 In $V_i''^{(1)}$, the following is $A_i$ assuming $B_{j}$ is 1:
-$$
-A_{i} = \frac{1}{\sum_j 1* O_j f(c_{ij})}\\
-A_{i=1} = \frac{1}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\
-.., i=2, i=3\\
+
+$$
+A_{i} = \frac{1}{\sum_j O_j f(c_{ij})}\\
+$$
+
+$$
+\begin{array}{l}
 A_{1} =  \frac{1}{160/2+437/15+193/5} = 0.006768953 \\ 
 A_{2} =  \frac{1}{160/15+437/2+193/10} = 0.004024685 \\
-A_{3} =  \frac{1}{160/5+437/10+193/2} = 0.005807201 \\
-$$
+A_{3} =  \frac{1}{160/5+437/10+193/2} = 0.005807201
+\end{array}
+$$
+
 ```{r}
@@
 ```
+
 (AGAIN NOTE: when $A_i$ is multiplied by $O_j*f(c_{ij})$, it is a proportion (like spatial availability's total balancing factor) representing how much of $P_i$ is allocated from each $i$, I think.
@@
 
-### Doubly-constrained case (population-opportunity constrained)
-
-In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
-In this case, the units of the population and opportunities must be able to be understood as the same.
-E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are saught by population at each $i$ must both be known to match.
-
-In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
-In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
-If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
-
-Literature has explored 'doubly-constrained' but -- not the same: ,etc. 
-ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020
+## Doubly-constrained case (population-opportunity constrained)
 
@@
 
+<!--
 ### Hybrid singly-constrained case: 
-<!--NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA? -->
+NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA?
 Balanced floating catchment areas (not truly doubly constrained)
 
+-->
+
 # Empirical examples
@@
 
-# Discussion
+# A note on interpretation and possible applications
 
@@
 
+<!--
 # APPENDIX (some musings... destined for the commented-out graveyard)
@@
 ```
+
 Starting off with this opportunity matrix, namely - at i=1, only a fixed number of opportunities are accessible to this origin. same for i=2, i=3.
@@
 ```
+
+-->
 

### 2025-02-10 paper/manuscript.qmd
Adjusted the simple numerical example

@@
 library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
-# library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
-# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
-# library(patchwork) # The Composer of Plots
-# library(RColorBrewer) # ColorBrewer Palettes
-# library(sf) # Simple Features for R
-# library(scales) # Scale Functions for Visualization
-# library(shadowtext) # Shadow Text Grob and Layer
-# library(skimr) # Compact and Flexible Summaries of Data
-# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
-# library(tidyr) # Tidy Messy Data
-# library(tmap) # Thematic Maps
-# library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
-# library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
-# library(ggspatial) # Spatial Data Framework for ggplot2
-# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
-# library(here) # A Simpler Way to Find Your Files
 ```
@@
 The TOTALs are the $V_i$ and $V_j$, the total number of jobs spatially available at $i$ and the number of jobs spatially available from $j$.
-```{r}
-known_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE)
-#unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE)
-P_i <- known_matrix |> rowSums()
-O_j <- known_matrix |> colSums()
-V <- ifelse(sum(P_i)==sum(O_j), sum(O_j),NA)
-knowns_longer <- cbind(known_matrix,P_i)
-knowns_longer <- rbind(knowns_longer,cbind(t(O_j),T))
-
-rownames(knowns_longer) <- c("1","2","3", "TOTAL") 
-colnames(knowns_longer) <- c("1","2","3", "TOTAL")
-
-knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE)
-```
-
-And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows:
-```{r}
-C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE)
-C_longer <-C
-rownames(C_longer) <- c("1","2","3") 
-colnames(C_longer) <- c("1","2","3") 
-C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE)
-```
+
+The following vectors are the population by origin ($O_i$) and the opportunities by destination $D_j$:
+
+<!--Check here, the number of physicians per 10,000 in Canada in 2022 was 24.97
+https://en.wikipedia.org/wiki/List_of_countries_by_number_of_physicians -->
+```{r}
+#| label: simple-numerical-example-opportunities-and-population
+
+id <- c("1", "2", "3")
+O_i <- c(4, 10, 6) 
+D_j <- c(160, 150, 180)
+LU <- data.frame(id, O_i, D_j)
+```
+
+```{r}
+LU
+```
+
+Think of $O_i$ as population in 10,000 and $D_j$ as number of physicians. The Provider-to-Population-Ratio (PPR) is:
+```{r}
+sum(LU$D_j) / sum(LU$O_i)
+```
+
+And the cost of travel from $i$ to $j$ is as follows:
+```{r}
+#| label: simple-numerical-example-cost
+
+# Simulate a table with OD cost for the simple example, beginning with the ids, the cost, and then two impedance functions 
+C <- expand.grid(oid = c("1", "2", "3"),
+                 did = c("1", "2", "3")) |>
+  mutate(cost = c(2, 15, 5, 15, 2, 10, 5, 10, 2)) |>
+  mutate(f1 = exp(-1.00000 * cost^2), 
+         f2 = exp(-0.01000 * cost^2),
+         f3 = exp(-0.00001 * cost^2))
+```
+
+```{r}
+C |> select(oid, did, cost)
+```
+
+For comparison, use three different impedance functions:
+$$
+\begin{array}{l}
+f_1(c_{ij}) = exp(-c_{ij}^2)\\
+f_2(c_{ij}) = exp(-0.01\cdot c_{ij}^2)\\
+f_3(c_{ij}) = exp(-0.00001\cdot c_{ij}^2)
+\end{array}
+$$
 
@@
 ```{r}
-O_P_non <- t(matrix(rep(O_j, 3), 3, 3))
-O_P_non_longer <- O_P_non
-rownames(O_P_non_longer) <- c("1","2","3") 
-colnames(O_P_non_longer) <- c("1","2","3")
-O_P_non_longer
-```
-
-The resulting matrix of $S_{ij}$ along with the $S_i$ and $S_j$ totals for this example:
-```{r}
-S_non <- O_P_non / C
-
-P_i_non <- S_non |> rowSums()
-O_j_non <- S_non |> colSums()
-S_TOTAL_non <- ifelse(sum(P_i_non|>round())==sum(O_j_non|>round()), sum(S_non), NA)
-S_non_longer <- cbind(S_non,P_i_non)
-S_non_longer <- rbind(S_non_longer, cbind(t(O_j_non),S_TOTAL_non))
-
-rownames(S_non_longer) <- c("1","2","3", "TOTAL") 
-colnames(S_non_longer) <- c("1","2","3", "TOTAL")
-
-S_non_longer |> round(digits = 2) #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
+#| label: simple-numerical-example-OD-table
+
+OD <- C |>
+  left_join(LU |>
+              select(-D_j),
+            by = c("oid" = "id")) |>
+  left_join(LU |>
+              select(-O_i),
+            by = c("did" = "id"))
+```
+
+```{r}
+unc_acc_ij <- OD |>
+  group_by(oid) |>
+  reframe(V_unc_ij_1 = D_j * f1,
+            V_unc_ij_2 = D_j * f2,
+            V_unc_ij_3 = D_j * f3)
+
+unc_acc <- unc_acc_ij |>
+  group_by(oid) |>
+  summarize(V_unc_i_1 = sum(V_unc_ij_1),
+            V_unc_i_2 = sum(V_unc_ij_2),
+            V_unc_i_3 = sum(V_unc_ij_3))
+
+unc_acc
+```
+
+Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
+```{r}
+unc_acc$V_unc_i_1[2] - unc_acc$V_unc_i_1[1]
+unc_acc$V_unc_i_2[2] - unc_acc$V_unc_i_2[1]
+unc_acc$V_unc_i_3[2] - unc_acc$V_unc_i_3[1]
+
+unc_acc$V_unc_i_1[2]/unc_acc$V_unc_i_1[1]
+unc_acc$V_unc_i_2[2]/unc_acc$V_unc_i_2[1]
+unc_acc$V_unc_i_3[2]/unc_acc$V_unc_i_3[1]
+```
+
+This illustrates how the differences and ratios are not meaningful.
+
+The total accessibility compared to the number of opportunities:
+```{r}
+unc_acc <- unc_acc |> 
+  left_join(LU |>
+              select(-D_j),
+            by = c("oid" = "id"))
+
+unc_acc  |>
+  summarize(V_unc_1 = sum(V_unc_i_1),
+            V_unc_2 = sum(V_unc_i_2),
+            V_unc_3 = sum(V_unc_i_3))
+
+LU |>
+  summarize(O = sum(D_j))
+```
+
+And the ratio of accessibility to population compared to the ratio of opportunities to population:
+```{r}
+# Accessibility divided by population
+unc_acc |> 
+  mutate(v_unc_1 = V_unc_i_1/O_i,
+            v_unc_2 = V_unc_i_2/O_i,
+            v_unc_3 = V_unc_i_3/O_i)
+
+# Sum of the accessibility divided by the total population
+unc_acc |> 
+  summarize(v_unc_1 = sum(V_unc_i_1)/sum(O_i),
+            v_unc_2 = sum(V_unc_i_2)/sum(O_i),
+            v_unc_3 = sum(V_unc_i_3)/sum(O_i))
+
+# Sum of the opportunities divided by the total population
+LU |> 
+  summarize(PPR = sum(D_j)/sum(O_i))
+```
+
+Notice the wild differences.
+Accessibility according to impedance function 3 basically says that everyone has access to all opportunities all at once.
+Accessibility according to function 1, on the other hand, says that each population has accessibility to around 3 opportunities, which is absurd: the number of opportunities available locally are in the order of 150-180.
+One should not have to adjust the impedance function to obtain "reasonable" results [EXAMPLE]; instead, the impedance function should reflect the travel behavior.
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
 
@@
 
+**END OF PREVIOUS TEXT**
+
 ## Total constrained accessibility
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
 
@@
 
-```{r}
-K <- V/sum(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3))
-V_tot <- K * S_non
-
-P_i_tot <- V_tot |> rowSums()
-D_j_tot <- V_tot |> colSums()
-V_TOTAL_tot <- ifelse(sum(P_i_tot)==sum(D_j_tot), sum(V_tot), NA)
-V_tot_longer <- cbind(V_tot,P_i_tot)
-V_tot_longer <- rbind(V_tot_longer, cbind(t(D_j_tot),V_TOTAL_tot))
-
-rownames(V_tot_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_tot_longer) <- c("1","2","3", "TOTAL")
-
-V_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-While we can see that the 'number of potentially reachable opportunities' `r V` results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model).
+**END OF PREVIOUS TEXT**
+
+Calculate the balancing factors/proportionality constants:
+```{r}
+k_tot <- unc_acc_ij |> 
+  summarize(k1 = sum(LU$D_j)/sum(V_unc_ij_1),
+            k2 = sum(LU$D_j)/sum(V_unc_ij_2),
+            k3 = sum(LU$D_j)/sum(V_unc_ij_3))
+```
+
+Use these proportionality constants to calculate the total opportunity constrained accessibility:
+```{r}
+tot_acc_ij <- unc_acc_ij |>
+  mutate(V_tot_ij_1 = k_tot$k1 * V_unc_ij_1,
+         V_tot_ij_2 = k_tot$k2 * V_unc_ij_2,
+         V_tot_ij_3 = k_tot$k3 * V_unc_ij_3)
+
+tot_acc <- tot_acc_ij |>
+  group_by(oid) |>
+  summarize(V_tot_i_1 = sum(V_tot_ij_1),
+            V_tot_i_2 = sum(V_tot_ij_2),
+            V_tot_i_3 = sum(V_tot_ij_3))
+
+tot_acc |>
+  select(oid, starts_with("V_"))
+```
+
+Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
+```{r}
+tot_acc$V_tot_i_1[2] - tot_acc$V_tot_i_1[1]
+tot_acc$V_tot_i_2[2] - tot_acc$V_tot_i_2[1]
+tot_acc$V_tot_i_3[2] - tot_acc$V_tot_i_3[1]
+
+tot_acc$V_tot_i_1[2]/tot_acc$V_tot_i_1[1]
+tot_acc$V_tot_i_2[2]/tot_acc$V_tot_i_2[1]
+tot_acc$V_tot_i_3[2]/tot_acc$V_tot_i_3[1]
+```
+
+This illustrates how the differences and ratios are now meaningful.
+
+The total accessibility compared to the number of opportunities:
+```{r}
+tot_acc <- tot_acc |> 
+  left_join(LU |>
+              select(-D_j),
+            by = c("oid" = "id"))
+
+tot_acc  |>
+  summarize(V_tot_1 = sum(V_tot_i_1),
+            V_tot_2 = sum(V_tot_i_2),
+            V_tot_3 = sum(V_tot_i_3))
+
+LU |>
+  summarize(O = sum(D_j))
+```
+
+This shows how the number of opportunities is preserved.
+
+And the ratio of accessibility to population compared to the ratio of opportunities to population:
+```{r}
+# Accessibility divided by population
+tot_acc |> 
+  mutate(v_tot_1 = V_tot_i_1/O_i,
+            v_tot_2 = V_tot_i_2/O_i,
+            v_tot_3 = V_tot_i_3/O_i)
+
+# Sum of the accessibility divided by the total population
+tot_acc |> 
+  summarize(v_tot_1 = sum(V_tot_i_1)/sum(O_i),
+            v_tot_2 = sum(V_tot_i_2)/sum(O_i),
+            v_tot_3 = sum(V_tot_i_3)/sum(O_i))
+
+# Sum of the opportunities divided by the total population
+LU |> 
+  summarize(PPR = sum(D_j)/sum(O_i))
+```
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
+
+While we can see that the 'number of potentially reachable opportunities' V results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model).
 This approach is not yet seen in accessibility literature, but could be an intelligible way to compare accessibility values ($V_i$ or $V_j$) without considering the population matrix.
@@
 This case can be reformulated if the spatial weights matrix reflects only weights at the origin $W_i^(1)$ such as population $P_i$ and $W_j^(2)$ is unknown/irrelevant.
+
+**END OF PREVIOUS TEXT**
 
@@
 
-Returning to our simple example, the opportunity-constrained singly-constrained case would yield the following $B_{j}$ values assuming $A_{i}$ is 1:
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
+
+Returning to our simple example, the attraction-constrained case would yield the following $B_{j}$ values assuming $A_{i}$ is 1:
 
@@
 
-```{r}
-B_j <- 1/(P_i %*% (1/C))
-B_j_longer <- B_j
-rownames(B_j_longer) <- c("i..") 
-colnames(B_j_longer) <- c("1","2","3") 
-B_j_longer 
-```
+**END OF PREVIOUS TEXT**
+
+Balancing factors $B_j$:
+```{r}
+B_j <- OD |> 
+  group_by(did) |>
+  summarize(B_j_1 = 1/sum(O_i * f1),
+            B_j_2 = 1/sum(O_i * f2),
+            B_j_3 = 1/sum(O_i * f3))
+
+B_j
+```
+
+Use these proportionality constants to calculate the total opportunity constrained accessibility:
+```{r}
+opp_acc_ij <- OD |>
+  left_join(B_j,
+            by = "did") |>
+  mutate(V_opp_ij_1 = B_j_1  * O_i * D_j * f1,
+         V_opp_ij_2 = B_j_2  * O_i * D_j * f2,
+         V_opp_ij_3 = B_j_3  * O_i * D_j * f3)
+
+opp_acc <- opp_acc_ij |>
+  group_by(oid) |>
+  summarize(V_opp_i_1 = sum(V_opp_ij_1),
+            V_opp_i_2 = sum(V_opp_ij_2),
+            V_opp_i_3 = sum(V_opp_ij_3))
+
+opp_acc  |>
+  select(oid, starts_with("V_"))
+```
+
+Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
+```{r}
+opp_acc$V_opp_i_1[2] - opp_acc$V_opp_i_1[1]
+opp_acc$V_opp_i_2[2] - opp_acc$V_opp_i_2[1]
+opp_acc$V_opp_i_3[2] - opp_acc$V_opp_i_3[1]
+
+opp_acc$V_opp_i_1[2] / opp_acc$V_opp_i_1[1]
+opp_acc$V_opp_i_2[2] / opp_acc$V_opp_i_2[1]
+opp_acc$V_opp_i_3[2] / opp_acc$V_opp_i_3[1]
+```
+
+This illustrates how the differences and ratios are now meaningful.
+
+The total accessibility compared to the number of opportunities:
+```{r}
+opp_acc <- opp_acc |> 
+  left_join(LU |>
+              select(-D_j),
+            by = c("oid" = "id"))
+
+opp_acc  |>
+  summarize(V_opp_1 = sum(V_opp_i_1),
+            V_opp_2 = sum(V_opp_i_2),
+            V_opp_3 = sum(V_opp_i_3))
+
+LU |>
+  summarize(O = sum(D_j))
+```
+
+This shows how the number of opportunities is preserved.
+
+And the ratio of accessibility to population compared to the ratio of opportunities to population:
+```{r}
+# Accessibility divided by population
+opp_acc |> 
+  mutate(v_opp_1 = V_opp_i_1/O_i,
+            v_opp_2 = V_opp_i_2/O_i,
+            v_opp_3 = V_opp_i_3/O_i)
+
+# Sum of the accessibility divided by the total population
+opp_acc |> 
+  summarize(v_opp_1 = sum(V_opp_i_1)/sum(O_i),
+            v_opp_2 = sum(V_opp_i_2)/sum(O_i),
+            v_opp_3 = sum(V_opp_i_3)/sum(O_i))
+
+# Sum of the opportunities divided by the total population
+LU |> 
+  summarize(PPR = sum(D_j)/sum(O_i))
+```
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
 
 (NOTE: notably, when $B_j$ is multiplied by $P_i*f(c_{ij})$, it is a proportion (like spatial availability's total population&travel-cost balancing factor) representing how much of $O_j$ is allocated to each $i$)
-
-$V_{ij}''^{(2)}$ values:
-```{r}
-V_dest <- outer(P_i, as.numeric(B_j) * O_j) * (1/C) #we're not doing standard matrix multiplication here.
-We use 'outer' for element-wise multiplication, then element-wise multiplication of the resulting 3*3 matrix with the 3*3 impedance table
-
-P_i_dest <- V_dest |> rowSums()
-O_j_dest <- V_dest |> colSums()
-
-V_TOTAL_dest <- ifelse(all.equal(sum(P_i_dest),sum(O_j_dest)), sum(V_dest), NA)
-V_dest_longer <- cbind(V_dest,P_i_dest)
-V_dest_longer <- rbind(V_dest_longer, cbind(t(O_j_dest),V_TOTAL_dest))
-
-rownames(V_dest_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_dest_longer) <- c("1","2","3", "TOTAL")
-
-V_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
 
@@
 
-In $V_i''^{(1)}$, the following is $A_i$ assuming $B_{j}$ is 1:
+**END OF PREVIOUS TEXT**
+
+## Singly-constrained accessibility: production-constrained
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
+
+Returning to our simple example, the production-constrained case would yield the following $A_{j}$:
 
@@
 
-```{r}
-A_i <- 1/((1/C) %*% O_j)
-A_i_longer <- A_i
-rownames(A_i_longer) <- c("1","2","3") 
-colnames(A_i_longer) <- c("j...") 
-A_i_longer 
-```
-
-(AGAIN NOTE: when $A_i$ is multiplied by $O_j*f(c_{ij})$, it is a proportion (like spatial availability's total balancing factor) representing how much of $P_i$ is allocated from each $i$, I think.
-e.g.
-each $V_{ij}$ is the proportional opportunity&travel-cost balancing factor of $A_{i} W_j^{(2)}f(c_{ij})$ multiplied by $P_i$)
-
-The resulting $V_{ij}''^{(1)}$ values:
-```{r}
-V_orig <- outer(as.numeric(A_i) * P_i, O_j) * (1/C) #we're not doing standard matrix multiplication here.
-We use 'outer' for element-wise multiplication, then element-wise multiplication of the resulting 3*3 matrix with the 3*3 impedance table
-
-P_i_orig <- V_orig |> rowSums()
-O_j_orig <- V_orig |> colSums()
-
-V_TOTAL_orig <- ifelse(all.equal(sum(P_i_orig),sum(O_j_orig)), sum(V_orig), NA)
-V_orig_longer <- cbind(V_orig,P_i_orig)
-V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig))
-
-rownames(V_orig_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_orig_longer) <- c("1","2","3", "TOTAL")
-
-V_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
+**END OF PREVIOUS TEXT**
+
+Balancing factors $A_i$:
+```{r}
+A_i <- OD |> 
+  group_by(oid) |>
+  summarize(A_i_1 = 1/sum(D_j * f1),
+            A_i_2 = 1/sum(D_j * f2),
+            A_i_3 = 1/sum(D_j * f3))
+
+A_i
+```
+
+Use these proportionality constants to calculate the total opportunity constrained accessibility:
+```{r}
+pop_acc_ij <- OD |>
+  left_join(A_i,
+            by = "oid") |>
+  mutate(V_pop_ij_1 = A_i_1  * O_i * D_j * f1,
+         V_pop_ij_2 = A_i_2  * O_i * D_j * f2,
+         V_pop_ij_3 = A_i_3  * O_i * D_j * f3)
+
+pop_acc <- pop_acc_ij |>
+  group_by(did) |>
+  summarize(V_pop_i_1 = sum(V_pop_ij_1),
+            V_pop_i_2 = sum(V_pop_ij_2),
+            V_pop_i_3 = sum(V_pop_ij_3))
+
+pop_acc  |>
+  select(did, starts_with("V_"))
+```
+
+Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
+```{r}
+pop_acc$V_pop_i_1[2] - pop_acc$V_pop_i_1[1]
+pop_acc$V_pop_i_2[2] - pop_acc$V_pop_i_2[1]
+pop_acc$V_pop_i_3[2] - pop_acc$V_pop_i_3[1]
+
+pop_acc$V_pop_i_1[2] / pop_acc$V_pop_i_1[1]
+pop_acc$V_pop_i_2[2] / pop_acc$V_pop_i_2[1]
+pop_acc$V_pop_i_3[2] / pop_acc$V_pop_i_3[1]
+```
+
+This illustrates how the differences and ratios are now meaningful.
+
+The total accessibility compared to the number of opportunities:
+```{r}
+pop_acc <- pop_acc |> 
+  left_join(LU |>
+              select(-O_i),
+            by = c("did" = "id"))
+
+pop_acc  |>
+  summarize(V_pop_1 = sum(V_pop_i_1),
+            V_pop_2 = sum(V_pop_i_2),
+            V_pop_3 = sum(V_pop_i_3))
+
+LU |>
+  summarize(P = sum(O_i))
+```
+
+This shows how the population is preserved.
+
+And the ratio of accessibility to population compared to the ratio of opportunities to population:
+```{r}
+# Accessibility divided by population
+pop_acc |> 
+  mutate(v_pop_1 = V_pop_i_1/O_i,
+            v_pop_2 = V_pop_i_2/O_i,
+            v_pop_3 = V_pop_i_3/O_i)
+
+# Sum of the accessibility divided by the total population
+pop_acc |> 
+  summarize(v_pop_1 = sum(V_pop_i_1)/sum(D_j),
+            v_pop_2 = sum(V_pop_i_2)/sum(D_j),
+            v_pop_3 = sum(V_pop_i_3)/sum(D_j))
+
+# Sum of the opportunities divided by the total population
+LU |> 
+  summarize(PPR = sum(O_i))/sum(D_j)
+```
+
+**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
 
@@
 
+**END OF PREVIOUS TEXT**
+
 ## Doubly-constrained case (population-opportunity constrained)
-
+NOTE: The example needs to be revised in this case to have the population and opportunities match, as in, one person-one opportunity
+
+<!--
 In this case, $V_{ij}$ would be:
 ```{r}
+#| eval: false
+
 doubly_constrained_gravity_model <- function(
@@
     # Calculate Tij
-    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix #Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix -->
+    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix #Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
 
@@
 ```{r}
+#| eval: false
+
 A_i_doubly <- dc_results$Ai
@@
 ```{r}
+#| eval: false
+
 V_doubly <- dc_results$Tij
@@
 Nice. 
+-->
 
@@
 ```{r}
+#| eval: false
+
 sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){
@@
 ```{r}
+#| eval: false
 
@@
 ```{r}
+#| eval: false
+
 A_ij <- matrix(O_j) %*% (1/(t((matrix(O_j))) %*% (1/C))) |> t()
@@
 ```{r}
+#| eval: false
+
 (matrix(rep(O_j, 3), 3, 3))
@@
 ```{r}
+#| eval: false
+
 V_orig <- (A_ij) * t(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.

### 2025-02-10 TRB/Musings-a-draft-pres.qmd
Edits to new text added

@@
 ```{r}
-library(here)
-library(knitr)
-library(glue)
+library(here) # A Simpler Way to Find Your Files
+library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
+library(glue) # Interpreted String Literals
+#annotater::annotate_active_file() #Cool!
 ```

### 2025-02-10 paper/manuscript.qmd
Edits to new text added

@@
 
-In this respect, we welcome @wuUnifyingAccess2020's contribution for the way it provides a unifying framework to think about accessibility measures.
-By considering the key features of accessibility, namely travel-cost and the distribution of opportunities, these authors demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula.
+In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
+By considering the key features of accessibility, namely travel-cost $f(c{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula.
 One needs only to judiciously change the way travel-cost and opportunities are formulated, to derive almost any known accessibility measure.
-
-With their research, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
-Still, we feel that there is room to further clarify another relevant aspect.
-@wuUnifyingAccess2020's treatment of competition deserves further scrutiny, since the unifying framework does not appear to accommodate many (or any) of the hugely popular floating catchment area methods.
-In recent research, however, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn an accessibility measure of the following type (the most general in Wu and Levinson's framework) into a competitive measure of accessibility _à la_ 2-Step Floating Catchment Area approach:
-
-$$
-S_i \propto \sum_j g(O_i)f(c_{ij})
-$$
-
-In this paper we contend that accessibility research needs to reconnect with its roots in spatial interaction modelling.
-It is by looking to the past that we believe accessibility analysis can take a new step to the future.
-In particular, we argue that an important aspect of spatial interaction modelling, namely constraining the results to match empirical observations, was never properly reincorporated into accessibility analysis after these two streams of research began to diverge after the work of @hansen1959 and @wilson1971.
-As Wilson demonstrated, applying empirical constraints led to various spatial interaction models, and in this work we aim to show that the same can be done for accessibility measures.
-On the contrary, the lack of constraints in accessibility analysis seems to have contributed to some of the interpretability issues that plague it, particularly the fuzziness of insights beyond making statements of proportion such as higher-than/lower-than.
-
-With this background in mind, the focus of the present paper is on the empirical constraints and the attendant proportionality constant missing from accessibility measures.
-We propose to map accessibility onto Wilson's family of spatial interactions models.
-To achieve this, we begin by tracing the development of accessibility from its origins in spatial interaction, beginning with @ravensteinLawsMigration1889 to the accessibility research of @hansen1959.
-We also provide evidence of a marked divergence of accessibility and spatial interaction modelling research after the work of @wilson1971.
-After this, we hark back to Wilson's spatial interaction model, and use it to derive a family of accessibility measures based on different types of constraints.
+In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
+
+$$
+S_i \propto \sum_j g(O_j)f(c_{ij})
+$${#eq-access-Lu-Levinson}
+
+However, there is room to further clarify another relevant aspect: competition.
+@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular method within the competitive accessibility literature.
+There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach.
+Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971.
+
+Taken this motivation of incorporating competition in the unified framework and recent developments in methodology, this paper contends that accessibility research must reconnect with its spatial interaction origins.
+Particularly, we argue an important aspect of spatial interaction modelling -namely, constraining the results to match empirical observations- was never effectively reincorporated into accessibility analysis.
+Empirical constraints were embraced by early spatial interaction literature until the research stream diverged following the work of @hansen1959 and @wilson1971; at which point 'accessibility' and 'spatial interaction modelling' began to flow more distinctly.
+The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
+However, the same cannot be traced for contemporary accessibility literature, where empirical constraints were not explicitly adopted.
+We argue the absence of empirical constraints or attendance proportionality constants has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
+
+These streams of literature share common headwaters.
+It is by looking to the past that we believe accessibility analysis can newly wade into the future.
+Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility.
+We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the "seminal" accessibility work of @hansen1959.
+We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
+Next, we hark back to @wilson1971's spatial interaction models, and use it to derive a family of accessibility measures based on different types of constraints.
 We illustrate various members of this family with a simple numerical example and a real world data set.
@@
 From as far back as Henry C.
-Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858] a concern can be observed with the scientific study of human spatial interaction.
+Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858], a concern with the scientific study of human spatial interaction can be observed.
 It was in this work where Carey stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
@@
 Likewise, inverse proportionality means that as the distance increases, the force decreases.
-@eq-phys-grav-prop, however, does not quantify the magnitude of the force, for to do so an empirical constant is required to convert the proportionality into an equality, thus ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
+@eq-phys-grav-prop, however, does not quantify the magnitude of the force.
+To do so, an empirical constant is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
 In other words, @eq-phys-grav-prop needs to be _constrained_ using empirical data.
-Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically callibrated proportionality constant:
+Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:
 
@@
 Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts.
-In the late 1880s Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
+In the late 1880s, Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
 In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
@@
 
-\noindent where:
+\noindent Where:
 
@@
 
-\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
+\noindent Where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
 In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
@@
 34 and then later as _demographic energy_ on p.
-53.--> More seriously, though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism.
+53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism.
 For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" [p.
@@
 @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
-@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns .
+@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
 In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
@@
 
-A modern re-writing of the above accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
+Or a contemporary rewriting of the above @eq-accessibility accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
 
@@
 # Wilson's family of spatial interaction models
+
+<!--SUMMARY: Introduce Wilson's spatial interaction model.
+Then discuss how it could relate to gravity analogy -- but is tactically derived using entropy-maximisation, obtaining the interaction variable as a statistical average.
+Then introduce the constraints and cases.
+-->
 
@@
 
-Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced by the actual mass, as measured by the trips produced.
+Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced by the actual mass, as measured by the trips produced $O_i$.
 Also, there is no longer a single system-wide proportionality constant, but rather a balancing factor specific to each origin, namely $A_i$, which Wilson found to be as follows:
@@
 
-Notice how now the proxy for the mass at the destination $W_j^{(2)}$ is replaced by the actual mass, as measured by the trips attracted.
+Notice how now the proxy for the mass at the destination $W_j^{(2)}$ is replaced by the actual mass, as measured by the trips attracted $D_j$.
 As before, destination-specific balancing factors $B_j$ were derived by Wilson as:
@@
 
-In this model both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination. And, now the balancing factors are:
+In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination. And, now the balancing factors are:
 
@@
 Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-Accessibility at the time did not quite have that power: it did not quantify trips, but rather something fuzzier: the less tangible "potential for interaction".
+Accessibility at the time did not quite have that power: it did not quantify trips, but rather something more elusive: the less tangible "potential for interaction".
 In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
@@
 
-To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959 and/or @wilson1971.
+To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959, @wilson1971 or both.
 We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
@@
 After compiling this corpus of documents, we used the {bibliometrix} package [@bibliometrix2017] to create a bibliographical coupling matrix.
-In this matrix two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common.
+In this matrix, two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common.
 @fig-Fig1 presents the results.
-In this figure each symbol represents a document, and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents, meaning those that share more items in their lists of references, are plotted closer to each other.
+In this figure, each symbol represents a document and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents suggests those that share more items in their lists of references are plotted closer to each other.
 
@@
 
-As we argued above, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact over the years.
+As we argued in the preceding sections, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact.
 This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis.
@@
 $$
-V^T_i = k\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{Of(c_{ij})}{\sum_i\sum_j V^0_{ij}} = \sum_j W^{(2)}_j\frac{Of(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
+V^T_i = k\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{O*f(c_{ij})}{\sum_i\sum_j V^0_{ij}} = \sum_j W^{(2)}_j\frac{O*f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
 $$
@@
 $$
-\kappa = \frac{Of(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+\kappa = \frac{O*f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $$
@@
 In this case, the units of the population and opportunities must be able to be understood as the same.
-E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are saught by population at each $i$ must both be known to match.
+E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are sought by population at each $i$ must both be known to match.
 

### 2025-02-14 paper/manuscript.qmd
Edits to discussion and example, re. doubly constrained

@@
 bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
+includes:
+      in_header: 'gt_packages.sty'
 ---
@@
 
-Historically, the focus of transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement.
+Historically, the focus of modern transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement.
 This has had problematic consequences: with the car seen as the ultimate mobility tool, this approach has led to the emergence and continued dominance of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013].
-Decades of mobility-based planning (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often have had limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+Decades of planning for this mono-culture (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often having only limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
 In response to this situation, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020] using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020].
@@
 
-As a result, scholarly research on accessibility has experienced a remarkable boom, and has grown to include employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
-In other words, accessibility analysis is used today to understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
+An ascendant interest in accessibility has been accompanied by a remarkable boom on scholarly research, which has grown to include employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
+In other words, accessibility analysis is used today to broadly understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
 However, despite its growth in popularity in scholarly works, challenges remain to more widespread adoption of accessibility in planning practice.
@@
 In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
-By considering the key features of accessibility, namely travel-cost $f(c{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula.
+By considering the key features of accessibility, namely travel-cost $f(c{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson).
 One needs only to judiciously change the way travel-cost and opportunities are formulated, to derive almost any known accessibility measure.
@@
 However, there is room to further clarify another relevant aspect: competition.
-@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular method within the competitive accessibility literature.
+@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach within the competitive accessibility literature.
 There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach.
@@
 
-Taken this motivation of incorporating competition in the unified framework and recent developments in methodology, this paper contends that accessibility research must reconnect with its spatial interaction origins.
-Particularly, we argue an important aspect of spatial interaction modelling -namely, constraining the results to match empirical observations- was never effectively reincorporated into accessibility analysis.
-Empirical constraints were embraced by early spatial interaction literature until the research stream diverged following the work of @hansen1959 and @wilson1971; at which point 'accessibility' and 'spatial interaction modelling' began to flow more distinctly.
+Motivated by the challenge of incorporating competition into a unified framework for accessibility, this paper contends that accessibility research must reconnect with its spatial interaction origins.
+Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
+Empirical constraints were embraced by early spatial interaction literature following the work of @wilson1971, but this stream of literature tended to flow separately from research inspired by @hansen1959' accessibility.
 The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
-However, the same cannot be traced for contemporary accessibility literature, where empirical constraints were not explicitly adopted.
-We argue the absence of empirical constraints or attendance proportionality constants has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
+However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted.
+We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
 
@@
 Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility.
-We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the "seminal" accessibility work of @hansen1959.
+We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
 We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
@@
 
-The patterns of people's movement in space have been a subject of scientific inquiry for almost a century and a half.
+The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half.
 From as far back as Henry C.
@@
 In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
-As with Carey, these propositions are similar to Newton's gravitational laws.
+As with Carey, these propositions echo Newton's gravitational laws.
 Over time, other researchers discovered similar relationships.
-<!--Rilley does not cite Newton, nor Carey, nor Ravenstein, but I removed this curiosity as it's not central to our point here--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations.
-Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of goods movement by railways [@zipfHypothesisCaseRailway1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and information [@zipfDeterminantsCirculationInformation1946].
+<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations.
+Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of information [@zipfDeterminantsCirculationInformation1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and goods movement by railways [@zipfHypothesisCaseRailway1946].
 The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
@@
 After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
-We suspect that this relates to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist.
-Besides awareness of preceding research (Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
+This may relate to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist.
+Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
 
@@
 
-\noindent Where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
+\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
 In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
@@
 @hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
-@hansen1959 recast @eq-stewart-population-potential-sum to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-In this equation, @hansen1959 replaced $M_j$ with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
+@hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
+@hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
 
@@
 
-Or a contemporary rewriting of the above @eq-accessibility accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
+A contemporary rewriting of @eq-accessibility accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
 
@@
 Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships between opportunities in one area, opportunities at another, and their separation distance.
-At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
+There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
 
@@
 Alas, without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
-Hansen-style accessibility, therefore, is technically an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].
+For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$.
+Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].
 
@@
 Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
-In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of the single proportionality constant $k$ is to ensure the total sum of $T_{ij}$ represents the total flows in the data.
+In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of the single proportionality constant $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
 In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
@@
 But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
-Entropy maximisation in this cases achieves stable results as a statistical average that represents the population.
+Entropy maximisation in this case achieves stable results as a statistical average that represents the population.
 The approach works by assuming undifferentiated individual interactions, and assessing their probabilities of making a particular journey.
@@
 Parting from this, it is possible to derive a family of spatial interaction models.
-In Wilson's terms, there is the "unconstrained" (but actually total flow constrained) model in equation @eq-phys-gravity-model.
+In Wilson's terms, there is the "unconstrained" (but actually total flow constrained) model in @eq-phys-gravity-model.
 The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
@@
 Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced by the actual mass, as measured by the trips produced $O_i$.
-Also, there is no longer a single system-wide proportionality constant, but rather a balancing factor specific to each origin, namely $A_i$, which Wilson found to be as follows:
+Also, there is no longer a single system-wide proportionality constant, but rather a set of proportionality constants (i.e., balancing factors) specific to origins.
+According to Wilson, these constants, namely $A_i$, are:
 
@@
 Notice how now the proxy for the mass at the destination $W_j^{(2)}$ is replaced by the actual mass, as measured by the trips attracted $D_j$.
-As before, destination-specific balancing factors $B_j$ were derived by Wilson as:
+As before, destination-specific proportionality constants (i.e., balancing factors) $B_j$ were derived by Wilson as:
 
@@
 
-In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination. And, now the balancing factors are:
+In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination.
+And, now there are two sets of mutually dependent proportionality constants:
 
@@
 Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
-It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.
-Why could this be?
+It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been ignored in almost all subsequent accessibility research.
+Why was this key element of spatial interaction models ignored in accessibility research?
 In the next section we aim to address this question.
@@
 Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-Accessibility at the time did not quite have that power: it did not quantify trips, but rather something more elusive: the less tangible "potential for interaction".
+Accessibility at the time did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: the less tangible "potential for interaction".
 In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
@@
 
-To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959, @wilson1971 or both.
+To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959, @wilson1971, or both.
 We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
@@
 @fig-Fig1 presents the results.
-In this figure, each symbol represents a document and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents suggests those that share more items in their lists of references are plotted closer to each other.
+In this figure, each symbol represents a document and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents (i.e., those sharing more items in their lists of references) are plotted closer to each other.
 
@@
 
-As we argued in the preceding sections, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact.
+As we argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact after @hansen1959 and @wilson1971.
 This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis.
-This is intriguing, considering that Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the similarity between the denominator in @eq-production-constrained-balancing-factor and Hansen's accessibility indicator.
-In fact, Wilson noted that Hansen's accessibility was the inverse of $A_i$, which offered a potential interpretation for the balancing factor (p.
+This is intriguing, considering that Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the similarity between the denominator of @eq-production-constrained-balancing-factor and Hansen's accessibility indicator.
+In fact, Wilson noted that Hansen's accessibility was the inverse of $A_i$, which offered a potential interpretation for the proportionality constants $A_i$ (p.
 10):
@@
 
-\noindent and from there to the system-wide population potential:
+\noindent and from there to the system-wide population potential (after losing $G$):
 $$
@@
 
-Then, we define the _potential for interaction_ (i.e., accessibility) as follows:
+Then, a way to define the _potential for interaction_ (i.e., accessibility) is as follows:
 
@@
 V_{ij} = k W_j^{(2)} f(c_{ij})
-$${#eq-access}
-
-\noindent where $V_{ij}$ is the potential for interaction.
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ is the mass at the destination, and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+$${#eq-access-01}
+
+\noindent where $V_{ij}$ is the potential for interaction, and therefore:
+
+$$
+V_{i} = k \sum_j W_j^{(2)} f(c_{ij})
+$${#eq-accesssibility-01}
+
+Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination, and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
 
@@
 
-From here, we can define various accessibility measures depending on how we choose to constrain them.
+Next, we show how various accessibility measures can be defined depending on how we choose to constrain them.
 
@@
 
-Unconstrained accessibility results when we set the proportionality constant/balancing factor to one in @eq-access:
+Unconstrained accessibility results when we set the proportionality constant/balancing factor to one in @eq-access-01:
 
@@
 
-We can choose to retain the proportionality constant in @eq-access to obtain the following:
-
-$$
-V^T_{ij} = kW_j^{(2)} f(c_{ij})
+We can choose to retain the proportionality constant in @eq-access-01 to obtain the following:
+
+$$
+V^T_{ij} = k^DW_j^{(2)} f(c_{ij})
 $${#eq-total-constrained-access}
@@
 $$
-V^T_i = \sum_j V^T_{ij} = k\sum_j W^{(2)}_jf(c_{ij}) = kV^0_i
+V^T_i = \sum_j V^T_{ij} = k^D\sum_j W^{(2)}_jf(c_{ij}) = k^DV^0_i
 $${#eq-total-constrained-accessibility}
 
-The constraint that we impose in this case is the total number of opportunities in the region $O$ (compare to @eq-constraint0-gravitymodel):
-
-$$
-\sum_i\sum_j V^T_{ij} = O
+The constraint that we impose in this case is the total number of opportunities $D$ in the region (compare to @eq-constraint0-gravitymodel):
+
+$$
+\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
 $${#eq-total-constraint-accessibility}
@@
 $$
-k=\frac{O}{\sum_i\sum_j V^0_{ij}}
+k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $${#eq-total-opportunity-balancing-factor}
@@
 $$
-V^T_i = k\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{O*f(c_{ij})}{\sum_i\sum_j V^0_{ij}} = \sum_j W^{(2)}_j\frac{O*f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
+V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \frac{D}{\sum_i\sum_j V^0_{ij}}\sum_j W^{(2)}_j f(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
 $$
@@
 $$
-\kappa = \frac{O*f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+\kappa^D_{ij} = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $$
@@
 $$
-V^T_i = \kappa\sum_j W^{(2)}_j 
-$$
-
-Notice that the role of $\kappa$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
-Constant $\kappa$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij} 
+$$
+
+Notice that the role of $\kappa^0_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+Constant $\kappa^0_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
 For this reason we call it a proportional allocation factor.
 
-## Singly-constrained accessibility: attraction-constrained
-
-If we decide to introduce additional information into our analysis, say, the number of opportunities _by destination_ (i.e., $O_j$), we can impose the following constraint (compare to @eq-constraint2-gravitymodel):
+## Total population constrained accessibility: market potential
+
+One might wish to measure accessibility to the mass of the _population_ at $i$ instead of the opportunities at $j$. This is equivalent to the concept of market potential:
+
+$$
+M^T_{ij} = k^OW_i^{(1)} f(c_{ij}) = k^OM_{ij}^0
+$${#eq-total-constrained-market}
+
+\noindent with:
+
+$$
+M_i^0 = W_i^{(1)} f(c_{ij})
+$$
+
+The population-constrained market potential becomes:
+
+$$
+M^T_j = \sum_i M^T_{ij} = k^O\sum_i W^{(1)}_if(c_{ij})
+$${#eq-total-constrained-market}
+
+The constraint that we impose in this case is the total market potential equals the total population $O$ in the region:
+
+$$
+\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
+$${#eq-total-constraint-market}
+
+Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $k$, we obtain:
+
+$$
+k^O=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
+$${#eq-total-population-balancing-factor}
+
+The constrained market potential then takes the following form:
+
+$$
+M^T_i = k\sum_i W^{(1)}_if(c_{ij}) = \frac{O}{\sum_i\sum_j M^T_{ij}}\sum_i W^{(i)}_i f(c_{ij})
+$$
+
+## Singly-constrained accessibility: opportunity-constrained
+
+If we decide to introduce additional information into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (compare to @eq-constraint2-gravitymodel):
 
@@
 V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
-$${#eq-attraction-constrained-accessibility}
+$${#eq-opportunity-constrained-accessibility}
 
@@
 B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$$
-
-If we introduce this balancing factor in @eq-attraction-constrained-accessibility we obtain:
+$${#eq-opportunity-constrained-proportionality-constants}
+
+If we introduce this balancing factor in @eq-opportunity-constrained-accessibility we obtain:
 
@@
 V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-attraction-constrained-accessibility-with-balancing-factor}
+$${#eq-opportunity-constrained-accessibility-with-balancing-factor}
 
@@
 \kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-attraction-constrained-proportional-allocation-factor}
-
-After this, it is possible to rewrite @eq-attraction-constrained-accessibility-with-balancing-factor as:
+$${#eq-opportunity-constrained-proportional-allocation-factor}
+
+After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as:
 
@@
 $$
-\sum_i V^D_{i} = O
+\sum_i V^D_{i} = D
 $$
@@
 
-## Singly-constrained accessibility: production-constrained
-
-Not the most common application of accessibility analysis, but if the origins and destinations are transposed, we can define a measure of market potential that preserves the population.
+## Singly-constrained accessibility: population-constrained
+
+Not the most common application of accessibility analysis, but if the origins and destinations are transposed, we can define a measure of market potential that preserves the population, similar to @eq-total-population-balancing-factor.
 The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $V^P_i$ becomes:
@@
 $$
-V^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
-$${#eq-attraction-constrained-accessibility}
-
-In this case the measure is constrained by the population _by origin_ (i.e., $P_i$). Compare to @eq-constraint2-gravitymodel:
-
-$$
-\sum_j V^P_{ij} =  P_i
-$${#eq-attraction-constraint}
+M^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
+$${#eq-population-constrained-accessibility}
+
+In this case the measure is constrained by the population _by origin_ (i.e., $O_i$). Compare to @eq-constraint2-gravitymodel:
+
+$$
+\sum_j M^P_{ij} =  O_i
+$${#eq-population-constraint}
 
@@
 A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$$
+$${#eq-population-constrained-proportionality-constants}
 
@@
 $$
-V^P_{j} = \sum_i  \kappa^P_{ij} P_i
+M^P_{j} = \sum_i  \kappa^O_{ij} O_i
 $${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
@@
 $$
-\kappa^P _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+\kappa^O _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
 $${#eq-attraction-constrained-proportional-allocation-factor}
 
-In addition to market potential, $V_j^P$ can be interpreted as a constrained measure of "facility crowdedness" as in @wang_inverted_2018.
+In addition to market potential, $M_j^P$ can be interpreted as a constrained measure of "facility crowdedness" as in @wang_inverted_2018.
 This "crowdedness" now satisfies, in addition to @eq-attraction-constraint, the following:
@@
 $$
-\sum_j V^P_{j} = P
-$$
-
-\noindent where $P$ is the total population of the region. In this way, $\frac{V_j^P}{P}$ can be interpreted as the proportion of the total population serviced by location $j$.
+\sum_j M^P_{j} = O
+$$
+
+\noindent where $O$ is the total population of the region. In this way, $\frac{M_j^O}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
 
@@
 
+Readers will have noticed that the accessibility measures above have used either $O_i$ or $D_j$ but not both.
+When $D_j$, the opportunities, were used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
+Contrariwise, When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
+The reason is that the population and the opportunities are typically different things without a common metric.
+For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
+There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most common one being employment, i.e., one person, one job.
+
+A doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities.
+In this case, the two constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel are imposed.
+Furthermore, the two constraints must match:
+
+$$
+\sum_iO_i = \sum_j D_j
+$$
+
+This is because now the proportionality constants are mutually dependent and cannot be estimated if the constraints diverge.
+For this reason, the applicability seems to be relatively limited, because even in the case of jobs, there might be imbalances between $O$ and $D$, say, where there are more jobs than people or vice-versa.
+
+Another important point to note about the doubly-constrained approach is as follows.
+Recall that accessibility is defined as "the potential for interaction".
+Additional constraints leave less room for that potential.
+In fact, a doubly constrained approach essentially converges on the spatial interaction model, where one person represents one trip to one opportunity, and it is probably better thought of as an estimate of "access" instead of "accessibility".
+
+Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
+The values of $B_j$ are used to update the underlying trip matrix, before calibrating $A_i$.
+This process continues to update $A_i$ and $B_j$ until a convergence criterion is met [see @ortuzar_2011_modelling, p.
+193-195].
+
+<!--
 In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
@@
 Literature has explored 'doubly-constrained' but -- not the same: , etc. Ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
+-->
 
@@
 
-The following is the known accessibility matrix.
-The $V_{ij}$, the number of jobs spatially accessible at $i$ from destination $j$.
-The TOTALs are the $V_i$ and $V_j$, the total number of jobs spatially available at $i$ and the number of jobs spatially available from $j$.
-
-The following vectors are the population by origin ($O_i$) and the opportunities by destination $D_j$:
-
-<!--Check here, the number of physicians per 10,000 in Canada in 2022 was 24.97
-https://en.wikipedia.org/wiki/List_of_countries_by_number_of_physicians -->
 ```{r}
@@
 
-```{r}
-LU
-```
-
-Think of $O_i$ as population in 10,000 and $D_j$ as number of physicians. The Provider-to-Population-Ratio (PPR) is:
-```{r}
-sum(LU$D_j) / sum(LU$O_i)
-```
-
-And the cost of travel from $i$ to $j$ is as follows:
+In this section we use a simple numerical example to illustrate the various members of the family of accessibility measures introduced above.
+Consider a very simple system with three zones, as shown in @tbl-small-system-land-use.
+It can help to think of these numbers as population in 10,000s and number of physicians (for comparison, the number of physicians per 10,000 in Canada in 2022 was 24.97).
+If so, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
+The population is $W_i^{(1)}$ when used as a proxy for the mass at the origin, and $O_i$ when used as a constraint.
+Similarly, the opportunities are represented by $W_j^{(2)}$ when used as a proxy for the mass at the destination, and $D_j$ when used as a constraint.
+
+To complete the set-up of this example, the cost of movement between origins and destinations is as shown in @tbl-small-system-cost.
+
+<!--The number of physicians per 10,000 in Canada in 2022 was 24.97 according to: https://en.wikipedia.org/wiki/List_of_countries_by_number_of_physicians -->
+
+```{r}
+#| label: tbl-small-system-land-use
+#| tbl-cap: "Simple system with three zones. Population is in 10,000 persons and opportunities in number of physicians."
+
+LU |>
+  gt() |>
+  # tab_spanner("{{W[_i^(1)]}}",
+  #             columns = 2) |>
+  # tab_spanner("{{W[_j^(2)]}}",
+  #             columns = 3) |>
+  cols_label(id = "ID",
+             O_i = "Population", 
+             D_j = "Opportunities")  |>
+  cols_align(align = "center",
+             columns = 2:3) |>
+  tab_footnote(footnote = md("Population is *Wi^(1)^* when used as a proxy for the mass at the origin, and *Oi* when used as a constraint."),
+               locations = cells_column_labels(columns = O_i)) |>
+  tab_footnote(footnote = md("Opportunities is *Wj^(2)^* when used as a proxy for the mass at the destination, and *Dj* when used as a constraint."),
+               locations = cells_column_labels(columns = D_j)) |>
+  as_latex()
+```
+
 ```{r}
@@
 
-# Simulate a table with OD cost for the simple example, beginning with the ids, the cost, and then two impedance functions 
 C <- expand.grid(oid = c("1", "2", "3"),
                   did = c("1", "2", "3")) |>
-  mutate(cost = c(2, 15, 5, 15, 2, 10, 5, 10, 2)) |>
-  mutate(f1 = exp(-1.00000 * cost^2), 
-         f2 = exp(-0.01000 * cost^2),
-         f3 = exp(-0.00001 * cost^2))
-```
-
-```{r}
-C |> select(oid, did, cost)
-```
-
-For comparison, use three different impedance functions:
-$$
-\begin{array}{l}
-f_1(c_{ij}) = exp(-c_{ij}^2)\\
-f_2(c_{ij}) = exp(-0.01\cdot c_{ij}^2)\\
-f_3(c_{ij}) = exp(-0.00001\cdot c_{ij}^2)
-\end{array}
-$$
-
-## Unconstrained accessibility
-
-The following case is standard practice, the HAM.
-It is expressed as a value of the origin, a summary of the destination weight multiplied by the travel impedance.
-There is no explicit consideration of a proportionality constant ($k$), we can presume it is still set to 1 as in @stewartPrinciples1947.
-Neither are there origin weights $W_i^{(1)}$, we assume is also wrapped up in this proportionality and implicitly set to 1.
-Working backward from this expression and assumptions, $S_{ij}$ can be expressed more generally as well.
-
-$S_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O_j$ and/or population $P_i$ (only in some case when $k$ and $W_i^{(1)}$ equal 1).
-
-In solving the example, each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin: 
+  mutate(cost = c(10, 30, 15, 30, 10, 25, 10, 25, 10)) |>
+  mutate(f1 = cost^-3,
+         f2 = cost^-2,
+         f3 = cost^-0.1)
+```
+
+```{r}
+#| label: tbl-small-system-cost
+#| tbl-cap: "Cost matrix for system with three zones (travel time in minutes)."
+
+C |> 
+  select(oid, did, cost) |>
+  tidyr::pivot_wider(names_from = did, values_from = cost) |>
+  gt() |>
+  tab_spanner(label = "Destination ID",
+              columns = 2:4) |>
+  cols_label(oid = "Origin ID")  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  as_latex()
+```
+
 ```{r}
@@
 
-```{r}
+For comparison, we use three different impedance functions:
+$$
+\begin{array}{l}
+f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
+f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
+f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
+\end{array}
+$$
+
+## Unconstrained accessibility
+
+The following case is standard practice, the HAM.
+It is expressed as a value of the origin, a summary of the destination weight multiplied by the travel impedance.
+There is no explicit consideration of a proportionality constant ($k$), we can presume it is still set to 1 as in @stewartPrinciples1947.
+Neither are there origin weights $W_i^{(1)}$, we assume is also wrapped up in this proportionality and implicitly set to 1.
+Working backward from this expression and assumptions, $S_{ij}$ can be expressed more generally as well.
+
+$S_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O_j$ and/or population $P_i$ (only in some case when $k$ and $W_i^{(1)}$ equal 1).
+
+In solving the example, each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin:
+
+```{r}
+#| label: tab-simple-example-unconstrained-accessibility
+#| tbl-cap: "Simple system: unconstrained accessibility."
+
 unc_acc_ij <- OD |>
@@
 
-unc_acc
+unc_acc |>
+  gt(rowname_col = "oid") |>
+  tab_stubhead(label = "Origin") |>
+  tab_spanner(label = "{{V[_i^0]}}",
+              columns = 2:4,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 2,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 4,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+  cols_label(V_unc_i_1 = md("units: *physicians per min*^3"),
+             V_unc_i_2 = md("units: *physicians per min*^2"),
+             V_unc_i_3 = md("units: *physicians per min*^0.1"))  |>
+  # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+  #            V_unc_i_2 = "{{f_2 (c_ij )  = 1/c[_ij^2]}}<br>*physicians min^-2*",
+  #            V_unc_i_3 = "{{f_3 (c_ij )  = 1/c[_ij^0.1]}}<br>*physicians min^-0.1*",
+  #            .fn = md)  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  cols_width(
+    stub() ~ px(60),
+    everything() ~ px(180)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
 ```
@@
 
-## Total constrained accessibility
-
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-Solved for our example, $K$ is equal to the total accessibility $V$, the only thing we know: either all the opportunities $O$ or all the population $P$, divided by remaining part of the equation.
-In the following, we assume our spatial weights matrix only reflects $W_j^(2)$ as $O_j$ and $W_i^{(1)}$ is uniform (1):
+## Total opportunity-constrained accessibility
+
+Proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$:
+
 $$
 \begin{array}{l}
-k = \frac{V}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
-k = \frac{V}{\frac{O_1}{c_{11}}+\frac{O_1}{c_{21}}+\frac{O_1}{c_{31}}+\frac{O_2}{c_{12}} +... \frac{O_3}{c_{33}}
+k^D = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
+k^D = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
 }\\
-k = \frac{790}{568.4} \\
-k = 1.389866
+k^D = \frac{490}{0.75} \\
+k^D = 653.3256
 \end{array}
@@
 
-**END OF PREVIOUS TEXT**
-
-Calculate the balancing factors/proportionality constants:
-```{r}
+```{r}
+# Calculate the balancing factors/proportionality constants for each of three impedance functions
+
 k_tot <- unc_acc_ij |> 
-  summarize(k1 = sum(LU$D_j)/sum(V_unc_ij_1),
-            k2 = sum(LU$D_j)/sum(V_unc_ij_2),
-            k3 = sum(LU$D_j)/sum(V_unc_ij_3))
+  summarize(k1 = sum(LU$D_j)/sum(V_unc_ij_1), # f_1
+            k2 = sum(LU$D_j)/sum(V_unc_ij_2), # f_2
+            k3 = sum(LU$D_j)/sum(V_unc_ij_3)) # f_3
 ```
@@
 ```{r}
+#| label: tab-simple-example-total-opportunity-accessibility
+#| tbl-cap: "Simple system: total opportunity constrained accessibility."
+
 tot_acc_ij <- unc_acc_ij |>
@@
 tot_acc |>
-  select(oid, starts_with("V_"))
+  select(oid, starts_with("V_")) |>
+  gt(rowname_col = "oid") |>
+  tab_stubhead(label = "Origin") |>
+  tab_spanner(label = "{{V[_i^T]}}",
+              columns = 2:4,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 2,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 4,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+  cols_label(V_tot_i_1 = md("units: *physicians*"),
+             V_tot_i_2 = md("units: *physicians*"),
+             V_tot_i_3 = md("units: *physicians*"))  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  cols_width(
+    stub() ~ px(60),
+    everything() ~ px(180)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
 ```
@@
 
+## Total population-constrained accessibility (market potential)
+
+Proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$:
+
+$$
+\begin{array}{l}
+k^O = \frac{O}{\sum_{i}\sum_{j} O_i f(c_{ij})}\\
+k^O = \frac{O}{\frac{O_1}{c_{11}^3}+\frac{O_1}{c_{21}^3} + \frac{O_1}{c_{31}^3} + \cdots + \frac{O_3}{c_{31}^3} + \frac{O_3}{c_{32}^3} + \frac{O_3}{c_{33}^3}
+}\\
+k^O = \frac{490}{0.75} \\
+k^O = 653.3256
+\end{array}
+$$
+
+```{r}
+# Calculate the balancing factors/proportionality constants for each of three impedance functions
+unc_market_ij <- OD |>
+  mutate(M_unc_ij_1 = O_i * f1,
+         M_unc_ij_2 = O_i * f2,
+         M_unc_ij_3 = O_i * f3)
+
+k_tot_pop <- unc_market_ij |> 
+  summarize(k1 = sum(LU$O_i)/sum(M_unc_ij_1), # f_1
+            k2 = sum(LU$O_i)/sum(M_unc_ij_2), # f_2
+            k3 = sum(LU$O_i)/sum(M_unc_ij_3)) # f_3
+```
+
+Use these proportionality constants to calculate the total opportunity constrained accessibility:
+```{r}
+#| label: tab-simple-example-total-population-accessibility
+#| tbl-cap: "Simple system: total opportunity constrained accessibility."
+
+tot_pop_ij <- unc_market_ij |>
+  mutate(M_tot_ij_1 = k_tot_pop$k1 * M_unc_ij_1,
+         M_tot_ij_2 = k_tot_pop$k2 * M_unc_ij_2,
+         M_tot_ij_3 = k_tot_pop$k3 * M_unc_ij_3)
+
+tot_market <- tot_pop_ij |>
+  group_by(did) |>
+  summarize(M_tot_i_1 = sum(M_tot_ij_1),
+            M_tot_i_2 = sum(M_tot_ij_2),
+            M_tot_i_3 = sum(M_tot_ij_3))
+
+tot_market |>
+  select(did, starts_with("M_")) |>
+  gt(rowname_col = "did") |>
+  tab_stubhead(label = "Destination") |>
+  tab_spanner(label = "{{M[_i^D]}}",
+              columns = 2:4,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 2,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 4,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
+             M_tot_i_2 = md("units: *population in 10,000s*"),
+             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  cols_width(
+    stub() ~ px(90),
+    everything() ~ px(150)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
+```
+
+<!--
 **PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
@@
 **END OF PREVIOUS TEXT**
-
-## Singly-constrained accessibility: attraction-constrained
+-->
+
+## Singly-constrained accessibility: opportunity-constrained
 
@@
 
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-Returning to our simple example, the attraction-constrained case would yield the following $B_{j}$ values assuming $A_{i}$ is 1:
-
-$$
-B_{j} = \frac{1}{\sum_i P_i f(c_{ij})}\\
+Returning to our simple example, the opportunity-constrained case would yield the following $B_{j}$:
+
+$$
+B_{j} = \frac{1}{\sum_i O_i f(c_{ij})}\\
 $$
@@
 \begin{array}{l}
-B_{1} =  \frac{1}{160/2+450/15+180/5} = 0.006849315 \\ 
-B_{2} =  \frac{1}{160/15+450/2+180/10} = 0.003942181 \\
-B_{3} =  \frac{1}{160/5+450/10+180/2} = 0.005988024
+B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
+B_{2} =  \frac{1}{\frac{4}{30^3} + \frac{10}{10^3} + \frac{6}{25^3}} = 94.9474\\
+B_{3} =  \frac{1}{\frac{4}{10^3} + \frac{10}{25^3} + \frac{6}{10^3}} = 93.9850
 \end{array}
 $$
-
-**END OF PREVIOUS TEXT**
 
@@
 
-Use these proportionality constants to calculate the total opportunity constrained accessibility:
-```{r}
+Use these proportionality constants to calculate the singly-constrained opportunity-constrained accessibility:
+```{r}
+#| label: tab-simple-example-attraction-constrained-accessibility
+#| tbl-cap: "Simple system: singly-constrained opportunity constrained accessibility."
+
 opp_acc_ij <- OD |>
@@
 opp_acc  |>
-  select(oid, starts_with("V_"))
+  select(oid, starts_with("V_")) |>
+  gt(rowname_col = "oid") |>
+  tab_stubhead(label = "Origin") |>
+  tab_spanner(label = "{{V[_i^D]}}",
+              columns = 2:4,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 2,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 4,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+  cols_label(V_opp_i_1 = md("units: *physicians*"),
+             V_opp_i_2 = md("units: *physicians*"),
+             V_opp_i_3 = md("units: *physicians*"))  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  cols_width(
+    stub() ~ px(60),
+    everything() ~ px(180)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
 ```
@@
 
+<!--
 **PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
@@
 **END OF PREVIOUS TEXT**
-
-## Singly-constrained accessibility: production-constrained
-
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-Returning to our simple example, the production-constrained case would yield the following $A_{j}$:
+-->
+
+## Singly-constrained accessibility: population-constrained
+
+Returning to our simple example, the production-constrained case would yield the following $A_{j}$ when we use $f_1(c_{ij})=1/c_{ij}^3$:
 
@@
 \begin{array}{l}
-A_{1} =  \frac{1}{160/2+437/15+193/5} = 0.006768953 \\ 
-A_{2} =  \frac{1}{160/15+437/2+193/10} = 0.004024685 \\
-A_{3} =  \frac{1}{160/5+437/10+193/2} = 0.005807201
+A_{1} =  \frac{1}{\frac{160}{10^3} + \frac{150}{30^3} + \frac{180}{15^3}} = 2.8939 \\ 
+A_{2} =  \frac{1}{\frac{160}{30^3} + \frac{150}{10^3} + \frac{180}{25^3}} = 5.9721\\
+A_{3} =  \frac{1}{\frac{160}{15^3} + \frac{150}{25^3} + \frac{180}{10^3}} = 4.2193
 \end{array}
@@
 
-**END OF PREVIOUS TEXT**
-
-Balancing factors $A_i$:
-```{r}
+```{r}
+# Balancing factors $A_i$:
+
 A_i <- OD |> 
@@
 
-Use these proportionality constants to calculate the total opportunity constrained accessibility:
-```{r}
+Use these proportionality constants to calculate the singly-constrained population constrained accessibility:
+```{r}
+#| label: tab-simple-example-production-constrained-accessibility
+#| tbl-cap: "Simple system: singly-constrained population-constrained accessibility (market potential)."
+
 pop_acc_ij <- OD |>
@@
             by = "oid") |>
-  mutate(V_pop_ij_1 = A_i_1  * O_i * D_j * f1,
-         V_pop_ij_2 = A_i_2  * O_i * D_j * f2,
-         V_pop_ij_3 = A_i_3  * O_i * D_j * f3)
+  mutate(M_pop_ij_1 = A_i_1  * O_i * D_j * f1,
+         M_pop_ij_2 = A_i_2  * O_i * D_j * f2,
+         M_pop_ij_3 = A_i_3  * O_i * D_j * f3)
 
@@
   group_by(did) |>
-  summarize(V_pop_i_1 = sum(V_pop_ij_1),
-            V_pop_i_2 = sum(V_pop_ij_2),
-            V_pop_i_3 = sum(V_pop_ij_3))
+  summarize(M_pop_i_1 = sum(M_pop_ij_1),
+            M_pop_i_2 = sum(M_pop_ij_2),
+            M_pop_i_3 = sum(M_pop_ij_3))
 
 pop_acc  |>
-  select(did, starts_with("V_"))
+  select(did, starts_with("M_")) |>
+  gt(rowname_col = "did") |>
+  tab_stubhead(label = "Destination") |>
+  tab_spanner(label = "{{M[_j^P]}}",
+              columns = 2:4,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 2,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 4,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+  cols_label(M_pop_i_1 = md("units: *population in 10,000s*"),
+             M_pop_i_2 = md("units: *population in 10,000s*"),
+             M_pop_i_3 = md("units: *population in 10,000s*"))  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  cols_width(
+    stub() ~ px(90),
+    everything() ~ px(150)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
 ```
@@
 ```{r}
-pop_acc$V_pop_i_1[2] - pop_acc$V_pop_i_1[1]
-pop_acc$V_pop_i_2[2] - pop_acc$V_pop_i_2[1]
-pop_acc$V_pop_i_3[2] - pop_acc$V_pop_i_3[1]
-
-pop_acc$V_pop_i_1[2] / pop_acc$V_pop_i_1[1]
-pop_acc$V_pop_i_2[2] / pop_acc$V_pop_i_2[1]
-pop_acc$V_pop_i_3[2] / pop_acc$V_pop_i_3[1]
+pop_acc$M_pop_i_1[2] - pop_acc$M_pop_i_1[1]
+pop_acc$M_pop_i_2[2] - pop_acc$M_pop_i_2[1]
+pop_acc$M_pop_i_3[2] - pop_acc$M_pop_i_3[1]
+
+pop_acc$M_pop_i_1[2] / pop_acc$M_pop_i_1[1]
+pop_acc$M_pop_i_2[2] / pop_acc$M_pop_i_2[1]
+pop_acc$M_pop_i_3[2] / pop_acc$M_pop_i_3[1]
 ```
@@
 pop_acc  |>
-  summarize(V_pop_1 = sum(V_pop_i_1),
-            V_pop_2 = sum(V_pop_i_2),
-            V_pop_3 = sum(V_pop_i_3))
+  summarize(M_pop_1 = sum(M_pop_i_1),
+            M_pop_2 = sum(M_pop_i_2),
+            M_pop_3 = sum(M_pop_i_3))
 
 LU |>
-  summarize(P = sum(O_i))
+  summarize(O = sum(O_i))
 ```
@@
 pop_acc |> 
-  mutate(v_pop_1 = V_pop_i_1/O_i,
-            v_pop_2 = V_pop_i_2/O_i,
-            v_pop_3 = V_pop_i_3/O_i)
+  mutate(m_pop_1 = M_pop_i_1/O_i,
+         m_pop_2 = M_pop_i_2/O_i,
+         m_pop_3 = M_pop_i_3/O_i)
 
@@
 pop_acc |> 
-  summarize(v_pop_1 = sum(V_pop_i_1)/sum(D_j),
-            v_pop_2 = sum(V_pop_i_2)/sum(D_j),
-            v_pop_3 = sum(V_pop_i_3)/sum(D_j))
+  summarize(m_pop_1 = sum(M_pop_i_1)/sum(D_j),
+            m_pop_2 = sum(M_pop_i_2)/sum(D_j),
+            m_pop_3 = sum(M_pop_i_3)/sum(D_j))
 
@@
 **END OF PREVIOUS TEXT**
+-->
 
 ## Doubly-constrained case (population-opportunity constrained)
+
+```{r}
+#| label: simple-numerical-example-doubly-constrained
+
+# Same as before
+id <- c("1", "2", "3")
+O_i2 <- c(4, 10, 6) 
+D_j2 <- c(160, 150, 180)
+
+# Balance the sum of origins and destinations
+f <- sum(O_i2 * 10000)/sum(D_j2)
+D_j2 <- (f * D_j2/10000) |>
+  round()
+
+LU2 <- data.frame(id, O_i2, D_j2)
+```
+
 NOTE: The example needs to be revised in this case to have the population and opportunities match, as in, one person-one opportunity
 
-<!--
+```{r}
+#| label: tbl-small-system-land-use-doubly-constrained
+#| tbl-cap: "Simple system with three zones. Population is in 10,000 persons and opportunities in 10,000s physicians."
+
+LU2 |>
+  gt() |>
+  cols_label(id = "ID",
+             O_i2 = "Population", 
+             D_j2 = "Opportunities")  |>
+  cols_align(align = "center",
+             columns = 2:3) |>
+  tab_footnote(footnote = md("Population is *Wi^(1)^* when used as a proxy for the mass at the origin, and *Oi* when used as a constraint."),
+               locations = cells_column_labels(columns = O_i2)) |>
+  tab_footnote(footnote = md("Opportunities is *Wj^(2)^* when used as a proxy for the mass at the destination, and *Dj* when used as a constraint."),
+               locations = cells_column_labels(columns = D_j2)) #|>
+  #as_latex()
+```
+
+```{r}
+#| label: simple-numerical-example-OD-table-doubly-constrained
+
+OD2 <- C |>
+  left_join(LU2 |>
+              select(-D_j2),
+            by = c("oid" = "id")) |>
+  left_join(LU2 |>
+              select(-O_i2),
+            by = c("did" = "id"))
+```
+
 In this case, $V_{ij}$ would be:
 ```{r}
-#| eval: false
 
@@
 }
-
-dc_results <- doubly_constrained_gravity_model(O=P_i, D=O_j, impedance_matrix = 1/C)
-dc_results$Tij
+```
+
+```{r}
+Impf1 <- C$f1 |> 
+  matrix(nrow = 3, ncol = 3)
+
+Impf2 <- C$f2 |> 
+  matrix(nrow = 3, ncol = 3)
+
+Impf3 <- C$f3 |> 
+  matrix(nrow = 3, ncol = 3)
+```
+
+```{r}
+dc_acc_1 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf1)
+```
+
+```{r}
+dc_acc_2 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf2)
+```
+
+```{r}
+dc_acc_3 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf3)
 ```
@@
 ```{r}
-#| eval: false
-
-A_i_doubly <- dc_results$Ai
-B_j_doubly <- dc_results$Bj
+
+A_i_doubly <- dc_acc_1$Ai
+B_j_doubly <- dc_acc_2$Bj
 
@@
 
-V_doubly <- dc_results$Tij
-
-P_i_doubly <- V_doubly |> rowSums()
-O_j_doubly <- V_doubly |> colSums()
+V_dc_1 <- dc_acc_1$Tij
+
+P_i_doubly <- V_dc_1 |> rowSums()
+O_j_doubly <- V_dc_1 |> colSums()
 V_TOTAL_dest <- ifelse(sum(P_i_doubly)==sum(O_j_doubly), sum(V_doubly), NA)
@@
 
-Nice. 
--->
+Nice. But notice how the accessibility is simply the number of opportunities at the destination! In other words, it is simply the access.
 

### 2025-02-21 paper/manuscript.qmd
cleaned up draft up until singly-constrained access. Consolidated sections, removed some commented-out work, decided against including an emp. example for now.

@@
       in_header: 'gt_packages.sty'
+      #\usepackage{float}
 ---
@@
   message = FALSE,
+  fig.pos = 'H', #to print figures on the line specified in knitted output
   comment = '', 
@@
 Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
+There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space.
+At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
 
@@
 
-Parting from this, it is possible to derive a family of spatial interaction models.
-In Wilson's terms, there is the "unconstrained" (but actually total flow constrained) model in @eq-phys-gravity-model.
+Building on this, a family of spatial interaction models can be derived.
+In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model.
 The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
@@
 Said dictates and needs were far from trivial.
-In the US alone, the Federal-Aid Highway Act of 1956 authorized the creation of the US Interstate Highway System, with a budget of more than one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S.
+Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
 Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
@@
 In this matrix, two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common.
-@fig-Fig1 presents the results.
-In this figure, each symbol represents a document and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents (i.e., those sharing more items in their lists of references) are plotted closer to each other.
+@fig-Fig1 presents the results, where each symbol represents a document.
+The distance between symbols on the plot indicates the strength of bibliographical coupling: documents that are more strongly coupled (i.e., those sharing more references) are plotted closer to each other.
 
@@
 #| out-width: 70%
-#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959) and/or Wilson (1971)."
+#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959), Wilson (1971), or both."
 
@@
 
+As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
+However, we can discern from a manual reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., _potential for_ spatial interaction) and another on spatial interaction.
+The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected.
+
+The spatial-interaction-focused stream of literature inspired by @wilson1971 that cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
+Specifically, works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
+99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
+Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the _competition destination_ model).
+Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research d[@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
+Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+
+On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons:
+- Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+- Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+While accessibility is an expression of _potential for_ spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
+This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues.
+Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
+
 <!-- 
@@
 
-Using these constraints, the SIM @eq-phys-gravity-model is distinguished into three general cases.
-These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling].
 -->
@@
 
-As we argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact after @hansen1959 and @wilson1971.
+As argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with little contact after @hansen1959 and @wilson1971.
 This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis.
-This is intriguing, considering that Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the similarity between the denominator of @eq-production-constrained-balancing-factor and Hansen's accessibility indicator.
-In fact, Wilson noted that Hansen's accessibility was the inverse of $A_i$, which offered a potential interpretation for the proportionality constants $A_i$ (p.
-10):
+This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of @eq-production-constrained-balancing-factor and is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
+10]:
 
@@
 This, however, only helps to illustrate the ease with which confusion may arise in this context.
-The relationship in @eq-Ai-as-accessibility appears to be misguided when we note that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force:
-
-$$
-F = G\frac{M_1\cdot M_2}{d_{12}^2}
-$$
-
-\noindent to the pairwise population potential:
-$$
-V_1 = G\frac{M_2}{d_{12}}
-$$
-
-\noindent and from there to the system-wide population potential (after losing $G$):
-$$
-V_i = \sum_j \frac{M_j}{d_{ij}}
-$$
-
-It is useful to think of the accessibility as a partial sum of the spatial interaction, before we return to Wilson's general model:
-
-$$
-T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$
-
-Then, a way to define the _potential for interaction_ (i.e., accessibility) is as follows:
+The relationship in @eq-Ai-as-accessibility appears to be misaligned.
+In the preceding section on early research, we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the pairwise population potential $V_1 = G\frac{M_2}{d_{12}}$ and from there to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
+While the inverse of $A_i$ may appear to be accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
+
+Hence, we propose a more useful definition: accessibility as a partial sum of the spatial interaction. A way to define the _potential for spatial interaction_ (i.e., accessibility) is as follows:
 
@@
 
-<!--
-.
-Potential interaction is the preoccupation of accessibility measures.
-The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of SIMs, however with different interpretation.
-In this section, we explicitly define these cases.
--->
-
-Next, we show how various accessibility measures can be defined depending on how we choose to constrain them.
-
-## Unconstrained accessibility
-
-Unconstrained accessibility results when we set the proportionality constant/balancing factor to one in @eq-access-01:
-
-$$
-V^0_{ij} = W_j^{(2)} f(c_{ij})
-$${#eq-unconstrained-access}
-
-Then, the partial sum becomes:
-
-$$
-V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
-$${#eq-unconstrained-accessibility}
-
-We can see that in this case we simply have Hansen's accessibility. In general, the following does not hold true:
-
-$$
-\sum_i V^0_{i} = O
-$${#eq-no-constraint0-accessibility}
-
-\noindent where $O$ is the total number of opportunities in the system.
-In other words, the sum of all accessibility scores will generally be different from the total of the opportunities available in the region.
-To complicate matters, the difference will depend on the choice of impedance function $f(c_{ij})$ and any associated parameters, as well as the number of locations $n$ for which an accessibility score is calculated.
-For this reason, unconstrained accessibility can only be appropriately used as an ordinal variable to make comparisons of size (greater than, less than, equal to), but not to calculate intervals (the magnitude of differences) or ratios.
-
-## Total opportunity constrained accessibility
-
-We can choose to retain the proportionality constant in @eq-access-01 to obtain the following:
-
-$$
-V^T_{ij} = k^DW_j^{(2)} f(c_{ij})
-$${#eq-total-constrained-access}
-
-Our accessibility measure now becomes:
-
-$$
-V^T_i = \sum_j V^T_{ij} = k^D\sum_j W^{(2)}_jf(c_{ij}) = k^DV^0_i
-$${#eq-total-constrained-accessibility}
-
-The constraint that we impose in this case is the total number of opportunities $D$ in the region (compare to @eq-constraint0-gravitymodel):
-
-$$
-\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
-$${#eq-total-constraint-accessibility}
-
-Substituting @eq-total-constrained-accessibility in @eq-total-constraint-accessibility, and solving for $k$, we have that (compare to the constant in the total flow spatial interaction model, Equation 2.11 in @cliff_evaluating_1974):
-
-$$
-k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$${#eq-total-opportunity-balancing-factor}
-
-Let us write our total opportunity constrained accessibility as follows:
-
-$$
-V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \frac{D}{\sum_i\sum_j V^0_{ij}}\sum_j W^{(2)}_j f(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
-$$
-
-Further, we can see that, since $O$ and $W^{(2)}_j$ are opportunities, the following term is dimensionless:
-
-$$
-\kappa^D_{ij} = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination:
-
-$$
-V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij} 
-$$
-
-Notice that the role of $\kappa^0_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
-Constant $\kappa^0_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-For this reason we call it a proportional allocation factor.
-
-## Total population constrained accessibility: market potential
-
-One might wish to measure accessibility to the mass of the _population_ at $i$ instead of the opportunities at $j$. This is equivalent to the concept of market potential:
-
-$$
-M^T_{ij} = k^OW_i^{(1)} f(c_{ij}) = k^OM_{ij}^0
-$${#eq-total-constrained-market}
-
-\noindent with:
-
-$$
-M_i^0 = W_i^{(1)} f(c_{ij})
-$$
-
-The population-constrained market potential becomes:
-
-$$
-M^T_j = \sum_i M^T_{ij} = k^O\sum_i W^{(1)}_if(c_{ij})
-$${#eq-total-constrained-market}
-
-The constraint that we impose in this case is the total market potential equals the total population $O$ in the region:
-
-$$
-\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
-$${#eq-total-constraint-market}
-
-Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $k$, we obtain:
-
-$$
-k^O=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
-$${#eq-total-population-balancing-factor}
-
-The constrained market potential then takes the following form:
-
-$$
-M^T_i = k\sum_i W^{(1)}_if(c_{ij}) = \frac{O}{\sum_i\sum_j M^T_{ij}}\sum_i W^{(i)}_i f(c_{ij})
-$$
-
-## Singly-constrained accessibility: opportunity-constrained
-
-If we decide to introduce additional information into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (compare to @eq-constraint2-gravitymodel):
-
-$$
-\sum_i V^D_{ij} =  D_j
-$${#eq-attraction-constraint}
-
-The underlying spatial interaction model is now the attraction-constrained model in @eq-attraction-constrained-gravitymodel, and our accessibility measure becomes:
-
-$$
-V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
-$${#eq-opportunity-constrained-accessibility}
-
-\noindent where $W_i^{(1)}$ is a measure of the mass at origin $i$ (i.e., the opportunity-seeking population). The corresponding balancing factor, as per Wilson, is:
-
-$$
-B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-proportionality-constants}
-
-If we introduce this balancing factor in @eq-opportunity-constrained-accessibility we obtain:
-
-$$
-V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-accessibility-with-balancing-factor}
-
-Further, we define the following proportional allocation factor:
-
-$$
-\kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-proportional-allocation-factor}
-
-After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as:
-
-$$
-V^D_{i} = \sum_j  \kappa^D_{ij} D_j
-$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
-
-@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^D_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As before, $\kappa^D_{ij}$ is dimensionless and $V_i^D$ is in the units of opportunities $D_j$.
-The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that there is no competition for them.
-These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
-
-$$
-v^D_{i} = \frac{V^D_{i}}{W^{(1)}_i}
-$$
-
-As an additional point, the constraint in @eq-attraction-constraint ensures that the following also holds:
-
-$$
-\sum_i V^D_{i} = D
-$$
-
-For this reason, $\frac{V_i^D}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system
-
-## Singly-constrained accessibility: population-constrained
-
-Not the most common application of accessibility analysis, but if the origins and destinations are transposed, we can define a measure of market potential that preserves the population, similar to @eq-total-population-balancing-factor.
-The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $V^P_i$ becomes:
-
-$$
-M^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
-$${#eq-population-constrained-accessibility}
-
-In this case the measure is constrained by the population _by origin_ (i.e., $O_i$). Compare to @eq-constraint2-gravitymodel:
-
-$$
-\sum_j M^P_{ij} =  O_i
-$${#eq-population-constraint}
-
-The corresponding balancing factor, as per Wilson, is:
-
-$$
-A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$${#eq-population-constrained-proportionality-constants}
-
-Following the same logic as in the preceding section, one arrives at the following expression:
-
-$$
-M^P_{j} = \sum_i  \kappa^O_{ij} O_i
-$${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
-
-\noindent with:
-
-$$
-\kappa^O _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
-$${#eq-attraction-constrained-proportional-allocation-factor}
-
-In addition to market potential, $M_j^P$ can be interpreted as a constrained measure of "facility crowdedness" as in @wang_inverted_2018.
-This "crowdedness" now satisfies, in addition to @eq-attraction-constraint, the following:
-
-$$
-\sum_j M^P_{j} = O
-$$
-
-\noindent where $O$ is the total population of the region. In this way, $\frac{M_j^O}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
-
-<!--
-
-## Gravity-based accessibility's cousin: Wilson's spatial interaction model
-Today's gravity-based accessibility measures (@eq-accessibility-general) share the same early 20th century Newtonian-based origins as the model's of Reilly, Zipf, and Stewart, versions of which continue to be extensively used today in transportation demand modelling [@ortuzar_2011_modelling;@wilson1971].
-However, in many ways these spatial interaction models, commonly refered to as gravity models, continued to evolve independently from accessibility (@eq-accessibility-general).
-Our paper aims to demonstrate that Hansen-type accessibility can be interpreted as a SIM.
-This connection can enrich the accessibility literature by linking it to the extensive practical application of SIMs.
-This paper's focus is on how SIMs utilize "proportionality constants" and constraints to interpret results, and how Hansen-type accessibility can transit to _constrained_ accessibility.
-
--->
-
-<!--
-## Interpreting SIM constraints for accessibility
-
-SUMMARY: Introduce the SIM constraints generally, and explain them in terms of accessibility (e.g., P_i as weight at origin and O_j as weight at destination)
-
-Then, detail a synthetic example for 'non-constrained' (e.g., HAM), Total constrained, Singly-constrained (both versions), Doubly-constrained and finally Hybrid constrained (?maybe?).
-
-Instead of an interaction variable as in the SIM (e.g., trips, spending, sales), we focus on _potential_ interaction (e.g., accessible opportunities).
-Potential interaction is the preoccupation of accessibility measures.
-The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of SIMs, however with different interpretation.
-In this section, we explicitly define these cases.
-
-In this context, accessibility is our _interaction_ variable of choice, so $T_{ij}$ in @eq-phys-gravity-model is replaced with $V_{ij}$.
-As the interaction variable represents _potential_, i.e., the number of potential reachable opportunities, as opposed to trips:
-
-$$
-V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
-$${#eq-access}
-
-Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.
-Notably, the mass vectors $W_i^{(1)}$ and $W_j^{(2)}$ can be replaced by $P_i$ and $O_j$, representing the weight of the origin (population) and destination (opportunities) as known, the weights common in accessibility literature.
-As accessibility is the sum of opportunities for a $j$ from $i$, when the mass of the origin ($P_i$) is relevant, it enters @eq-access through the $k$ constant.
-Accessibility is also most often presented as a locational measure, so the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone:
-
-$$
-V_{i} = k W_i^{(1)} \sum_j W_j^{(2)} f(c_{ij})
-$${#eq-access-sum}
-
-Like with @eq-phys-gravity-model, we distinguish @eq-access into three cases based on versions of the gravity-model building constraints (@eq-constraint0-access,  @eq-constraint1-access, and @eq-constraint2-access) to reflect macro-level knowledge, i.e., the sum of the number of potential reachable opportunities.
-
-$$
-\sum_i\sum_j V_{ij} = V
-$${#eq-constraint0-access}
-\noindent Where the sum of all potentially reached opportunities in the region is equal to a total opportunities in the region $V$.
-
-$$
-\sum_j V_{ij} = P_i
-$${#eq-constraint1-access}
-\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.
-
-$$
-\sum_i V_{ij} =  O_j
-$${#eq-constraint2-access}
-
-\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.
-
-All three cases are differentiated as follows:
-- The total constrained case: only @eq-constraint0-access hold. In this case, $V$ is the number of potential for interaction 
-- The singly-constrained case: either @eq-constraint1-access holds (origin-constrained)  or @eq-constraint2-access holds (destination-constrained) -- can only be in any combination of "potential for interaction" units, e.g., units of "opportunities" or "population" or "population for opportunities" or "opportunities for population;
-- The doubly-constrained case: both @eq-constraint1-access and @eq-constraint2-access hold simultaneously production-attraction
-
-Notably, $V_{ij}$ can be in any combination of "potential for interaction" units that reflect the meaning of the spatial weights matrix e.g., units of "opportunities" from $i$ to $j$, or "population" from $i$ to $j$, or "population for opportunities" from $i$ to $j$, or "opportunities for population" from $i$ to $j$.
-The meaning of the units depends on the constraints: e.g., if singly-constrained - the spatial weights matrix represents units of opportunity or population while in doubly-constrained population and opportunities are two sides of this same matrix so take on the meaning of "population for opportunities" or "opportunities for population" from $i$ to $j$.
-
-Also it is relevant to describe previous accessibility literature relation to this framework:
-- Nonconstrained cases: current practice, the HAM, where @eq-constraint0-access, @eq-constraint1-access, nor  @eq-constraint2-access hold.
-- HYBRID CASE: examples of where @eq-constraint1-access or @eq-constraint2-access partly hold BFCA? Others?
-
-In these cases, $V_{ij}$ is never explicitly defined but we work backwards to develop it and place it within the SIM framework.
-
-Like the conventional SIM, $V_{ij}$ can be solved using a travel cost matrix and a spatial weights matrix but with alternative meanings.
-$P_i$ (our $W_i^{(1)}$ if known) is population at $i$ who seek opportunities.
-$O_j$ (our $W_j^{(2)}$ if known) are the opportunities being sought.
-For this simple example, we can say that the opportunities are the population's jobs.
-In our region, each person has a job - but we only know at an aggregate how many people are at each $i$, the jobs that are located at $j$ and the travel cost $c_{ij}$.
-We assume the the travel impedance ($f(c_{ij})$) is the associated travel propensity of $f(c_{ij}) = 1/C$.
--->
-
-## Doubly-constrained
-
-Readers will have noticed that the accessibility measures above have used either $O_i$ or $D_j$ but not both.
-When $D_j$, the opportunities, were used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
-Contrariwise, When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
-The reason is that the population and the opportunities are typically different things without a common metric.
-For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
-There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most common one being employment, i.e., one person, one job.
-
-A doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities.
-In this case, the two constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel are imposed.
-Furthermore, the two constraints must match:
-
-$$
-\sum_iO_i = \sum_j D_j
-$$
-
-This is because now the proportionality constants are mutually dependent and cannot be estimated if the constraints diverge.
-For this reason, the applicability seems to be relatively limited, because even in the case of jobs, there might be imbalances between $O$ and $D$, say, where there are more jobs than people or vice-versa.
-
-Another important point to note about the doubly-constrained approach is as follows.
-Recall that accessibility is defined as "the potential for interaction".
-Additional constraints leave less room for that potential.
-In fact, a doubly constrained approach essentially converges on the spatial interaction model, where one person represents one trip to one opportunity, and it is probably better thought of as an estimate of "access" instead of "accessibility".
-
-Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
-The values of $B_j$ are used to update the underlying trip matrix, before calibrating $A_i$.
-This process continues to update $A_i$ and $B_j$ until a convergence criterion is met [see @ortuzar_2011_modelling, p.
-193-195].
-
-<!--
-In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
-In this case, the units of the population and opportunities must be able to be understood as the same.
-E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are sought by population at each $i$ must both be known to match.
-
-In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
-In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
-If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
-
-Literature has explored 'doubly-constrained' but -- not the same: , etc. Ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
--->
-
-# Simple numerical example
+In the following subsections, we detail how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus.
+Together, these constraints delineate different members of the family of accessibility measures.
+Along with a simple numerical example, we detail the unconstrained case (@hansen1959's formulation), then total constrained cases (opportunity-constrained and population-constrained), then singly-constrained cases (opportunity-constrained and population-constrained), and lastly the doubly-constrained case.
+These cases are presented in increasing order of constraint restrictiveness and information availability.
+
+## Simple numeric example setup
 
@@
 #| label: simple-numerical-example-opportunities-and-population
-
 id <- c("1", "2", "3")
@@
 
-In this section we use a simple numerical example to illustrate the various members of the family of accessibility measures introduced above.
-Consider a very simple system with three zones, as shown in @tbl-small-system-land-use.
-It can help to think of these numbers as population in 10,000s and number of physicians (for comparison, the number of physicians per 10,000 in Canada in 2022 was 24.97).
-If so, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
+Consider a simple region with three zones, IDs 1, 2 and 3.
+Each zone is both an origin $i$ and a destination $j$.
+The following three pieces of information are defined: zonal population and opportunities, zonal cost matrix, and travel impedance functions for three types of travel behaviour.
+In this example, the population are people, the opportunities are physicians, and the region can be described by three possible travel behaviours.
+
+Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone.
+^[For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]].
+Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
 The population is $W_i^{(1)}$ when used as a proxy for the mass at the origin, and $O_i$ when used as a constraint.
 Similarly, the opportunities are represented by $W_j^{(2)}$ when used as a proxy for the mass at the destination, and $D_j$ when used as a constraint.
-
-To complete the set-up of this example, the cost of movement between origins and destinations is as shown in @tbl-small-system-cost.
-
-<!--The number of physicians per 10,000 in Canada in 2022 was 24.97 according to: https://en.wikipedia.org/wiki/List_of_countries_by_number_of_physicians -->
 
@@
   #             columns = 3) |>
-  cols_label(id = "ID",
+  cols_label(id = "ID (i or j)",
              O_i = "Population", 
@@
 
+Secondly, to pair with the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
+From both @tbl-small-system-land-use and @tbl-small-system-cost, it can be interpreted that Zones 1 and 3 are more proximate to each other than to Zone 2.
+Combined, Zone 1 and 3 have the same amount of population as Zone 2 but more than double the opportunities.
+<!--for the readers, what sort of set-up could this reflect practically?
+-->
+
 ```{r}
@@
                   did = c("1", "2", "3")) |>
-  mutate(cost = c(10, 30, 15, 30, 10, 25, 10, 25, 10)) |>
+  mutate(cost = c(10, 30, 15, 30, 10, 25, 15, 25, 10)) |>
   mutate(f1 = cost^-3,
@@
 
+And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region:
+$$
+\begin{array}{l}
+f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
+f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
+f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
+\end{array}
+$$
+
+Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
+A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility.
+Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
+As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.
+
+Any set of concepts representing population, opportunities, and their associated travel behaviour -whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups- can be substituted into our simple example, depending on the research question.
+The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behavior groups and members of the accessibility family.
+
+## Unconstrained accessibility
+
+Setting the balancing factor $k$ to one in @eq-access-01 results in the unconstrained accessibility case:
+
+$$
+V^0_{ij} = W_j^{(2)} f(c_{ij})
+$${#eq-unconstrained-access}
+
+Then, the partial sum becomes the following: simply Hansen's accessibility $S_i$ [@hansen1959], standard practice in accessibility measurement:
+
+$$
+V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
+$${#eq-unconstrained-accessibility}
+
+Where the sum of $V^0_{i}$ in the region generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), as setting $k$ to 1 to arbitrarily eliminates the term and imposes no meaningful interpretation to the resulting values.
+To complicate matters, when comparing $V^0_{i}$ values between travel behaviour contexts, their difference depends on the choice of impedance function $f(c_{ij})$, associated parameters, and the number of zones for which an accessibility score is calculated.
+For this reason, the unconstrained accessibility should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
+
+Returning to the simple example, consider the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in @tbl-simple-example-unconstrained-accessibility.
+
 ```{r}
@@
 
-For comparison, we use three different impedance functions:
-$$
-\begin{array}{l}
-f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
-f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
-f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
-\end{array}
-$$
-
-## Unconstrained accessibility
-
-The following case is standard practice, the HAM.
-It is expressed as a value of the origin, a summary of the destination weight multiplied by the travel impedance.
-There is no explicit consideration of a proportionality constant ($k$), we can presume it is still set to 1 as in @stewartPrinciples1947.
-Neither are there origin weights $W_i^{(1)}$, we assume is also wrapped up in this proportionality and implicitly set to 1.
-Working backward from this expression and assumptions, $S_{ij}$ can be expressed more generally as well.
-
-$S_i$ reflects the magnitude of opportunity accessibility but it is not scaled to equal the number opportunities $O_j$ and/or population $P_i$ (only in some case when $k$ and $W_i^{(1)}$ equal 1).
-
-In solving the example, each origin (row) all the opportunities at each destination are available. Likewise, each destination (column) offers all opportunities to each origin:
-
-```{r}
-#| label: tab-simple-example-unconstrained-accessibility
+```{r}
+#| label: tbl-simple-example-unconstrained-accessibility
 #| tbl-cap: "Simple system: unconstrained accessibility."
@@
   gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Origin") |>
+  tab_stubhead(label = "Zone") |>
   tab_spanner(label = "{{V[_i^0]}}",
@@
   fmt_number(decimals = 3) |>
-  cols_label(V_unc_i_1 = md("units: *physicians per min*^3"),
-             V_unc_i_2 = md("units: *physicians per min*^2"),
-             V_unc_i_3 = md("units: *physicians per min*^0.1"))  |>
+  cols_label(V_unc_i_1 = md("units: *f_1(c_ij)-weighted physicians"),
+             V_unc_i_2 = md("units: *f_2(c_ij)-weighted physicians"),
+             V_unc_i_3 = md("units: *f_3(c_ij)-weighted physicians"))  |>
   # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
@@
 
-Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
-```{r}
-unc_acc$V_unc_i_1[2] - unc_acc$V_unc_i_1[1]
-unc_acc$V_unc_i_2[2] - unc_acc$V_unc_i_2[1]
-unc_acc$V_unc_i_3[2] - unc_acc$V_unc_i_3[1]
-
-unc_acc$V_unc_i_1[2]/unc_acc$V_unc_i_1[1]
-unc_acc$V_unc_i_2[2]/unc_acc$V_unc_i_2[1]
-unc_acc$V_unc_i_3[2]/unc_acc$V_unc_i_3[1]
-```
-
-This illustrates how the differences and ratios are not meaningful.
-
-The total accessibility compared to the number of opportunities:
-```{r}
-unc_acc <- unc_acc |> 
-  left_join(LU |>
-              select(-D_j),
-            by = c("oid" = "id"))
-
-unc_acc  |>
-  summarize(V_unc_1 = sum(V_unc_i_1),
-            V_unc_2 = sum(V_unc_i_2),
-            V_unc_3 = sum(V_unc_i_3))
-
-LU |>
-  summarize(O = sum(D_j))
-```
-
-And the ratio of accessibility to population compared to the ratio of opportunities to population:
-```{r}
-# Accessibility divided by population
-unc_acc |> 
-  mutate(v_unc_1 = V_unc_i_1/O_i,
-            v_unc_2 = V_unc_i_2/O_i,
-            v_unc_3 = V_unc_i_3/O_i)
-
-# Sum of the accessibility divided by the total population
-unc_acc |> 
-  summarize(v_unc_1 = sum(V_unc_i_1)/sum(O_i),
-            v_unc_2 = sum(V_unc_i_2)/sum(O_i),
-            v_unc_3 = sum(V_unc_i_3)/sum(O_i))
-
-# Sum of the opportunities divided by the total population
-LU |> 
-  summarize(PPR = sum(D_j)/sum(O_i))
-```
-
-Notice the wild differences.
+As the different impedance functions represent different travel behaviours, comparing unconstrained accessibility values across groups meaningless.
+For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1]`, but in what units?
+These two values, though representing the same zone but different groups in that zone, have different units so this difference value cannot be meaningfully interpreted.
+
+However, we can compare values within the same travel behavior scenario across different zones but values are not inherently meaningful due to the units.
+For example, considering the most decaying scenario at zone 1, it captures `r unc_acc$V_unc_i_1[1] - unc_acc$V_unc_i_1[2]` more $f_1(c_{ij})$-weighted-physicians than zone 2, and `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer than zone 3.
+In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3.
+Both scenarios at zone 1 captures an intermediate $V_i^0$ value between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians.
+The ratios between the zones in the two scenarios different.
+These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared.
+<!--maybe a better way of saying this?--> This simple example illustrates that the differences and ratios of unconstrained accessibility values are not inherently meaningful.
+While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select comparable impedance functions (at the cost of accurately reflecting travel behavior), such adjustments may introduce bias.
+
+<!-- Notice the wild differences.
 Accessibility according to impedance function 3 basically says that everyone has access to all opportunities all at once.
@@
 One should not have to adjust the impedance function to obtain "reasonable" results [EXAMPLE]; instead, the impedance function should reflect the travel behavior.
-
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-We'd report origin 1 as having 148 accessibility (i.e., $S_{i=1} = 148$), origin 2 as 248 accessibility, and origin 3 as 172 accessibility.
-The total accessibility in the region is 568 ($\sum_i S_{i} = 568$), not equal to any meaningful value.
-This value is not often reported in the accessibility literature for this reason.
-One could also report the accessibility values that destinations offers ($S_{j}$), namely destination 1 offers 123, destination 2 offer 291, and destination 3 offers 154.
-However, these values are also not reported in the literature currently...
-as they aren't constrained to match up to anything known, including the number of opportunities at each j (i.e., 160, 437 and 193).
-Assuming $k=1$, $S_i$ and $S_j$ represent magnitudes of accessibility concepts, but in an nonconstrained way.
-The population matrix is not incorporated in the $S_{ij}$ calculation, so $S_i$ cannot be directly compared to population.
-From a different perspective, $S_j$ values also cannot be compared to known $O_j$ values as the equation unconstrained.
-
-OF NOTE: in the context of spatial interaction modelling, this nonconstrained version of model is seldom useful.
-While the resulting $T_{ij}$ values are technically units of 'trips' - there are no constraints, so the assumption of $k$ and $W_i^(1)$ equaling 1 is practically meaningless.
-
-Accessibility, or the _potential_ for interaction is different than interaction ($V_{ij}$ vs.
-$T_{ij}$), namely since the number of opportunities that are accessible at an origin from a destination cannot be known in the same way as a trip from an origin to a destination.
-However, we would argue that getting closer to knowing this number and planning for it - is what accessibility research should strive to capture.
-An origin-destination based accessibility value $V_{ij}$.
-We demonstrate the following three cases sub-sequently.
-
-**END OF PREVIOUS TEXT**
-
-## Total opportunity-constrained accessibility
-
-Proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$:
-
+-->
+
+## Totally-constrained accessibility
+
+A total constraint proportionally adjusts zonal accessibility values based on the total population and/or opportunities in the region.
+The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case.
+We define two cases: a _totally-constrained opportunity_ accessibility case, interpreted as Hansen's accessibility with a constraint, and a _totally-constrained population_ accessibility case, interpreted as constrained market potential.
+
+### Totally-constrained opportunity: Hansen's accessibility with a constraint
+
+Unlike in @eq-access-01, the proportionality constant $k$ is retained and not set arbitrarily to 1:
+
+$$
+V^T_{ij} = k^DW_j^{(2)} f(c_{ij})
+$${#eq-total-constrained-access}
+
+Where the total opportunity constrained accessibility measure now becomes Hansen's accessibility with a proportionality constant:
+
+$$
+V^T_i = \sum_j V^T_{ij} = k^D\sum_j W^{(2)}_jf(c_{ij}) = k^DV^0_i
+$${#eq-total-constrained-accessibility}
+
+The constraint that we impose in this case is the total number of opportunities $D$ in the region.
+
+$$
+\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
+$${#eq-total-constraint-accessibility}
+
+This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework.
+We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $k$ to yield:
+$$
+k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$${#eq-total-opportunity-balancing-factor}
+
+Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
+Hence, our totally-constrained opportunity accessibility model is:
+
+$$
+V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
+$$
+
+Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term is dimensionless:
+
+$$
+\kappa^D_{ij} = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij}$).
+The role of $\kappa^D_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+Constant $\kappa^D_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+For this reason, we refer to it as a proportional allocation factor.
+
+Referring to our simple numeric example, proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
 $$
@@
 # Calculate the balancing factors/proportionality constants for each of three impedance functions
-
 k_tot <- unc_acc_ij |> 
@@
 
-Use these proportionality constants to calculate the total opportunity constrained accessibility:
-```{r}
-#| label: tab-simple-example-total-opportunity-accessibility
+Using the calculated proportionality constants for all zones, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
+
+```{r}
+#| label: tbl-simple-example-total-opportunity-accessibility
 #| tbl-cap: "Simple system: total opportunity constrained accessibility."
@@
 
-Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
-```{r}
-tot_acc$V_tot_i_1[2] - tot_acc$V_tot_i_1[1]
-tot_acc$V_tot_i_2[2] - tot_acc$V_tot_i_2[1]
-tot_acc$V_tot_i_3[2] - tot_acc$V_tot_i_3[1]
-
-tot_acc$V_tot_i_1[2]/tot_acc$V_tot_i_1[1]
-tot_acc$V_tot_i_2[2]/tot_acc$V_tot_i_2[1]
-tot_acc$V_tot_i_3[2]/tot_acc$V_tot_i_3[1]
-```
-
-This illustrates how the differences and ratios are now meaningful.
-
-The total accessibility compared to the number of opportunities:
-```{r}
-tot_acc <- tot_acc |> 
-  left_join(LU |>
-              select(-D_j),
-            by = c("oid" = "id"))
-
-tot_acc  |>
-  summarize(V_tot_1 = sum(V_tot_i_1),
-            V_tot_2 = sum(V_tot_i_2),
-            V_tot_3 = sum(V_tot_i_3))
-
-LU |>
-  summarize(O = sum(D_j))
-```
-
-This shows how the number of opportunities is preserved.
-
-And the ratio of accessibility to population compared to the ratio of opportunities to population:
-```{r}
-# Accessibility divided by population
-tot_acc |> 
-  mutate(v_tot_1 = V_tot_i_1/O_i,
-            v_tot_2 = V_tot_i_2/O_i,
-            v_tot_3 = V_tot_i_3/O_i)
-
-# Sum of the accessibility divided by the total population
-tot_acc |> 
-  summarize(v_tot_1 = sum(V_tot_i_1)/sum(O_i),
-            v_tot_2 = sum(V_tot_i_2)/sum(O_i),
-            v_tot_3 = sum(V_tot_i_3)/sum(O_i))
-
-# Sum of the opportunities divided by the total population
-LU |> 
-  summarize(PPR = sum(D_j)/sum(O_i))
-```
-
-## Total population-constrained accessibility (market potential)
-
-Proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$:
-
-$$
-\begin{array}{l}
-k^O = \frac{O}{\sum_{i}\sum_{j} O_i f(c_{ij})}\\
-k^O = \frac{O}{\frac{O_1}{c_{11}^3}+\frac{O_1}{c_{21}^3} + \frac{O_1}{c_{31}^3} + \cdots + \frac{O_3}{c_{31}^3} + \frac{O_3}{c_{32}^3} + \frac{O_3}{c_{33}^3}
-}\\
-k^O = \frac{490}{0.75} \\
-k^O = 653.3256
-\end{array}
-$$
+In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios.
+Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^D_{ij}$.
+
+Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`).
+However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
+Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$)  scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
+Between zonal comparisons can be more interpretably made for these travel behaviour scenarios as well.
+
+One can also directly compare values at a specific zone due to the consistent units.
+For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case.
+However, the difference between travel behaviour scenarios differ in direction.
+Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively.
+Why?
+$k^D$ ensures proportional allocation for each travel behaviour scenario.
+Meaning, while the unconstrained accessibility increases, $k^D$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
+As the decay behaviour decreases, more opportunities are accessible for all zones.
+In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones.
+However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).
+
+Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @han1959's accessibility measure.
+Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
+
+### Totally-constrained population: market potential
+
+Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974].
+The unconstrained form of market potential has been used in recent research  [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road] to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects.
+To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.
+
+$$
+M^T_{ij} = k^OW_i^{(1)} f(c_{ij}) = k^OM_{ij}^0
+$${#eq-total-constrained-market}
+
+\noindent with:
+
+$$
+M_j^0 = W_i^{(1)} f(c_{ij})
+$$
+
+The population-constrained accessibility (market potential) becomes:
+
+$$
+M^T_j = \sum_i M^T_{ij} = k^O\sum_i W^{(1)}_if(c_{ij})
+$${#eq-total-constrained-market}
+
+The constraint that we impose in this case is the total market potential equals the total population $O$ in the region :
+
+$$
+\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
+$${#eq-total-constraint-market}
+
+Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $k$, we obtain:
+
+$$
+k^O=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
+$${#eq-total-population-balancing-factor}
+
+The constrained market potential then takes the following form:
+
+$$
+M^T_i = k\sum_i W^{(1)}_if(c_{ij}) = \frac{O}{\sum_i\sum_j M^T_{ij}}\sum_i W^{(i)}_i f(c_{ij})
+$$
+
+Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility.
+Readers should not, zonals trends in accessible population and accessible physicians.
+in the preceding subsection (i.e., calculated using totally-constrained opportunity) are not the same: zone 1, 2, 3 represent destinations and the accessibility values represent the number of accessible people from the vantage of physicians.
+
+<!--maybe we should say a little more? -->
 
@@
 
-Use these proportionality constants to calculate the total opportunity constrained accessibility:
-```{r}
-#| label: tab-simple-example-total-population-accessibility
+```{r}
+#| label: tbl-simple-example-total-population-accessibility
 #| tbl-cap: "Simple system: total opportunity constrained accessibility."
@@
 
-<!--
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-While we can see that the 'number of potentially reachable opportunities' V results, the total access at a each $i$ or $j$ is not equal to the known known total access ends (because these constraints are not specified and thus not satisfied by the model).
-This approach is not yet seen in accessibility literature, but could be an intelligible way to compare accessibility values ($V_i$ or $V_j$) without considering the population matrix.
-
-This case can be reformulated if the spatial weights matrix reflects only weights at the origin $W_i^(1)$ such as population $P_i$ and $W_j^(2)$ is unknown/irrelevant.
-
-**END OF PREVIOUS TEXT**
--->
-
-## Singly-constrained accessibility: opportunity-constrained
-
-<!--
-The singly-constrained case can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$.
-For the origin- and destination- constrained versions, $k$ is replaced with $A_{i}$ and $B_{j}$ respectively (@eq-access enforcing constraint @eq-constraint1-access or @eq-constraint2-access, and implicitly @eq-constraint0-access):
-
-$$
-V_{ij}''^{(1)} = A_{i} W_j^{(2)} P_i f(c_{ij})\\
-V_{ij}''^{(2)} = B_{j} O_j W_i^{(1)} f(c_{ij})\\
-$${#eq-access-singlyconstrained-origin}
-
-\noindent Where $A_{i}$ and $B_{j}$ are the same as the balancing factors of the SIM for the singly-constrained cases:
-$$
-A_{i} = \frac{1}{\sum_j B_{j} O_j f(c_{ij})}\\
-B_{j} = \frac{1}{\sum_i A_{i} P_i f(c_{ij})}\\
-$$
-
-\noindent NOTE: if the unknown spatial weight ($W_j^{(2)}$ for $A_{i}$ and $W_i^{(1)}$ for $B_{j}$) is moved from @eq-access-singlyconstrained-origin's numerator to $A_{i}$ or $B_{j}$ it is equivalent to a simplified version of Spatial Availability's formulation which is also demonstrated to be related to @shen1998 and the 2SFCA (see appendix @soukhovIntroducingSpatialAvailability2023).
-
-In solving the origin-constrained $V_{ij}^{''(1)}$ case (e.g., @eq-constraint1-access of $\sum_j V_{ij} = P_i$ is enforced), $B_{j} = 1$.
-In solving the destination-constrained $V_{ij}^{''(2)}$ case (e.g., @eq-constraint2-access of $\sum_i V_{ij} = O_j$ is enforced), $A_{i} = 1$.
-
-Solving for the destination-constrained case $V_{ij}^{''(2)}$ may be helpful in an example where the number of people seeking opportunities is known ($P_i$), along with the number of opportunities $O_j$ offered by a destination as in total accessibility.
-An example could be the number of people at each residence seeking hospital-beds (origin) and hospital-beds (opportunities) at each healthcare location (destination) to calculate the opportunity access at each origin.
-Another case could be calculating the opportunity access at each origin using the quantity of goods available to purchase (opportunities) at each commercial location (destination).
-
-From our review, related versions of this defined destination-constrained accessibility are formulaically similar to retail accessibility models e.g., decision model for shops to optimally spatially locate to maximize their potential consumer interaction.
-Starting with reilly's law of gravitational retail [@reilly1929methods], formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.
-Other literature also demonstrates  similar _competition_ considerations through its analytical formulation; for instance, in the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGOPPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, and competition consideration expressed as a per population rate particularly in healthcare and job opportunity accessibility analysis [@shen1998], and 2FCA methods in general [@soukhovIntroducingSpatialAvailability2023].
-Most of these works do not explicitly focus on balancing accessibility units; rather, they center on _competition_ considerations.
-We suspect this may be why they have been used in case studies examining access to opportunities that are inherently competitive or where the concept of supply and demand resonates.
-It was only until the recent work of [@soukhovIntroducingSpatialAvailability2023], where the units of accessibility given competition, e.g., the spatial availability, was introduced with the focus on both competition and balanced units.
--->
-
-Returning to our simple example, the opportunity-constrained case would yield the following $B_{j}$:
-
-$$
+## Singly-constrained accessibility
+
+Increasing in restrictiveness relative to the totally-constrained accessibility measure, a single constraint proportionally adjusts the origin-destination zonal accessibility values based on the population or opportunities in the zone for all zones, while still maintaining the total system population or opportunity constraint, depending on the case.
+We define two cases: a _singly-constrained opportunity_ accessibility case and a _singly-constrained population_ accessibility case.
+
+### Singly-constrained opportunity: spatial availability(?)
+
+If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-attraction-constraint)- comparable to the single constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
+This constraint implicitly maintains the total system constraint (@eq-total-constraint-accessibility).
+
+$$
+\sum_i V^D_{ij} =  D_j
+$${#eq-attraction-constraint}
+
+The underlying spatial interaction model is now the attraction-constrained model in @eq-attraction-constrained-gravitymodel, and our accessibility measure becomes:
+
+$$
+V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
+$${#eq-opportunity-constrained-accessibility}
+
+\noindent where $W_i^{(1)}$ is a measure of the mass at origin $i$ (i.e., the opportunity-seeking population). The corresponding balancing factor, as per Wilson, is:
+
+$$
+B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-opportunity-constrained-proportionality-constants}
+
+If we introduce this balancing factor in @eq-opportunity-constrained-accessibility we obtain:
+
+$$
+V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-opportunity-constrained-accessibility-with-balancing-factor}
+
+Further, we define the following proportional allocation factor:
+
+$$
+\kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+$${#eq-opportunity-constrained-proportional-allocation-factor}
+
+After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as:
+
+$$
+V^D_{i} = \sum_j  \kappa^D_{ij} D_j
+$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
+
+@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^D_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
+As before, $\kappa^D_{ij}$ is dimensionless and $V_i^D$ is in the units of opportunities $D_j$.
+The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that there is no competition for them.
+These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
+
+$$
+v^D_{i} = \frac{V^D_{i}}{W^{(1)}_i}
+$$
+
+As an additional point, @eq-attraction-constraint also implies that the total constraint (e.g., $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$) is maintained.
+
+Due to the constraints, $\frac{V_i^D}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.
+
+Returning to the simple numeric example, the opportunity-constrained case would yield the following $B_{j}$ for $f_1(c_{ij})$:
+
+$$
+\begin{array}{l}
 B_{j} = \frac{1}{\sum_i O_i f(c_{ij})}\\
-$$
-
-$$
-\begin{array}{l}
 B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
@@
 $$
-
-Balancing factors $B_j$:
 ```{r}
@@
             B_j_3 = 1/sum(O_i * f3))
-
-B_j
-```
-
-Use these proportionality constants to calculate the singly-constrained opportunity-constrained accessibility:
-```{r}
-#| label: tab-simple-example-attraction-constrained-accessibility
-#| tbl-cap: "Simple system: singly-constrained opportunity constrained accessibility."
+# B_j_longer <- B_j
+# colnames(B_j_longer) <- c("ID","B_1","B_2","B_3")
+# B_j_longer |> as.data.frame()  |> gt(rownames_to_stub=FALSE)
+```
+
+The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
+Using these these proportionality constants, we can calculate the singly-constrained opportunity accessibility:
+
+```{r}
+#| label: tbl-simple-example-attraction-constrained-accessibility
+#| tbl-cap: "Simple system: singly-constrained opportunity accessibility."
 
@@
 
-Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
+<!--COMEBACKHERE -->
+Similar to the accessibility differences and between pairs of zone using the two different impedance functions: 
 ```{r}
@@
 
+### Singly-constrained population: market potential
+
+Similar to @eq-total-population-balancing-factor, if the origins and destinations are transposed, we can define a _singly-constrained_ measure of market potential that preserves the population.
+In it's per-capita expression, i.e., equivalent to 2SFCA, has been used to express "facility crowdedness" as in @wang_inverted_2018.
+<!-- Anything else like this in the ltierature?-->
+
+The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^P_i$ becomes:
+
+$$
+M^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
+$${#eq-population-constrained-accessibility}
+
+In this case, the measure is constrained by the population _by origin_ (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+
+$$
+\sum_j M^P_{ij} =  O_i \\
+$${#eq-population-constraint}
+
+And the corresponding balancing factor, as per Wilson, is:
+$$
+A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
+$${#eq-population-constrained-proportionality-constants}
+
+Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression:
+$$
+M^P_{j} = \sum_i  \kappa^O_{ij} O_i
+$${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
+
+\noindent with:
+
+$$
+\kappa^O _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+$${#eq-attraction-constrained-proportional-allocation-factor}
+
+As well, the constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^P_{j} = \sum_i\sum_j  M^P_{ij} = O$) is maintained.
+
+With these constraints, $\frac{M_j^O}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
+
+<!-- In this version of the singly-constrained case, the population is constrained so the $P_i$ values (right-most column) are equal to the known population, whereas the $O_j$ values (bottom row) is not.
+The total accessibility in the region is maintained at XX  -->
+<!-- To reiterate, each  $V_{ij}$ can be interpreted as  "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.  -->
+
+## Doubly-constrained accessibility
+
+Readers will have noticed that the accessibility measures above have used either $O_i$ or $D_j$ but not both.
+When $D_j$, the opportunities, were used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
+Contrariwise, When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
+The reason is that the population and the opportunities are typically different things without a common metric.
+For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
+There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most common one being employment, i.e., one person, one job.
+
+A doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities.
+In this case, the two constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel are imposed.
+Furthermore, the two constraints must match:
+
+$$
+\sum_iO_i = \sum_j D_j
+$$
+
+This is because now the proportionality constants are mutually dependent and cannot be estimated if the constraints diverge.
+For this reason, the applicability seems to be relatively limited, because even in the case of jobs, there might be imbalances between $O$ and $D$, say, where there are more jobs than people or vice-versa.
+
+Another important point to note about the doubly-constrained approach is as follows.
+Recall that accessibility is defined as "the potential for interaction".
+Additional constraints leave less room for that potential.
+In fact, a doubly constrained approach essentially converges on the spatial interaction model, where one person represents one trip to one opportunity, and it is probably better thought of as an estimate of "access" instead of "accessibility".
+
+Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
+The values of $B_j$ are used to update the underlying trip matrix, before calibrating $A_i$.
+This process continues to update $A_i$ and $B_j$ until a convergence criterion is met [see @ortuzar_2011_modelling, p.
+193-195].
+
 <!--
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-(NOTE: notably, when $B_j$ is multiplied by $P_i*f(c_{ij})$, it is a proportion (like spatial availability's total population&travel-cost balancing factor) representing how much of $O_j$ is allocated to each $i$)
-
-The $O_j$ values (bottom row) are equal to the known opportunities at each $j$, whereas the $P_i$ values (right-most column) are not.
-The total accessibility in the region is maintained at 790.
-Each $V_{ij}$ can be interpreted as $O_j$ weighted by the proportional balancing factor e.g., "the number of opportunities accessible from $j$ by the population at $i$" all weighted by travel cost, of course.
-Use this formulation of the singly-constrained case when the number of opportunities, at each $j$, that are being competed for by the population is known but the precise number of population that is competiting for opportunities is not.
-For example, public schools as a destination in a region where all children attend them.
-All school-seats at $j$ are open to all children at all $i$s.
-However, not all population $i$ are equally interested in school-seats at all $j$s (e.g., some prefer special programs, teachers, etc.); and we do not know this precise population $i$ to $j$ number.
-Hence, an opportunity-constrained version of the singly-constrained case is appropriate.
-
-However, the population-constrained version of the singly-constrained case can be appropriate in the converse situation.
-Consider the case of public urgent care clinics in a region where all people use them equally.
-Unlike schools, all population at $i$ are equally interested in urgent-care-seats at all clinics $j$.
-If there's an emergency, or not, they will visit.
-However, from an operational perspective all clinics at $j$ are _not_ interested in serving all people (e.g., they'd rather serve people who have access to primary care medicine as they are more likely to seek urgent care in actual medical emergencies).
-From this perspective, using a _population-constrained_ version of the singly-constrained case $V_i''^{(1)}$, is appropriate.
-Each $V_{ij}$ can be interpreted as $P_i$ weighted by the proportional balancing factor e.g., "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
-
-**END OF PREVIOUS TEXT**
+In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
+In this case, the units of the population and opportunities must be able to be understood as the same.
+E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are sought by population at each $i$ must both be known to match.
+
+In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
+In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
+If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
+
+Literature has explored 'doubly-constrained' but -- not the same: , etc. Ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
 -->
 
-## Singly-constrained accessibility: population-constrained
-
-Returning to our simple example, the production-constrained case would yield the following $A_{j}$ when we use $f_1(c_{ij})=1/c_{ij}^3$:
-
-$$
-A_{i} = \frac{1}{\sum_j O_j f(c_{ij})}\\
-$$
-
-$$
-\begin{array}{l}
-A_{1} =  \frac{1}{\frac{160}{10^3} + \frac{150}{30^3} + \frac{180}{15^3}} = 2.8939 \\ 
-A_{2} =  \frac{1}{\frac{160}{30^3} + \frac{150}{10^3} + \frac{180}{25^3}} = 5.9721\\
-A_{3} =  \frac{1}{\frac{160}{15^3} + \frac{150}{25^3} + \frac{180}{10^3}} = 4.2193
-\end{array}
-$$
-
-```{r}
-# Balancing factors $A_i$:
-
-A_i <- OD |> 
-  group_by(oid) |>
-  summarize(A_i_1 = 1/sum(D_j * f1),
-            A_i_2 = 1/sum(D_j * f2),
-            A_i_3 = 1/sum(D_j * f3))
-
-A_i
-```
-
-Use these proportionality constants to calculate the singly-constrained population constrained accessibility:
-```{r}
-#| label: tab-simple-example-production-constrained-accessibility
-#| tbl-cap: "Simple system: singly-constrained population-constrained accessibility (market potential)."
-
-pop_acc_ij <- OD |>
-  left_join(A_i,
-            by = "oid") |>
-  mutate(M_pop_ij_1 = A_i_1  * O_i * D_j * f1,
-         M_pop_ij_2 = A_i_2  * O_i * D_j * f2,
-         M_pop_ij_3 = A_i_3  * O_i * D_j * f3)
-
-pop_acc <- pop_acc_ij |>
-  group_by(did) |>
-  summarize(M_pop_i_1 = sum(M_pop_ij_1),
-            M_pop_i_2 = sum(M_pop_ij_2),
-            M_pop_i_3 = sum(M_pop_ij_3))
-
-pop_acc  |>
-  select(did, starts_with("M_")) |>
-  gt(rowname_col = "did") |>
-  tab_stubhead(label = "Destination") |>
-  tab_spanner(label = "{{M[_j^P]}}",
-              columns = 2:4,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 2,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 4,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-  cols_label(M_pop_i_1 = md("units: *population in 10,000s*"),
-             M_pop_i_2 = md("units: *population in 10,000s*"),
-             M_pop_i_3 = md("units: *population in 10,000s*"))  |>
-  cols_align(align = "center",
-             columns = 2:4) |>
-  cols_width(
-    stub() ~ px(90),
-    everything() ~ px(150)
-  ) |>
-  tab_options(column_labels.font.size = "small") |>
-  as_latex()
-```
-
-Compare the accessibility differences and between pairs of zone using the two different impedance functions: 
-```{r}
-pop_acc$M_pop_i_1[2] - pop_acc$M_pop_i_1[1]
-pop_acc$M_pop_i_2[2] - pop_acc$M_pop_i_2[1]
-pop_acc$M_pop_i_3[2] - pop_acc$M_pop_i_3[1]
-
-pop_acc$M_pop_i_1[2] / pop_acc$M_pop_i_1[1]
-pop_acc$M_pop_i_2[2] / pop_acc$M_pop_i_2[1]
-pop_acc$M_pop_i_3[2] / pop_acc$M_pop_i_3[1]
-```
-
-This illustrates how the differences and ratios are now meaningful.
-
-The total accessibility compared to the number of opportunities:
-```{r}
-pop_acc <- pop_acc |> 
-  left_join(LU |>
-              select(-O_i),
-            by = c("did" = "id"))
-
-pop_acc  |>
-  summarize(M_pop_1 = sum(M_pop_i_1),
-            M_pop_2 = sum(M_pop_i_2),
-            M_pop_3 = sum(M_pop_i_3))
-
-LU |>
-  summarize(O = sum(O_i))
-```
-
-This shows how the population is preserved.
-
-And the ratio of accessibility to population compared to the ratio of opportunities to population:
-```{r}
-# Accessibility divided by population
-pop_acc |> 
-  mutate(m_pop_1 = M_pop_i_1/O_i,
-         m_pop_2 = M_pop_i_2/O_i,
-         m_pop_3 = M_pop_i_3/O_i)
-
-# Sum of the accessibility divided by the total population
-pop_acc |> 
-  summarize(m_pop_1 = sum(M_pop_i_1)/sum(D_j),
-            m_pop_2 = sum(M_pop_i_2)/sum(D_j),
-            m_pop_3 = sum(M_pop_i_3)/sum(D_j))
-
-# Sum of the opportunities divided by the total population
-LU |> 
-  summarize(PPR = sum(O_i))/sum(D_j)
-```
-
-<!--
-**PREVIOUS TEXT FOLLOWS: REVISE DISCUSSION**
-
-In this version of the singly-constrained case, the population is constrained so the $P_i$ values (right-most column) are equal to the known population, whereas the $O_j$ values (bottom row) is not.
-The total accessibility in the region is maintained at 790.
-To reiterate, each  $V_{ij}$ can be interpreted as  "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
-
-**END OF PREVIOUS TEXT**
--->
-
-## Doubly-constrained case (population-opportunity constrained)
+Returning to the simple numeric example
 
@@
 
-<!--
-### Hybrid singly-constrained case: 
-NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA?
-Balanced floating catchment areas (not truly doubly constrained)
-
+# Discussion
+when to use what, their embedded assumptions, and dangers of using one over the other
+
+Wu and Lev say comp. can be considered in the imped function, but no details. we argue its' more. unconstrained - no comp,
+-- constraints consider competition
+	- total - regionally; not fixing it on what zones will be matched, it's just regional,
+	- singly- with zone and regional,
+	- doubly- per origin and destination zone.
+
+A sort of 'accessible' provider to population ratio, or accessible opportunities per capita, can also be easily calculated by dividing each constrained value by the number of opportunities. !!!
+
+# A note on interpretation and possible applications
+
+When to use each of these measures?
+<!--AS: CHECK MY OLD VERSION!  -->
+
+**Totally-constrained**
+
+**Singly-constrained**
+
+<!-- Each $V_{ij}$ can be interpreted as $O_j$ weighted by the proportional balancing factor e.g., "the number of opportunities accessible from $j$ by the population at $i$" all weighted by travel cost, of course.
+Use this formulation of the singly-constrained case when the number of opportunities, at each $j$, that are being competed for by the population is known but the precise number of population that is competiting for opportunities is not.
+For example, public schools as a destination in a region where all children attend them.
+All school-seats at $j$ are open to all children at all $i$s.
+However, not all population $i$ are equally interested in school-seats at all $j$s (e.g., some prefer special programs, teachers, etc.); and we do not know this precise population $i$ to $j$ number.
+Hence, an opportunity-constrained version of the singly-constrained case is appropriate.
 -->
 
-# Empirical examples
-
-Use the TTS2016R package to illustrate every measure.
-
-# A note on interpretation and possible applications
-
-When to use each of these measures?
+<!-- However, the population-constrained version of the singly-constrained case can be appropriate in the converse situation.
+Consider the case of public urgent care clinics in a region where all people use them equally.
+Unlike schools, all population at $i$ are equally interested in urgent-care-seats at all clinics $j$.
+If there's an emergency, or not, they will visit.
+However, from an operational perspective all clinics at $j$ are _not_ interested in serving all people (e.g., they'd rather serve people who have access to primary care medicine as they are more likely to seek urgent care in actual medical emergencies).
+From this perspective, using a _population-constrained_ version of the singly-constrained case $V_i''^{(1)}$, is appropriate.
+Each $V_{ij}$ can be interpreted as $P_i$ weighted by the proportional balancing factor e.g., "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
+-->
+
+**Doubly-constrained**
 
@@
 <!--
+### Hybrid singly-constrained case: 
+NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA?
+Balanced floating catchment areas (not truly doubly constrained)
+
+-->
+
+<!-- 
 ## QUESTIONS:
@@
 
-<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. -->
+<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. We don't have to comment on it. -->
 

### 2025-02-28 paper/manuscript.qmd
Full draft done - ready for other co-authors review!

@@
 ```
+```{r options}
+suppressMessages(library(gt))
+options(gt.html_tag_check = FALSE)
+```
 
@@
 Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
-In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of the single proportionality constant $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
-In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
+In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
+In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled: it balances the units.
 
@@
 In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model.
-The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
-
-$$
-k=\frac{T}{\sum_i\sum_j T_{ij}}
+The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
+
+$$
+K=\frac{T}{\sum_i\sum_j T_{ij}}
 $${#eq-total-flow-balancing-factor}
@@
 
-And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region:
+And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios:
 $$
@@
 \end{array}
-$$
+$${#eq-travel-behaviour-scenarios}
 
@@
   gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Zone") |>
+  tab_stubhead(label = "Origin") |>
   tab_spanner(label = "{{V[_i^0]}}",
@@
   fmt_number(decimals = 3) |>
-  cols_label(V_unc_i_1 = md("units: *f_1(c_ij)-weighted physicians"),
-             V_unc_i_2 = md("units: *f_2(c_ij)-weighted physicians"),
-             V_unc_i_3 = md("units: *f_3(c_ij)-weighted physicians"))  |>
+  cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
+             V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
+             V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
+             .fn = md)  |>
   # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
@@
 
-Unlike in @eq-access-01, the proportionality constant $k$ is retained and not set arbitrarily to 1:
-
-$$
-V^T_{ij} = k^DW_j^{(2)} f(c_{ij})
+Unlike in @eq-access-01, the proportionality constant $k$ is retained, not set arbitrarily to 1, and represented as $K^T$:
+
+$$
+V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
 $${#eq-total-constrained-access}
@@
 $$
-V^T_i = \sum_j V^T_{ij} = k^D\sum_j W^{(2)}_jf(c_{ij}) = k^DV^0_i
+V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
 $${#eq-total-constrained-accessibility}
@@
 This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework.
-We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $k$ to yield:
-$$
-k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $K^T$ to yield:
+$$
+K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $${#eq-total-opportunity-balancing-factor}
@@
 Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
-Hence, our totally-constrained opportunity accessibility model is:
-
-$$
-V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
-$$
-
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term is dimensionless:
-
-$$
-\kappa^D_{ij} = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij}$).
-The role of $\kappa^D_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
-Constant $\kappa^D_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our totally-constrained opportunity accessibility model is:
+
+$$
+V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
+$$
+
+Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term, the proportional allocation factor for the totally-constrained opportunity case $\kappa^T$ is dimensionless:
+
+$$
+\kappa^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$).
+The role of $\kappa^T$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
 For this reason, we refer to it as a proportional allocation factor.
 
-Referring to our simple numeric example, proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
+Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be:
 $$
 \begin{array}{l}
-k^D = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
-k^D = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
+K^T = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
+K^T = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
 }\\
-k^D = \frac{490}{0.75} \\
-k^D = 653.3256
+K^T = \frac{490}{0.6233422} \\
+K^T = 786.085
 \end{array}
@@
 In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios.
-Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^D_{ij}$.
+Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^T$.
 
@@
 Why?
-$k^D$ ensures proportional allocation for each travel behaviour scenario.
-Meaning, while the unconstrained accessibility increases, $k^D$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
+$\kappa^T$ ensures proportional allocation for each travel behaviour scenario.
+Meaning, while the unconstrained accessibility increases, $\kappa^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
 As the decay behaviour decreases, more opportunities are accessible for all zones.
@@
 
-Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @han1959's accessibility measure.
+Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.
 Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
@@
 Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974].
-The unconstrained form of market potential has been used in recent research  [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road] to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects.
+The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$,  has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
 To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.
@@
 $$
-M^T_{ij} = k^OW_i^{(1)} f(c_{ij}) = k^OM_{ij}^0
+M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
 $${#eq-total-constrained-market}
 
-\noindent with:
+\noindent with $M_j^0$ being the $i$ $j$ transpose of $V^0_{ij}$:
 
@@
 
-The population-constrained accessibility (market potential) becomes:
-
-$$
-M^T_j = \sum_i M^T_{ij} = k^O\sum_i W^{(1)}_if(c_{ij})
+The totally-constrained population accessibility (market potential) becomes:
+
+$$
+M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
 $${#eq-total-constrained-market}
@@
 
-Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $k$, we obtain:
-
-$$
-k^O=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
+Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $\hat K^T$, we obtain:
+
+$$
+\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
 $${#eq-total-population-balancing-factor}
@@
 $$
-M^T_i = k\sum_i W^{(1)}_if(c_{ij}) = \frac{O}{\sum_i\sum_j M^T_{ij}}\sum_i W^{(i)}_i f(c_{ij})
+M^T_i = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+$$
+Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
+$$
+\hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 $$
@@
 Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility.
-Readers should not, zonals trends in accessible population and accessible physicians.
-in the preceding subsection (i.e., calculated using totally-constrained opportunity) are not the same: zone 1, 2, 3 represent destinations and the accessibility values represent the number of accessible people from the vantage of physicians.
-
-<!--maybe we should say a little more? -->
+Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility).
+In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
+Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
+However, similar to the totally-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
+As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
 
@@
   tab_stubhead(label = "Destination") |>
-  tab_spanner(label = "{{M[_i^D]}}",
+  tab_spanner(label = "{{M[_i^S]}}",
               columns = 2:4,
@@
 
-### Singly-constrained opportunity: spatial availability(?)
-
-If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-attraction-constraint)- comparable to the single constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
-This constraint implicitly maintains the total system constraint (@eq-total-constraint-accessibility).
-
-$$
-\sum_i V^D_{ij} =  D_j
-$${#eq-attraction-constraint}
+### Singly-constrained opportunity: A.K.A spatial availability
+
+If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-opportunity-constraint)- comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
+
+$$
+\sum_i V^S_{ij} =  D_j
+$${#eq-opportunity-constraint}
 
@@
 $$
-V^D_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
+V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
 $${#eq-opportunity-constrained-accessibility}
@@
 
-If we introduce this balancing factor in @eq-opportunity-constrained-accessibility we obtain:
-
-$$
-V^D_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+Introducing the balancing factor in @eq-opportunity-constrained-accessibility, we obtain:
+
+$$
+V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
 $${#eq-opportunity-constrained-accessibility-with-balancing-factor}
@@
 $$
-\kappa^D _{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
 $${#eq-opportunity-constrained-proportional-allocation-factor}
 
-After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as:
-
-$$
-V^D_{i} = \sum_j  \kappa^D_{ij} D_j
+After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as an origin summary expression of proportionally allocated known opportunities (i.e., $D_j$):
+
+$$
+V^S_{i} = \kappa^S_{i} \sum_j   D_j
 $${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
 
-@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^D_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As before, $\kappa^D_{ij}$ is dimensionless and $V_i^D$ is in the units of opportunities $D_j$.
-The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that there is no competition for them.
+@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
+As in the totally=constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
+The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
 These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
@@
 $$
-v^D_{i} = \frac{V^D_{i}}{W^{(1)}_i}
-$$
-
-As an additional point, @eq-attraction-constraint also implies that the total constraint (e.g., $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$) is maintained.
-
-Due to the constraints, $\frac{V_i^D}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.
+v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
+$$
+
+To reinforce this point, @eq-opportunity-constraint also implies that the total system constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.
+
+Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.
 
@@
             by = "did") |>
-  mutate(V_opp_ij_1 = B_j_1 * O_i * D_j * f1,
+  mutate(k_opp_ij_1 = B_j_1 * O_i * f1 /3,
+         k_opp_ij_2 = B_j_2 * O_i  * f2 /3,
+         k_opp_ij_3 = B_j_3 * O_i  * f3 /3,
+         V_opp_ij_1 = B_j_1 * O_i * D_j * f1,
          V_opp_ij_2 = B_j_2 * O_i  * D_j * f2,
@@
   group_by(oid) |>
-  summarize(V_opp_i_1 = sum(V_opp_ij_1),
+  summarize(O_i = first(O_i) |> as.character(), 
+            V_opp_i_1 = sum(V_opp_ij_1),
             V_opp_i_2 = sum(V_opp_ij_2),
-            V_opp_i_3 = sum(V_opp_ij_3))
+            V_opp_i_3 = sum(V_opp_ij_3), 
+            .groups = "drop")
+
+opp_acc_with_ks <-  opp_acc_ij |>
+  group_by(oid) |>
+  summarize(O_i = first(O_i) |> as.character(), 
+            k_opp_ij_1 = sum(k_opp_ij_1),
+            k_opp_ij_2 = sum(k_opp_ij_2),
+            k_opp_ij_3 = sum(k_opp_ij_3))
 
 opp_acc |>
-  select(oid, starts_with("V_")) |>
+  #select(oid, O_i, starts_with("V_")) |>
   gt(rowname_col = "oid") |>
   tab_stubhead(label = "Origin") |>
-  tab_spanner(label = "{{V[_i^D]}}",
-              columns = 2:4,
+
+  tab_spanner(label = "{{V[_i^S]}}",
+              columns = 3:5,
               level = 2) |>
   tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 2,
+              columns = 3,
               level = 1) |>
   tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 3,
+              columns = 4,
               level = 1) |>
   tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 4,
+              columns = 5,
               level = 1) |>
   fmt_number(decimals = 3) |>
-  cols_label(V_opp_i_1 = md("units: *physicians*"),
+
+  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
+             V_opp_i_1 = md("units: *physicians*"),
              V_opp_i_2 = md("units: *physicians*"),
-             V_opp_i_3 = md("units: *physicians*"))  |>
+             V_opp_i_3 = md("units: *physicians*"),
+             )  |>
   cols_align(align = "center",
-             columns = 2:4) |>
+             columns = 2:5) |>
   cols_width(
-    stub() ~ px(60),
-    everything() ~ px(180)
+    stub() ~ px(60*0.8),
+    everything() ~ px(180*0.8)
   ) |>
@@
 
-<!--COMEBACKHERE -->
-Similar to the accessibility differences and between pairs of zone using the two different impedance functions: 
-```{r}
-opp_acc$V_opp_i_1[2] - opp_acc$V_opp_i_1[1]
-opp_acc$V_opp_i_2[2] - opp_acc$V_opp_i_2[1]
-opp_acc$V_opp_i_3[2] - opp_acc$V_opp_i_3[1]
-
-opp_acc$V_opp_i_1[2] / opp_acc$V_opp_i_1[1]
-opp_acc$V_opp_i_2[2] / opp_acc$V_opp_i_2[1]
-opp_acc$V_opp_i_3[2] / opp_acc$V_opp_i_3[1]
-```
-
-This illustrates how the differences and ratios are now meaningful.
-
-The total accessibility compared to the number of opportunities:
-```{r}
-opp_acc <- opp_acc |> 
-  left_join(LU |>
-              select(-D_j),
-            by = c("oid" = "id"))
-
-opp_acc  |>
-  summarize(V_opp_1 = sum(V_opp_i_1),
-            V_opp_2 = sum(V_opp_i_2),
-            V_opp_3 = sum(V_opp_i_3))
-
-LU |>
-  summarize(O = sum(D_j))
-```
-
-This shows how the number of opportunities is preserved.
-
-And the ratio of accessibility to population compared to the ratio of opportunities to population:
-```{r}
-# Accessibility divided by population
-opp_acc |> 
-  mutate(v_opp_1 = V_opp_i_1/O_i,
-            v_opp_2 = V_opp_i_2/O_i,
-            v_opp_3 = V_opp_i_3/O_i)
-
-# Sum of the accessibility divided by the total population
-opp_acc |> 
-  summarize(v_opp_1 = sum(V_opp_i_1)/sum(O_i),
-            v_opp_2 = sum(V_opp_i_2)/sum(O_i),
-            v_opp_3 = sum(V_opp_i_3)/sum(O_i))
-
-# Sum of the opportunities divided by the total population
-LU |> 
-  summarize(PPR = sum(D_j)/sum(O_i))
-```
+Imposing the single constraint $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values like previously discussed in the case of the totally-constrained opportunity case.
+Namely, when the single constraint is applied to the opportunities at $j$, each resulting accessibility value at $i$ is in units of physicians, with the impedance units already accounted for.
+
+However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ captures competition from the population.
+Again, let's consider the highest decay scenario $f_1(c_{ij})$.
+In this case, zone 1 no longer captures an intermediate amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
+Why may zone 1 capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
+This difference is due to the single-opportunity constraint $\kappa^S_{i}$'s role in $V_i^S$.
+The only inputs needed in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
+In other words, opportunities are proportionally allocated to each $i$ based on the proportion of impedance-weighted opportunities at all $j$ for that $i$ to the total sum of impedance-weighted opportunities in the entire region.
+Opportunities are allocated to $i$s, regardless of the mass weights of origin.
+Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input.
+In fact, $\kappa^S_{i}$ is the proportion of the sum of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
+Hence, since zone 1 has the lowest amount of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.
+
+From a different perspective: $V_{2}^S$ is intermediate in rank when considering the highest decay scenario ($f_1(c_{ij})$) but is the highest ranked zone under $f_3(c_{ij})$, from `r opp_acc$V_opp_i_1[2]` to `r opp_acc$V_opp_i_3[2]` accessible physicians out of `r opp_acc$V_opp_i_3|>sum() ` in the region; a `r opp_acc$V_opp_i_3[2]/opp_acc$V_opp_i_1[2]` times increase.
+Zone 2 has the highest amount of population and the least travel-impedance proximate to other zones.
+So when the the decay is high - it can most effectively capture opportunities at Zone 2, in fact it captures ($\kappa_2 ^T$) is `r opp_acc_with_ks$k_opp_ij_1[2]`, or `r opp_acc_with_ks$k_opp_ij_1[2] |> scales::percent()` of the opportunities in the region, with the vast majority coming from its own zone (`r (opp_acc_ij$k_opp_ij_1[5]/opp_acc_with_ks$k_opp_ij_1[2])|>scales::percent()` out of the total  opportunities that are allocated to Zone 2).
+This is unlike other zones, that capture opportunities from their own zones at lower rates (i.e., ($\kappa_{1}^S$ corresponds to a capture of `r (opp_acc_ij$k_opp_ij_1[1]/opp_acc_with_ks$k_opp_ij_1[1])|> scales::percent()` of its allocation from Zone 1 and $\kappa_{3}^S$ corresponds to `r (opp_acc_ij$k_opp_ij_1[9]/opp_acc_with_ks$k_opp_ij_1[3])|> scales::percent()` of its allocation from Zone 3)).
+
+In scenarios with low decay (e.g., $f_3(c_{ij})$), the connection between opportunities and population changes: spatial impedance among zones is more equal - zonal populations can get to zonal opportunities at relatively more similar travel costs.
+In this sense, spatial impedance has less relative impact on the effect of zonal population.
+For this reason, under the low decay scenario of $f_3(c_{ij})$, Zone 2 's competitive population advantage is more evident.
+Zone 1 captures more opportunities from Zones 1 and 3 than Zone 1 and 3 capture themselves (i.e., Zone 2 captures `r (opp_acc_ij$k_opp_ij_3[2])|> scales::percent()` from both Zone 1 and 3, while Zone 1 and Zone 3 only captures `r (opp_acc_ij$k_opp_ij_3[1])|> scales::percent()` and `r (opp_acc_ij$k_opp_ij_3[9])|> scales::percent()` from itself respectively.
+
+Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel scenarios.
+Again, returning to Zone 2 as an example, while its competitive population advantage is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios: this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
+In this sense, other zones are not _that_ disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
+Zones 1 and 3 are allocated opportunities at relates similar to their relative population size.
+
+In this way, the consideration of accessibility _per capita_ may be clarifying.
+Often, accessibility values are reported as raw scores without the consideration for population.
+But, as we introduced constraints, these constrained accessibility values can be normalized using anything that is relevant to the zone.
+In  @tbl-simple-example-attraction-constrained-accessibility-per-capita), we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
+These per capita rates are equivalent to the 2SFCA values.
+
+```{r}
+#| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
+#| tbl-cap: "Simple system: singly-constrained opportunity accessibility per capita."
+
+opp_acc_percapita <- opp_acc |>
+  transmute(
+    oid = id,
+    O_i = O_i,
+    v_opp_i_1 = V_opp_i_1 / (O_i|>as.numeric()),
+    v_opp_i_2 = V_opp_i_2 / (O_i|>as.numeric()),
+    v_opp_i_3 = V_opp_i_3 / (O_i|>as.numeric()))
+
+opp_acc_percapita  |>
+  gt(rowname_col = "oid") |>
+  tab_stubhead(label = "Origin") |>
+
+  tab_spanner(label = "{{v[_i^S]}}",
+              columns = 3:5,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 3,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 4,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 5,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+
+  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
+             v_opp_i_1 = md("units: *physicians per capita*"),
+             v_opp_i_2 = md("units: *physicians per capita*"),
+             v_opp_i_3 = md("units: *physicians per capita*"),
+             )  |>
+  cols_align(align = "center",
+             columns = 2:5) |>
+  cols_width(
+    stub() ~ px(60*0.8),
+    everything() ~ px(180*0.8)
+  ) |>
+  tab_options(column_labels.font.size = "small")  |>
+  as_latex()
+```
+
+Recall: our simple example was designed such that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
+And as the decay decreases and the effect of population is more independent: per capita values begin to stabilise to this regional per capita average, similar to how the accessibility values in the totally-constrained opportunity case stables to the $V_i^T$ regional average.
+These responses make intuitive sense: the use of constraints are an regional and/or zonal averaging methodology.
+By equalising the impact of travel impedance, the impact of the remaining constraint variable remains.
+In the case of the total opportunity constraint, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constraint, this is the population at a zone relative to the regional population.
+
+Like the totally-constrained opportunity case, this case offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.
+Building on this, the singly-constrained opportunity case also offers a way to consider origin-side competition while constraining the opportunities to match the input number of regional opportunities and zonal opportunity totals.
+This case builds on the previous intuition of totally-constrained opportunity case by providing additional insights from the proportional allocation factor and per capita considerations (that are mathematically equivalent to the 2SFCA, but with additional intuition).
 
@@
 
-Similar to @eq-total-population-balancing-factor, if the origins and destinations are transposed, we can define a _singly-constrained_ measure of market potential that preserves the population.
-In it's per-capita expression, i.e., equivalent to 2SFCA, has been used to express "facility crowdedness" as in @wang_inverted_2018.
-<!-- Anything else like this in the ltierature?-->
-
-The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^P_i$ becomes:
-
-$$
-M^P_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
+Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a _singly-constrained_ measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^(1)$ is now represented by $O_i$).
+In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
+<!-- Anything else like this in the literature?-->
+
+The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^S_i$ becomes:
+
+$$
+M^S_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
 $${#eq-population-constrained-accessibility}
 
-In this case, the measure is constrained by the population _by origin_ (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
-
-$$
-\sum_j M^P_{ij} =  O_i \\
+In this case, the measure is singly-constrained by the population _by origin_ (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+
+$$
+\sum_j M^S_{ij} =  O_i \\
 $${#eq-population-constraint}
@@
 $$
-M^P_{j} = \sum_i  \kappa^O_{ij} O_i
+M^S_{j} = \sum_i \hat \kappa^S_{ij} O_i
 $${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
@@
 $$
-\kappa^O _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+\kappa^S _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
 $${#eq-attraction-constrained-proportional-allocation-factor}
 
-As well, the constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^P_{j} = \sum_i\sum_j  M^P_{ij} = O$) is maintained.
-
-With these constraints, $\frac{M_j^O}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
-
-<!-- In this version of the singly-constrained case, the population is constrained so the $P_i$ values (right-most column) are equal to the known population, whereas the $O_j$ values (bottom row) is not.
-The total accessibility in the region is maintained at XX  -->
-<!-- To reiterate, each  $V_{ij}$ can be interpreted as  "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.  -->
+As well, the constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O$) is maintained.
+
+With these constraints, $\frac{M_j^S}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
+
+For the sake of brevity, we'll move forward onto the doubly-constrained case.
 
@@
 
-Readers will have noticed that the accessibility measures above have used either $O_i$ or $D_j$ but not both.
-When $D_j$, the opportunities, were used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
+Readers will have noticed that the accessibility measures above have used either $O_i$, $D_j$, or the regional sums of either, but not both.
+When $D_j$, the opportunities, was used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
 Contrariwise, When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
@@
 For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
-There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most common one being employment, i.e., one person, one job.
-
-A doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities.
-In this case, the two constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel are imposed.
-Furthermore, the two constraints must match:
-
-$$
-\sum_iO_i = \sum_j D_j
-$$
-
-This is because now the proportionality constants are mutually dependent and cannot be estimated if the constraints diverge.
-For this reason, the applicability seems to be relatively limited, because even in the case of jobs, there might be imbalances between $O$ and $D$, say, where there are more jobs than people or vice-versa.
-
-Another important point to note about the doubly-constrained approach is as follows.
-Recall that accessibility is defined as "the potential for interaction".
-Additional constraints leave less room for that potential.
-In fact, a doubly constrained approach essentially converges on the spatial interaction model, where one person represents one trip to one opportunity, and it is probably better thought of as an estimate of "access" instead of "accessibility".
+There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most commonly studied opportunity in the competitive accessibility literature being employment, e.g., one person, one job.
+Another possible opportunity type is healthcare, e.g., one person, one physician-capacity.
+
+So: a doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities is present.
+This is required because, the production- and attraction- constraints in Wilson's doubly-constrained case (@eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel) are imposed simultaneously.
+The accessibility-specific forms are used in the singly-constrained opportunities and population cases in the preceding section (@eq-opportunity-constraint and @eq-population-constraint).
+To simultaneously impose both population- and destination- constraints, they must match (@eq-opportunity-population-equality):
+
+$$
+\sum_i O_i = \sum_j D_j
+$${#eq-opportunity-population-equality}
+
+Like before, the imposition of these two constraints implies that the total system constraint is maintained e.g., the sum of all doubly-constrained accessibility $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ in @eq-total-constraint-accessibility is also maintained, and also equal to the total opportunities in the region $O$.
+
+To explicitly define the underlying accessibility measure in this case, it follows that of the production-attraction constrained (doubly-constrained) spatial interaction model in @eq-doubly-constrained-gravitymodel, hence our accessibility measure becomes:
+
+$$
+V_{i}^D = \sum_j A_i B_j O_i D_j f(c_{ij})
+$${#eq-doubly-constrained-accessibility}
+
+\noindent where the corresponding balancing factors, as per Wilson, are:
+
+$$
+\begin{array}{l}
+A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
+B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
+\end{array}
+$${#eq-doubly-constrained-balancing-factors}
 
 Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
-The values of $B_j$ are used to update the underlying trip matrix, before calibrating $A_i$.
+The values of $B_j$ are used to update the underlying $V_{ij}^D$ matrix, before calibrating $A_i$.
 This process continues to update $A_i$ and $B_j$ until a convergence criterion is met [see @ortuzar_2011_modelling, p.
@@
 
-<!--
-In the case of SIM calibrated for accessibility, the doubly-constrained case has more limited applications than the singly-constrained case because the double constraints (both marginals) are so specific.
-In this case, the units of the population and opportunities must be able to be understood as the same.
-E.g., the number of population at $i$ that seek opportunities at each $j$ and the number of opportunities at $j$ that are sought by population at each $i$ must both be known to match.
-
-In our examples of children and schools, and urgent care clincs and all people, they could be solved as doubly-constrained if additional information was known.
-In the case of urgent care children and schools, if the number of children at $i$ that visit specific schools $j$ was additionally known: then the doubly-constrained $V_{ij}$ could be interpreted as "the number of school-seats accessible from $j$ by the students at $i$ " OR "the number of students accessible from $i$ by the school-seats at $j$" - they would be equivalent statements.
-If you sub in 'urgent-care-seats' for 'school-seats' and 'population' for 'students', you'd have the analogous interpretation for the doubly-constrained case for urgent care clinics described.
-
-Literature has explored 'doubly-constrained' but -- not the same: , etc. Ex. @hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
--->
-
-Returning to the simple numeric example
+The proportional allocation factor $\kappa_{ij}^D$ would then be:
+$$
+\kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
+$${eq-doubly-constrained-proportional-allocation-factor-forpop}
+
+Rewriting @eq-doubly-constrained-accessibility as an origin summary expression of proportionally allocated opportunities:
+
+$$
+V^D_{i} = \kappa^D_{ij} \sum_j   D_j
+$${#eq-doubly-constrained-accessibility-with-proportional-allocation-factor}
+
+Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the the value _already_ matches population to opportunities, e.g., the number of opportunities accessed by the population.
+Similarly, the market potential form $M^D_j$ would be the transposed way of summarizing $V_{ij}^D$.
+In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to bother opportunities and population.
+As opportunities match population and both marginals are constrained, $M^D_j$ would simply be a summary of $V^D_{ij}$ values at each $j$, and would represent the number of people accessible to opportunities.
+The formulation is as follows:
+
+$$
+M^D_j = \sum_i A_i B_j O_i D_j f(c_{ij})\\
+$${#eq-doubly-constrained-market-potential}
 
@@
 id <- c("1", "2", "3")
-O_i2 <- c(4, 10, 6) 
-D_j2 <- c(160, 150, 180)
-
-# Balance the sum of origins and destinations
-f <- sum(O_i2 * 10000)/sum(D_j2)
-D_j2 <- (f * D_j2/10000) |>
-  round()
-
+O_i2 <- c(4, 10, 6) #in units of 10,000 people
+#D_j <- c(160, 150, 180) #the old
+#D_j2 <- c(6.530612, 6.122449, 7.346939) #this would be exactly linearly scaled down.
+D_j2 <- c(7, 5, 8) #in units of 10,000 physician-capacity
 LU2 <- data.frame(id, O_i2, D_j2)
@@
 
-NOTE: The example needs to be revised in this case to have the population and opportunities match, as in, one person-one opportunity
-
-```{r}
-#| label: tbl-small-system-land-use-doubly-constrained
-#| tbl-cap: "Simple system with three zones. Population is in 10,000 persons and opportunities in 10,000s physicians."
+Let's return to the numeric example with a new objective.
+Say, we're interested in measuring the zonal values of accessed physician-capacity.
+And we have new information: given the same zonal populations, a region-wide survey indicates the number of people that live in each zone, and the number of physician capacity per zone.
+The survey does not tell use what person is matched with what physician (assume that's confidential), we only have the numbers of population at $i$ and the number of physician-capacity at $j$.
+Lucky for us, this survey keeps our example simple: we find out that the physician-capacity at each zone is a approximately scaled version of the number of destination-side physicians at each zone.
+With this new information, we are interested in understanding what does the zonal access to physician-capacity for the population looks under different travel behaviour scenarios.
+@tbl-small-system-land-use-doubly-constrained-case summarises this adjusted simple example: with the population (in units of 10,000s of people) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+Considering this new definition of 'provider', the new (realized) Provider-to-Population-Ratio is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`.
+The zonal cost matrix, and travel impedance functions for three types of travel behaviour are the same as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
+
+```{r}
+#| label: tbl-small-system-land-use-doubly-constrained-case
+#| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."
 
@@
   gt() |>
-  cols_label(id = "ID",
+  # tab_spanner("{{W[_i^(1)]}}",
+  #             columns = 2) |>
+  # tab_spanner("{{W[_j^(2)]}}",
+  #             columns = 3) |>
+  cols_label(id = "ID (i or j)",
              O_i2 = "Population", 
@@
              columns = 2:3) |>
-  tab_footnote(footnote = md("Population is *Wi^(1)^* when used as a proxy for the mass at the origin, and *Oi* when used as a constraint."),
-               locations = cells_column_labels(columns = O_i2)) |>
-  tab_footnote(footnote = md("Opportunities is *Wj^(2)^* when used as a proxy for the mass at the destination, and *Dj* when used as a constraint."),
-               locations = cells_column_labels(columns = D_j2)) #|>
-  #as_latex()
+  as_latex()
 ```
@@
 
-OD2 <- C |>
+OD2 <- C |># mutate(f3=100) |>
   left_join(LU2 |>
@@
 
-In this case, $V_{ij}$ would be:
-```{r}
-
+In this case, $V_{ij}^D$ for the most decay travel behaviour scenario is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
+```{r, quiet = TRUE}
 doubly_constrained_gravity_model <- function(
@@
   sum_D <- sum(D)
-  cat("Checking production, attraction balancing:\n")
-  cat("Production: ", sum_O, "\n")
-  cat("Attraction: ", sum_D, "\n")
+  # cat("Checking production, attraction balancing:\n") #let's disable the messages..
+  # cat("Production: ", sum_O, "\n")
+  # cat("Attraction: ", sum_D, "\n")
 
   if (sum_O != sum_D) {
-    cat("Productions and attractions do not balance, scaling to match.\n")
+    # cat("Productions and attractions do not balance, scaling to match.\n") #let's disable the messages..
     if (sum_O < sum_D) {
@@
   } else {
-    cat("Production, attraction balancing OK.\n")
+    # cat("Production, attraction balancing OK.\n") #let's disable the messages..
   }
@@
     # Calculate Tij
-    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix #Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
-
-    # Print Ai and Bj for this iteration
-    cat("\nIteration:", iteration_count, "\n")
-    cat("Ai:", round(Ai, 3), "\n")
-    cat("Bj:", round(Bj_new, 3), "\n")
+    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
+
+    # # Print Ai and Bj for this iteration #let's disable the messages..
+    # cat("\nIteration:", iteration_count, "\n")
+    # cat("Ai:", round(Ai, 3), "\n")
+    # cat("Bj:", round(Bj_new, 3), "\n")
 
@@
 
-  # report stopping condition and final error threshold
-  cat("\nStopping Condition: ", stop_reason, "\n")
-  cat(sprintf("Final Error: %.3f%%\n", error * 100))
+  # report stopping condition and final error threshold #let's disable the messages..
+  #cat("\nStopping Condition: ", stop_reason, "\n")
+  #cat(sprintf("Final Error: %.3f%%\n", error * 100))
 
@@
 ```
-
-```{r}
+```{r}
+#| label: calculating balancing-factors and resulting Tij for each travel behaviour scenario
+
+#isolating impedance
 Impf1 <- C$f1 |> 
@@
   matrix(nrow = 3, ncol = 3)
-```
-
-```{r}
+# caclulating it
 dc_acc_1 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf1)
-```
-
-```{r}
 dc_acc_2 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf2)
-```
-
-```{r}
 dc_acc_3 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf3)
-```
-
-The $A_i$ and $B_j$ matrices:
-```{r}
-
-A_i_doubly <- dc_acc_1$Ai
-B_j_doubly <- dc_acc_2$Bj
-
-A_i_doubly |> matrix()
-B_j_doubly |> matrix() |> t()
-```
-
-The trip matrix (notice, all summed up opportunities and populations are equivalent to our knowns)
-```{r}
-#| eval: false
-
+
+#assigning the balancing factors to their own variables
+A_i_doubly1 <- dc_acc_1$Ai
+B_j_doubly1 <- dc_acc_1$Bj
+
+A_i_doubly2 <- dc_acc_2$Ai
+B_j_doubly2 <- dc_acc_2$Bj
+
+A_i_doubly3 <- dc_acc_3$Ai
+B_j_doubly3 <- dc_acc_3$Bj
+
+#assigning the Tijs to their own variables
 V_dc_1 <- dc_acc_1$Tij
-
-P_i_doubly <- V_dc_1 |> rowSums()
-O_j_doubly <- V_dc_1 |> colSums()
-V_TOTAL_dest <- ifelse(sum(P_i_doubly)==sum(O_j_doubly), sum(V_doubly), NA)
-V_doubly_longer <- cbind(V_doubly,P_i_doubly)
-V_doubly_longer <- rbind(V_doubly_longer, cbind(t(O_j_doubly),V_TOTAL_dest))
-
-rownames(V_doubly_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_doubly_longer) <- c("1","2","3", "TOTAL")
-
-V_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
-Nice. But notice how the accessibility is simply the number of opportunities at the destination! In other words, it is simply the access.
+V_dc_2 <- dc_acc_2$Tij
+V_dc_3 <- dc_acc_3$Tij
+
+#reformatting the matrix into long format, joining it to OD2
+OD2<-OD2 |> mutate(V_dc_ij_1 = c(V_dc_1[,1], V_dc_1[,2], V_dc_1[,3]),
+              V_dc_ij_2 = c(V_dc_2[,1], V_dc_2[,2], V_dc_2[,3]),
+              V_dc_ij_3 = c(V_dc_3[,1], V_dc_3[,2], V_dc_3[,3]))
+
+OD2$k_dc_ij_1 = c((outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,1],
+                   (outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,2],
+                   (outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,3]) / 3
+
+OD2$k_dc_ij_2 = c((outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,1],
+                   (outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,2],
+                   (outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,3]) / 3
+
+OD2$k_dc_ij_3 = c((outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,1],
+                   (outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,2],
+                   (outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,3]) / 3
+
+#Checks, should equal 1.
+# OD2$k_dc_ij_1 |> sum()
+# OD2$k_dc_ij_2 |> sum()
+# OD2$k_dc_ij_3 |> sum()
+```
+
+```{r}
+#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
+#| tbl-cap: "Adjusted simple system: doubly-constrained opportunity accessibility for the most decay travel behaviour scenario."
+
+OD2 |> 
+  select(c(oid, did, V_dc_ij_1)) |>
+  tidyr::pivot_wider(names_from = did, values_from = V_dc_ij_1) |>
+  gt() |>
+  tab_spanner(label = "Destination ID",
+              columns = 2:4) |>
+  cols_label(oid = "Origin ID")  |>
+  cols_align(align = "center",
+             columns = 2:4) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
+```
+
+As mentioned, accessibility can be interpreted as a summary of the proportionally allocated opportunities at each $i$.
+In interpreting the doubly-constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$, these values would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity for Zones 1, 2 and 3 respectively, approximately equal to the number of population at each of these zones.
+Conversely, $M_j^D$ values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of _opportunities_  (physician-capacities) at each of these zones.
+And that's precisely the function of the double constraint - to have the mass weight at the origin equal the mass weight of the destination.
+In other words, $V_i^D$ is simply accessed opportunities and $M_j^D$ population accessed.
+And following the same logic, the same $V_i^D$ and $M_j^D$ values are calculated for the other two travel behaviour scenarios.
+
+What may be interesting are the differences in $V_{ij}^D$ values  between travel behaviour scenarios, these values can be directly compared to discuss their mass and distance decay impacts.
+Returning to Zone 2,  @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for our more relatively remote, higher-populated and lower-opportunity rich zone.
+It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the ~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
+Following the intuition discussed in the singly-constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
+
+<!--NOTE FOR FUTURE EXPLORATION: Interestingly...
+for the doubly- and singly- constrained case, kappas are very similar *maybe close to identical?* when decay is low.
+This is because distance difference doesn't matter, everyone can reach everything and everything is given to everyone.
+But when decay is high, they are much different: since in the doubly-constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
+This relationship should be investigated further some other time..-->
+
+```{r}
+#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
+#| tbl-cap: "Adjusted simple system: doubly-constrained opportunity accessibility for all travel behaviour scenarios for Zone 2."
+
+OD2  |>
+  filter(oid == 2) |>
+  select(c(did, O_i2, D_j2, starts_with("V_"))) |>
+  gt(rowname_col = "did") |>
+  tab_stubhead(label = "Dest.") |>
+
+  tab_spanner(label = "{{V[_{ij}^D]}}",
+              columns = 4:6,
+              level = 2) |>
+  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
+              columns = 4,
+              level = 1) |>
+  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
+              columns = 5,
+              level = 1) |>
+  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
+              columns = 6,
+              level = 1) |>
+  fmt_number(decimals = 3) |>
+
+  cols_label(O_i2 = md("Population at 2 (units: *people in 10,000s*)"),
+             D_j2 = md("Opportunities (units: *capacity in 10,000s*)"),
+             V_dc_ij_1 = md("units: *physician-capacity in 10,000s*"),
+             V_dc_ij_2 = md("units: *physician-capacity in 10,000s*"),
+             V_dc_ij_3 = md("units: *physician-capacity in 10,000s*"),
+             )  |>
+  cols_align(align = "center",
+             columns = 2:6) |>
+  cols_width(
+    stub() ~ px(60*0.5),
+    everything() ~ px(180*0.5)
+  ) |>
+  tab_options(column_labels.font.size = "small") |>
+  as_latex()
+```
+
+Recall that accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
+In the doubly-constrained case,  we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
+However, if what interests readers is the "potential for interaction" in the case population and opportunities do match one-to-one, perhaps reframing the investigation is needed.
+In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ij}^D$).
+In this sense, the doubly-constrained case it better thought of as an estimate of "access" instead of "accessibility".
+It is also formulaically identical to the doubly-constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity' respectively.
+
+As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model _can_ be defined using any unit.
+Accessibility, however, is a summary of _potential_ for interaction: it's typically been used and understood as a zonal summary.
+Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained case make the connection between the _potential_ for spatial interaction (accessibility) and spatial interaction (access) more clear(?).
+
+<!--
+NOTE: competitive accessibility Literature has explored 'doubly-constrained' but -- it's not the same as this: they they just use the Wilson balancing factors , etc.
+Ex.
+@hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
+-->
+
+<!--ANOTHER NOTE:  -->
 
 # Discussion
-when to use what, their embedded assumptions, and dangers of using one over the other
-
-Wu and Lev say comp. can be considered in the imped function, but no details. we argue its' more. unconstrained - no comp,
--- constraints consider competition
-	- total - regionally; not fixing it on what zones will be matched, it's just regional,
-	- singly- with zone and regional,
-	- doubly- per origin and destination zone.
-
-A sort of 'accessible' provider to population ratio, or accessible opportunities per capita, can also be easily calculated by dividing each constrained value by the number of opportunities. !!!
-
-# A note on interpretation and possible applications
-
-When to use each of these measures?
-<!--AS: CHECK MY OLD VERSION!  -->
-
-**Totally-constrained**
-
-**Singly-constrained**
-
-<!-- Each $V_{ij}$ can be interpreted as $O_j$ weighted by the proportional balancing factor e.g., "the number of opportunities accessible from $j$ by the population at $i$" all weighted by travel cost, of course.
-Use this formulation of the singly-constrained case when the number of opportunities, at each $j$, that are being competed for by the population is known but the precise number of population that is competiting for opportunities is not.
-For example, public schools as a destination in a region where all children attend them.
-All school-seats at $j$ are open to all children at all $i$s.
-However, not all population $i$ are equally interested in school-seats at all $j$s (e.g., some prefer special programs, teachers, etc.); and we do not know this precise population $i$ to $j$ number.
-Hence, an opportunity-constrained version of the singly-constrained case is appropriate.
+
+## SUMMARY
+
+By introducing constraints, we fold in competition considerations into a family of accessibility measures.
+While previous methodologically clarifying work, such as @wuUnifyingAccess2020, has largely overlooked competition, this paper demonstrates how applying constraints consistent with the spatial interaction model allows for incorporating competition into accessibility.
+The constraints used depends on the research question, the suitability of assumptions, the granularity of data, but ultimately: the addition of constraints re-introduces unit into accessibility, make comparisons between scenarios and regions more intuitive.
+
+This work builds on spatial interaction modeling principles, following a review of earlier spatial interaction literature.
+We show that, while recent accessibility research has diverged from its spatial interaction foundations, it can be explicitly linked back.
+We demonstrate that accessibility can be consistently calculated for an $i$-$j$ pair by stacking constraints that preserve the units of "accessibility" while maintaining zonal or regional constraints.
+
+We first place the popular Hansen accessibility measure within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
+We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
+In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if regional competition is relevant; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, _without_ considering the population size at the origins.
+<!-- A.S.
+NOTE: Hansen's measure would converge to some global total if the decay effect was low.
+Imagine everyone can teleport.
+In the unconstrained case: all zones can reach all opportunities.
+In the totally constrained case: all zones can reach all opportunities divided by the number of zones.
+Alternatively, if 1 zone can teleport and no other zone can travel at all, that 1 zone will get all the opportunities in the region.
 -->
 
-<!-- However, the population-constrained version of the singly-constrained case can be appropriate in the converse situation.
-Consider the case of public urgent care clinics in a region where all people use them equally.
-Unlike schools, all population at $i$ are equally interested in urgent-care-seats at all clinics $j$.
-If there's an emergency, or not, they will visit.
-However, from an operational perspective all clinics at $j$ are _not_ interested in serving all people (e.g., they'd rather serve people who have access to primary care medicine as they are more likely to seek urgent care in actual medical emergencies).
-From this perspective, using a _population-constrained_ version of the singly-constrained case $V_i''^{(1)}$, is appropriate.
-Each $V_{ij}$ can be interpreted as $P_i$ weighted by the proportional balancing factor e.g., "the number of population accessible from $i$ by the opportunities at $j$" all weighted by travel cost, of course.
--->
-
-**Doubly-constrained**
+We then introduce the singly-constrained case, which _does_ takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
+In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
+The singly-constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
+We also introduce an 'accessible' PPR (e.g., opportunities per capita), calculated by dividing each accessibility value by the zonal population.
+To clarify, this per capita expression of the singly-constrained case is equivalent to the 2SFCA, hence linking this literature back to spatial interaction principles.
+
+Lastly, the doubly-constrained case is introduced.
+In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available.
+Specifically, accessibility values for each $i$-$j$ pair must be a proportion of he zonal opportunity and population values simultaneously.
+For example, the accessibility at zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at zone 1.
+Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest.
+In this sense, the research question should be concerned with realized access (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential 'access' (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
+
+## A note on interpretation and possible applications
+
+When to use each of these measures? Embedded assumptions? and dangers of using one over the other?
 
@@
 <!--
+BELOW IS THE TEXT GRAVEYARD!
+
 # APPENDIX (some musings... destined for the commented-out graveyard)

### 2025-03-07 paper/manuscript.qmd
edits
general typographic edits by cdh

@@
 title: "A family of accessibility measures derived from spatial interaction principles"
-format: pdf
+format:
+  pdf:
+    header-includes:
+        - \usepackage{float}
+        - \floatplacement{table}{H}
+        - \floatplacement{figure}{H}
 bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-includes:
-      in_header: 'gt_packages.sty'
+#includes:
+#      in_header: 'gt_packages.sty'
       #\usepackage{float}
@@
 ```
+
 ```{r options}
@@
 
-An ascendant interest in accessibility has been accompanied by a remarkable boom on scholarly research, which has grown to include employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
+An ascendant interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
 In other words, accessibility analysis is used today to broadly understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
-However, despite its growth in popularity in scholarly works, challenges remain to more widespread adoption of accessibility in planning practice.
+However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
 Several barriers to bridging the scholarly-practice accessibility gap have been identified.
@@
 
-Adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
-The menu of accessibility measures has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
+The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
+At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
+Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
 How is a practitioner to choose among this myriad options?
@@
 In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
-By considering the key features of accessibility, namely travel-cost $f(c{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson).
-One needs only to judiciously change the way travel-cost and opportunities are formulated, to derive almost any known accessibility measure.
+By considering the key features of accessibility, namely travel-cost $f(c_{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson).
+One needs only to judiciously change the way travel costs and opportunities are formulated to derive almost any known place-based accessibility measure.
 In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
@@
 S_i \propto \sum_j g(O_j)f(c_{ij})
-$${#eq-access-Lu-Levinson}
+$$ {#eq-access-Lu-Levinson}
 
@@
 It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility.
+Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility (REITERATE WHY).
 We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
@@
 
+```{=html}
 <!--
@@
 -->
+```
 
@@
 From as far back as Henry C.
-Carey's _Principles of Social Science_ [@careyPrinciplesSocialScience1858], a concern with the scientific study of human spatial interaction can be observed.
-It was in this work where Carey stated that "man [is] the molecule of society [and their interaction is subject to] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
+Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], a concern with the scientific study of human spatial interaction can be observed.
+It was in this work where Carey stated that "man \[is\] the molecule of society \[and their interaction is subject to\] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
 37-38].
-This statement shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation, first posited in 1687's _Principia Mathematica_ and expressed as in @eq-phys-grav-prop.
+This statement shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation, first posited in 1687's *Principia Mathematica* and expressed as in @eq-phys-grav-prop.
 
@@
 F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
-$${#eq-phys-grav-prop}
+$$ {#eq-phys-grav-prop}
 
 To be certain, the expression above, a proportionality, is one of the most famous in all of science.
-In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly _proportional_ to the product of their masses $M_i$ and $M_j$, and inversely _proportional_ to the square of the distance between them $D_{ij}$.
+In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$.
 Direct proportionality means that as the product of the masses increases, so does the force.
@@
 To do so, an empirical constant is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
-In other words, @eq-phys-grav-prop needs to be _constrained_ using empirical data.
+In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data.
 Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:
@@
 F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
-$${#eq-phys-grav}
+$$ {#eq-phys-grav}
 
@@
 783].
-Still, it took over a century from the publication of _Principia_ to refine the estimate of the proportionality constant to within 1\% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
+Still, it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant to within 1% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
 
@@
 
+```{=html}
 <!-- 
@@
 -->
-
-Following Carey's _Principles_ of 1858, research into human spatial interaction continued in different contexts.
+```
+
+Following Carey's *Principles* of 1858, research into human spatial interaction continued in different contexts.
 In the late 1880s, Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
@@
 B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
-$${#eq-reilly}
-
-\noindent Where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
-Quantity $N$ was chosen by Reilly in a somewhat _ad hoc_ fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
+$$ {#eq-reilly}
+
+\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
+Quantity $N$ was chosen by Reilly in a somewhat *ad hoc* fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
 
@@
 C^2 = \frac{P_1\cdot P_2}{D_{12}}
-$${#eq-zipf}
+$$ {#eq-zipf}
 
@@
 F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
-$${#eq-stewart-force}
+$$ {#eq-stewart-force}
 
@@
 
-- $F$ is the _demographic force_
+-   $F$ is the *demographic force*
 -   $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
-- $\mu_1$ and $\mu_2$ are so-called _molecular weights_
+-   $\mu_1$ and $\mu_2$ are so-called *molecular weights*
 -   $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
@@
 
-In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
-$$
+In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows: $$
 V_1 = G\frac{M_2}{d_{12}}
-$${#eq-stewart-population-potential}
-
-For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
-$$
+$$ {#eq-stewart-population-potential}
+
+For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$): $$
 V_i = \int\frac{D}{r} ds
-$${#eq-stewart-population-potential-integral}
+$$ {#eq-stewart-population-potential-integral}
 
 \noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
-In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
-$$
+In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as: $$
 V_i = \sum_j \frac{M_j}{d_{ij}}
-$${#eq-stewart-population-potential-sum}
-
-Alerted readers will notice that @eq-stewart-population-potential-sum is formally equivalent to our modern definition of accessibility.
+$$ {#eq-stewart-population-potential-sum}
+
+Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility:
+
+$$
+V_i = \sum_j M_jd_{ij}^{-1}
+$$ {#eq-stewart-population-potential-sum-rewrite}
+
+<!--I say we spell it out as in this eq^, what you think?-->
 
@@
 53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism.
-For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" [p.
-35].
+For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" \[p.
+35\].
 Stewart's ideas about "social physics" soon fell out of favour among social scientists, but not before influencing the nascent field of accessibility research, as detailed next.
@@
 In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
-a generalization of the population-over-distance relationship or _population potential_ concept developed by @stewartDemographicGravitationEvidence1948" (p.
+a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p.
 73).
 As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-@hansen1959 focused on @stewartDemographicGravitationEvidence1948's _population potential_ (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
+@hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
 @hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-@hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with _opportunities_ to derive an _opportunity potential_, or more accurately, a _potential of opportunities for interaction_ as follows:
+@hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:
 
@@
 S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
-$${#eq-accessibility}
+$$ {#eq-accessibility}
 
@@
 S_{i} = \sum_j O_j \cdot f(d_{ij})
-$${#eq-accessibility-general}
+$$ {#eq-accessibility-general}
 
@@
 Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space.
-At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".
-
-After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly determine it.
+There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about *directly* and *inversely proportional* relationships when discussing population, opportunities, and their separation in space.
+At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity".
+
+After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.
 In this way, $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
@@
 
+<!-- very brief narrative to introduce - is this happening separately from hansen? "On the other side of the atlantic, wilson was doing..." :) -->
+
 In his groundbreaking work, Wilson [-@wilson1971] defined a general spatial interaction model as follows:
@@
 T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
-$${#eq-phys-gravity-model}
+$$ {#eq-phys-gravity-model}
 
@@
 Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
-In this way, $T_{ij}$ explicitly measures _interaction_ in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
+In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
 In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled: it balances the units.
 
-Traditionally, development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967;@wilson1971].
+Traditionally, development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
 But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
@@
 \sum_i\sum_j T_{ij} = T
-$${#eq-constraint0-gravitymodel}
+$$ {#eq-constraint0-gravitymodel}
 
@@
 \sum_j T_{ij} = O_i
-$${#eq-constraint1-gravitymodel}
-
-Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used:
-$$
+$$ {#eq-constraint1-gravitymodel}
+
+Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used: $$
 \sum_i T_{ij} = D_j
-$${#eq-constraint2-gravitymodel}
+$$ {#eq-constraint2-gravitymodel}
 
@@
 K=\frac{T}{\sum_i\sum_j T_{ij}}
-$${#eq-total-flow-balancing-factor}
+$$ {#eq-total-flow-balancing-factor}
 
@@
 T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
-$${#eq-production-constrained-gravitymodel}
+$$ {#eq-production-constrained-gravitymodel}
 
@@
 A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$${#eq-production-constrained-balancing-factor}
+$$ {#eq-production-constrained-balancing-factor}
 
@@
 T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
-$${#eq-attraction-constrained-gravitymodel}
+$$ {#eq-attraction-constrained-gravitymodel}
 
@@
 B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-attraction-constrained-balancing-factor}
+$$ {#eq-attraction-constrained-balancing-factor}
 
@@
 T_{ij} = A_i B_j O_i D_j f(c_{ij})
-$${#eq-doubly-constrained-gravitymodel}
+$$ {#eq-doubly-constrained-gravitymodel}
 
@@
 \end{array}
-$${#eq-doubly-constrained-balancing-factors}
+$$ {#eq-doubly-constrained-balancing-factors}
 
@@
 
+````{=html}
 <!--
@@
 -->
+````
 
@@
 As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
-However, we can discern from a manual reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., _potential for_ spatial interaction) and another on spatial interaction.
+However, we can discern from a manual reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
 The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected.
@@
 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the _competition destination_ model).
-Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research d[@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
+Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
+Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
 Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
 
-On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons:
-- Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 - Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-While accessibility is an expression of _potential for_ spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
+While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
 This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
 However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues.
-Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
-
+Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways for maintaining constraints in opportunities in the context of competitive accessibility.
+
+```{=html}
 <!-- 
@@
 -->
-
+```
+
+```{=html}
 <!--
@@
 S_i = \frac{1}{A_i} = \sum_j W_j^{(2)} f(c_{ij})
-$${#eq-Ai-as-accessibility}
+$$ {#eq-Ai-as-accessibility}
 
@@
 
-Hence, we propose a more useful definition: accessibility as a partial sum of the spatial interaction. A way to define the _potential for spatial interaction_ (i.e., accessibility) is as follows:
+Hence, we propose a more useful definition: accessibility as a partial sum of the spatial interaction. A way to define the *potential for spatial interaction* (i.e., accessibility) is as follows:
 
@@
 V_{ij} = k W_j^{(2)} f(c_{ij})
-$${#eq-access-01}
+$$ {#eq-access-01}
 
@@
 V_{i} = k \sum_j W_j^{(2)} f(c_{ij})
-$${#eq-accesssibility-01}
+$$ {#eq-accesssibility-01}
 
@@
 Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone.
-^[For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]].
+[^1].
 Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
@@
 Similarly, the opportunities are represented by $W_j^{(2)}$ when used as a proxy for the mass at the destination, and $D_j$ when used as a constraint.
+
+[^1]: For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]
 
@@
 
-And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios:
-$$
+And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios: $$
 \begin{array}{l}
@@
 \end{array}
-$${#eq-travel-behaviour-scenarios}
+$$ {#eq-travel-behaviour-scenarios}
 
@@
 V^0_{ij} = W_j^{(2)} f(c_{ij})
-$${#eq-unconstrained-access}
+$$ {#eq-unconstrained-access}
 
@@
 V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
-$${#eq-unconstrained-accessibility}
+$$ {#eq-unconstrained-accessibility}
 
@@
   ) |>
+  grand_summary_rows(
+    columns = c(V_unc_i_1, V_unc_i_2, V_unc_i_3),
+    fns = list(label = "Sum", fn = "sum")) |> 
   tab_options(column_labels.font.size = "small") |>
@@
 
-As the different impedance functions represent different travel behaviours, comparing unconstrained accessibility values across groups meaningless.
-For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1]`, but in what units?
-These two values, though representing the same zone but different groups in that zone, have different units so this difference value cannot be meaningfully interpreted.
+As the different impedance functions represent different travel behaviours, comparing unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
+For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
+These two values have different units so this difference value cannot be meaningfully interpreted.
+
+<!-- not like terms? -->
 
@@
 Both scenarios at zone 1 captures an intermediate $V_i^0$ value between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians.
-The ratios between the zones in the two scenarios different.
+The ratios between the zones in the two scenarios are different.
 These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared.
@@
 
+<!-- CDH: I am not sold on the problem here, yes the values are tied to the impedance, one could use a calibrated impedance...
+but still meaningful comparisons on ordinal scale despite critiques of ratios - i would not expect ratios to be the same...
+need to better draw out in comparison to below?
+-->
+
 ## Totally-constrained accessibility
@@
 The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case.
-We define two cases: a _totally-constrained opportunity_ accessibility case, interpreted as Hansen's accessibility with a constraint, and a _totally-constrained population_ accessibility case, interpreted as constrained market potential.
+We define two cases: a *totally-constrained opportunity* accessibility case, interpreted as Hansen's accessibility with a constraint, and a *totally-constrained population* accessibility case, interpreted as constrained market potential.
 
@@
 V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
-$${#eq-total-constrained-access}
+$$ {#eq-total-constrained-access}
 
@@
 V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
-$${#eq-total-constrained-accessibility}
+$$ {#eq-total-constrained-accessibility}
 
@@
 \sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
-$${#eq-total-constraint-accessibility}
+$$ {#eq-total-constraint-accessibility}
 
 This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework.
-We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $K^T$ to yield:
-$$
+We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $K^T$ to yield: $$
 K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$${#eq-total-opportunity-balancing-factor}
+$$ {#eq-total-opportunity-balancing-factor}
 
@@
 
-Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be:
-$$
+<!-- kappa in equations now? -->
+
+Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be: $$
 \begin{array}{l}
@@
   tab_options(column_labels.font.size = "small") |>
+  grand_summary_rows(
+    columns = c(V_tot_i_1, V_tot_i_2, V_tot_i_3),
+    fns = list(label = "Sum", fn = "sum")) |> 
   as_latex()
@@
 
-Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974].
+<!-- need to improve flow here, why are we jumping into market potential, a bit of duplication in the effectively transpose part -->
+
+Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
 The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
-To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.
+To formulate this case, the total constraint may instead be applied to the mass of the *population* at $i$ instead of the opportunities at $j$.
 
@@
 M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
-$${#eq-total-constrained-market}
+$$ {#eq-total-constrained-market}
 
@@
 M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
-$${#eq-total-constrained-market}
+$$
 
@@
 \sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
-$${#eq-total-constraint-market}
+$$ {#eq-total-constraint-market}
 
@@
 \hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
-$${#eq-total-population-balancing-factor}
+$$ {#eq-total-population-balancing-factor}
 
@@
 M^T_i = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$
-Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
-$$
+$$ Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless: $$
 \hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
@@
   ) |>
+  grand_summary_rows(
+    columns = c(M_tot_i_1, M_tot_i_2, M_tot_i_3),
+    fns = list(label = "Sum", fn = "sum")) |> 
   tab_options(column_labels.font.size = "small") |>
@@
 Increasing in restrictiveness relative to the totally-constrained accessibility measure, a single constraint proportionally adjusts the origin-destination zonal accessibility values based on the population or opportunities in the zone for all zones, while still maintaining the total system population or opportunity constraint, depending on the case.
-We define two cases: a _singly-constrained opportunity_ accessibility case and a _singly-constrained population_ accessibility case.
+We define two cases: a *singly-constrained opportunity* accessibility case and a *singly-constrained population* accessibility case.
 
@@
 
-If additional information is introduced into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (@eq-opportunity-constraint)- comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
+If additional information is introduced into our analysis, say, the number of opportunities *by destination* (i.e., $D_j$), we can impose the following constraint (@eq-opportunity-constraint)- comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
 
@@
 \sum_i V^S_{ij} =  D_j
-$${#eq-opportunity-constraint}
+$$ {#eq-opportunity-constraint}
 
@@
 V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
-$${#eq-opportunity-constrained-accessibility}
+$$ {#eq-opportunity-constrained-accessibility}
 
@@
 B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-proportionality-constants}
+$$ {#eq-opportunity-constrained-proportionality-constants}
 
@@
 V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-accessibility-with-balancing-factor}
+$$ {#eq-opportunity-constrained-accessibility-with-balancing-factor}
 
@@
 \kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$${#eq-opportunity-constrained-proportional-allocation-factor}
+$$ {#eq-opportunity-constrained-proportional-allocation-factor}
 
@@
 V^S_{i} = \kappa^S_{i} \sum_j   D_j
-$${#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
+$$ {#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
 
 @soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As in the totally=constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
-The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached _and_ are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
+As in the totally-constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
+The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
 These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
@@
 
-Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible _and_ available to location $i$ out of the total number of opportunities in the system.
+Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible *and* available to location $i$ out of the total number of opportunities in the system.
 
@@
 $$
+
 ```{r}
@@
   ) |>
+  grand_summary_rows(
+    columns = c(V_opp_i_1, V_opp_i_2, V_opp_i_3),
+    fns = list(label = "Sum", fn = "sum")) |> 
   tab_options(column_labels.font.size = "small") |>
@@
 
-From a different perspective: $V_{2}^S$ is intermediate in rank when considering the highest decay scenario ($f_1(c_{ij})$) but is the highest ranked zone under $f_3(c_{ij})$, from `r opp_acc$V_opp_i_1[2]` to `r opp_acc$V_opp_i_3[2]` accessible physicians out of `r opp_acc$V_opp_i_3|>sum() ` in the region; a `r opp_acc$V_opp_i_3[2]/opp_acc$V_opp_i_1[2]` times increase.
+<!-- check sentence ^ -->
+
+From a different perspective: $V_{2}^S$ is intermediate in rank when considering the highest decay scenario ($f_1(c_{ij})$) but is the highest ranked zone under $f_3(c_{ij})$, from `r opp_acc$V_opp_i_1[2]` to `r opp_acc$V_opp_i_3[2]` accessible physicians out of `r opp_acc$V_opp_i_3|>sum()` in the region; a `r opp_acc$V_opp_i_3[2]/opp_acc$V_opp_i_1[2]` times increase.
 Zone 2 has the highest amount of population and the least travel-impedance proximate to other zones.
@@
 Again, returning to Zone 2 as an example, while its competitive population advantage is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios: this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
-In this sense, other zones are not _that_ disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
+In this sense, other zones are not *that* disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
 Zones 1 and 3 are allocated opportunities at relates similar to their relative population size.
 
-In this way, the consideration of accessibility _per capita_ may be clarifying.
+In this way, the consideration of accessibility *per capita* may be clarifying.
 Often, accessibility values are reported as raw scores without the consideration for population.
 But, as we introduced constraints, these constrained accessibility values can be normalized using anything that is relevant to the zone.
-In  @tbl-simple-example-attraction-constrained-accessibility-per-capita), we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
+In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
 These per capita rates are equivalent to the 2SFCA values.
@@
   ) |>
+  #grand_summary_rows(
+  #  columns = c(v_opp_i_1, v_opp_i_2, v_opp_i_3),
+  #  fns = list(label = "Sum", fn = "sum")) |> 
   tab_options(column_labels.font.size = "small")  |>
@@
 
-Recall: our simple example was designed such that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
-And as the decay decreases and the effect of population is more independent: per capita values begin to stabilise to this regional per capita average, similar to how the accessibility values in the totally-constrained opportunity case stables to the $V_i^T$ regional average.
+<!-- check wording in paragraph below, few issues -->
+
+Our simple example was designed such that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
+As the decay decreases and the effect of population is more independent: per capita values begin to stabilise to this regional per capita average, similar to how the accessibility values in the totally-constrained opportunity case stabilizes to the $V_i^T$ regional average.
 These responses make intuitive sense: the use of constraints are an regional and/or zonal averaging methodology.
@@
 
-Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a _singly-constrained_ measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^(1)$ is now represented by $O_i$).
+Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly-constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^(1)$ is now represented by $O_i$).
 In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
@@
 M^S_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
-$${#eq-population-constrained-accessibility}
-
-In this case, the measure is singly-constrained by the population _by origin_ (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+$$ {#eq-population-constrained-accessibility}
+
+In this case, the measure is singly-constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
 
@@
 \sum_j M^S_{ij} =  O_i \\
-$${#eq-population-constraint}
-
-And the corresponding balancing factor, as per Wilson, is:
-$$
+$$ {#eq-population-constraint}
+
+And the corresponding balancing factor, as per Wilson, is: $$
 A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$${#eq-population-constrained-proportionality-constants}
-
-Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression:
-$$
+$$ {#eq-population-constrained-proportionality-constants}
+
+Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression: $$
 M^S_{j} = \sum_i \hat \kappa^S_{ij} O_i
-$${#eq-production-constrained-accessibility-with-proportional-allocation-factor}
+$$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}
 
@@
 \kappa^S _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
-$${#eq-attraction-constrained-proportional-allocation-factor}
+$$ {#eq-attraction-constrained-proportional-allocation-factor}
 
@@
 
-Readers will have noticed that the accessibility measures above have used either $O_i$, $D_j$, or the regional sums of either, but not both.
+The accessibility measures above have used either $O_i$, $D_j$, or the regional sums of either, but not both.
 When $D_j$, the opportunities, was used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
-Contrariwise, When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
+When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
 The reason is that the population and the opportunities are typically different things without a common metric.
 For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
-There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most commonly studied opportunity in the competitive accessibility literature being employment, e.g., one person, one job.
-Another possible opportunity type is healthcare, e.g., one person, one physician-capacity.
-
-So: a doubly-constrained approach can be implemented when such one-to-one relationship between population and opportunities is present.
-This is required because, the production- and attraction- constraints in Wilson's doubly-constrained case (@eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel) are imposed simultaneously.
+There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most commonly studied opportunity in the competitive accessibility literature being employment, e.g., one person with one job.
+Another possible opportunity type is healthcare, e.g., one person and one unit of capacity.
+
+In these latter cases, a doubly-constrained approach can be implemented when one-to-one relationships between population and opportunities are present.
+This is required because the production- and attraction- constraints in Wilson's doubly-constrained case (@eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel) are imposed simultaneously.
 The accessibility-specific forms are used in the singly-constrained opportunities and population cases in the preceding section (@eq-opportunity-constraint and @eq-population-constraint).
@@
 \sum_i O_i = \sum_j D_j
-$${#eq-opportunity-population-equality}
+$$ {#eq-opportunity-population-equality}
 
@@
 V_{i}^D = \sum_j A_i B_j O_i D_j f(c_{ij})
-$${#eq-doubly-constrained-accessibility}
+$$ {#eq-doubly-constrained-accessibility}
 
@@
 \end{array}
-$${#eq-doubly-constrained-balancing-factors}
+$$
 
@@
 
-The proportional allocation factor $\kappa_{ij}^D$ would then be:
-$$
+The proportional allocation factor $\kappa_{ij}^D$ would then be: $$
 \kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
@@
 V^D_{i} = \kappa^D_{ij} \sum_j   D_j
-$${#eq-doubly-constrained-accessibility-with-proportional-allocation-factor}
-
-Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the the value _already_ matches population to opportunities, e.g., the number of opportunities accessed by the population.
+$$ {#eq-doubly-constrained-accessibility-w-paf}
+
+Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the the value *already* matches population to opportunities, e.g., the number of opportunities accessed by the population.
 Similarly, the market potential form $M^D_j$ would be the transposed way of summarizing $V_{ij}^D$.
-In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to bother opportunities and population.
+In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both <!-- is it "both" or "other" - word was "bother" --> opportunities and population.
 As opportunities match population and both marginals are constrained, $M^D_j$ would simply be a summary of $V^D_{ij}$ values at each $j$, and would represent the number of people accessible to opportunities.
@@
 M^D_j = \sum_i A_i B_j O_i D_j f(c_{ij})\\
-$${#eq-doubly-constrained-market-potential}
+$$ {#eq-doubly-constrained-market-potential}
 
@@
 With this new information, we are interested in understanding what does the zonal access to physician-capacity for the population looks under different travel behaviour scenarios.
-@tbl-small-system-land-use-doubly-constrained-case summarises this adjusted simple example: with the population (in units of 10,000s of people) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+@tbl-small-system-land-use-doubly-constrained-case summarises this adjusted simple example: with the population (in units of 10,000s of people) and the opportunities (in units of 10,00s of physician-capacity) per zone <!-- check -->.
 Considering this new definition of 'provider', the new (realized) Provider-to-Population-Ratio is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`.
@@
 In this case, $V_{ij}^D$ for the most decay travel behaviour scenario is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
+
 ```{r, quiet = TRUE}
@@
 ```
+
 ```{r}
@@
   tidyr::pivot_wider(names_from = did, values_from = V_dc_ij_1) |>
+  group_by(oid) |> 
+  mutate(sum = round(`1` + `2` + `3`, 0)) |> 
+  ungroup() |> 
   gt() |>
@@
   cols_align(align = "center",
-             columns = 2:4) |>
+             columns = 2:5) |>
+  grand_summary_rows(
+    columns = 2:4,
+    fns = list(label = "Sum", fn = "sum")) |>
+
   tab_options(column_labels.font.size = "small") |>
@@
 In interpreting the doubly-constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$, these values would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity for Zones 1, 2 and 3 respectively, approximately equal to the number of population at each of these zones.
-Conversely, $M_j^D$ values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of _opportunities_  (physician-capacities) at each of these zones.
+Conversely, $M_j^D$ values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of *opportunities* (physician-capacities) at each of these zones.
 And that's precisely the function of the double constraint - to have the mass weight at the origin equal the mass weight of the destination.
@@
 Returning to Zone 2, @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for our more relatively remote, higher-populated and lower-opportunity rich zone.
-It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the ~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
+It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
 Following the intuition discussed in the singly-constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
@@
 
-As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model _can_ be defined using any unit.
-Accessibility, however, is a summary of _potential_ for interaction: it's typically been used and understood as a zonal summary.
-Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained case make the connection between the _potential_ for spatial interaction (accessibility) and spatial interaction (access) more clear(?).
-
+As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
+Accessibility, however, is a summary of *potential* for interaction: it's typically been used and understood as a zonal summary.
+Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained case make the connection between the *potential* for spatial interaction (accessibility) and spatial interaction (access) more clear(?).
+
+```{=html}
 <!--
@@
 We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
-In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if regional competition is relevant; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, _without_ considering the population size at the origins.
+In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if regional competition is relevant; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, *without* considering the population size at the origins.
 <!-- A.S.
@@
 
-We then introduce the singly-constrained case, which _does_ takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
+We then introduce the singly-constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
@@
 -->
-
+```
+
+```{=html}
 <!-- 
@@
 
+```{=html}
 <!--TO DISCUSS:
@@
   -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->
+```
 
 <!--THE FOLLOWING ARE SOLVED EXAMPLES OF ALL FOUR CASES OF THE SIM -->
+
 <!-- The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.
 For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below: -->
+
 <!-- ```{r} -->
+
 <!-- known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE) -->
+
 <!-- #unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE) -->
+
 <!-- W_i <- known_trip_matrix |> rowSums() -->
+
 <!-- W_j <- known_trip_matrix |> colSums() -->
+
 <!-- T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA) -->
+
 <!-- knowns_longer <- cbind(known_trip_matrix,W_i) -->
+
 <!-- knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T)) -->
@@
 <!-- rownames(knowns_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(knowns_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows: -->
+
 <!-- ```{r} -->
+
 <!-- C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE) -->
+
 <!-- C_longer <-C -->
+
 <!-- rownames(C_longer) <- c("1","2","3")  -->
+
 <!-- colnames(C_longer) <- c("1","2","3")  -->
+
 <!-- C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- $$ -->
+
 <!-- T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+
 <!-- $$ -->
+
 <!-- ```{r} -->
+
 <!-- T_unc <- W_i %*% t(W_j) / C -->
@@
 <!-- W_i_unc <- T_unc |> rowSums() -->
+
 <!-- W_j_unc <- T_unc |> colSums() -->
+
 <!-- T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA) -->
+
 <!-- T_unc_longer <- cbind(T_unc,W_i_unc) -->
+
 <!-- T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc)) -->
@@
 <!-- rownames(T_unc_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(T_unc_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- T_unc_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
+
 <!-- $$ -->
+
 <!-- T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+
 <!-- $$ -->
+
 <!-- \where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made: -->
+
 <!-- $$ -->
+
 <!-- K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\ -->
+
 <!-- K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +...
 \frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}} -->
+
 <!-- }\\ -->
+
 <!-- K = \frac{790}{166443} \\ -->
+
 <!-- K = 0.00474636 -->
+
 <!-- $$ -->
@@
 <!-- ```{r} -->
+
 <!-- K <- T/sum(sum(T_unc)) -->
+
 <!-- T_tot <- K * T_unc -->
@@
 <!-- W_i_tot <- T_tot |> rowSums() -->
+
 <!-- W_j_tot <- T_tot |> colSums() -->
+
 <!-- T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA) -->
+
 <!-- T_tot_longer <- cbind(T_tot,W_i_tot) -->
+
 <!-- T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot)) -->
@@
 <!-- rownames(T_tot_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(T_tot_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- T_tot_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
+
 <!-- While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model. -->
@@
 <!-- For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
+
 <!-- $$ -->
+
 <!-- T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+
 <!-- $$ -->
+
 <!-- \where In our example, $A_i$ is: -->
+
 <!-- $$ -->
+
 <!-- A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\ -->
+
 <!-- A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\ -->
+
 <!-- ...\\ -->
+
 <!-- A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\ -->
+
 <!-- A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\ -->
+
 <!-- A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201  -->
+
 <!-- $$ -->
@@
 <!-- ```{r} -->
+
 <!-- A_i <- 1/t((matrix(W_j))) %*% (1/C) -->
+
 <!-- #A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns -->
+
 <!-- A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3) -->
@@
 <!-- W_i_orig <- T_orig |> rowSums() -->
+
 <!-- W_j_orig <- T_orig |> colSums() -->
+
 <!-- T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA) -->
+
 <!-- T_orig_longer <- cbind(T_orig,W_i_orig) -->
+
 <!-- T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig)) -->
@@
 <!-- rownames(T_orig_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(T_orig_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- T_orig_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel): -->
+
 <!-- $$ -->
+
 <!-- T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+
 <!-- $$ -->
+
 <!-- \where $B_j$ is, and equal to: -->
+
 <!-- $$ -->
+
 <!-- B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\ -->
+
 <!-- $$ -->
+
 <!-- ```{r} -->
+
 <!-- B_j <- 1/(t(matrix(W_i))) %*% (1/C)  -->
@@
 <!-- B_j <- matrix(B_j) |> t() -->
+
 <!-- B_j_longer <- B_j -->
+
 <!-- colnames(B_j_longer) <- c("1","2","3") -->
+
 <!-- B_j_longer  #|> as.data.frame()  |> gt() -->
+
 <!-- #B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns -->
+
 <!-- ``` -->
@@
 <!-- And trips are: -->
+
 <!-- ```{r} -->
+
 <!-- B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3) -->
@@
 <!-- W_i_dest <- T_dest |> rowSums() -->
+
 <!-- W_j_dest <- T_dest |> colSums() -->
+
 <!-- T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA) -->
+
 <!-- T_dest_longer <- cbind(T_dest,W_i_dest) -->
+
 <!-- T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest)) -->
@@
 <!-- rownames(T_dest_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(T_dest_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- T_dest_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel,  @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
+
 <!-- $$ -->
+
 <!-- T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
+
 <!-- $$ -->
+
 <!-- \where In our example, $A_i$ and $B_j$ is: -->
+
 <!-- $$ -->
+
 <!-- A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\ -->
+
 <!-- B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})} -->
+
 <!-- $$ -->
+
 <!-- ```{r} -->
+
 <!-- doubly_constrained_gravity_model <- function( -->
+
 <!--     #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R -->
+
 <!--   O,  # Origin production weights vector -->
+
 <!--   D,  # Destination attraction weights vector -->
+
 <!--   impedance_matrix,  # the resulting travel cost impedance matrix -->
+
 <!--   error_threshold = 0.001,  # Error threshold for stopping condition -->
+
 <!--   improvement_threshold = 1e-6  # Improvement threshold for stopping condition -->
+
 <!-- ) { -->
@@
 <!--   # Normalize O and D -->
+
 <!--   sum_O <- sum(O) -->
+
 <!--   sum_D <- sum(D) -->
+
 <!--   cat("Checking production, attraction balancing:\n") -->
+
 <!--   cat("Production: ", sum_O, "\n") -->
+
 <!--   cat("Attraction: ", sum_D, "\n") -->
@@
 <!--   if (sum_O != sum_D) { -->
+
 <!--     cat("Productions and attractions do not balance, scaling to match.\n") -->
+
 <!--     if (sum_O < sum_D) { -->
+
 <!--       O <- O * (sum_D / sum_O) -->
+
 <!--     } else { -->
+
 <!--       D <- D * (sum_O / sum_D) -->
+
 <!--     } -->
+
 <!--   } else { -->
+
 <!--     cat("Production, attraction balancing OK.\n") -->
+
 <!--   } -->
@@
 <!--   n <- length(O)  # Number of i -->
+
 <!--   T <- sum(O)  # Total trips -->
+
 <!--   Ai <- rep(1, n)  # Initialize Ai -->
+
 <!--   Bj <- rep(1, n)  # Initialize Bj -->
@@
 <!--   previous_error <- Inf -->
+
 <!--   iteration_count <- 0 -->
+
 <!--   stop_reason <- "" -->
@@
 <!--   while (TRUE) { -->
+
 <!--     iteration_count <- iteration_count + 1 -->
@@
 <!--     # Update Ai -->
+
 <!--     for (i in 1:n) { -->
+
 <!--       Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9) -->
+
 <!--     } -->
@@
 <!--     # Update Bj -->
+
 <!--     Bj_new <- rep(1, n) -->
+
 <!--     for (j in 1:n) { -->
+
 <!--       Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9) -->
+
 <!--     } -->
@@
 <!--     # Calculate Tij -->
+
 <!--     Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix -->
@@
 <!--     # Print Ai and Bj for this iteration -->
+
 <!--     cat("\nIteration:", iteration_count, "\n") -->
+
 <!--     cat("Ai:", round(Ai, 3), "\n") -->
+
 <!--     cat("Bj:", round(Bj_new, 3), "\n") -->
@@
 <!--     # Calculate error -->
+
 <!--     error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T -->
@@
 <!--     # Calculate change in error -->
+
 <!--     error_change <- abs(previous_error - error) -->
@@
 <!--     # Check stopping conditions -->
+
 <!--     if (error < error_threshold) { -->
+
 <!--       stop_reason <- "Error threshold met" -->
+
 <!--       break -->
+
 <!--     } else if (error_change < improvement_threshold) { -->
+
 <!--       stop_reason <- "Slow improvement" -->
+
 <!--       break -->
+
 <!--     } -->
@@
 <!--     previous_error <- error -->
+
 <!--     Bj <- Bj_new -->
+
 <!--   } -->
@@
 <!--   # report stopping condition and final error threshold -->
+
 <!--   cat("\nStopping Condition: ", stop_reason, "\n") -->
+
 <!--   cat(sprintf("Final Error: %.3f%%\n", error * 100)) -->
@@
 <!--   return(list(Ai = Ai, Bj = Bj, Tij = Tij)) -->
+
 <!-- } -->
@@
 <!-- dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C) -->
+
 <!-- ``` -->
@@
 <!-- The $A_i$ and $B_j$ matrices: -->
+
 <!-- ```{r} -->
+
 <!-- A_i_doubly <- dc_results$Ai -->
+
 <!-- B_j_doubly <- dc_results$Bj -->
@@
 <!-- A_i_doubly |> matrix() -->
+
 <!-- B_j_doubly |> matrix() |> t() -->
+
 <!-- ``` -->
@@
 <!-- The trip matrix (notice, all trip ends are equivalent to the known trip ends) -->
+
 <!-- ```{r} -->
+
 <!-- T_doubly <- dc_results$Tij -->
@@
 <!-- W_i_doubly <- T_doubly |> rowSums() -->
+
 <!-- W_j_doubly <- T_doubly |> colSums() -->
+
 <!-- T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA) -->
+
 <!-- T_doubly_longer <- cbind(T_doubly,W_i_dest) -->
+
 <!-- T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest)) -->
@@
 <!-- rownames(T_doubly_longer) <- c("1","2","3", "TOTAL")  -->
+
 <!-- colnames(T_doubly_longer) <- c("1","2","3", "TOTAL") -->
@@
 <!-- T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+
 <!-- ``` -->
@@
 <!-- ```{r} -->
+
 <!-- #unconstrained -->
+
 <!-- ((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
@@
 <!-- #unconstrained -- rescaled! -->
+
 <!-- W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums() -->
+
 <!-- W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums() -->
@@
 <!-- W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums() -->
+
 <!-- W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums() -->
@@
 <!-- ((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01) -->
+
 <!-- ``` -->
@@
 <!-- ```{r} -->
+
 <!-- #total constrained -->
+
 <!-- ((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
@@
 <!-- #singly-constrained (orig) -->
+
 <!-- ((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
@@
 <!-- #singly-constrained (dest) -->
+
 <!-- ((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
@@
 <!-- #doubly-constrained (dest) -->
+
 <!-- ((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
+
 <!-- ``` -->
@@
 <!-- - doubly c is lowest error -->
+
 <!-- - singly constrained error is mid -->
+
 <!-- - total constrained has high error -->
+
 <!-- - notably, unconstrained unscaled is obviously much higher in error - there are no constraints!

### 2025-04-04 paper/manuscript.qmd
definitio of competition

@@
 
-An ascendant interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
+\<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above --!\>
+
+A growing interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
 In other words, accessibility analysis is used today to broadly understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
@@
 
-However, there is room to further clarify another relevant aspect: competition.
+However, there is room to further clarify another relevant aspect: competition, i.e.
+when finite resources can be concurrently accessed by many people.
 @wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach within the competitive accessibility literature.

### 2025-04-04 paper/manuscript.qmd
comments to intro

@@
 
+\<!-- the first two paragraphs set the context / expectaction for the paper.
+In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice.
+However, the paper is very much a theoretical and methodological contribution.
+the first paragraphs should reflect this, drawing attention to the common origin of spatial interaction models and accessibility , and how they diverged over time and we got lost along the way .
+Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ.
+this could be briefly explained --!\>
+
 A growing interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
@@
 
+\<!-- the paragraph above is really good --!\>
+
 $$
@@
 when finite resources can be concurrently accessed by many people.
-@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach within the competitive accessibility literature.
+@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach in the accessibility literature that accounts for competition.
 There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach.
 Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971.
+
+\<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'.
+These ideas need to be explained in this paper.
+I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then --!\>
 
@@
 We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
+
+\<!--
+
+when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' .
+perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections
+
+--!\>
 

### 2025-04-04 paper/manuscript.qmd
fixing equations
the double dollar signs got messed up after my edits

--- 2025-03-07 paper/manuscript.qmd
@@
 
-In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows: $$
+In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
+
+$$
 V_1 = G\frac{M_2}{d_{12}}
@@
 
-For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$): $$
+For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
+
+$$
 V_i = \int\frac{D}{r} ds
@@
 \noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
-In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as: $$
+In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
+
+$$
 V_i = \sum_j \frac{M_j}{d_{ij}}
@@
 
-Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used: $$
+Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used:
+
+$$
 \sum_i T_{ij} = D_j
@@
 
-And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios: $$
+And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios:
+
+$$
 \begin{array}{l}
@@
 
-Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be: $$
+Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be:
+
+$$
 \begin{array}{l}
@@
 M^T_i = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$ Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless: $$
+$$
+
+Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
+
+$$
 \hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
@@
 
-And the corresponding balancing factor, as per Wilson, is: $$
+And the corresponding balancing factor, as per Wilson, is:
+
+$$
 A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
@@
 
-Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression: $$
+Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression:
+
+$$
 M^S_{j} = \sum_i \hat \kappa^S_{ij} O_i
@@
 
-The proportional allocation factor $\kappa_{ij}^D$ would then be: $$
+The proportional allocation factor $\kappa_{ij}^D$ would then be:
+
+$$
 \kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})

### 2025-04-05 paper/manuscript.qmd
oh Hansen, you broke my heart

--- 2025-04-04 paper/manuscript.qmd
@@
 
-\<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above --!\>
-
-\<!-- the first two paragraphs set the context / expectaction for the paper.
+<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above -->
+
+<!-- the first two paragraphs set the context / expectaction for the paper.
 In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice.
@@
 Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ.
-this could be briefly explained --!\>
+this could be briefly explained -->
 
@@
 
-\<!-- the paragraph above is really good --!\>
+<!--  the paragraph above is really good -->
 
@@
 
-\<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'.
+<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'.
 These ideas need to be explained in this paper.
-I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then --!\>
+I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then -->
 
@@
 
-\<!--
-
-when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' .
-perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections
-
---!\>
+<!-- when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' .
+perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections -->
 
@@
 @eq-phys-grav-prop, however, does not quantify the magnitude of the force.
-To do so, an empirical constant is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
+To do so, an empirical constant, aka gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
 In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data.
@@
 This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
--->
-```
+```
+
+Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
+This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
 
@@
 After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
-This may relate to academic backgrounds; where Ravenstein, Reilly, and Zipf were social scientists, Stewart was a physicist.
+This may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
 Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
@@
 
-In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows: $$
+<!-- what exactly did he mean by 'molecular weights' ? -->
+
+In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
+
+<!-- in plain language, what do you mean by 'the population potential of 2 with respect to 1' ?
+the potential number of people from location 2 that would visit location 1 ?
+but also, I'm not sure we need this equation 7 -->
+
+\
+$$
 V_1 = G\frac{M_2}{d_{12}}
@@
 
-<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up.
-Is now cited as the father of accessibility.
+<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up and is now cited as the father of accessibility.
 However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day.
@@
 73).
-As well as being a student of city and regional planning at Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
+As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
 @hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
@@
 
-Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about *directly* and *inversely proportional* relationships when discussing population, opportunities, and their separation in space.
+However, Hansen' use of Stewart's *population potential* measure included one crucial omission that afflict the literature to this day.
+The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
+This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
+Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.
+
+Paper, the omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis.
 At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity".
-
-After @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.
-In this way, $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
-Alas, without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
+In practice, the persistent omission of the costant in accessibility analyses  means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
 For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$.
-Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility [@millerAccessibilityMeasurementApplication2018].
+This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
+Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].
+
+<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English ->
 

### 2025-04-06 paper/manuscript.qmd
comments and remove bibliographic analysis

@@
 
-<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English ->
+<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English -->
 
@@
 
+<!-- rafa: I agree it would be nice to have a sentence or two that sets the context of wheter wilson was aware of stewart's and hansen's work -->
+
 In his groundbreaking work, Wilson [-@wilson1971] defined a general spatial interaction model as follows:
@@
 The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations.
-Finally, $f(c_{ij})$ is the $i-j$ element of an $n \times m$ matrix representing some function of travel cost $c_{ij}$ which reflects travel impedance.
+Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
 In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
-In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled: it balances the units.
+In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
+In other words, it balances the units, just like the gravitational constant in Newton’s Law of Universal Gravitation.
 
@@
 Building on this, a family of spatial interaction models can be derived.
-In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model.
+In the framework introduced in @wilson1971, this includes the "unconstrained" model in @eq-phys-gravity-model.
+In practical terms, this means that the total number of trips predicted by the mode must be equal to sum of all flows from origin i to destinations j.
 The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
+
+<!-- rafa: the terminology is confusing because first you say it is "unconstrained", and then you say it is "total flow constrained".
+I've adjusted the text trying say in plain English what means to be 'unconstrained' here.
+Not sure my interpretation is accurate, though -->
 
@@
 
+<!-- rafa: in plain English, what does it mean? that the total number of predicted trips must be equal to the sum of the all the masses in the origins ? -->
+
 $$
@@
 
+<!-- rafa: in plain English, what does it mean? that the total number of predicted trips must be equal to the sum of the all the masses in the destinations ? -->
+
 $$
@@
 When both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold, the resulting models is, in Wilson's terms, doubly-constrained, and takes the following form:
+
+<!-- rafa: in plain English, what does it mean?  -->
 
@@
 We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
-As a result of this search we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, `r n_w  |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
+As a result of this search we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, and only `r n_w  |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
 The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
 The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971 or both.
+In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971
+
+<!-- rafa: this  bibliographic analysis with Figure 2 is cool, but it looks unecessary and distractig from the core contribution of the paper. I'd suggest removing it  -->
 
@@
 
-After compiling this corpus of documents, we used the {bibliometrix} package [@bibliometrix2017] to create a bibliographical coupling matrix.
-In this matrix, two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common.
-@fig-Fig1 presents the results, where each symbol represents a document.
-The distance between symbols on the plot indicates the strength of bibliographical coupling: documents that are more strongly coupled (i.e., those sharing more references) are plotted closer to each other.
-
-```{r}
-#| label: layout-for-coupling-net
-#| cache: true
-
-layout <- create_layout(coupling_net,
-                        layout = "stress",
-                        weights = weight)
-```
-
-```{r}
-#| label: plot-coupling-network
-#| include: false
-
-coupling_net_plot <- ggplot(data = layout) +
-  geom_point(aes(x = x,
-                 y = y,
-                 shape = corpus,
-                 color = corpus),
-             size = 3,
-             alpha = 0.7) +
-  geom_point(data = layout |>
-               filter(corpus == "Both"),
-             aes(x = x,
-                 y = y),
-             shape = 1,
-             size = 3) +
-  geom_point(data = layout |>
-               filter(corpus == "Wilson"),
-             aes(x = x,
-                 y = y),
-             shape = 0,
-             size = 3,
-             alpha = 0.5) +
-  theme_void()
-
-ggsave(filename = glue::glue(here::here(), 
-                             "/figures/Han_Wilson_bib_coupling_plot.png"))
-```
-
-```{r}
-#| label: fig-Fig1
-#| out-width: 70%
-#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959), Wilson (1971), or both."
-
-coupling_net_plot
-```
-
-```{r}
-#| label: center-of-mass-of-Hansen
-
-# This is the center of mass of the documents that cite Hansen
-hansen_center_of_mass <- layout |>
-  filter(corpus == "Hansen") |>
-  summarize(x = mean(x),
-            y = mean(y))
-```
-
-```{r}
-#| label: center-of-mass-of-Wilson
-
-# This is the center of mass of the documents that cite Wilson
-wilson_center_of_mass <- layout |>
-  filter(corpus == "Wilson") |>
-  summarize(x = mean(x),
-            y = mean(y))
-```
-
-```{r}
-#| label: row-names-by-corpus
-
-# Extract the row names by corpus to be able to partition the coupling matrix
-hansen_names <- rownames(corpus |>
-                           filter(corpus == "Hansen"))
-
-wilson_names <- rownames(corpus |>
-                           filter(corpus == "Wilson"))
-
-both_names <- rownames(corpus |>
-                           filter(corpus == "Both"))
-```
-
-```{r}
-#| label: partition-of-coupling-matrix
-
-# With the row names we can extract submatrices to examine the strength of coupling within and between the three corpora, i.e., documents that cite only Hansen, only Wilson, or both
-hansen_to_wilson <- B_nd[hansen_names, wilson_names]
-
-hansen_to_hansen <- B_nd[hansen_names, hansen_names]
-
-wilson_to_wilson <- B_nd[wilson_names, wilson_names]
-
-both_to_both <- B_nd[both_names, both_names]
-
-both_to_hansen <- B_nd[both_names, hansen_names]
-
-both_to_wilson <- B_nd[both_names, wilson_names]
-```
-
-```{r}
-#| label: summary-of-coupling-within-and-between-corpora
-
-# The number of common citations per pair of documents can be calculated as the sum of all common citations in the matrix, divided by the number of pair-paper combinations:
-# Average coupling for the total corpora
-c_t <- sum(sum(B_nd))/(dim(B_nd)[1] * (dim(B_nd)[2]))
-# Average coupling for papers that cite both Hansen and Wilson
-c_b <- sum(sum(both_to_both))/(dim(both_to_both)[1] * (dim(both_to_both)[2]))
-# Average coupling for papers that cite only Hansen
-c_h <- sum(sum(hansen_to_hansen))/(dim(hansen_to_hansen)[1] * (dim(hansen_to_hansen)[2]))
-# Average coupling for papers that cite only Wilson
-c_w <- sum(sum(wilson_to_wilson))/(dim(wilson_to_wilson)[1] * (dim(wilson_to_wilson)[2]))
-# Average coupling between papers that cite both Hansen and Wilson, and those that cite only Hansen
-c_h_to_b <- sum(sum(both_to_hansen))/(dim(both_to_hansen)[1] * dim(both_to_hansen)[2])
-# Average coupling between papers that cite both Hansen and Wilson, and those that cite only Wilson
-c_w_to_b <- sum(sum(both_to_wilson))/(dim(both_to_wilson)[1] * dim(both_to_wilson)[2])
-# Average coupling between papers that cite only Hansen, and those that cite only Wilson
-c_h_to_w <- sum(sum(hansen_to_wilson))/(dim(hansen_to_wilson)[1] * dim(hansen_to_wilson)[2])
-```
-
-Further examination of the bibliographical coupling matrix allows us to check the coupling strength within and between subgroups of documents (i.e., those that cite only Hansen, only Wilson, or both papers).
-First, we note that when we look at the three subgroups of documents pooled together, the average number of references shared by a pair of documents is `r round(c_t, 2)`.
-@tbl-coupling-results presents the average number of items cited in common within each subgroup, as well as between subgroups.
-As seen in the table, the coupling within subgroup tends to be higher than when we consider the full corpus with all three groups together.
-The tightest coupling is seen for the group of documents that cite both Hansen and Wilson, with an average number of references in common of `r prettyNum(c_b, digits = 3)`.
-The weakest coupling is between the Hansen and Wilson subgroups, with only `r prettyNum(c_h_to_w, digits = 1)` references in common on average, which indicates a large degree of decoupling (independence) between these bodies of work.
-
-```{r}
-#| label: tbl-coupling-results
-#| tbl-cap: "Average coupling strength within and between groups of documents. The average coupling strength is the number of references shared on average by a pair of documents."
-
-data.frame(corpus = c(glue::glue("Hansen (n = {n_h |> prettyNum(big.mark = ",")})"), 
-                      glue::glue("Wilson (n = {n_w |> prettyNum(big.mark = ",")})"),
-                      glue::glue("Both (n = {n_b  |> prettyNum(big.mark = ",")})")),
-           Hansen = c(prettyNum(c_h, digits = 3),
-                      prettyNum(c_h_to_w, digits = 1),
-                      prettyNum(c_h_to_b, digits = 3)),
-           Wilson = c("-",
-                      prettyNum(c_w, digits = 3),
-                      prettyNum(c_w_to_b, digits = 3)),
-           Both = c("-",
-                    "-",
-                    prettyNum(c_b, digits = 3))
-           ) |>
-  gt() |>
-  tab_footnote(footnote = glue::glue("Average coupling for the pooled set of documents (n = {(n_h + n_w + n_b) |> prettyNum(big.mark = ",")}) is {round(c_t, 2)}")) |>
-  tab_options(footnotes.font.size = "small") |>
-  as_latex()
-```
-
 As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
@@
 The spatial-interaction-focused stream of literature inspired by @wilson1971 that cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
-Specifically, works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
+Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].

### 2025-04-06 paper/manuscript.qmd
comment on section two divergent research streams

@@
 
-The spatial-interaction-focused stream of literature inspired by @wilson1971 that cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
+The stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
 Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
@@
 
-On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-- Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+On the other hand, we discern that the stream of literature focused on accessibility inspired by @hansen1959 cites @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions <!-- no need to cite all of these references here.
+Keep it to 4 --> [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+- Secondly, associating spatial interaction with accessibility's potential for spatial interaction <!-- no need to cite all of these references here.
+Keep it to 4 --> [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
 While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
 This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts <!-- no need to cite all of these references here.
+Keep it to 4 --> [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
 However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues.
 Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways for maintaining constraints in opportunities in the context of competitive accessibility.
+
+<!-- check text of the paragraphp above for clarity -->
+
+<!-- if only soukhov et al 2023 and 2024 have previously used 'balancing factors as ways for maintaining constraints', it might, you might wanna explain in more detail what was dones in those papers.
+this is also important to position the contribution of this current paper compared to those previous studies -->
 

### 2025-04-06 paper/manuscript.qmd
constrained terminology

@@
 
-As argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with little contact after @hansen1959 and @wilson1971.
-This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis.
-This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of @eq-production-constrained-balancing-factor and is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
+As argued in the preceding section, the literatures on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
+This may explain why the constraints/proportionality constant(s) of spatial interaction models did not cross over to accessibility analysis.
+This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of @eq-production-constrained-balancing-factor is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
 10]:
+
+<!-- "offering a potential interpretation for the proportionality constant", which would be..... -->
 
@@
 In the following subsections, we detail how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus.
-Together, these constraints delineate different members of the family of accessibility measures.
-Along with a simple numerical example, we detail the unconstrained case (@hansen1959's formulation), then total constrained cases (opportunity-constrained and population-constrained), then singly-constrained cases (opportunity-constrained and population-constrained), and lastly the doubly-constrained case.
-These cases are presented in increasing order of constraint restrictiveness and information availability.
+Together, these constraints delineate different members of the family of accessibility measures and which fall under four categories: (1) the unconstrained case, which is equivalent to @hansen1959's formulation; (2) then total constrained cases, which is constrained by total population and opportunities.; (3) the singly-constrained cases, which is opportunity-constrained or population-constrained; and (4) the doubly-constrained case, which is constrained simultaneously by population and opportunities.
+In the next section, we use a simple numerical example to illustrate these cases in increasing order of constraint restrictiveness and information availability.
+
+<!-- confusing terms.
+what is the difference between total constrained and doubly-constrained case ?
+we probably need to think of better terms here.
+I also liked Chris' idea of including a table that summarizes the differences between these measures -->
 

### 2025-04-06 paper/manuscript.qmd
clarifications

@@
 In the framework introduced in @wilson1971, this includes the "unconstrained" model in @eq-phys-gravity-model.
-In practical terms, this means that the total number of trips predicted by the mode must be equal to sum of all flows from origin i to destinations j.
+In practical terms, this means that the total number of trips predicted by the model must be equal to sum of all flows from origins i to destinations j.
 The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
@@
 
-As argued in the preceding section, the literatures on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
+As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
 This may explain why the constraints/proportionality constant(s) of spatial interaction models did not cross over to accessibility analysis.
-This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of @eq-production-constrained-balancing-factor is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
+This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
 10]:
-
-<!-- "offering a potential interpretation for the proportionality constant", which would be..... -->
+<!-- "offering a potential interpretation for the proportionality constant...", which would be..... -->
 
@@
 In the preceding section on early research, we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the pairwise population potential $V_1 = G\frac{M_2}{d_{12}}$ and from there to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
-While the inverse of $A_i$ may appear to be accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
-
-Hence, we propose a more useful definition: accessibility as a partial sum of the spatial interaction. A way to define the *potential for spatial interaction* (i.e., accessibility) is as follows:
+While the inverse of $A_i$ is pretty much identical to Hansen's accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
+
+Hence, we propose a more general definition of accessibility as a partial sum of the spatial interaction.
+Once we bring back Wilson's proportionality constant into the picture, we can define the *potential for spatial interaction* between two locations i and j is as follows:
 
@@
 
-\noindent where $V_{ij}$ is the potential for interaction, and therefore:
+\noindent where $V_{ij}$ is the potential for interaction, and therefore the accessibility from origin i is defined as:
 
@@
 
-In the following subsections, we detail how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus.
-Together, these constraints delineate different members of the family of accessibility measures and which fall under four categories: (1) the unconstrained case, which is equivalent to @hansen1959's formulation; (2) then total constrained cases, which is constrained by total population and opportunities.; (3) the singly-constrained cases, which is opportunity-constrained or population-constrained; and (4) the doubly-constrained case, which is constrained simultaneously by population and opportunities.
+In the following subsections, we detail what are the implications of bringing this constraint into accessibility measures, and how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus.
+Together, these constraints delineate different members of the family of accessibility measures and which fall under four categories: (1) the unconstrained case, which is equivalent to @hansen1959's formulation; (2) the total constrained cases, which is constrained by total population and opportunities.; (3) the singly-constrained cases, which is opportunity-constrained or population-constrained; and (4) the doubly-constrained case, which is constrained simultaneously by population and opportunities.
 In the next section, we use a simple numerical example to illustrate these cases in increasing order of constraint restrictiveness and information availability.
@@
 
-Where the sum of $V^0_{i}$ in the region generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), as setting $k$ to 1 to arbitrarily eliminates the term and imposes no meaningful interpretation to the resulting values.
+In such case, the the sum of the accessibility from each origin $V^0_{i}$ in the region generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$).
+This is because setting $k$ to 1 to arbitrarily eliminates the constant term and imposes no meaningful interpretation to the resulting values.
 To complicate matters, when comparing $V^0_{i}$ values between travel behaviour contexts, their difference depends on the choice of impedance function $f(c_{ij})$, associated parameters, and the number of zones for which an accessibility score is calculated.
-For this reason, the unconstrained accessibility should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
-
-Returning to the simple example, consider the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in @tbl-simple-example-unconstrained-accessibility.
+<!-- double check this next sentence .
+there should be a more explicit mention to the units problem ?
+--> For this reason, the unconstrained accessibility is not comparable across different contexts and decay functions, and should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
+
+Returning to our numeric example, consider the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in @tbl-simple-example-unconstrained-accessibility.
 

### 2025-04-06 paper/manuscript.qmd
unconstrained access in one step

@@
 
+<!-- this paragraph could probably be merged with the previous one -->
+
 Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
@@
 $$
-V^0_{ij} = W_j^{(2)} f(c_{ij})
+V^0_{ij} = 1 *W_j^{(2)} f(c_{ij})
 $$ {#eq-unconstrained-access}
 
-Then, the partial sum becomes the following: simply Hansen's accessibility $S_i$ [@hansen1959], standard practice in accessibility measurement:
+In this case, the partial sum of spatial interaction is simply identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
 
@@
 
-unc_acc_ij <- OD |>
+unc_acc <- OD |>
   group_by(oid) |>
-  reframe(V_unc_ij_1 = D_j * f1,
-            V_unc_ij_2 = D_j * f2,
-            V_unc_ij_3 = D_j * f3)
-
-unc_acc <- unc_acc_ij |>
-  group_by(oid) |>
-  summarize(V_unc_i_1 = sum(V_unc_ij_1),
-            V_unc_i_2 = sum(V_unc_ij_2),
-            V_unc_i_3 = sum(V_unc_ij_3))
+  summarize(V_unc_i_1 = sum(D_j * f1),
+            V_unc_i_2 = sum(D_j * f2),
+            V_unc_i_3 = sum(D_j * f3))
 

### 2025-04-06 paper/manuscript.qmd
code

@@
 
-unc_acc <- OD |>
+unc_acc_ij <- OD |>
   group_by(oid) |>
-  summarize(V_unc_i_1 = sum(D_j * f1),
-            V_unc_i_2 = sum(D_j * f2),
-            V_unc_i_3 = sum(D_j * f3))
+  reframe(V_unc_ij_1 = D_j * f1,
+          V_unc_ij_2 = D_j * f2,
+          V_unc_ij_3 = D_j * f3)
+
+unc_acc <- unc_acc_ij |>
+  group_by(oid) |>
+  summarize(V_unc_i_1 = sum(V_unc_ij_1),
+            V_unc_i_2 = sum(V_unc_ij_2),
+            V_unc_i_3 = sum(V_unc_ij_3))
+
+## unconstrained access in one step
+# unc_acc <- OD |>
+#   group_by(oid) |>
+#   summarize(V_unc_i_1 = sum(D_j * f1),
+#             V_unc_i_2 = sum(D_j * f2),
+#             V_unc_i_3 = sum(D_j * f3))
 
@@
 In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3.
-Both scenarios at zone 1 captures an intermediate $V_i^0$ value between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians.
+Both scenarios at zone 1 captures an intermediate accessibility value $V_i^0$ between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians.
 The ratios between the zones in the two scenarios are different.
@@
 The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case.
-We define two cases: a *totally-constrained opportunity* accessibility case, interpreted as Hansen's accessibility with a constraint, and a *totally-constrained population* accessibility case, interpreted as constrained market potential.
+We define two cases: (a) a case where accessibility is constrained by the total number of opportunities (aka *totally-constrained opportunity*) and which is interpreted as Hansen's accessibility with a constraint, and (b) a case where accessibility is constrained by the total population (aka *totally-constrained population*) and which is interpreted as constrained market potential.
 
@@
 $$ {#eq-total-constraint-accessibility}
+
+<!-- rafa: do we need to keep equations 31 and 32 -->
 
@@
 \noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$).
-The role of $\kappa^T$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+The role of $\kappa^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
 Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
@@
 <!-- kappa in equations now? -->
+
+<!-- rafa: 'The role of $\kappa^T$ in this reformulation of accessibility is to..'.
+This sentence is gold and it should be said earlier on.
+My personal preference is that first we need to explain to the reader in plain language what the constraint is, and what it means in practical terms (why it matters).
+Then we can show this with the equations.
+-->
 

### 2025-04-07 paper/manuscript.qmd
Singly-constrained opportunity

@@
 
-In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios.
+In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
 Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa^T$.
@@
 Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`).
-However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
+However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
 Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
-Between zonal comparisons can be more interpretably made for these travel behaviour scenarios as well.
 
@@
 
+<!-- rafa: This is a very important takeaway. But this is true for any metric using the constant, right? 66666666 -->
+
 ### Totally-constrained population: market potential
@@
 
-Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
+By transposing $i$ and $j$, we have an  accessibility measure constrained by total population (@eq-total-constrained-market), which effectively gives us an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
 The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
@@
 $$
-
-The constraint that we impose in this case is the total market potential equals the total population $O$ in the region :
+This measure is simply a form of passive accessibility, and indicates that each destination can be reached by an x number of people.
+The constraint that we impose in this case in practice means that the total market potential equals the total population $O$ in the region:
 
@@
 In fact, $\kappa^S_{i}$ is the proportion of the sum of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
-Hence, since zone 1 has the lowest amount of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.
+Hence, since zone 1 has the lowest number of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.
 
 <!-- check sentence ^ -->
+
+<!-- rafa: I think we don't need the paragraph below -->
 
@@
 This is unlike other zones, that capture opportunities from their own zones at lower rates (i.e., ($\kappa_{1}^S$ corresponds to a capture of `r (opp_acc_ij$k_opp_ij_1[1]/opp_acc_with_ks$k_opp_ij_1[1])|> scales::percent()` of its allocation from Zone 1 and $\kappa_{3}^S$ corresponds to `r (opp_acc_ij$k_opp_ij_1[9]/opp_acc_with_ks$k_opp_ij_1[3])|> scales::percent()` of its allocation from Zone 3)).
+
+<!-- rafa: I think we don't need the paragraph below -->
 
@@
 
+<!-- rafa: in the last sentence, you clarify the difference between 'total opportunity constraint' Vs  'single opportunity constraint' .
+What is confusing here is how the single opportunity constraint compares to 'total population constraint' .
+Again, the idea of including a table would be very helpful.
+The table shoud include one indicator in each row.
+The columns could include (1) name of the indicator, (2) explanation in plain language of the constraint, (3) equation of the constraint, (4) equation of the indicator, (5) example of the interpretation of the indicator -->
+
 Like the totally-constrained opportunity case, this case offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.

### 2025-04-07 paper/manuscript.qmd
Singly-constrained

@@
 
-Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly-constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^(1)$ is now represented by $O_i$).
+Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly-constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
 In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.

### 2025-04-30 paper/manuscript.qmd
Another revision of the draft done --- off to AP!

@@
 ---
-title: "A family of accessibility measures derived from spatial interaction principles"
+title: "A family of accessibility measures and the concept of potential derived from spatial interaction principles"
 format:
@@
 
-<!-- SUMMARY: Accessibility is related to mobility, however, institutionalized mobility-based transportation planning practice has treated accessibility as it's proxy.
+<!-- SUMMARY *AND OUTLINE FOR ABSTRACT*: Transportation planning, historically, has been mobility-oriented.
+Accessibility, however, should be the primary focus, and institutionalized mobility-based transportation practice has treated accessibility as its proxy.
 This has had problematic consequences.
-Conceptually, accessibility is a more complete and comprehensive evolution of mobility.
-As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are 1) too many definitions and 2) uncertainty in meaning of the outputs-->
-
-Historically, the focus of modern transportation planning has been to prioritize mobility while treating access to destinations as a by-product of movement.
-This has had problematic consequences: with the car seen as the ultimate mobility tool, this approach has led to the emergence and continued dominance of an automobility mono-culture [@miller_collaborative_2011; @lavery_driving_2013].
-Decades of planning for this mono-culture (often characterized by road and highway expansion) have been marked by increased travel cost and environmental burdens, but often having only limited impact on the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
-In response to this situation, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020] using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020].
-Accessibility, a central concept in transport geography, planning, and engineering, is conceptually related to human mobility and the opportunity landscape.
-In simple terms, it is defined as the "potential of opportunities for interaction" [@hansen1959].
-Compared to other measures of performance used in transportation that benchmark movement (e.g., VKT, PKT, etc.), accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997].
-
-<!-- I suggest removing 'using both positive and normative approaches [@paez2012measuring; @levineCenturyEvolutionAccessibility2020]' from the paragraph above -->
-
-<!-- the first two paragraphs set the context / expectaction for the paper.
-In the current form, it the first two paragraphs seem to be more focused on the importance of accessibility for planning, and the challenges about adopting accessibility into planning practice.
-However, the paper is very much a theoretical and methodological contribution.
-the first paragraphs should reflect this, drawing attention to the common origin of spatial interaction models and accessibility , and how they diverged over time and we got lost along the way .
-Some pepople might not have a clear understanding of how the two concepts connect or how exactly they differ.
-this could be briefly explained -->
-
-A growing interest in accessibility has been accompanied by a remarkable boom in scholarly research, which has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
-In other words, accessibility analysis is used today to broadly understand the potential to reach opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
+Conceptually, accessibility is a more complete and comprehensive evolution of mobility: from movement, to the potential movement to destinations.
+As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are  too many definitions and uncertainty in meaning of the outputs.
+This work's goal is to enhance interpretation of values, clarify the concept of accessibility, and unify other measures of accessibility such as the popular Hansen-type measure the 2SFCA.
+
+To do so, this work demonstrates that accessibility and mobility oriented models (i.e., gravity-based accessibility and the spatial interaction model) share the same roots in spatial interaction modelling, and a family of accessibility measures can be specified within the spatial interaction modelling principles.
+In demonstrating these developments, three significant contributions emerge: A) we demonstrate numerically how constrained accessibility can be specified, and doing so brings interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'.
+B) we provide a clarifying framework for understanding the concept of 'potential' through constraints.
+Namely, demonstrating accessibility's maximum potential is captured by the the 'total constraint' (constrained version of Hansen-type measure), a restricted amount of potential is captured by the 'single constraint' (i.e., competition, such as the 2SFCA), and potential is fully restricted in the 'double constraint', effectively restricting potential such that this case is equivalent to access or simply spatial interaction.
+C) Demonstrate how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the  family of accessibility measures.
+
+In fewer words: this work makes an important contribution to accessibility literature, by providing accessibility with a more interpretable, clarified and unified definition through the introduction of constraints from spatial interaction modelling, that reintroduce units to enhance interpretability of raw values, clarify the concept of potential, and loops the most popular forms of accessibility measures into the family.
+-->
+
+As industrializing cities grew rapidly in the early nineteenth century, traffic congestion became a major issue, prompting the development of transportation planning focused primarily on mobility.
+In this newly founded practice, access to destinations was treated as a by-product of movement.
+Following the rise of the automobile and significant investments in transportation infrastructure after World War II, this mobility-oriented approach has led to some problematic outcomes.
+Specifically, the car became seen as the ultimate mobility tool, helping to foster the development of low-density, single-use residential neighborhoods and entrenching an automobility mono-culture within transportation systems [@miller_collaborative_2011; @lavery_driving_2013].
+Decades of planning for this automobility mono-culture, still often characterized by road and highway expansion, have been marked by increased travel costs and environmental burdens, with limited impact on enhancing the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic, etc.) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].
+
+Accessibility can be seen to represent an evolution of mobility; accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'.
+Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling].
+While accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997], by incorporating both the explicit consideration of destination as opportunity and the concept of _potential_.
+
+The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
+In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
 However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
@@
 
-The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
+Furthermore, the adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
 At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
@@
 
-In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
+More specifically, this work seeks to address the breath of place-based accessibility's breath of definitions and lack of interpretability by first demonstrating that and mobility-oriented models, such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971], respectively, are rooted in the same spatial interaction modeling foundation.
+In fact, accessibility can be specified using spatial interaction principles.
+And a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constraint values based on known information about the land-use.
+
+In this way, this work aims to introduce constrained accessibility measures, namely the totally-constrained accessibility measure and the singly-constrained accessibility measure, and demonstrate their connection to the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+These Wilson-analogous constraints can also be defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
+We also aim to demonstrate the intuitive clarity in restoring units to resulting accessibility values; using the constrained measures each zonal value retains interpretable units, namely the 'number of opportunities'.
+This is in contrast to conventional accessibility measures that yield a value that is in units of some 'opportunities-weighted-by-some-representation-of-travel-friction'.
+
+<!-- In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
 By considering the key features of accessibility, namely travel-cost $f(c_{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson).
@@
 In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
-
-<!--  the paragraph above is really good -->
-
-$$
-S_i \propto \sum_j g(O_j)f(c_{ij})
-$$ {#eq-access-Lu-Levinson}
-
-However, there is room to further clarify another relevant aspect: competition, i.e.
+-->
+
+<!-- $$ -->
+<!-- S_i \propto \sum_j g(O_j)f(c_{ij}) -->
+<!-- $$ {#eq-access-Lu-Levinson} -->
+
+<!-- However, there is room to further clarify another relevant aspect: competition, i.e.
 when finite resources can be concurrently accessed by many people.
@@
 Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971.
-
-<!-- Most readers will not be familiar with your previous papers, so they will not understand what you mean with 'single constraint' or 'balancing factors'.
-These ideas need to be explained in this paper.
-I suggest the text 'There have also been more recent.....' in the paragraph is moved to the literature review section, where we have proper space to explain those ideas and how the ideas proposed in this paper expand on then -->
-
-Motivated by the challenge of incorporating competition into a unified framework for accessibility, this paper contends that accessibility research must reconnect with its spatial interaction origins.
+-->
+
+<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To narratively reiterate: this paper contends that accessibility research must reconnect with its spatial interaction origins.
 Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
@@
 We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
-
-<!-- when you say that the absence of empirical constraints has led to interpretation issues in accessibility analysis, and think the example should be stronger that simply saying it leads to the 'fuzziness of insights' .
-perhaps mention the problem that we loose track of measurement units, that comparability between indicators, across cities and transport modes also become compromised but most of us don't even notice it, and this should be explained in the next sections -->
-
-These streams of literature share common headwaters.
+Moreover, without constraints: we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with.
+Without a clear sense of measurement units, comparability between accessibility measures, across cities, and transport modes, may also become compromised.
+
+Fortunately, accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
 It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-Hence, this work's primary focus is to show that the same empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility (REITERATE WHY).
-We begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
+The primary focus of this work, therefore, is to demonstrate how the empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility.
+Both share similar origins in spatial interaction, and popular accessibility measures can be traced directly back to these roots, offering a solution for reintroducing units to the outputs.
+In the following section, we begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
 We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
-Next, we hark back to @wilson1971's spatial interaction models, and use it to derive a family of accessibility measures based on different types of constraints.
-We illustrate various members of this family with a simple numerical example and a real world data set.
+Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on the total and single constraints.
+We illustrate various members of this family with a simple numerical example.
 We then conclude by discussing the uses of these measure and their interpretation.
 
-```{=html}
-<!--
-We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values, as well as converting accessibility indicators from ordinal variables into interval/ration ones.
-To address this matter, we need not reinvent the wheel.
-Stewart left the determination of $G$ for the future, possibly expecting it to be some sort of universal constant.
-However, the solution to this problem was found by @wilson1971, who discovered that there was not a single $G$, but rather various balancing factors that depend on 1) empirical observations of population and/or opportunities; and 2) whether (and how) they were used to constrain the analysis.
-This development is outlined next.
--->
-```
+Using the newly defined family of accessibility measures, this work aims to: (1) demonstrate how constrained accessibility can be specified, and by doing so, bring interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'; (2) provide a clarifying framework for understanding the concept of _potential_ through the introduced constraints; and (3) highlight how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the family of accessibility measures.
 
@@
 @eq-phys-grav-prop, however, does not quantify the magnitude of the force.
-To do so, an empirical constant, aka gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
+To do so, an empirical constant, a.k.a.
+the gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
 In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data.
@@
 # Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
-
-```{=html}
 <!-- 
@@
 This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-```
-
-Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
+-->
+
+Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance.
 This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
@@
 -   $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
--   $\mu_1$ and $\mu_2$ are so-called *molecular weights*
+-   $\mu_1$ and $\mu_2$ are so-called *molecular weights*, the attractive weight of groups 1 and 2
 -   $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
@@
 
-<!-- what exactly did he mean by 'molecular weights' ? -->
-
-In addition to demographic force, Stewart defined a measure of the "population potential" of $2$ with respect to $1$ as follows:
-
-<!-- in plain language, what do you mean by 'the population potential of 2 with respect to 1' ?
-the potential number of people from location 2 that would visit location 1 ?
-but also, I'm not sure we need this equation 7 -->
-
-\
+In addition to demographic force, Stewart defined a measure of the "population potential" of group $2$ with respect to group $1$.
+In other words, the potential number of people from location $2$ that could visit location $1$, as follows:
+
 $$
@@
 
-<!--I say we spell it out as in this eq^, what you think?-->
+<!-- RP: I say we spell it out as in this eq^, what you think? AS: Maybe! -->
 
@@
 
-# Hansen's gravity-based accessibility to today
+# Hansen's gravity-based accessibility to today {#grav-to-today}
 
@@
 
-However, Hansen' use of Stewart's *population potential* measure included one crucial omission that afflict the literature to this day.
-The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished.
+However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day.
+The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished.
 This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
-Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly determine it.
-
-Paper, the omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis.
-At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity".
-In practice, the persistent omission of the costant in accessibility analyses  means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i}  \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly acknowledge or determine it.
+
+The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis.
+Those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
+34].
+In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
 The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
@@
 
-<!-- jargon ? 'a suitable choice of other units can reduce it to unity' I think this could be explained in plain English -->
-
 # Wilson's family of spatial interaction models
@@
 
-<!-- very brief narrative to introduce - is this happening separately from hansen? "On the other side of the atlantic, wilson was doing..." :) -->
-
-<!-- rafa: I agree it would be nice to have a sentence or two that sets the context of wheter wilson was aware of stewart's and hansen's work -->
-
-In his groundbreaking work, Wilson [-@wilson1971] defined a general spatial interaction model as follows:
+On the other side of the Atlantic, Alan G.
+Wilson was developing related, yet parallel work.
+In the groundbreaking [@wilson1971], Wilson defined a general spatial interaction model as follows.
+While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling spatial interaction:
+
+On the other side of the Atlantic, Alan G.
+Wilson was developing related, yet parallel work.
+In his groundbreaking study [@wilson1971], accessibility is acknowledged as an associated concept, though the work's primary focuses on defining an introducing a general spatial interaction model as follows:
 
@@
 In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
-In other words, it balances the units, just like the gravitational constant in Newton’s Law of Universal Gravitation.
+In other words, it balances the units, in a conceptually similar sense as the gravitational constant in Newton’s Law of Universal Gravitation.
 
@@
 
-To ensure that $T_{ij}$ in @eq-phys-gravity-model is in the unit of trips, additional knowledge about the system is required.
+To ensure that $T_{ij}$ in @eq-phys-gravity-model is in the unit of trips (our unit of origin-destination spatial interaction), additional knowledge about the system is required.
 At the very least, this framework assumes that the total number of trips in the system $T$ is known, and therefore:
@@
 
-Additional information can be introduced. For example, when information is available about the total number of trips produced by each origin ($O_i$), the following constraint can be used:
+Additional information can be introduced.
+For example, when information is available about the total number of trips produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
 
@@
 
-Alternatively, if there is information available about the total number of trips attracted by each destination ($D_j$), the following constraint can be used: $$
+Alternatively, if there is information available about the total number of trips attracted by each destination, $W_j^{(2)}$ is represented as $D_j$ and the following constraint can be used:
+
+$$
 \sum_i T_{ij} = D_j
@@
 
-It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints could be imposed on the model.
-
-Building on this, a family of spatial interaction models can be derived.
-In the framework introduced in @wilson1971, this includes the "unconstrained" model in @eq-phys-gravity-model.
+It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints could be imposed on the model at once.
+
+Using information about the system that satisfy these constraints fully, partially or not at all, a family of spatial interaction models can be derived based on @eq-phys-gravity-model.
+$K$ is specified depending on the applied constraint(s).
+In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where no constraints apply, the second being a singly-constrained case, and the third a doubly-constrained case.
+
+In the first, @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel do not hold.
 In practical terms, this means that the total number of trips predicted by the model must be equal to sum of all flows from origins i to destinations j.
@@
 
-<!-- rafa: the terminology is confusing because first you say it is "unconstrained", and then you say it is "total flow constrained".
-I've adjusted the text trying say in plain English what means to be 'unconstrained' here.
-Not sure my interpretation is accurate, though -->
-
 $$
@@
 
-When only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel holds, the resulting models are, in Wilson's terms, singly-constrained.
-Entropy maximisation leads to the following production-constrained model:
+In the second case, only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold.
+The resulting models are, in Wilson's terms, singly-constrained.
+When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained model:
 
@@
 
-Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced by the actual mass, as measured by the trips produced $O_i$.
+Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$, representing what we know about the system, the spatial interaction outbound flow, i.e., outbound trips produced at $i$.
 Also, there is no longer a single system-wide proportionality constant, but rather a set of proportionality constants (i.e., balancing factors) specific to origins.
-According to Wilson, these constants, namely $A_i$, are:
-
-<!-- rafa: in plain English, what does it mean? that the total number of predicted trips must be equal to the sum of the all the masses in the origins ? -->
+For this model, the balancing factors ensure that @eq-constraint1-gravitymodel is satisfied, meaning that the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$ i.e., the total number of outbound trips.
+Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of trips.
+This model is useful when trips ends are unknown but the number of trips originating from each location is known and the total of these trips represents all trips in the system.
+The balancing factors for the production-constrained model are solved for each origin $A_i$, and according to Wilson are:
 
@@
 
-The attraction-constrained model, in turn, takes the following form:
-
-<!-- rafa: in plain English, what does it mean? that the total number of predicted trips must be equal to the sum of the all the masses in the destinations ? -->
+The attraction-constrained model is similar to the production-constrained model as it is also singly-constrained but from the perspective of the mass at the destination.
+From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow, i.e., trips attracted at the destination and takes the following form:
 
@@
 
-Notice how now the proxy for the mass at the destination $W_j^{(2)}$ is replaced by the actual mass, as measured by the trips attracted $D_j$.
+In this model (@eq-attraction-constrained-gravitymodel), the balancing factors ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$ i.e., the total number of inbound trips to $j$.
+This should hold for all destinations.
 As before, destination-specific proportionality constants (i.e., balancing factors) $B_j$ were derived by Wilson as:
@@
 
-When both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold, the resulting models is, in Wilson's terms, doubly-constrained, and takes the following form:
-
-<!-- rafa: in plain English, what does it mean?  -->
+In the third case in the family of spatial interaction models is the production-attraction constrained model.
+In this case, both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
+These constraints ensure that the sum of predicted flows from one origin to all destination, and the predicted flows going to one destination from all origins must equal the known mass of the origin $O_i$ and of the destination $D_j$.
+This should hold for all origins and destinations.
+The resulting model is, in Wilson's terms, doubly-constrained, and takes the following form:
 
@@
 
-In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by origin and the trips attracted by destination.
-And, now there are two sets of mutually dependent proportionality constants:
+In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by the origin and the trips attracted by the destination.
+There are now two sets of mutually dependent proportionality constants:
 
@@
 
-```{=html}
 <!--
@@
 -->
-```
 
@@
 
-The work of @hansen1959 and @wilson1971 responded to important developments in the Anglo-American world at the time, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
-Said dictates and needs were far from trivial.
+The work of @hansen1959 and @wilson1971 responded to important developments, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
+These dictates and needs were far from trivial.
 In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S.
 Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-Accessibility at the time did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: the less tangible "potential for interaction".
+Spatial interaction modelling was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+Accessibility, at the time, did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: it predicts the less tangible "potential" for spatial interaction with opportunities.
 In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
@@
 We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
-As a result of this search we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, and only `r n_w  |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
+As a result of this search, we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, only `r n_w  |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
 The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
 The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971
+In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971.
 
@@
 
-````{=html}
 <!--
-Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-However, within the same time period, accessibility remained largely within the academic realm.
-For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.].
-We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
-Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
-
-However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
-A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
-Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
-While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
-
 ```{r}
@@
 ```
-
 -->
-````
 
 As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
-However, we can discern from a manual reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
+However, we can discern from a reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
 The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected.
 
-The stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
+On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
 Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
@@
 Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
-Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
+Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatiall interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
 Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
 
-On the other hand, we discern that the stream of literature focused on accessibility inspired by @hansen1959 cites @wilson1971 for three general reasons: - Firstly, as attribution for using using different context-dependent travel cost functions <!-- no need to cite all of these references here.
-Keep it to 4 --> [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-- Secondly, associating spatial interaction with accessibility's potential for spatial interaction <!-- no need to cite all of these references here.
-Keep it to 4 --> [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+On the other hand, we discern that the stream of literature focused on accessibility inspired by @hansen1959 cite @wilson1971 for three general reasons.
+Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
 While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
 This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts <!-- no need to cite all of these references here.
-Keep it to 4 --> [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues.
-Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways for maintaining constraints in opportunities in the context of competitive accessibility.
-
-<!-- check text of the paragraphp above for clarity -->
-
-<!-- if only soukhov et al 2023 and 2024 have previously used 'balancing factors as ways for maintaining constraints', it might, you might wanna explain in more detail what was dones in those papers.
-this is also important to position the contribution of this current paper compared to those previous studies -->
-
-```{=html}
-<!-- 
-In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
-Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
-
-Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
-Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
-Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
-
-However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
-This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
-Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
-Some works went beyond HAM and considered other accessibility expressions of individual choice.
-For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
-
-Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility -eek?
-I don't think I can say this--.
-We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
-With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
-
-In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
-@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
-
-Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-
-Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
-Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-
-And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
-However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
-Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
-
-Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
-Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
-However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
-A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
-How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
--->
-```
-
-```{=html}
-<!--
-The approach is based on partition functions from statistical mechanics (the physics of gases), not Newtonian gravitation.
-In short, the approach accepts that there are macro, meso, and micro states.
-In the case of 'trips' being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
-Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
-From this approach, the SIM can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
-There is then a most probable distribution of trips between zones (meso-states) that satisfies all the macro-state constraints.
-
--->
-```
+Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+However, as outlined in preceding sections, we argue the resulting accessibility values are plagued by similar interpretability issues.
+Of the literature in this stream (i.e., accessibility works inspired by @hansen1959 and who cite @wilson1971), only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as method for maintaining constraints on opportunities within the context of competitive accessibility.
+These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
+This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region.
+However, these balancing factors were deduced intuitively.
+These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly-constrained model (derived from entropy maximization), as will be demonstrated in this study.
+They also do not discuss other constrained cases.
 
 # A family of accessibility measures
+
+In this section, we introduce a family of accessibility measures, using Wilson's spatial interaction modelling framework.
 
@@
 This may explain why the constraints/proportionality constant(s) of spatial interaction models did not cross over to accessibility analysis.
-This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p.
+This is intriguing since Wilson made an effort to connect developments in spatial interaction modelling to accessibility, noting for instance, the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of balancing factor $A_i$ [@wilson1971 p.
 10]:
-<!-- "offering a potential interpretation for the proportionality constant...", which would be..... -->
 
@@
 
-This, however, only helps to illustrate the ease with which confusion may arise in this context.
-The relationship in @eq-Ai-as-accessibility appears to be misaligned.
-In the preceding section on early research, we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the pairwise population potential $V_1 = G\frac{M_2}{d_{12}}$ and from there to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
-While the inverse of $A_i$ is pretty much identical to Hansen's accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
+Understanding $S_i$ as the inverse of $A_i$ offers a potential interpretation for the balancing factor, however, $S_i$ itself is not unclear.
+Also, the relationship in @eq-Ai-as-accessibility appears to be misaligned.
+In the preceding Section \@ref(grav-to-today), we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
+While the inverse of Wilson's $A_i$ is pretty much identical to Hansen's accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
 
 Hence, we propose a more general definition of accessibility as a partial sum of the spatial interaction.
-Once we bring back Wilson's proportionality constant into the picture, we can define the *potential for spatial interaction* between two locations i and j is as follows:
+Once we bring back Wilson's proportionality constant into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
 
@@
 
-\noindent where $V_{ij}$ is the potential for interaction, and therefore the accessibility from origin i is defined as:
+\noindent where $V_{ij}$ is the potential for interaction, and therefore the accessibility from origin $i$ is defined as:
 
@@
 
-In the following subsections, we detail what are the implications of bringing this constraint into accessibility measures, and how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus.
-Together, these constraints delineate different members of the family of accessibility measures and which fall under four categories: (1) the unconstrained case, which is equivalent to @hansen1959's formulation; (2) the total constrained cases, which is constrained by total population and opportunities.; (3) the singly-constrained cases, which is opportunity-constrained or population-constrained; and (4) the doubly-constrained case, which is constrained simultaneously by population and opportunities.
-In the next section, we use a simple numerical example to illustrate these cases in increasing order of constraint restrictiveness and information availability.
-
-<!-- confusing terms.
-what is the difference between total constrained and doubly-constrained case ?
-we probably need to think of better terms here.
-I also liked Chris' idea of including a table that summarizes the differences between these measures -->
+In the following subsections, we detail the implications of introducing the constraint $k$ into accessibility measures and how different _constrained_ accessibility measures can be defined based on the focus of measurement and the input population and opportunity data.
+As an overview, the way each constraint $k$ is defined delineates one of four distinct members of the constrained accessibility measure family:
+
+1.
+The unconstrained case.
+This case is equivalent to @hansen1959's formulation whereby a balancing factor is neglected, so units of zonal values are simply in 'opportunities-by-travel-impedance-value';
+2. The totally-constrained case; maximum potential. Two variants are introduced. 
+  - The first, the _totally-constrained accessible opportunities measure_, resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
+In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region, requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
+  - The second variant is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _totally-constrained accessible population measure_.
+In this variant each zonal accessibility value is a proportion of the population in the region, requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible;
+3. The singly-constrained case; medium potential, potential is limited by either an opportunity-constraint or population-constraint. Similarly, two variants are introduced. 
+  - The first is _attraction-constrained_ from the perspective of spatial interaction modelling.
+This variant is refereed to as the _singly-constrained accessible opportunities measure_.
+It can also be formulated to resembles @hansen1959's formulation but with an additional term for $k$, a set of balancing factors $B_j$ which ensure opportunities to each zone are allocated based on population and travel impedance regionally and zonally.
+$B_j$ ensures that a balanced proportion of the total opportunities at each destination is allocated based on zonal population and travel impedance, so that each zonal accessibility value reflects both a share of the total opportunities in the region _and_ is a sum of a _balanced_ proportions of opportunities allocated from each destinations.
+Similar to the totally-constrained case, each zonal accessibility value is expressed in units of 'opportunities' accessible.
+But unlike the totally-constrained case, the singly-constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
+This variant is mathematically equivalent to spatial availability [@soukhovIntroducingSpatialAvailability2023] and its per capita version is equivalent to the 2SFCA [@shenLocationCharacteristicsInnercity1998; @luo2003].
+  -  The second variant is _production-constrained_ from the perspective of spatial interaction modeling.
+It can also be interpreted as a constrained form of market potential (i.e., the transpose of the singly-constrained variant), incorporating the assumptions of the single constraint.
+Specifically, each zonal accessibility value represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance.
+Each zonal value is expressed in units of 'population' accessible;
+4.
+The doubly-constrained case.
+It is constrained simultaneously by population and opportunities, so each zonal value is expressed in units of 'population-opportunity capacity' that is accessible.
+This case requires the number of opportunities and population to match, e.g., the user must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
+By using this one-to-one matching data, this case's calculated values especially limits the 'potential' to spatially interact compared to the data used in the singly-constrained and totally-constrained cases.
+Instead, the doubly-constrained values reflect the number of predicted interactions between the opportunities and population.
+Effectively, this case is equivalent formulaically as Wilson's doubly-constrained spatial interaction model (i.e., _attraction-production constrained_) and can be understood as predicting a value of access , not accessibility.
+
+In the next section, we illustrate each case using a simple numerical example.
 
@@
 Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
-The population is $W_i^{(1)}$ when used as a proxy for the mass at the origin, and $O_i$ when used as a constraint.
-Similarly, the opportunities are represented by $W_j^{(2)}$ when used as a proxy for the mass at the destination, and $D_j$ when used as a constraint.
-
-[^1]: For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]
+For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025].
+Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
 
@@
 #| label: tbl-small-system-land-use
-#| tbl-cap: "Simple system with three zones. Population is in 10,000 persons and opportunities in number of physicians."
+#| tbl-cap: "Simple system with three zones (ID 1, 2 and 3). Population is in 10,000 persons and opportunities in number of physicians."
 
@@
 
-Secondly, to pair with the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
-From both @tbl-small-system-land-use and @tbl-small-system-cost, it can be interpreted that Zones 1 and 3 are more proximate to each other than to Zone 2.
-Combined, Zone 1 and 3 have the same amount of population as Zone 2 but more than double the opportunities.
-<!--for the readers, what sort of set-up could this reflect practically?
--->
-
 ```{r}
@@
 
-And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region are in @eq-travel-behaviour-scenarios: $$
+From both @tbl-small-system-land-use and @tbl-small-system-cost, Zone 3 and 1 can be interpreted as of an urban core with major healthcare institutions and a healthcare cluster on the edge of the city respectively, each with a moderate residing population (with zone 3 having both a higher population and physician count).
+Zone 2, can be interpreted as a more distant bedroom community, with a relatively high population and fewer physicians.
+In sum, Zones 1 and 3 are more proximate to each other than to Zone 2, and together match Zone 2’s population while offering more than twice the physician availability.
+
+And lastly, in @eq-travel-behaviour-scenarios we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities.
+Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
+A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility.
+Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
+As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.
+
+$$
 \begin{array}{l}
@@
 
-<!-- this paragraph could probably be merged with the previous one -->
-
-Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
-A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility.
-Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
-As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.
-
-Any set of concepts representing population, opportunities, and their associated travel behaviour -whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups- can be substituted into our simple example, depending on the research question.
-The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behavior groups and members of the accessibility family.
+Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups, can be substituted into our simple example, depending on the research question.
+The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behaviour groups and cases of the family of accessibility neasures.
 
@@
 
-Setting the balancing factor $k$ to one in @eq-access-01 results in the unconstrained accessibility case:
+Setting the balancing factor $k$ to 1 or omitting it completely in @eq-access-01 results in the unconstrained accessibility case:
 
@@
 
-In such case, the the sum of the accessibility from each origin $V^0_{i}$ in the region generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$).
-This is because setting $k$ to 1 to arbitrarily eliminates the constant term and imposes no meaningful interpretation to the resulting values.
-To complicate matters, when comparing $V^0_{i}$ values between travel behaviour contexts, their difference depends on the choice of impedance function $f(c_{ij})$, associated parameters, and the number of zones for which an accessibility score is calculated.
-<!-- double check this next sentence .
-there should be a more explicit mention to the units problem ?
---> For this reason, the unconstrained accessibility is not comparable across different contexts and decay functions, and should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
-
-Returning to our numeric example, consider the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in @tbl-simple-example-unconstrained-accessibility.
+The sum of the unconstrained accessibility values for each origin $V^0_{i}$ generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), since arbitrarily setting $k$ to 1 (or neglecting $k$ all together) strips the values of any meaningful unit-based interpretation, the units are in 'summed opportunities by some travel impedance value'.
+Moreover, comparisons of these $V^0_{i}$ values across different contexts such as different impedance functions $f(c_{ij})$ or varying number of zones  exacerbates this issue as the units between $V^0_{i}$ values change i.e., comparisons between a value of 'summed opportunities by a travel impedance value' to a value of 'summed opportunities by another travel impedance value' are not directly interpretable.
+From this perspective, the raw unconstrained accessibility scores are not intuitively comparable across different contexts and decay functions.
+They should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
+
+Returning to our numeric example, the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in displayed in @tbl-simple-example-unconstrained-accessibility.
 
@@
   fmt_number(decimals = 3) |>
-  cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
-             V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
-             V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
+  # cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
+  #            V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
+  #            V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
+  #            .fn = md)  |>
+  cols_label(V_unc_i_1 = md("units: *physicians-minute^-3*"),
+             V_unc_i_2 = md("units: *physicians-minute^-2*"),
+             V_unc_i_3 = md("units: *physicians-minute^-0.1*"),
              .fn = md)  |>
-  # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-  #            V_unc_i_2 = "{{f_2 (c_ij )  = 1/c[_ij^2]}}<br>*physicians min^-2*",
-  #            V_unc_i_3 = "{{f_3 (c_ij )  = 1/c[_ij^0.1]}}<br>*physicians min^-0.1*",
-  #            .fn = md)  |>
   cols_align(align = "center",
@@
 
-As the different impedance functions represent different travel behaviours, comparing unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
+As the different impedance functions represent different travel behaviours, comparing the raw unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
 For instance, at zone 1 the  difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
-These two values have different units so this difference value cannot be meaningfully interpreted.
-
-<!-- not like terms? -->
-
-However, we can compare values within the same travel behavior scenario across different zones but values are not inherently meaningful due to the units.
-For example, considering the most decaying scenario at zone 1, it captures `r unc_acc$V_unc_i_1[1] - unc_acc$V_unc_i_1[2]` more $f_1(c_{ij})$-weighted-physicians than zone 2, and `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer than zone 3.
-In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3.
-Both scenarios at zone 1 captures an intermediate accessibility value $V_i^0$ between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians.
-The ratios between the zones in the two scenarios are different.
-These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared.
-<!--maybe a better way of saying this?--> This simple example illustrates that the differences and ratios of unconstrained accessibility values are not inherently meaningful.
-While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select comparable impedance functions (at the cost of accurately reflecting travel behavior), such adjustments may introduce bias.
-
-<!-- Notice the wild differences.
-Accessibility according to impedance function 3 basically says that everyone has access to all opportunities all at once.
-Accessibility according to function 1, on the other hand, says that each population has accessibility to around 3 opportunities, which is absurd: the number of opportunities available locally are in the order of 150-180.
-One should not have to adjust the impedance function to obtain "reasonable" results [EXAMPLE]; instead, the impedance function should reflect the travel behavior.
--->
-
-<!-- CDH: I am not sold on the problem here, yes the values are tied to the impedance, one could use a calibrated impedance...
-but still meaningful comparisons on ordinal scale despite critiques of ratios - i would not expect ratios to be the same...
-need to better draw out in comparison to below?
--->
+These two values are a product of different impedance functions, making the comparison uninterpretable in absolute terms.
+Likewise, we could compare values within the same travel behaviour scenario across different zones, as they are in the same units, however the issue of unit interpretability will also be apparent.
+Considering the most decaying scenario $f_1(c_{ij})$ and zone 1 (the zone with a healthcare cluster at the edge of the city): zone 1 captures `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer physicians-minute$^{-1}$ than zone 3 (urban core).
+Again, the fundamental uninterpretability of what is a _physicians-minute$^{-1}$_ or _opportunity-weighted-travel-impedance_ unit remains.
+
+While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select impedance functions to facilitate comparison across scenarios (potentially at the expense of accurately reflecting travel behavior), such adjustments may introduce bias.
+Hence, the raw unconstrained accessibility values themselves are challenging to compare due to their units.
+To enable more meaningful comparison, the following sections will detail the introduction of constraints to ensure consistent units across scenarios and demonstrate results on the same numeric example.
 
@@
 
-A total constraint proportionally adjusts zonal accessibility values based on the total population and/or opportunities in the region.
-The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case.
-We define two cases: (a) a case where accessibility is constrained by the total number of opportunities (aka *totally-constrained opportunity*) and which is interpreted as Hansen's accessibility with a constraint, and (b) a case where accessibility is constrained by the total population (aka *totally-constrained population*) and which is interpreted as constrained market potential.
-
-### Totally-constrained opportunity: Hansen's accessibility with a constraint
-
-Unlike in @eq-access-01, the proportionality constant $k$ is retained, not set arbitrarily to 1, and represented as $K^T$:
+In the totally-constrained accessibility case, a total constraint proportionally adjusts unconstrained zonal accessibility values $V^0_i$ based on the regional sum of $V^0_i$ and the total population or opportunities in the region.
+Alternatively, we reformulate this case using a proportional allocation constant, which allocates opportunities (or population) proportionally based on the the travel impedance and the total population or opportunities in the region.
+In both formulations, all zonal values become a proportion of a known system total, be it the regional opportunities or regional population depending on the variant.
+
+We define two variants for this case: (a) case where accessibility is constrained by the total number of opportunities (totally-constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraint, and (b) a variant where $i$ and $j$ of the totally-constrained accessibility measure is transposed (totally-constrained accessible population) hence is constrained by the total number of population and interpreted as constrained 'market potential'.
+
+### Totally-constrained accessible opportunities: Hansen's accessibility with a constraint
+
+Unlike in @eq-access-01, the proportionality constant $k$ is retained. For the totally-constrained case, it is represented as $K^T$:
 
@@
 
-Where the total opportunity constrained accessibility measure now becomes Hansen's accessibility with a proportionality constant:
+In this way, the totally-constrained accessibility measure now becomes Hansen's accessibility with a proportionality constant $K^T$:
 
@@
 
-The constraint that we impose in this case is the total number of opportunities $D$ in the region.
-
+Imagine that the only system known is the total number of opportunities $D$ in the region.
+Accordingly, the constraint we impose in this case ensures the regional sum of totally-constrained accessibility is equal to the total number of opportunities as follows:
 $$
@@
 
-<!-- rafa: do we need to keep equations 31 and 32 -->
-
-This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework.
-We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $K^T$ to yield: $$
+<!-- rafa: do we need to keep equations 31 and 32. AS: I think yes. -->
+
+This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel, congruent with Wilson's framework.
+Given the total number of opportunities in the region, we can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility to solve for $K^T$:
+
+$$
 K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
@@
 Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
-Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our totally-constrained opportunity accessibility model is:
+Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our totally-constrained accessibility model is:
 
@@
 
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term, the proportional allocation factor for the totally-constrained opportunity case $\kappa^T$ is dimensionless:
+Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the totally-constrained opportunity case $\kappa^T$ is dimensionless:
 
@@
 \noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$).
-The role of $\kappa^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region.
+The role of $\kappa^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
 Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-For this reason, we refer to it as a proportional allocation factor.
-
-<!-- kappa in equations now? -->
-
-<!-- rafa: 'The role of $\kappa^T$ in this reformulation of accessibility is to..'.
-This sentence is gold and it should be said earlier on.
-My personal preference is that first we need to explain to the reader in plain language what the constraint is, and what it means in practical terms (why it matters).
-Then we can show this with the equations.
--->
-
-Referring to our simple numeric example, proportionality constant for the $f_1(c_{ij}) = 1/c_{ij}^3$ travel behaviour scenario would then be: $$
+For this reason, we refer to $\kappa^T$ as a proportional allocation factor.
+On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the totally-constrained accessibility measure.
+
+Referring back to our simple numeric example, proportionality constant $K^T$ for the most decay travel behaviour scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
+
+$$
 \begin{array}{l}
@@
 
-Using the calculated proportionality constants for all zones, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
+Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
+$\kappa^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
 
@@
 
-In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
+In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
 Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa^T$.
 
-Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`).
-However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
+Considering the highest decay scenario ($f_1(c_{ij})$), zone 1 (a healthcare cluster at the edge of the city) captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`) like in the unconstrained accessibility case.
+However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
 Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
@@
 Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
-
-<!-- rafa: This is a very important takeaway. But this is true for any metric using the constant, right? 66666666 -->
-
-### Totally-constrained population: market potential
-
-<!-- need to improve flow here, why are we jumping into market potential, a bit of duplication in the effectively transpose part -->
-
-By transposing $i$ and $j$, we have an  accessibility measure constrained by total population (@eq-total-constrained-market), which effectively gives us an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
-The unconstrained form of market potential, effectively the $i$ $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
-To formulate this case, the total constraint may instead be applied to the mass of the *population* at $i$ instead of the opportunities at $j$.
+This is true for other constrained cases of the family of accessibility measures.
+
+### Totally-constrained accessible population: totally-constrained market potential
+
+Another variant of the totally-constrained accessibility measure is the _totally-constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the totally-constrained accessible opportunities measure.
+This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
+This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
+Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination.
+However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form.
+To address this, the constrained variant, the totally-constrained accessible population measure $M^T_{j}$, is introduced.
+To formulate this variant, the total constraint $K^T $ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
 
@@
 M_j^0 = W_i^{(1)} f(c_{ij})
-$$
-
-The totally-constrained population accessibility (market potential) becomes:
+$$ {#eq-unconstrained-market}
+
+The totally-constrained accessible population measure (market potential) then becomes:
 
@@
 $$
-This measure is simply a form of passive accessibility, and indicates that each destination can be reached by an x number of people.
-The constraint that we impose in this case in practice means that the total market potential equals the total population $O$ in the region:
+
+Where we impose the total system known as a constraint, i.e., that the total market potential equals the total population $O$ in the region:
 
@@
 $$
-M^T_i = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$ Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless: $$
+M^T_j = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+$$
+
+Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
+
+$$
 \hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
@@
 
-Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility.
-Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility).
-In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
-Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
-However, similar to the totally-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
-As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
+Returning back to the numerical example, the proportionality constant $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
 
@@
 
+Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility).
+In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
+Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
+However, similar to the totally-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
+As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
+Overall: like the totally-constrained accessible opportunities case, this variant allows for the interpretation of both ordinal and interval comparisons of the raw values themselves.
+
 ## Singly-constrained accessibility
 
-Increasing in restrictiveness relative to the totally-constrained accessibility measure, a single constraint proportionally adjusts the origin-destination zonal accessibility values based on the population or opportunities in the zone for all zones, while still maintaining the total system population or opportunity constraint, depending on the case.
-We define two cases: a *singly-constrained opportunity* accessibility case and a *singly-constrained population* accessibility case.
-
-### Singly-constrained opportunity: A.K.A spatial availability
-
-If additional information is introduced into our analysis, say, the number of opportunities *by destination* (i.e., $D_j$), we can impose the following constraint (@eq-opportunity-constraint)- comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
+Similar to the totally constrained accessibility measure, the singly-constrained case includes a constraint that proportionally adjusts the unconstrained zonal accessibility values $V^0_i$.
+Two variants are defined: (a) the _singly-constrained accessible opportunities_  case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly-constrained accessible population_ case, its transpose, or singly-constrained market potential.
+
+Unlike the total constraint, the single constraint incorporates additional information.
+In the opportunities-accessible variant, the balancing factor constrains the 'potential' for spatial interaction by ensuring that only a proportional amount of opportunities at each destination are allocated to 'demanding' origins.
+This allocation is informed by demand, or the relative amount of population, and associated travel impedance connecting the zones.
+In the population-accessible variant (the transpose), population at each origin is proportionally allocated to destinations based on the share of opportunities and travel impedance.
+
+In both variants, the balancing factor allocates opportunities (or population) based on the relevant supply-demand weights of the origins and destinations, and the travel impedance between zones.
+The singly-constrained accessibility measure introduces population-based (or opportunity-based) competition at the zonal level, unlike the total constraint, which more simply allocates a fixed regional total of opportunities (or population depending on the variant).
+
+In sum, all singly-constrained zonal accessibility values are both a proportion of the known regional opportunity (or population) total _and_  a sum of a _balanced_ proportion of opportunities allocated from each destinations (or population allocated from each origin).
+Each zonal value remains in units of opportunities accessible (or population accessible), but in contrast to the totally constrained case, the singly constrained measures explicitly incorporates both spatial weights (i.e., population and opportunities) into the allocation process.
+
+### Singly-constrained accessible opportunities: spatial availability
+
+To demonstrate this variant formulaically, we begin with the opportunity constraint (@eq-opportunity-constraint) and our known piece of information, i.e., the sum of accessible opportunities from a destination should equal the number of opportunities $D_j$ at that destination.
+As the number of opportunities at each $j$ are known, it is represented as $D_j$ instead of $W_j^(2)$ as in the totally-constrained case.
+This constraint should hold for all destinations in the region.
+This is comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
 
@@
 
-After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as an origin summary expression of proportionally allocated known opportunities (i.e., $D_j$):
+After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as an origin summary expression of proportionally allocated known opportunities (i.e., $D_j$).
 
@@
 
-Imposing the single constraint $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values like previously discussed in the case of the totally-constrained opportunity case.
-Namely, when the single constraint is applied to the opportunities at $j$, each resulting accessibility value at $i$ is in units of physicians, with the impedance units already accounted for.
-
-However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ captures competition from the population.
-Again, let's consider the highest decay scenario $f_1(c_{ij})$.
-In this case, zone 1 no longer captures an intermediate amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
-Why may zone 1 capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
+Imposing the single constraint $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the totally-constrained accessible opportunities case.
+Namely, when the single constraint is applied to the opportunities at $j$, each resulting accessibility value at $i$ is in units of _physicians_, with the impedance units already accounted for.
+
+However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ reflects zonal competition based on the mass of the origin, i.e., population.
+Again, consider the highest decay scenario $f_1(c_{ij})$.
+Under this scenario, zone 1 no longer captures a medium amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
+Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
+
 This difference is due to the single-opportunity constraint $\kappa^S_{i}$'s role in $V_i^S$.
-The only inputs needed in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
-In other words, opportunities are proportionally allocated to each $i$ based on the proportion of impedance-weighted opportunities at all $j$ for that $i$ to the total sum of impedance-weighted opportunities in the entire region.
+The only inputs required in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
 Opportunities are allocated to $i$s, regardless of the mass weights of origin.
 Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input.
-In fact, $\kappa^S_{i}$ is the proportion of the sum of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
-Hence, since zone 1 has the lowest number of people in the region, and is not located -travel cost effectively- to other opportunities, $V_{1}^S$ is lowest.
-
-<!-- check sentence ^ -->
-
-<!-- rafa: I think we don't need the paragraph below -->
-
-From a different perspective: $V_{2}^S$ is intermediate in rank when considering the highest decay scenario ($f_1(c_{ij})$) but is the highest ranked zone under $f_3(c_{ij})$, from `r opp_acc$V_opp_i_1[2]` to `r opp_acc$V_opp_i_3[2]` accessible physicians out of `r opp_acc$V_opp_i_3|>sum()` in the region; a `r opp_acc$V_opp_i_3[2]/opp_acc$V_opp_i_1[2]` times increase.
-Zone 2 has the highest amount of population and the least travel-impedance proximate to other zones.
-So when the the decay is high - it can most effectively capture opportunities at Zone 2, in fact it captures ($\kappa_2 ^T$) is `r opp_acc_with_ks$k_opp_ij_1[2]`, or `r opp_acc_with_ks$k_opp_ij_1[2] |> scales::percent()` of the opportunities in the region, with the vast majority coming from its own zone (`r (opp_acc_ij$k_opp_ij_1[5]/opp_acc_with_ks$k_opp_ij_1[2])|>scales::percent()` out of the total opportunities that are allocated to Zone 2).
-This is unlike other zones, that capture opportunities from their own zones at lower rates (i.e., ($\kappa_{1}^S$ corresponds to a capture of `r (opp_acc_ij$k_opp_ij_1[1]/opp_acc_with_ks$k_opp_ij_1[1])|> scales::percent()` of its allocation from Zone 1 and $\kappa_{3}^S$ corresponds to `r (opp_acc_ij$k_opp_ij_1[9]/opp_acc_with_ks$k_opp_ij_1[3])|> scales::percent()` of its allocation from Zone 3)).
-
-<!-- rafa: I think we don't need the paragraph below -->
-
-In scenarios with low decay (e.g., $f_3(c_{ij})$), the connection between opportunities and population changes: spatial impedance among zones is more equal - zonal populations can get to zonal opportunities at relatively more similar travel costs.
-In this sense, spatial impedance has less relative impact on the effect of zonal population.
-For this reason, under the low decay scenario of $f_3(c_{ij})$, Zone 2 's competitive population advantage is more evident.
-Zone 1 captures more opportunities from Zones 1 and 3 than Zone 1 and 3 capture themselves (i.e., Zone 2 captures `r (opp_acc_ij$k_opp_ij_3[2])|> scales::percent()` from both Zone 1 and 3, while Zone 1 and Zone 3 only captures `r (opp_acc_ij$k_opp_ij_3[1])|> scales::percent()` and `r (opp_acc_ij$k_opp_ij_3[9])|> scales::percent()` from itself respectively.
-
-Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel scenarios.
-Again, returning to Zone 2 as an example, while its competitive population advantage is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios: this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
+In fact, $\kappa^S_{i}$ is  calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
+Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.
+
+Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
+For instance, consider Zone 2 which has the highest population.
+It is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios that this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
 In this sense, other zones are not *that* disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
@@
 
-In this way, the consideration of accessibility *per capita* may be clarifying.
+Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
+For instance, Zone 2 has the highest zonal population, representing 50% of the 200,000 regional population.
+However, due to its high relative travel distance from the other zones, its population is less competitive in capturing opportunities in the high-decay travel scenario ($f_1(c_{ij})$): Under this scenario, zone 2 captures almost exclusively opportunities from its own zone.
+However, in $f_3(c_{ij})$, the scenario with unfettered travel cost, Zone 2 captures by far the most number of physicians.
+But in this scenario, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
+Zones 1 and 3 are allocated opportunities at rates similar to their relative population size.
+
+In this way, the consideration of constrained accessibility *per capita* may be clarifying.
 Often, accessibility values are reported as raw scores without the consideration for population.
@@
 In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
-These per capita rates are equivalent to the 2SFCA values.
+Notabely, these per capita rates are equivalent to the 2SFCA values.
 
@@
 
-<!-- check wording in paragraph below, few issues -->
-
-Our simple example was designed such that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
-As the decay decreases and the effect of population is more independent: per capita values begin to stabilise to this regional per capita average, similar to how the accessibility values in the totally-constrained opportunity case stabilizes to the $V_i^T$ regional average.
-These responses make intuitive sense: the use of constraints are an regional and/or zonal averaging methodology.
-By equalising the impact of travel impedance, the impact of the remaining constraint variable remains.
-In the case of the total opportunity constraint, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constraint, this is the population at a zone relative to the regional population.
-
-<!-- rafa: in the last sentence, you clarify the difference between 'total opportunity constraint' Vs  'single opportunity constraint' .
-What is confusing here is how the single opportunity constraint compares to 'total population constraint' .
-Again, the idea of including a table would be very helpful.
-The table shoud include one indicator in each row.
-The columns could include (1) name of the indicator, (2) explanation in plain language of the constraint, (3) equation of the constraint, (4) equation of the indicator, (5) example of the interpretation of the indicator -->
-
-Like the totally-constrained opportunity case, this case offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.
-Building on this, the singly-constrained opportunity case also offers a way to consider origin-side competition while constraining the opportunities to match the input number of regional opportunities and zonal opportunity totals.
-This case builds on the previous intuition of totally-constrained opportunity case by providing additional insights from the proportional allocation factor and per capita considerations (that are mathematically equivalent to the 2SFCA, but with additional intuition).
-
-### Singly-constrained population: market potential
+This simple example was constructed so that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
+As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
+Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all around 24 physicans accessible per capita).
+
+This trend mirrors how the accessibility values in the totally-constrained opportunity case stabilizes to the $V_i^T$ regional average (e.g., the accessible opportunities allocated to each of the three zones approaches a third of 490 physicians, or 163.33, under the unfettered mobility scenario $f_3(c_{ij})$).
+These patterns make intuitive sense: constraints act as regional and/or zonal averaging mechanisms.
+In distance decay travel behaviour scenarios that are uniform (i.e., low for all like in $f_3(c_{ij})$), what remains is the relative effect of the other variables in the constraint variable.
+In the totally constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constraint, this is the population at a zone relative to the regional population.
+
+### Singly-constrained accessible population: singly-constrained market potential
 
@@
 In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
-<!-- Anything else like this in the literature?-->
-
-The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^S_i$ becomes:
+
+The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^S_j$ becomes:
 
@@
 
-Following the same logic as in the preceding section on totally-constrained population accessibility (market potential), one arrives at the following expression: $$
-M^S_{j} = \sum_i \hat \kappa^S_{ij} O_i
+Following the same logic as in the preceding section on totally-constrained market potential, one arrives at the following expression:
+
+$$
+M^S_{j} = \hat \kappa^S_{j} \sum_i  O_i
 $$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}
@@
 $$
-\kappa^S _{ij} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+\kappa^S _{j} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
 $$ {#eq-attraction-constrained-proportional-allocation-factor}
@@
 
-The accessibility measures above have used either $O_i$, $D_j$, or the regional sums of either, but not both.
-When $D_j$, the opportunities, was used to constrain @eq-opportunity-constrained-accessibility, the mass of the population at the origin was given by $W_i^{(1)}$ (also in @eq-opportunity-constrained-proportionality-constants).
-When $O_i$, the population, was used as a constraint in @eq-population-constrained-accessibility, the mass at the destination was given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants).
-The reason is that the population and the opportunities are typically different things without a common metric.
-For example, on the side of the population we usually have people, but on the side of the opportunities we can have physicians, clinics, grocery stores, green space, schools, libraries, and so on.
-There are a few cases where there might be a one-to-one relationship between the population and the opportunities, with the most commonly studied opportunity in the competitive accessibility literature being employment, e.g., one person with one job.
-Another possible opportunity type is healthcare, e.g., one person and one unit of capacity.
-
-In these latter cases, a doubly-constrained approach can be implemented when one-to-one relationships between population and opportunities are present.
-This is required because the production- and attraction- constraints in Wilson's doubly-constrained case (@eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel) are imposed simultaneously.
-The accessibility-specific forms are used in the singly-constrained opportunities and population cases in the preceding section (@eq-opportunity-constraint and @eq-population-constraint).
-To simultaneously impose both population- and destination- constraints, they must match (@eq-opportunity-population-equality):
+This accessibility case requires zonal populations and opportunities to match one-to-one, like the doubly-constrained spatial interaction model.
+In this way, doubly-constrained accessibility can be thought as "access", or the realized potential for spatial interaction.
+
+To contextualize this point, the totally- and singly-constrained accessibility measures discussed thus far have used either $O_i$, $D_j$, or the regional sums of either, but never both simultaneously.
+For example, when opportunities $D_j$ are used to constrain @eq-opportunity-constrained-accessibility in the _singly-constrained accessible opportunities measure_, the specific mass of the population at origin $i$ demanding only those opportunities at that $j$ is unknown.
+Instead, only the population at $i$ demanding opportunities in the region is known, and this is more generally represented as $W_i^{(1)}$ (as in @eq-opportunity-constrained-proportionality-constants).
+Similarly, when the population $O_i$ is used as a constraint in @eq-population-constrained-accessibility in the _singly-constrained accessible population measure_, the mass at the destination is given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants) since only the mass of opportunities at each $j$ is known and information on what opportunities are allocated to what zone is unknown.
+
+By contrast, the double-constrained accessibility case requires that both populations and opportunities match.
+Meaning, opportunities at each destination can be accessed by all populations, while at the same time, the population at each origin can be accessed by all opportunities.
+However, this requirement is often unintuitive in traditional accessibility analysis.
+Namely, the distinction between population (origin masses) and opportunities (destination masses) typically represent different entities without a shared unit of measurement.
+On the population side, we usually count people; on the opportunity side, we may be referring to physicians, clinics, grocery stores, schools, parks, or libraries.
+In a few cases, a one-to-one correspondence or their own capacities at which they interact and are interacted with may exist e.g., one person with one job.
+Another similar opportunity type example is healthcare, e.g., one person and one unit of capacity, e.g.
+a vaccine shot.
+
+A doubly-constrained approach to accessibility calculation needs a one-to-one relationships between population and opportunities to be present.
+Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly-constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of destinations in all destinations (@eq-opportunity-population-equality):
 
@@
 
-Like before, the imposition of these two constraints implies that the total system constraint is maintained e.g., the sum of all doubly-constrained accessibility $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ in @eq-total-constraint-accessibility is also maintained, and also equal to the total opportunities in the region $O$.
-
-To explicitly define the underlying accessibility measure in this case, it follows that of the production-attraction constrained (doubly-constrained) spatial interaction model in @eq-doubly-constrained-gravitymodel, hence our accessibility measure becomes:
+As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly-constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$ as shown in @eq-total-constraint-accessibility.
+
+As the doubly-constrained accessibility measure $V_{i}^D$ takes the form of the production-attraction (doubly-constrained) spatial interaction model, as shown in @eq-doubly-constrained-gravitymodel, $V_{i}^D$ is as follows:
 
@@
 
-\noindent where the corresponding balancing factors, as per Wilson, are:
+\noindent where the corresponding balancing factors $A_i$ and $B_j$, as per Wilson, are:
 
@@
 193-195].
-
-The proportional allocation factor $\kappa_{ij}^D$ would then be: $$
+The proportional allocation factor $\kappa_{ij}^D$ would then be:
+
+$$
 \kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
@@
 
-Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the the value *already* matches population to opportunities, e.g., the number of opportunities accessed by the population.
+Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the value *already* matches population to opportunities, e.g., the number of opportunities accessed by the population.
 Similarly, the market potential form $M^D_j$ would be the transposed way of summarizing $V_{ij}^D$.
-In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both <!-- is it "both" or "other" - word was "bother" --> opportunities and population.
-As opportunities match population and both marginals are constrained, $M^D_j$ would simply be a summary of $V^D_{ij}$ values at each $j$, and would represent the number of people accessible to opportunities.
+In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both opportunities and population.
+As opportunities match population and both marginals are constrained, $M^D_j$ would be a summary of $V^D_{ij}$ values at each $j$, and would represent the number of people accessible to opportunities.
 The formulation is as follows:
@@
 
-Let's return to the numeric example with a new objective.
-Say, we're interested in measuring the zonal values of accessed physician-capacity.
-And we have new information: given the same zonal populations, a region-wide survey indicates the number of people that live in each zone, and the number of physician capacity per zone.
-The survey does not tell use what person is matched with what physician (assume that's confidential), we only have the numbers of population at $i$ and the number of physician-capacity at $j$.
-Lucky for us, this survey keeps our example simple: we find out that the physician-capacity at each zone is a approximately scaled version of the number of destination-side physicians at each zone.
-With this new information, we are interested in understanding what does the zonal access to physician-capacity for the population looks under different travel behaviour scenarios.
-@tbl-small-system-land-use-doubly-constrained-case summarises this adjusted simple example: with the population (in units of 10,000s of people) and the opportunities (in units of 10,00s of physician-capacity) per zone <!-- check -->.
-Considering this new definition of 'provider', the new (realized) Provider-to-Population-Ratio is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`.
-The zonal cost matrix, and travel impedance functions for three types of travel behaviour are the same as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
+To calculate doubly-constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be modified.
+Namely, a count of physician _capacity_ per destination is needed instead of just the number of physicians, as used to calculated totally- and singly-constrained cases.
+We also must be able to clearly state that the population is a count of people seeking opportunities at the new capacities, i.e., the population must reflect the _capacity_ of the population to interact with opportunities.
+
+So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+For the population, we leave this unchanged numerically but theoretically know that each person interacts with one physician capacity (i.e., our opportunities with a _capacity_).
+Hence, the number of opportunities per destination is new: the example is modified such that the physician-capacity at each zone is an approximately scaled version of the number of destination-side physicians at each zone from the unmodified example (@tbl-small-system-land-use).
+To emphasis the new definition of 'provider' as physician capacity,  the new system PPR is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
+We keep the same zonal cost matrix, and travel impedance functions for three types of travel behaviour as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
 
@@
 
-In this case, $V_{ij}^D$ for the most decay travel behaviour scenario is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
+However, despite the modifications to the example, our objective remains the same as in previous cases: to measure accessibility under different travel behavior scenarios.
+Specifically, we aim to quantify the number of _potential_ spatial interactions between the physician-seeking population in each zone $i$ to the physician capacity in a zone $j$ i.e., $V_{ij}^D$.
+The highest decay travel behaviour scenario ($f_1(c_ij)$) is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
 
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
-#| tbl-cap: "Adjusted simple system: doubly-constrained opportunity accessibility for the most decay travel behaviour scenario."
+#| tbl-cap: "Doubly-constrained opportunity accessibility assuming the most decay travel behaviour scenario in the modified simple system."
 
@@
 
-As mentioned, accessibility can be interpreted as a summary of the proportionally allocated opportunities at each $i$.
-In interpreting the doubly-constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$, these values would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity for Zones 1, 2 and 3 respectively, approximately equal to the number of population at each of these zones.
-Conversely, $M_j^D$ values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of *opportunities* (physician-capacities) at each of these zones.
-And that's precisely the function of the double constraint - to have the mass weight at the origin equal the mass weight of the destination.
-In other words, $V_i^D$ is simply accessed opportunities and $M_j^D$ population accessed.
-And following the same logic, the same $V_i^D$ and $M_j^D$ values are calculated for the other two travel behaviour scenarios.
-
-What may be interesting are the differences in $V_{ij}^D$ values between travel behaviour scenarios, these values can be directly compared to discuss their mass and distance decay impacts.
-Returning to Zone 2, @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for our more relatively remote, higher-populated and lower-opportunity rich zone.
+As mentioned, accessibility values are typically interpreted as a summary of the proportionally allocated opportunities at each $i$.
+Hence, in interpreting the doubly-constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$ values (i.e., the sum of values at all three $j$ destinations for each origin $i$) ould be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity accessible for Zones 1, 2 and 3 respectively.
+This approximately equal to the number of population at each of these zones.
+Conversely, the market potential $M_j^D$ interpretation of these values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of *opportunities* (physician-capacities accessible) at each of these zones.
+Notice, the mass weight at the origin equals the mass weight of the destination: this is precisely the function of the double constraint.
+In other words, $V_i^D$ is the number of accessed opportunities and $M_j^D$ is the number of population accessed.
+For the other two travel behaviour, identical $V_i^D$ and $M_j^D$ values are calculated, following the same logic.
+Hence, the usefulness of the doubly-constrained measure lies in the interpretation as $V_{ij}^D$ values.
+
+For instance, differences in $V_{ij}^D$ values between travel behaviour scenarios are notable.
+These values can be directly compared to discuss mass and distance decay impacts.
+Examining Zone 2 (the bedroom community), @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for this more relatively remote, higher-populated and lower-opportunity rich zone.
 It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
@@
 
-<!--NOTE FOR FUTURE EXPLORATION: Interestingly...
-for the doubly- and singly- constrained case, kappas are very similar *maybe close to identical?* when decay is low.
-This is because distance difference doesn't matter, everyone can reach everything and everything is given to everyone.
-But when decay is high, they are much different: since in the doubly-constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
-This relationship should be investigated further some other time..-->
-
 ```{r}
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
-#| tbl-cap: "Adjusted simple system: doubly-constrained opportunity accessibility for all travel behaviour scenarios for Zone 2."
+#| tbl-cap: "Doubly-constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."
 
@@
 
-Recall that accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
+Recall, accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
 In the doubly-constrained case, we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
-However, if what interests readers is the "potential for interaction" in the case population and opportunities do match one-to-one, perhaps reframing the investigation is needed.
+However, if what interests readers is the "potential for interaction" in the case population and opportunities to match one-to-one, reframing the investigation may be needed.
 In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ij}^D$).
-In this sense, the doubly-constrained case it better thought of as an estimate of "access" instead of "accessibility".
-It is also formulaically identical to the doubly-constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity' respectively.
+In this sense, the doubly-constrained case can be thought of as an estimate of _realized_ accessibility or access.
+It is also formulaically identical to the doubly-constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity', respectively.
 
 As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
-Accessibility, however, is a summary of *potential* for interaction: it's typically been used and understood as a zonal summary.
-Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained case make the connection between the *potential* for spatial interaction (accessibility) and spatial interaction (access) more clear(?).
-
-```{=html}
-<!--
-NOTE: competitive accessibility Literature has explored 'doubly-constrained' but -- it's not the same as this: they they just use the Wilson balancing factors , etc.
-Ex.
-@hornerExploringMetropolitanAccessibility2004 and @allenMeasureCompetitiveAccess2020.
+Accessibility, however, is often presented and understood as a zonal summary of *potential* for interaction between origins and destinations that contain inherently different units.
+Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
+Namely, by demonstrating that these members of the accessibility family all derive from the same root and can be derived from Wilson's original formulation, this more clearly demonstrates what *potential* is, within the context of spatial interaction modelling.
+
+Namely, *potential* depends on the framing of the masses at the origin and destination, and how similar they are in their units.
+The appropriate constraint should be decided based on the the input data and their similarity.
+In increasing unit similarity from the perspective of opportunity-constraint: the total-constraint can be used if population (origin mass) is not known, the single constraint can be used if population is known and matters to *potential* but it does not match the capacity of opportunities, and lastly, the double constraint is appropriate if population is known, matters and population matches the capacity of opportunities.
+These constraint measures can reflect the opportunities accessible at each zone, but reflects the assumptions embedded in the input data about the potential to spatially interact.
+
+# Conclusions
+
+In this work, we examined the historical and mathematical commonalities between spatial interaction models and place-based accessibility measures.
+As accessibility research evolved largely influenced by @hansen1959, researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
+(1) We have demonstrated that by reintroducing such constraints into accessibility analyses, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values, make comparisons between scenarios and regions more intuitive.
+(2) Also, these constraints, depending on the case (totally-, singly- or doubly- constrained),  restrict the degree of _potential_ to spatially interact calculated, linking access to accessibility on the same continuum based on the constraint used.
+Lastly, we discussed how popular measures such as the one used in @hansen1959 and the 2SFCA link into the family.
+We summarise the family of accessibility measures, as follows:
+<!-- While methodologically clarifying work, such as @wuUnifyingAccess2020, has largely overlooked competition, this paper demonstrates how applying constraints consistent with the spatial interaction model allows for incorporating competition into accessibility.
 -->
-```
-
-<!--ANOTHER NOTE:  -->
-
-# Discussion
-
-## SUMMARY
-
-By introducing constraints, we fold in competition considerations into a family of accessibility measures.
-While previous methodologically clarifying work, such as @wuUnifyingAccess2020, has largely overlooked competition, this paper demonstrates how applying constraints consistent with the spatial interaction model allows for incorporating competition into accessibility.
-The constraints used depends on the research question, the suitability of assumptions, the granularity of data, but ultimately: the addition of constraints re-introduces unit into accessibility, make comparisons between scenarios and regions more intuitive.
-
-This work builds on spatial interaction modeling principles, following a review of earlier spatial interaction literature.
-We show that, while recent accessibility research has diverged from its spatial interaction foundations, it can be explicitly linked back.
-We demonstrate that accessibility can be consistently calculated for an $i$-$j$ pair by stacking constraints that preserve the units of "accessibility" while maintaining zonal or regional constraints.
-
-We first place the popular Hansen accessibility measure within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
+
+We first place the popular Hansen-type accessibility measure within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
 We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
-In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if regional competition is relevant; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, *without* considering the population size at the origins.
+In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if population-competition is not relevant and one is interested in capturing the maximum _potential_; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, *without* considering the population size at the origins.
 <!-- A.S.
@@
 We then introduce the singly-constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
+It can also mathematically equivalent to spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
@@
 We also introduce an 'accessible' PPR (e.g., opportunities per capita), calculated by dividing each accessibility value by the zonal population.
-To clarify, this per capita expression of the singly-constrained case is equivalent to the 2SFCA, hence linking this literature back to spatial interaction principles.
+To clarify, this per capita expression of the singly-constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
+
+<!-- Rafa asks, BFCA worth talking about? -->
 
@@
 Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest.
-In this sense, the research question should be concerned with realized access (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential 'access' (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
-
-## A note on interpretation and possible applications
-
-When to use each of these measures? Embedded assumptions? and dangers of using one over the other?
-
-# Conclusions
-
-Accessibility development has a rich history.
-There have been developments in the cases/breath of travel impedance functions, competition considerations, and in incorporating individual choice as part of activity-based accessibility.
-However, no work yet has examined accessibilities explicit connecting to SIM.
-We examine this connection, and provide a origin-destination ($i$ to $j$) formulation of accessibility as part of the SIM framework.
-Novel stuff.
-
-````{=html}
-<!--
-BELOW IS THE TEXT GRAVEYARD!
-
-# APPENDIX (some musings... destined for the commented-out graveyard)
-
-## Singly-constrained case is spatial availability
-
-The destination-constrained case is also spatial availability, see below!
+In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
+
+## A note on interpretation and possible applications/ summary table?
+
+<!-- Some Table here. AS example:
+(col 1) type of measure, (col 2) explanation in plain language of the constraint, (col 3) continuum of potential (high to low), (4) equation of the constraint, (5) equation of the indicator, (col 6) input data needed?, (col 8)  opp-constrained value and interpretation of value (col 9)  pop-constrained value and interpretation of value -->
+
+<!-- rafa: in the last sentence, you clarify the difference between 'total opportunity constraint' Vs  'single opportunity constraint' .
+What is confusing here is how the single opportunity constraint compares to 'total population constraint' .
+Again, the idea of including a table would be very helpful.
+The table shoud include one indicator in each row.
+The columns could include (1) name of the indicator, (2) explanation in plain language of the constraint, (3) equation of the constraint, (4) equation of the indicator, (5) example of the interpretation of the indicator.-->
+
+unconstrainted: doesn't match the anything.
+
+totally-constrained (opportunity): opportunities at all destinations can be accessed by all zones.
+singly-constrained (opportunity): opportunities at each destination can be accessed by all populations.
+
+totally-constrained (population): population at all origins can be accessed by all zones.
+singly-constrained (population): population at each origin can be accessed by all opportunities.
+
+doubly-constrained: opportunities at each destination can be accessed by all populations, while at the same time, the population at each origin can be accessed by all opportunities.
+i.e., a proportional amount of opportunities are allocated from each dest to origins (based on population demanding and travel impedance) and a proportional amount of population is allocated from each origin to dests (based on the opportunity supply and travel impedance).
+
+# References
+
+<!-- Of note RE DOUBLE CONSTRAINT: the accessibility case methodogy should be based selected based on the input data and the theory.
+For instance, the doubly-constrained accessibility method requires pportunities and population to matched one-to-one both mathematically (i.e., the balancing factors will not converge if they do not match) and theoretically (i.e., each person should seek one opportunity).
+By forcing these assumptions, the resulting accessibility value that is calculated can be interpreted with less 'potential', than in the totally- and singly- constrained cases using the unmodified example.
+This is because, the unmodified example inherently carries more 'potential', i.e., we only know where people are (who can potentially interact) and we know where there are opportunities (who can be potentially interacted with), but not the capacities of which this interaction happens.
+As mentioned earlier, in accessibility work, the capacities of which population and opportunities interact is not often known: in fact, accessibility research is often interested in capturing potential at a less granular level: the _potential_ for a population (whether the have interacted, or have not interacted) to interact with opportunities - the capacity of this interaction is not often known, and not the focus.
+-->
+
+<!-- Using the doubly-constrained measure with the unmodified example is both mathematically impossible (as there is no one-to-one match) but also theoretically uncompitable as the capacities at both ends are not known.
+Conversely, the totally- and singly- constrained cases can be used to calculate values using inputs from the modified example, but, the results will be more impresice.
+If capacities of population and opportunities are both known, and the objective of the exercise is to calculate the number of _capacity_ of the opportunity (or population) that is accessible, than by using the singly-constrained case, capacities at  either the opportunity or population would be neglected.
+In using the totally-constrained case, capacities at both would be neglected.
+In this way, if the goal was to calculated acccessibility to opportunity or population capacity, the answer would be only a more imprecise estimation than if the doubly-constrained method was used.
+-->
+
+<!-- BELOW IS THE TEXT GRAVEYARD!  -->
+
+<!-- # APPENDIX (some musings... destined for the commented-out graveyard) -->
+
+<!-- ## Singly-constrained case is spatial availability -->
+
+<!-- The destination-constrained case is also spatial availability, see below!
 As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{j}$ is removed).
-
-Spatial availability is formulated like so:
-$$
-V_{ij}''^{(2)}  = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
-$$
-It turns out that the denominators of the Population balancing factor $F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}$ and Travel cost balancing factor $F^c_{ij} = F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}$ cancel out, so spatial availability can be expressed as:
-$$
-V_{ij}''^{(2)} =O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
-$$
-
-Reformatted in the general family of accessibility measure form, it looks like:
-$$
-V_{ij}''^{(2)} =k O_j f(c_{ij})\\
-V_{ij}''^{(2)} =\frac{ P_i}{\sum_i P_i f(c_{ij})} O_j f(c_{ij})\\
-V_{ij}''^{(2)} =  O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} = O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\
-$$
-
-## Checking singly-constrained case against spatial availability
-```{r}
-#| eval: false
-
-sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){
-
-  o_id <- rlang::enquo(o_id)
-  d_id <- rlang::enquo(d_id)
-  pop <- rlang::enquo(pop)
-  opp <- rlang::enquo(opp)
-  r <- rlang::enquo(r)
-  f <- rlang::enquo(f)
-
-  sum_pop <- x %>%
-    dplyr::distinct(!!o_id,
-                    .keep_all = TRUE) %>%
-    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
-    dplyr::pull(sum_pop) %>%
-    sum()
-
-  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop
-
-  sum_impedance <- x %>%
-    dplyr::group_by(!!d_id) %>%
-    dplyr::summarize(sum_impedance = sum(!!f))
-
-  x <- x %>%
-    dplyr::left_join(sum_impedance,
-                     by = rlang::as_name(d_id))
-
-  f_c <- dplyr::pull(x, !!f) / x$sum_impedance
-
-  x$f_c <- f_c
-  x$f_p <- f_p
-
-  sum_pa <- x %>%
-    dplyr::group_by(!!d_id) %>%
-    dplyr::summarize(sum_pa= sum(f_p * f_c))
-
-  x <- x %>%
-    dplyr::left_join(sum_pa,
-                     by = rlang::as_name(d_id))
-  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)
-
-  dplyr::pull(x, !!opp) * f_t
-}
-```
-
-```{r}
-#| eval: false
-
-df <- data.frame(i = c(1,1,1,2,2,2,3,3,3), 
-                 j= c(1,2,3,1,2,3,1,2,3),
-                 f = c(1/C[1,], 1/C[2,], 1/C[3,]),
-                 pop = c(P_i[1], P_i[1], P_i[1], P_i[2], P_i[2], P_i[2], P_i[3], P_i[3], P_i[3]),
-                 opp = c(O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3]),
-                 catch = c(1,1,1,1,1,1,1,1,1))
-
-calced <- sp_avail(df,
-                         o_id = i,
-                         d_id = j,
-                         pop = pop,
-                         opp = opp,
-                         r = catch,
-                         f = f)
-df$V_ij <- calced
-
-df |> group_by(j) |> summarise(V_j = sum(V_ij))
-```
-
-## Destination- (or ATTRACTION) constrained is the only singly-constrained case... population- (or PRODUCTION) constrained doesn't make sense!
-
-Then the origin-constrained (singly-constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
+-->
+
+<!-- Spatial availability is formulated like so: -->
+<!-- $$ -->
+<!-- V_{ij}''^{(2)}  = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} -->
+<!-- $$ -->
+<!-- It turns out that the denominators of the Population balancing factor $F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}$ and Travel cost balancing factor $F^c_{ij} = F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}$ cancel out, so spatial availability can be expressed as: -->
+<!-- $$ -->
+<!-- V_{ij}''^{(2)} =O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\ -->
+<!-- $$ -->
+
+<!-- Reformatted in the general family of accessibility measure form, it looks like: -->
+<!-- $$ -->
+<!-- V_{ij}''^{(2)} =k O_j f(c_{ij})\\ -->
+<!-- V_{ij}''^{(2)} =\frac{ P_i}{\sum_i P_i f(c_{ij})} O_j f(c_{ij})\\ -->
+<!-- V_{ij}''^{(2)} =  O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} = O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\ -->
+<!-- $$ -->
+
+<!-- ## Checking singly-constrained case against spatial availability -->
+<!-- ```{r, eval=FALSE} -->
+
+<!-- sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){ -->
+
+<!--   o_id <- rlang::enquo(o_id) -->
+<!--   d_id <- rlang::enquo(d_id) -->
+<!--   pop <- rlang::enquo(pop) -->
+<!--   opp <- rlang::enquo(opp) -->
+<!--   r <- rlang::enquo(r) -->
+<!--   f <- rlang::enquo(f) -->
+
+<!--   sum_pop <- x %>% -->
+<!--     dplyr::distinct(!!o_id, -->
+<!--                     .keep_all = TRUE) %>% -->
+<!--     dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>% -->
+<!--     dplyr::pull(sum_pop) %>% -->
+<!--     sum() -->
+
+<!--   f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop -->
+
+<!--   sum_impedance <- x %>% -->
+<!--     dplyr::group_by(!!d_id) %>% -->
+<!--     dplyr::summarize(sum_impedance = sum(!!f)) -->
+
+<!--   x <- x %>% -->
+<!--     dplyr::left_join(sum_impedance, -->
+<!--                      by = rlang::as_name(d_id)) -->
+
+<!--   f_c <- dplyr::pull(x, !!f) / x$sum_impedance -->
+
+<!--   x$f_c <- f_c -->
+<!--   x$f_p <- f_p -->
+
+<!--   sum_pa <- x %>% -->
+<!--     dplyr::group_by(!!d_id) %>% -->
+<!--     dplyr::summarize(sum_pa= sum(f_p * f_c)) -->
+
+<!--   x <- x %>% -->
+<!--     dplyr::left_join(sum_pa, -->
+<!--                      by = rlang::as_name(d_id)) -->
+<!--   f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa) -->
+
+<!--   dplyr::pull(x, !!opp) * f_t -->
+<!-- } -->
+<!-- ``` -->
+
+<!-- ```{r, eval=false} -->
+
+<!-- df <- data.frame(i = c(1,1,1,2,2,2,3,3,3),  -->
+<!--                  j= c(1,2,3,1,2,3,1,2,3), -->
+<!--                  f = c(1/C[1,], 1/C[2,], 1/C[3,]), -->
+<!--                  pop = c(P_i[1], P_i[1], P_i[1], P_i[2], P_i[2], P_i[2], P_i[3], P_i[3], P_i[3]), -->
+<!--                  opp = c(O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3]), -->
+<!--                  catch = c(1,1,1,1,1,1,1,1,1)) -->
+
+<!-- calced <- sp_avail(df, -->
+<!--                          o_id = i, -->
+<!--                          d_id = j, -->
+<!--                          pop = pop, -->
+<!--                          opp = opp, -->
+<!--                          r = catch, -->
+<!--                          f = f) -->
+<!-- df$V_ij <- calced -->
+
+<!-- df |> group_by(j) |> summarise(V_j = sum(V_ij)) -->
+<!-- ``` -->
+
+<!-- ## Destination- (or ATTRACTION) constrained is the only singly-constrained case... population- (or PRODUCTION) constrained doesn't make sense! -->
+
+<!-- Then the origin-constrained (singly-constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
 Using this method, V_i values equal P_i values...?
@@
 This could be a normative minimum threshold, such as 100 opportunities at each origin.
-
-Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{i}$ values:
-$$
-A_{i} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\
-A_{1,1} = \frac{O_{j}}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\
-...\\
-A_{1,1} = 160 *\frac{1}{160/2+437/15+193/5} = 1.083032 \\ 
-A_{2,1} = 160 *\frac{1}{160/15+437/2+193/10} =0.6439496 \\
-A_{3,1} = 160 * \frac{1}{160/5+437/10+193/2} =0.9291521 \\
-...\\
-A_{1,2} = 437 *\frac{1}{160/2+437/15+193/5} =2.958032 \\
-A_{2,2} = 437 *\frac{1}{160/15+437/2+193/10} =1.7587872 \\
-A_{3,2} = 437 * \frac{1}{160/5+437/10+193/2} =2.5377468 \\
-...\\
-A_{1,3} = 193 *\frac{1}{160/2+437/15+193/5} = 1.306408 \\
-A_{2,3} = 193 *\frac{1}{160/15+437/2+193/10} = 0.7767642 \\
-A_{3,3} = 193 *\frac{1}{160/5+437/10+193/2} = 1.1207898 \\
-$$
-
-```{r}
-#| eval: false
-
-A_ij <- matrix(O_j) %*% (1/(t((matrix(O_j))) %*% (1/C))) |> t()
-A_ij_longer <- A_ij
-rownames(A_ij_longer) <- c("1","2","3") 
-colnames(A_ij_longer) <- c("1","2","3") 
-A_ij_longer 
-```
-
-Starting off with this opportunity matrix, namely - at i=1, only a fixed number of opportunities are accessible to this origin. same for i=2, i=3.
-```{r}
-#| eval: false
-
-(matrix(rep(O_j, 3), 3, 3))
-```
-
-$V_{ij}$ values:
-```{r}
-#| eval: false
-
-V_orig <- (A_ij) * t(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.
-
-P_i_orig <- V_orig |> rowSums()
-O_j_orig<- V_orig |> colSums()
-V_TOTAL_orig <- ifelse(sum(P_i_orig)==sum(O_j_orig), sum(V_orig), NA)
-V_orig_longer <- cbind(V_orig,P_i_orig)
-V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig))
-
-rownames(V_orig_longer) <- c("1","2","3", "TOTAL") 
-colnames(V_orig_longer) <- c("1","2","3", "TOTAL")
-
-V_orig_longer #|> as.data.frame()  |> gt(rownames_to_stub=TRUE)
-```
-
 -->
-````
-
-# References
-
-```{=html}
+
+<!-- Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{i}$ values: -->
+<!-- $$ -->
+<!-- A_{i} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\ -->
+<!-- A_{1,1} = \frac{O_{j}}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\ -->
+<!-- ...\\ -->
+<!-- A_{1,1} = 160 *\frac{1}{160/2+437/15+193/5} = 1.083032 \\  -->
+<!-- A_{2,1} = 160 *\frac{1}{160/15+437/2+193/10} =0.6439496 \\ -->
+<!-- A_{3,1} = 160 * \frac{1}{160/5+437/10+193/2} =0.9291521 \\ -->
+<!-- ...\\ -->
+<!-- A_{1,2} = 437 *\frac{1}{160/2+437/15+193/5} =2.958032 \\ -->
+<!-- A_{2,2} = 437 *\frac{1}{160/15+437/2+193/10} =1.7587872 \\ -->
+<!-- A_{3,2} = 437 * \frac{1}{160/5+437/10+193/2} =2.5377468 \\ -->
+<!-- ...\\ -->
+<!-- A_{1,3} = 193 *\frac{1}{160/2+437/15+193/5} = 1.306408 \\ -->
+<!-- A_{2,3} = 193 *\frac{1}{160/15+437/2+193/10} = 0.7767642 \\ -->
+<!-- A_{3,3} = 193 *\frac{1}{160/5+437/10+193/2} = 1.1207898 \\ -->
+<!-- $$ -->
+
+<!-- ```{r, eval=false} -->
+
+<!-- A_ij <- matrix(O_j) %*% (1/(t((matrix(O_j))) %*% (1/C))) |> t() -->
+<!-- A_ij_longer <- A_ij -->
+<!-- rownames(A_ij_longer) <- c("1","2","3")  -->
+<!-- colnames(A_ij_longer) <- c("1","2","3")  -->
+<!-- A_ij_longer  -->
+<!-- ``` -->
+
+<!-- Starting off with this opportunity matrix, namely - at i=1, only a fixed number of opportunities are accessible to this origin. same for i=2, i=3. -->
+<!-- ```{r, eval=false} -->
+
+<!-- (matrix(rep(O_j, 3), 3, 3)) -->
+<!-- ``` -->
+
+<!-- $V_{ij}$ values: -->
+<!-- ```{r, eval=false} -->
+
+<!-- V_orig <- (A_ij) * t(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.  -->
+
+<!-- P_i_orig <- V_orig |> rowSums() -->
+<!-- O_j_orig<- V_orig |> colSums() -->
+<!-- V_TOTAL_orig <- ifelse(sum(P_i_orig)==sum(O_j_orig), sum(V_orig), NA) -->
+<!-- V_orig_longer <- cbind(V_orig,P_i_orig) -->
+<!-- V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig)) -->
+
+<!-- rownames(V_orig_longer) <- c("1","2","3", "TOTAL")  -->
+<!-- colnames(V_orig_longer) <- c("1","2","3", "TOTAL") -->
+
+<!-- V_orig_longer #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
+<!-- ``` -->
+
+<!--NOTE FOR FUTURE EXPLORATION: Interestingly...
+for the doubly- and singly- constrained case, kappas are very similar *maybe close to identical?* when decay is low.
+This is because distance difference doesn't matter, everyone can reach everything and everything is given to everyone.
+But when decay is high, they are much different: since in the doubly-constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
+This relationship should be investigated further some other time..-->
+
 <!--
@@
 Balanced floating catchment areas (not truly doubly constrained)
-
 -->
-```
-
-```{=html}
+
 <!-- 
@@
 
-```{=html}
 <!--TO DISCUSS:
@@
   -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->
+
+<!-- REJECTED TEXT: 
+Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+However, within the same time period, accessibility remained largely within the academic realm.
+For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.].
+We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
+Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
+
+However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
+A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
+Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
+While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
+-->
+
+<!-- MORE REJECTED TEXT --> 
+```{=html}
+<!-- 
+In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
+Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
+
+Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
+Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
+Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
+
+However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
+This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
+Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
+Some works went beyond HAM and considered other accessibility expressions of individual choice.
+For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
+
+Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility -eek?
+I don't think I can say this--.
+We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
+With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
+
+In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
+@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
+
+Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+
+Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
+Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+
+And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
+However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
+Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
+
+Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
+Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
+However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
+A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
+How these balancing factors are applied in the context of _interaction_ like trips  can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
+-->
+```
+
+```{=html}
+<!--
+The approach is based on partition functions from statistical mechanics (the physics of gases), not Newtonian gravitation.
+In short, the approach accepts that there are macro, meso, and micro states.
+In the case of 'trips' being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
+Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
+From this approach, the SIM can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
+There is then a most probable distribution of trips between zones (meso-states) that satisfies all the macro-state constraints.
+
+-->
 ```
@@
 
-<!-- ## Total constrained interaction case -->
-
-<!-- In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
+<!-- ## totally-constrained interaction case -->
+
+<!-- In the totally-constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
 
@@
 
-<!-- #total constrained -->
+<!-- #totally-constrained -->
 
@@
 
-<!-- - total constrained has high error -->
+<!-- - totally-constrained has high error -->
 

### 2025-05-16 paper/manuscript.qmd
draft AS - ready for preprint

@@
 ---
-title: "A family of accessibility measures and the concept of potential derived from spatial interaction principles"
+title: "A family of accessibility measures derived from spatial interaction principles"
+abstract: |
+  Transportation planning has long prioritized the effiency of movement, or mobility.
+However, the concept of accessibility represents a more comprehensive evolution, shifting focus from mere movement to the potential to reach (i.e., spatially interact) with desired destinations.
+Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations.
+This work's aim is to clarify and unify the concept of accessibility by connecting it into spatial interaction modeling.
+We demonstrate that widely used mobility and accessibility models, such as gravity-based accessibility and spatial interaction models, share common theoretical roots.
+From this foundation, we define a family of accessibility measures within the principles of spatial interaction.
+In demonstrating these developments, three contributions emerge: A) the definition and linkages between four members of the family: specifically, the 'unconstrained' measure (i.e., Hansen-type measure), the 'total constrained' measure  (i.e., a constrained version of the Hansen-type measure), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach (2SFCA)), and the 'doubly constrained' measure representing realized interactions or 'access', effectively equal to the doubly constrained spatial interaction model in formulation.
+B) How to formulaically specify the total-, singly- and doubly- constrained accessibility measures.
+C) The interpretability advantages of the family, as these constrained accessibility measures yield values in units of the number of potential "opportunities for spatial interaction" or "population for spatial interaction" for each zone and zonal flow.
+By clarifying the concept of 'potential' in accessibility, demonstrating theoretical and formulaic linkages across popular accessibility and spatial interaction models, and reintroducing units, this work contributes a family of measures that unlock a clearer, more interpretable, and cohesive foundation for accessibility analysis.
+
 format:
@@
 
-As industrializing cities grew rapidly in the early nineteenth century, traffic congestion became a major issue, prompting the development of transportation planning focused primarily on mobility.
+In the early nineteenth century, industrializing cities grew rapidly: so did traffic congestion and the development of transportation planning practice focused primarily on mobility.
 In this newly founded practice, access to destinations was treated as a by-product of movement.
@@
 Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling].
-While accessibility brings a more holistic understanding of transportation and land use systems combined [@handyMeasuringAccessibilityExploration1997], by incorporating both the explicit consideration of destination as opportunity and the concept of _potential_.
+While accessibility brings a more holistic understanding of combined transportation and land use systems, incorporating both the explicit consideration of destination as opportunity and the concept of _potential_ [@handyMeasuringAccessibilityExploration1997].
 
@@
 However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
-Several barriers to bridging the scholarly-practice accessibility gap have been identified.
-For example, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
+For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
 Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
@@
 
-More specifically, this work seeks to address the breath of place-based accessibility's breath of definitions and lack of interpretability by first demonstrating that and mobility-oriented models, such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971], respectively, are rooted in the same spatial interaction modeling foundation.
-In fact, accessibility can be specified using spatial interaction principles.
-And a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constraint values based on known information about the land-use.
-
-In this way, this work aims to introduce constrained accessibility measures, namely the totally-constrained accessibility measure and the singly-constrained accessibility measure, and demonstrate their connection to the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
-These Wilson-analogous constraints can also be defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
-We also aim to demonstrate the intuitive clarity in restoring units to resulting accessibility values; using the constrained measures each zonal value retains interpretable units, namely the 'number of opportunities'.
-This is in contrast to conventional accessibility measures that yield a value that is in units of some 'opportunities-weighted-by-some-representation-of-travel-friction'.
-
-<!-- In this respect, @wuUnifyingAccess2020's contribution provides a unifying framework to think about accessibility measures, the most general of which they describe as $S_i$ in @eq-access-Lu-Levinson.
-By considering the key features of accessibility, namely travel-cost $f(c_{ij})$ and the distribution of opportunities $g(O_i)$, @wuUnifyingAccess2020 demonstrate that a majority of the concepts found scattered throughout the accessibility literature can be seen as particular cases of a general accessibility formula (@eq-access-Lu-Levinson).
-One needs only to judiciously change the way travel costs and opportunities are formulated to derive almost any known place-based accessibility measure.
-In this way, @wuUnifyingAccess2020 have taken a considerable step towards clarifying the differences between various accessibility indicators.
--->
-
-<!-- $$ -->
-<!-- S_i \propto \sum_j g(O_j)f(c_{ij}) -->
-<!-- $$ {#eq-access-Lu-Levinson} -->
-
-<!-- However, there is room to further clarify another relevant aspect: competition, i.e.
-when finite resources can be concurrently accessed by many people.
-@wuUnifyingAccess2020's unifying framework does not appear to accommodate many (or any) of the floating catchment area methods, a popular approach in the accessibility literature that accounts for competition.
-There have also been more recent developments in the literature: for instance, @soukhovIntroducingSpatialAvailability2023 demonstrate that the introduction of a single constraint is sufficient to turn the general accessibility measure (i.e, @eq-access-Lu-Levinson) into a competitive measure of accessibility in accordance with the 2-Step Floating Catchment Area approach.
-Moreover, the proportional allocation balancing factors in @soukhovIntroducingSpatialAvailability2023 apply a singly-constraint to the accessibility measure, akin to the constraints introduced within the spatial interaction modelling framework outlined in @wilson1971.
--->
-
-<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To narratively reiterate: this paper contends that accessibility research must reconnect with its spatial interaction origins.
+More specifically, this work seeks to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
+In fact, accessibility can be specified using spatial interaction principles and a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constrains values based on known information about the land-use.
+
+In this way, this work aims to offer three contributions: (1) introduce and (2) formulaically define a family of _constrained_ accessibility measures, namely the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure.
+These measures are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential, or what Reilly called 'potential trade territories' [@reilly1929methods] and further developed in other examples [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
+(3) This work also aims to demonstrate the intuitive value of restoring units to the resulting raw accessibility values.
+Each zonal and zonal flow value from a constrained accessibility measure is in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'.
+This is in contrast to conventional accessibility measures that yield values in units of 'opportunities weighted by some representation of travel friction'.
+
+<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To achieve these stated objectives, this paper contends that accessibility research must reconnect with its spatial interaction origins.
 Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
@@
 It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-The primary focus of this work, therefore, is to demonstrate how the empirical constraints used in @wilson1971's family of spatial interaction models can be mapped onto accessibility.
-Both share similar origins in spatial interaction, and popular accessibility measures can be traced directly back to these roots, offering a solution for reintroducing units to the outputs.
-In the following section, we begin by tracing the development of accessibility from its origins in spatial interaction, from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
+We continue this work in the following section by tracing the development of accessibility from its origins in spatial interaction: from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
 We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
-Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on the total and single constraints.
-We illustrate various members of this family with a simple numerical example.
+Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based analogous constraints.
+We illustrate members of this family of constrained accessibility measures with a simple numerical example.
 We then conclude by discussing the uses of these measure and their interpretation.
 
-Using the newly defined family of accessibility measures, this work aims to: (1) demonstrate how constrained accessibility can be specified, and by doing so, bring interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'; (2) provide a clarifying framework for understanding the concept of _potential_ through the introduced constraints; and (3) highlight how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the family of accessibility measures.
-
 # Newtonian's roots of human spatial interaction research
@@
 <!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
-
 The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half.
@@
 
-To be certain, the expression above, a proportionality, is one of the most famous in all of science.
-In brief, it states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$.
+To be certain, the expression above is one of proportionality, and is also the most famous in all of science.
+It states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$.
 Direct proportionality means that as the product of the masses increases, so does the force.
@@
 -->
-
-Since the 1880s to the 1940, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance.
+Since the 1880s to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance.
 This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
@@
 Over time, other researchers discovered similar relationships.
-<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations.
+<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations that could be understood as 'potential trade territories'.
 Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of information [@zipfDeterminantsCirculationInformation1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and goods movement by railways [@zipfHypothesisCaseRailway1946].
@@
 $$
-V_i = \sum_j \frac{M_j}{d_{ij}}
+V_i = \sum_j M_jd_{ij}^{-1}
 $$ {#eq-stewart-population-potential-sum}
 
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility:
-
-$$
-V_i = \sum_j M_jd_{ij}^{-1}
-$$ {#eq-stewart-population-potential-sum-rewrite}
-
-<!-- RP: I say we spell it out as in this eq^, what you think? AS: Maybe! -->
+Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility.
 
@@
 34 and then later as _demographic energy_ on p.
-53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific ideas that were unadultered racism.
+53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific and racist ideas.
 For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" \[p.
@@
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
-In his seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
+In this seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
 a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p.
@@
 As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-@hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), leaving Stewart's other formulaic contributions and objectionable aspects of "social physics" behind.
+@hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), and not on the other formulaic contributions and objectionable aspects of "social physics".
 @hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
@@
 However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day.
-The omission is that that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished.
+The omission is that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished.
 This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
@@
 Wilson was developing related, yet parallel work.
-In the groundbreaking [@wilson1971], Wilson defined a general spatial interaction model as follows.
+In his groundbreaking study [@wilson1971], Wilson defined a general spatial interaction model as follows.
 While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling spatial interaction:
-
-On the other side of the Atlantic, Alan G.
-Wilson was developing related, yet parallel work.
-In his groundbreaking study [@wilson1971], accessibility is acknowledged as an associated concept, though the work's primary focuses on defining an introducing a general spatial interaction model as follows:
-
 $$
@@
 The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations.
+The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., population as $W_i^(1)$ and hectares of parkspace as $W_j^(2)$.
 Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
@@
 $K$ is specified depending on the applied constraint(s).
-In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where no constraints apply, the second being a singly-constrained case, and the third a doubly-constrained case.
+In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where no constraints apply, the second being a singly constrained case, and the third a doubly constrained case.
 
@@
 In the second case, only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold.
-The resulting models are, in Wilson's terms, singly-constrained.
+The resulting models are, in Wilson's terms, singly constrained.
 When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained model:
@@
 
-The attraction-constrained model is similar to the production-constrained model as it is also singly-constrained but from the perspective of the mass at the destination.
+The attraction-constrained model is similar to the production-constrained model as it is also singly constrained but from the perspective of the mass at the destination.
 From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow, i.e., trips attracted at the destination and takes the following form:
@@
 This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly-constrained, and takes the following form:
+The resulting model is, in Wilson's terms, doubly constrained, and takes the following form:
 
@@
 It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been ignored in almost all subsequent accessibility research.
-Why was this key element of spatial interaction models ignored in accessibility research?
+Why was this key element of spatial interaction models potentially ignored in accessibility research?
 In the next section we aim to address this question.
@@
 
-<!-- rafa: this  bibliographic analysis with Figure 2 is cool, but it looks unecessary and distractig from the core contribution of the paper. I'd suggest removing it  -->
-
-<!--
-```{r}
-#| label: community-detection
-#| cache: true
-
-# Detect communities based on the bibliographic coupling. For repeatability set the random seed
-set.seed(979834789)
-coupling_net <- coupling_net |>
-  mutate(community = as.factor(tidygraph::group_infomap(weights = weight)))
-```
--->
-
 As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
@@
 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
-Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatiall interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
+Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model that may remedy the issue (e.g., the *competition destination* model).
+Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
 Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
@@
 However, these balancing factors were deduced intuitively.
-These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly-constrained model (derived from entropy maximization), as will be demonstrated in this study.
+These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization), as will be demonstrated in this study.
 They also do not discuss other constrained cases.
@@
 As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
-This may explain why the constraints/proportionality constant(s) of spatial interaction models did not cross over to accessibility analysis.
+This may explain why the constraints and associated balancing factors of spatial interaction models did not cross over to accessibility analysis.
 This is intriguing since Wilson made an effort to connect developments in spatial interaction modelling to accessibility, noting for instance, the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of balancing factor $A_i$ [@wilson1971 p.
@@
 
-Hence, we propose a more general definition of accessibility as a partial sum of the spatial interaction.
+Hence, we propose a more general definition of accessibility: the partial sum of the potential for spatial interaction.
 Once we bring back Wilson's proportionality constant into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
@@
 
-\noindent where $V_{ij}$ is the potential for interaction, and therefore the accessibility from origin $i$ is defined as:
+\noindent where $V_{ij}$ is the potential for interaction from $i$ to $j$.
+$X$ denotes a sub-indices for the member of the constrained accessibility value, either unconstrained $0$, total constrained $T$, singly constrained $S$ or doubly constrained $D$.
+The accessibility from origin $i$ is defined as:
 
@@
 
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination, and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
-
-In the following subsections, we detail the implications of introducing the constraint $k$ into accessibility measures and how different _constrained_ accessibility measures can be defined based on the focus of measurement and the input population and opportunity data.
-As an overview, the way each constraint $k$ is defined delineates one of four distinct members of the constrained accessibility measure family:
+Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+
+The market potential variant can also be generally defined, which is effectively the transpose of $i$ and $j$ in @eq-access-01 and @eq-accesssibility-01 as follows:
+$$
+M_{ji} = k W_i^{(1)} f(c_{ij})
+$$ {#eq-market-01}
+
+$$
+M_{j} = k \sum_j W_i^{(1)} f(c_{ji})
+$$ {#eq-market-potential-01}
+
+Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+
+To demonstrate the anatomy of $V_{ij}$ and $M_{ij}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility demonstrates this work's accessibility analytical device for a simple 3 zone system.
+Each measure's most disaggregate value is $X_{ij}$, potential for spatial interaction from $i$ to $j$.
+The $X$ is a stand in for the $ij$ values of all the cases and their variants that will be described (i.e., $V_{ij}^0, M_{ij}^0, V_{ij}^T, M_{ij}^T, V_{ij}^S, M_{ij}^S$ and $V_{ij}^D, M_{ij}^D$).
+The single marginals represent the origin-side and destination-side weights of the zones.
+The total marginal represent the sum of a single marginal.
+
+```{r}
+#| label: fig-analytical-device-conc-accessibility
+#| out-width: 70%
+#| fig-cap: "The family of accessibility measures analytical device: labelling and associating ij flows, zonal weights, the single marginals, and the total marginal."
+
+knitr::include_graphics(glue::glue(here::here(),
+                                   "/figures/access-analytical-device.png"))
+```
+
+As an additional overview, we define four distinct members of the constrained accessibility measure family, all delineated based on their constant $k$; either balancing factors $K^T$, $B_j$, $A_i$ depending on their case and variant:
 
 1.
-The unconstrained case.
-This case is equivalent to @hansen1959's formulation whereby a balancing factor is neglected, so units of zonal values are simply in 'opportunities-by-travel-impedance-value';
-2. The totally-constrained case; maximum potential. Two variants are introduced. 
-  - The first, the _totally-constrained accessible opportunities measure_, resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
-In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region, requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
-  - The second variant is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _totally-constrained accessible population measure_.
-In this variant each zonal accessibility value is a proportion of the population in the region, requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible;
-3. The singly-constrained case; medium potential, potential is limited by either an opportunity-constraint or population-constraint. Similarly, two variants are introduced. 
-  - The first is _attraction-constrained_ from the perspective of spatial interaction modelling.
-This variant is refereed to as the _singly-constrained accessible opportunities measure_.
-It can also be formulated to resembles @hansen1959's formulation but with an additional term for $k$, a set of balancing factors $B_j$ which ensure opportunities to each zone are allocated based on population and travel impedance regionally and zonally.
-$B_j$ ensures that a balanced proportion of the total opportunities at each destination is allocated based on zonal population and travel impedance, so that each zonal accessibility value reflects both a share of the total opportunities in the region _and_ is a sum of a _balanced_ proportions of opportunities allocated from each destinations.
-Similar to the totally-constrained case, each zonal accessibility value is expressed in units of 'opportunities' accessible.
-But unlike the totally-constrained case, the singly-constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
-This variant is mathematically equivalent to spatial availability [@soukhovIntroducingSpatialAvailability2023] and its per capita version is equivalent to the 2SFCA [@shenLocationCharacteristicsInnercity1998; @luo2003].
-  -  The second variant is _production-constrained_ from the perspective of spatial interaction modeling.
-It can also be interpreted as a constrained form of market potential (i.e., the transpose of the singly-constrained variant), incorporating the assumptions of the single constraint.
-Specifically, each zonal accessibility value represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance.
-Each zonal value is expressed in units of 'population' accessible;
+**The unconstrained case** with variants $V_i^0$ and $M_i^0$.
+This case is equivalent to @hansen1959's formulation (first variant) and @reilly1929methods's market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'.
+In this case, no constraints are introduced to ensure that values of marginals are preserved.
+
+2.
+**The total constrained case** with variant $V_i^T$ and variant $M_j^T$.
+The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
+In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
+The second variant $M_j^T$ is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _total constrained accessible population measure_.
+In this variant, each zonal accessibility value is a proportion of the total population in the region (maintained by $\hat K^T$), requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible.
+  - When conceptualising _potential_ for spatial interaction--whether with opportunities or population--the total constrained case reflects the lowest level of restriction and hence maximum potential.
+The balancing factors $K^T$ and $\hat K^T$ only ensure that $V_{ij}^T$ and $M_{ji}^T$ values end up matching the _total sum_ of one of the single marginals, not the individual marginals themselves.
+For example, $K^T$ does not guarantee that all the opportunities accessibility at $V_{i=1}^T$ reflect a proportional allocation of the destination mass (i.e., _number_ of opportunities) at $j=1, 2, 3$.
+In some cases, an allocation of opportunities from a destination $j$ to $i=1$ could _exceed_ the number of opportunities at that $j$ (i.e., meaning that destination $j$ is very attractive and reachable for an $i$, relative other flows in the region).
+However, what $K^T$ _does_ ensure is that each $V_{ij}^T$ does not exceed the overall number of opportunities in the region--the total marginal.
+In other words, $V_{i=1}^T$ expresses the number of opportunities that origin $i=1$ could potentially interact with, as drawn from the entire regional opportunity mass, rather than constrained by individual destination totals.
+In this way, the _total constrained_ case reflects the maximum amount of _potential_ while still maintaining interpretable units.
+
+3.
+**The singly constrained case**, with two variants $V_i^S$ (opportunity-constrained) and $M_j^S$ (population-constrained).
+The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998].
+Both variants can also be defined using either @hansen1959's or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$.
+These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility).
+In this way, the singly constrained case reflects a medium amount of potential, restricted by either values fitting the single destination-mass marginal or the single origin-mass marginal.
+  - The first variant $V_i^S$ includes a set of destination-side balancing factors $B_j$ which ensure opportunities (or destination masses) are allocated to each origin $i$ based on the population (i.e, origin mass) of the origin $i$ and travel impedance from that $j$ to all $i$s.
+Hence, $B_j$ ensures that each $V_i^S$ is the product sum of a _proportional share_ of opportunities allocated from each destination $j$ implicitly also being a share of the total opportunities in the region.
+Similar to the total constrained case, each $V_i^S$ and $V_ij^S$ accessibility value is expressed in units of 'opportunities' accessible.
+But unlike the total constrained case, the singly constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
+  - The second variant $M_j^S$ includes a set of opportunity-side balancing factors $A_i$ which ensure population (or origin masses) are allocated to each destination $j$ based on the destination mass of the destination $j$ and all possible travel impedance from that $i$ to all $j$s.
+$A_i$ ensures similar, but transposed, so constraints are satisfied.
+Specifically, each $M_j^S$ represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance.
+Each zonal value is expressed in units of 'population' accessible.
+
 4.
-The doubly-constrained case.
-It is constrained simultaneously by population and opportunities, so each zonal value is expressed in units of 'population-opportunity capacity' that is accessible.
-This case requires the number of opportunities and population to match, e.g., the user must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
-By using this one-to-one matching data, this case's calculated values especially limits the 'potential' to spatially interact compared to the data used in the singly-constrained and totally-constrained cases.
-Instead, the doubly-constrained values reflect the number of predicted interactions between the opportunities and population.
-Effectively, this case is equivalent formulaically as Wilson's doubly-constrained spatial interaction model (i.e., _attraction-production constrained_) and can be understood as predicting a value of access , not accessibility.
-
-In the next section, we illustrate each case using a simple numerical example.
-
-## Simple numeric example setup
+**The doubly constrained case**.
+It is constrained simultaneously by population (origin masses) and opportunities (destination masses), so each $V_{ij}^D$ (equivalent to $M_{ij}^D$) is expressed in units of 'population-opportunity capacity' that is accessible between $i$ to $j$.
+This case requires the number of opportunities and population to match, e.g., the analyst must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
+By using this one-to-one matching data and the double constraints (i.e., $A_i$ and $B_j$ at once, ensuring both the double constraint maintaints both marginals in @fig-analytical-device-conc-accessibility), this case restricts the _potential_ to spatially interact completely.
+In other words, the doubly constrained accessibility values reflect the number of predicted interactions between the opportunities and population, effectively, this case is equivalent formulaically as Wilson's doubly constrained spatial interaction model (i.e., _attraction-production constrained_).
+It can be understood as predicting a value of 'access', and not accessibility.
+
+In the next sections, we will assign values to the simple 3x3 numeric example shown in the analytical device (@fig-analytical-device-conc-accessibility).
+
+## Setup of the simple numeric example
 
@@
 
-Consider a simple region with three zones, IDs 1, 2 and 3.
+Consider a simple region as shown in @fig-analytical-device-conc-accessibility, with zone IDs 1, 2 and 3.
 Each zone is both an origin $i$ and a destination $j$.
@@
 Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups, can be substituted into our simple example, depending on the research question.
-The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behaviour groups and cases of the family of accessibility neasures.
+The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behaviour groups and members of the family of accessibility measures.
 
@@
             by = c("did" = "id"))
+
 ```
@@
             V_unc_i_3 = sum(V_unc_ij_3))
-
-## unconstrained access in one step
-# unc_acc <- OD |>
-#   group_by(oid) |>
-#   summarize(V_unc_i_1 = sum(D_j * f1),
-#             V_unc_i_2 = sum(D_j * f2),
-#             V_unc_i_3 = sum(D_j * f3))
 
@@
 Hence, the raw unconstrained accessibility values themselves are challenging to compare due to their units.
-To enable more meaningful comparison, the following sections will detail the introduction of constraints to ensure consistent units across scenarios and demonstrate results on the same numeric example.
-
-## Totally-constrained accessibility
-
-In the totally-constrained accessibility case, a total constraint proportionally adjusts unconstrained zonal accessibility values $V^0_i$ based on the regional sum of $V^0_i$ and the total population or opportunities in the region.
+To enable more meaningful comparison, the following sections will detail the introduction of constraining constants to ensure consistent units across scenarios and demonstrate results on the same numeric example.
+
+## Total constrained accessibility
+
+In the total constrained accessibility case, a total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ based on the regional sum of $V^0_i$ and the total population or opportunities in the region.
 Alternatively, we reformulate this case using a proportional allocation constant, which allocates opportunities (or population) proportionally based on the the travel impedance and the total population or opportunities in the region.
@@
 
-We define two variants for this case: (a) case where accessibility is constrained by the total number of opportunities (totally-constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraint, and (b) a variant where $i$ and $j$ of the totally-constrained accessibility measure is transposed (totally-constrained accessible population) hence is constrained by the total number of population and interpreted as constrained 'market potential'.
-
-### Totally-constrained accessible opportunities: Hansen's accessibility with a constraint
-
-Unlike in @eq-access-01, the proportionality constant $k$ is retained. For the totally-constrained case, it is represented as $K^T$:
+We define two variants for this case: (a) $V_i^T$ where accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and (b) $M_j^T$, where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total number of population and to be interpreted as constrained 'market potential'.
+
+### Total constrained accessible opportunities: Hansen's accessibility with a total constraint
+
+Unlike in @eq-access-01, the proportionality constant $k$ is retained. For the total constrained case, it is represented as $K^T$:
 
@@
 
-In this way, the totally-constrained accessibility measure now becomes Hansen's accessibility with a proportionality constant $K^T$:
+In this way, the total constrained accessibility measure now becomes Hansen's accessibility with a balancing factor $K^T$:
 
@@
 Imagine that the only system known is the total number of opportunities $D$ in the region.
-Accordingly, the constraint we impose in this case ensures the regional sum of totally-constrained accessibility is equal to the total number of opportunities as follows:
+Accordingly, the constant we impose in this case ensures the regional sum of total constrained accessibility is equal to the total number of opportunities as follows:
 $$
@@
 Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
-Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our totally-constrained accessibility model is:
+Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:
 
@@
 
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the totally-constrained opportunity case $\kappa^T$ is dimensionless:
-
-$$
-\kappa^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$).
-The role of $\kappa^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
-Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-For this reason, we refer to $\kappa^T$ as a proportional allocation factor.
-On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the totally-constrained accessibility measure.
-
-Referring back to our simple numeric example, proportionality constant $K^T$ for the most decay travel behaviour scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
+Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_i^T$ is dimensionless:
+
+$$
+\kappa_i^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa_i^T\sum_j W^{(2)}_j$).
+The role of $\kappa_i^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
+$\kappa_i^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+This is why we refer to $\kappa_i^T$ as a proportional allocation factor.
+On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.
+
+Referring back to our simple numeric example, balancing factor $K^T$ for the most decay travel behaviour scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
 
@@
 \begin{array}{l}
-K^T = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
-K^T = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
+K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})}\\
+K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + \cdots + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
 }\\
@@
 Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
-$\kappa^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
+$\kappa_i^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
 
@@
             V_tot_i_3 = sum(V_tot_ij_3))
+
+# tot_acc_containing_kappa <- unc_acc_ij |>
+#   mutate(V_tot_ij_1 = k_tot$k1 * V_unc_ij_1,
+#          V_tot_ij_2 = k_tot$k2 * V_unc_ij_2,
+#          V_tot_ij_3 = k_tot$k3 * V_unc_ij_3) |>
+#   left_join(LU |> rename("oid" = "id") |> select(-c("O_i")), by = "oid") |>
+#   group_by(oid) |>
+#   summarize(V_tot_i_1 = sum(V_tot_ij_1),
+#             V_tot_i_2 = sum(V_tot_ij_2),
+#             V_tot_i_3 = sum(V_tot_ij_3),
+#             D_j = first(D_j)) |>
+#   mutate(kappa_tot_i_1 = V_tot_i_1/D_j, 
+#          kappa_tot_i_2 = V_tot_i_2/D_j, 
+#          kappa_tot_i_3 = V_tot_i_3/D_j)
 
@@
 In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
-Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa^T$.
+Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_i^T$.
 
@@
 Why?
-$\kappa^T$ ensures proportional allocation for each travel behaviour scenario.
-Meaning, while the unconstrained accessibility increases, $\kappa^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
+$\kappa_i^T$ ensures proportional allocation for each travel behaviour scenario.
+Meaning, while the unconstrained accessibility increases, $\kappa_i^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
 As the decay behaviour decreases, more opportunities are accessible for all zones.
@@
 
-### Totally-constrained accessible population: totally-constrained market potential
-
-Another variant of the totally-constrained accessibility measure is the _totally-constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the totally-constrained accessible opportunities measure.
+### Total constrained accessible population: Reilly's potential trade territories with a total constraint
+
+Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure.
 This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
@@
 However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form.
-To address this, the constrained variant, the totally-constrained accessible population measure $M^T_{j}$, is introduced.
-To formulate this variant, the total constraint $K^T $ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
+To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced.
+To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
 
@@
 
-The totally-constrained accessible population measure (market potential) then becomes:
+The total constrained accessible population measure (market potential) then becomes:
 
@@
 
-Where the following $\hat \kappa^T$ proportional allocation factor is dimensionless:
-
-$$
-\hat \kappa^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+Where the following $\hat \kappa_i^T$ proportional allocation factor is dimensionless:
+
+$$
+\hat \kappa_i^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 $$
@@
 Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
-However, similar to the totally-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
+However, similar to the total constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
 As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
-Overall: like the totally-constrained accessible opportunities case, this variant allows for the interpretation of both ordinal and interval comparisons of the raw values themselves.
-
-## Singly-constrained accessibility
-
-Similar to the totally constrained accessibility measure, the singly-constrained case includes a constraint that proportionally adjusts the unconstrained zonal accessibility values $V^0_i$.
-Two variants are defined: (a) the _singly-constrained accessible opportunities_  case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly-constrained accessible population_ case, its transpose, or singly-constrained market potential.
-
-Unlike the total constraint, the single constraint incorporates additional information.
-In the opportunities-accessible variant, the balancing factor constrains the 'potential' for spatial interaction by ensuring that only a proportional amount of opportunities at each destination are allocated to 'demanding' origins.
+Overall: like the total constrained accessible opportunities case, this variant allows for the interpretation of both ordinal and interval comparisons of the raw values themselves.
+
+## Singly constrained accessibility
+
+Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied.
+Two variants are defined: (a) the _singly constrained accessible opportunities_  case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.
+
+Unlike the total constraint (i.e., @eq-total-constraint-accessibility), the single constraint--as will be defined in @eq-opportunity-constraint and @eq-population-constraint for the first and second variants-- incorporates additional information.
+In the opportunities-accessible variant, the associated balancing factor constrains the 'potential' for spatial interaction by ensuring that only a proportional amount of opportunities at each destination are allocated to 'demanding' origins.
 This allocation is informed by demand, or the relative amount of population, and associated travel impedance connecting the zones.
-In the population-accessible variant (the transpose), population at each origin is proportionally allocated to destinations based on the share of opportunities and travel impedance.
-
-In both variants, the balancing factor allocates opportunities (or population) based on the relevant supply-demand weights of the origins and destinations, and the travel impedance between zones.
-The singly-constrained accessibility measure introduces population-based (or opportunity-based) competition at the zonal level, unlike the total constraint, which more simply allocates a fixed regional total of opportunities (or population depending on the variant).
-
-In sum, all singly-constrained zonal accessibility values are both a proportion of the known regional opportunity (or population) total _and_  a sum of a _balanced_ proportion of opportunities allocated from each destinations (or population allocated from each origin).
-Each zonal value remains in units of opportunities accessible (or population accessible), but in contrast to the totally constrained case, the singly constrained measures explicitly incorporates both spatial weights (i.e., population and opportunities) into the allocation process.
-
-### Singly-constrained accessible opportunities: spatial availability
-
-To demonstrate this variant formulaically, we begin with the opportunity constraint (@eq-opportunity-constraint) and our known piece of information, i.e., the sum of accessible opportunities from a destination should equal the number of opportunities $D_j$ at that destination.
-As the number of opportunities at each $j$ are known, it is represented as $D_j$ instead of $W_j^(2)$ as in the totally-constrained case.
+In the population-accessible variant (i.e., the $i$ -> $j$ transpose of the first variant), population at each origin is proportionally allocated to destinations based on the share of opportunities and travel impedance.
+
+In both variants, the singly constrained accessibility measure introduces population-based (or opportunity-based) competition at the zonal level, unlike the total constraint, which more simply allocates a fixed regional total of opportunities (or population depending on the variant).
+In sum, all singly constrained zonal accessibility values are both a proportion of the known regional opportunity (or population) total _and_ a sum of a _balanced_ proportion of opportunities allocated from each destinations (or population allocated from each origin).
+Each zonal value remains in units of opportunities accessible (or population accessible).
+
+### Singly constrained accessible opportunities: spatial availability
+
+To demonstrate this variant formulaically, we begin with the opportunity constraint (@eq-opportunity-constraint) as our known piece of information.
+Namely, the sum of accessible opportunities from a destination should equal the number of opportunities $D_j$ at that destination.
+As the number of opportunities at each $j$ are known, it is represented as $D_j$ instead of $W_j^(2)$ as in the total constrained case.
 This constraint should hold for all destinations in the region.
@@
 @soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As in the totally-constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
-The singly-constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
+As in the total constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
+The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
 These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
@@
 v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
-$$
-
-To reinforce this point, @eq-opportunity-constraint also implies that the total system constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.
-
-Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible *and* available to location $i$ out of the total number of opportunities in the system.
+$${#eq-opportunity-constrained-accessibility-per-capita}
+
+<!-- To reinforce this point, @eq-opportunity-constraint also implies that the total constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.
+-->
+<!-- Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible *and* available to location $i$ out of the total number of opportunities in the system.
+-->
 
@@
 \begin{array}{l}
-B_{j} = \frac{1}{\sum_i O_i f(c_{ij})}\\
+B_{j} = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
 B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
@@
 The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
-Using these these proportionality constants, we can calculate the singly-constrained opportunity accessibility:
+Using these these balancing constants, we can calculate the singly constrained opportunity accessibility:
 
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility
-#| tbl-cap: "Simple system: singly-constrained opportunity accessibility."
+#| tbl-cap: "Simple system: singly constrained opportunity accessibility."
 
@@
 
-Imposing the single constraint $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the totally-constrained accessible opportunities case.
-Namely, when the single constraint is applied to the opportunities at $j$, each resulting accessibility value at $i$ is in units of _physicians_, with the impedance units already accounted for.
-
-However, unlike the totally-constrained opportunity case, $\kappa^S_{i}$ reflects zonal competition based on the mass of the origin, i.e., population.
+Imposing the single proportional allocation factor $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case.
+The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.
+
+However, unlike the total constrained opportunity case, $\kappa^S_{i}$ reflects zonal competition based on the mass of the origin, i.e., population.
 Again, consider the highest decay scenario $f_1(c_{ij})$.
-Under this scenario, zone 1 no longer captures a medium amount of physicians as in the totally-constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
+Under this scenario, zone 1 no longer captures a medium amount of physicians as in the total constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
 Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
 
-This difference is due to the single-opportunity constraint $\kappa^S_{i}$'s role in $V_i^S$.
-The only inputs required in the total-opportunity constraint $\kappa^T$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
+This difference is due to the single-opportunity factor $\kappa^S_{i}$'s role in $V_i^S$.
+The only inputs required in the total-opportunity factor $\kappa^T_i$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
 Opportunities are allocated to $i$s, regardless of the mass weights of origin.
@@
 In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
-Notabely, these per capita rates are equivalent to the 2SFCA values.
+Notably, these per capita rates are equivalent to the 2SFCA values.
 
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
-#| tbl-cap: "Simple system: singly-constrained opportunity accessibility per capita."
+#| tbl-cap: "Simple system: singly constrained opportunity accessibility per capita."
 
@@
 As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
-Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all around 24 physicans accessible per capita).
-
-This trend mirrors how the accessibility values in the totally-constrained opportunity case stabilizes to the $V_i^T$ regional average (e.g., the accessible opportunities allocated to each of the three zones approaches a third of 490 physicians, or 163.33, under the unfettered mobility scenario $f_3(c_{ij})$).
-These patterns make intuitive sense: constraints act as regional and/or zonal averaging mechanisms.
-In distance decay travel behaviour scenarios that are uniform (i.e., low for all like in $f_3(c_{ij})$), what remains is the relative effect of the other variables in the constraint variable.
-In the totally constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constraint, this is the population at a zone relative to the regional population.
-
-### Singly-constrained accessible population: singly-constrained market potential
-
-Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly-constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
+Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all around 24 physicians accessible per capita).
+
+This trend mirrors how the accessibility values in the total constrained opportunity case stabilises to the $V_i^T$ regional average (e.g., the accessible opportunities allocated to each of the three zones approaches a third of 490 physicians, or 163.33, under the unfettered mobility scenario $f_3(c_{ij})$).
+These patterns make intuitive sense: the balancing factors act as regional and/or zonal averaging mechanisms.
+In distance decay travel behaviour scenarios that are more relatively uniform (i.e., low for all zones like in $f_3(c_{ij})$), what remains is the relative effect of the other variables in the balancing factor.
+In the total constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constrained case, this is the population at a zone relative to the regional population.
+
+### Singly constrained accessible population: market availability
+
+Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
 In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
@@
 
-In this case, the measure is singly-constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+In this case, the measure is singly constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
 
@@
 
-Following the same logic as in the preceding section on totally-constrained market potential, one arrives at the following expression:
+Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression:
 
@@
 
-As well, the constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O$) is maintained.
+As well, the single (population) constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O$) is maintained.
 
@@
 
-For the sake of brevity, we'll move forward onto the doubly-constrained case.
-
-## Doubly-constrained accessibility
-
-This accessibility case requires zonal populations and opportunities to match one-to-one, like the doubly-constrained spatial interaction model.
-In this way, doubly-constrained accessibility can be thought as "access", or the realized potential for spatial interaction.
-
-To contextualize this point, the totally- and singly-constrained accessibility measures discussed thus far have used either $O_i$, $D_j$, or the regional sums of either, but never both simultaneously.
-For example, when opportunities $D_j$ are used to constrain @eq-opportunity-constrained-accessibility in the _singly-constrained accessible opportunities measure_, the specific mass of the population at origin $i$ demanding only those opportunities at that $j$ is unknown.
+For the sake of brevity, we'll move forward onto the doubly constrained case.
+
+## Doubly constrained accessibility
+
+This accessibility case requires zonal populations and opportunities to match one-to-one, like the doubly constrained spatial interaction model.
+In this way, doubly constrained accessibility can be thought as "access" or simply spatial interaction (no potential).
+
+To contextualize this point, the total and singly constrained accessibility measures discussed thus far have used either $O_i$, $D_j$, or the regional sums of either, but never both simultaneously.
+For example, when opportunities $D_j$ are used to constrain @eq-opportunity-constrained-accessibility in the _singly constrained accessible opportunities measure_, the specific mass of the population at origin $i$ demanding only those opportunities at that $j$ is unknown.
 Instead, only the population at $i$ demanding opportunities in the region is known, and this is more generally represented as $W_i^{(1)}$ (as in @eq-opportunity-constrained-proportionality-constants).
-Similarly, when the population $O_i$ is used as a constraint in @eq-population-constrained-accessibility in the _singly-constrained accessible population measure_, the mass at the destination is given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants) since only the mass of opportunities at each $j$ is known and information on what opportunities are allocated to what zone is unknown.
+Similarly, when the population $O_i$ is used as a constraint in @eq-population-constrained-accessibility in the _singly constrained accessible population measure_, the mass at the destination is given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants) since only the mass of opportunities at each $j$ is known and information on what opportunities are allocated to what zone is unknown.
 
@@
 
-A doubly-constrained approach to accessibility calculation needs a one-to-one relationships between population and opportunities to be present.
-Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly-constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of destinations in all destinations (@eq-opportunity-population-equality):
+A doubly constrained approach to accessibility calculation needs a one-to-one relationships between population and opportunities to be present.
+Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality):
 
@@
 
-As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly-constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$ as shown in @eq-total-constraint-accessibility.
-
-As the doubly-constrained accessibility measure $V_{i}^D$ takes the form of the production-attraction (doubly-constrained) spatial interaction model, as shown in @eq-doubly-constrained-gravitymodel, $V_{i}^D$ is as follows:
-
-$$
-V_{i}^D = \sum_j A_i B_j O_i D_j f(c_{ij})
+As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$ as shown in @eq-total-constraint-accessibility.
+
+As the doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model, as shown in @eq-doubly-constrained-gravitymodel, $V_{i}^D$ is as follows:
+
+$$
+V_{ij}^D = A_i B_j O_i D_j f(c_{ij})
 $$ {#eq-doubly-constrained-accessibility}
@@
 
-Rewriting @eq-doubly-constrained-accessibility as an origin summary expression of proportionally allocated opportunities:
+One could rewrite @eq-doubly-constrained-accessibility as an origin summary expression of proportionally allocated opportunities:
 
@@
 
-Unlike in the totally- and singly-constrained cases, representing $V_i^D$ per capita is not as meaningful, as the value *already* matches population to opportunities, e.g., the number of opportunities accessed by the population.
-Similarly, the market potential form $M^D_j$ would be the transposed way of summarizing $V_{ij}^D$.
-In other words, the inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both opportunities and population.
-As opportunities match population and both marginals are constrained, $M^D_j$ would be a summary of $V^D_{ij}$ values at each $j$, and would represent the number of people accessible to opportunities.
-The formulation is as follows:
-
-$$
-M^D_j = \sum_i A_i B_j O_i D_j f(c_{ij})\\
-$$ {#eq-doubly-constrained-market-potential}
-
-```{r}
-#| label: simple-numerical-example-doubly-constrained
+However, $V^D_{i}$ is not wholy helpful, as it will simply equal the origin-mass marginal.
+In this way, only $V^D{ij}$ values are interpretable.
+Furthermore, unlike in the total and singly constrained cases, the doubly-constrained case does not have an interpretable per capita version.
+For instance, representing $V_{ij}^D$ per capita is not meaningful, as the value *already* matches population to opportunities.
+Following this logic, the market potential form $M^D_{ij}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: i.e., the opportunities accessed from $j$ at an $i$ vs.
+the population accessed from $i$ at a $j$.
+The inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both opportunities and population.
+
+```{r}
+#| label: simple-numerical-example-doubly constrained
 
@@
 
-To calculate doubly-constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be modified.
-Namely, a count of physician _capacity_ per destination is needed instead of just the number of physicians, as used to calculated totally- and singly-constrained cases.
+To calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be modified.
+Namely, a count of physician _capacity_ per destination is needed instead of just the number of physicians, as used to calculated total and singly constrained cases.
 We also must be able to clearly state that the population is a count of people seeking opportunities at the new capacities, i.e., the population must reflect the _capacity_ of the population to interact with opportunities.
 
-So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
 For the population, we leave this unchanged numerically but theoretically know that each person interacts with one physician capacity (i.e., our opportunities with a _capacity_).
@@
 ```{r}
-#| label: tbl-small-system-land-use-doubly-constrained-case
+#| label: tbl-small-system-land-use-doubly constrained-case
 #| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."
@@
 ```{r}
-#| label: simple-numerical-example-OD-table-doubly-constrained
+#| label: simple-numerical-example-OD-table-doubly constrained
 
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
-#| tbl-cap: "Doubly-constrained opportunity accessibility assuming the most decay travel behaviour scenario in the modified simple system."
+#| tbl-cap: "Doubly constrained opportunity accessibility assuming the most decay travel behaviour scenario in the modified simple system."
 
@@
 As mentioned, accessibility values are typically interpreted as a summary of the proportionally allocated opportunities at each $i$.
-Hence, in interpreting the doubly-constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$ values (i.e., the sum of values at all three $j$ destinations for each origin $i$) ould be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity accessible for Zones 1, 2 and 3 respectively.
+Hence, in interpreting the doubly constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$ values (i.e., the sum of values at all three $j$ destinations for each origin $i$) would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity accessible for Zones 1, 2 and 3 respectively.
 This approximately equal to the number of population at each of these zones.
@@
 For the other two travel behaviour, identical $V_i^D$ and $M_j^D$ values are calculated, following the same logic.
-Hence, the usefulness of the doubly-constrained measure lies in the interpretation as $V_{ij}^D$ values.
+Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
 
@@
 It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
-Following the intuition discussed in the singly-constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
-
-```{r}
-#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
-#| tbl-cap: "Doubly-constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."
+Following the intuition discussed in the singly constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
+
+```{r}
+#| label: tbl-adjusted-small-system-land-use-doubly constrained-case-allfs-access-values-for-zone2
+#| tbl-cap: "Doubly constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."
 
@@
 Recall, accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
-In the doubly-constrained case, we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
+In the doubly constrained case, we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
 However, if what interests readers is the "potential for interaction" in the case population and opportunities to match one-to-one, reframing the investigation may be needed.
 In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ij}^D$).
-In this sense, the doubly-constrained case can be thought of as an estimate of _realized_ accessibility or access.
-It is also formulaically identical to the doubly-constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity', respectively.
+In this sense, the doubly constrained case can be thought of as an estimate of _realized_ accessibility or access: reflecting spatial interaction.
+It is also formulaically identical to the doubly constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity', respectively.
 
@@
 Accessibility, however, is often presented and understood as a zonal summary of *potential* for interaction between origins and destinations that contain inherently different units.
-Arriving at the doubly-constrained case through the unconstrained, totally, and singly-constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
+Arriving at the doubly constrained case through the unconstrained, totally, and singly constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
 Namely, by demonstrating that these members of the accessibility family all derive from the same root and can be derived from Wilson's original formulation, this more clearly demonstrates what *potential* is, within the context of spatial interaction modelling.
@@
 As accessibility research evolved largely influenced by @hansen1959, researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
-(1) We have demonstrated that by reintroducing such constraints into accessibility analyses, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values, make comparisons between scenarios and regions more intuitive.
-(2) Also, these constraints, depending on the case (totally-, singly- or doubly- constrained),  restrict the degree of _potential_ to spatially interact calculated, linking access to accessibility on the same continuum based on the constraint used.
-Lastly, we discussed how popular measures such as the one used in @hansen1959 and the 2SFCA link into the family.
-We summarise the family of accessibility measures, as follows:
-<!-- While methodologically clarifying work, such as @wuUnifyingAccess2020, has largely overlooked competition, this paper demonstrates how applying constraints consistent with the spatial interaction model allows for incorporating competition into accessibility.
--->
-
-We first place the popular Hansen-type accessibility measure within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
+Firstly, we have demonstrated that by reintroducing such constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
+These values become more intuitive for the purpose of analysis and comparison.
+Secondly, the constraining constants, depending on the case (i.e., total, singly- or doubly- constrained), restrict the degree of _potential_, linking accessibility (the potential for spatial interaction) with access (spatial interaction) on the same continuum based on the constraint used.
+Lastly, we discussed how popular measures such as the one used in @hansen1959 and the 2SFCA link into the family of accessibility measures.
+We summarise the family of accessibility measures, as follows.
+
+We first place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
 We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
@@
 In the unconstrained case: all zones can reach all opportunities.
-In the totally constrained case: all zones can reach all opportunities divided by the number of zones.
+In the total constrained case: all zones can reach all opportunities divided by the number of zones.
 Alternatively, if 1 zone can teleport and no other zone can travel at all, that 1 zone will get all the opportunities in the region.
@@
 
-We then introduce the singly-constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
+We then introduce the singly constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
 It can also mathematically equivalent to spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
-The singly-constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
+The singly constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
 We also introduce an 'accessible' PPR (e.g., opportunities per capita), calculated by dividing each accessibility value by the zonal population.
-To clarify, this per capita expression of the singly-constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
-
-<!-- Rafa asks, BFCA worth talking about? -->
-
-Lastly, the doubly-constrained case is introduced.
+To clarify, this per capita expression of the singly constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
+
+Lastly, the doubly constrained case is introduced.
 In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available.
@@
 
-## A note on interpretation and possible applications/ summary table?
-
-<!-- Some Table here. AS example:
-(col 1) type of measure, (col 2) explanation in plain language of the constraint, (col 3) continuum of potential (high to low), (4) equation of the constraint, (5) equation of the indicator, (col 6) input data needed?, (col 8)  opp-constrained value and interpretation of value (col 9)  pop-constrained value and interpretation of value -->
-
-<!-- rafa: in the last sentence, you clarify the difference between 'total opportunity constraint' Vs  'single opportunity constraint' .
-What is confusing here is how the single opportunity constraint compares to 'total population constraint' .
-Again, the idea of including a table would be very helpful.
-The table shoud include one indicator in each row.
-The columns could include (1) name of the indicator, (2) explanation in plain language of the constraint, (3) equation of the constraint, (4) equation of the indicator, (5) example of the interpretation of the indicator.-->
-
-unconstrainted: doesn't match the anything.
-
-totally-constrained (opportunity): opportunities at all destinations can be accessed by all zones.
-singly-constrained (opportunity): opportunities at each destination can be accessed by all populations.
-
-totally-constrained (population): population at all origins can be accessed by all zones.
-singly-constrained (population): population at each origin can be accessed by all opportunities.
-
-doubly-constrained: opportunities at each destination can be accessed by all populations, while at the same time, the population at each origin can be accessed by all opportunities.
-i.e., a proportional amount of opportunities are allocated from each dest to origins (based on population demanding and travel impedance) and a proportional amount of population is allocated from each origin to dests (based on the opportunity supply and travel impedance).
+And as a summary: each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations in Table 8.
+<!-- @tbl:summary-family-of-accessibility-measures.-->
+
+::: {=latex}
+\input{summary-family-access-measures-table.tex}
+:::
+
+Table 8: Summary of constrained accessibility measure types and interpretations
+
+<!-- {#tbl:summary-family-of-accessibility-measures} -->
 
@@
 <!-- Of note RE DOUBLE CONSTRAINT: the accessibility case methodogy should be based selected based on the input data and the theory.
-For instance, the doubly-constrained accessibility method requires pportunities and population to matched one-to-one both mathematically (i.e., the balancing factors will not converge if they do not match) and theoretically (i.e., each person should seek one opportunity).
-By forcing these assumptions, the resulting accessibility value that is calculated can be interpreted with less 'potential', than in the totally- and singly- constrained cases using the unmodified example.
+For instance, the doubly constrained accessibility method requires pportunities and population to matched one-to-one both mathematically (i.e., the balancing factors will not converge if they do not match) and theoretically (i.e., each person should seek one opportunity).
+By forcing these assumptions, the resulting accessibility value that is calculated can be interpreted with less 'potential', than in the total and singly- constrained cases using the unmodified example.
 This is because, the unmodified example inherently carries more 'potential', i.e., we only know where people are (who can potentially interact) and we know where there are opportunities (who can be potentially interacted with), but not the capacities of which this interaction happens.
@@
 
-<!-- Using the doubly-constrained measure with the unmodified example is both mathematically impossible (as there is no one-to-one match) but also theoretically uncompitable as the capacities at both ends are not known.
-Conversely, the totally- and singly- constrained cases can be used to calculate values using inputs from the modified example, but, the results will be more impresice.
-If capacities of population and opportunities are both known, and the objective of the exercise is to calculate the number of _capacity_ of the opportunity (or population) that is accessible, than by using the singly-constrained case, capacities at  either the opportunity or population would be neglected.
-In using the totally-constrained case, capacities at both would be neglected.
-In this way, if the goal was to calculated acccessibility to opportunity or population capacity, the answer would be only a more imprecise estimation than if the doubly-constrained method was used.
+<!-- Using the doubly constrained measure with the unmodified example is both mathematically impossible (as there is no one-to-one match) but also theoretically uncompitable as the capacities at both ends are not known.
+Conversely, the total and singly- constrained cases can be used to calculate values using inputs from the modified example, but, the results will be more impresice.
+If capacities of population and opportunities are both known, and the objective of the exercise is to calculate the number of _capacity_ of the opportunity (or population) that is accessible, than by using the singly constrained case, capacities at  either the opportunity or population would be neglected.
+In using the total constrained case, capacities at both would be neglected.
+In this way, if the goal was to calculated acccessibility to opportunity or population capacity, the answer would be only a more imprecise estimation than if the doubly constrained method was used.
 -->
@@
 
-<!-- ## Singly-constrained case is spatial availability -->
+<!-- ## Singly constrained case is spatial availability -->
 
@@
 
-<!-- ## Checking singly-constrained case against spatial availability -->
+<!-- ## Checking singly constrained case against spatial availability -->
 <!-- ```{r, eval=FALSE} -->
@@
 
-<!-- ## Destination- (or ATTRACTION) constrained is the only singly-constrained case... population- (or PRODUCTION) constrained doesn't make sense! -->
-
-<!-- Then the origin-constrained (singly-constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
+<!-- ## Destination- (or ATTRACTION) constrained is the only singly constrained case... population- (or PRODUCTION) constrained doesn't make sense! -->
+
+<!-- Then the origin-constrained (singly constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
 Using this method, V_i values equal P_i values...?
@@
 
-<!-- Returning to our simple example, the origin-constrained singly-constrained case would yield the following $A_{i}$ values: -->
+<!-- Returning to our simple example, the origin-constrained singly constrained case would yield the following $A_{i}$ values: -->
 <!-- $$ -->
@@
 This is because distance difference doesn't matter, everyone can reach everything and everything is given to everyone.
-But when decay is high, they are much different: since in the doubly-constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
+But when decay is high, they are much different: since in the doubly constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
 This relationship should be investigated further some other time..-->
@@
 <!--
-### Hybrid singly-constrained case: 
+### Hybrid singly constrained case: 
 NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA?
@@
 - what does k=1 mean? What does assuming k=1, so the units can be kept mean for the model? Is it meaningless?
-- not sure if the singly-constrained (PRODUCTION case) makes sense.. It sort of does..? I believe this needs to be figured out before the doubly constrained case. Pi needs to be integrated somehow. 
+- not sure if the singly constrained (PRODUCTION case) makes sense.. It sort of does..? I believe this needs to be figured out before the doubly constrained case. Pi needs to be integrated somehow. 
 - doubly constrained still needs to be solved.
@@
 For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly-constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
+Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
 
@@
 
-<!-- ## totally-constrained interaction case -->
-
-<!-- In the totally-constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
+<!-- ## total constrained interaction case -->
+
+<!-- In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
 
@@
 
-<!-- ## Singly-constrained -->
+<!-- ## Singly constrained -->
 
@@
 
-<!-- Similarly, the singly-constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel): -->
+<!-- Similarly, the singly constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel): -->
 
@@
 
-<!-- ## Doubly-constrained case -->
+<!-- ## Doubly constrained case -->
 
@@
 
-<!-- #totally-constrained -->
+<!-- #total constrained -->
 
@@
 
-<!-- #singly-constrained (orig) -->
+<!-- #singly constrained (orig) -->
 
@@
 
-<!-- #singly-constrained (dest) -->
+<!-- #singly constrained (dest) -->
 
@@
 
-<!-- #doubly-constrained (dest) -->
+<!-- #doubly constrained (dest) -->
 
@@
 
-<!-- - totally-constrained has high error -->
+<!-- - total constrained has high error -->
 

### 2025-05-16 paper/summary-family-access-measures-table.tex
draft AS - ready for preprint

{\tiny
\begin{longtable}{|p{2.5cm}|p{2.5cm}|p{2.5cm}|p{3cm}|p{3cm}|}
\hline
\textbf{Name of Case and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
\hline
\endfirsthead

\hline
\textbf{Name of Case and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
\hline
\endhead

Unconstrained Accessible Opportunities or Population ($V_i^0$, $M_j^0$)
& No constraints; marginals not preserved.
& None
& $V_i^0 = \sum_j O_j f(c_{ij})$;

$M_j^0 = \sum_i P_i f(c_{ij})$

& Impedance-weighted values (e.g., "opportunities × decay"); no marginal control. \\
\hline

Total Constrained Accessible Opportunities or Population ($V_i^T$, $M_j^T$)
& Single balancing factor preserves total marginal.

$K^T = \frac{D}{\sum_i V_i^0}$;

$\hat{K}^T = \frac{O}{\sum_j M_j^0}$

& $\kappa_i^T = \sum_j K^T f(c_{ij})$; $\hat{\kappa}_j^T = \sum_i \hat K^T f(c_{ij})$
& $V^T_i = \kappa_i^T \sum_j W^{(2)}_j$;

$M_j^T = \hat{\kappa}_j^T \sum_i W^{(1)}_i$

& Values reflect total regional opportunities or population. \\
\hline

Singly Constrained Accessible Opportunities ($V_i^S$)
& Destination-side constraint:

$B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}$
& $\kappa^S_i = W_i^{(1)} B_j$
& $V^S_i = \kappa^S_i \sum_j D_j$
& Shares of destination-side opportunities based on origin population and impedance. \\
\hline

Singly Constrained Accessible Population ($M_j^S$)
& Origin-side constraint:

$A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}$

& $\hat \kappa^S_j = W_j^{(2)} A_i$
& $M_j^S = \hat \kappa^S_j \sum_i P_i$
& Shares of origin population based on destination mass and impedance. \\
\hline

Doubly Constrained Access ($V_{ij}^D$ or $M_{ij}^D$)
& Both marginals constrained via $A_i$, $B_j$ simultaneously.
& —
& $V_{ij}^D = A_i B_j P_i O_j f(c_{ij})$
& Realized interactions between population and opportunities (i.e., access). \\
\hline

\end{longtable}
}

### 2025-05-17 paper/manuscript.qmd
final draft! To be submitted for peer review

@@
 title: "A family of accessibility measures derived from spatial interaction principles"
+subtitle: "Preprint submitted to peer-review in May 2025"
+author:
+  - name: Anastasia Soukhov 
+    email: soukhoa@mcmaster.ca
+    affiliations:
+        - name: McMaster University
+          department: School of Earth, Environment & Society
+          city: Hamilton, Canada
+  - name: Rafael H. M. Pereira
+    email: rafa.pereira.br@gmail.com
+    affiliations:
+        - name: Institute for Applied Economic Research
+          city: Brasília, Brazil
+  - name: Christopher D. Higgins
+    email: cd.higgins@utoronto.ca
+    affiliations:
+        - name: University of Toronto Scarborough
+          department: Department of Human Geography
+          city: Toronto, Canada
+    attributes:
+        corresponding: true
+  - name: Antonio Páez
+    email: paezha@mcmaster.ca
+    affiliations:
+        - name: McMaster University
+          department: School of Earth, Environment & Society
+          city: Hamilton, Canada
+format:
+  pdf:
+    documentclass: article
+    header-includes: |
+      \usepackage{draftwatermark} 
+      \SetWatermarkText{DRAFT}
+      \SetWatermarkScale{1.5}
+      \SetWatermarkColor[gray]{0.95}
+      \usepackage{float}
+      \floatplacement{table}{H}
+      \floatplacement{figure}{H}
+bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
+#includes:
+#      in_header: 'gt_packages.sty'
+      #\usepackage{float}
+
 abstract: |
-  Transportation planning has long prioritized the effiency of movement, or mobility.
+  Transportation planning has long prioritized the efficiency of movement, or mobility.
 However, the concept of accessibility represents a more comprehensive evolution, shifting focus from mere movement to the potential to reach (i.e., spatially interact) with desired destinations.
@@
 
-format:
-  pdf:
-    header-includes:
-        - \usepackage{float}
-        - \floatplacement{table}{H}
-        - \floatplacement{figure}{H}
-bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-#includes:
-#      in_header: 'gt_packages.sty'
-      #\usepackage{float}
+keywords: [accessibility; spatial interaction model; balancing factors; competitive accessibility; constraints; singly constrained; doubly constrained; total constrained]
 ---
@@
 
-In this way, this work aims to offer three contributions: (1) introduce and (2) formulaically define a family of _constrained_ accessibility measures, namely the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure.
+In this way, this work aims to offer three contributions: (1) introduce and (2) define a family of _constrained_ accessibility measures, namely the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure.
 These measures are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential, or what Reilly called 'potential trade territories' [@reilly1929methods] and further developed in other examples [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
@@
 The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
-For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{\beta\cdot m}$.
+For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{-\beta\cdot m}$.
 This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
@@
 The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations.
-The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., population as $W_i^(1)$ and hectares of parkspace as $W_j^(2)$.
+The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., $W_i^{(1)}$ could be populations, and $W_j^{(2)}$ hectares of park space.
 Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
@@
 $K$ is specified depending on the applied constraint(s).
-In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where no constraints apply, the second being a singly constrained case, and the third a doubly constrained case.
+In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where to results only match the total volume of interaction, the second being a singly constrained case, and the third a doubly constrained case.
 
@@
                locations = cells_column_labels(columns = D_j)) |>
+  tab_options(table.font.size = px(10))|>
   as_latex()
@@
              columns = 2:4) |>
+  tab_options(table.font.size = px(10))|>
   as_latex()
@@
     fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small") |>
+  tab_options(column_labels.font.size = "small",
+              table.font.size = px(10)) |>
   as_latex()
@@
 $$
-\kappa_i^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+\kappa_i^T = \frac{\sum_j D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $$
@@
   ) |>
-  tab_options(column_labels.font.size = "small") |>
+  tab_options(column_labels.font.size = "small",
+              table.font.size = px(10)) |>
   grand_summary_rows(
@@
 $$
-\hat \kappa_i^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+\hat \kappa_i^T = \frac{\sum_i O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 $$
@@
   #  fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small")  |>
+  tab_options(column_labels.font.size = "small",
+              table.font.size = px(10))  |>
   as_latex()
@@
 \kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
-$${eq-doubly-constrained-proportional-allocation-factor-forpop}
+$$
 
@@
 
-However, $V^D_{i}$ is not wholy helpful, as it will simply equal the origin-mass marginal.
+However, $V^D_{i}$ is not wholly helpful, as it will simply equal the origin-mass marginal.
 In this way, only $V^D{ij}$ values are interpretable.
@@
 
-So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
 For the population, we leave this unchanged numerically but theoretically know that each person interacts with one physician capacity (i.e., our opportunities with a _capacity_).
@@
 ```{r}
-#| label: tbl-small-system-land-use-doubly constrained-case
+#| label: tbl-small-system-land-use-doubly-constrained-case
 #| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."
@@
              columns = 2:3) |>
+  tab_options(table.font.size = px(10)) |>
   as_latex()
@@
 
-  tab_options(column_labels.font.size = "small") |>
+  tab_options(column_labels.font.size = "small",
+              table.font.size = px(10)) |>
   as_latex()
@@
 ```{r}
-#| label: tbl-adjusted-small-system-land-use-doubly constrained-case-allfs-access-values-for-zone2
+#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
 #| tbl-cap: "Doubly constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."
@@
   ) |>
-  tab_options(column_labels.font.size = "small") |>
+  tab_options(column_labels.font.size = "small",
+              table.font.size = px(10)) |>
   as_latex()
@@
 We then introduce the singly constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
-It can also mathematically equivalent to spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
+It is also mathematically equivalent to the spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.

### 2025-05-17 paper/summary-family-access-measures-table.tex
final draft! To be submitted for peer review

@@
 
-Unconstrained Accessible Opportunities or Population ($V_i^0$, $M_j^0$)
-& No constraints; marginals not preserved.
+Unconstrained Accessible Opportunities ($V_i^0$) and Unconstrained Accessibile Population ($M_j^0$)
+& No constraints; marginals not equal to any regional or zonal knowns.
 & None
@@
 
-& Impedance-weighted values (e.g., "opportunities × decay"); no marginal control. \\
+& Values in various units depending on the impedance and destination-mass (e.g., "opportunities x decay") for $V_i^0$ andimpedance and origin-mass (e.g., "population x decay"); no total or marginal constraint \\
 \hline
 
-Total Constrained Accessible Opportunities or Population ($V_i^T$, $M_j^T$)
-& Single balancing factor preserves total marginal.
+Total Constrained Accessible Opportunities ($V_i^T$) and Total Constrained Accessibile Population ($M_j^T$)
+& Balancing factors $K^T$ and $\hat{K}^T$ ensures the sum of $ij$ values equals the total marginal, where:
 
@@
 
-& $\kappa_i^T = \sum_j K^T f(c_{ij})$; $\hat{\kappa}_j^T = \sum_i \hat K^T f(c_{ij})$
+& Allocates the total marginal as opportunities based on $\kappa_i^T = \sum_j K^T f(c_{ij})$ and as population based on  $\hat{\kappa}_j^T = \sum_i \hat K^T f(c_{ij})$
+
 & $V^T_i = \kappa_i^T \sum_j W^{(2)}_j$;
@@
 
-& Values reflect total regional opportunities or population. \\
+& Values reflect a share of total regional opportunities ($V^T_i$) or population ($M^T_j$). \\
 \hline
 
-Singly Constrained Accessible Opportunities ($V_i^S$)
-& Destination-side constraint:
+Singly Constrained Accessible Opportunities ($V_i^S$) and Singly Constrained Accessible Population ($M_j^S$)
+& Single balancing factor $B_j$ (for $V_i^S$) that ensures the destination-mass marginal is constrained, and $A_i$ (for $M_j^S$) ensures the origin-mass marginal is constrained:
 
-$B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}$
-& $\kappa^S_i = W_i^{(1)} B_j$
-& $V^S_i = \kappa^S_i \sum_j D_j$
-& Shares of destination-side opportunities based on origin population and impedance. \\
-\hline
-
-Singly Constrained Accessible Population ($M_j^S$)
-& Origin-side constraint:
+$B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}$;
 
@@
 
-& $\hat \kappa^S_j = W_j^{(2)} A_i$
-& $M_j^S = \hat \kappa^S_j \sum_i P_i$
-& Shares of origin population based on destination mass and impedance. \\
+& Allocates the single opportunities marginal proportionally based on $\kappa^S_i = W_i^{(1)} B_j$ in the case of $V_i^S$ and the single population marginal proroportionally based on $\hat \kappa^S_j = W_j^{(2)} A_i$ in the case of $M_j^S$
+
+& $V^S_i = \kappa^S_i \sum_j D_j$;
+
+$M_j^S = \hat \kappa^S_j \sum_i P_i$
+
+& $V^S_i$ values reflect a share of the opportunities at each destination based on origin population 'demand' and impedance; $M^S_j$ values reflect a share of the population at each origin based on destination opportunities 'supply' and impedance.\\
 \hline
@@
 Doubly Constrained Access ($V_{ij}^D$ or $M_{ij}^D$)
-& Both marginals constrained via $A_i$, $B_j$ simultaneously.
+& Values reflect both single marginals simutaneously, maintained via $A_i$ and $B_j$.
 & —
 & $V_{ij}^D = A_i B_j P_i O_j f(c_{ij})$
-& Realized interactions between population and opportunities (i.e., access). \\
+& The spatial interactions between population and opportunities (i.e., access). \\
 \hline

### 2025-05-17 paper/manuscript.qmd
added .tex file

@@
   pdf:
+    keep-tex: true
     documentclass: article

### 2025-05-17 paper/manuscript.tex
added .tex file

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1.5}
\SetWatermarkColor[gray]{0.95}
\usepackage{float}
\floatplacement{table}{H}
\floatplacement{figure}{H}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  pdfkeywords={accessibility; spatial interaction model; balancing
factors; competitive accessibility; constraints; singly constrained;
doubly constrained; total constrained},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A family of accessibility measures derived from spatial
interaction principles}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Preprint submitted to peer-review in May 2025}
\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
Higgins \and Antonio Páez}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Transportation planning has long prioritized the efficiency of movement,
or mobility. However, the concept of accessibility represents a more
comprehensive evolution, shifting focus from mere movement to the
potential to reach (i.e., spatially interact) with desired destinations.
Despite growing recognition of accessibility-based planning approaches,
the concept remains fragmented, with inconsistent definitions and
unclear interpretations. This work's aim is to clarify and unify the
concept of accessibility by connecting it into spatial interaction
modeling. We demonstrate that widely used mobility and accessibility
models, such as gravity-based accessibility and spatial interaction
models, share common theoretical roots. From this foundation, we define
a family of accessibility measures within the principles of spatial
interaction. In demonstrating these developments, three contributions
emerge: A) the definition and linkages between four members of the
family: specifically, the `unconstrained' measure (i.e., Hansen-type
measure), the `total constrained' measure (i.e., a constrained version
of the Hansen-type measure), the `singly constrained' measure (i.e.,
related to the popular two step floating catchment approach (2SFCA)),
and the `doubly constrained' measure representing realized interactions
or `access', effectively equal to the doubly constrained spatial
interaction model in formulation. B) How to formulaically specify the
total-, singly- and doubly- constrained accessibility measures. C) The
interpretability advantages of the family, as these constrained
accessibility measures yield values in units of the number of potential
``opportunities for spatial interaction'' or ``population for spatial
interaction'' for each zone and zonal flow. By clarifying the concept of
`potential' in accessibility, demonstrating theoretical and formulaic
linkages across popular accessibility and spatial interaction models,
and reintroducing units, this work contributes a family of measures that
unlock a clearer, more interpretable, and cohesive foundation for
accessibility analysis.
\end{abstract}

\section{Introduction}\label{introduction}

In the early nineteenth century, industrializing cities grew rapidly: so
did traffic congestion and the development of transportation planning
practice focused primarily on mobility. In this newly founded practice,
access to destinations was treated as a by-product of movement.
Following the rise of the automobile and significant investments in
transportation infrastructure after World War II, this mobility-oriented
approach has led to some problematic outcomes. Specifically, the car
became seen as the ultimate mobility tool, helping to foster the
development of low-density, single-use residential neighborhoods and
entrenching an automobility mono-culture within transportation systems
(H. J. Miller 2011; Lavery, Páez, and Kanaroglou 2013). Decades of
planning for this automobility mono-culture, still often characterized
by road and highway expansion, have been marked by increased travel
costs and environmental burdens, with limited impact on enhancing the
ease with which people can reach destinations (Steven Farber and Páez
2011; S. Handy 2002; Páez et al. 2010). In response, transportation
researchers have increasingly advocated for the adoption of
accessibility as a planning criterion, in contrast to traditional
mobility-oriented transportation planning approaches which translate
into indicators that benchmark movement (e.g., vehicle kilometres
traveled, intersection through traffic, etc.) which are not necessarily
linked to improved accessibility (Silva et al. 2017; Paez et al. 2013;
S. Handy 2020; El-Geneidy and Levinson 2022).

Accessibility can be seen to represent an evolution of mobility;
accessibility may be defined as the ``potential of opportunities for
{[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
is simply, movement in itself or `spatial interaction'. Mobility has
been the basis of traditional transportation planning approaches
(Ortúzar and Willumsen 2011). While accessibility brings a more holistic
understanding of combined transportation and land use systems,
incorporating both the explicit consideration of destination as
opportunity and the concept of \emph{potential} (S. L. Handy and
Niemeier 1997).

The growing interest in accessibility has been accompanied by a boom in
scholarly research using different methods and focusing on different
research contexts; it has grown to include studies of access to
employment (e.g., Karst and Van Eck 2003; Grengs 2010; Páez et al. 2013;
Merlin and Hu 2017; Tao et al. 2020), health care (e.g., Luo and Wang
2003; Páez et al. 2010; Wan, Zou, and Sternberg 2012; Delamater 2013;
Boisjoly, Moreno-Monroy, and El-Geneidy 2017; Pereira et al. 2021),
green spaces (Reyes, Paez, and Morency 2014; Rojas et al. 2016; Liang,
Yan, and Yan 2024), schools (e.g., Williams and Wang 2014; Romanillos
and Garcia-Palomares 2018; Marques, Wolf, and Feitosa 2021), social
contacts (e.g., Neutens et al. 2007; S. Farber, Páez, and Morency 2012;
S. Farber et al. 2013), and regional economic analysis (e.g., R.
Vickerman, Spiekermann, and Wegener 1999; Lopez, Gutierrez, and Gomez
2008; Ribeiro, Antunes, and Páez 2010; Gutierrez et al. 2011) among many
other domains of application. In other words, accessibility analysis is
used today to broadly understand the potential to reach (or spatially
interact with) opportunities that are important to people (Ferreira and
Papa 2020). However, despite its growth in popularity in scholarly
works, challenges remain with respect to the more widespread adoption of
accessibility in planning practice. For instance, the diversity of
accessibility definitions has been flagged by van Wee (2016), S. Handy
(2020), and Kapatsila et al. (2023). Further, difficulties in the
interpretability and communicability of outputs has also been noticed by
Karst T. Geurs and van Wee (2004), van Wee (2016), and Ferreira and Papa
(2020).

Furthermore, the adoption of accessibility in planning practice is not
necessarily made easier when potential adopters have to contend with a
plethora of definitions, each seemingly more sophisticated but less
intuitive than the last (Kapatsila et al. 2023). At a high level, Karst
T. Geurs and van Wee (2004) identify four families of accessibility
measures: infrastructure-, place-, person-, and utility-based. Of the
place-based family, which is the focus of this paper, the menu has grown
to include gravity-based accessibility (e.g., Hansen 1959; Pirie 1979),
cumulative opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye
et al. 2018), modified gravity (e.g., Schuurman, Berube, and Crooks
2010), 2-Step Floating Catchment Areas (e.g., Luo and Wang 2003),
Enhanced 2-Step Floating Catchment Areas (e.g., Luo and Qi 2009),
3-Stage Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012),
Modified 2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse
2-Step Floating Catchment Areas (e.g., F. Wang 2021), and n-steps
Floating Catchment Areas (Liang, Yan, and Yan 2024). How is a
practitioner to choose among this myriad options? What differences in
accessibility scores should matter, and how should they be communicated?
{[}see van Wee (2016); p.~14{]}.

More specifically, this work seeks to address the breath of place-based
accessibility's definitions and lack of interpretability by
demonstrating that mobility-oriented models--such as the commonly used
gravity-based accessibility (e.g., Hansen 1959) and the spatial
interaction model (e.g., Wilson 1971)-- are rooted in the same spatial
interaction modeling foundation. In fact, accessibility can be specified
using spatial interaction principles and a \emph{family of measures},
akin to the family of spatial interaction models introduced in Wilson
(1971), can be defined using balancing factors that constrains values
based on known information about the land-use.

In this way, this work aims to offer three contributions: (1) introduce
and (2) define a family of \emph{constrained} accessibility measures,
namely the total constrained accessibility measure, the singly
constrained accessibility measure, and the doubly constrained measure.
These measures are explicitly connected to popular measures such as the
Hansen-type accessibility (Hansen 1959), the popular competition
approach of the 2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a;
Luo and Wang 2003), and the concept of market potential, or what Reilly
called `potential trade territories' (Reilly 1929) and further developed
in other examples (C. D. Harris 1954; R. W. Vickerman 1974). These
Wilson-analogous balancing factors introduced are also defined in order
of increasing restrictiveness to shed light on the role of
\emph{potential} in accessibility and access. (3) This work also aims to
demonstrate the intuitive value of restoring units to the resulting raw
accessibility values. Each zonal and zonal flow value from a constrained
accessibility measure is in interpretable units, namely the number of
`opportunities for spatial interaction' or `population for spatial
interaction'. This is in contrast to conventional accessibility measures
that yield values in units of `opportunities weighted by some
representation of travel friction'.

To achieve these stated objectives, this paper contends that
accessibility research must reconnect with its spatial interaction
origins. Particularly, we argue that an important aspect of spatial
interaction modelling--namely, constraining the results to match
empirical observations--was never effectively reincorporated into
accessibility analysis. Empirical constraints were embraced by early
spatial interaction literature following the work of Wilson (1971), but
this stream of literature tended to flow separately from research
inspired by Hansen (1959)' accessibility. The application of Wilson
(1971)`s empirical constraints supported the development of various
spatial interaction models that remain relevant in research and practice
today (Ortúzar and Willumsen 2011). However, the same cannot be said of
the contemporary accessibility literature, where empirical constraints
were not explicitly adopted. We argue that the absence of empirical
constraints (and their attendant proportionality constants) has
contributed to some of accessibility analysis' interpretability issues;
for instance, the fuzziness of insights beyond simple proportional
statements like `higher-than' or `lower-than' (E. J. Miller 2018).
Moreover, without constraints: we lose track of what accessibility
measures, that is, the number of opportunities that can be spatially
interacted with. Without a clear sense of measurement units,
comparability between accessibility measures, across cities, and
transport modes, may also become compromised.

Fortunately, accessibility and spatial interaction modelling literature
share common headwaters, and the latter has given careful attention to
measurement units and their interpretability. It is by looking to the
past that we believe accessibility analysis can newly wade into the
future. We continue this work in the following section by tracing the
development of accessibility from its origins in spatial interaction:
from the Newtonian gravitational expression in Ravenstein (1889) through
to the seminal accessibility work of Hansen (1959). We then present
evidence for a narrative highlighting the marked divergence between
accessibility and spatial interaction modelling research after the work
of Wilson (1971). Next, we hark back to Wilson (1971)'s spatial
interaction models, and use them to derive a family of accessibility
measures based analogous constraints. We illustrate members of this
family of constrained accessibility measures with a simple numerical
example. We then conclude by discussing the uses of these measure and
their interpretation.

\section{Newtonian's roots of human spatial interaction
research}\label{newtonians-roots-of-human-spatial-interaction-research}

The patterns of people's movement in space have been a subject of
scientific inquiry for at least a century and a half. From as far back
as Henry C. Carey's \emph{Principles of Social Science} (Carey 1858), a
concern with the scientific study of human spatial interaction can be
observed. It was in this work where Carey stated that ``man {[}is{]} the
molecule of society {[}and their interaction is subject to{]} the direct
ratio of the mass and the inverse one of distance'' (McKean 1883,
37--38). This statement shows how investigations into human spatial
interaction have often been explicitly coloured by the features of
Newton's Law of Universal Gravitation, first posited in 1687's
\emph{Principia Mathematica} and expressed as in
Equation~\ref{eq-phys-grav-prop}.

\begin{equation}\phantomsection\label{eq-phys-grav-prop}{
F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
}\end{equation}

To be certain, the expression above is one of proportionality, and is
also the most famous in all of science. It states that the force of
attraction \(F\) between a pair of bodies \(i\) and \(j\) is directly
\emph{proportional} to the product of their masses \(M_i\) and \(M_j\),
and inversely \emph{proportional} to the square of the distance between
them \(D_{ij}\). Direct proportionality means that as the product of the
masses increases, so does the force. Likewise, inverse proportionality
means that as the distance increases, the force decreases.
Equation~\ref{eq-phys-grav-prop}, however, does not quantify the
magnitude of the force. To do so, an empirical constant, a.k.a. the
gravitational constant, is required to convert the proportionality into
an equality, ensuring that values of the force \(F\) in
Equation~\ref{eq-phys-grav-prop} match the observed force of attraction
between masses. In other words, Equation~\ref{eq-phys-grav-prop} needs
to be \emph{constrained} using empirical data. Ultimately, the equation
for the force is as seen in Equation~\ref{eq-phys-grav}, where \(G\) is
an empirically calibrated proportionality constant:

\begin{equation}\phantomsection\label{eq-phys-grav}{
F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
}\end{equation}

Newton's initial estimate of \(G\) was based on a speculation that the
mean density of earth was between five or six times that of water, an
assumption that received support after Hutton's experiments of 1778
(Hutton 1778, 783). Still, it took over a century from the publication
of \emph{Principia} to refine the estimate of the proportionality
constant to within 1\% accuracy, with Cavendish's 1798 experiment
(Cavendish 1798).

\section{Early research on human spatial interaction: from Ravenstein
(1889) to Stewart
(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1889-to-stewart-1948}

Since the 1880s to the 1940s, a number of researchers theoretically and
empirically attempted to characterise human spatial interaction as some
force of attraction \(F\) that is directly proportion to the masses
\(M_i\) and \(M_j\) and inversely proportional by their separation
distance. This concept was captured with different expressions, but all
tie back to the same Newtonian gravity analogy, although not all of them
included a proportionality constant in their formulation.

Following Carey's \emph{Principles} of 1858, research into human spatial
interaction continued in different contexts. In the late 1880s,
Ravenstein proposed some ``Laws of Migration'' based on his empirical
analysis of migration flows in various countries (Ravenstein 1885,
1889). In these works, Ravenstein posited 1) a directly proportional
relationship between migration flows and the size of destinations (i.e.,
centres of commerce and industry), and 2) an inversely proportional
relationship between the size of flows and the separation between
origins and destinations. As with Carey, these propositions echo
Newton's gravitational laws. Over time, other researchers discovered
similar relationships. For example, Reilly (1929) formulated a law of
retail gravitation, expressed in terms of equal attraction to competing
retail destinations that could be understood as `potential trade
territories'. Later, Zipf proposed a \(\frac{P_1P_2}{D}\) hypothesis for
the case of information (Zipf 1946a), intercity personal movement (Zipf
1946b), and goods movement by railways (Zipf 1946c). The
\(\frac{P_1P_2}{D}\) hypothesis stated that the magnitude of flows was
proportional to the product of the populations of settlements, and
inversely proportional to the distance between them.

A common feature of these early investigations of human spatial
interaction is that a proportionality constant similar to \(G\) in
Equation~\ref{eq-phys-grav} was never considered. Of the researchers
cited above, only Reilly and Zipf expressed their hypotheses in
mathematical terms. Reilly's hypothesis was presented in the following
form:

\begin{equation}\phantomsection\label{eq-reilly}{
B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
}\end{equation}

\noindent where \(B_a\) is the amount of business drawn to \(a\) from
\(T\), \(P_a\) and \(P_T\) are the populations of the two settlements,
and \(D_{aT}\) is the distance between them. Quantity \(N\) was chosen
by Reilly in a somewhat \emph{ad hoc} fashion as 1, and he used
empirical observations of shoppers to choose a value of \(n = 2\).

Zipf, on the other hand, wrote his hypothesis in mathematical form as:

\begin{equation}\phantomsection\label{eq-zipf}{
C^2 = \frac{P_1\cdot P_2}{D_{12}}
}\end{equation}

\noindent where \(C\) is the volume of goods exchanged between \(1\) and
\(2\), \(P_1\) and \(P_2\) are the populations of the two settlements,
and \(D_{12}\) is the distance between them.

After Carey, it is in Stewart's work on the principles of demographic
gravitation that we find the strongest connection yet to Newton's law
(Stewart 1948). This may relate to academic backgrounds; where Stewart
was a physicist while Ravenstein, Reilly, and Zipf were social
scientists. Besides awareness of preceding research (he cites both
Reilly and Zipf as predecessors in the analysis of human spatial
interaction), Stewart appears to have been the first author to express
his theorized relationships for human spatial interaction with a
proportionality constant \(G\), as follows:

\begin{equation}\phantomsection\label{eq-stewart-force}{
F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
}\end{equation}

\noindent Where:

\begin{itemize}
\tightlist
\item
  \(F\) is the \emph{demographic force}
\item
  \(N_1\) and \(N_2\) are the numbers of people of in groups 1 and 2
\item
  \(\mu_1\) and \(\mu_2\) are so-called \emph{molecular weights}, the
  attractive weight of groups 1 and 2
\item
  \(M_1 = \mu_1N_1\) and \(M_2 = \mu_2N_2\) are the demographic masses
  at 1 and 2
\item
  \(d_{12}^2\) is the distance between \(1\) and \(2\)
\item
  And finally \(G\), a constant that Stewart ``left for future
  determination'' (1948, 34)
\end{itemize}

In addition to demographic force, Stewart defined a measure of the
``population potential'' of group \(2\) with respect to group \(1\). In
other words, the potential number of people from location \(2\) that
could visit location \(1\), as follows:

\begin{equation}\phantomsection\label{eq-stewart-population-potential}{
V_1 = G\frac{M_2}{d_{12}}
}\end{equation}

For a system with more than two population bodies, Stewart formulated
the population potential at \(i\) as follows (after arbitrarily assuming
that \(G=1\)):

\begin{equation}\phantomsection\label{eq-stewart-population-potential-integral}{
V_i = \int\frac{D}{r} ds
}\end{equation}

\noindent where \(D\) is the population density over an infinitesimal
area \(ds\) and \(r\) is the distance to \(i\). In
Equation~\ref{eq-stewart-population-potential-integral}, \(D\cdot ds\)
gives an infinitesimal count of the population, say \(dm\), and so,
after discretizing space,
Equation~\ref{eq-stewart-population-potential-integral} can be rewritten
as:

\begin{equation}\phantomsection\label{eq-stewart-population-potential-sum}{
V_i = \sum_j M_jd_{ij}^{-1}
}\end{equation}

Alerted readers will notice that
Equation~\ref{eq-stewart-population-potential-sum}, with some
re-organization of terms, is formally equivalent to our modern
definition of accessibility.

Stewart's formulation of demographic force, developed in the context of
what he called ``social physics'' (Stewart 1947), was problematic. It
had issues with inconsistent mathematical notation. More seriously
though, Stewart's work was permeated by a view of humans as particles
following physical laws, but tinted by unscientific and racist ideas.
For instance, he assumed that the molecular weight \(\mu\) of the
average American was one, but ``presumably\ldots much less than
one\ldots.for an Australian aborigine'' {[}p.~35{]}. Stewart's ideas
about ``social physics'' soon fell out of favour among social
scientists, but not before influencing the nascent field of
accessibility research, as detailed next.

\section{Hansen's gravity-based accessibility to
today}\label{grav-to-today}

From Stewart (1948), we arrive to 1959 and Walter G. Hansen, whose work
proved to be exceptionally influential in the accessibility literature
(Hansen 1959). In this seminal paper, Hansen defined accessibility as
``the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart (1948)''
(p.~73). As well as being a student of city and regional planning at the
Massachusetts Institute of Technology, Hansen was also an engineer with
the Bureau of Roads, and preoccupied with the power of transportation to
shape land uses in a very practical sense. Hansen (1959) focused on
Stewart (1948)'s \emph{population potential} (expressed in
Equation~\ref{eq-stewart-population-potential-sum}), and not on the
other formulaic contributions and objectionable aspects of ``social
physics''. Hansen (1959) recast Stewart's population potential to
reflect accessibility, a model of human behaviour useful to capture
regularities in mobility patterns. Hansen (1959) replaced \(M_j\) in
Equation~\ref{eq-stewart-population-potential-sum} with
\emph{opportunities} to derive an \emph{opportunity potential}, or more
accurately, a \emph{potential of opportunities for interaction} as
follows:

\begin{equation}\phantomsection\label{eq-accessibility}{
S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
}\end{equation}

A contemporary rewriting of Equation~\ref{eq-accessibility} accounts for
a variety of impedance functions beyond the inverse power
\(d^{-\beta}\):

\begin{equation}\phantomsection\label{eq-accessibility-general}{
S_{i} = \sum_j O_j \cdot f(d_{ij})
}\end{equation}

\(S_{i}\) in Equation~\ref{eq-accessibility} is a measure of the
accessibility of site \(i\). This is a function of \(O_j\) (the mass of
opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or travel
time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a parameter
that modulates the friction of cost). Today, Hansen is frequently cited
as the father of modern accessibility analysis (e.g., Reggiani and
Martín 2011), and Hansen-type accessibility is commonly referred to as
the gravity-based accessibility measure.

However, Hansen's use of Stewart's \emph{population potential} measure
included one crucial omission that afflicts the literature to this day.
The omission is that between Stewart (1948) and Hansen (1959), the
proportionality constant \(G\) in
Equation~\ref{eq-stewart-population-potential} vanished. This constant
was not explicitly addressed in Hansen (1959) and accessibility research
continues to evolve without it. Since Hansen (1959), accessibility
analysis has been widely used in numerous disciplines but, to our
knowledge, the proportionality constant has remained implicit or
forgotten, with no notable developments to explicitly acknowledge or
determine it.

The omission of this constant generates a fundamental problem for the
measurement unit of accessibility estimates, which has fundamental
implications for the interpretation, communication and comparability of
accessibility analysis. Those reading Hansen (1959) must recall that
Stewart (1948) had set the proportionality constant \(G\) to 1, with a
note that ``\(G\) {[}was{]} left for future determination: a suitable
choice of other units can reduce it to unity'' {[}p.~34{]}. In practice,
the persistent omission of the constant in accessibility analyses means
that \(G\) continues to be implicitly set to \(1\), even when the
fundamental relationship in accessibility is proportionality (e.g.,
\(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality (for
instance, see the formula for accessibility at the top of Figure 1 in Wu
and Levinson 2020). The direct consequence is that without a
proportionality constant, the units of \(S_i\) remain unclear: the unit
of ``potential of opportunity for interaction'' is left free to change
as \(\beta\) is calibrated. For example, if \(c_{ij}\) is distance in
meters, it will be number of opportunities per \(m^{\beta}\) when
\(f(c_{ij}) = d^{-\beta}\) but number of opportunities per
\(e^{-\beta\cdot m}\) when \(f(c_{ij}) = e^{-\beta\cdot m}\). This
undermines the comparability of accessibility metrics with different
decay functions, and renders their results difficult to understand and
communicate. Hansen-style accessibility, therefore, is better thought of
as an ordinal measure of potential that can only be interpreted in terms
of higher and lower accessibility, but which has not palpable meaning
(E. J. Miller 2018).

\section{Wilson's family of spatial interaction
models}\label{wilsons-family-of-spatial-interaction-models}

On the other side of the Atlantic, Alan G. Wilson was developing
related, yet parallel work. In his groundbreaking study (Wilson 1971),
Wilson defined a general spatial interaction model as follows. While
accessibility was characterised as an associated concept of `potential',
the primary focus was on modelling spatial interaction:
\begin{equation}\phantomsection\label{eq-phys-gravity-model}{
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
}\end{equation}

The model in Equation~\ref{eq-phys-gravity-model} posits a quantity
\(T_{ij}\) that represents a value in a matrix of flows of size
\(n \times m\), that is, between \(i = 1,\cdots, n\) origins and
\(j = 1,\cdots, m\) destinations. The quantities \(W_i^{(1)}\) and
\(W_j^{(2)}\) are proxies for the masses at \(i=1,\cdots,n\) origins and
\(j=1,\cdots,m\) destinations. The super-indices \((1)\) and \((2)\) are
meant to indicate that these masses could be different things, i.e.,
\(W_i^{(1)}\) could be populations, and \(W_j^{(2)}\) hectares of park
space. Finally, \(f(c_{ij})\) is some function of travel cost \(c_{ij}\)
which reflects travel impedance. In this way, \(T_{ij}\) explicitly
measures \emph{interaction} in the unit of trips, and the role of \(k\)
is to ensure that the system-wide sum of \(T_{ij}\) represents the total
flows in the data. In other words, \(k\) is a scale parameter that makes
the overall amount of flows identical to the magnitude of the phenomenon
being modeled. In other words, it balances the units, in a conceptually
similar sense as the gravitational constant in Newton's Law of Universal
Gravitation.

Traditionally, development of the spatial interaction model put an
emphasis on the interpretability of the results (Kirby 1970; Wilson
1967, 1971). But instead of relying on the heuristic of Newtonian
gravity (e.g., some interaction between a mass at \(i\) and a mass at
\(j\) separated by some distance), Wilson's approach was to maximise the
entropy of the system. Entropy maximisation in this case achieves stable
results as a statistical average that represents the population. The
approach works by assuming undifferentiated individual interactions, and
assessing their probabilities of making a particular journey. The result
of Equation~\ref{eq-phys-gravity-model} then is a statistical average
(Wilson 1971; Senior 1979).

To ensure that \(T_{ij}\) in Equation~\ref{eq-phys-gravity-model} is in
the unit of trips (our unit of origin-destination spatial interaction),
additional knowledge about the system is required. At the very least,
this framework assumes that the total number of trips in the system
\(T\) is known, and therefore:

\begin{equation}\phantomsection\label{eq-constraint0-gravitymodel}{
\sum_i\sum_j T_{ij} = T
}\end{equation}

Additional information can be introduced. For example, when information
is available about the total number of trips produced by each origin,
\(W_i^{(1)}\) is represented as \(O_i\) and the following constraint can
be used:

\begin{equation}\phantomsection\label{eq-constraint1-gravitymodel}{
\sum_j T_{ij} = O_i
}\end{equation}

Alternatively, if there is information available about the total number
of trips attracted by each destination, \(W_j^{(2)}\) is represented as
\(D_j\) and the following constraint can be used:

\begin{equation}\phantomsection\label{eq-constraint2-gravitymodel}{
\sum_i T_{ij} = D_j
}\end{equation}

It is also possible to have information about both \(O_i\) and \(D_j\),
in which case both constraints could be imposed on the model at once.

Using information about the system that satisfy these constraints fully,
partially or not at all, a family of spatial interaction models can be
derived based on Equation~\ref{eq-phys-gravity-model}. \(K\) is
specified depending on the applied constraint(s). In the framework
introduced in Wilson (1971), three constrained versions are specified:
the first being a case where to results only match the total volume of
interaction, the second being a singly constrained case, and the third a
doubly constrained case.

In the first, Equation~\ref{eq-constraint1-gravitymodel} and
Equation~\ref{eq-constraint2-gravitymodel} do not hold. In practical
terms, this means that the total number of trips predicted by the model
must be equal to sum of all flows from origins i to destinations j. The
balancing constant \(K\) in this case is (see Cliff, Martin, and Ord
1974; A. S. Fotheringham 1984):

\begin{equation}\phantomsection\label{eq-total-flow-balancing-factor}{
K=\frac{T}{\sum_i\sum_j T_{ij}}
}\end{equation}

In the second case, only one of
Equation~\ref{eq-constraint1-gravitymodel} or
Equation~\ref{eq-constraint2-gravitymodel} hold. The resulting models
are, in Wilson's terms, singly constrained. When only
Equation~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation
leads to the following production-constrained model:

\begin{equation}\phantomsection\label{eq-production-constrained-gravitymodel}{
T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
}\end{equation}

Notice how, in this model, the proxy for the mass at the origin
\(W_i^{(1)}\) is replaced with \(O_i\), representing what we know about
the system, the spatial interaction outbound flow, i.e., outbound trips
produced at \(i\). Also, there is no longer a single system-wide
proportionality constant, but rather a set of proportionality constants
(i.e., balancing factors) specific to origins. For this model, the
balancing factors ensure that Equation~\ref{eq-constraint1-gravitymodel}
is satisfied, meaning that the sum of predicted flows from one origin
going to all destinations must equal the known mass at that origin
\(O_i\) i.e., the total number of outbound trips. Satisfying this
constraint also implicitly fulfills the total constraint
(Equation~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\)
values across all origins equals the total number of trips. This model
is useful when trips ends are unknown but the number of trips
originating from each location is known and the total of these trips
represents all trips in the system. The balancing factors for the
production-constrained model are solved for each origin \(A_i\), and
according to Wilson are:

\begin{equation}\phantomsection\label{eq-production-constrained-balancing-factor}{
A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
}\end{equation}

The attraction-constrained model is similar to the
production-constrained model as it is also singly constrained but from
the perspective of the mass at the destination. From the
attraction-constrained model, the proxy for the mass at the destination
\(W_j^{(2)}\) is now replaced with \(D_j\), representing the spatial
interaction inbound flow, i.e., trips attracted at the destination and
takes the following form:

\begin{equation}\phantomsection\label{eq-attraction-constrained-gravitymodel}{
T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
}\end{equation}

In this model (Equation~\ref{eq-attraction-constrained-gravitymodel}),
the balancing factors ensure that
Equation~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
constraint Equation~\ref{eq-constraint0-gravitymodel} is as well),
meaning that the sum of predicted flows going to one destination from
all origins must equal the known mass of that destination \(D_j\) i.e.,
the total number of inbound trips to \(j\). This should hold for all
destinations. As before, destination-specific proportionality constants
(i.e., balancing factors) \(B_j\) were derived by Wilson as:

\begin{equation}\phantomsection\label{eq-attraction-constrained-balancing-factor}{
B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

In the third case in the family of spatial interaction models is the
production-attraction constrained model. In this case, both
Equation~\ref{eq-constraint1-gravitymodel} and
Equation~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and takes
the following form:

\begin{equation}\phantomsection\label{eq-doubly-constrained-gravitymodel}{
T_{ij} = A_i B_j O_i D_j f(c_{ij})
}\end{equation}

In this model, both proxies for the masses are replaced with the known
masses, that is, the trips produced by the origin and the trips
attracted by the destination. There are now two sets of mutually
dependent proportionality constants:

\begin{equation}\phantomsection\label{eq-doubly-constrained-balancing-factors}{
\begin{array}{l}
A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
\end{array}
}\end{equation}

Derivation of these models is demonstrated in detail elsewhere (e.g.,
Ortúzar and Willumsen 2011; Wilson 1967). It is worth noting, however,
that although Wilson's approach is built on a different conceptual
foundation than the old reference to Newtonian gravity, the work
succeeded at identifying the steps from proportionality to equality to
yield variations of proportionality constants, including the one that
eluded Stewart (1948) and that has been ignored in almost all subsequent
accessibility research. Why was this key element of spatial interaction
models potentially ignored in accessibility research? In the next
section we aim to address this question.

\section{Accessibility and spatial interaction modelling: two divergent
research
streams}\label{accessibility-and-spatial-interaction-modelling-two-divergent-research-streams}

The work of Hansen (1959) and Wilson (1971) responded to important
developments, in particular a need ``to meet the dictates and needs of
public policy for strategic land use and transportation planning''
(Michael Batty 1994). These dictates and needs were far from trivial. In
the United States alone, the Federal-Aid Highway Act of 1956 authorized
the creation of the U.S. Interstate Highway System, with a budget that
ultimately exceeded one hundred billion dollars (Weiner 2016; MDOT
2007). Spatial interaction modelling was incorporated into institutional
modelling practices meant to ``predict and provide'', i.e., predict
travel demand and supply transportation infrastructure (Kovatch, Zames,
et al. 1971; Weiner 2016). Accessibility, at the time, did not quite
have that power, as it did not quantify trips, but rather something
somewhat more elusive: it predicts the less tangible ``potential'' for
spatial interaction with opportunities. In this way, where spatial
interaction modelling became a key element of transportation planning
practice, accessibility remained a somewhat more academic pursuit, and
the two streams of literature only rarely connected.

\begin{figure}[H]

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{manuscript_files/figure-pdf/fig-docs-per-year-1.pdf}

}

\caption{\label{fig-docs-per-year}Historical pattern of publication:
documents per year.}

\end{figure}%

To illustrate this point, we conducted a bibliographic analysis of the
literature that cites Hansen (1959), Wilson (1971), or both. We
retrieved all relevant documents using the Web of Science ``Cited
References'' functionality, and the digital object identifiers of Hansen
(1959) and Wilson (1971). As a result of this search, we identified
1,788 documents that cite Hansen (1959), only 258 documents that cite
Wilson (1971), and 76 that cite both. The earliest document in this
corpus dates to 1976 and the most recent is from 2025. The number of
documents per year appears in Figure~\ref{fig-docs-per-year}, where we
see the frequency of documents over a span of almost fifty years. In
particular, we notice the remarkable growth in the number of papers that
cite Hansen (1959) compared to those that cite Wilson (1971).

As noted, literature that cites both Wilson (1971) and Hansen (1959) are
sparse (only 3.6\% of the corpus) but highly coupled. However, we can
discern from a reading of these works that they too are divergent, with
one stream focused on developing accessibility (i.e., \emph{potential
for} spatial interaction) and another on spatial interaction. The
focuses of these divergent streams contribute this paper's broader
hypothesis that the concepts of accessibility and spatial interaction
have remained largely disconnected.

On one hand, the stream of literature focused on spatial interaction
models inspired by Wilson (1971) and which cites both Hansen (1959) and
Wilson (1971), tends to contribute to understanding how accessibility is
interpreted and incorporated in spatial interaction models, treating it
as a separate (but related) phenomenon or variable. Specifically, some
early works interpret the spatial interaction model's balancing factors
(Equation~\ref{eq-production-constrained-balancing-factor} or
Equation~\ref{eq-doubly-constrained-balancing-factors}) as the inverse
of Hansen (1959)`s model (B. Harris and Wilson 1978; G. Leonardi 1978;
A. Stewart Fotheringham 1981; A. S. Fotheringham 1985), recognizing it
as a ``common sense'' approach (Morris, Dumble, and Wigan 1979, 99) to
including accessibility in the spatial interaction model, though further
exploration of its relationship is warranted (M. Batty and March 1976).
Some authors have explored this relationship, for instance as in A. S.
Fotheringham (1985) who demonstrates how the spatial interaction model
may insufficiently explain spatial patterns and suggest that explicitly
defining destinations' accessibility as a variable within the model that
may remedy the issue (e.g., the \emph{competition destination} model).
Other works used both Hansen (1959) and Wilson (1971)'s framework in
conjunction, such as in defining location-allocation problems in
operations research (G. Leonardi 1978; Beaumont 1981), estimating trip
flows (or some other spatial interaction flows) alongside accessibility
(e.g., Clarke, Eyre, and Guy 2002; Grengs 2004; Türk 2019), or
considering accessibility within spatial interaction models, in line
with A. S. Fotheringham (1985)'s demonstration (e.g., Beckers et al.
2022). Other works departed from Hansen (1959)'s definition and aligned
with spatial interaction in different ways, such as using micro-economic
consumer behaviour concepts to express potential for spatial interaction
(Morris, Dumble, and Wigan 1979; Giorgio Leonardi and Tadei 1984).

On the other hand, we discern that the stream of literature focused on
accessibility inspired by Hansen (1959) cite Wilson (1971) for three
general reasons. Firstly, as attribution for using using different
context-dependent travel cost functions (Weibull 1980; S. L. Handy and
Niemeier 1997; Kwan 1998; Q. Shen 1998a; Ashiru, Polak, and Noland 2003;
Rau and Vega 2012; Pan 2013; Margarida Condeço Melhorado et al. 2016;
Caschili, De Montis, and Trogu 2015; Grengs 2015; Pan, Jin, and Liu
2020; Chia and Lee 2020; Roblot et al. 2021; Sharifiasl, Kharel, and Pan
2023; Kharel, Sharifiasl, and Pan 2024). Secondly, associating spatial
interaction with accessibility's potential for spatial interaction
(e.g., Giuliano et al. 2010; Grengs et al. 2010; Grengs 2010, 2012;
Levine et al. 2012; Levinson and Huang 2012; Tong, Zhou, and Miller
2015; X. Liu and Zhou 2015; He et al. 2017; Wu and Levinson 2020; Ng et
al. 2022; Naqavi et al. 2023; Suel et al. 2024). While accessibility is
an expression of \emph{potential for} spatial interaction and Wilson
(1971) touches on the concept H. J. Miller (1999), it is distinct
concept from spatial interaction. This distinction, however, is often
unclear, as Hansen (1959) and Wilson (1971) are frequently co-cited as
both being `gravity models' (e.g., S. Liu and Zhu 2004; Dai, Wan, and
Gai 2017; Y. Shen 2019; Chia and Lee 2020). Thirdly, interpretation of
Hansen (1959)'s model as an inverse balancing factor (R. W. Vickerman
1974), especially in competitive opportunity or population contexts
(Karst and Van Eck 2003; Karst T. Geurs, van Wee, and Rietveld 2006;
Willigers, Floor, and van Wee 2007; El-Geneidy and Levinson 2011; Curtis
and Scheurer 2010; Manaugh and El-Geneidy 2012; Chen and Silva 2013;
Alonso et al. 2014; Albacete et al. 2017; Sahebgharani, Mohammadi, and
Haghshenas 2019; Mayaud et al. 2019; Allen and Farber 2020; Levinson and
Wu 2020; Marwal and Silva 2022; Su and Goulias 2023). However, as
outlined in preceding sections, we argue the resulting accessibility
values are plagued by similar interpretability issues. Of the literature
in this stream (i.e., accessibility works inspired by Hansen (1959) and
who cite Wilson (1971)), only the works of Soukhov et al. (2023) and
Soukhov et al. (2024) cite Wilson (1971)'s use of balancing factors as
method for maintaining constraints on opportunities within the context
of competitive accessibility. These works introduce the balancing
factors as a mechanisms to ensure that opportunities at each destination
are proportionally allocated to each zone (based on the proportion of
population seeking opportunities and the relative travel impedance).
This is to ensure that each zonal accessibility value is the sum of this
proportional allocation from each destination, and that all zonal values
ultimately sum to the total number of opportunities in the region.
However, these balancing factors were deduced intuitively. These works
do not explicitly state that the mathematical formulation of the
equations are effectively equivalent to Wilson's singly constrained
model (derived from entropy maximization), as will be demonstrated in
this study. They also do not discuss other constrained cases.

\section{A family of accessibility
measures}\label{a-family-of-accessibility-measures}

In this section, we introduce a family of accessibility measures, using
Wilson's spatial interaction modelling framework.

As argued in the preceding section, the streams of research on
accessibility and spatial interaction modelling have evolved as largely
separate streams with little contact since Hansen (1959) and Wilson
(1971). This may explain why the constraints and associated balancing
factors of spatial interaction models did not cross over to
accessibility analysis. This is intriguing since Wilson made an effort
to connect developments in spatial interaction modelling to
accessibility, noting for instance, the denominator of the
proportionality constants specific to origins
(Equation~\ref{eq-production-constrained-balancing-factor}) is the
inverse of balancing factor \(A_i\) (Wilson 1971, 10):

\begin{equation}\phantomsection\label{eq-Ai-as-accessibility}{
S_i = \frac{1}{A_i} = \sum_j W_j^{(2)} f(c_{ij})
}\end{equation}

Understanding \(S_i\) as the inverse of \(A_i\) offers a potential
interpretation for the balancing factor, however, \(S_i\) itself is not
unclear. Also, the relationship in Equation~\ref{eq-Ai-as-accessibility}
appears to be misaligned. In the preceding Section @ref(grav-to-today),
we noted that Hansen, and Stewart before him, defined accessibility as a
partial sum of the demographic force \(F\), essentially going from the
demographic force (\(F = G\frac{M_1\cdot M_2}{d_{12}^2}\) to the
system-wide population potential (after losing \(G\))
\(V_i = \sum_j \frac{M_j}{d_{ij}}\). While the inverse of Wilson's
\(A_i\) is pretty much identical to Hansen's accessibility, its
functional role in Wilson's general model
\(T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})\) is that of a balancing
factor \(k\) (maintaining proportionality).

Hence, we propose a more general definition of accessibility: the
partial sum of the potential for spatial interaction. Once we bring back
Wilson's proportionality constant into the picture, we can define the
\emph{potential for spatial interaction} between two locations \(i\) and
\(j\) is as follows:

\begin{equation}\phantomsection\label{eq-access-01}{
V_{ij} = k W_j^{(2)} f(c_{ij})
}\end{equation}

\noindent where \(V_{ij}\) is the potential for interaction from \(i\)
to \(j\). \(X\) denotes a sub-indices for the member of the constrained
accessibility value, either unconstrained \(0\), total constrained
\(T\), singly constrained \(S\) or doubly constrained \(D\). The
accessibility from origin \(i\) is defined as:

\begin{equation}\phantomsection\label{eq-accesssibility-01}{
V_{i} = k \sum_j W_j^{(2)} f(c_{ij})
}\end{equation}

Similar to Equation~\ref{eq-phys-gravity-model}, \(W_j^{(2)}\) above is
the mass at the destination and the sub-indices are for
\(i = 1,\cdots, n\) origins and \(j = 1,\cdots, m\) destinations.

The market potential variant can also be generally defined, which is
effectively the transpose of \(i\) and \(j\) in
Equation~\ref{eq-access-01} and Equation~\ref{eq-accesssibility-01} as
follows: \begin{equation}\phantomsection\label{eq-market-01}{
M_{ji} = k W_i^{(1)} f(c_{ij})
}\end{equation}

\begin{equation}\phantomsection\label{eq-market-potential-01}{
M_{j} = k \sum_j W_i^{(1)} f(c_{ji})
}\end{equation}

Similar to Equation~\ref{eq-phys-gravity-model}, \(W_j^{(2)}\) above is
the mass at the destination and the sub-indices are for
\(i = 1,\cdots, n\) origins and \(j = 1,\cdots, m\) destinations.

To demonstrate the anatomy of \(V_{ij}\) and \(M_{ij}\) along with the
partial sums of \(V_i\) and \(M_j\), Figure
Figure~\ref{fig-analytical-device-conc-accessibility} demonstrates this
work's accessibility analytical device for a simple 3 zone system. Each
measure's most disaggregate value is \(X_{ij}\), potential for spatial
interaction from \(i\) to \(j\). The \(X\) is a stand in for the \(ij\)
values of all the cases and their variants that will be described (i.e.,
\(V_{ij}^0, M_{ij}^0, V_{ij}^T, M_{ij}^T, V_{ij}^S, M_{ij}^S\) and
\(V_{ij}^D, M_{ij}^D\)). The single marginals represent the origin-side
and destination-side weights of the zones. The total marginal represent
the sum of a single marginal.

\begin{figure}[H]

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{../figures/access-analytical-device.png}

}

\caption{\label{fig-analytical-device-conc-accessibility}The family of
accessibility measures analytical device: labelling and associating ij
flows, zonal weights, the single marginals, and the total marginal.}

\end{figure}%

As an additional overview, we define four distinct members of the
constrained accessibility measure family, all delineated based on their
constant \(k\); either balancing factors \(K^T\), \(B_j\), \(A_i\)
depending on their case and variant:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{The unconstrained case} with variants \(V_i^0\) and \(M_i^0\).
  This case is equivalent to Hansen (1959)`s formulation (first variant)
  and Reilly (1929)'s market potential formulation (second variant),
  whereby a balancing factor is neglected, so units of zonal values are
  in units of 'opportunities-by-travel-impedance-value' and
  `population-by-travel-impedance-value'. In this case, no constraints
  are introduced to ensure that values of marginals are preserved.
\item
  \textbf{The total constrained case} with variant \(V_i^T\) and variant
  \(M_j^T\). The first variant \(V_i^T\)--the \emph{total constrained
  accessible opportunities measure}--resembles Hansen (1959)`s
  formulation but with an additional regional balancing factor \(K^T\)
  term, defined as the ratio of the total number of opportunities in the
  region to the total sum of unconstrained accessibility values in the
  region. In this way, \(K^T\) ensures that each zonal accessibility
  value is a proportion of the total opportunities in the region (i.e.,
  the total marginal in
  Figure~\ref{fig-analytical-device-conc-accessibility}), requires no
  information about the population seeking opportunities, and is in
  units of 'opportunities' accessible. The second variant \(M_j^T\) is
  the transpose of \(i\) to \(j\) of the first variant, effectively a
  constrained version of market potential and is referred to as
  \emph{total constrained accessible population measure}. In this
  variant, each zonal accessibility value is a proportion of the total
  population in the region (maintained by \(\hat K^T\)), requires no
  information about the opportunities that are sought, and each zonal
  value is in units of `population' accessible.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  When conceptualising \emph{potential} for spatial interaction--whether
  with opportunities or population--the total constrained case reflects
  the lowest level of restriction and hence maximum potential. The
  balancing factors \(K^T\) and \(\hat K^T\) only ensure that
  \(V_{ij}^T\) and \(M_{ji}^T\) values end up matching the \emph{total
  sum} of one of the single marginals, not the individual marginals
  themselves. For example, \(K^T\) does not guarantee that all the
  opportunities accessibility at \(V_{i=1}^T\) reflect a proportional
  allocation of the destination mass (i.e., \emph{number} of
  opportunities) at \(j=1, 2, 3\). In some cases, an allocation of
  opportunities from a destination \(j\) to \(i=1\) could \emph{exceed}
  the number of opportunities at that \(j\) (i.e., meaning that
  destination \(j\) is very attractive and reachable for an \(i\),
  relative other flows in the region). However, what \(K^T\) \emph{does}
  ensure is that each \(V_{ij}^T\) does not exceed the overall number of
  opportunities in the region--the total marginal. In other words,
  \(V_{i=1}^T\) expresses the number of opportunities that origin
  \(i=1\) could potentially interact with, as drawn from the entire
  regional opportunity mass, rather than constrained by individual
  destination totals. In this way, the \emph{total constrained} case
  reflects the maximum amount of \emph{potential} while still
  maintaining interpretable units.
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  \textbf{The singly constrained case}, with two variants \(V_i^S\)
  (opportunity-constrained) and \(M_j^S\) (population-constrained). The
  first variant is mathematically equivalent to the spatial availability
  measure (Soukhov et al. 2023), and this variant's per capita form is
  equivalent to the popular 2SFCA measure (Luo and Wang 2003; Q. Shen
  1998b). Both variants can also be defined using either Hansen (1959)'s
  or the market potential formulation but with balancing factor \(B_j\)
  for \(V_i^S\) or \(A_i\) for \(M_j^S\). These balancing factors ensure
  that the total marginal (green box in
  Figure~\ref{fig-analytical-device-conc-accessibility}) is maintained
  \emph{as well as} the values at one of the single marginals
  (destination-mass marginal for opportunity-constrained and origin-mass
  marginal for population constrained in
  Figure~\ref{fig-analytical-device-conc-accessibility}). In this way,
  the singly constrained case reflects a medium amount of potential,
  restricted by either values fitting the single destination-mass
  marginal or the single origin-mass marginal.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  The first variant \(V_i^S\) includes a set of destination-side
  balancing factors \(B_j\) which ensure opportunities (or destination
  masses) are allocated to each origin \(i\) based on the population
  (i.e, origin mass) of the origin \(i\) and travel impedance from that
  \(j\) to all \(i\)s. Hence, \(B_j\) ensures that each \(V_i^S\) is the
  product sum of a \emph{proportional share} of opportunities allocated
  from each destination \(j\) implicitly also being a share of the total
  opportunities in the region. Similar to the total constrained case,
  each \(V_i^S\) and \(V_ij^S\) accessibility value is expressed in
  units of `opportunities' accessible. But unlike the total constrained
  case, the singly constrained case explicitly incorporates population
  by allocating a balanced share of a destination's total opportunities
  to populations at reachable origins.
\item
  The second variant \(M_j^S\) includes a set of opportunity-side
  balancing factors \(A_i\) which ensure population (or origin masses)
  are allocated to each destination \(j\) based on the destination mass
  of the destination \(j\) and all possible travel impedance from that
  \(i\) to all \(j\)s. \(A_i\) ensures similar, but transposed, so
  constraints are satisfied. Specifically, each \(M_j^S\) represents
  both a share of the total regional population and the sum of balanced
  proportions of population from all origins, allocated to each
  destination based on the opportunities sought and travel impedance.
  Each zonal value is expressed in units of `population' accessible.
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  \textbf{The doubly constrained case}. It is constrained simultaneously
  by population (origin masses) and opportunities (destination masses),
  so each \(V_{ij}^D\) (equivalent to \(M_{ij}^D\)) is expressed in
  units of `population-opportunity capacity' that is accessible between
  \(i\) to \(j\). This case requires the number of opportunities and
  population to match, e.g., the analyst must know the matching spatial
  interaction capacity of the population (demand) and opportunities
  (supply). By using this one-to-one matching data and the double
  constraints (i.e., \(A_i\) and \(B_j\) at once, ensuring both the
  double constraint maintaints both marginals in
  Figure~\ref{fig-analytical-device-conc-accessibility}), this case
  restricts the \emph{potential} to spatially interact completely. In
  other words, the doubly constrained accessibility values reflect the
  number of predicted interactions between the opportunities and
  population, effectively, this case is equivalent formulaically as
  Wilson's doubly constrained spatial interaction model (i.e.,
  \emph{attraction-production constrained}). It can be understood as
  predicting a value of `access', and not accessibility.
\end{enumerate}

In the next sections, we will assign values to the simple 3x3 numeric
example shown in the analytical device
(Figure~\ref{fig-analytical-device-conc-accessibility}).

\subsection{Setup of the simple numeric
example}\label{setup-of-the-simple-numeric-example}

Consider a simple region as shown in
Figure~\ref{fig-analytical-device-conc-accessibility}, with zone IDs 1,
2 and 3. Each zone is both an origin \(i\) and a destination \(j\). The
following three pieces of information are defined: zonal population and
opportunities, zonal cost matrix, and travel impedance functions for
three types of travel behaviour. In this example, the population are
people, the opportunities are physicians, and the region can be
described by three possible travel behaviours.

Firstly, Table~\ref{tbl-small-system-land-use} summarises the population
(in units of 10,000s of people) and the opportunities (the number of
physicians) per zone. {[}\^{}1{]}. Considering
Table~\ref{tbl-small-system-land-use}'s values, the
Provider-to-Population-Ratio (PPR) in this system is 24.5. For
reference, the number of physicians per 10,000 in Canada in 2022 was
24.97 (WHO 2025). Secondly, to pair the zonal population and opportunity
information, the assumed cost of movement (in minutes of travel time)
between origins and destinations is as shown in
Table~\ref{tbl-small-system-cost}.

\begin{table}

\caption{\label{tbl-small-system-land-use}Simple system with three zones
(ID 1, 2 and 3). Population is in 10,000 persons and opportunities in
number of physicians.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}rcc}
\toprule
ID (i or j) & Population\textsuperscript{\textit{1}} & Opportunities\textsuperscript{\textit{2}} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 160 \\ 
2 & 10 & 150 \\ 
3 & 6 & 180 \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Population is \emph{Wi\textsuperscript{(1)}} when used as a proxy for the mass at the origin, and \emph{Oi} when used as a constraint.\\
\textsuperscript{\textit{2}}Opportunities is \emph{Wj\textsuperscript{(2)}} when used as a proxy for the mass at the destination, and \emph{Dj} when used as a constraint.\\
\end{minipage}

}

\end{table}%

\begin{table}

\caption{\label{tbl-small-system-cost}Cost matrix for system with three
zones (travel time in minutes).}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}cccc}
\toprule
 & \multicolumn{3}{c}{Destination ID} \\ 
\cmidrule(lr){2-4}
Origin ID & 1 & 2 & 3 \\ 
\midrule\addlinespace[2.5pt]
1 & 10 & 30 & 15 \\ 
2 & 30 & 10 & 25 \\ 
3 & 15 & 25 & 10 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

From both Table~\ref{tbl-small-system-land-use} and
Table~\ref{tbl-small-system-cost}, Zone 3 and 1 can be interpreted as of
an urban core with major healthcare institutions and a healthcare
cluster on the edge of the city respectively, each with a moderate
residing population (with zone 3 having both a higher population and
physician count). Zone 2, can be interpreted as a more distant bedroom
community, with a relatively high population and fewer physicians. In
sum, Zones 1 and 3 are more proximate to each other than to Zone 2, and
together match Zone 2's population while offering more than twice the
physician availability.

And lastly, in Equation~\ref{eq-travel-behaviour-scenarios} we
distinguish accessibility measure values for the following three
impedance functions that represent the potential for spatial interaction
travel behaviour of the population to opportunities. Accessibility will
be calculated three times for each case, one assuming the most decay
(\(f_1(c_{ij})\)), another assuming medium decay (\(f_2(c_{ij})\)), and
a third assuming the least decaying travel behaviour (\(f_3(c_{ij})\))
for the entire region. A helpful analogy may be tying travel behaviour
to the used mode's mobility potential, i.e., the most decaying travel
behaviour (\(f_1(c_{ij})\)) would assume all travel in the region being
done by foot, while calculating accessibility assuming the least decay
(\(f_3(c_{ij})\)) would assume unfettered automobility. Or
alternatively, these functions could represent travel behaviour on
snowstorm-affected day (\(f_1(c_{ij})\)) for the entire region versus a
clear, ideal travel day (\(f_3(c_{ij})\)). As an example of a discussion
on how travel behaviour has been considered in accessibility measures
cost of travel see Paez, Scott, and Morency (2012).

\begin{equation}\phantomsection\label{eq-travel-behaviour-scenarios}{
\begin{array}{l}
f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
\end{array}
}\end{equation}

Any set of concepts representing population, opportunities, and their
associated travel behaviour, whether representing the entire region
uniformly (as we will demonstrate) or representing specific subgroups,
can be substituted into our simple example, depending on the research
question. The purpose of the following simple example is to demonstrate
the calculation of each member of the accessibility measure family,
interpret the values, and compare them both within and across travel
behaviour groups and members of the family of accessibility measures.

\subsection{Unconstrained
accessibility}\label{unconstrained-accessibility}

Setting the balancing factor \(k\) to 1 or omitting it completely in
Equation~\ref{eq-access-01} results in the unconstrained accessibility
case:

\begin{equation}\phantomsection\label{eq-unconstrained-access}{
V^0_{ij} = 1 *W_j^{(2)} f(c_{ij})
}\end{equation}

In this case, the partial sum of spatial interaction is simply identical
to Hansen's accessibility \(S_i\) (Hansen 1959), the current standard
practice in accessibility measurement:

\begin{equation}\phantomsection\label{eq-unconstrained-accessibility}{
V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
}\end{equation}

The sum of the unconstrained accessibility values for each origin
\(V^0_{i}\) generally does not equal the total number of opportunities
\(O\) (e.g., \(\sum_i V^0_{i} \not= O\)), since arbitrarily setting
\(k\) to 1 (or neglecting \(k\) all together) strips the values of any
meaningful unit-based interpretation, the units are in `summed
opportunities by some travel impedance value'. Moreover, comparisons of
these \(V^0_{i}\) values across different contexts such as different
impedance functions \(f(c_{ij})\) or varying number of zones exacerbates
this issue as the units between \(V^0_{i}\) values change i.e.,
comparisons between a value of `summed opportunities by a travel
impedance value' to a value of `summed opportunities by another travel
impedance value' are not directly interpretable. From this perspective,
the raw unconstrained accessibility scores are not intuitively
comparable across different contexts and decay functions. They should
more appropriately be used as an ordinal variable to make comparisons of
size (i.e., greater than, less than, equal to), not to calculate ratios
or intervals (i.e., the magnitude of differences).

Returning to our numeric example, the calculated unconstrained
accessibility \(V^0_{i}\) for each origin, a sum of all the travel
impedance weighted opportunities at each destination
(\(\sum_i V^0_{i}\)), in displayed in
Table~\ref{tbl-simple-example-unconstrained-accessibility}.

\begin{table}

\caption{\label{tbl-simple-example-unconstrained-accessibility}Simple
system: unconstrained accessibility.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 45.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 405.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{0}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Origin & units: \emph{physicians-minute\^{}-3} & units: \emph{physicians-minute\^{}-2} & units: \emph{physicians-minute\^{}-0.1} \\ 
\midrule\addlinespace[2.5pt]
1 & 0.219 & 2.567 & 371.143 \\ 
2 & 0.167 & 1.966 & 363.479 \\ 
3 & 0.237 & 2.751 & 373.738 \\ 
\midrule 
\midrule 
Sum & 0.6233422 & 7.283556 & 1108.361 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

As the different impedance functions represent different travel
behaviours, comparing the raw unconstrained accessibility values across
groups is meaningless beyond notions of higher or lower. For instance,
at zone 1 the difference between the least decay (\(f_3(c_{ij})\)) and
most decay (\(f_1(c_{ij})\)) groups is 370.92, but in what units? These
two values are a product of different impedance functions, making the
comparison uninterpretable in absolute terms. Likewise, we could compare
values within the same travel behaviour scenario across different zones,
as they are in the same units, however the issue of unit
interpretability will also be apparent. Considering the most decaying
scenario \(f_1(c_{ij})\) and zone 1 (the zone with a healthcare cluster
at the edge of the city): zone 1 captures 0.0181185 fewer
physicians-minute\(^{-1}\) than zone 3 (urban core). Again, the
fundamental uninterpretability of what is a
\emph{physicians-minute\(^{-1}\)} or
\emph{opportunity-weighted-travel-impedance} unit remains.

While one could attempt to adjust the units post-calculation (e.g.,
scaling, population normalization) or select impedance functions to
facilitate comparison across scenarios (potentially at the expense of
accurately reflecting travel behavior), such adjustments may introduce
bias. Hence, the raw unconstrained accessibility values themselves are
challenging to compare due to their units. To enable more meaningful
comparison, the following sections will detail the introduction of
constraining constants to ensure consistent units across scenarios and
demonstrate results on the same numeric example.

\subsection{Total constrained
accessibility}\label{total-constrained-accessibility}

In the total constrained accessibility case, a total balancing factor
proportionally adjusts unconstrained zonal accessibility values
\(V^0_i\) based on the regional sum of \(V^0_i\) and the total
population or opportunities in the region. Alternatively, we reformulate
this case using a proportional allocation constant, which allocates
opportunities (or population) proportionally based on the the travel
impedance and the total population or opportunities in the region. In
both formulations, all zonal values become a proportion of a known
system total, be it the regional opportunities or regional population
depending on the variant.

We define two variants for this case: (a) \(V_i^T\) where accessibility
is constrained by the total number of opportunities (total constrained
accessible opportunity) and which is interpreted as Hansen's
accessibility with a constraining constant, and (b) \(M_j^T\), where
\(i\) and \(j\) of the first variant is transposed, yielding a measure
constrained by the total number of population and to be interpreted as
constrained `market potential'.

\subsubsection{Total constrained accessible opportunities: Hansen's
accessibility with a total
constraint}\label{total-constrained-accessible-opportunities-hansens-accessibility-with-a-total-constraint}

Unlike in Equation~\ref{eq-access-01}, the proportionality constant
\(k\) is retained. For the total constrained case, it is represented as
\(K^T\):

\begin{equation}\phantomsection\label{eq-total-constrained-access}{
V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
}\end{equation}

In this way, the total constrained accessibility measure now becomes
Hansen's accessibility with a balancing factor \(K^T\):

\begin{equation}\phantomsection\label{eq-total-constrained-accessibility}{
V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
}\end{equation}

Imagine that the only system known is the total number of opportunities
\(D\) in the region. Accordingly, the constant we impose in this case
ensures the regional sum of total constrained accessibility is equal to
the total number of opportunities as follows:
\begin{equation}\phantomsection\label{eq-total-constraint-accessibility}{
\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
}\end{equation}

This constraint is analogous to the total constraint of
Equation~\ref{eq-constraint0-gravitymodel}, congruent with Wilson's
framework. Given the total number of opportunities in the region, we can
then substitute Equation~\ref{eq-total-constrained-accessibility} in
Equation~\ref{eq-total-constraint-accessibility} to solve for \(K^T\):

\begin{equation}\phantomsection\label{eq-total-opportunity-balancing-factor}{
K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
}\end{equation}

Which is also congruent with Wilson's framework as it comparable to the
total flow spatial interaction model (e.g., Equation 2.11 in Cliff,
Martin, and Ord (1974)). Hence, rearranging the equation to have
opportunities and the proportional constant distinctly represented, our
total constrained accessibility model is:

\[
V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
\]

Further, we can see that, since \(D\) and \(W^{(2)}_j\) are both in
units of opportunities, the proportional allocation factor for the total
constrained opportunity case \(\kappa_i^T\) is dimensionless:

\[
\kappa_i^T = \frac{\sum_j D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
\]

\noindent and therefore \(V^T_i\) is now in the units of \(W^{(2)}_j\),
that is, the mass at the destination
(\(V^T_i = \kappa_i^T\sum_j W^{(2)}_j\)). The role of \(\kappa_i^T\) in
this reformulation of accessibility is to transform between units and to
adjust the number of opportunities accessible from \(i\) so they
represent a proportion of the total number of opportunities in the
region. \(\kappa_i^T\) then assigns opportunities in proportion to the
impedance between \(i\) and \(j\). This is why we refer to
\(\kappa_i^T\) as a proportional allocation factor. On the other hand,
the proportionality constant \(K^T\) balances the units of \(V^0_i\),
the Hansen-type accessibility values, and is an alternative expression
of the total constrained accessibility measure.

Referring back to our simple numeric example, balancing factor \(K^T\)
for the most decay travel behaviour scenario
\(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:

\[
\begin{array}{l}
K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})}\\
K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + \cdots + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
}\\
K^T = \frac{490}{0.6233422} \\
K^T = 786.085
\end{array}
\]

Using the calculated proportionality constants for all zones and
multiplying them by the unconstrained accessibility value \(V^0_i\), the
total opportunity constrained accessibility values for all zones and
different travel behaviour scenarios is presented in
Table~\ref{tbl-simple-example-total-opportunity-accessibility}.
\(\kappa_i^T\) for each zone is not shown, but can be understood to be
the unitless proportion of opportunities (of the total opportunities)
allocated to each zone based on travel impedance.

\begin{table}

\caption{\label{tbl-simple-example-total-opportunity-accessibility}Simple
system: total opportunity constrained accessibility.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 45.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 405.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{T}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Origin & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
\midrule\addlinespace[2.5pt]
1 & 172.065 & 172.672 & 164.080 \\ 
2 & 131.627 & 132.247 & 160.692 \\ 
3 & 186.308 & 185.081 & 165.228 \\ 
\midrule 
\midrule 
Sum & 490 & 490 & 490 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

In contrast to unconstrained accessibility \(V^0_i\), imposing a
constraint -in this case a total opportunity constraint for this
variant- allows for the comparison of differences and ratios between
regions and across different travel behaviour scenarios as well. Each
value is effectively in units of physicians, with the impedance units
already accounted for by \(\kappa_i^T\).

Considering the highest decay scenario (\(f_1(c_{ij})\)), zone 1 (a
healthcare cluster at the edge of the city) captures an intermediate
amount of physicians (172.0652825) like in the unconstrained
accessibility case. However, unlike in the unconstrained case, we can
say that this value is out of the 490 physicians in the region, which
allows us also to deduce that zone 1 captures 1.3072213 and 0.9235529
times more than zone 2 and 3. Values for the lesser decay
(\(f_2(c_{ij})\)) and lowest decay (\(f_3(c_{ij})\)) scenarios are
calculated separately, with decay scenario values also summing to equal
490 physicians accessible in the region.

One can also directly compare values at a specific zone due to the
consistent units. For instance, zone 1 remains intermediate in capturing
accessible physicians relative to zones 2 and 3 across scenarios,
similar to the unconstrained case. However, the difference between
travel behaviour scenarios differ in direction. Specifically, Zone 1
captures 0.6067946 more and 7.9850478 fewer opportunities than the
lesser decay scenarios \(f_2(c_{ij})\) and \(f_3(c_{ij})\) respectively.
Why? \(\kappa_i^T\) ensures proportional allocation for each travel
behaviour scenario. Meaning, while the unconstrained accessibility
increases, \(\kappa_i^T\) adjusts the values to remain proportional to
the total number of opportunities (490 physicians accessible in the
region). As the decay behaviour decreases, more opportunities are
accessible for all zones. In the medium decay scenario \(f_2(c_{ij})\),
zone 1 sees a slight increase in values (relative to the highest decay
scenario) as the zone can accessible more opportunities relative to
increases seen in other zones. However, in the lowest decay scenario,
zone 1 sees a decrease, as it is outpaced by increases in other zones -
namely zone 2 (recall: zone 2 has the lowest number of opportunities,
hence the increases in opportunity gains is much higher in a low decay
scenario).

Using the total opportunity constrained formulation of accessibility
offers a solution to the unit interpretability issue of Hansen (1959)'s
accessibility measure. Intuitively, the use of the constraint
illustrates how the differences and ratios of values between zones and
decay groups can be compared. This is true for other constrained cases
of the family of accessibility measures.

\subsubsection{Total constrained accessible population: Reilly's
potential trade territories with a total
constraint}\label{total-constrained-accessible-population-reillys-potential-trade-territories-with-a-total-constraint}

Another variant of the total constrained accessibility measure is the
\emph{total constrained accessible population measure}, which represents
the transpose of \(i\) to \(j\) of the total constrained accessible
opportunities measure. This variant, expressed in
Equation~\ref{eq-total-constrained-market}, represents an expression of
the concept of market potential (i.e., potential users) as proposed in
C. D. Harris (1954) and R. W. Vickerman (1974). This unconstrained form
of market potential \(M_j^0\) (Equation~\ref{eq-unconstrained-market}),
effectively the \(i\) to \(j\) transpose of \(V^0_{ij}\), has been used
in recent research to express the potentially accessible population
(i.e., users) as a result of regional transportation infrastructure
investment projects (e.g., Gutiérrez 2001; Holl 2007; Condeço-Melhorado
and Christidis 2018). Put another way, market potential can also be
thought of as a form of \emph{passive accessibility}, indicating the
number of people that can reach each destination. However, like
\(V_{ij}^0\), issues of unit interpretability arise in \(M_j^0\)'s
unconstrained form. To address this, the constrained variant, the total
constrained accessible population measure \(M^T_{j}\), is introduced. To
formulate this variant, the total balancing factor \(K^T\) is applied to
the mass of the \emph{population} at \(i\) (\(W_i^{(1)}\)) instead of
the opportunities at \(j\).

\begin{equation}\phantomsection\label{eq-total-constrained-market}{
M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
}\end{equation}

\noindent with \(M_j^0\) being the \(i\) \(j\) transpose of
\(V^0_{ij}\):

\begin{equation}\phantomsection\label{eq-unconstrained-market}{
M_j^0 = W_i^{(1)} f(c_{ij})
}\end{equation}

The total constrained accessible population measure (market potential)
then becomes:

\[
M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
\]

Where we impose the total system known as a constraint, i.e., that the
total market potential equals the total population \(O\) in the region:

\begin{equation}\phantomsection\label{eq-total-constraint-market}{
\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
}\end{equation}

Substituting Equation~\ref{eq-total-constrained-market} in
Equation~\ref{eq-total-constraint-market}, and solving for \(\hat K^T\),
we obtain:

\begin{equation}\phantomsection\label{eq-total-population-balancing-factor}{
\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
}\end{equation}

The constrained market potential then takes the following form:

\[
M^T_j = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
\]

Where the following \(\hat \kappa_i^T\) proportional allocation factor
is dimensionless:

\[
\hat \kappa_i^T = \frac{\sum_i O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
\]

Returning back to the numerical example, the proportionality constant
\(\hat K^T\) is solved for each travel behaviour scenario, and the
market potential of each zone \(M^T_j\) is expressed as units of
population (e.g., the number of people accessible from each origin at
that destination) in
Table~\ref{tbl-simple-example-total-population-accessibility}.

\begin{table}

\caption{\label{tbl-simple-example-total-population-accessibility}Simple
system: total opportunity constrained accessibility.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 337.50pt -2\tabcolsep-1.5\arrayrulewidth}}{M\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Destination & units: \emph{population in 10,000s} & units: \emph{population in 10,000s} & units: \emph{population in 10,000s} \\ 
\midrule\addlinespace[2.5pt]
1 & 5.018 & 5.447 & 6.598 \\ 
2 & 8.596 & 7.986 & 6.717 \\ 
3 & 6.386 & 6.567 & 6.684 \\ 
\midrule 
\midrule 
Sum & 20 & 20 & 20 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Readers may note the difference in trends in accessible population
(Table~\ref{tbl-simple-example-total-population-accessibility}) and
accessible physicians (i.e., the preceding subsection,
Table~\ref{tbl-simple-example-total-opportunity-accessibility}). In
Table~\ref{tbl-simple-example-total-opportunity-accessibility}, zone 1,
2, 3 represent destinations and the accessibility values reflect the
number of accessible people from the vantage of physicians. Zone 1, in
its role as a destination, is no longer intermediately-ranked relative
to other zones; it now attracts the fewest number of people across all
three travel behaviour scenarios. However, similar to the total
constrained opportunity case, as travel decay reduces, the availability
of population begins to converge (though Zone 1 continues as the
lowest-ranked) for similar reasons. As decay reduces, the population's
travel impedance to all zones become more similar, making the relative
location of the zones less important and all people in the region more
equally accessible. Overall: like the total constrained accessible
opportunities case, this variant allows for the interpretation of both
ordinal and interval comparisons of the raw values themselves.

\subsection{Singly constrained
accessibility}\label{singly-constrained-accessibility}

Similar to the total constrained accessibility measure, the singly
constrained case includes a balancing factor that adjusts the
unconstrained zonal accessibility values \(V^0_i\) such that a
\emph{single} constraint is satisfied. Two variants are defined: (a) the
\emph{singly constrained accessible opportunities} case (an alternative
formula to spatial availability in Soukhov et al. (2023)), and (b) the
\emph{singly constrained accessible population} case, its transpose, or
singly constrained market potential.

Unlike the total constraint (i.e.,
Equation~\ref{eq-total-constraint-accessibility}), the single
constraint--as will be defined in
Equation~\ref{eq-opportunity-constraint} and
Equation~\ref{eq-population-constraint} for the first and second
variants-- incorporates additional information. In the
opportunities-accessible variant, the associated balancing factor
constrains the `potential' for spatial interaction by ensuring that only
a proportional amount of opportunities at each destination are allocated
to `demanding' origins. This allocation is informed by demand, or the
relative amount of population, and associated travel impedance
connecting the zones. In the population-accessible variant (i.e., the
\(i\) -\textgreater{} \(j\) transpose of the first variant), population
at each origin is proportionally allocated to destinations based on the
share of opportunities and travel impedance.

In both variants, the singly constrained accessibility measure
introduces population-based (or opportunity-based) competition at the
zonal level, unlike the total constraint, which more simply allocates a
fixed regional total of opportunities (or population depending on the
variant). In sum, all singly constrained zonal accessibility values are
both a proportion of the known regional opportunity (or population)
total \emph{and} a sum of a \emph{balanced} proportion of opportunities
allocated from each destinations (or population allocated from each
origin). Each zonal value remains in units of opportunities accessible
(or population accessible).

\subsubsection{Singly constrained accessible opportunities: spatial
availability}\label{singly-constrained-accessible-opportunities-spatial-availability}

To demonstrate this variant formulaically, we begin with the opportunity
constraint (Equation~\ref{eq-opportunity-constraint}) as our known piece
of information. Namely, the sum of accessible opportunities from a
destination should equal the number of opportunities \(D_j\) at that
destination. As the number of opportunities at each \(j\) are known, it
is represented as \(D_j\) instead of \(W_j^(2)\) as in the total
constrained case. This constraint should hold for all destinations in
the region. This is comparable to the single attraction-constraint
(Equation~\ref{eq-constraint2-gravitymodel}) from Wilson's framework.

\begin{equation}\phantomsection\label{eq-opportunity-constraint}{
\sum_i V^S_{ij} =  D_j
}\end{equation}

The underlying spatial interaction model is now the
attraction-constrained model in
Equation~\ref{eq-attraction-constrained-gravitymodel}, and our
accessibility measure becomes:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility}{
V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
}\end{equation}

\noindent where \(W_i^{(1)}\) is a measure of the mass at origin \(i\)
(i.e., the opportunity-seeking population). The corresponding balancing
factor, as per Wilson, is:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-proportionality-constants}{
B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

Introducing the balancing factor in
Equation~\ref{eq-opportunity-constrained-accessibility}, we obtain:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility-with-balancing-factor}{
V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

Further, we define the following proportional allocation factor:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-proportional-allocation-factor}{
\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

After this, it is possible to rewrite
Equation~\ref{eq-opportunity-constrained-accessibility-with-balancing-factor}
as an origin summary expression of proportionally allocated known
opportunities (i.e., \(D_j\)).

\begin{equation}\phantomsection\label{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}{
V^S_{i} = \kappa^S_{i} \sum_j   D_j
}\end{equation}

Soukhov et al. (2023) have shown that the role of \(\kappa^S_{i}\) is to
allocate opportunities \(D_j\) proportionally to the mass at each origin
\(i\) and the impedance between \(i\) and \(j\). As in the total
constrained opportunity case, \(\kappa^S_{i}\) is dimensionless and
\(V_i^S\) is in the units of opportunities \(D_j\). The singly
constrained accessibility measure in
Equation~\ref{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
is called spatial availability by Soukhov et al. (2023), because it
represents the number of opportunities that can be reached \emph{and}
are available, in the sense that accessible opportunities have been
proportionally allocated based on relative demand, travel impedance and
the regional total number of opportunities, i.e., spatial competition
for them has been considered. These authors also show that the following
expression (accessibility per capita) is a constrained version of the
popular two-stage floating catchment area measure of Q. Shen (1998b) and
Luo and Wang (2003):

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility-per-capita}{
v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
}\end{equation}

Returning to the simple numeric example, the opportunity-constrained
case would yield the following \(B_{j}\) for \(f_1(c_{ij})\):

\[
\begin{array}{l}
B_{j} = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
B_{2} =  \frac{1}{\frac{4}{30^3} + \frac{10}{10^3} + \frac{6}{25^3}} = 94.9474\\
B_{3} =  \frac{1}{\frac{4}{10^3} + \frac{10}{25^3} + \frac{6}{10^3}} = 93.9850
\end{array}
\]

The balancing factors \(B_j\) for the \(f_2(c_{ij})\) decay group for
zones 1, 2 and 3 is 12.8571429, 8.7685113 and 10.6635071, and for
\(f_3(c_{ij})\) decay group is 0.0672461, 0.0660559 and 0.0663798. Using
these these balancing constants, we can calculate the singly constrained
opportunity accessibility:

\begin{table}

\caption{\label{tbl-simple-example-attraction-constrained-accessibility}Simple
system: singly constrained opportunity accessibility.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 36.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 324.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){3-5}
 &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
Origin & Population (units: \emph{people in 10,000s}) & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 133.469 & 122.255 & 98.848 \\ 
2 & 10 & 166.781 & 185.096 & 241.877 \\ 
3 & 6 & 189.750 & 182.650 & 149.275 \\ 
\midrule 
\midrule 
Sum & — & 490 & 490 & 490 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Imposing the single proportional allocation factor \(\kappa^S_{i}\)
allows for the comparison of differences and ratios of the accessibility
values, like previously discussed in the total constrained accessible
opportunities case. The proportional allocation factor ensures that
resulting values are in units of \emph{physicians}, with the impedance
units already accounted for in the allocation process.

However, unlike the total constrained opportunity case, \(\kappa^S_{i}\)
reflects zonal competition based on the mass of the origin, i.e.,
population. Again, consider the highest decay scenario \(f_1(c_{ij})\).
Under this scenario, zone 1 no longer captures a medium amount of
physicians as in the total constrained opportunity case: it now captures
the fewest in the region i.e., 133.4687282 at zone 1, 166.7813387 at
zone 2, and 189.7499331 at zone 3. Why may zone 1 (healthcare cluster at
the edge of the city) capture 27\% of the physicians regionally while
this same zone captures 35\% in the total opportunity constrained case?

This difference is due to the single-opportunity factor
\(\kappa^S_{i}\)'s role in \(V_i^S\). The only inputs required in the
total-opportunity factor \(\kappa^T_i\) is the total number of
opportunities in the region \(D\) as well as the associated
opportunities at each \(j\) and travel impedance. Opportunities are
allocated to \(i\)s, regardless of the mass weights of origin. Whereas
the single opportunity constraint \(\kappa^S_{i}\) requires the
population at \(i\) as an input. In fact, \(\kappa^S_{i}\) is calculated
as the proportion of impedance-weighted population at an \(i\) to the
sum of impedance-weighted population for the entire region. Hence, since
zone 1 has the lowest population in the region, is in close proximity to
a more populated zone (zone 3, the `urban core'), and is not well
connected (in terms of travel impedance) to other zones with
opportunities, \(V_{1}^S\) values, or the number of physicians
accessible, is lowest.

Readers may also notice the change in the proportion of opportunities
drawn from different zones depending on the travel behaviour scenarios
considered. For instance, consider Zone 2 which has the highest
population. It is more evident in the \(f_3(c_{ij})\) scenario than in
higher decay scenarios that this zone does not have an exceptionally
large population for the region - Zone 2 only represents 50\% of the
population in the three zone region. In this sense, other zones are not
\emph{that} disadvantaged, and in this scenario with unfettered travel
cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1
and Zone 3 takes 6\% and 8\% more from Zone 2 between \(f_3(c_{ij})\)
and \(f_1(c_{ij})\) scenarios hence \(\kappa_{2,2}^S\) decreases by
14\%). Zones 1 and 3 are allocated opportunities at relates similar to
their relative population size.

Readers may also notice the change in the proportion of opportunities
drawn from different zones depending on the travel behaviour scenarios
considered. For instance, Zone 2 has the highest zonal population,
representing 50\% of the 200,000 regional population. However, due to
its high relative travel distance from the other zones, its population
is less competitive in capturing opportunities in the high-decay travel
scenario (\(f_1(c_{ij})\)): Under this scenario, zone 2 captures almost
exclusively opportunities from its own zone. However, in
\(f_3(c_{ij})\), the scenario with unfettered travel cost, Zone 2
captures by far the most number of physicians. But in this scenario,
Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone
3 takes 6\% and 8\% more from Zone 2 between \(f_3(c_{ij})\) and
\(f_1(c_{ij})\) scenarios hence \(\kappa_{2,2}^S\) decreases by 14\%).
Zones 1 and 3 are allocated opportunities at rates similar to their
relative population size.

In this way, the consideration of constrained accessibility \emph{per
capita} may be clarifying. Often, accessibility values are reported as
raw scores without the consideration for population. But, as we
introduced constraints, these constrained accessibility values can be
normalized using anything that is relevant to the zone. In
Table~\ref{tbl-simple-example-attraction-constrained-accessibility-per-capita},
we present per capita accessibility for the numeric example, simply in
units of number of physicians accessible per population at each zone.
Notably, these per capita rates are equivalent to the 2SFCA values.

\begin{table}

\caption{\label{tbl-simple-example-attraction-constrained-accessibility-per-capita}Simple
system: singly constrained opportunity accessibility per capita.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 36.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 324.00pt -2\tabcolsep-1.5\arrayrulewidth}}{v\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){3-5}
 &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
Origin & Population (units: \emph{people in 10,000s}) & units: \emph{physicians per capita} & units: \emph{physicians per capita} & units: \emph{physicians per capita} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 33.367 & 30.564 & 24.712 \\ 
2 & 10 & 16.678 & 18.510 & 24.188 \\ 
3 & 6 & 31.625 & 30.442 & 24.879 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

This simple example was constructed so that the regional average equals
24.5 physicians per 10,000 people. As distance decay decreases and
becomes \emph{relatively} uniform (all zones can reach all zones), the
effect of population drives the proportional allocation of
opportunities. Consequently, per capita accessibility values begin to
stabilise to the regional per capita average (e.g., in the lowest
distance decay \(f_3(c_{ij})\), per capita values are all around 24
physicians accessible per capita).

This trend mirrors how the accessibility values in the total constrained
opportunity case stabilises to the \(V_i^T\) regional average (e.g., the
accessible opportunities allocated to each of the three zones approaches
a third of 490 physicians, or 163.33, under the unfettered mobility
scenario \(f_3(c_{ij})\)). These patterns make intuitive sense: the
balancing factors act as regional and/or zonal averaging mechanisms. In
distance decay travel behaviour scenarios that are more relatively
uniform (i.e., low for all zones like in \(f_3(c_{ij})\)), what remains
is the relative effect of the other variables in the balancing factor.
In the total constrained case, this is the proportion of opportunities
relative to the regional opportunities, and in the case of the single
opportunity constrained case, this is the population at a zone relative
to the regional population.

\subsubsection{Singly constrained accessible population: market
availability}\label{singly-constrained-accessible-population-market-availability}

Similar to Equation~\ref{eq-total-population-balancing-factor} in
transposing the origins and destinations, we can define a \emph{singly
constrained} measure of market potential that preserves the known
population (i.e., the mass weight at the origin \(W_i^{(1)}\) is now
represented by \(O_i\)). In it's per-capita expression, i.e., equivalent
to 2SFCA, this constrained concept of market potential been used to
express ``facility crowdedness'' as in F. H. Wang (2018).

The underlying spatial interaction model is now the
production-constrained model in
Equation~\ref{eq-production-constrained-gravitymodel}, and our market
potential measure \(M^S_j\) becomes:

\begin{equation}\phantomsection\label{eq-population-constrained-accessibility}{
M^S_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
}\end{equation}

In this case, the measure is singly constrained by the population
\emph{by origin} (i.e., \(O_i\)), like
Equation~\ref{eq-constraint2-gravitymodel} from Wilson's framework:

\begin{equation}\phantomsection\label{eq-population-constraint}{
\sum_j M^S_{ij} =  O_i \\
}\end{equation}

And the corresponding balancing factor, as per Wilson, is:

\begin{equation}\phantomsection\label{eq-population-constrained-proportionality-constants}{
A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
}\end{equation}

Following the same logic as in the preceding section on total
constrained market potential, one arrives at the following expression:

\begin{equation}\phantomsection\label{eq-production-constrained-accessibility-with-proportional-allocation-factor}{
M^S_{j} = \hat \kappa^S_{j} \sum_i  O_i
}\end{equation}

\noindent with:

\begin{equation}\phantomsection\label{eq-attraction-constrained-proportional-allocation-factor}{
\kappa^S _{j} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
}\end{equation}

As well, the single (population) constraint in
Equation~\ref{eq-population-constraint} ensures that the the total
constraint (e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O\)) is
maintained.

With these constraints, \(\frac{M_j^S}{O}\) can be interpreted as the
proportion of the total population serviced by location \(j\).

For the sake of brevity, we'll move forward onto the doubly constrained
case.

\subsection{Doubly constrained
accessibility}\label{doubly-constrained-accessibility}

This accessibility case requires zonal populations and opportunities to
match one-to-one, like the doubly constrained spatial interaction model.
In this way, doubly constrained accessibility can be thought as
``access'' or simply spatial interaction (no potential).

To contextualize this point, the total and singly constrained
accessibility measures discussed thus far have used either \(O_i\),
\(D_j\), or the regional sums of either, but never both simultaneously.
For example, when opportunities \(D_j\) are used to constrain
Equation~\ref{eq-opportunity-constrained-accessibility} in the
\emph{singly constrained accessible opportunities measure}, the specific
mass of the population at origin \(i\) demanding only those
opportunities at that \(j\) is unknown. Instead, only the population at
\(i\) demanding opportunities in the region is known, and this is more
generally represented as \(W_i^{(1)}\) (as in
Equation~\ref{eq-opportunity-constrained-proportionality-constants}).
Similarly, when the population \(O_i\) is used as a constraint in
Equation~\ref{eq-population-constrained-accessibility} in the
\emph{singly constrained accessible population measure}, the mass at the
destination is given by \(W_j^{(2)}\) (also in
Equation~\ref{eq-population-constrained-proportionality-constants})
since only the mass of opportunities at each \(j\) is known and
information on what opportunities are allocated to what zone is unknown.

By contrast, the double-constrained accessibility case requires that
both populations and opportunities match. Meaning, opportunities at each
destination can be accessed by all populations, while at the same time,
the population at each origin can be accessed by all opportunities.
However, this requirement is often unintuitive in traditional
accessibility analysis. Namely, the distinction between population
(origin masses) and opportunities (destination masses) typically
represent different entities without a shared unit of measurement. On
the population side, we usually count people; on the opportunity side,
we may be referring to physicians, clinics, grocery stores, schools,
parks, or libraries. In a few cases, a one-to-one correspondence or
their own capacities at which they interact and are interacted with may
exist e.g., one person with one job. Another similar opportunity type
example is healthcare, e.g., one person and one unit of capacity, e.g.~a
vaccine shot.

A doubly constrained approach to accessibility calculation needs a
one-to-one relationships between population and opportunities to be
present. Mathematically, this model requires the simultaneous imposition
of both the population- and opportunity- constraints in the preceding
singly constrained variants (Equation~\ref{eq-opportunity-constraint}
and Equation~\ref{eq-population-constraint}), namely the sum of
population in all origins should match the sum of opportunities in all
destinations (Equation~\ref{eq-opportunity-population-equality}):

\begin{equation}\phantomsection\label{eq-opportunity-population-equality}{
\sum_i O_i = \sum_j D_j
}\end{equation}

As before, the simultaneous imposition of both constraints ensures the
total system constraint is maintained i.e., the sum of all doubly
constrained accessibility values
\(\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D\) remains equal to the
total number of opportunities in the region \(O\) as shown in
Equation~\ref{eq-total-constraint-accessibility}.

As the doubly constrained accessibility measure \(V_{ij}^D\) takes the
form of the production-attraction (doubly constrained) spatial
interaction model, as shown in
Equation~\ref{eq-doubly-constrained-gravitymodel}, \(V_{i}^D\) is as
follows:

\begin{equation}\phantomsection\label{eq-doubly-constrained-accessibility}{
V_{ij}^D = A_i B_j O_i D_j f(c_{ij})
}\end{equation}

\noindent where the corresponding balancing factors \(A_i\) and \(B_j\),
as per Wilson, are:

\[
\begin{array}{l}
A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
\end{array}
\]

Calibration of the two sets of proportionality constants is accomplished
by means of iterative proportional fitting, whereby the values of
\(A_i\) are initialized as one for all i to obtain an initial estimate
of \(B_j\). The values of \(B_j\) are used to update the underlying
\(V_{ij}^D\) matrix, before calibrating \(A_i\). This process continues
to update \(A_i\) and \(B_j\) until a convergence criterion is met (see
Ortúzar and Willumsen 2011, 193--95). The proportional allocation factor
\(\kappa_{ij}^D\) would then be:

\[
\kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
\]

One could rewrite Equation~\ref{eq-doubly-constrained-accessibility} as
an origin summary expression of proportionally allocated opportunities:

\begin{equation}\phantomsection\label{eq-doubly-constrained-accessibility-w-paf}{
V^D_{i} = \kappa^D_{ij} \sum_j   D_j
}\end{equation}

However, \(V^D_{i}\) is not wholly helpful, as it will simply equal the
origin-mass marginal. In this way, only \(V^D{ij}\) values are
interpretable. Furthermore, unlike in the total and singly constrained
cases, the doubly-constrained case does not have an interpretable per
capita version. For instance, representing \(V_{ij}^D\) per capita is
not meaningful, as the value \emph{already} matches population to
opportunities. Following this logic, the market potential form
\(M^D_{ij}\) is effectively equivalent to \(V_{ij}^D\), but can be read
with a different interpretation: i.e., the opportunities accessed from
\(j\) at an \(i\) vs.~the population accessed from \(i\) at a \(j\). The
inputs of `opportunities accessed' and `accessed population' can already
be interpreted as inherently being sensitive to both opportunities and
population.

To calculate doubly constrained accessibility, the interpretation of the
population data and the counts of the opportunity data in the numeric
example must be modified. Namely, a count of physician \emph{capacity}
per destination is needed instead of just the number of physicians, as
used to calculated total and singly constrained cases. We also must be
able to clearly state that the population is a count of people seeking
opportunities at the new capacities, i.e., the population must reflect
the \emph{capacity} of the population to interact with opportunities.

So, this adjusted simple example is summarised in
Table~\ref{tbl-small-system-land-use-doubly-constrained-case}: with the
population (in units of 10,000s of people seeking physicians) and the
opportunities (in units of 10,00s of physician-capacity) per zone. For
the population, we leave this unchanged numerically but theoretically
know that each person interacts with one physician capacity (i.e., our
opportunities with a \emph{capacity}). Hence, the number of
opportunities per destination is new: the example is modified such that
the physician-capacity at each zone is an approximately scaled version
of the number of destination-side physicians at each zone from the
unmodified example (Table~\ref{tbl-small-system-land-use}). To emphasis
the new definition of `provider' as physician capacity, the new system
PPR is simply 1, this is compared to the unmodified example which yields
system PPR of24.5. We keep the same zonal cost matrix, and travel
impedance functions for three types of travel behaviour as before
(Table~\ref{tbl-small-system-cost} and
Equation~\ref{eq-travel-behaviour-scenarios}).

\begin{table}

\caption{\label{tbl-small-system-land-use-doubly-constrained-case}Modified
simple system with three zones reflecting matched population and
opportunities. Population is in 10,000 persons and opportunities in
10,000 of physician-capacity.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}rcc}
\toprule
ID (i or j) & Population & Opportunities \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 7 \\ 
2 & 10 & 5 \\ 
3 & 6 & 8 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

However, despite the modifications to the example, our objective remains
the same as in previous cases: to measure accessibility under different
travel behavior scenarios. Specifically, we aim to quantify the number
of \emph{potential} spatial interactions between the physician-seeking
population in each zone \(i\) to the physician capacity in a zone \(j\)
i.e., \(V_{ij}^D\). The highest decay travel behaviour scenario
(\(f_1(c_ij)\)) is presented in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}.

\begin{table}

\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}Doubly
constrained opportunity accessibility assuming the most decay travel
behaviour scenario in the modified simple system.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|rcccc}
\toprule
 &  & \multicolumn{3}{c}{Destination ID} &  \\ 
\cmidrule(lr){3-5}
 & Origin ID & 1 & 2 & 3 & sum \\ 
\midrule\addlinespace[2.5pt]
 & 1 & 3.235859 & 0.01032226 & 0.7556568 & 4 \\ 
 & 2 & 2.132602 & 4.95932483 & 2.9044391 & 10 \\ 
 & 3 & 1.631539 & 0.03035291 & 4.3399040 & 6 \\ 
\midrule 
\midrule 
Sum & — & 7 & 5 & 8 & — \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

As mentioned, accessibility values are typically interpreted as a
summary of the proportionally allocated opportunities at each \(i\).
Hence, in interpreting the doubly constrained accessibility from
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values},
\(V_i^D\) values (i.e., the sum of values at all three \(j\)
destinations for each origin \(i\)) would be 4.0018381, 9.9963656, and
6.0017963 physician-capacity accessible for Zones 1, 2 and 3
respectively. This approximately equal to the number of population at
each of these zones. Conversely, the market potential \(M_j^D\)
interpretation of these values would be 7, 5, and 8 people accessible
from Zones 1, 2 and 3 respectively, equal to the number of
\emph{opportunities} (physician-capacities accessible) at each of these
zones. Notice, the mass weight at the origin equals the mass weight of
the destination: this is precisely the function of the double
constraint. In other words, \(V_i^D\) is the number of accessed
opportunities and \(M_j^D\) is the number of population accessed. For
the other two travel behaviour, identical \(V_i^D\) and \(M_j^D\) values
are calculated, following the same logic. Hence, the usefulness of the
doubly constrained measure lies in the interpretation as \(V_{ij}^D\)
values.

For instance, differences in \(V_{ij}^D\) values between travel
behaviour scenarios are notable. These values can be directly compared
to discuss mass and distance decay impacts. Examining Zone 2 (the
bedroom community),
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}
demonstrates these \(i\) to \(j\) access values for this more relatively
remote, higher-populated and lower-opportunity rich zone. It can be
observed that the number of intrazonal opportunities proportionally
allocated decreases as the assumed distance decay decreases e.g., from
4.9593248 to 2.6672837 out of the \textasciitilde10 opportunities
allocated to Zone 2 (a population of 10). Following the intuition
discussed in the singly constrained opportunity case, as decay
decreases, the mass effect of the population (at origin) and opportunity
(at destination) is more evident: zonal opportunities are supplied and
zonal populations demand at the weights assigned to these zones, with
minimal decay adjustment, reflected by the proportional allocation
factors \(\kappa_{ij}^D\).

\begin{table}

\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}Doubly
constrained opportunity accessibility for all travel behaviour scenarios
for Zone 2 in the modified simple system.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 22.50pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 &  &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 202.50pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{\{ij\}}\textsuperscript{D}} \\ 
\cmidrule(lr){4-6}
 &  &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){4-4} \cmidrule(lr){5-5} \cmidrule(lr){6-6}
Dest. & Population at 2 (units: \emph{people in 10,000s}) & Opportunities (units: \emph{capacity in 10,000s}) & units: \emph{physician-capacity in 10,000s} & units: \emph{physician-capacity in 10,000s} & units: \emph{physician-capacity in 10,000s} \\ 
\midrule\addlinespace[2.5pt]
1 & 10.000 & 7.000 & 2.133 & 2.272 & 3.411 \\ 
2 & 10.000 & 5.000 & 4.959 & 4.766 & 2.667 \\ 
3 & 10.000 & 8.000 & 2.904 & 2.958 & 3.919 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Recall, accessibility is defined as ``the potential for interaction''
and is traditionally presented as a summary zonal measure. In the doubly
constrained case, we force the zonal population and zonal opportunities
to match one-to-one, hence providing a zonal summary is no longer
relevant: the sum of \(V_{ij}^D\) for all \(i\)s ends up being equal to
the population at the zone. However, if what interests readers is the
``potential for interaction'' in the case population and opportunities
to match one-to-one, reframing the investigation may be needed. In this
sense, it would be examining ``interaction'' (much less room for
potential) through the values of \(V_{ij}^D\) e.g., how many
opportunities are being allocated to an origin from a destination (or in
a transposed sense for \(M_{ij}^D\)). In this sense, the doubly
constrained case can be thought of as an estimate of \emph{realized}
accessibility or access: reflecting spatial interaction. It is also
formulaically identical to the doubly constrained spatial interaction
model, but with specific interpretations of the origin and destination
weights as `population' and `opportunity', respectively.

As Wilson explicitly noted, origin and destination weights defined in
the spatial interaction model \emph{can} be defined using any unit.
Accessibility, however, is often presented and understood as a zonal
summary of \emph{potential} for interaction between origins and
destinations that contain inherently different units. Arriving at the
doubly constrained case through the unconstrained, totally, and singly
constrained cases make the connection between the \emph{potential} for
spatial interaction (accessibility) and \emph{realized} potential for
spatial interaction more interpretable. Namely, by demonstrating that
these members of the accessibility family all derive from the same root
and can be derived from Wilson's original formulation, this more clearly
demonstrates what \emph{potential} is, within the context of spatial
interaction modelling.

Namely, \emph{potential} depends on the framing of the masses at the
origin and destination, and how similar they are in their units. The
appropriate constraint should be decided based on the the input data and
their similarity. In increasing unit similarity from the perspective of
opportunity-constraint: the total-constraint can be used if population
(origin mass) is not known, the single constraint can be used if
population is known and matters to \emph{potential} but it does not
match the capacity of opportunities, and lastly, the double constraint
is appropriate if population is known, matters and population matches
the capacity of opportunities. These constraint measures can reflect the
opportunities accessible at each zone, but reflects the assumptions
embedded in the input data about the potential to spatially interact.

\section{Conclusions}\label{conclusions}

In this work, we examined the historical and mathematical commonalities
between spatial interaction models and place-based accessibility
measures. As accessibility research evolved largely influenced by Hansen
(1959), researchers in the field neglected the proportionality constant
that was originally present in gravity-based models, and is still
present in spatial interaction modeling. Firstly, we have demonstrated
that by reintroducing such constraints and defining associated balancing
factors and proportional allocation factors, we can derive a unifying
family of accessibility measures that reintroduces units to the
resulting values. These values become more intuitive for the purpose of
analysis and comparison. Secondly, the constraining constants, depending
on the case (i.e., total, singly- or doubly- constrained), restrict the
degree of \emph{potential}, linking accessibility (the potential for
spatial interaction) with access (spatial interaction) on the same
continuum based on the constraint used. Lastly, we discussed how popular
measures such as the one used in Hansen (1959) and the 2SFCA link into
the family of accessibility measures. We summarise the family of
accessibility measures, as follows.

We first place the popular Hansen-type accessibility measure (Hansen
1959) within this family of measures as an ``unconstrained'' case,
demonstrating that resulting values cannot be directly compared across
different travel scenarios without ad-hoc adjustments. We then show how
applying a total constraint balances the units and produces a
statistically averaged solution that converges to the regional average
for each zone as the decay effect decreases. In other words, the
total-constraint model could be a more interpretable alternative for the
unconstrained case if population-competition is not relevant and one is
interested in capturing the maximum \emph{potential}; specifically, if
there is a fixed number of opportunities in the region, and if it makes
sense to assume that people accessing proximate opportunities leave
fewer for others, \emph{without} considering the population size at the
origins.

We then introduce the singly constrained case, which \emph{does} takes
into account the population size at the origin in the allocation of
opportunities (unlike the total constraint). It is also mathematically
equivalent to the spatial availability introduced in Soukhov et al.
(2023). In this case, all accessibility values are fixed to sum to a
known zonal opportunity-size value (implicitly, the regional total of
opportunities), but they are not required to sum to any population-based
values at the zone or regional level. The singly constrained model could
be useful if regional competition is a factor and if the acknowledgment
that only a finite number of opportunities can be allocated from each
destination (with those allocations distributed based on origin
population size) is suitable. We also introduce an `accessible' PPR
(e.g., opportunities per capita), calculated by dividing each
accessibility value by the zonal population. To clarify, this per capita
expression of the singly constrained case is equivalent to the 2SFCA
(Luo and Wang 2003; Q. Shen 1998b), hence linking this literature back
to spatial interaction principles.

Lastly, the doubly constrained case is introduced. In this case, the
sums must equal both the regional total and ensure that no zone
allocates more opportunities than it has available. Specifically,
accessibility values for each \(i\)-\(j\) pair must be a proportion of
he zonal opportunity and population values simultaneously. For example,
the accessibility at zone 1 must equal the sum of opportunities from
zones 1, 2, and 3, as well as the sum of the population at zone 1.
Satisfying the double constraint means the opportunities and population
data must match one-to-one, so working with the accessibility
\(i\)-\(j\) pair values should be of interest. In this sense, the
research question should be concerned with `access' (how many
opportunities accessed from \(j\) at \(i\) based on given zonal
opportunities and populations) instead of potential spatial interaction
(e.g., typically expressed as a zonal summary measure of how many
opportunities one could reach (out of a regional total and/or
zonal-allocation).

And as a summary: each member of the family of accessibility measure is
named, explained in plain language, alongside their balancing factor,
proportional allocation factor and mathematical equation and value
interpretations in Table 8.

\input{summary-family-access-measures-table.tex}

Table 8: Summary of constrained accessibility measure types and
interpretations

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-albacete2017measuring}
Albacete, Xavier, Doina Olaru, Valerià Paül, and Sharon Biermann. 2017.
{``Measuring the Accessibility of Public Transport: A Critical
Comparison Between Methods in Helsinki.''} \emph{Applied Spatial
Analysis and Policy} 10: 161--88.

\bibitem[\citeproctext]{ref-allenMeasureCompetitiveAccess2020}
Allen, Jeff, and Steven Farber. 2020. {``A Measure of Competitive Access
to Destinations for Comparing Across Multiple Study Regions.''}
\emph{Geographical Analysis} 52 (1): 69--86.
\url{https://doi.org/10.1111/gean.12188}.

\bibitem[\citeproctext]{ref-alonso2014labour}
Alonso, M Pilar, MA Beamonte, Pilar Gargallo, and Manuel J Salvador.
2014. {``Labour and Residential Accessibility: A Bayesian Analysis Based
on Poisson Gravity Models with Spatial Effects.''} \emph{Journal of
Geographical Systems} 16: 409--39.

\bibitem[\citeproctext]{ref-ashiru2003space}
Ashiru, Olu, John W Polak, and Robert B Noland. 2003. {``Space-Time User
Benefit and Utility Accessibility Measures for Individual Activity
Schedules.''} \emph{Transportation Research Record} 1854 (1): 62--73.

\bibitem[\citeproctext]{ref-battyChronicleScientificPlanning1994}
Batty, Michael. 1994. {``A {Chronicle} of {Scientific} {Planning}: {The}
{Anglo}-{American} {Modeling} {Experience}.''} \emph{Journal of the
American Planning Association} 60 (1): 7--16.
\url{https://doi.org/10.1080/01944369408975546}.

\bibitem[\citeproctext]{ref-battyMethodResiduesUrban1976}
Batty, M, and L March. 1976. {``The Method of Residues in Urban
Modelling.''} \emph{Environment and Planning A: Economy and Space} 8
(2): 189--214. \url{https://doi.org/10.1068/a080189}.

\bibitem[\citeproctext]{ref-beaumontLocationallocationProblemsPlane1981}
Beaumont, John R. 1981. {``Location-Allocation Problems in a Plane a
Review of Some Models.''} \emph{Socio-Economic Planning Sciences} 15
(5): 217--29. \url{https://doi.org/10.1016/0038-0121(81)90042-2}.

\bibitem[\citeproctext]{ref-beckers2022incorporating}
Beckers, Joris, Mark Birkin, Graham Clarke, Nick Hood, Andy Newing, and
Ryan Urquhart. 2022. {``Incorporating e-Commerce into Retail Location
Models.''} \emph{Geographical Analysis} 54 (2): 274--93.

\bibitem[\citeproctext]{ref-boisjoly2017informality}
Boisjoly, Geneviève, Ana Isabel Moreno-Monroy, and Ahmed El-Geneidy.
2017. {``Informality and Accessibility to Health by Public Transit:
Evidence from the São Paulo Metropolitan Region.''} Journal Article.
\emph{Journal of Transport Geography} 64: 89--96.
\url{https://doi.org/10.1016/j.jtrangeo.2017.08.005}.

\bibitem[\citeproctext]{ref-careyPrinciplesSocialScience1858}
Carey, Henry Charles. 1858. \emph{Principles of Social Science}.
University of Michigan Library Digital Collections: In the digital
collection Making of America Books.
\url{https://name.umdl.umich.edu/AFR1829.0001.001.}

\bibitem[\citeproctext]{ref-caschili2015accessibility}
Caschili, Simone, Andrea De Montis, and Daniele Trogu. 2015.
{``Accessibility and Rurality Indicators for Regional Development.''}
\emph{Computers, Environment and Urban Systems} 49: 98--114.

\bibitem[\citeproctext]{ref-cavendish_xxi_1798}
Cavendish, Henry. 1798. {``{XXI}. {Experiments} to Determine the Density
of the Earth.''} \emph{Philosophical Transactions of the Royal Society}
88: 469--526. \url{https://doi.org/10.1098/rstl.1798.0022}.

\bibitem[\citeproctext]{ref-chen2013regional}
Chen, Guineng, and João de Abreu e Silva. 2013. {``Regional Impacts of
High-Speed Rail: A Review of Methods and Models.''} \emph{Transportation
Letters} 5 (3): 131--43.

\bibitem[\citeproctext]{ref-chia2020extending}
Chia, Jason, and Jinwoo Brian Lee. 2020. {``Extending Public Transit
Accessibility Models to Recognise Transfer Location.''} \emph{Journal of
Transport Geography} 82: 102618.

\bibitem[\citeproctext]{ref-clarke2002deriving}
Clarke, Graham, Heather Eyre, and Cliff Guy. 2002. {``Deriving
Indicators of Access to Food Retail Provision in British Cities: Studies
of Cardiff, Leeds and Bradford.''} \emph{Urban Studies} 39 (11):
2041--60.

\bibitem[\citeproctext]{ref-cliff_evaluating_1974}
Cliff, A. D., R. L. Martin, and J. K. Ord. 1974. {``Evaluating the
Friction of Distance Parameter in Gravity Models.''} \emph{Regional
Studies} 8 (3-4): 281--86.
\url{https://doi.org/10.1080/09595237400185281}.

\bibitem[\citeproctext]{ref-condecco2018road}
Condeço-Melhorado, Ana, and Panayotis Christidis. 2018. {``Road
Accessibility in Border Regions: A Joint Approach.''} \emph{Networks and
Spatial Economics} 18 (2): 363--83.

\bibitem[\citeproctext]{ref-curtis2010planning}
Curtis, Carey, and Jan Scheurer. 2010. {``Planning for Sustainable
Accessibility: Developing Tools to Aid Discussion and
Decision-Making.''} \emph{Progress in Planning} 74 (2): 53--106.

\bibitem[\citeproctext]{ref-dai2017visualization}
Dai, Linlin, Li Wan, and Shijie Gai. 2017. {``A Visualization Framework
for Synthesizing Spatial Impacts from Multiple Site Factors.''}
\emph{Journal of Asian Architecture and Building Engineering} 16 (2):
311--15.

\bibitem[\citeproctext]{ref-delamater2013spatial}
Delamater, P. L. 2013. {``Spatial Accessibility in Suboptimally
Configured Health Care Systems: A Modified Two-Step Floating Catchment
Area (M2SFCA) Metric.''} Journal Article. \emph{Health \& Place} 24:
30--43. \url{https://doi.org/10.1016/j.healthplace.2013.07.012}.

\bibitem[\citeproctext]{ref-el2011place}
El-Geneidy, Ahmed, and David Levinson. 2011. {``Place Rank: Valuing
Spatial Interactions.''} \emph{Networks and Spatial Economics} 11:
643--59.

\bibitem[\citeproctext]{ref-elgeneidyMakingAccessibilityWork2022}
---------. 2022. {``Making Accessibility Work in Practice.''}
\emph{Transport Reviews} 42 (2): 129--33.
\url{https://doi.org/10.1080/01441647.2021.1975954}.

\bibitem[\citeproctext]{ref-farber_2013_social}
Farber, S., T. Neutens, H. J. Miller, and X. Li. 2013. {``The Social
Interaction Potential of Metropolitan Regions: A Time-Geographic
Measurement Approach Using Joint Accessibility.''} Journal Article.
\emph{Annals of the Association of American Geographers} 103 (3):
483--504. \url{https://doi.org/10.1080/00045608.2012.689238}.

\bibitem[\citeproctext]{ref-farberActivitySpacesMeasurement2012}
Farber, S., A. Páez, and C. Morency. 2012. {``Activity Spaces and the
Measurement of Clustering and Exposure: A Case Study of Linguistic
Groups in {Montreal}.''} \emph{Environment and Planning A} 44 (2):
315--32.

\bibitem[\citeproctext]{ref-farber_running_2011}
Farber, Steven, and Antonio Páez. 2011. {``Running to Stay in Place: The
Time-Use Implications of Automobile Oriented Land-Use and Travel.''}
\emph{Journal of Transport Geography} 19 (4): 782--93.
\url{https://doi.org/10.1016/j.jtrangeo.2010.09.008}.

\bibitem[\citeproctext]{ref-ferreiraReenactingMobilityAccessibility2020}
Ferreira, António, and Enrica Papa. 2020. {``Re-Enacting the Mobility
Versus Accessibility Debate: Moving Towards Collaborative Synergies
Among Experts.''} \emph{Case Studies on Transport Policy} 8 (3):
1002--9. \url{https://doi.org/10.1016/j.cstp.2020.04.006}.

\bibitem[\citeproctext]{ref-fotheringham_spatial_1984}
Fotheringham, A S. 1984. {``Spatial {Flows} and {Spatial} {Patterns}.''}
\emph{Environment and Planning A: Economy and Space} 16 (4): 529--43.
\url{https://doi.org/10.1068/a160529}.

\bibitem[\citeproctext]{ref-fotheringhamSpatialCompetitionAgglomeration1985}
---------. 1985. {``Spatial Competition and Agglomeration in Urban
Modelling.''} \emph{Environment and Planning A: Economy and Space} 17
(2): 213--30. \url{https://doi.org/10.1068/a170213}.

\bibitem[\citeproctext]{ref-fotheringhamSPATIALSTRUCTUREDISTANCE1981}
Fotheringham, A. Stewart. 1981. {``{SPATIAL} {STRUCTURE} {AND}
{DISTANCE}‐{DECAY} {PARAMETERS}.''} \emph{Annals of the Association of
American Geographers} 71 (3): 425--36.
\url{https://doi.org/10.1111/j.1467-8306.1981.tb01367.x}.

\bibitem[\citeproctext]{ref-geursAccessibilityEvaluationLanduse2004}
Geurs, Karst T., and Bert van Wee. 2004. {``Accessibility Evaluation of
Land-Use and Transport Strategies: Review and Research Directions.''}
\emph{Journal of Transport Geography} 12 (2): 127--40.
\url{https://doi.org/10.1016/j.jtrangeo.2003.10.005}.

\bibitem[\citeproctext]{ref-geurs2006accessibility}
Geurs, Karst T, Bert van Wee, and Piet Rietveld. 2006. {``Accessibility
Appraisal of Integrated Land-Use---Transport Strategies: Methodology and
Case Study for the Netherlands Randstad Area.''} \emph{Environment and
Planning B: Planning and Design} 33 (5): 639--60.

\bibitem[\citeproctext]{ref-giuliano2010accessibility}
Giuliano, Genevieve, Peter Gordon, Qisheng Pan, and JiYoung Park. 2010.
{``Accessibility and Residential Land Values: Some Tests with New
Measures.''} \emph{Urban Studies} 47 (14): 3103--30.

\bibitem[\citeproctext]{ref-grengs2004measuring}
Grengs, Joe. 2004. {``Measuring Change in Small-Scale Transit
Accessibility with Geographic Information Systems: Buffalo and
Rochester, New York.''} \emph{Transportation Research Record} 1887 (1):
10--17.

\bibitem[\citeproctext]{ref-grengs2010job}
---------. 2010. {``Job Accessibility and the Modal Mismatch in
Detroit.''} \emph{Journal of Transport Geography} 18 (1): 42--54.

\bibitem[\citeproctext]{ref-grengs2012equity}
---------. 2012. {``Equity and the Social Distribution of Job
Accessibility in Detroit.''} \emph{Environment and Planning B: Planning
and Design} 39 (5): 785--800.

\bibitem[\citeproctext]{ref-grengs2015nonwork}
---------. 2015. {``Nonwork Accessibility as a Social Equity
Indicator.''} \emph{International Journal of Sustainable Transportation}
9 (1): 1--14.

\bibitem[\citeproctext]{ref-grengs2010intermetropolitan}
Grengs, Joe, Jonathan Levine, Qing Shen, and Qingyun Shen. 2010.
{``Intermetropolitan Comparison of Transportation Accessibility: Sorting
Out Mobility and Proximity in San Francisco and Washington, DC.''}
\emph{Journal of Planning Education and Research} 29 (4): 427--43.

\bibitem[\citeproctext]{ref-gutierrez_evaluating_2011}
Gutierrez, J., A. Condeco-Melhorado, E. Lopez, and A. Monzon. 2011.
{``Evaluating the {European} Added Value of {TEN}-{T} Projects: A
Methodological Proposal Based on Spatial Spillovers, Accessibility and
{GIS}.''} \emph{Journal of Transport Geography} 19 (4): 840--50.
\url{https://doi.org/10.1016/j.jtrangeo.2010.10.011}.

\bibitem[\citeproctext]{ref-gutierrezLocationEconomicPotential2001}
Gutiérrez, Javier. 2001. {``Location, Economic Potential and Daily
Accessibility: An Analysis of the Accessibility Impact of the High-Speed
Line Madrid--Barcelona--French Border.''} \emph{Journal of Transport
Geography} 9 (4): 229--42.
\url{https://doi.org/10.1016/S0966-6923(01)00017-5}.

\bibitem[\citeproctext]{ref-handyMeasuringAccessibilityExploration1997}
Handy, Sandy L., and Debbie A. Niemeier. 1997. {``Measuring
Accessibility: An Exploration of Issues and Alternatives.''}
\emph{Environment and Planning A: Economy and Space} 29 (7): 1175--94.
\url{https://doi.org/10.1068/a291175}.

\bibitem[\citeproctext]{ref-handyACCESSIBILITYVSMOBILITYENHANCING2002}
Handy, Susan. 2002. {``{ACCESSIBILITY}- {VS}. {MOBILITY}-{ENHANCING}
{STRATEGIES} {FOR} {ADDRESSING} {AUTOMOBILE} {DEPENDENCE} {IN} {THE}
u.s.''}
\url{https://escholarship.org/content/qt5kn4s4pb/qt5kn4s4pb_noSplash_18f73162ff86f04dcb255331d63eeba8.pdf}.

\bibitem[\citeproctext]{ref-handy2020}
---------. 2020. {``Is Accessibility an Idea Whose Time Has Finally
Come?''} \emph{Transportation Research Part D: Transport and
Environment} 83 (June): 102319.
\url{https://doi.org/10.1016/j.trd.2020.102319}.

\bibitem[\citeproctext]{ref-hansen1959}
Hansen, Walter G. 1959. {``How Accessibility Shapes Land Use.''}
\emph{Journal of the American Institute of Planners} 25 (2): 73--76.
\url{https://doi.org/10.1080/01944365908978307}.

\bibitem[\citeproctext]{ref-harrisEquilibriumValuesDynamics1978}
Harris, B, and A G Wilson. 1978. {``Equilibrium Values and Dynamics of
Attractiveness Terms in Production-Constrained Spatial-Interaction
Models.''} \emph{Environment and Planning A: Economy and Space} 10 (4):
371--88. \url{https://doi.org/10.1068/a100371}.

\bibitem[\citeproctext]{ref-harris_market_1954}
Harris, Chauncy D. 1954. {``The {Market} as a {Factor} in the
{Localization} of {Industry} in the {United} {States}.''} \emph{Annals
of the Association of American Geographers} 44 (4): 315--48.
\url{https://www.jstor.org/stable/2561395}.

\bibitem[\citeproctext]{ref-he2017measuring}
He, Jianhua, Chun Li, Yan Yu, Yaolin Liu, and Junlong Huang. 2017.
{``Measuring Urban Spatial Interaction in Wuhan Urban Agglomeration,
Central China: A Spatially Explicit Approach.''} \emph{Sustainable
Cities and Society} 32: 569--83.

\bibitem[\citeproctext]{ref-holl2007twenty}
Holl, Adelheid. 2007. {``Twenty Years of Accessibility Improvements. The
Case of the Spanish Motorway Building Programme.''} \emph{Journal of
Transport Geography} 15 (4): 286--97.

\bibitem[\citeproctext]{ref-hutton_xxxiii_1778}
Hutton, Charles. 1778. {``{XXXIII}. {An} Account of the Calculations
Made from the Survey and Measures Taken at {Schehallien}, in Order to
Ascertain the Mean Density of the {Earth}.''} \emph{Philosophical
Transactions of the Royal Society} 68: 689--788.
\url{https://doi.org/10.1098/rstl.1778.0034}.

\bibitem[\citeproctext]{ref-kapatsila_resolving_2023}
Kapatsila, Bogdan, Manuel Santana Palacios, Emily Grisé, and Ahmed
El-Geneidy. 2023. {``Resolving the Accessibility Dilemma: {Comparing}
Cumulative and Gravity-Based Measures of Accessibility in Eight
{Canadian} Cities.''} \emph{Journal of Transport Geography} 107
(February): 103530.
\url{https://doi.org/10.1016/j.jtrangeo.2023.103530}.

\bibitem[\citeproctext]{ref-karstEvaluationAccessibilityImpacts2003}
Karst, T, and Jan R Ritsema Van Eck. 2003. {``Evaluation of
Accessibility Impacts of Land-Use Scenarios: The Implications of Job
Competition, Land-Use, and Infrastructure Developments for the
Netherlands.''} \emph{Environment and Planning B: Planning and Design}
30 (1): 69--87. \url{https://doi.org/10.1068/b12940}.

\bibitem[\citeproctext]{ref-kharel2024examining}
Kharel, Subham, Seyedsoheil Sharifiasl, and Qisheng Pan. 2024.
{``Examining Food Access Equity by Integrating Grocery Store Pricing
into Spatial Accessibility Measures.''} \emph{Transportation Research
Record}, 03611981241254382.

\bibitem[\citeproctext]{ref-kirbyNormalizingFactorsGravity1970}
Kirby, Howard R. 1970. {``Normalizing Factors of the Gravity Model---an
Interpretation.''} \emph{Transportation Research} 4 (1): 37--50.
\url{https://doi.org/10.1016/0041-1647(70)90073-0}.

\bibitem[\citeproctext]{ref-kovatch1971modeling}
Kovatch, George, George Zames, et al. 1971. {``Modeling Transportation
Systems: An Overview.''} \url{https://rosap.ntl.bts.gov/view/dot/11813}.

\bibitem[\citeproctext]{ref-kwan1998space}
Kwan, Mei-Po. 1998. {``Space-Time and Integral Measures of Individual
Accessibility: A Comparative Analysis Using a Point-Based Framework.''}
\emph{Geographical Analysis} 30 (3): 191--216.

\bibitem[\citeproctext]{ref-lavery_driving_2013}
Lavery, T. A., A. Páez, and P. S. Kanaroglou. 2013. {``Driving Out of
Choices: {An} Investigation of Transport Modality in a University
Sample.''} \emph{Transportation Research Part A: Policy and Practice} 57
(November): 37--46. \url{https://doi.org/10.1016/j.tra.2013.09.010}.

\bibitem[\citeproctext]{ref-leonardiOptimumFacilityLocation1978}
Leonardi, G. 1978. {``Optimum Facility Location by Accessibility
Maximizing.''} \emph{Environment and Planning A: Economy and Space} 10
(11): 1287--1305. \url{https://doi.org/10.1068/a101287}.

\bibitem[\citeproctext]{ref-leonardiRandomUtilityDemand1984}
Leonardi, Giorgio, and Roberto Tadei. 1984. {``Random Utility Demand
Models and Service Location.''} \emph{Regional Science and Urban
Economics} 14 (3): 399--431.
\url{https://doi.org/10.1016/0166-0462(84)90009-7}.

\bibitem[\citeproctext]{ref-levine2012does}
Levine, Jonathan, Joe Grengs, Qingyun Shen, and Qing Shen. 2012. {``Does
Accessibility Require Density or Speed? A Comparison of Fast Versus
Close in Getting Where You Want to Go in US Metropolitan Regions.''}
\emph{Journal of the American Planning Association} 78 (2): 157--72.

\bibitem[\citeproctext]{ref-levinson2012positive}
Levinson, David, and Arthur Huang. 2012. {``A Positive Theory of Network
Connectivity.''} \emph{Environment and Planning B: Planning and Design}
39 (2): 308--25.

\bibitem[\citeproctext]{ref-levinsonGeneralTheoryAccess2020}
Levinson, David, and Hao Wu. 2020. {``Towards a General Theory of
Access.''} \emph{Journal of Transport and Land Use} 13 (1): 129--58.
\url{https://www.jstor.org/stable/26967239}.

\bibitem[\citeproctext]{ref-liang_novel_2024}
Liang, Huilin, Qi Yan, and Yujia Yan. 2024. {``A Novel Spatiotemporal
Framework for Accessing Green Space Accessibility Change in Adequacy and
Equity: {Evidence} from a Rapidly Urbanizing {Chinese} {City} in
2012--2021.''} \emph{Cities} 151 (August): 105112.
\url{https://doi.org/10.1016/j.cities.2024.105112}.

\bibitem[\citeproctext]{ref-liu2004accessibility}
Liu, Suxia, and Xuan Zhu. 2004. {``Accessibility Analyst: An Integrated
GIS Tool for Accessibility Analysis in Urban Transportation Planning.''}
\emph{Environment and Planning B: Planning and Design} 31 (1): 105--24.

\bibitem[\citeproctext]{ref-liu2015spatial}
Liu, Xianteng, and Jiangping Zhou. 2015. {``Spatial Pattern of Land Use
and Its Implications for Mode-Based Accessibility: Case Study of
Nanjing, China.''} \emph{Journal of Urban Planning and Development} 141
(2): 05014012.

\bibitem[\citeproctext]{ref-lopezMeasuring2008}
Lopez, E., J. Gutierrez, and G. Gomez. 2008. {``Measuring Regional
Cohesion Effects of Large-Scale Transport Infrastructure Investments:
{An} Accessibility Approach.''} \emph{European Planning Studies} 16 (2):
277--301.

\bibitem[\citeproctext]{ref-luoEnhanced2009}
Luo, Wei, and Y. Qi. 2009. {``An Enhanced Two-Step Floating Catchment
Area ({E2SFCA}) Method for Measuring Spatial Accessibility to Primary
Care Physicians.''} \emph{Health \& Place} 15 (4): 1100--1107.

\bibitem[\citeproctext]{ref-luo2003}
Luo, Wei, and Fahui Wang. 2003. {``Measures of Spatial Accessibility to
Health Care in a GIS Environment: Synthesis and a Case Study in the
Chicago Region.''} \emph{Environment and Planning B: Planning and
Design} 30 (6): 865--84. \url{https://doi.org/10.1068/b29120}.

\bibitem[\citeproctext]{ref-manaugh2012makes}
Manaugh, Kevin, and Ahmed M El-Geneidy. 2012. {``What Makes
Travel'local' Defining and Understanding Local Travel Behavior.''}
\emph{Journal of Transport and Land Use} 5 (3): 15--27.

\bibitem[\citeproctext]{ref-margaridacondecomelhoradoImpactMeasuringInternal2016}
Margarida Condeço Melhorado, Ana, Hande Demirel, Mert Kompil, Elena
Navajas, and Panayotis Christidis. 2016. {``The Impact of Measuring
Internal Travel Distances on Selfpotentials and Accessibility.''}
\emph{European Journal of Transport and Infrastructure Research}.
\url{https://doi.org/10.18757/EJTIR.2016.16.2.3139}.

\bibitem[\citeproctext]{ref-marques_accessibility_2021}
Marques, João Lourenço, Jan Wolf, and Fillipe Feitosa. 2021.
{``Accessibility to Primary Schools in {Portugal}: A Case of Spatial
Inequity?''} \emph{Regional Science Policy \& Practice} 13 (3):
693--708. \url{https://doi.org/10.1111/rsp3.12303}.

\bibitem[\citeproctext]{ref-marwal2022literature}
Marwal, Aviral, and Elisabete Silva. 2022. {``Literature Review of
Accessibility Measures and Models Used in Land Use and Transportation
Planning in Last 5 Years.''} \emph{Journal of Geographical Sciences} 32
(3): 560--84.

\bibitem[\citeproctext]{ref-mayaud2019future}
Mayaud, Jerome R, Martino Tran, Rafael HM Pereira, and Rohan Nuttall.
2019. {``Future Access to Essential Services in a Growing Smart City:
The Case of Surrey, British Columbia.''} \emph{Computers, Environment
and Urban Systems} 73: 1--15.

\bibitem[\citeproctext]{ref-mckeanManual1883}
McKean, Kate. 1883. \emph{Manual of {Social} {Science} Being a
{Condensation} of the {Principles} of {Social} {Science} of {H}.{C}.
{Carey}}. Philadelphia: Henry Carey Baird; Co. Industrial Publishers.

\bibitem[\citeproctext]{ref-mdotMnDOTJoins2007}
MDOT. 2007. {``Mn/{DOT} Joins Interstate Highway System's 50th
Anniversary Celebration.''} December 4, 2007.
\url{https://web.archive.org/web/20071204072603/http://www.dot.state.mn.us/interstate50/50facts.html}.

\bibitem[\citeproctext]{ref-merlin2017competition}
Merlin, Louis A., and Lingqian Hu. 2017. {``Does Competition Matter in
Measures of Job Accessibility? Explaining Employment in Los Angeles.''}
Journal Article. \emph{Journal of Transport Geography} 64: 77--88.
\url{https://doi.org/10.1016/j.jtrangeo.2017.08.009}.

\bibitem[\citeproctext]{ref-millerAccessibilityMeasurementApplication2018}
Miller, Eric J. 2018. {``Accessibility: Measurement and Application in
Transportation Planning.''} \emph{Transport Reviews} 38 (5): 551--55.
\url{https://doi.org/10.1080/01441647.2018.1492778}.

\bibitem[\citeproctext]{ref-millerMeasuringSpaceTimeAccessibility1999}
Miller, Harvey J. 1999. {``Measuring Space‐time Accessibility Benefits
Within Transportation Networks: Basic Theory and Computational
Procedures.''} \emph{Geographical Analysis} 31 (1): 187--212.
\url{https://doi.org/10.1111/gean.1999.31.1.187}.

\bibitem[\citeproctext]{ref-miller_collaborative_2011}
---------. 2011. {``Collaborative Mobility: Using Geographic Information
Science to Cultivate Cooperative Transportation Systems.''}
\emph{Procedia - Social and Behavioral Sciences} 21 (0): 24--28.
https://doi.org/\url{http://dx.doi.org/10.1016/j.sbspro.2011.07.005}.

\bibitem[\citeproctext]{ref-morrisAccessibilityIndicatorsTransport1979}
Morris, J. M., P. L. Dumble, and M. R. Wigan. 1979. {``Accessibility
Indicators for Transport Planning.''} \emph{Transportation Research Part
A: General} 13 (2): 91--109.
\url{https://doi.org/10.1016/0191-2607(79)90012-8}.

\bibitem[\citeproctext]{ref-naqavi2023mobility}
Naqavi, Fatemeh, Marcus Sundberg, Oskar Blom Västberg, Anders Karlström,
and Muriel Beser Hugosson. 2023. {``Mobility Constraints and
Accessibility to Work: Application to Stockholm.''} \emph{Transportation
Research Part A: Policy and Practice} 175: 103790.

\bibitem[\citeproctext]{ref-neutens_human_2007}
Neutens, Tijs, Frank Witlox, Nico Van de Weghe, and Philippe De Maeyer.
2007. {``Human Interaction Spaces Under Uncertainty.''}
\emph{Transportation Research Record} 2021 (1): 28--35.

\bibitem[\citeproctext]{ref-ng2022reflection}
Ng, Matthew Kok Ming, Josephine Roper, Chyi Lin Lee, and Christopher
Pettit. 2022. {``The Reflection of Income Segregation and Accessibility
Cleavages in Sydney's House Prices.''} \emph{ISPRS International Journal
of Geo-Information} 11 (7): 413.

\bibitem[\citeproctext]{ref-ortuzar_2011_modelling}
Ortúzar, J. D., and L. G. Willumsen. 2011. \emph{Modelling Transport}.
Book. New York: Wiley.

\bibitem[\citeproctext]{ref-paez_developing_2013}
Paez, A., M. Moniruzzaman, P. L. Bourbonnais, and C. Morency. 2013.
{``Developing a Web-Based Accessibility Calculator Prototype for the
{Greater} {Montreal} {Area}.''} \emph{Transportation Research Part
A-Policy and Practice} 58 (December): 103--15.
\url{https://doi.org/10.1016/j.tra.2013.10.020}.

\bibitem[\citeproctext]{ref-paez2012measuring}
Paez, A., D. M. Scott, and C. Morency. 2012. {``Measuring Accessibility:
Positive and Normative Implementations of Various Accessibility
Indicators.''} Journal Article. \emph{Journal of Transport Geography}
25: 141--53. \url{https://doi.org/10.1016/j.jtrangeo.2012.03.016}.

\bibitem[\citeproctext]{ref-paez_jobs_2013}
Páez, Antonio, Steven Farber, Ruben Mercado, Matthew Roorda, and
Catherine Morency. 2013. {``Jobs and the {Single} {Parent}: {An}
{Analysis} of {Accessibility} to {Employment} in {Toronto}.''}
\emph{Urban Geography} 34 (6): 815--42.
\url{https://doi.org/10.1080/02723638.2013.778600}.

\bibitem[\citeproctext]{ref-paez_healthcare_2010}
Páez, Antonio, Ruben Mercado, Steven Farber, Catherine Morency, and
Matthew Roorda. 2010. {``Accessibility to Health Care Facilities in
Montreal Island: An Application of Relative Accessibility Indicators
from the Perspective of Senior and Non-Senior Residents.''} Journal
Article. \emph{International Journal of Health Geographics} 9 (52):
1--9. \url{http://www.ij-healthgeographics.com/content/9/1/52}.

\bibitem[\citeproctext]{ref-pan2013impacts}
Pan, Qisheng. 2013. {``The Impacts of an Urban Light Rail System on
Residential Property Values: A Case Study of the Houston METRORail
Transit Line.''} \emph{Transportation Planning and Technology} 36 (2):
145--69.

\bibitem[\citeproctext]{ref-pan2020measuring}
Pan, Qisheng, Zhonghua Jin, and Xuejun Liu. 2020. {``Measuring the
Effects of Job Competition and Matching on Employment Accessibility.''}
\emph{Transportation Research Part D: Transport and Environment} 87:
102535.

\bibitem[\citeproctext]{ref-pereira_2021_geographic}
Pereira, Rafael H. M., Carlos Kauê Vieira Braga, Luciana Mendes Servo,
Bernardo Serra, Pedro Amaral, Nelson Gouveia, and Antonio Paez. 2021.
{``Geographic Access to COVID-19 Healthcare in Brazil Using a Balanced
Float Catchment Area Approach.''} Journal Article. \emph{Social Science
\& Medicine} 273: 113773.
https://doi.org/\url{https://doi.org/10.1016/j.socscimed.2021.113773}.

\bibitem[\citeproctext]{ref-pirie_measuring_1979}
Pirie, G. H. 1979. {``Measuring {Accessibility}: {A} {Review} and
{Proposal}.''} \emph{Environment and Planning A: Economy and Space} 11
(3): 299--312. \url{https://doi.org/10.1068/a110299}.

\bibitem[\citeproctext]{ref-rau2012spatial}
Rau, Henrike, and Amaya Vega. 2012. {``Spatial (Im) Mobility and
Accessibility in i Reland: Implications for Transport Policy.''}
\emph{Growth and Change} 43 (4): 667--96.

\bibitem[\citeproctext]{ref-ravensteinLawsMigration1885}
Ravenstein, E. G. 1885. {``The Laws of Migration Paper 1.''}
\emph{Journal of the Royal Statistical Society} 48 (2): 167--227.

\bibitem[\citeproctext]{ref-ravensteinLawsMigration1889}
---------. 1889. {``The Laws of Migration Paper 2.''} \emph{Journal of
the Royal Statistical Society} 52 (2): 241--305.
\url{https://doi.org/10.2307/2979333}.

\bibitem[\citeproctext]{ref-reggianiGuestEditorialNew2011}
Reggiani, Aura, and Juan Carlos Martín. 2011. {``Guest Editorial: New
Frontiers in Accessibility Modelling: An Introduction.''} \emph{Networks
and Spatial Economics} 11 (4): 577--80.
\url{https://doi.org/10.1007/s11067-011-9155-x}.

\bibitem[\citeproctext]{ref-reilly1929methods}
Reilly, William J. 1929. {``Methods for the Study of Retail
Relationships.''} 2944.

\bibitem[\citeproctext]{ref-reyesAccessibility2014}
Reyes, M., A. Paez, and C. Morency. 2014. {``Walking Accessibility to
Urban Parks by Children: {A} Case Study of {Montreal}.''}
\emph{Landscape and Urban Planning} 125 (May): 38--47.
\url{https://doi.org/10.1016/j.landurbplan.2014.02.002}.

\bibitem[\citeproctext]{ref-ribeiro_road_2010}
Ribeiro, A., A. P. Antunes, and A. Páez. 2010. {``Road Accessibility and
Cohesion in Lagging Regions: {Empirical} Evidence from {Portugal} Based
on Spatial Econometric Models.''} \emph{Journal of Transport Geography}
18 (1): 125--32.

\bibitem[\citeproctext]{ref-roblot2021participation}
Roblot, Mathilde, Geneviève Boisjoly, Ciari Francesco, and Trépanier
Martin. 2021. {``Participation in Shared Mobility: An Analysis of the
Influence of Walking and Public Transport Accessibility to Vehicles on
Carsharing Membership in Montreal, Canada.''} \emph{Transportation
Research Record} 2675 (12): 1160--71.

\bibitem[\citeproctext]{ref-rojas_accessibility_2016}
Rojas, C., A. Paez, O. Barbosa, and J. Carrasco. 2016. {``Accessibility
to Urban Green Spaces in {Chilean} Cities Using Adaptive Thresholds.''}
\emph{Journal of Transport Geography} 57 (December): 227--40.
\url{https://doi.org/10.1016/j.jtrangeo.2016.10.012}.

\bibitem[\citeproctext]{ref-romanillosAccessibilitySchoolsSpatial2018}
Romanillos, G., and J. C. Garcia-Palomares. 2018. {``Accessibility to
{Schools}: {Spatial} and {Social} {Imbalances} and the {Impact} of
{Population} {Density} in {Four} {European} {Cities}.''} \emph{Journal
of Urban Planning and Development} 144 (4).
\url{https://doi.org/10.1061/(asce)up.1943-5444.0000491}.

\bibitem[\citeproctext]{ref-sahebgharani2019computing}
Sahebgharani, Alireza, Mahmoud Mohammadi, and Hossein Haghshenas. 2019.
{``Computing Spatiotemporal Accessibility to Urban Opportunities: A
Reliable Space-Time Prism Approach in Uncertain Urban Networks.''}
\emph{Computation} 7 (3): 51.

\bibitem[\citeproctext]{ref-SchuurmanMeasuring2010}
Schuurman, N., M. Berube, and V. A. Crooks. 2010. {``Measuring Potential
Spatial Access to Primary Health Care Physicians Using a Modified
Gravity Model.''} \emph{Canadian Geographer-Geographe Canadien} 54 (1):
29--45.

\bibitem[\citeproctext]{ref-seniorGravityModellingEntropy1979}
Senior, Martyn L. 1979. {``From Gravity Modelling to Entropy Maximizing:
A Pedagogic Guide.''} \emph{Progress in Human Geography} 3 (2):
175--210. \url{https://doi.org/10.1177/030913257900300218}.

\bibitem[\citeproctext]{ref-sharifiasl2023incorporating}
Sharifiasl, Seyedsoheil, Subham Kharel, and Qisheng Pan. 2023.
{``Incorporating Job Competition and Matching to an Indicator-Based
Transportation Equity Analysis for Auto and Transit in Dallas-Fort Worth
Area.''} \emph{Transportation Research Record} 2677 (12): 240--54.

\bibitem[\citeproctext]{ref-shen1998}
Shen, Q. 1998b. {``Location Characteristics of Inner-City Neighborhoods
and Employment Accessibility of Low-Wage Workers.''} \emph{Environment
and Planning B: Planning and Design} 25 (3): 345--65.
\url{https://doi.org/10.1068/b250345}.

\bibitem[\citeproctext]{ref-shenLocationCharacteristicsInnercity1998}
---------. 1998a. {``Location Characteristics of Inner-City
Neighborhoods and Employment Accessibility of Low-Wage Workers.''}
\emph{Environment and Planning B: Planning and Design} 25 (3): 345--65.
\url{https://doi.org/10.1068/b250345}.

\bibitem[\citeproctext]{ref-shen2019segregation}
Shen, Yao. 2019. {``Segregation Through Space: A Scope of the Flow-Based
Spatial Interaction Model.''} \emph{Journal of Transport Geography} 76:
10--23.

\bibitem[\citeproctext]{ref-silvaAccessibilityInstrumentsPlanning2017}
Silva, Cecília, Luca Bertolini, Marco Te Brömmelstroet, Dimitris
Milakis, and Enrica Papa. 2017. {``Accessibility Instruments in Planning
Practice: Bridging the Implementation Gap.''} \emph{Transport Policy} 53
(January): 135--45. \url{https://doi.org/10.1016/j.tranpol.2016.09.006}.

\bibitem[\citeproctext]{ref-soukhovIntroducingSpatialAvailability2023}
Soukhov, Anastasia, Antonio Paez, Christopher D. Higgins, and Moataz
Mohamed. 2023. {``Introducing Spatial Availability, a Singly-Constrained
Measure of Competitive Accessibility {\textbar} {PLOS} {ONE}.''}
\emph{{PLOS} {ONE}}, 1--30.
https://doi.org/\href{https://\%20doi.org/10.1371/journal.pone.0278468}{https://
doi.org/10.1371/journal.pone.0278468}.

\bibitem[\citeproctext]{ref-soukhovMultimodalSpatialAvailability2024}
Soukhov, Anastasia, Javier Tarriño-Ortiz, Julio A. Soria-Lara, and
Antonio Páez. 2024. {``Multimodal Spatial Availability: A
Singly-Constrained Measure of Accessibility Considering Multiple
Modes.''} \emph{{PLOS} {ONE}} 19 (2): e0299077.
\url{https://doi.org/10.1371/journal.pone.0299077}.

\bibitem[\citeproctext]{ref-stewartPrinciples1947}
Stewart, John Q. 1947. {``Suggested {Principles} of ''{Social}
{Physics}''.''} \emph{Science} 106 (2748): 179--80.

\bibitem[\citeproctext]{ref-stewartDemographicGravitationEvidence1948}
---------. 1948. {``Demographic Gravitation: Evidence and
Applications.''} \emph{Sociometry} 11 (1): 31--58.
\url{https://doi.org/10.2307/2785468}.

\bibitem[\citeproctext]{ref-su2023untangling}
Su, Rongxiang, and Konstadinos Goulias. 2023. {``Untangling the
Relationships Among Residential Environment, Destination Choice, and
Daily Walk Accessibility.''} \emph{Journal of Transport Geography} 109:
103595.

\bibitem[\citeproctext]{ref-suel2024measuring}
Suel, Esra, Claude Lynch, Maria Wood, Tom Murat, Gerard Casey, and Adam
Dennett. 2024. {``Measuring Transport-Associated Urban Inequalities:
Where Are We and Where Do We Go from Here?''} \emph{Transport Reviews}
44 (6): 1235--57.

\bibitem[\citeproctext]{ref-tao_investigating_2020}
Tao, ZL, JP Zhou, XB Lin, H Chao, and GC Li. 2020. {``Investigating the
Impacts of Public Transport on Job Accessibility in {Shenzhen}, {China}:
A Multi-Modal Approach.''} \emph{LAND USE POLICY} 99 (December).
\url{https://doi.org/10.1016/j.landusepol.2020.105025}.

\bibitem[\citeproctext]{ref-tong2015transportation}
Tong, Lu, Xuesong Zhou, and Harvey J Miller. 2015. {``Transportation
Network Design for Maximizing Space--Time Accessibility.''}
\emph{Transportation Research Part B: Methodological} 81: 555--76.

\bibitem[\citeproctext]{ref-turk2019socio}
Türk, Umut. 2019. {``Socio-Economic Determinants of Student Mobility and
Inequality of Access to Higher Education in Italy.''} \emph{Networks and
Spatial Economics} 19 (1): 125--48.

\bibitem[\citeproctext]{ref-vanweeAccessible2016}
van Wee, Bert. 2016. {``Accessible Accessibility Research Challenges.''}
\emph{Journal of Transport Geography} 51 (February): 9--16.
https://doi.org/\url{https://doi.org/10.1016/j.jtrangeo.2015.10.018}.

\bibitem[\citeproctext]{ref-vickermanAccessibilityAttractionPotential1974}
Vickerman, R W. 1974. {``Accessibility, Attraction, and Potential: A
Review of Some Concepts and Their Use in Determining Mobility.''}
\emph{Environment and Planning A} 6: 675--91.
\url{https://doi.org/10.1068/a060675}.

\bibitem[\citeproctext]{ref-vickermanAccessibility1999}
Vickerman, R., K. Spiekermann, and M. Wegener. 1999. {``Accessibility
and Economic Development in {Europe}.''} \emph{Regional Studies} 33 (1):
1--15.

\bibitem[\citeproctext]{ref-wachs_physical_1973}
Wachs, Martin, and T.Gordon Kumagai. 1973. {``Physical Accessibility as
a Social Indicator.''} \emph{Socio-Economic Planning Sciences} 7 (5):
437--56. \url{https://doi.org/10.1016/0038-0121(73)90041-4}.

\bibitem[\citeproctext]{ref-wan2012three}
Wan, N., B. Zou, and T. Sternberg. 2012. {``A Three-Step Floating
Catchment Area Method for Analyzing Spatial Access to Health
Services.''} Journal Article. \emph{International Journal of
Geographical Information Science} 26 (6): 1073--89.
\url{https://doi.org/10.1080/13658816.2011.624987}.

\bibitem[\citeproctext]{ref-wang_inverted_2018}
Wang, F. H. 2018. {``Inverted {Two}-{Step} {Floating} {Catchment} {Area}
{Method} for {Measuring} {Facility} {Crowdedness}.''} \emph{Professional
Geographer} 70 (2): 251--60.
\url{https://doi.org/10.1080/00330124.2017.1365308}.

\bibitem[\citeproctext]{ref-wang_2sfca_2021}
Wang, Fahui. 2021. {``From {2SFCA} to {i2SFCA}: Integration, Derivation
and Validation.''} \emph{International Journal of Geographical
Information Science} 35 (3): 628--38.
\url{https://doi.org/10.1080/13658816.2020.1811868}.

\bibitem[\citeproctext]{ref-weibullNumericalMeasurementAccessibility1980}
Weibull, J W. 1980. {``On the Numerical Measurement of Accessibility.''}
\emph{Environment and Planning A: Economy and Space} 12 (1): 53--67.
\url{https://doi.org/10.1068/a120053}.

\bibitem[\citeproctext]{ref-weinerUrbanTransportationPlanning2016}
Weiner, Edward. 2016. \emph{Urban Transportation Planning in the United
States}. Springer Cham: Springer International Publishing.
\url{https://doi.org/10.1007/978-3-319-39975-1}.

\bibitem[\citeproctext]{ref-whoMedicalDoctors102025}
WHO. 2025. {``Medical Doctors (Per 10 000 Population).''} 2025.
\url{https://www.who.int/data/gho/data/indicators/indicator-details/GHO/medical-doctors-(per-10-000-population)}.

\bibitem[\citeproctext]{ref-williams_disparities_2014}
Williams, S., and F. H. Wang. 2014. {``Disparities in Accessibility of
Public High Schools, in Metropolitan {Baton} {Rouge}, {Louisiana}
1990-2010.''} \emph{Urban Geography} 35 (7): 1066--83.
\url{https://doi.org/10.1080/02723638.2014.936668}.

\bibitem[\citeproctext]{ref-willigers2007accessibility}
Willigers, Jasper, Han Floor, and Bert van Wee. 2007. {``Accessibility
Indicators for Location Choices of Offices: An Application to the
Intraregional Distributive Effects of High-Speed Rail in the
Netherlands.''} \emph{Environment and Planning A} 39 (9): 2086--2898.

\bibitem[\citeproctext]{ref-wilsonSTATISTICALTHEORYSPATIAL1967}
Wilson, A. G. 1967. {``A {STATISTICAL} {THEORY} {OF} {SPATIAL}
{DISTRIBUTION} {MODELS}.''} \emph{Transportation Research} 1: 253--69.
\url{https://journals-scholarsportal-info.libaccess.lib.mcmaster.ca/pdf/00411647/v01i0003/253_astosdm.xml_en}.

\bibitem[\citeproctext]{ref-wilson1971}
---------. 1971. {``A Family of Spatial Interaction Models, and
Associated Developments.''} \emph{Environment and Planning A: Economy
and Space} 3 (1): 1--32. \url{https://doi.org/10.1068/a030001}.

\bibitem[\citeproctext]{ref-wuUnifyingAccess2020}
Wu, Hao, and David Levinson. 2020. {``Unifying Access.''}
\emph{Transportation Research Part D: Transport and Environment} 83
(June): 102355. \url{https://doi.org/10.1016/j.trd.2020.102355}.

\bibitem[\citeproctext]{ref-ye_spatial_2018}
Ye, Changdong, Yushu Zhu, Jiangxue Yang, and Qiang Fu. 2018. {``Spatial
Equity in Accessing Secondary Education: {Evidence} from a Gravity-Based
Model: {Spatial} Equity in Accessing Secondary Education.''} \emph{The
Canadian Geographer / Le Géographe Canadien} 62 (4): 452--69.
\url{https://doi.org/10.1111/cag.12482}.

\bibitem[\citeproctext]{ref-zipfDeterminantsCirculationInformation1946}
Zipf, George Kingsley. 1946a. {``Some Determinants of the Circulation of
Information.''} \emph{The American Journal of Psychology} 59 (3):
401--21. \url{https://doi.org/10.2307/1417611}.

\bibitem[\citeproctext]{ref-zipfHypothesisIntercityMovement1946}
---------. 1946b. {``The \emph{p} \(_{\textrm{1}}\) \emph{p}
\(_{\textrm{2}}\) / \emph{d} Hypothesis: On the Intercity Movement of
Persons''} 11 (6): 677--86.

\bibitem[\citeproctext]{ref-zipfHypothesisCaseRailway1946}
---------. 1946c. {``The \emph{p} \(_{\textrm{1}}\) \emph{p}
\(_{\textrm{2}}\) / \emph{d} Hypothesis: The Case of Railway Express.''}
\emph{The Journal of Psychology} 22 (1): 3--8.
\url{https://doi.org/10.1080/00223980.1946.9917292}.

\end{CSLReferences}

\end{document}

### 2025-05-17 paper/manuscript.qmd
rafa's email and affiliation

@@
   - name: Rafael H. M. Pereira
-    email: rafa.pereira.br@gmail.com
+    email: rafael.pereira@ipea.gov.br
     affiliations:
-        - name: Institute for Applied Economic Research
+        - name: Institute for Applied Economic Research - Ipea
           city: Brasília, Brazil

### 2025-05-17 paper/manuscript.qmd
abstract

@@
 We demonstrate that widely used mobility and accessibility models, such as gravity-based accessibility and spatial interaction models, share common theoretical roots.
-From this foundation, we define a family of accessibility measures within the principles of spatial interaction.
-In demonstrating these developments, three contributions emerge: A) the definition and linkages between four members of the family: specifically, the 'unconstrained' measure (i.e., Hansen-type measure), the 'total constrained' measure (i.e., a constrained version of the Hansen-type measure), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach (2SFCA)), and the 'doubly constrained' measure representing realized interactions or 'access', effectively equal to the doubly constrained spatial interaction model in formulation.
-B) How to formulaically specify the total-, singly- and doubly- constrained accessibility measures.
-C) The interpretability advantages of the family, as these constrained accessibility measures yield values in units of the number of potential "opportunities for spatial interaction" or "population for spatial interaction" for each zone and zonal flow.
-By clarifying the concept of 'potential' in accessibility, demonstrating theoretical and formulaic linkages across popular accessibility and spatial interaction models, and reintroducing units, this work contributes a family of measures that unlock a clearer, more interpretable, and cohesive foundation for accessibility analysis.
+From this foundation, this paper offers three contributions: (A) we introduce a family of accessibility measures within the principles of spatial interaction, and (B) formally define four members of the family, namely the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized interactions or 'access', effectively equal to the doubly constrained spatial interaction model; and (C) we demonstrate the interpretability advantages of the family, as these constrained accessibility measures yield values in units of the number of potential "opportunities for spatial interaction" or "population for spatial interaction" for each zone and zonal flow.
+The family of accessibility measures proposed here clarifies the concept of 'potential' in accessibility, demonstrates theoretical and formulaic linkages across popular accessibility and spatial interaction models, and reintroduces measurement units into accessibility measures.
+By doing so, we believe this family of measures can unlock a clearer, more interpretable, and cohesive foundation for accessibility analysis.
 

### 2025-05-17 paper/manuscript.qmd
intro part 1

@@
 
-Accessibility can be seen to represent an evolution of mobility; accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'.
+Accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'.
 Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling].
@@
 For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
-Further, difficulties in the interpretability and communicability of outputs has also been noticed by @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
-
-Furthermore, the adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
+Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
+
+The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
 At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
@@
 
-More specifically, this work seeks to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
-In fact, accessibility can be specified using spatial interaction principles and a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971, can be defined using balancing factors that constrains values based on known information about the land-use.
-
-In this way, this work aims to offer three contributions: (1) introduce and (2) define a family of _constrained_ accessibility measures, namely the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure.
-These measures are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential, or what Reilly called 'potential trade territories' [@reilly1929methods] and further developed in other examples [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+Here, we seek to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
+In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971 that can be defined using balancing factors that constrains values based on known information about the land-use.
+
+This work aims to offer three contributions: (1) it introduces a family of accessibility measures within the principles of spatial interaction; (2) it formally defines three _constrained_ accessibility measures by reintroducing Wilson-analogous balancing factors.
+These measures are the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure, which are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
 These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
-(3) This work also aims to demonstrate the intuitive value of restoring units to the resulting raw accessibility values.
-Each zonal and zonal flow value from a constrained accessibility measure is in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'.
-This is in contrast to conventional accessibility measures that yield values in units of 'opportunities weighted by some representation of travel friction'.
+(3) This work also demonstrates that the introduction of balancing factors makes accessibility measures easier to interpret and to communicate by restoring the measurement units to the resulting raw accessibility values.
+Each zonal and zonal flow value from a constrained accessibility measure is always in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'.
+This is in contrast to conventional accessibility measures, particularly gravity based measures, that yield values in units of 'opportunities weighted by some representation of travel friction'.
 
@@
 Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure.
-This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974.
+This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974, and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods].
 This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].

### 2025-05-17 paper/manuscript.qmd
intro part 2

@@
 We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
-Moreover, without constraints: we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with.
+Moreover, without constraints we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with.
 Without a clear sense of measurement units, comparability between accessibility measures, across cities, and transport modes, may also become compromised.
@@
 We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
-Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based analogous constraints.
+Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on analogous constraints.
 We illustrate members of this family of constrained accessibility measures with a simple numerical example.

### 2025-05-18 paper/manuscript.qmd
sections 2, 3, 4

@@
 
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility.
+Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility polularized by @hansen1959 in the late 1950s.
 
@@
 
-$S_{i}$ in @eq-accessibility is a measure of the accessibility of site $i$.
+$S_{i}$ in @eq-accessibility is a measure of the accessibility from site $i$.
 This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost).
@@
 This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
-Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained implicit or forgotten, with no notable developments to explicitly acknowledge or determine it.
-
-The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which has fundamental implications for the interpretation, communication and comparability of accessibility analysis.
+Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly acknowledge or determine it.
+
+The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which undermines the interpretation, communication and comparability of accessibility analysis.
 Those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
@@
 This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
-Hansen-style accessibility, therefore, is better thought of as an ordinal measure of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].
+The Hansen-style accessibility estimates found in the literature, therefore, are better thought of as ordinal measures of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].
 

### 2025-05-18 paper/manuscript.qmd
section 5

@@
 In his groundbreaking study [@wilson1971], Wilson defined a general spatial interaction model as follows.
-While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling spatial interaction:
+While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling observed spatial interaction:
 $$
@@
 
-Traditionally, development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
+Traditionally, the development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
 But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
@@
 
-In the third case in the family of spatial interaction models is the production-attraction constrained model.
+The third case in the family of spatial interaction models is the production-attraction constrained model.
 In this case, both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
@@
 Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
-It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been ignored in almost all subsequent accessibility research.
+It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been overlooked in almost all subsequent accessibility research.
 Why was this key element of spatial interaction models potentially ignored in accessibility research?

### 2025-05-18 paper/manuscript.qmd
section 6

@@
 In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S.
-Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars (equivalent to over $600 billion in 2023) [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
 Spatial interaction modelling was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
@@
 The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971.
+In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971 since the year 2000.
 
@@
 
-On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable.
+On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon.
 Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model that may remedy the issue (e.g., the *competition destination* model).
+Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
 Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
 Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+
+<!-- rafa 1: no need to keep all references in each citation block -->
+
+<!-- rafa 1: no need to keep all references in each citation block -->
 
@@
 Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction.
+While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept (e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999), accessibility is a distinct concept from spatial interaction.
 This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
@@
 However, as outlined in preceding sections, we argue the resulting accessibility values are plagued by similar interpretability issues.
-Of the literature in this stream (i.e., accessibility works inspired by @hansen1959 and who cite @wilson1971), only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as method for maintaining constraints on opportunities within the context of competitive accessibility.
+
+Within this literature stream on accessibility inspired by @hansen1959 but that also acknowledges @wilson1971 ideas, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as method for maintaining constraints on opportunities within the context of competitive accessibility.
 These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
@@
 These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization), as will be demonstrated in this study.
-They also do not discuss other constrained cases.
+They also do not discuss other constrained cases that we will address in the next section.
 
@@
 
-<!-- rafa: do we need to keep equations 31 and 32. AS: I think yes. -->
-
 This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel, congruent with Wilson's framework.

### 2025-05-18 paper/manuscript.qmd
section 6

@@
 
-<!-- rafa 1: no need to keep all references in each citation block -->
+<!-- rafa 2: no need to keep all references in each citation block -->
 
@@
 
-In this section, we introduce a family of accessibility measures, using Wilson's spatial interaction modelling framework.
+In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework.
 
@@
 
+<!-- rafa 3: check sentence 'however, $S_i$ itself is not unclear' -->
+
+<!-- rafa 4: check broken cross reference \@ref(grav-to-today) -->
+
 Understanding $S_i$ as the inverse of $A_i$ offers a potential interpretation for the balancing factor, however, $S_i$ itself is not unclear.
 Also, the relationship in @eq-Ai-as-accessibility appears to be misaligned.
-In the preceding Section \@ref(grav-to-today), we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
+In the preceding Section \@ref(grav-to-today), we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$) to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
 While the inverse of Wilson's $A_i$ is pretty much identical to Hansen's accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
@@
 $$ {#eq-access-01}
+
+<!-- rafa 4: there is no $X$ in equation 23-->
 
@@
 
-To demonstrate the anatomy of $V_{ij}$ and $M_{ij}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility demonstrates this work's accessibility analytical device for a simple 3 zone system.
+<!-- rafa 4: there is no $X$ in equation 23-->
+
+To demonstrate the anatomy of $V_{ij}$ and $M_{ij}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility illustrates the accessibility analytical framework we propose using a simple 3 zone system.
 Each measure's most disaggregate value is $X_{ij}$, potential for spatial interaction from $i$ to $j$.
@@
 #| out-width: 70%
-#| fig-cap: "The family of accessibility measures analytical device: labelling and associating ij flows, zonal weights, the single marginals, and the total marginal."
+#| fig-cap: "The family of accessibility measures analytical framework: labelling and associating ij flows, zonal weights, the single marginals, and the total marginal."
 
@@
 
-As an additional overview, we define four distinct members of the constrained accessibility measure family, all delineated based on their constant $k$; either balancing factors $K^T$, $B_j$, $A_i$ depending on their case and variant:
+As an additional overview, we define four distinct members of the constrained accessibility measure family, all delineated based on their constant $k$, which takes the form of either balancing factors $K^T$, $B_j$, $A_i$ depending on the indicator:
 

### 2025-05-18 paper/manuscript.qmd
table 8 and conclusion

@@
 In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework.
+At the end of this section, Table 8 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.
 
@@
 
-<!-- rafa 4: there is no $X$ in equation 23-->
+<!-- rafa 5: there is no $X$ in equation 23-->
 
@@
 Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
-
-<!-- rafa 4: there is no $X$ in equation 23-->
 
@@
 In this case, no constraints are introduced to ensure that values of marginals are preserved.
+
+<!-- rafa 6: glad to see that the paragraphs below use Figure 2 as an explanation aid. The text/explanation you sent me in your last email was quite helpful -->
 
@@
 It can be understood as predicting a value of 'access', and not accessibility.
+
+::: {=latex}
+\input{summary-family-access-measures-table.tex}
+:::
+
+Table 8: Summary of constrained accessibility measure types and interpretations
+
+<!-- {#tbl:summary-family-of-accessibility-measures} -->
 
@@
 
-And as a summary: each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations in Table 8.
-<!-- @tbl:summary-family-of-accessibility-measures.-->
-
-::: {=latex}
-\input{summary-family-access-measures-table.tex}
-:::
-
-Table 8: Summary of constrained accessibility measure types and interpretations
-
-<!-- {#tbl:summary-family-of-accessibility-measures} -->
+<!-- rafa 7: final paragraph with take aways -->
+666666
+
+In summary, building on @wilson1971's foundational work, this paper proposed a unified framework for accessibility that is able to account for competition.
+By reintroducing Wilson's proportionality constant, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.
+This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.
 

### 2025-05-20 paper/manuscript.qmd
typo
incorporated rafa's latest comments -- preprint draft ready.

@@
 title: "A family of accessibility measures derived from spatial interaction principles"
-subtitle: "Preprint submitted to peer-review in May 2025"
+subtitle: "Submitted preprint for peer-review (May 2025)"
 author:
@@
 
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility polularized by @hansen1959 in the late 1950s.
+Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility popularized by @hansen1959 in the late 1950s.
 
@@
 
-As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled.
-However, we can discern from a reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
-The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected.
-
-On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon.
+As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus, visualised in pink in @fig-docs-per-year).
+After reading the works, we can also discern that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
+The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected, and at times, improperly conflated.
+
+On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cite both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models.
+These works treat separate spatial interaction and accessibility as a seperate but related phenomenon.
 Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
@@
 
-<!-- rafa 1: no need to keep all references in each citation block -->
-
-<!-- rafa 2: no need to keep all references in each citation block -->
-
-On the other hand, we discern that the stream of literature focused on accessibility inspired by @hansen1959 cite @wilson1971 for three general reasons.
-Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-While accessibility is an expression of *potential for* spatial interaction and @wilson1971 touches on the concept (e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999), accessibility is a distinct concept from spatial interaction.
-This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-However, as outlined in preceding sections, we argue the resulting accessibility values are plagued by similar interpretability issues.
-
-Within this literature stream on accessibility inspired by @hansen1959 but that also acknowledges @wilson1971 ideas, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as method for maintaining constraints on opportunities within the context of competitive accessibility.
+<!-- rafa 1: no need to keep all references in each citation block. AS: I'll keep them for now. -->
+
+<!-- rafa 2: no need to keep all references in each citation block. AS: I'll keep them for now. -->
+
+On the other hand, there is another subset of literature that cite both @hansen1959 and @wilson1971 that is accessibility-focused.
+We categorise their citation of @wilson1971 for three general reasons.
+Firstly, a group of these works cite @wilson1971 as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+These works do not necessarily comment on spatial interaction explicitly.
+Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in @wilson1971 with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the @wilson1971 paper briefly touches on the concept.
+However, in some of this literature, @hansen1959 and @wilson1971 are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature.
+Thirdly, there is a group of accessibility-focused works that interpret @hansen1959's model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
+This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+As outlined in preceding sections, we argue interpreting the singly- or doubly- constrained spatial interaction model's balancing factor as accessibility yields output values that are similarly plagued by interpretability issues.
+
+Lastly, as an extension of the third reason within this group, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 use @wilson1971's balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility.
 These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
@@
 However, these balancing factors were deduced intuitively.
-These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization), as will be demonstrated in this study.
-They also do not discuss other constrained cases that we will address in the next section.
+These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
+This equivalence is only discovered in hindsight, as will be demonstrated in this study.
+These two works also do not discuss other constrained cases that will be addressed in the next section.
 
@@
 
-<!-- rafa 3: check sentence 'however, $S_i$ itself is not unclear' -->
-
-<!-- rafa 4: check broken cross reference \@ref(grav-to-today) -->
-
-Understanding $S_i$ as the inverse of $A_i$ offers a potential interpretation for the balancing factor, however, $S_i$ itself is not unclear.
-Also, the relationship in @eq-Ai-as-accessibility appears to be misaligned.
-In the preceding Section \@ref(grav-to-today), we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$) to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$.
-While the inverse of Wilson's $A_i$ is pretty much identical to Hansen's accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).
-
-Hence, we propose a more general definition of accessibility: the partial sum of the potential for spatial interaction.
-Once we bring back Wilson's proportionality constant into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
+Understanding $S_i$ as the inverse of Wilson's $A_i$ does not uncover any new meaning for $S_i$ itself.
+Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints.
+Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @q-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).
+
+Hence, we propose stepping back to introduce a revised definition of accessibility: the _constrained_ potential for spatial interaction.
+Once we bring back Wilson's proportionality constant $k$ into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
 
@@
 
-<!-- rafa 5: there is no $X$ in equation 23-->
-
-\noindent where $V_{ij}$ is the potential for interaction from $i$ to $j$.
-$X$ denotes a sub-indices for the member of the constrained accessibility value, either unconstrained $0$, total constrained $T$, singly constrained $S$ or doubly constrained $D$.
-The accessibility from origin $i$ is defined as:
+\noindent where $V_{ij}$ is the potential for interaction from $i$ to $j$. The accessibility from origin $i$ can then be summarised as as a partial sum of the potential at $i$:
 
@@
 
-To demonstrate the anatomy of $V_{ij}$ and $M_{ij}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility illustrates the accessibility analytical framework we propose using a simple 3 zone system.
+To detail the anatomy of $V_{ij}$ and $M_{ji}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility illustrates the accessibility analytical framework we propose using a simple 3 zone system.
 Each measure's most disaggregate value is $X_{ij}$, potential for spatial interaction from $i$ to $j$.
-The $X$ is a stand in for the $ij$ values of all the cases and their variants that will be described (i.e., $V_{ij}^0, M_{ij}^0, V_{ij}^T, M_{ij}^T, V_{ij}^S, M_{ij}^S$ and $V_{ij}^D, M_{ij}^D$).
+The $X$ is a stand in for the $ij$ values of all the cases and their variants that will be described (i.e., $V_{ij}^0, M_{ji}^0, V_{ij}^T, M_{ji}^T, V_{ij}^S, M_{ji}^S$ and $V_{ij}^D, M_{ji}^D$).
 The single marginals represent the origin-side and destination-side weights of the zones.
@@
 In this case, no constraints are introduced to ensure that values of marginals are preserved.
-
-<!-- rafa 6: glad to see that the paragraphs below use Figure 2 as an explanation aid. The text/explanation you sent me in your last email was quite helpful -->
 
@@
 **The doubly constrained case**.
-It is constrained simultaneously by population (origin masses) and opportunities (destination masses), so each $V_{ij}^D$ (equivalent to $M_{ij}^D$) is expressed in units of 'population-opportunity capacity' that is accessible between $i$ to $j$.
+It is constrained simultaneously by population (origin masses) and opportunities (destination masses), so each $V_{ij}^D$ (equivalent to $M_{ji}^D$) is expressed in units of 'population-opportunity capacity' that is accessible between $i$ to $j$.
 This case requires the number of opportunities and population to match, e.g., the analyst must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
-By using this one-to-one matching data and the double constraints (i.e., $A_i$ and $B_j$ at once, ensuring both the double constraint maintaints both marginals in @fig-analytical-device-conc-accessibility), this case restricts the _potential_ to spatially interact completely.
+By using this one-to-one matching data and the double constraints (i.e., $A_i$ and $B_j$ at once, ensuring both the double constraint maintains both marginals in @fig-analytical-device-conc-accessibility), this case restricts the _potential_ to spatially interact completely.
 In other words, the doubly constrained accessibility values reflect the number of predicted interactions between the opportunities and population, effectively, this case is equivalent formulaically as Wilson's doubly constrained spatial interaction model (i.e., _attraction-production constrained_).
 It can be understood as predicting a value of 'access', and not accessibility.
+
+And as a summary: each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
 
@@
 :::
-
-Table 8: Summary of constrained accessibility measure types and interpretations
-
-<!-- {#tbl:summary-family-of-accessibility-measures} -->
-
-In the next sections, we will assign values to the simple 3x3 numeric example shown in the analytical device (@fig-analytical-device-conc-accessibility).
+Table 1: Summary of constrained accessibility measure types and interpretations
 
@@
 $$
-M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
+M^T_{ji} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ji}^0
 $$ {#eq-total-constrained-market}
@@
 $$
-M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
+M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
 $$
@@
 $$
-\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
+\sum_j M^T_{j} = \sum_i\sum_j M^T_{ji} = O
 $$ {#eq-total-constraint-market}
@@
 $$
-\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
+\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ji}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
 $$ {#eq-total-population-balancing-factor}
@@
 $$
-\sum_j M^S_{ij} =  O_i \\
+\sum_j M^S_{ji} =  O_i \\
 $$ {#eq-population-constraint}
@@
 
-As well, the single (population) constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O$) is maintained.
+As well, the single (population) constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O$) is maintained.
 
@@
   impedance_matrix,  # the resulting travel cost impedance matrix
-  error_threshold = 0.001,  # Error threshold for stopping condition
-  improvement_threshold = 1e-6  # Improvement threshold for stopping condition
+  error_threshold = 0.001,  # error threshold for stopping condition
+  improvement_threshold = 1e-6  # improvement threshold for stopping condition
 ) {
@@
 However, if what interests readers is the "potential for interaction" in the case population and opportunities to match one-to-one, reframing the investigation may be needed.
-In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ij}^D$).
+In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ji}^D$).
 In this sense, the doubly constrained case can be thought of as an estimate of _realized_ accessibility or access: reflecting spatial interaction.
@@
 In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
-
-<!-- rafa 7: final paragraph with take aways -->
-666666
 

### 2025-05-20 paper/manuscript.tex
incorporated rafa's latest comments -- preprint draft ready.

@@
 \makeatother
-\subtitle{Preprint submitted to peer-review in May 2025}
+\subtitle{Submitted preprint for peer-review (May 2025)}
 \author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
@@
 models, such as gravity-based accessibility and spatial interaction
-models, share common theoretical roots. From this foundation, we define
-a family of accessibility measures within the principles of spatial
-interaction. In demonstrating these developments, three contributions
-emerge: A) the definition and linkages between four members of the
-family: specifically, the `unconstrained' measure (i.e., Hansen-type
-measure), the `total constrained' measure (i.e., a constrained version
-of the Hansen-type measure), the `singly constrained' measure (i.e.,
-related to the popular two step floating catchment approach (2SFCA)),
-and the `doubly constrained' measure representing realized interactions
-or `access', effectively equal to the doubly constrained spatial
-interaction model in formulation. B) How to formulaically specify the
-total-, singly- and doubly- constrained accessibility measures. C) The
-interpretability advantages of the family, as these constrained
-accessibility measures yield values in units of the number of potential
-``opportunities for spatial interaction'' or ``population for spatial
-interaction'' for each zone and zonal flow. By clarifying the concept of
-`potential' in accessibility, demonstrating theoretical and formulaic
+models, share common theoretical roots. From this foundation, this paper
+offers three contributions: (A) we introduce a family of accessibility
+measures within the principles of spatial interaction, and (B) formally
+define four members of the family, namely the `unconstrained' measure
+(i.e., Hansen-type accessibility), the `total constrained' measure
+(i.e., a constrained version of the Hansen-type accessibility), the
+`singly constrained' measure (i.e., related to the popular two step
+floating catchment approach - 2SFCA), and the `doubly constrained'
+measure representing realized interactions or `access', effectively
+equal to the doubly constrained spatial interaction model; and (C) we
+demonstrate the interpretability advantages of the family, as these
+constrained accessibility measures yield values in units of the number
+of potential ``opportunities for spatial interaction'' or ``population
+for spatial interaction'' for each zone and zonal flow. The family of
+accessibility measures proposed here clarifies the concept of
+`potential' in accessibility, demonstrates theoretical and formulaic
 linkages across popular accessibility and spatial interaction models,
-and reintroducing units, this work contributes a family of measures that
-unlock a clearer, more interpretable, and cohesive foundation for
-accessibility analysis.
+and reintroduces measurement units into accessibility measures. By doing
+so, we believe this family of measures can unlock a clearer, more
+interpretable, and cohesive foundation for accessibility analysis.
 \end{abstract}
@@
 
-Accessibility can be seen to represent an evolution of mobility;
-accessibility may be defined as the ``potential of opportunities for
+Accessibility may be defined as the ``potential of opportunities for
 {[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
@@
 interpretability and communicability of outputs has also been noticed by
-Karst T. Geurs and van Wee (2004), van Wee (2016), and Ferreira and Papa
-(2020).
-
-Furthermore, the adoption of accessibility in planning practice is not
-necessarily made easier when potential adopters have to contend with a
-plethora of definitions, each seemingly more sophisticated but less
-intuitive than the last (Kapatsila et al. 2023). At a high level, Karst
-T. Geurs and van Wee (2004) identify four families of accessibility
-measures: infrastructure-, place-, person-, and utility-based. Of the
-place-based family, which is the focus of this paper, the menu has grown
-to include gravity-based accessibility (e.g., Hansen 1959; Pirie 1979),
-cumulative opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye
-et al. 2018), modified gravity (e.g., Schuurman, Berube, and Crooks
-2010), 2-Step Floating Catchment Areas (e.g., Luo and Wang 2003),
-Enhanced 2-Step Floating Catchment Areas (e.g., Luo and Qi 2009),
-3-Stage Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012),
-Modified 2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse
-2-Step Floating Catchment Areas (e.g., F. Wang 2021), and n-steps
-Floating Catchment Areas (Liang, Yan, and Yan 2024). How is a
-practitioner to choose among this myriad options? What differences in
-accessibility scores should matter, and how should they be communicated?
-{[}see van Wee (2016); p.~14{]}.
-
-More specifically, this work seeks to address the breath of place-based
-accessibility's definitions and lack of interpretability by
-demonstrating that mobility-oriented models--such as the commonly used
-gravity-based accessibility (e.g., Hansen 1959) and the spatial
-interaction model (e.g., Wilson 1971)-- are rooted in the same spatial
-interaction modeling foundation. In fact, accessibility can be specified
-using spatial interaction principles and a \emph{family of measures},
-akin to the family of spatial interaction models introduced in Wilson
-(1971), can be defined using balancing factors that constrains values
-based on known information about the land-use.
-
-In this way, this work aims to offer three contributions: (1) introduce
-and (2) define a family of \emph{constrained} accessibility measures,
-namely the total constrained accessibility measure, the singly
-constrained accessibility measure, and the doubly constrained measure.
-These measures are explicitly connected to popular measures such as the
-Hansen-type accessibility (Hansen 1959), the popular competition
-approach of the 2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a;
-Luo and Wang 2003), and the concept of market potential, or what Reilly
-called `potential trade territories' (Reilly 1929) and further developed
-in other examples (C. D. Harris 1954; R. W. Vickerman 1974). These
-Wilson-analogous balancing factors introduced are also defined in order
-of increasing restrictiveness to shed light on the role of
-\emph{potential} in accessibility and access. (3) This work also aims to
-demonstrate the intuitive value of restoring units to the resulting raw
-accessibility values. Each zonal and zonal flow value from a constrained
-accessibility measure is in interpretable units, namely the number of
+many authores, including Karst T. Geurs and van Wee (2004), van Wee
+(2016), and Ferreira and Papa (2020).
+
+The adoption of accessibility in planning practice is not necessarily
+made easier when potential adopters have to contend with a plethora of
+definitions, each seemingly more sophisticated but less intuitive than
+the last (Kapatsila et al. 2023). At a high level, Karst T. Geurs and
+van Wee (2004) identify four families of accessibility measures:
+infrastructure-, place-, person-, and utility-based. Of the place-based
+family, which is the focus of this paper, the menu has grown to include
+gravity-based accessibility (e.g., Hansen 1959; Pirie 1979), cumulative
+opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye et al.
+2018), modified gravity (e.g., Schuurman, Berube, and Crooks 2010),
+2-Step Floating Catchment Areas (e.g., Luo and Wang 2003), Enhanced
+2-Step Floating Catchment Areas (e.g., Luo and Qi 2009), 3-Stage
+Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012), Modified
+2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse 2-Step
+Floating Catchment Areas (e.g., F. Wang 2021), and n-steps Floating
+Catchment Areas (Liang, Yan, and Yan 2024). How is a practitioner to
+choose among this myriad options? What differences in accessibility
+scores should matter, and how should they be communicated? {[}see van
+Wee (2016); p.~14{]}.
+
+Here, we seek to address the breath of place-based accessibility's
+definitions and lack of interpretability by demonstrating that
+mobility-oriented models--such as the commonly used gravity-based
+accessibility (e.g., Hansen 1959) and the spatial interaction model
+(e.g., Wilson 1971)-- are rooted in the same spatial interaction
+modeling foundation. In fact, we propose that accessibility can be
+specified using spatial interaction principles as a \emph{family of
+measures}, akin to the family of spatial interaction models introduced
+in Wilson (1971) that can be defined using balancing factors that
+constrains values based on known information about the land-use.
+
+This work aims to offer three contributions: (1) it introduces a family
+of accessibility measures within the principles of spatial interaction;
+(2) it formally defines three \emph{constrained} accessibility measures
+by reintroducing Wilson-analogous balancing factors. These measures are
+the total constrained accessibility measure, the singly constrained
+accessibility measure, and the doubly constrained measure, which are
+explicitly connected to popular measures such as the Hansen-type
+accessibility (Hansen 1959), the popular competition approach of the
+2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a; Luo and Wang
+2003), and the concept of market potential (C. D. Harris 1954; R. W.
+Vickerman 1974). These Wilson-analogous balancing factors introduced are
+also defined in order of increasing restrictiveness to shed light on the
+role of \emph{potential} in accessibility and access. (3) This work also
+demonstrates that the introduction of balancing factors makes
+accessibility measures easier to interpret and to communicate by
+restoring the measurement units to the resulting raw accessibility
+values. Each zonal and zonal flow value from a constrained accessibility
+measure is always in interpretable units, namely the number of
 `opportunities for spatial interaction' or `population for spatial
-interaction'. This is in contrast to conventional accessibility measures
-that yield values in units of `opportunities weighted by some
-representation of travel friction'.
+interaction'. This is in contrast to conventional accessibility
+measures, particularly gravity based measures, that yield values in
+units of `opportunities weighted by some representation of travel
+friction'.
 
@@
 statements like `higher-than' or `lower-than' (E. J. Miller 2018).
-Moreover, without constraints: we lose track of what accessibility
+Moreover, without constraints we lose track of what accessibility
 measures, that is, the number of opportunities that can be spatially
@@
 interaction models, and use them to derive a family of accessibility
-measures based analogous constraints. We illustrate members of this
+measures based on analogous constraints. We illustrate members of this
 family of constrained accessibility measures with a simple numerical
@@
 re-organization of terms, is formally equivalent to our modern
-definition of accessibility.
+definition of accessibility popularized by Hansen (1959) in the late
+1950s.
 
@@
 \(S_{i}\) in Equation~\ref{eq-accessibility} is a measure of the
-accessibility of site \(i\). This is a function of \(O_j\) (the mass of
-opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or travel
-time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a parameter
-that modulates the friction of cost). Today, Hansen is frequently cited
-as the father of modern accessibility analysis (e.g., Reggiani and
-Martín 2011), and Hansen-type accessibility is commonly referred to as
-the gravity-based accessibility measure.
+accessibility from site \(i\). This is a function of \(O_j\) (the mass
+of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
+travel time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a
+parameter that modulates the friction of cost). Today, Hansen is
+frequently cited as the father of modern accessibility analysis (e.g.,
+Reggiani and Martín 2011), and Hansen-type accessibility is commonly
+referred to as the gravity-based accessibility measure.
 
@@
 analysis has been widely used in numerous disciplines but, to our
-knowledge, the proportionality constant has remained implicit or
-forgotten, with no notable developments to explicitly acknowledge or
-determine it.
+knowledge, the proportionality constant has remained forgotten, with no
+notable developments to explicitly acknowledge or determine it.
 
 The omission of this constant generates a fundamental problem for the
-measurement unit of accessibility estimates, which has fundamental
-implications for the interpretation, communication and comparability of
-accessibility analysis. Those reading Hansen (1959) must recall that
-Stewart (1948) had set the proportionality constant \(G\) to 1, with a
-note that ``\(G\) {[}was{]} left for future determination: a suitable
-choice of other units can reduce it to unity'' {[}p.~34{]}. In practice,
-the persistent omission of the constant in accessibility analyses means
-that \(G\) continues to be implicitly set to \(1\), even when the
-fundamental relationship in accessibility is proportionality (e.g.,
+measurement unit of accessibility estimates, which undermines the
+interpretation, communication and comparability of accessibility
+analysis. Those reading Hansen (1959) must recall that Stewart (1948)
+had set the proportionality constant \(G\) to 1, with a note that
+``\(G\) {[}was{]} left for future determination: a suitable choice of
+other units can reduce it to unity'' {[}p.~34{]}. In practice, the
+persistent omission of the constant in accessibility analyses means that
+\(G\) continues to be implicitly set to \(1\), even when the fundamental
+relationship in accessibility is proportionality (e.g.,
 \(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality (for
@@
 decay functions, and renders their results difficult to understand and
-communicate. Hansen-style accessibility, therefore, is better thought of
-as an ordinal measure of potential that can only be interpreted in terms
-of higher and lower accessibility, but which has not palpable meaning
-(E. J. Miller 2018).
+communicate. The Hansen-style accessibility estimates found in the
+literature, therefore, are better thought of as ordinal measures of
+potential that can only be interpreted in terms of higher and lower
+accessibility, but which has not palpable meaning (E. J. Miller 2018).
 
@@
 accessibility was characterised as an associated concept of `potential',
-the primary focus was on modelling spatial interaction:
+the primary focus was on modelling observed spatial interaction:
 \begin{equation}\phantomsection\label{eq-phys-gravity-model}{
@@
 
-Traditionally, development of the spatial interaction model put an
+Traditionally, the development of the spatial interaction model put an
 emphasis on the interpretability of the results (Kirby 1970; Wilson
@@
 
-In the third case in the family of spatial interaction models is the
+The third case in the family of spatial interaction models is the
 production-attraction constrained model. In this case, both
@@
 yield variations of proportionality constants, including the one that
-eluded Stewart (1948) and that has been ignored in almost all subsequent
-accessibility research. Why was this key element of spatial interaction
-models potentially ignored in accessibility research? In the next
-section we aim to address this question.
+eluded Stewart (1948) and that has been overlooked in almost all
+subsequent accessibility research. Why was this key element of spatial
+interaction models potentially ignored in accessibility research? In the
+next section we aim to address this question.
 
@@
 the creation of the U.S. Interstate Highway System, with a budget that
-ultimately exceeded one hundred billion dollars (Weiner 2016; MDOT
-2007). Spatial interaction modelling was incorporated into institutional
-modelling practices meant to ``predict and provide'', i.e., predict
-travel demand and supply transportation infrastructure (Kovatch, Zames,
-et al. 1971; Weiner 2016). Accessibility, at the time, did not quite
-have that power, as it did not quantify trips, but rather something
-somewhat more elusive: it predicts the less tangible ``potential'' for
-spatial interaction with opportunities. In this way, where spatial
-interaction modelling became a key element of transportation planning
-practice, accessibility remained a somewhat more academic pursuit, and
-the two streams of literature only rarely connected.
+ultimately exceeded one hundred billion dollars (equivalent to over
+\$600 billion in 2023) (Weiner 2016; MDOT 2007). Spatial interaction
+modelling was incorporated into institutional modelling practices meant
+to ``predict and provide'', i.e., predict travel demand and supply
+transportation infrastructure (Kovatch, Zames, et al. 1971; Weiner
+2016). Accessibility, at the time, did not quite have that power, as it
+did not quantify trips, but rather something somewhat more elusive: it
+predicts the less tangible ``potential'' for spatial interaction with
+opportunities. In this way, where spatial interaction modelling became a
+key element of transportation planning practice, accessibility remained
+a somewhat more academic pursuit, and the two streams of literature only
+rarely connected.
 
@@
 particular, we notice the remarkable growth in the number of papers that
-cite Hansen (1959) compared to those that cite Wilson (1971).
+cite Hansen (1959) compared to those that cite Wilson (1971) since the
+year 2000.
 
 As noted, literature that cites both Wilson (1971) and Hansen (1959) are
-sparse (only 3.6\% of the corpus) but highly coupled. However, we can
-discern from a reading of these works that they too are divergent, with
-one stream focused on developing accessibility (i.e., \emph{potential
-for} spatial interaction) and another on spatial interaction. The
-focuses of these divergent streams contribute this paper's broader
-hypothesis that the concepts of accessibility and spatial interaction
-have remained largely disconnected.
+sparse (only 3.6\% of the corpus, visualised in pink in
+Figure~\ref{fig-docs-per-year}). After reading the works, we can also
+discern that they too are divergent, with one stream focused on
+developing accessibility (i.e., \emph{potential for} spatial
+interaction) and another on spatial interaction. The focuses of these
+divergent streams contribute this paper's broader hypothesis that the
+concepts of accessibility and spatial interaction have remained largely
+disconnected, and at times, improperly conflated.
 
 On one hand, the stream of literature focused on spatial interaction
-models inspired by Wilson (1971) and which cites both Hansen (1959) and
+models inspired by Wilson (1971) and which cite both Hansen (1959) and
 Wilson (1971), tends to contribute to understanding how accessibility is
-interpreted and incorporated in spatial interaction models, treating it
-as a separate (but related) phenomenon or variable. Specifically, some
-early works interpret the spatial interaction model's balancing factors
+interpreted and incorporated in spatial interaction models. These works
+treat separate spatial interaction and accessibility as a seperate but
+related phenomenon. Specifically, some early works interpret the spatial
+interaction model's balancing factors
 (Equation~\ref{eq-production-constrained-balancing-factor} or
@@
 Fotheringham (1985) who demonstrates how the spatial interaction model
-may insufficiently explain spatial patterns and suggest that explicitly
-defining destinations' accessibility as a variable within the model that
-may remedy the issue (e.g., the \emph{competition destination} model).
-Other works used both Hansen (1959) and Wilson (1971)'s framework in
-conjunction, such as in defining location-allocation problems in
-operations research (G. Leonardi 1978; Beaumont 1981), estimating trip
-flows (or some other spatial interaction flows) alongside accessibility
-(e.g., Clarke, Eyre, and Guy 2002; Grengs 2004; Türk 2019), or
-considering accessibility within spatial interaction models, in line
-with A. S. Fotheringham (1985)'s demonstration (e.g., Beckers et al.
-2022). Other works departed from Hansen (1959)'s definition and aligned
-with spatial interaction in different ways, such as using micro-economic
-consumer behaviour concepts to express potential for spatial interaction
-(Morris, Dumble, and Wigan 1979; Giorgio Leonardi and Tadei 1984).
-
-On the other hand, we discern that the stream of literature focused on
-accessibility inspired by Hansen (1959) cite Wilson (1971) for three
-general reasons. Firstly, as attribution for using using different
-context-dependent travel cost functions (Weibull 1980; S. L. Handy and
-Niemeier 1997; Kwan 1998; Q. Shen 1998a; Ashiru, Polak, and Noland 2003;
-Rau and Vega 2012; Pan 2013; Margarida Condeço Melhorado et al. 2016;
-Caschili, De Montis, and Trogu 2015; Grengs 2015; Pan, Jin, and Liu
-2020; Chia and Lee 2020; Roblot et al. 2021; Sharifiasl, Kharel, and Pan
-2023; Kharel, Sharifiasl, and Pan 2024). Secondly, associating spatial
-interaction with accessibility's potential for spatial interaction
-(e.g., Giuliano et al. 2010; Grengs et al. 2010; Grengs 2010, 2012;
-Levine et al. 2012; Levinson and Huang 2012; Tong, Zhou, and Miller
-2015; X. Liu and Zhou 2015; He et al. 2017; Wu and Levinson 2020; Ng et
-al. 2022; Naqavi et al. 2023; Suel et al. 2024). While accessibility is
-an expression of \emph{potential for} spatial interaction and Wilson
-(1971) touches on the concept H. J. Miller (1999), it is distinct
-concept from spatial interaction. This distinction, however, is often
-unclear, as Hansen (1959) and Wilson (1971) are frequently co-cited as
-both being `gravity models' (e.g., S. Liu and Zhu 2004; Dai, Wan, and
-Gai 2017; Y. Shen 2019; Chia and Lee 2020). Thirdly, interpretation of
-Hansen (1959)'s model as an inverse balancing factor (R. W. Vickerman
-1974), especially in competitive opportunity or population contexts
-(Karst and Van Eck 2003; Karst T. Geurs, van Wee, and Rietveld 2006;
-Willigers, Floor, and van Wee 2007; El-Geneidy and Levinson 2011; Curtis
-and Scheurer 2010; Manaugh and El-Geneidy 2012; Chen and Silva 2013;
-Alonso et al. 2014; Albacete et al. 2017; Sahebgharani, Mohammadi, and
-Haghshenas 2019; Mayaud et al. 2019; Allen and Farber 2020; Levinson and
-Wu 2020; Marwal and Silva 2022; Su and Goulias 2023). However, as
-outlined in preceding sections, we argue the resulting accessibility
-values are plagued by similar interpretability issues. Of the literature
-in this stream (i.e., accessibility works inspired by Hansen (1959) and
-who cite Wilson (1971)), only the works of Soukhov et al. (2023) and
-Soukhov et al. (2024) cite Wilson (1971)'s use of balancing factors as
-method for maintaining constraints on opportunities within the context
-of competitive accessibility. These works introduce the balancing
-factors as a mechanisms to ensure that opportunities at each destination
-are proportionally allocated to each zone (based on the proportion of
-population seeking opportunities and the relative travel impedance).
-This is to ensure that each zonal accessibility value is the sum of this
-proportional allocation from each destination, and that all zonal values
-ultimately sum to the total number of opportunities in the region.
-However, these balancing factors were deduced intuitively. These works
-do not explicitly state that the mathematical formulation of the
-equations are effectively equivalent to Wilson's singly constrained
-model (derived from entropy maximization), as will be demonstrated in
-this study. They also do not discuss other constrained cases.
+may insufficiently explain spatial patterns, and suggests that
+explicitly defining destinations' accessibility as a variable within the
+model may remedy the issue (e.g., the \emph{competition destination}
+model). Other works used both Hansen (1959) and Wilson (1971)'s
+framework in conjunction, such as in defining location-allocation
+problems in operations research (G. Leonardi 1978; Beaumont 1981),
+estimating trip flows (or some other spatial interaction flows)
+alongside accessibility (e.g., Clarke, Eyre, and Guy 2002; Grengs 2004;
+Türk 2019), or considering accessibility within spatial interaction
+models, in line with A. S. Fotheringham (1985)'s demonstration (e.g.,
+Beckers et al. 2022). Other works departed from Hansen (1959)'s
+definition and aligned with spatial interaction in different ways, such
+as using micro-economic consumer behaviour concepts to express potential
+for spatial interaction (Morris, Dumble, and Wigan 1979; Giorgio
+Leonardi and Tadei 1984).
+
+On the other hand, there is another subset of literature that cite both
+Hansen (1959) and Wilson (1971) that is accessibility-focused. We
+categorise their citation of Wilson (1971) for three general reasons.
+Firstly, a group of these works cite Wilson (1971) as attribution for
+using context-dependent travel cost functions (e.g., Weibull 1980; S. L.
+Handy and Niemeier 1997; Kwan 1998; Q. Shen 1998a; Ashiru, Polak, and
+Noland 2003; Rau and Vega 2012; Pan 2013; Margarida Condeço Melhorado et
+al. 2016; Caschili, De Montis, and Trogu 2015; Grengs 2015; Pan, Jin,
+and Liu 2020; Chia and Lee 2020; Roblot et al. 2021; Sharifiasl, Kharel,
+and Pan 2023; Kharel, Sharifiasl, and Pan 2024). These works do not
+necessarily comment on spatial interaction explicitly. Secondly, another
+group of this accessibility-focused ``both'' citing literature
+\emph{does} associate spatial interaction as defined in Wilson (1971)
+with accessibility's potential for spatial interaction more explicitly
+(e.g., H. J. Miller 1999; Giuliano et al. 2010; Grengs et al. 2010;
+Grengs 2010, 2012; Levine et al. 2012; Levinson and Huang 2012; Tong,
+Zhou, and Miller 2015; X. Liu and Zhou 2015; He et al. 2017; Wu and
+Levinson 2020; Ng et al. 2022; Naqavi et al. 2023; Suel et al. 2024). We
+agree accessibility and spatial interaction are related topics:
+accessibility is an expression of its \emph{potential} and the Wilson
+(1971) paper briefly touches on the concept. However, in some of this
+literature, Hansen (1959) and Wilson (1971) are co-cited as both being
+`gravity models' (e.g., S. Liu and Zhu 2004; Dai, Wan, and Gai 2017; Y.
+Shen 2019; Chia and Lee 2020), perhaps revealing the murkiness of the
+distinction between spatial interaction and the \emph{potential for}
+spatial interaction in the literature. Thirdly, there is a group of
+accessibility-focused works that interpret Hansen (1959)'s model as the
+singly- or doubly- constrained spatial interaction model's inverse
+balancing factor (e.g., R. W. Vickerman 1974). This group cities the
+earlier spatial interaction works that make this connection and is
+especially prominent in the investigation of competitive accessibility
+topics (e.g., Karst and Van Eck 2003; Karst T. Geurs, van Wee, and
+Rietveld 2006; Willigers, Floor, and van Wee 2007; El-Geneidy and
+Levinson 2011; Curtis and Scheurer 2010; Manaugh and El-Geneidy 2012;
+Chen and Silva 2013; Alonso et al. 2014; Albacete et al. 2017;
+Sahebgharani, Mohammadi, and Haghshenas 2019; Mayaud et al. 2019; Allen
+and Farber 2020; Levinson and Wu 2020; Marwal and Silva 2022; Su and
+Goulias 2023). As outlined in preceding sections, we argue interpreting
+the singly- or doubly- constrained spatial interaction model's balancing
+factor as accessibility yields output values that are similarly plagued
+by interpretability issues.
+
+Lastly, as an extension of the third reason within this group, only the
+works of Soukhov et al. (2023) and Soukhov et al. (2024) use Wilson
+(1971)'s balancing factors as a method for maintaining constraints on
+opportunities within the context of competitive accessibility. These
+works introduce the balancing factors as a mechanisms to ensure that
+opportunities at each destination are proportionally allocated to each
+zone (based on the proportion of population seeking opportunities and
+the relative travel impedance). This is to ensure that each zonal
+accessibility value is the sum of this proportional allocation from each
+destination, and that all zonal values ultimately sum to the total
+number of opportunities in the region. However, these balancing factors
+were deduced intuitively. These works do not explicitly state that the
+mathematical formulation of the equations are effectively equivalent to
+Wilson's singly constrained model (derived from entropy maximization).
+This equivalence is only discovered in hindsight, as will be
+demonstrated in this study. These two works also do not discuss other
+constrained cases that will be addressed in the next section.
 
@@
 
-In this section, we introduce a family of accessibility measures, using
-Wilson's spatial interaction modelling framework.
+In this section, we introduce a family of accessibility measures
+inspired by Wilson's spatial interaction modelling framework. At the end
+of this section, Table 8 presents a summary, where each member of the
+family of accessibility measure is explained in plain language,
+alongside their balancing factor, proportional allocation factor and
+mathematical equation and value interpretations.
 
@@
 
-Understanding \(S_i\) as the inverse of \(A_i\) offers a potential
-interpretation for the balancing factor, however, \(S_i\) itself is not
-unclear. Also, the relationship in Equation~\ref{eq-Ai-as-accessibility}
-appears to be misaligned. In the preceding Section @ref(grav-to-today),
-we noted that Hansen, and Stewart before him, defined accessibility as a
-partial sum of the demographic force \(F\), essentially going from the
-demographic force (\(F = G\frac{M_1\cdot M_2}{d_{12}^2}\) to the
-system-wide population potential (after losing \(G\))
-\(V_i = \sum_j \frac{M_j}{d_{ij}}\). While the inverse of Wilson's
-\(A_i\) is pretty much identical to Hansen's accessibility, its
-functional role in Wilson's general model
+Understanding \(S_i\) as the inverse of Wilson's \(A_i\) does not
+uncover any new meaning for \(S_i\) itself. Indeed, mathematically it is
+true, \(A_i\)'s role in Wilson's general model
 \(T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})\) is that of a balancing
-factor \(k\) (maintaining proportionality).
-
-Hence, we propose a more general definition of accessibility: the
-partial sum of the potential for spatial interaction. Once we bring back
-Wilson's proportionality constant into the picture, we can define the
-\emph{potential for spatial interaction} between two locations \(i\) and
-\(j\) is as follows:
+factor \(k\) i.e., keeping units balanced and proportionality based on
+constraints. Understanding \(S_i\) itself as a balancing factor is not
+wholly helpful as Hansen and (Stewart before him) defined accessibility
+as a partial sum of the demographic force \(F\) or the system-wide
+population potential of opportunities for spatial interaction (i.e.,
+(\textbf{q-stewart-population-potential-sum?}) representing
+\(V_i = \sum_j \frac{M_j}{d_{ij}}\) after losing \(G\)).
+
+Hence, we propose stepping back to introduce a revised definition of
+accessibility: the \emph{constrained} potential for spatial interaction.
+Once we bring back Wilson's proportionality constant \(k\) into the
+picture, we can define the \emph{potential for spatial interaction}
+between two locations \(i\) and \(j\) is as follows:
 
@@
 \noindent where \(V_{ij}\) is the potential for interaction from \(i\)
-to \(j\). \(X\) denotes a sub-indices for the member of the constrained
-accessibility value, either unconstrained \(0\), total constrained
-\(T\), singly constrained \(S\) or doubly constrained \(D\). The
-accessibility from origin \(i\) is defined as:
+to \(j\). The accessibility from origin \(i\) can then be summarised as
+as a partial sum of the potential at \(i\):
 
@@
 
-To demonstrate the anatomy of \(V_{ij}\) and \(M_{ij}\) along with the
+To detail the anatomy of \(V_{ij}\) and \(M_{ji}\) along with the
 partial sums of \(V_i\) and \(M_j\), Figure
-Figure~\ref{fig-analytical-device-conc-accessibility} demonstrates this
-work's accessibility analytical device for a simple 3 zone system. Each
-measure's most disaggregate value is \(X_{ij}\), potential for spatial
-interaction from \(i\) to \(j\). The \(X\) is a stand in for the \(ij\)
-values of all the cases and their variants that will be described (i.e.,
-\(V_{ij}^0, M_{ij}^0, V_{ij}^T, M_{ij}^T, V_{ij}^S, M_{ij}^S\) and
-\(V_{ij}^D, M_{ij}^D\)). The single marginals represent the origin-side
+Figure~\ref{fig-analytical-device-conc-accessibility} illustrates the
+accessibility analytical framework we propose using a simple 3 zone
+system. Each measure's most disaggregate value is \(X_{ij}\), potential
+for spatial interaction from \(i\) to \(j\). The \(X\) is a stand in for
+the \(ij\) values of all the cases and their variants that will be
+described (i.e.,
+\(V_{ij}^0, M_{ji}^0, V_{ij}^T, M_{ji}^T, V_{ij}^S, M_{ji}^S\) and
+\(V_{ij}^D, M_{ji}^D\)). The single marginals represent the origin-side
 and destination-side weights of the zones. The total marginal represent
@@
 \caption{\label{fig-analytical-device-conc-accessibility}The family of
-accessibility measures analytical device: labelling and associating ij
-flows, zonal weights, the single marginals, and the total marginal.}
+accessibility measures analytical framework: labelling and associating
+ij flows, zonal weights, the single marginals, and the total marginal.}
 
@@
 constrained accessibility measure family, all delineated based on their
-constant \(k\); either balancing factors \(K^T\), \(B_j\), \(A_i\)
-depending on their case and variant:
+constant \(k\), which takes the form of either balancing factors
+\(K^T\), \(B_j\), \(A_i\) depending on the indicator:
 
@@
   by population (origin masses) and opportunities (destination masses),
-  so each \(V_{ij}^D\) (equivalent to \(M_{ij}^D\)) is expressed in
+  so each \(V_{ij}^D\) (equivalent to \(M_{ji}^D\)) is expressed in
   units of `population-opportunity capacity' that is accessible between
@@
   constraints (i.e., \(A_i\) and \(B_j\) at once, ensuring both the
-  double constraint maintaints both marginals in
+  double constraint maintains both marginals in
   Figure~\ref{fig-analytical-device-conc-accessibility}), this case
@@
 
-In the next sections, we will assign values to the simple 3x3 numeric
-example shown in the analytical device
-(Figure~\ref{fig-analytical-device-conc-accessibility}).
+And as a summary: each member of the family of accessibility measure is
+named, explained in plain language, alongside their balancing factor(s),
+proportional allocation factor(s), and mathematical equation and value
+interpretations in Table 1.
+
+\input{summary-family-access-measures-table.tex}
+
+Table 1: Summary of constrained accessibility measure types and
+interpretations
 
@@
 the concept of market potential (i.e., potential users) as proposed in
-C. D. Harris (1954) and R. W. Vickerman (1974). This unconstrained form
-of market potential \(M_j^0\) (Equation~\ref{eq-unconstrained-market}),
-effectively the \(i\) to \(j\) transpose of \(V^0_{ij}\), has been used
-in recent research to express the potentially accessible population
-(i.e., users) as a result of regional transportation infrastructure
-investment projects (e.g., Gutiérrez 2001; Holl 2007; Condeço-Melhorado
-and Christidis 2018). Put another way, market potential can also be
-thought of as a form of \emph{passive accessibility}, indicating the
-number of people that can reach each destination. However, like
-\(V_{ij}^0\), issues of unit interpretability arise in \(M_j^0\)'s
-unconstrained form. To address this, the constrained variant, the total
-constrained accessible population measure \(M^T_{j}\), is introduced. To
-formulate this variant, the total balancing factor \(K^T\) is applied to
-the mass of the \emph{population} at \(i\) (\(W_i^{(1)}\)) instead of
-the opportunities at \(j\).
+C. D. Harris (1954) and R. W. Vickerman (1974), and which Reilly earlier
+referred to as `potential trade territories' (Reilly 1929). This
+unconstrained form of market potential \(M_j^0\)
+(Equation~\ref{eq-unconstrained-market}), effectively the \(i\) to \(j\)
+transpose of \(V^0_{ij}\), has been used in recent research to express
+the potentially accessible population (i.e., users) as a result of
+regional transportation infrastructure investment projects (e.g.,
+Gutiérrez 2001; Holl 2007; Condeço-Melhorado and Christidis 2018). Put
+another way, market potential can also be thought of as a form of
+\emph{passive accessibility}, indicating the number of people that can
+reach each destination. However, like \(V_{ij}^0\), issues of unit
+interpretability arise in \(M_j^0\)'s unconstrained form. To address
+this, the constrained variant, the total constrained accessible
+population measure \(M^T_{j}\), is introduced. To formulate this
+variant, the total balancing factor \(K^T\) is applied to the mass of
+the \emph{population} at \(i\) (\(W_i^{(1)}\)) instead of the
+opportunities at \(j\).
 
 \begin{equation}\phantomsection\label{eq-total-constrained-market}{
-M^T_{ij} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ij}^0
+M^T_{ji} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ji}^0
 }\end{equation}
@@
 \[
-M^T_j = \sum_i M^T_{ij} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
+M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
 \]
@@
 \begin{equation}\phantomsection\label{eq-total-constraint-market}{
-\sum_j M^T_{j} = \sum_i\sum_j M^T_{ij} = O
+\sum_j M^T_{j} = \sum_i\sum_j M^T_{ji} = O
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-total-population-balancing-factor}{
-\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ij}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
+\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ji}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})} 
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-population-constraint}{
-\sum_j M^S_{ij} =  O_i \\
+\sum_j M^S_{ji} =  O_i \\
 }\end{equation}
@@
 Equation~\ref{eq-population-constraint} ensures that the the total
-constraint (e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ij} = O\)) is
+constraint (e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O\)) is
 maintained.
@@
 opportunities are being allocated to an origin from a destination (or in
-a transposed sense for \(M_{ij}^D\)). In this sense, the doubly
+a transposed sense for \(M_{ji}^D\)). In this sense, the doubly
 constrained case can be thought of as an estimate of \emph{realized}
@@
 
-And as a summary: each member of the family of accessibility measure is
-named, explained in plain language, alongside their balancing factor,
-proportional allocation factor and mathematical equation and value
-interpretations in Table 8.
-
-\input{summary-family-access-measures-table.tex}
-
-Table 8: Summary of constrained accessibility measure types and
-interpretations
+In summary, building on Wilson (1971)'s foundational work, this paper
+proposed a unified framework for accessibility that is able to account
+for competition. By reintroducing Wilson's proportionality constant, the
+proposed family of constrained accessibility measures restores
+measurement units to accessibility estimates. This enhancement provides
+a more interpretable, consistent, and theoretically grounded basis for
+accessibility analysis, which could help advance the adoption of
+accessibility-oriented planning.
 

### 2025-05-20 paper/manuscript.qmd
deleted 'graveyard' commented out text in manuscript

@@
 # References
-
-<!-- Of note RE DOUBLE CONSTRAINT: the accessibility case methodogy should be based selected based on the input data and the theory.
-For instance, the doubly constrained accessibility method requires pportunities and population to matched one-to-one both mathematically (i.e., the balancing factors will not converge if they do not match) and theoretically (i.e., each person should seek one opportunity).
-By forcing these assumptions, the resulting accessibility value that is calculated can be interpreted with less 'potential', than in the total and singly- constrained cases using the unmodified example.
-This is because, the unmodified example inherently carries more 'potential', i.e., we only know where people are (who can potentially interact) and we know where there are opportunities (who can be potentially interacted with), but not the capacities of which this interaction happens.
-As mentioned earlier, in accessibility work, the capacities of which population and opportunities interact is not often known: in fact, accessibility research is often interested in capturing potential at a less granular level: the _potential_ for a population (whether the have interacted, or have not interacted) to interact with opportunities - the capacity of this interaction is not often known, and not the focus.
--->
-
-<!-- Using the doubly constrained measure with the unmodified example is both mathematically impossible (as there is no one-to-one match) but also theoretically uncompitable as the capacities at both ends are not known.
-Conversely, the total and singly- constrained cases can be used to calculate values using inputs from the modified example, but, the results will be more impresice.
-If capacities of population and opportunities are both known, and the objective of the exercise is to calculate the number of _capacity_ of the opportunity (or population) that is accessible, than by using the singly constrained case, capacities at either the opportunity or population would be neglected.
-In using the total constrained case, capacities at both would be neglected.
-In this way, if the goal was to calculated acccessibility to opportunity or population capacity, the answer would be only a more imprecise estimation than if the doubly constrained method was used.
--->
-
-<!-- BELOW IS THE TEXT GRAVEYARD! -->
-
-<!-- # APPENDIX (some musings... destined for the commented-out graveyard) -->
-
-<!-- ## Singly constrained case is spatial availability -->
-
-<!-- The destination-constrained case is also spatial availability, see below!
-As follows then, this case is also related to 2SFCA - if the value is divided by per capita (OR $P_i$ from the numerator in $B_{j}$ is removed).
--->
-
-<!-- Spatial availability is formulated like so: -->
-<!-- $$ -->
-<!-- V_{ij}''^{(2)}  = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} -->
-<!-- $$ -->
-<!-- It turns out that the denominators of the Population balancing factor $F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i=1}^K P_{i\in r}^\alpha}$ and Travel cost balancing factor $F^c_{ij} = F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^K f(c_{ij})}$ cancel out, so spatial availability can be expressed as: -->
-<!-- $$ -->
-<!-- V_{ij}''^{(2)} =O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\ -->
-<!-- $$ -->
-
-<!-- Reformatted in the general family of accessibility measure form, it looks like: -->
-<!-- $$ -->
-<!-- V_{ij}''^{(2)} =k O_j f(c_{ij})\\ -->
-<!-- V_{ij}''^{(2)} =\frac{ P_i}{\sum_i P_i f(c_{ij})} O_j f(c_{ij})\\ -->
-<!-- V_{ij}''^{(2)} =  O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}} = O_j \frac{ P_i  f(c_{ij})}{\sum_i P_i f(c_{ij})}\\ -->
-<!-- $$ -->
-
-<!-- ## Checking singly constrained case against spatial availability -->
-<!-- ```{r, eval=FALSE} -->
-
-<!-- sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){ -->
-
-<!--   o_id <- rlang::enquo(o_id) -->
-<!--   d_id <- rlang::enquo(d_id) -->
-<!--   pop <- rlang::enquo(pop) -->
-<!--   opp <- rlang::enquo(opp) -->
-<!--   r <- rlang::enquo(r) -->
-<!--   f <- rlang::enquo(f) -->
-
-<!--   sum_pop <- x %>% -->
-<!--     dplyr::distinct(!!o_id, -->
-<!--                     .keep_all = TRUE) %>% -->
-<!--     dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>% -->
-<!--     dplyr::pull(sum_pop) %>% -->
-<!--     sum() -->
-
-<!--   f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop -->
-
-<!--   sum_impedance <- x %>% -->
-<!--     dplyr::group_by(!!d_id) %>% -->
-<!--     dplyr::summarize(sum_impedance = sum(!!f)) -->
-
-<!--   x <- x %>% -->
-<!--     dplyr::left_join(sum_impedance, -->
-<!--                      by = rlang::as_name(d_id)) -->
-
-<!--   f_c <- dplyr::pull(x, !!f) / x$sum_impedance -->
-
-<!--   x$f_c <- f_c -->
-<!--   x$f_p <- f_p -->
-
-<!--   sum_pa <- x %>% -->
-<!--     dplyr::group_by(!!d_id) %>% -->
-<!--     dplyr::summarize(sum_pa= sum(f_p * f_c)) -->
-
-<!--   x <- x %>% -->
-<!--     dplyr::left_join(sum_pa, -->
-<!--                      by = rlang::as_name(d_id)) -->
-<!--   f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa) -->
-
-<!--   dplyr::pull(x, !!opp) * f_t -->
-<!-- } -->
-<!-- ``` -->
-
-<!-- ```{r, eval=false} -->
-
-<!-- df <- data.frame(i = c(1,1,1,2,2,2,3,3,3),  -->
-<!--                  j= c(1,2,3,1,2,3,1,2,3), -->
-<!--                  f = c(1/C[1,], 1/C[2,], 1/C[3,]), -->
-<!--                  pop = c(P_i[1], P_i[1], P_i[1], P_i[2], P_i[2], P_i[2], P_i[3], P_i[3], P_i[3]), -->
-<!--                  opp = c(O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3], O_j[1],O_j[2],O_j[3]), -->
-<!--                  catch = c(1,1,1,1,1,1,1,1,1)) -->
-
-<!-- calced <- sp_avail(df, -->
-<!--                          o_id = i, -->
-<!--                          d_id = j, -->
-<!--                          pop = pop, -->
-<!--                          opp = opp, -->
-<!--                          r = catch, -->
-<!--                          f = f) -->
-<!-- df$V_ij <- calced -->
-
-<!-- df |> group_by(j) |> summarise(V_j = sum(V_ij)) -->
-<!-- ``` -->
-
-<!-- ## Destination- (or ATTRACTION) constrained is the only singly constrained case... population- (or PRODUCTION) constrained doesn't make sense! -->
-
-<!-- Then the origin-constrained (singly constrained) case $V_{ij}^{(1)}$, no $P_i$ values are needed only $O_j$ values.
-Using this method, V_i values equal P_i values...?
-while O_j values are unconstrained.
-This case may be helpful in an example where the number of opportunities available to an origin is known and fixed.
-This could be a normative minimum threshold, such as 100 opportunities at each origin.
--->
-
-<!-- Returning to our simple example, the origin-constrained singly constrained case would yield the following $A_{i}$ values: -->
-<!-- $$ -->
-<!-- A_{i} = \frac{O_{j}}{\sum_j O_j f(c_{ij})}\\ -->
-<!-- A_{1,1} = \frac{O_{j}}{O_1 f(c_{1,1}) + O_2 f(c_{1,2}) + O_3 f(c_{1,3})}\\ -->
-<!-- ...\\ -->
-<!-- A_{1,1} = 160 *\frac{1}{160/2+437/15+193/5} = 1.083032 \\  -->
-<!-- A_{2,1} = 160 *\frac{1}{160/15+437/2+193/10} =0.6439496 \\ -->
-<!-- A_{3,1} = 160 * \frac{1}{160/5+437/10+193/2} =0.9291521 \\ -->
-<!-- ...\\ -->
-<!-- A_{1,2} = 437 *\frac{1}{160/2+437/15+193/5} =2.958032 \\ -->
-<!-- A_{2,2} = 437 *\frac{1}{160/15+437/2+193/10} =1.7587872 \\ -->
-<!-- A_{3,2} = 437 * \frac{1}{160/5+437/10+193/2} =2.5377468 \\ -->
-<!-- ...\\ -->
-<!-- A_{1,3} = 193 *\frac{1}{160/2+437/15+193/5} = 1.306408 \\ -->
-<!-- A_{2,3} = 193 *\frac{1}{160/15+437/2+193/10} = 0.7767642 \\ -->
-<!-- A_{3,3} = 193 *\frac{1}{160/5+437/10+193/2} = 1.1207898 \\ -->
-<!-- $$ -->
-
-<!-- ```{r, eval=false} -->
-
-<!-- A_ij <- matrix(O_j) %*% (1/(t((matrix(O_j))) %*% (1/C))) |> t() -->
-<!-- A_ij_longer <- A_ij -->
-<!-- rownames(A_ij_longer) <- c("1","2","3")  -->
-<!-- colnames(A_ij_longer) <- c("1","2","3")  -->
-<!-- A_ij_longer  -->
-<!-- ``` -->
-
-<!-- Starting off with this opportunity matrix, namely - at i=1, only a fixed number of opportunities are accessible to this origin. same for i=2, i=3. -->
-<!-- ```{r, eval=false} -->
-
-<!-- (matrix(rep(O_j, 3), 3, 3)) -->
-<!-- ``` -->
-
-<!-- $V_{ij}$ values: -->
-<!-- ```{r, eval=false} -->
-
-<!-- V_orig <- (A_ij) * t(S_non) #S_non -> (1/C)* t(matrix(rep(O_j, 3), 3, 3)) #if you transpose the matrix here, that's another way of generating the answer.  -->
-
-<!-- P_i_orig <- V_orig |> rowSums() -->
-<!-- O_j_orig<- V_orig |> colSums() -->
-<!-- V_TOTAL_orig <- ifelse(sum(P_i_orig)==sum(O_j_orig), sum(V_orig), NA) -->
-<!-- V_orig_longer <- cbind(V_orig,P_i_orig) -->
-<!-- V_orig_longer <- rbind(V_orig_longer, cbind(t(O_j_orig),V_TOTAL_orig)) -->
-
-<!-- rownames(V_orig_longer) <- c("1","2","3", "TOTAL")  -->
-<!-- colnames(V_orig_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- V_orig_longer #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
-<!-- ``` -->
-
-<!--NOTE FOR FUTURE EXPLORATION: Interestingly...
-for the doubly- and singly- constrained case, kappas are very similar *maybe close to identical?* when decay is low.
-This is because distance difference doesn't matter, everyone can reach everything and everything is given to everyone.
-But when decay is high, they are much different: since in the doubly constrained case, enough opps from other zones must be pulled such to match the population -- the kappa is pushed to be constrained to satisfy both the match at the weight at the origin as well as pull proportional from destinations as not to exceed their caps.
-This relationship should be investigated further some other time..-->
-
-<!--
-### Hybrid singly constrained case: 
-NOTE TO SELF: where does E2SFCA and 3SFCA fit?? Maybe here. Maybe BFCA?
-Balanced floating catchment areas (not truly doubly constrained)
--->
-
-<!-- 
-## QUESTIONS:
-- what does k=1 mean? What does assuming k=1, so the units can be kept mean for the model? Is it meaningless?
-- not sure if the singly constrained (PRODUCTION case) makes sense.. It sort of does..? I believe this needs to be figured out before the doubly constrained case. Pi needs to be integrated somehow. 
-- doubly constrained still needs to be solved.
-
-## next steps
-- grocery stores for example; maybe 2 grids?
-- bibliometric analysis 
-- - draft of this end of jan 2025
-
-- (future) intervening opportunities ? radiation model - deep gravity
-- (future) accessibility pyramids 
-- (future) MAUP addressed
-- (future) comparing cities 
--->
-
-<!--NOTE: what about activity-based accessibility? Answer: it is individual utility based, unlike traditional SIM which is origin-destination based. We don't have to comment on it. -->
-
-<!-- Despite the theoretical and observed findings that _both_ mobility and accessibility influence an individual’s capacity to travel [@morrisAccessibilityIndicatorsTransport1979], SIM literature that focuses on Hansen's accessibility is relatively scarce in the literature.
--->
-
-<!--the benefits of this formulation, and the inclusion of the propoartionality constant has been imense for spatial interaction modelling.
-For instance, the identification of the conovlution of spatial autocorrelation and and distance-decay @griffithSpatialStructureSpatial2007 -->
-
-<!-- Wilson in @wilsonSTATISTICALTHEORYSPATIAL1967 was the first to connect accessibility measures to balancing factors-->
-
-<!--
-@wilson1971 (pg. 28) summarises the following checklist for spatial interaction model builders:
-1. Nature of the spatial interaction variable; 
-2. Nature of the mass terms
-3. Measurement of mass terms
-4. Measurement of interaction cost
-5. Choice of cost function
-6. Other terms
-7. Design of spatial system
-8. Calibration
-9. Relation to general model framework
--->
-
-<!-- (cool paper that explores this @kirbyNormalizingFactorsGravity1970 - he "shows that the Ai and Bj balancing factors are the mean values of the inverse of the function e^-N*F(c_ij) calculated by summing the function over i values for Ai factors and over j values for Bj factors." -->
-
-<!--TO DISCUSS:
--- should we use entropy maximisation to actually derive the models? I'm hazy on that.
-  -- relatedly, in the unconstrained case...
-is k=1 or does k just not exist..?
-I assume k=1 and say that the interpretation in reality is confusing.
-a consistent derivation of all cases can help address this.
-  -- constraint 0 or 0.5..? a consistent derivation of all cases should help address this. -->
-
-<!-- REJECTED TEXT: 
-Current transportation planning practice was arguably born in the 1950s; fueled by advancements in computer technologies and catalyzed by the unprecedented Federal-Aid Highway Act of 1956 which authorized the creation of the US Interstate Highway System, costing over a hundred billion in public spending in the following 35 years [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-To plan this ambitious project, transportation planning practice adopted SIM and extended it into four-step transportation demand modelling; where tangible outputs of transportation systems such as _mobility_ (e.g., trips) were calibrated, modeled and applied to "predict and provide" transportation supply and demand [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-However, within the same time period, accessibility remained largely within the academic realm.
-For instance, accessibility measures were used to study a variety of transportation-related dimensions such as residential (or firm) location and city growth [e.g., @kain1962multiple]; where the importance of accessibility versus non-accessibility characteristics (quality of destination, composition of household or inherent individual characteristics) were debated [e.g., @redding1970quality; @chapin1968activity] and experimentation with different cases and travel behaviours emerged showing this nuance [e.g., studies that showcase nuanced preferences in accessibility based on household characteristics or destination type etc.].
-We perceive these debates as a significant point that defined accessibility's outgrowth from the mobility-oriented practical implementation of SIM in transportation planning.
-Accessibility's flexible definition, various mathematical formulations and increasing complexity in its application and interpretation made it a concept more difficult to institutionalize in the moments were planning "embraced" automobility-oriented regional development.
-
-However, in recent years, this embrace has been weakened by the mounting evidence of the burdens that automobility is placing on the planet and communities.
-A renewed wave of criticism levelled at traditional transportation modelling practice's focus on  _mobility_ instead of _accessibility_ have been growing [@handy2020].
-Accessibility, captures the concept of _potential_ interaction, more aligned with the true purpose of transportation systems, and can be used to plan for equitable and sustainable transportation systems in ways using _mobility_ cannot.
-While SIM and mobility-based planning has been institutionalized in transportation practice, accessibility has yet to do so [@papaAccessibilityInstrumentsPlanning2015; @boisjolyInsiderPlannersPerspective2017; @silvaProximitycentredAccessibilityConceptual2023].
--->
-
-<!-- MORE REJECTED TEXT --> 
-```{=html}
-<!-- 
-In summary, papers that cite only @hansen1959 do not focus on SIM and instead focus on _potential_ for interaction while papers that cite only @wilson1971 deal with SIM in the realm of _interaction_.
-Those that cite both @hansen1959 and @wilson1971 (77 journal articles) are of interest in this paper, as they often focus on accessibility's role within the SIM formulation (@eq-phys-gravity-model) but are often still seem to be more SIM than HAM, hence are overviewed as follows.
-
-Often, these SIM/HAM papers seek interpretation of the HAM within the SIM.
-Specifically, many early works in the 70s and 80s have indicated that HAM is represented within the balancing factors of the SIM [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985].
-Recognizing the balancing factors as HAM is a "common sense" approach at including accessibility in the SIM [@morrisAccessibilityIndicatorsTransport1979], though its relationship may be worth further exploring [@battyMethodResiduesUrban1976].
-
-However, some authors that did explore this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how SIM may insufficiently explain spatial patterns and how explicitly defining destination's accessibility, citing HAM as convenvient formulation, within the SIM framework may remedy the issue.
-This re-formulation of the SIM later came to be defined as _competition destination_ model, and is now one of a few techniques to address spatial variation in SIMs [@oshanSpatialStructureDebate2021].
-Other early SIM works incorporating accessibility also went implicitly beyond this understanding that accessibility is neatly considered within the SIM framework.
-Some works went beyond HAM and considered other accessibility expressions of individual choice.
-For instance, using concepts of micro-economic consumer behaviour such as random utility choice models [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-Other papers used the SIM and HAM in conjunction, such as in Operations research that define a location-allocation problem using the singly constrained SIM, and HAM as an objective function subject [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981].
-
-Interestingly from this work in the 70s and 80s, it can be understood that while the balancing factors within the SIM could be read as HAM, they should not be understood as capturing accessibility -eek?
-I don't think I can say this--.
-We also gather that in this moment of growing complexity of SIM (not) capturing accessibility, literature exploring accessibility sprouted in new mediums of budding GIS technologies and the conceptual emergence of space-time dimensions in the 90s onward [e.g., @millerMeasuringSpaceTimeAccessibility1999; @occelli2000revisiting].
-With the possibility for showcasing complex GIS methodologies for varied case studies, this accessibility-based literature became mature enough to branch off on its own without meting in the affairs of SIM.
-
-In papers published within and after the 1990s, @hansen1959 is consistently cited as providing a definition of accessibility or in using the HAM, however the attribution of @wilson1971 is various.
-@wilson1971 is cited alongside accessibility-based literature in three important ways; two of which incorrectly conflate and interpret the SIM and HAM.
-
-Firstly, many of these works attribute @wilson1971 for using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-
-Secondly, in the 2000s and beyond, the concept of spatial interaction and accessibility are conflated throughout the years [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]; indeed, accessibility is an expression of _potential_ spatial interaction, and should not be interpreted the same way as the SIM is commonly used.
-Surprisingly, @hansen1959 and @wilson1971 are further conflated by being co-cited as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
-
-And thirdly, accessibility-based literature cite @wilson1971's balancing factors, and re-iterate how they can be interpreted as the inverse of HAM and hence can capture competition [@vickermanAccessibilityAttractionPotential1974; @karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-However, we argue that this interpretation is fraut, considering @fotheringhamSpatialCompetitionAgglomeration1985's demonstration that SIM does not completely incorporate accessibility.
-However some work does consider competition more scrupulously; these works use SIMs to estimate trip (or some other interaction flows) alongside accessibility indicators [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio] or integrate accessibility within the SIM, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration and most often in context with retailing [@beckers2022incorporating].
-Only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.
-
-Indeed, @hansen1959 offers a relatively straightforward and practical application of accessibility, while @wilson1971's contribution is methodological and dense.
-Accessibility-based literature that has flourished since the 1990s, sprouting from the rise of GIS technologies, the plethora of _bigger_ data and different qualities of opportunities and populations to examine, it has developed in a different direction.
-However, HAM and SIM share similar roots, there are plenty of insights to be shared across these sibling branches.
-A relevant and contemporary connection between HAM and SIM is in the realm of competitive accessibility, and how the SIM can be reconceptualised as measuring _potential for interaction_ and competition can be interpreted through the regional and zonal constraints ensured by the balancing factors.
-How these balancing factors are applied in the context of _interaction_ like trips can be applied to accessibility outputs like the "number of opportunities potentially interacted with".
--->
-```
-
-```{=html}
-<!--
-The approach is based on partition functions from statistical mechanics (the physics of gases), not Newtonian gravitation.
-In short, the approach accepts that there are macro, meso, and micro states.
-In the case of 'trips' being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region.
-Using these assumptions, a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel).
-From this approach, the SIM can be interpreted as the given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level).
-There is then a most probable distribution of trips between zones (meso-states) that satisfies all the macro-state constraints.
-
--->
-```
-
-<!--THE FOLLOWING ARE SOLVED EXAMPLES OF ALL FOUR CASES OF THE SIM -->
-
-<!-- The following are simple numerical examples of each case assuming the following 1) $T_{ij}$ trip ends matrix, 2) $C_{ij}$ travel cost matrix, and 3) an assumed travel impedance function of $f(c_{ij}) = \frac{1}{c_{ij}}$.
-For all cases, we look to estimate each $T_{ij}$ value using these three inputs, we present the known trip matrix and the trip ends below: -->
-
-<!-- ```{r} -->
-
-<!-- known_trip_matrix <- matrix(c(95,23,42, 27,378,45, 38,36,106), ncol = 3, byrow = TRUE) -->
-
-<!-- #unknowns <- matrix(c(rep(NA,9)), ncol=3, byrow=TRUE) -->
-
-<!-- W_i <- known_trip_matrix |> rowSums() -->
-
-<!-- W_j <- known_trip_matrix |> colSums() -->
-
-<!-- T <- ifelse(sum(W_i)==sum(W_j), sum(W_j),NA) -->
-
-<!-- knowns_longer <- cbind(known_trip_matrix,W_i) -->
-
-<!-- knowns_longer <- rbind(knowns_longer,cbind(t(W_j),T)) -->
-
-<!-- rownames(knowns_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- colnames(knowns_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- knowns_longer #|> as.data.frame() |> round() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- And the matrix of cost of travel from $i$ (rows) to $j$ (columns) is as follows: -->
-
-<!-- ```{r} -->
-
-<!-- C <- matrix(c(2, 15, 5, 15, 2, 10, 5, 10, 2), ncol = 3, byrow = TRUE) -->
-
-<!-- C_longer <-C -->
-
-<!-- rownames(C_longer) <- c("1","2","3") -->
-
-<!-- colnames(C_longer) <- c("1","2","3") -->
-
-<!-- C_longer #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- For examples of how each type of case is solved, see many of the cited literutre, like @Wilson and @ortuza. -->
-
-<!-- ## Unconstrained interaction case -->
-
-<!-- $k$ can be understood as being set to 1, so the model looks as follows, with no constraints and the trip matrix is as follows. For example, $T_{1,1} = 160*160*\frac{1}{2}$ -->
-
-<!-- $$ -->
-
-<!-- T_{ij} = W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
-
-<!-- $$ -->
-
-<!-- ```{r} -->
-
-<!-- T_unc <- W_i %*% t(W_j) / C -->
-
-<!-- W_i_unc <- T_unc |> rowSums() -->
-
-<!-- W_j_unc <- T_unc |> colSums() -->
-
-<!-- T_TOTAL_unc <- ifelse(sum(W_i_unc|>round())==sum(W_j_unc|>round()), sum(T_unc), NA) -->
-
-<!-- T_unc_longer <- cbind(T_unc,W_i_unc) -->
-
-<!-- T_unc_longer <- rbind(T_unc_longer, cbind(t(W_j_unc),T_TOTAL_unc)) -->
-
-<!-- rownames(T_unc_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- colnames(T_unc_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- T_unc_longer |> round() #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- As we can surmise, the trip ends do not add up to anything we know -- they do not add up to the known trip ends.
-Instead, they take on the assumptions of all balancing factors $k$ being equal to 1.
-When would this happen?
-It's hard to say...
-as the balancing factors themselves do not have practical meaning; what has meaning are the constraints.
-By satisfying the constraints, the model gains meaning.
-In this way, the resulting unconstrained interaction matrix is simply an expression of the magnitude of interaction of $W_i$ to $W_j$ assuming $f(c_{ij})$.
-In the context of spatial interaction modelling, this model is not entirely useful; it is in units of 'trips' but trips that satisify the macro constraint of $k$ equaling 1, which is practically meaningless.
--->
-
-<!-- ## total constrained interaction case -->
-
-<!-- In the total constrained case, $k$ is replaced with $K$ (@eq-bf-K-gravitymodel satisfying only constraint @eq-constraint0-gravitymodel) so the model looks as follows: -->
-
-<!-- $$ -->
-
-<!-- T_{ij}' = K W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
-
-<!-- $$ -->
-
-<!-- \where Solved for our example, $K$ is equal to the total number of known trips divided by the sum of all the unconstrained trips made: -->
-
-<!-- $$ -->
-
-<!-- K = \frac{T}{\sum_{i}\sum_{j} W_i^{(1)} W_j^{(2)} f(c_{ij})}\\ -->
-
-<!-- K = \frac{T}{\frac{W_1^{(1)}W_1^{(2)}}{c_{1,1}}+\frac{W_2^{(1)}W_1^{(2)}}{c_{2,1}}+\frac{W_3^{(1)}W_1^{(2)}}{c_{3,1}}+\frac{W_1^{(1)}W_2^{(2)}}{c_{1,2}} +...
-\frac{W_3^{(1)}W_3^{(2)}}{c_{3,3}} -->
-
-<!-- }\\ -->
-
-<!-- K = \frac{790}{166443} \\ -->
-
-<!-- K = 0.00474636 -->
-
-<!-- $$ -->
-
-<!-- ```{r} -->
-
-<!-- K <- T/sum(sum(T_unc)) -->
-
-<!-- T_tot <- K * T_unc -->
-
-<!-- W_i_tot <- T_tot |> rowSums() -->
-
-<!-- W_j_tot <- T_tot |> colSums() -->
-
-<!-- T_TOTAL_tot <- ifelse(sum(W_i_tot|>round())==sum(W_j_tot|>round()), sum(T_tot), NA) -->
-
-<!-- T_tot_longer <- cbind(T_tot,W_i_tot) -->
-
-<!-- T_tot_longer <- rbind(T_tot_longer, cbind(t(W_j_tot),T_TOTAL_tot)) -->
-
-<!-- rownames(T_tot_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- colnames(T_tot_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- T_tot_longer |> round() #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- While we can see that the total trips `r T` results, the total trip ends are not equal to the known trip ends as these constraints are not specified and thus not satisfied by the model. -->
-
-<!-- ## Singly constrained -->
-
-<!-- Two versions of this case, either origin-constrained or destination-constrained. -->
-
-<!-- For the origin-constrained model, where $k$ is replaced with $A_i$ (@eq-bf-A-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
-
-<!-- $$ -->
-
-<!-- T_{ij}^{''(1)} = A_i W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
-
-<!-- $$ -->
-
-<!-- \where In our example, $A_i$ is: -->
-
-<!-- $$ -->
-
-<!-- A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}\\ -->
-
-<!-- A_1 = \frac{1}{W_1^{(2)} f(c_{1,1}) + W_2^{(2)} f(c_{1,2}) + W_3^{(2)} f(c_{1,3})}\\ -->
-
-<!-- ...\\ -->
-
-<!-- A_1 = \frac{1}{160/2+437/15+193/5} =0.006768953 \\ -->
-
-<!-- A_2 = \frac{1}{160/15+437/2+193/10} =0.004024685 \\ -->
-
-<!-- A_3 = \frac{1}{160/5+437/10+193/2} =0.005807201 -->
-
-<!-- $$ -->
-
-<!-- ```{r} -->
-
-<!-- A_i <- 1/t((matrix(W_j))) %*% (1/C) -->
-
-<!-- #A_i <- rowSums(known_trip_matrix)/rowSums(T_unc) #our answer with knowns -->
-
-<!-- A_i<- diag(A_i|> as.numeric(), nrow=3, ncol=3) -->
-
-<!-- T_orig <- A_i%*%T_unc -->
-
-<!-- W_i_orig <- T_orig |> rowSums() -->
-
-<!-- W_j_orig <- T_orig |> colSums() -->
-
-<!-- T_TOTAL_orig <- ifelse(sum(W_i_orig)==sum(W_j_orig), sum(T_orig), NA) -->
-
-<!-- T_orig_longer <- cbind(T_orig,W_i_orig) -->
-
-<!-- T_orig_longer <- rbind(T_orig_longer, cbind(t(W_j_orig),T_TOTAL_orig)) -->
-
-<!-- rownames(T_orig_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- colnames(T_orig_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- T_orig_longer |> round() #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- We can see that the $W_i^{(1)}$ values are equal to the trip ends, but not $W_j^{(2)}$ values. -->
-
-<!-- Similarly, the singly constrained case can be flipped to be origin-constrained (@eq-bf-B-gravitymodel satisfying only constraint @eq-constraint1-gravitymodel and implicitly @eq-constraint0-gravitymodel): -->
-
-<!-- $$ -->
-
-<!-- T_{ij}^{''(2)} = B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
-
-<!-- $$ -->
-
-<!-- \where $B_j$ is, and equal to: -->
-
-<!-- $$ -->
-
-<!-- B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\ -->
-
-<!-- $$ -->
-
-<!-- ```{r} -->
-
-<!-- B_j <- 1/(t(matrix(W_i))) %*% (1/C) -->
-
-<!-- B_j <- matrix(B_j) |> t() -->
-
-<!-- B_j_longer <- B_j -->
-
-<!-- colnames(B_j_longer) <- c("1","2","3") -->
-
-<!-- B_j_longer #|> as.data.frame() |> gt() -->
-
-<!-- #B_j <- colSums(known_trip_matrix)/colSums(T_unc) #our answer with knowns -->
-
-<!-- ``` -->
-
-<!-- And trips are: -->
-
-<!-- ```{r} -->
-
-<!-- B_j <- diag(B_j|> as.numeric(), nrow=3, ncol=3) -->
-
-<!-- T_dest <- (B_j)%*%t(T_unc) |> t() -->
-
-<!-- W_i_dest <- T_dest |> rowSums() -->
-
-<!-- W_j_dest <- T_dest |> colSums() -->
-
-<!-- T_TOTAL_dest <- ifelse(sum(W_i_dest)==sum(W_j_dest), sum(T_dest), NA) -->
-
-<!-- T_dest_longer <- cbind(T_dest,W_i_dest) -->
-
-<!-- T_dest_longer <- rbind(T_dest_longer, cbind(t(W_j_dest),T_TOTAL_dest)) -->
-
-<!-- rownames(T_dest_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- colnames(T_dest_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- T_dest_longer |> round() #|> as.data.frame() |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- And as we can observe, the $W_j^{(2)}$ values are equal to the known trip ends, while $W_i^{(1)}$ is not. -->
-
-<!-- ## Doubly constrained case -->
-
-<!-- The parameter $k$ is replaced with both $A_i$ and $B_j$ (@eq-bf-A-gravitymodel and @eq-bf-B-gravitymodel satisfying all three constraints @eq-constraint1-gravitymodel, @eq-constraint2-gravitymodel and implicitly @eq-constraint0-gravitymodel), the model looks as follows: -->
-
-<!-- $$ -->
-
-<!-- T_{ij}^{'''} = A_i B_j W_i^{(1)} W_j^{(2)} f(c_{ij}) -->
-
-<!-- $$ -->
-
-<!-- \where In our example, $A_i$ and $B_j$ is: -->
-
-<!-- $$ -->
-
-<!-- A_i = \frac{1}{\sum_j B_j W_j^{(2)} f(c_{ij})}\\ -->
-
-<!-- B_j = \frac{1}{\sum_i A_i W_i^{(1)} f(c_{ij})} -->
-
-<!-- $$ -->
-
-<!-- ```{r} -->
-
-<!-- doubly_constrained_gravity_model <- function( -->
-
-<!--   #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R -->
-
-<!--  O, # Origin production weights vector -->
-
-<!--  D, # Destination attraction weights vector -->
-
-<!--  impedance_matrix, # the resulting travel cost impedance matrix -->
-
-<!--  error_threshold = 0.001, # Error threshold for stopping condition -->
-
-<!--  improvement_threshold = 1e-6 # Improvement threshold for stopping condition -->
-
-<!-- ) { -->
-
-<!--  # Normalize O and D -->
-
-<!--  sum_O <- sum(O) -->
-
-<!--  sum_D <- sum(D) -->
-
-<!--  cat("Checking production, attraction balancing:\n") -->
-
-<!--  cat("Production: ", sum_O, "\n") -->
-
-<!--  cat("Attraction: ", sum_D, "\n") -->
-
-<!--  if (sum_O != sum_D) { -->
-
-<!--   cat("Productions and attractions do not balance, scaling to match.\n") -->
-
-<!--   if (sum_O < sum_D) { -->
-
-<!--    O <- O * (sum_D / sum_O) -->
-
-<!--   } else { -->
-
-<!--    D <- D * (sum_O / sum_D) -->
-
-<!--   } -->
-
-<!--  } else { -->
-
-<!--   cat("Production, attraction balancing OK.\n") -->
-
-<!--  } -->
-
-<!--  n <- length(O) # Number of i -->
-
-<!--  T <- sum(O) # Total trips -->
-
-<!--  Ai <- rep(1, n) # Initialize Ai -->
-
-<!--   Bj <- rep(1, n)  # Initialize Bj -->
-
-<!--   previous_error <- Inf -->
-
-<!--   iteration_count <- 0 -->
-
-<!--   stop_reason <- "" -->
-
-<!--   while (TRUE) { -->
-
-<!--     iteration_count <- iteration_count + 1 -->
-
-<!--     # Update Ai -->
-
-<!--     for (i in 1:n) { -->
-
-<!--       Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9) -->
-
-<!--     } -->
-
-<!--     # Update Bj -->
-
-<!--     Bj_new <- rep(1, n) -->
-
-<!--     for (j in 1:n) { -->
-
-<!--       Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9) -->
-
-<!--     } -->
-
-<!--     # Calculate Tij -->
-
-<!--     Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix -->
-
-<!--     # Print Ai and Bj for this iteration -->
-
-<!--     cat("\nIteration:", iteration_count, "\n") -->
-
-<!--     cat("Ai:", round(Ai, 3), "\n") -->
-
-<!--     cat("Bj:", round(Bj_new, 3), "\n") -->
-
-<!--     # Calculate error -->
-
-<!--     error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T -->
-
-<!--     # Calculate change in error -->
-
-<!--     error_change <- abs(previous_error - error) -->
-
-<!--     # Check stopping conditions -->
-
-<!--     if (error < error_threshold) { -->
-
-<!--       stop_reason <- "Error threshold met" -->
-
-<!--       break -->
-
-<!--     } else if (error_change < improvement_threshold) { -->
-
-<!--       stop_reason <- "Slow improvement" -->
-
-<!--       break -->
-
-<!--     } -->
-
-<!--     previous_error <- error -->
-
-<!--     Bj <- Bj_new -->
-
-<!--   } -->
-
-<!--   # report stopping condition and final error threshold -->
-
-<!--   cat("\nStopping Condition: ", stop_reason, "\n") -->
-
-<!--   cat(sprintf("Final Error: %.3f%%\n", error * 100)) -->
-
-<!--   return(list(Ai = Ai, Bj = Bj, Tij = Tij)) -->
-
-<!-- } -->
-
-<!-- dc_results <- doubly_constrained_gravity_model(O=W_i, D=W_j, impedance_matrix = 1/C) -->
-
-<!-- ``` -->
-
-<!-- The $A_i$ and $B_j$ matrices: -->
-
-<!-- ```{r} -->
-
-<!-- A_i_doubly <- dc_results$Ai -->
-
-<!-- B_j_doubly <- dc_results$Bj -->
-
-<!-- A_i_doubly |> matrix() -->
-
-<!-- B_j_doubly |> matrix() |> t() -->
-
-<!-- ``` -->
-
-<!-- The trip matrix (notice, all trip ends are equivalent to the known trip ends) -->
-
-<!-- ```{r} -->
-
-<!-- T_doubly <- dc_results$Tij -->
-
-<!-- W_i_doubly <- T_doubly |> rowSums() -->
-
-<!-- W_j_doubly <- T_doubly |> colSums() -->
-
-<!-- T_TOTAL_dest <- ifelse(sum(W_i_doubly)==sum(W_j_doubly), sum(T_doubly), NA) -->
-
-<!-- T_doubly_longer <- cbind(T_doubly,W_i_dest) -->
-
-<!-- T_doubly_longer <- rbind(T_doubly_longer, cbind(t(W_j_doubly),T_TOTAL_dest)) -->
-
-<!-- rownames(T_doubly_longer) <- c("1","2","3", "TOTAL")  -->
-
-<!-- colnames(T_doubly_longer) <- c("1","2","3", "TOTAL") -->
-
-<!-- T_doubly_longer |> round() #|> as.data.frame()  |> gt(rownames_to_stub=TRUE) -->
-
-<!-- ``` -->
-
-<!-- ## Mean absolute percentage error (MAPE) between all cases -->
-
-<!-- ```{r} -->
-
-<!-- #unconstrained -->
-
-<!-- ((sum(abs(W_i_unc - rowSums(known_trip_matrix))) + sum(abs(W_j_unc - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
-
-<!-- #unconstrained -- rescaled! -->
-
-<!-- W_i_unc_rescaled <- T_unc |> scales::rescale() |> rowSums() -->
-
-<!-- W_j_unc_rescaled <- T_unc |> scales::rescale() |> colSums() -->
-
-<!-- W_i_rescaled <- known_trip_matrix |> scales::rescale() |> rowSums() -->
-
-<!-- W_j_rescaled <- known_trip_matrix |> scales::rescale() |> colSums() -->
-
-<!-- ((sum(abs(W_i_unc_rescaled - W_i_rescaled)) + sum(abs(W_j_unc_rescaled - W_j_rescaled))) / T) |> scales::percent(0.01) -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-
-<!-- #total constrained -->
-
-<!-- ((sum(abs(W_i_tot - rowSums(known_trip_matrix))) + sum(abs(W_j_tot - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
-
-<!-- #singly constrained (orig) -->
-
-<!-- ((sum(abs(W_i_orig - rowSums(known_trip_matrix))) + sum(abs(W_j_orig - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
-
-<!-- #singly constrained (dest) -->
-
-<!-- ((sum(abs(W_i_dest - rowSums(known_trip_matrix))) + sum(abs(W_j_dest - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
-
-<!-- #doubly constrained (dest) -->
-
-<!-- ((sum(abs(W_i_doubly - rowSums(known_trip_matrix))) + sum(abs(W_j_doubly - colSums(known_trip_matrix)))) / T) |> scales::percent(0.01) -->
-
-<!-- ``` -->
-
-<!-- - doubly c is lowest error -->
-
-<!-- - singly constrained error is mid -->
-
-<!-- - total constrained has high error -->
-
-<!-- - notably, unconstrained unscaled is obviously much higher in error - there are no constraints!
-But if you scale the matrices, the error is both low (interesting!!
-why is this?).
-However as mentioned, the unconstrained case is hard to interpret literally -- in what situations would k=1..?
--->
-
-<!-- So...
-our dear Mac alumni Fotheringham, in the early 1980s as a fresh PhD graduate demonstrates that the gravity model is misspecified.
-Responding to other authors at the time saying there is a 'map pattern' left over in estimated gravity models -- he demonstrated that this residual spatial variation can be reduced by including an additional independent variable into the gravity model, accessibility at the destination!
-This is the 'competing destination' model.
-In transport modelling, the CD model is one of a few ways they address this spatial In the wake of the debate, four generalizable approaches were proposed to account for the spatial structure effect (others include: the Box–Cox transform, spatial econometric models, and eigenvector spatial filtering (ESF)).
-Failing to account for destination accessibility when spatial structure effects are present generally biases origin-specific distance-decay parameter estimates upwards for accessible origins and downward for inaccessible origin.
-The strength of this bias is shown by Fotheringham (1984) to depend on the strength of two relationships: that between the volume of flows and destination accessibility and that between destination accessibility and distance from the origin to each destination.
-It is this bias that causes the unintuitive spatial patterns observed in origin-specific distance-decay parameter estimates, such as positive estimates for the most accessible origins (Fotheringham, 1981) -->
-
-<!-- So Aij is a relevant explanatory variable that can increase the accuracy of a Tij model!
-There's a relationship there.
-It makes me wonder if those spatial effects would be there if you used a constrained accessibility measure.
-Maybe using Aij (Hansen-type accessibility) doesn't completely solve this issue.
--->

### 2025-05-20 paper/manuscript.qmd
fixed a typo

@@
 Accessibility, however, is often presented and understood as a zonal summary of *potential* for interaction between origins and destinations that contain inherently different units.
-Arriving at the doubly constrained case through the unconstrained, totally, and singly constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
+Arriving at the doubly constrained case through the unconstrained, total, and singly constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
 Namely, by demonstrating that these members of the accessibility family all derive from the same root and can be derived from Wilson's original formulation, this more clearly demonstrates what *potential* is, within the context of spatial interaction modelling.

### 2025-05-20 paper/manuscript.tex
fixed a typo

@@
 destinations that contain inherently different units. Arriving at the
-doubly constrained case through the unconstrained, totally, and singly
+doubly constrained case through the unconstrained, total, and singly
 constrained cases make the connection between the \emph{potential} for
@@
 
-\section{References}\label{references}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 

### 2025-05-21 paper/manuscript.qmd
added email to abstract -- for preprint

@@
 author:
-  - name: Anastasia Soukhov 
+  - name: Anastasia Soukhov*
     email: soukhoa@mcmaster.ca
@@
 
+  *Email of corresponding author for preprint: soukhoa@mcmaster.ca
+
 keywords: [accessibility; spatial interaction model; balancing factors; competitive accessibility; constraints; singly constrained; doubly constrained; total constrained]
@@
 In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework.
-At the end of this section, Table 8 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.
+At the end of this section, Table 1 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.
 

### 2025-05-21 paper/manuscript.tex
added email to abstract -- for preprint

@@
   pdftitle={A family of accessibility measures derived from spatial interaction principles},
-  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
+  pdfauthor={Anastasia Soukhov*; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
   pdfkeywords={accessibility; spatial interaction model; balancing
@@
 \subtitle{Submitted preprint for peer-review (May 2025)}
-\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
+\author{Anastasia Soukhov* \and Rafael H. M. Pereira \and Christopher D.
 Higgins \and Antonio Páez}
@@
 interpretable, and cohesive foundation for accessibility analysis.
+
+*Email of corresponding author for preprint: soukhoa@mcmaster.ca
 \end{abstract}
@@
 inspired by Wilson's spatial interaction modelling framework. At the end
-of this section, Table 8 presents a summary, where each member of the
+of this section, Table 1 presents a summary, where each member of the
 family of accessibility measure is explained in plain language,

### 2025-05-23 paper/manuscript.qmd
Cleaned up a few small equation errors - and posted preprint link to repo page.

@@
 title: "A family of accessibility measures derived from spatial interaction principles"
-subtitle: "Submitted preprint for peer-review (May 2025)"
+#subtitle: "Submitted preprint for peer-review (May 2025)"
 author:
-  - name: Anastasia Soukhov*
+  - name: Anastasia Soukhov
     email: soukhoa@mcmaster.ca
@@
     header-includes: |
-      \usepackage{draftwatermark} 
-      \SetWatermarkText{DRAFT}
-      \SetWatermarkScale{1.5}
-      \SetWatermarkColor[gray]{0.95}
       \usepackage{float}
@@
 By doing so, we believe this family of measures can unlock a clearer, more interpretable, and cohesive foundation for accessibility analysis.
-
-  *Email of corresponding author for preprint: soukhoa@mcmaster.ca
 
@@
 Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints.
-Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @q-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).
+Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @eq-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).
 
@@
 Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone.
-[^1].
-Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
-For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025].
+For reference, considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
+The number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025].
 Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
@@
 
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_i^T$ is dimensionless:
-
-$$
-\kappa_i^T = \frac{\sum_j D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa_i^T\sum_j W^{(2)}_j$).
-The role of $\kappa_i^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
-$\kappa_i^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-This is why we refer to $\kappa_i^T$ as a proportional allocation factor.
+Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_{ij}^T$ is dimensionless:
+
+$$
+\kappa_{ij}^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T W^{(2)}_j$).
+The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
+$\kappa_{ij}^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
+This is why we refer to $\kappa_{ij}^T$ as a proportional allocation factor.
 On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.
@@
 Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
-$\kappa_i^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
+$\kappa_{ij}^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
 
@@
 #| label: tbl-simple-example-total-opportunity-accessibility
-#| tbl-cap: "Simple system: total opportunity constrained accessibility."
+#| tbl-cap: "Simple system: total constrained accessible opportunities."
 
@@
 In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
-Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_i^T$.
+Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_{ij}^T$.
 
@@
 Why?
-$\kappa_i^T$ ensures proportional allocation for each travel behaviour scenario.
-Meaning, while the unconstrained accessibility increases, $\kappa_i^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
+$\kappa_{ij}^T$ ensures proportional allocation for each travel behaviour scenario.
+Meaning, while the unconstrained accessibility increases, $\kappa_{ij}^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
 As the decay behaviour decreases, more opportunities are accessible for all zones.
@@
 
-Where the following $\hat \kappa_i^T$ proportional allocation factor is dimensionless:
-
-$$
-\hat \kappa_i^T = \frac{\sum_i O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+Where the following $\hat \kappa_{ji}^T$ proportional allocation factor is dimensionless:
+
+$$
+\hat \kappa_{ji}^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 $$
@@
 #| label: tbl-simple-example-total-population-accessibility
-#| tbl-cap: "Simple system: total opportunity constrained accessibility."
+#| tbl-cap: "Simple system: total constrained accessible population."
 
@@
 $$
-\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+\kappa^S_{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
 $$ {#eq-opportunity-constrained-proportional-allocation-factor}
@@
 $$
-V^S_{i} = \kappa^S_{i} \sum_j   D_j
+V^S_{i} = \sum_j \kappa^S_{ij} D_j
 $$ {#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
 
-@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{i}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As in the total constrained opportunity case, $\kappa^S_{i}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
+@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
+As in the total constrained opportunity case, $\kappa^S_{ij}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
 The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility
-#| tbl-cap: "Simple system: singly constrained opportunity accessibility."
+#| tbl-cap: "Simple system: singly constrained accessible opportunities."
 
@@
 
-Imposing the single proportional allocation factor $\kappa^S_{i}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case.
+Imposing the single proportional allocation factor $\kappa^S_{ij}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case.
 The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.
 
-However, unlike the total constrained opportunity case, $\kappa^S_{i}$ reflects zonal competition based on the mass of the origin, i.e., population.
+However, unlike the total constrained opportunity case, $\kappa^S_{ij}$ reflects zonal competition based on the mass of the origin, i.e., population.
 Again, consider the highest decay scenario $f_1(c_{ij})$.
@@
 
-This difference is due to the single-opportunity factor $\kappa^S_{i}$'s role in $V_i^S$.
-The only inputs required in the total-opportunity factor $\kappa^T_i$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
+This difference is due to the single-opportunity factor $\kappa^S_{ij}$'s role in $V_i^S$.
+The only inputs required in the total-opportunity factor $\kappa^T_{ij}$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
 Opportunities are allocated to $i$s, regardless of the mass weights of origin.
-Whereas the single opportunity constraint $\kappa^S_{i}$ requires the population at $i$ as an input.
-In fact, $\kappa^S_{i}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
+Whereas the single opportunity constraint $\kappa^S_{ij}$ requires the population at $i$ as an input.
+In fact, $\kappa^S_{ij}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
 Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
-#| tbl-cap: "Simple system: singly constrained opportunity accessibility per capita."
+#| tbl-cap: "Simple system: singly constrained accessible opportunities per capita."
 
@@
 $$
-M^S_{j} = \hat \kappa^S_{j} \sum_i  O_i
+M^S_{j} = \hat \kappa^S_{ji} \sum_i  O_i
 $$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}
@@
 $$
-\kappa^S _{j} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
 $$ {#eq-attraction-constrained-proportional-allocation-factor}
@@
 $$
-V^D_{i} = \kappa^D_{ij} \sum_j D_j
+V^D_{i} = \sum_j \kappa^D_{ij} D_j
 $$ {#eq-doubly-constrained-accessibility-w-paf}
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
-#| tbl-cap: "Doubly constrained opportunity accessibility assuming the most decay travel behaviour scenario in the modified simple system."
+#| tbl-cap: "Doubly constrained accessible opportunities assuming highest travel decay in the modified simple system."
 
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
-#| tbl-cap: "Doubly constrained opportunity accessibility for all travel behaviour scenarios for Zone 2 in the modified simple system."
+#| tbl-cap: "Doubly constrained accessible opportunities at Zone 2 for all travel decay groups in the modified simple system."
 

### 2025-05-23 paper/manuscript.tex
Cleaned up a few small equation errors - and posted preprint link to repo page.

@@
 \usepackage{multirow}
-\usepackage{draftwatermark}
-\SetWatermarkText{DRAFT}
-\SetWatermarkScale{1.5}
-\SetWatermarkColor[gray]{0.95}
 \usepackage{float}
@@
   pdftitle={A family of accessibility measures derived from spatial interaction principles},
-  pdfauthor={Anastasia Soukhov*; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
+  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
   pdfkeywords={accessibility; spatial interaction model; balancing
@@
 interaction principles}
-\usepackage{etoolbox}
-\makeatletter
-\providecommand{\subtitle}[1]{% add subtitle to \maketitle
-  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
-}
-\makeatother
-\subtitle{Submitted preprint for peer-review (May 2025)}
-\author{Anastasia Soukhov* \and Rafael H. M. Pereira \and Christopher D.
+\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
 Higgins \and Antonio Páez}
@@
 interpretable, and cohesive foundation for accessibility analysis.
-
-*Email of corresponding author for preprint: soukhoa@mcmaster.ca
 \end{abstract}
@@
 population potential of opportunities for spatial interaction (i.e.,
-(\textbf{q-stewart-population-potential-sum?}) representing
+Equation~\ref{eq-stewart-population-potential-sum} representing
 \(V_i = \sum_j \frac{M_j}{d_{ij}}\) after losing \(G\)).
@@
 (in units of 10,000s of people) and the opportunities (the number of
-physicians) per zone. {[}\^{}1{]}. Considering
+physicians) per zone. For reference, considering
 Table~\ref{tbl-small-system-land-use}'s values, the
-Provider-to-Population-Ratio (PPR) in this system is 24.5. For
-reference, the number of physicians per 10,000 in Canada in 2022 was
-24.97 (WHO 2025). Secondly, to pair the zonal population and opportunity
-information, the assumed cost of movement (in minutes of travel time)
-between origins and destinations is as shown in
-Table~\ref{tbl-small-system-cost}.
+Provider-to-Population-Ratio (PPR) in this system is 24.5. The number of
+physicians per 10,000 in Canada in 2022 was 24.97 (WHO 2025). Secondly,
+to pair the zonal population and opportunity information, the assumed
+cost of movement (in minutes of travel time) between origins and
+destinations is as shown in Table~\ref{tbl-small-system-cost}.
 
@@
 units of opportunities, the proportional allocation factor for the total
-constrained opportunity case \(\kappa_i^T\) is dimensionless:
+constrained opportunity case \(\kappa_{ij}^T\) is dimensionless:
 
 \[
-\kappa_i^T = \frac{\sum_j D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+\kappa_{ij}^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 \]
@@
 that is, the mass at the destination
-(\(V^T_i = \kappa_i^T\sum_j W^{(2)}_j\)). The role of \(\kappa_i^T\) in
-this reformulation of accessibility is to transform between units and to
-adjust the number of opportunities accessible from \(i\) so they
-represent a proportion of the total number of opportunities in the
-region. \(\kappa_i^T\) then assigns opportunities in proportion to the
-impedance between \(i\) and \(j\). This is why we refer to
-\(\kappa_i^T\) as a proportional allocation factor. On the other hand,
-the proportionality constant \(K^T\) balances the units of \(V^0_i\),
-the Hansen-type accessibility values, and is an alternative expression
-of the total constrained accessibility measure.
+(\(V^T_i = \sum_j \kappa_{ij}^T W^{(2)}_j\)). The role of
+\(\kappa_{ij}^T\) in this reformulation of accessibility is to transform
+between units and to adjust the number of opportunities accessible from
+\(i\) so they represent a proportion of the total number of
+opportunities in the region. \(\kappa_{ij}^T\) then assigns
+opportunities in proportion to the impedance between \(i\) and \(j\).
+This is why we refer to \(\kappa_{ij}^T\) as a proportional allocation
+factor. On the other hand, the proportionality constant \(K^T\) balances
+the units of \(V^0_i\), the Hansen-type accessibility values, and is an
+alternative expression of the total constrained accessibility measure.
 
@@
 Table~\ref{tbl-simple-example-total-opportunity-accessibility}.
-\(\kappa_i^T\) for each zone is not shown, but can be understood to be
-the unitless proportion of opportunities (of the total opportunities)
+\(\kappa_{ij}^T\) for each zone is not shown, but can be understood to
+be the unitless proportion of opportunities (of the total opportunities)
 allocated to each zone based on travel impedance.
@@
 \caption{\label{tbl-simple-example-total-opportunity-accessibility}Simple
-system: total opportunity constrained accessibility.}
+system: total constrained accessible opportunities.}
 
@@
 value is effectively in units of physicians, with the impedance units
-already accounted for by \(\kappa_i^T\).
+already accounted for by \(\kappa_{ij}^T\).
 
@@
 lesser decay scenarios \(f_2(c_{ij})\) and \(f_3(c_{ij})\) respectively.
-Why? \(\kappa_i^T\) ensures proportional allocation for each travel
+Why? \(\kappa_{ij}^T\) ensures proportional allocation for each travel
 behaviour scenario. Meaning, while the unconstrained accessibility
-increases, \(\kappa_i^T\) adjusts the values to remain proportional to
-the total number of opportunities (490 physicians accessible in the
+increases, \(\kappa_{ij}^T\) adjusts the values to remain proportional
+to the total number of opportunities (490 physicians accessible in the
 region). As the decay behaviour decreases, more opportunities are
@@
 
-Where the following \(\hat \kappa_i^T\) proportional allocation factor
-is dimensionless:
+Where the following \(\hat \kappa_{ji}^T\) proportional allocation
+factor is dimensionless:
 
 \[
-\hat \kappa_i^T = \frac{\sum_i O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+\hat \kappa_{ji}^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 \]
@@
 \caption{\label{tbl-simple-example-total-population-accessibility}Simple
-system: total opportunity constrained accessibility.}
+system: total constrained accessible population.}
 
@@
 \begin{equation}\phantomsection\label{eq-opportunity-constrained-proportional-allocation-factor}{
-\kappa^S_{i} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
+\kappa^S_{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}{
-V^S_{i} = \kappa^S_{i} \sum_j   D_j
-}\end{equation}
-
-Soukhov et al. (2023) have shown that the role of \(\kappa^S_{i}\) is to
-allocate opportunities \(D_j\) proportionally to the mass at each origin
-\(i\) and the impedance between \(i\) and \(j\). As in the total
-constrained opportunity case, \(\kappa^S_{i}\) is dimensionless and
+V^S_{i} = \sum_j \kappa^S_{ij} D_j
+}\end{equation}
+
+Soukhov et al. (2023) have shown that the role of \(\kappa^S_{ij}\) is
+to allocate opportunities \(D_j\) proportionally to the mass at each
+origin \(i\) and the impedance between \(i\) and \(j\). As in the total
+constrained opportunity case, \(\kappa^S_{ij}\) is dimensionless and
 \(V_i^S\) is in the units of opportunities \(D_j\). The singly
@@
 \caption{\label{tbl-simple-example-attraction-constrained-accessibility}Simple
-system: singly constrained opportunity accessibility.}
+system: singly constrained accessible opportunities.}
 
@@
 
-Imposing the single proportional allocation factor \(\kappa^S_{i}\)
+Imposing the single proportional allocation factor \(\kappa^S_{ij}\)
 allows for the comparison of differences and ratios of the accessibility
@@
 
-However, unlike the total constrained opportunity case, \(\kappa^S_{i}\)
-reflects zonal competition based on the mass of the origin, i.e.,
-population. Again, consider the highest decay scenario \(f_1(c_{ij})\).
-Under this scenario, zone 1 no longer captures a medium amount of
-physicians as in the total constrained opportunity case: it now captures
-the fewest in the region i.e., 133.4687282 at zone 1, 166.7813387 at
-zone 2, and 189.7499331 at zone 3. Why may zone 1 (healthcare cluster at
-the edge of the city) capture 27\% of the physicians regionally while
-this same zone captures 35\% in the total opportunity constrained case?
+However, unlike the total constrained opportunity case,
+\(\kappa^S_{ij}\) reflects zonal competition based on the mass of the
+origin, i.e., population. Again, consider the highest decay scenario
+\(f_1(c_{ij})\). Under this scenario, zone 1 no longer captures a medium
+amount of physicians as in the total constrained opportunity case: it
+now captures the fewest in the region i.e., 133.4687282 at zone 1,
+166.7813387 at zone 2, and 189.7499331 at zone 3. Why may zone 1
+(healthcare cluster at the edge of the city) capture 27\% of the
+physicians regionally while this same zone captures 35\% in the total
+opportunity constrained case?
 
 This difference is due to the single-opportunity factor
-\(\kappa^S_{i}\)'s role in \(V_i^S\). The only inputs required in the
-total-opportunity factor \(\kappa^T_i\) is the total number of
+\(\kappa^S_{ij}\)'s role in \(V_i^S\). The only inputs required in the
+total-opportunity factor \(\kappa^T_{ij}\) is the total number of
 opportunities in the region \(D\) as well as the associated
@@
 allocated to \(i\)s, regardless of the mass weights of origin. Whereas
-the single opportunity constraint \(\kappa^S_{i}\) requires the
-population at \(i\) as an input. In fact, \(\kappa^S_{i}\) is calculated
-as the proportion of impedance-weighted population at an \(i\) to the
-sum of impedance-weighted population for the entire region. Hence, since
-zone 1 has the lowest population in the region, is in close proximity to
-a more populated zone (zone 3, the `urban core'), and is not well
-connected (in terms of travel impedance) to other zones with
+the single opportunity constraint \(\kappa^S_{ij}\) requires the
+population at \(i\) as an input. In fact, \(\kappa^S_{ij}\) is
+calculated as the proportion of impedance-weighted population at an
+\(i\) to the sum of impedance-weighted population for the entire region.
+Hence, since zone 1 has the lowest population in the region, is in close
+proximity to a more populated zone (zone 3, the `urban core'), and is
+not well connected (in terms of travel impedance) to other zones with
 opportunities, \(V_{1}^S\) values, or the number of physicians
@@
 \caption{\label{tbl-simple-example-attraction-constrained-accessibility-per-capita}Simple
-system: singly constrained opportunity accessibility per capita.}
+system: singly constrained accessible opportunities per capita.}
 
@@
 \begin{equation}\phantomsection\label{eq-production-constrained-accessibility-with-proportional-allocation-factor}{
-M^S_{j} = \hat \kappa^S_{j} \sum_i  O_i
+M^S_{j} = \hat \kappa^S_{ji} \sum_i  O_i
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-attraction-constrained-proportional-allocation-factor}{
-\kappa^S _{j} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
+\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-doubly-constrained-accessibility-w-paf}{
-V^D_{i} = \kappa^D_{ij} \sum_j D_j
+V^D_{i} = \sum_j \kappa^D_{ij} D_j
 }\end{equation}
@@
 \caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}Doubly
-constrained opportunity accessibility assuming the most decay travel
-behaviour scenario in the modified simple system.}
+constrained accessible opportunities assuming highest travel decay in
+the modified simple system.}
 
@@
 \caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}Doubly
-constrained opportunity accessibility for all travel behaviour scenarios
-for Zone 2 in the modified simple system.}
+constrained accessible opportunities at Zone 2 for all travel decay
+groups in the modified simple system.}
 

### 2025-05-23 paper/summary-family-access-measures-table.tex
Cleaned up a few small equation errors - and posted preprint link to repo page.

@@
 \hline
-\textbf{Name of Case and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
+\textbf{Name of Member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
 \hline
@@
 \hline
-\textbf{Name of Case and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
+\textbf{Name of member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
 \hline
@@
 & None
-& $V_i^0 = \sum_j O_j f(c_{ij})$;
+& $V_i^0 = \sum_j D_j f(c_{ij})$;
 
-$M_j^0 = \sum_i P_i f(c_{ij})$
+$M_j^0 = \sum_i O_i f(c_{ij})$
 
@@
 
-& Allocates the total marginal as opportunities based on $\kappa_i^T = \sum_j K^T f(c_{ij})$ and as population based on  $\hat{\kappa}_j^T = \sum_i \hat K^T f(c_{ij})$
+& Allocates the total marginal as opportunities based on $\kappa_{ij}^T = K^T f(c_{ij})$ and as population based on  $\hat{\kappa}_{ji}^T = \hat K^T f(c_{ij})$
 
-& $V^T_i = \kappa_i^T \sum_j W^{(2)}_j$;
+& $V^T_i =  \sum_j \kappa_{ij}^T W^{(2)}_j$;
 
-$M_j^T = \hat{\kappa}_j^T \sum_i W^{(1)}_i$
+$M_j^T =  \sum_i \hat{\kappa}_{ji}^T W^{(1)}_i$
 
@@
 
-& Allocates the single opportunities marginal proportionally based on $\kappa^S_i = W_i^{(1)} B_j$ in the case of $V_i^S$ and the single population marginal proroportionally based on $\hat \kappa^S_j = W_j^{(2)} A_i$ in the case of $M_j^S$
+& Allocates the single opportunities marginal proportionally based on $\kappa^S_{ij} =  B_j W_i^{(1)} f(c_{ij})$ in the case of $V_i^S$ and the single population marginal proroportionally based on $\hat \kappa^S_{ji} = A_i W_j^{(2)} f(c_{ij})$ in the case of $M_j^S$.
 
-& $V^S_i = \kappa^S_i \sum_j D_j$;
+& $V^S_i = \sum_j \kappa^S_{ij} D_j$;
 
-$M_j^S = \hat \kappa^S_j \sum_i P_i$
+$M_j^S =  \sum_i \hat \kappa^S_{ji} O_i$
 
@@
 & —
-& $V_{ij}^D = A_i B_j P_i O_j f(c_{ij})$
+& $V_{ij}^D = A_i B_j O_i D_j f(c_{ij})$
 & The spatial interactions between population and opportunities (i.e., access). \\

### 2025-05-26 paper/manuscript.qmd
additional small kappa related typos fixed

@@
 $$
-B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
+B_a = \frac{(P_a P_T)^N}{D_{aT}^n}
 $$ {#eq-reilly}
@@
 $$
-C^2 = \frac{P_1\cdot P_2}{D_{12}}
+C^2 = \frac{P_1 P_2}{D_{12}}
 $$ {#eq-zipf}
@@
 $$
-F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
+F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1 M_2}{d_{12}^2} 
 $$ {#eq-stewart-force}
@@
 \noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
-In @eq-stewart-population-potential-integral, $D\cdot ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
+In @eq-stewart-population-potential-integral, $D ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
 
@@
 $$
-S_{i} = \sum_j O_j \cdot f(d_{ij})
+S_{i} = \sum_j O_j  f(d_{ij})
 $$ {#eq-accessibility-general}
@@
 The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
-For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta\cdot m}$ when $f(c_{ij}) = e^{-\beta\cdot m}$.
+For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta m}$ when $f(c_{ij}) = e^{-\beta m}$.
 This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
@@
 
-The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
-The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,\cdots,n$ origins and $j=1,\cdots,m$ destinations.
+The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
+The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,s,n$ origins and $j=1,s,m$ destinations.
 The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., $W_i^{(1)}$ could be populations, and $W_j^{(2)}$ hectares of park space.
@@
 
-# The product of $A$ and its transpose gives the bibliographical coupling by cited references $B = A\cdot A^T$
+# The product of $A$ and its transpose gives the bibliographical coupling by cited references $B = A A^T$
 
@@
 
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,\cdots, n$ origins and $j = 1,\cdots, m$ destinations.
+Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
 
@@
 $$
-V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
+V^T_{ij} = K^T  W_j^{(2)}  f(c_{ij})
 $$ {#eq-total-constrained-access}
@@
 $$
-V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
+V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T  V^0_i
 $$ {#eq-total-constrained-accessibility}
@@
 $$
-V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j D\frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $$
@@
 $$
-\kappa_{ij}^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T W^{(2)}_j$).
-The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region.
+\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+
+\noindent and therefore $V^T_i$ is now in the units of $D$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T D$).
+The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region $D$.
 $\kappa_{ij}^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
@@
 K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})}\\
-K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + \cdots + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
+K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + s + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
 }\\
@@
 $$
-M^T_{ji} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ji}^0
+M^T_{ji} = \hat K^T  W_i^{(1)} f(c_{ij}) = \hat K^T  M_{ji}^0
 $$ {#eq-total-constrained-market}
@@
 \sum_j M^T_{j} = \sum_i\sum_j M^T_{ji} = O
-$$ {#eq-total-constraint-market}
+$${#eq-total-constraint-market}
 
@@
 \hat K^T=\frac{O}{\sum_i\sum_i M^T_{ji}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
-$$ {#eq-total-population-balancing-factor}
+$${#eq-total-population-balancing-factor}
 
 The constrained market potential then takes the following form:
-
-$$
-M^T_j = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$
-
-Where the following $\hat \kappa_{ji}^T$ proportional allocation factor is dimensionless:
-
-$$
-\hat \kappa_{ji}^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$
-
-Returning back to the numerical example, the proportionality constant $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
+$$
+M^T_j = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = O \sum_i \frac{W_i^{(1)} f(c_{ij})}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
+$$
+
+Where the following $\hat\kappa_{ji}^T$ proportional allocation factor is dimensionless, reflecting the proportional allocation of population in the region $O$ (i.e., $M^T_j = \sum_i \hat \kappa_{ji}^T O$)
+$$
+\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+$$
+
+Returning back to the numerical example, the balancing factor $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
 
@@
 $$
-M^S_{j} = \hat \kappa^S_{ji} \sum_i  O_i
+M^S_{j} = \sum_i \hat \kappa^S_{ji} O_i
 $$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}

### 2025-05-26 paper/manuscript.tex
additional small kappa related typos fixed

@@
 \begin{equation}\phantomsection\label{eq-reilly}{
-B_a = \frac{(P_a\cdot P_T)^N}{D_{aT}^n}
+B_a = \frac{(P_a P_T)^N}{D_{aT}^n}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-zipf}{
-C^2 = \frac{P_1\cdot P_2}{D_{12}}
+C^2 = \frac{P_1 P_2}{D_{12}}
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-stewart-force}{
-F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1\cdot M_2}{d_{12}^2} 
+F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1 M_2}{d_{12}^2} 
 }\end{equation}
@@
 area \(ds\) and \(r\) is the distance to \(i\). In
-Equation~\ref{eq-stewart-population-potential-integral}, \(D\cdot ds\)
-gives an infinitesimal count of the population, say \(dm\), and so,
-after discretizing space,
+Equation~\ref{eq-stewart-population-potential-integral}, \(D ds\) gives
+an infinitesimal count of the population, say \(dm\), and so, after
+discretizing space,
 Equation~\ref{eq-stewart-population-potential-integral} can be rewritten
@@
 \begin{equation}\phantomsection\label{eq-accessibility-general}{
-S_{i} = \sum_j O_j \cdot f(d_{ij})
+S_{i} = \sum_j O_j  f(d_{ij})
 }\end{equation}
@@
 \(f(c_{ij}) = d^{-\beta}\) but number of opportunities per
-\(e^{-\beta\cdot m}\) when \(f(c_{ij}) = e^{-\beta\cdot m}\). This
-undermines the comparability of accessibility metrics with different
-decay functions, and renders their results difficult to understand and
-communicate. The Hansen-style accessibility estimates found in the
-literature, therefore, are better thought of as ordinal measures of
-potential that can only be interpreted in terms of higher and lower
-accessibility, but which has not palpable meaning (E. J. Miller 2018).
+\(e^{-\beta m}\) when \(f(c_{ij}) = e^{-\beta m}\). This undermines the
+comparability of accessibility metrics with different decay functions,
+and renders their results difficult to understand and communicate. The
+Hansen-style accessibility estimates found in the literature, therefore,
+are better thought of as ordinal measures of potential that can only be
+interpreted in terms of higher and lower accessibility, but which has
+not palpable meaning (E. J. Miller 2018).
 
@@
 \(T_{ij}\) that represents a value in a matrix of flows of size
-\(n \times m\), that is, between \(i = 1,\cdots, n\) origins and
-\(j = 1,\cdots, m\) destinations. The quantities \(W_i^{(1)}\) and
-\(W_j^{(2)}\) are proxies for the masses at \(i=1,\cdots,n\) origins and
-\(j=1,\cdots,m\) destinations. The super-indices \((1)\) and \((2)\) are
+\(n \times m\), that is, between \(i = 1,s, n\) origins and
+\(j = 1,s, m\) destinations. The quantities \(W_i^{(1)}\) and
+\(W_j^{(2)}\) are proxies for the masses at \(i=1,s,n\) origins and
+\(j=1,s,m\) destinations. The super-indices \((1)\) and \((2)\) are
 meant to indicate that these masses could be different things, i.e.,
@@
 Similar to Equation~\ref{eq-phys-gravity-model}, \(W_j^{(2)}\) above is
-the mass at the destination and the sub-indices are for
-\(i = 1,\cdots, n\) origins and \(j = 1,\cdots, m\) destinations.
+the mass at the destination and the sub-indices are for \(i = 1,s, n\)
+origins and \(j = 1,s, m\) destinations.
 
@@
 \begin{equation}\phantomsection\label{eq-total-constrained-access}{
-V^T_{ij} = K^T \cdot W_j^{(2)} \cdot f(c_{ij})
+V^T_{ij} = K^T  W_j^{(2)}  f(c_{ij})
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-total-constrained-accessibility}{
-V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T \cdot V^0_i
+V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T  V^0_i
 }\end{equation}
@@
 \[
-V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j D\frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 \]
@@
 \[
-\kappa_{ij}^T = \frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 \]
 
-\noindent and therefore \(V^T_i\) is now in the units of \(W^{(2)}_j\),
-that is, the mass at the destination
-(\(V^T_i = \sum_j \kappa_{ij}^T W^{(2)}_j\)). The role of
-\(\kappa_{ij}^T\) in this reformulation of accessibility is to transform
-between units and to adjust the number of opportunities accessible from
-\(i\) so they represent a proportion of the total number of
-opportunities in the region. \(\kappa_{ij}^T\) then assigns
+\noindent and therefore \(V^T_i\) is now in the units of \(D\), that is,
+the mass at the destination (\(V^T_i = \sum_j \kappa_{ij}^T D\)). The
+role of \(\kappa_{ij}^T\) in this reformulation of accessibility is to
+transform between units and to adjust the number of opportunities
+accessible from \(i\) so they represent a proportion of the total number
+of opportunities in the region \(D\). \(\kappa_{ij}^T\) then assigns
 opportunities in proportion to the impedance between \(i\) and \(j\).
@@
 \begin{equation}\phantomsection\label{eq-total-constrained-market}{
-M^T_{ji} = \hat K^T \cdot W_i^{(1)} f(c_{ij}) = \hat K^T \cdot M_{ji}^0
+M^T_{ji} = \hat K^T  W_i^{(1)} f(c_{ij}) = \hat K^T  M_{ji}^0
 }\end{equation}
@@
 
-The constrained market potential then takes the following form:
-
-\[
-M^T_j = \hat K^T\sum_i W^{(1)}_if(c_{ij}) = \sum_i W^{(2)}_i \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+The constrained market potential then takes the following form: \[
+M^T_j = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = O \sum_i \frac{W_i^{(1)} f(c_{ij})}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
 \]
 
-Where the following \(\hat \kappa_{ji}^T\) proportional allocation
-factor is dimensionless:
-
-\[
-\hat \kappa_{ji}^T = \frac{O \cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
+Where the following \(\hat\kappa_{ji}^T\) proportional allocation factor
+is dimensionless, reflecting the proportional allocation of population
+in the region \(O\) (i.e., \(M^T_j = \sum_i \hat \kappa_{ji}^T O\)) \[
+\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
 \]
 
-Returning back to the numerical example, the proportionality constant
+Returning back to the numerical example, the balancing factor
 \(\hat K^T\) is solved for each travel behaviour scenario, and the
@@
 \begin{equation}\phantomsection\label{eq-production-constrained-accessibility-with-proportional-allocation-factor}{
-M^S_{j} = \hat \kappa^S_{ji} \sum_i  O_i
+M^S_{j} = \sum_i \hat \kappa^S_{ji} O_i
 }\end{equation}

### 2025-05-26 paper/summary-family-access-measures-table.tex
additional small kappa related typos fixed

@@
 
-& Allocates the total marginal as opportunities based on $\kappa_{ij}^T = K^T f(c_{ij})$ and as population based on  $\hat{\kappa}_{ji}^T = \hat K^T f(c_{ij})$
+& Allocates the total marginal as opportunities $D$ based on $\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}$ and as population $O$ based on $\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}$
 
-& $V^T_i =  \sum_j \kappa_{ij}^T W^{(2)}_j$;
+& $V^T_i =  \sum_j \kappa_{ij}^T D$;
 
-$M_j^T =  \sum_i \hat{\kappa}_{ji}^T W^{(1)}_i$
+$M_j^T =  \sum_i \hat{\kappa}_{ji}^T O$
 
@@
 
-& Allocates the single opportunities marginal proportionally based on $\kappa^S_{ij} =  B_j W_i^{(1)} f(c_{ij})$ in the case of $V_i^S$ and the single population marginal proroportionally based on $\hat \kappa^S_{ji} = A_i W_j^{(2)} f(c_{ij})$ in the case of $M_j^S$.
+& Allocates the single opportunities marginal $D_j$ proportionally based on $\kappa^S_{ij} =  \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}$ in the case of $V_i^S$ and the single population marginal $O_i$ proroportionally based on $\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}$ in the case of $M_j^S$.
 

### 2025-05-28 paper_plos/paper_plos.qmd
move to the plos template
most recent paper now in `paper_plos`

--- 2025-05-26 paper/manuscript.qmd
@@
 title: "A family of accessibility measures derived from spatial interaction principles"
-#subtitle: "Submitted preprint for peer-review (May 2025)"
+format:
+  plos-pdf:
+    number-sections: false
+    keep-tex: true
+    journal:
+      id: plosone
+    header-includes:
+      - \usepackage{float}
+      - \floatplacement{table}{H}
+      - \floatplacement{figure}{H}
+      - \usepackage{pdflscape}
+      - \newcommand{\blandscape}{\begin{landscape}}
+      - \newcommand{\elandscape}{\end{landscape}}
+    #include-in-header:
+    #  text: |
+    #    % Remove comment for double spacing
+    #    % \usepackage{setspace} 
+    ##    % \doublespacing
+    #    % \usepackage{pdflscape}
+    #    % \newcommand{\blandscape}{\begin{landscape}}
+    #    % \newcommand{\elandscape}{\end{landscape}}
 author:
@@
         - name: Institute for Applied Economic Research - Ipea
+          department: Data Science Division
           city: Brasília, Brazil
@@
           city: Hamilton, Canada
-format:
-  pdf:
-    keep-tex: true
-    documentclass: article
-    header-includes: |
-      \usepackage{float}
-      \floatplacement{table}{H}
-      \floatplacement{figure}{H}
-bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-#includes:
-#      in_header: 'gt_packages.sty'
-      #\usepackage{float}
+
+author-notes:
+  equal-contributor: These authors contributed equally to this work.
+  deceased: Deceased
+  group: Membership list can be found in the Acknowledgments sections
 
@@
 
-keywords: [accessibility; spatial interaction model; balancing factors; competitive accessibility; constraints; singly constrained; doubly constrained; total constrained]
+bibliography: bibliography.bib 
 ---
@@
 However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
-For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
-Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
+For instance, the diversity of accessibility definitions has been flagged by van Wee [@vanweeAccessible2016], Handy [@handy2020], and Kapatsila et al.
+[@kapatsila_resolving_2023].
+Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004], van Wee [@vanweeAccessible2016], and Ferreira and Papa [@ferreiraReenactingMobilityAccessibility2020].
 
 The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
-At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
+At a high level, Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
 Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
@@
 What differences in accessibility scores should matter, and how should they be communicated?
-[see @vanweeAccessible2016; p.
+[see @vanweeAccessible2016 p.
 14].
@@
 Here, we seek to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
-In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971 that can be defined using balancing factors that constrains values based on known information about the land-use.
+In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in Wilson [@wilson1971] that can be defined using balancing factors that constrains values based on known information about the land-use.
 
@@
 Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
-Empirical constraints were embraced by early spatial interaction literature following the work of @wilson1971, but this stream of literature tended to flow separately from research inspired by @hansen1959' accessibility.
-The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
+Empirical constraints were embraced by early spatial interaction literature following the work of Wilson [@wilson1971], but this stream of literature tended to flow separately from research inspired by Hansen's [@hansen1959] accessibility.
+The application of Wilson's [@wilson1971] empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
 However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted.
@@
 It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-We continue this work in the following section by tracing the development of accessibility from its origins in spatial interaction: from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
-We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
-Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on analogous constraints.
+We continue this work in the following section by tracing the development of accessibility from its origins in spatial interaction: from the Newtonian gravitational expression in Eavenstein [@ravensteinLawsMigration1889] through to the seminal accessibility work of Hansen [@hansen1959].
+We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of Wilson [@wilson1971].
+Next, we hark back to Wilson's [@wilson1971] spatial interaction models, and use them to derive a family of accessibility measures based on analogous constraints.
 We illustrate members of this family of constrained accessibility measures with a simple numerical example.
@@
 Over time, other researchers discovered similar relationships.
-<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations that could be understood as 'potential trade territories'.
+<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, Reilly [@reilly1929methods] formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations that could be understood as 'potential trade territories'.
 Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of information [@zipfDeterminantsCirculationInformation1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and goods movement by railways [@zipfHypothesisCaseRailway1946].
@@
 -   $d_{12}^2$ is the distance between $1$ and $2$
--   And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]
+-   And finally $G$, a constant that Stewart "left for future determination" [@stewartDemographicGravitationEvidence1948, p. 34]
 
@@
 
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility popularized by @hansen1959 in the late 1950s.
+Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility popularized by Hansen [@hansen1959] in the late 1950s.
 
@@
 
-From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G.
+From Stewart [@stewartDemographicGravitationEvidence1948], we arrive to 1959 and Walter G.
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
 In this seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
-a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p.
+a generalization of the population-over-distance relationship or *population potential* concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" (p.
 73).
 As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-@hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), and not on the other formulaic contributions and objectionable aspects of "social physics".
-@hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-@hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:
+Hansen [@hansen1959] focused on Stewart's [@stewartDemographicGravitationEvidence1948] *population potential* (expressed in @eq-stewart-population-potential-sum), and not on the other formulaic contributions and objectionable aspects of "social physics".
+Hansen [@hansen1959] recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
+Hansen [@hansen1959] replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:
 
@@
 However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day.
-The omission is that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
-Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly acknowledge or determine it.
+The omission is that between Stewart [@stewartDemographicGravitationEvidence1948] and Hansen [@hansen1959], the proportionality constant $G$ in @eq-stewart-population-potential vanished.
+This constant was not explicitly addressed in Hansen [@hansen1959] and accessibility research continues to evolve without it.
+Since Hansen [@hansen1959], accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly acknowledge or determine it.
 
 The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which undermines the interpretation, communication and comparability of accessibility analysis.
-Those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
+Those reading Hansen [@hansen1959] must recall that Stewart [@stewartDemographicGravitationEvidence1948] had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
 34].
-In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
+In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in Wu and Levinson [@wuUnifyingAccess2020]].
 The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
@@
 $K$ is specified depending on the applied constraint(s).
-In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where to results only match the total volume of interaction, the second being a singly constrained case, and the third a doubly constrained case.
+In the framework introduced in Wilson [@wilson1971], three constrained versions are specified: the first being a case where to results only match the total volume of interaction, the second being a singly constrained case, and the third a doubly constrained case.
 
@@
 Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
-It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been overlooked in almost all subsequent accessibility research.
+It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded Stewart [@stewartDemographicGravitationEvidence1948] and that has been overlooked in almost all subsequent accessibility research.
 Why was this key element of spatial interaction models potentially ignored in accessibility research?
@@
 
-The work of @hansen1959 and @wilson1971 responded to important developments, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
+The work of Hansen [@hansen1959] and Wilson [@wilson1971] responded to important developments, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
 These dictates and needs were far from trivial.
@@
 
-To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959, @wilson1971, or both.
-We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
-As a result of this search, we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, only `r n_w |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
+To illustrate this point, we conducted a bibliographic analysis of the literature that cites Hansen [@hansen1959], Wilson [@wilson1971], or both.
+We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971].
+As a result of this search, we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite Hansen [@hansen1959], only `r n_w |> prettyNum(big.mark = ",")` documents that cite Wilson [@wilson1971], and `r n_b |> prettyNum(big.mark = ",")` that cite both.
 The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
 The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971 since the year 2000.
-
-As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus, visualised in pink in @fig-docs-per-year).
+In particular, we notice the remarkable growth in the number of papers that cite Hansen [@hansen1959] compared to those that cite Wilson [@wilson1971] since the year 2000.
+
+As noted, literature that cites both Wilson [@wilson1971] and Hansen [@hansen1959] are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus, visualised in pink in @fig-docs-per-year).
 After reading the works, we can also discern that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
@@
 
-On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cite both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models.
+On one hand, the stream of literature focused on spatial interaction models inspired by Wilson [@wilson1971] and which cite both Hansen [@hansen1959] and Wilson [@wilson1971], tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models.
 These works treat separate spatial interaction and accessibility as a seperate but related phenomenon.
-Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
+Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of Hansen's [@hansen1959] model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
-Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
-Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+Some authors have explored this relationship, for instance as in Fotheringham [@fotheringhamSpatialCompetitionAgglomeration1985] who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
+Other works used both Hansen [@hansen1959] and Wilson's [@wilson1971] framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with Fotheringham's [@fotheringhamSpatialCompetitionAgglomeration1985] demonstration [e.g., @beckers2022incorporating].
+Other works departed from Hansen's [@hansen1959] definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
 
@@
 
-On the other hand, there is another subset of literature that cite both @hansen1959 and @wilson1971 that is accessibility-focused.
-We categorise their citation of @wilson1971 for three general reasons.
-Firstly, a group of these works cite @wilson1971 as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+On the other hand, there is another subset of literature that cite both Hansen [@hansen1959] and Wilson [@wilson1971] that is accessibility-focused.
+We categorise their citation of Wilson [@wilson1971] for three general reasons.
+Firstly, a group of these works cite Wilson [@wilson1971] as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 These works do not necessarily comment on spatial interaction explicitly.
-Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in @wilson1971 with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the @wilson1971 paper briefly touches on the concept.
-However, in some of this literature, @hansen1959 and @wilson1971 are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature.
-Thirdly, there is a group of accessibility-focused works that interpret @hansen1959's model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
+Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in Wilson [@wilson1971] with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the Wilson [@wilson1971] paper briefly touches on the concept.
+However, in some of this literature, Hansen [@hansen1959] and Wilson [@wilson1971] are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature.
+Thirdly, there is a group of accessibility-focused works that interpret Hansen's [@hansen1959] model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
 This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
@@
 
-Lastly, as an extension of the third reason within this group, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 use @wilson1971's balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility.
+Lastly, as an extension of the third reason within this group, only the works of Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] use Wilson's [@wilson1971] balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility.
 These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
@@
 
-As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
+As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since Hansen [@hansen1959] and Wilson [@wilson1971].
 This may explain why the constraints and associated balancing factors of spatial interaction models did not cross over to accessibility analysis.
@@
 **The unconstrained case** with variants $V_i^0$ and $M_i^0$.
-This case is equivalent to @hansen1959's formulation (first variant) and @reilly1929methods's market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'.
+This case is equivalent to Hansen's [@hansen1959] formulation (first variant) and Reilly's [@reilly1929methods] market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'.
 In this case, no constraints are introduced to ensure that values of marginals are preserved.
@@
 **The total constrained case** with variant $V_i^T$ and variant $M_j^T$.
-The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
+The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles Hansen's [@hansen1959] formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
 In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
@@
 The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998].
-Both variants can also be defined using either @hansen1959's or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$.
+Both variants can also be defined using either Hansen's [@hansen1959] or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$.
 These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility).
@@
 
+`\blandscape`{=latex}
+
 ::: {=latex}
@@
 Table 1: Summary of constrained accessibility measure types and interpretations
+
+`\elandscape`{=latex}
 
@@
 Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
-As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.
+As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see Páez et al.
+[@paez2012measuring].
 
@@
              columns = 2:4) |>
-  cols_width(
-    stub() ~ px(60),
-    everything() ~ px(180)
-  ) |>
+  #cols_width(
+  #  stub() ~ px(60),
+  #  everything() ~ px(180)
+  #) |>
   grand_summary_rows(
@@
   tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
+              table.font.size = px(8)) |>
   as_latex()
@@
 
-Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
+Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in [@cliff_evaluating_1974]).
 Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:
@@
              columns = 2:4) |>
-  cols_width(
-    stub() ~ px(60),
-    everything() ~ px(180)
-  ) |>
+  #cols_width(
+  #  stub() ~ px(60),
+  #  everything() ~ px(180)
+  #) |>
   tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
+              table.font.size = px(8)) |>
   grand_summary_rows(
@@
 
-Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.
+Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of Hansen's [@hansen1959] accessibility measure.
 Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
@@
 Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure.
-This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974, and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods].
+This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in Harris [@harris_market_1954] and Vickerman [@vickermanAccessibilityAttractionPotential1974], and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods].
 This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
@@
              columns = 2:4) |>
-  cols_width(
-    stub() ~ px(90),
-    everything() ~ px(150)
-  ) |>
+  #cols_width(
+  #  stub() ~ px(90),
+  #  everything() ~ px(150)
+  #) |>
   grand_summary_rows(
@@
 Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied.
-Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.
+Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023]), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.
 
@@
 
-@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
+Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023] have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
 As in the total constrained opportunity case, $\kappa^S_{ij}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
-The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
-These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
+The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023], because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
+These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of Shen [@shen1998] and Luo and Wang [@luo2003]:
 
@@
 The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
-Using these these balancing constants, we can calculate the singly constrained opportunity accessibility:
+Using these these balancing constants, we can calculate the singly constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).
 
@@
 
-  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
+  cols_label(O_i = md("Population (10k)"),
              V_opp_i_1 = md("units: *physicians*"),
@@
              columns = 2:5) |>
-  cols_width(
-    stub() ~ px(60*0.8),
-    everything() ~ px(180*0.8)
-  ) |>
+  #cols_width(
+  #  stub() ~ px(60*0.8),
+  #  everything() ~ px(180*0.8)
+  #) |>
   grand_summary_rows(
@@
 
-  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
+  cols_label(O_i = md("Population (10k)"),
              v_opp_i_1 = md("units: *physicians per capita*"),
@@
              columns = 2:5) |>
-  cols_width(
-    stub() ~ px(60*0.8),
-    everything() ~ px(180*0.8)
-  ) |>
+  #cols_width(
+  #  stub() ~ px(60*0.8),
+  #  everything() ~ px(180*0.8)
+  #) |>
   #grand_summary_rows(
@@
   tab_options(column_labels.font.size = "small",
-              table.font.size = px(10))  |>
+              table.font.size = px(8))  |>
   as_latex()
@@
 Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
-In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
+In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
 
@@
 $$
-\sum_j M^S_{ji} =  O_i \\
+\sum_j M^S_{ji} =  O_i 
 $$ {#eq-population-constraint}
@@
 In this work, we examined the historical and mathematical commonalities between spatial interaction models and place-based accessibility measures.
-As accessibility research evolved largely influenced by @hansen1959, researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
+As accessibility research evolved largely influenced by Hansen [@hansen1959], researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
 Firstly, we have demonstrated that by reintroducing such constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
@@
 Secondly, the constraining constants, depending on the case (i.e., total, singly- or doubly- constrained), restrict the degree of _potential_, linking accessibility (the potential for spatial interaction) with access (spatial interaction) on the same continuum based on the constraint used.
-Lastly, we discussed how popular measures such as the one used in @hansen1959 and the 2SFCA link into the family of accessibility measures.
+Lastly, we discussed how popular measures such as the one used in Hansen [@hansen1959] and the 2SFCA link into the family of accessibility measures.
 We summarise the family of accessibility measures, as follows.
@@
 We then introduce the singly constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
-It is also mathematically equivalent to the spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
+It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023].
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
@@
 
-In summary, building on @wilson1971's foundational work, this paper proposed a unified framework for accessibility that is able to account for competition.
+In summary, building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility that is able to account for competition.
 By reintroducing Wilson's proportionality constant, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.

### 2025-05-28 paper_plos/title_page.qmd
move to the plos template
most recent paper now in `paper_plos`

--- 2025-05-26 paper/manuscript.qmd
@@
   pdf:
-    keep-tex: true
     documentclass: article
-    header-includes: |
-      \usepackage{float}
-      \floatplacement{table}{H}
-      \floatplacement{figure}{H}
-bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-#includes:
-#      in_header: 'gt_packages.sty'
-      #\usepackage{float}
 
@@
 ---
-
-```{r knitr-setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  cache = FALSE,
-  warning = FALSE,
-  message = FALSE,
-  fig.pos = 'H', #to print figures on the line specified in knitted output
-  comment = '', 
-  out.width = "1\\linewidth")
-```
-
-```{r load-packages, include=FALSE, cache=FALSE}
-library(bibliometrix) # Comprehensive Science Mapping Analysis
-library(dplyr) # A Grammar of Data Manipulation
-library(flextable) # Functions for Tabular Reporting
-library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
-library(ggraph) # An Implementation of Grammar of Graphics for Graphs and Networks
-library(glue) # Interpreted String Literals
-library(gt) # Easily Create Presentation-Ready Display Tables
-library(here) # A Simpler Way to Find Your Files
-library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
-library(tidytext) # Text Mining using 'dplyr', 'ggplot2', and Other Tidy Tools
-```
-
-```{r options}
-suppressMessages(library(gt))
-options(gt.html_tag_check = FALSE)
-```
-
-# Introduction
-
-<!-- SUMMARY *AND OUTLINE FOR ABSTRACT*: Transportation planning, historically, has been mobility-oriented.
-Accessibility, however, should be the primary focus, and institutionalized mobility-based transportation practice has treated accessibility as its proxy.
-This has had problematic consequences.
-Conceptually, accessibility is a more complete and comprehensive evolution of mobility: from movement, to the potential movement to destinations.
-As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are too many definitions and uncertainty in meaning of the outputs.
-This work's goal is to enhance interpretation of values, clarify the concept of accessibility, and unify other measures of accessibility such as the popular Hansen-type measure the 2SFCA.
-
-To do so, this work demonstrates that accessibility and mobility oriented models (i.e., gravity-based accessibility and the spatial interaction model) share the same roots in spatial interaction modelling, and a family of accessibility measures can be specified within the spatial interaction modelling principles.
-In demonstrating these developments, three significant contributions emerge: A) we demonstrate numerically how constrained accessibility can be specified, and doing so brings interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'.
-B) we provide a clarifying framework for understanding the concept of 'potential' through constraints.
-Namely, demonstrating accessibility's maximum potential is captured by the the 'total constraint' (constrained version of Hansen-type measure), a restricted amount of potential is captured by the 'single constraint' (i.e., competition, such as the 2SFCA), and potential is fully restricted in the 'double constraint', effectively restricting potential such that this case is equivalent to access or simply spatial interaction.
-C) Demonstrate how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the family of accessibility measures.
-
-In fewer words: this work makes an important contribution to accessibility literature, by providing accessibility with a more interpretable, clarified and unified definition through the introduction of constraints from spatial interaction modelling, that reintroduce units to enhance interpretability of raw values, clarify the concept of potential, and loops the most popular forms of accessibility measures into the family.
--->
-
-In the early nineteenth century, industrializing cities grew rapidly: so did traffic congestion and the development of transportation planning practice focused primarily on mobility.
-In this newly founded practice, access to destinations was treated as a by-product of movement.
-Following the rise of the automobile and significant investments in transportation infrastructure after World War II, this mobility-oriented approach has led to some problematic outcomes.
-Specifically, the car became seen as the ultimate mobility tool, helping to foster the development of low-density, single-use residential neighborhoods and entrenching an automobility mono-culture within transportation systems [@miller_collaborative_2011; @lavery_driving_2013].
-Decades of planning for this automobility mono-culture, still often characterized by road and highway expansion, have been marked by increased travel costs and environmental burdens, with limited impact on enhancing the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
-In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic, etc.) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].
-
-Accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'.
-Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling].
-While accessibility brings a more holistic understanding of combined transportation and land use systems, incorporating both the explicit consideration of destination as opportunity and the concept of _potential_ [@handyMeasuringAccessibilityExploration1997].
-
-The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
-In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
-However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
-For instance, the diversity of accessibility definitions has been flagged by @vanweeAccessible2016, @handy2020, and @kapatsila_resolving_2023.
-Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including @geursAccessibilityEvaluationLanduse2004, @vanweeAccessible2016, and @ferreiraReenactingMobilityAccessibility2020.
-
-The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
-At a high level, @geursAccessibilityEvaluationLanduse2004 identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
-Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
-How is a practitioner to choose among this myriad options?
-What differences in accessibility scores should matter, and how should they be communicated?
-[see @vanweeAccessible2016; p.
-14].
-
-Here, we seek to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
-In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in @wilson1971 that can be defined using balancing factors that constrains values based on known information about the land-use.
-
-This work aims to offer three contributions: (1) it introduces a family of accessibility measures within the principles of spatial interaction; (2) it formally defines three _constrained_ accessibility measures by reintroducing Wilson-analogous balancing factors.
-These measures are the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure, which are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
-These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
-(3) This work also demonstrates that the introduction of balancing factors makes accessibility measures easier to interpret and to communicate by restoring the measurement units to the resulting raw accessibility values.
-Each zonal and zonal flow value from a constrained accessibility measure is always in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'.
-This is in contrast to conventional accessibility measures, particularly gravity based measures, that yield values in units of 'opportunities weighted by some representation of travel friction'.
-
-<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To achieve these stated objectives, this paper contends that accessibility research must reconnect with its spatial interaction origins.
-Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
-Empirical constraints were embraced by early spatial interaction literature following the work of @wilson1971, but this stream of literature tended to flow separately from research inspired by @hansen1959' accessibility.
-The application of @wilson1971's empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
-However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted.
-We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
-Moreover, without constraints we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with.
-Without a clear sense of measurement units, comparability between accessibility measures, across cities, and transport modes, may also become compromised.
-
-Fortunately, accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
-It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-We continue this work in the following section by tracing the development of accessibility from its origins in spatial interaction: from the Newtonian gravitational expression in @ravensteinLawsMigration1889 through to the seminal accessibility work of @hansen1959.
-We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of @wilson1971.
-Next, we hark back to @wilson1971's spatial interaction models, and use them to derive a family of accessibility measures based on analogous constraints.
-We illustrate members of this family of constrained accessibility measures with a simple numerical example.
-We then conclude by discussing the uses of these measure and their interpretation.
-
-# Newtonian's roots of human spatial interaction research
-
-<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
-The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half.
-From as far back as Henry C.
-Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], a concern with the scientific study of human spatial interaction can be observed.
-It was in this work where Carey stated that "man \[is\] the molecule of society \[and their interaction is subject to\] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
-37-38].
-This statement shows how investigations into human spatial interaction have often been explicitly coloured by the features of Newton's Law of Universal Gravitation, first posited in 1687's *Principia Mathematica* and expressed as in @eq-phys-grav-prop.
-
-$$
-F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
-$$ {#eq-phys-grav-prop}
-
-To be certain, the expression above is one of proportionality, and is also the most famous in all of science.
-It states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$.
-Direct proportionality means that as the product of the masses increases, so does the force.
-Likewise, inverse proportionality means that as the distance increases, the force decreases.
-@eq-phys-grav-prop, however, does not quantify the magnitude of the force.
-To do so, an empirical constant, a.k.a.
-the gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
-In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data.
-Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:
-
-$$
-F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
-$$ {#eq-phys-grav}
-
-Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p.
-783].
-Still, it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant to within 1% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
-
-# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
-<!-- 
-SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s.
-A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
-This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
--->
-Since the 1880s to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance.
-This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-
-Following Carey's *Principles* of 1858, research into human spatial interaction continued in different contexts.
-In the late 1880s, Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
-In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
-As with Carey, these propositions echo Newton's gravitational laws.
-Over time, other researchers discovered similar relationships.
-<!--Rilley does not cite Newton, nor Carey, nor Ravenstein.--> For example, @reilly1929methods formulated a law of retail gravitation, expressed in terms of equal attraction to competing retail destinations that could be understood as 'potential trade territories'.
-Later, Zipf proposed a $\frac{P_1P_2}{D}$ hypothesis for the case of information [@zipfDeterminantsCirculationInformation1946], intercity personal movement [@zipfHypothesisIntercityMovement1946], and goods movement by railways [@zipfHypothesisCaseRailway1946].
-The $\frac{P_1P_2}{D}$ hypothesis stated that the magnitude of flows was proportional to the product of the populations of settlements, and inversely proportional to the distance between them.
-
-A common feature of these early investigations of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav was never considered.
-Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms.
-Reilly's hypothesis was presented in the following form:
-
-$$
-B_a = \frac{(P_a P_T)^N}{D_{aT}^n}
-$$ {#eq-reilly}
-
-\noindent where $B_a$ is the amount of business drawn to $a$ from $T$, $P_a$ and $P_T$ are the populations of the two settlements, and $D_{aT}$ is the distance between them.
-Quantity $N$ was chosen by Reilly in a somewhat *ad hoc* fashion as 1, and he used empirical observations of shoppers to choose a value of $n = 2$.
-
-Zipf, on the other hand, wrote his hypothesis in mathematical form as:
-
-$$
-C^2 = \frac{P_1 P_2}{D_{12}}
-$$ {#eq-zipf}
-
-\noindent where $C$ is the volume of goods exchanged between $1$ and $2$, $P_1$ and $P_2$ are the populations of the two settlements, and $D_{12}$ is the distance between them.
-
-After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
-This may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
-Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
-
-$$
-F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1 M_2}{d_{12}^2} 
-$$ {#eq-stewart-force}
-
-\noindent Where:
-
--   $F$ is the *demographic force*
--   $N_1$ and $N_2$ are the numbers of people of in groups 1 and 2
--   $\mu_1$ and $\mu_2$ are so-called *molecular weights*, the attractive weight of groups 1 and 2
--   $M_1 = \mu_1N_1$ and $M_2 = \mu_2N_2$ are the demographic masses at 1 and 2
--   $d_{12}^2$ is the distance between $1$ and $2$
--   And finally $G$, a constant that Stewart "left for future determination" [-@stewartDemographicGravitationEvidence1948, p. 34]
-
-In addition to demographic force, Stewart defined a measure of the "population potential" of group $2$ with respect to group $1$.
-In other words, the potential number of people from location $2$ that could visit location $1$, as follows:
-
-$$
-V_1 = G\frac{M_2}{d_{12}}
-$$ {#eq-stewart-population-potential}
-
-For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
-
-$$
-V_i = \int\frac{D}{r} ds
-$$ {#eq-stewart-population-potential-integral}
-
-\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
-In @eq-stewart-population-potential-integral, $D ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
-
-$$
-V_i = \sum_j M_jd_{ij}^{-1}
-$$ {#eq-stewart-population-potential-sum}
-
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility popularized by @hansen1959 in the late 1950s.
-
-Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic.
-It had issues with inconsistent mathematical notation.
-<!--For example, in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant p.
-34 and then later as _demographic energy_ on p.
-53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific and racist ideas.
-For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" \[p.
-35\].
-Stewart's ideas about "social physics" soon fell out of favour among social scientists, but not before influencing the nascent field of accessibility research, as detailed next.
-
-# Hansen's gravity-based accessibility to today {#grav-to-today}
-
-<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up and is now cited as the father of accessibility.
-However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day.
-The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it.
-The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient.
-In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed.
--->
-
-From @stewartDemographicGravitationEvidence1948, we arrive to 1959 and Walter G.
-Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
-In this seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
-a generalization of the population-over-distance relationship or *population potential* concept developed by @stewartDemographicGravitationEvidence1948" (p.
-73).
-As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-@hansen1959 focused on @stewartDemographicGravitationEvidence1948's *population potential* (expressed in @eq-stewart-population-potential-sum), and not on the other formulaic contributions and objectionable aspects of "social physics".
-@hansen1959 recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-@hansen1959 replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:
-
-$$
-S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
-$$ {#eq-accessibility}
-
-A contemporary rewriting of @eq-accessibility accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
-
-$$
-S_{i} = \sum_j O_j  f(d_{ij})
-$$ {#eq-accessibility-general}
-
-$S_{i}$ in @eq-accessibility is a measure of the accessibility from site $i$.
-This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost).
-Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.
-
-However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day.
-The omission is that between @stewartDemographicGravitationEvidence1948 and @hansen1959, the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-This constant was not explicitly addressed in @hansen1959 and accessibility research continues to evolve without it.
-Since @hansen1959, accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly acknowledge or determine it.
-
-The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which undermines the interpretation, communication and comparability of accessibility analysis.
-Those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
-34].
-In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in @wuUnifyingAccess2020].
-The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
-For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta m}$ when $f(c_{ij}) = e^{-\beta m}$.
-This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
-The Hansen-style accessibility estimates found in the literature, therefore, are better thought of as ordinal measures of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].
-
-# Wilson's family of spatial interaction models
-
-<!--SUMMARY: Introduce Wilson's spatial interaction model.
-Then discuss how it could relate to gravity analogy -- but is tactically derived using entropy-maximisation, obtaining the interaction variable as a statistical average.
-Then introduce the constraints and cases.
--->
-
-On the other side of the Atlantic, Alan G.
-Wilson was developing related, yet parallel work.
-In his groundbreaking study [@wilson1971], Wilson defined a general spatial interaction model as follows.
-While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling observed spatial interaction:
-$$
-T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
-$$ {#eq-phys-gravity-model}
-
-The model in @eq-phys-gravity-model posits a quantity $T_{ij}$ that represents a value in a matrix of flows of size $n \times m$, that is, between $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
-The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,s,n$ origins and $j=1,s,m$ destinations.
-The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., $W_i^{(1)}$ could be populations, and $W_j^{(2)}$ hectares of park space.
-Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
-In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
-In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
-In other words, it balances the units, in a conceptually similar sense as the gravitational constant in Newton’s Law of Universal Gravitation.
-
-Traditionally, the development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
-But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
-Entropy maximisation in this case achieves stable results as a statistical average that represents the population.
-The approach works by assuming undifferentiated individual interactions, and assessing their probabilities of making a particular journey.
-The result of @eq-phys-gravity-model then is a statistical average [@wilson1971; @seniorGravityModellingEntropy1979].
-
-To ensure that $T_{ij}$ in @eq-phys-gravity-model is in the unit of trips (our unit of origin-destination spatial interaction), additional knowledge about the system is required.
-At the very least, this framework assumes that the total number of trips in the system $T$ is known, and therefore:
-
-$$
-\sum_i\sum_j T_{ij} = T
-$$ {#eq-constraint0-gravitymodel}
-
-Additional information can be introduced.
-For example, when information is available about the total number of trips produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
-
-$$
-\sum_j T_{ij} = O_i
-$$ {#eq-constraint1-gravitymodel}
-
-Alternatively, if there is information available about the total number of trips attracted by each destination, $W_j^{(2)}$ is represented as $D_j$ and the following constraint can be used:
-
-$$
-\sum_i T_{ij} = D_j
-$$ {#eq-constraint2-gravitymodel}
-
-It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints could be imposed on the model at once.
-
-Using information about the system that satisfy these constraints fully, partially or not at all, a family of spatial interaction models can be derived based on @eq-phys-gravity-model.
-$K$ is specified depending on the applied constraint(s).
-In the framework introduced in @wilson1971, three constrained versions are specified: the first being a case where to results only match the total volume of interaction, the second being a singly constrained case, and the third a doubly constrained case.
-
-In the first, @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel do not hold.
-In practical terms, this means that the total number of trips predicted by the model must be equal to sum of all flows from origins i to destinations j.
-The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
-
-$$
-K=\frac{T}{\sum_i\sum_j T_{ij}}
-$$ {#eq-total-flow-balancing-factor}
-
-In the second case, only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold.
-The resulting models are, in Wilson's terms, singly constrained.
-When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained model:
-
-$$
-T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
-$$ {#eq-production-constrained-gravitymodel}
-
-Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$, representing what we know about the system, the spatial interaction outbound flow, i.e., outbound trips produced at $i$.
-Also, there is no longer a single system-wide proportionality constant, but rather a set of proportionality constants (i.e., balancing factors) specific to origins.
-For this model, the balancing factors ensure that @eq-constraint1-gravitymodel is satisfied, meaning that the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$ i.e., the total number of outbound trips.
-Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of trips.
-This model is useful when trips ends are unknown but the number of trips originating from each location is known and the total of these trips represents all trips in the system.
-The balancing factors for the production-constrained model are solved for each origin $A_i$, and according to Wilson are:
-
-$$
-A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$$ {#eq-production-constrained-balancing-factor}
-
-The attraction-constrained model is similar to the production-constrained model as it is also singly constrained but from the perspective of the mass at the destination.
-From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow, i.e., trips attracted at the destination and takes the following form:
-
-$$
-T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
-$$ {#eq-attraction-constrained-gravitymodel}
-
-In this model (@eq-attraction-constrained-gravitymodel), the balancing factors ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$ i.e., the total number of inbound trips to $j$.
-This should hold for all destinations.
-As before, destination-specific proportionality constants (i.e., balancing factors) $B_j$ were derived by Wilson as:
-
-$$
-B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$$ {#eq-attraction-constrained-balancing-factor}
-
-The third case in the family of spatial interaction models is the production-attraction constrained model.
-In this case, both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
-These constraints ensure that the sum of predicted flows from one origin to all destination, and the predicted flows going to one destination from all origins must equal the known mass of the origin $O_i$ and of the destination $D_j$.
-This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly constrained, and takes the following form:
-
-$$
-T_{ij} = A_i B_j O_i D_j f(c_{ij})
-$$ {#eq-doubly-constrained-gravitymodel}
-
-In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by the origin and the trips attracted by the destination.
-There are now two sets of mutually dependent proportionality constants:
-
-$$
-\begin{array}{l}
-A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
-B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
-\end{array}
-$$ {#eq-doubly-constrained-balancing-factors}
-
-Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
-It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded @stewartDemographicGravitationEvidence1948 and that has been overlooked in almost all subsequent accessibility research.
-Why was this key element of spatial interaction models potentially ignored in accessibility research?
-In the next section we aim to address this question.
-
-# Accessibility and spatial interaction modelling: two divergent research streams
-
-<!--
-SUMMARY: Accessibility developed separate from SIM.
-
-SIM focuses on trips and was institutionalized to develop the US Interstate highway systems.
-Many guidelines and practical training was done - a lot of funding.
-SIM development has integrated accessibility in a few ways: 1) within the SIM itself as a balancing factor, 2) as an additional independent variable in the SIM, 3) SIM and HAM in parallel in the context of location allocation problems and within the steps of transportation demand modelling (trip distribution or trip generation).
-
-Accessibility research has separated SIM almost entirely.
-When Wilson is cited: 1) typically when using a different impedance function, 2) when using the SIM to predict trip flows for accessibility input or vice versa, 3) a few exceptional papers that interpret the doubly-constraints as an inverse HAM in the context of competitive accessibility, and 4) it was only until recently, that a singly-constrain was introduced in Soukhov et al.
-2023.
--->
-
-```{r}
-#| label: load-data
-
-load(glue::glue(here::here(),
-                "/data/corpus.rda"))
-```
-
-```{r}
-#| label: calculate-cocMatrix
-#| cache: true
-
-# $A$ is a sparse adjacency matrix with sources cited by each document in the corpus
-A <- cocMatrix(corpus, 
-               Field = "CR", 
-               sep = ";")
-```
-
-```{r}
-#| label: coupling-matrix
-
-# The product of $A$ and its transpose gives the bibliographical coupling by cited references $B = A A^T$
-
-# $B$ is a square symmetric matrix where element $b_{ij}$ is the total number of cited references in common between document $i$ and document $j$.
-A higher value of $b_{ij}$ indicates a stronger relationship between documents $i$ and $j$ (more references in common).
-A value of zero indicates no references in common.
-B <- tcrossprod(A)
-```
-
-```{r}
-#| label: clean-coupling-matrix
-
-# Remove the diagonal (papers cited within a single source)
-B_nd <- B - Matrix::band(B, 0, 0)
-```
-
-```{r}
-#| label: create-coupling-network
-
-#These matrices are sparse symmetric matrices of "adjacencies" between documents in the corpus. Convert `B` to an `igraph` network without the diagonal
-coupling_net <- igraph::graph_from_adjacency_matrix(B, 
-                                            mode = "directed", 
-                                            weighted = TRUE,
-                                            diag = FALSE)
-```
-
-```{r}
-#| label: remove-A-memory
-
-rm(A)
-```
-
-```{r}
-#| label: convert-to-tidygraph
-
-# For convenience, convert the [{igraph}](https://r.igraph.org) object to a [{tidygraph}](https://www.data-imaginist.com/posts/2017-07-07-introducing-tidygraph/) object:
-coupling_net <- tidygraph::as_tbl_graph(coupling_net)
-```
-
-```{r}
-#| label: add-corpus-label-to-coupling-network
-
-# Add the `corpus` variable to `coupling_net`. Remember to activate the nodes.
-coupling_net <- coupling_net |>
-  tidygraph::activate("nodes") |>
-  left_join(tibble::rownames_to_column(corpus, 
-                                       var = "rowname") |> 
-              transmute(rowname, 
-                        corpus = corpus),
-            by = c("name" = "rowname"))
-```
-
-```{r}
-#| label: corpus-summary
-
-summary_coupling_network <- coupling_net |>
-  tidygraph::activate("nodes") |>
-  as_tibble() |>
-  mutate(corpus = factor(corpus)) |>
-  group_by(corpus) |>
-  summarize(n = n())
-
-# Number of documents in each corpus
-# Hansen
-n_h <- summary_coupling_network |> 
-  filter(corpus == "Hansen") |> 
-  pull(n)
-# Wilson
-n_w <- summary_coupling_network |> 
-  filter(corpus == "Wilson") |> 
-  pull(n)
-# Both
-n_b <- summary_coupling_network |> 
-  filter(corpus == "Both") |> 
-  pull(n)
-```
-
-The work of @hansen1959 and @wilson1971 responded to important developments, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
-These dictates and needs were far from trivial.
-In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S.
-Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars (equivalent to over $600 billion in 2023) [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-Spatial interaction modelling was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-Accessibility, at the time, did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: it predicts the less tangible "potential" for spatial interaction with opportunities.
-In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
-
-```{r}
-#| label: plot-docs-per-year
-#| include: false
-
-docs_per_year_plot <- ggplot(data = corpus |> 
-         group_by(corpus, PY) |> 
-         summarize(n = n()),
-         .groups = "drop") + 
-  geom_col(aes(x = PY, 
-               y = n,
-               fill = corpus),
-           color = "black") + 
-  xlab("Publication year") + 
-  ylab("Number of documents") +
-  theme_minimal()
-
-ggsave(filename = glue::glue(here::here(), 
-                             "/figures/docs_per_year_plot.png"))
-```
-
-```{r}
-#| label: fig-docs-per-year
-#| out-width: 70%
-#| fig-cap: "Historical pattern of publication: documents per year."
-
-docs_per_year_plot
-```
-
-To illustrate this point, we conducted a bibliographic analysis of the literature that cites @hansen1959, @wilson1971, or both.
-We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of @hansen1959 and @wilson1971.
-As a result of this search, we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite @hansen1959, only `r n_w |> prettyNum(big.mark = ",")` documents that cite @wilson1971, and `r n_b |> prettyNum(big.mark = ",")` that cite both.
-The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
-The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite @hansen1959 compared to those that cite @wilson1971 since the year 2000.
-
-As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus, visualised in pink in @fig-docs-per-year).
-After reading the works, we can also discern that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
-The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected, and at times, improperly conflated.
-
-On one hand, the stream of literature focused on spatial interaction models inspired by @wilson1971 and which cite both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models.
-These works treat separate spatial interaction and accessibility as a seperate but related phenomenon.
-Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
-99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
-Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating].
-Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-
-<!-- rafa 1: no need to keep all references in each citation block. AS: I'll keep them for now. -->
-
-<!-- rafa 2: no need to keep all references in each citation block. AS: I'll keep them for now. -->
-
-On the other hand, there is another subset of literature that cite both @hansen1959 and @wilson1971 that is accessibility-focused.
-We categorise their citation of @wilson1971 for three general reasons.
-Firstly, a group of these works cite @wilson1971 as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-These works do not necessarily comment on spatial interaction explicitly.
-Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in @wilson1971 with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the @wilson1971 paper briefly touches on the concept.
-However, in some of this literature, @hansen1959 and @wilson1971 are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature.
-Thirdly, there is a group of accessibility-focused works that interpret @hansen1959's model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
-This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-As outlined in preceding sections, we argue interpreting the singly- or doubly- constrained spatial interaction model's balancing factor as accessibility yields output values that are similarly plagued by interpretability issues.
-
-Lastly, as an extension of the third reason within this group, only the works of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 use @wilson1971's balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility.
-These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
-This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region.
-However, these balancing factors were deduced intuitively.
-These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
-This equivalence is only discovered in hindsight, as will be demonstrated in this study.
-These two works also do not discuss other constrained cases that will be addressed in the next section.
-
-# A family of accessibility measures
-
-In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework.
-At the end of this section, Table 1 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.
-
-As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since @hansen1959 and @wilson1971.
-This may explain why the constraints and associated balancing factors of spatial interaction models did not cross over to accessibility analysis.
-This is intriguing since Wilson made an effort to connect developments in spatial interaction modelling to accessibility, noting for instance, the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of balancing factor $A_i$ [@wilson1971 p.
-10]:
-
-$$
-S_i = \frac{1}{A_i} = \sum_j W_j^{(2)} f(c_{ij})
-$$ {#eq-Ai-as-accessibility}
-
-Understanding $S_i$ as the inverse of Wilson's $A_i$ does not uncover any new meaning for $S_i$ itself.
-Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints.
-Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @eq-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).
-
-Hence, we propose stepping back to introduce a revised definition of accessibility: the _constrained_ potential for spatial interaction.
-Once we bring back Wilson's proportionality constant $k$ into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
-
-$$
-V_{ij} = k W_j^{(2)} f(c_{ij})
-$$ {#eq-access-01}
-
-\noindent where $V_{ij}$ is the potential for interaction from $i$ to $j$. The accessibility from origin $i$ can then be summarised as as a partial sum of the potential at $i$:
-
-$$
-V_{i} = k \sum_j W_j^{(2)} f(c_{ij})
-$$ {#eq-accesssibility-01}
-
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
-
-The market potential variant can also be generally defined, which is effectively the transpose of $i$ and $j$ in @eq-access-01 and @eq-accesssibility-01 as follows:
-$$
-M_{ji} = k W_i^{(1)} f(c_{ij})
-$$ {#eq-market-01}
-
-$$
-M_{j} = k \sum_j W_i^{(1)} f(c_{ji})
-$$ {#eq-market-potential-01}
-
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
-
-To detail the anatomy of $V_{ij}$ and $M_{ji}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility illustrates the accessibility analytical framework we propose using a simple 3 zone system.
-Each measure's most disaggregate value is $X_{ij}$, potential for spatial interaction from $i$ to $j$.
-The $X$ is a stand in for the $ij$ values of all the cases and their variants that will be described (i.e., $V_{ij}^0, M_{ji}^0, V_{ij}^T, M_{ji}^T, V_{ij}^S, M_{ji}^S$ and $V_{ij}^D, M_{ji}^D$).
-The single marginals represent the origin-side and destination-side weights of the zones.
-The total marginal represent the sum of a single marginal.
-
-```{r}
-#| label: fig-analytical-device-conc-accessibility
-#| out-width: 70%
-#| fig-cap: "The family of accessibility measures analytical framework: labelling and associating ij flows, zonal weights, the single marginals, and the total marginal."
-
-knitr::include_graphics(glue::glue(here::here(),
-                                   "/figures/access-analytical-device.png"))
-```
-
-As an additional overview, we define four distinct members of the constrained accessibility measure family, all delineated based on their constant $k$, which takes the form of either balancing factors $K^T$, $B_j$, $A_i$ depending on the indicator:
-
-1.
-**The unconstrained case** with variants $V_i^0$ and $M_i^0$.
-This case is equivalent to @hansen1959's formulation (first variant) and @reilly1929methods's market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'.
-In this case, no constraints are introduced to ensure that values of marginals are preserved.
-
-2.
-**The total constrained case** with variant $V_i^T$ and variant $M_j^T$.
-The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles @hansen1959's formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
-In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
-The second variant $M_j^T$ is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _total constrained accessible population measure_.
-In this variant, each zonal accessibility value is a proportion of the total population in the region (maintained by $\hat K^T$), requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible.
-  - When conceptualising _potential_ for spatial interaction--whether with opportunities or population--the total constrained case reflects the lowest level of restriction and hence maximum potential.
-The balancing factors $K^T$ and $\hat K^T$ only ensure that $V_{ij}^T$ and $M_{ji}^T$ values end up matching the _total sum_ of one of the single marginals, not the individual marginals themselves.
-For example, $K^T$ does not guarantee that all the opportunities accessibility at $V_{i=1}^T$ reflect a proportional allocation of the destination mass (i.e., _number_ of opportunities) at $j=1, 2, 3$.
-In some cases, an allocation of opportunities from a destination $j$ to $i=1$ could _exceed_ the number of opportunities at that $j$ (i.e., meaning that destination $j$ is very attractive and reachable for an $i$, relative other flows in the region).
-However, what $K^T$ _does_ ensure is that each $V_{ij}^T$ does not exceed the overall number of opportunities in the region--the total marginal.
-In other words, $V_{i=1}^T$ expresses the number of opportunities that origin $i=1$ could potentially interact with, as drawn from the entire regional opportunity mass, rather than constrained by individual destination totals.
-In this way, the _total constrained_ case reflects the maximum amount of _potential_ while still maintaining interpretable units.
-
-3.
-**The singly constrained case**, with two variants $V_i^S$ (opportunity-constrained) and $M_j^S$ (population-constrained).
-The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998].
-Both variants can also be defined using either @hansen1959's or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$.
-These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility).
-In this way, the singly constrained case reflects a medium amount of potential, restricted by either values fitting the single destination-mass marginal or the single origin-mass marginal.
-  - The first variant $V_i^S$ includes a set of destination-side balancing factors $B_j$ which ensure opportunities (or destination masses) are allocated to each origin $i$ based on the population (i.e, origin mass) of the origin $i$ and travel impedance from that $j$ to all $i$s.
-Hence, $B_j$ ensures that each $V_i^S$ is the product sum of a _proportional share_ of opportunities allocated from each destination $j$ implicitly also being a share of the total opportunities in the region.
-Similar to the total constrained case, each $V_i^S$ and $V_ij^S$ accessibility value is expressed in units of 'opportunities' accessible.
-But unlike the total constrained case, the singly constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
-  - The second variant $M_j^S$ includes a set of opportunity-side balancing factors $A_i$ which ensure population (or origin masses) are allocated to each destination $j$ based on the destination mass of the destination $j$ and all possible travel impedance from that $i$ to all $j$s.
-$A_i$ ensures similar, but transposed, so constraints are satisfied.
-Specifically, each $M_j^S$ represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance.
-Each zonal value is expressed in units of 'population' accessible.
-
-4.
-**The doubly constrained case**.
-It is constrained simultaneously by population (origin masses) and opportunities (destination masses), so each $V_{ij}^D$ (equivalent to $M_{ji}^D$) is expressed in units of 'population-opportunity capacity' that is accessible between $i$ to $j$.
-This case requires the number of opportunities and population to match, e.g., the analyst must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
-By using this one-to-one matching data and the double constraints (i.e., $A_i$ and $B_j$ at once, ensuring both the double constraint maintains both marginals in @fig-analytical-device-conc-accessibility), this case restricts the _potential_ to spatially interact completely.
-In other words, the doubly constrained accessibility values reflect the number of predicted interactions between the opportunities and population, effectively, this case is equivalent formulaically as Wilson's doubly constrained spatial interaction model (i.e., _attraction-production constrained_).
-It can be understood as predicting a value of 'access', and not accessibility.
-
-And as a summary: each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
-
-::: {=latex}
-\input{summary-family-access-measures-table.tex}
-:::
-Table 1: Summary of constrained accessibility measure types and interpretations
-
-## Setup of the simple numeric example
-
-```{r}
-#| label: simple-numerical-example-opportunities-and-population
-id <- c("1", "2", "3")
-O_i <- c(4, 10, 6) 
-D_j <- c(160, 150, 180)
-LU <- data.frame(id, O_i, D_j)
-```
-
-Consider a simple region as shown in @fig-analytical-device-conc-accessibility, with zone IDs 1, 2 and 3.
-Each zone is both an origin $i$ and a destination $j$.
-The following three pieces of information are defined: zonal population and opportunities, zonal cost matrix, and travel impedance functions for three types of travel behaviour.
-In this example, the population are people, the opportunities are physicians, and the region can be described by three possible travel behaviours.
-
-Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone.
-For reference, considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
-The number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025].
-Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
-
-```{r}
-#| label: tbl-small-system-land-use
-#| tbl-cap: "Simple system with three zones (ID 1, 2 and 3). Population is in 10,000 persons and opportunities in number of physicians."
-
-LU |>
-  gt() |>
-  # tab_spanner("{{W[_i^(1)]}}",
-  #             columns = 2) |>
-  # tab_spanner("{{W[_j^(2)]}}",
-  #             columns = 3) |>
-  cols_label(id = "ID (i or j)",
-             O_i = "Population", 
-             D_j = "Opportunities")  |>
-  cols_align(align = "center",
-             columns = 2:3) |>
-  tab_footnote(footnote = md("Population is *Wi^(1)^* when used as a proxy for the mass at the origin, and *Oi* when used as a constraint."),
-               locations = cells_column_labels(columns = O_i)) |>
-  tab_footnote(footnote = md("Opportunities is *Wj^(2)^* when used as a proxy for the mass at the destination, and *Dj* when used as a constraint."),
-               locations = cells_column_labels(columns = D_j)) |>
-  tab_options(table.font.size = px(10))|>
-  as_latex()
-```
-
-```{r}
-#| label: simple-numerical-example-cost
-
-C <- expand.grid(oid = c("1", "2", "3"),
-                  did = c("1", "2", "3")) |>
-  mutate(cost = c(10, 30, 15, 30, 10, 25, 15, 25, 10)) |>
-  mutate(f1 = cost^-3,
-         f2 = cost^-2,
-         f3 = cost^-0.1)
-```
-
-```{r}
-#| label: tbl-small-system-cost
-#| tbl-cap: "Cost matrix for system with three zones (travel time in minutes)."
-
-C |> 
-  select(oid, did, cost) |>
-  tidyr::pivot_wider(names_from = did, values_from = cost) |>
-  gt() |>
-  tab_spanner(label = "Destination ID",
-              columns = 2:4) |>
-  cols_label(oid = "Origin ID")  |>
-  cols_align(align = "center",
-             columns = 2:4) |>
-  tab_options(table.font.size = px(10))|>
-  as_latex()
-```
-
-From both @tbl-small-system-land-use and @tbl-small-system-cost, Zone 3 and 1 can be interpreted as of an urban core with major healthcare institutions and a healthcare cluster on the edge of the city respectively, each with a moderate residing population (with zone 3 having both a higher population and physician count).
-Zone 2, can be interpreted as a more distant bedroom community, with a relatively high population and fewer physicians.
-In sum, Zones 1 and 3 are more proximate to each other than to Zone 2, and together match Zone 2’s population while offering more than twice the physician availability.
-
-And lastly, in @eq-travel-behaviour-scenarios we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities.
-Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
-A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility.
-Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
-As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.
-
-$$
-\begin{array}{l}
-f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
-f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
-f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
-\end{array}
-$$ {#eq-travel-behaviour-scenarios}
-
-Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups, can be substituted into our simple example, depending on the research question.
-The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behaviour groups and members of the family of accessibility measures.
-
-## Unconstrained accessibility
-
-Setting the balancing factor $k$ to 1 or omitting it completely in @eq-access-01 results in the unconstrained accessibility case:
-
-$$
-V^0_{ij} = 1 *W_j^{(2)} f(c_{ij})
-$$ {#eq-unconstrained-access}
-
-In this case, the partial sum of spatial interaction is simply identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
-
-$$
-V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
-$$ {#eq-unconstrained-accessibility}
-
-The sum of the unconstrained accessibility values for each origin $V^0_{i}$ generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), since arbitrarily setting $k$ to 1 (or neglecting $k$ all together) strips the values of any meaningful unit-based interpretation, the units are in 'summed opportunities by some travel impedance value'.
-Moreover, comparisons of these $V^0_{i}$ values across different contexts such as different impedance functions $f(c_{ij})$ or varying number of zones exacerbates this issue as the units between $V^0_{i}$ values change i.e., comparisons between a value of 'summed opportunities by a travel impedance value' to a value of 'summed opportunities by another travel impedance value' are not directly interpretable.
-From this perspective, the raw unconstrained accessibility scores are not intuitively comparable across different contexts and decay functions.
-They should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
-
-Returning to our numeric example, the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in displayed in @tbl-simple-example-unconstrained-accessibility.
-
-```{r}
-#| label: simple-numerical-example-OD-table
-
-OD <- C |>
-  left_join(LU |>
-              select(-D_j),
-            by = c("oid" = "id")) |>
-  left_join(LU |>
-              select(-O_i),
-            by = c("did" = "id"))
-
-```
-
-```{r}
-#| label: tbl-simple-example-unconstrained-accessibility
-#| tbl-cap: "Simple system: unconstrained accessibility."
-
-unc_acc_ij <- OD |>
-  group_by(oid) |>
-  reframe(V_unc_ij_1 = D_j * f1,
-          V_unc_ij_2 = D_j * f2,
-          V_unc_ij_3 = D_j * f3)
-
-unc_acc <- unc_acc_ij |>
-  group_by(oid) |>
-  summarize(V_unc_i_1 = sum(V_unc_ij_1),
-            V_unc_i_2 = sum(V_unc_ij_2),
-            V_unc_i_3 = sum(V_unc_ij_3))
-
-unc_acc |>
-  gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Origin") |>
-  tab_spanner(label = "{{V[_i^0]}}",
-              columns = 2:4,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}", 
-              columns = 2,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 4,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-  # cols_label(V_unc_i_1 = md("units: *f<sub>1</sub>(c<sub>ij</sub>)-weighted physicians*"),
-  #            V_unc_i_2 = md("units: *f<sub>2</sub>(c<sub>ij</sub>)-weighted physicians*"),
-  #            V_unc_i_3 = md("units: *f<sub>3</sub>(c<sub>ij</sub>)-weighted physicians*"),
-  #            .fn = md)  |>
-  cols_label(V_unc_i_1 = md("units: *physicians-minute^-3*"),
-             V_unc_i_2 = md("units: *physicians-minute^-2*"),
-             V_unc_i_3 = md("units: *physicians-minute^-0.1*"),
-             .fn = md)  |>
-  cols_align(align = "center",
-             columns = 2:4) |>
-  cols_width(
-    stub() ~ px(60),
-    everything() ~ px(180)
-  ) |>
-  grand_summary_rows(
-    columns = c(V_unc_i_1, V_unc_i_2, V_unc_i_3),
-    fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  as_latex()
-```
-
-As the different impedance functions represent different travel behaviours, comparing the raw unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
-For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
-These two values are a product of different impedance functions, making the comparison uninterpretable in absolute terms.
-Likewise, we could compare values within the same travel behaviour scenario across different zones, as they are in the same units, however the issue of unit interpretability will also be apparent.
-Considering the most decaying scenario $f_1(c_{ij})$ and zone 1 (the zone with a healthcare cluster at the edge of the city): zone 1 captures `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer physicians-minute$^{-1}$ than zone 3 (urban core).
-Again, the fundamental uninterpretability of what is a _physicians-minute$^{-1}$_ or _opportunity-weighted-travel-impedance_ unit remains.
-
-While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select impedance functions to facilitate comparison across scenarios (potentially at the expense of accurately reflecting travel behavior), such adjustments may introduce bias.
-Hence, the raw unconstrained accessibility values themselves are challenging to compare due to their units.
-To enable more meaningful comparison, the following sections will detail the introduction of constraining constants to ensure consistent units across scenarios and demonstrate results on the same numeric example.
-
-## Total constrained accessibility
-
-In the total constrained accessibility case, a total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ based on the regional sum of $V^0_i$ and the total population or opportunities in the region.
-Alternatively, we reformulate this case using a proportional allocation constant, which allocates opportunities (or population) proportionally based on the the travel impedance and the total population or opportunities in the region.
-In both formulations, all zonal values become a proportion of a known system total, be it the regional opportunities or regional population depending on the variant.
-
-We define two variants for this case: (a) $V_i^T$ where accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and (b) $M_j^T$, where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total number of population and to be interpreted as constrained 'market potential'.
-
-### Total constrained accessible opportunities: Hansen's accessibility with a total constraint
-
-Unlike in @eq-access-01, the proportionality constant $k$ is retained. For the total constrained case, it is represented as $K^T$:
-
-$$
-V^T_{ij} = K^T  W_j^{(2)}  f(c_{ij})
-$$ {#eq-total-constrained-access}
-
-In this way, the total constrained accessibility measure now becomes Hansen's accessibility with a balancing factor $K^T$:
-
-$$
-V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T  V^0_i
-$$ {#eq-total-constrained-accessibility}
-
-Imagine that the only system known is the total number of opportunities $D$ in the region.
-Accordingly, the constant we impose in this case ensures the regional sum of total constrained accessibility is equal to the total number of opportunities as follows:
-$$
-\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
-$$ {#eq-total-constraint-accessibility}
-
-This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel, congruent with Wilson's framework.
-Given the total number of opportunities in the region, we can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility to solve for $K^T$:
-
-$$
-K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$ {#eq-total-opportunity-balancing-factor}
-
-Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974).
-Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:
-
-$$
-V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j D\frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_{ij}^T$ is dimensionless:
-
-$$
-\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $D$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T D$).
-The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region $D$.
-$\kappa_{ij}^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-This is why we refer to $\kappa_{ij}^T$ as a proportional allocation factor.
-On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.
-
-Referring back to our simple numeric example, balancing factor $K^T$ for the most decay travel behaviour scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
-
-$$
-\begin{array}{l}
-K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})}\\
-K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + s + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
-}\\
-K^T = \frac{490}{0.6233422} \\
-K^T = 786.085
-\end{array}
-$$
-
-```{r}
-# Calculate the balancing factors/proportionality constants for each of three impedance functions
-k_tot <- unc_acc_ij |> 
-  summarize(k1 = sum(LU$D_j)/sum(V_unc_ij_1), # f_1
-            k2 = sum(LU$D_j)/sum(V_unc_ij_2), # f_2
-            k3 = sum(LU$D_j)/sum(V_unc_ij_3)) # f_3
-```
-
-Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
-$\kappa_{ij}^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
-
-```{r}
-#| label: tbl-simple-example-total-opportunity-accessibility
-#| tbl-cap: "Simple system: total constrained accessible opportunities."
-
-tot_acc_ij <- unc_acc_ij |>
-  mutate(V_tot_ij_1 = k_tot$k1 * V_unc_ij_1,
-         V_tot_ij_2 = k_tot$k2 * V_unc_ij_2,
-         V_tot_ij_3 = k_tot$k3 * V_unc_ij_3)
-
-tot_acc <- tot_acc_ij |>
-  group_by(oid) |>
-  summarize(V_tot_i_1 = sum(V_tot_ij_1),
-            V_tot_i_2 = sum(V_tot_ij_2),
-            V_tot_i_3 = sum(V_tot_ij_3))
-
-# tot_acc_containing_kappa <- unc_acc_ij |>
-#   mutate(V_tot_ij_1 = k_tot$k1 * V_unc_ij_1,
-#          V_tot_ij_2 = k_tot$k2 * V_unc_ij_2,
-#          V_tot_ij_3 = k_tot$k3 * V_unc_ij_3) |>
-#   left_join(LU |> rename("oid" = "id") |> select(-c("O_i")), by = "oid") |>
-#   group_by(oid) |>
-#   summarize(V_tot_i_1 = sum(V_tot_ij_1),
-#             V_tot_i_2 = sum(V_tot_ij_2),
-#             V_tot_i_3 = sum(V_tot_ij_3),
-#             D_j = first(D_j)) |>
-#   mutate(kappa_tot_i_1 = V_tot_i_1/D_j, 
-#          kappa_tot_i_2 = V_tot_i_2/D_j, 
-#          kappa_tot_i_3 = V_tot_i_3/D_j)
-
-tot_acc |>
-  select(oid, starts_with("V_")) |>
-  gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Origin") |>
-  tab_spanner(label = "{{V[_i^T]}}",
-              columns = 2:4,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 2,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 4,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-  cols_label(V_tot_i_1 = md("units: *physicians*"),
-             V_tot_i_2 = md("units: *physicians*"),
-             V_tot_i_3 = md("units: *physicians*"))  |>
-  cols_align(align = "center",
-             columns = 2:4) |>
-  cols_width(
-    stub() ~ px(60),
-    everything() ~ px(180)
-  ) |>
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  grand_summary_rows(
-    columns = c(V_tot_i_1, V_tot_i_2, V_tot_i_3),
-    fns = list(label = "Sum", fn = "sum")) |> 
-  as_latex()
-```
-
-In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
-Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_{ij}^T$.
-
-Considering the highest decay scenario ($f_1(c_{ij})$), zone 1 (a healthcare cluster at the edge of the city) captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`) like in the unconstrained accessibility case.
-However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
-Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
-
-One can also directly compare values at a specific zone due to the consistent units.
-For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case.
-However, the difference between travel behaviour scenarios differ in direction.
-Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively.
-Why?
-$\kappa_{ij}^T$ ensures proportional allocation for each travel behaviour scenario.
-Meaning, while the unconstrained accessibility increases, $\kappa_{ij}^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
-As the decay behaviour decreases, more opportunities are accessible for all zones.
-In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones.
-However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).
-
-Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @hansen1959's accessibility measure.
-Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
-This is true for other constrained cases of the family of accessibility measures.
-
-### Total constrained accessible population: Reilly's potential trade territories with a total constraint
-
-Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure.
-This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 and @vickermanAccessibilityAttractionPotential1974, and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods].
-This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
-Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination.
-However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form.
-To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced.
-To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
-
-$$
-M^T_{ji} = \hat K^T  W_i^{(1)} f(c_{ij}) = \hat K^T  M_{ji}^0
-$$ {#eq-total-constrained-market}
-
-\noindent with $M_j^0$ being the $i$ $j$ transpose of $V^0_{ij}$:
-
-$$
-M_j^0 = W_i^{(1)} f(c_{ij})
-$$ {#eq-unconstrained-market}
-
-The total constrained accessible population measure (market potential) then becomes:
-
-$$
-M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
-$$
-
-Where we impose the total system known as a constraint, i.e., that the total market potential equals the total population $O$ in the region:
-
-$$
-\sum_j M^T_{j} = \sum_i\sum_j M^T_{ji} = O
-$${#eq-total-constraint-market}
-
-Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $\hat K^T$, we obtain:
-
-$$
-\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ji}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
-$${#eq-total-population-balancing-factor}
-
-The constrained market potential then takes the following form:
-$$
-M^T_j = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = O \sum_i \frac{W_i^{(1)} f(c_{ij})}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
-$$
-
-Where the following $\hat\kappa_{ji}^T$ proportional allocation factor is dimensionless, reflecting the proportional allocation of population in the region $O$ (i.e., $M^T_j = \sum_i \hat \kappa_{ji}^T O$)
-$$
-\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$
-
-Returning back to the numerical example, the balancing factor $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
-
-```{r}
-# Calculate the balancing factors/proportionality constants for each of three impedance functions
-unc_market_ij <- OD |>
-  mutate(M_unc_ij_1 = O_i * f1,
-         M_unc_ij_2 = O_i * f2,
-         M_unc_ij_3 = O_i * f3)
-
-k_tot_pop <- unc_market_ij |> 
-  summarize(k1 = sum(LU$O_i)/sum(M_unc_ij_1), # f_1
-            k2 = sum(LU$O_i)/sum(M_unc_ij_2), # f_2
-            k3 = sum(LU$O_i)/sum(M_unc_ij_3)) # f_3
-```
-
-```{r}
-#| label: tbl-simple-example-total-population-accessibility
-#| tbl-cap: "Simple system: total constrained accessible population."
-
-tot_pop_ij <- unc_market_ij |>
-  mutate(M_tot_ij_1 = k_tot_pop$k1 * M_unc_ij_1,
-         M_tot_ij_2 = k_tot_pop$k2 * M_unc_ij_2,
-         M_tot_ij_3 = k_tot_pop$k3 * M_unc_ij_3)
-
-tot_market <- tot_pop_ij |>
-  group_by(did) |>
-  summarize(M_tot_i_1 = sum(M_tot_ij_1),
-            M_tot_i_2 = sum(M_tot_ij_2),
-            M_tot_i_3 = sum(M_tot_ij_3))
-
-tot_market |>
-  select(did, starts_with("M_")) |>
-  gt(rowname_col = "did") |>
-  tab_stubhead(label = "Destination") |>
-  tab_spanner(label = "{{M[_i^S]}}",
-              columns = 2:4,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 2,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 4,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-  cols_label(M_tot_i_1 = md("units: *population in 10,000s*"),
-             M_tot_i_2 = md("units: *population in 10,000s*"),
-             M_tot_i_3 = md("units: *population in 10,000s*"))  |>
-  cols_align(align = "center",
-             columns = 2:4) |>
-  cols_width(
-    stub() ~ px(90),
-    everything() ~ px(150)
-  ) |>
-  grand_summary_rows(
-    columns = c(M_tot_i_1, M_tot_i_2, M_tot_i_3),
-    fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  as_latex()
-```
-
-Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility).
-In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
-Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
-However, similar to the total constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
-As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
-Overall: like the total constrained accessible opportunities case, this variant allows for the interpretation of both ordinal and interval comparisons of the raw values themselves.
-
-## Singly constrained accessibility
-
-Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied.
-Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in @soukhovIntroducingSpatialAvailability2023), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.
-
-Unlike the total constraint (i.e., @eq-total-constraint-accessibility), the single constraint--as will be defined in @eq-opportunity-constraint and @eq-population-constraint for the first and second variants-- incorporates additional information.
-In the opportunities-accessible variant, the associated balancing factor constrains the 'potential' for spatial interaction by ensuring that only a proportional amount of opportunities at each destination are allocated to 'demanding' origins.
-This allocation is informed by demand, or the relative amount of population, and associated travel impedance connecting the zones.
-In the population-accessible variant (i.e., the $i$ -> $j$ transpose of the first variant), population at each origin is proportionally allocated to destinations based on the share of opportunities and travel impedance.
-
-In both variants, the singly constrained accessibility measure introduces population-based (or opportunity-based) competition at the zonal level, unlike the total constraint, which more simply allocates a fixed regional total of opportunities (or population depending on the variant).
-In sum, all singly constrained zonal accessibility values are both a proportion of the known regional opportunity (or population) total _and_ a sum of a _balanced_ proportion of opportunities allocated from each destinations (or population allocated from each origin).
-Each zonal value remains in units of opportunities accessible (or population accessible).
-
-### Singly constrained accessible opportunities: spatial availability
-
-To demonstrate this variant formulaically, we begin with the opportunity constraint (@eq-opportunity-constraint) as our known piece of information.
-Namely, the sum of accessible opportunities from a destination should equal the number of opportunities $D_j$ at that destination.
-As the number of opportunities at each $j$ are known, it is represented as $D_j$ instead of $W_j^(2)$ as in the total constrained case.
-This constraint should hold for all destinations in the region.
-This is comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
-
-$$
-\sum_i V^S_{ij} =  D_j
-$$ {#eq-opportunity-constraint}
-
-The underlying spatial interaction model is now the attraction-constrained model in @eq-attraction-constrained-gravitymodel, and our accessibility measure becomes:
-
-$$
-V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
-$$ {#eq-opportunity-constrained-accessibility}
-
-\noindent where $W_i^{(1)}$ is a measure of the mass at origin $i$ (i.e., the opportunity-seeking population). The corresponding balancing factor, as per Wilson, is:
-
-$$
-B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
-$$ {#eq-opportunity-constrained-proportionality-constants}
-
-Introducing the balancing factor in @eq-opportunity-constrained-accessibility, we obtain:
-
-$$
-V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$$ {#eq-opportunity-constrained-accessibility-with-balancing-factor}
-
-Further, we define the following proportional allocation factor:
-
-$$
-\kappa^S_{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
-$$ {#eq-opportunity-constrained-proportional-allocation-factor}
-
-After this, it is possible to rewrite @eq-opportunity-constrained-accessibility-with-balancing-factor as an origin summary expression of proportionally allocated known opportunities (i.e., $D_j$).
-
-$$
-V^S_{i} = \sum_j \kappa^S_{ij} D_j
-$$ {#eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
-
-@soukhovIntroducingSpatialAvailability2023 have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As in the total constrained opportunity case, $\kappa^S_{ij}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
-The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by @soukhovIntroducingSpatialAvailability2023, because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
-These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of @shen1998 and @luo2003:
-
-$$
-v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
-$${#eq-opportunity-constrained-accessibility-per-capita}
-
-<!-- To reinforce this point, @eq-opportunity-constraint also implies that the total constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.
--->
-<!-- Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible *and* available to location $i$ out of the total number of opportunities in the system.
--->
-
-Returning to the simple numeric example, the opportunity-constrained case would yield the following $B_{j}$ for $f_1(c_{ij})$:
-
-$$
-\begin{array}{l}
-B_{j} = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
-B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
-B_{2} =  \frac{1}{\frac{4}{30^3} + \frac{10}{10^3} + \frac{6}{25^3}} = 94.9474\\
-B_{3} =  \frac{1}{\frac{4}{10^3} + \frac{10}{25^3} + \frac{6}{10^3}} = 93.9850
-\end{array}
-$$
-
-```{r}
-B_j <- OD |> 
-  group_by(did) |>
-  summarize(B_j_1 = 1/sum(O_i * f1),
-            B_j_2 = 1/sum(O_i * f2),
-            B_j_3 = 1/sum(O_i * f3))
-# B_j_longer <- B_j
-# colnames(B_j_longer) <- c("ID","B_1","B_2","B_3")
-# B_j_longer |> as.data.frame()  |> gt(rownames_to_stub=FALSE)
-```
-
-The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
-Using these these balancing constants, we can calculate the singly constrained opportunity accessibility:
-
-```{r}
-#| label: tbl-simple-example-attraction-constrained-accessibility
-#| tbl-cap: "Simple system: singly constrained accessible opportunities."
-
-opp_acc_ij <- OD |>
-  left_join(B_j,
-            by = "did") |>
-  mutate(k_opp_ij_1 = B_j_1 * O_i * f1 /3,
-         k_opp_ij_2 = B_j_2 * O_i  * f2 /3,
-         k_opp_ij_3 = B_j_3 * O_i  * f3 /3,
-         V_opp_ij_1 = B_j_1 * O_i * D_j * f1,
-         V_opp_ij_2 = B_j_2 * O_i  * D_j * f2,
-         V_opp_ij_3 = B_j_3 * O_i  * D_j * f3)
-
-opp_acc <- opp_acc_ij |>
-  group_by(oid) |>
-  summarize(O_i = first(O_i) |> as.character(), 
-            V_opp_i_1 = sum(V_opp_ij_1),
-            V_opp_i_2 = sum(V_opp_ij_2),
-            V_opp_i_3 = sum(V_opp_ij_3), 
-            .groups = "drop")
-
-opp_acc_with_ks <-  opp_acc_ij |>
-  group_by(oid) |>
-  summarize(O_i = first(O_i) |> as.character(), 
-            k_opp_ij_1 = sum(k_opp_ij_1),
-            k_opp_ij_2 = sum(k_opp_ij_2),
-            k_opp_ij_3 = sum(k_opp_ij_3))
-
-opp_acc |>
-  #select(oid, O_i, starts_with("V_")) |>
-  gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Origin") |>
-
-  tab_spanner(label = "{{V[_i^S]}}",
-              columns = 3:5,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 4,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 5,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-
-  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
-             V_opp_i_1 = md("units: *physicians*"),
-             V_opp_i_2 = md("units: *physicians*"),
-             V_opp_i_3 = md("units: *physicians*"),
-             )  |>
-  cols_align(align = "center",
-             columns = 2:5) |>
-  cols_width(
-    stub() ~ px(60*0.8),
-    everything() ~ px(180*0.8)
-  ) |>
-  grand_summary_rows(
-    columns = c(V_opp_i_1, V_opp_i_2, V_opp_i_3),
-    fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  as_latex()
-```
-
-Imposing the single proportional allocation factor $\kappa^S_{ij}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case.
-The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.
-
-However, unlike the total constrained opportunity case, $\kappa^S_{ij}$ reflects zonal competition based on the mass of the origin, i.e., population.
-Again, consider the highest decay scenario $f_1(c_{ij})$.
-Under this scenario, zone 1 no longer captures a medium amount of physicians as in the total constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
-Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
-
-This difference is due to the single-opportunity factor $\kappa^S_{ij}$'s role in $V_i^S$.
-The only inputs required in the total-opportunity factor $\kappa^T_{ij}$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
-Opportunities are allocated to $i$s, regardless of the mass weights of origin.
-Whereas the single opportunity constraint $\kappa^S_{ij}$ requires the population at $i$ as an input.
-In fact, $\kappa^S_{ij}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
-Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.
-
-Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
-For instance, consider Zone 2 which has the highest population.
-It is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios that this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
-In this sense, other zones are not *that* disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
-Zones 1 and 3 are allocated opportunities at relates similar to their relative population size.
-
-Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
-For instance, Zone 2 has the highest zonal population, representing 50% of the 200,000 regional population.
-However, due to its high relative travel distance from the other zones, its population is less competitive in capturing opportunities in the high-decay travel scenario ($f_1(c_{ij})$): Under this scenario, zone 2 captures almost exclusively opportunities from its own zone.
-However, in $f_3(c_{ij})$, the scenario with unfettered travel cost, Zone 2 captures by far the most number of physicians.
-But in this scenario, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
-Zones 1 and 3 are allocated opportunities at rates similar to their relative population size.
-
-In this way, the consideration of constrained accessibility *per capita* may be clarifying.
-Often, accessibility values are reported as raw scores without the consideration for population.
-But, as we introduced constraints, these constrained accessibility values can be normalized using anything that is relevant to the zone.
-In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
-Notably, these per capita rates are equivalent to the 2SFCA values.
-
-```{r}
-#| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
-#| tbl-cap: "Simple system: singly constrained accessible opportunities per capita."
-
-opp_acc_percapita <- opp_acc |>
-  transmute(
-    oid = id,
-    O_i = O_i,
-    v_opp_i_1 = V_opp_i_1 / (O_i|>as.numeric()),
-    v_opp_i_2 = V_opp_i_2 / (O_i|>as.numeric()),
-    v_opp_i_3 = V_opp_i_3 / (O_i|>as.numeric()))
-
-opp_acc_percapita  |>
-  gt(rowname_col = "oid") |>
-  tab_stubhead(label = "Origin") |>
-
-  tab_spanner(label = "{{v[_i^S]}}",
-              columns = 3:5,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 3,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 4,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 5,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-
-  cols_label(O_i = md("Population (units: *people in 10,000s*)"),
-             v_opp_i_1 = md("units: *physicians per capita*"),
-             v_opp_i_2 = md("units: *physicians per capita*"),
-             v_opp_i_3 = md("units: *physicians per capita*"),
-             )  |>
-  cols_align(align = "center",
-             columns = 2:5) |>
-  cols_width(
-    stub() ~ px(60*0.8),
-    everything() ~ px(180*0.8)
-  ) |>
-  #grand_summary_rows(
-  #  columns = c(v_opp_i_1, v_opp_i_2, v_opp_i_3),
-  #  fns = list(label = "Sum", fn = "sum")) |> 
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10))  |>
-  as_latex()
-```
-
-This simple example was constructed so that the regional average equals `r (opp_acc$V_opp_i_1|>sum()) /(opp_acc$O_i|>as.numeric()|>sum())` physicians per 10,000 people.
-As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
-Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all around 24 physicians accessible per capita).
-
-This trend mirrors how the accessibility values in the total constrained opportunity case stabilises to the $V_i^T$ regional average (e.g., the accessible opportunities allocated to each of the three zones approaches a third of 490 physicians, or 163.33, under the unfettered mobility scenario $f_3(c_{ij})$).
-These patterns make intuitive sense: the balancing factors act as regional and/or zonal averaging mechanisms.
-In distance decay travel behaviour scenarios that are more relatively uniform (i.e., low for all zones like in $f_3(c_{ij})$), what remains is the relative effect of the other variables in the balancing factor.
-In the total constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constrained case, this is the population at a zone relative to the regional population.
-
-### Singly constrained accessible population: market availability
-
-Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
-In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in @wang_inverted_2018.
-
-The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^S_j$ becomes:
-
-$$
-M^S_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
-$$ {#eq-population-constrained-accessibility}
-
-In this case, the measure is singly constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
-
-$$
-\sum_j M^S_{ji} =  O_i \\
-$$ {#eq-population-constraint}
-
-And the corresponding balancing factor, as per Wilson, is:
-
-$$
-A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
-$$ {#eq-population-constrained-proportionality-constants}
-
-Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression:
-
-$$
-M^S_{j} = \sum_i \hat \kappa^S_{ji} O_i
-$$ {#eq-production-constrained-accessibility-with-proportional-allocation-factor}
-
-\noindent with:
-
-$$
-\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
-$$ {#eq-attraction-constrained-proportional-allocation-factor}
-
-As well, the single (population) constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O$) is maintained.
-
-With these constraints, $\frac{M_j^S}{O}$ can be interpreted as the proportion of the total population serviced by location $j$.
-
-For the sake of brevity, we'll move forward onto the doubly constrained case.
-
-## Doubly constrained accessibility
-
-This accessibility case requires zonal populations and opportunities to match one-to-one, like the doubly constrained spatial interaction model.
-In this way, doubly constrained accessibility can be thought as "access" or simply spatial interaction (no potential).
-
-To contextualize this point, the total and singly constrained accessibility measures discussed thus far have used either $O_i$, $D_j$, or the regional sums of either, but never both simultaneously.
-For example, when opportunities $D_j$ are used to constrain @eq-opportunity-constrained-accessibility in the _singly constrained accessible opportunities measure_, the specific mass of the population at origin $i$ demanding only those opportunities at that $j$ is unknown.
-Instead, only the population at $i$ demanding opportunities in the region is known, and this is more generally represented as $W_i^{(1)}$ (as in @eq-opportunity-constrained-proportionality-constants).
-Similarly, when the population $O_i$ is used as a constraint in @eq-population-constrained-accessibility in the _singly constrained accessible population measure_, the mass at the destination is given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants) since only the mass of opportunities at each $j$ is known and information on what opportunities are allocated to what zone is unknown.
-
-By contrast, the double-constrained accessibility case requires that both populations and opportunities match.
-Meaning, opportunities at each destination can be accessed by all populations, while at the same time, the population at each origin can be accessed by all opportunities.
-However, this requirement is often unintuitive in traditional accessibility analysis.
-Namely, the distinction between population (origin masses) and opportunities (destination masses) typically represent different entities without a shared unit of measurement.
-On the population side, we usually count people; on the opportunity side, we may be referring to physicians, clinics, grocery stores, schools, parks, or libraries.
-In a few cases, a one-to-one correspondence or their own capacities at which they interact and are interacted with may exist e.g., one person with one job.
-Another similar opportunity type example is healthcare, e.g., one person and one unit of capacity, e.g.
-a vaccine shot.
-
-A doubly constrained approach to accessibility calculation needs a one-to-one relationships between population and opportunities to be present.
-Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality):
-
-$$
-\sum_i O_i = \sum_j D_j
-$$ {#eq-opportunity-population-equality}
-
-As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$ as shown in @eq-total-constraint-accessibility.
-
-As the doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model, as shown in @eq-doubly-constrained-gravitymodel, $V_{i}^D$ is as follows:
-
-$$
-V_{ij}^D = A_i B_j O_i D_j f(c_{ij})
-$$ {#eq-doubly-constrained-accessibility}
-
-\noindent where the corresponding balancing factors $A_i$ and $B_j$, as per Wilson, are:
-
-$$
-\begin{array}{l}
-A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
-B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
-\end{array}
-$$
-
-Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
-The values of $B_j$ are used to update the underlying $V_{ij}^D$ matrix, before calibrating $A_i$.
-This process continues to update $A_i$ and $B_j$ until a convergence criterion is met [see @ortuzar_2011_modelling, p.
-193-195].
-The proportional allocation factor $\kappa_{ij}^D$ would then be:
-
-$$
-\kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
-$$
-
-One could rewrite @eq-doubly-constrained-accessibility as an origin summary expression of proportionally allocated opportunities:
-
-$$
-V^D_{i} = \sum_j \kappa^D_{ij} D_j
-$$ {#eq-doubly-constrained-accessibility-w-paf}
-
-However, $V^D_{i}$ is not wholly helpful, as it will simply equal the origin-mass marginal.
-In this way, only $V^D{ij}$ values are interpretable.
-Furthermore, unlike in the total and singly constrained cases, the doubly-constrained case does not have an interpretable per capita version.
-For instance, representing $V_{ij}^D$ per capita is not meaningful, as the value *already* matches population to opportunities.
-Following this logic, the market potential form $M^D_{ij}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: i.e., the opportunities accessed from $j$ at an $i$ vs.
-the population accessed from $i$ at a $j$.
-The inputs of 'opportunities accessed' and 'accessed population' can already be interpreted as inherently being sensitive to both opportunities and population.
-
-```{r}
-#| label: simple-numerical-example-doubly constrained
-
-# Same as before
-id <- c("1", "2", "3")
-O_i2 <- c(4, 10, 6) #in units of 10,000 people
-#D_j <- c(160, 150, 180) #the old
-#D_j2 <- c(6.530612, 6.122449, 7.346939) #this would be exactly linearly scaled down.
-D_j2 <- c(7, 5, 8) #in units of 10,000 physician-capacity
-LU2 <- data.frame(id, O_i2, D_j2)
-```
-
-To calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be modified.
-Namely, a count of physician _capacity_ per destination is needed instead of just the number of physicians, as used to calculated total and singly constrained cases.
-We also must be able to clearly state that the population is a count of people seeking opportunities at the new capacities, i.e., the population must reflect the _capacity_ of the population to interact with opportunities.
-
-So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
-For the population, we leave this unchanged numerically but theoretically know that each person interacts with one physician capacity (i.e., our opportunities with a _capacity_).
-Hence, the number of opportunities per destination is new: the example is modified such that the physician-capacity at each zone is an approximately scaled version of the number of destination-side physicians at each zone from the unmodified example (@tbl-small-system-land-use).
-To emphasis the new definition of 'provider' as physician capacity, the new system PPR is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
-We keep the same zonal cost matrix, and travel impedance functions for three types of travel behaviour as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
-
-```{r}
-#| label: tbl-small-system-land-use-doubly-constrained-case
-#| tbl-cap: "Modified simple system with three zones reflecting matched population and opportunities. Population is in 10,000 persons and opportunities in 10,000 of physician-capacity."
-
-LU2 |>
-  gt() |>
-  # tab_spanner("{{W[_i^(1)]}}",
-  #             columns = 2) |>
-  # tab_spanner("{{W[_j^(2)]}}",
-  #             columns = 3) |>
-  cols_label(id = "ID (i or j)",
-             O_i2 = "Population", 
-             D_j2 = "Opportunities")  |>
-  cols_align(align = "center",
-             columns = 2:3) |>
-  tab_options(table.font.size = px(10)) |>
-  as_latex()
-```
-
-```{r}
-#| label: simple-numerical-example-OD-table-doubly constrained
-
-OD2 <- C |># mutate(f3=100) |>
-  left_join(LU2 |>
-              select(-D_j2),
-            by = c("oid" = "id")) |>
-  left_join(LU2 |>
-              select(-O_i2),
-            by = c("did" = "id"))
-```
-
-However, despite the modifications to the example, our objective remains the same as in previous cases: to measure accessibility under different travel behavior scenarios.
-Specifically, we aim to quantify the number of _potential_ spatial interactions between the physician-seeking population in each zone $i$ to the physician capacity in a zone $j$ i.e., $V_{ij}^D$.
-The highest decay travel behaviour scenario ($f_1(c_ij)$) is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
-
-```{r, quiet = TRUE}
-doubly_constrained_gravity_model <- function(
-    #CREDIT: I converted the python code from (https://github.com/SadraDaneshvar/Gravity_Model) to R
-  O,  # Origin production weights vector
-  D,  # Destination attraction weights vector
-  impedance_matrix,  # the resulting travel cost impedance matrix
-  error_threshold = 0.001,  # error threshold for stopping condition
-  improvement_threshold = 1e-6  # improvement threshold for stopping condition
-) {
-
-  # Normalize O and D
-  sum_O <- sum(O)
-  sum_D <- sum(D)
-  # cat("Checking production, attraction balancing:\n") #let's disable the messages..
-  # cat("Production: ", sum_O, "\n")
-  # cat("Attraction: ", sum_D, "\n")
-
-  if (sum_O != sum_D) {
-    # cat("Productions and attractions do not balance, scaling to match.\n") #let's disable the messages..
-    if (sum_O < sum_D) {
-      O <- O * (sum_D / sum_O)
-    } else {
-      D <- D * (sum_O / sum_D)
-    }
-  } else {
-    # cat("Production, attraction balancing OK.\n") #let's disable the messages..
-  }
-
-  n <- length(O)  # Number of i
-  T <- sum(O)  # Total trips
-  Ai <- rep(1, n)  # Initialize Ai
-  Bj <- rep(1, n)  # Initialize Bj
-
-  previous_error <- Inf
-  iteration_count <- 0
-  stop_reason <- ""
-
-  while (TRUE) {
-    iteration_count <- iteration_count + 1
-
-    # Update Ai
-    for (i in 1:n) {
-      Ai[i] <- 1 / (sum(Bj * D * impedance_matrix[i, ]) + 1e-9)
-    }
-
-    # Update Bj
-    Bj_new <- rep(1, n)
-    for (j in 1:n) {
-      Bj_new[j] <- 1 / (sum(Ai * O * impedance_matrix[, j]) + 1e-9)
-    }
-
-    # Calculate Tij
-    Tij <- outer(Ai * O, Bj_new * D) * impedance_matrix
-
-    # # Print Ai and Bj for this iteration #let's disable the messages..
-    # cat("\nIteration:", iteration_count, "\n")
-    # cat("Ai:", round(Ai, 3), "\n")
-    # cat("Bj:", round(Bj_new, 3), "\n")
-
-    # Calculate error
-    error <- (sum(abs(O - rowSums(Tij))) + sum(abs(D - colSums(Tij)))) / T
-
-    # Calculate change in error
-    error_change <- abs(previous_error - error)
-
-    # Check stopping conditions
-    if (error < error_threshold) {
-      stop_reason <- "Error threshold met"
-      break
-    } else if (error_change < improvement_threshold) {
-      stop_reason <- "Slow improvement"
-      break
-    }
-
-    previous_error <- error
-    Bj <- Bj_new
-  }
-
-  # report stopping condition and final error threshold #let's disable the messages..
-  #cat("\nStopping Condition: ", stop_reason, "\n")
-  #cat(sprintf("Final Error: %.3f%%\n", error * 100))
-
-  return(list(Ai = Ai, Bj = Bj, Tij = Tij))
-}
-```
-
-```{r}
-#| label: calculating balancing-factors and resulting Tij for each travel behaviour scenario
-
-#isolating impedance
-Impf1 <- C$f1 |> 
-  matrix(nrow = 3, ncol = 3)
-
-Impf2 <- C$f2 |> 
-  matrix(nrow = 3, ncol = 3)
-
-Impf3 <- C$f3 |> 
-  matrix(nrow = 3, ncol = 3)
-# caclulating it
-dc_acc_1 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf1)
-dc_acc_2 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf2)
-dc_acc_3 <- doubly_constrained_gravity_model(O_i2, D_j2, Impf3)
-
-#assigning the balancing factors to their own variables
-A_i_doubly1 <- dc_acc_1$Ai
-B_j_doubly1 <- dc_acc_1$Bj
-
-A_i_doubly2 <- dc_acc_2$Ai
-B_j_doubly2 <- dc_acc_2$Bj
-
-A_i_doubly3 <- dc_acc_3$Ai
-B_j_doubly3 <- dc_acc_3$Bj
-
-#assigning the Tijs to their own variables
-V_dc_1 <- dc_acc_1$Tij
-V_dc_2 <- dc_acc_2$Tij
-V_dc_3 <- dc_acc_3$Tij
-
-#reformatting the matrix into long format, joining it to OD2
-OD2<-OD2 |> mutate(V_dc_ij_1 = c(V_dc_1[,1], V_dc_1[,2], V_dc_1[,3]),
-              V_dc_ij_2 = c(V_dc_2[,1], V_dc_2[,2], V_dc_2[,3]),
-              V_dc_ij_3 = c(V_dc_3[,1], V_dc_3[,2], V_dc_3[,3]))
-
-OD2$k_dc_ij_1 = c((outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,1],
-                   (outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,2],
-                   (outer(A_i_doubly1 * O_i2, B_j_doubly1) * Impf1)[,3]) / 3
-
-OD2$k_dc_ij_2 = c((outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,1],
-                   (outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,2],
-                   (outer(A_i_doubly2 * O_i2, B_j_doubly2) * Impf2)[,3]) / 3
-
-OD2$k_dc_ij_3 = c((outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,1],
-                   (outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,2],
-                   (outer(A_i_doubly3 * O_i2, B_j_doubly3) * Impf3)[,3]) / 3
-
-#Checks, should equal 1.
-# OD2$k_dc_ij_1 |> sum()
-# OD2$k_dc_ij_2 |> sum()
-# OD2$k_dc_ij_3 |> sum()
-```
-
-```{r}
-#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
-#| tbl-cap: "Doubly constrained accessible opportunities assuming highest travel decay in the modified simple system."
-
-OD2 |> 
-  select(c(oid, did, V_dc_ij_1)) |>
-  tidyr::pivot_wider(names_from = did, values_from = V_dc_ij_1) |>
-  group_by(oid) |> 
-  mutate(sum = round(`1` + `2` + `3`, 0)) |> 
-  ungroup() |> 
-  gt() |>
-  tab_spanner(label = "Destination ID",
-              columns = 2:4) |>
-  cols_label(oid = "Origin ID")  |>
-  cols_align(align = "center",
-             columns = 2:5) |>
-  grand_summary_rows(
-    columns = 2:4,
-    fns = list(label = "Sum", fn = "sum")) |>
-
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  as_latex()
-```
-
-As mentioned, accessibility values are typically interpreted as a summary of the proportionally allocated opportunities at each $i$.
-Hence, in interpreting the doubly constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$ values (i.e., the sum of values at all three $j$ destinations for each origin $i$) would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity accessible for Zones 1, 2 and 3 respectively.
-This approximately equal to the number of population at each of these zones.
-Conversely, the market potential $M_j^D$ interpretation of these values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of *opportunities* (physician-capacities accessible) at each of these zones.
-Notice, the mass weight at the origin equals the mass weight of the destination: this is precisely the function of the double constraint.
-In other words, $V_i^D$ is the number of accessed opportunities and $M_j^D$ is the number of population accessed.
-For the other two travel behaviour, identical $V_i^D$ and $M_j^D$ values are calculated, following the same logic.
-Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
-
-For instance, differences in $V_{ij}^D$ values between travel behaviour scenarios are notable.
-These values can be directly compared to discuss mass and distance decay impacts.
-Examining Zone 2 (the bedroom community), @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for this more relatively remote, higher-populated and lower-opportunity rich zone.
-It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
-Following the intuition discussed in the singly constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
-
-```{r}
-#| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
-#| tbl-cap: "Doubly constrained accessible opportunities at Zone 2 for all travel decay groups in the modified simple system."
-
-OD2  |>
-  filter(oid == 2) |>
-  select(c(did, O_i2, D_j2, starts_with("V_"))) |>
-  gt(rowname_col = "did") |>
-  tab_stubhead(label = "Dest.") |>
-
-  tab_spanner(label = "{{V[_{ij}^D]}}",
-              columns = 4:6,
-              level = 2) |>
-  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
-              columns = 4,
-              level = 1) |>
-  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
-              columns = 5,
-              level = 1) |>
-  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
-              columns = 6,
-              level = 1) |>
-  fmt_number(decimals = 3) |>
-
-  cols_label(O_i2 = md("Population at 2 (units: *people in 10,000s*)"),
-             D_j2 = md("Opportunities (units: *capacity in 10,000s*)"),
-             V_dc_ij_1 = md("units: *physician-capacity in 10,000s*"),
-             V_dc_ij_2 = md("units: *physician-capacity in 10,000s*"),
-             V_dc_ij_3 = md("units: *physician-capacity in 10,000s*"),
-             )  |>
-  cols_align(align = "center",
-             columns = 2:6) |>
-  cols_width(
-    stub() ~ px(60*0.5),
-    everything() ~ px(180*0.5)
-  ) |>
-  tab_options(column_labels.font.size = "small",
-              table.font.size = px(10)) |>
-  as_latex()
-```
-
-Recall, accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
-In the doubly constrained case, we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
-However, if what interests readers is the "potential for interaction" in the case population and opportunities to match one-to-one, reframing the investigation may be needed.
-In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ji}^D$).
-In this sense, the doubly constrained case can be thought of as an estimate of _realized_ accessibility or access: reflecting spatial interaction.
-It is also formulaically identical to the doubly constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity', respectively.
-
-As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
-Accessibility, however, is often presented and understood as a zonal summary of *potential* for interaction between origins and destinations that contain inherently different units.
-Arriving at the doubly constrained case through the unconstrained, total, and singly constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
-Namely, by demonstrating that these members of the accessibility family all derive from the same root and can be derived from Wilson's original formulation, this more clearly demonstrates what *potential* is, within the context of spatial interaction modelling.
-
-Namely, *potential* depends on the framing of the masses at the origin and destination, and how similar they are in their units.
-The appropriate constraint should be decided based on the the input data and their similarity.
-In increasing unit similarity from the perspective of opportunity-constraint: the total-constraint can be used if population (origin mass) is not known, the single constraint can be used if population is known and matters to *potential* but it does not match the capacity of opportunities, and lastly, the double constraint is appropriate if population is known, matters and population matches the capacity of opportunities.
-These constraint measures can reflect the opportunities accessible at each zone, but reflects the assumptions embedded in the input data about the potential to spatially interact.
-
-# Conclusions
-
-In this work, we examined the historical and mathematical commonalities between spatial interaction models and place-based accessibility measures.
-As accessibility research evolved largely influenced by @hansen1959, researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
-Firstly, we have demonstrated that by reintroducing such constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
-These values become more intuitive for the purpose of analysis and comparison.
-Secondly, the constraining constants, depending on the case (i.e., total, singly- or doubly- constrained), restrict the degree of _potential_, linking accessibility (the potential for spatial interaction) with access (spatial interaction) on the same continuum based on the constraint used.
-Lastly, we discussed how popular measures such as the one used in @hansen1959 and the 2SFCA link into the family of accessibility measures.
-We summarise the family of accessibility measures, as follows.
-
-We first place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
-We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
-In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if population-competition is not relevant and one is interested in capturing the maximum _potential_; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, *without* considering the population size at the origins.
-<!-- A.S.
-NOTE: Hansen's measure would converge to some global total if the decay effect was low.
-Imagine everyone can teleport.
-In the unconstrained case: all zones can reach all opportunities.
-In the total constrained case: all zones can reach all opportunities divided by the number of zones.
-Alternatively, if 1 zone can teleport and no other zone can travel at all, that 1 zone will get all the opportunities in the region.
--->
-
-We then introduce the singly constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
-It is also mathematically equivalent to the spatial availability introduced in @soukhovIntroducingSpatialAvailability2023.
-In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
-The singly constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
-We also introduce an 'accessible' PPR (e.g., opportunities per capita), calculated by dividing each accessibility value by the zonal population.
-To clarify, this per capita expression of the singly constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
-
-Lastly, the doubly constrained case is introduced.
-In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available.
-Specifically, accessibility values for each $i$-$j$ pair must be a proportion of he zonal opportunity and population values simultaneously.
-For example, the accessibility at zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at zone 1.
-Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest.
-In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
-
-In summary, building on @wilson1971's foundational work, this paper proposed a unified framework for accessibility that is able to account for competition.
-By reintroducing Wilson's proportionality constant, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.
-This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.
-
-# References

### 2025-08-01 paper_plos/paper_plos.qmd
Made the existing draft as concise as possible -- shed 6 pages. Old version is paper_plos_V0.qmd and .pdf.

@@
 abstract: |
-  Transportation planning has long prioritized the efficiency of movement, or mobility.
-However, the concept of accessibility represents a more comprehensive evolution, shifting focus from mere movement to the potential to reach (i.e., spatially interact) with desired destinations.
+  Transportation planning has long prioritized the efficiency of movement.
+However, the concept of accessibility represents a more comprehensive evolution, shifting focus from movement to the potential to reach--or spatially interact with--desired destinations.
 Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations.
-This work's aim is to clarify and unify the concept of accessibility by connecting it into spatial interaction modeling.
-We demonstrate that widely used mobility and accessibility models, such as gravity-based accessibility and spatial interaction models, share common theoretical roots.
-From this foundation, this paper offers three contributions: (A) we introduce a family of accessibility measures within the principles of spatial interaction, and (B) formally define four members of the family, namely the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized interactions or 'access', effectively equal to the doubly constrained spatial interaction model; and (C) we demonstrate the interpretability advantages of the family, as these constrained accessibility measures yield values in units of the number of potential "opportunities for spatial interaction" or "population for spatial interaction" for each zone and zonal flow.
-The family of accessibility measures proposed here clarifies the concept of 'potential' in accessibility, demonstrates theoretical and formulaic linkages across popular accessibility and spatial interaction models, and reintroduces measurement units into accessibility measures.
-By doing so, we believe this family of measures can unlock a clearer, more interpretable, and cohesive foundation for accessibility analysis.
+To this end, this paper offers a methodological contribution by specifying a family of accessibility measures grounded in the shared theoretical roots of gravity-based accessibility and spatial interaction models, particularly their balancing factors.
+From this foundation, we outlined four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
+These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential).
+A toy example illustrates how they produce interpretable, unit-based values, offering a clearer and more coherent basis for accessibility analysis.
 
@@
 
-<!-- SUMMARY *AND OUTLINE FOR ABSTRACT*: Transportation planning, historically, has been mobility-oriented.
-Accessibility, however, should be the primary focus, and institutionalized mobility-based transportation practice has treated accessibility as its proxy.
-This has had problematic consequences.
-Conceptually, accessibility is a more complete and comprehensive evolution of mobility: from movement, to the potential movement to destinations.
-As such, there has been recognition that the adoption of accessibility planning approaches are needed; however there are too many definitions and uncertainty in meaning of the outputs.
-This work's goal is to enhance interpretation of values, clarify the concept of accessibility, and unify other measures of accessibility such as the popular Hansen-type measure the 2SFCA.
-
-To do so, this work demonstrates that accessibility and mobility oriented models (i.e., gravity-based accessibility and the spatial interaction model) share the same roots in spatial interaction modelling, and a family of accessibility measures can be specified within the spatial interaction modelling principles.
-In demonstrating these developments, three significant contributions emerge: A) we demonstrate numerically how constrained accessibility can be specified, and doing so brings interpretable units to each accessibility value, units of 'number of potential opportunities for spatial interaction'.
-B) we provide a clarifying framework for understanding the concept of 'potential' through constraints.
-Namely, demonstrating accessibility's maximum potential is captured by the the 'total constraint' (constrained version of Hansen-type measure), a restricted amount of potential is captured by the 'single constraint' (i.e., competition, such as the 2SFCA), and potential is fully restricted in the 'double constraint', effectively restricting potential such that this case is equivalent to access or simply spatial interaction.
-C) Demonstrate how the two most popular accessibility measures, the Hansen-type measure (non-competitive) and the 2SFCA (competitive), link into the family of accessibility measures.
-
-In fewer words: this work makes an important contribution to accessibility literature, by providing accessibility with a more interpretable, clarified and unified definition through the introduction of constraints from spatial interaction modelling, that reintroduce units to enhance interpretability of raw values, clarify the concept of potential, and loops the most popular forms of accessibility measures into the family.
--->
-
-In the early nineteenth century, industrializing cities grew rapidly: so did traffic congestion and the development of transportation planning practice focused primarily on mobility.
-In this newly founded practice, access to destinations was treated as a by-product of movement.
-Following the rise of the automobile and significant investments in transportation infrastructure after World War II, this mobility-oriented approach has led to some problematic outcomes.
-Specifically, the car became seen as the ultimate mobility tool, helping to foster the development of low-density, single-use residential neighborhoods and entrenching an automobility mono-culture within transportation systems [@miller_collaborative_2011; @lavery_driving_2013].
-Decades of planning for this automobility mono-culture, still often characterized by road and highway expansion, have been marked by increased travel costs and environmental burdens, with limited impact on enhancing the ease with which people can reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+In the early nineteenth century, rapid urbanisation driven by industrialisation expanded cities across Great Britain, Northern Europe, and North America.
+This expansion led to traffic congestion and the emergence of a transportation planning paradigm focused primarily on mobility.
+Within this new practice, access to destinations was treated as a by-product of movement.
+After World War II, major investments in automobile and transportation infrastructure cemented this mobility-oriented model, fostering flight from urban congested cities, lower-density sprawl, car-dependent development and entrenching automobility in planning practice [@miller_collaborative_2011; @lavery_driving_2013].
+Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people’s ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+
 In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic, etc.) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].
-
-Accessibility may be defined as the "potential of opportunities for [spatial] interaction" [@hansen1959], in contrast to mobility which is simply, movement in itself or 'spatial interaction'.
-Mobility has been the basis of traditional transportation planning approaches [@ortuzar_2011_modelling].
-While accessibility brings a more holistic understanding of combined transportation and land use systems, incorporating both the explicit consideration of destination as opportunity and the concept of _potential_ [@handyMeasuringAccessibilityExploration1997].
-
-The growing interest in accessibility has been accompanied by a boom in scholarly research using different methods and focusing on different research contexts; it has grown to include studies of access to employment [e.g., @karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], health care [e.g., @luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green spaces [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], schools [e.g., @williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contacts [e.g., @neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economic analysis [e.g., @vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011] among many other domains of application.
-In other words, accessibility analysis is used today to broadly understand the potential to reach (or spatially interact with) opportunities that are important to people [@ferreiraReenactingMobilityAccessibility2020].
-However, despite its growth in popularity in scholarly works, challenges remain with respect to the more widespread adoption of accessibility in planning practice.
-For instance, the diversity of accessibility definitions has been flagged by van Wee [@vanweeAccessible2016], Handy [@handy2020], and Kapatsila et al.
-[@kapatsila_resolving_2023].
-Further, difficulties in the interpretability and communicability of outputs has also been noticed by many authores, including Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004], van Wee [@vanweeAccessible2016], and Ferreira and Papa [@ferreiraReenactingMobilityAccessibility2020].
-
-The adoption of accessibility in planning practice is not necessarily made easier when potential adopters have to contend with a plethora of definitions, each seemingly more sophisticated but less intuitive than the last [@kapatsila_resolving_2023].
-At a high level, Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] identify four families of accessibility measures: infrastructure-, place-, person-, and utility-based.
-Of the place-based family, which is the focus of this paper, the menu has grown to include gravity-based accessibility [e.g., @hansen1959; @pirie_measuring_1979], cumulative opportunities [e.g., @wachs_physical_1973; @pirie_measuring_1979; @ye_spatial_2018], modified gravity [e.g., @SchuurmanMeasuring2010], 2-Step Floating Catchment Areas [e.g., @luo2003], Enhanced 2-Step Floating Catchment Areas [e.g., @luoEnhanced2009], 3-Stage Floating Catchment Areas [e.g., @wan2012three], Modified 2-Step Floating Catchment Areas [e.g., @delamater2013spatial], inverse 2-Step Floating Catchment Areas [e.g., @wang_2sfca_2021], and n-steps Floating Catchment Areas [@liang_novel_2024].
-How is a practitioner to choose among this myriad options?
-What differences in accessibility scores should matter, and how should they be communicated?
-[see @vanweeAccessible2016 p.
-14].
-
-Here, we seek to address the breath of place-based accessibility's definitions and lack of interpretability by demonstrating that mobility-oriented models--such as the commonly used gravity-based accessibility [e.g., @hansen1959] and the spatial interaction model [e.g., @wilson1971]-- are rooted in the same spatial interaction modeling foundation.
-In fact, we propose that accessibility can be specified using spatial interaction principles as a _family of measures_, akin to the family of spatial interaction models introduced in Wilson [@wilson1971] that can be defined using balancing factors that constrains values based on known information about the land-use.
-
-This work aims to offer three contributions: (1) it introduces a family of accessibility measures within the principles of spatial interaction; (2) it formally defines three _constrained_ accessibility measures by reintroducing Wilson-analogous balancing factors.
-These measures are the total constrained accessibility measure, the singly constrained accessibility measure, and the doubly constrained measure, which are explicitly connected to popular measures such as the Hansen-type accessibility [@hansen1959], the popular competition approach of the 2-Step Floating Catchment Area (2SFCA) [@shenLocationCharacteristicsInnercity1998; @luo2003], and the concept of market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
-These Wilson-analogous balancing factors introduced are also defined in order of increasing restrictiveness to shed light on the role of _potential_ in accessibility and access.
-(3) This work also demonstrates that the introduction of balancing factors makes accessibility measures easier to interpret and to communicate by restoring the measurement units to the resulting raw accessibility values.
-Each zonal and zonal flow value from a constrained accessibility measure is always in interpretable units, namely the number of 'opportunities for spatial interaction' or 'population for spatial interaction'.
-This is in contrast to conventional accessibility measures, particularly gravity based measures, that yield values in units of 'opportunities weighted by some representation of travel friction'.
-
-<!--Motivated by the challenge of incorporating competition into a unified framework for accessibility -->To achieve these stated objectives, this paper contends that accessibility research must reconnect with its spatial interaction origins.
-Particularly, we argue that an important aspect of spatial interaction modelling--namely, constraining the results to match empirical observations--was never effectively reincorporated into accessibility analysis.
-Empirical constraints were embraced by early spatial interaction literature following the work of Wilson [@wilson1971], but this stream of literature tended to flow separately from research inspired by Hansen's [@hansen1959] accessibility.
-The application of Wilson's [@wilson1971] empirical constraints supported the development of various spatial interaction models that remain relevant in research and practice today [@ortuzar_2011_modelling].
-However, the same cannot be said of the contemporary accessibility literature, where empirical constraints were not explicitly adopted.
-We argue that the absence of empirical constraints (and their attendant proportionality constants) has contributed to some of accessibility analysis' interpretability issues; for instance, the fuzziness of insights beyond simple proportional statements like 'higher-than' or 'lower-than' [@millerAccessibilityMeasurementApplication2018].
-Moreover, without constraints we lose track of what accessibility measures, that is, the number of opportunities that can be spatially interacted with.
-Without a clear sense of measurement units, comparability between accessibility measures, across cities, and transport modes, may also become compromised.
-
-Fortunately, accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
-It is by looking to the past that we believe accessibility analysis can newly wade into the future.
-We continue this work in the following section by tracing the development of accessibility from its origins in spatial interaction: from the Newtonian gravitational expression in Eavenstein [@ravensteinLawsMigration1889] through to the seminal accessibility work of Hansen [@hansen1959].
-We then present evidence for a narrative highlighting the marked divergence between accessibility and spatial interaction modelling research after the work of Wilson [@wilson1971].
-Next, we hark back to Wilson's [@wilson1971] spatial interaction models, and use them to derive a family of accessibility measures based on analogous constraints.
-We illustrate members of this family of constrained accessibility measures with a simple numerical example.
-We then conclude by discussing the uses of these measure and their interpretation.
+Accessibility, by contrast, is the "potential of opportunities for [spatial] interaction" [@hansen1959].
+While mobility reflects movement, accessibility captures the combined influence of transport and land use, emphasizing destinations and the potential to reach them [@handyMeasuringAccessibilityExploration1997].
+
+Accessibility research has expanded across diverse domains including: employment [@karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], healthcare [@luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green space [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], education [@williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contact [@neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economics [@vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011], among many other domains of application.
+Despite its popularity in scholarly works, accessibility still remains difficult to implement in planning due to definitional inconsistencies [@vanweeAccessible2016; @handy2020; @kapatsila_resolving_2023] and challenges in interpreting and communicating results [@geursAccessibilityEvaluationLanduse2004; @vanweeAccessible2016; @ferreiraReenactingMobilityAccessibility2020].
+
+More specifically, the wide range of accessibility definitions--with novel methods being more sophisticated but less intuitive [@kapatsila_resolving_2023]--can further hinder practical uptake [@vanweeAccessible2016].
+Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] classify accessibility measures into four categories: infrastructure-, place-, person-, and utility-based.
+Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024].
+While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in @vanweeAccessible2016: How should practitioners interpret difference in scores between modes, and how should results be communicated?
+
+Rather than propose a new measure, this work argues for a return to the spatial interaction foundations of accessibility.
+Specifically, we show how the family of spatial interaction models [@wilson1971] can be reformulated in the context of accessibility, namely as a "family of accessibility measures".
+This formulation results in constrained versions of gravity-based accessibility [e.g., @hansen1959], but in units of  _spatially reachable opportunities_.
+This approach offers a direct mathematical link to existing accessibility measures while restoring tangible meaning to zonal values.
+Instead of abstract proportional scores, constrained accessibility expresses the number of opportunities a population may potentially spatially interact with.
+
+In summary, this paper makes two contributions: (1) Reviews how spatial interaction modeling and accessibility share similar origins, but diverged in focus and interpretation.
+(2) Introduces a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
+These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+
+We contend that accessibility research should reengage with spatial interaction modeling, particularly the use of Wilson's system constraints.
+While spatial interaction models embraced such constraints to improve interpretability, most accessibility models have not.
+This lack of adoption contributes to fuzziness in current analyses, limiting interpretive clarity to simple proportional comparisons (e.g., “higher-than,” “lower-than”) [@millerAccessibilityMeasurementApplication2018].
+Without such constraints, accessibility scores lack clear units and comparability across cities or modes.
+In contrast, constrained measures yield values that can be tied to tangible values without any post-hoc treatment, making them more interpretable, communicable, and actionable in planning.
+
+This paper lays the historical and theoretical groundwork for a companion empirical study demonstrating how these constrained forms restore interpretability through concrete measurement units (e.g., number of accessible opportunities), as opposed to the abstract weighted outputs typical of conventional gravity models.
+
+Accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
+Following this introduction, this paper revisits that lineage--from Ravenstein’s Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959], to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
+We illustrate each case of the family with a numerical toy example and conclude by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.
 
@@
 
-<!--SUMMARY: Overview of the proportionality constant in Newton Law of Universal Gravitation - it takes the proportion relationship into one of equality. -->
 The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half.
@@
 
-To be certain, the expression above is one of proportionality, and is also the most famous in all of science.
-It states that the force of attraction $F$ between a pair of bodies $i$ and $j$ is directly *proportional* to the product of their masses $M_i$ and $M_j$, and inversely *proportional* to the square of the distance between them $D_{ij}$.
-Direct proportionality means that as the product of the masses increases, so does the force.
-Likewise, inverse proportionality means that as the distance increases, the force decreases.
-@eq-phys-grav-prop, however, does not quantify the magnitude of the force.
-To do so, an empirical constant, a.k.a.
-the gravitational constant, is required to convert the proportionality into an equality, ensuring that values of the force $F$ in @eq-phys-grav-prop match the observed force of attraction between masses.
-In other words, @eq-phys-grav-prop needs to be *constrained* using empirical data.
-Ultimately, the equation for the force is as seen in @eq-phys-grav, where $G$ is an empirically calibrated proportionality constant:
+This famous equation expresses that the attractive force $F$ between two bodies is directly proportional to the product of their masses and inversely proportional to the square of the distance between them.
+As mass increases, so does force; as distance increases, force decreases.
+However, @eq-phys-grav-prop only demonstrates a proportional relationship.
+To quantify the magnitude (not the proportional magnitude) of the force, it must be _constrained_ with an empirical constant.
+This constant $G$ converts @eq-phys-grav-prop from an expression of proportionality to the following expression of equality:
 
@@
 
+Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces.
 Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p.
@@
 
+This gravitational framing laid the conceptual groundwork for later empirical studies of human spatial interaction, such as those by Ravenstein, Stewart, and others, who began translating these principles into observed patterns of human movement.
+
 # Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
-<!-- 
-SUMMARY: a review of spatial interaction literature from the 1880s to the 1940s.
-A number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $Mi$ and $Mj$ and inversely proportional by their separation distance.
-This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
--->
-Since the 1880s to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as some force of attraction $F$ that is directly proportion to the masses $M_i$ and $M_j$ and inversely proportional by their separation distance.
-This concept was captured with different expressions, but all tie back to the same Newtonian gravity analogy, although not all of them included a proportionality constant in their formulation.
-
-Following Carey's *Principles* of 1858, research into human spatial interaction continued in different contexts.
-In the late 1880s, Ravenstein proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
-In these works, Ravenstein posited 1) a directly proportional relationship between migration flows and the size of destinations (i.e., centres of commerce and industry), and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
+
+Following Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts.
+Namely, since Carey to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
+
+Beginning with Ravenstein in the late 1880s, his works proposed some "Laws of Migration" based on his empirical analysis of migration flows in various countries [@ravensteinLawsMigration1885; @ravensteinLawsMigration1889].
+These works posited 1) a directly proportional relationship between migration flows and the attractive size of destinations, and 2) an inversely proportional relationship between the size of flows and the separation between origins and destinations.
 As with Carey, these propositions echo Newton's gravitational laws.
+
 Over time, other researchers discovered similar relationships.
@@
 
-A common feature of these early investigations of human spatial interaction is that a proportionality constant similar to $G$ in @eq-phys-grav was never considered.
+A common feature of these early investigations of human spatial interaction is that a proportionality constant, similar to $G$ in @eq-phys-grav was never considered.
 Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms.
@@
 
-After Carey, it is in Stewart's work on the principles of demographic gravitation that we find the strongest connection yet to Newton's law [@stewartDemographicGravitationEvidence1948].
-This may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
-Besides awareness of preceding research (he cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, as follows:
+These early formulations (@eq-reilly, @eq-zipf) clearly reflect the influence of Newtonian gravity on human spatial interaction theory, revealing a shared mathematical structure across migration, trade, and communication models.
+However, none included a proportionality constant--highlighting a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities
+
+After Carey and folowing Reilly and Zipf, it is only in @stewartDemographicGravitationEvidence1948 that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
+<!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
+--> Besides awareness of preceding research (i.e., Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, enabling his formulation to be interpreted as a measurable 'demographic' force:
 
@@
 -   $d_{12}^2$ is the distance between $1$ and $2$
--   And finally $G$, a constant that Stewart "left for future determination" [@stewartDemographicGravitationEvidence1948, p. 34]
-
-In addition to demographic force, Stewart defined a measure of the "population potential" of group $2$ with respect to group $1$.
-In other words, the potential number of people from location $2$ that could visit location $1$, as follows:
-
-$$
-V_1 = G\frac{M_2}{d_{12}}
-$$ {#eq-stewart-population-potential}
-
-For a system with more than two population bodies, Stewart formulated the population potential at $i$ as follows (after arbitrarily assuming that $G=1$):
-
-$$
-V_i = \int\frac{D}{r} ds
-$$ {#eq-stewart-population-potential-integral}
-
-\noindent where $D$ is the population density over an infinitesimal area $ds$ and $r$ is the distance to $i$.
-In @eq-stewart-population-potential-integral, $D ds$ gives an infinitesimal count of the population, say $dm$, and so, after discretizing space, @eq-stewart-population-potential-integral can be rewritten as:
+-   And finally proportionality constant $G$
+
+What is notable about @eq-stewart-force, however, is the proportionality constant $G$ was specified but "left for future determination" [@stewartDemographicGravitationEvidence1948, p.
+34].
+We can infer that it is crucial for ensuring $F$ is maintained in some units of demographic force.
+
+In addition to demographic force $F$, Stewart defined a measure of the "potential" of group $2$ with respect to group $1$.
+The partial sum of the demographic force experienced by group $1$, or the potential number of people from location $2$ that could visit location $1$, as $V_1 = G\frac{M_2}{d_{12}}$.
+For a system with more than two population bodies, Stewart formulated the population potential at $i$ by summing the contributions from each group $j$, after arbitrarily setting $G=1$:
 
@@
 
-Alerted readers will notice that @eq-stewart-population-potential-sum, with some re-organization of terms, is formally equivalent to our modern definition of accessibility popularized by Hansen [@hansen1959] in the late 1950s.
-
-Stewart's formulation of demographic force, developed in the context of what he called "social physics" [@stewartPrinciples1947], was problematic.
-It had issues with inconsistent mathematical notation.
-<!--For example, in @stewartDemographicGravitationEvidence1948, $G$ is used as a proportionality constant p.
+\noindent Where $M_j$ is the demographic mass at location $j$ and $d_{ij}$ is the distance between $i$ and $j$.
+A version of this discrete form is what was used in @hansen1959, going on to become a foundation of modern accessibility definitions, as will be discussed.
+
+Although Stewart’s concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p.
 34 and then later as _demographic energy_ on p.
-53.--> More seriously though, Stewart's work was permeated by a view of humans as particles following physical laws, but tinted by unscientific and racist ideas.
-For instance, he assumed that the molecular weight $\mu$ of the average American was one, but "presumably...much less than one....for an Australian aborigine" \[p.
-35\].
-Stewart's ideas about "social physics" soon fell out of favour among social scientists, but not before influencing the nascent field of accessibility research, as detailed next.
+53.) and its foundation in racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight $\mu$ of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
+35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration.
+In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., people^2-distance^-2) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).
+
+As will be discussed, when Hansen later adopted Stewart’s formula (@eq-stewart-population-potential-sum) for accessibility, he omitted any mention of $G$, effectively setting it to 1 arbitrarily as Stewart had done.
+This omission has persisted in accessibility research, leaving a conceptual gap in how such measures are interpreted and compared.
 
 # Hansen's gravity-based accessibility to today {#grav-to-today}
-
-<!-- SUMMARY: Stewart 1948 made an important contribution methodologically: Hansen picked it up and is now cited as the father of accessibility.
-However, Hansen work was applied, and he's use of Stewart's Population Potential measure included two crucial omissions that afflict the literature to this day.
-The first omission is 1) Stewart set the proportional constant $G$ to 1 _for future determination_, this constant is not explicitly addressed in @hansen1959 and accessibility continues to evolve without it.
-The second omission is 2) Stewart suggests that keeping the exponent value of $d_ij$ as 1 is sufficient.
-In @hansen1959, an exponent other than 1 for $d_ij$ is selected based on empirical findings, however, the issue of _opportunity potential_ units are never addressed.
--->
 
@@
 As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-Hansen [@hansen1959] focused on Stewart's [@stewartDemographicGravitationEvidence1948] *population potential* (expressed in @eq-stewart-population-potential-sum), and not on the other formulaic contributions and objectionable aspects of "social physics".
-Hansen [@hansen1959] recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-Hansen [@hansen1959] replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as follows:
-
-$$
-S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}
-$$ {#eq-accessibility}
-
-A contemporary rewriting of @eq-accessibility accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$:
+@hansen1959 drew directly from Stewart’s population potential (see @eq-stewart-population-potential-sum), but left aside the broader (and often problematic) framework of “social physics.”.
+
+Hansen recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
+Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting to accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$ as follows:
 
@@
 
-$S_{i}$ in @eq-accessibility is a measure of the accessibility from site $i$.
+$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from site $i$.
 This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost).
@@
 
-However, Hansen's use of Stewart's *population potential* measure included one crucial omission that afflicts the literature to this day.
-The omission is that between Stewart [@stewartDemographicGravitationEvidence1948] and Hansen [@hansen1959], the proportionality constant $G$ in @eq-stewart-population-potential vanished.
-This constant was not explicitly addressed in Hansen [@hansen1959] and accessibility research continues to evolve without it.
-Since Hansen [@hansen1959], accessibility analysis has been widely used in numerous disciplines but, to our knowledge, the proportionality constant has remained forgotten, with no notable developments to explicitly acknowledge or determine it.
-
-The omission of this constant generates a fundamental problem for the measurement unit of accessibility estimates, which undermines the interpretation, communication and comparability of accessibility analysis.
-Those reading Hansen [@hansen1959] must recall that Stewart [@stewartDemographicGravitationEvidence1948] had set the proportionality constant $G$ to 1, with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
-34].
-In practice, the persistent omission of the constant in accessibility analyses means that $G$ continues to be implicitly set to $1$, even when the fundamental relationship in accessibility is proportionality (e.g., $S_{i} \propto \sum_j g(O_j)f(d_{ij})$) and not equality [for instance, see the formula for accessibility at the top of Figure 1 in Wu and Levinson [@wuUnifyingAccess2020]].
-The direct consequence is that without a proportionality constant, the units of $S_i$ remain unclear: the unit of "potential of opportunity for interaction" is left free to change as $\beta$ is calibrated.
-For example, if $c_{ij}$ is distance in meters, it will be number of opportunities per $m^{\beta}$ when $f(c_{ij}) = d^{-\beta}$ but number of opportunities per $e^{-\beta m}$ when $f(c_{ij}) = e^{-\beta m}$.
-This undermines the comparability of accessibility metrics with different decay functions, and renders their results difficult to understand and communicate.
-The Hansen-style accessibility estimates found in the literature, therefore, are better thought of as ordinal measures of potential that can only be interpreted in terms of higher and lower accessibility, but which has not palpable meaning [@millerAccessibilityMeasurementApplication2018].
+However, Hansen’s formulation carried forward a crucial omission that continues to affect the literature: the proportionality constant $G$ included in Stewart’s original formulation (@eq-stewart-population-potential-sum) has vanished entirely.
+Although Stewart included $G$ explicitly (with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
+34].) Hansen made no mention of it.
+As a result, modern accessibility analysis has largely evolved without addressing the constant’s role, leaving $G$ effectively and implicitly fixed at 1.
+This omission has significant implications.
+Without a proportionality constant, the accessibility formula expresses only a proportional relationship: $S_i \propto \sum_j g(O_j)f(d_{ij})$, not of calibrated equality.
+Recognition of the nature of this relationship is not common in the literature, but is known, i.e., this proportional equation is shown in Figure 1 in Wu and Levinson [@wuUnifyingAccess2020].
+
+However, the omission of a proportionality constants and working with a proportional relationship generates fundamental issues in comparability between and, arguably within, studies.
+As the absence of $G$ means accessibility estimates have no fixed unit--this renders them sensitive to the choice of impedance functions.
+For instance, if travel cost $d_{ij}$ is measured in meters, then when the travel impedance function $f(d_{ij})$ equals $d_{ij}^{-\beta}$, the resulting $S_i$ has units of opportunities per $\text{metres}^{\beta}$; but when $f(d_{ij})$ is set to equal $e^{-\beta d_{ij}}$, the units become opportunities per $e^{\beta \text{metres}}$.
+Such variation impairs comparability across analysis and obscures the meaning of accessibility scores renders their results difficult to understand and communicate without post-hoc treatment.
+
+Therefore in practice, Hansen-type measures (of proportionality) are better understood as *ordinal indicators*--they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
+The continued absence of $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement.
 
 # Wilson's family of spatial interaction models
-
-<!--SUMMARY: Introduce Wilson's spatial interaction model.
-Then discuss how it could relate to gravity analogy -- but is tactically derived using entropy-maximisation, obtaining the interaction variable as a statistical average.
-Then introduce the constraints and cases.
--->
-
-On the other side of the Atlantic, Alan G.
-Wilson was developing related, yet parallel work.
-In his groundbreaking study [@wilson1971], Wilson defined a general spatial interaction model as follows.
-While accessibility was characterised as an associated concept of 'potential', the primary focus was on modelling observed spatial interaction:
+While accessibility research evolved in North America with @hansen1959, a parallel development was taking place across the Atlantic with Alan G.
+Wilson.
+In his groundbreaking study [@wilson1971], advanced a general framework for spatial interaction modeling--focused not on potential (as accessibility), but on modeling observed flows of interaction between places.
+@wilson1971 formalized the general spatial interaction model through the following equation:
+
 $$
@@
 The quantities $W_i^{(1)}$ and $W_j^{(2)}$ are proxies for the masses at $i=1,s,n$ origins and $j=1,s,m$ destinations.
-The super-indices $(1)$ and $(2)$ are meant to indicate that these masses could be different things, i.e., $W_i^{(1)}$ could be populations, and $W_j^{(2)}$ hectares of park space.
-Finally, $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
-In this way, $T_{ij}$ explicitly measures *interaction* in the unit of trips, and the role of $k$ is to ensure that the system-wide sum of $T_{ij}$ represents the total flows in the data.
-In other words, $k$ is a scale parameter that makes the overall amount of flows identical to the magnitude of the phenomenon being modeled.
-In other words, it balances the units, in a conceptually similar sense as the gravitational constant in Newton’s Law of Universal Gravitation.
-
-Traditionally, the development of the spatial interaction model put an emphasis on the interpretability of the results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
-But instead of relying on the heuristic of Newtonian gravity (e.g., some interaction between a mass at $i$ and a mass at $j$ separated by some distance), Wilson's approach was to maximise the entropy of the system.
-Entropy maximisation in this case achieves stable results as a statistical average that represents the population.
-The approach works by assuming undifferentiated individual interactions, and assessing their probabilities of making a particular journey.
-The result of @eq-phys-gravity-model then is a statistical average [@wilson1971; @seniorGravityModellingEntropy1979].
-
-To ensure that $T_{ij}$ in @eq-phys-gravity-model is in the unit of trips (our unit of origin-destination spatial interaction), additional knowledge about the system is required.
-At the very least, this framework assumes that the total number of trips in the system $T$ is known, and therefore:
+The super-indices $(1)$ and $(2)$ indicate that these masses can be any number of different things associated with the zones, i.e., $W_i^{(1)}$ could be population at a zone as an origin, and $W_j^{(2)}$ hectares of park space at a zone as a destination.
+$f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
+Finally, $k$ acts as a proportionality constant, shifting the equation from a proportional to an equal relationship by incorporating known system totals.
+In this specific sense, we infer that $k$ serves a role similar to the gravitational constant in Newton’s law--it calibrates the model so that outputs match real-world quantities.
+
+From the outset, spatial interaction models emphasized interpretability of results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
+But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson’s innovation was to ground the model in _entropy maximisation_.
+By maximizing the number of ways individual trip probabilities could be arranged under known constraints, Wilson derived models that estimate _statistical averages_ of flows between zones [@wilson1971; @seniorGravityModellingEntropy1979].
+
+Crucially, to ensure that $T_{ij}$ in @eq-phys-gravity-model is maintained in units of flow between $i$ and $j$, the model moves from proportionality to calibrated equality by incorporating empirical constraints.
+At a minimum, this requires knowledge of the total number of flows $T$ in the system, leading to the basic constraint:
 
@@
 Additional information can be introduced.
-For example, when information is available about the total number of trips produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
+For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
 
@@
 
-Alternatively, if there is information available about the total number of trips attracted by each destination, $W_j^{(2)}$ is represented as $D_j$ and the following constraint can be used:
+As well, if there is information available about the total number of flows attracted by each destination, $W_j^{(2)}$ is represented as $D_j$ and the following constraint can be used:
 
@@
 
-Using information about the system that satisfy these constraints fully, partially or not at all, a family of spatial interaction models can be derived based on @eq-phys-gravity-model.
-$K$ is specified depending on the applied constraint(s).
-In the framework introduced in Wilson [@wilson1971], three constrained versions are specified: the first being a case where to results only match the total volume of interaction, the second being a singly constrained case, and the third a doubly constrained case.
-
-In the first, @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel do not hold.
-In practical terms, this means that the total number of trips predicted by the model must be equal to sum of all flows from origins i to destinations j.
-The balancing constant $K$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:
+Depending on which system constraints are applied—fully, partially, or not at all—a family of spatial interaction models can be derived from @eq-phys-gravity-model.
+The proportionality constant $k$ is replaced with different _balancing factors_.
+This change in name is more useful as these factors not only preserve proportionality but also ensure that the predicted flows $T_{ij}$ align with the known constraints considered in the system.
+In other words, the different balancing factors adjust the model so that flows become statistical averages consistent with observed origin and/or destination data.
+In the framework introduced and inferred from @wilson1971, three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.
+
+In the unconstrained model, constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel do not hold.
+In practical terms, this means that the total number of flows predicted by the model must be equal to sum of all flows from origins $i$ to destinations $j$.
+The balancing factor $K$ takes the place of $k$ and is equal to the following [as specified in @cliff_evaluating_1974 and @fotheringham_spatial_1984]:
 
@@
 
-In the second case, only one of @eq-constraint1-gravitymodel or @eq-constraint2-gravitymodel hold.
-The resulting models are, in Wilson's terms, singly constrained.
-When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained model:
+In the singly-constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
+When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained singly-constrained model:
 
@@
 
-Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$, representing what we know about the system, the spatial interaction outbound flow, i.e., outbound trips produced at $i$.
-Also, there is no longer a single system-wide proportionality constant, but rather a set of proportionality constants (i.e., balancing factors) specific to origins.
-For this model, the balancing factors ensure that @eq-constraint1-gravitymodel is satisfied, meaning that the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$ i.e., the total number of outbound trips.
-Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of trips.
-This model is useful when trips ends are unknown but the number of trips originating from each location is known and the total of these trips represents all trips in the system.
+Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$, representing what we know about the system, the spatial interaction outbound flow, i.e., outbound flows produced at $i$.
+Also, there is no longer a single system-wide balancing factor $K$, but rather a set of balancing factors specific to origins $A_i$.
+
+For this model, the balancing factors ensure that constraint @eq-constraint1-gravitymodel is satisfied, meaning that the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$ e.g., the total number of outbound flows.
+Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of flows.
+
+The production-constrained singly-constrained model is useful when flows ends are unknown but the number of flows originating from each location is known.
 The balancing factors for the production-constrained model are solved for each origin $A_i$, and according to Wilson are:
@@
 The attraction-constrained model is similar to the production-constrained model as it is also singly constrained but from the perspective of the mass at the destination.
-From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow, i.e., trips attracted at the destination and takes the following form:
+From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow e.g., flows attracted to the destination, and takes the following form:
 
@@
 
-In this model (@eq-attraction-constrained-gravitymodel), the balancing factors ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$ i.e., the total number of inbound trips to $j$.
+In this model (@eq-attraction-constrained-gravitymodel), the balancing factors ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$ i.e., the total number of inbound flows to $j$.
 This should hold for all destinations.
-As before, destination-specific proportionality constants (i.e., balancing factors) $B_j$ were derived by Wilson as:
+As before, destination-specific balancing factors $B_j$ were derived by Wilson as:
 
@@
 
-The third case in the family of spatial interaction models is the production-attraction constrained model.
-In this case, both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
+Lastly, the doubly-constrained model is the production-attraction constrained model in @wilson1971.
+In this case, both constraints @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
 These constraints ensure that the sum of predicted flows from one origin to all destination, and the predicted flows going to one destination from all origins must equal the known mass of the origin $O_i$ and of the destination $D_j$.
@@
 
-In this model, both proxies for the masses are replaced with the known masses, that is, the trips produced by the origin and the trips attracted by the destination.
-There are now two sets of mutually dependent proportionality constants:
+In this model, both proxies for the masses are replaced with the known masses, that is, the flows produced by the origin and the flows attracted by the destination.
+There are now two sets of mutually dependent proportionality constants shown in @eq-doubly-constrained-balancing-factors.
+Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
 
@@
 
-Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
-It is worth noting, however, that although Wilson's approach is built on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants, including the one that eluded Stewart [@stewartDemographicGravitationEvidence1948] and that has been overlooked in almost all subsequent accessibility research.
-Why was this key element of spatial interaction models potentially ignored in accessibility research?
-In the next section we aim to address this question.
+Wilson’s work is notable for many reasons, one of which is his departure from the metaphor of Newtonian gravitational physics.
+Instead, he grounded his models in statistical mechanics, particularly the principle of entropy maximisation.
+For accessibility research, Wilson’s approach is especially significant because it can address a problem that has quietly plagued this field (as this work has argued): the transition from proportionality to equality.
+
+Rather than relying on a universal constant like $G$, Wilson’s models calibrate interaction flows through known empirical constraints (e.g., total system flows, or origin-side and destination-side flow totals).
+This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
+In short, when aspects of the systems are inputs into the balacing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled.
+In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
+While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that _indicate_ a proportional relationship and not one of equality nor yielding results without interpretable units.
+
+As a preview for the work ahead: Wilson's balancing factors can be manipulated into _proportional allocation factors_ and equations can be established to reflect accessibility.
+The resulting $ij$ flows can be understood as accessibility flows--and aggregated at the zone level to express accessibility in units of actual opportunities or population.
+
+In this way, these balancing factors can be used to recover what Stewart’s $G$ was meant to accomplish, but through a different conceptual path.
+But before demonstrating a family of accessibility measures that are sensitive to constraints, in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
+This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this work.
 
 # Accessibility and spatial interaction modelling: two divergent research streams
-
-<!--
-SUMMARY: Accessibility developed separate from SIM.
-
-SIM focuses on trips and was institutionalized to develop the US Interstate highway systems.
-Many guidelines and practical training was done - a lot of funding.
-SIM development has integrated accessibility in a few ways: 1) within the SIM itself as a balancing factor, 2) as an additional independent variable in the SIM, 3) SIM and HAM in parallel in the context of location allocation problems and within the steps of transportation demand modelling (trip distribution or trip generation).
-
-Accessibility research has separated SIM almost entirely.
-When Wilson is cited: 1) typically when using a different impedance function, 2) when using the SIM to predict trip flows for accessibility input or vice versa, 3) a few exceptional papers that interpret the doubly-constraints as an inverse HAM in the context of competitive accessibility, and 4) it was only until recently, that a singly-constrain was introduced in Soukhov et al.
-2023.
--->
 
@@
 
-The work of Hansen [@hansen1959] and Wilson [@wilson1971] responded to important developments, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994].
-These dictates and needs were far from trivial.
-In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S.
-Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars (equivalent to over $600 billion in 2023) [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
-Spatial interaction modelling was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
-Accessibility, at the time, did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: it predicts the less tangible "potential" for spatial interaction with opportunities.
-In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
-
 ```{r}
@@
 
-To illustrate this point, we conducted a bibliographic analysis of the literature that cites Hansen [@hansen1959], Wilson [@wilson1971], or both.
-We retrieved all relevant documents using the Web of Science "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971].
-As a result of this search, we identified `r n_h |> prettyNum(big.mark = ",")` documents that cite Hansen [@hansen1959], only `r n_w |> prettyNum(big.mark = ",")` documents that cite Wilson [@wilson1971], and `r n_b |> prettyNum(big.mark = ",")` that cite both.
-The earliest document in this corpus dates to `r min(corpus$PY)` and the most recent is from `r max(corpus$PY)`.
-The number of documents per year appears in @fig-docs-per-year, where we see the frequency of documents over a span of almost fifty years.
-In particular, we notice the remarkable growth in the number of papers that cite Hansen [@hansen1959] compared to those that cite Wilson [@wilson1971] since the year 2000.
-
-As noted, literature that cites both Wilson [@wilson1971] and Hansen [@hansen1959] are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus, visualised in pink in @fig-docs-per-year).
-After reading the works, we can also discern that they too are divergent, with one stream focused on developing accessibility (i.e., *potential for* spatial interaction) and another on spatial interaction.
-The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected, and at times, improperly conflated.
-
-On one hand, the stream of literature focused on spatial interaction models inspired by Wilson [@wilson1971] and which cite both Hansen [@hansen1959] and Wilson [@wilson1971], tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models.
-These works treat separate spatial interaction and accessibility as a seperate but related phenomenon.
-Specifically, some early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of Hansen's [@hansen1959] model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p.
-99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976].
-Some authors have explored this relationship, for instance as in Fotheringham [@fotheringhamSpatialCompetitionAgglomeration1985] who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the *competition destination* model).
-Other works used both Hansen [@hansen1959] and Wilson's [@wilson1971] framework in conjunction, such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with Fotheringham's [@fotheringhamSpatialCompetitionAgglomeration1985] demonstration [e.g., @beckers2022incorporating].
-Other works departed from Hansen's [@hansen1959] definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-
-<!-- rafa 1: no need to keep all references in each citation block. AS: I'll keep them for now. -->
-
-<!-- rafa 2: no need to keep all references in each citation block. AS: I'll keep them for now. -->
-
-On the other hand, there is another subset of literature that cite both Hansen [@hansen1959] and Wilson [@wilson1971] that is accessibility-focused.
-We categorise their citation of Wilson [@wilson1971] for three general reasons.
-Firstly, a group of these works cite Wilson [@wilson1971] as attribution for using context-dependent travel cost functions [e.g., @weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
-These works do not necessarily comment on spatial interaction explicitly.
-Secondly, another group of this accessibility-focused "both" citing literature _does_ associate spatial interaction as defined in Wilson [@wilson1971] with accessibility's potential for spatial interaction more explicitly [e.g., @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-We agree accessibility and spatial interaction are related topics: accessibility is an expression of its *potential* and the Wilson [@wilson1971] paper briefly touches on the concept.
-However, in some of this literature, Hansen [@hansen1959] and Wilson [@wilson1971] are co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending], perhaps revealing the murkiness of the distinction between spatial interaction and the *potential for* spatial interaction in the literature.
-Thirdly, there is a group of accessibility-focused works that interpret Hansen's [@hansen1959] model as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
-This group cities the earlier spatial interaction works that make this connection and is especially prominent in the investigation of competitive accessibility topics [e.g., @karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
-As outlined in preceding sections, we argue interpreting the singly- or doubly- constrained spatial interaction model's balancing factor as accessibility yields output values that are similarly plagued by interpretability issues.
-
-Lastly, as an extension of the third reason within this group, only the works of Soukhov et al.
+Despite their close conceptual ties, the accessibility and spatial interaction modeling literatures have developed along largely separate paths since the 1970s.
+Hansen’s [@hansen1959] formulation of accessibility became the method used for decades of work on transport equity, land use analysis, and urban accessibility planning.
+Meanwhile, Wilson’s [@wilson1971] entropy-maximising framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
+We argue the framework's quiet innovation--introducing empirically grounded constraints to shift from proportionality to calibrated equality--made the framework immediately relevant for policy applications as outputs were in tangible units.
+Yet, this mechanism was never widely adopted in accessibility analysis.
+
+This divergence is especially striking given the context in which both frameworks emerged.
+As noted in @battyChronicleScientificPlanning1994, large scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
+And these needs were far from trivial: in the U.S., for instance, the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today’s dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+In this context, spatial interaction models were incorporated into institutional practices focused on "predict and provide" travel demand forecasting [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
+Accessibility analysis, by contrast, remained more conceptually diffuse, focused on indicators of "potential" spatial interaction with opportunities rather than flows that could tangibly guide infrastructure decisions (e.g., roadway capacity expansion, new construction).
+Whereas spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.
+
+To explore why Wilson’s approach never crossed over to accessibility modeling, we conduct a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both (on Web of Science using the "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]) .
+Only `r (n_b) |> prettyNum(big.mark = ",")` out of the `r (n_h+n_w+n_b) |> prettyNum(big.mark = ",")` documents that emerged from our search cite both.
+The number of articles, by year and if they cite Hasen, Wilson, or both are shown in @fig-docs-per-year.
+
+Through the close analysis of _how_ articles that cite both works, we identify two distinct citation patterns: one group of articles focused on accessibility, the other on spatial interaction.
+In examining these articles, we uncover how the relationship between the two has often been misunderstood, underexplored, or entirely overlooked.
+
+In the first stream of literature--which cite both but are focused more so on spatial interaction models--they treat spatial interaction and accessibility as separate but related phenomenons.
+Four subsets of this stream emerge.
+
+First, some of the more early works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of Hansen's accessibility measure [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], likely following Wilson's own recognition of this similarity between balancing factor $A_i$ and Hansen-type measure $S_i$ on p.
+10 in Wilson [@wilson1971].
+In some ways, this relationship has been recognized as a "common sense" approach to incorporating accessibility in the spatial interaction model [@morrisAccessibilityIndicatorsTransport1979, p.
+99], though acknowledgment of its further exploration has been recommended [@battyMethodResiduesUrban1976].
+
+The second subset of articles within this stream use both Hansen [@hansen1959] and Wilson's [@wilson1971] framework in conjunction.
+For instance, some articles argue that spatial interaction models fail to explain certain spatial patterns on their own, for instance, as in Fotheringham [@fotheringhamSpatialCompetitionAgglomeration1985] who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility (Hansen-type accessibility) as a variable within the model may remedy the issue (e.g., the *competition destination* model).
+Other works take a more applied approach: such as in defining location-allocation problems in operations research [@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trips (or some other spatial interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility as a variable within spatial interaction models, in line with Fotheringham's [@fotheringhamSpatialCompetitionAgglomeration1985] demonstration [e.g., @beckers2022incorporating].
+
+The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+In sum though, across these works, they recognize Hansen-type accessibility as an indicator of 'potential' but as a separate but related concept to spatial interaction.
+
+Moving onto the group of accessibility-focused literature that cites both works, we categorise their citation of Wilson [@wilson1971] within three general groups.
+Overall though, these works do not engage, or only superficially engage with Wilson [@wilson1971].
+
+Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions.
+This trend is common: for instance it is done in [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+However, these works do not engage with spatial interaction beyond this attribution.
+
+Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility’s potential for spatial interaction--but only superficially.
+These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+Indeed, while accessibility can be seen as the *potential* for spatial interaction--and Wilson [@wilson1971] briefly touches on this--such mentions have not resulted in deeper analytical integration of these concepts.
+Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+This conflation reveals ongoing murkiness between the distinction of spatial interaction and the *potential for* spatial interaction in the literature.
+
+Thirdly, there is a group of accessibility-focused works that interprets the measure used in @hansen1959 as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
+This group often cites the spatial interaction works that make this connection (i.e., the first subset of the first stream of literature) and is especially prominent in the investigation of competitive accessibility topics e.g., [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
+Only the works of Soukhov et al.
 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] use Wilson's [@wilson1971] balancing factors as a method for maintaining constraints on opportunities within the context of competitive accessibility.
-These works introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
+
+On that note, Soukhov et al., 2023 and 2024 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
 This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region.
@@
 These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
-This equivalence is only discovered in hindsight, as will be demonstrated in this study.
-These two works also do not discuss other constrained cases that will be addressed in the next section.
-
-# A family of accessibility measures
-
-In this section, we introduce a family of accessibility measures inspired by Wilson's spatial interaction modelling framework.
-At the end of this section, Table 1 presents a summary, where each member of the family of accessibility measure is explained in plain language, alongside their balancing factor, proportional allocation factor and mathematical equation and value interpretations.
-
-As argued in the preceding section, the streams of research on accessibility and spatial interaction modelling have evolved as largely separate streams with little contact since Hansen [@hansen1959] and Wilson [@wilson1971].
-This may explain why the constraints and associated balancing factors of spatial interaction models did not cross over to accessibility analysis.
-This is intriguing since Wilson made an effort to connect developments in spatial interaction modelling to accessibility, noting for instance, the denominator of the proportionality constants specific to origins (@eq-production-constrained-balancing-factor) is the inverse of balancing factor $A_i$ [@wilson1971 p.
-10]:
-
-$$
-S_i = \frac{1}{A_i} = \sum_j W_j^{(2)} f(c_{ij})
-$$ {#eq-Ai-as-accessibility}
-
-Understanding $S_i$ as the inverse of Wilson's $A_i$ does not uncover any new meaning for $S_i$ itself.
-Indeed, mathematically it is true, $A_i$'s role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ i.e., keeping units balanced and proportionality based on constraints.
-Understanding $S_i$ itself as a balancing factor is not wholly helpful as Hansen and (Stewart before him) defined accessibility as a partial sum of the demographic force $F$ or the system-wide population potential of opportunities for spatial interaction (i.e., @eq-stewart-population-potential-sum representing $V_i = \sum_j \frac{M_j}{d_{ij}}$ after losing $G$).
-
-Hence, we propose stepping back to introduce a revised definition of accessibility: the _constrained_ potential for spatial interaction.
-Once we bring back Wilson's proportionality constant $k$ into the picture, we can define the *potential for spatial interaction* between two locations $i$ and $j$ is as follows:
-
-$$
-V_{ij} = k W_j^{(2)} f(c_{ij})
+This equivalence is only discovered in hindsight, as will be demonstrated in the following section.
+These two works also do not discuss other constrained cases that will also be addressed.
+
+In sum, despite the interpretative advantaged offered by the statistical logic of @wilson1971's framework, the spatial interaction-oriented @hansen1959 citing literature nor the accessibility-oriented literature that cite @wilson1971 have yet to meaningfully adopt Wilson's constraint-based logic to the concept of accessibility.
+So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
+This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
+This approach takes the same path of entropy-maximisation as in @wilson1971, and does not rely on specifying some universal constant $G$ like initially proposed in @stewartDemographicGravitationEvidence1948 (recall, @eq-stewart-population-potential-sum which @hansen1959 operationalised).
+
+# A family of accessibility measures: from proportionality to equality
+
+Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter.
+This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la @hansen1959) from a relationship describing proportionality to equality.
+
+This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
+Rather than relying on arbitrary or uncalibrated proportional relationships--or invoking a universal constant $G$ as proposed in @stewartDemographicGravitationEvidence1948 --we adopt Wilson’s approach grounded in statistical mechanics.
+
+<!-- fix this one-->
+To do so, we propose a revised definition of accessibility considerate of the constraint-based spatial intereaction model: the potential for spatial interaction with opportunities or population.We can bring $k$ as a proportional allocation factor which encorporates Wilson's balancing factor, to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potentnial for spatial intereaction with population_ $M_{ji}$.
+In effect, $\kappa$ is unitless and proportionally allocates (based on the constraints of the case) opportunities (for $V_{ij}$) and population (for $M_{ji}$).
+The equation is as follows:
+
+$$
+\begin{array}{l}
+V_{ij}^X = \kappa_{ij}^X W_X^{(2)}\\ 
+M_{ji}^X = \hat \kappa_{ji}^X W_X^{(1)}
+\end{array}
 $$ {#eq-access-01}
 
-\noindent where $V_{ij}$ is the potential for interaction from $i$ to $j$. The accessibility from origin $i$ can then be summarised as as a partial sum of the potential at $i$:
-
-$$
-V_{i} = k \sum_j W_j^{(2)} f(c_{ij})
+\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ is the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case.
+Accessibility flows can also be summarised as as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively, as often done in accessibility research:
+
+$$
+\begin{array}{l}
+V_{i}^X = \sum_j \kappa_{ij}^X W_X^{(2)}\\
+M_{j}^X = \sum_i \hat \kappa_{ji}^X W_X^{(1)}
+\end{array}
 $$ {#eq-accesssibility-01}
 
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
-
-The market potential variant can also be generally defined, which is effectively the transpose of $i$ and $j$ in @eq-access-01 and @eq-accesssibility-01 as follows:
-$$
-M_{ji} = k W_i^{(1)} f(c_{ij})
-$$ {#eq-market-01}
-
-$$
-M_{j} = k \sum_j W_i^{(1)} f(c_{ji})
-$$ {#eq-market-potential-01}
-
-Similar to @eq-phys-gravity-model, $W_j^{(2)}$ above is the mass at the destination and the sub-indices are for $i = 1,s, n$ origins and $j = 1,s, m$ destinations.
-
-To detail the anatomy of $V_{ij}$ and $M_{ji}$ along with the partial sums of $V_i$ and $M_j$, Figure @fig-analytical-device-conc-accessibility illustrates the accessibility analytical framework we propose using a simple 3 zone system.
-Each measure's most disaggregate value is $X_{ij}$, potential for spatial interaction from $i$ to $j$.
-The $X$ is a stand in for the $ij$ values of all the cases and their variants that will be described (i.e., $V_{ij}^0, M_{ji}^0, V_{ij}^T, M_{ji}^T, V_{ij}^S, M_{ji}^S$ and $V_{ij}^D, M_{ji}^D$).
-The single marginals represent the origin-side and destination-side weights of the zones.
-The total marginal represent the sum of a single marginal.
+Figure @fig-analytical-device-conc-accessibility illustrates our analytical framework using a simple 3-zone system.
+The most detailed values, $X_{ij}$, represent the potential for spatial interaction from origin $i$ to destination $j$.
+Here, $X$ stands for all cases and variants to be discussed (e.g., $V_{ij}^0$, $M_{ji}^0$, $V_{ij}^T$, $M_{ji}^T$, $V_{ij}^S$, $M_{ji}^S$, $V_{ij}^D$, and $M_{ji}^D$).
+Single marginals show origin and destination weights, while the total marginal sums these values.
 
@@
 
-As an additional overview, we define four distinct members of the constrained accessibility measure family, all delineated based on their constant $k$, which takes the form of either balancing factors $K^T$, $B_j$, $A_i$ depending on the indicator:
-
-1.
-**The unconstrained case** with variants $V_i^0$ and $M_i^0$.
-This case is equivalent to Hansen's [@hansen1959] formulation (first variant) and Reilly's [@reilly1929methods] market potential formulation (second variant), whereby a balancing factor is neglected, so units of zonal values are in units of 'opportunities-by-travel-impedance-value' and 'population-by-travel-impedance-value'.
-In this case, no constraints are introduced to ensure that values of marginals are preserved.
-
-2.
-**The total constrained case** with variant $V_i^T$ and variant $M_j^T$.
-The first variant $V_i^T$--the _total constrained accessible opportunities measure_--resembles Hansen's [@hansen1959] formulation but with an additional regional balancing factor $K^T$ term, defined as the ratio of the total number of opportunities in the region to the total sum of unconstrained accessibility values in the region.
-In this way, $K^T$ ensures that each zonal accessibility value is a proportion of the total opportunities in the region (i.e., the total marginal in @fig-analytical-device-conc-accessibility), requires no information about the population seeking opportunities, and is in units of 'opportunities' accessible.
-The second variant $M_j^T$ is the transpose of $i$ to $j$ of the first variant, effectively a constrained version of market potential and is referred to as _total constrained accessible population measure_.
-In this variant, each zonal accessibility value is a proportion of the total population in the region (maintained by $\hat K^T$), requires no information about the opportunities that are sought, and each zonal value is in units of 'population' accessible.
-  - When conceptualising _potential_ for spatial interaction--whether with opportunities or population--the total constrained case reflects the lowest level of restriction and hence maximum potential.
-The balancing factors $K^T$ and $\hat K^T$ only ensure that $V_{ij}^T$ and $M_{ji}^T$ values end up matching the _total sum_ of one of the single marginals, not the individual marginals themselves.
-For example, $K^T$ does not guarantee that all the opportunities accessibility at $V_{i=1}^T$ reflect a proportional allocation of the destination mass (i.e., _number_ of opportunities) at $j=1, 2, 3$.
-In some cases, an allocation of opportunities from a destination $j$ to $i=1$ could _exceed_ the number of opportunities at that $j$ (i.e., meaning that destination $j$ is very attractive and reachable for an $i$, relative other flows in the region).
-However, what $K^T$ _does_ ensure is that each $V_{ij}^T$ does not exceed the overall number of opportunities in the region--the total marginal.
-In other words, $V_{i=1}^T$ expresses the number of opportunities that origin $i=1$ could potentially interact with, as drawn from the entire regional opportunity mass, rather than constrained by individual destination totals.
-In this way, the _total constrained_ case reflects the maximum amount of _potential_ while still maintaining interpretable units.
-
-3.
-**The singly constrained case**, with two variants $V_i^S$ (opportunity-constrained) and $M_j^S$ (population-constrained).
-The first variant is mathematically equivalent to the spatial availability measure [@soukhovIntroducingSpatialAvailability2023], and this variant's per capita form is equivalent to the popular 2SFCA measure [@luo2003; @shen1998].
-Both variants can also be defined using either Hansen's [@hansen1959] or the market potential formulation but with balancing factor $B_j$ for $V_i^S$ or $A_i$ for $M_j^S$.
-These balancing factors ensure that the total marginal (green box in @fig-analytical-device-conc-accessibility) is maintained _as well as_ the values at one of the single marginals (destination-mass marginal for opportunity-constrained and origin-mass marginal for population constrained in @fig-analytical-device-conc-accessibility).
-In this way, the singly constrained case reflects a medium amount of potential, restricted by either values fitting the single destination-mass marginal or the single origin-mass marginal.
-  - The first variant $V_i^S$ includes a set of destination-side balancing factors $B_j$ which ensure opportunities (or destination masses) are allocated to each origin $i$ based on the population (i.e, origin mass) of the origin $i$ and travel impedance from that $j$ to all $i$s.
-Hence, $B_j$ ensures that each $V_i^S$ is the product sum of a _proportional share_ of opportunities allocated from each destination $j$ implicitly also being a share of the total opportunities in the region.
-Similar to the total constrained case, each $V_i^S$ and $V_ij^S$ accessibility value is expressed in units of 'opportunities' accessible.
-But unlike the total constrained case, the singly constrained case explicitly incorporates population by allocating a balanced share of a destination's total opportunities to populations at reachable origins.
-  - The second variant $M_j^S$ includes a set of opportunity-side balancing factors $A_i$ which ensure population (or origin masses) are allocated to each destination $j$ based on the destination mass of the destination $j$ and all possible travel impedance from that $i$ to all $j$s.
-$A_i$ ensures similar, but transposed, so constraints are satisfied.
-Specifically, each $M_j^S$ represents both a share of the total regional population and the sum of balanced proportions of population from all origins, allocated to each destination based on the opportunities sought and travel impedance.
-Each zonal value is expressed in units of 'population' accessible.
-
-4.
-**The doubly constrained case**.
-It is constrained simultaneously by population (origin masses) and opportunities (destination masses), so each $V_{ij}^D$ (equivalent to $M_{ji}^D$) is expressed in units of 'population-opportunity capacity' that is accessible between $i$ to $j$.
-This case requires the number of opportunities and population to match, e.g., the analyst must know the matching spatial interaction capacity of the population (demand) and opportunities (supply).
-By using this one-to-one matching data and the double constraints (i.e., $A_i$ and $B_j$ at once, ensuring both the double constraint maintains both marginals in @fig-analytical-device-conc-accessibility), this case restricts the _potential_ to spatially interact completely.
-In other words, the doubly constrained accessibility values reflect the number of predicted interactions between the opportunities and population, effectively, this case is equivalent formulaically as Wilson's doubly constrained spatial interaction model (i.e., _attraction-production constrained_).
-It can be understood as predicting a value of 'access', and not accessibility.
-
-And as a summary: each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
-
-`\blandscape`{=latex}
-
-::: {=latex}
-\input{summary-family-access-measures-table.tex}
-:::
-Table 1: Summary of constrained accessibility measure types and interpretations
-
-`\elandscape`{=latex}
-
-## Setup of the simple numeric example
+The proportional allocation constant $\kappa$ takes the form of a balancing factor that varies depending on the constraints applied.
+Each member of the accessibility measure family is defined by the constraints used, and can be grouped into the following four cases:
+
+1. **Unconstrained Case ($V_i^0$, $M_j^0$)**
+- Equivalent to Hansen’s [@hansen1959] and Reilly’s [@reilly1929methods] original formulations; the status quo of accessibility modelling.
+- No balancing factors applied; units are in "opportunities-by-impedance" for $V_i^0$ or "population-by-impedance" for $M_j^0$.
+- No constraints are applied, so values reflect proportionality only and are not calibrated to known system totals.
+2. **Total Constrained Case ($V_i^T$, $M_j^T$)**
+- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in @fig-analytical-device-conc-accessibility) i.e., total number of opportunities or population in the system.
+This ensures the sum of all values in the system match the total marginal.
+- Units of $V_i^T$: accessible opportunities from $i$, a value that is total constrained and linearly proportion to $V_i^0$.
+- Units of $M_j^T$: accessible population from $j$, a value that is total constrained and linearly proportion to $M_j^0$.
+3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
+- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
+Reflects how the literature calculates competitive accessibility.
+- Units of $V_i^S$: accessible opportunities from $i$, a value that is the sum of opportunity supply flows allocated proportionally based on demand at $i$.
+Mathematically equivalent in per-capita form to 2SFCA [@luo2003].
+- Units of  $M_j^S$: accessible population from $j$, a value that is the sum of population demand flows allocated proportionally based on supply at $j$.
+4. **Doubly Constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
+- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously; equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
+- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in @fig-analytical-device-conc-accessibility).
+- Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
+Represents ‘interaction capacity’ or ‘realized access,’ serving as predictions of real interaction flows.
+
+## Toy example setup
 
@@
 
-Consider a simple region as shown in @fig-analytical-device-conc-accessibility, with zone IDs 1, 2 and 3.
-Each zone is both an origin $i$ and a destination $j$.
-The following three pieces of information are defined: zonal population and opportunities, zonal cost matrix, and travel impedance functions for three types of travel behaviour.
-In this example, the population are people, the opportunities are physicians, and the region can be described by three possible travel behaviours.
-
-Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone.
-For reference, considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`.
-The number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025].
-Secondly, to pair the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost.
+Consider the simple 3-zone region in Figure @fig-analytical-device-conc-accessibility, where each zone serves as both origin ($i$) and destination ($j$).
+The system includes three inputs: zonal population and opportunities, a zonal cost (travel time) matrix, and three travel impedance functions representing different travel behaviours.
+
+First, Table @tbl-small-system-land-use summarizes population (in 10,000s) and physicians per zone.
+For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)` comparable to Canada’s 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
+Second, Table @tbl-small-system-cost shows travel times (minutes); it can be discerned that Zones 1 and 3 are closer to each other than to Zone 2.
+Zones 1 and 3 together have a population roughly equal to Zone 2 but offer more than twice the physician availability.
+We interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core, and Zone 2 as the Suburban.
 
@@
 
-From both @tbl-small-system-land-use and @tbl-small-system-cost, Zone 3 and 1 can be interpreted as of an urban core with major healthcare institutions and a healthcare cluster on the edge of the city respectively, each with a moderate residing population (with zone 3 having both a higher population and physician count).
-Zone 2, can be interpreted as a more distant bedroom community, with a relatively high population and fewer physicians.
-In sum, Zones 1 and 3 are more proximate to each other than to Zone 2, and together match Zone 2’s population while offering more than twice the physician availability.
-
-And lastly, in @eq-travel-behaviour-scenarios we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities.
-Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region.
+And third, @eq-travel-behaviour-scenarios presents the assumed travel impedance functions reflecting three different travel behaviours.
 A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility.
-Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$).
-As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see Páez et al.
-[@paez2012measuring].
 
@@
 
-Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups, can be substituted into our simple example, depending on the research question.
-The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behaviour groups and members of the family of accessibility measures.
+Any set of concepts representing population, opportunities, and their associated travel behaviour, whether representing the entire region uniformly (as will be demonstrated) or representing specific subgroups, can be substituted into our simple toy example.
+The purpose of the following simple example is to demonstrate the calculation and interpretation of the four accessibility measure variants.
 
 ## Unconstrained accessibility
-
-Setting the balancing factor $k$ to 1 or omitting it completely in @eq-access-01 results in the unconstrained accessibility case:
-
-$$
-V^0_{ij} = 1 *W_j^{(2)} f(c_{ij})
-$$ {#eq-unconstrained-access}
-
-In this case, the partial sum of spatial interaction is simply identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
+<!-- fix-->
+In $V^0_i$, no proportional allocation factor is defined, simply $f(c_{ij})$ is used to weight the number of opportunities at each $j$ and the weighted values for each $j$ are summed for each $i$, yielding the an expression identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
 
@@
 
-The sum of the unconstrained accessibility values for each origin $V^0_{i}$ generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), since arbitrarily setting $k$ to 1 (or neglecting $k$ all together) strips the values of any meaningful unit-based interpretation, the units are in 'summed opportunities by some travel impedance value'.
-Moreover, comparisons of these $V^0_{i}$ values across different contexts such as different impedance functions $f(c_{ij})$ or varying number of zones exacerbates this issue as the units between $V^0_{i}$ values change i.e., comparisons between a value of 'summed opportunities by a travel impedance value' to a value of 'summed opportunities by another travel impedance value' are not directly interpretable.
-From this perspective, the raw unconstrained accessibility scores are not intuitively comparable across different contexts and decay functions.
-They should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).
-
-Returning to our numeric example, the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in displayed in @tbl-simple-example-unconstrained-accessibility.
+However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are ‘opportunities weighted by travel impedance’ and lack meaningful scaling or direct interpretability.
+Comparing values across different decay functions or contexts (i.e., different number of zones) is therefore limited to ordinal statements (more vs.
+less), not intervals or ratios (i.e., the magnitude of differences).
 
@@
 
+For example, Table @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function.
+Comparing across decay types is meaningless in absolute terms.
+For instance, the difference in zone 1 (edge of urban core)’s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
+These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
+The fundamental uninterpretability of what is a _opportunity-weighted-travel-impedance_ unit remains.
+
 As the different impedance functions represent different travel behaviours, comparing the raw unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
-For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
-These two values are a product of different impedance functions, making the comparison uninterpretable in absolute terms.
-Likewise, we could compare values within the same travel behaviour scenario across different zones, as they are in the same units, however the issue of unit interpretability will also be apparent.
-Considering the most decaying scenario $f_1(c_{ij})$ and zone 1 (the zone with a healthcare cluster at the edge of the city): zone 1 captures `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer physicians-minute$^{-1}$ than zone 3 (urban core).
-Again, the fundamental uninterpretability of what is a _physicians-minute$^{-1}$_ or _opportunity-weighted-travel-impedance_ unit remains.
-
 While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select impedance functions to facilitate comparison across scenarios (potentially at the expense of accurately reflecting travel behavior), such adjustments may introduce bias.
-Hence, the raw unconstrained accessibility values themselves are challenging to compare due to their units.
-To enable more meaningful comparison, the following sections will detail the introduction of constraining constants to ensure consistent units across scenarios and demonstrate results on the same numeric example.
+The unconstrained scores are best used for ranking within a single context.
+
+The next sections introduce constraints to calibrate these measures for better interpretability and comparability, applying each to this example in turn.
 
@@
 
-In the total constrained accessibility case, a total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ based on the regional sum of $V^0_i$ and the total population or opportunities in the region.
-Alternatively, we reformulate this case using a proportional allocation constant, which allocates opportunities (or population) proportionally based on the the travel impedance and the total population or opportunities in the region.
+The total constrained accessibility case can be interpreted in a few ways.
+In the one that connects to the status quo: the total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ so their total sum of $V^0_i$ matches a known system total--either total opportunities or total population.
+Another interpretation is in reformulating the equation to use a a proportional allocation constant based on the total balancing factor.
+The proportional allocation constant distributes opportunities (or population) proportionally by travel impedance.
+
 In both formulations, all zonal values become a proportion of a known system total, be it the regional opportunities or regional population depending on the variant.
 
-We define two variants for this case: (a) $V_i^T$ where accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and (b) $M_j^T$, where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total number of population and to be interpreted as constrained 'market potential'.
+We define two variants for this case:
+
+- $V_i^T$: accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and
+- $M_j^T$: where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total number of population and to be interpreted as constrained 'market potential'.
 
 ### Total constrained accessible opportunities: Hansen's accessibility with a total constraint
-
-Unlike in @eq-access-01, the proportionality constant $k$ is retained. For the total constrained case, it is represented as $K^T$:
-
-$$
-V^T_{ij} = K^T  W_j^{(2)}  f(c_{ij})
+In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance.
+The total constrained accessibility from $i$ to $j$:
+$$
+V^T_{ij} = \kappa_{ij}^T D
 $$ {#eq-total-constrained-access}
 
-In this way, the total constrained accessibility measure now becomes Hansen's accessibility with a balancing factor $K^T$:
-
+This formulation satisfies the total constraint, analogous to the one in the Wilson framework:
+$$
+\sum_i \sum_j V^T_{ij} =  D
+$$ {#eq-total-constraint-D}
+
+Next, the proportional allocation factor $\kappa_{ij}^T$ determines the share of total opportunities assigned to each origin–destination pair, based on the relative proportion of opportunities-weighted travel impedance:
+$$
+\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+$$
+This renders $V_i^T$ (equal to $\sum_jV_{ij}^T$) into units of opportunities (e.g., physicians), and allows direct interpretation and comparison of results between zones and scenarios.
+
+Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $K^T$, which scales Hansen’s unconstrained accessibility $V_i^0$ to meet the total opportunity constraint:
 $$
@@
 
-Imagine that the only system known is the total number of opportunities $D$ in the region.
-Accordingly, the constant we impose in this case ensures the regional sum of total constrained accessibility is equal to the total number of opportunities as follows:
-$$
-\sum_i V^T_{i} = \sum_i\sum_j V^T_{ij} = D
-$$ {#eq-total-constraint-accessibility}
-
-This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel, congruent with Wilson's framework.
-Given the total number of opportunities in the region, we can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility to solve for $K^T$:
-
-$$
-K^T =\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
+Where the total constrained balancing factor $K^T$ is:
+$$
+K^T = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
 $$ {#eq-total-opportunity-balancing-factor}
 
-Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in [@cliff_evaluating_1974]).
-Hence, rearranging the equation to have opportunities and the proportional constant distinctly represented, our total constrained accessibility model is:
-
-$$
-V^T_i = K^T\sum_j W^{(2)}_jf(c_{ij}) = \sum_j D\frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the proportional allocation factor for the total constrained opportunity case $\kappa_{ij}^T$ is dimensionless:
-
-$$
-\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
-$$
-
-\noindent and therefore $V^T_i$ is now in the units of $D$, that is, the mass at the destination ($V^T_i = \sum_j \kappa_{ij}^T D$).
-The role of $\kappa_{ij}^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so they represent a proportion of the total number of opportunities in the region $D$.
-$\kappa_{ij}^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$.
-This is why we refer to $\kappa_{ij}^T$ as a proportional allocation factor.
-On the other hand, the proportionality constant $K^T$ balances the units of $V^0_i$, the Hansen-type accessibility values, and is an alternative expression of the total constrained accessibility measure.
-
-Referring back to our simple numeric example, balancing factor $K^T$ for the most decay travel behaviour scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
-
-$$
-\begin{array}{l}
-K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})}\\
-K^T = \frac{D}{\frac{W_1^{(2)}}{c_{11}^3}+\frac{W_1^{(2)}}{c_{21}^3} + \frac{W_1^{(2)}}{c_{31}^3} + s + \frac{W_3^{(2)}}{c_{31}^3} + \frac{W_3^{(2)}}{c_{32}^3} + \frac{W_3^{(2)}}{c_{33}^3}
-}\\
-K^T = \frac{490}{0.6233422} \\
-K^T = 786.085
-\end{array}
+This expression is consistent with Wilson's entropy-maximizing framework and analogous to the total flow spatial interaction model (e.g., Equation 2.11 in [@cliff_evaluating_1974]).
+
+In summary, $\kappa_{ij}^T$ proportionally allocates the total number of opportunities $D$ to each origin–destination pair based on relative opportunity-weighted travel impedance.
+These values can be aggregated across destinations to obtain total constrained accessibility at each origin.
+Alternatively, the measure can be expressed using the balancing factor  $K^T$, demonstrating that it is algebraically proportional to unconstrained accessibility $V_i^0$, but with interpretable units (i.e., opportunities).
+This allows for meaningful comparisons of differences across zones and travel behaviour scenarios.
+
+Referring back to our simple numeric example, $K^T$ for the highest decay t scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
+
+$$
+K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})} = \frac{490}{0.6233} = 786.085
 $$
@@
 
-Using the calculated proportionality constants for all zones and multiplying them by the unconstrained accessibility value $V^0_i$, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.
-$\kappa_{ij}^T$ for each zone is not shown, but can be understood to be the unitless proportion of opportunities (of the total opportunities) allocated to each zone based on travel impedance.
+$K^T$ for other decay scenarios are calculated similarly in code.
+Applying each $K^T$ to the unconstrained values $V_i^0$ yields total constrained accessibility values (Table @tbl-simple-example-total-opportunity-accessibility), all in units of physicians.
 
@@
 
-In contrast to unconstrained accessibility $V^0_i$, imposing a constraint -in this case a total opportunity constraint for this variant- allows for the comparison of differences and ratios between regions and across different travel behaviour scenarios as well.
-Each value is effectively in units of physicians, with the impedance units already accounted for by $\kappa_{ij}^T$.
-
-Considering the highest decay scenario ($f_1(c_{ij})$), zone 1 (a healthcare cluster at the edge of the city) captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`) like in the unconstrained accessibility case.
+Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
+For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (`r tot_acc$V_tot_i_1[1]`), like in the unconstrained accessibility case.
 However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
@@
 
-One can also directly compare values at a specific zone due to the consistent units.
-For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case.
-However, the difference between travel behaviour scenarios differ in direction.
-Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively.
-Why?
-$\kappa_{ij}^T$ ensures proportional allocation for each travel behaviour scenario.
-Meaning, while the unconstrained accessibility increases, $\kappa_{ij}^T$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region).
-As the decay behaviour decreases, more opportunities are accessible for all zones.
-In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones.
-However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).
-
-Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of Hansen's [@hansen1959] accessibility measure.
-Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.
-This is true for other constrained cases of the family of accessibility measures.
+One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
+As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
+This averaging of the total amount is how the total constrained proportional allocation factor functions.
+However, what's notable is how Zones change between scenarios.
+For instance, Zone 1’s share only declines slightly--these declines are outpaced by Zone 2’s relative gains.
+This shift reflects how $\kappa_{ij}^T$ redistributes opportunities in proportion to impedance under different travel behaviours.
+
+The total constrained accessibility measure resolves the interpretability issue of Hansen’s accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total--enabling robust comparisons across zones and scenarios--but also by keeping values proportional to $V_i^0$ so interpretation is similar.
 
@@
 
-Another variant of the total constrained accessibility measure is the _total constrained accessible population measure_, which represents the transpose of $i$ to $j$ of the total constrained accessible opportunities measure.
-This variant, expressed in @eq-total-constrained-market, represents an expression of the concept of market potential (i.e., potential users) as proposed in Harris [@harris_market_1954] and Vickerman [@vickermanAccessibilityAttractionPotential1974], and which Reilly earlier referred to as 'potential trade territories' [@reilly1929methods].
-This unconstrained form of market potential $M_j^0$ (@eq-unconstrained-market), effectively the $i$ to $j$ transpose of $V^0_{ij}$, has been used in recent research to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
-Put another way, market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination.
-However, like $V_{ij}^0$, issues of unit interpretability arise in $M_j^0$'s unconstrained form.
-To address this, the constrained variant, the total constrained accessible population measure $M^T_{j}$, is introduced.
-To formulate this variant, the total balancing factor $K^T$ is applied to the mass of the *population* at $i$ ($W_i^{(1)}$) instead of the opportunities at $j$.
-
-$$
-M^T_{ji} = \hat K^T  W_i^{(1)} f(c_{ij}) = \hat K^T  M_{ji}^0
+This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield s measure of market potential--the number of people who can spatially interact with a destination.
+
+Though not outlined in the "Unconstrained accessibility" section, the unconstrained form aligns with Reilly's "potential trade territories" [@reilly1929methods] and Harris’ and Vickerman’s formulations of regional market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+In its unconstrained form, market potential has also been more recently to estimate potentially accessible populations following infrastructure investments [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
+Market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination.
+
+However, like $V_{ij}^0$, issues of unit interpretability arise in unconstrained market potential $M_j^0$.
+To address this, we introduce the total constrained accessible population measure $M^T_{ji}$, which allocates the total population $O$ across all origin-destination pairs proportionally:
+
+$$
+M^T_{ji} = \kappa_{ji}^T O
 $$ {#eq-total-constrained-market}
 
-\noindent with $M_j^0$ being the $i$ $j$ transpose of $V^0_{ij}$:
-
-$$
-M_j^0 = W_i^{(1)} f(c_{ij})
-$$ {#eq-unconstrained-market}
-
-The total constrained accessible population measure (market potential) then becomes:
-
-$$
-M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij})
-$$
-
-Where we impose the total system known as a constraint, i.e., that the total market potential equals the total population $O$ in the region:
-
-$$
-\sum_j M^T_{j} = \sum_i\sum_j M^T_{ji} = O
-$${#eq-total-constraint-market}
-
-Substituting @eq-total-constrained-market in @eq-total-constraint-market, and solving for $\hat K^T$, we obtain:
-
-$$
-\hat K^T=\frac{O}{\sum_i\sum_i M^T_{ji}} = \frac{O}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
-$${#eq-total-population-balancing-factor}
-
-The constrained market potential then takes the following form:
-$$
-M^T_j = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = O \sum_i \frac{W_i^{(1)} f(c_{ij})}{\sum_i\sum_j W^{(1)}_i f(c_{ij})}
-$$
-
-Where the following $\hat\kappa_{ji}^T$ proportional allocation factor is dimensionless, reflecting the proportional allocation of population in the region $O$ (i.e., $M^T_j = \sum_i \hat \kappa_{ji}^T O$)
-$$
-\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}
-$$
-
-Returning back to the numerical example, the balancing factor $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
+\noindent Subject to the total constraint::
+$$
+\sum_i \sum_j M^T_{ji} =  O
+$$ {#eq-total-constraint-O}
+
+Next, $\hat \kappa_{ij}$ is the total constrained proportional allocation factor, a dimensionless term which distributes population based on impedance-weighted accessibility: 
+$$
+\hat \kappa_{ij}^T = \frac{W^{(1)}_i f(c_{ij})}{\sum_i\sum_j W^{(1)}_if(c_{ij})}
+$$
+
+This renders $M_j^T$ (equal to $\sum_jM_{ji}^T$) into units of population.
+Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $\hat K^T$, which scales unconstrained market potential $M_j^0$ to meet the total population constraint:
+$$
+M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = \hat K^T  M^0_j
+$$ {#eq-total-constrained-market-potential}
+
+Where the total constrained balancing factor $\hat K^T$ is:
+$$
+\hat K^T = \frac{D}{\sum_i\sum_j W^{(1)}_if(c_{ij})}
+$$ {#eq-total-population-balancing-factor}
+
+In summary, $\hat \kappa_{ij}^T$  allocates the total number of population $O$ proportionally to each origin–destination pair based on relative population-weighted travel impedance.
+As well, the measure can be expressed using the balancing factor  $\hat K^T$, demonstrating that it is algebraically proportional to unconstrained market potential, but also yielding interpretable units that allow for meaningful comparison.
+
+Returning to the numerical example, the balancing factor $\hat K^T$ is solved for each travel behaviour scenario, and the market potential of each zone $M^T_j$ is expressed as units of population (e.g., the number of people accessible from each origin at that destination) in @tbl-simple-example-total-population-accessibility.
 
@@
 Readers may note the difference in trends in accessible population (@tbl-simple-example-total-population-accessibility) and accessible physicians (i.e., the preceding subsection, @tbl-simple-example-total-opportunity-accessibility).
+
 In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
@@
 As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
-Overall: like the total constrained accessible opportunities case, this variant allows for the interpretation of both ordinal and interval comparisons of the raw values themselves.
+
+Like in the total constrained accessible opportunities variant, the total constrained accessible population enables direct comparison of raw values, supporting both ordinal and interval interpretations across space and travel behaviour scenarios.
 
@@
 
-Similar to the total constrained accessibility measure, the singly constrained case includes a balancing factor that adjusts the unconstrained zonal accessibility values $V^0_i$ such that a _single_ constraint is satisfied.
-Two variants are defined: (a) the _singly constrained accessible opportunities_ case (an alternative formula to spatial availability in Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023]), and (b) the _singly constrained accessible population_ case, its transpose, or singly constrained market potential.
-
-Unlike the total constraint (i.e., @eq-total-constraint-accessibility), the single constraint--as will be defined in @eq-opportunity-constraint and @eq-population-constraint for the first and second variants-- incorporates additional information.
-In the opportunities-accessible variant, the associated balancing factor constrains the 'potential' for spatial interaction by ensuring that only a proportional amount of opportunities at each destination are allocated to 'demanding' origins.
-This allocation is informed by demand, or the relative amount of population, and associated travel impedance connecting the zones.
-In the population-accessible variant (i.e., the $i$ -> $j$ transpose of the first variant), population at each origin is proportionally allocated to destinations based on the share of opportunities and travel impedance.
-
-In both variants, the singly constrained accessibility measure introduces population-based (or opportunity-based) competition at the zonal level, unlike the total constraint, which more simply allocates a fixed regional total of opportunities (or population depending on the variant).
-In sum, all singly constrained zonal accessibility values are both a proportion of the known regional opportunity (or population) total _and_ a sum of a _balanced_ proportion of opportunities allocated from each destinations (or population allocated from each origin).
-Each zonal value remains in units of opportunities accessible (or population accessible).
+The singly constrained accessibility case can also be expressed in two variants, each defined by the direction in which a constraint is applied:
+
+- $V_i^S$: accessibility constrained by opportunities at destinations (singly constrained accessible opportunities), and
+- $M_j^S$: its transpose, constrained by population at origins (singly constrained accessible population, or market potential).
+
+Similar to the total constrained case, the singly constrained measures adjust unconstrained zonal accessibility values ($V_i^0$ or $M_j^0$) using a balancing factor to satisfy the known system constraint.
+However, unlike the total constraint (which enforces a global/total sum), the singly constrained case applies a localized constraint at one end of the interaction--either origin or destination.
+
+In the opportunities-constrained variant, the balancing factor ensures that only a proportion of opportunities at each destination are allocated to origins, based on their relative demand (population) and travel impedance.
+This variant mirrors the concept of spatial availability as discussed in Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023].
+In the population-constrained variant, the logic is reversed: population at each origin is allocated proportionally across destinations, informed by the distribution of opportunities and impedance.
+
+In both cases, the singly constrained formulation introduces zonal-level competition, unlike the total constrained case which distributes a fixed regional sum.
+Each zonal accessibility value becomes not only a fraction of the regional total (opportunities or population), but also a balanced sum of interactions, weighted by impedance and relative competition.
+The result remains in interpretable units--accessible opportunities or accessible population--but reflects more complex spatial dynamics.
 
@@
 
-To demonstrate this variant formulaically, we begin with the opportunity constraint (@eq-opportunity-constraint) as our known piece of information.
-Namely, the sum of accessible opportunities from a destination should equal the number of opportunities $D_j$ at that destination.
-As the number of opportunities at each $j$ are known, it is represented as $D_j$ instead of $W_j^(2)$ as in the total constrained case.
-This constraint should hold for all destinations in the region.
-This is comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework.
+In this singly constrained variant, accessibility is constrained at the destination side: the sum of accessible opportunities allocated from each destination must equal the known number of opportunities $D_j$.
+This is comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework:
 
@@
 
-Further, we define the following proportional allocation factor:
+Further, we can express the formula even more simply, by defining the following proportional allocation factor:
 
@@
 
-Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] have shown that the role of $\kappa^S_{ij}$ is to allocate opportunities $D_j$ proportionally to the mass at each origin $i$ and the impedance between $i$ and $j$.
-As in the total constrained opportunity case, $\kappa^S_{ij}$ is dimensionless and $V_i^S$ is in the units of opportunities $D_j$.
-The singly constrained accessibility measure in @eq-attraction-constrained-accessibility-with-proportional-allocation-factor is called spatial availability by Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023], because it represents the number of opportunities that can be reached *and* are available, in the sense that accessible opportunities have been proportionally allocated based on relative demand, travel impedance and the regional total number of opportunities, i.e., spatial competition for them has been considered.
-These authors also show that the following expression (accessibility per capita) is a constrained version of the popular two-stage floating catchment area measure of Shen [@shen1998] and Luo and Wang [@luo2003]:
+This formulation has been referred to as **spatial availability** by Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023], since it incorporates spatial competition by allocating opportunities based on demand (population), impedance, and the known opportunity totals $D_j$.
+The dimensionless factor $\kappa^S_{ij}$ ensures that each destination’s opportunities are distributed proportionally to origins.
+As in the total constrained case, $V_i^S$ is expressed in the units of accessible opportunities.
+Finally, the **per capita** version of this measure:
+
+@soukhovIntroducingSpatialAvailability2023 also showed that the following expression (accessibility per capita) is a constrained version of the popular 2SFCA approach of Shen [@shen1998] and Luo and Wang [@luo2003]:
 
@@
 
-<!-- To reinforce this point, @eq-opportunity-constraint also implies that the total constraint (e.g., $\sum_i V^S_{i} = \sum_i\sum_j  V^S_{ij} = D$ in @eq-total-constraint-accessibility) is maintained.
--->
-<!-- Due to the constraints, $\frac{V_i^S}{O}$ can be interpreted as the proportion of opportunities accessible *and* available to location $i$ out of the total number of opportunities in the system.
--->
-
-Returning to the simple numeric example, the opportunity-constrained case would yield the following $B_{j}$ for $f_1(c_{ij})$:
+Returning to the simple numeric example, as an example of the solved $B_{j}$ for the highest decay travel behaviour $f_1(c_{ij})$:
 
@@
 
-The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 is `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, and for $f_3(c_{ij})$ decay group is `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
+The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 area `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, respectively.
+For the $f_3(c_{ij})$ decay group, they are  `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
 Using these these balancing constants, we can calculate the singly constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).
@@
 
-However, unlike the total constrained opportunity case, $\kappa^S_{ij}$ reflects zonal competition based on the mass of the origin, i.e., population.
-Again, consider the highest decay scenario $f_1(c_{ij})$.
-Under this scenario, zone 1 no longer captures a medium amount of physicians as in the total constrained opportunity case: it now captures the fewest in the region i.e., `r opp_acc$V_opp_i_1[1]` at zone 1, `r opp_acc$V_opp_i_1[2]` at zone 2, and `r opp_acc$V_opp_i_1[3]` at zone 3.
-Why may zone 1 (healthcare cluster at the edge of the city) capture `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()` of the physicians regionally while this same zone captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` in the total opportunity constrained case?
-
-This difference is due to the single-opportunity factor $\kappa^S_{ij}$'s role in $V_i^S$.
-The only inputs required in the total-opportunity factor $\kappa^T_{ij}$ is the total number of opportunities in the region $D$ as well as the associated opportunities at each $j$ and travel impedance.
-Opportunities are allocated to $i$s, regardless of the mass weights of origin.
-Whereas the single opportunity constraint $\kappa^S_{ij}$ requires the population at $i$ as an input.
-In fact, $\kappa^S_{ij}$ is calculated as the proportion of impedance-weighted population at an $i$ to the sum of impedance-weighted population for the entire region.
-Hence, since zone 1 has the lowest population in the region, is in close proximity to a more populated zone (zone 3, the 'urban core'), and is not well connected (in terms of travel impedance) to other zones with opportunities, $V_{1}^S$ values, or the number of physicians accessible, is lowest.
-
-Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
-For instance, consider Zone 2 which has the highest population.
-It is more evident in the $f_3(c_{ij})$ scenario than in higher decay scenarios that this zone does not have an exceptionally large population for the region - Zone 2 only represents `r (opp_acc$O_i[2] |> as.numeric() / opp_acc$O_i |> as.numeric() |>sum()) |>scales::percent()` of the population in the three zone region.
-In this sense, other zones are not *that* disadvantaged, and in this scenario with unfettered travel cost, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
-Zones 1 and 3 are allocated opportunities at relates similar to their relative population size.
-
-Readers may also notice the change in the proportion of opportunities drawn from different zones depending on the travel behaviour scenarios considered.
-For instance, Zone 2 has the highest zonal population, representing 50% of the 200,000 regional population.
-However, due to its high relative travel distance from the other zones, its population is less competitive in capturing opportunities in the high-decay travel scenario ($f_1(c_{ij})$): Under this scenario, zone 2 captures almost exclusively opportunities from its own zone.
-However, in $f_3(c_{ij})$, the scenario with unfettered travel cost, Zone 2 captures by far the most number of physicians.
-But in this scenario, Zones 1 and 3 also take opportunities from Zone 2 (i.e., Zone 1 and Zone 3 takes `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` and `r ((opp_acc_ij$k_opp_ij_3[6])-(opp_acc_ij$k_opp_ij_1[6]))|> scales::percent()` more from Zone 2 between $f_3(c_{ij})$ and $f_1(c_{ij})$ scenarios hence $\kappa_{2,2}^S$ decreases by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`).
-Zones 1 and 3 are allocated opportunities at rates similar to their relative population size.
+However, unlike $\kappa^T_{ij}$, $\kappa^S_{ij}$ introduces zonal competition based on the mass of the origin (population).
+In the total constrained case, opportunities are distributed based on impedance alone, regardless of population at $i$.
+In contrast, the singly constrained case allocates each zone’s opportunities proportionally across the region based on the relative impedance-weighted demand from all origins.
+
+This consideration has important implications.
+Consider the highest decay scenario $f_1(c_{ij})$.
+Under this scenario, Zone 1--despite hosting a medium amount of physicans--captures the fewest physicians (`r opp_acc$V_opp_i_1[1]`), compared to `r opp_acc$V_opp_i_1[2]` at Zone 2, and `r opp_acc$V_opp_i_1[3]` at Zone 3.
+Why?
+Zone 1 has the smallest population and is adjacent to Zone 3, the urban core.
+Its low impedance-weighted demand means $\kappa^S_{ij}$ allocates it fewer opportunities.
+By contrast, in the total constrained case, Zone 1 fares better, capturing  `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` of all physicians (compared to `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()`).
+
+As travel decay decreases (e.g., $f_3(c_{ij})$), competition becomes more diffuse.
+Zone 2, with the largest population, initially dominates accessibility—but under low decay, Zones 1 and 3 also begin drawing more opportunities from Zone 2.
+For instance, Zone 1 gains  `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` more from Zone 2 in $f_3(c_{ij})$ than in $f_1(c_{ij})$.
+This shift reflects a drop in $\kappa^S_{2,2}$ by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`,  showing Zone 2’s decreasing hold on its own opportunities as other zones gain accessibility 'parity'.
+
+This dynamic reveals how $\kappa^S_{ij}$ embeds both travel impedance and population competition.
+Unlike the total constraint lowest decay scenarios that allocate evenly, the singly constrained case reflects how competition evolves continues to influence allocation.
 
@@
 As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
-Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all around 24 physicians accessible per capita).
-
-This trend mirrors how the accessibility values in the total constrained opportunity case stabilises to the $V_i^T$ regional average (e.g., the accessible opportunities allocated to each of the three zones approaches a third of 490 physicians, or 163.33, under the unfettered mobility scenario $f_3(c_{ij})$).
-These patterns make intuitive sense: the balancing factors act as regional and/or zonal averaging mechanisms.
-In distance decay travel behaviour scenarios that are more relatively uniform (i.e., low for all zones like in $f_3(c_{ij})$), what remains is the relative effect of the other variables in the balancing factor.
+Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all ~24 physicians accessible per capita).
+
+This convergence mirrors the trend in the total constrained opportunity case, where accessibility values approach a third of the 490 physicians under the unfettered mobility scenario $f_3(c_{ij})$.
+In both cases, the balancing factors ($K^S$ and $B_j$) act as averaging mechanisms but at different scales.
+As distance decay becomes more _relatively_ uniform, the role of remaining variables (i.e., total population or opportunities) drive the proportional allocation differences.
 In the total constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constrained case, this is the population at a zone relative to the regional population.
@@
 
-In this case, the measure is singly constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+In this variant, the measure is singly constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
 
@@
 
-Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression:
+Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression of accessible population $M_j^S$ being the product of proportionally allocated ($\hat \kappa^S_{ji}$) popultion:
 
@@
 
-This accessibility case requires zonal populations and opportunities to match one-to-one, like the doubly constrained spatial interaction model.
-In this way, doubly constrained accessibility can be thought as "access" or simply spatial interaction (no potential).
-
-To contextualize this point, the total and singly constrained accessibility measures discussed thus far have used either $O_i$, $D_j$, or the regional sums of either, but never both simultaneously.
-For example, when opportunities $D_j$ are used to constrain @eq-opportunity-constrained-accessibility in the _singly constrained accessible opportunities measure_, the specific mass of the population at origin $i$ demanding only those opportunities at that $j$ is unknown.
-Instead, only the population at $i$ demanding opportunities in the region is known, and this is more generally represented as $W_i^{(1)}$ (as in @eq-opportunity-constrained-proportionality-constants).
-Similarly, when the population $O_i$ is used as a constraint in @eq-population-constrained-accessibility in the _singly constrained accessible population measure_, the mass at the destination is given by $W_j^{(2)}$ (also in @eq-population-constrained-proportionality-constants) since only the mass of opportunities at each $j$ is known and information on what opportunities are allocated to what zone is unknown.
-
-By contrast, the double-constrained accessibility case requires that both populations and opportunities match.
-Meaning, opportunities at each destination can be accessed by all populations, while at the same time, the population at each origin can be accessed by all opportunities.
-However, this requirement is often unintuitive in traditional accessibility analysis.
-Namely, the distinction between population (origin masses) and opportunities (destination masses) typically represent different entities without a shared unit of measurement.
-On the population side, we usually count people; on the opportunity side, we may be referring to physicians, clinics, grocery stores, schools, parks, or libraries.
-In a few cases, a one-to-one correspondence or their own capacities at which they interact and are interacted with may exist e.g., one person with one job.
-Another similar opportunity type example is healthcare, e.g., one person and one unit of capacity, e.g.
-a vaccine shot.
-
-A doubly constrained approach to accessibility calculation needs a one-to-one relationships between population and opportunities to be present.
-Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality):
+This accessibility case adopts the structure of the doubly constrained spatial interaction model, where $V_{ij}^D$ flows are constrained by both origin populations $O_i$ and destination opportunities $D_j$.
+That is, the resulting accessibility outflow from each origin must match the origin’s population demand, and the resulting accessibility inflow to each destination must match the number of opportunities supplied:
+
+$$
+\sum_j V_{ij}^D = O_i \text{ and }  \sum_i V_{ij}^D =  D_j
+$$ {#eq-opportunity-population-equality-1}
+
+Because both margins are fixed, this measure reflects fully allocated interaction flows, not latent potential.
+As such, accessibility cannot be interpreted as a traditional summary at $i$ or $j$ (e.g., "opportunities accessible from $i$")--those sums simply reproduce the constraint totals.
+Instead, the meaningful unit of analysis is the $ij$ flow itself.
+
+This distinguishes doubly constrained accessibility from the total and singly constrained cases discussed previously.
+In those cases, only one side of the interaction--either the total marginal or opportunity/population marginal--was constrained, while the other was treated as a demand/supply weight (e.g., $D$/$O$ or $W_j^{(2)}$/$W_i^{(1)}$), without any allocation details.
+As a result, total constrained models express more 'potential' from both sides of the system, and singly constrained models express potential access from one side.
+
+By contrast, the doubly constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly.
+This makes it less suitable for traditional accessibility analysis, namely because origin and destination masses often differ in kind and units.
+For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not known.
+A doubly constrained approach only makes conceptual sense when population and opportunities are comparable, have a one-to-one correspondence or are paired somehow--e.g., job per worker, student per school-seat, or vaccine doses per person.
+
+A doubly constrained approach to accessibility adopts the structure of the production-attraction spatial interaction model, where both population and opportunity totals are fixed.
+Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):
 
@@
 \sum_i O_i = \sum_j D_j
-$$ {#eq-opportunity-population-equality}
-
-As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$ as shown in @eq-total-constraint-accessibility.
+$$ {#eq-opportunity-population-equality-2}
+
+As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$.
 
@@
 193-195].
-The proportional allocation factor $\kappa_{ij}^D$ would then be:
+
+The doubly constrained model completely distributes origin populations to destination opportunities according to travel impedance and supply-demand balance.
+This ensures that: summing $V^D_{ij}$ across $j$ returns $O_i$; summing across $i$ returns $D_j$.
+Thus, aggregating over $i$ or $j$ yields only the known constraints.
+In this way, a per-capita form (e.g., $V^D_i / O_i$) is not meaningful--since the output already reflects population-normalized allocation.
+As such, the $ij$ matrix $V^D_{ij}$ is the only interpretable output.
+
+We could define the proportional allocation factor $\kappa_{ij}^D$ such that:
 
@@
 $$
-
-One could rewrite @eq-doubly-constrained-accessibility as an origin summary expression of proportionally allocated opportunities:
-
-$$
-V^D_{i} = \sum_j \kappa^D_{ij} D_j
-$$ {#eq-doubly-constrained-accessibility-w-paf}
-
-However, $V^D_{i}$ is not wholly helpful, as it will simply equal the origin-mass marginal.
-In this way, only $V^D{ij}$ values are interpretable.
-Furthermore, unlike in the total and singly constrained cases, the doubly-constrained case does not have an interpretable per capita version.
-For instance, representing $V_{ij}^D$ per capita is not meaningful, as the value *already* matches population to opportunities.
-Following this logic, the market potential form $M^D_{ij}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: i.e., the opportunities accessed from $j$ at an $i$ vs.
+\noindent and represent $V^D_{ij}$ as equal to $\kappa^D_{ij} D_j$, allowing the analyst to understand the proportional allocation of D_js to each $ij$ flow.
+
+Following this logic, the market potential form $M^D_{ji}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: i.e., the opportunities accessed from $j$ at an $i$ vs.
 the population accessed from $i$ at a $j$.
@@
 
-To calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be modified.
-Namely, a count of physician _capacity_ per destination is needed instead of just the number of physicians, as used to calculated total and singly constrained cases.
-We also must be able to clearly state that the population is a count of people seeking opportunities at the new capacities, i.e., the population must reflect the _capacity_ of the population to interact with opportunities.
-
-So, this adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
-For the population, we leave this unchanged numerically but theoretically know that each person interacts with one physician capacity (i.e., our opportunities with a _capacity_).
-Hence, the number of opportunities per destination is new: the example is modified such that the physician-capacity at each zone is an approximately scaled version of the number of destination-side physicians at each zone from the unmodified example (@tbl-small-system-land-use).
-To emphasis the new definition of 'provider' as physician capacity, the new system PPR is simply `r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
+Moving onto the toy example, to calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be reinterpreted.
+
+Namely, a count of physician _capacity_ per destination $D_j$ is needed instead of simply the number of physicians.
+Also, we must be able to clearly state that the population is the _capacity_ of the origin to interact with opportunities $O_i$, i.e., the count of people seeking opportunities.
+
+This adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+For the population, we leave this unchanged numerically but theoretically note that each person interacts with one physician capacity.
+However, the number of providers per destination is revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).The system-wide provider-to-population ratio (PPR) is now:`r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
+
 We keep the same zonal cost matrix, and travel impedance functions for three types of travel behaviour as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
@@
 
-However, despite the modifications to the example, our objective remains the same as in previous cases: to measure accessibility under different travel behavior scenarios.
-Specifically, we aim to quantify the number of _potential_ spatial interactions between the physician-seeking population in each zone $i$ to the physician capacity in a zone $j$ i.e., $V_{ij}^D$.
+However, despite the modifications to the example, our objective remains the same: to measure accessibility under different travel behavior scenarios.
+We aim to quantify the number of _potential_ spatial interactions between the physician-seeking population in each zone $i$ to the physician capacity in a zone $j$ i.e., $V_{ij}^D$.
 The highest decay travel behaviour scenario ($f_1(c_ij)$) is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
@@
 
-As mentioned, accessibility values are typically interpreted as a summary of the proportionally allocated opportunities at each $i$.
-Hence, in interpreting the doubly constrained accessibility from @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values, $V_i^D$ values (i.e., the sum of values at all three $j$ destinations for each origin $i$) would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()` physician-capacity accessible for Zones 1, 2 and 3 respectively.
-This approximately equal to the number of population at each of these zones.
-Conversely, the market potential $M_j^D$ interpretation of these values would be `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(did)|>summarize(M_dc_j_1=sum(V_dc_ij_1)))[3,2] |> pull()` people accessible from Zones 1, 2 and 3 respectively, equal to the number of *opportunities* (physician-capacities accessible) at each of these zones.
-Notice, the mass weight at the origin equals the mass weight of the destination: this is precisely the function of the double constraint.
-In other words, $V_i^D$ is the number of accessed opportunities and $M_j^D$ is the number of population accessed.
-For the other two travel behaviour, identical $V_i^D$ and $M_j^D$ values are calculated, following the same logic.
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physican-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()`.
 Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
-
-For instance, differences in $V_{ij}^D$ values between travel behaviour scenarios are notable.
-These values can be directly compared to discuss mass and distance decay impacts.
-Examining Zone 2 (the bedroom community), @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2 demonstrates these $i$ to $j$ access values for this more relatively remote, higher-populated and lower-opportunity rich zone.
-It can be observed that the number of intrazonal opportunities proportionally allocated decreases as the assumed distance decay decreases e.g., from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
-Following the intuition discussed in the singly constrained opportunity case, as decay decreases, the mass effect of the population (at origin) and opportunity (at destination) is more evident: zonal opportunities are supplied and zonal populations demand at the weights assigned to these zones, with minimal decay adjustment, reflected by the proportional allocation factors $\kappa_{ij}^D$.
+$V^D_{ij}$ values represent the number of opportunities from zone $j$ allocated to populations in zone $i$, shaped by both mass and travel impedance.
+
+To illustrate this, consider the results for Zone 2 (Suburban zone--a high-population, low-opportunity, relatively remote zone).
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal allocation (i.e., $V^D_{22}$) declines as travel impedance decay decreases--from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
+
+Following the intuition discussed in the singly constrained opportunity case, as decay decreases (i.e., more relatively uniform for all zones), the mass effects (effect of the population and opportunities magnitudes) become more relatively dominant in the spatial allocation.
 
@@
 
-Recall, accessibility is defined as "the potential for interaction" and is traditionally presented as a summary zonal measure.
-In the doubly constrained case, we force the zonal population and zonal opportunities to match one-to-one, hence providing a zonal summary is no longer relevant: the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone.
-However, if what interests readers is the "potential for interaction" in the case population and opportunities to match one-to-one, reframing the investigation may be needed.
-In this sense, it would be examining "interaction" (much less room for potential) through the values of $V_{ij}^D$ e.g., how many opportunities are being allocated to an origin from a destination (or in a transposed sense for $M_{ji}^D$).
-In this sense, the doubly constrained case can be thought of as an estimate of _realized_ accessibility or access: reflecting spatial interaction.
-It is also formulaically identical to the doubly constrained spatial interaction model, but with specific interpretations of the origin and destination weights as 'population' and 'opportunity', respectively.
-
-As Wilson explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
-Accessibility, however, is often presented and understood as a zonal summary of *potential* for interaction between origins and destinations that contain inherently different units.
-Arriving at the doubly constrained case through the unconstrained, total, and singly constrained cases make the connection between the *potential* for spatial interaction (accessibility) and *realized* potential for spatial interaction more interpretable.
-Namely, by demonstrating that these members of the accessibility family all derive from the same root and can be derived from Wilson's original formulation, this more clearly demonstrates what *potential* is, within the context of spatial interaction modelling.
-
-Namely, *potential* depends on the framing of the masses at the origin and destination, and how similar they are in their units.
-The appropriate constraint should be decided based on the the input data and their similarity.
-In increasing unit similarity from the perspective of opportunity-constraint: the total-constraint can be used if population (origin mass) is not known, the single constraint can be used if population is known and matters to *potential* but it does not match the capacity of opportunities, and lastly, the double constraint is appropriate if population is known, matters and population matches the capacity of opportunities.
-These constraint measures can reflect the opportunities accessible at each zone, but reflects the assumptions embedded in the input data about the potential to spatially interact.
+Recall, accessibility is defined as "the potential of spatial interaction" and is traditionally presented as a summary zonal measure.
+In the doubly constrained case, we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, hence the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone (and $\sum_i V_{ij}^D$ being the opportunities at the zone).
+
+So, in the doubly-constrained case, accessibility flows becomes simply equivalent to spatial interaction flows--spatial proportional allocations between population and opportunity capacity.
+In fact, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $kappa_{ij}^D$ and $\hat kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
+
+As Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit; but the focus of these models have typically on $ij$ flows and often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
+On the other hand, accessibility has been interpreted differently: often understood as a zonal summary of the *potential* for spatial interaction, conceptualising origins and destinations in inherently different units.
+
+Tracing the application of increasing more restrictive constraints: from the total, to the singly- and now in the doubly-constrained measure, sheds light on how the balancing factors embedded in the unitless proportional allocation factors restrict the range in "potential" spatial interaction values that can be taken.
 
@@
 
-In this work, we examined the historical and mathematical commonalities between spatial interaction models and place-based accessibility measures.
+In this paper, we examined the historical and mathematical commonalities between spatial interaction models and place-based accessibility measures.
 As accessibility research evolved largely influenced by Hansen [@hansen1959], researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
-Firstly, we have demonstrated that by reintroducing such constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
+
+Firstly, we have demonstrated that by reintroducing Wilson's system constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
 These values become more intuitive for the purpose of analysis and comparison.
@@
 Lastly, we discussed how popular measures such as the one used in Hansen [@hansen1959] and the 2SFCA link into the family of accessibility measures.
-We summarise the family of accessibility measures, as follows.
-
-We first place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
+
+A summary of the family of accessibility measures, is detailed; First, we place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
 We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
 In other words, the total-constraint model could be a more interpretable alternative for the unconstrained case if population-competition is not relevant and one is interested in capturing the maximum _potential_; specifically, if there is a fixed number of opportunities in the region, and if it makes sense to assume that people accessing proximate opportunities leave fewer for others, *without* considering the population size at the origins.
-<!-- A.S.
-NOTE: Hansen's measure would converge to some global total if the decay effect was low.
-Imagine everyone can teleport.
-In the unconstrained case: all zones can reach all opportunities.
-In the total constrained case: all zones can reach all opportunities divided by the number of zones.
-Alternatively, if 1 zone can teleport and no other zone can travel at all, that 1 zone will get all the opportunities in the region.
--->
-
-We then introduce the singly constrained case, which *does* takes into account the population size at the origin in the allocation of opportunities (unlike the total constraint).
+
+We then introduce the singly-constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
 It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.
@@
 This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.
-
-# References

### 2025-08-01 paper_plos/paper_plos.tex
Made the existing draft as concise as possible -- shed 6 pages. Old version is paper_plos_V0.qmd and .pdf.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%

\PassOptionsToPackage{table}{xcolor}

\documentclass[
  10pt,
  letterpaper,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% marvosym package for additional characters
\usepackage{marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% Using natbib instead
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{anyfontsize}
\usepackage{multirow}
\usepackage{float}
\floatplacement{table}{H}
\floatplacement{figure}{H}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[numbers,square,comma]{natbib}
\bibliographystyle{plos2015}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A family of accessibility measures derived from spatial interaction principles},
  pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{A family of accessibility measures derived from spatial
interaction
principles} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Anastasia Soukhov\textsuperscript{1}, Rafael H. M.
Pereira\textsuperscript{2}, Christopher D.
Higgins\textsuperscript{3*}, Antonio Páez\textsuperscript{1}
\\
\bigskip
\textbf{1} McMaster University, School of Earth, Environment \&
Society, Hamilton, Canada, \\ \textbf{2} Institute for Applied Economic
Research - Ipea, Data Science Division, Brasília,
Brazil, \\ \textbf{3} University of Toronto Scarborough, Department of
Human Geography, Toronto, Canada, 
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments
sections

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* cd.higgins@utoronto.ca

\end{flushleft}

\section*{Abstract}
Transportation planning has long prioritized the efficiency of movement.
However, the concept of accessibility represents a more comprehensive
evolution, shifting focus from movement to the potential to reach--or
spatially interact with--desired destinations. Despite growing
recognition of accessibility-based planning approaches, the concept
remains fragmented, with inconsistent definitions and unclear
interpretations. To this end, this paper offers a methodological
contribution by specifying a family of accessibility measures grounded
in the shared theoretical roots of gravity-based accessibility and
spatial interaction models, particularly their balancing factors. From
this foundation, we outlined four memebers of the family: the
`unconstrained' measure (i.e., Hansen-type accessibility), the `total
constrained' measure (i.e., a constrained version of the Hansen-type
accessibility), the `singly constrained' measure (i.e., related to the
popular two step floating catchment approach - 2SFCA), and the `doubly
constrained' measure representing realized access (i.e., equal to the
doubly constrained spatial interaction model). These measures can be
interpreted as either the number of accessible opportunities or
accessible population (i.e., market potential). A toy example
illustrates how they produce interpretable, unit-based values, offering
a clearer and more coherent basis for accessibility analysis.

\linenumbers
\section{Introduction}\label{introduction}

In the early nineteenth century, rapid urbanisation driven by
industrialisation expanded cities across Great Britain, Northern Europe,
and North America. This expansion led to traffic congestion and the
emergence of a transportation planning paradigm focused primarily on
mobility. Within this new practice, access to destinations was treated
as a by-product of movement. After World War II, major investments in
automobile and transportation infrastructure cemented this
mobility-oriented model, fostering flight from urban congested cities,
lower-density sprawl, car-dependent development and entrenching
automobility in planning practice
\citep{miller_collaborative_2011, lavery_driving_2013}. Despite
continued road and highway expansion, this automobility monoculture has
proven ineffective at reducing travel costs or environmental burdens,
and has not clearly improved people's ability to reach destinations
\citep{farber_running_2011, handyACCESSIBILITYVSMOBILITYENHANCING2002, paez_healthcare_2010}.

In response, transportation researchers have increasingly advocated for
the adoption of accessibility as a planning criterion, in contrast to
traditional mobility-oriented transportation planning approaches which
translate into indicators that benchmark movement (e.g., vehicle
kilometres traveled, intersection through traffic, etc.) which are not
necessarily linked to improved accessibility
\citep{silvaAccessibilityInstrumentsPlanning2017, paez_developing_2013, handy2020, elgeneidyMakingAccessibilityWork2022}.
Accessibility, by contrast, is the ``potential of opportunities for
{[}spatial{]} interaction'' \citep{hansen1959}. While mobility reflects
movement, accessibility captures the combined influence of transport and
land use, emphasizing destinations and the potential to reach them
\citep{handyMeasuringAccessibilityExploration1997}.

Accessibility research has expanded across diverse domains including:
employment
\citep{karstEvaluationAccessibilityImpacts2003, grengs2010job, paez_jobs_2013, merlin2017competition, tao_investigating_2020},
healthcare
\citep{luo2003, paez_healthcare_2010, wan2012three, delamater2013spatial, boisjoly2017informality, pereira_2021_geographic},
green space
\citep{reyesAccessibility2014, rojas_accessibility_2016, liang_novel_2024},
education
\citep{williams_disparities_2014, romanillosAccessibilitySchoolsSpatial2018, marques_accessibility_2021},
social contact
\citep{neutens_human_2007, farberActivitySpacesMeasurement2012, farber_2013_social},
and regional economics
\citep{vickermanAccessibility1999, lopezMeasuring2008, ribeiro_road_2010, gutierrez_evaluating_2011},
among many other domains of application. Despite its popularity in
scholarly works, accessibility still remains difficult to implement in
planning due to definitional inconsistencies
\citep{vanweeAccessible2016, handy2020, kapatsila_resolving_2023} and
challenges in interpreting and communicating results
\citep{geursAccessibilityEvaluationLanduse2004, vanweeAccessible2016, ferreiraReenactingMobilityAccessibility2020}.

More specifically, the wide range of accessibility definitions--with
novel methods being more sophisticated but less intuitive
\citep{kapatsila_resolving_2023}--can further hinder practical uptake
\citep{vanweeAccessible2016}. Geurs and van Wee
\citep{geursAccessibilityEvaluationLanduse2004} classify accessibility
measures into four categories: infrastructure-, place-, person-, and
utility-based. Among place-based measures (this work's focus), variants
include gravity-based \citep{hansen1959}, cumulative opportunity
\citep{pirie_measuring_1979}, the 2 Step Floating Catchment Area (FCA)
approach \citep{luo2003}, and a variety of modifications to these
approaches e.g., Enhanced 2-Step FCA \citep{luoEnhanced2009}, 3-Stage
FCA \citep{wan2012three}, Modified 2-Step FCA
\citep{delamater2013spatial}, inverse 2-Step FCA
\citep{wang_2sfca_2021}, and n-steps FCA \citep{liang_novel_2024}. While
these methods are tailored to address specific research contexts,
overall this diversity does not demystify existing questions like those
raised in \citet{vanweeAccessible2016}: How should practitioners
interpret difference in scores between modes, and how should results be
communicated?

Rather than propose a new measure, this work argues for a return to the
spatial interaction foundations of accessibility. Specifically, we show
how the family of spatial interaction models \citep{wilson1971} can be
reformulated in the context of accessibility, namely as a ``family of
accessibility measures''. This formulation results in constrained
versions of gravity-based accessibility \citep[e.g.,][]{hansen1959}, but
in units of \emph{spatially reachable opportunities}. This approach
offers a direct mathematical link to existing accessibility measures
while restoring tangible meaning to zonal values. Instead of abstract
proportional scores, constrained accessibility expresses the number of
opportunities a population may potentially spatially interact with.

In summary, this paper makes two contributions: (1) Reviews how spatial
interaction modeling and accessibility share similar origins, but
diverged in focus and interpretation. (2) Introduces a family of
accessibility measures grounded in spatial interaction principles,
including total, singly, and doubly constrained cases and variants of
``accessible opportunities'' and ``accessible population''. These cases
and variants align with common measures such as Hansen-type
accessibility \citep{hansen1959}, competitive accessibility measures
such as the 2SFCA method \citep{shen1998, luo2003}, and market potential
models
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.

We contend that accessibility research should reengage with spatial
interaction modeling, particularly the use of Wilson's system
constraints. While spatial interaction models embraced such constraints
to improve interpretability, most accessibility models have not. This
lack of adoption contributes to fuzziness in current analyses, limiting
interpretive clarity to simple proportional comparisons (e.g.,
``higher-than,'' ``lower-than'')
\citep{millerAccessibilityMeasurementApplication2018}. Without such
constraints, accessibility scores lack clear units and comparability
across cities or modes. In contrast, constrained measures yield values
that can be tied to tangible values without any post-hoc treatment,
making them more interpretable, communicable, and actionable in
planning.

This paper lays the historical and theoretical groundwork for a
companion empirical study demonstrating how these constrained forms
restore interpretability through concrete measurement units (e.g.,
number of accessible opportunities), as opposed to the abstract weighted
outputs typical of conventional gravity models.

Accessibility and spatial interaction modelling literature share common
headwaters, and the latter has given careful attention to measurement
units and their interpretability. Following this introduction, this
paper revisits that lineage--from Ravenstein's Newtonian gravitational
expression of spatial interaction \citep{ravensteinLawsMigration1889},
through Hansen \citep{hansen1959}, to Wilson \citep{wilson1971}--to
propose a unified family of constrained accessibility measures. We
illustrate each case of the family with a numerical toy example and
conclude by discussing how these constrained measures can inform
planning decisions and improve clarity in accessibility analysis.

\section{Newtonian's roots of human spatial interaction
research}\label{newtonians-roots-of-human-spatial-interaction-research}

The patterns of people's movement in space have been a subject of
scientific inquiry for at least a century and a half. From as far back
as Henry C. Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858}, a concern with the scientific
study of human spatial interaction can be observed. It was in this work
where Carey stated that ``man {[}is{]} the molecule of society {[}and
their interaction is subject to{]} the direct ratio of the mass and the
inverse one of distance'' \citep[pp.~37-38]{mckeanManual1883}. This
statement shows how investigations into human spatial interaction have
often been explicitly coloured by the features of Newton's Law of
Universal Gravitation, first posited in 1687's \emph{Principia
Mathematica} and expressed as in Eq~\ref{eq-phys-grav-prop}.

\begin{equation}\phantomsection\label{eq-phys-grav-prop}{
F_{ij} \propto \frac{M_i M_j} {D_{ij}^{2}}
}\end{equation}

This famous equation expresses that the attractive force \(F\) between
two bodies is directly proportional to the product of their masses and
inversely proportional to the square of the distance between them. As
mass increases, so does force; as distance increases, force decreases.
However, Eq~\ref{eq-phys-grav-prop} only demonstrates a proportional
relationship. To quantify the magnitude (not the proportional magnitude)
of the force, it must be \emph{constrained} with an empirical constant.
This constant \(G\) converts Eq~\ref{eq-phys-grav-prop} from an
expression of proportionality to the following expression of equality:

\begin{equation}\phantomsection\label{eq-phys-grav}{
F_{ij} = G \frac{M_i M_j} {D_{ij}^{2}}
}\end{equation}

Where \(G\) is the gravitational constant--an empirically calibrated
value that ensures the model reflects observed forces. Newton's initial
estimate of \(G\) was based on a speculation that the mean density of
earth was between five or six times that of water, an assumption that
received support after Hutton's experiments of 1778
\citep[p.~783]{hutton_xxxiii_1778}. Still, it took over a century from
the publication of \emph{Principia} to refine the estimate of the
proportionality constant to within 1\% accuracy, with Cavendish's 1798
experiment \citep{cavendish_xxi_1798}.

This gravitational framing laid the conceptual groundwork for later
empirical studies of human spatial interaction, such as those by
Ravenstein, Stewart, and others, who began translating these principles
into observed patterns of human movement.

\section{Early research on human spatial interaction: from Ravenstein
(1889) to Stewart
(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1889-to-stewart-1948}

Following Carey's \emph{Principles of Social Science}
\citep{careyPrinciplesSocialScience1858}, research into human spatial
interaction continued in different contexts. Namely, since Carey to the
1940s, a number of researchers theoretically and empirically attempted
to characterise human spatial interaction as a force \(F\) directly
proportional to the ``masses'' \(M_i\) and \(M_j\) of two locations, and
inversely proportional to their separation distance--conceptually
parallel to Newtonian gravity, but often omitting a proportionality
constant.

Beginning with Ravenstein in the late 1880s, his works proposed some
``Laws of Migration'' based on his empirical analysis of migration flows
in various countries
\citep{ravensteinLawsMigration1885, ravensteinLawsMigration1889}. These
works posited 1) a directly proportional relationship between migration
flows and the attractive size of destinations, and 2) an inversely
proportional relationship between the size of flows and the separation
between origins and destinations. As with Carey, these propositions echo
Newton's gravitational laws.

Over time, other researchers discovered similar relationships. For
example, Reilly \citep{reilly1929methods} formulated a law of retail
gravitation, expressed in terms of equal attraction to competing retail
destinations that could be understood as `potential trade territories'.
Later, Zipf proposed a \(\frac{P_1P_2}{D}\) hypothesis for the case of
information \citep{zipfDeterminantsCirculationInformation1946},
intercity personal movement \citep{zipfHypothesisIntercityMovement1946},
and goods movement by railways \citep{zipfHypothesisCaseRailway1946}.
The \(\frac{P_1P_2}{D}\) hypothesis stated that the magnitude of flows
was proportional to the product of the populations of settlements, and
inversely proportional to the distance between them.

A common feature of these early investigations of human spatial
interaction is that a proportionality constant, similar to \(G\) in
Eq~\ref{eq-phys-grav} was never considered. Of the researchers cited
above, only Reilly and Zipf expressed their hypotheses in mathematical
terms. Reilly's hypothesis was presented in the following form:

\begin{equation}\phantomsection\label{eq-reilly}{
B_a = \frac{(P_a P_T)^N}{D_{aT}^n}
}\end{equation}

\noindent where \(B_a\) is the amount of business drawn to \(a\) from
\(T\), \(P_a\) and \(P_T\) are the populations of the two settlements,
and \(D_{aT}\) is the distance between them. Quantity \(N\) was chosen
by Reilly in a somewhat \emph{ad hoc} fashion as 1, and he used
empirical observations of shoppers to choose a value of \(n = 2\).

Zipf, on the other hand, wrote his hypothesis in mathematical form as:

\begin{equation}\phantomsection\label{eq-zipf}{
C^2 = \frac{P_1 P_2}{D_{12}}
}\end{equation}

\noindent where \(C\) is the volume of goods exchanged between \(1\) and
\(2\), \(P_1\) and \(P_2\) are the populations of the two settlements,
and \(D_{12}\) is the distance between them.

These early formulations (Eq~\ref{eq-reilly}, Eq~\ref{eq-zipf}) clearly
reflect the influence of Newtonian gravity on human spatial interaction
theory, revealing a shared mathematical structure across migration,
trade, and communication models. However, none included a
proportionality constant--highlighting a consistent omission of the
empirical calibration necessary to convert these proportional
relationships into measurable and comparable quantities

After Carey and folowing Reilly and Zipf, it is only in
\citet{stewartDemographicGravitationEvidence1948} that we find the most
explicit connection yet to Newton's Gravitational law and the use of a
proportionality constant. Besides awareness of preceding research (i.e.,
Stewart cites both Reilly and Zipf as predecessors in the analysis of
human spatial interaction), Stewart appears to have been the first
author to express his theorized relationships for human spatial
interaction with a proportionality constant \(G\), enabling his
formulation to be interpreted as a measurable `demographic' force:

\begin{equation}\phantomsection\label{eq-stewart-force}{
F = G\frac{(\mu_1N_1)(\mu_2N_2)}{d_{12}^2} = G\frac{M_1 M_2}{d_{12}^2} 
}\end{equation}

\noindent Where:

\begin{itemize}
\tightlist
\item
  \(F\) is the \emph{demographic force}
\item
  \(N_1\) and \(N_2\) are the numbers of people of in groups 1 and 2
\item
  \(\mu_1\) and \(\mu_2\) are so-called \emph{molecular weights}, the
  attractive weight of groups 1 and 2
\item
  \(M_1 = \mu_1N_1\) and \(M_2 = \mu_2N_2\) are the demographic masses
  at 1 and 2
\item
  \(d_{12}^2\) is the distance between \(1\) and \(2\)
\item
  And finally proportionality constant \(G\)
\end{itemize}

What is notable about Eq~\ref{eq-stewart-force}, however, is the
proportionality constant \(G\) was specified but ``left for future
determination''
\citep[p.~34]{stewartDemographicGravitationEvidence1948}. We can infer
that it is crucial for ensuring \(F\) is maintained in some units of
demographic force.

In addition to demographic force \(F\), Stewart defined a measure of the
``potential'' of group \(2\) with respect to group \(1\). The partial
sum of the demographic force experienced by group \(1\), or the
potential number of people from location \(2\) that could visit location
\(1\), as \(V_1 = G\frac{M_2}{d_{12}}\). For a system with more than two
population bodies, Stewart formulated the population potential at \(i\)
by summing the contributions from each group \(j\), after arbitrarily
setting \(G=1\):

\begin{equation}\phantomsection\label{eq-stewart-population-potential-sum}{
V_i = \sum_j M_jd_{ij}^{-1}
}\end{equation}

\noindent Where \(M_j\) is the demographic mass at location \(j\) and
\(d_{ij}\) is the distance between \(i\) and \(j\). A version of this
discrete form is what was used in \citet{hansen1959}, going on to become
a foundation of modern accessibility definitions, as will be discussed.

Although Stewart's concept of ``social physics'' eventually fell out of
favour--due to inconsistent mathematical notation (e.g., \(G\) is used
as both a proportionality constant p.~34 and then later as
\emph{demographic energy} on p.~53.) and its foundation in racist and
unscientific assumptions (e.g., view of humans as particles following
physical laws and assumptions of the molecular weight \(\mu\) of the
average American being one, but ``presumably\ldots much less than
one\ldots.for an Australian aborigine'' {[}p.~35{]})--his introduction
of a proportionality constant \(G\) in modeling demographic force marks
an important conceptual step: recognizing that moving from
proportionality to equality requires empirical calibration. In other
words, the addition of \(G\) shifts results from being abstract
indicators of potential (i.e., people\textsuperscript{2-distance}-2) to
having interpretable units grounded in consistent, albeit still
abstract, quantities (i.e., units of demographic force).

As will be discussed, when Hansen later adopted Stewart's formula
(Eq~\ref{eq-stewart-population-potential-sum}) for accessibility, he
omitted any mention of \(G\), effectively setting it to 1 arbitrarily as
Stewart had done. This omission has persisted in accessibility research,
leaving a conceptual gap in how such measures are interpreted and
compared.

\section{Hansen's gravity-based accessibility to
today}\label{grav-to-today}

From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
arrive to 1959 and Walter G. Hansen, whose work proved to be
exceptionally influential in the accessibility literature
\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
as ``the potential of opportunities for interaction\ldots{} a
generalization of the population-over-distance relationship or
\emph{population potential} concept developed by Stewart
\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
being a student of city and regional planning at the Massachusetts
Institute of Technology, Hansen was also an engineer with the Bureau of
Roads, and preoccupied with the power of transportation to shape land
uses in a very practical sense. \citet{hansen1959} drew directly from
Stewart's population potential (see
Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
broader (and often problematic) framework of ``social physics.''.

Hansen recast Stewart's population potential to reflect accessibility, a
model of human behaviour useful to capture regularities in mobility
patterns. Hansen replaced \(M_j\) in
Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
to derive an \emph{opportunity potential}, or more accurately, a
\emph{potential of opportunities for interaction} as
\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
rewriting to accounts for a variety of impedance functions beyond the
inverse power \(d^{-\beta}\) as follows:

\begin{equation}\phantomsection\label{eq-accessibility-general}{
S_{i} = \sum_j O_j  f(d_{ij})
}\end{equation}

\(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
accessibility from site \(i\). This is a function of \(O_j\) (the mass
of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
travel time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a
parameter that modulates the friction of cost). Today, Hansen is
frequently cited as the father of modern accessibility analysis
\citep[e.g.,][]{reggianiGuestEditorialNew2011}, and Hansen-type
accessibility is commonly referred to as the gravity-based accessibility
measure.

However, Hansen's formulation carried forward a crucial omission that
continues to affect the literature: the proportionality constant \(G\)
included in Stewart's original formulation
(Eq~\ref{eq-stewart-population-potential-sum}) has vanished entirely.
Although Stewart included \(G\) explicitly (with a note that ``\(G\)
{[}was{]} left for future determination: a suitable choice of other
units can reduce it to unity'' {[}p.~34{]}.) Hansen made no mention of
it. As a result, modern accessibility analysis has largely evolved
without addressing the constant's role, leaving \(G\) effectively and
implicitly fixed at 1. This omission has significant implications.
Without a proportionality constant, the accessibility formula expresses
only a proportional relationship:
\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not of calibrated equality.
Recognition of the nature of this relationship is not common in the
literature, but is known, i.e., this proportional equation is shown in
Figure 1 in Wu and Levinson \citep{wuUnifyingAccess2020}.

However, the omission of a proportionality constants and working with a
proportional relationship generates fundamental issues in comparability
between and, arguably within, studies. As the absence of \(G\) means
accessibility estimates have no fixed unit--this renders them sensitive
to the choice of impedance functions. For instance, if travel cost
\(d_{ij}\) is measured in meters, then when the travel impedance
function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the resulting \(S_i\)
has units of opportunities per \(\text{metres}^{\beta}\); but when
\(f(d_{ij})\) is set to equal \(e^{-\beta d_{ij}}\), the units become
opportunities per \(e^{\beta \text{metres}}\). Such variation impairs
comparability across analysis and obscures the meaning of accessibility
scores renders their results difficult to understand and communicate
without post-hoc treatment.

Therefore in practice, Hansen-type measures (of proportionality) are
better understood as \emph{ordinal indicators}--they rank accessibility
but lack cardinal meaning or consistent units
\citep{millerAccessibilityMeasurementApplication2018}. The continued
absence of \(G\) leaves a conceptual and practical gap in accessibility
analysis: a missing link between theoretical form and empirical
measurement.

\section{Wilson's family of spatial interaction
models}\label{wilsons-family-of-spatial-interaction-models}

While accessibility research evolved in North America with
\citet{hansen1959}, a parallel development was taking place across the
Atlantic with Alan G. Wilson. In his groundbreaking study
\citep{wilson1971}, advanced a general framework for spatial interaction
modeling--focused not on potential (as accessibility), but on modeling
observed flows of interaction between places. \citet{wilson1971}
formalized the general spatial interaction model through the following
equation:

\begin{equation}\phantomsection\label{eq-phys-gravity-model}{
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
}\end{equation}

The model in Eq~\ref{eq-phys-gravity-model} posits a quantity \(T_{ij}\)
that represents a value in a matrix of flows of size \(n \times m\),
that is, between \(i = 1,s, n\) origins and \(j = 1,s, m\) destinations.
The quantities \(W_i^{(1)}\) and \(W_j^{(2)}\) are proxies for the
masses at \(i=1,s,n\) origins and \(j=1,s,m\) destinations. The
super-indices \((1)\) and \((2)\) indicate that these masses can be any
number of different things associated with the zones, i.e.,
\(W_i^{(1)}\) could be population at a zone as an origin, and
\(W_j^{(2)}\) hectares of park space at a zone as a destination.
\(f(c_{ij})\) is some function of travel cost \(c_{ij}\) which reflects
travel impedance. Finally, \(k\) acts as a proportionality constant,
shifting the equation from a proportional to an equal relationship by
incorporating known system totals. In this specific sense, we infer that
\(k\) serves a role similar to the gravitational constant in Newton's
law--it calibrates the model so that outputs match real-world
quantities.

From the outset, spatial interaction models emphasized interpretability
of results
\citep{kirbyNormalizingFactorsGravity1970, wilsonSTATISTICALTHEORYSPATIAL1967, wilson1971}.
But unlike earlier approaches that borrowed heuristically from Newtonian
gravity (i.e., interaction between masses over distance), Wilson's
innovation was to ground the model in \emph{entropy maximisation}. By
maximizing the number of ways individual trip probabilities could be
arranged under known constraints, Wilson derived models that estimate
\emph{statistical averages} of flows between zones
\citep{wilson1971, seniorGravityModellingEntropy1979}.

Crucially, to ensure that \(T_{ij}\) in Eq~\ref{eq-phys-gravity-model}
is maintained in units of flow between \(i\) and \(j\), the model moves
from proportionality to calibrated equality by incorporating empirical
constraints. At a minimum, this requires knowledge of the total number
of flows \(T\) in the system, leading to the basic constraint:

\begin{equation}\phantomsection\label{eq-constraint0-gravitymodel}{
\sum_i\sum_j T_{ij} = T
}\end{equation}

Additional information can be introduced. For example, when information
is available about the total number of flows produced by each origin,
\(W_i^{(1)}\) is represented as \(O_i\) and the following constraint can
be used:

\begin{equation}\phantomsection\label{eq-constraint1-gravitymodel}{
\sum_j T_{ij} = O_i
}\end{equation}

As well, if there is information available about the total number of
flows attracted by each destination, \(W_j^{(2)}\) is represented as
\(D_j\) and the following constraint can be used:

\begin{equation}\phantomsection\label{eq-constraint2-gravitymodel}{
\sum_i T_{ij} = D_j
}\end{equation}

It is also possible to have information about both \(O_i\) and \(D_j\),
in which case both constraints could be imposed on the model at once.

Depending on which system constraints are applied---fully, partially, or
not at all---a family of spatial interaction models can be derived from
Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
replaced with different \emph{balancing factors}. This change in name is
more useful as these factors not only preserve proportionality but also
ensure that the predicted flows \(T_{ij}\) align with the known
constraints considered in the system. In other words, the different
balancing factors adjust the model so that flows become statistical
averages consistent with observed origin and/or destination data. In the
framework introduced and inferred from \citet{wilson1971}, three types
of balancing factors are specified: (1) an unconstrained model that only
matches the total volume of interaction \(K\), (2) a singly constrained
model (either by origins \(A_i\) or destinations \(B_j\)), and (3) a
doubly constrained model that satisfies both.

In the unconstrained model, constraints in
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
this means that the total number of flows predicted by the model must be
equal to sum of all flows from origins \(i\) to destinations \(j\). The
balancing factor \(K\) takes the place of \(k\) and is equal to the
following \citep[as specified in][ and
\citet{fotheringham_spatial_1984}]{cliff_evaluating_1974}:

\begin{equation}\phantomsection\label{eq-total-flow-balancing-factor}{
K=\frac{T}{\sum_i\sum_j T_{ij}}
}\end{equation}

In the singly-constrained model, only constraint
Eq~\ref{eq-constraint1-gravitymodel} or constraint
Eq~\ref{eq-constraint2-gravitymodel} hold. When only
Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
to the following production-constrained singly-constrained model:

\begin{equation}\phantomsection\label{eq-production-constrained-gravitymodel}{
T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
}\end{equation}

Notice how, in this model, the proxy for the mass at the origin
\(W_i^{(1)}\) is replaced with \(O_i\), representing what we know about
the system, the spatial interaction outbound flow, i.e., outbound flows
produced at \(i\). Also, there is no longer a single system-wide
balancing factor \(K\), but rather a set of balancing factors specific
to origins \(A_i\).

For this model, the balancing factors ensure that constraint
Eq~\ref{eq-constraint1-gravitymodel} is satisfied, meaning that the sum
of predicted flows from one origin going to all destinations must equal
the known mass at that origin \(O_i\) e.g., the total number of outbound
flows. Satisfying this constraint also implicitly fulfills the total
constraint (Eq~\ref{eq-constraint0-gravitymodel}), since the sum of
\(O_i\) values across all origins equals the total number of flows.

The production-constrained singly-constrained model is useful when flows
ends are unknown but the number of flows originating from each location
is known. The balancing factors for the production-constrained model are
solved for each origin \(A_i\), and according to Wilson are:

\begin{equation}\phantomsection\label{eq-production-constrained-balancing-factor}{
A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
}\end{equation}

The attraction-constrained model is similar to the
production-constrained model as it is also singly constrained but from
the perspective of the mass at the destination. From the
attraction-constrained model, the proxy for the mass at the destination
\(W_j^{(2)}\) is now replaced with \(D_j\), representing the spatial
interaction inbound flow e.g., flows attracted to the destination, and
takes the following form:

\begin{equation}\phantomsection\label{eq-attraction-constrained-gravitymodel}{
T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
}\end{equation}

In this model (Eq~\ref{eq-attraction-constrained-gravitymodel}), the
balancing factors ensure that Eq~\ref{eq-constraint2-gravitymodel} is
satisfied (hence the total constraint
Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning that the sum
of predicted flows going to one destination from all origins must equal
the known mass of that destination \(D_j\) i.e., the total number of
inbound flows to \(j\). This should hold for all destinations. As
before, destination-specific balancing factors \(B_j\) were derived by
Wilson as:

\begin{equation}\phantomsection\label{eq-attraction-constrained-balancing-factor}{
B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

Lastly, the doubly-constrained model is the production-attraction
constrained model in \citet{wilson1971}. In this case, both constraints
Eq~\ref{eq-constraint1-gravitymodel} and
Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
constraints ensure that the sum of predicted flows from one origin to
all destination, and the predicted flows going to one destination from
all origins must equal the known mass of the origin \(O_i\) and of the
destination \(D_j\). This should hold for all origins and destinations.
The resulting model is, in Wilson's terms, doubly constrained, and takes
the following form:

\begin{equation}\phantomsection\label{eq-doubly-constrained-gravitymodel}{
T_{ij} = A_i B_j O_i D_j f(c_{ij})
}\end{equation}

In this model, both proxies for the masses are replaced with the known
masses, that is, the flows produced by the origin and the flows
attracted by the destination. There are now two sets of mutually
dependent proportionality constants shown in
Eq~\ref{eq-doubly-constrained-balancing-factors}. Derivation of these
models is demonstrated in detail elsewhere
\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.

\begin{equation}\phantomsection\label{eq-doubly-constrained-balancing-factors}{
\begin{array}{l}
A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
\end{array}
}\end{equation}

Wilson's work is notable for many reasons, one of which is his departure
from the metaphor of Newtonian gravitational physics. Instead, he
grounded his models in statistical mechanics, particularly the principle
of entropy maximisation. For accessibility research, Wilson's approach
is especially significant because it can address a problem that has
quietly plagued this field (as this work has argued): the transition
from proportionality to equality.

Rather than relying on a universal constant like \(G\), Wilson's models
calibrate interaction flows through known empirical constraints (e.g.,
total system flows, or origin-side and destination-side flow totals).
This results in interpretable, balanced (given the constraints), and
unit-consistent \(ij\) flows. In short, when aspects of the systems are
inputs into the balacing factors, \(ij\) flows can be robustly predicted
in tangible units of `flow', units that align with the system being
modelled. In this way, spatial interaction modeling tradition can be
seen to have succeeded where accessibility modeling stalled. While
Wilson's model produces results that are in units of flow tethered to
the system of analysis, Hansen-type measures (still widely used today)
yield outputs that \emph{indicate} a proportional relationship and not
one of equality nor yielding results without interpretable units.

As a preview for the work ahead: Wilson's balancing factors can be
manipulated into \emph{proportional allocation factors} and equations
can be established to reflect accessibility. The resulting \(ij\) flows
can be understood as accessibility flows--and aggregated at the zone
level to express accessibility in units of actual opportunities or
population.

In this way, these balancing factors can be used to recover what
Stewart's \(G\) was meant to accomplish, but through a different
conceptual path. But before demonstrating a family of accessibility
measures that are sensitive to constraints, in the next section we
review how conceptually intertwined the spatial interaction modelling
and accessibility literatures are, and where they began to diverge. This
investigation sheds light on why accessibility research may have failed
to adopt a comparable approach focused, until this work.

\section{Accessibility and spatial interaction modelling: two divergent
research
streams}\label{accessibility-and-spatial-interaction-modelling-two-divergent-research-streams}

\begin{figure}[H]

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}

}

\caption{\label{fig-docs-per-year}Historical pattern of publication:
documents per year.}

\end{figure}%

Despite their close conceptual ties, the accessibility and spatial
interaction modeling literatures have developed along largely separate
paths since the 1970s. Hansen's \citep{hansen1959} formulation of
accessibility became the method used for decades of work on transport
equity, land use analysis, and urban accessibility planning. Meanwhile,
Wilson's \citep{wilson1971} entropy-maximising framework reshaped how
spatial interaction models were constructed, particularly in
transportation demand forecasting. We argue the framework's quiet
innovation--introducing empirically grounded constraints to shift from
proportionality to calibrated equality--made the framework immediately
relevant for policy applications as outputs were in tangible units. Yet,
this mechanism was never widely adopted in accessibility analysis.

This divergence is especially striking given the context in which both
frameworks emerged. As noted in
\citet{battyChronicleScientificPlanning1994}, large scale spatial
interaction models (like Wilson's) responded to important developments
at the time, a need ``to meet the dictates and needs of public policy
for strategic land use and transportation planning''. And these needs
were far from trivial: in the U.S., for instance, the Federal-Aid
Highway Act of 1956 set in motion the construction of the Interstate
Highway System with an eventual budget exceeding \$600 billion in
today's dollars
\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
this context, spatial interaction models were incorporated into
institutional practices focused on ``predict and provide'' travel demand
forecasting
\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
Accessibility analysis, by contrast, remained more conceptually diffuse,
focused on indicators of ``potential'' spatial interaction with
opportunities rather than flows that could tangibly guide infrastructure
decisions (e.g., roadway capacity expansion, new construction). Whereas
spatial interaction modelling became a key element of transportation
planning practice, accessibility remained a somewhat more academic
pursuit, and the two streams of literature only rarely connected.

To explore why Wilson's approach never crossed over to accessibility
modeling, we conduct a review of the literature citing Hansen
\citep{hansen1959}, Wilson \citep{wilson1971}, or both (on Web of
Science using the ``Cited References'' functionality, and the digital
object identifiers of Hansen \citep{hansen1959} and Wilson
\citep{wilson1971}) . Only 76 out of the 2,122 documents that emerged
from our search cite both. The number of articles, by year and if they
cite Hasen, Wilson, or both are shown in Fig~\ref{fig-docs-per-year}.

Through the close analysis of \emph{how} articles that cite both works,
we identify two distinct citation patterns: one group of articles
focused on accessibility, the other on spatial interaction. In examining
these articles, we uncover how the relationship between the two has
often been misunderstood, underexplored, or entirely overlooked.

In the first stream of literature--which cite both but are focused more
so on spatial interaction models--they treat spatial interaction and
accessibility as separate but related phenomenons. Four subsets of this
stream emerge.

First, some of the more early works interpret the spatial interaction
model's balancing factors
(Eq~\ref{eq-production-constrained-balancing-factor} or
Eq~\ref{eq-doubly-constrained-balancing-factors}) as the inverse of
Hansen's accessibility measure
\citep{harrisEquilibriumValuesDynamics1978, leonardiOptimumFacilityLocation1978, fotheringhamSPATIALSTRUCTUREDISTANCE1981, fotheringhamSpatialCompetitionAgglomeration1985},
likely following Wilson's own recognition of this similarity between
balancing factor \(A_i\) and Hansen-type measure \(S_i\) on p.~10 in
Wilson \citep{wilson1971}. In some ways, this relationship has been
recognized as a ``common sense'' approach to incorporating accessibility
in the spatial interaction model
\citep[p.~99]{morrisAccessibilityIndicatorsTransport1979}, though
acknowledgment of its further exploration has been recommended
\citep{battyMethodResiduesUrban1976}.

The second subset of articles within this stream use both Hansen
\citep{hansen1959} and Wilson's \citep{wilson1971} framework in
conjunction. For instance, some articles argue that spatial interaction
models fail to explain certain spatial patterns on their own, for
instance, as in Fotheringham
\citep{fotheringhamSpatialCompetitionAgglomeration1985} who demonstrates
how the spatial interaction model may insufficiently explain spatial
patterns, and suggests that explicitly defining destinations'
accessibility (Hansen-type accessibility) as a variable within the model
may remedy the issue (e.g., the \emph{competition destination} model).
Other works take a more applied approach: such as in defining
location-allocation problems in operations research
\citep{leonardiOptimumFacilityLocation1978, beaumontLocationallocationProblemsPlane1981},
estimating trips (or some other spatial interaction flows) alongside
accessibility
\citep[e.g.,][]{clarke2002deriving, grengs2004measuring, turk2019socio},
or considering accessibility as a variable within spatial interaction
models, in line with Fotheringham's
\citep{fotheringhamSpatialCompetitionAgglomeration1985} demonstration
\citep[e.g.,][]{beckers2022incorporating}.

The third subset of the spatial-interaction focused literature, depart
from Hansen's \citep{hansen1959} definition, aligning instead with
microeconomic or utility-based interpretations of potential spatial
interaction e.g.,
\citep{morrisAccessibilityIndicatorsTransport1979, leonardiRandomUtilityDemand1984}.
In sum though, across these works, they recognize Hansen-type
accessibility as an indicator of `potential' but as a separate but
related concept to spatial interaction.

Moving onto the group of accessibility-focused literature that cites
both works, we categorise their citation of Wilson \citep{wilson1971}
within three general groups. Overall though, these works do not engage,
or only superficially engage with Wilson \citep{wilson1971}.

Firstly, there is a group of articles within this stream that cite
Wilson \citep{wilson1971} exclusively as attribution for using
context-dependent travel cost functions. This trend is common: for
instance it is done in
\citep{weibullNumericalMeasurementAccessibility1980, handyMeasuringAccessibilityExploration1997, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
However, these works do not engage with spatial interaction beyond this
attribution.

Secondly, a subset of literature explicitly associates spatial
interaction, as defined in Wilson \citep{wilson1971}, with
accessibility's potential for spatial interaction--but only
superficially. These works acknowledge the conceptual link, but do not
go beyond this recognition in the scope of their works e.g.,
\citep{millerMeasuringSpaceTimeAccessibility1999, giuliano2010accessibility, grengs2010intermetropolitan, grengs2010job, grengs2012equity, levine2012does, levinson2012positive, tong2015transportation, liu2015spatial, he2017measuring, wuUnifyingAccess2020, ng2022reflection, naqavi2023mobility, suel2024measuring}.
Indeed, while accessibility can be seen as the \emph{potential} for
spatial interaction--and Wilson \citep{wilson1971} briefly touches on
this--such mentions have not resulted in deeper analytical integration
of these concepts. Furthermore, some of this literature also
occasionally conflates or blurs the distinction entirely, for instance
by co-citing Hansen and Wilson as being `gravity models'
\citep[e.g.,][]{liu2004accessibility, dai2017visualization, shen2019segregation, chia2020extending}.
This conflation reveals ongoing murkiness between the distinction of
spatial interaction and the \emph{potential for} spatial interaction in
the literature.

Thirdly, there is a group of accessibility-focused works that interprets
the measure used in \citet{hansen1959} as the singly- or doubly-
constrained spatial interaction model's inverse balancing factor
\citep[e.g.,][]{vickermanAccessibilityAttractionPotential1974}. This
group often cites the spatial interaction works that make this
connection (i.e., the first subset of the first stream of literature)
and is especially prominent in the investigation of competitive
accessibility topics e.g.,
\citep{karstEvaluationAccessibilityImpacts2003, geurs2006accessibility, willigers2007accessibility, el2011place, curtis2010planning, manaugh2012makes, chen2013regional, alonso2014labour, albacete2017measuring, sahebgharani2019computing, mayaud2019future, allenMeasureCompetitiveAccess2020, levinsonGeneralTheoryAccess2020, marwal2022literature, su2023untangling}.
Only the works of Soukhov et al.
\citep{soukhovIntroducingSpatialAvailability2023, soukhovMultimodalSpatialAvailability2024}
use Wilson's \citep{wilson1971} balancing factors as a method for
maintaining constraints on opportunities within the context of
competitive accessibility.

On that note, Soukhov et al., 2023 and 2024
\citep{soukhovIntroducingSpatialAvailability2023, soukhovMultimodalSpatialAvailability2024}
introduce the balancing factors as a mechanisms to ensure that
opportunities at each destination are proportionally allocated to each
zone (based on the proportion of population seeking opportunities and
the relative travel impedance). This is to ensure that each zonal
accessibility value is the sum of this proportional allocation from each
destination, and that all zonal values ultimately sum to the total
number of opportunities in the region. However, these balancing factors
were deduced intuitively. These works do not explicitly state that the
mathematical formulation of the equations are effectively equivalent to
Wilson's singly constrained model (derived from entropy maximization).
This equivalence is only discovered in hindsight, as will be
demonstrated in the following section. These two works also do not
discuss other constrained cases that will also be addressed.

In sum, despite the interpretative advantaged offered by the statistical
logic of \citet{wilson1971}'s framework, the spatial
interaction-oriented \citet{hansen1959} citing literature nor the
accessibility-oriented literature that cite \citet{wilson1971} have yet
to meaningfully adopt Wilson's constraint-based logic to the concept of
accessibility. So, in the next section, we demonstrate how applying
Wilson's framework enables accessibility to move from a proportional
relationship to calibrated equality--tying outputs to tangible system
knowns. This re-expression of accessibility using constraints yields
interpretable, unit-consistent measures of opportunity. This approach
takes the same path of entropy-maximisation as in \citet{wilson1971},
and does not rely on specifying some universal constant \(G\) like
initially proposed in \citet{stewartDemographicGravitationEvidence1948}
(recall, Eq~\ref{eq-stewart-population-potential-sum} which
\citet{hansen1959} operationalised).

\section{A family of accessibility measures: from proportionality to
equality}\label{a-family-of-accessibility-measures-from-proportionality-to-equality}

Despite the close conceptual ties between accessibility and spatial
interaction modeling, the former has not meaningfully absorbed the
constraint-based logic of the latter. This section introduces this
reframing: defining a family of accessibility measures using Wilson's
approach grounded in statistical mechanics, shifting place-based
accessibility (a la \citet{hansen1959}) from a relationship describing
proportionality to equality.

This shift addresses the fundamental issue of interpretability
associated with the proportional nature of Hansen-type accessibility
indicators we previous outlined. Rather than relying on arbitrary or
uncalibrated proportional relationships--or invoking a universal
constant \(G\) as proposed in
\citet{stewartDemographicGravitationEvidence1948} --we adopt Wilson's
approach grounded in statistical mechanics.

To do so, we propose a revised definition of accessibility considerate
of the constraint-based spatial intereaction model: the potential for
spatial interaction with opportunities or population.We can bring \(k\)
as a proportional allocation factor which encorporates Wilson's
balancing factor, to define the \emph{potential for spatial interaction
with opportunities} \(V_{ij}\) and the \emph{potentnial for spatial
intereaction with population} \(M_{ji}\). In effect, \(\kappa\) is
unitless and proportionally allocates (based on the constraints of the
case) opportunities (for \(V_{ij}\)) and population (for \(M_{ji}\)).
The equation is as follows:

\begin{equation}\phantomsection\label{eq-access-01}{
\begin{array}{l}
V_{ij}^X = \kappa_{ij}^X W_X^{(2)}\\ 
M_{ji}^X = \hat \kappa_{ji}^X W_X^{(1)}
\end{array}
}\end{equation}

\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) is the mass of the
destination (i.e., opportunities) and mass of the origin (i.e.,
population) at either the zone or for the full region, depending on the
case. Accessibility flows can also be summarised as as a partial sum of
the potential at \(i\) and at \(j\) to express accessibility at the
origin zone or at the destination zone, respectively, as often done in
accessibility research:

\begin{equation}\phantomsection\label{eq-accesssibility-01}{
\begin{array}{l}
V_{i}^X = \sum_j \kappa_{ij}^X W_X^{(2)}\\
M_{j}^X = \sum_i \hat \kappa_{ji}^X W_X^{(1)}
\end{array}
}\end{equation}

Figure Fig~\ref{fig-analytical-device-conc-accessibility} illustrates
our analytical framework using a simple 3-zone system. The most detailed
values, \(X_{ij}\), represent the potential for spatial interaction from
origin \(i\) to destination \(j\). Here, \(X\) stands for all cases and
variants to be discussed (e.g., \(V_{ij}^0\), \(M_{ji}^0\),
\(V_{ij}^T\), \(M_{ji}^T\), \(V_{ij}^S\), \(M_{ji}^S\), \(V_{ij}^D\),
and \(M_{ji}^D\)). Single marginals show origin and destination weights,
while the total marginal sums these values.

\begin{figure}[H]

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{../figures/access-analytical-device.png}

}

\caption{\label{fig-analytical-device-conc-accessibility}The family of
accessibility measures analytical framework: labelling and associating
ij flows, zonal weights, the single marginals, and the total marginal.}

\end{figure}%

The proportional allocation constant \(\kappa\) takes the form of a
balancing factor that varies depending on the constraints applied. Each
member of the accessibility measure family is defined by the constraints
used, and can be grouped into the following four cases:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Unconstrained Case (\(V_i^0\), \(M_j^0\))}
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Equivalent to Hansen's \citep{hansen1959} and Reilly's
  \citep{reilly1929methods} original formulations; the status quo of
  accessibility modelling.
\item
  No balancing factors applied; units are in
  ``opportunities-by-impedance'' for \(V_i^0\) or
  ``population-by-impedance'' for \(M_j^0\).
\item
  No constraints are applied, so values reflect proportionality only and
  are not calibrated to known system totals.
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  \textbf{Total Constrained Case (\(V_i^T\), \(M_j^T\))}
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Applies a total proportional allocation factor (\(\kappa_{ij}^T\),
  \(\hat \kappa_{ji}^T\)) based only on the total marginal (green box in
  Fig~\ref{fig-analytical-device-conc-accessibility}) i.e., total number
  of opportunities or population in the system. This ensures the sum of
  all values in the system match the total marginal.
\item
  Units of \(V_i^T\): accessible opportunities from \(i\), a value that
  is total constrained and linearly proportion to \(V_i^0\).
\item
  Units of \(M_j^T\): accessible population from \(j\), a value that is
  total constrained and linearly proportion to \(M_j^0\).
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  \textbf{Singly Constrained Case (\(V_i^S\), \(M_j^S\))}
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Applies singly-constrained proportional allocation factors
  (\(\kappa_{ij}^S\), \(\hat \kappa_{ji}^S\)) based on Wilson's
  balancing factors (\(B_j\), \(A_i\)) to preserve either the
  destination-side or origin-side marginal totals (blue and red boxes in
  Fig~\ref{fig-analytical-device-conc-accessibility}) i.e., the number
  of opportunities or population at each zone. Reflects how the
  literature calculates competitive accessibility.
\item
  Units of \(V_i^S\): accessible opportunities from \(i\), a value that
  is the sum of opportunity supply flows allocated proportionally based
  on demand at \(i\). Mathematically equivalent in per-capita form to
  2SFCA \citep{luo2003}.
\item
  Units of \(M_j^S\): accessible population from \(j\), a value that is
  the sum of population demand flows allocated proportionally based on
  supply at \(j\).
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  \textbf{Doubly Constrained Case (\(V_{ij}^D\), \(M_{ji}^D\))}
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Constrained on both origin and destination sides using both \(A_i\)
  and \(B_j\) simultaneously; equivalent in interpretation to Wilson's
  \citep{wilson1971} doubly constrained spatial interaction model.
\item
  Simultaneous application ensures both the destination-side \emph{and}
  origin-side marginal totals are maintained (blue and red boxes in
  Fig~\ref{fig-analytical-device-conc-accessibility}).
\item
  Interpretable only as \(ij\) and \(ji\) flows, since aggregation at
  \(i\) and \(j\) simply reproduces known totals. Represents
  `interaction capacity' or `realized access,' serving as predictions of
  real interaction flows.
\end{itemize}

\subsection{Toy example setup}\label{toy-example-setup}

Consider the simple 3-zone region in Figure
Fig~\ref{fig-analytical-device-conc-accessibility}, where each zone
serves as both origin (\(i\)) and destination (\(j\)). The system
includes three inputs: zonal population and opportunities, a zonal cost
(travel time) matrix, and three travel impedance functions representing
different travel behaviours.

First, Table Table~\ref{tbl-small-system-land-use} summarizes population
(in 10,000s) and physicians per zone. For context, the
Provider-to-Population Ratio (PPR) is 24.5 comparable to Canada's 2022
PPR of 24.97 physicians per 10,000 \citep{whoMedicalDoctors102025}.
Second, Table Table~\ref{tbl-small-system-cost} shows travel times
(minutes); it can be discerned that Zones 1 and 3 are closer to each
other than to Zone 2. Zones 1 and 3 together have a population roughly
equal to Zone 2 but offer more than twice the physician availability. We
interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core,
and Zone 2 as the Suburban.

\begin{table}

\caption{\label{tbl-small-system-land-use}Simple system with three zones
(ID 1, 2 and 3). Population is in 10,000 persons and opportunities in
number of physicians.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}rcc}
\toprule
ID (i or j) & Population\textsuperscript{\textit{1}} & Opportunities\textsuperscript{\textit{2}} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 160 \\ 
2 & 10 & 150 \\ 
3 & 6 & 180 \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Population is \emph{Wi\textsuperscript{(1)}} when used as a proxy for the mass at the origin, and \emph{Oi} when used as a constraint.\\
\textsuperscript{\textit{2}}Opportunities is \emph{Wj\textsuperscript{(2)}} when used as a proxy for the mass at the destination, and \emph{Dj} when used as a constraint.\\
\end{minipage}

}

\end{table}%

\begin{table}

\caption{\label{tbl-small-system-cost}Cost matrix for system with three
zones (travel time in minutes).}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}cccc}
\toprule
 & \multicolumn{3}{c}{Destination ID} \\ 
\cmidrule(lr){2-4}
Origin ID & 1 & 2 & 3 \\ 
\midrule\addlinespace[2.5pt]
1 & 10 & 30 & 15 \\ 
2 & 30 & 10 & 25 \\ 
3 & 15 & 25 & 10 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

And third, Eq~\ref{eq-travel-behaviour-scenarios} presents the assumed
travel impedance functions reflecting three different travel behaviours.
A helpful analogy may be tying travel behaviour to the used mode's
mobility potential, i.e., the most decaying travel behaviour
(\(f_1(c_{ij})\)) would assume all travel in the region being done by
foot, while calculating accessibility assuming the least decay
(\(f_3(c_{ij})\)) would assume unfettered automobility.

\begin{equation}\phantomsection\label{eq-travel-behaviour-scenarios}{
\begin{array}{l}
f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
\end{array}
}\end{equation}

Any set of concepts representing population, opportunities, and their
associated travel behaviour, whether representing the entire region
uniformly (as will be demonstrated) or representing specific subgroups,
can be substituted into our simple toy example. The purpose of the
following simple example is to demonstrate the calculation and
interpretation of the four accessibility measure variants.

\subsection{Unconstrained
accessibility}\label{unconstrained-accessibility}

In \(V^0_i\), no proportional allocation factor is defined, simply
\(f(c_{ij})\) is used to weight the number of opportunities at each
\(j\) and the weighted values for each \(j\) are summed for each \(i\),
yielding the an expression identical to Hansen's accessibility \(S_i\)
\citep{hansen1959}, the current standard practice in accessibility
measurement:

\begin{equation}\phantomsection\label{eq-unconstrained-accessibility}{
V^0_i = \sum_j V^0_{ij} = \sum_j W^{(2)}_jf(c_{ij}) = S_i
}\end{equation}

However \(\sum_i V^0_{i}\) generally does not equal the total
opportunities \(O\), so units here are `opportunities weighted by travel
impedance' and lack meaningful scaling or direct interpretability.
Comparing values across different decay functions or contexts (i.e.,
different number of zones) is therefore limited to ordinal statements
(more vs.~less), not intervals or ratios (i.e., the magnitude of
differences).

\begin{table}

\caption{\label{tbl-simple-example-unconstrained-accessibility}Simple
system: unconstrained accessibility.}

\centering{

\fontsize{6.0pt}{7.2pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
\toprule
 & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{0}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Origin & units: \emph{physicians-minute\^{}-3} & units: \emph{physicians-minute\^{}-2} & units: \emph{physicians-minute\^{}-0.1} \\ 
\midrule\addlinespace[2.5pt]
1 & 0.219 & 2.567 & 371.143 \\ 
2 & 0.167 & 1.966 & 363.479 \\ 
3 & 0.237 & 2.751 & 373.738 \\ 
\midrule 
\midrule 
Sum & 0.6233422 & 7.283556 & 1108.361 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

For example, Table
Table~\ref{tbl-simple-example-unconstrained-accessibility} shows
\(V^0_{i}\) under each decay function. Comparing across decay types is
meaningless in absolute terms. For instance, the difference in zone 1
(edge of urban core)'s accessibility under \(f_3\) vs \(f_1\) is 370.92,
but in what units? These two values are a product of different impedance
functions ( \emph{physicians-minute\(^{-3}\)} and
\emph{physicians-minute\(^{-0.1}\)}), making the comparison
uninterpretable (and arguably incorrect). The fundamental
uninterpretability of what is a
\emph{opportunity-weighted-travel-impedance} unit remains.

As the different impedance functions represent different travel
behaviours, comparing the raw unconstrained accessibility values across
groups is meaningless beyond notions of higher or lower. While one could
attempt to adjust the units post-calculation (e.g., scaling, population
normalization) or select impedance functions to facilitate comparison
across scenarios (potentially at the expense of accurately reflecting
travel behavior), such adjustments may introduce bias. The unconstrained
scores are best used for ranking within a single context.

The next sections introduce constraints to calibrate these measures for
better interpretability and comparability, applying each to this example
in turn.

\subsection{Total constrained
accessibility}\label{total-constrained-accessibility}

The total constrained accessibility case can be interpreted in a few
ways. In the one that connects to the status quo: the total balancing
factor proportionally adjusts unconstrained zonal accessibility values
\(V^0_i\) so their total sum of \(V^0_i\) matches a known system
total--either total opportunities or total population. Another
interpretation is in reformulating the equation to use a a proportional
allocation constant based on the total balancing factor. The
proportional allocation constant distributes opportunities (or
population) proportionally by travel impedance.

In both formulations, all zonal values become a proportion of a known
system total, be it the regional opportunities or regional population
depending on the variant.

We define two variants for this case:

\begin{itemize}
\tightlist
\item
  \(V_i^T\): accessibility is constrained by the total number of
  opportunities (total constrained accessible opportunity) and which is
  interpreted as Hansen's accessibility with a constraining constant,
  and
\item
  \(M_j^T\): where \(i\) and \(j\) of the first variant is transposed,
  yielding a measure constrained by the total number of population and
  to be interpreted as constrained `market potential'.
\end{itemize}

\subsubsection{Total constrained accessible opportunities: Hansen's
accessibility with a total
constraint}\label{total-constrained-accessible-opportunities-hansens-accessibility-with-a-total-constraint}

In the total constrained case, accessibility is expressed as a share of
the total number of opportunities in the region \(D\), allocated based
on travel impedance. The total constrained accessibility from \(i\) to
\(j\):
\begin{equation}\phantomsection\label{eq-total-constrained-access}{
V^T_{ij} = \kappa_{ij}^T D
}\end{equation}

This formulation satisfies the total constraint, analogous to the one in
the Wilson framework:
\begin{equation}\phantomsection\label{eq-total-constraint-D}{
\sum_i \sum_j V^T_{ij} =  D
}\end{equation}

Next, the proportional allocation factor \(\kappa_{ij}^T\) determines
the share of total opportunities assigned to each origin--destination
pair, based on the relative proportion of opportunities-weighted travel
impedance: \[
\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
\] This renders \(V_i^T\) (equal to \(\sum_jV_{ij}^T\)) into units of
opportunities (e.g., physicians), and allows direct interpretation and
comparison of results between zones and scenarios.

Alternatively, this formulation can be rewritten to be expressed using a
total constrained balancing factor \(K^T\), which scales Hansen's
unconstrained accessibility \(V_i^0\) to meet the total opportunity
constraint:
\begin{equation}\phantomsection\label{eq-total-constrained-accessibility}{
V^T_i = \sum_j V^T_{ij} = K^T \sum_j W^{(2)}_jf(c_{ij}) = K^T  V^0_i
}\end{equation}

Where the total constrained balancing factor \(K^T\) is:
\begin{equation}\phantomsection\label{eq-total-opportunity-balancing-factor}{
K^T = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
}\end{equation}

This expression is consistent with Wilson's entropy-maximizing framework
and analogous to the total flow spatial interaction model (e.g.,
Equation 2.11 in \citep{cliff_evaluating_1974}).

In summary, \(\kappa_{ij}^T\) proportionally allocates the total number
of opportunities \(D\) to each origin--destination pair based on
relative opportunity-weighted travel impedance. These values can be
aggregated across destinations to obtain total constrained accessibility
at each origin. Alternatively, the measure can be expressed using the
balancing factor \(K^T\), demonstrating that it is algebraically
proportional to unconstrained accessibility \(V_i^0\), but with
interpretable units (i.e., opportunities). This allows for meaningful
comparisons of differences across zones and travel behaviour scenarios.

Referring back to our simple numeric example, \(K^T\) for the highest
decay t scenario \(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:

\[
K^T = \frac{D}{\sum_{i}\sum_{j} W_j^{(2)} f(c_{ij})} = \frac{490}{0.6233} = 786.085
\]

\(K^T\) for other decay scenarios are calculated similarly in code.
Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields total
constrained accessibility values (Table
Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
units of physicians.

\begin{table}

\caption{\label{tbl-simple-example-total-opportunity-accessibility}Simple
system: total constrained accessible opportunities.}

\centering{

\fontsize{6.0pt}{7.2pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
\toprule
 & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{T}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Origin & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
\midrule\addlinespace[2.5pt]
1 & 172.065 & 172.672 & 164.080 \\ 
2 & 131.627 & 132.247 & 160.692 \\ 
3 & 186.308 & 185.081 & 165.228 \\ 
\midrule 
\midrule 
Sum & 490 & 490 & 490 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Compared to the unconstrained case, values now sum to the known regional
total \(D\), allowing interpretation of absolute and relative
differences across zones and travel scenarios. For example, in the
highest decay case, Zone 1 (Urban Edge) captures an intermediate number
of physicians (172.0652825), like in the unconstrained accessibility
case. However, unlike in the unconstrained case, we can say that this
value is out of the 490 physicians in the region, which allows us also
to deduce that zone 1 captures 1.3072213 and 0.9235529 times more than
zone 2 and 3. Values for the lesser decay (\(f_2(c_{ij})\)) and lowest
decay (\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
scenario values also summing to equal 490 physicians accessible in the
region.

One can also directly compare values at a specific zone, across travel
impedance scenarios, due to the consistent units. As the decay scenario
decreases, all zones become more accessible to each other and the
differences between pairs diminishes (i.e., in \(f_3(c_{ij})\) each zone
captures close to an average amount of physicians, a third of 490 or
\textasciitilde163). This averaging of the total amount is how the total
constrained proportional allocation factor functions. However, what's
notable is how Zones change between scenarios. For instance, Zone 1's
share only declines slightly--these declines are outpaced by Zone 2's
relative gains. This shift reflects how \(\kappa_{ij}^T\) redistributes
opportunities in proportion to impedance under different travel
behaviours.

The total constrained accessibility measure resolves the
interpretability issue of Hansen's accessibility (i.e., unconstrained
accessibility) by grounding values in a meaningful total--enabling
robust comparisons across zones and scenarios--but also by keeping
values proportional to \(V_i^0\) so interpretation is similar.

\subsubsection{Total constrained accessible population: Reilly's
potential trade territories with a total
constraint}\label{total-constrained-accessible-population-reillys-potential-trade-territories-with-a-total-constraint}

This second total constrained variant is a transpose of the
opportunity-constrained formulation, switching indices \(i\) and \(j\)
to yield s measure of market potential--the number of people who can
spatially interact with a destination.

Though not outlined in the ``Unconstrained accessibility'' section, the
unconstrained form aligns with Reilly's ``potential trade territories''
\citep{reilly1929methods} and Harris' and Vickerman's formulations of
regional market potential
\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.
In its unconstrained form, market potential has also been more recently
to estimate potentially accessible populations following infrastructure
investments
\citep[e.g.,][]{gutierrezLocationEconomicPotential2001, holl2007twenty, condecco2018road}.
Market potential can also be thought of as a form of \emph{passive
accessibility}, indicating the number of people that can reach each
destination.

However, like \(V_{ij}^0\), issues of unit interpretability arise in
unconstrained market potential \(M_j^0\). To address this, we introduce
the total constrained accessible population measure \(M^T_{ji}\), which
allocates the total population \(O\) across all origin-destination pairs
proportionally:

\begin{equation}\phantomsection\label{eq-total-constrained-market}{
M^T_{ji} = \kappa_{ji}^T O
}\end{equation}

\noindent Subject to the total constraint::
\begin{equation}\phantomsection\label{eq-total-constraint-O}{
\sum_i \sum_j M^T_{ji} =  O
}\end{equation}

Next, \(\hat \kappa_{ij}\) is the total constrained proportional
allocation factor, a dimensionless term which distributes population
based on impedance-weighted accessibility: \[
\hat \kappa_{ij}^T = \frac{W^{(1)}_i f(c_{ij})}{\sum_i\sum_j W^{(1)}_if(c_{ij})}
\]

This renders \(M_j^T\) (equal to \(\sum_jM_{ji}^T\)) into units of
population. Alternatively, this formulation can be rewritten to be
expressed using a total constrained balancing factor \(\hat K^T\), which
scales unconstrained market potential \(M_j^0\) to meet the total
population constraint:
\begin{equation}\phantomsection\label{eq-total-constrained-market-potential}{
M^T_j = \sum_i M^T_{ji} = \hat K^T \sum_i W^{(1)}_if(c_{ij}) = \hat K^T  M^0_j
}\end{equation}

Where the total constrained balancing factor \(\hat K^T\) is:
\begin{equation}\phantomsection\label{eq-total-population-balancing-factor}{
\hat K^T = \frac{D}{\sum_i\sum_j W^{(1)}_if(c_{ij})}
}\end{equation}

In summary, \(\hat \kappa_{ij}^T\) allocates the total number of
population \(O\) proportionally to each origin--destination pair based
on relative population-weighted travel impedance. As well, the measure
can be expressed using the balancing factor \(\hat K^T\), demonstrating
that it is algebraically proportional to unconstrained market potential,
but also yielding interpretable units that allow for meaningful
comparison.

Returning to the numerical example, the balancing factor \(\hat K^T\) is
solved for each travel behaviour scenario, and the market potential of
each zone \(M^T_j\) is expressed as units of population (e.g., the
number of people accessible from each origin at that destination) in
Table~\ref{tbl-simple-example-total-population-accessibility}.

\begin{table}

\caption{\label{tbl-simple-example-total-population-accessibility}Simple
system: total constrained accessible population.}

\centering{

\fontsize{6.0pt}{7.2pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
\toprule
 & \multicolumn{3}{c}{M\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){2-4}
 & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){2-2} \cmidrule(lr){3-3} \cmidrule(lr){4-4}
Destination & units: \emph{population in 10,000s} & units: \emph{population in 10,000s} & units: \emph{population in 10,000s} \\ 
\midrule\addlinespace[2.5pt]
1 & 5.018 & 5.447 & 6.598 \\ 
2 & 8.596 & 7.986 & 6.717 \\ 
3 & 6.386 & 6.567 & 6.684 \\ 
\midrule 
\midrule 
Sum & 20 & 20 & 20 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Readers may note the difference in trends in accessible population
(Table~\ref{tbl-simple-example-total-population-accessibility}) and
accessible physicians (i.e., the preceding subsection,
Table~\ref{tbl-simple-example-total-opportunity-accessibility}).

In Table~\ref{tbl-simple-example-total-opportunity-accessibility}, zone
1, 2, 3 represent destinations and the accessibility values reflect the
number of accessible people from the vantage of physicians. Zone 1, in
its role as a destination, is no longer intermediately-ranked relative
to other zones; it now attracts the fewest number of people across all
three travel behaviour scenarios. However, similar to the total
constrained opportunity case, as travel decay reduces, the availability
of population begins to converge (though Zone 1 continues as the
lowest-ranked) for similar reasons. As decay reduces, the population's
travel impedance to all zones become more similar, making the relative
location of the zones less important and all people in the region more
equally accessible.

Like in the total constrained accessible opportunities variant, the
total constrained accessible population enables direct comparison of raw
values, supporting both ordinal and interval interpretations across
space and travel behaviour scenarios.

\subsection{Singly constrained
accessibility}\label{singly-constrained-accessibility}

The singly constrained accessibility case can also be expressed in two
variants, each defined by the direction in which a constraint is
applied:

\begin{itemize}
\tightlist
\item
  \(V_i^S\): accessibility constrained by opportunities at destinations
  (singly constrained accessible opportunities), and
\item
  \(M_j^S\): its transpose, constrained by population at origins (singly
  constrained accessible population, or market potential).
\end{itemize}

Similar to the total constrained case, the singly constrained measures
adjust unconstrained zonal accessibility values (\(V_i^0\) or \(M_j^0\))
using a balancing factor to satisfy the known system constraint.
However, unlike the total constraint (which enforces a global/total
sum), the singly constrained case applies a localized constraint at one
end of the interaction--either origin or destination.

In the opportunities-constrained variant, the balancing factor ensures
that only a proportion of opportunities at each destination are
allocated to origins, based on their relative demand (population) and
travel impedance. This variant mirrors the concept of spatial
availability as discussed in Soukhov et al.
\citep{soukhovIntroducingSpatialAvailability2023}. In the
population-constrained variant, the logic is reversed: population at
each origin is allocated proportionally across destinations, informed by
the distribution of opportunities and impedance.

In both cases, the singly constrained formulation introduces zonal-level
competition, unlike the total constrained case which distributes a fixed
regional sum. Each zonal accessibility value becomes not only a fraction
of the regional total (opportunities or population), but also a balanced
sum of interactions, weighted by impedance and relative competition. The
result remains in interpretable units--accessible opportunities or
accessible population--but reflects more complex spatial dynamics.

\subsubsection{Singly constrained accessible opportunities: spatial
availability}\label{singly-constrained-accessible-opportunities-spatial-availability}

In this singly constrained variant, accessibility is constrained at the
destination side: the sum of accessible opportunities allocated from
each destination must equal the known number of opportunities \(D_j\).
This is comparable to the single attraction-constraint
(Eq~\ref{eq-constraint2-gravitymodel}) from Wilson's framework:

\begin{equation}\phantomsection\label{eq-opportunity-constraint}{
\sum_i V^S_{ij} =  D_j
}\end{equation}

The underlying spatial interaction model is now the
attraction-constrained model in
Eq~\ref{eq-attraction-constrained-gravitymodel}, and our accessibility
measure becomes:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility}{
V^S_{i} = \sum_j B_j D_j W_i^{(1)} f(c_{ij})
}\end{equation}

\noindent where \(W_i^{(1)}\) is a measure of the mass at origin \(i\)
(i.e., the opportunity-seeking population). The corresponding balancing
factor, as per Wilson, is:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-proportionality-constants}{
B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

Introducing the balancing factor in
Eq~\ref{eq-opportunity-constrained-accessibility}, we obtain:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility-with-balancing-factor}{
V^S_{i} = \sum_j D_j \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

Further, we can express the formula even more simply, by defining the
following proportional allocation factor:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-proportional-allocation-factor}{
\kappa^S_{ij} = \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}
}\end{equation}

After this, it is possible to rewrite
Eq~\ref{eq-opportunity-constrained-accessibility-with-balancing-factor}
as an origin summary expression of proportionally allocated known
opportunities (i.e., \(D_j\)).

\begin{equation}\phantomsection\label{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}{
V^S_{i} = \sum_j \kappa^S_{ij} D_j
}\end{equation}

This formulation has been referred to as \textbf{spatial availability}
by Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023},
since it incorporates spatial competition by allocating opportunities
based on demand (population), impedance, and the known opportunity
totals \(D_j\). The dimensionless factor \(\kappa^S_{ij}\) ensures that
each destination's opportunities are distributed proportionally to
origins. As in the total constrained case, \(V_i^S\) is expressed in the
units of accessible opportunities. Finally, the \textbf{per capita}
version of this measure:

\citet{soukhovIntroducingSpatialAvailability2023} also showed that the
following expression (accessibility per capita) is a constrained version
of the popular 2SFCA approach of Shen \citep{shen1998} and Luo and Wang
\citep{luo2003}:

\begin{equation}\phantomsection\label{eq-opportunity-constrained-accessibility-per-capita}{
v^S_{i} = \frac{V^S_{i}}{W^{(1)}_i}
}\end{equation}

Returning to the simple numeric example, as an example of the solved
\(B_{j}\) for the highest decay travel behaviour \(f_1(c_{ij})\):

\[
\begin{array}{l}
B_{j} = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}\\
B_{1} =  \frac{1}{\frac{4}{10^3} + \frac{10}{30^3} + \frac{6}{15^3}} = 162.6506\\ 
B_{2} =  \frac{1}{\frac{4}{30^3} + \frac{10}{10^3} + \frac{6}{25^3}} = 94.9474\\
B_{3} =  \frac{1}{\frac{4}{10^3} + \frac{10}{25^3} + \frac{6}{10^3}} = 93.9850
\end{array}
\]

The balancing factors \(B_j\) for the \(f_2(c_{ij})\) decay group for
zones 1, 2 and 3 area 12.8571429, 8.7685113 and 10.6635071,
respectively. For the \(f_3(c_{ij})\) decay group, they are 0.0672461,
0.0660559 and 0.0663798. Using these these balancing constants, we can
calculate the singly constrained opportunity accessibility
(Table~\ref{tbl-simple-example-attraction-constrained-accessibility}).

\begin{table}

\caption{\label{tbl-simple-example-attraction-constrained-accessibility}Simple
system: singly constrained accessible opportunities.}

\centering{

\fontsize{6.0pt}{7.2pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|cccc}
\toprule
 &  & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){3-5}
 &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
Origin & Population (10k) & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 133.469 & 122.255 & 98.848 \\ 
2 & 10 & 166.781 & 185.096 & 241.877 \\ 
3 & 6 & 189.750 & 182.650 & 149.275 \\ 
\midrule 
\midrule 
Sum & — & 490 & 490 & 490 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Imposing the single proportional allocation factor \(\kappa^S_{ij}\)
allows for the comparison of differences and ratios of the accessibility
values, like previously discussed in the total constrained accessible
opportunities case. The proportional allocation factor ensures that
resulting values are in units of \emph{physicians}, with the impedance
units already accounted for in the allocation process.

However, unlike \(\kappa^T_{ij}\), \(\kappa^S_{ij}\) introduces zonal
competition based on the mass of the origin (population). In the total
constrained case, opportunities are distributed based on impedance
alone, regardless of population at \(i\). In contrast, the singly
constrained case allocates each zone's opportunities proportionally
across the region based on the relative impedance-weighted demand from
all origins.

This consideration has important implications. Consider the highest
decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1--despite
hosting a medium amount of physicans--captures the fewest physicians
(133.4687282), compared to 166.7813387 at Zone 2, and 189.7499331 at
Zone 3. Why? Zone 1 has the smallest population and is adjacent to Zone
3, the urban core. Its low impedance-weighted demand means
\(\kappa^S_{ij}\) allocates it fewer opportunities. By contrast, in the
total constrained case, Zone 1 fares better, capturing 35\% of all
physicians (compared to 27\%).

As travel decay decreases (e.g., \(f_3(c_{ij})\)), competition becomes
more diffuse. Zone 2, with the largest population, initially dominates
accessibility---but under low decay, Zones 1 and 3 also begin drawing
more opportunities from Zone 2. For instance, Zone 1 gains 6\% more from
Zone 2 in \(f_3(c_{ij})\) than in \(f_1(c_{ij})\). This shift reflects a
drop in \(\kappa^S_{2,2}\) by 14\%, showing Zone 2's decreasing hold on
its own opportunities as other zones gain accessibility `parity'.

This dynamic reveals how \(\kappa^S_{ij}\) embeds both travel impedance
and population competition. Unlike the total constraint lowest decay
scenarios that allocate evenly, the singly constrained case reflects how
competition evolves continues to influence allocation.

In this way, the consideration of constrained accessibility \emph{per
capita} may be clarifying. Often, accessibility values are reported as
raw scores without the consideration for population. But, as we
introduced constraints, these constrained accessibility values can be
normalized using anything that is relevant to the zone. In
Table~\ref{tbl-simple-example-attraction-constrained-accessibility-per-capita},
we present per capita accessibility for the numeric example, simply in
units of number of physicians accessible per population at each zone.
Notably, these per capita rates are equivalent to the 2SFCA values.

\begin{table}

\caption{\label{tbl-simple-example-attraction-constrained-accessibility-per-capita}Simple
system: singly constrained accessible opportunities per capita.}

\centering{

\fontsize{6.0pt}{7.2pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|cccc}
\toprule
 &  & \multicolumn{3}{c}{v\textsubscript{i}\textsuperscript{S}} \\ 
\cmidrule(lr){3-5}
 &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
Origin & Population (10k) & units: \emph{physicians per capita} & units: \emph{physicians per capita} & units: \emph{physicians per capita} \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 33.367 & 30.564 & 24.712 \\ 
2 & 10 & 16.678 & 18.510 & 24.188 \\ 
3 & 6 & 31.625 & 30.442 & 24.879 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

This simple example was constructed so that the regional average equals
24.5 physicians per 10,000 people. As distance decay decreases and
becomes \emph{relatively} uniform (all zones can reach all zones), the
effect of population drives the proportional allocation of
opportunities. Consequently, per capita accessibility values begin to
stabilise to the regional per capita average (e.g., in the lowest
distance decay \(f_3(c_{ij})\), per capita values are all
\textasciitilde24 physicians accessible per capita).

This convergence mirrors the trend in the total constrained opportunity
case, where accessibility values approach a third of the 490 physicians
under the unfettered mobility scenario \(f_3(c_{ij})\). In both cases,
the balancing factors (\(K^S\) and \(B_j\)) act as averaging mechanisms
but at different scales. As distance decay becomes more
\emph{relatively} uniform, the role of remaining variables (i.e., total
population or opportunities) drive the proportional allocation
differences. In the total constrained case, this is the proportion of
opportunities relative to the regional opportunities, and in the case of
the single opportunity constrained case, this is the population at a
zone relative to the regional population.

\subsubsection{Singly constrained accessible population: market
availability}\label{singly-constrained-accessible-population-market-availability}

Similar to Eq~\ref{eq-total-population-balancing-factor} in transposing
the origins and destinations, we can define a \emph{singly constrained}
measure of market potential that preserves the known population (i.e.,
the mass weight at the origin \(W_i^{(1)}\) is now represented by
\(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
constrained concept of market potential been used to express ``facility
crowdedness'' as in Wang \citep{wang_inverted_2018}.

The underlying spatial interaction model is now the
production-constrained model in
Eq~\ref{eq-production-constrained-gravitymodel}, and our market
potential measure \(M^S_j\) becomes:

\begin{equation}\phantomsection\label{eq-population-constrained-accessibility}{
M^S_j = \sum_i A_i O_i W_j^{(2)} f(c_{ij})
}\end{equation}

In this variant, the measure is singly constrained by the population
\emph{by origin} (i.e., \(O_i\)), like
Eq~\ref{eq-constraint2-gravitymodel} from Wilson's framework:

\begin{equation}\phantomsection\label{eq-population-constraint}{
\sum_j M^S_{ji} =  O_i 
}\end{equation}

And the corresponding balancing factor, as per Wilson, is:

\begin{equation}\phantomsection\label{eq-population-constrained-proportionality-constants}{
A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}
}\end{equation}

Following the same logic as in the preceding section on total
constrained market potential, one arrives at the following expression of
accessible population \(M_j^S\) being the product of proportionally
allocated (\(\hat \kappa^S_{ji}\)) popultion:

\begin{equation}\phantomsection\label{eq-production-constrained-accessibility-with-proportional-allocation-factor}{
M^S_{j} = \sum_i \hat \kappa^S_{ji} O_i
}\end{equation}

\noindent with:

\begin{equation}\phantomsection\label{eq-attraction-constrained-proportional-allocation-factor}{
\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}
}\end{equation}

As well, the single (population) constraint in
Eq~\ref{eq-population-constraint} ensures that the the total constraint
(e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O\)) is maintained.

With these constraints, \(\frac{M_j^S}{O}\) can be interpreted as the
proportion of the total population serviced by location \(j\).

For the sake of brevity, we'll move forward onto the doubly constrained
case.

\subsection{Doubly constrained
accessibility}\label{doubly-constrained-accessibility}

This accessibility case adopts the structure of the doubly constrained
spatial interaction model, where \(V_{ij}^D\) flows are constrained by
both origin populations \(O_i\) and destination opportunities \(D_j\).
That is, the resulting accessibility outflow from each origin must match
the origin's population demand, and the resulting accessibility inflow
to each destination must match the number of opportunities supplied:

\begin{equation}\phantomsection\label{eq-opportunity-population-equality-1}{
\sum_j V_{ij}^D = O_i \text{ and }  \sum_i V_{ij}^D =  D_j
}\end{equation}

Because both margins are fixed, this measure reflects fully allocated
interaction flows, not latent potential. As such, accessibility cannot
be interpreted as a traditional summary at \(i\) or \(j\) (e.g.,
``opportunities accessible from \(i\)'')--those sums simply reproduce
the constraint totals. Instead, the meaningful unit of analysis is the
\(ij\) flow itself.

This distinguishes doubly constrained accessibility from the total and
singly constrained cases discussed previously. In those cases, only one
side of the interaction--either the total marginal or
opportunity/population marginal--was constrained, while the other was
treated as a demand/supply weight (e.g., \(D\)/\(O\) or
\(W_j^{(2)}\)/\(W_i^{(1)}\)), without any allocation details. As a
result, total constrained models express more `potential' from both
sides of the system, and singly constrained models express potential
access from one side.

By contrast, the doubly constrained model assumes both the demand
(population) and the supply (opportunity) are known and bounded, and
allocates flows accordingly. This makes it less suitable for traditional
accessibility analysis, namely because origin and destination masses
often differ in kind and units. For instance, the number of people
accessing an opportunity such as park may be known, but the capacity of
each park is not known. A doubly constrained approach only makes
conceptual sense when population and opportunities are comparable, have
a one-to-one correspondence or are paired somehow--e.g., job per worker,
student per school-seat, or vaccine doses per person.

A doubly constrained approach to accessibility adopts the structure of
the production-attraction spatial interaction model, where both
population and opportunity totals are fixed. Mathematically, this model
requires the simultaneous imposition of both the population- and
opportunity- constraints in the preceding singly constrained variants
(Eq~\ref{eq-opportunity-constraint} and
Eq~\ref{eq-population-constraint}), namely the sum of population in all
origins should match the sum of opportunities in all destinations
(Eq~\ref{eq-opportunity-population-equality-2}):

\begin{equation}\phantomsection\label{eq-opportunity-population-equality-2}{
\sum_i O_i = \sum_j D_j
}\end{equation}

As before, the simultaneous imposition of both constraints ensures the
total system constraint is maintained i.e., the sum of all doubly
constrained accessibility values
\(\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D\) remains equal to the
total number of opportunities in the region \(O\).

As the doubly constrained accessibility measure \(V_{ij}^D\) takes the
form of the production-attraction (doubly constrained) spatial
interaction model, as shown in
Eq~\ref{eq-doubly-constrained-gravitymodel}, \(V_{i}^D\) is as follows:

\begin{equation}\phantomsection\label{eq-doubly-constrained-accessibility}{
V_{ij}^D = A_i B_j O_i D_j f(c_{ij})
}\end{equation}

\noindent where the corresponding balancing factors \(A_i\) and \(B_j\),
as per Wilson, are:

\[
\begin{array}{l}
A_i = \frac{1}{\sum_j B_j D_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i A_i O_i f(c_{ij})}
\end{array}
\]

Calibration of the two sets of proportionality constants is accomplished
by means of iterative proportional fitting, whereby the values of
\(A_i\) are initialized as one for all i to obtain an initial estimate
of \(B_j\). The values of \(B_j\) are used to update the underlying
\(V_{ij}^D\) matrix, before calibrating \(A_i\). This process continues
to update \(A_i\) and \(B_j\) until a convergence criterion is met
\citep[see][p.~193-195]{ortuzar_2011_modelling}.

The doubly constrained model completely distributes origin populations
to destination opportunities according to travel impedance and
supply-demand balance. This ensures that: summing \(V^D_{ij}\) across
\(j\) returns \(O_i\); summing across \(i\) returns \(D_j\). Thus,
aggregating over \(i\) or \(j\) yields only the known constraints. In
this way, a per-capita form (e.g., \(V^D_i / O_i\)) is not
meaningful--since the output already reflects population-normalized
allocation. As such, the \(ij\) matrix \(V^D_{ij}\) is the only
interpretable output.

We could define the proportional allocation factor \(\kappa_{ij}^D\)
such that:

\[
\kappa_{ij}^D = \sum_j \frac{1}{\sum_j B_j D_j f(c_{ij})} \frac{1}{\sum_i A_i O_i f(c_{ij})} O_i f(c_{ij})
\] \noindent and represent \(V^D_{ij}\) as equal to
\(\kappa^D_{ij} D_j\), allowing the analyst to understand the
proportional allocation of D\_js to each \(ij\) flow.

Following this logic, the market potential form \(M^D_{ji}\) is
effectively equivalent to \(V_{ij}^D\), but can be read with a different
interpretation: i.e., the opportunities accessed from \(j\) at an \(i\)
vs.~the population accessed from \(i\) at a \(j\). The inputs of
`opportunities accessed' and `accessed population' can already be
interpreted as inherently being sensitive to both opportunities and
population.

Moving onto the toy example, to calculate doubly constrained
accessibility, the interpretation of the population data and the counts
of the opportunity data in the numeric example must be reinterpreted.

Namely, a count of physician \emph{capacity} per destination \(D_j\) is
needed instead of simply the number of physicians. Also, we must be able
to clearly state that the population is the \emph{capacity} of the
origin to interact with opportunities \(O_i\), i.e., the count of people
seeking opportunities.

This adjusted simple example is summarised in
Table~\ref{tbl-small-system-land-use-doubly-constrained-case}: with the
population (in units of 10,000s of people seeking physicians) and the
opportunities (in units of 10,00s of physician-capacity) per zone. For
the population, we leave this unchanged numerically but theoretically
note that each person interacts with one physician capacity. However,
the number of providers per destination is revised to represent
physician capacity, scaled approximately from the original number of
physicians used in previous cases
(Table~\ref{tbl-small-system-land-use}).The system-wide
provider-to-population ratio (PPR) is now:1, this is compared to the
unmodified example which yields system PPR of24.5.

We keep the same zonal cost matrix, and travel impedance functions for
three types of travel behaviour as before
(Table~\ref{tbl-small-system-cost} and
Eq~\ref{eq-travel-behaviour-scenarios}).

\begin{table}

\caption{\label{tbl-small-system-land-use-doubly-constrained-case}Modified
simple system with three zones reflecting matched population and
opportunities. Population is in 10,000 persons and opportunities in
10,000 of physician-capacity.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}rcc}
\toprule
ID (i or j) & Population & Opportunities \\ 
\midrule\addlinespace[2.5pt]
1 & 4 & 7 \\ 
2 & 10 & 5 \\ 
3 & 6 & 8 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

However, despite the modifications to the example, our objective remains
the same: to measure accessibility under different travel behavior
scenarios. We aim to quantify the number of \emph{potential} spatial
interactions between the physician-seeking population in each zone \(i\)
to the physician capacity in a zone \(j\) i.e., \(V_{ij}^D\). The
highest decay travel behaviour scenario (\(f_1(c_ij)\)) is presented in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}.

\begin{table}

\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}Doubly
constrained accessible opportunities assuming highest travel decay in
the modified simple system.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|rcccc}
\toprule
 &  & \multicolumn{3}{c}{Destination ID} &  \\ 
\cmidrule(lr){3-5}
 & Origin ID & 1 & 2 & 3 & sum \\ 
\midrule\addlinespace[2.5pt]
 & 1 & 3.235859 & 0.01032226 & 0.7556568 & 4 \\ 
 & 2 & 2.132602 & 4.95932483 & 2.9044391 & 10 \\ 
 & 3 & 1.631539 & 0.03035291 & 4.3399040 & 6 \\ 
\midrule 
\midrule 
Sum & — & 7 & 5 & 8 & — \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}
for the highest-decay scenario \(f_1(c_{ij})\), accessibility is no
longer meaningfully represented as zonal summaries like \(V^D_i\) or
\(M^D_j\), since these values reproduce the original constraints--i.e.,
\(V^D_i = O_i\) so the physican-capacity accessible for Zones 1, 2, and
3 would be 4.0018381, 9.9963656, and 6.0017963. Hence, the usefulness of
the doubly constrained measure lies in the interpretation as
\(V_{ij}^D\) values. \(V^D_{ij}\) values represent the number of
opportunities from zone \(j\) allocated to populations in zone \(i\),
shaped by both mass and travel impedance.

To illustrate this, consider the results for Zone 2 (Suburban zone--a
high-population, low-opportunity, relatively remote zone). As shown in
Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2},
its intrazonal allocation (i.e., \(V^D_{22}\)) declines as travel
impedance decay decreases--from 4.9593248 under \(f_1(c_{ij})\) to
2.6672837 under \(f_3(c_{ij})\), out of the \textasciitilde10
opportunities allocated to Zone 2 (a population of 10).

Following the intuition discussed in the singly constrained opportunity
case, as decay decreases (i.e., more relatively uniform for all zones),
the mass effects (effect of the population and opportunities magnitudes)
become more relatively dominant in the spatial allocation.

\begin{table}

\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}Doubly
constrained accessible opportunities at Zone 2 for all travel decay
groups in the modified simple system.}

\centering{

\fontsize{7.5pt}{9.0pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 22.50pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}}
\toprule
 &  &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 202.50pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{\{ij\}}\textsuperscript{D}} \\ 
\cmidrule(lr){4-6}
 &  &  & f\textsubscript{1} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{3} & f\textsubscript{2} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{2} & f\textsubscript{3} (c\textsubscript{ij}) = 1/c\textsubscript{ij}\textsuperscript{0.1} \\ 
\cmidrule(lr){4-4} \cmidrule(lr){5-5} \cmidrule(lr){6-6}
Dest. & Population at 2 (units: \emph{people in 10,000s}) & Opportunities (units: \emph{capacity in 10,000s}) & units: \emph{physician-capacity in 10,000s} & units: \emph{physician-capacity in 10,000s} & units: \emph{physician-capacity in 10,000s} \\ 
\midrule\addlinespace[2.5pt]
1 & 10.000 & 7.000 & 2.133 & 2.272 & 3.411 \\ 
2 & 10.000 & 5.000 & 4.959 & 4.766 & 2.667 \\ 
3 & 10.000 & 8.000 & 2.904 & 2.958 & 3.919 \\ 
\bottomrule
\end{tabular*}

}

\end{table}%

Recall, accessibility is defined as ``the potential of spatial
interaction'' and is traditionally presented as a summary zonal measure.
In the doubly constrained case, we force the allocation of zonal
population demand and zonal opportunities supplied to be paired and
allocation to be proportional, hence the sum of \(V_{ij}^D\) for all
\(i\)s ends up being equal to the population at the zone (and
\(\sum_i V_{ij}^D\) being the opportunities at the zone).

So, in the doubly-constrained case, accessibility flows becomes simply
equivalent to spatial interaction flows--spatial proportional
allocations between population and opportunity capacity. In fact,
\(V^D_{ij}\) and its transposed counterpart \(M^D_{ji}\) are
structurally identical, differing only in interpretation (referring to
\(kappa_{ij}^D\) and \(\hat kappa_{ij}^D\)): one reflects the
proportional allocation of opportunity to population flows; the other,
population to opportunity flows.

As Wilson \citep{wilson1971} explicitly noted, origin and destination
weights defined in the spatial interaction model \emph{can} be defined
using any unit; but the focus of these models have typically on \(ij\)
flows and often calibrated using trips (i.e., outbound and inbound
trips, inherently in the same units). On the other hand, accessibility
has been interpreted differently: often understood as a zonal summary of
the \emph{potential} for spatial interaction, conceptualising origins
and destinations in inherently different units.

Tracing the application of increasing more restrictive constraints: from
the total, to the singly- and now in the doubly-constrained measure,
sheds light on how the balancing factors embedded in the unitless
proportional allocation factors restrict the range in ``potential''
spatial interaction values that can be taken.

\section{Conclusions}\label{conclusions}

In this paper, we examined the historical and mathematical commonalities
between spatial interaction models and place-based accessibility
measures. As accessibility research evolved largely influenced by Hansen
\citep{hansen1959}, researchers in the field neglected the
proportionality constant that was originally present in gravity-based
models, and is still present in spatial interaction modeling.

Firstly, we have demonstrated that by reintroducing Wilson's system
constraints and defining associated balancing factors and proportional
allocation factors, we can derive a unifying family of accessibility
measures that reintroduces units to the resulting values. These values
become more intuitive for the purpose of analysis and comparison.
Secondly, the constraining constants, depending on the case (i.e.,
total, singly- or doubly- constrained), restrict the degree of
\emph{potential}, linking accessibility (the potential for spatial
interaction) with access (spatial interaction) on the same continuum
based on the constraint used. Lastly, we discussed how popular measures
such as the one used in Hansen \citep{hansen1959} and the 2SFCA link
into the family of accessibility measures.

A summary of the family of accessibility measures, is detailed; First,
we place the popular Hansen-type accessibility measure
\citep{hansen1959} within this family of measures as an
``unconstrained'' case, demonstrating that resulting values cannot be
directly compared across different travel scenarios without ad-hoc
adjustments. We then show how applying a total constraint balances the
units and produces a statistically averaged solution that converges to
the regional average for each zone as the decay effect decreases. In
other words, the total-constraint model could be a more interpretable
alternative for the unconstrained case if population-competition is not
relevant and one is interested in capturing the maximum
\emph{potential}; specifically, if there is a fixed number of
opportunities in the region, and if it makes sense to assume that people
accessing proximate opportunities leave fewer for others, \emph{without}
considering the population size at the origins.

We then introduce the singly-constrained case, which \emph{does} take
into account the population size at the origin in the allocation of
opportunities. It is also mathematically equivalent to the spatial
availability introduced in Soukhov et al.
\citep{soukhovIntroducingSpatialAvailability2023}. In this case, all
accessibility values are fixed to sum to a known zonal opportunity-size
value (implicitly, the regional total of opportunities), but they are
not required to sum to any population-based values at the zone or
regional level. The singly constrained model could be useful if regional
competition is a factor and if the acknowledgment that only a finite
number of opportunities can be allocated from each destination (with
those allocations distributed based on origin population size) is
suitable. We also introduce an `accessible' PPR (e.g., opportunities per
capita), calculated by dividing each accessibility value by the zonal
population. To clarify, this per capita expression of the singly
constrained case is equivalent to the 2SFCA \citep{luo2003, shen1998},
hence linking this literature back to spatial interaction principles.

Lastly, the doubly constrained case is introduced. In this case, the
sums must equal both the regional total and ensure that no zone
allocates more opportunities than it has available. Specifically,
accessibility values for each \(i\)-\(j\) pair must be a proportion of
he zonal opportunity and population values simultaneously. For example,
the accessibility at zone 1 must equal the sum of opportunities from
zones 1, 2, and 3, as well as the sum of the population at zone 1.
Satisfying the double constraint means the opportunities and population
data must match one-to-one, so working with the accessibility
\(i\)-\(j\) pair values should be of interest. In this sense, the
research question should be concerned with `access' (how many
opportunities accessed from \(j\) at \(i\) based on given zonal
opportunities and populations) instead of potential spatial interaction
(e.g., typically expressed as a zonal summary measure of how many
opportunities one could reach (out of a regional total and/or
zonal-allocation).

In summary, building on Wilson's \citep{wilson1971} foundational work,
this paper proposed a unified framework for accessibility that is able
to account for competition. By reintroducing Wilson's proportionality
constant, the proposed family of constrained accessibility measures
restores measurement units to accessibility estimates. This enhancement
provides a more interpretable, consistent, and theoretically grounded
basis for accessibility analysis, which could help advance the adoption
of accessibility-oriented planning.

\nolinenumbers
  \bibliography{bibliography.bib}

\end{document}

### 2025-08-01 paper_plos/paper_plos_V0.tex
Made the existing draft as concise as possible -- shed 6 pages. Old version is paper_plos_V0.qmd and .pdf.

--- 2025-05-26 paper/manuscript.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
+
+\PassOptionsToPackage{table}{xcolor}
+
 \documentclass[
+  10pt,
+  letterpaper,
 ]{article}
@@
 \usepackage{xcolor}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% marvosym package for additional characters
+\usepackage{marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+% Using natbib instead
+% \usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Remove brackets from numbering in List of References
 \makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\renewcommand{\@biblabel}[1]{\quad#1.}
 \makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
 \usepackage{booktabs}
@@
 \floatplacement{figure}{H}
+\usepackage{pdflscape}
+\newcommand{\blandscape}{\begin{landscape}}
+\newcommand{\elandscape}{\end{landscape}}
 \makeatletter
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
+\usepackage[numbers,square,comma]{natbib}
+\bibliographystyle{plos2015}
 \usepackage{bookmark}
@@
   pdfauthor={Anastasia Soukhov; Rafael H. M. Pereira; Christopher D. Higgins; Antonio Páez},
-  pdfkeywords={accessibility; spatial interaction model; balancing
-factors; competitive accessibility; constraints; singly constrained;
-doubly constrained; total constrained},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
+  hidelinks,
   pdfcreator={LaTeX via pandoc}}
 
-\title{A family of accessibility measures derived from spatial
-interaction principles}
-\author{Anastasia Soukhov \and Rafael H. M. Pereira \and Christopher D.
-Higgins \and Antonio Páez}
-\date{}
-
 \begin{document}
-\maketitle
-\begin{abstract}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{A family of accessibility measures derived from spatial
+interaction
+principles} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+\\
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+Anastasia Soukhov\textsuperscript{1}, Rafael H. M.
+Pereira\textsuperscript{2}, Christopher D.
+Higgins\textsuperscript{3*}, Antonio Páez\textsuperscript{1}
+\\
+\bigskip
+\textbf{1} McMaster University, School of Earth, Environment \&
+Society, Hamilton, Canada, \\ \textbf{2} Institute for Applied Economic
+Research - Ipea, Data Science Division, Brasília,
+Brazil, \\ \textbf{3} University of Toronto Scarborough, Department of
+Human Geography, Toronto, Canada, 
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+%\ddag These authors also contributed equally to this work.
+
+% Current address notes
+\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+\dag Deceased
+
+% Group/Consortium Author Note
+\textpilcrow Membership list can be found in the Acknowledgments
+sections
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* cd.higgins@utoronto.ca
+
+\end{flushleft}
+
+\section*{Abstract}
 Transportation planning has long prioritized the efficiency of movement,
@@
 interpretable, and cohesive foundation for accessibility analysis.
-\end{abstract}
+
+\linenumbers
 
@@
 entrenching an automobility mono-culture within transportation systems
-(H. J. Miller 2011; Lavery, Páez, and Kanaroglou 2013). Decades of
+\citep{miller_collaborative_2011, lavery_driving_2013}. Decades of
 planning for this automobility mono-culture, still often characterized
@@
 costs and environmental burdens, with limited impact on enhancing the
-ease with which people can reach destinations (Steven Farber and Páez
-2011; S. Handy 2002; Páez et al. 2010). In response, transportation
-researchers have increasingly advocated for the adoption of
-accessibility as a planning criterion, in contrast to traditional
-mobility-oriented transportation planning approaches which translate
-into indicators that benchmark movement (e.g., vehicle kilometres
-traveled, intersection through traffic, etc.) which are not necessarily
-linked to improved accessibility (Silva et al. 2017; Paez et al. 2013;
-S. Handy 2020; El-Geneidy and Levinson 2022).
+ease with which people can reach destinations
+\citep{farber_running_2011, handyACCESSIBILITYVSMOBILITYENHANCING2002, paez_healthcare_2010}.
+In response, transportation researchers have increasingly advocated for
+the adoption of accessibility as a planning criterion, in contrast to
+traditional mobility-oriented transportation planning approaches which
+translate into indicators that benchmark movement (e.g., vehicle
+kilometres traveled, intersection through traffic, etc.) which are not
+necessarily linked to improved accessibility
+\citep{silvaAccessibilityInstrumentsPlanning2017, paez_developing_2013, handy2020, elgeneidyMakingAccessibilityWork2022}.
 
 Accessibility may be defined as the ``potential of opportunities for
-{[}spatial{]} interaction'' (Hansen 1959), in contrast to mobility which
-is simply, movement in itself or `spatial interaction'. Mobility has
-been the basis of traditional transportation planning approaches
-(Ortúzar and Willumsen 2011). While accessibility brings a more holistic
-understanding of combined transportation and land use systems,
+{[}spatial{]} interaction'' \citep{hansen1959}, in contrast to mobility
+which is simply, movement in itself or `spatial interaction'. Mobility
+has been the basis of traditional transportation planning approaches
+\citep{ortuzar_2011_modelling}. While accessibility brings a more
+holistic understanding of combined transportation and land use systems,
 incorporating both the explicit consideration of destination as
-opportunity and the concept of \emph{potential} (S. L. Handy and
-Niemeier 1997).
+opportunity and the concept of \emph{potential}
+\citep{handyMeasuringAccessibilityExploration1997}.
 
@@
 research contexts; it has grown to include studies of access to
-employment (e.g., Karst and Van Eck 2003; Grengs 2010; Páez et al. 2013;
-Merlin and Hu 2017; Tao et al. 2020), health care (e.g., Luo and Wang
-2003; Páez et al. 2010; Wan, Zou, and Sternberg 2012; Delamater 2013;
-Boisjoly, Moreno-Monroy, and El-Geneidy 2017; Pereira et al. 2021),
-green spaces (Reyes, Paez, and Morency 2014; Rojas et al. 2016; Liang,
-Yan, and Yan 2024), schools (e.g., Williams and Wang 2014; Romanillos
-and Garcia-Palomares 2018; Marques, Wolf, and Feitosa 2021), social
-contacts (e.g., Neutens et al. 2007; S. Farber, Páez, and Morency 2012;
-S. Farber et al. 2013), and regional economic analysis (e.g., R.
-Vickerman, Spiekermann, and Wegener 1999; Lopez, Gutierrez, and Gomez
-2008; Ribeiro, Antunes, and Páez 2010; Gutierrez et al. 2011) among many
-other domains of application. In other words, accessibility analysis is
-used today to broadly understand the potential to reach (or spatially
-interact with) opportunities that are important to people (Ferreira and
-Papa 2020). However, despite its growth in popularity in scholarly
-works, challenges remain with respect to the more widespread adoption of
-accessibility in planning practice. For instance, the diversity of
-accessibility definitions has been flagged by van Wee (2016), S. Handy
-(2020), and Kapatsila et al. (2023). Further, difficulties in the
+employment
+\citep[e.g.,][]{karstEvaluationAccessibilityImpacts2003, grengs2010job, paez_jobs_2013, merlin2017competition, tao_investigating_2020},
+health care
+\citep[e.g.,][]{luo2003, paez_healthcare_2010, wan2012three, delamater2013spatial, boisjoly2017informality, pereira_2021_geographic},
+green spaces
+\citep{reyesAccessibility2014, rojas_accessibility_2016, liang_novel_2024},
+schools
+\citep[e.g.,][]{williams_disparities_2014, romanillosAccessibilitySchoolsSpatial2018, marques_accessibility_2021},
+social contacts
+\citep[e.g.,][]{neutens_human_2007, farberActivitySpacesMeasurement2012, farber_2013_social},
+and regional economic analysis
+\citep[e.g.,][]{vickermanAccessibility1999, lopezMeasuring2008, ribeiro_road_2010, gutierrez_evaluating_2011}
+among many other domains of application. In other words, accessibility
+analysis is used today to broadly understand the potential to reach (or
+spatially interact with) opportunities that are important to people
+\citep{ferreiraReenactingMobilityAccessibility2020}. However, despite
+its growth in popularity in scholarly works, challenges remain with
+respect to the more widespread adoption of accessibility in planning
+practice. For instance, the diversity of accessibility definitions has
+been flagged by van Wee \citep{vanweeAccessible2016}, Handy
+\citep{handy2020}, and Kapatsila et al.
+\citep{kapatsila_resolving_2023}. Further, difficulties in the
 interpretability and communicability of outputs has also been noticed by
-many authores, including Karst T. Geurs and van Wee (2004), van Wee
-(2016), and Ferreira and Papa (2020).
+many authores, including Geurs and van Wee
+\citep{geursAccessibilityEvaluationLanduse2004}, van Wee
+\citep{vanweeAccessible2016}, and Ferreira and Papa
+\citep{ferreiraReenactingMobilityAccessibility2020}.
 
@@
 definitions, each seemingly more sophisticated but less intuitive than
-the last (Kapatsila et al. 2023). At a high level, Karst T. Geurs and
-van Wee (2004) identify four families of accessibility measures:
-infrastructure-, place-, person-, and utility-based. Of the place-based
-family, which is the focus of this paper, the menu has grown to include
-gravity-based accessibility (e.g., Hansen 1959; Pirie 1979), cumulative
-opportunities (e.g., Wachs and Kumagai 1973; Pirie 1979; Ye et al.
-2018), modified gravity (e.g., Schuurman, Berube, and Crooks 2010),
-2-Step Floating Catchment Areas (e.g., Luo and Wang 2003), Enhanced
-2-Step Floating Catchment Areas (e.g., Luo and Qi 2009), 3-Stage
-Floating Catchment Areas (e.g., Wan, Zou, and Sternberg 2012), Modified
-2-Step Floating Catchment Areas (e.g., Delamater 2013), inverse 2-Step
-Floating Catchment Areas (e.g., F. Wang 2021), and n-steps Floating
-Catchment Areas (Liang, Yan, and Yan 2024). How is a practitioner to
-choose among this myriad options? What differences in accessibility
-scores should matter, and how should they be communicated? {[}see van
-Wee (2016); p.~14{]}.
+the last \citep{kapatsila_resolving_2023}. At a high level, Geurs and
+van Wee \citep{geursAccessibilityEvaluationLanduse2004} identify four
+families of accessibility measures: infrastructure-, place-, person-,
+and utility-based. Of the place-based family, which is the focus of this
+paper, the menu has grown to include gravity-based accessibility
+\citep[e.g.,][]{hansen1959, pirie_measuring_1979}, cumulative
+opportunities
+\citep[e.g.,][]{wachs_physical_1973, pirie_measuring_1979, ye_spatial_2018},
+modified gravity \citep[e.g.,][]{SchuurmanMeasuring2010}, 2-Step
+Floating Catchment Areas \citep[e.g.,][]{luo2003}, Enhanced 2-Step
+Floating Catchment Areas \citep[e.g.,][]{luoEnhanced2009}, 3-Stage
+Floating Catchment Areas \citep[e.g.,][]{wan2012three}, Modified 2-Step
+Floating Catchment Areas \citep[e.g.,][]{delamater2013spatial}, inverse
+2-Step Floating Catchment Areas \citep[e.g.,][]{wang_2sfca_2021}, and
+n-steps Floating Catchment Areas \citep{liang_novel_2024}. How is a
+practitioner to choose among this myriad options? What differences in
+accessibility scores should matter, and how should they be communicated?
+\citep[see][ p.~14]{vanweeAccessible2016}.
 
@@
 mobility-oriented models--such as the commonly used gravity-based
-accessibility (e.g., Hansen 1959) and the spatial interaction model
-(e.g., Wilson 1971)-- are rooted in the same spatial interaction
-modeling foundation. In fact, we propose that accessibility can be
-specified using spatial interaction principles as a \emph{family of
-measures}, akin to the family of spatial interaction models introduced
-in Wilson (1971) that can be defined using balancing factors that
-constrains values based on known information about the land-use.
+accessibility \citep[e.g.,][]{hansen1959} and the spatial interaction
+model \citep[e.g.,][]{wilson1971}-- are rooted in the same spatial
+interaction modeling foundation. In fact, we propose that accessibility
+can be specified using spatial interaction principles as a \emph{family
+of measures}, akin to the family of spatial interaction models
+introduced in Wilson \citep{wilson1971} that can be defined using
+balancing factors that constrains values based on known information
+about the land-use.
 
@@
 explicitly connected to popular measures such as the Hansen-type
-accessibility (Hansen 1959), the popular competition approach of the
-2-Step Floating Catchment Area (2SFCA) (Q. Shen 1998a; Luo and Wang
-2003), and the concept of market potential (C. D. Harris 1954; R. W.
-Vickerman 1974). These Wilson-analogous balancing factors introduced are
-also defined in order of increasing restrictiveness to shed light on the
-role of \emph{potential} in accessibility and access. (3) This work also
+accessibility \citep{hansen1959}, the popular competition approach of
+the 2-Step Floating Catchment Area (2SFCA)
+\citep{shenLocationCharacteristicsInnercity1998, luo2003}, and the
+concept of market potential
+\citep{harris_market_1954, vickermanAccessibilityAttractionPotential1974}.
+These Wilson-analogous balancing factors introduced are also defined in
+order of increasing restrictiveness to shed light on the role of
+\emph{potential} in accessibility and access. (3) This work also
 demonstrates that the introduction of balancing factors makes
@@
 accessibility analysis. Empirical constraints were embraced by early
-spatial interaction literature following the work of Wilson (1971), but
-this stream of literature tended to flow separately from research
-inspired by Hansen (1959)' accessibility. The application of Wilson
-(1971)`s empirical constraints supported the development of various
-spatial interaction models that remain relevant in research and practice
-today (Ortúzar and Willumsen 2011). However, the same cannot be said of
-the contemporary accessibility literature, where empirical constraints
-were not explicitly adopted. We argue that the absence of empirical
+spatial interaction literature following the work of Wilson
+\citep{wilson1971}, but this stream of literature tended to flow
+separately from research inspired by Hansen's \citep{hansen1959}
+accessibility. The application of Wilson's \citep{wilson1971} empirical
+constraints supported the development of various spatial interaction
+models that remain relevant in research and practice today
+\citep{ortuzar_2011_modelling}. However, the same cannot be said of the
+contemporary accessibility literature, where empirical constraints were
+not explicitly adopted. We argue that the absence of empirical
 constraints (and their attendant proportionality constants) has
@@
 for instance, the fuzziness of insights beyond simple proportional
-statements like `higher-than' or `lower-than' (E. J. Miller 2018).
-Moreover, without constraints we lose track of what accessibility
-measures, that is, the number of opportunities that can be spatially
-interacted with. Without a clear sense of measurement units,
-comparability between accessibility measures, across cities, and
-transport modes, may also become compromised.
+statements like `higher-than' or `lower-than'
+\citep{millerAccessibilityMeasurementApplication2018}. Moreover, without
+constraints we lose track of what accessibility measures, that is, the
+number of opportunities that can be spatially interacted with. Without a
+clear sense of measurement units, comparability between accessibility
+measures, across cities, and transport modes, may also become
+compromised.
 
@@
 development of accessibility from its origins in spatial interaction:
-from the Newtonian gravitational expression in Ravenstein (1889) through
-to the seminal accessibility work of Hansen (1959). We then present
-evidence for a narrative highlighting the marked divergence between
-accessibility and spatial interaction modelling research after the work
-of Wilson (1971). Next, we hark back to Wilson (1971)'s spatial
-interaction models, and use them to derive a family of accessibility
-measures based on analogous constraints. We illustrate members of this
-family of constrained accessibility measures with a simple numerical
-example. We then conclude by discussing the uses of these measure and
-their interpretation.
+from the Newtonian gravitational expression in Eavenstein
+\citep{ravensteinLawsMigration1889} through to the seminal accessibility
+work of Hansen \citep{hansen1959}. We then present evidence for a
+narrative highlighting the marked divergence between accessibility and
+spatial interaction modelling research after the work of Wilson
+\citep{wilson1971}. Next, we hark back to Wilson's \citep{wilson1971}
+spatial interaction models, and use them to derive a family of
+accessibility measures based on analogous constraints. We illustrate
+members of this family of constrained accessibility measures with a
+simple numerical example. We then conclude by discussing the uses of
+these measure and their interpretation.
 
@@
 scientific inquiry for at least a century and a half. From as far back
-as Henry C. Carey's \emph{Principles of Social Science} (Carey 1858), a
-concern with the scientific study of human spatial interaction can be
-observed. It was in this work where Carey stated that ``man {[}is{]} the
-molecule of society {[}and their interaction is subject to{]} the direct
-ratio of the mass and the inverse one of distance'' (McKean 1883,
-37--38). This statement shows how investigations into human spatial
-interaction have often been explicitly coloured by the features of
-Newton's Law of Universal Gravitation, first posited in 1687's
-\emph{Principia Mathematica} and expressed as in
-Equation~\ref{eq-phys-grav-prop}.
+as Henry C. Carey's \emph{Principles of Social Science}
+\citep{careyPrinciplesSocialScience1858}, a concern with the scientific
+study of human spatial interaction can be observed. It was in this work
+where Carey stated that ``man {[}is{]} the molecule of society {[}and
+their interaction is subject to{]} the direct ratio of the mass and the
+inverse one of distance'' \citep[pp.~37-38]{mckeanManual1883}. This
+statement shows how investigations into human spatial interaction have
+often been explicitly coloured by the features of Newton's Law of
+Universal Gravitation, first posited in 1687's \emph{Principia
+Mathematica} and expressed as in Eq~\ref{eq-phys-grav-prop}.
 
@@
 means that as the distance increases, the force decreases.
-Equation~\ref{eq-phys-grav-prop}, however, does not quantify the
-magnitude of the force. To do so, an empirical constant, a.k.a. the
-gravitational constant, is required to convert the proportionality into
-an equality, ensuring that values of the force \(F\) in
-Equation~\ref{eq-phys-grav-prop} match the observed force of attraction
-between masses. In other words, Equation~\ref{eq-phys-grav-prop} needs
-to be \emph{constrained} using empirical data. Ultimately, the equation
-for the force is as seen in Equation~\ref{eq-phys-grav}, where \(G\) is
-an empirically calibrated proportionality constant:
+Eq~\ref{eq-phys-grav-prop}, however, does not quantify the magnitude of
+the force. To do so, an empirical constant, a.k.a. the gravitational
+constant, is required to convert the proportionality into an equality,
+ensuring that values of the force \(F\) in Eq~\ref{eq-phys-grav-prop}
+match the observed force of attraction between masses. In other words,
+Eq~\ref{eq-phys-grav-prop} needs to be \emph{constrained} using
+empirical data. Ultimately, the equation for the force is as seen in
+Eq~\ref{eq-phys-grav}, where \(G\) is an empirically calibrated
+proportionality constant:
 
@@
 assumption that received support after Hutton's experiments of 1778
-(Hutton 1778, 783). Still, it took over a century from the publication
-of \emph{Principia} to refine the estimate of the proportionality
-constant to within 1\% accuracy, with Cavendish's 1798 experiment
-(Cavendish 1798).
+\citep[p.~783]{hutton_xxxiii_1778}. Still, it took over a century from
+the publication of \emph{Principia} to refine the estimate of the
+proportionality constant to within 1\% accuracy, with Cavendish's 1798
+experiment \citep{cavendish_xxi_1798}.
 
@@
 Ravenstein proposed some ``Laws of Migration'' based on his empirical
-analysis of migration flows in various countries (Ravenstein 1885,
-1889). In these works, Ravenstein posited 1) a directly proportional
-relationship between migration flows and the size of destinations (i.e.,
-centres of commerce and industry), and 2) an inversely proportional
-relationship between the size of flows and the separation between
-origins and destinations. As with Carey, these propositions echo
-Newton's gravitational laws. Over time, other researchers discovered
-similar relationships. For example, Reilly (1929) formulated a law of
-retail gravitation, expressed in terms of equal attraction to competing
-retail destinations that could be understood as `potential trade
-territories'. Later, Zipf proposed a \(\frac{P_1P_2}{D}\) hypothesis for
-the case of information (Zipf 1946a), intercity personal movement (Zipf
-1946b), and goods movement by railways (Zipf 1946c). The
+analysis of migration flows in various countries
+\citep{ravensteinLawsMigration1885, ravensteinLawsMigration1889}. In
+these works, Ravenstein posited 1) a directly proportional relationship
+between migration flows and the size of destinations (i.e., centres of
+commerce and industry), and 2) an inversely proportional relationship
+between the size of flows and the separation between origins and
+destinations. As with Carey, these propositions echo Newton's
+gravitational laws. Over time, other researchers discovered similar
+relationships. For example, Reilly \citep{reilly1929methods} formulated
+a law of retail gravitation, expressed in terms of equal attraction to
+competing retail destinations that could be understood as `potential
+trade territories'. Later, Zipf proposed a \(\frac{P_1P_2}{D}\)
+hypothesis for the case of information
+\citep{zipfDeterminantsCirculationInformation1946}, intercity personal
+movement \citep{zipfHypothesisIntercityMovement1946}, and goods movement
+by railways \citep{zipfHypothesisCaseRailway1946}. The
 \(\frac{P_1P_2}{D}\) hypothesis stated that the magnitude of flows was
@@
 interaction is that a proportionality constant similar to \(G\) in
-Equation~\ref{eq-phys-grav} was never considered. Of the researchers
-cited above, only Reilly and Zipf expressed their hypotheses in
-mathematical terms. Reilly's hypothesis was presented in the following
-form:
+Eq~\ref{eq-phys-grav} was never considered. Of the researchers cited
+above, only Reilly and Zipf expressed their hypotheses in mathematical
+terms. Reilly's hypothesis was presented in the following form:
 
@@
 gravitation that we find the strongest connection yet to Newton's law
-(Stewart 1948). This may relate to academic backgrounds; where Stewart
-was a physicist while Ravenstein, Reilly, and Zipf were social
-scientists. Besides awareness of preceding research (he cites both
-Reilly and Zipf as predecessors in the analysis of human spatial
-interaction), Stewart appears to have been the first author to express
-his theorized relationships for human spatial interaction with a
-proportionality constant \(G\), as follows:
+\citep{stewartDemographicGravitationEvidence1948}. This may relate to
+academic backgrounds; where Stewart was a physicist while Ravenstein,
+Reilly, and Zipf were social scientists. Besides awareness of preceding
+research (he cites both Reilly and Zipf as predecessors in the analysis
+of human spatial interaction), Stewart appears to have been the first
+author to express his theorized relationships for human spatial
+interaction with a proportionality constant \(G\), as follows:
 
@@
   And finally \(G\), a constant that Stewart ``left for future
-  determination'' (1948, 34)
+  determination''
+  \citep[p.~34]{stewartDemographicGravitationEvidence1948}
 \end{itemize}
@@
 area \(ds\) and \(r\) is the distance to \(i\). In
-Equation~\ref{eq-stewart-population-potential-integral}, \(D ds\) gives
-an infinitesimal count of the population, say \(dm\), and so, after
-discretizing space,
-Equation~\ref{eq-stewart-population-potential-integral} can be rewritten
-as:
+Eq~\ref{eq-stewart-population-potential-integral}, \(D ds\) gives an
+infinitesimal count of the population, say \(dm\), and so, after
+discretizing space, Eq~\ref{eq-stewart-population-potential-integral}
+can be rewritten as:
 
@@
 Alerted readers will notice that
-Equation~\ref{eq-stewart-population-potential-sum}, with some
-re-organization of terms, is formally equivalent to our modern
-definition of accessibility popularized by Hansen (1959) in the late
+Eq~\ref{eq-stewart-population-potential-sum}, with some re-organization
+of terms, is formally equivalent to our modern definition of
+accessibility popularized by Hansen \citep{hansen1959} in the late
 1950s.
@@
 Stewart's formulation of demographic force, developed in the context of
-what he called ``social physics'' (Stewart 1947), was problematic. It
-had issues with inconsistent mathematical notation. More seriously
-though, Stewart's work was permeated by a view of humans as particles
-following physical laws, but tinted by unscientific and racist ideas.
-For instance, he assumed that the molecular weight \(\mu\) of the
+what he called ``social physics'' \citep{stewartPrinciples1947}, was
+problematic. It had issues with inconsistent mathematical notation. More
+seriously though, Stewart's work was permeated by a view of humans as
+particles following physical laws, but tinted by unscientific and racist
+ideas. For instance, he assumed that the molecular weight \(\mu\) of the
 average American was one, but ``presumably\ldots much less than
@@
 
-From Stewart (1948), we arrive to 1959 and Walter G. Hansen, whose work
-proved to be exceptionally influential in the accessibility literature
-(Hansen 1959). In this seminal paper, Hansen defined accessibility as
-``the potential of opportunities for interaction\ldots{} a
+From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
+arrive to 1959 and Walter G. Hansen, whose work proved to be
+exceptionally influential in the accessibility literature
+\citep{hansen1959}. In this seminal paper, Hansen defined accessibility
+as ``the potential of opportunities for interaction\ldots{} a
 generalization of the population-over-distance relationship or
-\emph{population potential} concept developed by Stewart (1948)''
-(p.~73). As well as being a student of city and regional planning at the
-Massachusetts Institute of Technology, Hansen was also an engineer with
-the Bureau of Roads, and preoccupied with the power of transportation to
-shape land uses in a very practical sense. Hansen (1959) focused on
-Stewart (1948)'s \emph{population potential} (expressed in
-Equation~\ref{eq-stewart-population-potential-sum}), and not on the
-other formulaic contributions and objectionable aspects of ``social
-physics''. Hansen (1959) recast Stewart's population potential to
+\emph{population potential} concept developed by Stewart
+\citep{stewartDemographicGravitationEvidence1948}'' (p.~73). As well as
+being a student of city and regional planning at the Massachusetts
+Institute of Technology, Hansen was also an engineer with the Bureau of
+Roads, and preoccupied with the power of transportation to shape land
+uses in a very practical sense. Hansen \citep{hansen1959} focused on
+Stewart's \citep{stewartDemographicGravitationEvidence1948}
+\emph{population potential} (expressed in
+Eq~\ref{eq-stewart-population-potential-sum}), and not on the other
+formulaic contributions and objectionable aspects of ``social physics''.
+Hansen \citep{hansen1959} recast Stewart's population potential to
 reflect accessibility, a model of human behaviour useful to capture
-regularities in mobility patterns. Hansen (1959) replaced \(M_j\) in
-Equation~\ref{eq-stewart-population-potential-sum} with
+regularities in mobility patterns. Hansen \citep{hansen1959} replaced
+\(M_j\) in Eq~\ref{eq-stewart-population-potential-sum} with
 \emph{opportunities} to derive an \emph{opportunity potential}, or more
@@
 
-A contemporary rewriting of Equation~\ref{eq-accessibility} accounts for
-a variety of impedance functions beyond the inverse power
-\(d^{-\beta}\):
+A contemporary rewriting of Eq~\ref{eq-accessibility} accounts for a
+variety of impedance functions beyond the inverse power \(d^{-\beta}\):
 
@@
 
-\(S_{i}\) in Equation~\ref{eq-accessibility} is a measure of the
-accessibility from site \(i\). This is a function of \(O_j\) (the mass
-of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
-travel time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a
-parameter that modulates the friction of cost). Today, Hansen is
-frequently cited as the father of modern accessibility analysis (e.g.,
-Reggiani and Martín 2011), and Hansen-type accessibility is commonly
-referred to as the gravity-based accessibility measure.
+\(S_{i}\) in Eq~\ref{eq-accessibility} is a measure of the accessibility
+from site \(i\). This is a function of \(O_j\) (the mass of
+opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or travel
+time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a parameter
+that modulates the friction of cost). Today, Hansen is frequently cited
+as the father of modern accessibility analysis
+\citep[e.g.,][]{reggianiGuestEditorialNew2011}, and Hansen-type
+accessibility is commonly referred to as the gravity-based accessibility
+measure.
 
@@
 included one crucial omission that afflicts the literature to this day.
-The omission is that between Stewart (1948) and Hansen (1959), the
-proportionality constant \(G\) in
-Equation~\ref{eq-stewart-population-potential} vanished. This constant
-was not explicitly addressed in Hansen (1959) and accessibility research
-continues to evolve without it. Since Hansen (1959), accessibility
-analysis has been widely used in numerous disciplines but, to our
-knowledge, the proportionality constant has remained forgotten, with no
-notable developments to explicitly acknowledge or determine it.
+The omission is that between Stewart
+\citep{stewartDemographicGravitationEvidence1948} and Hansen
+\citep{hansen1959}, the proportionality constant \(G\) in
+Eq~\ref{eq-stewart-population-potential} vanished. This constant was not
+explicitly addressed in Hansen \citep{hansen1959} and accessibility
+research continues to evolve without it. Since Hansen
+\citep{hansen1959}, accessibility analysis has been widely used in
+numerous disciplines but, to our knowledge, the proportionality constant
+has remained forgotten, with no notable developments to explicitly
+acknowledge or determine it.
 
@@
 interpretation, communication and comparability of accessibility
-analysis. Those reading Hansen (1959) must recall that Stewart (1948)
-had set the proportionality constant \(G\) to 1, with a note that
-``\(G\) {[}was{]} left for future determination: a suitable choice of
-other units can reduce it to unity'' {[}p.~34{]}. In practice, the
-persistent omission of the constant in accessibility analyses means that
-\(G\) continues to be implicitly set to \(1\), even when the fundamental
-relationship in accessibility is proportionality (e.g.,
-\(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality (for
+analysis. Those reading Hansen \citep{hansen1959} must recall that
+Stewart \citep{stewartDemographicGravitationEvidence1948} had set the
+proportionality constant \(G\) to 1, with a note that ``\(G\) {[}was{]}
+left for future determination: a suitable choice of other units can
+reduce it to unity'' {[}p.~34{]}. In practice, the persistent omission
+of the constant in accessibility analyses means that \(G\) continues to
+be implicitly set to \(1\), even when the fundamental relationship in
+accessibility is proportionality (e.g.,
+\(S_{i} \propto \sum_j g(O_j)f(d_{ij})\)) and not equality {[}for
 instance, see the formula for accessibility at the top of Figure 1 in Wu
-and Levinson 2020). The direct consequence is that without a
-proportionality constant, the units of \(S_i\) remain unclear: the unit
-of ``potential of opportunity for interaction'' is left free to change
-as \(\beta\) is calibrated. For example, if \(c_{ij}\) is distance in
-meters, it will be number of opportunities per \(m^{\beta}\) when
-\(f(c_{ij}) = d^{-\beta}\) but number of opportunities per
-\(e^{-\beta m}\) when \(f(c_{ij}) = e^{-\beta m}\). This undermines the
-comparability of accessibility metrics with different decay functions,
-and renders their results difficult to understand and communicate. The
-Hansen-style accessibility estimates found in the literature, therefore,
-are better thought of as ordinal measures of potential that can only be
-interpreted in terms of higher and lower accessibility, but which has
-not palpable meaning (E. J. Miller 2018).
+and Levinson \citep{wuUnifyingAccess2020}{]}. The direct consequence is
+that without a proportionality constant, the units of \(S_i\) remain
+unclear: the unit of ``potential of opportunity for interaction'' is
+left free to change as \(\beta\) is calibrated. For example, if
+\(c_{ij}\) is distance in meters, it will be number of opportunities per
+\(m^{\beta}\) when \(f(c_{ij}) = d^{-\beta}\) but number of
+opportunities per \(e^{-\beta m}\) when \(f(c_{ij}) = e^{-\beta m}\).
+This undermines the comparability of accessibility metrics with
+different decay functions, and renders their results difficult to
+understand and communicate. The Hansen-style accessibility estimates
+found in the literature, therefore, are better thought of as ordinal
+measures of potential that can only be interpreted in terms of higher
+and lower accessibility, but which has not palpable meaning
+\citep{millerAccessibilityMeasurementApplication2018}.
 
@@
 On the other side of the Atlantic, Alan G. Wilson was developing
-related, yet parallel work. In his groundbreaking study (Wilson 1971),
-Wilson defined a general spatial interaction model as follows. While
-accessibility was characterised as an associated concept of `potential',
-the primary focus was on modelling observed spatial interaction:
+related, yet parallel work. In his groundbreaking study
+\citep{wilson1971}, Wilson defined a general spatial interaction model
+as follows. While accessibility was characterised as an associated
+concept of `potential', the primary focus was on modelling observed
+spatial interaction:
 \begin{equation}\phantomsection\label{eq-phys-gravity-model}{
@@
 
-The model in Equation~\ref{eq-phys-gravity-model} posits a quantity
-\(T_{ij}\) that represents a value in a matrix of flows of size
-\(n \times m\), that is, between \(i = 1,s, n\) origins and
-\(j = 1,s, m\) destinations. The quantities \(W_i^{(1)}\) and
-\(W_j^{(2)}\) are proxies for the masses at \(i=1,s,n\) origins and
-\(j=1,s,m\) destinations. The super-indices \((1)\) and \((2)\) are
-meant to indicate that these masses could be different things, i.e.,
-\(W_i^{(1)}\) could be populations, and \(W_j^{(2)}\) hectares of park
-space. Finally, \(f(c_{ij})\) is some function of travel cost \(c_{ij}\)
-which reflects travel impedance. In this way, \(T_{ij}\) explicitly
-measures \emph{interaction} in the unit of trips, and the role of \(k\)
-is to ensure that the system-wide sum of \(T_{ij}\) represents the total
-flows in the data. In other words, \(k\) is a scale parameter that makes
-the overall amount of flows identical to the magnitude of the phenomenon
+The model in Eq~\ref{eq-phys-gravity-model} posits a quantity \(T_{ij}\)
+that represents a value in a matrix of flows of size \(n \times m\),
+that is, between \(i = 1,s, n\) origins and \(j = 1,s, m\) destinations.
+The quantities \(W_i^{(1)}\) and \(W_j^{(2)}\) are proxies for the
+masses at \(i=1,s,n\) origins and \(j=1,s,m\) destinations. The
+super-indices \((1)\) and \((2)\) are meant to indicate that these
+masses could be different things, i.e., \(W_i^{(1)}\) could be
+populations, and \(W_j^{(2)}\) hectares of park space. Finally,
+\(f(c_{ij})\) is some function of travel cost \(c_{ij}\) which reflects
+travel impedance. In this way, \(T_{ij}\) explicitly measures
+\emph{interaction} in the unit of trips, and the role of \(k\) is to
+ensure that the system-wide sum of \(T_{ij}\) represents the total flows
+in the data. In other words, \(k\) is a scale parameter that makes the
+overall amount of flows identical to the magnitude of the phenomenon
 being modeled. In other words, it balances the units, in a conceptually
@@
 Traditionally, the development of the spatial interaction model put an
-emphasis on the interpretability of the results (Kirby 1970; Wilson
-1967, 1971). But instead of relying on the heuristic of Newtonian
-gravity (e.g., some interaction between a mass at \(i\) and a mass at
-\(j\) separated by some distance), Wilson's approach was to maximise the
-entropy of the system. Entropy maximisation in this case achieves stable
-results as a statistical average that represents the population. The
-approach works by assuming undifferentiated individual interactions, and
-assessing their probabilities of making a particular journey. The result
-of Equation~\ref{eq-phys-gravity-model} then is a statistical average
-(Wilson 1971; Senior 1979).
-
-To ensure that \(T_{ij}\) in Equation~\ref{eq-phys-gravity-model} is in
-the unit of trips (our unit of origin-destination spatial interaction),
+emphasis on the interpretability of the results
+\citep{kirbyNormalizingFactorsGravity1970, wilsonSTATISTICALTHEORYSPATIAL1967, wilson1971}.
+But instead of relying on the heuristic of Newtonian gravity (e.g., some
+interaction between a mass at \(i\) and a mass at \(j\) separated by
+some distance), Wilson's approach was to maximise the entropy of the
+system. Entropy maximisation in this case achieves stable results as a
+statistical average that represents the population. The approach works
+by assuming undifferentiated individual interactions, and assessing
+their probabilities of making a particular journey. The result of
+Eq~\ref{eq-phys-gravity-model} then is a statistical average
+\citep{wilson1971, seniorGravityModellingEntropy1979}.
+
+To ensure that \(T_{ij}\) in Eq~\ref{eq-phys-gravity-model} is in the
+unit of trips (our unit of origin-destination spatial interaction),
 additional knowledge about the system is required. At the very least,
@@
 partially or not at all, a family of spatial interaction models can be
-derived based on Equation~\ref{eq-phys-gravity-model}. \(K\) is
-specified depending on the applied constraint(s). In the framework
-introduced in Wilson (1971), three constrained versions are specified:
-the first being a case where to results only match the total volume of
+derived based on Eq~\ref{eq-phys-gravity-model}. \(K\) is specified
+depending on the applied constraint(s). In the framework introduced in
+Wilson \citep{wilson1971}, three constrained versions are specified: the
+first being a case where to results only match the total volume of
 interaction, the second being a singly constrained case, and the third a
@@
 
-In the first, Equation~\ref{eq-constraint1-gravitymodel} and
-Equation~\ref{eq-constraint2-gravitymodel} do not hold. In practical
-terms, this means that the total number of trips predicted by the model
-must be equal to sum of all flows from origins i to destinations j. The
-balancing constant \(K\) in this case is (see Cliff, Martin, and Ord
-1974; A. S. Fotheringham 1984):
+In the first, Eq~\ref{eq-constraint1-gravitymodel} and
+Eq~\ref{eq-constraint2-gravitymodel} do not hold. In practical terms,
+this means that the total number of trips predicted by the model must be
+equal to sum of all flows from origins i to destinations j. The
+balancing constant \(K\) in this case is
+\citep[see][]{cliff_evaluating_1974, fotheringham_spatial_1984}:
 
@@
 
-In the second case, only one of
-Equation~\ref{eq-constraint1-gravitymodel} or
-Equation~\ref{eq-constraint2-gravitymodel} hold. The resulting models
-are, in Wilson's terms, singly constrained. When only
-Equation~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation
-leads to the following production-constrained model:
+In the second case, only one of Eq~\ref{eq-constraint1-gravitymodel} or
+Eq~\ref{eq-constraint2-gravitymodel} hold. The resulting models are, in
+Wilson's terms, singly constrained. When only
+Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
+to the following production-constrained model:
 
@@
 (i.e., balancing factors) specific to origins. For this model, the
-balancing factors ensure that Equation~\ref{eq-constraint1-gravitymodel}
-is satisfied, meaning that the sum of predicted flows from one origin
-going to all destinations must equal the known mass at that origin
-\(O_i\) i.e., the total number of outbound trips. Satisfying this
-constraint also implicitly fulfills the total constraint
-(Equation~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\)
-values across all origins equals the total number of trips. This model
-is useful when trips ends are unknown but the number of trips
-originating from each location is known and the total of these trips
-represents all trips in the system. The balancing factors for the
+balancing factors ensure that Eq~\ref{eq-constraint1-gravitymodel} is
+satisfied, meaning that the sum of predicted flows from one origin going
+to all destinations must equal the known mass at that origin \(O_i\)
+i.e., the total number of outbound trips. Satisfying this constraint
+also implicitly fulfills the total constraint
+(Eq~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\) values
+across all origins equals the total number of trips. This model is
+useful when trips ends are unknown but the number of trips originating
+from each location is known and the total of these trips represents all
+trips in the system. The balancing factors for the
 production-constrained model are solved for each origin \(A_i\), and
@@
 
-In this model (Equation~\ref{eq-attraction-constrained-gravitymodel}),
-the balancing factors ensure that
-Equation~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
-constraint Equation~\ref{eq-constraint0-gravitymodel} is as well),
-meaning that the sum of predicted flows going to one destination from
-all origins must equal the known mass of that destination \(D_j\) i.e.,
-the total number of inbound trips to \(j\). This should hold for all
-destinations. As before, destination-specific proportionality constants
-(i.e., balancing factors) \(B_j\) were derived by Wilson as:
+In this model (Eq~\ref{eq-attraction-constrained-gravitymodel}), the
+balancing factors ensure that Eq~\ref{eq-constraint2-gravitymodel} is
+satisfied (hence the total constraint
+Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning that the sum
+of predicted flows going to one destination from all origins must equal
+the known mass of that destination \(D_j\) i.e., the total number of
+inbound trips to \(j\). This should hold for all destinations. As
+before, destination-specific proportionality constants (i.e., balancing
+factors) \(B_j\) were derived by Wilson as:
 
@@
 production-attraction constrained model. In this case, both
-Equation~\ref{eq-constraint1-gravitymodel} and
-Equation~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
+Eq~\ref{eq-constraint1-gravitymodel} and
+Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
 constraints ensure that the sum of predicted flows from one origin to
@@
 
-Derivation of these models is demonstrated in detail elsewhere (e.g.,
-Ortúzar and Willumsen 2011; Wilson 1967). It is worth noting, however,
-that although Wilson's approach is built on a different conceptual
-foundation than the old reference to Newtonian gravity, the work
-succeeded at identifying the steps from proportionality to equality to
-yield variations of proportionality constants, including the one that
-eluded Stewart (1948) and that has been overlooked in almost all
-subsequent accessibility research. Why was this key element of spatial
-interaction models potentially ignored in accessibility research? In the
-next section we aim to address this question.
+Derivation of these models is demonstrated in detail elsewhere
+\citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.
+It is worth noting, however, that although Wilson's approach is built on
+a different conceptual foundation than the old reference to Newtonian
+gravity, the work succeeded at identifying the steps from
+proportionality to equality to yield variations of proportionality
+constants, including the one that eluded Stewart
+\citep{stewartDemographicGravitationEvidence1948} and that has been
+overlooked in almost all subsequent accessibility research. Why was this
+key element of spatial interaction models potentially ignored in
+accessibility research? In the next section we aim to address this
+question.
 
@@
 
-The work of Hansen (1959) and Wilson (1971) responded to important
-developments, in particular a need ``to meet the dictates and needs of
-public policy for strategic land use and transportation planning''
-(Michael Batty 1994). These dictates and needs were far from trivial. In
-the United States alone, the Federal-Aid Highway Act of 1956 authorized
-the creation of the U.S. Interstate Highway System, with a budget that
-ultimately exceeded one hundred billion dollars (equivalent to over
-\$600 billion in 2023) (Weiner 2016; MDOT 2007). Spatial interaction
-modelling was incorporated into institutional modelling practices meant
-to ``predict and provide'', i.e., predict travel demand and supply
-transportation infrastructure (Kovatch, Zames, et al. 1971; Weiner
-2016). Accessibility, at the time, did not quite have that power, as it
-did not quantify trips, but rather something somewhat more elusive: it
-predicts the less tangible ``potential'' for spatial interaction with
+The work of Hansen \citep{hansen1959} and Wilson \citep{wilson1971}
+responded to important developments, in particular a need ``to meet the
+dictates and needs of public policy for strategic land use and
+transportation planning'' \citep{battyChronicleScientificPlanning1994}.
+These dictates and needs were far from trivial. In the United States
+alone, the Federal-Aid Highway Act of 1956 authorized the creation of
+the U.S. Interstate Highway System, with a budget that ultimately
+exceeded one hundred billion dollars (equivalent to over \$600 billion
+in 2023)
+\citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}.
+Spatial interaction modelling was incorporated into institutional
+modelling practices meant to ``predict and provide'', i.e., predict
+travel demand and supply transportation infrastructure
+\citep{kovatch1971modeling, weinerUrbanTransportationPlanning2016}.
+Accessibility, at the time, did not quite have that power, as it did not
+quantify trips, but rather something somewhat more elusive: it predicts
+the less tangible ``potential'' for spatial interaction with
 opportunities. In this way, where spatial interaction modelling became a
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{manuscript_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 To illustrate this point, we conducted a bibliographic analysis of the
-literature that cites Hansen (1959), Wilson (1971), or both. We
-retrieved all relevant documents using the Web of Science ``Cited
-References'' functionality, and the digital object identifiers of Hansen
-(1959) and Wilson (1971). As a result of this search, we identified
-1,788 documents that cite Hansen (1959), only 258 documents that cite
-Wilson (1971), and 76 that cite both. The earliest document in this
-corpus dates to 1976 and the most recent is from 2025. The number of
-documents per year appears in Figure~\ref{fig-docs-per-year}, where we
-see the frequency of documents over a span of almost fifty years. In
-particular, we notice the remarkable growth in the number of papers that
-cite Hansen (1959) compared to those that cite Wilson (1971) since the
-year 2000.
-
-As noted, literature that cites both Wilson (1971) and Hansen (1959) are
-sparse (only 3.6\% of the corpus, visualised in pink in
-Figure~\ref{fig-docs-per-year}). After reading the works, we can also
-discern that they too are divergent, with one stream focused on
-developing accessibility (i.e., \emph{potential for} spatial
+literature that cites Hansen \citep{hansen1959}, Wilson
+\citep{wilson1971}, or both. We retrieved all relevant documents using
+the Web of Science ``Cited References'' functionality, and the digital
+object identifiers of Hansen \citep{hansen1959} and Wilson
+\citep{wilson1971}. As a result of this search, we identified 1,788
+documents that cite Hansen \citep{hansen1959}, only 258 documents that
+cite Wilson \citep{wilson1971}, and 76 that cite both. The earliest
+document in this corpus dates to 1976 and the most recent is from 2025.
+The number of documents per year appears in Fig~\ref{fig-docs-per-year},
+where we see the frequency of documents over a span of almost fifty
+years. In particular, we notice the remarkable growth in the number of
+papers that cite Hansen \citep{hansen1959} compared to those that cite
+Wilson \citep{wilson1971} since the year 2000.
+
+As noted, literature that cites both Wilson \citep{wilson1971} and
+Hansen \citep{hansen1959} are sparse (only 3.6\% of the corpus,
+visualised in pink in Fig~\ref{fig-docs-per-year}). After reading the
+works, we can also discern that they too are divergent, with one stream
+focused on developing accessibility (i.e., \emph{potential for} spatial
 interaction) and another on spatial interaction. The focuses of these
@@
 On one hand, the stream of literature focused on spatial interaction
-models inspired by Wilson (1971) and which cite both Hansen (1959) and
-Wilson (1971), tends to contribute to understanding how accessibility is
-interpreted and incorporated in spatial interaction models. These works
-treat separate spatial interaction and accessibility as a seperate but
-related phenomenon. Specifically, some early works interpret the spatial
-interaction model's balancing factors
-(Equation~\ref{eq-production-constrained-balancing-factor} or
-Equation~\ref{eq-doubly-constrained-balancing-factors}) as the inverse
-of Hansen (1959)`s model (B. Harris and Wilson 1978; G. Leonardi 1978;
-A. Stewart Fotheringham 1981; A. S. Fotheringham 1985), recognizing it
-as a ``common sense'' approach (Morris, Dumble, and Wigan 1979, 99) to
-including accessibility in the spatial interaction model, though further
-exploration of its relationship is warranted (M. Batty and March 1976).
-Some authors have explored this relationship, for instance as in A. S.
-Fotheringham (1985) who demonstrates how the spatial interaction model
-may insufficiently explain spatial patterns, and suggests that
-explicitly defining destinations' accessibility as a variable within the
-model may remedy the issue (e.g., the \emph{competition destination}
-model). Other works used both Hansen (1959) and Wilson (1971)'s
-framework in conjunction, such as in defining location-allocation
-problems in operations research (G. Leonardi 1978; Beaumont 1981),
+models inspired by Wilson \citep{wilson1971} and which cite both Hansen
+\citep{hansen1959} and Wilson \citep{wilson1971}, tends to contribute to
+understanding how accessibility is interpreted and incorporated in
+spatial interaction models. These works treat separate spatial
+interaction and accessibility as a seperate but related phenomenon.
+Specifically, some early works interpret the spatial interaction model's
+balancing factors (Eq~\ref{eq-production-constrained-balancing-factor}
+or Eq~\ref{eq-doubly-constrained-balancing-factors}) as the inverse of
+Hansen's \citep{hansen1959} model
+\citep{harrisEquilibriumValuesDynamics1978, leonardiOptimumFacilityLocation1978, fotheringhamSPATIALSTRUCTUREDISTANCE1981, fotheringhamSpatialCompetitionAgglomeration1985},
+recognizing it as a ``common sense'' approach
+\citep[p.~99]{morrisAccessibilityIndicatorsTransport1979} to including
+accessibility in the spatial interaction model, though further
+exploration of its relationship is warranted
+\citep{battyMethodResiduesUrban1976}. Some authors have explored this
+relationship, for instance as in Fotheringham
+\citep{fotheringhamSpatialCompetitionAgglomeration1985} who demonstrates
+how the spatial interaction model may insufficiently explain spatial
+patterns, and suggests that explicitly defining destinations'
+accessibility as a variable within the model may remedy the issue (e.g.,
+the \emph{competition destination} model). Other works used both Hansen
+\citep{hansen1959} and Wilson's \citep{wilson1971} framework in
+conjunction, such as in defining location-allocation problems in
+operations research
+\citep{leonardiOptimumFacilityLocation1978, beaumontLocationallocationProblemsPlane1981},
 estimating trip flows (or some other spatial interaction flows)
-alongside accessibility (e.g., Clarke, Eyre, and Guy 2002; Grengs 2004;
-Türk 2019), or considering accessibility within spatial interaction
-models, in line with A. S. Fotheringham (1985)'s demonstration (e.g.,
-Beckers et al. 2022). Other works departed from Hansen (1959)'s
-definition and aligned with spatial interaction in different ways, such
-as using micro-economic consumer behaviour concepts to express potential
-for spatial interaction (Morris, Dumble, and Wigan 1979; Giorgio
-Leonardi and Tadei 1984).
+alongside accessibility
+\citep[e.g.,][]{clarke2002deriving, grengs2004measuring, turk2019socio},
+or considering accessibility within spatial interaction models, in line
+with Fotheringham's
+\citep{fotheringhamSpatialCompetitionAgglomeration1985} demonstration
+\citep[e.g.,][]{beckers2022incorporating}. Other works departed from
+Hansen's \citep{hansen1959} definition and aligned with spatial
+interaction in different ways, such as using micro-economic consumer
+behaviour concepts to express potential for spatial interaction
+\citep{morrisAccessibilityIndicatorsTransport1979, leonardiRandomUtilityDemand1984}.
 
 On the other hand, there is another subset of literature that cite both
-Hansen (1959) and Wilson (1971) that is accessibility-focused. We
-categorise their citation of Wilson (1971) for three general reasons.
-Firstly, a group of these works cite Wilson (1971) as attribution for
-using context-dependent travel cost functions (e.g., Weibull 1980; S. L.
-Handy and Niemeier 1997; Kwan 1998; Q. Shen 1998a; Ashiru, Polak, and
-Noland 2003; Rau and Vega 2012; Pan 2013; Margarida Condeço Melhorado et
-al. 2016; Caschili, De Montis, and Trogu 2015; Grengs 2015; Pan, Jin,
-and Liu 2020; Chia and Lee 2020; Roblot et al. 2021; Sharifiasl, Kharel,
-and Pan 2023; Kharel, Sharifiasl, and Pan 2024). These works do not
-necessarily comment on spatial interaction explicitly. Secondly, another
-group of this accessibility-focused ``both'' citing literature
-\emph{does} associate spatial interaction as defined in Wilson (1971)
-with accessibility's potential for spatial interaction more explicitly
-(e.g., H. J. Miller 1999; Giuliano et al. 2010; Grengs et al. 2010;
-Grengs 2010, 2012; Levine et al. 2012; Levinson and Huang 2012; Tong,
-Zhou, and Miller 2015; X. Liu and Zhou 2015; He et al. 2017; Wu and
-Levinson 2020; Ng et al. 2022; Naqavi et al. 2023; Suel et al. 2024). We
-agree accessibility and spatial interaction are related topics:
+Hansen \citep{hansen1959} and Wilson \citep{wilson1971} that is
+accessibility-focused. We categorise their citation of Wilson
+\citep{wilson1971} for three general reasons. Firstly, a group of these
+works cite Wilson \citep{wilson1971} as attribution for using
+context-dependent travel cost functions
+\citep[e.g.,][]{weibullNumericalMeasurementAccessibility1980, handyMeasuringAccessibilityExploration1997, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
+These works do not necessarily comment on spatial interaction
+explicitly. Secondly, another group of this accessibility-focused
+``both'' citing literature \emph{does} associate spatial interaction as
+defined in Wilson \citep{wilson1971} with accessibility's potential for
+spatial interaction more explicitly
+\citep[e.g.,][]{millerMeasuringSpaceTimeAccessibility1999, giuliano2010accessibility, grengs2010intermetropolitan, grengs2010job, grengs2012equity, levine2012does, levinson2012positive, tong2015transportation, liu2015spatial, he2017measuring, wuUnifyingAccess2020, ng2022reflection, naqavi2023mobility, suel2024measuring}.
+We agree accessibility and spatial interaction are related topics:
 accessibility is an expression of its \emph{potential} and the Wilson
-(1971) paper briefly touches on the concept. However, in some of this
-literature, Hansen (1959) and Wilson (1971) are co-cited as both being
-`gravity models' (e.g., S. Liu and Zhu 2004; Dai, Wan, and Gai 2017; Y.
-Shen 2019; Chia and Lee 2020), perhaps revealing the murkiness of the
-distinction between spatial interaction and the \emph{potential for}
-spatial interaction in the literature. Thirdly, there is a group of
-accessibility-focused works that interpret Hansen (1959)'s model as the
-singly- or doubly- constrained spatial interaction model's inverse
-balancing factor (e.g., R. W. Vickerman 1974). This group cities the
-earlier spatial interaction works that make this connection and is
-especially prominent in the investigation of competitive accessibility
-topics (e.g., Karst and Van Eck 2003; Karst T. Geurs, van Wee, and
-Rietveld 2006; Willigers, Floor, and van Wee 2007; El-Geneidy and
-Levinson 2011; Curtis and Scheurer 2010; Manaugh and El-Geneidy 2012;
-Chen and Silva 2013; Alonso et al. 2014; Albacete et al. 2017;
-Sahebgharani, Mohammadi, and Haghshenas 2019; Mayaud et al. 2019; Allen
-and Farber 2020; Levinson and Wu 2020; Marwal and Silva 2022; Su and
-Goulias 2023). As outlined in preceding sections, we argue interpreting
-the singly- or doubly- constrained spatial interaction model's balancing
-factor as accessibility yields output values that are similarly plagued
-by interpretability issues.
+\citep{wilson1971} paper briefly touches on the concept. However, in
+some of this literature, Hansen \citep{hansen1959} and Wilson
+\citep{wilson1971} are co-cited as both being `gravity models'
+\citep[e.g.,][]{liu2004accessibility, dai2017visualization, shen2019segregation, chia2020extending},
+perhaps revealing the murkiness of the distinction between spatial
+interaction and the \emph{potential for} spatial interaction in the
+literature. Thirdly, there is a group of accessibility-focused works
+that interpret Hansen's \citep{hansen1959} model as the singly- or
+doubly- constrained spatial interaction model's inverse balancing factor
+\citep[e.g.,][]{vickermanAccessibilityAttractionPotential1974}. This
+group cities the earlier spatial interaction works that make this
+connection and is especially prominent in the investigation of
+competitive accessibility topics
+\citep[e.g.,][]{karstEvaluationAccessibilityImpacts2003, geurs2006accessibility, willigers2007accessibility, el2011place, curtis2010planning, manaugh2012makes, chen2013regional, alonso2014labour, albacete2017measuring, sahebgharani2019computing, mayaud2019future, allenMeasureCompetitiveAccess2020, levinsonGeneralTheoryAccess2020, marwal2022literature, su2023untangling}.
+As outlined in preceding sections, we argue interpreting the singly- or
+doubly- constrained spatial interaction model's balancing factor as
+accessibility yields output values that are similarly plagued by
+interpretability issues.
 
 Lastly, as an extension of the third reason within this group, only the
-works of Soukhov et al. (2023) and Soukhov et al. (2024) use Wilson
-(1971)'s balancing factors as a method for maintaining constraints on
-opportunities within the context of competitive accessibility. These
-works introduce the balancing factors as a mechanisms to ensure that
-opportunities at each destination are proportionally allocated to each
-zone (based on the proportion of population seeking opportunities and
-the relative travel impedance). This is to ensure that each zonal
-accessibility value is the sum of this proportional allocation from each
-destination, and that all zonal values ultimately sum to the total
-number of opportunities in the region. However, these balancing factors
-were deduced intuitively. These works do not explicitly state that the
-mathematical formulation of the equations are effectively equivalent to
-Wilson's singly constrained model (derived from entropy maximization).
-This equivalence is only discovered in hindsight, as will be
-demonstrated in this study. These two works also do not discuss other
-constrained cases that will be addressed in the next section.
+works of Soukhov et al.
+\citep{soukhovIntroducingSpatialAvailability2023, soukhovMultimodalSpatialAvailability2024}
+use Wilson's \citep{wilson1971} balancing factors as a method for
+maintaining constraints on opportunities within the context of
+competitive accessibility. These works introduce the balancing factors
+as a mechanisms to ensure that opportunities at each destination are
+proportionally allocated to each zone (based on the proportion of
+population seeking opportunities and the relative travel impedance).
+This is to ensure that each zonal accessibility value is the sum of this
+proportional allocation from each destination, and that all zonal values
+ultimately sum to the total number of opportunities in the region.
+However, these balancing factors were deduced intuitively. These works
+do not explicitly state that the mathematical formulation of the
+equations are effectively equivalent to Wilson's singly constrained
+model (derived from entropy maximization). This equivalence is only
+discovered in hindsight, as will be demonstrated in this study. These
+two works also do not discuss other constrained cases that will be
+addressed in the next section.
 
@@
 accessibility and spatial interaction modelling have evolved as largely
-separate streams with little contact since Hansen (1959) and Wilson
-(1971). This may explain why the constraints and associated balancing
-factors of spatial interaction models did not cross over to
-accessibility analysis. This is intriguing since Wilson made an effort
-to connect developments in spatial interaction modelling to
+separate streams with little contact since Hansen \citep{hansen1959} and
+Wilson \citep{wilson1971}. This may explain why the constraints and
+associated balancing factors of spatial interaction models did not cross
+over to accessibility analysis. This is intriguing since Wilson made an
+effort to connect developments in spatial interaction modelling to
 accessibility, noting for instance, the denominator of the
 proportionality constants specific to origins
-(Equation~\ref{eq-production-constrained-balancing-factor}) is the
-inverse of balancing factor \(A_i\) (Wilson 1971, 10):
+(Eq~\ref{eq-production-constrained-balancing-factor}) is the inverse of
+balancing factor \(A_i\) \citep[ p.~10]{wilson1971}:
 
@@
 population potential of opportunities for spatial interaction (i.e.,
-Equation~\ref{eq-stewart-population-potential-sum} representing
+Eq~\ref{eq-stewart-population-potential-sum} representing
 \(V_i = \sum_j \frac{M_j}{d_{ij}}\) after losing \(G\)).
@@
 
-Similar to Equation~\ref{eq-phys-gravity-model}, \(W_j^{(2)}\) above is
-the mass at the destination and the sub-indices are for \(i = 1,s, n\)
+Similar to Eq~\ref{eq-phys-gravity-model}, \(W_j^{(2)}\) above is the
+mass at the destination and the sub-indices are for \(i = 1,s, n\)
 origins and \(j = 1,s, m\) destinations.
@@
 The market potential variant can also be generally defined, which is
-effectively the transpose of \(i\) and \(j\) in
-Equation~\ref{eq-access-01} and Equation~\ref{eq-accesssibility-01} as
-follows: \begin{equation}\phantomsection\label{eq-market-01}{
+effectively the transpose of \(i\) and \(j\) in Eq~\ref{eq-access-01}
+and Eq~\ref{eq-accesssibility-01} as follows:
+\begin{equation}\phantomsection\label{eq-market-01}{
 M_{ji} = k W_i^{(1)} f(c_{ij})
@@
 partial sums of \(V_i\) and \(M_j\), Figure
-Figure~\ref{fig-analytical-device-conc-accessibility} illustrates the
+Fig~\ref{fig-analytical-device-conc-accessibility} illustrates the
 accessibility analytical framework we propose using a simple 3 zone
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{../figures/access-analytical-device.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{../figures/access-analytical-device.png}
 
@@
   \textbf{The unconstrained case} with variants \(V_i^0\) and \(M_i^0\).
-  This case is equivalent to Hansen (1959)`s formulation (first variant)
-  and Reilly (1929)'s market potential formulation (second variant),
-  whereby a balancing factor is neglected, so units of zonal values are
-  in units of 'opportunities-by-travel-impedance-value' and
+  This case is equivalent to Hansen's \citep{hansen1959} formulation
+  (first variant) and Reilly's \citep{reilly1929methods} market
+  potential formulation (second variant), whereby a balancing factor is
+  neglected, so units of zonal values are in units of
+  `opportunities-by-travel-impedance-value' and
   `population-by-travel-impedance-value'. In this case, no constraints
@@
   \(M_j^T\). The first variant \(V_i^T\)--the \emph{total constrained
-  accessible opportunities measure}--resembles Hansen (1959)`s
-  formulation but with an additional regional balancing factor \(K^T\)
-  term, defined as the ratio of the total number of opportunities in the
-  region to the total sum of unconstrained accessibility values in the
-  region. In this way, \(K^T\) ensures that each zonal accessibility
-  value is a proportion of the total opportunities in the region (i.e.,
-  the total marginal in
-  Figure~\ref{fig-analytical-device-conc-accessibility}), requires no
+  accessible opportunities measure}--resembles Hansen's
+  \citep{hansen1959} formulation but with an additional regional
+  balancing factor \(K^T\) term, defined as the ratio of the total
+  number of opportunities in the region to the total sum of
+  unconstrained accessibility values in the region. In this way, \(K^T\)
+  ensures that each zonal accessibility value is a proportion of the
+  total opportunities in the region (i.e., the total marginal in
+  Fig~\ref{fig-analytical-device-conc-accessibility}), requires no
   information about the population seeking opportunities, and is in
-  units of 'opportunities' accessible. The second variant \(M_j^T\) is
+  units of `opportunities' accessible. The second variant \(M_j^T\) is
   the transpose of \(i\) to \(j\) of the first variant, effectively a
@@
   first variant is mathematically equivalent to the spatial availability
-  measure (Soukhov et al. 2023), and this variant's per capita form is
-  equivalent to the popular 2SFCA measure (Luo and Wang 2003; Q. Shen
-  1998b). Both variants can also be defined using either Hansen (1959)'s
-  or the market potential formulation but with balancing factor \(B_j\)
-  for \(V_i^S\) or \(A_i\) for \(M_j^S\). These balancing factors ensure
-  that the total marginal (green box in
-  Figure~\ref{fig-analytical-device-conc-accessibility}) is maintained
-  \emph{as well as} the values at one of the single marginals
+  measure \citep{soukhovIntroducingSpatialAvailability2023}, and this
+  variant's per capita form is equivalent to the popular 2SFCA measure
+  \citep{luo2003, shen1998}. Both variants can also be defined using
+  either Hansen's \citep{hansen1959} or the market potential formulation
+  but with balancing factor \(B_j\) for \(V_i^S\) or \(A_i\) for
+  \(M_j^S\). These balancing factors ensure that the total marginal
+  (green box in Fig~\ref{fig-analytical-device-conc-accessibility}) is
+  maintained \emph{as well as} the values at one of the single marginals
   (destination-mass marginal for opportunity-constrained and origin-mass
   marginal for population constrained in
-  Figure~\ref{fig-analytical-device-conc-accessibility}). In this way,
-  the singly constrained case reflects a medium amount of potential,
+  Fig~\ref{fig-analytical-device-conc-accessibility}). In this way, the
+  singly constrained case reflects a medium amount of potential,
   restricted by either values fitting the single destination-mass
@@
   double constraint maintains both marginals in
-  Figure~\ref{fig-analytical-device-conc-accessibility}), this case
+  Fig~\ref{fig-analytical-device-conc-accessibility}), this case
   restricts the \emph{potential} to spatially interact completely. In
@@
 
+\blandscape
+
 \input{summary-family-access-measures-table.tex}
@@
 
+\elandscape
+
 \subsection{Setup of the simple numeric
@@
 Consider a simple region as shown in
-Figure~\ref{fig-analytical-device-conc-accessibility}, with zone IDs 1,
-2 and 3. Each zone is both an origin \(i\) and a destination \(j\). The
+Fig~\ref{fig-analytical-device-conc-accessibility}, with zone IDs 1, 2
+and 3. Each zone is both an origin \(i\) and a destination \(j\). The
 following three pieces of information are defined: zonal population and
@@
 Provider-to-Population-Ratio (PPR) in this system is 24.5. The number of
-physicians per 10,000 in Canada in 2022 was 24.97 (WHO 2025). Secondly,
-to pair the zonal population and opportunity information, the assumed
-cost of movement (in minutes of travel time) between origins and
-destinations is as shown in Table~\ref{tbl-small-system-cost}.
+physicians per 10,000 in Canada in 2022 was 24.97
+\citep{whoMedicalDoctors102025}. Secondly, to pair the zonal population
+and opportunity information, the assumed cost of movement (in minutes of
+travel time) between origins and destinations is as shown in
+Table~\ref{tbl-small-system-cost}.
 
@@
 
-And lastly, in Equation~\ref{eq-travel-behaviour-scenarios} we
-distinguish accessibility measure values for the following three
-impedance functions that represent the potential for spatial interaction
-travel behaviour of the population to opportunities. Accessibility will
-be calculated three times for each case, one assuming the most decay
-(\(f_1(c_{ij})\)), another assuming medium decay (\(f_2(c_{ij})\)), and
-a third assuming the least decaying travel behaviour (\(f_3(c_{ij})\))
-for the entire region. A helpful analogy may be tying travel behaviour
-to the used mode's mobility potential, i.e., the most decaying travel
-behaviour (\(f_1(c_{ij})\)) would assume all travel in the region being
-done by foot, while calculating accessibility assuming the least decay
+And lastly, in Eq~\ref{eq-travel-behaviour-scenarios} we distinguish
+accessibility measure values for the following three impedance functions
+that represent the potential for spatial interaction travel behaviour of
+the population to opportunities. Accessibility will be calculated three
+times for each case, one assuming the most decay (\(f_1(c_{ij})\)),
+another assuming medium decay (\(f_2(c_{ij})\)), and a third assuming
+the least decaying travel behaviour (\(f_3(c_{ij})\)) for the entire
+region. A helpful analogy may be tying travel behaviour to the used
+mode's mobility potential, i.e., the most decaying travel behaviour
+(\(f_1(c_{ij})\)) would assume all travel in the region being done by
+foot, while calculating accessibility assuming the least decay
 (\(f_3(c_{ij})\)) would assume unfettered automobility. Or
@@
 on how travel behaviour has been considered in accessibility measures
-cost of travel see Paez, Scott, and Morency (2012).
+cost of travel see Páez et al. \citep{paez2012measuring}.
 
@@
 Setting the balancing factor \(k\) to 1 or omitting it completely in
-Equation~\ref{eq-access-01} results in the unconstrained accessibility
-case:
+Eq~\ref{eq-access-01} results in the unconstrained accessibility case:
 
@@
 In this case, the partial sum of spatial interaction is simply identical
-to Hansen's accessibility \(S_i\) (Hansen 1959), the current standard
-practice in accessibility measurement:
+to Hansen's accessibility \(S_i\) \citep{hansen1959}, the current
+standard practice in accessibility measurement:
 
@@
 
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 45.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}}
+\fontsize{6.0pt}{7.2pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
 \toprule
- & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 405.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{0}} \\ 
+ & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{0}} \\ 
 \cmidrule(lr){2-4}
@@
 
-Unlike in Equation~\ref{eq-access-01}, the proportionality constant
-\(k\) is retained. For the total constrained case, it is represented as
-\(K^T\):
+Unlike in Eq~\ref{eq-access-01}, the proportionality constant \(k\) is
+retained. For the total constrained case, it is represented as \(K^T\):
 
@@
 This constraint is analogous to the total constraint of
-Equation~\ref{eq-constraint0-gravitymodel}, congruent with Wilson's
-framework. Given the total number of opportunities in the region, we can
-then substitute Equation~\ref{eq-total-constrained-accessibility} in
-Equation~\ref{eq-total-constraint-accessibility} to solve for \(K^T\):
+Eq~\ref{eq-constraint0-gravitymodel}, congruent with Wilson's framework.
+Given the total number of opportunities in the region, we can then
+substitute Eq~\ref{eq-total-constrained-accessibility} in
+Eq~\ref{eq-total-constraint-accessibility} to solve for \(K^T\):
 
@@
 Which is also congruent with Wilson's framework as it comparable to the
-total flow spatial interaction model (e.g., Equation 2.11 in Cliff,
-Martin, and Ord (1974)). Hence, rearranging the equation to have
+total flow spatial interaction model (e.g., Equation 2.11 in
+\citep{cliff_evaluating_1974}). Hence, rearranging the equation to have
 opportunities and the proportional constant distinctly represented, our
@@
 
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 45.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 135.00pt -2\tabcolsep-1.5\arrayrulewidth}}
+\fontsize{6.0pt}{7.2pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
 \toprule
- & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 405.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{T}} \\ 
+ & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{T}} \\ 
 \cmidrule(lr){2-4}
@@
 Using the total opportunity constrained formulation of accessibility
-offers a solution to the unit interpretability issue of Hansen (1959)'s
-accessibility measure. Intuitively, the use of the constraint
-illustrates how the differences and ratios of values between zones and
-decay groups can be compared. This is true for other constrained cases
-of the family of accessibility measures.
+offers a solution to the unit interpretability issue of Hansen's
+\citep{hansen1959} accessibility measure. Intuitively, the use of the
+constraint illustrates how the differences and ratios of values between
+zones and decay groups can be compared. This is true for other
+constrained cases of the family of accessibility measures.
 
@@
 opportunities measure. This variant, expressed in
-Equation~\ref{eq-total-constrained-market}, represents an expression of
-the concept of market potential (i.e., potential users) as proposed in
-C. D. Harris (1954) and R. W. Vickerman (1974), and which Reilly earlier
-referred to as `potential trade territories' (Reilly 1929). This
-unconstrained form of market potential \(M_j^0\)
-(Equation~\ref{eq-unconstrained-market}), effectively the \(i\) to \(j\)
-transpose of \(V^0_{ij}\), has been used in recent research to express
-the potentially accessible population (i.e., users) as a result of
-regional transportation infrastructure investment projects (e.g.,
-Gutiérrez 2001; Holl 2007; Condeço-Melhorado and Christidis 2018). Put
-another way, market potential can also be thought of as a form of
+Eq~\ref{eq-total-constrained-market}, represents an expression of the
+concept of market potential (i.e., potential users) as proposed in
+Harris \citep{harris_market_1954} and Vickerman
+\citep{vickermanAccessibilityAttractionPotential1974}, and which Reilly
+earlier referred to as `potential trade territories'
+\citep{reilly1929methods}. This unconstrained form of market potential
+\(M_j^0\) (Eq~\ref{eq-unconstrained-market}), effectively the \(i\) to
+\(j\) transpose of \(V^0_{ij}\), has been used in recent research to
+express the potentially accessible population (i.e., users) as a result
+of regional transportation infrastructure investment projects
+\citep[e.g.,][]{gutierrezLocationEconomicPotential2001, holl2007twenty, condecco2018road}.
+Put another way, market potential can also be thought of as a form of
 \emph{passive accessibility}, indicating the number of people that can
@@
 
-Substituting Equation~\ref{eq-total-constrained-market} in
-Equation~\ref{eq-total-constraint-market}, and solving for \(\hat K^T\),
-we obtain:
+Substituting Eq~\ref{eq-total-constrained-market} in
+Eq~\ref{eq-total-constraint-market}, and solving for \(\hat K^T\), we
+obtain:
 
@@
 
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 67.50pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 112.50pt -2\tabcolsep-1.5\arrayrulewidth}}
+\fontsize{6.0pt}{7.2pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|ccc}
 \toprule
- & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 337.50pt -2\tabcolsep-1.5\arrayrulewidth}}{M\textsubscript{i}\textsuperscript{S}} \\ 
+ & \multicolumn{3}{c}{M\textsubscript{i}\textsuperscript{S}} \\ 
 \cmidrule(lr){2-4}
@@
 \emph{singly constrained accessible opportunities} case (an alternative
-formula to spatial availability in Soukhov et al. (2023)), and (b) the
+formula to spatial availability in Soukhov et al.
+\citep{soukhovIntroducingSpatialAvailability2023}), and (b) the
 \emph{singly constrained accessible population} case, its transpose, or
@@
 Unlike the total constraint (i.e.,
-Equation~\ref{eq-total-constraint-accessibility}), the single
-constraint--as will be defined in
-Equation~\ref{eq-opportunity-constraint} and
-Equation~\ref{eq-population-constraint} for the first and second
-variants-- incorporates additional information. In the
-opportunities-accessible variant, the associated balancing factor
-constrains the `potential' for spatial interaction by ensuring that only
-a proportional amount of opportunities at each destination are allocated
-to `demanding' origins. This allocation is informed by demand, or the
-relative amount of population, and associated travel impedance
-connecting the zones. In the population-accessible variant (i.e., the
-\(i\) -\textgreater{} \(j\) transpose of the first variant), population
-at each origin is proportionally allocated to destinations based on the
-share of opportunities and travel impedance.
+Eq~\ref{eq-total-constraint-accessibility}), the single constraint--as
+will be defined in Eq~\ref{eq-opportunity-constraint} and
+Eq~\ref{eq-population-constraint} for the first and second variants--
+incorporates additional information. In the opportunities-accessible
+variant, the associated balancing factor constrains the `potential' for
+spatial interaction by ensuring that only a proportional amount of
+opportunities at each destination are allocated to `demanding' origins.
+This allocation is informed by demand, or the relative amount of
+population, and associated travel impedance connecting the zones. In the
+population-accessible variant (i.e., the \(i\) -\textgreater{} \(j\)
+transpose of the first variant), population at each origin is
+proportionally allocated to destinations based on the share of
+opportunities and travel impedance.
 
@@
 To demonstrate this variant formulaically, we begin with the opportunity
-constraint (Equation~\ref{eq-opportunity-constraint}) as our known piece
-of information. Namely, the sum of accessible opportunities from a
+constraint (Eq~\ref{eq-opportunity-constraint}) as our known piece of
+information. Namely, the sum of accessible opportunities from a
 destination should equal the number of opportunities \(D_j\) at that
@@
 the region. This is comparable to the single attraction-constraint
-(Equation~\ref{eq-constraint2-gravitymodel}) from Wilson's framework.
+(Eq~\ref{eq-constraint2-gravitymodel}) from Wilson's framework.
 
@@
 attraction-constrained model in
-Equation~\ref{eq-attraction-constrained-gravitymodel}, and our
-accessibility measure becomes:
+Eq~\ref{eq-attraction-constrained-gravitymodel}, and our accessibility
+measure becomes:
 
@@
 Introducing the balancing factor in
-Equation~\ref{eq-opportunity-constrained-accessibility}, we obtain:
+Eq~\ref{eq-opportunity-constrained-accessibility}, we obtain:
 
@@
 After this, it is possible to rewrite
-Equation~\ref{eq-opportunity-constrained-accessibility-with-balancing-factor}
+Eq~\ref{eq-opportunity-constrained-accessibility-with-balancing-factor}
 as an origin summary expression of proportionally allocated known
@@
 
-Soukhov et al. (2023) have shown that the role of \(\kappa^S_{ij}\) is
-to allocate opportunities \(D_j\) proportionally to the mass at each
-origin \(i\) and the impedance between \(i\) and \(j\). As in the total
-constrained opportunity case, \(\kappa^S_{ij}\) is dimensionless and
-\(V_i^S\) is in the units of opportunities \(D_j\). The singly
-constrained accessibility measure in
-Equation~\ref{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
-is called spatial availability by Soukhov et al. (2023), because it
-represents the number of opportunities that can be reached \emph{and}
-are available, in the sense that accessible opportunities have been
+Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023} have
+shown that the role of \(\kappa^S_{ij}\) is to allocate opportunities
+\(D_j\) proportionally to the mass at each origin \(i\) and the
+impedance between \(i\) and \(j\). As in the total constrained
+opportunity case, \(\kappa^S_{ij}\) is dimensionless and \(V_i^S\) is in
+the units of opportunities \(D_j\). The singly constrained accessibility
+measure in
+Eq~\ref{eq-attraction-constrained-accessibility-with-proportional-allocation-factor}
+is called spatial availability by Soukhov et al.
+\citep{soukhovIntroducingSpatialAvailability2023}, because it represents
+the number of opportunities that can be reached \emph{and} are
+available, in the sense that accessible opportunities have been
 proportionally allocated based on relative demand, travel impedance and
@@
 expression (accessibility per capita) is a constrained version of the
-popular two-stage floating catchment area measure of Q. Shen (1998b) and
-Luo and Wang (2003):
+popular two-stage floating catchment area measure of Shen
+\citep{shen1998} and Luo and Wang \citep{luo2003}:
 
@@
 these these balancing constants, we can calculate the singly constrained
-opportunity accessibility:
+opportunity accessibility
+(Table~\ref{tbl-simple-example-attraction-constrained-accessibility}).
 
@@
 
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 36.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}}
+\fontsize{6.0pt}{7.2pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|cccc}
 \toprule
- &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 324.00pt -2\tabcolsep-1.5\arrayrulewidth}}{V\textsubscript{i}\textsuperscript{S}} \\ 
+ &  & \multicolumn{3}{c}{V\textsubscript{i}\textsuperscript{S}} \\ 
 \cmidrule(lr){3-5}
@@
 \cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
-Origin & Population (units: \emph{people in 10,000s}) & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
+Origin & Population (10k) & units: \emph{physicians} & units: \emph{physicians} & units: \emph{physicians} \\ 
 \midrule\addlinespace[2.5pt]
@@
 
-\fontsize{7.5pt}{9.0pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 36.00pt -2\tabcolsep-1.5\arrayrulewidth}|>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}>{\centering\arraybackslash}p{\dimexpr 108.00pt -2\tabcolsep-1.5\arrayrulewidth}}
+\fontsize{6.0pt}{7.2pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}l|cccc}
 \toprule
- &  & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 324.00pt -2\tabcolsep-1.5\arrayrulewidth}}{v\textsubscript{i}\textsuperscript{S}} \\ 
+ &  & \multicolumn{3}{c}{v\textsubscript{i}\textsuperscript{S}} \\ 
 \cmidrule(lr){3-5}
@@
 \cmidrule(lr){3-3} \cmidrule(lr){4-4} \cmidrule(lr){5-5}
-Origin & Population (units: \emph{people in 10,000s}) & units: \emph{physicians per capita} & units: \emph{physicians per capita} & units: \emph{physicians per capita} \\ 
+Origin & Population (10k) & units: \emph{physicians per capita} & units: \emph{physicians per capita} & units: \emph{physicians per capita} \\ 
 \midrule\addlinespace[2.5pt]
@@
 
-Similar to Equation~\ref{eq-total-population-balancing-factor} in
-transposing the origins and destinations, we can define a \emph{singly
-constrained} measure of market potential that preserves the known
-population (i.e., the mass weight at the origin \(W_i^{(1)}\) is now
-represented by \(O_i\)). In it's per-capita expression, i.e., equivalent
-to 2SFCA, this constrained concept of market potential been used to
-express ``facility crowdedness'' as in F. H. Wang (2018).
+Similar to Eq~\ref{eq-total-population-balancing-factor} in transposing
+the origins and destinations, we can define a \emph{singly constrained}
+measure of market potential that preserves the known population (i.e.,
+the mass weight at the origin \(W_i^{(1)}\) is now represented by
+\(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
+constrained concept of market potential been used to express ``facility
+crowdedness'' as in Wang \citep{wang_inverted_2018}.
 
@@
 production-constrained model in
-Equation~\ref{eq-production-constrained-gravitymodel}, and our market
+Eq~\ref{eq-production-constrained-gravitymodel}, and our market
 potential measure \(M^S_j\) becomes:
@@
 \emph{by origin} (i.e., \(O_i\)), like
-Equation~\ref{eq-constraint2-gravitymodel} from Wilson's framework:
+Eq~\ref{eq-constraint2-gravitymodel} from Wilson's framework:
 
 \begin{equation}\phantomsection\label{eq-population-constraint}{
-\sum_j M^S_{ji} =  O_i \\
+\sum_j M^S_{ji} =  O_i 
 }\end{equation}
@@
 As well, the single (population) constraint in
-Equation~\ref{eq-population-constraint} ensures that the the total
-constraint (e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O\)) is
-maintained.
+Eq~\ref{eq-population-constraint} ensures that the the total constraint
+(e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O\)) is maintained.
 
@@
 For example, when opportunities \(D_j\) are used to constrain
-Equation~\ref{eq-opportunity-constrained-accessibility} in the
-\emph{singly constrained accessible opportunities measure}, the specific
-mass of the population at origin \(i\) demanding only those
-opportunities at that \(j\) is unknown. Instead, only the population at
-\(i\) demanding opportunities in the region is known, and this is more
-generally represented as \(W_i^{(1)}\) (as in
-Equation~\ref{eq-opportunity-constrained-proportionality-constants}).
+Eq~\ref{eq-opportunity-constrained-accessibility} in the \emph{singly
+constrained accessible opportunities measure}, the specific mass of the
+population at origin \(i\) demanding only those opportunities at that
+\(j\) is unknown. Instead, only the population at \(i\) demanding
+opportunities in the region is known, and this is more generally
+represented as \(W_i^{(1)}\) (as in
+Eq~\ref{eq-opportunity-constrained-proportionality-constants}).
 Similarly, when the population \(O_i\) is used as a constraint in
-Equation~\ref{eq-population-constrained-accessibility} in the
-\emph{singly constrained accessible population measure}, the mass at the
-destination is given by \(W_j^{(2)}\) (also in
-Equation~\ref{eq-population-constrained-proportionality-constants})
-since only the mass of opportunities at each \(j\) is known and
-information on what opportunities are allocated to what zone is unknown.
+Eq~\ref{eq-population-constrained-accessibility} in the \emph{singly
+constrained accessible population measure}, the mass at the destination
+is given by \(W_j^{(2)}\) (also in
+Eq~\ref{eq-population-constrained-proportionality-constants}) since only
+the mass of opportunities at each \(j\) is known and information on what
+opportunities are allocated to what zone is unknown.
 
@@
 of both the population- and opportunity- constraints in the preceding
-singly constrained variants (Equation~\ref{eq-opportunity-constraint}
-and Equation~\ref{eq-population-constraint}), namely the sum of
-population in all origins should match the sum of opportunities in all
-destinations (Equation~\ref{eq-opportunity-population-equality}):
+singly constrained variants (Eq~\ref{eq-opportunity-constraint} and
+Eq~\ref{eq-population-constraint}), namely the sum of population in all
+origins should match the sum of opportunities in all destinations
+(Eq~\ref{eq-opportunity-population-equality}):
 
@@
 total number of opportunities in the region \(O\) as shown in
-Equation~\ref{eq-total-constraint-accessibility}.
+Eq~\ref{eq-total-constraint-accessibility}.
 
@@
 interaction model, as shown in
-Equation~\ref{eq-doubly-constrained-gravitymodel}, \(V_{i}^D\) is as
-follows:
+Eq~\ref{eq-doubly-constrained-gravitymodel}, \(V_{i}^D\) is as follows:
 
@@
 \(V_{ij}^D\) matrix, before calibrating \(A_i\). This process continues
-to update \(A_i\) and \(B_j\) until a convergence criterion is met (see
-Ortúzar and Willumsen 2011, 193--95). The proportional allocation factor
-\(\kappa_{ij}^D\) would then be:
+to update \(A_i\) and \(B_j\) until a convergence criterion is met
+\citep[see][p.~193-195]{ortuzar_2011_modelling}. The proportional
+allocation factor \(\kappa_{ij}^D\) would then be:
 
@@
 
-One could rewrite Equation~\ref{eq-doubly-constrained-accessibility} as
-an origin summary expression of proportionally allocated opportunities:
+One could rewrite Eq~\ref{eq-doubly-constrained-accessibility} as an
+origin summary expression of proportionally allocated opportunities:
 
@@
 (Table~\ref{tbl-small-system-cost} and
-Equation~\ref{eq-travel-behaviour-scenarios}).
+Eq~\ref{eq-travel-behaviour-scenarios}).
 
@@
 measures. As accessibility research evolved largely influenced by Hansen
-(1959), researchers in the field neglected the proportionality constant
-that was originally present in gravity-based models, and is still
-present in spatial interaction modeling. Firstly, we have demonstrated
-that by reintroducing such constraints and defining associated balancing
-factors and proportional allocation factors, we can derive a unifying
-family of accessibility measures that reintroduces units to the
-resulting values. These values become more intuitive for the purpose of
-analysis and comparison. Secondly, the constraining constants, depending
-on the case (i.e., total, singly- or doubly- constrained), restrict the
-degree of \emph{potential}, linking accessibility (the potential for
-spatial interaction) with access (spatial interaction) on the same
-continuum based on the constraint used. Lastly, we discussed how popular
-measures such as the one used in Hansen (1959) and the 2SFCA link into
-the family of accessibility measures. We summarise the family of
-accessibility measures, as follows.
-
-We first place the popular Hansen-type accessibility measure (Hansen
-1959) within this family of measures as an ``unconstrained'' case,
-demonstrating that resulting values cannot be directly compared across
-different travel scenarios without ad-hoc adjustments. We then show how
-applying a total constraint balances the units and produces a
-statistically averaged solution that converges to the regional average
-for each zone as the decay effect decreases. In other words, the
-total-constraint model could be a more interpretable alternative for the
-unconstrained case if population-competition is not relevant and one is
-interested in capturing the maximum \emph{potential}; specifically, if
-there is a fixed number of opportunities in the region, and if it makes
-sense to assume that people accessing proximate opportunities leave
-fewer for others, \emph{without} considering the population size at the
-origins.
+\citep{hansen1959}, researchers in the field neglected the
+proportionality constant that was originally present in gravity-based
+models, and is still present in spatial interaction modeling. Firstly,
+we have demonstrated that by reintroducing such constraints and defining
+associated balancing factors and proportional allocation factors, we can
+derive a unifying family of accessibility measures that reintroduces
+units to the resulting values. These values become more intuitive for
+the purpose of analysis and comparison. Secondly, the constraining
+constants, depending on the case (i.e., total, singly- or doubly-
+constrained), restrict the degree of \emph{potential}, linking
+accessibility (the potential for spatial interaction) with access
+(spatial interaction) on the same continuum based on the constraint
+used. Lastly, we discussed how popular measures such as the one used in
+Hansen \citep{hansen1959} and the 2SFCA link into the family of
+accessibility measures. We summarise the family of accessibility
+measures, as follows.
+
+We first place the popular Hansen-type accessibility measure
+\citep{hansen1959} within this family of measures as an
+``unconstrained'' case, demonstrating that resulting values cannot be
+directly compared across different travel scenarios without ad-hoc
+adjustments. We then show how applying a total constraint balances the
+units and produces a statistically averaged solution that converges to
+the regional average for each zone as the decay effect decreases. In
+other words, the total-constraint model could be a more interpretable
+alternative for the unconstrained case if population-competition is not
+relevant and one is interested in capturing the maximum
+\emph{potential}; specifically, if there is a fixed number of
+opportunities in the region, and if it makes sense to assume that people
+accessing proximate opportunities leave fewer for others, \emph{without}
+considering the population size at the origins.
 
@@
 equivalent to the spatial availability introduced in Soukhov et al.
-(2023). In this case, all accessibility values are fixed to sum to a
-known zonal opportunity-size value (implicitly, the regional total of
-opportunities), but they are not required to sum to any population-based
-values at the zone or regional level. The singly constrained model could
-be useful if regional competition is a factor and if the acknowledgment
-that only a finite number of opportunities can be allocated from each
-destination (with those allocations distributed based on origin
-population size) is suitable. We also introduce an `accessible' PPR
-(e.g., opportunities per capita), calculated by dividing each
-accessibility value by the zonal population. To clarify, this per capita
-expression of the singly constrained case is equivalent to the 2SFCA
-(Luo and Wang 2003; Q. Shen 1998b), hence linking this literature back
-to spatial interaction principles.
+\citep{soukhovIntroducingSpatialAvailability2023}. In this case, all
+accessibility values are fixed to sum to a known zonal opportunity-size
+value (implicitly, the regional total of opportunities), but they are
+not required to sum to any population-based values at the zone or
+regional level. The singly constrained model could be useful if regional
+competition is a factor and if the acknowledgment that only a finite
+number of opportunities can be allocated from each destination (with
+those allocations distributed based on origin population size) is
+suitable. We also introduce an `accessible' PPR (e.g., opportunities per
+capita), calculated by dividing each accessibility value by the zonal
+population. To clarify, this per capita expression of the singly
+constrained case is equivalent to the 2SFCA \citep{luo2003, shen1998},
+hence linking this literature back to spatial interaction principles.
 
@@
 
-In summary, building on Wilson (1971)'s foundational work, this paper
-proposed a unified framework for accessibility that is able to account
-for competition. By reintroducing Wilson's proportionality constant, the
-proposed family of constrained accessibility measures restores
-measurement units to accessibility estimates. This enhancement provides
-a more interpretable, consistent, and theoretically grounded basis for
-accessibility analysis, which could help advance the adoption of
-accessibility-oriented planning.
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-albacete2017measuring}
-Albacete, Xavier, Doina Olaru, Valerià Paül, and Sharon Biermann. 2017.
-{``Measuring the Accessibility of Public Transport: A Critical
-Comparison Between Methods in Helsinki.''} \emph{Applied Spatial
-Analysis and Policy} 10: 161--88.
-
-\bibitem[\citeproctext]{ref-allenMeasureCompetitiveAccess2020}
-Allen, Jeff, and Steven Farber. 2020. {``A Measure of Competitive Access
-to Destinations for Comparing Across Multiple Study Regions.''}
-\emph{Geographical Analysis} 52 (1): 69--86.
-\url{https://doi.org/10.1111/gean.12188}.
-
-\bibitem[\citeproctext]{ref-alonso2014labour}
-Alonso, M Pilar, MA Beamonte, Pilar Gargallo, and Manuel J Salvador.
-2014. {``Labour and Residential Accessibility: A Bayesian Analysis Based
-on Poisson Gravity Models with Spatial Effects.''} \emph{Journal of
-Geographical Systems} 16: 409--39.
-
-\bibitem[\citeproctext]{ref-ashiru2003space}
-Ashiru, Olu, John W Polak, and Robert B Noland. 2003. {``Space-Time User
-Benefit and Utility Accessibility Measures for Individual Activity
-Schedules.''} \emph{Transportation Research Record} 1854 (1): 62--73.
-
-\bibitem[\citeproctext]{ref-battyChronicleScientificPlanning1994}
-Batty, Michael. 1994. {``A {Chronicle} of {Scientific} {Planning}: {The}
-{Anglo}-{American} {Modeling} {Experience}.''} \emph{Journal of the
-American Planning Association} 60 (1): 7--16.
-\url{https://doi.org/10.1080/01944369408975546}.
-
-\bibitem[\citeproctext]{ref-battyMethodResiduesUrban1976}
-Batty, M, and L March. 1976. {``The Method of Residues in Urban
-Modelling.''} \emph{Environment and Planning A: Economy and Space} 8
-(2): 189--214. \url{https://doi.org/10.1068/a080189}.
-
-\bibitem[\citeproctext]{ref-beaumontLocationallocationProblemsPlane1981}
-Beaumont, John R. 1981. {``Location-Allocation Problems in a Plane a
-Review of Some Models.''} \emph{Socio-Economic Planning Sciences} 15
-(5): 217--29. \url{https://doi.org/10.1016/0038-0121(81)90042-2}.
-
-\bibitem[\citeproctext]{ref-beckers2022incorporating}
-Beckers, Joris, Mark Birkin, Graham Clarke, Nick Hood, Andy Newing, and
-Ryan Urquhart. 2022. {``Incorporating e-Commerce into Retail Location
-Models.''} \emph{Geographical Analysis} 54 (2): 274--93.
-
-\bibitem[\citeproctext]{ref-boisjoly2017informality}
-Boisjoly, Geneviève, Ana Isabel Moreno-Monroy, and Ahmed El-Geneidy.
-2017. {``Informality and Accessibility to Health by Public Transit:
-Evidence from the São Paulo Metropolitan Region.''} Journal Article.
-\emph{Journal of Transport Geography} 64: 89--96.
-\url{https://doi.org/10.1016/j.jtrangeo.2017.08.005}.
-
-\bibitem[\citeproctext]{ref-careyPrinciplesSocialScience1858}
-Carey, Henry Charles. 1858. \emph{Principles of Social Science}.
-University of Michigan Library Digital Collections: In the digital
-collection Making of America Books.
-\url{https://name.umdl.umich.edu/AFR1829.0001.001.}
-
-\bibitem[\citeproctext]{ref-caschili2015accessibility}
-Caschili, Simone, Andrea De Montis, and Daniele Trogu. 2015.
-{``Accessibility and Rurality Indicators for Regional Development.''}
-\emph{Computers, Environment and Urban Systems} 49: 98--114.
-
-\bibitem[\citeproctext]{ref-cavendish_xxi_1798}
-Cavendish, Henry. 1798. {``{XXI}. {Experiments} to Determine the Density
-of the Earth.''} \emph{Philosophical Transactions of the Royal Society}
-88: 469--526. \url{https://doi.org/10.1098/rstl.1798.0022}.
-
-\bibitem[\citeproctext]{ref-chen2013regional}
-Chen, Guineng, and João de Abreu e Silva. 2013. {``Regional Impacts of
-High-Speed Rail: A Review of Methods and Models.''} \emph{Transportation
-Letters} 5 (3): 131--43.
-
-\bibitem[\citeproctext]{ref-chia2020extending}
-Chia, Jason, and Jinwoo Brian Lee. 2020. {``Extending Public Transit
-Accessibility Models to Recognise Transfer Location.''} \emph{Journal of
-Transport Geography} 82: 102618.
-
-\bibitem[\citeproctext]{ref-clarke2002deriving}
-Clarke, Graham, Heather Eyre, and Cliff Guy. 2002. {``Deriving
-Indicators of Access to Food Retail Provision in British Cities: Studies
-of Cardiff, Leeds and Bradford.''} \emph{Urban Studies} 39 (11):
-2041--60.
-
-\bibitem[\citeproctext]{ref-cliff_evaluating_1974}
-Cliff, A. D., R. L. Martin, and J. K. Ord. 1974. {``Evaluating the
-Friction of Distance Parameter in Gravity Models.''} \emph{Regional
-Studies} 8 (3-4): 281--86.
-\url{https://doi.org/10.1080/09595237400185281}.
-
-\bibitem[\citeproctext]{ref-condecco2018road}
-Condeço-Melhorado, Ana, and Panayotis Christidis. 2018. {``Road
-Accessibility in Border Regions: A Joint Approach.''} \emph{Networks and
-Spatial Economics} 18 (2): 363--83.
-
-\bibitem[\citeproctext]{ref-curtis2010planning}
-Curtis, Carey, and Jan Scheurer. 2010. {``Planning for Sustainable
-Accessibility: Developing Tools to Aid Discussion and
-Decision-Making.''} \emph{Progress in Planning} 74 (2): 53--106.
-
-\bibitem[\citeproctext]{ref-dai2017visualization}
-Dai, Linlin, Li Wan, and Shijie Gai. 2017. {``A Visualization Framework
-for Synthesizing Spatial Impacts from Multiple Site Factors.''}
-\emph{Journal of Asian Architecture and Building Engineering} 16 (2):
-311--15.
-
-\bibitem[\citeproctext]{ref-delamater2013spatial}
-Delamater, P. L. 2013. {``Spatial Accessibility in Suboptimally
-Configured Health Care Systems: A Modified Two-Step Floating Catchment
-Area (M2SFCA) Metric.''} Journal Article. \emph{Health \& Place} 24:
-30--43. \url{https://doi.org/10.1016/j.healthplace.2013.07.012}.
-
-\bibitem[\citeproctext]{ref-el2011place}
-El-Geneidy, Ahmed, and David Levinson. 2011. {``Place Rank: Valuing
-Spatial Interactions.''} \emph{Networks and Spatial Economics} 11:
-643--59.
-
-\bibitem[\citeproctext]{ref-elgeneidyMakingAccessibilityWork2022}
----------. 2022. {``Making Accessibility Work in Practice.''}
-\emph{Transport Reviews} 42 (2): 129--33.
-\url{https://doi.org/10.1080/01441647.2021.1975954}.
-
-\bibitem[\citeproctext]{ref-farber_2013_social}
-Farber, S., T. Neutens, H. J. Miller, and X. Li. 2013. {``The Social
-Interaction Potential of Metropolitan Regions: A Time-Geographic
-Measurement Approach Using Joint Accessibility.''} Journal Article.
-\emph{Annals of the Association of American Geographers} 103 (3):
-483--504. \url{https://doi.org/10.1080/00045608.2012.689238}.
-
-\bibitem[\citeproctext]{ref-farberActivitySpacesMeasurement2012}
-Farber, S., A. Páez, and C. Morency. 2012. {``Activity Spaces and the
-Measurement of Clustering and Exposure: A Case Study of Linguistic
-Groups in {Montreal}.''} \emph{Environment and Planning A} 44 (2):
-315--32.
-
-\bibitem[\citeproctext]{ref-farber_running_2011}
-Farber, Steven, and Antonio Páez. 2011. {``Running to Stay in Place: The
-Time-Use Implications of Automobile Oriented Land-Use and Travel.''}
-\emph{Journal of Transport Geography} 19 (4): 782--93.
-\url{https://doi.org/10.1016/j.jtrangeo.2010.09.008}.
-
-\bibitem[\citeproctext]{ref-ferreiraReenactingMobilityAccessibility2020}
-Ferreira, António, and Enrica Papa. 2020. {``Re-Enacting the Mobility
-Versus Accessibility Debate: Moving Towards Collaborative Synergies
-Among Experts.''} \emph{Case Studies on Transport Policy} 8 (3):
-1002--9. \url{https://doi.org/10.1016/j.cstp.2020.04.006}.
-
-\bibitem[\citeproctext]{ref-fotheringham_spatial_1984}
-Fotheringham, A S. 1984. {``Spatial {Flows} and {Spatial} {Patterns}.''}
-\emph{Environment and Planning A: Economy and Space} 16 (4): 529--43.
-\url{https://doi.org/10.1068/a160529}.
-
-\bibitem[\citeproctext]{ref-fotheringhamSpatialCompetitionAgglomeration1985}
----------. 1985. {``Spatial Competition and Agglomeration in Urban
-Modelling.''} \emph{Environment and Planning A: Economy and Space} 17
-(2): 213--30. \url{https://doi.org/10.1068/a170213}.
-
-\bibitem[\citeproctext]{ref-fotheringhamSPATIALSTRUCTUREDISTANCE1981}
-Fotheringham, A. Stewart. 1981. {``{SPATIAL} {STRUCTURE} {AND}
-{DISTANCE}‐{DECAY} {PARAMETERS}.''} \emph{Annals of the Association of
-American Geographers} 71 (3): 425--36.
-\url{https://doi.org/10.1111/j.1467-8306.1981.tb01367.x}.
-
-\bibitem[\citeproctext]{ref-geursAccessibilityEvaluationLanduse2004}
-Geurs, Karst T., and Bert van Wee. 2004. {``Accessibility Evaluation of
-Land-Use and Transport Strategies: Review and Research Directions.''}
-\emph{Journal of Transport Geography} 12 (2): 127--40.
-\url{https://doi.org/10.1016/j.jtrangeo.2003.10.005}.
-
-\bibitem[\citeproctext]{ref-geurs2006accessibility}
-Geurs, Karst T, Bert van Wee, and Piet Rietveld. 2006. {``Accessibility
-Appraisal of Integrated Land-Use---Transport Strategies: Methodology and
-Case Study for the Netherlands Randstad Area.''} \emph{Environment and
-Planning B: Planning and Design} 33 (5): 639--60.
-
-\bibitem[\citeproctext]{ref-giuliano2010accessibility}
-Giuliano, Genevieve, Peter Gordon, Qisheng Pan, and JiYoung Park. 2010.
-{``Accessibility and Residential Land Values: Some Tests with New
-Measures.''} \emph{Urban Studies} 47 (14): 3103--30.
-
-\bibitem[\citeproctext]{ref-grengs2004measuring}
-Grengs, Joe. 2004. {``Measuring Change in Small-Scale Transit
-Accessibility with Geographic Information Systems: Buffalo and
-Rochester, New York.''} \emph{Transportation Research Record} 1887 (1):
-10--17.
-
-\bibitem[\citeproctext]{ref-grengs2010job}
----------. 2010. {``Job Accessibility and the Modal Mismatch in
-Detroit.''} \emph{Journal of Transport Geography} 18 (1): 42--54.
-
-\bibitem[\citeproctext]{ref-grengs2012equity}
----------. 2012. {``Equity and the Social Distribution of Job
-Accessibility in Detroit.''} \emph{Environment and Planning B: Planning
-and Design} 39 (5): 785--800.
-
-\bibitem[\citeproctext]{ref-grengs2015nonwork}
----------. 2015. {``Nonwork Accessibility as a Social Equity
-Indicator.''} \emph{International Journal of Sustainable Transportation}
-9 (1): 1--14.
-
-\bibitem[\citeproctext]{ref-grengs2010intermetropolitan}
-Grengs, Joe, Jonathan Levine, Qing Shen, and Qingyun Shen. 2010.
-{``Intermetropolitan Comparison of Transportation Accessibility: Sorting
-Out Mobility and Proximity in San Francisco and Washington, DC.''}
-\emph{Journal of Planning Education and Research} 29 (4): 427--43.
-
-\bibitem[\citeproctext]{ref-gutierrez_evaluating_2011}
-Gutierrez, J., A. Condeco-Melhorado, E. Lopez, and A. Monzon. 2011.
-{``Evaluating the {European} Added Value of {TEN}-{T} Projects: A
-Methodological Proposal Based on Spatial Spillovers, Accessibility and
-{GIS}.''} \emph{Journal of Transport Geography} 19 (4): 840--50.
-\url{https://doi.org/10.1016/j.jtrangeo.2010.10.011}.
-
-\bibitem[\citeproctext]{ref-gutierrezLocationEconomicPotential2001}
-Gutiérrez, Javier. 2001. {``Location, Economic Potential and Daily
-Accessibility: An Analysis of the Accessibility Impact of the High-Speed
-Line Madrid--Barcelona--French Border.''} \emph{Journal of Transport
-Geography} 9 (4): 229--42.
-\url{https://doi.org/10.1016/S0966-6923(01)00017-5}.
-
-\bibitem[\citeproctext]{ref-handyMeasuringAccessibilityExploration1997}
-Handy, Sandy L., and Debbie A. Niemeier. 1997. {``Measuring
-Accessibility: An Exploration of Issues and Alternatives.''}
-\emph{Environment and Planning A: Economy and Space} 29 (7): 1175--94.
-\url{https://doi.org/10.1068/a291175}.
-
-\bibitem[\citeproctext]{ref-handyACCESSIBILITYVSMOBILITYENHANCING2002}
-Handy, Susan. 2002. {``{ACCESSIBILITY}- {VS}. {MOBILITY}-{ENHANCING}
-{STRATEGIES} {FOR} {ADDRESSING} {AUTOMOBILE} {DEPENDENCE} {IN} {THE}
-u.s.''}
-\url{https://escholarship.org/content/qt5kn4s4pb/qt5kn4s4pb_noSplash_18f73162ff86f04dcb255331d63eeba8.pdf}.
-
-\bibitem[\citeproctext]{ref-handy2020}
----------. 2020. {``Is Accessibility an Idea Whose Time Has Finally
-Come?''} \emph{Transportation Research Part D: Transport and
-Environment} 83 (June): 102319.
-\url{https://doi.org/10.1016/j.trd.2020.102319}.
-
-\bibitem[\citeproctext]{ref-hansen1959}
-Hansen, Walter G. 1959. {``How Accessibility Shapes Land Use.''}
-\emph{Journal of the American Institute of Planners} 25 (2): 73--76.
-\url{https://doi.org/10.1080/01944365908978307}.
-
-\bibitem[\citeproctext]{ref-harrisEquilibriumValuesDynamics1978}
-Harris, B, and A G Wilson. 1978. {``Equilibrium Values and Dynamics of
-Attractiveness Terms in Production-Constrained Spatial-Interaction
-Models.''} \emph{Environment and Planning A: Economy and Space} 10 (4):
-371--88. \url{https://doi.org/10.1068/a100371}.
-
-\bibitem[\citeproctext]{ref-harris_market_1954}
-Harris, Chauncy D. 1954. {``The {Market} as a {Factor} in the
-{Localization} of {Industry} in the {United} {States}.''} \emph{Annals
-of the Association of American Geographers} 44 (4): 315--48.
-\url{https://www.jstor.org/stable/2561395}.
-
-\bibitem[\citeproctext]{ref-he2017measuring}
-He, Jianhua, Chun Li, Yan Yu, Yaolin Liu, and Junlong Huang. 2017.
-{``Measuring Urban Spatial Interaction in Wuhan Urban Agglomeration,
-Central China: A Spatially Explicit Approach.''} \emph{Sustainable
-Cities and Society} 32: 569--83.
-
-\bibitem[\citeproctext]{ref-holl2007twenty}
-Holl, Adelheid. 2007. {``Twenty Years of Accessibility Improvements. The
-Case of the Spanish Motorway Building Programme.''} \emph{Journal of
-Transport Geography} 15 (4): 286--97.
-
-\bibitem[\citeproctext]{ref-hutton_xxxiii_1778}
-Hutton, Charles. 1778. {``{XXXIII}. {An} Account of the Calculations
-Made from the Survey and Measures Taken at {Schehallien}, in Order to
-Ascertain the Mean Density of the {Earth}.''} \emph{Philosophical
-Transactions of the Royal Society} 68: 689--788.
-\url{https://doi.org/10.1098/rstl.1778.0034}.
-
-\bibitem[\citeproctext]{ref-kapatsila_resolving_2023}
-Kapatsila, Bogdan, Manuel Santana Palacios, Emily Grisé, and Ahmed
-El-Geneidy. 2023. {``Resolving the Accessibility Dilemma: {Comparing}
-Cumulative and Gravity-Based Measures of Accessibility in Eight
-{Canadian} Cities.''} \emph{Journal of Transport Geography} 107
-(February): 103530.
-\url{https://doi.org/10.1016/j.jtrangeo.2023.103530}.
-
-\bibitem[\citeproctext]{ref-karstEvaluationAccessibilityImpacts2003}
-Karst, T, and Jan R Ritsema Van Eck. 2003. {``Evaluation of
-Accessibility Impacts of Land-Use Scenarios: The Implications of Job
-Competition, Land-Use, and Infrastructure Developments for the
-Netherlands.''} \emph{Environment and Planning B: Planning and Design}
-30 (1): 69--87. \url{https://doi.org/10.1068/b12940}.
-
-\bibitem[\citeproctext]{ref-kharel2024examining}
-Kharel, Subham, Seyedsoheil Sharifiasl, and Qisheng Pan. 2024.
-{``Examining Food Access Equity by Integrating Grocery Store Pricing
-into Spatial Accessibility Measures.''} \emph{Transportation Research
-Record}, 03611981241254382.
-
-\bibitem[\citeproctext]{ref-kirbyNormalizingFactorsGravity1970}
-Kirby, Howard R. 1970. {``Normalizing Factors of the Gravity Model---an
-Interpretation.''} \emph{Transportation Research} 4 (1): 37--50.
-\url{https://doi.org/10.1016/0041-1647(70)90073-0}.
-
-\bibitem[\citeproctext]{ref-kovatch1971modeling}
-Kovatch, George, George Zames, et al. 1971. {``Modeling Transportation
-Systems: An Overview.''} \url{https://rosap.ntl.bts.gov/view/dot/11813}.
-
-\bibitem[\citeproctext]{ref-kwan1998space}
-Kwan, Mei-Po. 1998. {``Space-Time and Integral Measures of Individual
-Accessibility: A Comparative Analysis Using a Point-Based Framework.''}
-\emph{Geographical Analysis} 30 (3): 191--216.
-
-\bibitem[\citeproctext]{ref-lavery_driving_2013}
-Lavery, T. A., A. Páez, and P. S. Kanaroglou. 2013. {``Driving Out of
-Choices: {An} Investigation of Transport Modality in a University
-Sample.''} \emph{Transportation Research Part A: Policy and Practice} 57
-(November): 37--46. \url{https://doi.org/10.1016/j.tra.2013.09.010}.
-
-\bibitem[\citeproctext]{ref-leonardiOptimumFacilityLocation1978}
-Leonardi, G. 1978. {``Optimum Facility Location by Accessibility
-Maximizing.''} \emph{Environment and Planning A: Economy and Space} 10
-(11): 1287--1305. \url{https://doi.org/10.1068/a101287}.
-
-\bibitem[\citeproctext]{ref-leonardiRandomUtilityDemand1984}
-Leonardi, Giorgio, and Roberto Tadei. 1984. {``Random Utility Demand
-Models and Service Location.''} \emph{Regional Science and Urban
-Economics} 14 (3): 399--431.
-\url{https://doi.org/10.1016/0166-0462(84)90009-7}.
-
-\bibitem[\citeproctext]{ref-levine2012does}
-Levine, Jonathan, Joe Grengs, Qingyun Shen, and Qing Shen. 2012. {``Does
-Accessibility Require Density or Speed? A Comparison of Fast Versus
-Close in Getting Where You Want to Go in US Metropolitan Regions.''}
-\emph{Journal of the American Planning Association} 78 (2): 157--72.
-
-\bibitem[\citeproctext]{ref-levinson2012positive}
-Levinson, David, and Arthur Huang. 2012. {``A Positive Theory of Network
-Connectivity.''} \emph{Environment and Planning B: Planning and Design}
-39 (2): 308--25.
-
-\bibitem[\citeproctext]{ref-levinsonGeneralTheoryAccess2020}
-Levinson, David, and Hao Wu. 2020. {``Towards a General Theory of
-Access.''} \emph{Journal of Transport and Land Use} 13 (1): 129--58.
-\url{https://www.jstor.org/stable/26967239}.
-
-\bibitem[\citeproctext]{ref-liang_novel_2024}
-Liang, Huilin, Qi Yan, and Yujia Yan. 2024. {``A Novel Spatiotemporal
-Framework for Accessing Green Space Accessibility Change in Adequacy and
-Equity: {Evidence} from a Rapidly Urbanizing {Chinese} {City} in
-2012--2021.''} \emph{Cities} 151 (August): 105112.
-\url{https://doi.org/10.1016/j.cities.2024.105112}.
-
-\bibitem[\citeproctext]{ref-liu2004accessibility}
-Liu, Suxia, and Xuan Zhu. 2004. {``Accessibility Analyst: An Integrated
-GIS Tool for Accessibility Analysis in Urban Transportation Planning.''}
-\emph{Environment and Planning B: Planning and Design} 31 (1): 105--24.
-
-\bibitem[\citeproctext]{ref-liu2015spatial}
-Liu, Xianteng, and Jiangping Zhou. 2015. {``Spatial Pattern of Land Use
-and Its Implications for Mode-Based Accessibility: Case Study of
-Nanjing, China.''} \emph{Journal of Urban Planning and Development} 141
-(2): 05014012.
-
-\bibitem[\citeproctext]{ref-lopezMeasuring2008}
-Lopez, E., J. Gutierrez, and G. Gomez. 2008. {``Measuring Regional
-Cohesion Effects of Large-Scale Transport Infrastructure Investments:
-{An} Accessibility Approach.''} \emph{European Planning Studies} 16 (2):
-277--301.
-
-\bibitem[\citeproctext]{ref-luoEnhanced2009}
-Luo, Wei, and Y. Qi. 2009. {``An Enhanced Two-Step Floating Catchment
-Area ({E2SFCA}) Method for Measuring Spatial Accessibility to Primary
-Care Physicians.''} \emph{Health \& Place} 15 (4): 1100--1107.
-
-\bibitem[\citeproctext]{ref-luo2003}
-Luo, Wei, and Fahui Wang. 2003. {``Measures of Spatial Accessibility to
-Health Care in a GIS Environment: Synthesis and a Case Study in the
-Chicago Region.''} \emph{Environment and Planning B: Planning and
-Design} 30 (6): 865--84. \url{https://doi.org/10.1068/b29120}.
-
-\bibitem[\citeproctext]{ref-manaugh2012makes}
-Manaugh, Kevin, and Ahmed M El-Geneidy. 2012. {``What Makes
-Travel'local' Defining and Understanding Local Travel Behavior.''}
-\emph{Journal of Transport and Land Use} 5 (3): 15--27.
-
-\bibitem[\citeproctext]{ref-margaridacondecomelhoradoImpactMeasuringInternal2016}
-Margarida Condeço Melhorado, Ana, Hande Demirel, Mert Kompil, Elena
-Navajas, and Panayotis Christidis. 2016. {``The Impact of Measuring
-Internal Travel Distances on Selfpotentials and Accessibility.''}
-\emph{European Journal of Transport and Infrastructure Research}.
-\url{https://doi.org/10.18757/EJTIR.2016.16.2.3139}.
-
-\bibitem[\citeproctext]{ref-marques_accessibility_2021}
-Marques, João Lourenço, Jan Wolf, and Fillipe Feitosa. 2021.
-{``Accessibility to Primary Schools in {Portugal}: A Case of Spatial
-Inequity?''} \emph{Regional Science Policy \& Practice} 13 (3):
-693--708. \url{https://doi.org/10.1111/rsp3.12303}.
-
-\bibitem[\citeproctext]{ref-marwal2022literature}
-Marwal, Aviral, and Elisabete Silva. 2022. {``Literature Review of
-Accessibility Measures and Models Used in Land Use and Transportation
-Planning in Last 5 Years.''} \emph{Journal of Geographical Sciences} 32
-(3): 560--84.
-
-\bibitem[\citeproctext]{ref-mayaud2019future}
-Mayaud, Jerome R, Martino Tran, Rafael HM Pereira, and Rohan Nuttall.
-2019. {``Future Access to Essential Services in a Growing Smart City:
-The Case of Surrey, British Columbia.''} \emph{Computers, Environment
-and Urban Systems} 73: 1--15.
-
-\bibitem[\citeproctext]{ref-mckeanManual1883}
-McKean, Kate. 1883. \emph{Manual of {Social} {Science} Being a
-{Condensation} of the {Principles} of {Social} {Science} of {H}.{C}.
-{Carey}}. Philadelphia: Henry Carey Baird; Co. Industrial Publishers.
-
-\bibitem[\citeproctext]{ref-mdotMnDOTJoins2007}
-MDOT. 2007. {``Mn/{DOT} Joins Interstate Highway System's 50th
-Anniversary Celebration.''} December 4, 2007.
-\url{https://web.archive.org/web/20071204072603/http://www.dot.state.mn.us/interstate50/50facts.html}.
-
-\bibitem[\citeproctext]{ref-merlin2017competition}
-Merlin, Louis A., and Lingqian Hu. 2017. {``Does Competition Matter in
-Measures of Job Accessibility? Explaining Employment in Los Angeles.''}
-Journal Article. \emph{Journal of Transport Geography} 64: 77--88.
-\url{https://doi.org/10.1016/j.jtrangeo.2017.08.009}.
-
-\bibitem[\citeproctext]{ref-millerAccessibilityMeasurementApplication2018}
-Miller, Eric J. 2018. {``Accessibility: Measurement and Application in
-Transportation Planning.''} \emph{Transport Reviews} 38 (5): 551--55.
-\url{https://doi.org/10.1080/01441647.2018.1492778}.
-
-\bibitem[\citeproctext]{ref-millerMeasuringSpaceTimeAccessibility1999}
-Miller, Harvey J. 1999. {``Measuring Space‐time Accessibility Benefits
-Within Transportation Networks: Basic Theory and Computational
-Procedures.''} \emph{Geographical Analysis} 31 (1): 187--212.
-\url{https://doi.org/10.1111/gean.1999.31.1.187}.
-
-\bibitem[\citeproctext]{ref-miller_collaborative_2011}
----------. 2011. {``Collaborative Mobility: Using Geographic Information
-Science to Cultivate Cooperative Transportation Systems.''}
-\emph{Procedia - Social and Behavioral Sciences} 21 (0): 24--28.
-https://doi.org/\url{http://dx.doi.org/10.1016/j.sbspro.2011.07.005}.
-
-\bibitem[\citeproctext]{ref-morrisAccessibilityIndicatorsTransport1979}
-Morris, J. M., P. L. Dumble, and M. R. Wigan. 1979. {``Accessibility
-Indicators for Transport Planning.''} \emph{Transportation Research Part
-A: General} 13 (2): 91--109.
-\url{https://doi.org/10.1016/0191-2607(79)90012-8}.
-
-\bibitem[\citeproctext]{ref-naqavi2023mobility}
-Naqavi, Fatemeh, Marcus Sundberg, Oskar Blom Västberg, Anders Karlström,
-and Muriel Beser Hugosson. 2023. {``Mobility Constraints and
-Accessibility to Work: Application to Stockholm.''} \emph{Transportation
-Research Part A: Policy and Practice} 175: 103790.
-
-\bibitem[\citeproctext]{ref-neutens_human_2007}
-Neutens, Tijs, Frank Witlox, Nico Van de Weghe, and Philippe De Maeyer.
-2007. {``Human Interaction Spaces Under Uncertainty.''}
-\emph{Transportation Research Record} 2021 (1): 28--35.
-
-\bibitem[\citeproctext]{ref-ng2022reflection}
-Ng, Matthew Kok Ming, Josephine Roper, Chyi Lin Lee, and Christopher
-Pettit. 2022. {``The Reflection of Income Segregation and Accessibility
-Cleavages in Sydney's House Prices.''} \emph{ISPRS International Journal
-of Geo-Information} 11 (7): 413.
-
-\bibitem[\citeproctext]{ref-ortuzar_2011_modelling}
-Ortúzar, J. D., and L. G. Willumsen. 2011. \emph{Modelling Transport}.
-Book. New York: Wiley.
-
-\bibitem[\citeproctext]{ref-paez_developing_2013}
-Paez, A., M. Moniruzzaman, P. L. Bourbonnais, and C. Morency. 2013.
-{``Developing a Web-Based Accessibility Calculator Prototype for the
-{Greater} {Montreal} {Area}.''} \emph{Transportation Research Part
-A-Policy and Practice} 58 (December): 103--15.
-\url{https://doi.org/10.1016/j.tra.2013.10.020}.
-
-\bibitem[\citeproctext]{ref-paez2012measuring}
-Paez, A., D. M. Scott, and C. Morency. 2012. {``Measuring Accessibility:
-Positive and Normative Implementations of Various Accessibility
-Indicators.''} Journal Article. \emph{Journal of Transport Geography}
-25: 141--53. \url{https://doi.org/10.1016/j.jtrangeo.2012.03.016}.
-
-\bibitem[\citeproctext]{ref-paez_jobs_2013}
-Páez, Antonio, Steven Farber, Ruben Mercado, Matthew Roorda, and
-Catherine Morency. 2013. {``Jobs and the {Single} {Parent}: {An}
-{Analysis} of {Accessibility} to {Employment} in {Toronto}.''}
-\emph{Urban Geography} 34 (6): 815--42.
-\url{https://doi.org/10.1080/02723638.2013.778600}.
-
-\bibitem[\citeproctext]{ref-paez_healthcare_2010}
-Páez, Antonio, Ruben Mercado, Steven Farber, Catherine Morency, and
-Matthew Roorda. 2010. {``Accessibility to Health Care Facilities in
-Montreal Island: An Application of Relative Accessibility Indicators
-from the Perspective of Senior and Non-Senior Residents.''} Journal
-Article. \emph{International Journal of Health Geographics} 9 (52):
-1--9. \url{http://www.ij-healthgeographics.com/content/9/1/52}.
-
-\bibitem[\citeproctext]{ref-pan2013impacts}
-Pan, Qisheng. 2013. {``The Impacts of an Urban Light Rail System on
-Residential Property Values: A Case Study of the Houston METRORail
-Transit Line.''} \emph{Transportation Planning and Technology} 36 (2):
-145--69.
-
-\bibitem[\citeproctext]{ref-pan2020measuring}
-Pan, Qisheng, Zhonghua Jin, and Xuejun Liu. 2020. {``Measuring the
-Effects of Job Competition and Matching on Employment Accessibility.''}
-\emph{Transportation Research Part D: Transport and Environment} 87:
-102535.
-
-\bibitem[\citeproctext]{ref-pereira_2021_geographic}
-Pereira, Rafael H. M., Carlos Kauê Vieira Braga, Luciana Mendes Servo,
-Bernardo Serra, Pedro Amaral, Nelson Gouveia, and Antonio Paez. 2021.
-{``Geographic Access to COVID-19 Healthcare in Brazil Using a Balanced
-Float Catchment Area Approach.''} Journal Article. \emph{Social Science
-\& Medicine} 273: 113773.
-https://doi.org/\url{https://doi.org/10.1016/j.socscimed.2021.113773}.
-
-\bibitem[\citeproctext]{ref-pirie_measuring_1979}
-Pirie, G. H. 1979. {``Measuring {Accessibility}: {A} {Review} and
-{Proposal}.''} \emph{Environment and Planning A: Economy and Space} 11
-(3): 299--312. \url{https://doi.org/10.1068/a110299}.
-
-\bibitem[\citeproctext]{ref-rau2012spatial}
-Rau, Henrike, and Amaya Vega. 2012. {``Spatial (Im) Mobility and
-Accessibility in i Reland: Implications for Transport Policy.''}
-\emph{Growth and Change} 43 (4): 667--96.
-
-\bibitem[\citeproctext]{ref-ravensteinLawsMigration1885}
-Ravenstein, E. G. 1885. {``The Laws of Migration Paper 1.''}
-\emph{Journal of the Royal Statistical Society} 48 (2): 167--227.
-
-\bibitem[\citeproctext]{ref-ravensteinLawsMigration1889}
----------. 1889. {``The Laws of Migration Paper 2.''} \emph{Journal of
-the Royal Statistical Society} 52 (2): 241--305.
-\url{https://doi.org/10.2307/2979333}.
-
-\bibitem[\citeproctext]{ref-reggianiGuestEditorialNew2011}
-Reggiani, Aura, and Juan Carlos Martín. 2011. {``Guest Editorial: New
-Frontiers in Accessibility Modelling: An Introduction.''} \emph{Networks
-and Spatial Economics} 11 (4): 577--80.
-\url{https://doi.org/10.1007/s11067-011-9155-x}.
-
-\bibitem[\citeproctext]{ref-reilly1929methods}
-Reilly, William J. 1929. {``Methods for the Study of Retail
-Relationships.''} 2944.
-
-\bibitem[\citeproctext]{ref-reyesAccessibility2014}
-Reyes, M., A. Paez, and C. Morency. 2014. {``Walking Accessibility to
-Urban Parks by Children: {A} Case Study of {Montreal}.''}
-\emph{Landscape and Urban Planning} 125 (May): 38--47.
-\url{https://doi.org/10.1016/j.landurbplan.2014.02.002}.
-
-\bibitem[\citeproctext]{ref-ribeiro_road_2010}
-Ribeiro, A., A. P. Antunes, and A. Páez. 2010. {``Road Accessibility and
-Cohesion in Lagging Regions: {Empirical} Evidence from {Portugal} Based
-on Spatial Econometric Models.''} \emph{Journal of Transport Geography}
-18 (1): 125--32.
-
-\bibitem[\citeproctext]{ref-roblot2021participation}
-Roblot, Mathilde, Geneviève Boisjoly, Ciari Francesco, and Trépanier
-Martin. 2021. {``Participation in Shared Mobility: An Analysis of the
-Influence of Walking and Public Transport Accessibility to Vehicles on
-Carsharing Membership in Montreal, Canada.''} \emph{Transportation
-Research Record} 2675 (12): 1160--71.
-
-\bibitem[\citeproctext]{ref-rojas_accessibility_2016}
-Rojas, C., A. Paez, O. Barbosa, and J. Carrasco. 2016. {``Accessibility
-to Urban Green Spaces in {Chilean} Cities Using Adaptive Thresholds.''}
-\emph{Journal of Transport Geography} 57 (December): 227--40.
-\url{https://doi.org/10.1016/j.jtrangeo.2016.10.012}.
-
-\bibitem[\citeproctext]{ref-romanillosAccessibilitySchoolsSpatial2018}
-Romanillos, G., and J. C. Garcia-Palomares. 2018. {``Accessibility to
-{Schools}: {Spatial} and {Social} {Imbalances} and the {Impact} of
-{Population} {Density} in {Four} {European} {Cities}.''} \emph{Journal
-of Urban Planning and Development} 144 (4).
-\url{https://doi.org/10.1061/(asce)up.1943-5444.0000491}.
-
-\bibitem[\citeproctext]{ref-sahebgharani2019computing}
-Sahebgharani, Alireza, Mahmoud Mohammadi, and Hossein Haghshenas. 2019.
-{``Computing Spatiotemporal Accessibility to Urban Opportunities: A
-Reliable Space-Time Prism Approach in Uncertain Urban Networks.''}
-\emph{Computation} 7 (3): 51.
-
-\bibitem[\citeproctext]{ref-SchuurmanMeasuring2010}
-Schuurman, N., M. Berube, and V. A. Crooks. 2010. {``Measuring Potential
-Spatial Access to Primary Health Care Physicians Using a Modified
-Gravity Model.''} \emph{Canadian Geographer-Geographe Canadien} 54 (1):
-29--45.
-
-\bibitem[\citeproctext]{ref-seniorGravityModellingEntropy1979}
-Senior, Martyn L. 1979. {``From Gravity Modelling to Entropy Maximizing:
-A Pedagogic Guide.''} \emph{Progress in Human Geography} 3 (2):
-175--210. \url{https://doi.org/10.1177/030913257900300218}.
-
-\bibitem[\citeproctext]{ref-sharifiasl2023incorporating}
-Sharifiasl, Seyedsoheil, Subham Kharel, and Qisheng Pan. 2023.
-{``Incorporating Job Competition and Matching to an Indicator-Based
-Transportation Equity Analysis for Auto and Transit in Dallas-Fort Worth
-Area.''} \emph{Transportation Research Record} 2677 (12): 240--54.
-
-\bibitem[\citeproctext]{ref-shen1998}
-Shen, Q. 1998b. {``Location Characteristics of Inner-City Neighborhoods
-and Employment Accessibility of Low-Wage Workers.''} \emph{Environment
-and Planning B: Planning and Design} 25 (3): 345--65.
-\url{https://doi.org/10.1068/b250345}.
-
-\bibitem[\citeproctext]{ref-shenLocationCharacteristicsInnercity1998}
----------. 1998a. {``Location Characteristics of Inner-City
-Neighborhoods and Employment Accessibility of Low-Wage Workers.''}
-\emph{Environment and Planning B: Planning and Design} 25 (3): 345--65.
-\url{https://doi.org/10.1068/b250345}.
-
-\bibitem[\citeproctext]{ref-shen2019segregation}
-Shen, Yao. 2019. {``Segregation Through Space: A Scope of the Flow-Based
-Spatial Interaction Model.''} \emph{Journal of Transport Geography} 76:
-10--23.
-
-\bibitem[\citeproctext]{ref-silvaAccessibilityInstrumentsPlanning2017}
-Silva, Cecília, Luca Bertolini, Marco Te Brömmelstroet, Dimitris
-Milakis, and Enrica Papa. 2017. {``Accessibility Instruments in Planning
-Practice: Bridging the Implementation Gap.''} \emph{Transport Policy} 53
-(January): 135--45. \url{https://doi.org/10.1016/j.tranpol.2016.09.006}.
-
-\bibitem[\citeproctext]{ref-soukhovIntroducingSpatialAvailability2023}
-Soukhov, Anastasia, Antonio Paez, Christopher D. Higgins, and Moataz
-Mohamed. 2023. {``Introducing Spatial Availability, a Singly-Constrained
-Measure of Competitive Accessibility {\textbar} {PLOS} {ONE}.''}
-\emph{{PLOS} {ONE}}, 1--30.
-https://doi.org/\href{https://\%20doi.org/10.1371/journal.pone.0278468}{https://
-doi.org/10.1371/journal.pone.0278468}.
-
-\bibitem[\citeproctext]{ref-soukhovMultimodalSpatialAvailability2024}
-Soukhov, Anastasia, Javier Tarriño-Ortiz, Julio A. Soria-Lara, and
-Antonio Páez. 2024. {``Multimodal Spatial Availability: A
-Singly-Constrained Measure of Accessibility Considering Multiple
-Modes.''} \emph{{PLOS} {ONE}} 19 (2): e0299077.
-\url{https://doi.org/10.1371/journal.pone.0299077}.
-
-\bibitem[\citeproctext]{ref-stewartPrinciples1947}
-Stewart, John Q. 1947. {``Suggested {Principles} of ''{Social}
-{Physics}''.''} \emph{Science} 106 (2748): 179--80.
-
-\bibitem[\citeproctext]{ref-stewartDemographicGravitationEvidence1948}
----------. 1948. {``Demographic Gravitation: Evidence and
-Applications.''} \emph{Sociometry} 11 (1): 31--58.
-\url{https://doi.org/10.2307/2785468}.
-
-\bibitem[\citeproctext]{ref-su2023untangling}
-Su, Rongxiang, and Konstadinos Goulias. 2023. {``Untangling the
-Relationships Among Residential Environment, Destination Choice, and
-Daily Walk Accessibility.''} \emph{Journal of Transport Geography} 109:
-103595.
-
-\bibitem[\citeproctext]{ref-suel2024measuring}
-Suel, Esra, Claude Lynch, Maria Wood, Tom Murat, Gerard Casey, and Adam
-Dennett. 2024. {``Measuring Transport-Associated Urban Inequalities:
-Where Are We and Where Do We Go from Here?''} \emph{Transport Reviews}
-44 (6): 1235--57.
-
-\bibitem[\citeproctext]{ref-tao_investigating_2020}
-Tao, ZL, JP Zhou, XB Lin, H Chao, and GC Li. 2020. {``Investigating the
-Impacts of Public Transport on Job Accessibility in {Shenzhen}, {China}:
-A Multi-Modal Approach.''} \emph{LAND USE POLICY} 99 (December).
-\url{https://doi.org/10.1016/j.landusepol.2020.105025}.
-
-\bibitem[\citeproctext]{ref-tong2015transportation}
-Tong, Lu, Xuesong Zhou, and Harvey J Miller. 2015. {``Transportation
-Network Design for Maximizing Space--Time Accessibility.''}
-\emph{Transportation Research Part B: Methodological} 81: 555--76.
-
-\bibitem[\citeproctext]{ref-turk2019socio}
-Türk, Umut. 2019. {``Socio-Economic Determinants of Student Mobility and
-Inequality of Access to Higher Education in Italy.''} \emph{Networks and
-Spatial Economics} 19 (1): 125--48.
-
-\bibitem[\citeproctext]{ref-vanweeAccessible2016}
-van Wee, Bert. 2016. {``Accessible Accessibility Research Challenges.''}
-\emph{Journal of Transport Geography} 51 (February): 9--16.
-https://doi.org/\url{https://doi.org/10.1016/j.jtrangeo.2015.10.018}.
-
-\bibitem[\citeproctext]{ref-vickermanAccessibilityAttractionPotential1974}
-Vickerman, R W. 1974. {``Accessibility, Attraction, and Potential: A
-Review of Some Concepts and Their Use in Determining Mobility.''}
-\emph{Environment and Planning A} 6: 675--91.
-\url{https://doi.org/10.1068/a060675}.
-
-\bibitem[\citeproctext]{ref-vickermanAccessibility1999}
-Vickerman, R., K. Spiekermann, and M. Wegener. 1999. {``Accessibility
-and Economic Development in {Europe}.''} \emph{Regional Studies} 33 (1):
-1--15.
-
-\bibitem[\citeproctext]{ref-wachs_physical_1973}
-Wachs, Martin, and T.Gordon Kumagai. 1973. {``Physical Accessibility as
-a Social Indicator.''} \emph{Socio-Economic Planning Sciences} 7 (5):
-437--56. \url{https://doi.org/10.1016/0038-0121(73)90041-4}.
-
-\bibitem[\citeproctext]{ref-wan2012three}
-Wan, N., B. Zou, and T. Sternberg. 2012. {``A Three-Step Floating
-Catchment Area Method for Analyzing Spatial Access to Health
-Services.''} Journal Article. \emph{International Journal of
-Geographical Information Science} 26 (6): 1073--89.
-\url{https://doi.org/10.1080/13658816.2011.624987}.
-
-\bibitem[\citeproctext]{ref-wang_inverted_2018}
-Wang, F. H. 2018. {``Inverted {Two}-{Step} {Floating} {Catchment} {Area}
-{Method} for {Measuring} {Facility} {Crowdedness}.''} \emph{Professional
-Geographer} 70 (2): 251--60.
-\url{https://doi.org/10.1080/00330124.2017.1365308}.
-
-\bibitem[\citeproctext]{ref-wang_2sfca_2021}
-Wang, Fahui. 2021. {``From {2SFCA} to {i2SFCA}: Integration, Derivation
-and Validation.''} \emph{International Journal of Geographical
-Information Science} 35 (3): 628--38.
-\url{https://doi.org/10.1080/13658816.2020.1811868}.
-
-\bibitem[\citeproctext]{ref-weibullNumericalMeasurementAccessibility1980}
-Weibull, J W. 1980. {``On the Numerical Measurement of Accessibility.''}
-\emph{Environment and Planning A: Economy and Space} 12 (1): 53--67.
-\url{https://doi.org/10.1068/a120053}.
-
-\bibitem[\citeproctext]{ref-weinerUrbanTransportationPlanning2016}
-Weiner, Edward. 2016. \emph{Urban Transportation Planning in the United
-States}. Springer Cham: Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-319-39975-1}.
-
-\bibitem[\citeproctext]{ref-whoMedicalDoctors102025}
-WHO. 2025. {``Medical Doctors (Per 10 000 Population).''} 2025.
-\url{https://www.who.int/data/gho/data/indicators/indicator-details/GHO/medical-doctors-(per-10-000-population)}.
-
-\bibitem[\citeproctext]{ref-williams_disparities_2014}
-Williams, S., and F. H. Wang. 2014. {``Disparities in Accessibility of
-Public High Schools, in Metropolitan {Baton} {Rouge}, {Louisiana}
-1990-2010.''} \emph{Urban Geography} 35 (7): 1066--83.
-\url{https://doi.org/10.1080/02723638.2014.936668}.
-
-\bibitem[\citeproctext]{ref-willigers2007accessibility}
-Willigers, Jasper, Han Floor, and Bert van Wee. 2007. {``Accessibility
-Indicators for Location Choices of Offices: An Application to the
-Intraregional Distributive Effects of High-Speed Rail in the
-Netherlands.''} \emph{Environment and Planning A} 39 (9): 2086--2898.
-
-\bibitem[\citeproctext]{ref-wilsonSTATISTICALTHEORYSPATIAL1967}
-Wilson, A. G. 1967. {``A {STATISTICAL} {THEORY} {OF} {SPATIAL}
-{DISTRIBUTION} {MODELS}.''} \emph{Transportation Research} 1: 253--69.
-\url{https://journals-scholarsportal-info.libaccess.lib.mcmaster.ca/pdf/00411647/v01i0003/253_astosdm.xml_en}.
-
-\bibitem[\citeproctext]{ref-wilson1971}
----------. 1971. {``A Family of Spatial Interaction Models, and
-Associated Developments.''} \emph{Environment and Planning A: Economy
-and Space} 3 (1): 1--32. \url{https://doi.org/10.1068/a030001}.
-
-\bibitem[\citeproctext]{ref-wuUnifyingAccess2020}
-Wu, Hao, and David Levinson. 2020. {``Unifying Access.''}
-\emph{Transportation Research Part D: Transport and Environment} 83
-(June): 102355. \url{https://doi.org/10.1016/j.trd.2020.102355}.
-
-\bibitem[\citeproctext]{ref-ye_spatial_2018}
-Ye, Changdong, Yushu Zhu, Jiangxue Yang, and Qiang Fu. 2018. {``Spatial
-Equity in Accessing Secondary Education: {Evidence} from a Gravity-Based
-Model: {Spatial} Equity in Accessing Secondary Education.''} \emph{The
-Canadian Geographer / Le Géographe Canadien} 62 (4): 452--69.
-\url{https://doi.org/10.1111/cag.12482}.
-
-\bibitem[\citeproctext]{ref-zipfDeterminantsCirculationInformation1946}
-Zipf, George Kingsley. 1946a. {``Some Determinants of the Circulation of
-Information.''} \emph{The American Journal of Psychology} 59 (3):
-401--21. \url{https://doi.org/10.2307/1417611}.
-
-\bibitem[\citeproctext]{ref-zipfHypothesisIntercityMovement1946}
----------. 1946b. {``The \emph{p} \(_{\textrm{1}}\) \emph{p}
-\(_{\textrm{2}}\) / \emph{d} Hypothesis: On the Intercity Movement of
-Persons''} 11 (6): 677--86.
-
-\bibitem[\citeproctext]{ref-zipfHypothesisCaseRailway1946}
----------. 1946c. {``The \emph{p} \(_{\textrm{1}}\) \emph{p}
-\(_{\textrm{2}}\) / \emph{d} Hypothesis: The Case of Railway Express.''}
-\emph{The Journal of Psychology} 22 (1): 3--8.
-\url{https://doi.org/10.1080/00223980.1946.9917292}.
-
-\end{CSLReferences}
+In summary, building on Wilson's \citep{wilson1971} foundational work,
+this paper proposed a unified framework for accessibility that is able
+to account for competition. By reintroducing Wilson's proportionality
+constant, the proposed family of constrained accessibility measures
+restores measurement units to accessibility estimates. This enhancement
+provides a more interpretable, consistent, and theoretically grounded
+basis for accessibility analysis, which could help advance the adoption
+of accessibility-oriented planning.
+
+\nolinenumbers
+\renewcommand\refname{References}
+  \bibliography{bibliography.bib}
 

### 2025-08-01 paper_plos/reviewer_respose_letter_V1.qmd
Made the existing draft as concise as possible -- shed 6 pages. Old version is paper_plos_V0.qmd and .pdf.

---
title: Response to reviewer
---

Thank you for your responses. Below are replies, with our response in black font and the reviewers comments in \textcolor{blue}{blue}.

# Review 1

\textcolor{blue}{Thanks for the opportunity to review a paper that handles an interesting topic. I think the major contribution of the paper lies in 1) combining the concept of accessibility and spatial interaction modeling and 2) classifying the numerous accessibility measures into four categories based on the constraints. While the topic is really interesting and handles the important topic, I would like to ask the authors to revise the paper due to the following reasons.}

\textcolor{blue}{Most importantly, the manuscript is too lengthy and verbose. I am having a hard time evaluating the manuscript since the length is about 30 pages. Given that the typical length of the paper is about 10 pages, I would like to request a summary of the key points for better deliverability.}

\textcolor{blue}{Related to the first issue, I am not sure about the point that the authors explain the details of Newtonian’s roots and early research on human spatial interactions from Ravenstein to Stewart. Of course, I think mentioning those contents is beneficial, but it is hard to get the key point given that it is verbose.}

\textcolor{blue}{Third, I believe the main contribution of this paper lies in classifying the accessibility measures into four groups. However, I also had a hard time finding the definition of those four groups. Given that the research articles aim to reveal previously unknown facts, the literacy level of this manuscript is relatively low. Additionally, the introduction of too many equations makes it difficult to establish connections between them. Therefore, I recommend that the authors revise and reconstruct the manuscript.}

THank you for these comments. We've heavily revised the manuscript, focusing now on explaining things concisely and putting forward the focus. We did this by:

- reducing the length of the manuscript, it is now 10 pages
- removed tangental details (e.g., )
- streamlined the historical section (e.g., )
- highlight the main contribution of the paper, namely how it provides constrained definitions to accessibility
- reduced the number of equations where possible, and make it clear why those that remain are there.

# Reviewer #2

\textcolor{blue}{This manuscript presents a theoretically motivated and conceptually rich framework that bridges spatial interaction modeling and accessibility measures. The authors introduce a family of accessibility metrics grounded in Wilson’s spatial interaction models and argue convincingly for the reintroduction of interpretability and measurement units in accessibility analysis. The historical tracing of accessibility theory development and the synthesis of divergent literature streams are timely and valuable contributions to the field.}

\textcolor{blue}{1. While the authors provide a detailed numerical example, the manuscript lacks any empirical case study or application to real-world accessibility data. This limits the reader’s ability to evaluate the practical utility and robustness of the proposed framework. The authors can draw inspiration from several studies such as "Factoring in temporal variations of public transit-based healthcare accessibility and equity" to better understand this concept.}

\textcolor{blue}{2. The manuscript frequently shifts between concepts such as "potential for spatial interaction," "accessibility," and "access," especially in the discussion of the doubly constrained model. However, the terminological distinction remains conceptually blurry.}

\textcolor{blue}{3. The historical literature review is valuable but disproportionately long. Pages are devoted to recapping classical gravitational thinking, while the more innovative contributions of the paper (e.g., balancing factor reinterpretation) are condensed. The authors should also focus on more recent publications such as "Evaluating temporal variations in access to multi-tier hospitals using personal vehicles and public transit: Implications for healthcare equity" instead of older ones.}

\textcolor{blue}{4. The paper pays limited attention to recent methodological developments such as time-sensitive, network-based, or multi-modal accessibility models, as well as behavioral or utility-based approaches.}

\textcolor{blue}{5. The authors argue that the constrained models restore interpretability by attaching meaningful units (e.g., number of people or opportunities). However, this is not substantiated with user- or practitioner-oriented tests.}

\textcolor{blue}{6. Some sections suffer from repetition and overly technical language without summarizing takeaways (e.g., Table 1 is helpful, but the text before and after it repeats much of the same information).}

### 2025-08-05 paper_plos/paper_plos.qmd
made some additional, smaller, cuts. still 37 pages.

@@
 Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024].
-While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in @vanweeAccessible2016: How should practitioners interpret difference in scores between modes, and how should results be communicated?
+While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret difference in scores between modes, and how should results be communicated?
 
@@
 
-We contend that accessibility research should reengage with spatial interaction modeling, particularly the use of Wilson's system constraints.
+We contend that accessibility research should re-engage with spatial interaction modeling, particularly the use of Wilson's system constraints.
 While spatial interaction models embraced such constraints to improve interpretability, most accessibility models have not.
-This lack of adoption contributes to fuzziness in current analyses, limiting interpretive clarity to simple proportional comparisons (e.g., “higher-than,” “lower-than”) [@millerAccessibilityMeasurementApplication2018].
+This lack of adoption contributes to fuzziness in current analyses, limiting interpretive clarity to simple proportional comparisons (e.g., “higher-than”, “lower-than”) [@millerAccessibilityMeasurementApplication2018].
 Without such constraints, accessibility scores lack clear units and comparability across cities or modes.
-In contrast, constrained measures yield values that can be tied to tangible values without any post-hoc treatment, making them more interpretable, communicable, and actionable in planning.
-
-This paper lays the historical and theoretical groundwork for a companion empirical study demonstrating how these constrained forms restore interpretability through concrete measurement units (e.g., number of accessible opportunities), as opposed to the abstract weighted outputs typical of conventional gravity models.
+In contrast, constrained measures yield values that can be tied to tangible values without any post-hoc treatment, theoretically making them more interpretable, communicable, and actionable in planning.
 
 Accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
-Following this introduction, this paper revisits that lineage--from Ravenstein’s Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959], to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
+Following this introduction, this paper revisits that lineage--from Ravenstein’s Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959] to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
 We illustrate each case of the family with a numerical toy example and conclude by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.
@@
 
-The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half.
-From as far back as Henry C.
-Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], a concern with the scientific study of human spatial interaction can be observed.
+The patterns of people's movement in space have been a subject of scientific inquiry for at least a century and a half, from as far back as Henry C.
+Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858].
 It was in this work where Carey stated that "man \[is\] the molecule of society \[and their interaction is subject to\] the direct ratio of the mass and the inverse one of distance" [@mckeanManual1883, pp.
@@
 As mass increases, so does force; as distance increases, force decreases.
-However, @eq-phys-grav-prop only demonstrates a proportional relationship.
-To quantify the magnitude (not the proportional magnitude) of the force, it must be _constrained_ with an empirical constant.
+However, @eq-phys-grav-prop only demonstrates a _proportional_ relationship.
+To quantify the magnitude (not the _proportional_ magnitude) of the force, it must be _constrained_ with an empirical constant.
 This constant $G$ converts @eq-phys-grav-prop from an expression of proportionality to the following expression of equality:
@@
 Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p.
-783].
-Still, it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant to within 1% accuracy, with Cavendish's 1798 experiment [@cavendish_xxi_1798].
-
-This gravitational framing laid the conceptual groundwork for later empirical studies of human spatial interaction, such as those by Ravenstein, Stewart, and others, who began translating these principles into observed patterns of human movement.
+783] and Cavendish's 1798 experiment [@cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy.
+In sum: it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant.
+
+The Newtonian gravitational relationship laid the conceptual groundwork for later empirical studies of human spatial interaction, such as those by Ravenstein, Stewart, and others, who began translating these principles into observed patterns of human movement.
+However, the majority of these attempts described a proportional relationship, or one arbitrarily set to equality.
 
@@
 These early formulations (@eq-reilly, @eq-zipf) clearly reflect the influence of Newtonian gravity on human spatial interaction theory, revealing a shared mathematical structure across migration, trade, and communication models.
-However, none included a proportionality constant--highlighting a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities
-
-After Carey and folowing Reilly and Zipf, it is only in @stewartDemographicGravitationEvidence1948 that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
+However, none included a proportionality constant--highlighting a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities.
+
+After Carey and folowing Reilly and Zipf, it is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
 <!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
@@
 \noindent Where $M_j$ is the demographic mass at location $j$ and $d_{ij}$ is the distance between $i$ and $j$.
-A version of this discrete form is what was used in @hansen1959, going on to become a foundation of modern accessibility definitions, as will be discussed.
+A version of this discrete form is what was used in Hansen [@hansen1959], going on to become a foundation of modern accessibility definitions, as will be discussed.
 
@@
 34 and then later as _demographic energy_ on p.
-53.) and its foundation in racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight $\mu$ of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
+53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight $\mu$ of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
 35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration.
-In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., people^2-distance^-2) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).
-
-As will be discussed, when Hansen later adopted Stewart’s formula (@eq-stewart-population-potential-sum) for accessibility, he omitted any mention of $G$, effectively setting it to 1 arbitrarily as Stewart had done.
+In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).
+
+As will be discussed, when Hansen [@hansen1959] later adopted Stewart’s formula (@eq-stewart-population-potential-sum) for accessibility, he omitted any mention of $G$, effectively setting it to 1 arbitrarily as Stewart had done.
 This omission has persisted in accessibility research, leaving a conceptual gap in how such measures are interpreted and compared.
@@
 As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-@hansen1959 drew directly from Stewart’s population potential (see @eq-stewart-population-potential-sum), but left aside the broader (and often problematic) framework of “social physics.”.
+Hansen [@hansen1959] drew directly from Stewart’s population potential formula (see @eq-stewart-population-potential-sum), but left aside the broader (and often problematic) framework of “social physics”.
 
 Hansen recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more accurately, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting to accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$ as follows:
+Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more specifically, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting to accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$ as follows:
 
@@
 
-However, the omission of a proportionality constants and working with a proportional relationship generates fundamental issues in comparability between and, arguably within, studies.
-As the absence of $G$ means accessibility estimates have no fixed unit--this renders them sensitive to the choice of impedance functions.
-For instance, if travel cost $d_{ij}$ is measured in meters, then when the travel impedance function $f(d_{ij})$ equals $d_{ij}^{-\beta}$, the resulting $S_i$ has units of opportunities per $\text{metres}^{\beta}$; but when $f(d_{ij})$ is set to equal $e^{-\beta d_{ij}}$, the units become opportunities per $e^{\beta \text{metres}}$.
-Such variation impairs comparability across analysis and obscures the meaning of accessibility scores renders their results difficult to understand and communicate without post-hoc treatment.
-
-Therefore in practice, Hansen-type measures (of proportionality) are better understood as *ordinal indicators*--they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
-The continued absence of $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement.
+Furthermore, working with a proportional relationship generates fundamental issues in comparability between and, arguably within, studies.
+Namely, accessibility estimates have no fixed unit, rendering them sensitive to the choice of impedance functions.
+For instance, if travel cost $d_{ij}$ is measured in meters, then when the travel impedance function $f(d_{ij})$ equals $d_{ij}^{-\beta}$, the resulting $S_i$ has units of opportunities per $\text{metres}^{\beta}$.
+However, when $f(d_{ij})$ is set to equal $e^{-\beta d_{ij}}$, the units become opportunities per $e^{\beta \text{metres}}$.
+Such variation impairs comparability across analysis and obscures the meaning of accessibility scores, making them difficult to understand and communicate without post-hoc treatment.
+
+Therefore in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
+The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement.
 
 # Wilson's family of spatial interaction models
-While accessibility research evolved in North America with @hansen1959, a parallel development was taking place across the Atlantic with Alan G.
+While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G.
 Wilson.
-In his groundbreaking study [@wilson1971], advanced a general framework for spatial interaction modeling--focused not on potential (as accessibility), but on modeling observed flows of interaction between places.
-@wilson1971 formalized the general spatial interaction model through the following equation:
+Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling--focused on modeling flows of interaction between places.
+This work was not focused on the 'potential' concept as accessibility has been.
+Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:
 
@@
 $f(c_{ij})$ is some function of travel cost $c_{ij}$ which reflects travel impedance.
-Finally, $k$ acts as a proportionality constant, shifting the equation from a proportional to an equal relationship by incorporating known system totals.
-In this specific sense, we infer that $k$ serves a role similar to the gravitational constant in Newton’s law--it calibrates the model so that outputs match real-world quantities.
+
+Of important note for this paper, $k$ in @eq-phys-gravity-model acts as a proportionality constant, shifting the equation from a proportional to an equal relationship by incorporating known system totals.
+In this sense, we infer that $k$ serves a role similar to the gravitational constant in Newton’s law--it calibrates the model so that outputs match real-world quantities.
+However, these real-world quantities was not a set empirical constant (like Newton's $G$) but instead, sensitive to the system and known information about the system.
 
@@
 Additional information can be introduced.
-For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ is represented as $O_i$ and the following constraint can be used:
+For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ in @eq-phys-gravity-model is represented as $O_i$ and the following constraint can be used:
 
@@
 
-It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints could be imposed on the model at once.
-
-Depending on which system constraints are applied—fully, partially, or not at all—a family of spatial interaction models can be derived from @eq-phys-gravity-model.
+It is also possible to have information about both $O_i$ and $D_j$, in which case both constraints can be imposed on the model at once.
+
+Depending on which of the three system constraints are applied a family of spatial interaction models can be derived from @eq-phys-gravity-model.
 The proportionality constant $k$ is replaced with different _balancing factors_.
@@
 In other words, the different balancing factors adjust the model so that flows become statistical averages consistent with observed origin and/or destination data.
-In the framework introduced and inferred from @wilson1971, three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.
+In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.
 
@@
 In practical terms, this means that the total number of flows predicted by the model must be equal to sum of all flows from origins $i$ to destinations $j$.
-The balancing factor $K$ takes the place of $k$ and is equal to the following [as specified in @cliff_evaluating_1974 and @fotheringham_spatial_1984]:
+The balancing factor $K$ takes the place of $k$ and is equal to the following (as specified in [@cliff_evaluating_1974] and [@fotheringham_spatial_1984]):
 
@@
 In the singly-constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
-When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the following production-constrained singly-constrained model:
-
-$$
-T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
-$$ {#eq-production-constrained-gravitymodel}
-
-Notice how, in this model, the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$, representing what we know about the system, the spatial interaction outbound flow, i.e., outbound flows produced at $i$.
-Also, there is no longer a single system-wide balancing factor $K$, but rather a set of balancing factors specific to origins $A_i$.
-
-For this model, the balancing factors ensure that constraint @eq-constraint1-gravitymodel is satisfied, meaning that the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$ e.g., the total number of outbound flows.
+When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the production-constrained singly-constrained version of @eq-phys-gravity-model, where the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$.
+Also, $k$ is replaced with a set of balancing factors specific to origins $A_i$ which ensures that constraint @eq-constraint1-gravitymodel is satisfied (i.e., the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$).
 Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of flows.
-
-The production-constrained singly-constrained model is useful when flows ends are unknown but the number of flows originating from each location is known.
-The balancing factors for the production-constrained model are solved for each origin $A_i$, and according to Wilson are:
+$A_i$ takes the following form:
 
@@
 
-The attraction-constrained model is similar to the production-constrained model as it is also singly constrained but from the perspective of the mass at the destination.
-From the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is now replaced with $D_j$, representing the spatial interaction inbound flow e.g., flows attracted to the destination, and takes the following form:
-
-$$
-T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
-$$ {#eq-attraction-constrained-gravitymodel}
-
-In this model (@eq-attraction-constrained-gravitymodel), the balancing factors ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$ i.e., the total number of inbound flows to $j$.
-This should hold for all destinations.
+The singly-constrained attraction-constrained model is similar to the production-constrained version but from the perspective of the mass at the destination.
+For the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is replaced with $D_j$ in @eq-phys-gravity-model, representing the spatial interaction inbound flow.
+Also, $k$ is replaced with a set of destination-specific balancing factors $B_j$ that ensure that @eq-constraint2-gravitymodel is satisfied (hence the total constraint @eq-constraint0-gravitymodel is as well), meaning that the sum of predicted flows going to one destination from all origins must equal the known mass of that destination $D_j$.
 As before, destination-specific balancing factors $B_j$ were derived by Wilson as:
@@
 
-Lastly, the doubly-constrained model is the production-attraction constrained model in @wilson1971.
+Lastly, the doubly-constrained model is the production-attraction constrained model in Wilson [@wilson1971].
 In this case, both constraints @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
@@
 This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly constrained, and takes the following form:
-
-$$
-T_{ij} = A_i B_j O_i D_j f(c_{ij})
-$$ {#eq-doubly-constrained-gravitymodel}
-
-In this model, both proxies for the masses are replaced with the known masses, that is, the flows produced by the origin and the flows attracted by the destination.
-There are now two sets of mutually dependent proportionality constants shown in @eq-doubly-constrained-balancing-factors.
+The resulting model is, in Wilson's terms, doubly constrained, and from @eq-phys-gravity-model, $k$ becomes both $A_i$ and $B_j$ shown in @eq-doubly-constrained-balancing-factors, and $W_i^{(1)}$ and $W_j^{(2)}$ is replaced with $O_i$ and $D_j$.
 Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
@@
 Instead, he grounded his models in statistical mechanics, particularly the principle of entropy maximisation.
-For accessibility research, Wilson’s approach is especially significant because it can address a problem that has quietly plagued this field (as this work has argued): the transition from proportionality to equality.
+For accessibility research, Wilson’s approach is especially significant because it can address a problem that has quietly afflicted this field (as this paper has argued): the transition from proportionality to equality.
 
@@
 This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-In short, when aspects of the systems are inputs into the balacing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled.
+In short, when aspects of the systems are inputs into the balancing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled.
 In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
@@
 But before demonstrating a family of accessibility measures that are sensitive to constraints, in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
-This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this work.
+This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper.
 
@@
 This divergence is especially striking given the context in which both frameworks emerged.
-As noted in @battyChronicleScientificPlanning1994, large scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
+As noted in [@battyChronicleScientificPlanning1994], large scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
 And these needs were far from trivial: in the U.S., for instance, the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today’s dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
@@
 
-Thirdly, there is a group of accessibility-focused works that interprets the measure used in @hansen1959 as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
+Thirdly, there is a group of accessibility-focused works that interprets the measure used in Hansen [@hansen1959] as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
 This group often cites the spatial interaction works that make this connection (i.e., the first subset of the first stream of literature) and is especially prominent in the investigation of competitive accessibility topics e.g., [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
@@
 
-In sum, despite the interpretative advantaged offered by the statistical logic of @wilson1971's framework, the spatial interaction-oriented @hansen1959 citing literature nor the accessibility-oriented literature that cite @wilson1971 have yet to meaningfully adopt Wilson's constraint-based logic to the concept of accessibility.
+In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, the spatial interaction-oriented Hansen [@hansen1959] citing literature nor the accessibility-oriented literature that cite Wilson [@wilson1971] have yet to meaningfully adopt Wilson's constraint-based logic to the concept of accessibility.
 So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
 This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
-This approach takes the same path of entropy-maximisation as in @wilson1971, and does not rely on specifying some universal constant $G$ like initially proposed in @stewartDemographicGravitationEvidence1948 (recall, @eq-stewart-population-potential-sum which @hansen1959 operationalised).
+This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially proposed in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).
 
@@
 Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter.
-This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la @hansen1959) from a relationship describing proportionality to equality.
+This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
 
 This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
-Rather than relying on arbitrary or uncalibrated proportional relationships--or invoking a universal constant $G$ as proposed in @stewartDemographicGravitationEvidence1948 --we adopt Wilson’s approach grounded in statistical mechanics.
-
-<!-- fix this one-->
-To do so, we propose a revised definition of accessibility considerate of the constraint-based spatial intereaction model: the potential for spatial interaction with opportunities or population.We can bring $k$ as a proportional allocation factor which encorporates Wilson's balancing factor, to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potentnial for spatial intereaction with population_ $M_{ji}$.
+Rather than relying on arbitrary or uncalibrated proportional relationships (or invoking a universal constant $G$ as proposed in Stewart [@stewartDemographicGravitationEvidence1948]) we adopt Wilson’s approach grounded in statistical mechanics.
+
+To do so, we propose a revised definition of accessibility considerate of the constraint-based spatial interaction model: the potential for spatial interaction with opportunities (or population).
+We can specify $k$ as a type of proportional allocation factor $\kappa$ which incorporates Wilson's balancing factor(s) to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potential for spatial interaction with population_ $M_{ji}$.
 In effect, $\kappa$ is unitless and proportionally allocates (based on the constraints of the case) opportunities (for $V_{ij}$) and population (for $M_{ji}$).
@@
 
-\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ is the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case.
+\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ is the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case (hence represented by a $X$).
+
 Accessibility flows can also be summarised as as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively, as often done in accessibility research:
@@
 4. **Doubly Constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
-- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously; equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
+- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously, which can also be expressed as proportional allocation factors ($\kappa_{ij}^D$, $\hat \kappa_{ji}^D$); equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
 - Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in @fig-analytical-device-conc-accessibility).
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
-Represents ‘interaction capacity’ or ‘realized access,’ serving as predictions of real interaction flows.
+Represents ‘interaction capacity’ or ‘realized access’ serving as predictions of real interaction flows.
 
@@
 ## Unconstrained accessibility
-<!-- fix-->
+
 In $V^0_i$, no proportional allocation factor is defined, simply $f(c_{ij})$ is used to weight the number of opportunities at each $j$ and the weighted values for each $j$ are summed for each $i$, yielding the an expression identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
@@
 
-The total constrained accessibility measure resolves the interpretability issue of Hansen’s accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total--enabling robust comparisons across zones and scenarios--but also by keeping values proportional to $V_i^0$ so interpretation is similar.
+The total constrained accessibility measure resolves the interpretability issue of Hansen’s accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total, enabling robust comparisons across zones and scenarios but also by keeping values proportional to $V_i^0$ so interpretation is similar.
 
@@
 
-The underlying spatial interaction model is now the attraction-constrained model in @eq-attraction-constrained-gravitymodel, and our accessibility measure becomes:
+The underlying spatial interaction model is now the attraction-constrained model and our accessibility measure becomes:
 
@@
 
-@soukhovIntroducingSpatialAvailability2023 also showed that the following expression (accessibility per capita) is a constrained version of the popular 2SFCA approach of Shen [@shen1998] and Luo and Wang [@luo2003]:
+Soukhov et al., 2023 [@soukhovIntroducingSpatialAvailability2023] also showed that the following expression (accessibility per capita) is a constrained version of the popular 2SFCA approach of Shen 1998 [@shen1998] and Luo and Wang 2003 [@luo2003]:
 
@@
 
-The underlying spatial interaction model is now the production-constrained model in @eq-production-constrained-gravitymodel, and our market potential measure $M^S_j$ becomes:
+The underlying spatial interaction model is now the production-constrained model version of @eq-phys-gravity-model, and our market potential measure $M^S_j$ becomes:
 
@@
 
-Because both margins are fixed, this measure reflects fully allocated interaction flows, not latent potential.
-As such, accessibility cannot be interpreted as a traditional summary at $i$ or $j$ (e.g., "opportunities accessible from $i$")--those sums simply reproduce the constraint totals.
+Because results are made to match both margins, the results cannot be interpreted as a traditional summary at $i$ or $j$ (e.g., "opportunities accessible from $i$")--those sums simply reproduce the constraint totals.
 Instead, the meaningful unit of analysis is the $ij$ flow itself.
@@
 This distinguishes doubly constrained accessibility from the total and singly constrained cases discussed previously.
-In those cases, only one side of the interaction--either the total marginal or opportunity/population marginal--was constrained, while the other was treated as a demand/supply weight (e.g., $D$/$O$ or $W_j^{(2)}$/$W_i^{(1)}$), without any allocation details.
-As a result, total constrained models express more 'potential' from both sides of the system, and singly constrained models express potential access from one side.
+In those cases, only one side of the interaction--either the total marginal or opportunity/population marginal--was constrained, while the other was treated as a demand/supply weight (e.g., $D$ or $O$ for total constrained and  $W_j^{(2)}$ or $W_i^{(1)}$ for singly constrained).
 
@@
 
-As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., the sum of all doubly constrained accessibility values $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$.
-
-As the doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model, as shown in @eq-doubly-constrained-gravitymodel, $V_{i}^D$ is as follows:
+As before, the simultaneous imposition of both constraints ensures the total system constraint is maintained i.e., $\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D$ remains equal to the total number of opportunities in the region $O$.
+
+As the doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model as follows:
 
@@
 
-Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as one for all i to obtain an initial estimate of $B_j$.
+Calibration of the two sets of proportionality constants is accomplished by means of iterative proportional fitting, whereby the values of $A_i$ are initialized as 1 for all i to obtain an initial estimate of $B_j$.
 The values of $B_j$ are used to update the underlying $V_{ij}^D$ matrix, before calibrating $A_i$.
@@
 $$
-\noindent and represent $V^D_{ij}$ as equal to $\kappa^D_{ij} D_j$, allowing the analyst to understand the proportional allocation of D_js to each $ij$ flow.
+\noindent and represent $V^D_{ij}$ as equal to $\kappa^D_{ij} D_j$, allowing the analyst to understand the proportional allocation of $D_j$s to each $ij$ flow.
 
@@
 Moving onto the toy example, to calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be reinterpreted.
-
 Namely, a count of physician _capacity_ per destination $D_j$ is needed instead of simply the number of physicians.
@@
 This adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
-For the population, we leave this unchanged numerically but theoretically note that each person interacts with one physician capacity.
-However, the number of providers per destination is revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).The system-wide provider-to-population ratio (PPR) is now:`r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
+For the population, we leave this unchanged numerically but we now must keep in mind that each person interacts with one physician capacity.
+The number of providers per destination is however revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).
+The system-wide PPR is now:`r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
 
@@
 
-However, despite the modifications to the example, our objective remains the same: to measure accessibility under different travel behavior scenarios.
-We aim to quantify the number of _potential_ spatial interactions between the physician-seeking population in each zone $i$ to the physician capacity in a zone $j$ i.e., $V_{ij}^D$.
+And with these modifications to the example, our objective is slightly different: to predict the flows from $j$ knowing that the amount of physician-capacity at each $j$ must be preserved and all flows to $i$ should match the number of people at $i$, under different travel behaviour scenarios.
+Put another way, we're interested in the $ij$ flows assuming we already know accessibility at each $i$.
 The highest decay travel behaviour scenario ($f_1(c_ij)$) is presented in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values.
@@
 
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physican-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()`.
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()`.
 Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
@@
 
-Recall, accessibility is defined as "the potential of spatial interaction" and is traditionally presented as a summary zonal measure.
-In the doubly constrained case, we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, hence the sum of $V_{ij}^D$ for all $i$s ends up being equal to the population at the zone (and $\sum_i V_{ij}^D$ being the opportunities at the zone).
-
-So, in the doubly-constrained case, accessibility flows becomes simply equivalent to spatial interaction flows--spatial proportional allocations between population and opportunity capacity.
-In fact, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $kappa_{ij}^D$ and $\hat kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
-
-As Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit; but the focus of these models have typically on $ij$ flows and often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
-On the other hand, accessibility has been interpreted differently: often understood as a zonal summary of the *potential* for spatial interaction, conceptualising origins and destinations in inherently different units.
-
-Tracing the application of increasing more restrictive constraints: from the total, to the singly- and now in the doubly-constrained measure, sheds light on how the balancing factors embedded in the unitless proportional allocation factors restrict the range in "potential" spatial interaction values that can be taken.
+Recall, accessibility is traditionally presented as a summary zonal measure.
+However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$.
+So following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
+Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $kappa_{ij}^D$ and $\hat kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
+
+Readers should recall, $V_{ij}^D$ are also mathematically equivalent to Wilson's spatial interaction flows.
+And as Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
+However, traditionally the focus of these models have typically on $ij$ flows and often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
+
+Accessibility is often understood as a zonal summary of potential spatial interaction, often involving origin and destination masses in different units.
+These units also often misaligned--for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
+When they do align--such as people to physician capacity--we’re essentially modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
+In such cases, the traditional accessibility question is already answered by the known information (i.e., how many opportunities can be reached by a zone?
+the number of people at that zone).
+This is why we do not foresee the doubly constrained measure being widely used in accessibility analysis: the literature has largely focused on questions of potential access, not on predicting flows of realised access.
 
@@
 It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023].
+2023 [@soukhovIntroducingSpatialAvailability2023].
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
@@
 
-In summary, building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility that is able to account for competition.
+Building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility..
 By reintroducing Wilson's proportionality constant, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.

### 2025-08-05 paper_plos/paper_plos.tex
made some additional, smaller, cuts. still 37 pages.

@@
 overall this diversity does not demystify existing questions like those
-raised in \citet{vanweeAccessible2016}: How should practitioners
+raised in \citep{vanweeAccessible2016}: How should practitioners
 interpret difference in scores between modes, and how should results be
@@
 
-We contend that accessibility research should reengage with spatial
+We contend that accessibility research should re-engage with spatial
 interaction modeling, particularly the use of Wilson's system
@@
 interpretive clarity to simple proportional comparisons (e.g.,
-``higher-than,'' ``lower-than'')
+``higher-than'', ``lower-than'')
 \citep{millerAccessibilityMeasurementApplication2018}. Without such
@@
 that can be tied to tangible values without any post-hoc treatment,
-making them more interpretable, communicable, and actionable in
-planning.
-
-This paper lays the historical and theoretical groundwork for a
-companion empirical study demonstrating how these constrained forms
-restore interpretability through concrete measurement units (e.g.,
-number of accessible opportunities), as opposed to the abstract weighted
-outputs typical of conventional gravity models.
+theoretically making them more interpretable, communicable, and
+actionable in planning.
 
@@
 expression of spatial interaction \citep{ravensteinLawsMigration1889},
-through Hansen \citep{hansen1959}, to Wilson \citep{wilson1971}--to
+through Hansen \citep{hansen1959} to Wilson \citep{wilson1971}--to
 propose a unified family of constrained accessibility measures. We
@@
 The patterns of people's movement in space have been a subject of
-scientific inquiry for at least a century and a half. From as far back
+scientific inquiry for at least a century and a half, from as far back
 as Henry C. Carey's \emph{Principles of Social Science}
-\citep{careyPrinciplesSocialScience1858}, a concern with the scientific
-study of human spatial interaction can be observed. It was in this work
-where Carey stated that ``man {[}is{]} the molecule of society {[}and
-their interaction is subject to{]} the direct ratio of the mass and the
+\citep{careyPrinciplesSocialScience1858}. It was in this work where
+Carey stated that ``man {[}is{]} the molecule of society {[}and their
+interaction is subject to{]} the direct ratio of the mass and the
 inverse one of distance'' \citep[pp.~37-38]{mckeanManual1883}. This
@@
 mass increases, so does force; as distance increases, force decreases.
-However, Eq~\ref{eq-phys-grav-prop} only demonstrates a proportional
-relationship. To quantify the magnitude (not the proportional magnitude)
-of the force, it must be \emph{constrained} with an empirical constant.
-This constant \(G\) converts Eq~\ref{eq-phys-grav-prop} from an
-expression of proportionality to the following expression of equality:
+However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
+\emph{proportional} relationship. To quantify the magnitude (not the
+\emph{proportional} magnitude) of the force, it must be
+\emph{constrained} with an empirical constant. This constant \(G\)
+converts Eq~\ref{eq-phys-grav-prop} from an expression of
+proportionality to the following expression of equality:
 
@@
 received support after Hutton's experiments of 1778
-\citep[p.~783]{hutton_xxxiii_1778}. Still, it took over a century from
-the publication of \emph{Principia} to refine the estimate of the
-proportionality constant to within 1\% accuracy, with Cavendish's 1798
-experiment \citep{cavendish_xxi_1798}.
-
-This gravitational framing laid the conceptual groundwork for later
-empirical studies of human spatial interaction, such as those by
-Ravenstein, Stewart, and others, who began translating these principles
-into observed patterns of human movement.
+\citep[p.~783]{hutton_xxxiii_1778} and Cavendish's 1798 experiment
+\citep{cavendish_xxi_1798} to estimate of the proportionality constant
+to within 1\% accuracy. In sum: it took over a century from the
+publication of \emph{Principia} to refine the estimate of the
+proportionality constant.
+
+The Newtonian gravitational relationship laid the conceptual groundwork
+for later empirical studies of human spatial interaction, such as those
+by Ravenstein, Stewart, and others, who began translating these
+principles into observed patterns of human movement. However, the
+majority of these attempts described a proportional relationship, or one
+arbitrarily set to equality.
 
@@
 empirical calibration necessary to convert these proportional
-relationships into measurable and comparable quantities
-
-After Carey and folowing Reilly and Zipf, it is only in
-\citet{stewartDemographicGravitationEvidence1948} that we find the most
+relationships into measurable and comparable quantities.
+
+After Carey and folowing Reilly and Zipf, it is only in Stewart
+\citep{stewartDemographicGravitationEvidence1948} that we find the most
 explicit connection yet to Newton's Gravitational law and the use of a
@@
 \(d_{ij}\) is the distance between \(i\) and \(j\). A version of this
-discrete form is what was used in \citet{hansen1959}, going on to become
-a foundation of modern accessibility definitions, as will be discussed.
+discrete form is what was used in Hansen \citep{hansen1959}, going on to
+become a foundation of modern accessibility definitions, as will be
+discussed.
 
@@
 as both a proportionality constant p.~34 and then later as
-\emph{demographic energy} on p.~53.) and its foundation in racist and
-unscientific assumptions (e.g., view of humans as particles following
-physical laws and assumptions of the molecular weight \(\mu\) of the
-average American being one, but ``presumably\ldots much less than
-one\ldots.for an Australian aborigine'' {[}p.~35{]})--his introduction
-of a proportionality constant \(G\) in modeling demographic force marks
-an important conceptual step: recognizing that moving from
-proportionality to equality requires empirical calibration. In other
-words, the addition of \(G\) shifts results from being abstract
-indicators of potential (i.e., people\textsuperscript{2-distance}-2) to
-having interpretable units grounded in consistent, albeit still
-abstract, quantities (i.e., units of demographic force).
-
-As will be discussed, when Hansen later adopted Stewart's formula
-(Eq~\ref{eq-stewart-population-potential-sum}) for accessibility, he
-omitted any mention of \(G\), effectively setting it to 1 arbitrarily as
-Stewart had done. This omission has persisted in accessibility research,
-leaving a conceptual gap in how such measures are interpreted and
-compared.
+\emph{demographic energy} on p.~53.) and its racist and unscientific
+assumptions (e.g., view of humans as particles following physical laws
+and assumptions of the molecular weight \(\mu\) of the average American
+being one, but ``presumably\ldots much less than one\ldots.for an
+Australian aborigine'' {[}p.~35{]})--his introduction of a
+proportionality constant \(G\) in modeling demographic force marks an
+important conceptual step: recognizing that moving from proportionality
+to equality requires empirical calibration. In other words, the addition
+of \(G\) shifts results from being abstract indicators of potential
+(i.e., \(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having
+interpretable units grounded in consistent, albeit still abstract,
+quantities (i.e., units of demographic force).
+
+As will be discussed, when Hansen \citep{hansen1959} later adopted
+Stewart's formula (Eq~\ref{eq-stewart-population-potential-sum}) for
+accessibility, he omitted any mention of \(G\), effectively setting it
+to 1 arbitrarily as Stewart had done. This omission has persisted in
+accessibility research, leaving a conceptual gap in how such measures
+are interpreted and compared.
 
@@
 Roads, and preoccupied with the power of transportation to shape land
-uses in a very practical sense. \citet{hansen1959} drew directly from
-Stewart's population potential (see
+uses in a very practical sense. Hansen \citep{hansen1959} drew directly
+from Stewart's population potential formula (see
 Eq~\ref{eq-stewart-population-potential-sum}), but left aside the
-broader (and often problematic) framework of ``social physics.''.
+broader (and often problematic) framework of ``social physics''.
 
@@
 Eq~\ref{eq-stewart-population-potential-sum} with \emph{opportunities}
-to derive an \emph{opportunity potential}, or more accurately, a
+to derive an \emph{opportunity potential}, or more specifically, a
 \emph{potential of opportunities for interaction} as
@@
 
-However, the omission of a proportionality constants and working with a
-proportional relationship generates fundamental issues in comparability
-between and, arguably within, studies. As the absence of \(G\) means
-accessibility estimates have no fixed unit--this renders them sensitive
-to the choice of impedance functions. For instance, if travel cost
-\(d_{ij}\) is measured in meters, then when the travel impedance
-function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the resulting \(S_i\)
-has units of opportunities per \(\text{metres}^{\beta}\); but when
-\(f(d_{ij})\) is set to equal \(e^{-\beta d_{ij}}\), the units become
-opportunities per \(e^{\beta \text{metres}}\). Such variation impairs
-comparability across analysis and obscures the meaning of accessibility
-scores renders their results difficult to understand and communicate
-without post-hoc treatment.
-
-Therefore in practice, Hansen-type measures (of proportionality) are
-better understood as \emph{ordinal indicators}--they rank accessibility
-but lack cardinal meaning or consistent units
+Furthermore, working with a proportional relationship generates
+fundamental issues in comparability between and, arguably within,
+studies. Namely, accessibility estimates have no fixed unit, rendering
+them sensitive to the choice of impedance functions. For instance, if
+travel cost \(d_{ij}\) is measured in meters, then when the travel
+impedance function \(f(d_{ij})\) equals \(d_{ij}^{-\beta}\), the
+resulting \(S_i\) has units of opportunities per
+\(\text{metres}^{\beta}\). However, when \(f(d_{ij})\) is set to equal
+\(e^{-\beta d_{ij}}\), the units become opportunities per
+\(e^{\beta \text{metres}}\). Such variation impairs comparability across
+analysis and obscures the meaning of accessibility scores, making them
+difficult to understand and communicate without post-hoc treatment.
+
+Therefore in practice, Hansen-type measures are ones of proportionality
+and are better understood as \emph{ordinal indicators}; they rank
+accessibility but lack cardinal meaning or consistent units
 \citep{millerAccessibilityMeasurementApplication2018}. The continued
-absence of \(G\) leaves a conceptual and practical gap in accessibility
-analysis: a missing link between theoretical form and empirical
-measurement.
+absence of a sort of proportionality constant \(G\) leaves a conceptual
+and practical gap in accessibility analysis: a missing link between
+theoretical form and empirical measurement.
 
@@
 
-While accessibility research evolved in North America with
-\citet{hansen1959}, a parallel development was taking place across the
-Atlantic with Alan G. Wilson. In his groundbreaking study
+While accessibility research evolved in North America with Hansen
+\citep{hansen1959}, a parallel development was taking place across the
+Atlantic with Alan G. Wilson. Wilson's ground breaking paper
 \citep{wilson1971}, advanced a general framework for spatial interaction
-modeling--focused not on potential (as accessibility), but on modeling
-observed flows of interaction between places. \citet{wilson1971}
-formalized the general spatial interaction model through the following
-equation:
+modeling--focused on modeling flows of interaction between places. This
+work was not focused on the `potential' concept as accessibility has
+been. Wilson \citep{wilson1971} formalized the general spatial
+interaction model through the following equation:
 
@@
 \(f(c_{ij})\) is some function of travel cost \(c_{ij}\) which reflects
-travel impedance. Finally, \(k\) acts as a proportionality constant,
+travel impedance.
+
+Of important note for this paper, \(k\) in
+Eq~\ref{eq-phys-gravity-model} acts as a proportionality constant,
 shifting the equation from a proportional to an equal relationship by
-incorporating known system totals. In this specific sense, we infer that
-\(k\) serves a role similar to the gravitational constant in Newton's
-law--it calibrates the model so that outputs match real-world
-quantities.
+incorporating known system totals. In this sense, we infer that \(k\)
+serves a role similar to the gravitational constant in Newton's law--it
+calibrates the model so that outputs match real-world quantities.
+However, these real-world quantities was not a set empirical constant
+(like Newton's \(G\)) but instead, sensitive to the system and known
+information about the system.
 
@@
 is available about the total number of flows produced by each origin,
-\(W_i^{(1)}\) is represented as \(O_i\) and the following constraint can
-be used:
+\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model} is represented as
+\(O_i\) and the following constraint can be used:
 
@@
 It is also possible to have information about both \(O_i\) and \(D_j\),
-in which case both constraints could be imposed on the model at once.
-
-Depending on which system constraints are applied---fully, partially, or
-not at all---a family of spatial interaction models can be derived from
+in which case both constraints can be imposed on the model at once.
+
+Depending on which of the three system constraints are applied a family
+of spatial interaction models can be derived from
 Eq~\ref{eq-phys-gravity-model}. The proportionality constant \(k\) is
@@
 averages consistent with observed origin and/or destination data. In the
-framework introduced and inferred from \citet{wilson1971}, three types
-of balancing factors are specified: (1) an unconstrained model that only
-matches the total volume of interaction \(K\), (2) a singly constrained
-model (either by origins \(A_i\) or destinations \(B_j\)), and (3) a
-doubly constrained model that satisfies both.
+framework introduced and inferred from Wilson \citep{wilson1971}, three
+types of balancing factors are specified: (1) an unconstrained model
+that only matches the total volume of interaction \(K\), (2) a singly
+constrained model (either by origins \(A_i\) or destinations \(B_j\)),
+and (3) a doubly constrained model that satisfies both.
 
@@
 balancing factor \(K\) takes the place of \(k\) and is equal to the
-following \citep[as specified in][ and
-\citet{fotheringham_spatial_1984}]{cliff_evaluating_1974}:
+following (as specified in \citep{cliff_evaluating_1974} and
+\citep{fotheringham_spatial_1984}):
 
@@
 Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
-to the following production-constrained singly-constrained model:
-
-\begin{equation}\phantomsection\label{eq-production-constrained-gravitymodel}{
-T_{ij} = A_i O_i W_j^{(2)} f(c_{ij})
-}\end{equation}
-
-Notice how, in this model, the proxy for the mass at the origin
-\(W_i^{(1)}\) is replaced with \(O_i\), representing what we know about
-the system, the spatial interaction outbound flow, i.e., outbound flows
-produced at \(i\). Also, there is no longer a single system-wide
-balancing factor \(K\), but rather a set of balancing factors specific
-to origins \(A_i\).
-
-For this model, the balancing factors ensure that constraint
-Eq~\ref{eq-constraint1-gravitymodel} is satisfied, meaning that the sum
-of predicted flows from one origin going to all destinations must equal
-the known mass at that origin \(O_i\) e.g., the total number of outbound
-flows. Satisfying this constraint also implicitly fulfills the total
-constraint (Eq~\ref{eq-constraint0-gravitymodel}), since the sum of
-\(O_i\) values across all origins equals the total number of flows.
-
-The production-constrained singly-constrained model is useful when flows
-ends are unknown but the number of flows originating from each location
-is known. The balancing factors for the production-constrained model are
-solved for each origin \(A_i\), and according to Wilson are:
+to the production-constrained singly-constrained version of
+Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
+origin \(W_i^{(1)}\) is replaced with \(O_i\). Also, \(k\) is replaced
+with a set of balancing factors specific to origins \(A_i\) which
+ensures that constraint Eq~\ref{eq-constraint1-gravitymodel} is
+satisfied (i.e., the sum of predicted flows from one origin going to all
+destinations must equal the known mass at that origin \(O_i\)).
+Satisfying this constraint also implicitly fulfills the total constraint
+(Eq~\ref{eq-constraint0-gravitymodel}), since the sum of \(O_i\) values
+across all origins equals the total number of flows. \(A_i\) takes the
+following form:
 
@@
 
-The attraction-constrained model is similar to the
-production-constrained model as it is also singly constrained but from
-the perspective of the mass at the destination. From the
-attraction-constrained model, the proxy for the mass at the destination
-\(W_j^{(2)}\) is now replaced with \(D_j\), representing the spatial
-interaction inbound flow e.g., flows attracted to the destination, and
-takes the following form:
-
-\begin{equation}\phantomsection\label{eq-attraction-constrained-gravitymodel}{
-T_{ij} = B_j D_j W_i^{(1)} f(c_{ij})
-}\end{equation}
-
-In this model (Eq~\ref{eq-attraction-constrained-gravitymodel}), the
-balancing factors ensure that Eq~\ref{eq-constraint2-gravitymodel} is
-satisfied (hence the total constraint
-Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning that the sum
-of predicted flows going to one destination from all origins must equal
-the known mass of that destination \(D_j\) i.e., the total number of
-inbound flows to \(j\). This should hold for all destinations. As
+The singly-constrained attraction-constrained model is similar to the
+production-constrained version but from the perspective of the mass at
+the destination. For the attraction-constrained model, the proxy for the
+mass at the destination \(W_j^{(2)}\) is replaced with \(D_j\) in
+Eq~\ref{eq-phys-gravity-model}, representing the spatial interaction
+inbound flow. Also, \(k\) is replaced with a set of destination-specific
+balancing factors \(B_j\) that ensure that
+Eq~\ref{eq-constraint2-gravitymodel} is satisfied (hence the total
+constraint Eq~\ref{eq-constraint0-gravitymodel} is as well), meaning
+that the sum of predicted flows going to one destination from all
+origins must equal the known mass of that destination \(D_j\). As
 before, destination-specific balancing factors \(B_j\) were derived by
@@
 Lastly, the doubly-constrained model is the production-attraction
-constrained model in \citet{wilson1971}. In this case, both constraints
-Eq~\ref{eq-constraint1-gravitymodel} and
+constrained model in Wilson \citep{wilson1971}. In this case, both
+constraints Eq~\ref{eq-constraint1-gravitymodel} and
 Eq~\ref{eq-constraint2-gravitymodel} hold simultaneously. These
@@
 destination \(D_j\). This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly constrained, and takes
-the following form:
-
-\begin{equation}\phantomsection\label{eq-doubly-constrained-gravitymodel}{
-T_{ij} = A_i B_j O_i D_j f(c_{ij})
-}\end{equation}
-
-In this model, both proxies for the masses are replaced with the known
-masses, that is, the flows produced by the origin and the flows
-attracted by the destination. There are now two sets of mutually
-dependent proportionality constants shown in
-Eq~\ref{eq-doubly-constrained-balancing-factors}. Derivation of these
-models is demonstrated in detail elsewhere
+The resulting model is, in Wilson's terms, doubly constrained, and from
+Eq~\ref{eq-phys-gravity-model}, \(k\) becomes both \(A_i\) and \(B_j\)
+shown in Eq~\ref{eq-doubly-constrained-balancing-factors}, and
+\(W_i^{(1)}\) and \(W_j^{(2)}\) is replaced with \(O_i\) and \(D_j\).
+Derivation of these models is demonstrated in detail elsewhere
 \citep[e.g.,][]{ortuzar_2011_modelling, wilsonSTATISTICALTHEORYSPATIAL1967}.
@@
 is especially significant because it can address a problem that has
-quietly plagued this field (as this work has argued): the transition
+quietly afflicted this field (as this paper has argued): the transition
 from proportionality to equality.
@@
 unit-consistent \(ij\) flows. In short, when aspects of the systems are
-inputs into the balacing factors, \(ij\) flows can be robustly predicted
-in tangible units of `flow', units that align with the system being
-modelled. In this way, spatial interaction modeling tradition can be
-seen to have succeeded where accessibility modeling stalled. While
+inputs into the balancing factors, \(ij\) flows can be robustly
+predicted in tangible units of `flow', units that align with the system
+being modelled. In this way, spatial interaction modeling tradition can
+be seen to have succeeded where accessibility modeling stalled. While
 Wilson's model produces results that are in units of flow tethered to
@@
 investigation sheds light on why accessibility research may have failed
-to adopt a comparable approach focused, until this work.
+to adopt a comparable approach focused, until this paper.
 
@@
 frameworks emerged. As noted in
-\citet{battyChronicleScientificPlanning1994}, large scale spatial
+\citep{battyChronicleScientificPlanning1994}, large scale spatial
 interaction models (like Wilson's) responded to important developments
@@
 Thirdly, there is a group of accessibility-focused works that interprets
-the measure used in \citet{hansen1959} as the singly- or doubly-
+the measure used in Hansen \citep{hansen1959} as the singly- or doubly-
 constrained spatial interaction model's inverse balancing factor
@@
 In sum, despite the interpretative advantaged offered by the statistical
-logic of \citet{wilson1971}'s framework, the spatial
-interaction-oriented \citet{hansen1959} citing literature nor the
-accessibility-oriented literature that cite \citet{wilson1971} have yet
-to meaningfully adopt Wilson's constraint-based logic to the concept of
-accessibility. So, in the next section, we demonstrate how applying
-Wilson's framework enables accessibility to move from a proportional
-relationship to calibrated equality--tying outputs to tangible system
-knowns. This re-expression of accessibility using constraints yields
-interpretable, unit-consistent measures of opportunity. This approach
-takes the same path of entropy-maximisation as in \citet{wilson1971},
-and does not rely on specifying some universal constant \(G\) like
-initially proposed in \citet{stewartDemographicGravitationEvidence1948}
-(recall, Eq~\ref{eq-stewart-population-potential-sum} which
-\citet{hansen1959} operationalised).
+logic of Wilson \citep{wilson1971}'s framework, the spatial
+interaction-oriented Hansen \citep{hansen1959} citing literature nor the
+accessibility-oriented literature that cite Wilson \citep{wilson1971}
+have yet to meaningfully adopt Wilson's constraint-based logic to the
+concept of accessibility. So, in the next section, we demonstrate how
+applying Wilson's framework enables accessibility to move from a
+proportional relationship to calibrated equality--tying outputs to
+tangible system knowns. This re-expression of accessibility using
+constraints yields interpretable, unit-consistent measures of
+opportunity. This approach takes the same path of entropy-maximisation
+as in Wilson \citep{wilson1971}, and does not rely on specifying some
+universal constant \(G\) like initially proposed in Stewart
+\citep{stewartDemographicGravitationEvidence1948} (recall,
+Eq~\ref{eq-stewart-population-potential-sum} which Hansen
+\citep{hansen1959} operationalised).
 
@@
 approach grounded in statistical mechanics, shifting place-based
-accessibility (a la \citet{hansen1959}) from a relationship describing
-proportionality to equality.
+accessibility (a la Hansen \citep{hansen1959}) from a relationship
+describing proportionality to equality.
 
@@
 indicators we previous outlined. Rather than relying on arbitrary or
-uncalibrated proportional relationships--or invoking a universal
-constant \(G\) as proposed in
-\citet{stewartDemographicGravitationEvidence1948} --we adopt Wilson's
+uncalibrated proportional relationships (or invoking a universal
+constant \(G\) as proposed in Stewart
+\citep{stewartDemographicGravitationEvidence1948}) we adopt Wilson's
 approach grounded in statistical mechanics.
@@
 To do so, we propose a revised definition of accessibility considerate
-of the constraint-based spatial intereaction model: the potential for
-spatial interaction with opportunities or population.We can bring \(k\)
-as a proportional allocation factor which encorporates Wilson's
-balancing factor, to define the \emph{potential for spatial interaction
-with opportunities} \(V_{ij}\) and the \emph{potentnial for spatial
-intereaction with population} \(M_{ji}\). In effect, \(\kappa\) is
-unitless and proportionally allocates (based on the constraints of the
-case) opportunities (for \(V_{ij}\)) and population (for \(M_{ji}\)).
-The equation is as follows:
+of the constraint-based spatial interaction model: the potential for
+spatial interaction with opportunities (or population). We can specify
+\(k\) as a type of proportional allocation factor \(\kappa\) which
+incorporates Wilson's balancing factor(s) to define the \emph{potential
+for spatial interaction with opportunities} \(V_{ij}\) and the
+\emph{potential for spatial interaction with population} \(M_{ji}\). In
+effect, \(\kappa\) is unitless and proportionally allocates (based on
+the constraints of the case) opportunities (for \(V_{ij}\)) and
+population (for \(M_{ji}\)). The equation is as follows:
 
@@
 population) at either the zone or for the full region, depending on the
-case. Accessibility flows can also be summarised as as a partial sum of
-the potential at \(i\) and at \(j\) to express accessibility at the
-origin zone or at the destination zone, respectively, as often done in
+case (hence represented by a \(X\)).
+
+Accessibility flows can also be summarised as as a partial sum of the
+potential at \(i\) and at \(j\) to express accessibility at the origin
+zone or at the destination zone, respectively, as often done in
 accessibility research:
@@
   Constrained on both origin and destination sides using both \(A_i\)
-  and \(B_j\) simultaneously; equivalent in interpretation to Wilson's
+  and \(B_j\) simultaneously, which can also be expressed as
+  proportional allocation factors (\(\kappa_{ij}^D\),
+  \(\hat \kappa_{ji}^D\)); equivalent in interpretation to Wilson's
   \citep{wilson1971} doubly constrained spatial interaction model.
@@
   \(i\) and \(j\) simply reproduces known totals. Represents
-  `interaction capacity' or `realized access,' serving as predictions of
+  `interaction capacity' or `realized access' serving as predictions of
   real interaction flows.
@@
 interpretability issue of Hansen's accessibility (i.e., unconstrained
-accessibility) by grounding values in a meaningful total--enabling
-robust comparisons across zones and scenarios--but also by keeping
-values proportional to \(V_i^0\) so interpretation is similar.
+accessibility) by grounding values in a meaningful total, enabling
+robust comparisons across zones and scenarios but also by keeping values
+proportional to \(V_i^0\) so interpretation is similar.
 
@@
 The underlying spatial interaction model is now the
-attraction-constrained model in
-Eq~\ref{eq-attraction-constrained-gravitymodel}, and our accessibility
-measure becomes:
+attraction-constrained model and our accessibility measure becomes:
 
@@
 
-\citet{soukhovIntroducingSpatialAvailability2023} also showed that the
-following expression (accessibility per capita) is a constrained version
-of the popular 2SFCA approach of Shen \citep{shen1998} and Luo and Wang
-\citep{luo2003}:
+Soukhov et al., 2023 \citep{soukhovIntroducingSpatialAvailability2023}
+also showed that the following expression (accessibility per capita) is
+a constrained version of the popular 2SFCA approach of Shen 1998
+\citep{shen1998} and Luo and Wang 2003 \citep{luo2003}:
 
@@
 The underlying spatial interaction model is now the
-production-constrained model in
-Eq~\ref{eq-production-constrained-gravitymodel}, and our market
-potential measure \(M^S_j\) becomes:
+production-constrained model version of Eq~\ref{eq-phys-gravity-model},
+and our market potential measure \(M^S_j\) becomes:
 
@@
 
-Because both margins are fixed, this measure reflects fully allocated
-interaction flows, not latent potential. As such, accessibility cannot
-be interpreted as a traditional summary at \(i\) or \(j\) (e.g.,
+Because results are made to match both margins, the results cannot be
+interpreted as a traditional summary at \(i\) or \(j\) (e.g.,
 ``opportunities accessible from \(i\)'')--those sums simply reproduce
@@
 opportunity/population marginal--was constrained, while the other was
-treated as a demand/supply weight (e.g., \(D\)/\(O\) or
-\(W_j^{(2)}\)/\(W_i^{(1)}\)), without any allocation details. As a
-result, total constrained models express more `potential' from both
-sides of the system, and singly constrained models express potential
-access from one side.
+treated as a demand/supply weight (e.g., \(D\) or \(O\) for total
+constrained and \(W_j^{(2)}\) or \(W_i^{(1)}\) for singly constrained).
 
@@
 As before, the simultaneous imposition of both constraints ensures the
-total system constraint is maintained i.e., the sum of all doubly
-constrained accessibility values
+total system constraint is maintained i.e.,
 \(\sum_i V^D_{i} = \sum_i\sum_j  V^D_{ij} = D\) remains equal to the
@@
 form of the production-attraction (doubly constrained) spatial
-interaction model, as shown in
-Eq~\ref{eq-doubly-constrained-gravitymodel}, \(V_{i}^D\) is as follows:
+interaction model as follows:
 
@@
 by means of iterative proportional fitting, whereby the values of
-\(A_i\) are initialized as one for all i to obtain an initial estimate
-of \(B_j\). The values of \(B_j\) are used to update the underlying
+\(A_i\) are initialized as 1 for all i to obtain an initial estimate of
+\(B_j\). The values of \(B_j\) are used to update the underlying
 \(V_{ij}^D\) matrix, before calibrating \(A_i\). This process continues
@@
 \(\kappa^D_{ij} D_j\), allowing the analyst to understand the
-proportional allocation of D\_js to each \(ij\) flow.
+proportional allocation of \(D_j\)s to each \(ij\) flow.
 
@@
 of the opportunity data in the numeric example must be reinterpreted.
-
 Namely, a count of physician \emph{capacity} per destination \(D_j\) is
@@
 opportunities (in units of 10,00s of physician-capacity) per zone. For
-the population, we leave this unchanged numerically but theoretically
-note that each person interacts with one physician capacity. However,
-the number of providers per destination is revised to represent
+the population, we leave this unchanged numerically but we now must keep
+in mind that each person interacts with one physician capacity. The
+number of providers per destination is however revised to represent
 physician capacity, scaled approximately from the original number of
 physicians used in previous cases
-(Table~\ref{tbl-small-system-land-use}).The system-wide
-provider-to-population ratio (PPR) is now:1, this is compared to the
-unmodified example which yields system PPR of24.5.
+(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now:1,
+this is compared to the unmodified example which yields system PPR
+of24.5.
 
@@
 
-However, despite the modifications to the example, our objective remains
-the same: to measure accessibility under different travel behavior
-scenarios. We aim to quantify the number of \emph{potential} spatial
-interactions between the physician-seeking population in each zone \(i\)
-to the physician capacity in a zone \(j\) i.e., \(V_{ij}^D\). The
-highest decay travel behaviour scenario (\(f_1(c_ij)\)) is presented in
+And with these modifications to the example, our objective is slightly
+different: to predict the flows from \(j\) knowing that the amount of
+physician-capacity at each \(j\) must be preserved and all flows to
+\(i\) should match the number of people at \(i\), under different travel
+behaviour scenarios. Put another way, we're interested in the \(ij\)
+flows assuming we already know accessibility at each \(i\). The highest
+decay travel behaviour scenario (\(f_1(c_ij)\)) is presented in
 Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}.
@@
 \(M^D_j\), since these values reproduce the original constraints--i.e.,
-\(V^D_i = O_i\) so the physican-capacity accessible for Zones 1, 2, and
+\(V^D_i = O_i\) so the physician-capacity accessible for Zones 1, 2, and
 3 would be 4.0018381, 9.9963656, and 6.0017963. Hence, the usefulness of
@@
 
-Recall, accessibility is defined as ``the potential of spatial
-interaction'' and is traditionally presented as a summary zonal measure.
-In the doubly constrained case, we force the allocation of zonal
-population demand and zonal opportunities supplied to be paired and
-allocation to be proportional, hence the sum of \(V_{ij}^D\) for all
-\(i\)s ends up being equal to the population at the zone (and
-\(\sum_i V_{ij}^D\) being the opportunities at the zone).
-
-So, in the doubly-constrained case, accessibility flows becomes simply
-equivalent to spatial interaction flows--spatial proportional
-allocations between population and opportunity capacity. In fact,
-\(V^D_{ij}\) and its transposed counterpart \(M^D_{ji}\) are
-structurally identical, differing only in interpretation (referring to
-\(kappa_{ij}^D\) and \(\hat kappa_{ij}^D\)): one reflects the
-proportional allocation of opportunity to population flows; the other,
-population to opportunity flows.
-
-As Wilson \citep{wilson1971} explicitly noted, origin and destination
-weights defined in the spatial interaction model \emph{can} be defined
-using any unit; but the focus of these models have typically on \(ij\)
-flows and often calibrated using trips (i.e., outbound and inbound
-trips, inherently in the same units). On the other hand, accessibility
-has been interpreted differently: often understood as a zonal summary of
-the \emph{potential} for spatial interaction, conceptualising origins
-and destinations in inherently different units.
-
-Tracing the application of increasing more restrictive constraints: from
-the total, to the singly- and now in the doubly-constrained measure,
-sheds light on how the balancing factors embedded in the unitless
-proportional allocation factors restrict the range in ``potential''
-spatial interaction values that can be taken.
+Recall, accessibility is traditionally presented as a summary zonal
+measure. However, in the doubly constrained case, since we force the
+allocation of zonal population demand and zonal opportunities supplied
+to be paired and allocation to be proportional, \(V_i^D\) is simply the
+number of opportunities that matches our known population at \(i\). So
+following the logic of the family of accessibility measures, in the
+doubly constrained case, \(V_{ij}^D\) flows are the only relevant unit
+of analysis: spatial proportional allocations between population and
+opportunity capacity. Furthermore, \(V^D_{ij}\) and its transposed
+counterpart \(M^D_{ji}\) are structurally identical, differing only in
+interpretation (referring to \(kappa_{ij}^D\) and
+\(\hat kappa_{ij}^D\)): one reflects the proportional allocation of
+opportunity to population flows; the other, population to opportunity
+flows.
+
+Readers should recall, \(V_{ij}^D\) are also mathematically equivalent
+to Wilson's spatial interaction flows. And as Wilson \citep{wilson1971}
+explicitly noted, origin and destination weights defined in the spatial
+interaction model \emph{can} be defined using any unit. However,
+traditionally the focus of these models have typically on \(ij\) flows
+and often calibrated using trips (i.e., outbound and inbound trips,
+inherently in the same units).
+
+Accessibility is often understood as a zonal summary of potential
+spatial interaction, often involving origin and destination masses in
+different units. These units also often misaligned--for instance, we may
+not know how much park space, grocery area, or childcare capacity is
+accessible per person. When they do align--such as people to physician
+capacity--we're essentially modeling realised access flows based on
+known quantities of \emph{those that interact} and the
+\emph{interacted}. In such cases, the traditional accessibility question
+is already answered by the known information (i.e., how many
+opportunities can be reached by a zone? the number of people at that
+zone). This is why we do not foresee the doubly constrained measure
+being widely used in accessibility analysis: the literature has largely
+focused on questions of potential access, not on predicting flows of
+realised access.
 
@@
 opportunities. It is also mathematically equivalent to the spatial
-availability introduced in Soukhov et al.
+availability introduced in Soukhov et al.~2023
 \citep{soukhovIntroducingSpatialAvailability2023}. In this case, all
@@
 
-In summary, building on Wilson's \citep{wilson1971} foundational work,
-this paper proposed a unified framework for accessibility that is able
-to account for competition. By reintroducing Wilson's proportionality
-constant, the proposed family of constrained accessibility measures
-restores measurement units to accessibility estimates. This enhancement
-provides a more interpretable, consistent, and theoretically grounded
-basis for accessibility analysis, which could help advance the adoption
-of accessibility-oriented planning.
+Building on Wilson's \citep{wilson1971} foundational work, this paper
+proposed a unified framework for accessibility.. By reintroducing
+Wilson's proportionality constant, the proposed family of constrained
+accessibility measures restores measurement units to accessibility
+estimates. This enhancement provides a more interpretable, consistent,
+and theoretically grounded basis for accessibility analysis, which could
+help advance the adoption of accessibility-oriented planning.
 

### 2025-08-05 paper_plos/paper_plos.qmd
draft of response to reviewers

@@
 As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
-This averaging of the total amount is how the total constrained proportional allocation factor functions.
+In terms of proportional magnitude, this can also be observed in the unconstrained measure for this scenario.
+However, for the total constrained measure, this plateauing of results have meaning.
+In fact, each zone is allocated an average of the total amount in the region, as a result of the total constrained proportional allocation factor.
+
 However, what's notable is how Zones change between scenarios.

### 2025-08-05 paper_plos/reviewer_respose_letter_V1.qmd
draft of response to reviewers

@@
 ---
-title: Response to reviewer
+title: Response to reviewers
+format: pdf
 ---
@@
 
+Thank you for the kind words and for this helpful feedback.
+We've signfigicantly revised the manuscript, focusing on explaining things concisely and ensuring each paragraph is move forward the key points of each section.
+In summary:
+
+- We've reduced the length of the manuscript from 34 to 28 pages (without references).
+- We've removed tangential details and minimised the equations where possible, particularly in the “Early Research” section.
+- We've streamlined the manuscript's text--to make each sections purpose more clear.
+- We've removed redundancies where possible, especially in the "Family of Accessibility" section. 
+- We've more clearly highlighted the paper's main contribution: namely, a review of the shared origins of spatial interaction modeling and accessibility, how their interpretations have diverged over time, and the introduction of a family of accessibility measures grounded in spatial interaction principles--including total, singly, and doubly constrained cases, as well as variants of “accessible opportunities” and “accessible population".
+
 \textcolor{blue}{Related to the first issue, I am not sure about the point that the authors explain the details of Newtonian’s roots and early research on human spatial interactions from Ravenstein to Stewart.
 Of course, I think mentioning those contents is beneficial, but it is hard to get the key point given that it is verbose.}
+
+The main aim of these two sections is to demonstrate how Newton's gravitational equation (with its proportionality factor $G$) inspired early research on human spatial interactions.
+The majority of early researchers relied on the underlying proportional mass/distance relationship to model human spatial interaction, but did not use $G$--except for Stewart.
+However, Stewart didn't have an answer to what $G$ should be, just that it is needed to move from a relationship of proportionality to one of equality.
+
+Stewart is especially important because Hansen later adopted it into his equation but ommitted both the $G$ (i.e., kept it set to 1) and was quiet on the broader issue of moving from proportionality to equality.
+
+We've revised these sections, hopefully we've clarifed these points and the purpose of these sections in the process!
 
@@
 
-THank you for these comments. We've heavily revised the manuscript, focusing now on explaining things concisely and putting forward the focus. We did this by:
+Good point.
+We now clearly summarize all four groups of constrained measures at the start of the “A Family of Accessibility Measures: From Proportionality to Equality” section.
+These definitions build on the preceding sections, which trace how spatial interaction modeling followed the Newtonian analogy while neglecting $G$, how Wilson shifted to an entropy-maximisation formulation without requiring an empirical $G$ (but instead, a constraint that functions to tether the results to system knowns), and how the place-based accessibility literature has largely overlooked this solution.
+The main section then presents the mathematical formulations and a toy example to illustrate the full family of measures.
 
-- reducing the length of the manuscript, it is now 10 pages
-- removed tangental details (e.g., )
-- streamlined the historical section (e.g., )
-- highlight the main contribution of the paper, namely how it provides constrained definitions to accessibility
-- reduced the number of equations where possible, and make it clear why those that remain are there.
+Furthermore, regarding the equations--many have now been removed when possible if they don't directly further the argument of the section or are redundant.
+For instance, some equations in "Early research on human spatial interaction" have been removed, and rewritings of Wilson's general formula from "Wilson's family of spatial interaction models" section have been removed.
 
@@
 
+Thannk you for your time and effort in reviewing our manuscript. We appreciate the constructive feedback and comments.
+
+Thank you for this point.
+We agree, the inclusion of an empirical example would demonstrate the real-world applicability of the measures.
+However, due to the existing length of the document, the addition of an empirical example (even if we replaced the toy example) would still add additional length: we would need to add a condensed methods and data section, and dedicate more room for tests.
+
+The length of the current document is an issue, so the addition of additional pages isn't feasible--especially given the concern over paper length that was raised by Reviewer 1 and how we’ve already reduced it by six pages in response to these concerns.
+However, we've put great thought into how the manuscript could be restructured to included an additional empirical example, but it just isn't possible given the manuscript's current format.
+
+As such, we’ve clarified that the contribution of this work is primarily theoretical and methodological.
+We’ve also tempered earlier claims about the measure’s real-world applicability, instead allowing readers to consider potential uses and the potential value of reuniting accessibility measures with interpretable units.
+We plan to explore an empirical application in a future publication which make this contribution more practically clear--where we may have enough space to develop and explore an empirical example without shortchanging the historical context.
+
 \textcolor{blue}{2.
@@
 However, the terminological distinction remains conceptually blurry.}
+
+Agreed, the use of these words--especially in the doubly constrained model subsection should be tuned up.
+We've substantially revised this subsection.
+Namely, we've removed the discussion about the meaning of "potential", and instead focus on the embedded assumption of the doubly constrained measure i.e., how it simultaneously considers both origin-side and destination-side constraints--or that all people are matched with opportunities, and vice versa.
+We discuss how this is not a frequent situation in place-based accessibility analysis (i.e., we don't know how much parkland space capacity is matched to each person).
+
+Also, the double constraint ends up fixomg the aggregated zonal value to match the 'mass' at the zone itself (i.e., reproducing what we already imposed onto the model).
+In this way, the $ij$ flows are the only useful thing for analysis.
+This focus has not tradtionally occupied the research context of accessibility-based work.
 
@@
 
+Thank you for this comment.
+The historical literature review section (Accessibility and spatial interaction modelling: two divergent research streams) covers upward of 76 papers citing Wilson and Hansen, condensed into about 3 pages.
+This section sets the stage by outlining Wilson’s balancing factors and highlighting how place-based accessibility literature (largely based on Hansen-type measures) has not adopted these balancing factors.
+This context is needed for the subsequent, more detailed 16-page section, which introduces the newly proposed family of accessibility measures which incorporate Wilson's balancing factors.
+We believe this structure balances breadth and depth appropriately--we've tried to reduce redundancies and improve when possible.
+We hope you find the revised manuscript easier to follow.
+
 \textcolor{blue}{4.
 The paper pays limited attention to recent methodological developments such as time-sensitive, network-based, or multi-modal accessibility models, as well as behavioral or utility-based approaches.}
+
+Thank you for this fair point.
+In the newly revised manuscript, we make it more clear that our focus is on place-based accessibility, which remains widely used by practitioners and prominent in the literature, as well as Wilson's spatial interaction model which is also popular in spatial interaction modeling still to this day.
+Given the broad scope and space constraints (the current manuscript is already quite long and conceptually broad), incorporating additional detailed discussions of novel approaches is challenging, and somewhat tangential.
+
+However, we do acknowledge some utility-based approaches within the spatial interaction literature--for example, "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
+
+Our point with the historical review section is to demonstrate how even the papers that _do_ cite both Wilson and Hansen do not operationalise Wilson's balancing factors.
+Utility-based approaches take a different direction--not based on Newtonian gravitation--but also not based on Wilson's balancing factors.
+Indeed recent developments such as time-sensitive, network-based, or multi-modal accessibility models, as well as behavioral or utility-based approaches take on new approaches, but we believe these are tangential to our scope.
+
+Thoughts???
 
@@
 
+This is a very fair point. Inside of explaining results with an empirical example, we theoretical demonstrate this throughout the paper.
+
+First, this idea is brought up more concretely on page 5, when discussing Stewart (1948)'s use of his proportionality constant $G$: "In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force)."
+
+Later, we develop this point deeper when discussing Hansen-type accessibility, and the impact of not having a proportionality constant:
+pg.
+6: "Furthermore, working with a proportional relationship generates fundamental issues in comparability between and, arguably within, studies.
+Namely, accessibility estimates have no fixed unit, rendering them sensitive to the choice of impedance functions.
+For instance, if travel cost $d_{ij}$ is measured in meters, then when the travel impedance function $f(d_{ij})$ equals $d_{ij}^{-\beta}$, the resulting $S_i$ has units of opportunities per $\text{metres}^{\beta}$.
+However, when $f(d_{ij})$ is set to equal $e^{-\beta d_{ij}}$, the units become opportunities per $e^{\beta \text{metres}}$.
+Such variation impairs comparability across analysis and obscures the meaning of accessibility scores, making them difficult to understand and communicate without post-hoc treatment."
+
+And to solidify this point, again more theoretically but still practically, we develop and solve the toy example.
+
+Regarding the unconstrained example:
+pg.
+16: "For example, Table @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function.
+Comparing across decay types is meaningless in absolute terms.
+For instance, the difference in zone 1 (edge of urban core)’s accessibility under $f_3$ vs $f_1$ is 370.92, but in what units?
+These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
+The fundamental uninterpretability of what is a _opportunity-weighted-travel-impedance_ unit remains."
+
+Regarding the total constrained example:
+pg.
+18: "Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
+For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (172.06), like in the unconstrained accessibility case.
+However, unlike in the unconstrained case, we can say that this value is out of the 490 physicians in the region, which allows us also to deduce that zone 1 captures 1.31 and 0.92 times more than zone 2 and 3."
+
+pg.
+18:"One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
+As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
+In terms of proportional magnitude, this can also be observed in the unconstrained measure for this scenario.
+However, for the total constrained measure, this plateauing of results have meaning.
+In fact, each zone is allocated an average of the total amount in the region, as a result of the total constrained proportional allocation factor."
+
 \textcolor{blue}{6.
 Some sections suffer from repetition and overly technical language without summarizing takeaways (e.g., Table 1 is helpful, but the text before and after it repeats much of the same information).}
+
+Excellent point. We've decided to remove Table 1 all together, and provide a more condensed version of it in text on pg 13/14. Namely:
+
+"The proportional allocation constant $\kappa$ takes the form of a balancing factor that varies depending on the constraints applied.
+Each member of the accessibility measure family is defined by the constraints used, and can be grouped into the following four cases:
+
+1. **Unconstrained Case ($V_i^0$, $M_j^0$)**
+- Equivalent to Hansen’s [@hansen1959] and Reilly’s [@reilly1929methods] original formulations; the status quo of accessibility modelling.
+- No balancing factors applied; units are in "opportunities-by-impedance" for $V_i^0$ or "population-by-impedance" for $M_j^0$.
+- No constraints are applied, so values reflect proportionality only and are not calibrated to known system totals.
+2. **Total Constrained Case ($V_i^T$, $M_j^T$)**
+- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in @fig-analytical-device-conc-accessibility) i.e., total number of opportunities or population in the system.
+This ensures the sum of all values in the system match the total marginal.
+- Units of $V_i^T$: accessible opportunities from $i$, a value that is total constrained and linearly proportion to $V_i^0$.
+- Units of $M_j^T$: accessible population from $j$, a value that is total constrained and linearly proportion to $M_j^0$.
+3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
+- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
+Reflects how the literature calculates competitive accessibility.
+- Units of $V_i^S$: accessible opportunities from $i$, a value that is the sum of opportunity supply flows allocated proportionally based on demand at $i$.
+Mathematically equivalent in per-capita form to 2SFCA [@luo2003].
+- Units of  $M_j^S$: accessible population from $j$, a value that is the sum of population demand flows allocated proportionally based on supply at $j$.
+4. **Doubly Constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
+- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously, which can also be expressed as proportional allocation factors ($\kappa_{ij}^D$, $\hat \kappa_{ji}^D$); equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
+- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in @fig-analytical-device-conc-accessibility).
+- Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
+Represents ‘interaction capacity’ or ‘realized access’ serving as predictions of real interaction flows."
+
+We hope this addresses your concern! As well, we've streamlined redundancies throughout the manuscript, especially in the "A family of accessibility measures" section.

### 2025-08-06 paper_plos/reviewer_respose_letter_V1.qmd
small edits to the response letter

@@
 
-Also, the double constraint ends up fixomg the aggregated zonal value to match the 'mass' at the zone itself (i.e., reproducing what we already imposed onto the model).
-In this way, the $ij$ flows are the only useful thing for analysis.
-This focus has not tradtionally occupied the research context of accessibility-based work.
-
 \textcolor{blue}{3.
@@
 This context is needed for the subsequent, more detailed 16-page section, which introduces the newly proposed family of accessibility measures which incorporate Wilson's balancing factors.
-We believe this structure balances breadth and depth appropriately--we've tried to reduce redundancies and improve when possible.
+We believe this structure balances breadth and depth appropriately; we've tried to reduce redundancies and improve when possible.
 We hope you find the revised manuscript easier to follow.
@@
 
-However, we do acknowledge some utility-based approaches within the spatial interaction literature--for example, "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
+However, we do acknowledge some utility-based approaches within the spatial interaction literature.
+For example, "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
 

### 2025-08-15 paper_plos/reviewer_respose_letter_V1.qmd
A few updates to draft response to reviewers.

@@
 
-Thank you for the kind words and for this helpful feedback.
-We've signfigicantly revised the manuscript, focusing on explaining things concisely and ensuring each paragraph is move forward the key points of each section.
-In summary:
+Thank you for this point. We've significantly revised the manuscript, focusing on explaining things concisely and ensuring each paragraph is move forward the key points of each section. In summary:
 
 - We've reduced the length of the manuscript from 34 to 28 pages (without references).
-- We've removed tangential details and minimised the equations where possible, particularly in the “Early Research” section.
-- We've streamlined the manuscript's text--to make each sections purpose more clear.
+- We've removed tangential details and minimized the equations where possible, particularly in the “Early Research” section.
+- We've streamlined the manuscript's text--to make each sections purpose more clear. The introductions and conclusions of sections now have a better flow.
 - We've removed redundancies where possible, especially in the "Family of Accessibility" section. 
 - We've more clearly highlighted the paper's main contribution: namely, a review of the shared origins of spatial interaction modeling and accessibility, how their interpretations have diverged over time, and the introduction of a family of accessibility measures grounded in spatial interaction principles--including total, singly, and doubly constrained cases, as well as variants of “accessible opportunities” and “accessible population".
+
+The key points of the manuscript can be summarised as follows:
+- First, it explicitly links accessibility to spatial interaction through their shared intellectual foundations (pre- @hansen1959), highlighting how contemporary accessibility research has diverged and showing the mathematical, intuitive, and interpretative advantages of preserving units through constrained formulations, as done in spatial interaction modeling.
+- Second, it introduces 'constraints' to accessibility measures.
+First: the total constrained accessibility measure as an alternative to the Hansen-type formulation, producing zonal values that represent proportions of the known number of destination masses (i.e., opportunities) or origin masses (i.e., population) in the region.
+Second: the singly constrained accessibility measure as a competitive method that proportionally allocates opportunities from the destination based on population demand (from the origin), or in the case of market potential, proportional allocation of population from the origin based on the supply of opportunities at the destination.
+Its per-capita zonal rate is shown to be mathematically equivalent to the widely used two-step floating catchment area (2SFCA) method [@luo2003; @shen1998].
 
@@
 
-Thannk you for your time and effort in reviewing our manuscript. We appreciate the constructive feedback and comments.
-
 Thank you for this point.
 We agree, the inclusion of an empirical example would demonstrate the real-world applicability of the measures.
-However, due to the existing length of the document, the addition of an empirical example (even if we replaced the toy example) would still add additional length: we would need to add a condensed methods and data section, and dedicate more room for tests.
+However, the length of the current document is an issue, so the addition of additional pages isn't feasible--especially given the concern over paper length that was raised by Reviewer 1 and how we’ve already reduced it by six pages in response to these concerns.
 
-The length of the current document is an issue, so the addition of additional pages isn't feasible--especially given the concern over paper length that was raised by Reviewer 1 and how we’ve already reduced it by six pages in response to these concerns.
-However, we've put great thought into how the manuscript could be restructured to included an additional empirical example, but it just isn't possible given the manuscript's current format.
+But, we've put great thought into how the manuscript could be restructured to included an additional empirical example.
+However, many of these ideas are not possible given the manuscript's current format.
+For instance, even if we replaced the toy example with an empirical example, it would still add quite significant additional length (i.e., explanation of data, room for tests).
+Further, a single empirical example is not sufficient to cover all cases and variants of the family.
 
-As such, we’ve clarified that the contribution of this work is primarily theoretical and methodological.
+As such, we’ve clarified that the contribution of this work: primarily theoretical and methodological.
 We’ve also tempered earlier claims about the measure’s real-world applicability, instead allowing readers to consider potential uses and the potential value of reuniting accessibility measures with interpretable units.
-We plan to explore an empirical application in a future publication which make this contribution more practically clear--where we may have enough space to develop and explore an empirical example without shortchanging the historical context.
+While we believe these measures utility will be highlighted through real-world examples--we plan to explore these in future publications which make this contribution more practically clear (i.e., where we may have enough space to develop and explore an empirical example without shortchanging the historical context).
 

### 2025-08-18 paper_plos/reviewer_respose_letter_V1.qmd
additional changes to the draft response letter

@@
 
-Thank you for this point. We've significantly revised the manuscript, focusing on explaining things concisely and ensuring each paragraph is move forward the key points of each section. In summary:
+Thank you for this point. We've significantly revised the manuscript to make it shorter and to ensure we articulate the key points of each section:
 
-- We've reduced the length of the manuscript from 34 to 28 pages (without references).
+- We've substantially reduced the length of the manuscript from 34 to 28 pages (without references).
 - We've removed tangential details and minimized the equations where possible, particularly in the “Early Research” section.
@@
 - We've removed redundancies where possible, especially in the "Family of Accessibility" section. 
-- We've more clearly highlighted the paper's main contribution: namely, a review of the shared origins of spatial interaction modeling and accessibility, how their interpretations have diverged over time, and the introduction of a family of accessibility measures grounded in spatial interaction principles--including total, singly, and doubly constrained cases, as well as variants of “accessible opportunities” and “accessible population".
+- We've more clearly highlighted the paper's main contribution: namely, a review of the shared origins of spatial interaction modeling and accessibility, how their interpretations have diverged over time, and the introduction of a family of accessibility measures grounded in spatial interaction principles.
 
-The key points of the manuscript can be summarised as follows:
+The key contributions of the manuscript can be summarised as follows:
 - First, it explicitly links accessibility to spatial interaction through their shared intellectual foundations (pre- @hansen1959), highlighting how contemporary accessibility research has diverged and showing the mathematical, intuitive, and interpretative advantages of preserving units through constrained formulations, as done in spatial interaction modeling.
-- Second, it introduces 'constraints' to accessibility measures.
-First: the total constrained accessibility measure as an alternative to the Hansen-type formulation, producing zonal values that represent proportions of the known number of destination masses (i.e., opportunities) or origin masses (i.e., population) in the region.
+- Second, it introduces 'constraints' to accessibility measures grounded in spatial interaction principles.
+<!-- First: the total constrained accessibility measure as an alternative to the Hansen-type formulation, producing zonal values that represent proportions of the known number of destination masses (i.e., opportunities) or origin masses (i.e., population) in the region.
 Second: the singly constrained accessibility measure as a competitive method that proportionally allocates opportunities from the destination based on population demand (from the origin), or in the case of market potential, proportional allocation of population from the origin based on the supply of opportunities at the destination.
-Its per-capita zonal rate is shown to be mathematically equivalent to the widely used two-step floating catchment area (2SFCA) method [@luo2003; @shen1998].
+Its per-capita zonal rate is shown to be mathematically equivalent to the widely used two-step floating catchment area (2SFCA) method [@luo2003; @shen1998].-->
 
@@
 
-As such, we’ve clarified that the contribution of this work: primarily theoretical and methodological.
+As such, we’ve worked to clarify that the contribution of this work which is primarily theoretical and methodological.
 We’ve also tempered earlier claims about the measure’s real-world applicability, instead allowing readers to consider potential uses and the potential value of reuniting accessibility measures with interpretable units.
-While we believe these measures utility will be highlighted through real-world examples--we plan to explore these in future publications which make this contribution more practically clear (i.e., where we may have enough space to develop and explore an empirical example without shortchanging the historical context).
+While we believe these measures utility would be highlighted through real-world examples--we plan to explore these in future publications which make this contribution more practically clear (i.e., where we may have enough space to develop and explore an empirical example without shortchanging the historical context).
+In fact, we are currently in conversation with the editor about the possibility to submit a separate paper demonstrating the proposed measures with an empirical case study.
 

### 2025-08-28 paper_plos/reviewer_respose_letter_V1.qmd
rafa on response letter

@@
 
-Thank you for your responses. Below are replies, with our response in black font and the reviewers comments in \textcolor{blue}{blue}.
+Dear editors,
+
+We appreciate the valuable comments from the reviewers.
+We have done our best to address all comments and suggestions as much as possible, which we believe has significantly improved the paper.
+We address the individual comments of the reviewers one by one (list below), but we would also like to note three major points in response to the reviewers’ recommendations.
+
+- We have substantially shortened the paper from 34 to 28 pages.
+This involved streamlining the manuscript's text, removing equations and text redundancies where possible, especially in the "Family of Accessibility" section.
+- We have also edited the text to more clearly articulate the paper’s main contribution and to make each section’s purpose clearer.
+- While we agree with reviewer’s #2  that including an empirical example would demonstrate the real-world applicability of our proposed measures, we were not able to include an empirical case study in the paper due to manuscript length limits.
+
+As a whole, we believe the paper is more consistent after heeding the reviewer’s comments.
+We hope you and the reviewers will agree.
+Please see below the point-by-point replies, with our response in black font and the reviewers comments in \textcolor{blue}{blue}.
 
@@
 
-The key contributions of the manuscript can be summarised as follows:
+The key contributions of the manuscript are highlighted in the introduction of the paper (see paragraph starting at line 48) and can be summarised as follows:
 - First, it explicitly links accessibility to spatial interaction through their shared intellectual foundations (pre- @hansen1959), highlighting how contemporary accessibility research has diverged and showing the mathematical, intuitive, and interpretative advantages of preserving units through constrained formulations, as done in spatial interaction modeling.
@@
 
-Furthermore, regarding the equations--many have now been removed when possible if they don't directly further the argument of the section or are redundant.
+Furthermore, regarding the equations -- 10 equations have now been removed because they did not directly further the argument of the section or are redundant.
 For instance, some equations in "Early research on human spatial interaction" have been removed, and rewritings of Wilson's general formula from "Wilson's family of spatial interaction models" section have been removed.
@@
 Thank you for this comment.
-The historical literature review section (Accessibility and spatial interaction modelling: two divergent research streams) covers upward of 76 papers citing Wilson and Hansen, condensed into about 3 pages.
+
+We’ve removed tangential details and minimized the equations where possible, particularly in the “Early Research” section.
+The historical literature review section (Accessibility and spatial interaction modelling: two divergent research streams) now covers upward of 76 papers citing Wilson and Hansen, condensed into about 3 pages.
 This section sets the stage by outlining Wilson’s balancing factors and highlighting how place-based accessibility literature (largely based on Hansen-type measures) has not adopted these balancing factors.
@@
 Thank you for this fair point.
-In the newly revised manuscript, we make it more clear that our focus is on place-based accessibility, which remains widely used by practitioners and prominent in the literature, as well as Wilson's spatial interaction model which is also popular in spatial interaction modeling still to this day.
-Given the broad scope and space constraints (the current manuscript is already quite long and conceptually broad), incorporating additional detailed discussions of novel approaches is challenging, and somewhat tangential.
+In the revised manuscript, we state explicitly that our focus is on place-based accessibility, which remains widely used in practice and prominent in the literature, as well as Wilson's spatial interaction model which is also popular in spatial interaction modeling still to this day.
+Given the broad scope of the paper and space constraints, incorporating additional detailed discussions of novel approaches is challenging, and somewhat tangential.
 
 However, we do acknowledge some utility-based approaches within the spatial interaction literature.
-For example, "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
+For example, on page 11 "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
 
@@
 
+\textcolor{red}{Rafa: perhaps we can add a sentence in the conclusion saying somehting like: "Recent developments in the literature include novel approaches to utility and person-based accessibility, including time-sensitive behavioral or utility-based approaches, that are not based on Newtonian gravitation nor on Wilson's balancing factors.
+Future studies could further explore how these ideas of balacing factors/spatial constraints could helo inform the new models."}
+
 \textcolor{blue}{5.
@@
 
-This is a very fair point. Inside of explaining results with an empirical example, we theoretical demonstrate this throughout the paper.
+Thank you for giving us the opportunity to clarify this point.
+The argument that constrained models attaches meaningful units to accessibility measure is theoretically demonstrated in the paper, and can be checked with careful examination of the equations.
+Nonetheless, we also believe this is demonstrated with the explanation of the toy example used in the paper.
 
@@
 
-Excellent point. We've decided to remove Table 1 all together, and provide a more condensed version of it in text on pg 13/14. Namely:
+\textcolor{red}{Rafa: I actually liked table 1 and thought perhaps we should keep the table and remove the redundancy of the text around it. But I'm Ok with this solution below."}
+
+Excellent point.
+We've streamlined redundancies throughout the manuscript, especially in the "A family of accessibility measures" section.
+Specifically regarding table 1, we've decided to remove Table 1 all together, and provide a more condensed version of it in text on pg 13/14.
+Namely:
 
@@
 
-We hope this addresses your concern! As well, we've streamlined redundancies throughout the manuscript, especially in the "A family of accessibility measures" section.
+We hope this addresses your concern!

### 2025-08-28 paper_plos/paper_plos.qmd
rafa up to page 10

@@
 
+\textcolor{red}{Rafa: Do we have to shorthen the abstract?
+I actually preferred the abstract we had in the preprint.
+I like that the abstract mentions the toy example though, which was not done in the 1st version.}
+
 abstract: |
@@
 
-In the early nineteenth century, rapid urbanisation driven by industrialisation expanded cities across Great Britain, Northern Europe, and North America.
-This expansion led to traffic congestion and the emergence of a transportation planning paradigm focused primarily on mobility.
+In the early twentieth century, the emergence of a transportation planning paradigm focused primarily on mobility cemented major investments in automobile and transportation infrastructure, fostering lower-density sprawl, car-dependent development and entrenching automobility in planning practice [@miller_collaborative_2011; @lavery_driving_2013].
 Within this new practice, access to destinations was treated as a by-product of movement.
-After World War II, major investments in automobile and transportation infrastructure cemented this mobility-oriented model, fostering flight from urban congested cities, lower-density sprawl, car-dependent development and entrenching automobility in planning practice [@miller_collaborative_2011; @lavery_driving_2013].
 Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people’s ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
@@
 Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024].
-While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret difference in scores between modes, and how should results be communicated?
+While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret differences in accessibility scores between modes, and how should results be communicated?
 
@@
 Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces.
-Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's experiments of 1778 [@hutton_xxxiii_1778, p.
-783] and Cavendish's 1798 experiment [@cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy.
+Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's and
+Cavendish's experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy.
 In sum: it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant.
@@
 
-Following Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts.
+Following Carey's work [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts.
 Namely, since Carey to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
@@
 
-A common feature of these early investigations of human spatial interaction is that a proportionality constant, similar to $G$ in @eq-phys-grav was never considered.
-Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms.
-Reilly's hypothesis was presented in the following form:
+Of the researchers cited above, only Reilly and Zipf expressed their hypotheses in mathematical terms. Reilly's hypothesis was presented in the following form:
 
@@
 These early formulations (@eq-reilly, @eq-zipf) clearly reflect the influence of Newtonian gravity on human spatial interaction theory, revealing a shared mathematical structure across migration, trade, and communication models.
-However, none included a proportionality constant--highlighting a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities.
-
-After Carey and folowing Reilly and Zipf, it is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
+
+However, a common feature of these early investigations is that none of them included a proportionality constant (similar to $G$ in @eq-phys-grav), a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities.
+It is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
 <!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
---> Besides awareness of preceding research (i.e., Stewart cites both Reilly and Zipf as predecessors in the analysis of human spatial interaction), Stewart appears to have been the first author to express his theorized relationships for human spatial interaction with a proportionality constant $G$, enabling his formulation to be interpreted as a measurable 'demographic' force:
+--> While acknowledging predecessors like Reilly and Zipf, the physicist Stewart was likely the first author to formalize human spatial interaction using an explicit proportionality constant $G$, enabling his formulation to be interpreted as a measurable 'demographic' force:
 
@@
 34 and then later as _demographic energy_ on p.
-53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight $\mu$ of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
+53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
 35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration.
@@
 
+\textcolor{red}{Rafa: there is no $\beta$ in this equation. Check text in paragraph below.}
+
 $S_{i}$ in @eq-accessibility-general is a measure of the accessibility from site $i$.
-This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $\beta$ (a parameter that modulates the friction of cost).
+This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $f(d_{ij})$ (a function that modulates the friction of cost).
 Today, Hansen is frequently cited as the father of modern accessibility analysis [e.g., @reggianiGuestEditorialNew2011], and Hansen-type accessibility is commonly referred to as the gravity-based accessibility measure.
@@
 Therefore in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
-The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement.
+The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement that hinders the interpretability of accessibility measures.
 
@@
 Wilson.
-Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling--focused on modeling flows of interaction between places.
+Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling focused on flows of interaction between places.
 This work was not focused on the 'potential' concept as accessibility has been.
@@
 In other words, the different balancing factors adjust the model so that flows become statistical averages consistent with observed origin and/or destination data.
+
 In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.
@@
 In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
-While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that _indicate_ a proportional relationship and not one of equality nor yielding results without interpretable units.
+While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that reflect proportional—not equal—relationships and typically lack interpretable units.
 

### 2025-08-28 paper_plos/paper_plos.qmd
rafa's comments and edits

@@
 Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions.
-This trend is common: for instance it is done in [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+This trend is common: for instance it is done in [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 However, these works do not engage with spatial interaction beyond this attribution.
@@
 Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility’s potential for spatial interaction--but only superficially.
-These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@grengs2010job; @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
 Indeed, while accessibility can be seen as the *potential* for spatial interaction--and Wilson [@wilson1971] briefly touches on this--such mentions have not resulted in deeper analytical integration of these concepts.
@@
 However, these balancing factors were deduced intuitively.
-These works do not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
+These works did not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
 This equivalence is only discovered in hindsight, as will be demonstrated in the following section.
@@
 
-In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, the spatial interaction-oriented Hansen [@hansen1959] citing literature nor the accessibility-oriented literature that cite Wilson [@wilson1971] have yet to meaningfully adopt Wilson's constraint-based logic to the concept of accessibility.
+In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility.
 So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
@@
 This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
-
 This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
-Rather than relying on arbitrary or uncalibrated proportional relationships (or invoking a universal constant $G$ as proposed in Stewart [@stewartDemographicGravitationEvidence1948]) we adopt Wilson’s approach grounded in statistical mechanics.
 

### 2025-08-28 paper_plos/paper_plos.tex
rafa's comments and edits

@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
 \usepackage[numbers,square,comma]{natbib}
@@
 \linenumbers
+
 \section{Introduction}\label{introduction}
 
-In the early nineteenth century, rapid urbanisation driven by
-industrialisation expanded cities across Great Britain, Northern Europe,
-and North America. This expansion led to traffic congestion and the
-emergence of a transportation planning paradigm focused primarily on
-mobility. Within this new practice, access to destinations was treated
-as a by-product of movement. After World War II, major investments in
-automobile and transportation infrastructure cemented this
-mobility-oriented model, fostering flight from urban congested cities,
+In the early twentieth century, the emergence of a transportation
+planning paradigm focused primarily on mobility cemented major
+investments in automobile and transportation infrastructure, fostering
 lower-density sprawl, car-dependent development and entrenching
 automobility in planning practice
-\citep{miller_collaborative_2011, lavery_driving_2013}. Despite
-continued road and highway expansion, this automobility monoculture has
-proven ineffective at reducing travel costs or environmental burdens,
-and has not clearly improved people's ability to reach destinations
+\citep{miller_collaborative_2011, lavery_driving_2013}. Within this new
+practice, access to destinations was treated as a by-product of
+movement. Despite continued road and highway expansion, this
+automobility monoculture has proven ineffective at reducing travel costs
+or environmental burdens, and has not clearly improved people's ability
+to reach destinations
 \citep{farber_running_2011, handyACCESSIBILITYVSMOBILITYENHANCING2002, paez_healthcare_2010}.
@@
 raised in \citep{vanweeAccessible2016}: How should practitioners
-interpret difference in scores between modes, and how should results be
-communicated?
+interpret differences in accessibility scores between modes, and how
+should results be communicated?
 
@@
 earth was between five or six times that of water, an assumption that
-received support after Hutton's experiments of 1778
-\citep[p.~783]{hutton_xxxiii_1778} and Cavendish's 1798 experiment
-\citep{cavendish_xxi_1798} to estimate of the proportionality constant
-to within 1\% accuracy. In sum: it took over a century from the
-publication of \emph{Principia} to refine the estimate of the
-proportionality constant.
+received support after Hutton's and Cavendish's experiments in the late
+1700s \citep{hutton_xxxiii_1778, cavendish_xxi_1798} to estimate of the
+proportionality constant to within 1\% accuracy. In sum: it took over a
+century from the publication of \emph{Principia} to refine the estimate
+of the proportionality constant.
 
@@
 
-Following Carey's \emph{Principles of Social Science}
-\citep{careyPrinciplesSocialScience1858}, research into human spatial
-interaction continued in different contexts. Namely, since Carey to the
-1940s, a number of researchers theoretically and empirically attempted
-to characterise human spatial interaction as a force \(F\) directly
-proportional to the ``masses'' \(M_i\) and \(M_j\) of two locations, and
-inversely proportional to their separation distance--conceptually
-parallel to Newtonian gravity, but often omitting a proportionality
-constant.
+Following Carey's work \citep{careyPrinciplesSocialScience1858},
+research into human spatial interaction continued in different contexts.
+Namely, since Carey to the 1940s, a number of researchers theoretically
+and empirically attempted to characterise human spatial interaction as a
+force \(F\) directly proportional to the ``masses'' \(M_i\) and \(M_j\)
+of two locations, and inversely proportional to their separation
+distance--conceptually parallel to Newtonian gravity, but often omitting
+a proportionality constant.
 
@@
 
-A common feature of these early investigations of human spatial
-interaction is that a proportionality constant, similar to \(G\) in
-Eq~\ref{eq-phys-grav} was never considered. Of the researchers cited
-above, only Reilly and Zipf expressed their hypotheses in mathematical
-terms. Reilly's hypothesis was presented in the following form:
+Of the researchers cited above, only Reilly and Zipf expressed their
+hypotheses in mathematical terms. Reilly's hypothesis was presented in
+the following form:
 
@@
 theory, revealing a shared mathematical structure across migration,
-trade, and communication models. However, none included a
-proportionality constant--highlighting a consistent omission of the
-empirical calibration necessary to convert these proportional
-relationships into measurable and comparable quantities.
-
-After Carey and folowing Reilly and Zipf, it is only in Stewart
+trade, and communication models.
+
+However, a common feature of these early investigations is that none of
+them included a proportionality constant (similar to \(G\) in
+Eq~\ref{eq-phys-grav}), a consistent omission of the empirical
+calibration necessary to convert these proportional relationships into
+measurable and comparable quantities. It is only in Stewart
 \citep{stewartDemographicGravitationEvidence1948} that we find the most
 explicit connection yet to Newton's Gravitational law and the use of a
-proportionality constant. Besides awareness of preceding research (i.e.,
-Stewart cites both Reilly and Zipf as predecessors in the analysis of
-human spatial interaction), Stewart appears to have been the first
-author to express his theorized relationships for human spatial
-interaction with a proportionality constant \(G\), enabling his
-formulation to be interpreted as a measurable `demographic' force:
+proportionality constant. While acknowledging predecessors like Reilly
+and Zipf, the physicist Stewart was likely the first author to formalize
+human spatial interaction using an explicit proportionality constant
+\(G\), enabling his formulation to be interpreted as a measurable
+`demographic' force:
 
@@
 assumptions (e.g., view of humans as particles following physical laws
-and assumptions of the molecular weight \(\mu\) of the average American
-being one, but ``presumably\ldots much less than one\ldots.for an
-Australian aborigine'' {[}p.~35{]})--his introduction of a
-proportionality constant \(G\) in modeling demographic force marks an
-important conceptual step: recognizing that moving from proportionality
-to equality requires empirical calibration. In other words, the addition
-of \(G\) shifts results from being abstract indicators of potential
-(i.e., \(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having
-interpretable units grounded in consistent, albeit still abstract,
-quantities (i.e., units of demographic force).
+and assumptions of the molecular weight of the average American being
+one, but ``presumably\ldots much less than one\ldots.for an Australian
+aborigine'' {[}p.~35{]})--his introduction of a proportionality constant
+\(G\) in modeling demographic force marks an important conceptual step:
+recognizing that moving from proportionality to equality requires
+empirical calibration. In other words, the addition of \(G\) shifts
+results from being abstract indicators of potential (i.e.,
+\(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having interpretable
+units grounded in consistent, albeit still abstract, quantities (i.e.,
+units of demographic force).
 
@@
 
+\textcolor{red}{Rafa: there is no $\beta$ in this equation. Check text in paragraph below.}
+
 \(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
@@
 of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
-travel time, incurred to reach \(j\) from \(i\)), and \(\beta\) (a
-parameter that modulates the friction of cost). Today, Hansen is
+travel time, incurred to reach \(j\) from \(i\)), and \(f(d_{ij})\) (a
+function that modulates the friction of cost). Today, Hansen is
 frequently cited as the father of modern accessibility analysis
@@
 and practical gap in accessibility analysis: a missing link between
-theoretical form and empirical measurement.
+theoretical form and empirical measurement that hinders the
+interpretability of accessibility measures.
 
@@
 \citep{wilson1971}, advanced a general framework for spatial interaction
-modeling--focused on modeling flows of interaction between places. This
-work was not focused on the `potential' concept as accessibility has
-been. Wilson \citep{wilson1971} formalized the general spatial
-interaction model through the following equation:
+modeling focused on flows of interaction between places. This work was
+not focused on the `potential' concept as accessibility has been. Wilson
+\citep{wilson1971} formalized the general spatial interaction model
+through the following equation:
 
@@
 balancing factors adjust the model so that flows become statistical
-averages consistent with observed origin and/or destination data. In the
-framework introduced and inferred from Wilson \citep{wilson1971}, three
-types of balancing factors are specified: (1) an unconstrained model
-that only matches the total volume of interaction \(K\), (2) a singly
-constrained model (either by origins \(A_i\) or destinations \(B_j\)),
-and (3) a doubly constrained model that satisfies both.
+averages consistent with observed origin and/or destination data.
+
+In the framework introduced and inferred from Wilson \citep{wilson1971},
+three types of balancing factors are specified: (1) an unconstrained
+model that only matches the total volume of interaction \(K\), (2) a
+singly constrained model (either by origins \(A_i\) or destinations
+\(B_j\)), and (3) a doubly constrained model that satisfies both.
 
@@
 the system of analysis, Hansen-type measures (still widely used today)
-yield outputs that \emph{indicate} a proportional relationship and not
-one of equality nor yielding results without interpretable units.
+yield outputs that reflect proportional---not equal---relationships and
+typically lack interpretable units.
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 instance it is done in
-\citep{weibullNumericalMeasurementAccessibility1980, handyMeasuringAccessibilityExploration1997, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
+\citep{handyMeasuringAccessibilityExploration1997, weibullNumericalMeasurementAccessibility1980, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
 However, these works do not engage with spatial interaction beyond this
@@
 go beyond this recognition in the scope of their works e.g.,
-\citep{millerMeasuringSpaceTimeAccessibility1999, giuliano2010accessibility, grengs2010intermetropolitan, grengs2010job, grengs2012equity, levine2012does, levinson2012positive, tong2015transportation, liu2015spatial, he2017measuring, wuUnifyingAccess2020, ng2022reflection, naqavi2023mobility, suel2024measuring}.
+\citep{grengs2010job, millerMeasuringSpaceTimeAccessibility1999, giuliano2010accessibility, grengs2010intermetropolitan, grengs2012equity, levine2012does, levinson2012positive, tong2015transportation, liu2015spatial, he2017measuring, wuUnifyingAccess2020, ng2022reflection, naqavi2023mobility, suel2024measuring}.
 Indeed, while accessibility can be seen as the \emph{potential} for
@@
 number of opportunities in the region. However, these balancing factors
-were deduced intuitively. These works do not explicitly state that the
+were deduced intuitively. These works did not explicitly state that the
 mathematical formulation of the equations are effectively equivalent to
@@
 In sum, despite the interpretative advantaged offered by the statistical
-logic of Wilson \citep{wilson1971}'s framework, the spatial
-interaction-oriented Hansen \citep{hansen1959} citing literature nor the
-accessibility-oriented literature that cite Wilson \citep{wilson1971}
-have yet to meaningfully adopt Wilson's constraint-based logic to the
-concept of accessibility. So, in the next section, we demonstrate how
-applying Wilson's framework enables accessibility to move from a
-proportional relationship to calibrated equality--tying outputs to
-tangible system knowns. This re-expression of accessibility using
-constraints yields interpretable, unit-consistent measures of
-opportunity. This approach takes the same path of entropy-maximisation
-as in Wilson \citep{wilson1971}, and does not rely on specifying some
-universal constant \(G\) like initially proposed in Stewart
+logic of Wilson \citep{wilson1971}'s framework, neither the spatial
+interaction literature that cites Hansen \citep{hansen1959} nor the
+accessibility literature citing Wilson \citep{wilson1971} has
+meaningfully applied Wilson's constraint-based logic to the concept of
+accessibility. So, in the next section, we demonstrate how applying
+Wilson's framework enables accessibility to move from a proportional
+relationship to calibrated equality--tying outputs to tangible system
+knowns. This re-expression of accessibility using constraints yields
+interpretable, unit-consistent measures of opportunity. This approach
+takes the same path of entropy-maximisation as in Wilson
+\citep{wilson1971}, and does not rely on specifying some universal
+constant \(G\) like initially proposed in Stewart
 \citep{stewartDemographicGravitationEvidence1948} (recall,
@@
 accessibility (a la Hansen \citep{hansen1959}) from a relationship
-describing proportionality to equality.
-
-This shift addresses the fundamental issue of interpretability
-associated with the proportional nature of Hansen-type accessibility
-indicators we previous outlined. Rather than relying on arbitrary or
-uncalibrated proportional relationships (or invoking a universal
-constant \(G\) as proposed in Stewart
-\citep{stewartDemographicGravitationEvidence1948}) we adopt Wilson's
-approach grounded in statistical mechanics.
+describing proportionality to equality. This shift addresses the
+fundamental issue of interpretability associated with the proportional
+nature of Hansen-type accessibility indicators we previous outlined.
 
@@
 captures close to an average amount of physicians, a third of 490 or
-\textasciitilde163). This averaging of the total amount is how the total
-constrained proportional allocation factor functions. However, what's
-notable is how Zones change between scenarios. For instance, Zone 1's
-share only declines slightly--these declines are outpaced by Zone 2's
-relative gains. This shift reflects how \(\kappa_{ij}^T\) redistributes
-opportunities in proportion to impedance under different travel
-behaviours.
+\textasciitilde163). In terms of proportional magnitude, this can also
+be observed in the unconstrained measure for this scenario. However, for
+the total constrained measure, this plateauing of results have meaning.
+In fact, each zone is allocated an average of the total amount in the
+region, as a result of the total constrained proportional allocation
+factor.
+
+However, what's notable is how Zones change between scenarios. For
+instance, Zone 1's share only declines slightly--these declines are
+outpaced by Zone 2's relative gains. This shift reflects how
+\(\kappa_{ij}^T\) redistributes opportunities in proportion to impedance
+under different travel behaviours.
 

### 2025-08-28 paper_plos/paper_plos.qmd
Ready for Chris to submit

@@
 Within this new practice, access to destinations was treated as a by-product of movement.
-Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people’s ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
-
-In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic, etc.) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].
+Despite continued road and highway expansion, this automobility monoculture has proven ineffective at reducing travel costs or environmental burdens, and has not clearly improved people's ability to reach destinations [@farber_running_2011; @handyACCESSIBILITYVSMOBILITYENHANCING2002; @paez_healthcare_2010].
+
+In response, transportation researchers have increasingly advocated for the adoption of accessibility as a planning criterion, in contrast to traditional mobility-oriented transportation planning approaches which translate into indicators that benchmark movement (e.g., vehicle kilometres traveled, intersection through traffic) which are not necessarily linked to improved accessibility [@silvaAccessibilityInstrumentsPlanning2017; @paez_developing_2013; @handy2020; @elgeneidyMakingAccessibilityWork2022].
 Accessibility, by contrast, is the "potential of opportunities for [spatial] interaction" [@hansen1959].
@@
 
-Accessibility research has expanded across diverse domains including: employment [@karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], healthcare [@luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic], green space [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], education [@williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contact [@neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economics [@vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011], among many other domains of application.
+Accessibility research has expanded across diverse domains including: employment [@karstEvaluationAccessibilityImpacts2003; @grengs2010job; @paez_jobs_2013; @merlin2017competition; @tao_investigating_2020], healthcare [@luo2003; @paez_healthcare_2010; @wan2012three; @delamater2013spatial; @boisjoly2017informality; @pereira_2021_geographic; @yang2024evaluating], green space [@reyesAccessibility2014; @rojas_accessibility_2016; @liang_novel_2024], education [@williams_disparities_2014; @romanillosAccessibilitySchoolsSpatial2018; @marques_accessibility_2021], social contact [@neutens_human_2007; @farberActivitySpacesMeasurement2012; @farber_2013_social], and regional economics [@vickermanAccessibility1999; @lopezMeasuring2008; @ribeiro_road_2010; @gutierrez_evaluating_2011], among many other domains of application.
 Despite its popularity in scholarly works, accessibility still remains difficult to implement in planning due to definitional inconsistencies [@vanweeAccessible2016; @handy2020; @kapatsila_resolving_2023] and challenges in interpreting and communicating results [@geursAccessibilityEvaluationLanduse2004; @vanweeAccessible2016; @ferreiraReenactingMobilityAccessibility2020].
 
-More specifically, the wide range of accessibility definitions--with novel methods being more sophisticated but less intuitive [@kapatsila_resolving_2023]--can further hinder practical uptake [@vanweeAccessible2016].
+More specifically, the wide range of accessibility definitions, with novel methods being more sophisticated but less intuitive [@kapatsila_resolving_2023], can further hinder practical uptake [@vanweeAccessible2016].
 Geurs and van Wee [@geursAccessibilityEvaluationLanduse2004] classify accessibility measures into four categories: infrastructure-, place-, person-, and utility-based.
@@
 While spatial interaction models embraced such constraints to improve interpretability, most accessibility models have not.
-This lack of adoption contributes to fuzziness in current analyses, limiting interpretive clarity to simple proportional comparisons (e.g., “higher-than”, “lower-than”) [@millerAccessibilityMeasurementApplication2018].
+This lack of adoption contributes to fuzziness in current analyses, limiting interpretive clarity to simple proportional comparisons (e.g., "higher-than", "lower-than") [@millerAccessibilityMeasurementApplication2018].
 Without such constraints, accessibility scores lack clear units and comparability across cities or modes.
@@
 Accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
-Following this introduction, this paper revisits that lineage--from Ravenstein’s Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959] to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
+Following this introduction, this paper revisits that lineage--from Ravenstein's Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959] to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
 We illustrate each case of the family with a numerical toy example and conclude by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.
@@
 Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces.
-Newton's initial estimate of $G$ was based on a speculation that the mean density of earth was between five or six times that of water, an assumption that received support after Hutton's and
-Cavendish's experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] to estimate of the proportionality constant to within 1% accuracy.
-In sum: it took over a century from the publication of *Principia* to refine the estimate of the proportionality constant.
-
-The Newtonian gravitational relationship laid the conceptual groundwork for later empirical studies of human spatial interaction, such as those by Ravenstein, Stewart, and others, who began translating these principles into observed patterns of human movement.
-However, the majority of these attempts described a proportional relationship, or one arbitrarily set to equality.
+Newton's initial estimate of $G$ was based on a speculation <!--that the mean density of earth was between five or six times that of water, an assumption that -->but received empirical support after Hutton's and Cavendish' experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] which estimated it to within 1% accuracy.
+That is, it took over a century from the publication of Newton's *Principia Mathematica* to refine the estimate of the proportionality constant.
+
+While the Newtonian gravitational relationship laid the conceptual groundwork for later empirical studies of human spatial interaction, the majority of these early attempts described a proportional relationship or one arbitrarily set to equality.
+They did not establish an empirical $G$ as in the Newtownian tradition, as will be discussed in the next subsection.
 
@@
 
-Following Carey's work [@careyPrinciplesSocialScience1858], research into human spatial interaction continued in different contexts.
-Namely, since Carey to the 1940s, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
+Henry C.
+Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858] inspired empirical spatial intereaction research in different contexts.
+Namely, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
 
@@
 However, a common feature of these early investigations is that none of them included a proportionality constant (similar to $G$ in @eq-phys-grav), a consistent omission of the empirical calibration necessary to convert these proportional relationships into measurable and comparable quantities.
-It is only in Stewart [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
+It is only in Stewart (1948) [@stewartDemographicGravitationEvidence1948] that we find the most explicit connection yet to Newton's Gravitational law and the use of a proportionality constant.
 <!-- If we were to ponder as of to _why_: it may relate to academic backgrounds; where Stewart was a physicist while Ravenstein, Reilly, and Zipf were social scientists.
@@
 
-Although Stewart’s concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p.
+Although Stewart's concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p.
 34 and then later as _demographic energy_ on p.
@@
 
-As will be discussed, when Hansen [@hansen1959] later adopted Stewart’s formula (@eq-stewart-population-potential-sum) for accessibility, he omitted any mention of $G$, effectively setting it to 1 arbitrarily as Stewart had done.
+As will be discussed, when Hansen (1959) [@hansen1959] later adopted Stewart's formula (@eq-stewart-population-potential-sum) for accessibility, he omitted any mention of $G$, effectively setting it to 1 arbitrarily as Stewart had done.
 This omission has persisted in accessibility research, leaving a conceptual gap in how such measures are interpreted and compared.
@@
 As well as being a student of city and regional planning at the Massachusetts Institute of Technology, Hansen was also an engineer with the Bureau of Roads, and preoccupied with the power of transportation to shape land uses in a very practical sense.
-Hansen [@hansen1959] drew directly from Stewart’s population potential formula (see @eq-stewart-population-potential-sum), but left aside the broader (and often problematic) framework of “social physics”.
+Hansen [@hansen1959] drew directly from Stewart's population potential formula (see @eq-stewart-population-potential-sum), but left aside the broader (and often problematic) framework of "social physics".
 
 Hansen recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more specifically, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting to accounts for a variety of impedance functions beyond the inverse power $d^{-\beta}$ as follows:
+Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more specifically, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting of $d^{-\beta}$ to $f(d_{ij})$ accounts for a variety of impedance functions beyond the inverse power as follows:
 
@@
 $$ {#eq-accessibility-general}
-
-\textcolor{red}{Rafa: there is no $\beta$ in this equation. Check text in paragraph below.}
 
@@
 
-However, Hansen’s formulation carried forward a crucial omission that continues to affect the literature: the proportionality constant $G$ included in Stewart’s original formulation (@eq-stewart-population-potential-sum) has vanished entirely.
+However, Hansen's formulation carried forward a crucial omission that continues to affect the literature: the proportionality constant $G$ included in Stewart's original formulation (@eq-stewart-population-potential-sum) has vanished entirely.
 Although Stewart included $G$ explicitly (with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
 34].) Hansen made no mention of it.
-As a result, modern accessibility analysis has largely evolved without addressing the constant’s role, leaving $G$ effectively and implicitly fixed at 1.
+As a result, modern accessibility analysis has largely evolved without addressing the constant's role, leaving $G$ effectively and implicitly fixed at 1.
 This omission has significant implications.
@@
 Wilson.
-Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling focused on flows of interaction between places.
-This work was not focused on the 'potential' concept as accessibility has been.
+Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
+Additionally, this work was not focused on the 'potential' concept, as accessibility has been.
 Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:
@@
 Of important note for this paper, $k$ in @eq-phys-gravity-model acts as a proportionality constant, shifting the equation from a proportional to an equal relationship by incorporating known system totals.
-In this sense, we infer that $k$ serves a role similar to the gravitational constant in Newton’s law--it calibrates the model so that outputs match real-world quantities.
-However, these real-world quantities was not a set empirical constant (like Newton's $G$) but instead, sensitive to the system and known information about the system.
+In some sense, $k$ serves a role similar to the gravitational constant in Newton's law--it calibrates the model so that outputs match real-world quantities.
+However, these real-world quantities are not a set empirical constant (like Newton's $G$) but instead, sensitive to the system and known information about the system.
 
 From the outset, spatial interaction models emphasized interpretability of results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
-But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson’s innovation was to ground the model in _entropy maximisation_.
+But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson's innovation was to ground and generalise the model using _entropy maximisation_.
 By maximizing the number of ways individual trip probabilities could be arranged under known constraints, Wilson derived models that estimate _statistical averages_ of flows between zones [@wilson1971; @seniorGravityModellingEntropy1979].
@@
 
-Wilson’s work is notable for many reasons, one of which is his departure from the metaphor of Newtonian gravitational physics.
-Instead, he grounded his models in statistical mechanics, particularly the principle of entropy maximisation.
-For accessibility research, Wilson’s approach is especially significant because it can address a problem that has quietly afflicted this field (as this paper has argued): the transition from proportionality to equality.
-
-Rather than relying on a universal constant like $G$, Wilson’s models calibrate interaction flows through known empirical constraints (e.g., total system flows, or origin-side and destination-side flow totals).
+Wilson's work is notable for many reasons.
+Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation.
 This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-In short, when aspects of the systems are inputs into the balancing factors, $ij$ flows can be robustly predicted in tangible units of 'flow', units that align with the system being modelled.
 In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
-While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today) yield outputs that reflect proportional—not equal—relationships and typically lack interpretable units.
+While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional--not equal--relationships and typically lack interpretable units.
 
@@
 
-In this way, these balancing factors can be used to recover what Stewart’s $G$ was meant to accomplish, but through a different conceptual path.
+In this way, these balancing factors can be used to recover what Stewart's $G$ was meant to accomplish, but through a different conceptual path.
 But before demonstrating a family of accessibility measures that are sensitive to constraints, in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
@@
 Despite their close conceptual ties, the accessibility and spatial interaction modeling literatures have developed along largely separate paths since the 1970s.
-Hansen’s [@hansen1959] formulation of accessibility became the method used for decades of work on transport equity, land use analysis, and urban accessibility planning.
-Meanwhile, Wilson’s [@wilson1971] entropy-maximising framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
+Hansen's [@hansen1959] formulation of accessibility became the method used for decades of work on transport equity, land use analysis, and urban accessibility planning.
+Meanwhile, Wilson's [@wilson1971] entropy-maximising framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
 We argue the framework's quiet innovation--introducing empirically grounded constraints to shift from proportionality to calibrated equality--made the framework immediately relevant for policy applications as outputs were in tangible units.
@@
 As noted in [@battyChronicleScientificPlanning1994], large scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
-And these needs were far from trivial: in the U.S., for instance, the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today’s dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+And these needs were far from trivial: for instance in the U.S., the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today's dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
 In this context, spatial interaction models were incorporated into institutional practices focused on "predict and provide" travel demand forecasting [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
@@
 
-To explore why Wilson’s approach never crossed over to accessibility modeling, we conduct a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both (on Web of Science using the "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]) .
+To explore why Wilson's approach never crossed over to accessibility modeling, we conduct a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both (on Web of Science using the "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]) .
 Only `r (n_b) |> prettyNum(big.mark = ",")` out of the `r (n_h+n_w+n_b) |> prettyNum(big.mark = ",")` documents that emerged from our search cite both.
@@
 
-Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility’s potential for spatial interaction--but only superficially.
+Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility's potential for spatial interaction--but only superficially.
 These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@grengs2010job; @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
@@
 However, these balancing factors were deduced intuitively.
-These works did not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson’s singly constrained model (derived from entropy maximization).
+These works did not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson's singly constrained model (derived from entropy maximization).
 This equivalence is only discovered in hindsight, as will be demonstrated in the following section.
@@
 In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility.
-So, in the next section, we demonstrate how applying Wilson’s framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
+So, in the next section, we demonstrate how applying Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
 This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
@@
 Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter.
-This section introduces this reframing: defining a family of accessibility measures using Wilson’s approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
+This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
 This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
@@
 1. **Unconstrained Case ($V_i^0$, $M_j^0$)**
-- Equivalent to Hansen’s [@hansen1959] and Reilly’s [@reilly1929methods] original formulations; the status quo of accessibility modelling.
+- Equivalent to Hansen's [@hansen1959] and Reilly's [@reilly1929methods] original formulations; the status quo of accessibility modelling.
 - No balancing factors applied; units are in "opportunities-by-impedance" for $V_i^0$ or "population-by-impedance" for $M_j^0$.
@@
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
-Represents ‘interaction capacity’ or ‘realized access’ serving as predictions of real interaction flows.
+Represents ‘interaction capacity' or ‘realized access' serving as predictions of real interaction flows.
 
@@
 First, Table @tbl-small-system-land-use summarizes population (in 10,000s) and physicians per zone.
-For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)` comparable to Canada’s 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
+For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)` comparable to Canada's 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
 Second, Table @tbl-small-system-cost shows travel times (minutes); it can be discerned that Zones 1 and 3 are closer to each other than to Zone 2.
@@
 
-However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are ‘opportunities weighted by travel impedance’ and lack meaningful scaling or direct interpretability.
+However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are ‘opportunities weighted by travel impedance' and lack meaningful scaling or direct interpretability.
 Comparing values across different decay functions or contexts (i.e., different number of zones) is therefore limited to ordinal statements (more vs.
@@
 Comparing across decay types is meaningless in absolute terms.
-For instance, the difference in zone 1 (edge of urban core)’s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
+For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
 These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
@@
 
-Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $K^T$, which scales Hansen’s unconstrained accessibility $V_i^0$ to meet the total opportunity constraint:
+Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $K^T$, which scales Hansen's unconstrained accessibility $V_i^0$ to meet the total opportunity constraint:
 $$
@@
 However, what's notable is how Zones change between scenarios.
-For instance, Zone 1’s share only declines slightly--these declines are outpaced by Zone 2’s relative gains.
+For instance, Zone 1's share only declines slightly--these declines are outpaced by Zone 2's relative gains.
 This shift reflects how $\kappa_{ij}^T$ redistributes opportunities in proportion to impedance under different travel behaviours.
 
-The total constrained accessibility measure resolves the interpretability issue of Hansen’s accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total, enabling robust comparisons across zones and scenarios but also by keeping values proportional to $V_i^0$ so interpretation is similar.
+The total constrained accessibility measure resolves the interpretability issue of Hansen's accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total, enabling robust comparisons across zones and scenarios but also by keeping values proportional to $V_i^0$ so interpretation is similar.
 
@@
 
-Though not outlined in the "Unconstrained accessibility" section, the unconstrained form aligns with Reilly's "potential trade territories" [@reilly1929methods] and Harris’ and Vickerman’s formulations of regional market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
+Though not outlined in the "Unconstrained accessibility" section, the unconstrained form aligns with Reilly's "potential trade territories" [@reilly1929methods] and Harris' and Vickerman's formulations of regional market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
 In its unconstrained form, market potential has also been more recently to estimate potentially accessible populations following infrastructure investments [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
@@
 [@soukhovIntroducingSpatialAvailability2023], since it incorporates spatial competition by allocating opportunities based on demand (population), impedance, and the known opportunity totals $D_j$.
-The dimensionless factor $\kappa^S_{ij}$ ensures that each destination’s opportunities are distributed proportionally to origins.
+The dimensionless factor $\kappa^S_{ij}$ ensures that each destination's opportunities are distributed proportionally to origins.
 As in the total constrained case, $V_i^S$ is expressed in the units of accessible opportunities.
@@
 In the total constrained case, opportunities are distributed based on impedance alone, regardless of population at $i$.
-In contrast, the singly constrained case allocates each zone’s opportunities proportionally across the region based on the relative impedance-weighted demand from all origins.
+In contrast, the singly constrained case allocates each zone's opportunities proportionally across the region based on the relative impedance-weighted demand from all origins.
 
@@
 As travel decay decreases (e.g., $f_3(c_{ij})$), competition becomes more diffuse.
-Zone 2, with the largest population, initially dominates accessibility—but under low decay, Zones 1 and 3 also begin drawing more opportunities from Zone 2.
+Zone 2, with the largest population, initially dominates accessibility--but under low decay, Zones 1 and 3 also begin drawing more opportunities from Zone 2.
 For instance, Zone 1 gains  `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` more from Zone 2 in $f_3(c_{ij})$ than in $f_1(c_{ij})$.
-This shift reflects a drop in $\kappa^S_{2,2}$ by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`,  showing Zone 2’s decreasing hold on its own opportunities as other zones gain accessibility 'parity'.
+This shift reflects a drop in $\kappa^S_{2,2}$ by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`,  showing Zone 2's decreasing hold on its own opportunities as other zones gain accessibility 'parity'.
 
@@
 This accessibility case adopts the structure of the doubly constrained spatial interaction model, where $V_{ij}^D$ flows are constrained by both origin populations $O_i$ and destination opportunities $D_j$.
-That is, the resulting accessibility outflow from each origin must match the origin’s population demand, and the resulting accessibility inflow to each destination must match the number of opportunities supplied:
+That is, the resulting accessibility outflow from each origin must match the origin's population demand, and the resulting accessibility inflow to each destination must match the number of opportunities supplied:
 
@@
 These units also often misaligned--for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
-When they do align--such as people to physician capacity--we’re essentially modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
+When they do align--such as people to physician capacity--we're essentially modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
 In such cases, the traditional accessibility question is already answered by the known information (i.e., how many opportunities can be reached by a zone?
@@
 As accessibility research evolved largely influenced by Hansen [@hansen1959], researchers in the field neglected the proportionality constant that was originally present in gravity-based models, and is still present in spatial interaction modeling.
-
-Firstly, we have demonstrated that by reintroducing Wilson's system constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces units to the resulting values.
-These values become more intuitive for the purpose of analysis and comparison.
-Secondly, the constraining constants, depending on the case (i.e., total, singly- or doubly- constrained), restrict the degree of _potential_, linking accessibility (the potential for spatial interaction) with access (spatial interaction) on the same continuum based on the constraint used.
-Lastly, we discussed how popular measures such as the one used in Hansen [@hansen1959] and the 2SFCA link into the family of accessibility measures.
+This work has demonstrated theoretically, and through a simple numeric example, that by reintroducing Wilson's system constraints and defining associated balancing factors and proportional allocation factors, we can derive a unifying family of accessibility measures that reintroduces tangible units to the resulting values.
+These values may be more intuitive for the purpose of analysis and comparison.
 
@@
 
-We then introduce the singly-constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
+We then introduced the singly-constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
 It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.
@@
 
-Building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility..
-By reintroducing Wilson's proportionality constant, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.
+Building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility.
+By reintroducing Wilson's proportionality constants, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.
 This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.
+
+That said, there have been more recent developments in the accessibility literature which include novel approaches such as person-based approaches, including those that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete], that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure.
+Future work could further explore how these ideas of proportionality constants and balanced units could help inform new modeling approaches.

### 2025-08-28 paper_plos/paper_plos.tex
Ready for Chris to submit

@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
@@
 \makeatother
-
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage[numbers,square,comma]{natbib}
@@
 \linenumbers
-
 \section{Introduction}\label{introduction}
@@
 translate into indicators that benchmark movement (e.g., vehicle
-kilometres traveled, intersection through traffic, etc.) which are not
+kilometres traveled, intersection through traffic) which are not
 necessarily linked to improved accessibility
@@
 healthcare
-\citep{luo2003, paez_healthcare_2010, wan2012three, delamater2013spatial, boisjoly2017informality, pereira_2021_geographic},
+\citep{luo2003, paez_healthcare_2010, wan2012three, delamater2013spatial, boisjoly2017informality, pereira_2021_geographic, yang2024evaluating},
 green space
@@
 
-More specifically, the wide range of accessibility definitions--with
+More specifically, the wide range of accessibility definitions, with
 novel methods being more sophisticated but less intuitive
-\citep{kapatsila_resolving_2023}--can further hinder practical uptake
+\citep{kapatsila_resolving_2023}, can further hinder practical uptake
 \citep{vanweeAccessible2016}. Geurs and van Wee
@@
 value that ensures the model reflects observed forces. Newton's initial
-estimate of \(G\) was based on a speculation that the mean density of
-earth was between five or six times that of water, an assumption that
-received support after Hutton's and Cavendish's experiments in the late
-1700s \citep{hutton_xxxiii_1778, cavendish_xxi_1798} to estimate of the
-proportionality constant to within 1\% accuracy. In sum: it took over a
-century from the publication of \emph{Principia} to refine the estimate
-of the proportionality constant.
-
-The Newtonian gravitational relationship laid the conceptual groundwork
-for later empirical studies of human spatial interaction, such as those
-by Ravenstein, Stewart, and others, who began translating these
-principles into observed patterns of human movement. However, the
-majority of these attempts described a proportional relationship, or one
-arbitrarily set to equality.
+estimate of \(G\) was based on a speculation but received empirical
+support after Hutton's and Cavendish' experiments in the late 1700s
+\citep{hutton_xxxiii_1778, cavendish_xxi_1798} which estimated it to
+within 1\% accuracy. That is, it took over a century from the
+publication of Newton's \emph{Principia Mathematica} to refine the
+estimate of the proportionality constant.
+
+While the Newtonian gravitational relationship laid the conceptual
+groundwork for later empirical studies of human spatial interaction, the
+majority of these early attempts described a proportional relationship
+or one arbitrarily set to equality. They did not establish an empirical
+\(G\) as in the Newtownian tradition, as will be discussed in the next
+subsection.
 
@@
 
-Following Carey's work \citep{careyPrinciplesSocialScience1858},
-research into human spatial interaction continued in different contexts.
-Namely, since Carey to the 1940s, a number of researchers theoretically
-and empirically attempted to characterise human spatial interaction as a
-force \(F\) directly proportional to the ``masses'' \(M_i\) and \(M_j\)
-of two locations, and inversely proportional to their separation
-distance--conceptually parallel to Newtonian gravity, but often omitting
-a proportionality constant.
+Henry C. Carey's \emph{Principles of Social Science}
+\citep{careyPrinciplesSocialScience1858} inspired empirical spatial
+intereaction research in different contexts. Namely, a number of
+researchers theoretically and empirically attempted to characterise
+human spatial interaction as a force \(F\) directly proportional to the
+``masses'' \(M_i\) and \(M_j\) of two locations, and inversely
+proportional to their separation distance--conceptually parallel to
+Newtonian gravity, but often omitting a proportionality constant.
 
@@
 calibration necessary to convert these proportional relationships into
-measurable and comparable quantities. It is only in Stewart
+measurable and comparable quantities. It is only in Stewart (1948)
 \citep{stewartDemographicGravitationEvidence1948} that we find the most
@@
 
-As will be discussed, when Hansen \citep{hansen1959} later adopted
-Stewart's formula (Eq~\ref{eq-stewart-population-potential-sum}) for
-accessibility, he omitted any mention of \(G\), effectively setting it
-to 1 arbitrarily as Stewart had done. This omission has persisted in
+As will be discussed, when Hansen (1959) \citep{hansen1959} later
+adopted Stewart's formula (Eq~\ref{eq-stewart-population-potential-sum})
+for accessibility, he omitted any mention of \(G\), effectively setting
+it to 1 arbitrarily as Stewart had done. This omission has persisted in
 accessibility research, leaving a conceptual gap in how such measures
@@
 \(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
-rewriting to accounts for a variety of impedance functions beyond the
-inverse power \(d^{-\beta}\) as follows:
+rewriting of \(d^{-\beta}\) to \(f(d_{ij})\) accounts for a variety of
+impedance functions beyond the inverse power as follows:
 
@@
 }\end{equation}
-
-\textcolor{red}{Rafa: there is no $\beta$ in this equation. Check text in paragraph below.}
 
@@
 \citep{wilson1971}, advanced a general framework for spatial interaction
-modeling focused on flows of interaction between places. This work was
-not focused on the `potential' concept as accessibility has been. Wilson
+modeling focused on flows of interaction between places and derived from
+a non-Newtonian analogy. Additionally, this work was not focused on the
+`potential' concept, as accessibility has been. Wilson
 \citep{wilson1971} formalized the general spatial interaction model
@@
 shifting the equation from a proportional to an equal relationship by
-incorporating known system totals. In this sense, we infer that \(k\)
-serves a role similar to the gravitational constant in Newton's law--it
-calibrates the model so that outputs match real-world quantities.
-However, these real-world quantities was not a set empirical constant
-(like Newton's \(G\)) but instead, sensitive to the system and known
-information about the system.
+incorporating known system totals. In some sense, \(k\) serves a role
+similar to the gravitational constant in Newton's law--it calibrates the
+model so that outputs match real-world quantities. However, these
+real-world quantities are not a set empirical constant (like Newton's
+\(G\)) but instead, sensitive to the system and known information about
+the system.
 
@@
 gravity (i.e., interaction between masses over distance), Wilson's
-innovation was to ground the model in \emph{entropy maximisation}. By
-maximizing the number of ways individual trip probabilities could be
-arranged under known constraints, Wilson derived models that estimate
-\emph{statistical averages} of flows between zones
+innovation was to ground and generalise the model using \emph{entropy
+maximisation}. By maximizing the number of ways individual trip
+probabilities could be arranged under known constraints, Wilson derived
+models that estimate \emph{statistical averages} of flows between zones
 \citep{wilson1971, seniorGravityModellingEntropy1979}.
@@
 
-Wilson's work is notable for many reasons, one of which is his departure
-from the metaphor of Newtonian gravitational physics. Instead, he
-grounded his models in statistical mechanics, particularly the principle
-of entropy maximisation. For accessibility research, Wilson's approach
-is especially significant because it can address a problem that has
-quietly afflicted this field (as this paper has argued): the transition
-from proportionality to equality.
-
-Rather than relying on a universal constant like \(G\), Wilson's models
-calibrate interaction flows through known empirical constraints (e.g.,
-total system flows, or origin-side and destination-side flow totals).
-This results in interpretable, balanced (given the constraints), and
-unit-consistent \(ij\) flows. In short, when aspects of the systems are
-inputs into the balancing factors, \(ij\) flows can be robustly
-predicted in tangible units of `flow', units that align with the system
-being modelled. In this way, spatial interaction modeling tradition can
-be seen to have succeeded where accessibility modeling stalled. While
+Wilson's work is notable for many reasons. Rather than relying on a
+universal constant like \(G\) or a scaling factor to balance units,
+Wilson's models calibrate interaction flows through known empirical
+constraints using principles of entropy maximisation. This results in
+interpretable, balanced (given the constraints), and unit-consistent
+\(ij\) flows. In this way, spatial interaction modeling tradition can be
+seen to have succeeded where accessibility modeling stalled. While
 Wilson's model produces results that are in units of flow tethered to
-the system of analysis, Hansen-type measures (still widely used today)
-yield outputs that reflect proportional---not equal---relationships and
-typically lack interpretable units.
+the system of analysis, Hansen-type measures (still widely used today in
+accessibility work) yield outputs that reflect proportional--not
+equal--relationships and typically lack interpretable units.
 
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 for strategic land use and transportation planning''. And these needs
-were far from trivial: in the U.S., for instance, the Federal-Aid
-Highway Act of 1956 set in motion the construction of the Interstate
-Highway System with an eventual budget exceeding \$600 billion in
-today's dollars
+were far from trivial: for instance in the U.S., the Federal-Aid Highway
+Act of 1956 set in motion the construction of the Interstate Highway
+System with an eventual budget exceeding \$600 billion in today's
+dollars
 \citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{../figures/access-analytical-device.png}
+\includegraphics[width=0.7\textwidth,height=\textheight]{../figures/access-analytical-device.png}
 
@@
 more diffuse. Zone 2, with the largest population, initially dominates
-accessibility---but under low decay, Zones 1 and 3 also begin drawing
+accessibility--but under low decay, Zones 1 and 3 also begin drawing
 more opportunities from Zone 2. For instance, Zone 1 gains 6\% more from
@@
 proportionality constant that was originally present in gravity-based
-models, and is still present in spatial interaction modeling.
-
-Firstly, we have demonstrated that by reintroducing Wilson's system
-constraints and defining associated balancing factors and proportional
-allocation factors, we can derive a unifying family of accessibility
-measures that reintroduces units to the resulting values. These values
-become more intuitive for the purpose of analysis and comparison.
-Secondly, the constraining constants, depending on the case (i.e.,
-total, singly- or doubly- constrained), restrict the degree of
-\emph{potential}, linking accessibility (the potential for spatial
-interaction) with access (spatial interaction) on the same continuum
-based on the constraint used. Lastly, we discussed how popular measures
-such as the one used in Hansen \citep{hansen1959} and the 2SFCA link
-into the family of accessibility measures.
+models, and is still present in spatial interaction modeling. This work
+has demonstrated theoretically, and through a simple numeric example,
+that by reintroducing Wilson's system constraints and defining
+associated balancing factors and proportional allocation factors, we can
+derive a unifying family of accessibility measures that reintroduces
+tangible units to the resulting values. These values may be more
+intuitive for the purpose of analysis and comparison.
 
@@
 
-We then introduce the singly-constrained case, which \emph{does} take
+We then introduced the singly-constrained case, which \emph{does} take
 into account the population size at the origin in the allocation of
@@
 Building on Wilson's \citep{wilson1971} foundational work, this paper
-proposed a unified framework for accessibility.. By reintroducing
-Wilson's proportionality constant, the proposed family of constrained
+proposed a unified framework for accessibility. By reintroducing
+Wilson's proportionality constants, the proposed family of constrained
 accessibility measures restores measurement units to accessibility
@@
 
+That said, there have been more recent developments in the accessibility
+literature which include novel approaches such as person-based
+approaches, including those that are time-sensitive
+\citep{yang2024evaluating, braga2023evaluating}, behavioural
+\citep{kar2024inclusive, lu2014effects} or utility-based
+\citep{guzman2023much, ben1985discrete}, that are not based on top-down
+approaches such as Wilson's model formalized through the
+entropy-maximisation procedure. Future work could further explore how
+these ideas of proportionality constants and balanced units could help
+inform new modeling approaches.
+
 \nolinenumbers

### 2025-08-28 paper_plos/reviewer_respose_letter_V1.qmd
Ready for Chris to submit

@@
 We have done our best to address all comments and suggestions as much as possible, which we believe has significantly improved the paper.
-We address the individual comments of the reviewers one by one (list below), but we would also like to note three major points in response to the reviewers’ recommendations.
+We address each comment of the reviewers point by point (as follows), but we would also like to note three significant points in response to the reviewers' recommendations.
 
@@
 This involved streamlining the manuscript's text, removing equations and text redundancies where possible, especially in the "Family of Accessibility" section.
-- We have also edited the text to more clearly articulate the paper’s main contribution and to make each section’s purpose clearer.
-- While we agree with reviewer’s #2  that including an empirical example would demonstrate the real-world applicability of our proposed measures, we were not able to include an empirical case study in the paper due to manuscript length limits.
+- We have also edited the text to more clearly articulate the paper's main contribution and to make each section's purpose clearer.
+- While we agree with Reviewer #2's comment that suggested including an empirical example to demonstrate the real-world applicability of our proposed measures, we were not able to include such a case study in the paper due to manuscript length limits.
+As such, we've toned down language concerning real-world applicability and instead point to how the theoretical clarity of the resulting values can be helpful practically.
+An empirical case study will be left for future work.
 
-As a whole, we believe the paper is more consistent after heeding the reviewer’s comments.
+As a whole, we believe the paper is more consistent after heeding the reviewer's comments.
 We hope you and the reviewers will agree.
@@
 
-# Review 1
+# Response to Reviewer #1
 
@@
 - We've substantially reduced the length of the manuscript from 34 to 28 pages (without references).
-- We've removed tangential details and minimized the equations where possible, particularly in the “Early Research” section.
+- We've removed tangential details and minimized the equations where possible, particularly in the "Early Research" section.
 - We've streamlined the manuscript's text--to make each sections purpose more clear. The introductions and conclusions of sections now have a better flow.
@@
 - Second, it introduces 'constraints' to accessibility measures grounded in spatial interaction principles.
-<!-- First: the total constrained accessibility measure as an alternative to the Hansen-type formulation, producing zonal values that represent proportions of the known number of destination masses (i.e., opportunities) or origin masses (i.e., population) in the region.
-Second: the singly constrained accessibility measure as a competitive method that proportionally allocates opportunities from the destination based on population demand (from the origin), or in the case of market potential, proportional allocation of population from the origin based on the supply of opportunities at the destination.
-Its per-capita zonal rate is shown to be mathematically equivalent to the widely used two-step floating catchment area (2SFCA) method [@luo2003; @shen1998].-->
 
-\textcolor{blue}{Related to the first issue, I am not sure about the point that the authors explain the details of Newtonian’s roots and early research on human spatial interactions from Ravenstein to Stewart.
+\textcolor{blue}{Related to the first issue, I am not sure about the point that the authors explain the details of Newtonian's roots and early research on human spatial interactions from Ravenstein to Stewart.
 Of course, I think mentioning those contents is beneficial, but it is hard to get the key point given that it is verbose.}
@@
 The main aim of these two sections is to demonstrate how Newton's gravitational equation (with its proportionality factor $G$) inspired early research on human spatial interactions.
-The majority of early researchers relied on the underlying proportional mass/distance relationship to model human spatial interaction, but did not use $G$--except for Stewart.
+The majority of early researchers relied on the underlying proportional mass/distance relationship to model human spatial interaction, but did not use $G$--except for Stewart (1947).
 However, Stewart didn't have an answer to what $G$ should be, just that it is needed to move from a relationship of proportionality to one of equality.
 
-Stewart is especially important because Hansen later adopted it into his equation but ommitted both the $G$ (i.e., kept it set to 1) and was quiet on the broader issue of moving from proportionality to equality.
+Stewart is especially important because Hansen (1959) later adopted it into his equation but omitted both the $G$ (i.e., kept it set to 1) and was quiet on the broader issue of moving from proportionality to equality.
 
-We've revised these sections, hopefully we've clarifed these points and the purpose of these sections in the process!
+We've revised these sections. We hope we've sufficiently clarified these points and the purpose of these sections in the process.
 
@@
 
-Good point.
-We now clearly summarize all four groups of constrained measures at the start of the “A Family of Accessibility Measures: From Proportionality to Equality” section.
+Fair points, thank you for raising them.
+To assist with comprehension, we now clearly summarize all four groups of constrained measures at the start of the "A Family of Accessibility Measures: From Proportionality to Equality" section.
 These definitions build on the preceding sections, which trace how spatial interaction modeling followed the Newtonian analogy while neglecting $G$, how Wilson shifted to an entropy-maximisation formulation without requiring an empirical $G$ (but instead, a constraint that functions to tether the results to system knowns), and how the place-based accessibility literature has largely overlooked this solution.
@@
 
-# Reviewer #2
+# Response to Reviewer #2
 
 \textcolor{blue}{This manuscript presents a theoretically motivated and conceptually rich framework that bridges spatial interaction modeling and accessibility measures.
-The authors introduce a family of accessibility metrics grounded in Wilson’s spatial interaction models and argue convincingly for the reintroduction of interpretability and measurement units in accessibility analysis.
+The authors introduce a family of accessibility metrics grounded in Wilson's spatial interaction models and argue convincingly for the reintroduction of interpretability and measurement units in accessibility analysis.
 The historical tracing of accessibility theory development and the synthesis of divergent literature streams are timely and valuable contributions to the field.}
@@
 While the authors provide a detailed numerical example, the manuscript lacks any empirical case study or application to real-world accessibility data.
-This limits the reader’s ability to evaluate the practical utility and robustness of the proposed framework.
+This limits the reader's ability to evaluate the practical utility and robustness of the proposed framework.
 The authors can draw inspiration from several studies such as "Factoring in temporal variations of public transit-based healthcare accessibility and equity" to better understand this concept.}
@@
 We agree, the inclusion of an empirical example would demonstrate the real-world applicability of the measures.
-However, the length of the current document is an issue, so the addition of additional pages isn't feasible--especially given the concern over paper length that was raised by Reviewer 1 and how we’ve already reduced it by six pages in response to these concerns.
+However, the length of the current document is an issue, so the inclusion of additional pages isn't feasible--especially given the concern over paper length that was raised by Reviewer #1 and how we've reduced the manuscript by six pages in response to these concerns.
 
-But, we've put great thought into how the manuscript could be restructured to included an additional empirical example.
+As well, we put great thought into how the manuscript could be restructured to included an additional empirical example.
 However, many of these ideas are not possible given the manuscript's current format.
-For instance, even if we replaced the toy example with an empirical example, it would still add quite significant additional length (i.e., explanation of data, room for tests).
+For instance, even if we replaced the toy example with an empirical example, it would still add quite significant additional length (i.e., explanation of data, additional space needed for tests).
 Further, a single empirical example is not sufficient to cover all cases and variants of the family.
 
-As such, we’ve worked to clarify that the contribution of this work which is primarily theoretical and methodological.
-We’ve also tempered earlier claims about the measure’s real-world applicability, instead allowing readers to consider potential uses and the potential value of reuniting accessibility measures with interpretable units.
-While we believe these measures utility would be highlighted through real-world examples--we plan to explore these in future publications which make this contribution more practically clear (i.e., where we may have enough space to develop and explore an empirical example without shortchanging the historical context).
-In fact, we are currently in conversation with the editor about the possibility to submit a separate paper demonstrating the proposed measures with an empirical case study.
+As such, we've worked to clarify that the contribution of this work which is primarily theoretical and methodological.
+We've also tempered earlier claims about the measure's real-world applicability, instead allowing readers to consider potential uses and the potential value of reuniting accessibility measures with interpretable units.
+While we believe these measures usefulness would be highlighted through real-world examples--we plan to explore these in future publications which make this contribution more practically clear (i.e., where we may have enough space to develop and explore an empirical example without shortchanging the historical context).
+
+In fact, this comment inspired action.
+We are currently in conversation with the editor about the possibility to submit a separate paper demonstrating the proposed measures with an empirical case study.
+We really appreciate the perspective this comment added.
 
@@
 
-Agreed, the use of these words--especially in the doubly constrained model subsection should be tuned up.
-We've substantially revised this subsection.
+Completely agree, the use of these words--especially in the doubly constrained model subsection should be tuned up.
+In response, we've substantially revised this subsection.
 Namely, we've removed the discussion about the meaning of "potential", and instead focus on the embedded assumption of the doubly constrained measure i.e., how it simultaneously considers both origin-side and destination-side constraints--or that all people are matched with opportunities, and vice versa.
@@
 
-Thank you for this comment.
+Thank you for these valuable points.
+In response, we've made significant efforts to streamline the introductory and review sections (i.e., before _"A family of accessibility measures: from proportionality to equality"_ section), to ensure a better balance with the paper's original contributions.
+These sections are now 12 pages instead of the 15 pages in the previous version, and now include additional text at the end and start of each section to better summarise and bridge the ideas.
+We also removed 6 equations that we viewed as non-essential in these earlier sections along with removing redundant/tangential text to reduce the overall length while still covering all necessary literature and the over 70 papers that cite both Wilson (1971) and Hansen (1959).
 
-We’ve removed tangential details and minimized the equations where possible, particularly in the “Early Research” section.
-The historical literature review section (Accessibility and spatial interaction modelling: two divergent research streams) now covers upward of 76 papers citing Wilson and Hansen, condensed into about 3 pages.
-This section sets the stage by outlining Wilson’s balancing factors and highlighting how place-based accessibility literature (largely based on Hansen-type measures) has not adopted these balancing factors.
-This context is needed for the subsequent, more detailed 16-page section, which introduces the newly proposed family of accessibility measures which incorporate Wilson's balancing factors.
-We believe this structure balances breadth and depth appropriately; we've tried to reduce redundancies and improve when possible.
-We hope you find the revised manuscript easier to follow.
+We agree that these early sections must not overshadow the other core contributions.
+Accordingly, we also improved the clarity of the _"A family of accessibility measures: from proportionality to equality"_ section by streamling the _"Toy example setup"_ subsection and significantly improving the flow of the discussion of toy example results for each case.
+We reduced redundant text and removed an additional 4 equations.
+We believe the paper now strikes a more appropriate balance between historical context and novel insights.
+
+Regarding the suggested inclusion of more recent work, we have now added "Evaluating temporal variations in access to multi-tier hospitals..." to the Introduction as an example of recent work.
+Furthermore, while it offers important insights on temporal accessibility and equity, the review section of our work is scoped specifically to place-based accessibility measures that engage with spatial interaction models.
+As such, we found this work outside the direct narrative arch of the work's earlier section.
+Nonetheless, we've also included a brief acknowledgement of the broader literature on temporal considerations in accessibility approarches in the concluding paragraph of the Conclusion section (see response to the following question).
+
+We hope these revisions address your concerns and improve the clarity and balance of the manuscript. Thank you again.
 
@@
 Thank you for this fair point.
-In the revised manuscript, we state explicitly that our focus is on place-based accessibility, which remains widely used in practice and prominent in the literature, as well as Wilson's spatial interaction model which is also popular in spatial interaction modeling still to this day.
+In the revised manuscript, we state explicitly that our focus is on place-based accessibility, which remains widely used in practice and prominent in the literature, as well as Wilson's spatial interaction model which is also popular in spatial interaction modeling to this day.
 Given the broad scope of the paper and space constraints, incorporating additional detailed discussions of novel approaches is challenging, and somewhat tangential.
@@
 However, we do acknowledge some utility-based approaches within the spatial interaction literature.
-For example, on page 11 "The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984]."
+For example, on page 11 "The third subset of the spatial-interaction focused literature, depart from Hansen's definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., Morris et al.
+(1979) and Leonardi et al.
+(1984)".
 
 Our point with the historical review section is to demonstrate how even the papers that _do_ cite both Wilson and Hansen do not operationalise Wilson's balancing factors.
-Utility-based approaches take a different direction--not based on Newtonian gravitation--but also not based on Wilson's balancing factors.
-Indeed recent developments such as time-sensitive, network-based, or multi-modal accessibility models, as well as behavioral or utility-based approaches take on new approaches, but we believe these are tangential to our scope.
+Utility-based approaches take a different direction--not based explicitly on Newtonian gravitation--but also not based on Wilson's balancing factors.
+Indeed recent developments such as time-sensitive, network-based, or multi-modal accessibility models, as well as behavioral or utility-based approaches take on new approaches, but we believe these are tangential to our scope but are valuable to mention.
 
-Thoughts???
-
-\textcolor{red}{Rafa: perhaps we can add a sentence in the conclusion saying somehting like: "Recent developments in the literature include novel approaches to utility and person-based accessibility, including time-sensitive behavioral or utility-based approaches, that are not based on Newtonian gravitation nor on Wilson's balancing factors.
-Future studies could further explore how these ideas of balacing factors/spatial constraints could helo inform the new models."}
+In response, we've added a paragraph at the end of the conclusion that mentions this explicitly: "That said, there have been more recent developments in the accessibility literature which include novel approaches such as person-based approaches, including those that are time-sensitive (Yang et al.
+2024; Braga 2024 et al.
+2023), behavioural (Kar et al.
+2024; Lu et al.
+2014), or utility-based (Guzman et al.
+2023; Ben-Akiva and Lerman 1985), that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure.
+Future work could further explore how these ideas of proportionality constants and balanced units could help inform new modeling approaches."
 
@@
 And to solidify this point, again more theoretically but still practically, we develop and solve the toy example.
+We discuss the advantage of interpreting the units when constraints are used, and the challenge of doing so across toy example scenarios in the case of the unconstrained example.
 
-Regarding the unconstrained example:
-pg.
-16: "For example, Table @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function.
+Namely, concerning the unconstrained example on pg.
+16 we write: "For example, Table 3 shows $V^0_{i}$ under each decay function.
 Comparing across decay types is meaningless in absolute terms.
-For instance, the difference in zone 1 (edge of urban core)’s accessibility under $f_3$ vs $f_1$ is 370.92, but in what units?
+For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is 370.92, but in what units?
 These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
@@
 
-Regarding the total constrained example:
-pg.
+Building on this idea, we discuss the interpretive clarity in the total constrained numeric example on pg.
 18: "Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
@@
 
-pg.
-18:"One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
+We also go on to write on pg.
+18: "One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
 As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
@@
 
+Through the discussion of the toy example, we hope we make clear the interpretative clarity that using the constrained framework offers.
+
 \textcolor{blue}{6.
@@
 \textcolor{red}{Rafa: I actually liked table 1 and thought perhaps we should keep the table and remove the redundancy of the text around it. But I'm Ok with this solution below."}
+\textcolor{red}{Ana: What do you think Chris?
+I prefer my approach as opposed to the table, but I could easily add it back in.
+The Table looks a bit confusing in my eyes, I personally don't like reading text in tables, I like reading text in paragraphs.}
 
@@
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
-Represents ‘interaction capacity’ or ‘realized access’ serving as predictions of real interaction flows."
+Represents ‘interaction capacity' or ‘realized access' serving as predictions of real interaction flows."
 

### 2025-09-03 paper_plos/paper_plos_V0.qmd
final changes to R1

--- 2025-05-26 paper/manuscript.qmd
@@
 title: "A family of accessibility measures derived from spatial interaction principles"
-#subtitle: "Submitted preprint for peer-review (May 2025)"
+format:
+  docx:
+    toc: false
+  plos-pdf:
+    number-sections: false
+    keep-tex: true
+    journal:
+      id: plosone
+    header-includes:
+      - \usepackage{float}
+      - \floatplacement{table}{H}
+      - \floatplacement{figure}{H}
+      - \usepackage{pdflscape}
+      - \newcommand{\blandscape}{\begin{landscape}}
+      - \newcommand{\elandscape}{\end{landscape}}
+    #include-in-header:
+    #  text: |
+    #    % Remove comment for double spacing
+    #    % \usepackage{setspace} 
+    ##    % \doublespacing
+    #    % \usepackage{pdflscape}
+    #    % \newcommand{\blandscape}{\begin{landscape}}
+    #    % \newcommand{\elandscape}{\end{landscape}}
 author:

### 2025-09-03 paper_plos/paper_plos_V1.qmd
final changes to R1

--- 2025-08-28 paper_plos/paper_plos.qmd
@@
 format:
+  #docx:
+  #  toc: false
   plos-pdf:
@@
 
-\textcolor{red}{Rafa: Do we have to shorthen the abstract?
-I actually preferred the abstract we had in the preprint.
-I like that the abstract mentions the toy example though, which was not done in the 1st version.}
-
 abstract: |
@@
 To this end, this paper offers a methodological contribution by specifying a family of accessibility measures grounded in the shared theoretical roots of gravity-based accessibility and spatial interaction models, particularly their balancing factors.
-From this foundation, we outlined four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
+From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
 These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential).
@@
 Among place-based measures (this work's focus), variants include gravity-based [@hansen1959], cumulative opportunity [@pirie_measuring_1979], the 2 Step Floating Catchment Area (FCA) approach [@luo2003], and a variety of modifications to these approaches e.g., Enhanced 2-Step FCA [@luoEnhanced2009], 3-Stage FCA [@wan2012three], Modified 2-Step FCA [@delamater2013spatial], inverse 2-Step FCA [@wang_2sfca_2021], and n-steps FCA [@liang_novel_2024].
-While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised in [@vanweeAccessible2016]: How should practitioners interpret differences in accessibility scores between modes, and how should results be communicated?
+While these methods are tailored to address specific research contexts, overall this diversity does not demystify existing questions like those raised by van Wee [@vanweeAccessible2016]: How should practitioners interpret differences in accessibility scores between modes, and how should results be communicated?
 
@@
 
-In summary, this paper makes two contributions: (1) Reviews how spatial interaction modeling and accessibility share similar origins, but diverged in focus and interpretation.
-(2) Introduces a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
+In summary, this paper makes two contributions: (1) we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation.
+(2) we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
 These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
 
-We contend that accessibility research should re-engage with spatial interaction modeling, particularly the use of Wilson's system constraints.
+We contend that accessibility research should re-engage with spatial interaction modeling, particularly the use of Wilson's [@wilson1971] system constraints.
 While spatial interaction models embraced such constraints to improve interpretability, most accessibility models have not.
@@
 
-This famous equation expresses that the attractive force $F$ between two bodies is directly proportional to the product of their masses and inversely proportional to the square of the distance between them.
+This famous equation expresses that the attractive force $F$ between two bodies $i$ and $j$ is directly proportional to the product of their masses and inversely proportional to the square of the distance between them.
 As mass increases, so does force; as distance increases, force decreases.
@@
 Hansen recast Stewart's population potential to reflect accessibility, a model of human behaviour useful to capture regularities in mobility patterns.
-Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more specifically, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$ with a contemporary rewriting of $d^{-\beta}$ to $f(d_{ij})$ accounts for a variety of impedance functions beyond the inverse power as follows:
+Hansen replaced $M_j$ in @eq-stewart-population-potential-sum with *opportunities* to derive an *opportunity potential*, or more specifically, a *potential of opportunities for interaction* as $S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}$.
+A contemporary rewriting of $d^{-\beta}$ to $f(d_{ij})$ accounts for the variety of impedance functions beyond the inverse power used in the applied literature:
 
@@
 This omission has significant implications.
-Without a proportionality constant, the accessibility formula expresses only a proportional relationship: $S_i \propto \sum_j g(O_j)f(d_{ij})$, not of calibrated equality.
+Without a proportionality constant, the accessibility formula expresses only a proportional relationship: $S_i \propto \sum_j g(O_j)f(d_{ij})$, not one of calibrated equality.
 Recognition of the nature of this relationship is not common in the literature, but is known, i.e., this proportional equation is shown in Figure 1 in Wu and Levinson [@wuUnifyingAccess2020].
@@
 Wilson.
-Wilson's ground breaking paper [@wilson1971], advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
+Wilson's ground breaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
 Additionally, this work was not focused on the 'potential' concept, as accessibility has been.
@@
 Additional information can be introduced.
-For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ in @eq-phys-gravity-model is represented as $O_i$ and the following constraint can be used:
+For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ in @eq-phys-gravity-model, represented as $O_i$, then following constraint can be used:
 
@@
 
-Depending on which of the three system constraints are applied a family of spatial interaction models can be derived from @eq-phys-gravity-model.
+Depending on which of the three system constraints are applied, a family of spatial interaction models can be derived from @eq-phys-gravity-model.
 The proportionality constant $k$ is replaced with different _balancing factors_.
@@
 
-In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$), and (3) a doubly constrained model that satisfies both.
+In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$ - see below), and (3) a doubly constrained model that satisfies both.
 
@@
 In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
-While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional--not equal--relationships and typically lack interpretable units.
+While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.
 
@@
 
-\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ is the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case (hence represented by a $X$).
+\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ reflect the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case (hence represented by a $X$).
 
@@
 - No constraints are applied, so values reflect proportionality only and are not calibrated to known system totals.
+
 2. **Total Constrained Case ($V_i^T$, $M_j^T$)**
@@
 - Units of $M_j^T$: accessible population from $j$, a value that is total constrained and linearly proportion to $M_j^0$.
+
 3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
@@
 - Units of  $M_j^S$: accessible population from $j$, a value that is the sum of population demand flows allocated proportionally based on supply at $j$.
+
 4. **Doubly Constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
@@
 
+As a summary, each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
+
+`\blandscape`{=latex}
+
+Table 1: Summary of constrained accessibility measure types and interpretations
+
+::: {=latex}
+\input{summary-family-access-measures-table.tex}
+
+:::
+
+`\elandscape`{=latex}
+
 ## Toy example setup
@@
 
-Consider the simple 3-zone region in Figure @fig-analytical-device-conc-accessibility, where each zone serves as both origin ($i$) and destination ($j$).
+Consider the simple 3-zone region in @fig-analytical-device-conc-accessibility, where each zone serves as both origin ($i$) and destination ($j$).
 The system includes three inputs: zonal population and opportunities, a zonal cost (travel time) matrix, and three travel impedance functions representing different travel behaviours.
 
-First, Table @tbl-small-system-land-use summarizes population (in 10,000s) and physicians per zone.
+First, @tbl-small-system-land-use summarizes population (in 10,000s) and physicians per zone.
 For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)` comparable to Canada's 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
-Second, Table @tbl-small-system-cost shows travel times (minutes); it can be discerned that Zones 1 and 3 are closer to each other than to Zone 2.
+Second, @tbl-small-system-cost shows travel times (minutes); it can be discerned that Zones 1 and 3 are closer to each other than to Zone 2.
 Zones 1 and 3 together have a population roughly equal to Zone 2 but offer more than twice the physician availability.
-We interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core, and Zone 2 as the Suburban.
+We interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core, and Zone 2 as Suburban.
 
@@
 
-For example, Table @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function.
+For example, @tbl-simple-example-unconstrained-accessibility shows $V^0_{i}$ under each decay function.
 Comparing across decay types is meaningless in absolute terms.
@@
 These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
-The fundamental uninterpretability of what is a _opportunity-weighted-travel-impedance_ unit remains.
+The fundamental uninterpretability of what is an _opportunity-weighted-travel-impedance_ unit remains.
 
@@
 - $V_i^T$: accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and
-- $M_j^T$: where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total number of population and to be interpreted as constrained 'market potential'.
+- $M_j^T$: where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total population and to be interpreted as constrained 'market potential'.
 
@@
 In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance.
-The total constrained accessibility from $i$ to $j$:
+The total constrained accessibility from $i$ to $j$ takes the form:
 $$
@@
 
-Referring back to our simple numeric example, $K^T$ for the highest decay t scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
+Referring back to our simple numeric example, $K^T$ for the highest decay travel scenario $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
 
@@
 $K^T$ for other decay scenarios are calculated similarly in code.
-Applying each $K^T$ to the unconstrained values $V_i^0$ yields total constrained accessibility values (Table @tbl-simple-example-total-opportunity-accessibility), all in units of physicians.
+Applying each $K^T$ to the unconstrained values $V_i^0$ yields total constrained accessibility values (@tbl-simple-example-total-opportunity-accessibility), all in units of physicians.
 
@@
 Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
-For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (`r tot_acc$V_tot_i_1[1]`), like in the unconstrained accessibility case.
-However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum()` physicians in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3.
+For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (`r tot_acc$V_tot_i_1[1] |> round(digits = 3)`), like in the unconstrained accessibility case.
+However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum() |> round(digits = 3)` physicians in the region, which allows us also to deduce that zone 1 captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]) |> round(digits = 3)` and `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]) |> round(digits = 3)` times more than zone 2 and 3.
 Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
@@
 In terms of proportional magnitude, this can also be observed in the unconstrained measure for this scenario.
-However, for the total constrained measure, this plateauing of results have meaning.
+However, for the total constrained measure, this plateauing of results has meaning.
 In fact, each zone is allocated an average of the total amount in the region, as a result of the total constrained proportional allocation factor.
 
-However, what's notable is how Zones change between scenarios.
+However, what's notable is how zones change between scenarios.
 For instance, Zone 1's share only declines slightly--these declines are outpaced by Zone 2's relative gains.
@@
 
-This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield s measure of market potential--the number of people who can spatially interact with a destination.
+This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield a measure of market potential--the number of people who can spatially interact with a destination.
 
@@
 As in the total constrained case, $V_i^S$ is expressed in the units of accessible opportunities.
-Finally, the **per capita** version of this measure:
-
-Soukhov et al., 2023 [@soukhovIntroducingSpatialAvailability2023] also showed that the following expression (accessibility per capita) is a constrained version of the popular 2SFCA approach of Shen 1998 [@shen1998] and Luo and Wang 2003 [@luo2003]:
+
+Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023] also showed that the following expression (accessibility per capita) is a constrained version of the popular 2SFCA approach of Shen [@shen1998] and Luo and Wang [@luo2003]:
 
@@
 
-The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 area `r B_j$B_j_2[1]`, `r B_j$B_j_2[2]` and `r B_j$B_j_2[3]`, respectively.
-For the $f_3(c_{ij})$ decay group, they are  `r B_j$B_j_3[1]`, `r B_j$B_j_3[2]` and `r B_j$B_j_3[3]`.
+The balancing factors $B_j$ for the $f_2(c_{ij})$ decay group for zones 1, 2 and 3 are `r B_j$B_j_2[1] |> round(digits = 3)`, `r B_j$B_j_2[2]  |> round(digits = 3)` and `r B_j$B_j_2[3]  |> round(digits = 3)`, respectively.
+For the $f_3(c_{ij})$ decay group, they are  `r B_j$B_j_3[1]  |> round(digits = 3)`, `r B_j$B_j_3[2]  |> round(digits = 3)` and `r B_j$B_j_3[3]  |> round(digits = 3)`.
 Using these these balancing constants, we can calculate the singly constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).
@@
 Consider the highest decay scenario $f_1(c_{ij})$.
-Under this scenario, Zone 1--despite hosting a medium amount of physicans--captures the fewest physicians (`r opp_acc$V_opp_i_1[1]`), compared to `r opp_acc$V_opp_i_1[2]` at Zone 2, and `r opp_acc$V_opp_i_1[3]` at Zone 3.
+Under this scenario, Zone 1--despite hosting a medium amount of physicians--captures the fewest physicians (`r opp_acc$V_opp_i_1[1]  |> round(digits = 3)`), compared to `r opp_acc$V_opp_i_1[2]  |> round(digits = 3)` at Zone 2, and `r opp_acc$V_opp_i_1[3]  |> round(digits = 3)` at Zone 3.
 Why?
@@
 In this way, the consideration of constrained accessibility *per capita* may be clarifying.
-Often, accessibility values are reported as raw scores without the consideration for population.
+Often, accessibility values are reported as raw scores without considering the population potentially accessing them.
 But, as we introduced constraints, these constrained accessibility values can be normalized using anything that is relevant to the zone.
-In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example, simply in units of number of physicians accessible per population at each zone.
+In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example simply in units of number of physicians accessible per population at each zone.
 Notably, these per capita rates are equivalent to the 2SFCA values.
@@
 Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
-In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
+In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential has been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
 
@@
 
-For the sake of brevity, we'll move forward onto the doubly constrained case.
+For the sake of brevity, we'll move forward on to the doubly constrained case.
 
@@
 
-As the doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model as follows:
+The doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model as follows:
 
@@
 
-Moving onto the toy example, to calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be reinterpreted.
+Moving on to the toy example, to calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be reinterpreted.
 Namely, a count of physician _capacity_ per destination $D_j$ is needed instead of simply the number of physicians.
@@
 
-This adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case: with the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,00s of physician-capacity) per zone.
+This adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case.
+With the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,000s of physician-capacity) per zone.
 For the population, we leave this unchanged numerically but we now must keep in mind that each person interacts with one physician capacity.
 The number of providers per destination is however revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).
-The system-wide PPR is now:`r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of`r sum(LU$D_j) / sum(LU$O_i)`.
+The system-wide PPR is now `r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of `r sum(LU$D_j) / sum(LU$O_i)`.
 
@@
 
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull()`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull()`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull()`.
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
 Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
@@
 To illustrate this, consider the results for Zone 2 (Suburban zone--a high-population, low-opportunity, relatively remote zone).
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal allocation (i.e., $V^D_{22}$) declines as travel impedance decay decreases--from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal allocation (i.e., $V^D_{22}$) declines as travel impedance decay decreases--from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
 
@@
 So following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
-Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $kappa_{ij}^D$ and $\hat kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
-
-Readers should recall, $V_{ij}^D$ are also mathematically equivalent to Wilson's spatial interaction flows.
+Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $\kappa_{ij}^D$ and $\hat \kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
+
+$V_{ij}^D$ are also mathematically equivalent to Wilson's spatial interaction flows.
 And as Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
-However, traditionally the focus of these models have typically on $ij$ flows and often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
+However, the focus of these models has typically been on $ij$ flows, often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
 
@@
 
-A summary of the family of accessibility measures, is detailed; First, we place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
+To summarize the contributions of this work, first, we place the popular Hansen-type accessibility measure [@hansen1959] within this family of measures as an "unconstrained" case, demonstrating that resulting values cannot be directly compared across different travel scenarios without ad-hoc adjustments.
 We then show how applying a total constraint balances the units and produces a statistically averaged solution that converges to the regional average for each zone as the decay effect decreases.
@@
 In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available.
-Specifically, accessibility values for each $i$-$j$ pair must be a proportion of he zonal opportunity and population values simultaneously.
-For example, the accessibility at zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at zone 1.
+Specifically, accessibility values for each $i$-$j$ pair must be a proportion of the zonal opportunity and population values simultaneously.
+For example, the accessibility at Zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at Zone 1.
 Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest.
-In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach (out of a regional total and/or zonal-allocation).
-
-Building on Wilson's [@wilson1971] foundational work, this paper proposed a unified framework for accessibility.
-By reintroducing Wilson's proportionality constants, the proposed family of constrained accessibility measures restores measurement units to accessibility estimates.
+In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach out of a regional total and/or zonal-allocation).
+
+Overall, and building on Wilson's [@wilson1971] foundational work, this paper proposes a unified framework for analyzing gravity-based accessibility.
+By reintroducing Wilson's proportionality constants, the family of constrained accessibility measures restores measurement units to accessibility estimates.
 This enhancement provides a more interpretable, consistent, and theoretically grounded basis for accessibility analysis, which could help advance the adoption of accessibility-oriented planning.
 
-That said, there have been more recent developments in the accessibility literature which include novel approaches such as person-based approaches, including those that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete], that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure.
-Future work could further explore how these ideas of proportionality constants and balanced units could help inform new modeling approaches.
+That said, there have been more recent developments in the accessibility literature, including novel person-based approaches that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete], that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure.
+While gravity-based measures of accessibility dominate the applied literature, future work could further explore how these ideas of proportionality constants and balanced units could also help inform these and other new modeling approaches.

### 2025-09-03 paper_plos/paper_plos_V1.tex
final changes to R1

--- 2025-08-28 paper_plos/paper_plos.tex
@@
 spatial interaction models, particularly their balancing factors. From
-this foundation, we outlined four memebers of the family: the
+this foundation, we outline four memebers of the family: the
 `unconstrained' measure (i.e., Hansen-type accessibility), the `total
@@
 \linenumbers
+
 \section{Introduction}\label{introduction}
@@
 overall this diversity does not demystify existing questions like those
-raised in \citep{vanweeAccessible2016}: How should practitioners
+raised by van Wee \citep{vanweeAccessible2016}: How should practitioners
 interpret differences in accessibility scores between modes, and how
@@
 
-In summary, this paper makes two contributions: (1) Reviews how spatial
-interaction modeling and accessibility share similar origins, but
-diverged in focus and interpretation. (2) Introduces a family of
+In summary, this paper makes two contributions: (1) we review how
+spatial interaction modeling and accessibility share similar origins but
+have diverged in focus and interpretation. (2) we introduce a family of
 accessibility measures grounded in spatial interaction principles,
@@
 We contend that accessibility research should re-engage with spatial
-interaction modeling, particularly the use of Wilson's system
-constraints. While spatial interaction models embraced such constraints
-to improve interpretability, most accessibility models have not. This
-lack of adoption contributes to fuzziness in current analyses, limiting
-interpretive clarity to simple proportional comparisons (e.g.,
-``higher-than'', ``lower-than'')
+interaction modeling, particularly the use of Wilson's
+\citep{wilson1971} system constraints. While spatial interaction models
+embraced such constraints to improve interpretability, most
+accessibility models have not. This lack of adoption contributes to
+fuzziness in current analyses, limiting interpretive clarity to simple
+proportional comparisons (e.g., ``higher-than'', ``lower-than'')
 \citep{millerAccessibilityMeasurementApplication2018}. Without such
@@
 This famous equation expresses that the attractive force \(F\) between
-two bodies is directly proportional to the product of their masses and
-inversely proportional to the square of the distance between them. As
-mass increases, so does force; as distance increases, force decreases.
-However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
+two bodies \(i\) and \(j\) is directly proportional to the product of
+their masses and inversely proportional to the square of the distance
+between them. As mass increases, so does force; as distance increases,
+force decreases. However, Eq~\ref{eq-phys-grav-prop} only demonstrates a
 \emph{proportional} relationship. To quantify the magnitude (not the
@@
 \emph{potential of opportunities for interaction} as
-\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\) with a contemporary
-rewriting of \(d^{-\beta}\) to \(f(d_{ij})\) accounts for a variety of
-impedance functions beyond the inverse power as follows:
+\(S_{i} = \sum_j \frac{O_j }{d_{ij}^\beta}\). A contemporary rewriting
+of \(d^{-\beta}\) to \(f(d_{ij})\) accounts for the variety of impedance
+functions beyond the inverse power used in the applied literature:
 
@@
 only a proportional relationship:
-\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not of calibrated equality.
+\(S_i \propto \sum_j g(O_j)f(d_{ij})\), not one of calibrated equality.
 Recognition of the nature of this relationship is not common in the
@@
 Atlantic with Alan G. Wilson. Wilson's ground breaking paper
-\citep{wilson1971}, advanced a general framework for spatial interaction
+\citep{wilson1971} advanced a general framework for spatial interaction
 modeling focused on flows of interaction between places and derived from
@@
 is available about the total number of flows produced by each origin,
-\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model} is represented as
-\(O_i\) and the following constraint can be used:
+\(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model}, represented as \(O_i\),
+then following constraint can be used:
 
@@
 
-Depending on which of the three system constraints are applied a family
+Depending on which of the three system constraints are applied, a family
 of spatial interaction models can be derived from
@@
 singly constrained model (either by origins \(A_i\) or destinations
-\(B_j\)), and (3) a doubly constrained model that satisfies both.
+\(B_j\) - see below), and (3) a doubly constrained model that satisfies
+both.
 
@@
 the system of analysis, Hansen-type measures (still widely used today in
-accessibility work) yield outputs that reflect proportional--not
-equal--relationships and typically lack interpretable units.
+accessibility work) yield outputs that reflect proportional -- not equal
+-- relationships and typically lack interpretable units.
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_V1_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 
-\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) is the mass of the
+\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) reflect the mass of the
 destination (i.e., opportunities) and mass of the origin (i.e.,
@@
 
+As a summary, each member of the family of accessibility measure is
+named, explained in plain language, alongside their balancing factor(s),
+proportional allocation factor(s), and mathematical equation and value
+interpretations in Table 1.
+
+\blandscape
+
+Table 1: Summary of constrained accessibility measure types and
+interpretations
+
+\input{summary-family-access-measures-table.tex}
+
+\elandscape
+
 \subsection{Toy example setup}\label{toy-example-setup}
 
-Consider the simple 3-zone region in Figure
+Consider the simple 3-zone region in
 Fig~\ref{fig-analytical-device-conc-accessibility}, where each zone
@@
 
-First, Table Table~\ref{tbl-small-system-land-use} summarizes population
-(in 10,000s) and physicians per zone. For context, the
+First, Table~\ref{tbl-small-system-land-use} summarizes population (in
+10,000s) and physicians per zone. For context, the
 Provider-to-Population Ratio (PPR) is 24.5 comparable to Canada's 2022
 PPR of 24.97 physicians per 10,000 \citep{whoMedicalDoctors102025}.
-Second, Table Table~\ref{tbl-small-system-cost} shows travel times
-(minutes); it can be discerned that Zones 1 and 3 are closer to each
-other than to Zone 2. Zones 1 and 3 together have a population roughly
-equal to Zone 2 but offer more than twice the physician availability. We
-interpret Zone 1 as the Urban Edge, Zone 3 as part of the Urban Core,
-and Zone 2 as the Suburban.
+Second, Table~\ref{tbl-small-system-cost} shows travel times (minutes);
+it can be discerned that Zones 1 and 3 are closer to each other than to
+Zone 2. Zones 1 and 3 together have a population roughly equal to Zone 2
+but offer more than twice the physician availability. We interpret Zone
+1 as the Urban Edge, Zone 3 as part of the Urban Core, and Zone 2 as
+Suburban.
 
@@
 
-For example, Table
-Table~\ref{tbl-simple-example-unconstrained-accessibility} shows
-\(V^0_{i}\) under each decay function. Comparing across decay types is
-meaningless in absolute terms. For instance, the difference in zone 1
-(edge of urban core)'s accessibility under \(f_3\) vs \(f_1\) is 370.92,
-but in what units? These two values are a product of different impedance
-functions ( \emph{physicians-minute\(^{-3}\)} and
+For example, Table~\ref{tbl-simple-example-unconstrained-accessibility}
+shows \(V^0_{i}\) under each decay function. Comparing across decay
+types is meaningless in absolute terms. For instance, the difference in
+zone 1 (edge of urban core)'s accessibility under \(f_3\) vs \(f_1\) is
+370.92, but in what units? These two values are a product of different
+impedance functions ( \emph{physicians-minute\(^{-3}\)} and
 \emph{physicians-minute\(^{-0.1}\)}), making the comparison
 uninterpretable (and arguably incorrect). The fundamental
-uninterpretability of what is a
+uninterpretability of what is an
 \emph{opportunity-weighted-travel-impedance} unit remains.
@@
   \(M_j^T\): where \(i\) and \(j\) of the first variant is transposed,
-  yielding a measure constrained by the total number of population and
-  to be interpreted as constrained `market potential'.
+  yielding a measure constrained by the total population and to be
+  interpreted as constrained `market potential'.
 \end{itemize}
@@
 on travel impedance. The total constrained accessibility from \(i\) to
-\(j\):
+\(j\) takes the form:
 \begin{equation}\phantomsection\label{eq-total-constrained-access}{
@@
 Referring back to our simple numeric example, \(K^T\) for the highest
-decay t scenario \(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:
+decay travel scenario \(f_1(c_{ij}) = 1/c_{ij}^3\) would then be:
 
@@
 Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields total
-constrained accessibility values (Table
-Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
+constrained accessibility values
+(Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
 units of physicians.
@@
 highest decay case, Zone 1 (Urban Edge) captures an intermediate number
-of physicians (172.0652825), like in the unconstrained accessibility
-case. However, unlike in the unconstrained case, we can say that this
-value is out of the 490 physicians in the region, which allows us also
-to deduce that zone 1 captures 1.3072213 and 0.9235529 times more than
-zone 2 and 3. Values for the lesser decay (\(f_2(c_{ij})\)) and lowest
-decay (\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
+of physicians (172.065), like in the unconstrained accessibility case.
+However, unlike in the unconstrained case, we can say that this value is
+out of the 490 physicians in the region, which allows us also to deduce
+that zone 1 captures 1.307 and 0.924 times more than zone 2 and 3.
+Values for the lesser decay (\(f_2(c_{ij})\)) and lowest decay
+(\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
 scenario values also summing to equal 490 physicians accessible in the
@@
 be observed in the unconstrained measure for this scenario. However, for
-the total constrained measure, this plateauing of results have meaning.
+the total constrained measure, this plateauing of results has meaning.
 In fact, each zone is allocated an average of the total amount in the
@@
 
-However, what's notable is how Zones change between scenarios. For
+However, what's notable is how zones change between scenarios. For
 instance, Zone 1's share only declines slightly--these declines are
@@
 opportunity-constrained formulation, switching indices \(i\) and \(j\)
-to yield s measure of market potential--the number of people who can
+to yield a measure of market potential--the number of people who can
 spatially interact with a destination.
@@
 origins. As in the total constrained case, \(V_i^S\) is expressed in the
-units of accessible opportunities. Finally, the \textbf{per capita}
-version of this measure:
-
-Soukhov et al., 2023 \citep{soukhovIntroducingSpatialAvailability2023}
-also showed that the following expression (accessibility per capita) is
-a constrained version of the popular 2SFCA approach of Shen 1998
-\citep{shen1998} and Luo and Wang 2003 \citep{luo2003}:
+units of accessible opportunities.
+
+Soukhov et al. \citep{soukhovIntroducingSpatialAvailability2023} also
+showed that the following expression (accessibility per capita) is a
+constrained version of the popular 2SFCA approach of Shen
+\citep{shen1998} and Luo and Wang \citep{luo2003}:
 
@@
 The balancing factors \(B_j\) for the \(f_2(c_{ij})\) decay group for
-zones 1, 2 and 3 area 12.8571429, 8.7685113 and 10.6635071,
-respectively. For the \(f_3(c_{ij})\) decay group, they are 0.0672461,
-0.0660559 and 0.0663798. Using these these balancing constants, we can
-calculate the singly constrained opportunity accessibility
+zones 1, 2 and 3 are 12.857, 8.769 and 10.664, respectively. For the
+\(f_3(c_{ij})\) decay group, they are 0.067, 0.066 and 0.066. Using
+these these balancing constants, we can calculate the singly constrained
+opportunity accessibility
 (Table~\ref{tbl-simple-example-attraction-constrained-accessibility}).
@@
 decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1--despite
-hosting a medium amount of physicans--captures the fewest physicians
-(133.4687282), compared to 166.7813387 at Zone 2, and 189.7499331 at
-Zone 3. Why? Zone 1 has the smallest population and is adjacent to Zone
-3, the urban core. Its low impedance-weighted demand means
-\(\kappa^S_{ij}\) allocates it fewer opportunities. By contrast, in the
-total constrained case, Zone 1 fares better, capturing 35\% of all
-physicians (compared to 27\%).
+hosting a medium amount of physicians--captures the fewest physicians
+(133.469), compared to 166.781 at Zone 2, and 189.75 at Zone 3. Why?
+Zone 1 has the smallest population and is adjacent to Zone 3, the urban
+core. Its low impedance-weighted demand means \(\kappa^S_{ij}\)
+allocates it fewer opportunities. By contrast, in the total constrained
+case, Zone 1 fares better, capturing 35\% of all physicians (compared to
+27\%).
 
@@
 capita} may be clarifying. Often, accessibility values are reported as
-raw scores without the consideration for population. But, as we
-introduced constraints, these constrained accessibility values can be
-normalized using anything that is relevant to the zone. In
+raw scores without considering the population potentially accessing
+them. But, as we introduced constraints, these constrained accessibility
+values can be normalized using anything that is relevant to the zone. In
 Table~\ref{tbl-simple-example-attraction-constrained-accessibility-per-capita},
-we present per capita accessibility for the numeric example, simply in
+we present per capita accessibility for the numeric example simply in
 units of number of physicians accessible per population at each zone.
@@
 \(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
-constrained concept of market potential been used to express ``facility
-crowdedness'' as in Wang \citep{wang_inverted_2018}.
+constrained concept of market potential has been used to express
+``facility crowdedness'' as in Wang \citep{wang_inverted_2018}.
 
@@
 
-For the sake of brevity, we'll move forward onto the doubly constrained
+For the sake of brevity, we'll move forward on to the doubly constrained
 case.
@@
 
-As the doubly constrained accessibility measure \(V_{ij}^D\) takes the
-form of the production-attraction (doubly constrained) spatial
-interaction model as follows:
+The doubly constrained accessibility measure \(V_{ij}^D\) takes the form
+of the production-attraction (doubly constrained) spatial interaction
+model as follows:
 
@@
 
-Moving onto the toy example, to calculate doubly constrained
+Moving on to the toy example, to calculate doubly constrained
 accessibility, the interpretation of the population data and the counts
@@
 This adjusted simple example is summarised in
-Table~\ref{tbl-small-system-land-use-doubly-constrained-case}: with the
+Table~\ref{tbl-small-system-land-use-doubly-constrained-case}. With the
 population (in units of 10,000s of people seeking physicians) and the
-opportunities (in units of 10,00s of physician-capacity) per zone. For
+opportunities (in units of 10,000s of physician-capacity) per zone. For
 the population, we leave this unchanged numerically but we now must keep
@@
 physicians used in previous cases
-(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now:1,
-this is compared to the unmodified example which yields system PPR
-of24.5.
+(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now 1,
+this is compared to the unmodified example which yields system PPR of
+24.5.
 
@@
 \(V^D_i = O_i\) so the physician-capacity accessible for Zones 1, 2, and
-3 would be 4.0018381, 9.9963656, and 6.0017963. Hence, the usefulness of
-the doubly constrained measure lies in the interpretation as
-\(V_{ij}^D\) values. \(V^D_{ij}\) values represent the number of
-opportunities from zone \(j\) allocated to populations in zone \(i\),
-shaped by both mass and travel impedance.
+3 would be 4.002, 9.996, and 6.002. Hence, the usefulness of the doubly
+constrained measure lies in the interpretation as \(V_{ij}^D\) values.
+\(V^D_{ij}\) values represent the number of opportunities from zone
+\(j\) allocated to populations in zone \(i\), shaped by both mass and
+travel impedance.
 
@@
 its intrazonal allocation (i.e., \(V^D_{22}\)) declines as travel
-impedance decay decreases--from 4.9593248 under \(f_1(c_{ij})\) to
-2.6672837 under \(f_3(c_{ij})\), out of the \textasciitilde10
-opportunities allocated to Zone 2 (a population of 10).
+impedance decay decreases--from 4.959 under \(f_1(c_{ij})\) to 2.667
+under \(f_3(c_{ij})\), out of the \textasciitilde10 opportunities
+allocated to Zone 2 (a population of 10).
 
@@
 counterpart \(M^D_{ji}\) are structurally identical, differing only in
-interpretation (referring to \(kappa_{ij}^D\) and
-\(\hat kappa_{ij}^D\)): one reflects the proportional allocation of
+interpretation (referring to \(\kappa_{ij}^D\) and
+\(\hat \kappa_{ij}^D\)): one reflects the proportional allocation of
 opportunity to population flows; the other, population to opportunity
@@
 
-Readers should recall, \(V_{ij}^D\) are also mathematically equivalent
-to Wilson's spatial interaction flows. And as Wilson \citep{wilson1971}
-explicitly noted, origin and destination weights defined in the spatial
-interaction model \emph{can} be defined using any unit. However,
-traditionally the focus of these models have typically on \(ij\) flows
-and often calibrated using trips (i.e., outbound and inbound trips,
-inherently in the same units).
+\(V_{ij}^D\) are also mathematically equivalent to Wilson's spatial
+interaction flows. And as Wilson \citep{wilson1971} explicitly noted,
+origin and destination weights defined in the spatial interaction model
+\emph{can} be defined using any unit. However, the focus of these models
+has typically been on \(ij\) flows, often calibrated using trips (i.e.,
+outbound and inbound trips, inherently in the same units).
 
@@
 
-A summary of the family of accessibility measures, is detailed; First,
-we place the popular Hansen-type accessibility measure
-\citep{hansen1959} within this family of measures as an
-``unconstrained'' case, demonstrating that resulting values cannot be
-directly compared across different travel scenarios without ad-hoc
-adjustments. We then show how applying a total constraint balances the
-units and produces a statistically averaged solution that converges to
-the regional average for each zone as the decay effect decreases. In
-other words, the total-constraint model could be a more interpretable
-alternative for the unconstrained case if population-competition is not
-relevant and one is interested in capturing the maximum
-\emph{potential}; specifically, if there is a fixed number of
-opportunities in the region, and if it makes sense to assume that people
-accessing proximate opportunities leave fewer for others, \emph{without}
-considering the population size at the origins.
+To summarize the contributions of this work, first, we place the popular
+Hansen-type accessibility measure \citep{hansen1959} within this family
+of measures as an ``unconstrained'' case, demonstrating that resulting
+values cannot be directly compared across different travel scenarios
+without ad-hoc adjustments. We then show how applying a total constraint
+balances the units and produces a statistically averaged solution that
+converges to the regional average for each zone as the decay effect
+decreases. In other words, the total-constraint model could be a more
+interpretable alternative for the unconstrained case if
+population-competition is not relevant and one is interested in
+capturing the maximum \emph{potential}; specifically, if there is a
+fixed number of opportunities in the region, and if it makes sense to
+assume that people accessing proximate opportunities leave fewer for
+others, \emph{without} considering the population size at the origins.
 
@@
 accessibility values for each \(i\)-\(j\) pair must be a proportion of
-he zonal opportunity and population values simultaneously. For example,
-the accessibility at zone 1 must equal the sum of opportunities from
-zones 1, 2, and 3, as well as the sum of the population at zone 1.
+the zonal opportunity and population values simultaneously. For example,
+the accessibility at Zone 1 must equal the sum of opportunities from
+zones 1, 2, and 3, as well as the sum of the population at Zone 1.
 Satisfying the double constraint means the opportunities and population
@@
 (e.g., typically expressed as a zonal summary measure of how many
-opportunities one could reach (out of a regional total and/or
+opportunities one could reach out of a regional total and/or
 zonal-allocation).
 
-Building on Wilson's \citep{wilson1971} foundational work, this paper
-proposed a unified framework for accessibility. By reintroducing
-Wilson's proportionality constants, the proposed family of constrained
-accessibility measures restores measurement units to accessibility
-estimates. This enhancement provides a more interpretable, consistent,
-and theoretically grounded basis for accessibility analysis, which could
-help advance the adoption of accessibility-oriented planning.
+Overall, and building on Wilson's \citep{wilson1971} foundational work,
+this paper proposes a unified framework for analyzing gravity-based
+accessibility. By reintroducing Wilson's proportionality constants, the
+family of constrained accessibility measures restores measurement units
+to accessibility estimates. This enhancement provides a more
+interpretable, consistent, and theoretically grounded basis for
+accessibility analysis, which could help advance the adoption of
+accessibility-oriented planning.
 
 That said, there have been more recent developments in the accessibility
-literature which include novel approaches such as person-based
-approaches, including those that are time-sensitive
-\citep{yang2024evaluating, braga2023evaluating}, behavioural
-\citep{kar2024inclusive, lu2014effects} or utility-based
+literature, including novel person-based approaches that are
+time-sensitive \citep{yang2024evaluating, braga2023evaluating},
+behavioural \citep{kar2024inclusive, lu2014effects} or utility-based
 \citep{guzman2023much, ben1985discrete}, that are not based on top-down
 approaches such as Wilson's model formalized through the
-entropy-maximisation procedure. Future work could further explore how
-these ideas of proportionality constants and balanced units could help
-inform new modeling approaches.
+entropy-maximisation procedure. While gravity-based measures of
+accessibility dominate the applied literature, future work could further
+explore how these ideas of proportionality constants and balanced units
+could also help inform these and other new modeling approaches.
 

### 2025-09-03 paper_plos/reviewer_respose_letter_V1.qmd
final changes to R1

@@
 format: pdf
+bibliography: bibliography.bib 
 ---
@@
 
-# Response to Reviewer #1
+# Response to Reviewer 1
 
@@
 
-# Response to Reviewer #2
+# Response to Reviewer 2
 
@@
 Namely, concerning the unconstrained example on pg.
-16 we write: "For example, Table 3 shows $V^0_{i}$ under each decay function.
+17 we write: "For example, Table 4 shows $V^0_{i}$ under each decay function.
 Comparing across decay types is meaningless in absolute terms.
@@
 Building on this idea, we discuss the interpretive clarity in the total constrained numeric example on pg.
-18: "Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
+19: "Compared to the unconstrained case, values now sum to the known regional total $D$, allowing interpretation of absolute and relative differences across zones and travel scenarios.
 For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (172.06), like in the unconstrained accessibility case.
@@
 We also go on to write on pg.
-18: "One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
+19: "One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
 As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
@@
 
-\textcolor{red}{Rafa: I actually liked table 1 and thought perhaps we should keep the table and remove the redundancy of the text around it. But I'm Ok with this solution below."}
-\textcolor{red}{Ana: What do you think Chris?
-I prefer my approach as opposed to the table, but I could easily add it back in.
-The Table looks a bit confusing in my eyes, I personally don't like reading text in tables, I like reading text in paragraphs.}
-
-Excellent point.
+A fair point.
 We've streamlined redundancies throughout the manuscript, especially in the "A family of accessibility measures" section.
-Specifically regarding table 1, we've decided to remove Table 1 all together, and provide a more condensed version of it in text on pg 13/14.
-Namely:
+Specifically regarding table 1, we've gone back-and-forth on it and ultimately opt to include it as a helpful and more plain-language summary.
+We also provide more condensed descriptions in text, namely:
 
@@
 2. **Total Constrained Case ($V_i^T$, $M_j^T$)**
-- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in @fig-analytical-device-conc-accessibility) i.e., total number of opportunities or population in the system.
+- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in Figure 2) i.e., total number of opportunities or population in the system.
 This ensures the sum of all values in the system match the total marginal.
@@
 3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
-- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
+- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in Figure 2) i.e., the number of opportunities or population at each zone.
 Reflects how the literature calculates competitive accessibility.
@@
 - Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously, which can also be expressed as proportional allocation factors ($\kappa_{ij}^D$, $\hat \kappa_{ji}^D$); equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
-- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in @fig-analytical-device-conc-accessibility).
+- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in Figure 2).
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
@@
 We hope this addresses your concern!
+
+# References

### 2025-10-15 paper_plos/paper_plos_V2.qmd
V2 revision, first draft

--- 2025-09-03 paper_plos/paper_plos_V1.qmd
@@
   Transportation planning has long prioritized the efficiency of movement.
-However, the concept of accessibility represents a more comprehensive evolution, shifting focus from movement to the potential to reach--or spatially interact with--desired destinations.
+However, the concept of accessibility represents a more comprehensive evolution, shifting focus from movement (i.e., trips) to the potential to spatially interact with desired destinations.
 Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations.
-To this end, this paper offers a methodological contribution by specifying a family of accessibility measures grounded in the shared theoretical roots of gravity-based accessibility and spatial interaction models, particularly their balancing factors.
+To this end, this paper offers a methodological contribution by specifying a family of accessibility measures that are grounded in the shared 'gravity-based' theoretical roots of spatial interaction models, particularly their balancing factors.
 From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
@@
 
-Accessibility and spatial interaction modelling literature share common headwaters, and the latter has given careful attention to measurement units and their interpretability.
-Following this introduction, this paper revisits that lineage--from Ravenstein's Newtonian gravitational expression of spatial interaction [@ravensteinLawsMigration1889], through Hansen [@hansen1959] to Wilson [@wilson1971]--to propose a unified family of constrained accessibility measures.
-We illustrate each case of the family with a numerical toy example and conclude by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.
+The remainder of this paper proceeds as follows.
+Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4).
+Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
+Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations.
+Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied--unconstrained, singly-constrained, and doubly-constrained cases--demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices.
+We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions.
+Section 8 concludes by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis.
+
+The key insight is that accessibility need not be unit-inconsistent.
+By building from spatial interaction theory's constraint framework rather than the gravitational analogy alone, we show how to construct accessibility measures that are simultaneously grounded in behavioural principles (the gravitational analogy) as well as being expressed in interpretable units (opportunities or population) that are sensitive to the known to the region through empirically-set constraints.
 
@@
 
-# Early research on human spatial interaction: from Ravenstein (1889) to Stewart (1948)
+# Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)
 
@@
 This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-In this way, spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
-While Wilson's model produces results that are in units of flow tethered to the system of analysis, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.
-
-As a preview for the work ahead: Wilson's balancing factors can be manipulated into _proportional allocation factors_ and equations can be established to reflect accessibility.
-The resulting $ij$ flows can be understood as accessibility flows--and aggregated at the zone level to express accessibility in units of actual opportunities or population.
-
-In this way, these balancing factors can be used to recover what Stewart's $G$ was meant to accomplish, but through a different conceptual path.
-But before demonstrating a family of accessibility measures that are sensitive to constraints, in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
+The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
+In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
+While Wilson's model produces results that are in units of flow tethered to the system of analysis, which facilitates such interpretations, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units.
+
+Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
 This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper.
@@
 This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
-This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially proposed in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).
+This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially suggested in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).
 
@@
 Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter.
-This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility (a la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
+This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility ($\grave{a}$ la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
 This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.

### 2025-10-15 paper_plos/paper_plos_V2.tex
V2 revision, first draft

--- 2025-09-03 paper_plos/paper_plos_V1.tex
@@
 However, the concept of accessibility represents a more comprehensive
-evolution, shifting focus from movement to the potential to reach--or
-spatially interact with--desired destinations. Despite growing
+evolution, shifting focus from movement (i.e., trips) to the potential
+to spatially interact with desired destinations. Despite growing
 recognition of accessibility-based planning approaches, the concept
@@
 interpretations. To this end, this paper offers a methodological
-contribution by specifying a family of accessibility measures grounded
-in the shared theoretical roots of gravity-based accessibility and
-spatial interaction models, particularly their balancing factors. From
-this foundation, we outline four memebers of the family: the
-`unconstrained' measure (i.e., Hansen-type accessibility), the `total
-constrained' measure (i.e., a constrained version of the Hansen-type
-accessibility), the `singly constrained' measure (i.e., related to the
-popular two step floating catchment approach - 2SFCA), and the `doubly
-constrained' measure representing realized access (i.e., equal to the
-doubly constrained spatial interaction model). These measures can be
+contribution by specifying a family of accessibility measures that are
+grounded in the shared `gravity-based' theoretical roots of spatial
+interaction models, particularly their balancing factors. From this
+foundation, we outline four memebers of the family: the `unconstrained'
+measure (i.e., Hansen-type accessibility), the `total constrained'
+measure (i.e., a constrained version of the Hansen-type accessibility),
+the `singly constrained' measure (i.e., related to the popular two step
+floating catchment approach - 2SFCA), and the `doubly constrained'
+measure representing realized access (i.e., equal to the doubly
+constrained spatial interaction model). These measures can be
 interpreted as either the number of accessible opportunities or
@@
 
-Accessibility and spatial interaction modelling literature share common
-headwaters, and the latter has given careful attention to measurement
-units and their interpretability. Following this introduction, this
-paper revisits that lineage--from Ravenstein's Newtonian gravitational
-expression of spatial interaction \citep{ravensteinLawsMigration1889},
-through Hansen \citep{hansen1959} to Wilson \citep{wilson1971}--to
-propose a unified family of constrained accessibility measures. We
-illustrate each case of the family with a numerical toy example and
-conclude by discussing how these constrained measures can inform
-planning decisions and improve clarity in accessibility analysis.
+The remainder of this paper proceeds as follows. Sections 2-4 trace the
+historical development of spatial interaction and accessibility
+research, beginning with Newtonian gravitational analogies and
+\citet{careyPrinciplesSocialScience1858} (Section 2), moving through
+early researchers like \citet{ravensteinLawsMigration1885} to
+\citet{stewartDemographicGravitationEvidence1948} who theorised and
+formalized spatial interaction patterns (Section 3), and examining how
+the ``gravity-based'' accessibility approach in \citet{hansen1959}
+became the dominant approach in planning practice (Section 4). Section 5
+presents the entropy-based family of spatial interaction models in
+\citet{wilson1971}, explaining how the introduction of ``constraints''
+(based on top-down known information as part of entropy-maximisation)
+produce interpretable, unit-consistent flow estimates. Section 6
+explores why accessibility and spatial interaction literature diverged
+into separate branches despite their shared conceptual foundations.
+Section 7 constitutes the paper's core contribution: we derive a family
+of accessibility measures corresponding to different constraints
+applied--unconstrained, singly-constrained, and doubly-constrained
+cases--demonstrating how each yields zone-level accessibility values
+expressed in meaningful units (opportunities or population) rather than
+unit-inconsistent indices. We illustrate each measure with numerical
+examples that clarify the practical implications of different constraint
+assumptions. Section 8 concludes by discussing how these constrained
+measures can inform planning decisions and improve clarity in
+accessibility analysis.
+
+The key insight is that accessibility need not be unit-inconsistent. By
+building from spatial interaction theory's constraint framework rather
+than the gravitational analogy alone, we show how to construct
+accessibility measures that are simultaneously grounded in behavioural
+principles (the gravitational analogy) as well as being expressed in
+interpretable units (opportunities or population) that are sensitive to
+the known to the region through empirically-set constraints.
 
@@
 \section{Early research on human spatial interaction: from Ravenstein
-(1889) to Stewart
-(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1889-to-stewart-1948}
+(1885) to Stewart
+(1948)}\label{early-research-on-human-spatial-interaction-from-ravenstein-1885-to-stewart-1948}
 
@@
 interpretable, balanced (given the constraints), and unit-consistent
-\(ij\) flows. In this way, spatial interaction modeling tradition can be
-seen to have succeeded where accessibility modeling stalled. While
-Wilson's model produces results that are in units of flow tethered to
-the system of analysis, Hansen-type measures (still widely used today in
+\(ij\) flows. The balancing factors themselves have been subject to
+various interpretations--as terminal costs
+\citep{dieter1962distribution}, weighted mean values
+\citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
+themselves \citep{cesario1977new} (as suggested in \citet{Wilson1971})
+and as rents \citep{morphetGravityModelCalibration2023}--reflecting
+ongoing efforts to understand what these mathematical constructs
+represent behaviourally. In this way, the spatial interaction modeling
+tradition can be seen to have succeeded where accessibility modeling
+stalled. While Wilson's model produces results that are in units of flow
+tethered to the system of analysis, which facilitates such
+interpretations, Hansen-type measures (still widely used today in
 accessibility work) yield outputs that reflect proportional -- not equal
@@
 
-As a preview for the work ahead: Wilson's balancing factors can be
-manipulated into \emph{proportional allocation factors} and equations
-can be established to reflect accessibility. The resulting \(ij\) flows
-can be understood as accessibility flows--and aggregated at the zone
-level to express accessibility in units of actual opportunities or
-population.
-
-In this way, these balancing factors can be used to recover what
-Stewart's \(G\) was meant to accomplish, but through a different
-conceptual path. But before demonstrating a family of accessibility
-measures that are sensitive to constraints, in the next section we
+Before demonstrating the derivation of family of accessibility measures
+that are sensitive to constraints (Section 7), in the next section we
 review how conceptually intertwined the spatial interaction modelling
@@
 
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_V1_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V2_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 \citep{wilson1971}, and does not rely on specifying some universal
-constant \(G\) like initially proposed in Stewart
+constant \(G\) like initially suggested in Stewart
 \citep{stewartDemographicGravitationEvidence1948} (recall,
@@
 approach grounded in statistical mechanics, shifting place-based
-accessibility (a la Hansen \citep{hansen1959}) from a relationship
-describing proportionality to equality. This shift addresses the
-fundamental issue of interpretability associated with the proportional
-nature of Hansen-type accessibility indicators we previous outlined.
+accessibility (\(\grave{a}\) la Hansen \citep{hansen1959}) from a
+relationship describing proportionality to equality. This shift
+addresses the fundamental issue of interpretability associated with the
+proportional nature of Hansen-type accessibility indicators we previous
+outlined.
 

### 2025-10-15 paper_plos/reviewer_respose_letter_V2.qmd
V2 revision, first draft

---
title: Response to reviewers V2
format: pdf
bibliography: bibliography.bib 
---

Dear editors,

We appreciate the valuable comments from the reviewers on this second round of revisions. Changes to the manuscript are much lighter than the previous round, reflecting the quality of the manuscript as is and the lighter points raised by the reviewers.

We address each comment of the reviewers point-by-point, with our response in black font and the reviewers comments in \textcolor{blue}{blue}.

# Response to Reviewer 1

\textcolor{blue}{NA}

Thank you for your efforts in the previous round of revision Reviewer 1!

# Response to Reviewer 2

\textcolor{blue}{NA}

Thank you for your efforts in the previous round of revision Reviewer 2!

# Response to Reviewer 3

\textcolor{blue}{
This is my first time seeing the paper, though I can see it has already been through a major revision. Looking at the previous comments and changes it looks like you have responded to the comments of the reviewers. The paper makes a valuable contribution in drawing together these different accessibility measures that many researchers know are somehow related but wouldn’t be able to explain how. The review of literature included also adds value and will make this a valuable paper for people working in the field to read.}

\textcolor{blue}{
I was hoping not to have to ask for any further revision because I know it can be frustrating when a new reviewer is brought in after a revision has been made. However, reading your paper reminded me of Cesario (1977) who also discussed the interpretation of balancing factors from spatial interaction models as accessibility measures. I can’t see that this is cited in the work, but given the nature of what you are doing and the historical approach taken in parts of your paper I think it should be included in the discussion.}

\textcolor{blue}{
Cesario, F. J. (1977). A new interpretation of the "normalizing" or "balancing" factors of gravity-type spatial models. Socio-Economic Planning Sciences, 11(3), 131-136.}

Thank you for these positive comments and for the helpful suggestion regarding Cesario 1977. We have incorporated this citation in the section preceding the literature review Section 6 (i.e., "Accessibility and spatial interaction modelling: two divergent research streams"). Specifically, the final paragraphs of "Wilson's family of spatial interaction models" section 5 now read:

_Wilson's work is notable for many reasons. Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation. This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows. The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally. In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled. While Wilson's model produces results that are in units of flow tethered to the system of analysis, which facilitates such interpretations, Hansen-type measures (still widely used today in accessibility work) yield outputs that reflect proportional -- not equal -- relationships and typically lack interpretable units._

_Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge. This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper._

# Response to Reviewer 4

\textcolor{blue}{The manuscript presents a theoretically rich and mathematically detailed framework that attempts to bridge spatial interaction modelling and accessibility analysis. While the topic is interesting and potentially valuable, I found the paper quite challenging to read due to its length and the density of the mathematical exposition. The presentation is heavy and at times difficult to follow, which limits accessibility to a broader audience.}

We appreciate the recognition that our work addresses an important theoretical gap. In the previous revision round, we substantially shortened the manuscript from 34 to 29 pages, streamlined the use of mathematical exposition, and clarified each section's contribution in response to concerns about length from other reviewers. This reviewer now reports to be satisfied with our changes in this round.

While we hear you that the mathematical detail may be dense, we believe it is inherent to our contribution. Our manuscript establishes formal relationships between spatial interaction models and accessibility measures--specifically focusing on the use of the balancing factors. This manuscript offers a methodological and theoretical advancement that we view requires mathematical logic that matches what is common in the field of spatial interaction.

To address the general concern with reader navigation, we have enhanced the paper's overview (Section 1, final paragraphs). We now explicitly preview the paper's logical structure, explaining what each section accomplishes and how Sections 2-6 build the conceptual foundations for the derivations (core contribution of the paper) in Section 7. We believe this will help the readers navigate the text with clearer orientation. The following is the revised text:

_The remainder of this paper proceeds as follows. Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4). Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates. Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations. Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied--unconstrained, singly-constrained, and doubly-constrained cases--demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices. We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions. Section 8 concludes by discussing how these constrained measures can inform planning decisions and improve clarity in accessibility analysis._

_The key insight is that accessibility need not be unit-inconsistent. By building from spatial interaction theory's constraint framework rather than the gravitational analogy alone, we show how to construct accessibility measures that are simultaneously grounded in behavioural principles (the gravitational analogy) as well as being expressed in interpretable units (opportunities or population) that are sensitive to the known to the region through empirically-set constraints._

As well, we'd like to note that the manuscript already underwent extension revision that included the addition of intuitive explanations before mathematical developments, removal/consolidation of tangential equations throughout to improve readability, and transitional text at the end of each section.

Together, we believe the addition to the Introduction and previous changes made should address the Reviewer's concerns. We believe the manuscript strikes a balance between readability and the detail offered by the mathematical expressions. We would welcome more specific and directed feedback from the reviewer regarding particular sections or derivations that remain unclear, as this would allow us to target improvements more effectively.

# References

### 2025-10-16 paper_plos/paper_plos_V2.qmd
cdh edits for v2 submission

@@
 The remainder of this paper proceeds as follows.
-Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4).
-Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
+Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and Carey [@careyPrinciplesSocialScience1858] (Section 2), moving through early researchers like Ravenstein [@ravensteinLawsMigration1885] to Stewart [@stewartDemographicGravitationEvidence1948] who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in Hansen [@hansen1959] became the dominant approach in planning practice (Section 4).
+Section 5 presents the entropy-based family of spatial interaction models in Wilson [@wilson1971], explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
 Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations.
@@
 This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
+The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves ([@cesario1977new], as suggested in Wilson [@wilson1971]) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
 In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.

### 2025-10-16 paper_plos/paper_plos_V2.tex
cdh edits for v2 submission

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 ]{article}
-
+\usepackage{xcolor}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage[numbers,square,comma]{natbib}
+\bibliographystyle{plos2015}
 
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[numbers,square,comma]{natbib}
-\bibliographystyle{plos2015}
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 historical development of spatial interaction and accessibility
-research, beginning with Newtonian gravitational analogies and
-\citet{careyPrinciplesSocialScience1858} (Section 2), moving through
-early researchers like \citet{ravensteinLawsMigration1885} to
-\citet{stewartDemographicGravitationEvidence1948} who theorised and
-formalized spatial interaction patterns (Section 3), and examining how
-the ``gravity-based'' accessibility approach in \citet{hansen1959}
-became the dominant approach in planning practice (Section 4). Section 5
-presents the entropy-based family of spatial interaction models in
-\citet{wilson1971}, explaining how the introduction of ``constraints''
-(based on top-down known information as part of entropy-maximisation)
-produce interpretable, unit-consistent flow estimates. Section 6
-explores why accessibility and spatial interaction literature diverged
-into separate branches despite their shared conceptual foundations.
-Section 7 constitutes the paper's core contribution: we derive a family
-of accessibility measures corresponding to different constraints
-applied--unconstrained, singly-constrained, and doubly-constrained
-cases--demonstrating how each yields zone-level accessibility values
-expressed in meaningful units (opportunities or population) rather than
-unit-inconsistent indices. We illustrate each measure with numerical
-examples that clarify the practical implications of different constraint
-assumptions. Section 8 concludes by discussing how these constrained
-measures can inform planning decisions and improve clarity in
-accessibility analysis.
+research, beginning with Newtonian gravitational analogies and Carey
+\citep{careyPrinciplesSocialScience1858} (Section 2), moving through
+early researchers like Ravenstein \citep{ravensteinLawsMigration1885} to
+Stewart \citep{stewartDemographicGravitationEvidence1948} who theorised
+and formalized spatial interaction patterns (Section 3), and examining
+how the ``gravity-based'' accessibility approach in Hansen
+\citep{hansen1959} became the dominant approach in planning practice
+(Section 4). Section 5 presents the entropy-based family of spatial
+interaction models in Wilson \citep{wilson1971}, explaining how the
+introduction of ``constraints'' (based on top-down known information as
+part of entropy-maximisation) produce interpretable, unit-consistent
+flow estimates. Section 6 explores why accessibility and spatial
+interaction literature diverged into separate branches despite their
+shared conceptual foundations. Section 7 constitutes the paper's core
+contribution: we derive a family of accessibility measures corresponding
+to different constraints applied--unconstrained, singly-constrained, and
+doubly-constrained cases--demonstrating how each yields zone-level
+accessibility values expressed in meaningful units (opportunities or
+population) rather than unit-inconsistent indices. We illustrate each
+measure with numerical examples that clarify the practical implications
+of different constraint assumptions. Section 8 concludes by discussing
+how these constrained measures can inform planning decisions and improve
+clarity in accessibility analysis.
 
@@
 \citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
-themselves \citep{cesario1977new} (as suggested in \citet{Wilson1971})
-and as rents \citep{morphetGravityModelCalibration2023}--reflecting
-ongoing efforts to understand what these mathematical constructs
-represent behaviourally. In this way, the spatial interaction modeling
-tradition can be seen to have succeeded where accessibility modeling
-stalled. While Wilson's model produces results that are in units of flow
-tethered to the system of analysis, which facilitates such
-interpretations, Hansen-type measures (still widely used today in
-accessibility work) yield outputs that reflect proportional -- not equal
--- relationships and typically lack interpretable units.
+themselves (\citep{cesario1977new}, as suggested in Wilson
+\citep{wilson1971}) and as rents
+\citep{morphetGravityModelCalibration2023}--reflecting ongoing efforts
+to understand what these mathematical constructs represent
+behaviourally. In this way, the spatial interaction modeling tradition
+can be seen to have succeeded where accessibility modeling stalled.
+While Wilson's model produces results that are in units of flow tethered
+to the system of analysis, which facilitates such interpretations,
+Hansen-type measures (still widely used today in accessibility work)
+yield outputs that reflect proportional -- not equal -- relationships
+and typically lack interpretable units.
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V2_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{paper_plos_V2_files/figure-pdf/fig-docs-per-year-1.pdf}
 

### 2025-10-22 paper_plos/paper_plos_V3.qmd
V3 AS revision; small gramar changes here and there.

--- 2025-10-15 paper_plos/paper_plos_V2.qmd
@@
 To this end, this paper offers a methodological contribution by specifying a family of accessibility measures that are grounded in the shared 'gravity-based' theoretical roots of spatial interaction models, particularly their balancing factors.
-From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach - 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
+From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach -- 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
 These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential).
-A toy example illustrates how they produce interpretable, unit-based values, offering a clearer and more coherent basis for accessibility analysis.
+A toy example illustrates how they produce interpretable unit-based values, offering a clearer and more coherent basis for accessibility analysis.
 
@@
 Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations.
-Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied--unconstrained, singly-constrained, and doubly-constrained cases--demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices.
+Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied -- unconstrained, singly constrained, and doubly constrained cases -- demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices.
 We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions.
@@
 
-Where $G$ is the gravitational constant--an empirically calibrated value that ensures the model reflects observed forces.
+Where $G$ is the gravitational constant -- an empirically calibrated value that ensures the model reflects observed forces.
 Newton's initial estimate of $G$ was based on a speculation <!--that the mean density of earth was between five or six times that of water, an assumption that -->but received empirical support after Hutton's and Cavendish' experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] which estimated it to within 1% accuracy.
@@
 Henry C.
-Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858] inspired empirical spatial intereaction research in different contexts.
-Namely, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance--conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
+Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858] inspired empirical spatial interaction research in different contexts.
+Namely, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance -- conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
 
@@
 
-Although Stewart's concept of "social physics" eventually fell out of favour--due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p.
+Although Stewart's concept of "social physics" eventually fell out of favour, potentially in part due to inconsistent mathematical notation (e.g., $G$ is used as both a proportionality constant p.
 34 and then later as _demographic energy_ on p.
-53.) and its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
-35\])--his introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration.
-In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having interpretable units grounded in consistent, albeit still abstract, quantities (i.e., units of demographic force).
+53.) as well as its racist and unscientific assumptions (e.g., view of humans as particles following physical laws and assumptions of the molecular weight of the average American being one, but "presumably...much less than one....for an Australian aborigine" \[p.
+35\]).
+However, Stewart's introduction of a proportionality constant $G$ in modeling demographic force marks an important conceptual step: recognizing that moving from proportionality to equality requires empirical calibration.
+In other words, the addition of $G$ shifts results from being abstract indicators of potential (i.e., $\frac{\text{people}^2}{\text{distance}^{2}}$) to having units grounded in consistent, albeit still abstract, quantities (i.e., units of some sort of demographic force).
 
@@
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
-In this seminal paper, Hansen defined accessibility as "the potential of opportunities for interaction...
+In this seminal paper, Hansen defined accessibility as the potential of opportunities for interaction...
 a generalization of the population-over-distance relationship or *population potential* concept developed by Stewart [@stewartDemographicGravitationEvidence1948]" (p.
@@
 
-$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from site $i$.
+$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from a zone $i$.
 This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $f(d_{ij})$ (a function that modulates the friction of cost).
@@
 Wilson's ground breaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
-Additionally, this work was not focused on the 'potential' concept, as accessibility has been.
+Additionally, this work was not focused on the 'potential' concept as accessibility had been.
 Wilson [@wilson1971] formalized the general spatial interaction model through the following equation:
@@
 Of important note for this paper, $k$ in @eq-phys-gravity-model acts as a proportionality constant, shifting the equation from a proportional to an equal relationship by incorporating known system totals.
-In some sense, $k$ serves a role similar to the gravitational constant in Newton's law--it calibrates the model so that outputs match real-world quantities.
+In some sense, $k$ serves a role similar to the gravitational constant in Newton's law -- it calibrates the model so that outputs match real-world quantities.
 However, these real-world quantities are not a set empirical constant (like Newton's $G$) but instead, sensitive to the system and known information about the system.
@@
 From the outset, spatial interaction models emphasized interpretability of results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
-But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson's innovation was to ground and generalise the model using _entropy maximisation_.
+But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson's innovation was to generalise the model using _entropy maximisation_.
 By maximizing the number of ways individual trip probabilities could be arranged under known constraints, Wilson derived models that estimate _statistical averages_ of flows between zones [@wilson1971; @seniorGravityModellingEntropy1979].
@@
 
-In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$ - see below), and (3) a doubly constrained model that satisfies both.
+In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$ -- explained below), and (3) a doubly constrained model that satisfies both.
 
@@
 
-In the singly-constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
-When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the production-constrained singly-constrained version of @eq-phys-gravity-model, where the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$.
+In the singly constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
+When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the production-constrained singly constrained version of @eq-phys-gravity-model, where the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$.
 Also, $k$ is replaced with a set of balancing factors specific to origins $A_i$ which ensures that constraint @eq-constraint1-gravitymodel is satisfied (i.e., the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$).
@@
 
-The singly-constrained attraction-constrained model is similar to the production-constrained version but from the perspective of the mass at the destination.
+The singly constrained attraction-constrained model is similar to the production-constrained version but from the perspective of the mass at the destination.
 For the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is replaced with $D_j$ in @eq-phys-gravity-model, representing the spatial interaction inbound flow.
@@
 
-Lastly, the doubly-constrained model is the production-attraction constrained model in Wilson [@wilson1971].
+Lastly, the doubly constrained model is the production-attraction constrained model in Wilson [@wilson1971].
 In this case, both constraints @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
@@
 Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation.
-This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
+This results in interpretable, balanced (given the system knowns are expressed in the constraints), and unit-consistent $ij$ flows.
+The balancing factors themselves have been subject to various interpretations -- as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023] -- reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
 In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
@@
 Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
-This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper.
+This investigation sheds light on why accessibility research may have failed to adopt a comparable approach, until this paper.
 
@@
 Meanwhile, Wilson's [@wilson1971] entropy-maximising framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
-We argue the framework's quiet innovation--introducing empirically grounded constraints to shift from proportionality to calibrated equality--made the framework immediately relevant for policy applications as outputs were in tangible units.
+We argue the framework's quiet innovation -- introducing empirically grounded constraints to shift from proportionality to calibrated equality -- made the framework immediately relevant for policy applications as outputs were in tangible units.
 Yet, this mechanism was never widely adopted in accessibility analysis.
@@
 
-To explore why Wilson's approach never crossed over to accessibility modeling, we conduct a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both (on Web of Science using the "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]) .
-Only `r (n_b) |> prettyNum(big.mark = ",")` out of the `r (n_h+n_w+n_b) |> prettyNum(big.mark = ",")` documents that emerged from our search cite both.
+To explore why Wilson's approach may not have crossed over to accessibility modeling sooner, we conducted a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both.
+This was done using Web of Science's "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]).
+Only `r (n_b) |> prettyNum(big.mark = ",")` out of the `r (n_h+n_w+n_b) |> prettyNum(big.mark = ",")` documents that emerged from our search of the General Database cite both.
 The number of articles, by year and if they cite Hasen, Wilson, or both are shown in @fig-docs-per-year.
 
-Through the close analysis of _how_ articles that cite both works, we identify two distinct citation patterns: one group of articles focused on accessibility, the other on spatial interaction.
-In examining these articles, we uncover how the relationship between the two has often been misunderstood, underexplored, or entirely overlooked.
-
-In the first stream of literature--which cite both but are focused more so on spatial interaction models--they treat spatial interaction and accessibility as separate but related phenomenons.
+Through the close analysis of why articles that cite both works cite each work, we identify two distinct patterns: one group of articles focused on accessibility, the other on spatial interaction.
+In examining these groups of articles, we uncover how the relationship between Wilson and Hansen has often been misunderstood, underexplored, or entirely overlooked.
+
+In the first stream of literature -- which cite both but are focused more so on spatial interaction models -- they treat spatial interaction and accessibility as separate but related phenomenons.
 Four subsets of this stream emerge.
@@
 
-The third subset of the spatial-interaction focused literature, depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
-In sum though, across these works, they recognize Hansen-type accessibility as an indicator of 'potential' but as a separate but related concept to spatial interaction.
+The third subset of the spatial-interaction focused literature depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+Though across works in this subset, they recognize Hansen-type accessibility as an indicator of 'potential' but as a separate but related concept to spatial interaction.
 
 Moving onto the group of accessibility-focused literature that cites both works, we categorise their citation of Wilson [@wilson1971] within three general groups.
-Overall though, these works do not engage, or only superficially engage with Wilson [@wilson1971].
+Overall these works do not engage -- or only superficially engage -- with Wilson [@wilson1971].
 
 Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions.
-This trend is common: for instance it is done in [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+This trend is common: for instance it is done in the following papers: [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 However, these works do not engage with spatial interaction beyond this attribution.
 
-Secondly, a subset of literature explicitly associates spatial interaction, as defined in Wilson [@wilson1971], with accessibility's potential for spatial interaction--but only superficially.
-These works acknowledge the conceptual link, but do not go beyond this recognition in the scope of their works e.g., [@grengs2010job; @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
-Indeed, while accessibility can be seen as the *potential* for spatial interaction--and Wilson [@wilson1971] briefly touches on this--such mentions have not resulted in deeper analytical integration of these concepts.
+Secondly, a subset of literature explicitly acknowledge the conceptual link between spatial interaction and with Hansen-type accessibility -- but only superficially, not going beyond the recognition that they both relate to spatial interaction e.g., [@grengs2010job; @millerMeasuringSpaceTimeAccessibility1999; @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring].
+Indeed, while accessibility can be seen as the *potential* for spatial interaction -- and Wilson [@wilson1971] briefly touches on this -- such mentions have not resulted in deeper analytical integration of these concepts.
 Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
@@
 This equivalence is only discovered in hindsight, as will be demonstrated in the following section.
-These two works also do not discuss other constrained cases that will also be addressed.
-
-In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based logic to the concept of accessibility.
-So, in the next section, we demonstrate how applying Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality--tying outputs to tangible system knowns.
+These two works also do not discuss other constrained cases that will also be addressed in this paper.
+
+In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based intuition to the concept of accessibility.
+So, in the next section, we do so by demonstrating how the application of Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality -- tying outputs to tangible system knowns.
 This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
@@
 
-Despite the close conceptual ties between accessibility and spatial interaction modeling, the former has not meaningfully absorbed the constraint-based logic of the latter.
-This section introduces this reframing: defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility ($\grave{a}$ la Hansen [@hansen1959]) from a relationship describing proportionality to equality.
-This shift addresses the fundamental issue of interpretability associated with the proportional nature of Hansen-type accessibility indicators we previous outlined.
-
-To do so, we propose a revised definition of accessibility considerate of the constraint-based spatial interaction model: the potential for spatial interaction with opportunities (or population).
+Despite their close conceptual ties, accessibility has not meaningfully absorbed the constraint-based logic of spatial interaction modeling.
+This section introduces this conceptual connection by defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility ($\grave{a}$ la Hansen [@hansen1959]) from a relationship describing proportionality to one of equality.
+This shift addresses the issue of unit interpretability (associated with the proportional nature of Hansen-type accessibility indicators previous outlined).
+
+We propose a revised definition of accessibility considerate of the constraint-based spatial interaction model: _the potential for spatial interaction with opportunities (or population)_.
 We can specify $k$ as a type of proportional allocation factor $\kappa$ which incorporates Wilson's balancing factor(s) to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potential for spatial interaction with population_ $M_{ji}$.
 In effect, $\kappa$ is unitless and proportionally allocates (based on the constraints of the case) opportunities (for $V_{ij}$) and population (for $M_{ji}$).
-The equation is as follows:
+The equations are generally expressed as follows:
 
@@
 
-\noindent Where $W_X^{(2)}$ and $W_X^{(1)}$ reflect the mass of the destination (i.e., opportunities) and mass of the origin (i.e., population) at either the zone or for the full region, depending on the case (hence represented by a $X$).
-
-Accessibility flows can also be summarised as as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively, as often done in accessibility research:
+\noindent Where $W_X^{(2)}$ is the mass of the destination (i.e., opportunities $O_j$ or $O$) and $W_X^{(1)}$ is the mass of the origin (i.e., population $D_i$ or $D$) for either the zone or full region depending on the case (hence represented by with a stand-in $X$ subindex).
+Accessibility flows can also be summarised as as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively.
+This form is common in accessibility research:
 
@@
 
-Figure @fig-analytical-device-conc-accessibility illustrates our analytical framework using a simple 3-zone system.
+Figure @fig-analytical-device-conc-accessibility illustrates our analytical framework using a 3-zone system.
 The most detailed values, $X_{ij}$, represent the potential for spatial interaction from origin $i$ to destination $j$.
 Here, $X$ stands for all cases and variants to be discussed (e.g., $V_{ij}^0$, $M_{ji}^0$, $V_{ij}^T$, $M_{ji}^T$, $V_{ij}^S$, $M_{ji}^S$, $V_{ij}^D$, and $M_{ji}^D$).
-Single marginals show origin and destination weights, while the total marginal sums these values.
+Single marginals show the origin and destination weights and the total marginal is a sum of these values.
 
@@
 3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
-- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
-Reflects how the literature calculates competitive accessibility.
+- Applies singly constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
+Reflects how te literature calculates competitive accessibility.
 - Units of $V_i^S$: accessible opportunities from $i$, a value that is the sum of opportunity supply flows allocated proportionally based on demand at $i$.
@@
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
-Represents ‘interaction capacity' or ‘realized access' serving as predictions of real interaction flows.
+Represents 'interaction capacity' or 'realized access' serving as predictions of real interaction flows.
 
@@
 
-However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are ‘opportunities weighted by travel impedance' and lack meaningful scaling or direct interpretability.
+However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are 'opportunities weighted by travel impedance' and lack meaningful scaling or direct interpretability.
 Comparing values across different decay functions or contexts (i.e., different number of zones) is therefore limited to ordinal statements (more vs.
@@
 The total constrained accessibility case can be interpreted in a few ways.
-In the one that connects to the status quo: the total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ so their total sum of $V^0_i$ matches a known system total--either total opportunities or total population.
+In the one that connects to the status quo: the total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ so their total sum of $V^0_i$ matches a known system total -- either total opportunities or total population.
 Another interpretation is in reformulating the equation to use a a proportional allocation constant based on the total balancing factor.
@@
 However, what's notable is how zones change between scenarios.
-For instance, Zone 1's share only declines slightly--these declines are outpaced by Zone 2's relative gains.
+For instance, Zone 1's share only declines slightly -- these declines are outpaced by Zone 2's relative gains.
 This shift reflects how $\kappa_{ij}^T$ redistributes opportunities in proportion to impedance under different travel behaviours.
@@
 
-This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield a measure of market potential--the number of people who can spatially interact with a destination.
+This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield a measure of market potential -- the number of people who can spatially interact with a destination.
 
@@
 Similar to the total constrained case, the singly constrained measures adjust unconstrained zonal accessibility values ($V_i^0$ or $M_j^0$) using a balancing factor to satisfy the known system constraint.
-However, unlike the total constraint (which enforces a global/total sum), the singly constrained case applies a localized constraint at one end of the interaction--either origin or destination.
-
-In the opportunities-constrained variant, the balancing factor ensures that only a proportion of opportunities at each destination are allocated to origins, based on their relative demand (population) and travel impedance.
+However, unlike the total constraint (which enforces a regional sum), the singly constrained case applies a localized constraint at one end of the interaction -- either origin or destination.
+
+In the opportunities-constrained variant $V_i^S$, the balancing factor ensures that only a proportion of opportunities at each destination are allocated to origins, based on their relative demand (population) and travel impedance.
 This variant mirrors the concept of spatial availability as discussed in Soukhov et al.
 [@soukhovIntroducingSpatialAvailability2023].
-In the population-constrained variant, the logic is reversed: population at each origin is allocated proportionally across destinations, informed by the distribution of opportunities and impedance.
+In the population-constrained variant $M_j^S$, the logic is reversed: population at each origin is allocated proportionally across destinations, informed by the distribution of opportunities and impedance.
 
@@
 Each zonal accessibility value becomes not only a fraction of the regional total (opportunities or population), but also a balanced sum of interactions, weighted by impedance and relative competition.
-The result remains in interpretable units--accessible opportunities or accessible population--but reflects more complex spatial dynamics.
+The result remains in interpretable units -- accessible opportunities or accessible population -- but reflect these more complex spatial dynamics.
 
@@
 Consider the highest decay scenario $f_1(c_{ij})$.
-Under this scenario, Zone 1--despite hosting a medium amount of physicians--captures the fewest physicians (`r opp_acc$V_opp_i_1[1]  |> round(digits = 3)`), compared to `r opp_acc$V_opp_i_1[2]  |> round(digits = 3)` at Zone 2, and `r opp_acc$V_opp_i_1[3]  |> round(digits = 3)` at Zone 3.
+Under this scenario, Zone 1 -- despite hosting a medium amount of physicians -- captures the fewest physicians (`r opp_acc$V_opp_i_1[1]  |> round(digits = 3)`), compared to `r opp_acc$V_opp_i_1[2]  |> round(digits = 3)` at Zone 2, and `r opp_acc$V_opp_i_1[3]  |> round(digits = 3)` at Zone 3.
 Why?
@@
 
-As travel decay decreases (e.g., $f_3(c_{ij})$), competition becomes more diffuse.
-Zone 2, with the largest population, initially dominates accessibility--but under low decay, Zones 1 and 3 also begin drawing more opportunities from Zone 2.
-For instance, Zone 1 gains  `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` more from Zone 2 in $f_3(c_{ij})$ than in $f_1(c_{ij})$.
-This shift reflects a drop in $\kappa^S_{2,2}$ by `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`,  showing Zone 2's decreasing hold on its own opportunities as other zones gain accessibility 'parity'.
+As travel decay increases (e.g., from $f_3(c_{ij})$ to $f_1(c_{ij})$), competition becomes less diffuse.
+Zone 2, with the largest population, claims the most opportunities under $f_3(c_{ij})$.
+But under lower decay, Zones 1 and 3 draw relatively more opportunities from Zone 2.
+For instance, Zone 1 gains `r ((opp_acc_ij$k_opp_ij_3[4])-(opp_acc_ij$k_opp_ij_1[4]) )|> scales::percent()` more from Zone 2 in $f_1(c_{ij})$ than in $f_3(c_{ij})$.
+This shift reflects a drop in $\kappa^S_{2,2}$ of `r -((opp_acc_ij$k_opp_ij_3[5])-(opp_acc_ij$k_opp_ij_1[5])) |> scales::percent()`, reflecting Zone 2's decreasing hold on its own opportunities as other zones gain accessibility 'parity'.
 
 This dynamic reveals how $\kappa^S_{ij}$ embeds both travel impedance and population competition.
-Unlike the total constraint lowest decay scenarios that allocate evenly, the singly constrained case reflects how competition evolves continues to influence allocation.
+Unlike the total constraint lowest decay scenarios that allocate evenly (as localised supply and demand is assumed to be unknown), the singly constrained case reflects competition's influence on allocation.
 
@@
 As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
-Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the lowest distance decay $f_3(c_{ij})$, per capita values are all ~24 physicians accessible per capita).
-
-This convergence mirrors the trend in the total constrained opportunity case, where accessibility values approach a third of the 490 physicians under the unfettered mobility scenario $f_3(c_{ij})$.
+Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the highest distance decay $f_1(c_{ij})$, per capita values are all ~24 physicians accessible per capita).
+
+This convergence mirrors the trend in the total constrained opportunity case, where accessibility values approach a third of the 490 physicians under the lowest-decay unfettered mobility scenario $f_3(c_{ij})$.
 In both cases, the balancing factors ($K^S$ and $B_j$) act as averaging mechanisms but at different scales.
@@
 Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
-In it's per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential has been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
+In its per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential has been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
 
@@
 
-Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression of accessible population $M_j^S$ being the product of proportionally allocated ($\hat \kappa^S_{ji}$) popultion:
+Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression of accessible population $M_j^S$ being the product of proportionally allocated ($\hat \kappa^S_{ji}$) population:
 
@@
 
-For the sake of brevity, we'll move forward on to the doubly constrained case.
+For the sake of brevity, we'll move onto the doubly constrained case.
 
@@
 
-Because results are made to match both margins, the results cannot be interpreted as a traditional summary at $i$ or $j$ (e.g., "opportunities accessible from $i$")--those sums simply reproduce the constraint totals.
+Because results are made to match values in both marginals, the results cannot be interpreted as a traditional summary at $i$ or $j$ (e.g., "opportunities accessible from $i$"), as evidently those sums simply reproduce themselves.
 Instead, the meaningful unit of analysis is the $ij$ flow itself.
@@
 This distinguishes doubly constrained accessibility from the total and singly constrained cases discussed previously.
-In those cases, only one side of the interaction--either the total marginal or opportunity/population marginal--was constrained, while the other was treated as a demand/supply weight (e.g., $D$ or $O$ for total constrained and  $W_j^{(2)}$ or $W_i^{(1)}$ for singly constrained).
+In those cases, only one side of the interaction -- either the total marginal or opportunity/population marginal -- was constrained, while the other was treated as a demand/supply weight (e.g., $D$ or $O$ for total constrained and  $W_j^{(2)}$ or $W_i^{(1)}$ for singly constrained).
 
 By contrast, the doubly constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly.
-This makes it less suitable for traditional accessibility analysis, namely because origin and destination masses often differ in kind and units.
-For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not known.
-A doubly constrained approach only makes conceptual sense when population and opportunities are comparable, have a one-to-one correspondence or are paired somehow--e.g., job per worker, student per school-seat, or vaccine doses per person.
-
-A doubly constrained approach to accessibility adopts the structure of the production-attraction spatial interaction model, where both population and opportunity totals are fixed.
+This makes it less suitable for traditional accessibility analysis; namely because origin and destination masses often differ in kind and units.
+For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not.
+A doubly constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school-seat, or vaccine doses per person).
+
+As mentioned, doubly constrained approach to accessibility adopts the structure of the production-attraction spatial interaction model, where both population and opportunity totals are fixed.
 Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):
@@
 
-Following this logic, the market potential form $M^D_{ji}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: i.e., the opportunities accessed from $j$ at an $i$ vs.
+Following this logic, the market potential form $M^D_{ji}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: the opportunities accessed from $j$ at an $i$ vs.
 the population accessed from $i$ at a $j$.
@@
 ```{r}
-#| label: simple-numerical-example-doubly constrained
+#| label: simple-numerical-example-doubly-constrained
 
@@
 
-Moving on to the toy example, to calculate doubly constrained accessibility, the interpretation of the population data and the counts of the opportunity data in the numeric example must be reinterpreted.
+To calculate doubly constrained accessibility using the toy example, the interpretation of the population data and the counts of the opportunity data must be reworked.
 Namely, a count of physician _capacity_ per destination $D_j$ is needed instead of simply the number of physicians.
-Also, we must be able to clearly state that the population is the _capacity_ of the origin to interact with opportunities $O_i$, i.e., the count of people seeking opportunities.
-
-This adjusted simple example is summarised in @tbl-small-system-land-use-doubly-constrained-case.
+We need to also be able to clearly state that the population is the _capacity_ of the origin to interact with opportunities $O_i$, i.e., the count of people seeking opportunities.
+
+This adjustment to the example is summarised in @tbl-small-system-land-use-doubly-constrained-case.
 With the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,000s of physician-capacity) per zone.
@@
 The number of providers per destination is however revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).
-The system-wide PPR is now `r sum(LU2$D_j2) / sum(LU2$O_i2)`, this is compared to the unmodified example which yields system PPR of `r sum(LU$D_j) / sum(LU$O_i)`.
+The system-wide PPR is now `r sum(LU2$D_j2) / sum(LU2$O_i2)` (recall: the unmodified example's system PPR is `r sum(LU$D_j) / sum(LU$O_i)`).
 
@@
 ```{r}
-#| label: simple-numerical-example-OD-table-doubly constrained
+#| label: simple-numerical-example-OD-table-doubly-constrained
 
@@
 
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints--i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints -- i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
 Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
@@
 
-To illustrate this, consider the results for Zone 2 (Suburban zone--a high-population, low-opportunity, relatively remote zone).
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal allocation (i.e., $V^D_{22}$) declines as travel impedance decay decreases--from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
-
-Following the intuition discussed in the singly constrained opportunity case, as decay decreases (i.e., more relatively uniform for all zones), the mass effects (effect of the population and opportunities magnitudes) become more relatively dominant in the spatial allocation.
+To illustrate this, consider the results for Zone 2 (recall: a suburban zone -- a high-population, low-opportunity, relatively remote).
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal flow (i.e., $V^D_{22}$) declines as travel impedance decay decreases -- from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
+
+Following the intuition discussed in the singly constrained opportunity case, as decay decreases, the mass effects (effect of the population and opportunities magnitudes) become more relatively dominant in the spatial allocation.
 
@@
 
-Recall, accessibility is traditionally presented as a summary zonal measure.
+Recall, accessibility is conventionally presented as a zonal value not a flow.
 However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$.
@@
 However, the focus of these models has typically been on $ij$ flows, often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
-
-Accessibility is often understood as a zonal summary of potential spatial interaction, often involving origin and destination masses in different units.
-These units also often misaligned--for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
-When they do align--such as people to physician capacity--we're essentially modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
-In such cases, the traditional accessibility question is already answered by the known information (i.e., how many opportunities can be reached by a zone?
-the number of people at that zone).
-This is why we do not foresee the doubly constrained measure being widely used in accessibility analysis: the literature has largely focused on questions of potential access, not on predicting flows of realised access.
+With the family of accessibility measures, it is made clear that we are working in units of opportunities and population.
+
+And as mentioned, these units are often misaligned -- for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
+When they do align -- such as people to physician capacity -- we are modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
+In such cases, the traditional accessibility question can be seen to be already answered by the known information (i.e., how many opportunities can be reached by a zone?
+Answer: the number of people at that zone).
+For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' access, not on predicting flows of realised access.
 
@@
 
-We then introduced the singly-constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
+We then introduced the singly constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
 It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.

### 2025-10-22 paper_plos/paper_plos_V3.tex
V3 AS revision; small gramar changes here and there.

--- 2025-10-15 paper_plos/paper_plos_V2.tex
@@
 the `singly constrained' measure (i.e., related to the popular two step
-floating catchment approach - 2SFCA), and the `doubly constrained'
+floating catchment approach -- 2SFCA), and the `doubly constrained'
 measure representing realized access (i.e., equal to the doubly
@@
 accessible population (i.e., market potential). A toy example
-illustrates how they produce interpretable, unit-based values, offering
-a clearer and more coherent basis for accessibility analysis.
+illustrates how they produce interpretable unit-based values, offering a
+clearer and more coherent basis for accessibility analysis.
 
@@
 Section 7 constitutes the paper's core contribution: we derive a family
-of accessibility measures corresponding to different constraints
-applied--unconstrained, singly-constrained, and doubly-constrained
-cases--demonstrating how each yields zone-level accessibility values
-expressed in meaningful units (opportunities or population) rather than
+of accessibility measures corresponding to different constraints applied
+-- unconstrained, singly constrained, and doubly constrained cases --
+demonstrating how each yields zone-level accessibility values expressed
+in meaningful units (opportunities or population) rather than
 unit-inconsistent indices. We illustrate each measure with numerical
@@
 
-Where \(G\) is the gravitational constant--an empirically calibrated
+Where \(G\) is the gravitational constant -- an empirically calibrated
 value that ensures the model reflects observed forces. Newton's initial
@@
 \citep{careyPrinciplesSocialScience1858} inspired empirical spatial
-intereaction research in different contexts. Namely, a number of
+interaction research in different contexts. Namely, a number of
 researchers theoretically and empirically attempted to characterise
@@
 ``masses'' \(M_i\) and \(M_j\) of two locations, and inversely
-proportional to their separation distance--conceptually parallel to
+proportional to their separation distance -- conceptually parallel to
 Newtonian gravity, but often omitting a proportionality constant.
@@
 Although Stewart's concept of ``social physics'' eventually fell out of
-favour--due to inconsistent mathematical notation (e.g., \(G\) is used
-as both a proportionality constant p.~34 and then later as
-\emph{demographic energy} on p.~53.) and its racist and unscientific
-assumptions (e.g., view of humans as particles following physical laws
-and assumptions of the molecular weight of the average American being
-one, but ``presumably\ldots much less than one\ldots.for an Australian
-aborigine'' {[}p.~35{]})--his introduction of a proportionality constant
-\(G\) in modeling demographic force marks an important conceptual step:
-recognizing that moving from proportionality to equality requires
-empirical calibration. In other words, the addition of \(G\) shifts
-results from being abstract indicators of potential (i.e.,
-\(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having interpretable
-units grounded in consistent, albeit still abstract, quantities (i.e.,
-units of demographic force).
+favour, potentially in part due to inconsistent mathematical notation
+(e.g., \(G\) is used as both a proportionality constant p.~34 and then
+later as \emph{demographic energy} on p.~53.) as well as its racist and
+unscientific assumptions (e.g., view of humans as particles following
+physical laws and assumptions of the molecular weight of the average
+American being one, but ``presumably\ldots much less than one\ldots.for
+an Australian aborigine'' {[}p.~35{]}). However, Stewart's introduction
+of a proportionality constant \(G\) in modeling demographic force marks
+an important conceptual step: recognizing that moving from
+proportionality to equality requires empirical calibration. In other
+words, the addition of \(G\) shifts results from being abstract
+indicators of potential (i.e.,
+\(\frac{\text{people}^2}{\text{distance}^{2}}\)) to having units
+grounded in consistent, albeit still abstract, quantities (i.e., units
+of some sort of demographic force).
 
@@
 \citep{hansen1959}. In this seminal paper, Hansen defined accessibility
-as ``the potential of opportunities for interaction\ldots{} a
+as the potential of opportunities for interaction\ldots{} a
 generalization of the population-over-distance relationship or
@@
 \(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
-accessibility from site \(i\). This is a function of \(O_j\) (the mass
+accessibility from a zone \(i\). This is a function of \(O_j\) (the mass
 of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
@@
 a non-Newtonian analogy. Additionally, this work was not focused on the
-`potential' concept, as accessibility has been. Wilson
-\citep{wilson1971} formalized the general spatial interaction model
-through the following equation:
+`potential' concept as accessibility had been. Wilson \citep{wilson1971}
+formalized the general spatial interaction model through the following
+equation:
 
@@
 incorporating known system totals. In some sense, \(k\) serves a role
-similar to the gravitational constant in Newton's law--it calibrates the
-model so that outputs match real-world quantities. However, these
+similar to the gravitational constant in Newton's law -- it calibrates
+the model so that outputs match real-world quantities. However, these
 real-world quantities are not a set empirical constant (like Newton's
@@
 gravity (i.e., interaction between masses over distance), Wilson's
-innovation was to ground and generalise the model using \emph{entropy
+innovation was to generalise the model using \emph{entropy
 maximisation}. By maximizing the number of ways individual trip
@@
 singly constrained model (either by origins \(A_i\) or destinations
-\(B_j\) - see below), and (3) a doubly constrained model that satisfies
-both.
+\(B_j\) -- explained below), and (3) a doubly constrained model that
+satisfies both.
 
@@
 
-In the singly-constrained model, only constraint
+In the singly constrained model, only constraint
 Eq~\ref{eq-constraint1-gravitymodel} or constraint
@@
 Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
-to the production-constrained singly-constrained version of
+to the production-constrained singly constrained version of
 Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
@@
 
-The singly-constrained attraction-constrained model is similar to the
+The singly constrained attraction-constrained model is similar to the
 production-constrained version but from the perspective of the mass at
@@
 
-Lastly, the doubly-constrained model is the production-attraction
+Lastly, the doubly constrained model is the production-attraction
 constrained model in Wilson \citep{wilson1971}. In this case, both
@@
 constraints using principles of entropy maximisation. This results in
-interpretable, balanced (given the constraints), and unit-consistent
-\(ij\) flows. The balancing factors themselves have been subject to
-various interpretations--as terminal costs
-\citep{dieter1962distribution}, weighted mean values
+interpretable, balanced (given the system knowns are expressed in the
+constraints), and unit-consistent \(ij\) flows. The balancing factors
+themselves have been subject to various interpretations -- as terminal
+costs \citep{dieter1962distribution}, weighted mean values
 \citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
 themselves \citep{cesario1977new} (as suggested in \citet{Wilson1971})
-and as rents \citep{morphetGravityModelCalibration2023}--reflecting
+and as rents \citep{morphetGravityModelCalibration2023} -- reflecting
 ongoing efforts to understand what these mathematical constructs
@@
 investigation sheds light on why accessibility research may have failed
-to adopt a comparable approach focused, until this paper.
+to adopt a comparable approach, until this paper.
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V2_files/figure-pdf/fig-docs-per-year-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V3_files/figure-pdf/fig-docs-per-year-1.pdf}
 
@@
 transportation demand forecasting. We argue the framework's quiet
-innovation--introducing empirically grounded constraints to shift from
-proportionality to calibrated equality--made the framework immediately
+innovation -- introducing empirically grounded constraints to shift from
+proportionality to calibrated equality -- made the framework immediately
 relevant for policy applications as outputs were in tangible units. Yet,
@@
 
-To explore why Wilson's approach never crossed over to accessibility
-modeling, we conduct a review of the literature citing Hansen
-\citep{hansen1959}, Wilson \citep{wilson1971}, or both (on Web of
-Science using the ``Cited References'' functionality, and the digital
-object identifiers of Hansen \citep{hansen1959} and Wilson
-\citep{wilson1971}) . Only 76 out of the 2,122 documents that emerged
-from our search cite both. The number of articles, by year and if they
-cite Hasen, Wilson, or both are shown in Fig~\ref{fig-docs-per-year}.
-
-Through the close analysis of \emph{how} articles that cite both works,
-we identify two distinct citation patterns: one group of articles
+To explore why Wilson's approach may not have crossed over to
+accessibility modeling sooner, we conducted a review of the literature
+citing Hansen \citep{hansen1959}, Wilson \citep{wilson1971}, or both.
+This was done using Web of Science's ``Cited References'' functionality,
+and the digital object identifiers of Hansen \citep{hansen1959} and
+Wilson \citep{wilson1971}). Only 76 out of the 2,122 documents that
+emerged from our search of the General Database cite both. The number of
+articles, by year and if they cite Hasen, Wilson, or both are shown in
+Fig~\ref{fig-docs-per-year}.
+
+Through the close analysis of why articles that cite both works cite
+each work, we identify two distinct patterns: one group of articles
 focused on accessibility, the other on spatial interaction. In examining
-these articles, we uncover how the relationship between the two has
-often been misunderstood, underexplored, or entirely overlooked.
-
-In the first stream of literature--which cite both but are focused more
-so on spatial interaction models--they treat spatial interaction and
-accessibility as separate but related phenomenons. Four subsets of this
-stream emerge.
+these groups of articles, we uncover how the relationship between Wilson
+and Hansen has often been misunderstood, underexplored, or entirely
+overlooked.
+
+In the first stream of literature -- which cite both but are focused
+more so on spatial interaction models -- they treat spatial interaction
+and accessibility as separate but related phenomenons. Four subsets of
+this stream emerge.
 
@@
 
-The third subset of the spatial-interaction focused literature, depart
+The third subset of the spatial-interaction focused literature depart
 from Hansen's \citep{hansen1959} definition, aligning instead with
@@
 \citep{morrisAccessibilityIndicatorsTransport1979, leonardiRandomUtilityDemand1984}.
-In sum though, across these works, they recognize Hansen-type
+Though across works in this subset, they recognize Hansen-type
 accessibility as an indicator of `potential' but as a separate but
@@
 both works, we categorise their citation of Wilson \citep{wilson1971}
-within three general groups. Overall though, these works do not engage,
-or only superficially engage with Wilson \citep{wilson1971}.
+within three general groups. Overall these works do not engage -- or
+only superficially engage -- with Wilson \citep{wilson1971}.
 
@@
 context-dependent travel cost functions. This trend is common: for
-instance it is done in
+instance it is done in the following papers:
 \citep{handyMeasuringAccessibilityExploration1997, weibullNumericalMeasurementAccessibility1980, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
@@
 
-Secondly, a subset of literature explicitly associates spatial
-interaction, as defined in Wilson \citep{wilson1971}, with
-accessibility's potential for spatial interaction--but only
-superficially. These works acknowledge the conceptual link, but do not
-go beyond this recognition in the scope of their works e.g.,
+Secondly, a subset of literature explicitly acknowledge the conceptual
+link between spatial interaction and with Hansen-type accessibility --
+but only superficially, not going beyond the recognition that they both
+relate to spatial interaction e.g.,
 \citep{grengs2010job, millerMeasuringSpaceTimeAccessibility1999, giuliano2010accessibility, grengs2010intermetropolitan, grengs2012equity, levine2012does, levinson2012positive, tong2015transportation, liu2015spatial, he2017measuring, wuUnifyingAccess2020, ng2022reflection, naqavi2023mobility, suel2024measuring}.
 Indeed, while accessibility can be seen as the \emph{potential} for
-spatial interaction--and Wilson \citep{wilson1971} briefly touches on
-this--such mentions have not resulted in deeper analytical integration
+spatial interaction -- and Wilson \citep{wilson1971} briefly touches on
+this -- such mentions have not resulted in deeper analytical integration
 of these concepts. Furthermore, some of this literature also
@@
 demonstrated in the following section. These two works also do not
-discuss other constrained cases that will also be addressed.
+discuss other constrained cases that will also be addressed in this
+paper.
 
@@
 accessibility literature citing Wilson \citep{wilson1971} has
-meaningfully applied Wilson's constraint-based logic to the concept of
-accessibility. So, in the next section, we demonstrate how applying
-Wilson's framework enables accessibility to move from a proportional
-relationship to calibrated equality--tying outputs to tangible system
-knowns. This re-expression of accessibility using constraints yields
-interpretable, unit-consistent measures of opportunity. This approach
-takes the same path of entropy-maximisation as in Wilson
-\citep{wilson1971}, and does not rely on specifying some universal
-constant \(G\) like initially suggested in Stewart
+meaningfully applied Wilson's constraint-based intuition to the concept
+of accessibility. So, in the next section, we do so by demonstrating how
+the application of Wilson's framework enables accessibility to move from
+a proportional relationship to calibrated equality -- tying outputs to
+tangible system knowns. This re-expression of accessibility using
+constraints yields interpretable, unit-consistent measures of
+opportunity. This approach takes the same path of entropy-maximisation
+as in Wilson \citep{wilson1971}, and does not rely on specifying some
+universal constant \(G\) like initially suggested in Stewart
 \citep{stewartDemographicGravitationEvidence1948} (recall,
@@
 
-Despite the close conceptual ties between accessibility and spatial
-interaction modeling, the former has not meaningfully absorbed the
-constraint-based logic of the latter. This section introduces this
-reframing: defining a family of accessibility measures using Wilson's
-approach grounded in statistical mechanics, shifting place-based
-accessibility (\(\grave{a}\) la Hansen \citep{hansen1959}) from a
-relationship describing proportionality to equality. This shift
-addresses the fundamental issue of interpretability associated with the
-proportional nature of Hansen-type accessibility indicators we previous
-outlined.
-
-To do so, we propose a revised definition of accessibility considerate
-of the constraint-based spatial interaction model: the potential for
-spatial interaction with opportunities (or population). We can specify
+Despite their close conceptual ties, accessibility has not meaningfully
+absorbed the constraint-based logic of spatial interaction modeling.
+This section introduces this conceptual connection by defining a family
+of accessibility measures using Wilson's approach grounded in
+statistical mechanics, shifting place-based accessibility (\(\grave{a}\)
+la Hansen \citep{hansen1959}) from a relationship describing
+proportionality to one of equality. This shift addresses the issue of
+unit interpretability (associated with the proportional nature of
+Hansen-type accessibility indicators previous outlined).
+
+We propose a revised definition of accessibility considerate of the
+constraint-based spatial interaction model: \emph{the potential for
+spatial interaction with opportunities (or population)}. We can specify
 \(k\) as a type of proportional allocation factor \(\kappa\) which
@@
 the constraints of the case) opportunities (for \(V_{ij}\)) and
-population (for \(M_{ji}\)). The equation is as follows:
+population (for \(M_{ji}\)). The equations are generally expressed as
+follows:
 
@@
 
-\noindent Where \(W_X^{(2)}\) and \(W_X^{(1)}\) reflect the mass of the
-destination (i.e., opportunities) and mass of the origin (i.e.,
-population) at either the zone or for the full region, depending on the
-case (hence represented by a \(X\)).
-
-Accessibility flows can also be summarised as as a partial sum of the
-potential at \(i\) and at \(j\) to express accessibility at the origin
-zone or at the destination zone, respectively, as often done in
-accessibility research:
+\noindent Where \(W_X^{(2)}\) is the mass of the destination (i.e.,
+opportunities \(O_j\) or \(O\)) and \(W_X^{(1)}\) is the mass of the
+origin (i.e., population \(D_i\) or \(D\)) for either the zone or full
+region depending on the case (hence represented by with a stand-in \(X\)
+subindex). Accessibility flows can also be summarised as as a partial
+sum of the potential at \(i\) and at \(j\) to express accessibility at
+the origin zone or at the destination zone, respectively. This form is
+common in accessibility research:
 
@@
 Figure Fig~\ref{fig-analytical-device-conc-accessibility} illustrates
-our analytical framework using a simple 3-zone system. The most detailed
+our analytical framework using a 3-zone system. The most detailed
 values, \(X_{ij}\), represent the potential for spatial interaction from
@@
 \(V_{ij}^T\), \(M_{ji}^T\), \(V_{ij}^S\), \(M_{ji}^S\), \(V_{ij}^D\),
-and \(M_{ji}^D\)). Single marginals show origin and destination weights,
-while the total marginal sums these values.
+and \(M_{ji}^D\)). Single marginals show the origin and destination
+weights and the total marginal is a sum of these values.
 
@@
 \item
-  Applies singly-constrained proportional allocation factors
+  Applies singly constrained proportional allocation factors
   (\(\kappa_{ij}^S\), \(\hat \kappa_{ji}^S\)) based on Wilson's
@@
   Fig~\ref{fig-analytical-device-conc-accessibility}) i.e., the number
-  of opportunities or population at each zone. Reflects how the
+  of opportunities or population at each zone. Reflects how te
   literature calculates competitive accessibility.
@@
 factor proportionally adjusts unconstrained zonal accessibility values
-\(V^0_i\) so their total sum of \(V^0_i\) matches a known system
-total--either total opportunities or total population. Another
+\(V^0_i\) so their total sum of \(V^0_i\) matches a known system total
+-- either total opportunities or total population. Another
 interpretation is in reformulating the equation to use a a proportional
@@
 However, what's notable is how zones change between scenarios. For
-instance, Zone 1's share only declines slightly--these declines are
+instance, Zone 1's share only declines slightly -- these declines are
 outpaced by Zone 2's relative gains. This shift reflects how
@@
 opportunity-constrained formulation, switching indices \(i\) and \(j\)
-to yield a measure of market potential--the number of people who can
+to yield a measure of market potential -- the number of people who can
 spatially interact with a destination.
@@
 using a balancing factor to satisfy the known system constraint.
-However, unlike the total constraint (which enforces a global/total
-sum), the singly constrained case applies a localized constraint at one
-end of the interaction--either origin or destination.
-
-In the opportunities-constrained variant, the balancing factor ensures
-that only a proportion of opportunities at each destination are
+However, unlike the total constraint (which enforces a regional sum),
+the singly constrained case applies a localized constraint at one end of
+the interaction -- either origin or destination.
+
+In the opportunities-constrained variant \(V_i^S\), the balancing factor
+ensures that only a proportion of opportunities at each destination are
 allocated to origins, based on their relative demand (population) and
@@
 \citep{soukhovIntroducingSpatialAvailability2023}. In the
-population-constrained variant, the logic is reversed: population at
-each origin is allocated proportionally across destinations, informed by
-the distribution of opportunities and impedance.
+population-constrained variant \(M_j^S\), the logic is reversed:
+population at each origin is allocated proportionally across
+destinations, informed by the distribution of opportunities and
+impedance.
 
@@
 sum of interactions, weighted by impedance and relative competition. The
-result remains in interpretable units--accessible opportunities or
-accessible population--but reflects more complex spatial dynamics.
+result remains in interpretable units -- accessible opportunities or
+accessible population -- but reflect these more complex spatial
+dynamics.
 
@@
 This consideration has important implications. Consider the highest
-decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1--despite
-hosting a medium amount of physicians--captures the fewest physicians
+decay scenario \(f_1(c_{ij})\). Under this scenario, Zone 1 -- despite
+hosting a medium amount of physicians -- captures the fewest physicians
 (133.469), compared to 166.781 at Zone 2, and 189.75 at Zone 3. Why?
@@
 
-As travel decay decreases (e.g., \(f_3(c_{ij})\)), competition becomes
-more diffuse. Zone 2, with the largest population, initially dominates
-accessibility--but under low decay, Zones 1 and 3 also begin drawing
-more opportunities from Zone 2. For instance, Zone 1 gains 6\% more from
-Zone 2 in \(f_3(c_{ij})\) than in \(f_1(c_{ij})\). This shift reflects a
-drop in \(\kappa^S_{2,2}\) by 14\%, showing Zone 2's decreasing hold on
-its own opportunities as other zones gain accessibility `parity'.
+As travel decay increases (e.g., from \(f_3(c_{ij})\) to
+\(f_1(c_{ij})\)), competition becomes less diffuse. Zone 2, with the
+largest population, claims the most opportunities under \(f_3(c_{ij})\).
+But under lower decay, Zones 1 and 3 draw relatively more opportunities
+from Zone 2. For instance, Zone 1 gains 6\% more from Zone 2 in
+\(f_1(c_{ij})\) than in \(f_3(c_{ij})\). This shift reflects a drop in
+\(\kappa^S_{2,2}\) of 14\%, reflecting Zone 2's decreasing hold on its
+own opportunities as other zones gain accessibility `parity'.
 
@@
 and population competition. Unlike the total constraint lowest decay
-scenarios that allocate evenly, the singly constrained case reflects how
-competition evolves continues to influence allocation.
+scenarios that allocate evenly (as localised supply and demand is
+assumed to be unknown), the singly constrained case reflects
+competition's influence on allocation.
 
@@
 opportunities. Consequently, per capita accessibility values begin to
-stabilise to the regional per capita average (e.g., in the lowest
-distance decay \(f_3(c_{ij})\), per capita values are all
+stabilise to the regional per capita average (e.g., in the highest
+distance decay \(f_1(c_{ij})\), per capita values are all
 \textasciitilde24 physicians accessible per capita).
@@
 case, where accessibility values approach a third of the 490 physicians
-under the unfettered mobility scenario \(f_3(c_{ij})\). In both cases,
-the balancing factors (\(K^S\) and \(B_j\)) act as averaging mechanisms
-but at different scales. As distance decay becomes more
+under the lowest-decay unfettered mobility scenario \(f_3(c_{ij})\). In
+both cases, the balancing factors (\(K^S\) and \(B_j\)) act as averaging
+mechanisms but at different scales. As distance decay becomes more
 \emph{relatively} uniform, the role of remaining variables (i.e., total
@@
 the mass weight at the origin \(W_i^{(1)}\) is now represented by
-\(O_i\)). In it's per-capita expression, i.e., equivalent to 2SFCA, this
+\(O_i\)). In its per-capita expression, i.e., equivalent to 2SFCA, this
 constrained concept of market potential has been used to express
@@
 accessible population \(M_j^S\) being the product of proportionally
-allocated (\(\hat \kappa^S_{ji}\)) popultion:
+allocated (\(\hat \kappa^S_{ji}\)) population:
 
@@
 
-For the sake of brevity, we'll move forward on to the doubly constrained
-case.
+For the sake of brevity, we'll move onto the doubly constrained case.
 
@@
 
-Because results are made to match both margins, the results cannot be
-interpreted as a traditional summary at \(i\) or \(j\) (e.g.,
-``opportunities accessible from \(i\)'')--those sums simply reproduce
-the constraint totals. Instead, the meaningful unit of analysis is the
+Because results are made to match values in both marginals, the results
+cannot be interpreted as a traditional summary at \(i\) or \(j\) (e.g.,
+``opportunities accessible from \(i\)''), as evidently those sums simply
+reproduce themselves. Instead, the meaningful unit of analysis is the
 \(ij\) flow itself.
@@
 singly constrained cases discussed previously. In those cases, only one
-side of the interaction--either the total marginal or
-opportunity/population marginal--was constrained, while the other was
+side of the interaction -- either the total marginal or
+opportunity/population marginal -- was constrained, while the other was
 treated as a demand/supply weight (e.g., \(D\) or \(O\) for total
@@
 allocates flows accordingly. This makes it less suitable for traditional
-accessibility analysis, namely because origin and destination masses
+accessibility analysis; namely because origin and destination masses
 often differ in kind and units. For instance, the number of people
 accessing an opportunity such as park may be known, but the capacity of
-each park is not known. A doubly constrained approach only makes
-conceptual sense when population and opportunities are comparable, have
-a one-to-one correspondence or are paired somehow--e.g., job per worker,
-student per school-seat, or vaccine doses per person.
-
-A doubly constrained approach to accessibility adopts the structure of
-the production-attraction spatial interaction model, where both
-population and opportunity totals are fixed. Mathematically, this model
-requires the simultaneous imposition of both the population- and
+each park is not. A doubly constrained approach may make conceptual
+sense if the ``potential'' should be contained only in the flows
+themselves, meaning the units of population and opportunities are
+comparable, have a one-to-one correspondence or are otherwise paired
+(e.g., job per worker, student per school-seat, or vaccine doses per
+person).
+
+As mentioned, doubly constrained approach to accessibility adopts the
+structure of the production-attraction spatial interaction model, where
+both population and opportunity totals are fixed. Mathematically, this
+model requires the simultaneous imposition of both the population- and
 opportunity- constraints in the preceding singly constrained variants
@@
 effectively equivalent to \(V_{ij}^D\), but can be read with a different
-interpretation: i.e., the opportunities accessed from \(j\) at an \(i\)
+interpretation: the opportunities accessed from \(j\) at an \(i\)
 vs.~the population accessed from \(i\) at a \(j\). The inputs of
@@
 
-Moving on to the toy example, to calculate doubly constrained
-accessibility, the interpretation of the population data and the counts
-of the opportunity data in the numeric example must be reinterpreted.
-Namely, a count of physician \emph{capacity} per destination \(D_j\) is
-needed instead of simply the number of physicians. Also, we must be able
-to clearly state that the population is the \emph{capacity} of the
-origin to interact with opportunities \(O_i\), i.e., the count of people
-seeking opportunities.
-
-This adjusted simple example is summarised in
-Table~\ref{tbl-small-system-land-use-doubly-constrained-case}. With the
+To calculate doubly constrained accessibility using the toy example, the
+interpretation of the population data and the counts of the opportunity
+data must be reworked. Namely, a count of physician \emph{capacity} per
+destination \(D_j\) is needed instead of simply the number of
+physicians. We need to also be able to clearly state that the population
+is the \emph{capacity} of the origin to interact with opportunities
+\(O_i\), i.e., the count of people seeking opportunities.
+
+This adjustment to the example is summarised in
+\textbf{?@tbl-small-system-land-use-doubly-constrained-case}. With the
 population (in units of 10,000s of people seeking physicians) and the
@@
 physicians used in previous cases
-(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now 1,
-this is compared to the unmodified example which yields system PPR of
-24.5.
+(Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now 1
+(recall: the unmodified example's system PPR is 24.5).
 
@@
 
+::: \{\#tbl-small-system-land-use-doubly constrained-case .cell
+tbl-cap=`Modified simple system with three zones reflecting matched
+population and opportunities. Population is in 10,000 persons and
+opportunities in 10,000 of physician-capacity.'\} :::
+\{.cell-output-display\}
+
 \begin{table}
-
-\caption{\label{tbl-small-system-land-use-doubly-constrained-case}Modified
-simple system with three zones reflecting matched population and
-opportunities. Population is in 10,000 persons and opportunities in
-10,000 of physician-capacity.}
-
-\centering{
-
 \fontsize{7.5pt}{9.0pt}\selectfont
@@
 \end{tabular*}
-
-}
-
-\end{table}%
+\end{table}
+
+::: :::
 
@@
 longer meaningfully represented as zonal summaries like \(V^D_i\) or
-\(M^D_j\), since these values reproduce the original constraints--i.e.,
-\(V^D_i = O_i\) so the physician-capacity accessible for Zones 1, 2, and
-3 would be 4.002, 9.996, and 6.002. Hence, the usefulness of the doubly
-constrained measure lies in the interpretation as \(V_{ij}^D\) values.
-\(V^D_{ij}\) values represent the number of opportunities from zone
-\(j\) allocated to populations in zone \(i\), shaped by both mass and
-travel impedance.
-
-To illustrate this, consider the results for Zone 2 (Suburban zone--a
-high-population, low-opportunity, relatively remote zone). As shown in
+\(M^D_j\), since these values reproduce the original constraints --
+i.e., \(V^D_i = O_i\) so the physician-capacity accessible for Zones 1,
+2, and 3 would be 4.002, 9.996, and 6.002. Hence, the usefulness of the
+doubly constrained measure lies in the interpretation as \(V_{ij}^D\)
+values. \(V^D_{ij}\) values represent the number of opportunities from
+zone \(j\) allocated to populations in zone \(i\), shaped by both mass
+and travel impedance.
+
+To illustrate this, consider the results for Zone 2 (recall: a suburban
+zone -- a high-population, low-opportunity, relatively remote). As shown
+in
 Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2},
-its intrazonal allocation (i.e., \(V^D_{22}\)) declines as travel
-impedance decay decreases--from 4.959 under \(f_1(c_{ij})\) to 2.667
-under \(f_3(c_{ij})\), out of the \textasciitilde10 opportunities
-allocated to Zone 2 (a population of 10).
+its intrazonal flow (i.e., \(V^D_{22}\)) declines as travel impedance
+decay decreases -- from 4.959 under \(f_1(c_{ij})\) to 2.667 under
+\(f_3(c_{ij})\), out of the \textasciitilde10 opportunities allocated to
+Zone 2 (a population of 10).
 
 Following the intuition discussed in the singly constrained opportunity
-case, as decay decreases (i.e., more relatively uniform for all zones),
-the mass effects (effect of the population and opportunities magnitudes)
-become more relatively dominant in the spatial allocation.
+case, as decay decreases, the mass effects (effect of the population and
+opportunities magnitudes) become more relatively dominant in the spatial
+allocation.
 
@@
 
-Recall, accessibility is traditionally presented as a summary zonal
-measure. However, in the doubly constrained case, since we force the
+Recall, accessibility is conventionally presented as a zonal value not a
+flow. However, in the doubly constrained case, since we force the
 allocation of zonal population demand and zonal opportunities supplied
@@
 has typically been on \(ij\) flows, often calibrated using trips (i.e.,
-outbound and inbound trips, inherently in the same units).
-
-Accessibility is often understood as a zonal summary of potential
-spatial interaction, often involving origin and destination masses in
-different units. These units also often misaligned--for instance, we may
-not know how much park space, grocery area, or childcare capacity is
-accessible per person. When they do align--such as people to physician
-capacity--we're essentially modeling realised access flows based on
-known quantities of \emph{those that interact} and the
-\emph{interacted}. In such cases, the traditional accessibility question
-is already answered by the known information (i.e., how many
-opportunities can be reached by a zone? the number of people at that
-zone). This is why we do not foresee the doubly constrained measure
-being widely used in accessibility analysis: the literature has largely
-focused on questions of potential access, not on predicting flows of
+outbound and inbound trips, inherently in the same units). With the
+family of accessibility measures, it is made clear that we are working
+in units of opportunities and population.
+
+And as mentioned, these units are often misaligned -- for instance, we
+may not know how much park space, grocery area, or childcare capacity is
+accessible per person. When they do align -- such as people to physician
+capacity -- we are modeling realised access flows based on known
+quantities of \emph{those that interact} and the \emph{interacted}. In
+such cases, the traditional accessibility question can be seen to be
+already answered by the known information (i.e., how many opportunities
+can be reached by a zone? Answer: the number of people at that zone).
+For this reason, we do not foresee the doubly constrained measure being
+widely used in accessibility analysis as the literature has largely
+focused on questions of `potential' access, not on predicting flows of
 realised access.
@@
 
-We then introduced the singly-constrained case, which \emph{does} take
+We then introduced the singly constrained case, which \emph{does} take
 into account the population size at the origin in the allocation of

### 2025-10-22 paper_plos/summary-family-access-measures-table.tex
V3 AS revision; small gramar changes here and there.

--- 2025-05-26 paper/summary-family-access-measures-table.tex
@@
 \hline
-\textbf{Name of Member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
+\textbf{Name of Member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
 \hline
@@
 \hline
-\textbf{Name of member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Measure Equation} & \textbf{Interpretation} \\
+\textbf{Name of member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
 \hline
@@
 
-& Values in various units depending on the impedance and destination-mass (e.g., "opportunities x decay") for $V_i^0$ andimpedance and origin-mass (e.g., "population x decay"); no total or marginal constraint \\
+& Values in various units depending on the impedance and destination-mass (e.g., "opportunities x decay") for $V_i^0$ and impedance and origin-mass (e.g., "population x decay") for $M_j^0$; no total or marginal constraint \\
 \hline
@@
 
-& Values reflect a share of total regional opportunities ($V^T_i$) or population ($M^T_j$). \\
+& Values reflect a share of the total opportunities in the region $D$ for $V^T_i$ or total population in the region $O$ for $M^T_j$. \\
 \hline
@@
 
-& $V^S_i$ values reflect a share of the opportunities at each destination based on origin population 'demand' and impedance; $M^S_j$ values reflect a share of the population at each origin based on destination opportunities 'supply' and impedance.\\
+& $V^S_i$ values reflect a share of the opportunities at each destination $D_j$ based on origin population 'demand' and impedance; $M^S_j$ values reflect a share of the population at each origin $O_i$ based on destination opportunities 'supply' and impedance.\\
 \hline

### 2025-10-22 paper_plos/reviewer_respose_letter_V2.qmd

@@
 
-_Wilson's work is notable for many reasons.
+>_Wilson's work is notable for many reasons.
 Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation.
 This results in interpretable, balanced (given the constraints), and unit-consistent $ij$ flows.
-The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @wilson1971) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
+The balancing factors themselves have been subject to various interpretations--as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves ([@cesario1977new] as suggested in Wilson [@wilson1971]) and as rents [@morphetGravityModelCalibration2023]--reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
 In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
@@
 
-_Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
+>_Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
 This investigation sheds light on why accessibility research may have failed to adopt a comparable approach focused, until this paper._
@@
 
-While we hear you that the mathematical detail may be dense, we believe it is inherent to our contribution.
+While we acknowledge you that the mathematical detail may be dense, we believe it is inherent to our contribution.
 Our manuscript establishes formal relationships between spatial interaction models and accessibility measures--specifically focusing on the use of the balancing factors.
@@
 
-_The remainder of this paper proceeds as follows.
+>_The remainder of this paper proceeds as follows.
 Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4).
@@
 
-_The key insight is that accessibility need not be unit-inconsistent.
+>_The key insight is that accessibility need not be unit-inconsistent.
 By building from spatial interaction theory's constraint framework rather than the gravitational analogy alone, we show how to construct accessibility measures that are simultaneously grounded in behavioural principles (the gravitational analogy) as well as being expressed in interpretable units (opportunities or population) that are sensitive to the known to the region through empirically-set constraints._
 
-As well, we'd like to note that the manuscript already underwent extension revision that included the addition of intuitive explanations before mathematical developments, removal/consolidation of tangential equations throughout to improve readability, and transitional text at the end of each section.
+As well, we'd like to note that the manuscript already underwent extensive revision that included the addition of intuitive explanations before mathematical developments, removal/consolidation of tangential equations throughout to improve readability, and transitional text at the end of each section.
 

### 2025-10-22 paper_plos/paper_plos_V3.qmd
V3 final edits AS

@@
 
-# Introduction
+# 1. Introduction
 
@@
 
-In summary, this paper makes two contributions: (1) we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation.
-(2) we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
+This paper makes two contributions.
+First, we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation.
+Second, we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
 These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
@@
 The remainder of this paper proceeds as follows.
-Sections 2-4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and @careyPrinciplesSocialScience1858 (Section 2), moving through early researchers like @ravensteinLawsMigration1885 to @stewartDemographicGravitationEvidence1948 who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in @hansen1959 became the dominant approach in planning practice (Section 4).
-Section 5 presents the entropy-based family of spatial interaction models in @wilson1971, explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
+Sections 2 through 4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and Carey (1858) [@careyPrinciplesSocialScience1858] (Section 2), moving through early researchers like Ravenstein (1885) [@ravensteinLawsMigration1885] to Stewart (1948) [@stewartDemographicGravitationEvidence1948] who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in Hansen (1959) [@hansen1959] became the dominant approach in planning practice (Section 4).
+Section 5 presents the entropy-based family of spatial interaction models in Wilson (1971) [@wilson1971], explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
 Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations.
@@
 
-# Newtonian's roots of human spatial interaction research
+# 2. Newtonian's roots of human spatial interaction research
 
@@
 
-# Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)
+# 3. Early research on human spatial interaction: from Ravenstein (1885) to Stewart (1948)
 
@@
 
-# Hansen's gravity-based accessibility to today {#grav-to-today}
+# 4. Hansen's gravity-based accessibility to today {#grav-to-today}
 
@@
 
-# Wilson's family of spatial interaction models
+# 5. Wilson's family of spatial interaction models
 While accessibility research evolved in North America with Hansen [@hansen1959], a parallel development was taking place across the Atlantic with Alan G.
@@
 This results in interpretable, balanced (given the system knowns are expressed in the constraints), and unit-consistent $ij$ flows.
-The balancing factors themselves have been subject to various interpretations -- as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in @Wilson1971) and as rents [@morphetGravityModelCalibration2023] -- reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
+The balancing factors themselves have been subject to various interpretations -- as terminal costs [@dieter1962distribution], weighted mean values [@kirbyNormalizingFactorsGravity1970], or as accessibility measures themselves [@cesario1977new] (as suggested in Wilson [@wilson1971]) and as rents [@morphetGravityModelCalibration2023] -- reflecting ongoing efforts to understand what these mathematical constructs represent behaviourally.
 In this way, the spatial interaction modeling tradition can be seen to have succeeded where accessibility modeling stalled.
@@
 
-# Accessibility and spatial interaction modelling: two divergent research streams
+# 6. Accessibility and spatial interaction modelling: two divergent research streams
 
@@
 
-# A family of accessibility measures: from proportionality to equality
+# 7. A family of accessibility measures: from proportionality to equality
 
@@
 
-## Toy example setup
+## 7.1. Toy example setup
 
@@
 The purpose of the following simple example is to demonstrate the calculation and interpretation of the four accessibility measure variants.
-
-## Unconstrained accessibility
+As well, the family of measures proposed will be available in the [{accessibility} R package](https://ipeagit.github.io/accessibility/) for convenient use in future studies.
+
+## 7.2. Unconstrained accessibility
 
@@
 For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
-These two values are a product of different impedance functions ( _physicians-minute$^{-3}$_ and _physicians-minute$^{-0.1}$_), making the comparison uninterpretable (and arguably incorrect).
+These two values are a product of different impedance functions _physicians-minute$^{-0.1}$_ and _physicians-minute$^{-3}$_), making the direct comparison uninterpretable (and arguably incorrect).
 The fundamental uninterpretability of what is an _opportunity-weighted-travel-impedance_ unit remains.
@@
 
-## Total constrained accessibility
+## 7.3. Total constrained accessibility
 
@@
 
-### Total constrained accessible opportunities: Hansen's accessibility with a total constraint
+### 7.3.1. Total constrained accessible opportunities: Hansen's accessibility with a total constraint
 In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance.
@@
 
-### Total constrained accessible population: Reilly's potential trade territories with a total constraint
+### 7.3.2. Total constrained accessible population: Reilly's potential trade territories with a total constraint
 
@@
 
-## Singly constrained accessibility
+## 7.4. Singly constrained accessibility
 
@@
 
-### Singly constrained accessible opportunities: spatial availability
+### 7.4.1. Singly constrained accessible opportunities: spatial availability
 
@@
 
-### Singly constrained accessible population: market availability
+### 7.4.2. Singly constrained accessible population: market availability
 
@@
 
-## Doubly constrained accessibility
+## 7.5. Doubly constrained accessibility
 
@@
 A doubly constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school-seat, or vaccine doses per person).
-
-As mentioned, doubly constrained approach to accessibility adopts the structure of the production-attraction spatial interaction model, where both population and opportunity totals are fixed.
 Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):
@@
 
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints -- i.e., $V^D_i = O_i$ so the physician-capacity accessible for Zones 1, 2, and 3 would be `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
-Hence, the usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values.
-$V^D_{ij}$ values represent the number of opportunities from zone $j$ allocated to populations in zone $i$, shaped by both mass and travel impedance.
-
-To illustrate this, consider the results for Zone 2 (recall: a suburban zone -- a high-population, low-opportunity, relatively remote).
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints i.e., $V^D_i = O_i$ hence the physician-capacity accessible for Zones 1, 2, and 3 are `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
+The usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values; the number of opportunities from zone $j$ allocated to populations in zone $i$ shaped by both mass and travel impedance.
+
+To illustrate this concept, consider the results for Zone 2 (recall: a suburban type zone, with a higher population, lower amount of opportunities, and relatively remote).
 As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2, its intrazonal flow (i.e., $V^D_{22}$) declines as travel impedance decay decreases -- from `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_1)|> round(digits = 3)` under $f_1(c_{ij})$ to `r OD2|>filter(oid == 2 & did == 2) |> pull(V_dc_ij_3) |> round(digits = 3)` under $f_3(c_{ij})$, out of the \~`r OD2|>filter(oid == 2) |> summarize(V_dc_i_1 = sum(V_dc_ij_1)) |> pull(V_dc_i_1) |>round()` opportunities allocated to Zone 2 (a population of `r OD2|>filter(oid == 2) |> summarize(O_i2 = first(O_i2)) |> pull(O_i2) |>round()`).
@@
 
-Recall, accessibility is conventionally presented as a zonal value not a flow.
+Accessibility is conventionally presented as a zonal value not a flow.
 However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$.
-So following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
+Following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
 Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $\kappa_{ij}^D$ and $\hat \kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
@@
 
-And as mentioned, these units are often misaligned -- for instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
-When they do align -- such as people to physician capacity -- we are modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
+And as mentioned, these units are often misaligned.
+For instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
+When they do align, such as people to physician capacity, we would be modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
 In such cases, the traditional accessibility question can be seen to be already answered by the known information (i.e., how many opportunities can be reached by a zone?
 Answer: the number of people at that zone).
-For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' access, not on predicting flows of realised access.
-
-# Conclusions
+For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' not on predicting flows of realised access.
+
+# 8. Conclusions
 
@@
 Specifically, accessibility values for each $i$-$j$ pair must be a proportion of the zonal opportunity and population values simultaneously.
-For example, the accessibility at Zone 1 must equal the sum of opportunities from zones 1, 2, and 3, as well as the sum of the population at Zone 1.
-Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair values should be of interest.
-In this sense, the research question should be concerned with 'access' (how many opportunities accessed from $j$ at $i$ based on given zonal opportunities and populations) instead of potential spatial interaction (e.g., typically expressed as a zonal summary measure of how many opportunities one could reach out of a regional total and/or zonal-allocation).
-
-Overall, and building on Wilson's [@wilson1971] foundational work, this paper proposes a unified framework for analyzing gravity-based accessibility.
+For example, the accessibility at Zone 1 must equal the sum of opportunities from Zones 1, 2, and 3, as well as the sum of the population at Zone 1.
+Satisfying the double constraint means the opportunities and population data must match one-to-one, so working with the accessibility $i$-$j$ pair value should be of research interest.
+
+Building on Wilson's [@wilson1971] foundational work, this paper proposes a unified framework for analyzing gravity-based accessibility.
 By reintroducing Wilson's proportionality constants, the family of constrained accessibility measures restores measurement units to accessibility estimates.
@@
 
-That said, there have been more recent developments in the accessibility literature, including novel person-based approaches that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete], that are not based on top-down approaches such as Wilson's model formalized through the entropy-maximisation procedure.
-While gravity-based measures of accessibility dominate the applied literature, future work could further explore how these ideas of proportionality constants and balanced units could also help inform these and other new modeling approaches.
+While this work exclusively focused on top-down gravity-based accessibility measures, there have been recent developments in the accessibility literature that include person-based approaches that are time-sensitive [@yang2024evaluating; @braga2023evaluating], behavioural [@kar2024inclusive; @lu2014effects] or utility-based [@guzman2023much; @ben1985discrete].
+While gravity-based measures of accessibility still dominate the applied literature, future work could further explore how these ideas of proportionality constants and balanced units could also help inform these and other new modeling approaches.

### 2025-10-22 paper_plos/paper_plos_V3.tex
V3 final edits AS

@@
 \linenumbers
-\section{Introduction}\label{introduction}
+\section{1. Introduction}\label{introduction}
 
@@
 
-In summary, this paper makes two contributions: (1) we review how
-spatial interaction modeling and accessibility share similar origins but
-have diverged in focus and interpretation. (2) we introduce a family of
+This paper makes two contributions. First, we review how spatial
+interaction modeling and accessibility share similar origins but have
+diverged in focus and interpretation. Second, we introduce a family of
 accessibility measures grounded in spatial interaction principles,
@@
 
-The remainder of this paper proceeds as follows. Sections 2-4 trace the
-historical development of spatial interaction and accessibility
-research, beginning with Newtonian gravitational analogies and
-\citet{careyPrinciplesSocialScience1858} (Section 2), moving through
-early researchers like \citet{ravensteinLawsMigration1885} to
-\citet{stewartDemographicGravitationEvidence1948} who theorised and
+The remainder of this paper proceeds as follows. Sections 2 through 4
+trace the historical development of spatial interaction and
+accessibility research, beginning with Newtonian gravitational analogies
+and Carey (1858) \citep{careyPrinciplesSocialScience1858} (Section 2),
+moving through early researchers like Ravenstein (1885)
+\citep{ravensteinLawsMigration1885} to Stewart (1948)
+\citep{stewartDemographicGravitationEvidence1948} who theorised and
 formalized spatial interaction patterns (Section 3), and examining how
-the ``gravity-based'' accessibility approach in \citet{hansen1959}
-became the dominant approach in planning practice (Section 4). Section 5
-presents the entropy-based family of spatial interaction models in
-\citet{wilson1971}, explaining how the introduction of ``constraints''
-(based on top-down known information as part of entropy-maximisation)
-produce interpretable, unit-consistent flow estimates. Section 6
-explores why accessibility and spatial interaction literature diverged
-into separate branches despite their shared conceptual foundations.
-Section 7 constitutes the paper's core contribution: we derive a family
-of accessibility measures corresponding to different constraints applied
--- unconstrained, singly constrained, and doubly constrained cases --
-demonstrating how each yields zone-level accessibility values expressed
-in meaningful units (opportunities or population) rather than
-unit-inconsistent indices. We illustrate each measure with numerical
-examples that clarify the practical implications of different constraint
-assumptions. Section 8 concludes by discussing how these constrained
-measures can inform planning decisions and improve clarity in
-accessibility analysis.
+the ``gravity-based'' accessibility approach in Hansen (1959)
+\citep{hansen1959} became the dominant approach in planning practice
+(Section 4). Section 5 presents the entropy-based family of spatial
+interaction models in Wilson (1971) \citep{wilson1971}, explaining how
+the introduction of ``constraints'' (based on top-down known information
+as part of entropy-maximisation) produce interpretable, unit-consistent
+flow estimates. Section 6 explores why accessibility and spatial
+interaction literature diverged into separate branches despite their
+shared conceptual foundations. Section 7 constitutes the paper's core
+contribution: we derive a family of accessibility measures corresponding
+to different constraints applied -- unconstrained, singly constrained,
+and doubly constrained cases -- demonstrating how each yields zone-level
+accessibility values expressed in meaningful units (opportunities or
+population) rather than unit-inconsistent indices. We illustrate each
+measure with numerical examples that clarify the practical implications
+of different constraint assumptions. Section 8 concludes by discussing
+how these constrained measures can inform planning decisions and improve
+clarity in accessibility analysis.
 
@@
 
-\section{Newtonian's roots of human spatial interaction
+\section{2. Newtonian's roots of human spatial interaction
 research}\label{newtonians-roots-of-human-spatial-interaction-research}
@@
 
-\section{Early research on human spatial interaction: from Ravenstein
+\section{3. Early research on human spatial interaction: from Ravenstein
 (1885) to Stewart
@@
 
-\section{Hansen's gravity-based accessibility to
+\section{4. Hansen's gravity-based accessibility to
 today}\label{grav-to-today}
@@
 
-\section{Wilson's family of spatial interaction
+\section{5. Wilson's family of spatial interaction
 models}\label{wilsons-family-of-spatial-interaction-models}
@@
 \citep{kirbyNormalizingFactorsGravity1970}, or as accessibility measures
-themselves \citep{cesario1977new} (as suggested in \citet{Wilson1971})
-and as rents \citep{morphetGravityModelCalibration2023} -- reflecting
-ongoing efforts to understand what these mathematical constructs
-represent behaviourally. In this way, the spatial interaction modeling
-tradition can be seen to have succeeded where accessibility modeling
-stalled. While Wilson's model produces results that are in units of flow
-tethered to the system of analysis, which facilitates such
-interpretations, Hansen-type measures (still widely used today in
-accessibility work) yield outputs that reflect proportional -- not equal
--- relationships and typically lack interpretable units.
+themselves \citep{cesario1977new} (as suggested in Wilson
+\citep{wilson1971}) and as rents
+\citep{morphetGravityModelCalibration2023} -- reflecting ongoing efforts
+to understand what these mathematical constructs represent
+behaviourally. In this way, the spatial interaction modeling tradition
+can be seen to have succeeded where accessibility modeling stalled.
+While Wilson's model produces results that are in units of flow tethered
+to the system of analysis, which facilitates such interpretations,
+Hansen-type measures (still widely used today in accessibility work)
+yield outputs that reflect proportional -- not equal -- relationships
+and typically lack interpretable units.
 
@@
 
-\section{Accessibility and spatial interaction modelling: two divergent
-research
+\section{6. Accessibility and spatial interaction modelling: two
+divergent research
 streams}\label{accessibility-and-spatial-interaction-modelling-two-divergent-research-streams}
@@
 
-\section{A family of accessibility measures: from proportionality to
+\section{7. A family of accessibility measures: from proportionality to
 equality}\label{a-family-of-accessibility-measures-from-proportionality-to-equality}
@@
 
-\subsection{Toy example setup}\label{toy-example-setup}
+\subsection{7.1. Toy example setup}\label{toy-example-setup}
 
@@
 following simple example is to demonstrate the calculation and
-interpretation of the four accessibility measure variants.
-
-\subsection{Unconstrained
+interpretation of the four accessibility measure variants. As well, the
+family of measures proposed will be available in the
+\href{https://ipeagit.github.io/accessibility/}{\{accessibility\} R
+package} for convenient use in future studies.
+
+\subsection{7.2. Unconstrained
 accessibility}\label{unconstrained-accessibility}
@@
 370.92, but in what units? These two values are a product of different
-impedance functions ( \emph{physicians-minute\(^{-3}\)} and
-\emph{physicians-minute\(^{-0.1}\)}), making the comparison
+impedance functions \emph{physicians-minute\(^{-0.1}\)} and
+\emph{physicians-minute\(^{-3}\)}), making the direct comparison
 uninterpretable (and arguably incorrect). The fundamental
@@
 
-\subsection{Total constrained
+\subsection{7.3. Total constrained
 accessibility}\label{total-constrained-accessibility}
@@
 
-\subsubsection{Total constrained accessible opportunities: Hansen's
-accessibility with a total
+\subsubsection{7.3.1. Total constrained accessible opportunities:
+Hansen's accessibility with a total
 constraint}\label{total-constrained-accessible-opportunities-hansens-accessibility-with-a-total-constraint}
@@
 
-\subsubsection{Total constrained accessible population: Reilly's
+\subsubsection{7.3.2. Total constrained accessible population: Reilly's
 potential trade territories with a total
@@
 
-\subsection{Singly constrained
+\subsection{7.4. Singly constrained
 accessibility}\label{singly-constrained-accessibility}
@@
 
-\subsubsection{Singly constrained accessible opportunities: spatial
+\subsubsection{7.4.1. Singly constrained accessible opportunities:
+spatial
 availability}\label{singly-constrained-accessible-opportunities-spatial-availability}
@@
 
-\subsubsection{Singly constrained accessible population: market
+\subsubsection{7.4.2. Singly constrained accessible population: market
 availability}\label{singly-constrained-accessible-population-market-availability}
@@
 
-\subsection{Doubly constrained
+\subsection{7.5. Doubly constrained
 accessibility}\label{doubly-constrained-accessibility}
@@
 (e.g., job per worker, student per school-seat, or vaccine doses per
-person).
-
-As mentioned, doubly constrained approach to accessibility adopts the
-structure of the production-attraction spatial interaction model, where
-both population and opportunity totals are fixed. Mathematically, this
-model requires the simultaneous imposition of both the population- and
-opportunity- constraints in the preceding singly constrained variants
-(Eq~\ref{eq-opportunity-constraint} and
+person). Mathematically, this model requires the simultaneous imposition
+of both the population- and opportunity- constraints in the preceding
+singly constrained variants (Eq~\ref{eq-opportunity-constraint} and
 Eq~\ref{eq-population-constraint}), namely the sum of population in all
@@
 This adjustment to the example is summarised in
-\textbf{?@tbl-small-system-land-use-doubly-constrained-case}. With the
+Table~\ref{tbl-small-system-land-use-doubly-constrained-case}. With the
 population (in units of 10,000s of people seeking physicians) and the
@@
 
-::: \{\#tbl-small-system-land-use-doubly constrained-case .cell
-tbl-cap=`Modified simple system with three zones reflecting matched
-population and opportunities. Population is in 10,000 persons and
-opportunities in 10,000 of physician-capacity.'\} :::
-\{.cell-output-display\}
-
 \begin{table}
+
+\caption{\label{tbl-small-system-land-use-doubly-constrained-case}Modified
+simple system with three zones reflecting matched population and
+opportunities. Population is in 10,000 persons and opportunities in
+10,000 of physician-capacity.}
+
+\centering{
+
 \fontsize{7.5pt}{9.0pt}\selectfont
@@
 \end{tabular*}
-\end{table}
-
-::: :::
+
+}
+
+\end{table}%
 
@@
 longer meaningfully represented as zonal summaries like \(V^D_i\) or
-\(M^D_j\), since these values reproduce the original constraints --
-i.e., \(V^D_i = O_i\) so the physician-capacity accessible for Zones 1,
-2, and 3 would be 4.002, 9.996, and 6.002. Hence, the usefulness of the
-doubly constrained measure lies in the interpretation as \(V_{ij}^D\)
-values. \(V^D_{ij}\) values represent the number of opportunities from
-zone \(j\) allocated to populations in zone \(i\), shaped by both mass
-and travel impedance.
-
-To illustrate this, consider the results for Zone 2 (recall: a suburban
-zone -- a high-population, low-opportunity, relatively remote). As shown
-in
+\(M^D_j\), since these values reproduce the original constraints i.e.,
+\(V^D_i = O_i\) hence the physician-capacity accessible for Zones 1, 2,
+and 3 are 4.002, 9.996, and 6.002. The usefulness of the doubly
+constrained measure lies in the interpretation as \(V_{ij}^D\) values;
+the number of opportunities from zone \(j\) allocated to populations in
+zone \(i\) shaped by both mass and travel impedance.
+
+To illustrate this concept, consider the results for Zone 2 (recall: a
+suburban type zone, with a higher population, lower amount of
+opportunities, and relatively remote). As shown in
 Table~\ref{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2},
@@
 
-Recall, accessibility is conventionally presented as a zonal value not a
-flow. However, in the doubly constrained case, since we force the
-allocation of zonal population demand and zonal opportunities supplied
-to be paired and allocation to be proportional, \(V_i^D\) is simply the
-number of opportunities that matches our known population at \(i\). So
-following the logic of the family of accessibility measures, in the
-doubly constrained case, \(V_{ij}^D\) flows are the only relevant unit
-of analysis: spatial proportional allocations between population and
-opportunity capacity. Furthermore, \(V^D_{ij}\) and its transposed
-counterpart \(M^D_{ji}\) are structurally identical, differing only in
-interpretation (referring to \(\kappa_{ij}^D\) and
-\(\hat \kappa_{ij}^D\)): one reflects the proportional allocation of
-opportunity to population flows; the other, population to opportunity
-flows.
+Accessibility is conventionally presented as a zonal value not a flow.
+However, in the doubly constrained case, since we force the allocation
+of zonal population demand and zonal opportunities supplied to be paired
+and allocation to be proportional, \(V_i^D\) is simply the number of
+opportunities that matches our known population at \(i\). Following the
+logic of the family of accessibility measures, in the doubly constrained
+case, \(V_{ij}^D\) flows are the only relevant unit of analysis: spatial
+proportional allocations between population and opportunity capacity.
+Furthermore, \(V^D_{ij}\) and its transposed counterpart \(M^D_{ji}\)
+are structurally identical, differing only in interpretation (referring
+to \(\kappa_{ij}^D\) and \(\hat \kappa_{ij}^D\)): one reflects the
+proportional allocation of opportunity to population flows; the other,
+population to opportunity flows.
 
@@
 
-And as mentioned, these units are often misaligned -- for instance, we
-may not know how much park space, grocery area, or childcare capacity is
-accessible per person. When they do align -- such as people to physician
-capacity -- we are modeling realised access flows based on known
+And as mentioned, these units are often misaligned. For instance, we may
+not know how much park space, grocery area, or childcare capacity is
+accessible per person. When they do align, such as people to physician
+capacity, we would be modeling realised access flows based on known
 quantities of \emph{those that interact} and the \emph{interacted}. In
@@
 widely used in accessibility analysis as the literature has largely
-focused on questions of `potential' access, not on predicting flows of
-realised access.
-
-\section{Conclusions}\label{conclusions}
+focused on questions of `potential' not on predicting flows of realised
+access.
+
+\section{8. Conclusions}\label{conclusions}
 
@@
 the accessibility at Zone 1 must equal the sum of opportunities from
-zones 1, 2, and 3, as well as the sum of the population at Zone 1.
+Zones 1, 2, and 3, as well as the sum of the population at Zone 1.
 Satisfying the double constraint means the opportunities and population
 data must match one-to-one, so working with the accessibility
-\(i\)-\(j\) pair values should be of interest. In this sense, the
-research question should be concerned with `access' (how many
-opportunities accessed from \(j\) at \(i\) based on given zonal
-opportunities and populations) instead of potential spatial interaction
-(e.g., typically expressed as a zonal summary measure of how many
-opportunities one could reach out of a regional total and/or
-zonal-allocation).
-
-Overall, and building on Wilson's \citep{wilson1971} foundational work,
-this paper proposes a unified framework for analyzing gravity-based
-accessibility. By reintroducing Wilson's proportionality constants, the
-family of constrained accessibility measures restores measurement units
-to accessibility estimates. This enhancement provides a more
-interpretable, consistent, and theoretically grounded basis for
-accessibility analysis, which could help advance the adoption of
-accessibility-oriented planning.
-
-That said, there have been more recent developments in the accessibility
-literature, including novel person-based approaches that are
+\(i\)-\(j\) pair value should be of research interest.
+
+Building on Wilson's \citep{wilson1971} foundational work, this paper
+proposes a unified framework for analyzing gravity-based accessibility.
+By reintroducing Wilson's proportionality constants, the family of
+constrained accessibility measures restores measurement units to
+accessibility estimates. This enhancement provides a more interpretable,
+consistent, and theoretically grounded basis for accessibility analysis,
+which could help advance the adoption of accessibility-oriented
+planning.
+
+While this work exclusively focused on top-down gravity-based
+accessibility measures, there have been recent developments in the
+accessibility literature that include person-based approaches that are
 time-sensitive \citep{yang2024evaluating, braga2023evaluating},
 behavioural \citep{kar2024inclusive, lu2014effects} or utility-based
-\citep{guzman2023much, ben1985discrete}, that are not based on top-down
-approaches such as Wilson's model formalized through the
-entropy-maximisation procedure. While gravity-based measures of
-accessibility dominate the applied literature, future work could further
-explore how these ideas of proportionality constants and balanced units
-could also help inform these and other new modeling approaches.
+\citep{guzman2023much, ben1985discrete}. While gravity-based measures of
+accessibility still dominate the applied literature, future work could
+further explore how these ideas of proportionality constants and
+balanced units could also help inform these and other new modeling
+approaches.
 

### 2025-10-23 paper_plos/paper_plos_V3.qmd
cdh final proofread for v3

@@
       - \newcommand{\elandscape}{\end{landscape}}
-    #include-in-header:
-    #  text: |
-    #    % Remove comment for double spacing
-    #    % \usepackage{setspace} 
-    ##    % \doublespacing
-    #    % \usepackage{pdflscape}
-    #    % \newcommand{\blandscape}{\begin{landscape}}
-    #    % \newcommand{\elandscape}{\end{landscape}}
 author:
@@
 Despite growing recognition of accessibility-based planning approaches, the concept remains fragmented, with inconsistent definitions and unclear interpretations.
-To this end, this paper offers a methodological contribution by specifying a family of accessibility measures that are grounded in the shared 'gravity-based' theoretical roots of spatial interaction models, particularly their balancing factors.
-From this foundation, we outline four memebers of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly constrained' measure (i.e., related to the popular two step floating catchment approach -- 2SFCA), and the 'doubly constrained' measure representing realized access (i.e., equal to the doubly constrained spatial interaction model).
+To this end, this paper makes a methodological contribution by specifying a family of accessibility measures that are grounded in the shared 'gravity-based' theoretical roots of spatial interaction models, particularly their balancing factors.
+From this foundation, we outline four members of the family: the 'unconstrained' measure (i.e., Hansen-type accessibility), the 'total-constrained' measure (i.e., a constrained version of the Hansen-type accessibility), the 'singly-constrained' measure (i.e., related to the popular two-step floating catchment approach -- 2SFCA), and the 'doubly-constrained' measure representing realized access (i.e., equal to the doubly-constrained spatial interaction model).
 These measures can be interpreted as either the number of accessible opportunities or accessible population (i.e., market potential).
@@
 First, we review how spatial interaction modeling and accessibility share similar origins but have diverged in focus and interpretation.
-Second, we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly constrained cases and variants of "accessible opportunities" and "accessible population".
+Second, we introduce a family of accessibility measures grounded in spatial interaction principles, including total, singly, and doubly-constrained cases and variants of "accessible opportunities" and "accessible population".
 These cases and variants align with common measures such as Hansen-type accessibility [@hansen1959], competitive accessibility measures such as the 2SFCA method [@shen1998; @luo2003], and market potential models [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
@@
 The remainder of this paper proceeds as follows.
-Sections 2 through 4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and Carey (1858) [@careyPrinciplesSocialScience1858] (Section 2), moving through early researchers like Ravenstein (1885) [@ravensteinLawsMigration1885] to Stewart (1948) [@stewartDemographicGravitationEvidence1948] who theorised and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in Hansen (1959) [@hansen1959] became the dominant approach in planning practice (Section 4).
-Section 5 presents the entropy-based family of spatial interaction models in Wilson (1971) [@wilson1971], explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximisation) produce interpretable, unit-consistent flow estimates.
+Sections 2 through 4 trace the historical development of spatial interaction and accessibility research, beginning with Newtonian gravitational analogies and Carey (1858) [@careyPrinciplesSocialScience1858] (Section 2), moving through early researchers like Ravenstein (1885) [@ravensteinLawsMigration1885] to Stewart (1948) [@stewartDemographicGravitationEvidence1948] who theorized and formalized spatial interaction patterns (Section 3), and examining how the "gravity-based" accessibility approach in Hansen (1959) [@hansen1959] became the dominant approach in planning practice (Section 4).
+Section 5 presents the entropy-based family of spatial interaction models in Wilson (1971) [@wilson1971], explaining how the introduction of "constraints" (based on top-down known information as part of entropy-maximization) produces interpretable, unit-consistent flow estimates.
 Section 6 explores why accessibility and spatial interaction literature diverged into separate branches despite their shared conceptual foundations.
-Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied -- unconstrained, singly constrained, and doubly constrained cases -- demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices.
+Section 7 constitutes the paper's core contribution: we derive a family of accessibility measures corresponding to different constraints applied -- unconstrained, singly-constrained, and doubly-constrained cases -- demonstrating how each yields zone-level accessibility values expressed in meaningful units (opportunities or population) rather than unit-inconsistent indices.
 We illustrate each measure with numerical examples that clarify the practical implications of different constraint assumptions.
@@
 Where $G$ is the gravitational constant -- an empirically calibrated value that ensures the model reflects observed forces.
-Newton's initial estimate of $G$ was based on a speculation <!--that the mean density of earth was between five or six times that of water, an assumption that -->but received empirical support after Hutton's and Cavendish' experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798] which estimated it to within 1% accuracy.
+Newton's initial estimate of $G$ was based on a speculation <!--that the mean density of earth was between five or six times that of water, an assumption that -->but received empirical support after Hutton's and Cavendish's experiments in the late 1700s [@hutton_xxxiii_1778; @cavendish_xxi_1798], which estimated it to within 1% accuracy.
 That is, it took over a century from the publication of Newton's *Principia Mathematica* to refine the estimate of the proportionality constant.
@@
 Carey's *Principles of Social Science* [@careyPrinciplesSocialScience1858] inspired empirical spatial interaction research in different contexts.
-Namely, a number of researchers theoretically and empirically attempted to characterise human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance -- conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
+Namely, a number of researchers theoretically and empirically attempted to characterize human spatial interaction as a force $F$ directly proportional to the "masses" $M_i$ and $M_j$ of two locations, and inversely proportional to their separation distance -- conceptually parallel to Newtonian gravity, but often omitting a proportionality constant.
 
@@
 
-What is notable about @eq-stewart-force, however, is the proportionality constant $G$ was specified but "left for future determination" [@stewartDemographicGravitationEvidence1948, p.
+What is notable about @eq-stewart-force, however, is that the proportionality constant $G$ was specified but "left for future determination" [@stewartDemographicGravitationEvidence1948, p.
 34].
@@
 
-From Stewart [@stewartDemographicGravitationEvidence1948], we arrive to 1959 and Walter G.
+From Stewart [@stewartDemographicGravitationEvidence1948], we arrive at 1959 and Walter G.
 Hansen, whose work proved to be exceptionally influential in the accessibility literature [@hansen1959].
@@
 
-$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from a zone $i$.
+$S_{i}$ in @eq-accessibility-general is a measure of the accessibility from zone $i$.
 This is a function of $O_j$ (the mass of opportunities at $j$), $d_{ij}$ (the cost, e.g., distance or travel time, incurred to reach $j$ from $i$), and $f(d_{ij})$ (a function that modulates the friction of cost).
@@
 Although Stewart included $G$ explicitly (with a note that "$G$ \[was\] left for future determination: a suitable choice of other units can reduce it to unity" [p.
-34].) Hansen made no mention of it.
+34]).
+Hansen made no mention of it.
 As a result, modern accessibility analysis has largely evolved without addressing the constant's role, leaving $G$ effectively and implicitly fixed at 1.
@@
 
-Furthermore, working with a proportional relationship generates fundamental issues in comparability between and, arguably within, studies.
+Furthermore, working with a proportional relationship generates fundamental issues in comparability between and, arguably, within studies.
 Namely, accessibility estimates have no fixed unit, rendering them sensitive to the choice of impedance functions.
@@
 However, when $f(d_{ij})$ is set to equal $e^{-\beta d_{ij}}$, the units become opportunities per $e^{\beta \text{metres}}$.
-Such variation impairs comparability across analysis and obscures the meaning of accessibility scores, making them difficult to understand and communicate without post-hoc treatment.
-
-Therefore in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
+Such variation impairs comparability across analyses and obscures the meaning of accessibility scores, making them difficult to understand and communicate without post-hoc treatment.
+
+Therefore, in practice, Hansen-type measures are ones of proportionality and are better understood as *ordinal indicators*; they rank accessibility but lack cardinal meaning or consistent units [@millerAccessibilityMeasurementApplication2018].
 The continued absence of a sort of proportionality constant $G$ leaves a conceptual and practical gap in accessibility analysis: a missing link between theoretical form and empirical measurement that hinders the interpretability of accessibility measures.
@@
 Wilson.
-Wilson's ground breaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
+Wilson's groundbreaking paper [@wilson1971] advanced a general framework for spatial interaction modeling focused on flows of interaction between places and derived from a non-Newtonian analogy.
 Additionally, this work was not focused on the 'potential' concept as accessibility had been.
@@
 In some sense, $k$ serves a role similar to the gravitational constant in Newton's law -- it calibrates the model so that outputs match real-world quantities.
-However, these real-world quantities are not a set empirical constant (like Newton's $G$) but instead, sensitive to the system and known information about the system.
+However, these real-world quantities are not a set empirical constant (like Newton's $G$) but are instead sensitive to the system and known information about the system.
 
 From the outset, spatial interaction models emphasized interpretability of results [@kirbyNormalizingFactorsGravity1970; @wilsonSTATISTICALTHEORYSPATIAL1967; @wilson1971].
-But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson's innovation was to generalise the model using _entropy maximisation_.
+But unlike earlier approaches that borrowed heuristically from Newtonian gravity (i.e., interaction between masses over distance), Wilson's innovation was to generalize the model using _entropy maximization_.
 By maximizing the number of ways individual trip probabilities could be arranged under known constraints, Wilson derived models that estimate _statistical averages_ of flows between zones [@wilson1971; @seniorGravityModellingEntropy1979].
@@
 Additional information can be introduced.
-For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ in @eq-phys-gravity-model, represented as $O_i$, then following constraint can be used:
+For example, when information is available about the total number of flows produced by each origin, $W_i^{(1)}$ in @eq-phys-gravity-model, represented as $O_i$, then the following constraint can be used:
 
@@
 
-In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly constrained model (either by origins $A_i$ or destinations $B_j$ -- explained below), and (3) a doubly constrained model that satisfies both.
+In the framework introduced and inferred from Wilson [@wilson1971], three types of balancing factors are specified: (1) an unconstrained model that only matches the total volume of interaction $K$, (2) a singly-constrained model (either by origins $A_i$ or destinations $B_j$ -- explained below), and (3) a doubly-constrained model that satisfies both.
 
 In the unconstrained model, constraints in @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel do not hold.
-In practical terms, this means that the total number of flows predicted by the model must be equal to sum of all flows from origins $i$ to destinations $j$.
+In practical terms, this means that the total number of flows predicted by the model must be equal to the sum of all flows from origins $i$ to destinations $j$.
 The balancing factor $K$ takes the place of $k$ and is equal to the following (as specified in [@cliff_evaluating_1974] and [@fotheringham_spatial_1984]):
@@
 
-In the singly constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
-When only @eq-constraint1-gravitymodel holds, entropy maximisation leads to the production-constrained singly constrained version of @eq-phys-gravity-model, where the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$.
-Also, $k$ is replaced with a set of balancing factors specific to origins $A_i$ which ensures that constraint @eq-constraint1-gravitymodel is satisfied (i.e., the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$).
+In the singly-constrained model, only constraint @eq-constraint1-gravitymodel or constraint @eq-constraint2-gravitymodel hold.
+When only @eq-constraint1-gravitymodel holds, entropy maximization leads to the production-constrained singly-constrained version of @eq-phys-gravity-model, where the proxy for the mass at the origin $W_i^{(1)}$ is replaced with $O_i$.
+Also, $k$ is replaced with a set of balancing factors specific to origins $A_i$, which ensures that constraint @eq-constraint1-gravitymodel is satisfied (i.e., the sum of predicted flows from one origin going to all destinations must equal the known mass at that origin $O_i$).
 Satisfying this constraint also implicitly fulfills the total constraint (@eq-constraint0-gravitymodel), since the sum of $O_i$ values across all origins equals the total number of flows.
@@
 
-The singly constrained attraction-constrained model is similar to the production-constrained version but from the perspective of the mass at the destination.
+The singly-constrained attraction-constrained model is similar to the production-constrained version but from the perspective of the mass at the destination.
 For the attraction-constrained model, the proxy for the mass at the destination $W_j^{(2)}$ is replaced with $D_j$ in @eq-phys-gravity-model, representing the spatial interaction inbound flow.
@@
 
-Lastly, the doubly constrained model is the production-attraction constrained model in Wilson [@wilson1971].
+Lastly, the doubly-constrained model is the production-attraction constrained model in Wilson [@wilson1971].
 In this case, both constraints @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously.
-These constraints ensure that the sum of predicted flows from one origin to all destination, and the predicted flows going to one destination from all origins must equal the known mass of the origin $O_i$ and of the destination $D_j$.
+These constraints ensure that the sum of predicted flows from one origin to all destinations, and the predicted flows going to one destination from all origins, must equal the known mass of the origin $O_i$ and of the destination $D_j$.
 This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly constrained, and from @eq-phys-gravity-model, $k$ becomes both $A_i$ and $B_j$ shown in @eq-doubly-constrained-balancing-factors, and $W_i^{(1)}$ and $W_j^{(2)}$ is replaced with $O_i$ and $D_j$.
+The resulting model is, in Wilson's terms, doubly-constrained, and from @eq-phys-gravity-model, $k$ becomes both $A_i$ and $B_j$ shown in @eq-doubly-constrained-balancing-factors, and $W_i^{(1)}$ and $W_j^{(2)}$ is replaced with $O_i$ and $D_j$.
 Derivation of these models is demonstrated in detail elsewhere [e.g., @ortuzar_2011_modelling; @wilsonSTATISTICALTHEORYSPATIAL1967].
@@
 Wilson's work is notable for many reasons.
-Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximisation.
+Rather than relying on a universal constant like $G$ or a scaling factor to balance units, Wilson's models calibrate interaction flows through known empirical constraints using principles of entropy maximization.
 This results in interpretable, balanced (given the system knowns are expressed in the constraints), and unit-consistent $ij$ flows.
@@
 
-Before demonstrating the derivation of family of accessibility measures that are sensitive to constraints (Section 7), in the next section we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
-This investigation sheds light on why accessibility research may have failed to adopt a comparable approach, until this paper.
+Before demonstrating the derivation of the family of accessibility measures that are sensitive to constraints (Section 7), in the next section, we review how conceptually intertwined the spatial interaction modelling and accessibility literatures are, and where they began to diverge.
+This investigation sheds light on why accessibility research may have failed to adopt a comparable approach until this paper.
 
@@
 #| out-width: 70%
-#| fig-cap: "Historical pattern of publication: documents per year."
+#| fig-cap: "Historical pattern of publication: documents per year"
+#| eval: false
 
@@
 Hansen's [@hansen1959] formulation of accessibility became the method used for decades of work on transport equity, land use analysis, and urban accessibility planning.
-Meanwhile, Wilson's [@wilson1971] entropy-maximising framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
+Meanwhile, Wilson's [@wilson1971] entropy-maximizing framework reshaped how spatial interaction models were constructed, particularly in transportation demand forecasting.
 We argue the framework's quiet innovation -- introducing empirically grounded constraints to shift from proportionality to calibrated equality -- made the framework immediately relevant for policy applications as outputs were in tangible units.
@@
 This divergence is especially striking given the context in which both frameworks emerged.
-As noted in [@battyChronicleScientificPlanning1994], large scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
-And these needs were far from trivial: for instance in the U.S., the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today's dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
+As noted in [@battyChronicleScientificPlanning1994], large-scale spatial interaction models (like Wilson's) responded to important developments at the time, a need "to meet the dictates and needs of public policy for strategic land use and transportation planning".
+And these needs were far from trivial: for instance, in the U.S., the Federal-Aid Highway Act of 1956 set in motion the construction of the Interstate Highway System with an eventual budget exceeding $600 billion in today's dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007].
 In this context, spatial interaction models were incorporated into institutional practices focused on "predict and provide" travel demand forecasting [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016].
@@
 To explore why Wilson's approach may not have crossed over to accessibility modeling sooner, we conducted a review of the literature citing Hansen [@hansen1959], Wilson [@wilson1971], or both.
-This was done using Web of Science's "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971]).
+This was done using Web of Science's "Cited References" functionality, and the digital object identifiers of Hansen [@hansen1959] and Wilson [@wilson1971].
 Only `r (n_b) |> prettyNum(big.mark = ",")` out of the `r (n_h+n_w+n_b) |> prettyNum(big.mark = ",")` documents that emerged from our search of the General Database cite both.
-The number of articles, by year and if they cite Hasen, Wilson, or both are shown in @fig-docs-per-year.
+The number of articles, by year and if they cite Hasen, Wilson, or both, are shown in Figure 1.
 
 Through the close analysis of why articles that cite both works cite each work, we identify two distinct patterns: one group of articles focused on accessibility, the other on spatial interaction.
-In examining these groups of articles, we uncover how the relationship between Wilson and Hansen has often been misunderstood, underexplored, or entirely overlooked.
-
-In the first stream of literature -- which cite both but are focused more so on spatial interaction models -- they treat spatial interaction and accessibility as separate but related phenomenons.
+In examining these groups of articles, we uncover how the relationship between Wilson and Hansen has often been misunderstood, under-explored, or entirely overlooked.
+
+In the first stream of literature -- which cite both but are focused more on spatial interaction models -- they treat spatial interaction and accessibility as separate but related phenomena.
 Four subsets of this stream emerge.
@@
 
-The second subset of articles within this stream use both Hansen [@hansen1959] and Wilson's [@wilson1971] framework in conjunction.
+The second subset of articles within this stream uses both Hansen [@hansen1959] and Wilson's [@wilson1971] framework in conjunction.
 For instance, some articles argue that spatial interaction models fail to explain certain spatial patterns on their own, for instance, as in Fotheringham [@fotheringhamSpatialCompetitionAgglomeration1985] who demonstrates how the spatial interaction model may insufficiently explain spatial patterns, and suggests that explicitly defining destinations' accessibility (Hansen-type accessibility) as a variable within the model may remedy the issue (e.g., the *competition destination* model).
@@
 
-The third subset of the spatial-interaction focused literature depart from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
+The third subset of the spatial-interaction focused literature departs from Hansen's [@hansen1959] definition, aligning instead with microeconomic or utility-based interpretations of potential spatial interaction e.g., [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].
 Though across works in this subset, they recognize Hansen-type accessibility as an indicator of 'potential' but as a separate but related concept to spatial interaction.
 
-Moving onto the group of accessibility-focused literature that cites both works, we categorise their citation of Wilson [@wilson1971] within three general groups.
-Overall these works do not engage -- or only superficially engage -- with Wilson [@wilson1971].
+Moving on to the group of accessibility-focused literature that cites both works, we categorize their citation of Wilson [@wilson1971] within three general groups.
+Overall, these works do not engage -- or only superficially engage -- with Wilson [@wilson1971].
 
 Firstly, there is a group of articles within this stream that cite Wilson [@wilson1971] exclusively as attribution for using context-dependent travel cost functions.
-This trend is common: for instance it is done in the following papers: [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
+This trend is common: for instance, it is done in the following papers: [@handyMeasuringAccessibilityExploration1997; @weibullNumericalMeasurementAccessibility1980; @kwan1998space; @shenLocationCharacteristicsInnercity1998; @ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining].
 However, these works do not engage with spatial interaction beyond this attribution.
@@
 Indeed, while accessibility can be seen as the *potential* for spatial interaction -- and Wilson [@wilson1971] briefly touches on this -- such mentions have not resulted in deeper analytical integration of these concepts.
-Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
+Furthermore, some of this literature also occasionally conflates or blurs the distinction entirely, for instance, by co-citing Hansen and Wilson as being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
 This conflation reveals ongoing murkiness between the distinction of spatial interaction and the *potential for* spatial interaction in the literature.
 
-Thirdly, there is a group of accessibility-focused works that interprets the measure used in Hansen [@hansen1959] as the singly- or doubly- constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
+Thirdly, there is a group of accessibility-focused works that interprets the measure used in Hansen [@hansen1959] as the singly- or doubly-constrained spatial interaction model's inverse balancing factor [e.g., @vickermanAccessibilityAttractionPotential1974].
 This group often cites the spatial interaction works that make this connection (i.e., the first subset of the first stream of literature) and is especially prominent in the investigation of competitive accessibility topics e.g., [@karstEvaluationAccessibilityImpacts2003; @geurs2006accessibility; @willigers2007accessibility; @el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling].
@@
 
-On that note, Soukhov et al., 2023 and 2024 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] introduce the balancing factors as a mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
+On that note, Soukhov et al., 2023 and 2024 [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024] introduce the balancing factors as mechanisms to ensure that opportunities at each destination are proportionally allocated to each zone (based on the proportion of population seeking opportunities and the relative travel impedance).
 This is to ensure that each zonal accessibility value is the sum of this proportional allocation from each destination, and that all zonal values ultimately sum to the total number of opportunities in the region.
 However, these balancing factors were deduced intuitively.
-These works did not explicitly state that the mathematical formulation of the equations are effectively equivalent to Wilson's singly constrained model (derived from entropy maximization).
+These works did not explicitly state that the mathematical formulations of the equations are effectively equivalent to Wilson's singly-constrained model (derived from entropy maximization).
 This equivalence is only discovered in hindsight, as will be demonstrated in the following section.
@@
 
-In sum, despite the interpretative advantaged offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based intuition to the concept of accessibility.
+In sum, despite the interpretative advantages offered by the statistical logic of Wilson [@wilson1971]'s framework, neither the spatial interaction literature that cites Hansen [@hansen1959] nor the accessibility literature citing Wilson [@wilson1971] has meaningfully applied Wilson's constraint-based intuition to the concept of accessibility.
 So, in the next section, we do so by demonstrating how the application of Wilson's framework enables accessibility to move from a proportional relationship to calibrated equality -- tying outputs to tangible system knowns.
 This re-expression of accessibility using constraints yields interpretable, unit-consistent measures of opportunity.
-This approach takes the same path of entropy-maximisation as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially suggested in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalised).
+This approach takes the same path of entropy-maximization as in Wilson [@wilson1971], and does not rely on specifying some universal constant $G$ like initially suggested in Stewart [@stewartDemographicGravitationEvidence1948] (recall, @eq-stewart-population-potential-sum which Hansen [@hansen1959] operationalized).
 
@@
 This section introduces this conceptual connection by defining a family of accessibility measures using Wilson's approach grounded in statistical mechanics, shifting place-based accessibility ($\grave{a}$ la Hansen [@hansen1959]) from a relationship describing proportionality to one of equality.
-This shift addresses the issue of unit interpretability (associated with the proportional nature of Hansen-type accessibility indicators previous outlined).
+This shift addresses the issue of unit interpretability (associated with the proportional nature of Hansen-type accessibility indicators previously outlined).
 
 We propose a revised definition of accessibility considerate of the constraint-based spatial interaction model: _the potential for spatial interaction with opportunities (or population)_.
-We can specify $k$ as a type of proportional allocation factor $\kappa$ which incorporates Wilson's balancing factor(s) to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potential for spatial interaction with population_ $M_{ji}$.
+We can specify $k$ as a type of proportional allocation factor $\kappa$, which incorporates Wilson's balancing factor(s) to define the *potential for spatial interaction with opportunities* $V_{ij}$ and the _potential for spatial interaction with population_ $M_{ji}$.
 In effect, $\kappa$ is unitless and proportionally allocates (based on the constraints of the case) opportunities (for $V_{ij}$) and population (for $M_{ji}$).
@@
 
-\noindent Where $W_X^{(2)}$ is the mass of the destination (i.e., opportunities $O_j$ or $O$) and $W_X^{(1)}$ is the mass of the origin (i.e., population $D_i$ or $D$) for either the zone or full region depending on the case (hence represented by with a stand-in $X$ subindex).
-Accessibility flows can also be summarised as as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively.
+\noindent Where $W_X^{(2)}$ is the mass of the destination (i.e., opportunities $O_j$ or $O$) and $W_X^{(1)}$ is the mass of the origin (i.e., population $D_i$ or $D$) for either the zone or full region, depending on the case (hence represented by a stand-in $X$ subindex).
+Accessibility flows can also be summarised as a partial sum of the potential at $i$ and at $j$ to express accessibility at the origin zone or at the destination zone, respectively.
 This form is common in accessibility research:
@@
 
-Figure @fig-analytical-device-conc-accessibility illustrates our analytical framework using a 3-zone system.
+Figure 2 illustrates our analytical framework using a 3-zone system.
 The most detailed values, $X_{ij}$, represent the potential for spatial interaction from origin $i$ to destination $j$.
 Here, $X$ stands for all cases and variants to be discussed (e.g., $V_{ij}^0$, $M_{ji}^0$, $V_{ij}^T$, $M_{ji}^T$, $V_{ij}^S$, $M_{ji}^S$, $V_{ij}^D$, and $M_{ji}^D$).
-Single marginals show the origin and destination weights and the total marginal is a sum of these values.
+Single marginals show the origin, and destination weights and the total marginal is a sum of these values.
 
@@
 #| fig-cap: "The family of accessibility measures analytical framework: labelling and associating ij flows, zonal weights, the single marginals, and the total marginal."
-
-knitr::include_graphics(glue::glue(here::here(),
-                                   "/figures/access-analytical-device.png"))
+#| eval: false
+
+knitr::include_graphics(glue::glue(here::here(), "/figures/access-analytical-device.png"))
 ```
@@
 1. **Unconstrained Case ($V_i^0$, $M_j^0$)**
-- Equivalent to Hansen's [@hansen1959] and Reilly's [@reilly1929methods] original formulations; the status quo of accessibility modelling.
+- Equivalent to Hansen's [@hansen1959] and Reilly's [@reilly1929methods] original formulations, the status quo of accessibility modelling.
 - No balancing factors applied; units are in "opportunities-by-impedance" for $V_i^0$ or "population-by-impedance" for $M_j^0$.
@@
 
-2. **Total Constrained Case ($V_i^T$, $M_j^T$)**
-- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in @fig-analytical-device-conc-accessibility) i.e., total number of opportunities or population in the system.
+2. **Total-constrained Case ($V_i^T$, $M_j^T$)**
+- Applies a total proportional allocation factor ($\kappa_{ij}^T$, $\hat \kappa_{ji}^T$) based only on the total marginal (green box in Figure 2) i.e., total number of opportunities or population in the system.
 This ensures the sum of all values in the system match the total marginal.
-- Units of $V_i^T$: accessible opportunities from $i$, a value that is total constrained and linearly proportion to $V_i^0$.
-- Units of $M_j^T$: accessible population from $j$, a value that is total constrained and linearly proportion to $M_j^0$.
-
-3. **Singly Constrained Case ($V_i^S$, $M_j^S$)**
-- Applies singly constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in @fig-analytical-device-conc-accessibility) i.e., the number of opportunities or population at each zone.
-Reflects how te literature calculates competitive accessibility.
+- Units of $V_i^T$: accessible opportunities from $i$, a value that is total-constrained and linearly proportional to $V_i^0$.
+- Units of $M_j^T$: accessible population from $j$, a value that is total-constrained and linearly proportional to $M_j^0$.
+
+3. **Singly-constrained Case ($V_i^S$, $M_j^S$)**
+- Applies singly-constrained proportional allocation factors ($\kappa_{ij}^S$, $\hat \kappa_{ji}^S$) based on Wilson's balancing factors ($B_j$, $A_i$) to preserve either the destination-side or origin-side marginal totals (blue and red boxes in Figure 2) i.e., the number of opportunities or population at each zone.
+Reflects how the literature calculates competitive accessibility.
 - Units of $V_i^S$: accessible opportunities from $i$, a value that is the sum of opportunity supply flows allocated proportionally based on demand at $i$.
@@
 
-4. **Doubly Constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
-- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously, which can also be expressed as proportional allocation factors ($\kappa_{ij}^D$, $\hat \kappa_{ji}^D$); equivalent in interpretation to Wilson's [@wilson1971] doubly constrained spatial interaction model.
-- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in @fig-analytical-device-conc-accessibility).
+4. **Doubly-constrained Case ($V_{ij}^D$, $M_{ji}^D$)**
+- Constrained on both origin and destination sides using both $A_i$ and $B_j$ simultaneously, which can also be expressed as proportional allocation factors ($\kappa_{ij}^D$, $\hat \kappa_{ji}^D$); equivalent in interpretation to Wilson's [@wilson1971] doubly-constrained spatial interaction model.
+- Simultaneous application ensures both the destination-side _and_ origin-side marginal totals are maintained (blue and red boxes in Figure 2).
 - Interpretable only as $ij$ and $ji$ flows, since aggregation at $i$ and $j$ simply reproduces known totals.
@@
 
-As a summary, each member of the family of accessibility measure is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
+As a summary, each member of the family of accessibility measures is named, explained in plain language, alongside their balancing factor(s), proportional allocation factor(s), and mathematical equation and value interpretations in Table 1.
 
@@
 
-Consider the simple 3-zone region in @fig-analytical-device-conc-accessibility, where each zone serves as both origin ($i$) and destination ($j$).
+Consider the simple 3-zone region in Figure 2, where each zone serves as both origin ($i$) and destination ($j$).
 The system includes three inputs: zonal population and opportunities, a zonal cost (travel time) matrix, and three travel impedance functions representing different travel behaviours.
@@
 First, @tbl-small-system-land-use summarizes population (in 10,000s) and physicians per zone.
-For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)` comparable to Canada's 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
+For context, the Provider-to-Population Ratio (PPR) is `r sum(LU$D_j) / sum(LU$O_i)`, comparable to Canada's 2022 PPR of 24.97 physicians per 10,000 [@whoMedicalDoctors102025].
 Second, @tbl-small-system-cost shows travel times (minutes); it can be discerned that Zones 1 and 3 are closer to each other than to Zone 2.
@@
 
-In $V^0_i$, no proportional allocation factor is defined, simply $f(c_{ij})$ is used to weight the number of opportunities at each $j$ and the weighted values for each $j$ are summed for each $i$, yielding the an expression identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
+In $V^0_i$, no proportional allocation factor is defined; simply $f(c_{ij})$ is used to weight the number of opportunities at each $j$ and the weighted values for each $j$ are summed for each $i$, yielding an expression identical to Hansen's accessibility $S_i$ [@hansen1959], the current standard practice in accessibility measurement:
 
@@
 
-However $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are 'opportunities weighted by travel impedance' and lack meaningful scaling or direct interpretability.
-Comparing values across different decay functions or contexts (i.e., different number of zones) is therefore limited to ordinal statements (more vs.
+However, $\sum_i V^0_{i}$ generally does not equal the total opportunities $O$, so units here are 'opportunities weighted by travel impedance' and lack meaningful scaling or direct interpretability.
+Comparing values across different decay functions or contexts (i.e., different numbers of zones) is therefore limited to ordinal statements (more vs.
 less), not intervals or ratios (i.e., the magnitude of differences).
@@
 For instance, the difference in zone 1 (edge of urban core)'s accessibility under $f_3$ vs $f_1$ is `r round(unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1], 2)`, but in what units?
-These two values are a product of different impedance functions _physicians-minute$^{-0.1}$_ and _physicians-minute$^{-3}$_), making the direct comparison uninterpretable (and arguably incorrect).
+These two values are a product of different impedance functions ($\text{physicians-minute}^{-0.1}$ and $\text{physicians-minute}^{-3}$), making the direct comparison uninterpretable (and arguably incorrect).
 The fundamental uninterpretability of what is an _opportunity-weighted-travel-impedance_ unit remains.
@@
 As the different impedance functions represent different travel behaviours, comparing the raw unconstrained accessibility values across groups is meaningless beyond notions of higher or lower.
-While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select impedance functions to facilitate comparison across scenarios (potentially at the expense of accurately reflecting travel behavior), such adjustments may introduce bias.
+While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select impedance functions to facilitate comparison across scenarios (potentially at the expense of accurately reflecting travel behaviour), such adjustments may introduce bias.
 The unconstrained scores are best used for ranking within a single context.
@@
 
-## 7.3. Total constrained accessibility
-
-The total constrained accessibility case can be interpreted in a few ways.
+## 7.3. Total-constrained accessibility
+
+The total-constrained accessibility case can be interpreted in a few ways.
 In the one that connects to the status quo: the total balancing factor proportionally adjusts unconstrained zonal accessibility values $V^0_i$ so their total sum of $V^0_i$ matches a known system total -- either total opportunities or total population.
-Another interpretation is in reformulating the equation to use a a proportional allocation constant based on the total balancing factor.
+Another interpretation is in reformulating the equation to use a proportional allocation constant based on the total balancing factor.
 The proportional allocation constant distributes opportunities (or population) proportionally by travel impedance.
 
-In both formulations, all zonal values become a proportion of a known system total, be it the regional opportunities or regional population depending on the variant.
+In both formulations, all zonal values become a proportion of a known system total, be it the regional opportunities or regional population, depending on the variant.
 
@@
 
-- $V_i^T$: accessibility is constrained by the total number of opportunities (total constrained accessible opportunity) and which is interpreted as Hansen's accessibility with a constraining constant, and
-- $M_j^T$: where $i$ and $j$ of the first variant is transposed, yielding a measure constrained by the total population and to be interpreted as constrained 'market potential'.
-
-### 7.3.1. Total constrained accessible opportunities: Hansen's accessibility with a total constraint
-In the total constrained case, accessibility is expressed as a share of the total number of opportunities in the region $D$, allocated based on travel impedance.
-The total constrained accessibility from $i$ to $j$ takes the form:
+- $V_i^T$: accessibility is constrained by the total number of opportunities (total-constrained accessible opportunity), which is interpreted as Hansen's accessibility with a constraining constant, and
+- $M_j^T$: where $i$ and $j$ of the first variant are transposed, yielding a measure constrained by the total population and to be interpreted as constrained 'market potential'.
+
+### 7.3.1. Total-constrained accessible opportunities: Hansen's accessibility with a total constraint
+In the total-constrained case, accessibility is expressed as a share of the total number of opportunities in region $D$, allocated based on travel impedance.
+The total-constrained accessibility from $i$ to $j$ takes the form:
 $$
@@
 
-Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $K^T$, which scales Hansen's unconstrained accessibility $V_i^0$ to meet the total opportunity constraint:
+Alternatively, this formulation can be rewritten to be expressed using a total-constrained balancing factor $K^T$, which scales Hansen's unconstrained accessibility $V_i^0$ to meet the total opportunity constraint:
 $$
@@
 
-Where the total constrained balancing factor $K^T$ is:
+Where the total-constrained balancing factor $K^T$ is:
 $$
@@
 In summary, $\kappa_{ij}^T$ proportionally allocates the total number of opportunities $D$ to each origin–destination pair based on relative opportunity-weighted travel impedance.
-These values can be aggregated across destinations to obtain total constrained accessibility at each origin.
+These values can be aggregated across destinations to obtain total-constrained accessibility at each origin.
 Alternatively, the measure can be expressed using the balancing factor  $K^T$, demonstrating that it is algebraically proportional to unconstrained accessibility $V_i^0$, but with interpretable units (i.e., opportunities).
@@
 $K^T$ for other decay scenarios are calculated similarly in code.
-Applying each $K^T$ to the unconstrained values $V_i^0$ yields total constrained accessibility values (@tbl-simple-example-total-opportunity-accessibility), all in units of physicians.
+Applying each $K^T$ to the unconstrained values $V_i^0$ yields total-constrained accessibility values (@tbl-simple-example-total-opportunity-accessibility), all in units of physicians.
 
@@
 #| label: tbl-simple-example-total-opportunity-accessibility
-#| tbl-cap: "Simple system: total constrained accessible opportunities."
+#| tbl-cap: "Simple system: total-constrained accessible opportunities."
 
@@
 For example, in the highest decay case, Zone 1 (Urban Edge) captures an intermediate number of physicians (`r tot_acc$V_tot_i_1[1] |> round(digits = 3)`), like in the unconstrained accessibility case.
-However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum() |> round(digits = 3)` physicians in the region, which allows us also to deduce that zone 1 captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]) |> round(digits = 3)` and `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]) |> round(digits = 3)` times more than zone 2 and 3.
-Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
+However, unlike in the unconstrained case, we can say that this value is out of the `r tot_acc$V_tot_i_1 |> sum() |> round(digits = 3)` physicians in the region, which allows us to deduce that zone 1 captures `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]) |> round(digits = 3)` and `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]) |> round(digits = 3)` times more than zone 2 and 3.
+Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$) scenarios are calculated separately, with decay scenario values also summing to `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region.
 
 One can also directly compare values at a specific zone, across travel impedance scenarios, due to the consistent units.
-As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminishes (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
+As the decay scenario decreases, all zones become more accessible to each other and the differences between pairs diminish (i.e., in $f_3(c_{ij})$ each zone captures close to an average amount of physicians, a third of 490 or ~163).
 In terms of proportional magnitude, this can also be observed in the unconstrained measure for this scenario.
-However, for the total constrained measure, this plateauing of results has meaning.
-In fact, each zone is allocated an average of the total amount in the region, as a result of the total constrained proportional allocation factor.
+However, for the total-constrained measure, this plateauing of results has meaning.
+In fact, each zone is allocated an average of the total amount in the region, as a result of the total-constrained proportional allocation factor.
 
@@
 
-The total constrained accessibility measure resolves the interpretability issue of Hansen's accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total, enabling robust comparisons across zones and scenarios but also by keeping values proportional to $V_i^0$ so interpretation is similar.
-
-### 7.3.2. Total constrained accessible population: Reilly's potential trade territories with a total constraint
-
-This second total constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield a measure of market potential -- the number of people who can spatially interact with a destination.
+The total-constrained accessibility measure resolves the interpretability issue of Hansen's accessibility (i.e., unconstrained accessibility) by grounding values in a meaningful total, enabling robust comparisons across zones and scenarios, but also by keeping values proportional to $V_i^0$, so interpretation is similar.
+
+### 7.3.2. Total-constrained accessible population: Reilly's potential trade territories with a total constraint
+
+This second total-constrained variant is a transpose of the opportunity-constrained formulation, switching indices $i$ and $j$ to yield a measure of market potential -- the number of people who can spatially interact with a destination.
 
 Though not outlined in the "Unconstrained accessibility" section, the unconstrained form aligns with Reilly's "potential trade territories" [@reilly1929methods] and Harris' and Vickerman's formulations of regional market potential [@harris_market_1954; @vickermanAccessibilityAttractionPotential1974].
-In its unconstrained form, market potential has also been more recently to estimate potentially accessible populations following infrastructure investments [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
+In its unconstrained form, market potential has also been more recently used to estimate potentially accessible populations following infrastructure investments [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road].
 Market potential can also be thought of as a form of _passive accessibility_, indicating the number of people that can reach each destination.
@@
 However, like $V_{ij}^0$, issues of unit interpretability arise in unconstrained market potential $M_j^0$.
-To address this, we introduce the total constrained accessible population measure $M^T_{ji}$, which allocates the total population $O$ across all origin-destination pairs proportionally:
+To address this, we introduce the total-constrained accessible population measure $M^T_{ji}$, which allocates the total population $O$ across all origin-destination pairs proportionally:
 
@@
 
-Next, $\hat \kappa_{ij}$ is the total constrained proportional allocation factor, a dimensionless term which distributes population based on impedance-weighted accessibility: 
+Next, $\hat \kappa_{ij}$ is the total-constrained proportional allocation factor, a dimensionless term which distributes population based on impedance-weighted accessibility: 
 $$
@@
 This renders $M_j^T$ (equal to $\sum_jM_{ji}^T$) into units of population.
-Alternatively, this formulation can be rewritten to be expressed using a total constrained balancing factor $\hat K^T$, which scales unconstrained market potential $M_j^0$ to meet the total population constraint:
+Alternatively, this formulation can be rewritten to be expressed using a total-constrained balancing factor $\hat K^T$, which scales unconstrained market potential $M_j^0$ to meet the total population constraint:
 $$
@@
 
-Where the total constrained balancing factor $\hat K^T$ is:
+Where the total-constrained balancing factor $\hat K^T$ is:
 $$
@@
 #| label: tbl-simple-example-total-population-accessibility
-#| tbl-cap: "Simple system: total constrained accessible population."
+#| tbl-cap: "Simple system: Total-constrained accessible population."
 
@@
 
-In @tbl-simple-example-total-opportunity-accessibility, zone 1, 2, 3 represent destinations and the accessibility values reflect the number of accessible people from the vantage of physicians.
-Zone 1, in its role as a destination, is no longer intermediately-ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
-However, similar to the total constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
-As decay reduces, the population's travel impedance to all zones become more similar, making the relative location of the zones less important and all people in the region more equally accessible.
-
-Like in the total constrained accessible opportunities variant, the total constrained accessible population enables direct comparison of raw values, supporting both ordinal and interval interpretations across space and travel behaviour scenarios.
-
-## 7.4. Singly constrained accessibility
-
-The singly constrained accessibility case can also be expressed in two variants, each defined by the direction in which a constraint is applied:
-
-- $V_i^S$: accessibility constrained by opportunities at destinations (singly constrained accessible opportunities), and
-- $M_j^S$: its transpose, constrained by population at origins (singly constrained accessible population, or market potential).
-
-Similar to the total constrained case, the singly constrained measures adjust unconstrained zonal accessibility values ($V_i^0$ or $M_j^0$) using a balancing factor to satisfy the known system constraint.
-However, unlike the total constraint (which enforces a regional sum), the singly constrained case applies a localized constraint at one end of the interaction -- either origin or destination.
+In @tbl-simple-example-total-opportunity-accessibility, zones 1-3 represent destinations, and the accessibility values reflect the number of accessible people from the vantage of physicians.
+Zone 1, in its role as a destination, is no longer intermediately ranked relative to other zones; it now attracts the fewest number of people across all three travel behaviour scenarios.
+However, similar to the total-constrained opportunity case, as travel decay reduces, the availability of population begins to converge (though Zone 1 continues as the lowest-ranked) for similar reasons.
+As decay reduces, the population's travel impedance to all zones becomes more similar, making the relative location of the zones less important and all people in the region more equally accessible.
+
+Like in the total-constrained accessible opportunities variant, the total-constrained accessible population enables direct comparison of raw values, supporting both ordinal and interval interpretations across space and travel behaviour scenarios.
+
+## 7.4. Singly-constrained accessibility
+
+The singly-constrained accessibility case can also be expressed in two variants, each defined by the direction in which a constraint is applied:
+
+- $V_i^S$: accessibility constrained by opportunities at destinations (singly-constrained accessible opportunities), and
+- $M_j^S$: its transpose, constrained by population at origins (singly-constrained accessible population, or market potential).
+
+Similar to the total-constrained case, the singly-constrained measures adjust unconstrained zonal accessibility values ($V_i^0$ or $M_j^0$) using a balancing factor to satisfy the known system constraint.
+However, unlike the total constraint (which enforces a regional sum), the singly-constrained case applies a localized constraint at one end of the interaction -- either origin or destination.
 
@@
 
-In both cases, the singly constrained formulation introduces zonal-level competition, unlike the total constrained case which distributes a fixed regional sum.
+In both cases, the singly-constrained formulation introduces zonal-level competition, unlike the total-constrained case, which distributes a fixed regional sum.
 Each zonal accessibility value becomes not only a fraction of the regional total (opportunities or population), but also a balanced sum of interactions, weighted by impedance and relative competition.
-The result remains in interpretable units -- accessible opportunities or accessible population -- but reflect these more complex spatial dynamics.
-
-### 7.4.1. Singly constrained accessible opportunities: spatial availability
-
-In this singly constrained variant, accessibility is constrained at the destination side: the sum of accessible opportunities allocated from each destination must equal the known number of opportunities $D_j$.
+The result remains in interpretable units -- accessible opportunities or accessible population -- but reflects these more complex spatial dynamics.
+
+### 7.4.1. Singly-constrained accessible opportunities: spatial availability
+
+In this singly-constrained variant, accessibility is constrained at the destination side: the sum of accessible opportunities allocated from each destination must equal the known number of opportunities $D_j$.
 This is comparable to the single attraction-constraint (@eq-constraint2-gravitymodel) from Wilson's framework:
@@
 
-The underlying spatial interaction model is now the attraction-constrained model and our accessibility measure becomes:
+The underlying spatial interaction model is now the attraction-constrained model, and our accessibility measure becomes:
 
@@
 
-Further, we can express the formula even more simply, by defining the following proportional allocation factor:
+Further, we can express the formula even more simply by defining the following proportional allocation factor:
 
@@
 The dimensionless factor $\kappa^S_{ij}$ ensures that each destination's opportunities are distributed proportionally to origins.
-As in the total constrained case, $V_i^S$ is expressed in the units of accessible opportunities.
+As in the total-constrained case, $V_i^S$ is expressed in the units of accessible opportunities.
 
@@
 For the $f_3(c_{ij})$ decay group, they are  `r B_j$B_j_3[1]  |> round(digits = 3)`, `r B_j$B_j_3[2]  |> round(digits = 3)` and `r B_j$B_j_3[3]  |> round(digits = 3)`.
-Using these these balancing constants, we can calculate the singly constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).
+Using these balancing constants, we can calculate the singly-constrained opportunity accessibility (@tbl-simple-example-attraction-constrained-accessibility).
 
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility
-#| tbl-cap: "Simple system: singly constrained accessible opportunities."
+#| tbl-cap: "Simple system: singly-constrained accessible opportunities."
 
@@
 
-Imposing the single proportional allocation factor $\kappa^S_{ij}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total constrained accessible opportunities case.
+Imposing the single proportional allocation factor $\kappa^S_{ij}$ allows for the comparison of differences and ratios of the accessibility values, like previously discussed in the total-constrained accessible opportunities case.
 The proportional allocation factor ensures that resulting values are in units of _physicians_, with the impedance units already accounted for in the allocation process.
@@
 However, unlike $\kappa^T_{ij}$, $\kappa^S_{ij}$ introduces zonal competition based on the mass of the origin (population).
-In the total constrained case, opportunities are distributed based on impedance alone, regardless of population at $i$.
-In contrast, the singly constrained case allocates each zone's opportunities proportionally across the region based on the relative impedance-weighted demand from all origins.
+In the total-constrained case, opportunities are distributed based on impedance alone, regardless of population at $i$.
+In contrast, the singly-constrained case allocates each zone's opportunities proportionally across the region based on the relative impedance-weighted demand from all origins.
 
@@
 Its low impedance-weighted demand means $\kappa^S_{ij}$ allocates it fewer opportunities.
-By contrast, in the total constrained case, Zone 1 fares better, capturing  `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` of all physicians (compared to `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()`).
+By contrast, in the total-constrained case, Zone 1 fares better, capturing  `r (tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1 |> sum()) |> scales::percent()` of all physicians (compared to `r (opp_acc$V_opp_i_1[1]/opp_acc$V_opp_i_1 |> sum()) |> scales::percent()`).
 
@@
 This dynamic reveals how $\kappa^S_{ij}$ embeds both travel impedance and population competition.
-Unlike the total constraint lowest decay scenarios that allocate evenly (as localised supply and demand is assumed to be unknown), the singly constrained case reflects competition's influence on allocation.
+Unlike the total constraint lowest decay scenarios that allocate evenly (as localized supply and demand are assumed to be unknown), the singly-constrained case reflects competition's influence on allocation.
 
@@
 But, as we introduced constraints, these constrained accessibility values can be normalized using anything that is relevant to the zone.
-In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example simply in units of number of physicians accessible per population at each zone.
+In @tbl-simple-example-attraction-constrained-accessibility-per-capita, we present per capita accessibility for the numeric example simply in units reflecting the number of physicians accessible per population at each zone.
 Notably, these per capita rates are equivalent to the 2SFCA values.
@@
 #| label: tbl-simple-example-attraction-constrained-accessibility-per-capita
-#| tbl-cap: "Simple system: singly constrained accessible opportunities per capita."
+#| tbl-cap: "Simple system: Singly-constrained accessible opportunities per capita."
 
@@
 As distance decay decreases and becomes _relatively_ uniform (all zones can reach all zones), the effect of population drives the proportional allocation of opportunities.
-Consequently, per capita accessibility values begin to stabilise to the regional per capita average (e.g., in the highest distance decay $f_1(c_{ij})$, per capita values are all ~24 physicians accessible per capita).
-
-This convergence mirrors the trend in the total constrained opportunity case, where accessibility values approach a third of the 490 physicians under the lowest-decay unfettered mobility scenario $f_3(c_{ij})$.
+Consequently, per capita accessibility values begin to stabilize to the regional per capita average (e.g., in the highest distance decay $f_1(c_{ij})$, per capita values are all ~24 physicians accessible).
+
+This convergence mirrors the trend in the total-constrained opportunity case, where accessibility values approach a third of the 490 physicians under the lowest-decay unfettered mobility scenario $f_3(c_{ij})$.
 In both cases, the balancing factors ($K^S$ and $B_j$) act as averaging mechanisms but at different scales.
-As distance decay becomes more _relatively_ uniform, the role of remaining variables (i.e., total population or opportunities) drive the proportional allocation differences.
-In the total constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constrained case, this is the population at a zone relative to the regional population.
-
-### 7.4.2. Singly constrained accessible population: market availability
-
-Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
+As distance decay becomes _relatively_ more uniform, the role of remaining variables (i.e., total population or opportunities) drive the proportional allocation differences.
+In the total-constrained case, this is the proportion of opportunities relative to the regional opportunities, and in the case of the single opportunity constrained case, this is the population at a zone relative to the regional population.
+
+### 7.4.2. Singly-constrained accessible population: market availability
+
+Similar to @eq-total-population-balancing-factor in transposing the origins and destinations, we can define a *singly-constrained* measure of market potential that preserves the known population (i.e., the mass weight at the origin $W_i^{(1)}$ is now represented by $O_i$).
 In its per-capita expression, i.e., equivalent to 2SFCA, this constrained concept of market potential has been used to express "facility crowdedness" as in Wang [@wang_inverted_2018].
@@
 
-In this variant, the measure is singly constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
+In this variant, the measure is singly-constrained by the population *by origin* (i.e., $O_i$), like @eq-constraint2-gravitymodel from Wilson's framework:
 
@@
 
-Following the same logic as in the preceding section on total constrained market potential, one arrives at the following expression of accessible population $M_j^S$ being the product of proportionally allocated ($\hat \kappa^S_{ji}$) population:
+Following the same logic as in the preceding section on total-constrained market potential, one arrives at the following expression of accessible population $M_j^S$ being the product of proportionally allocated ($\hat \kappa^S_{ji}$) population:
 
@@
 
-As well, the single (population) constraint in @eq-population-constraint ensures that the the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O$) is maintained.
+As well, the single (population) constraint in @eq-population-constraint ensures that the total constraint (e.g., $\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O$) is maintained.
 
@@
 
-For the sake of brevity, we'll move onto the doubly constrained case.
-
-## 7.5. Doubly constrained accessibility
-
-This accessibility case adopts the structure of the doubly constrained spatial interaction model, where $V_{ij}^D$ flows are constrained by both origin populations $O_i$ and destination opportunities $D_j$.
+For the sake of brevity, we'll move on to the doubly-constrained case.
+
+## 7.5. Doubly-constrained accessibility
+
+This accessibility case adopts the structure of the doubly-constrained spatial interaction model, where $V_{ij}^D$ flows are constrained by both origin populations $O_i$ and destination opportunities $D_j$.
 That is, the resulting accessibility outflow from each origin must match the origin's population demand, and the resulting accessibility inflow to each destination must match the number of opportunities supplied:
@@
 
-This distinguishes doubly constrained accessibility from the total and singly constrained cases discussed previously.
-In those cases, only one side of the interaction -- either the total marginal or opportunity/population marginal -- was constrained, while the other was treated as a demand/supply weight (e.g., $D$ or $O$ for total constrained and  $W_j^{(2)}$ or $W_i^{(1)}$ for singly constrained).
-
-By contrast, the doubly constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly.
-This makes it less suitable for traditional accessibility analysis; namely because origin and destination masses often differ in kind and units.
-For instance, the number of people accessing an opportunity such as park may be known, but the capacity of each park is not.
-A doubly constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school-seat, or vaccine doses per person).
-Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):
+This distinguishes doubly-constrained accessibility from the total and singly-constrained cases discussed previously.
+In those cases, only one side of the interaction -- either the total marginal or opportunity/population marginal -- was constrained, while the other was treated as a demand/supply weight (e.g., $D$ or $O$ for total-constrained and  $W_j^{(2)}$ or $W_i^{(1)}$ for singly-constrained).
+
+By contrast, the doubly-constrained model assumes both the demand (population) and the supply (opportunity) are known and bounded, and allocates flows accordingly.
+This makes it less suitable for traditional accessibility analysis, namely because origin and destination masses often differ in kind and units.
+For instance, the number of people accessing an opportunity, such as a park, may be known, but the capacity of each park is not.
+A doubly-constrained approach may make conceptual sense if the "potential" should be contained only in the flows themselves, meaning the units of population and opportunities are comparable, have a one-to-one correspondence or are otherwise paired (e.g., job per worker, student per school seat, or vaccine doses per person).
+Mathematically, this model requires the simultaneous imposition of both the population- and opportunity- constraints in the preceding singly-constrained variants (@eq-opportunity-constraint and @eq-population-constraint), namely the sum of population in all origins should match the sum of opportunities in all destinations (@eq-opportunity-population-equality-2):
 
@@
 
-The doubly constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly constrained) spatial interaction model as follows:
+The doubly-constrained accessibility measure $V_{ij}^D$ takes the form of the production-attraction (doubly-constrained) spatial interaction model as follows:
 
@@
 
-The doubly constrained model completely distributes origin populations to destination opportunities according to travel impedance and supply-demand balance.
+The doubly-constrained model completely distributes origin populations to destination opportunities according to travel impedance and supply-demand balance.
 This ensures that: summing $V^D_{ij}$ across $j$ returns $O_i$; summing across $i$ returns $D_j$.
@@
 
-Following this logic, the market potential form $M^D_{ji}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: the opportunities accessed from $j$ at an $i$ vs.
+Following this logic, the market potential from $M^D_{ji}$ is effectively equivalent to $V_{ij}^D$, but can be read with a different interpretation: the opportunities accessed from $j$ at an $i$ vs.
 the population accessed from $i$ at a $j$.
@@
 
-To calculate doubly constrained accessibility using the toy example, the interpretation of the population data and the counts of the opportunity data must be reworked.
+To calculate doubly-constrained accessibility using the toy example, the interpretation of the population data and the counts of the opportunity data must be reworked.
 Namely, a count of physician _capacity_ per destination $D_j$ is needed instead of simply the number of physicians.
-We need to also be able to clearly state that the population is the _capacity_ of the origin to interact with opportunities $O_i$, i.e., the count of people seeking opportunities.
+We also need to be able to clearly state that the population is the _capacity_ of the origin to interact with opportunities $O_i$, i.e., the count of people seeking opportunities.
 
@@
 With the population (in units of 10,000s of people seeking physicians) and the opportunities (in units of 10,000s of physician-capacity) per zone.
-For the population, we leave this unchanged numerically but we now must keep in mind that each person interacts with one physician capacity.
-The number of providers per destination is however revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).
+For the population, we leave this unchanged numerically, but we now must keep in mind that each person interacts with one unit of physician capacity.
+The number of providers per destination is, however, revised to represent physician capacity, scaled approximately from the original number of physicians used in previous cases (@tbl-small-system-land-use).
 The system-wide PPR is now `r sum(LU2$D_j2) / sum(LU2$O_i2)` (recall: the unmodified example's system PPR is `r sum(LU$D_j) / sum(LU$O_i)`).
 
-We keep the same zonal cost matrix, and travel impedance functions for three types of travel behaviour as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
+We keep the same zonal cost matrix and travel impedance functions for three types of travel behaviour as before (@tbl-small-system-cost and @eq-travel-behaviour-scenarios).
 
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values
-#| tbl-cap: "Doubly constrained accessible opportunities assuming highest travel decay in the modified simple system."
+#| tbl-cap: "Doubly-constrained accessible opportunities assuming highest travel decay in the modified simple system."
 
@@
 
-As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints i.e., $V^D_i = O_i$ hence the physician-capacity accessible for Zones 1, 2, and 3 are `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
-The usefulness of the doubly constrained measure lies in the interpretation as $V_{ij}^D$ values; the number of opportunities from zone $j$ allocated to populations in zone $i$ shaped by both mass and travel impedance.
+As shown in @tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values for the highest-decay scenario $f_1(c_{ij})$, accessibility is no longer meaningfully represented as zonal summaries like $V^D_i$ or $M^D_j$, since these values reproduce the original constraints, i.e., $V^D_i = O_i$, hence the physician-capacity accessible for Zones 1, 2, and 3 are `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[1,2] |> pull() |> round(digits = 3)`, `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[2,2] |> pull() |> round(digits = 3)`, and `r (OD2|>group_by(oid)|>summarize(V_dc_i_1=sum(V_dc_ij_1)))[3,2] |> pull() |> round(digits = 3)`.
+The usefulness of the doubly-constrained measure lies in the interpretation as $V_{ij}^D$ values; the number of opportunities from zone $j$ allocated to populations in zone $i$ is shaped by both mass and travel impedance.
 
@@
 
-Following the intuition discussed in the singly constrained opportunity case, as decay decreases, the mass effects (effect of the population and opportunities magnitudes) become more relatively dominant in the spatial allocation.
+Following the intuition discussed in the singly-constrained opportunity case, as decay decreases, the mass effects (effect of the population and opportunities magnitudes) become relatively more dominant in the spatial allocation.
 
@@
 #| label: tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2
-#| tbl-cap: "Doubly constrained accessible opportunities at Zone 2 for all travel decay groups in the modified simple system."
+#| tbl-cap: "Doubly-constrained accessible opportunities at Zone 2 for all travel decay groups in the modified simple system."
 
@@
 
-Accessibility is conventionally presented as a zonal value not a flow.
-However, in the doubly constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matches our known population at $i$.
-Following the logic of the family of accessibility measures, in the doubly constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
+Accessibility is conventionally presented as a zonal value, not a flow.
+However, in the doubly-constrained case, since we force the allocation of zonal population demand and zonal opportunities supplied to be paired and allocation to be proportional, $V_i^D$ is simply the number of opportunities that matche our known population at $i$.
+Following the logic of the family of accessibility measures, in the doubly-constrained case, $V_{ij}^D$ flows are the only relevant unit of analysis: spatial proportional allocations between population and opportunity capacity.
 Furthermore, $V^D_{ij}$ and its transposed counterpart $M^D_{ji}$ are structurally identical, differing only in interpretation (referring to $\kappa_{ij}^D$ and $\hat \kappa_{ij}^D$): one reflects the proportional allocation of opportunity to population flows; the other, population to opportunity flows.
@@
 $V_{ij}^D$ are also mathematically equivalent to Wilson's spatial interaction flows.
-And as Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
+And, as Wilson [@wilson1971] explicitly noted, origin and destination weights defined in the spatial interaction model *can* be defined using any unit.
 However, the focus of these models has typically been on $ij$ flows, often calibrated using trips (i.e., outbound and inbound trips, inherently in the same units).
@@
 For instance, we may not know how much park space, grocery area, or childcare capacity is accessible per person.
-When they do align, such as people to physician capacity, we would be modeling realised access flows based on known quantities of _those that interact_ and the _interacted_.
+When they do align, such as people to physician capacity, we would be modeling realized access flows based on known quantities of _those that interact_ and the _interacted_.
 In such cases, the traditional accessibility question can be seen to be already answered by the known information (i.e., how many opportunities can be reached by a zone?
 Answer: the number of people at that zone).
-For this reason, we do not foresee the doubly constrained measure being widely used in accessibility analysis as the literature has largely focused on questions of 'potential' not on predicting flows of realised access.
+For this reason, we do not foresee the doubly-constrained measure being widely used in accessibility analysis, as the literature has largely focused on questions of 'potential', not on predicting flows of realized access.
 
@@
 
-We then introduced the singly constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
+We then introduced the singly-constrained case, which *does* take into account the population size at the origin in the allocation of opportunities.
 It is also mathematically equivalent to the spatial availability introduced in Soukhov et al.
@@
 In this case, all accessibility values are fixed to sum to a known zonal opportunity-size value (implicitly, the regional total of opportunities), but they are not required to sum to any population-based values at the zone or regional level.
-The singly constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
+The singly-constrained model could be useful if regional competition is a factor and if the acknowledgment that only a finite number of opportunities can be allocated from each destination (with those allocations distributed based on origin population size) is suitable.
 We also introduce an 'accessible' PPR (e.g., opportunities per capita), calculated by dividing each accessibility value by the zonal population.
-To clarify, this per capita expression of the singly constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
-
-Lastly, the doubly constrained case is introduced.
+To clarify, this per capita expression of the singly-constrained case is equivalent to the 2SFCA [@luo2003; @shen1998], hence linking this literature back to spatial interaction principles.
+
+Lastly, the doubly-constrained case is introduced.
 In this case, the sums must equal both the regional total and ensure that no zone allocates more opportunities than it has available.

### 2025-10-23 paper_plos/paper_plos_V3.tex
cdh final proofread for v3

@@
 remains fragmented, with inconsistent definitions and unclear
-interpretations. To this end, this paper offers a methodological
+interpretations. To this end, this paper makes a methodological
 contribution by specifying a family of accessibility measures that are
@@
 interaction models, particularly their balancing factors. From this
-foundation, we outline four memebers of the family: the `unconstrained'
-measure (i.e., Hansen-type accessibility), the `total constrained'
+foundation, we outline four members of the family: the `unconstrained'
+measure (i.e., Hansen-type accessibility), the `total-constrained'
 measure (i.e., a constrained version of the Hansen-type accessibility),
-the `singly constrained' measure (i.e., related to the popular two step
-floating catchment approach -- 2SFCA), and the `doubly constrained'
-measure representing realized access (i.e., equal to the doubly
-constrained spatial interaction model). These measures can be
+the `singly-constrained' measure (i.e., related to the popular two-step
+floating catchment approach -- 2SFCA), and the `doubly-constrained'
+measure representing realized access (i.e., equal to the
+doubly-constrained spatial interaction model). These measures can be
 interpreted as either the number of accessible opportunities or
@@
 \linenumbers
+
 \section{1. Introduction}\label{introduction}
@@
 accessibility measures grounded in spatial interaction principles,
-including total, singly, and doubly constrained cases and variants of
+including total, singly, and doubly-constrained cases and variants of
 ``accessible opportunities'' and ``accessible population''. These cases
@@
 \citep{ravensteinLawsMigration1885} to Stewart (1948)
-\citep{stewartDemographicGravitationEvidence1948} who theorised and
+\citep{stewartDemographicGravitationEvidence1948} who theorized and
 formalized spatial interaction patterns (Section 3), and examining how
@@
 the introduction of ``constraints'' (based on top-down known information
-as part of entropy-maximisation) produce interpretable, unit-consistent
+as part of entropy-maximization) produces interpretable, unit-consistent
 flow estimates. Section 6 explores why accessibility and spatial
@@
 contribution: we derive a family of accessibility measures corresponding
-to different constraints applied -- unconstrained, singly constrained,
-and doubly constrained cases -- demonstrating how each yields zone-level
+to different constraints applied -- unconstrained, singly-constrained,
+and doubly-constrained cases -- demonstrating how each yields zone-level
 accessibility values expressed in meaningful units (opportunities or
@@
 estimate of \(G\) was based on a speculation but received empirical
-support after Hutton's and Cavendish' experiments in the late 1700s
-\citep{hutton_xxxiii_1778, cavendish_xxi_1798} which estimated it to
+support after Hutton's and Cavendish's experiments in the late 1700s
+\citep{hutton_xxxiii_1778, cavendish_xxi_1798}, which estimated it to
 within 1\% accuracy. That is, it took over a century from the
@@
 interaction research in different contexts. Namely, a number of
-researchers theoretically and empirically attempted to characterise
+researchers theoretically and empirically attempted to characterize
 human spatial interaction as a force \(F\) directly proportional to the
@@
 
-What is notable about Eq~\ref{eq-stewart-force}, however, is the
+What is notable about Eq~\ref{eq-stewart-force}, however, is that the
 proportionality constant \(G\) was specified but ``left for future
@@
 From Stewart \citep{stewartDemographicGravitationEvidence1948}, we
-arrive to 1959 and Walter G. Hansen, whose work proved to be
+arrive at 1959 and Walter G. Hansen, whose work proved to be
 exceptionally influential in the accessibility literature
@@
 \(S_{i}\) in Eq~\ref{eq-accessibility-general} is a measure of the
-accessibility from a zone \(i\). This is a function of \(O_j\) (the mass
+accessibility from zone \(i\). This is a function of \(O_j\) (the mass
 of opportunities at \(j\)), \(d_{ij}\) (the cost, e.g., distance or
@@
 {[}was{]} left for future determination: a suitable choice of other
-units can reduce it to unity'' {[}p.~34{]}.) Hansen made no mention of
+units can reduce it to unity'' {[}p.~34{]}). Hansen made no mention of
 it. As a result, modern accessibility analysis has largely evolved
@@
 Furthermore, working with a proportional relationship generates
-fundamental issues in comparability between and, arguably within,
+fundamental issues in comparability between and, arguably, within
 studies. Namely, accessibility estimates have no fixed unit, rendering
@@
 \(e^{\beta \text{metres}}\). Such variation impairs comparability across
-analysis and obscures the meaning of accessibility scores, making them
+analyses and obscures the meaning of accessibility scores, making them
 difficult to understand and communicate without post-hoc treatment.
 
-Therefore in practice, Hansen-type measures are ones of proportionality
+Therefore, in practice, Hansen-type measures are ones of proportionality
 and are better understood as \emph{ordinal indicators}; they rank
@@
 \citep{hansen1959}, a parallel development was taking place across the
-Atlantic with Alan G. Wilson. Wilson's ground breaking paper
+Atlantic with Alan G. Wilson. Wilson's groundbreaking paper
 \citep{wilson1971} advanced a general framework for spatial interaction
@@
 real-world quantities are not a set empirical constant (like Newton's
-\(G\)) but instead, sensitive to the system and known information about
-the system.
+\(G\)) but are instead sensitive to the system and known information
+about the system.
 
@@
 gravity (i.e., interaction between masses over distance), Wilson's
-innovation was to generalise the model using \emph{entropy
-maximisation}. By maximizing the number of ways individual trip
+innovation was to generalize the model using \emph{entropy
+maximization}. By maximizing the number of ways individual trip
 probabilities could be arranged under known constraints, Wilson derived
@@
 \(W_i^{(1)}\) in Eq~\ref{eq-phys-gravity-model}, represented as \(O_i\),
-then following constraint can be used:
+then the following constraint can be used:
 
@@
 model that only matches the total volume of interaction \(K\), (2) a
-singly constrained model (either by origins \(A_i\) or destinations
-\(B_j\) -- explained below), and (3) a doubly constrained model that
+singly-constrained model (either by origins \(A_i\) or destinations
+\(B_j\) -- explained below), and (3) a doubly-constrained model that
 satisfies both.
@@
 this means that the total number of flows predicted by the model must be
-equal to sum of all flows from origins \(i\) to destinations \(j\). The
-balancing factor \(K\) takes the place of \(k\) and is equal to the
+equal to the sum of all flows from origins \(i\) to destinations \(j\).
+The balancing factor \(K\) takes the place of \(k\) and is equal to the
 following (as specified in \citep{cliff_evaluating_1974} and
@@
 
-In the singly constrained model, only constraint
+In the singly-constrained model, only constraint
 Eq~\ref{eq-constraint1-gravitymodel} or constraint
 Eq~\ref{eq-constraint2-gravitymodel} hold. When only
-Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximisation leads
-to the production-constrained singly constrained version of
+Eq~\ref{eq-constraint1-gravitymodel} holds, entropy maximization leads
+to the production-constrained singly-constrained version of
 Eq~\ref{eq-phys-gravity-model}, where the proxy for the mass at the
 origin \(W_i^{(1)}\) is replaced with \(O_i\). Also, \(k\) is replaced
-with a set of balancing factors specific to origins \(A_i\) which
+with a set of balancing factors specific to origins \(A_i\), which
 ensures that constraint Eq~\ref{eq-constraint1-gravitymodel} is
@@
 
-The singly constrained attraction-constrained model is similar to the
+The singly-constrained attraction-constrained model is similar to the
 production-constrained version but from the perspective of the mass at
@@
 
-Lastly, the doubly constrained model is the production-attraction
+Lastly, the doubly-constrained model is the production-attraction
 constrained model in Wilson \citep{wilson1971}. In this case, both
@@
 constraints ensure that the sum of predicted flows from one origin to
-all destination, and the predicted flows going to one destination from
-all origins must equal the known mass of the origin \(O_i\) and of the
+all destinations, and the predicted flows going to one destination from
+all origins, must equal the known mass of the origin \(O_i\) and of the
 destination \(D_j\). This should hold for all origins and destinations.
-The resulting model is, in Wilson's terms, doubly constrained, and from
+The resulting model is, in Wilson's terms, doubly-constrained, and from
 Eq~\ref{eq-phys-gravity-model}, \(k\) becomes both \(A_i\) and \(B_j\)
@@
 Wilson's models calibrate interaction flows through known empirical
-constraints using principles of entropy maximisation. This results in
+constraints using principles of entropy maximization. This results in
 interpretable, balanced (given the system knowns are expressed in the
@@
 
-Before demonstrating the derivation of family of accessibility measures
-that are sensitive to constraints (Section 7), in the next section we
-review how conceptually intertwined the spatial interaction modelling
-and accessibility literatures are, and where they began to diverge. This
-investigation sheds light on why accessibility research may have failed
-to adopt a comparable approach, until this paper.
+Before demonstrating the derivation of the family of accessibility
+measures that are sensitive to constraints (Section 7), in the next
+section, we review how conceptually intertwined the spatial interaction
+modelling and accessibility literatures are, and where they began to
+diverge. This investigation sheds light on why accessibility research
+may have failed to adopt a comparable approach until this paper.
 
@@
 streams}\label{accessibility-and-spatial-interaction-modelling-two-divergent-research-streams}
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{paper_plos_V3_files/figure-pdf/fig-docs-per-year-1.pdf}
-
-}
-
-\caption{\label{fig-docs-per-year}Historical pattern of publication:
-documents per year.}
-
-\end{figure}%
 
@@
 equity, land use analysis, and urban accessibility planning. Meanwhile,
-Wilson's \citep{wilson1971} entropy-maximising framework reshaped how
+Wilson's \citep{wilson1971} entropy-maximizing framework reshaped how
 spatial interaction models were constructed, particularly in
@@
 frameworks emerged. As noted in
-\citep{battyChronicleScientificPlanning1994}, large scale spatial
+\citep{battyChronicleScientificPlanning1994}, large-scale spatial
 interaction models (like Wilson's) responded to important developments
@@
 for strategic land use and transportation planning''. And these needs
-were far from trivial: for instance in the U.S., the Federal-Aid Highway
-Act of 1956 set in motion the construction of the Interstate Highway
-System with an eventual budget exceeding \$600 billion in today's
-dollars
+were far from trivial: for instance, in the U.S., the Federal-Aid
+Highway Act of 1956 set in motion the construction of the Interstate
+Highway System with an eventual budget exceeding \$600 billion in
+today's dollars
 \citep{weinerUrbanTransportationPlanning2016, mdotMnDOTJoins2007}. In
@@
 and the digital object identifiers of Hansen \citep{hansen1959} and
-Wilson \citep{wilson1971}). Only 76 out of the 2,122 documents that
+Wilson \citep{wilson1971}. Only 76 out of the 2,122 documents that
 emerged from our search of the General Database cite both. The number of
-articles, by year and if they cite Hasen, Wilson, or both are shown in
-Fig~\ref{fig-docs-per-year}.
+articles, by year and if they cite Hasen, Wilson, or both, are shown in
+Figure 1.
 
@@
 these groups of articles, we uncover how the relationship between Wilson
-and Hansen has often been misunderstood, underexplored, or entirely
+and Hansen has often been misunderstood, under-explored, or entirely
 overlooked.
@@
 In the first stream of literature -- which cite both but are focused
-more so on spatial interaction models -- they treat spatial interaction
-and accessibility as separate but related phenomenons. Four subsets of
-this stream emerge.
+more on spatial interaction models -- they treat spatial interaction and
+accessibility as separate but related phenomena. Four subsets of this
+stream emerge.
 
@@
 
-The second subset of articles within this stream use both Hansen
+The second subset of articles within this stream uses both Hansen
 \citep{hansen1959} and Wilson's \citep{wilson1971} framework in
@@
 
-The third subset of the spatial-interaction focused literature depart
+The third subset of the spatial-interaction focused literature departs
 from Hansen's \citep{hansen1959} definition, aligning instead with
@@
 
-Moving onto the group of accessibility-focused literature that cites
-both works, we categorise their citation of Wilson \citep{wilson1971}
-within three general groups. Overall these works do not engage -- or
+Moving on to the group of accessibility-focused literature that cites
+both works, we categorize their citation of Wilson \citep{wilson1971}
+within three general groups. Overall, these works do not engage -- or
 only superficially engage -- with Wilson \citep{wilson1971}.
@@
 context-dependent travel cost functions. This trend is common: for
-instance it is done in the following papers:
+instance, it is done in the following papers:
 \citep{handyMeasuringAccessibilityExploration1997, weibullNumericalMeasurementAccessibility1980, kwan1998space, shenLocationCharacteristicsInnercity1998, ashiru2003space, rau2012spatial, pan2013impacts, margaridacondecomelhoradoImpactMeasuringInternal2016, caschili2015accessibility, grengs2015nonwork, pan2020measuring, chia2020extending, roblot2021participation, sharifiasl2023incorporating, kharel2024examining}.
@@
 of these concepts. Furthermore, some of this literature also
-occasionally conflates or blurs the distinction entirely, for instance
+occasionally conflates or blurs the distinction entirely, for instance,
 by co-citing Hansen and Wilson as being `gravity models'
@@
 Thirdly, there is a group of accessibility-focused works that interprets
-the measure used in Hansen \citep{hansen1959} as the singly- or doubly-
-constrained spatial interaction model's inverse balancing factor
+the measure used in Hansen \citep{hansen1959} as the singly- or
+doubly-constrained spatial interaction model's inverse balancing factor
 \citep[e.g.,][]{vickermanAccessibilityAttractionPotential1974}. This
@@
 \citep{soukhovIntroducingSpatialAvailability2023, soukhovMultimodalSpatialAvailability2024}
-introduce the balancing factors as a mechanisms to ensure that
+introduce the balancing factors as mechanisms to ensure that
 opportunities at each destination are proportionally allocated to each
@@
 were deduced intuitively. These works did not explicitly state that the
-mathematical formulation of the equations are effectively equivalent to
-Wilson's singly constrained model (derived from entropy maximization).
+mathematical formulations of the equations are effectively equivalent to
+Wilson's singly-constrained model (derived from entropy maximization).
 This equivalence is only discovered in hindsight, as will be
@@
 
-In sum, despite the interpretative advantaged offered by the statistical
+In sum, despite the interpretative advantages offered by the statistical
 logic of Wilson \citep{wilson1971}'s framework, neither the spatial
@@
 constraints yields interpretable, unit-consistent measures of
-opportunity. This approach takes the same path of entropy-maximisation
+opportunity. This approach takes the same path of entropy-maximization
 as in Wilson \citep{wilson1971}, and does not rely on specifying some
@@
 Eq~\ref{eq-stewart-population-potential-sum} which Hansen
-\citep{hansen1959} operationalised).
+\citep{hansen1959} operationalized).
 
@@
 unit interpretability (associated with the proportional nature of
-Hansen-type accessibility indicators previous outlined).
+Hansen-type accessibility indicators previously outlined).
 
@@
 spatial interaction with opportunities (or population)}. We can specify
-\(k\) as a type of proportional allocation factor \(\kappa\) which
+\(k\) as a type of proportional allocation factor \(\kappa\), which
 incorporates Wilson's balancing factor(s) to define the \emph{potential
@@
 origin (i.e., population \(D_i\) or \(D\)) for either the zone or full
-region depending on the case (hence represented by with a stand-in \(X\)
-subindex). Accessibility flows can also be summarised as as a partial
-sum of the potential at \(i\) and at \(j\) to express accessibility at
-the origin zone or at the destination zone, respectively. This form is
+region, depending on the case (hence represented by a stand-in \(X\)
+subindex). Accessibility flows can also be summarised as a partial sum
+of the potential at \(i\) and at \(j\) to express accessibility at the
+origin zone or at the destination zone, respectively. This form is
 common in accessibility research:
@@
 
-Figure Fig~\ref{fig-analytical-device-conc-accessibility} illustrates
-our analytical framework using a 3-zone system. The most detailed
-values, \(X_{ij}\), represent the potential for spatial interaction from
-origin \(i\) to destination \(j\). Here, \(X\) stands for all cases and
-variants to be discussed (e.g., \(V_{ij}^0\), \(M_{ji}^0\),
-\(V_{ij}^T\), \(M_{ji}^T\), \(V_{ij}^S\), \(M_{ji}^S\), \(V_{ij}^D\),
-and \(M_{ji}^D\)). Single marginals show the origin and destination
-weights and the total marginal is a sum of these values.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{../figures/access-analytical-device.png}
-
-}
-
-\caption{\label{fig-analytical-device-conc-accessibility}The family of
-accessibility measures analytical framework: labelling and associating
-ij flows, zonal weights, the single marginals, and the total marginal.}
-
-\end{figure}%
+Figure 2 illustrates our analytical framework using a 3-zone system. The
+most detailed values, \(X_{ij}\), represent the potential for spatial
+interaction from origin \(i\) to destination \(j\). Here, \(X\) stands
+for all cases and variants to be discussed (e.g., \(V_{ij}^0\),
+\(M_{ji}^0\), \(V_{ij}^T\), \(M_{ji}^T\), \(V_{ij}^S\), \(M_{ji}^S\),
+\(V_{ij}^D\), and \(M_{ji}^D\)). Single marginals show the origin, and
+destination weights and the total marginal is a sum of these values.
 
@@
   Equivalent to Hansen's \citep{hansen1959} and Reilly's
-  \citep{reilly1929methods} original formulations; the status quo of
+  \citep{reilly1929methods} original formulations, the status quo of
   accessibility modelling.
@@
 \item
-  \textbf{Total Constrained Case (\(V_i^T\), \(M_j^T\))}
+  \textbf{Total-constrained Case (\(V_i^T\), \(M_j^T\))}
 \end{enumerate}
@@
   \(\hat \kappa_{ji}^T\)) based only on the total marginal (green box in
-  Fig~\ref{fig-analytical-device-conc-accessibility}) i.e., total number
-  of opportunities or population in the system. This ensures the sum of
-  all values in the system match the total marginal.
+  Figure 2) i.e., total number of opportunities or population in the
+  system. This ensures the sum of all values in the system match the
+  total marginal.
 \item
   Units of \(V_i^T\): accessible opportunities from \(i\), a value that
-  is total constrained and linearly proportion to \(V_i^0\).
+  is total-constrained and linearly proportional to \(V_i^0\).
 \item
   Units of \(M_j^T\): accessible population from \(j\), a value that is
-  total constrained and linearly proportion to \(M_j^0\).
+  total-constrained and linearly proportional to \(M_j^0\).
 \end{itemize}
@@
 \item
-  \textbf{Singly Constrained Case (\(V_i^S\), \(M_j^S\))}
+  \textbf{Singly-constrained Case (\(V_i^S\), \(M_j^S\))}
 \end{enumerate}
@@
 \item
-  Applies singly constrained proportional allocation factors
+  Applies singly-constrained proportional allocation factors
   (\(\kappa_{ij}^S\), \(\hat \kappa_{ji}^S\)) based on Wilson's
@@
   destination-side or origin-side marginal totals (blue and red boxes in
-  Fig~\ref{fig-analytical-device-conc-accessibility}) i.e., the number
-  of opportunities or population at each zone. Reflects how te
-  literature calculates competitive accessibility.
+  Figure 2) i.e., the number of opportunities or population at each
+  zone. Reflects how the literature calculates competitive
+  accessibility.
 \item
@@
 \item
-  \textbf{Doubly Constrained Case (\(V_{ij}^D\), \(M_{ji}^D\))}
+  \textbf{Doubly-constrained Case (\(V_{ij}^D\), \(M_{ji}^D\))}
 \end{enumerate}
@@
   \(\hat \kappa_{ji}^D\)); equivalent in interpretation to Wilson's
-  \citep{wilson1971} doubly constrained spatial interaction model.
+  \citep{wilson1971} doubly-constrained spatial interaction model.
 \item
@@
   origin-side marginal totals are maintained (blue and red boxes in
-  Fig~\ref{fig-analytical-device-conc-accessibility}).
+  Figure 2).
 \item
@@
 
-As a summary, each member of the family of accessibility measure is
+As a summary, each member of the family of accessibility measures is
 named, explained in plain language, alongside their balancing factor(s),
@@
 
-Consider the simple 3-zone region in
-Fig~\ref{fig-analytical-device-conc-accessibility}, where each zone
-serves as both origin (\(i\)) and destination (\(j\)). The system
-includes three inputs: zonal population and opportunities, a zonal cost
-(travel time) matrix, and three travel impedance functions representing
-different travel behaviours.
+Consider the simple 3-zone region in Figure 2, where each zone serves as
+both origin (\(i\)) and destination (\(j\)). The system includes three
+inputs: zonal population and opportunities, a zonal cost (travel time)
+matrix, and three travel impedance functions representing different
+travel behaviours.
 
@@
 10,000s) and physicians per zone. For context, the
-Provider-to-Population Ratio (PPR) is 24.5 comparable to Canada's 2022
+Provider-to-Population Ratio (PPR) is 24.5, comparable to Canada's 2022
 PPR of 24.97 physicians per 10,000 \citep{whoMedicalDoctors102025}.
@@
 
-In \(V^0_i\), no proportional allocation factor is defined, simply
+In \(V^0_i\), no proportional allocation factor is defined; simply
 \(f(c_{ij})\) is used to weight the number of opportunities at each
 \(j\) and the weighted values for each \(j\) are summed for each \(i\),
-yielding the an expression identical to Hansen's accessibility \(S_i\)
+yielding an expression identical to Hansen's accessibility \(S_i\)
 \citep{hansen1959}, the current standard practice in accessibility
@@
 
-However \(\sum_i V^0_{i}\) generally does not equal the total
+However, \(\sum_i V^0_{i}\) generally does not equal the total
 opportunities \(O\), so units here are `opportunities weighted by travel
@@
 Comparing values across different decay functions or contexts (i.e.,
-different number of zones) is therefore limited to ordinal statements
+different numbers of zones) is therefore limited to ordinal statements
 (more vs.~less), not intervals or ratios (i.e., the magnitude of
@@
 370.92, but in what units? These two values are a product of different
-impedance functions \emph{physicians-minute\(^{-0.1}\)} and
-\emph{physicians-minute\(^{-3}\)}), making the direct comparison
+impedance functions (\(\text{physicians-minute}^{-0.1}\) and
+\(\text{physicians-minute}^{-3}\)), making the direct comparison
 uninterpretable (and arguably incorrect). The fundamental
@@
 across scenarios (potentially at the expense of accurately reflecting
-travel behavior), such adjustments may introduce bias. The unconstrained
-scores are best used for ranking within a single context.
+travel behaviour), such adjustments may introduce bias. The
+unconstrained scores are best used for ranking within a single context.
 
@@
 
-\subsection{7.3. Total constrained
+\subsection{7.3. Total-constrained
 accessibility}\label{total-constrained-accessibility}
 
-The total constrained accessibility case can be interpreted in a few
+The total-constrained accessibility case can be interpreted in a few
 ways. In the one that connects to the status quo: the total balancing
@@
 -- either total opportunities or total population. Another
-interpretation is in reformulating the equation to use a a proportional
+interpretation is in reformulating the equation to use a proportional
 allocation constant based on the total balancing factor. The
@@
 In both formulations, all zonal values become a proportion of a known
-system total, be it the regional opportunities or regional population
+system total, be it the regional opportunities or regional population,
 depending on the variant.
@@
   \(V_i^T\): accessibility is constrained by the total number of
-  opportunities (total constrained accessible opportunity) and which is
+  opportunities (total-constrained accessible opportunity), which is
   interpreted as Hansen's accessibility with a constraining constant,
@@
 \item
-  \(M_j^T\): where \(i\) and \(j\) of the first variant is transposed,
+  \(M_j^T\): where \(i\) and \(j\) of the first variant are transposed,
   yielding a measure constrained by the total population and to be
@@
 
-\subsubsection{7.3.1. Total constrained accessible opportunities:
+\subsubsection{7.3.1. Total-constrained accessible opportunities:
 Hansen's accessibility with a total
@@
 
-In the total constrained case, accessibility is expressed as a share of
-the total number of opportunities in the region \(D\), allocated based
-on travel impedance. The total constrained accessibility from \(i\) to
+In the total-constrained case, accessibility is expressed as a share of
+the total number of opportunities in region \(D\), allocated based on
+travel impedance. The total-constrained accessibility from \(i\) to
 \(j\) takes the form:
@@
 Alternatively, this formulation can be rewritten to be expressed using a
-total constrained balancing factor \(K^T\), which scales Hansen's
+total-constrained balancing factor \(K^T\), which scales Hansen's
 unconstrained accessibility \(V_i^0\) to meet the total opportunity
@@
 
-Where the total constrained balancing factor \(K^T\) is:
+Where the total-constrained balancing factor \(K^T\) is:
 \begin{equation}\phantomsection\label{eq-total-opportunity-balancing-factor}{
@@
 relative opportunity-weighted travel impedance. These values can be
-aggregated across destinations to obtain total constrained accessibility
+aggregated across destinations to obtain total-constrained accessibility
 at each origin. Alternatively, the measure can be expressed using the
@@
 \(K^T\) for other decay scenarios are calculated similarly in code.
-Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields total
-constrained accessibility values
+Applying each \(K^T\) to the unconstrained values \(V_i^0\) yields
+total-constrained accessibility values
 (Table~\ref{tbl-simple-example-total-opportunity-accessibility}), all in
@@
 \caption{\label{tbl-simple-example-total-opportunity-accessibility}Simple
-system: total constrained accessible opportunities.}
+system: total-constrained accessible opportunities.}
 
@@
 However, unlike in the unconstrained case, we can say that this value is
-out of the 490 physicians in the region, which allows us also to deduce
-that zone 1 captures 1.307 and 0.924 times more than zone 2 and 3.
-Values for the lesser decay (\(f_2(c_{ij})\)) and lowest decay
-(\(f_3(c_{ij})\)) scenarios are calculated separately, with decay
-scenario values also summing to equal 490 physicians accessible in the
-region.
+out of the 490 physicians in the region, which allows us to deduce that
+zone 1 captures 1.307 and 0.924 times more than zone 2 and 3. Values for
+the lesser decay (\(f_2(c_{ij})\)) and lowest decay (\(f_3(c_{ij})\))
+scenarios are calculated separately, with decay scenario values also
+summing to 490 physicians accessible in the region.
 
@@
 decreases, all zones become more accessible to each other and the
-differences between pairs diminishes (i.e., in \(f_3(c_{ij})\) each zone
+differences between pairs diminish (i.e., in \(f_3(c_{ij})\) each zone
 captures close to an average amount of physicians, a third of 490 or
@@
 be observed in the unconstrained measure for this scenario. However, for
-the total constrained measure, this plateauing of results has meaning.
+the total-constrained measure, this plateauing of results has meaning.
 In fact, each zone is allocated an average of the total amount in the
-region, as a result of the total constrained proportional allocation
+region, as a result of the total-constrained proportional allocation
 factor.
@@
 
-The total constrained accessibility measure resolves the
+The total-constrained accessibility measure resolves the
 interpretability issue of Hansen's accessibility (i.e., unconstrained
 accessibility) by grounding values in a meaningful total, enabling
-robust comparisons across zones and scenarios but also by keeping values
-proportional to \(V_i^0\) so interpretation is similar.
-
-\subsubsection{7.3.2. Total constrained accessible population: Reilly's
+robust comparisons across zones and scenarios, but also by keeping
+values proportional to \(V_i^0\), so interpretation is similar.
+
+\subsubsection{7.3.2. Total-constrained accessible population: Reilly's
 potential trade territories with a total
@@
 
-This second total constrained variant is a transpose of the
+This second total-constrained variant is a transpose of the
 opportunity-constrained formulation, switching indices \(i\) and \(j\)
@@
 In its unconstrained form, market potential has also been more recently
-to estimate potentially accessible populations following infrastructure
-investments
+used to estimate potentially accessible populations following
+infrastructure investments
 \citep[e.g.,][]{gutierrezLocationEconomicPotential2001, holl2007twenty, condecco2018road}.
@@
 unconstrained market potential \(M_j^0\). To address this, we introduce
-the total constrained accessible population measure \(M^T_{ji}\), which
+the total-constrained accessible population measure \(M^T_{ji}\), which
 allocates the total population \(O\) across all origin-destination pairs
@@
 
-Next, \(\hat \kappa_{ij}\) is the total constrained proportional
+Next, \(\hat \kappa_{ij}\) is the total-constrained proportional
 allocation factor, a dimensionless term which distributes population
@@
 population. Alternatively, this formulation can be rewritten to be
-expressed using a total constrained balancing factor \(\hat K^T\), which
+expressed using a total-constrained balancing factor \(\hat K^T\), which
 scales unconstrained market potential \(M_j^0\) to meet the total
@@
 
-Where the total constrained balancing factor \(\hat K^T\) is:
+Where the total-constrained balancing factor \(\hat K^T\) is:
 \begin{equation}\phantomsection\label{eq-total-population-balancing-factor}{
@@
 \caption{\label{tbl-simple-example-total-population-accessibility}Simple
-system: total constrained accessible population.}
+system: Total-constrained accessible population.}
 
@@
 
-In Table~\ref{tbl-simple-example-total-opportunity-accessibility}, zone
-1, 2, 3 represent destinations and the accessibility values reflect the
+In Table~\ref{tbl-simple-example-total-opportunity-accessibility}, zones
+1-3 represent destinations, and the accessibility values reflect the
 number of accessible people from the vantage of physicians. Zone 1, in
-its role as a destination, is no longer intermediately-ranked relative
+its role as a destination, is no longer intermediately ranked relative
 to other zones; it now attracts the fewest number of people across all
-three travel behaviour scenarios. However, similar to the total
-constrained opportunity case, as travel decay reduces, the availability
-of population begins to converge (though Zone 1 continues as the
-lowest-ranked) for similar reasons. As decay reduces, the population's
-travel impedance to all zones become more similar, making the relative
-location of the zones less important and all people in the region more
-equally accessible.
-
-Like in the total constrained accessible opportunities variant, the
-total constrained accessible population enables direct comparison of raw
+three travel behaviour scenarios. However, similar to the
+total-constrained opportunity case, as travel decay reduces, the
+availability of population begins to converge (though Zone 1 continues
+as the lowest-ranked) for similar reasons. As decay reduces, the
+population's travel impedance to all zones becomes more similar, making
+the relative location of the zones less important and all people in the
+region more equally accessible.
+
+Like in the total-constrained accessible opportunities variant, the
+total-constrained accessible population enables direct comparison of raw
 values, supporting both ordinal and interval interpretations across
@@
 
-\subsection{7.4. Singly constrained
+\subsection{7.4. Singly-constrained
 accessibility}\label{singly-constrained-accessibility}
 
-The singly constrained accessibility case can also be expressed in two
+The singly-constrained accessibility case can also be expressed in two
 variants, each defined by the direction in which a constraint is
@@
   \(V_i^S\): accessibility constrained by opportunities at destinations
-  (singly constrained accessible opportunities), and
-\item
-  \(M_j^S\): its transpose, constrained by population at origins (singly
-  constrained accessible population, or market potential).
+  (singly-constrained accessible opportunities), and
+\item
+  \(M_j^S\): its transpose, constrained by population at origins
+  (singly-constrained accessible population, or market potential).
 \end{itemize}
 
-Similar to the total constrained case, the singly constrained measures
+Similar to the total-constrained case, the singly-constrained measures
 adjust unconstrained zonal accessibility values (\(V_i^0\) or \(M_j^0\))
@@
 However, unlike the total constraint (which enforces a regional sum),
-the singly constrained case applies a localized constraint at one end of
+the singly-constrained case applies a localized constraint at one end of
 the interaction -- either origin or destination.
@@
 
-In both cases, the singly constrained formulation introduces zonal-level
-competition, unlike the total constrained case which distributes a fixed
-regional sum. Each zonal accessibility value becomes not only a fraction
-of the regional total (opportunities or population), but also a balanced
-sum of interactions, weighted by impedance and relative competition. The
-result remains in interpretable units -- accessible opportunities or
-accessible population -- but reflect these more complex spatial
-dynamics.
-
-\subsubsection{7.4.1. Singly constrained accessible opportunities:
+In both cases, the singly-constrained formulation introduces zonal-level
+competition, unlike the total-constrained case, which distributes a
+fixed regional sum. Each zonal accessibility value becomes not only a
+fraction of the regional total (opportunities or population), but also a
+balanced sum of interactions, weighted by impedance and relative
+competition. The result remains in interpretable units -- accessible
+opportunities or accessible population -- but reflects these more
+complex spatial dynamics.
+
+\subsubsection{7.4.1. Singly-constrained accessible opportunities:
 spatial
@@
 
-In this singly constrained variant, accessibility is constrained at the
+In this singly-constrained variant, accessibility is constrained at the
 destination side: the sum of accessible opportunities allocated from
@@
 The underlying spatial interaction model is now the
-attraction-constrained model and our accessibility measure becomes:
+attraction-constrained model, and our accessibility measure becomes:
 
@@
 
-Further, we can express the formula even more simply, by defining the
+Further, we can express the formula even more simply by defining the
 following proportional allocation factor:
@@
 each destination's opportunities are distributed proportionally to
-origins. As in the total constrained case, \(V_i^S\) is expressed in the
+origins. As in the total-constrained case, \(V_i^S\) is expressed in the
 units of accessible opportunities.
@@
 \(f_3(c_{ij})\) decay group, they are 0.067, 0.066 and 0.066. Using
-these these balancing constants, we can calculate the singly constrained
+these balancing constants, we can calculate the singly-constrained
 opportunity accessibility
@@
 \caption{\label{tbl-simple-example-attraction-constrained-accessibility}Simple
-system: singly constrained accessible opportunities.}
+system: singly-constrained accessible opportunities.}
 
@@
 allows for the comparison of differences and ratios of the accessibility
-values, like previously discussed in the total constrained accessible
+values, like previously discussed in the total-constrained accessible
 opportunities case. The proportional allocation factor ensures that
@@
 However, unlike \(\kappa^T_{ij}\), \(\kappa^S_{ij}\) introduces zonal
-competition based on the mass of the origin (population). In the total
-constrained case, opportunities are distributed based on impedance
-alone, regardless of population at \(i\). In contrast, the singly
-constrained case allocates each zone's opportunities proportionally
-across the region based on the relative impedance-weighted demand from
-all origins.
+competition based on the mass of the origin (population). In the
+total-constrained case, opportunities are distributed based on impedance
+alone, regardless of population at \(i\). In contrast, the
+singly-constrained case allocates each zone's opportunities
+proportionally across the region based on the relative
+impedance-weighted demand from all origins.
 
@@
 core. Its low impedance-weighted demand means \(\kappa^S_{ij}\)
-allocates it fewer opportunities. By contrast, in the total constrained
+allocates it fewer opportunities. By contrast, in the total-constrained
 case, Zone 1 fares better, capturing 35\% of all physicians (compared to
@@
 and population competition. Unlike the total constraint lowest decay
-scenarios that allocate evenly (as localised supply and demand is
-assumed to be unknown), the singly constrained case reflects
+scenarios that allocate evenly (as localized supply and demand are
+assumed to be unknown), the singly-constrained case reflects
 competition's influence on allocation.
@@
 we present per capita accessibility for the numeric example simply in
-units of number of physicians accessible per population at each zone.
-Notably, these per capita rates are equivalent to the 2SFCA values.
+units reflecting the number of physicians accessible per population at
+each zone. Notably, these per capita rates are equivalent to the 2SFCA
+values.
 
@@
 \caption{\label{tbl-simple-example-attraction-constrained-accessibility-per-capita}Simple
-system: singly constrained accessible opportunities per capita.}
+system: Singly-constrained accessible opportunities per capita.}
 
@@
 opportunities. Consequently, per capita accessibility values begin to
-stabilise to the regional per capita average (e.g., in the highest
+stabilize to the regional per capita average (e.g., in the highest
 distance decay \(f_1(c_{ij})\), per capita values are all
-\textasciitilde24 physicians accessible per capita).
-
-This convergence mirrors the trend in the total constrained opportunity
+\textasciitilde24 physicians accessible).
+
+This convergence mirrors the trend in the total-constrained opportunity
 case, where accessibility values approach a third of the 490 physicians
@@
 both cases, the balancing factors (\(K^S\) and \(B_j\)) act as averaging
-mechanisms but at different scales. As distance decay becomes more
-\emph{relatively} uniform, the role of remaining variables (i.e., total
-population or opportunities) drive the proportional allocation
-differences. In the total constrained case, this is the proportion of
+mechanisms but at different scales. As distance decay becomes
+\emph{relatively} more uniform, the role of remaining variables (i.e.,
+total population or opportunities) drive the proportional allocation
+differences. In the total-constrained case, this is the proportion of
 opportunities relative to the regional opportunities, and in the case of
@@
 
-\subsubsection{7.4.2. Singly constrained accessible population: market
+\subsubsection{7.4.2. Singly-constrained accessible population: market
 availability}\label{singly-constrained-accessible-population-market-availability}
@@
 Similar to Eq~\ref{eq-total-population-balancing-factor} in transposing
-the origins and destinations, we can define a \emph{singly constrained}
+the origins and destinations, we can define a \emph{singly-constrained}
 measure of market potential that preserves the known population (i.e.,
@@
 
-In this variant, the measure is singly constrained by the population
+In this variant, the measure is singly-constrained by the population
 \emph{by origin} (i.e., \(O_i\)), like
@@
 
-Following the same logic as in the preceding section on total
-constrained market potential, one arrives at the following expression of
-accessible population \(M_j^S\) being the product of proportionally
-allocated (\(\hat \kappa^S_{ji}\)) population:
+Following the same logic as in the preceding section on
+total-constrained market potential, one arrives at the following
+expression of accessible population \(M_j^S\) being the product of
+proportionally allocated (\(\hat \kappa^S_{ji}\)) population:
 
@@
 As well, the single (population) constraint in
-Eq~\ref{eq-population-constraint} ensures that the the total constraint
+Eq~\ref{eq-population-constraint} ensures that the total constraint
 (e.g., \(\sum_j M^S_{j} = \sum_i\sum_j  M^S_{ji} = O\)) is maintained.
@@
 
-For the sake of brevity, we'll move onto the doubly constrained case.
-
-\subsection{7.5. Doubly constrained
+For the sake of brevity, we'll move on to the doubly-constrained case.
+
+\subsection{7.5. Doubly-constrained
 accessibility}\label{doubly-constrained-accessibility}
 
-This accessibility case adopts the structure of the doubly constrained
+This accessibility case adopts the structure of the doubly-constrained
 spatial interaction model, where \(V_{ij}^D\) flows are constrained by
@@
 
-This distinguishes doubly constrained accessibility from the total and
-singly constrained cases discussed previously. In those cases, only one
+This distinguishes doubly-constrained accessibility from the total and
+singly-constrained cases discussed previously. In those cases, only one
 side of the interaction -- either the total marginal or
 opportunity/population marginal -- was constrained, while the other was
-treated as a demand/supply weight (e.g., \(D\) or \(O\) for total
-constrained and \(W_j^{(2)}\) or \(W_i^{(1)}\) for singly constrained).
-
-By contrast, the doubly constrained model assumes both the demand
+treated as a demand/supply weight (e.g., \(D\) or \(O\) for
+total-constrained and \(W_j^{(2)}\) or \(W_i^{(1)}\) for
+singly-constrained).
+
+By contrast, the doubly-constrained model assumes both the demand
 (population) and the supply (opportunity) are known and bounded, and
 allocates flows accordingly. This makes it less suitable for traditional
-accessibility analysis; namely because origin and destination masses
+accessibility analysis, namely because origin and destination masses
 often differ in kind and units. For instance, the number of people
-accessing an opportunity such as park may be known, but the capacity of
-each park is not. A doubly constrained approach may make conceptual
+accessing an opportunity, such as a park, may be known, but the capacity
+of each park is not. A doubly-constrained approach may make conceptual
 sense if the ``potential'' should be contained only in the flows
@@
 comparable, have a one-to-one correspondence or are otherwise paired
-(e.g., job per worker, student per school-seat, or vaccine doses per
+(e.g., job per worker, student per school seat, or vaccine doses per
 person). Mathematically, this model requires the simultaneous imposition
 of both the population- and opportunity- constraints in the preceding
-singly constrained variants (Eq~\ref{eq-opportunity-constraint} and
+singly-constrained variants (Eq~\ref{eq-opportunity-constraint} and
 Eq~\ref{eq-population-constraint}), namely the sum of population in all
@@
 
-The doubly constrained accessibility measure \(V_{ij}^D\) takes the form
-of the production-attraction (doubly constrained) spatial interaction
+The doubly-constrained accessibility measure \(V_{ij}^D\) takes the form
+of the production-attraction (doubly-constrained) spatial interaction
 model as follows:
@@
 
-The doubly constrained model completely distributes origin populations
+The doubly-constrained model completely distributes origin populations
 to destination opportunities according to travel impedance and
@@
 
-Following this logic, the market potential form \(M^D_{ji}\) is
+Following this logic, the market potential from \(M^D_{ji}\) is
 effectively equivalent to \(V_{ij}^D\), but can be read with a different
@@
 
-To calculate doubly constrained accessibility using the toy example, the
+To calculate doubly-constrained accessibility using the toy example, the
 interpretation of the population data and the counts of the opportunity
@@
 destination \(D_j\) is needed instead of simply the number of
-physicians. We need to also be able to clearly state that the population
+physicians. We also need to be able to clearly state that the population
 is the \emph{capacity} of the origin to interact with opportunities
@@
 opportunities (in units of 10,000s of physician-capacity) per zone. For
-the population, we leave this unchanged numerically but we now must keep
-in mind that each person interacts with one physician capacity. The
-number of providers per destination is however revised to represent
-physician capacity, scaled approximately from the original number of
-physicians used in previous cases
+the population, we leave this unchanged numerically, but we now must
+keep in mind that each person interacts with one unit of physician
+capacity. The number of providers per destination is, however, revised
+to represent physician capacity, scaled approximately from the original
+number of physicians used in previous cases
 (Table~\ref{tbl-small-system-land-use}). The system-wide PPR is now 1
@@
 
-We keep the same zonal cost matrix, and travel impedance functions for
+We keep the same zonal cost matrix and travel impedance functions for
 three types of travel behaviour as before
@@
 
-\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}Doubly
-constrained accessible opportunities assuming highest travel decay in
-the modified simple system.}
+\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-f1cij-access-values}Doubly-constrained
+accessible opportunities assuming highest travel decay in the modified
+simple system.}
 
@@
 longer meaningfully represented as zonal summaries like \(V^D_i\) or
-\(M^D_j\), since these values reproduce the original constraints i.e.,
-\(V^D_i = O_i\) hence the physician-capacity accessible for Zones 1, 2,
-and 3 are 4.002, 9.996, and 6.002. The usefulness of the doubly
-constrained measure lies in the interpretation as \(V_{ij}^D\) values;
-the number of opportunities from zone \(j\) allocated to populations in
-zone \(i\) shaped by both mass and travel impedance.
+\(M^D_j\), since these values reproduce the original constraints, i.e.,
+\(V^D_i = O_i\), hence the physician-capacity accessible for Zones 1, 2,
+and 3 are 4.002, 9.996, and 6.002. The usefulness of the
+doubly-constrained measure lies in the interpretation as \(V_{ij}^D\)
+values; the number of opportunities from zone \(j\) allocated to
+populations in zone \(i\) is shaped by both mass and travel impedance.
 
@@
 
-Following the intuition discussed in the singly constrained opportunity
+Following the intuition discussed in the singly-constrained opportunity
 case, as decay decreases, the mass effects (effect of the population and
-opportunities magnitudes) become more relatively dominant in the spatial
+opportunities magnitudes) become relatively more dominant in the spatial
 allocation.
@@
 
-\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}Doubly
-constrained accessible opportunities at Zone 2 for all travel decay
-groups in the modified simple system.}
+\caption{\label{tbl-adjusted-small-system-land-use-doubly-constrained-case-allfs-access-values-for-zone2}Doubly-constrained
+accessible opportunities at Zone 2 for all travel decay groups in the
+modified simple system.}
 
@@
 
-Accessibility is conventionally presented as a zonal value not a flow.
-However, in the doubly constrained case, since we force the allocation
+Accessibility is conventionally presented as a zonal value, not a flow.
+However, in the doubly-constrained case, since we force the allocation
 of zonal population demand and zonal opportunities supplied to be paired
 and allocation to be proportional, \(V_i^D\) is simply the number of
-opportunities that matches our known population at \(i\). Following the
-logic of the family of accessibility measures, in the doubly constrained
+opportunities that matche our known population at \(i\). Following the
+logic of the family of accessibility measures, in the doubly-constrained
 case, \(V_{ij}^D\) flows are the only relevant unit of analysis: spatial
@@
 \(V_{ij}^D\) are also mathematically equivalent to Wilson's spatial
-interaction flows. And as Wilson \citep{wilson1971} explicitly noted,
+interaction flows. And, as Wilson \citep{wilson1971} explicitly noted,
 origin and destination weights defined in the spatial interaction model
@@
 accessible per person. When they do align, such as people to physician
-capacity, we would be modeling realised access flows based on known
+capacity, we would be modeling realized access flows based on known
 quantities of \emph{those that interact} and the \emph{interacted}. In
@@
 can be reached by a zone? Answer: the number of people at that zone).
-For this reason, we do not foresee the doubly constrained measure being
-widely used in accessibility analysis as the literature has largely
-focused on questions of `potential' not on predicting flows of realised
+For this reason, we do not foresee the doubly-constrained measure being
+widely used in accessibility analysis, as the literature has largely
+focused on questions of `potential', not on predicting flows of realized
 access.
@@
 
-We then introduced the singly constrained case, which \emph{does} take
+We then introduced the singly-constrained case, which \emph{does} take
 into account the population size at the origin in the allocation of
@@
 not required to sum to any population-based values at the zone or
-regional level. The singly constrained model could be useful if regional
+regional level. The singly-constrained model could be useful if regional
 competition is a factor and if the acknowledgment that only a finite
@@
 capita), calculated by dividing each accessibility value by the zonal
-population. To clarify, this per capita expression of the singly
-constrained case is equivalent to the 2SFCA \citep{luo2003, shen1998},
-hence linking this literature back to spatial interaction principles.
-
-Lastly, the doubly constrained case is introduced. In this case, the
+population. To clarify, this per capita expression of the
+singly-constrained case is equivalent to the 2SFCA
+\citep{luo2003, shen1998}, hence linking this literature back to spatial
+interaction principles.
+
+Lastly, the doubly-constrained case is introduced. In this case, the
 sums must equal both the regional total and ensure that no zone

### 2025-10-24 paper_plos/paper_plos_V3.qmd
fixing tex for table 1

@@
 ::: {=latex}
-\input{summary-family-access-measures-table.tex}
-
+\tiny
+\begin{longtable}{|p{2.5cm}|p{2.5cm}|p{2.5cm}|p{3cm}|p{3cm}|}
+\hline
+\textbf{Name of Member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
+\hline
+\endfirsthead
+
+\hline
+\textbf{Name of member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
+\hline
+\endhead
+
+Unconstrained Accessible Opportunities ($V_i^0$) and Unconstrained Accessibile Population ($M_j^0$)
+& No constraints; marginals not equal to any regional or zonal knowns.
+& None
+& $V_i^0 = \sum_j D_j f(c_{ij})$;
+
+$M_j^0 = \sum_i O_i f(c_{ij})$
+
+& Values in various units depending on the impedance and destination-mass (e.g., opportunities $\times$ decay) for $V_i^0$ and impedance and origin-mass (e.g., population $\times$ decay) for $M_j^0$; no total or marginal constraint \\
+\hline
+
+Total-constrained Accessible Opportunities ($V_i^T$) and Total Constrained Accessibile Population ($M_j^T$)
+& Balancing factors $K^T$ and $\hat{K}^T$ ensures the sum of $ij$ values equals the total marginal, where:
+
+$K^T = \frac{D}{\sum_i V_i^0}$;
+
+$\hat{K}^T = \frac{O}{\sum_j M_j^0}$
+
+& Allocates the total marginal as opportunities $D$ based on $\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}$ and as population $O$ based on $\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}$
+
+& $V^T_i =  \sum_j \kappa_{ij}^T D$;
+
+$M_j^T =  \sum_i \hat{\kappa}_{ji}^T O$
+
+& Values reflect a share of the total opportunities in the region $D$ for $V^T_i$ or total population in the region $O$ for $M^T_j$. \\
+\hline
+
+Singly-constrained Accessible Opportunities ($V_i^S$) and Singly Constrained Accessible Population ($M_j^S$)
+& Single balancing factor $B_j$ (for $V_i^S$) that ensures the destination-mass marginal is constrained, and $A_i$ (for $M_j^S$) ensures the origin-mass marginal is constrained:
+
+$B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}$;
+
+$A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}$
+
+& Allocates the single opportunities marginal $D_j$ proportionally based on $\kappa^S_{ij} =  \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}$ in the case of $V_i^S$ and the single population marginal $O_i$ proroportionally based on $\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}$ in the case of $M_j^S$.
+
+& $V^S_i = \sum_j \kappa^S_{ij} D_j$;
+
+$M_j^S =  \sum_i \hat \kappa^S_{ji} O_i$
+
+& $V^S_i$ values reflect a share of the opportunities at each destination $D_j$ based on origin population 'demand' and impedance; $M^S_j$ values reflect a share of the population at each origin $O_i$ based on destination opportunities 'supply' and impedance.\\
+\hline
+
+Doubly-constrained Access ($V_{ij}^D$ or $M_{ij}^D$)
+& Values reflect both single marginals simutaneously, maintained via $A_i$ and $B_j$.
+& —
+& $V_{ij}^D = A_i B_j O_i D_j f(c_{ij})$
+& The spatial interactions between population and opportunities (i.e., access). \\
+\hline
+
+\end{longtable}
 :::

### 2025-10-24 paper_plos/paper_plos_V3.tex
fixing tex for table 1

@@
 
-\input{summary-family-access-measures-table.tex}
+\tiny
+\begin{longtable*}{|p{2.5cm}|p{2.5cm}|p{2.5cm}|p{3cm}|p{3cm}|}
+\hline
+\textbf{Name of Member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
+\hline
+\endfirsthead
+
+\hline
+\textbf{Name of member and Variant} & \textbf{Constraint Explanation and Balancing Factor} & \textbf{Proportional Allocation Factor} & \textbf{Accessibility Measure Equation} & \textbf{Interpretation} \\
+\hline
+\endhead
+
+Unconstrained Accessible Opportunities ($V_i^0$) and Unconstrained Accessibile Population ($M_j^0$)
+& No constraints; marginals not equal to any regional or zonal knowns.
+& None
+& $V_i^0 = \sum_j D_j f(c_{ij})$;
+
+$M_j^0 = \sum_i O_i f(c_{ij})$
+
+& Values in various units depending on the impedance and destination-mass (e.g., opportunities $\times$ decay) for $V_i^0$ and impedance and origin-mass (e.g., population $\times$ decay) for $M_j^0$; no total or marginal constraint \\
+\hline
+
+Total-constrained Accessible Opportunities ($V_i^T$) and Total Constrained Accessibile Population ($M_j^T$)
+& Balancing factors $K^T$ and $\hat{K}^T$ ensures the sum of $ij$ values equals the total marginal, where:
+
+$K^T = \frac{D}{\sum_i V_i^0}$;
+
+$\hat{K}^T = \frac{O}{\sum_j M_j^0}$
+
+& Allocates the total marginal as opportunities $D$ based on $\kappa_{ij}^T = \frac{W^{(2)}_j f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}$ and as population $O$ based on $\hat \kappa_{ji}^T = \frac{W^{(2)}_i  f(c_{ij})}{\sum_i\sum_j W^{(2)}_if(c_{ij})}$
+
+& $V^T_i =  \sum_j \kappa_{ij}^T D$;
+
+$M_j^T =  \sum_i \hat{\kappa}_{ji}^T O$
+
+& Values reflect a share of the total opportunities in the region $D$ for $V^T_i$ or total population in the region $O$ for $M^T_j$. \\
+\hline
+
+Singly-constrained Accessible Opportunities ($V_i^S$) and Singly Constrained Accessible Population ($M_j^S$)
+& Single balancing factor $B_j$ (for $V_i^S$) that ensures the destination-mass marginal is constrained, and $A_i$ (for $M_j^S$) ensures the origin-mass marginal is constrained:
+
+$B_j = \frac{1}{\sum_i W_i^{(1)} f(c_{ij})}$;
+
+$A_i = \frac{1}{\sum_j W_j^{(2)} f(c_{ij})}$
+
+& Allocates the single opportunities marginal $D_j$ proportionally based on $\kappa^S_{ij} =  \frac{W_i^{(1)} f(c_{ij})}{\sum_i W_i^{(1)} f(c_{ij})}$ in the case of $V_i^S$ and the single population marginal $O_i$ proroportionally based on $\hat \kappa^S_{ji} = \frac{W_j^{(2)} f(c_{ij})}{\sum_i W_j^{(2)} f(c_{ij})}$ in the case of $M_j^S$.
+
+& $V^S_i = \sum_j \kappa^S_{ij} D_j$;
+
+$M_j^S =  \sum_i \hat \kappa^S_{ji} O_i$
+
+& $V^S_i$ values reflect a share of the opportunities at each destination $D_j$ based on origin population 'demand' and impedance; $M^S_j$ values reflect a share of the population at each origin $O_i$ based on destination opportunities 'supply' and impedance.\\
+\hline
+
+Doubly-constrained Access ($V_{ij}^D$ or $M_{ij}^D$)
+& Values reflect both single marginals simutaneously, maintained via $A_i$ and $B_j$.
+& —
+& $V_{ij}^D = A_i B_j O_i D_j f(c_{ij})$
+& The spatial interactions between population and opportunities (i.e., access). \\
+\hline
+
+\end{longtable*}
 
